### 2010-09-24 par_instable.tex
Directory for measurement invariance paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@2 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
\title{Studying measurement invariance in factor analysis models
  via stochastic processes}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{Wichita State University}{Universit\"{a}t Innsbruck}

\abstract{Common methods for the study of measurement invariance in a
  factor-analytic context include
  likelihood ratio tests, Lagrange multiplier tests, and
  Wald tests. 
  These methods require advance definition of the number of groups, group
  membership, and offending model parameters.
In this paper, we construct tests of parameter
invariance based on stochastic processes of casewise
derivatives of the likelihood function.  These tests can be viewed as
generalizations of the Lagrange multiplier test, and they are especially
useful for: (1) isolating specific parameters affected by measurement
invariance violations, and (2) identifying subgroups of individuals
that violated measurement invariance based on a continuous auxiliary
variable.  The tests are presented and illustrated in detail, along
with simulations examining the tests' abilities in controlled conditions.}

\acknowledgements{Correspondence to Edgar C.\
  Merkle, Department of 
  Psychology, Wichita State University, Wichita, KS 67260-0034.
  Email: {\texttt{edgar.merkle@wichita.edu}}.}
\shorttitle{Parameter instability tests}
\rightheader{Parameter instability tests}

\begin{document}
\maketitle

The assumption that parameters are
invariant across observations is a
fundamental tenet of many statistical models.  A specific type of 
parameter invariance, measurement invariance, has implications for the
general design and use of psychometric scales.
A set of scales are defined to be measurement invariant with respect to
an auxiliary variable $V$ if:
\begin{equation}
    \label{eq:midef}
      f({\bf X} | T, V) = f({\bf X} | T),
\end{equation}
where ${\bf X}$ is a matrix of scale scores, $T$ is the latent
variable that the scales purport to measure, and $f$ is some assumed
distributional form.  This concept is 
particularly important because violations can render the scales
useless.  
That is, if a set of scales violates the condition described by
Equation~\eqref{eq:midef}, then individuals with the same
score on $T$ can systematically receive 
different scale scores.
This may lead researchers to conclude subgroup
differences on a wide variety of interesting constructs 
when, in reality, the scales are the sole cause of the 
differences.  Further, it can be inappropriate to incorporate scales
violating measurement invariance into structural equation
models, where relationships between constructs are hypothesized.  Horn
and McArdle \citeyear{HorMca92} concisely summarize the impact of
these issues, stating ``Lack of evidence of measurement 
invariance equivocates conclusions and casts doubt on theory in the
behavioral sciences'' (p.\ 141).  Borsboom \citeyear{Bor06a} also
notes that researchers often fail to assess whether measurement
invariance holds.

Focusing on applications, measurement
invariance is commonly assessed with respect to categorical variables
such as ethnicity. 
These assessments are often carried out using nested Multiple Groups
Models (e.g., \citeNP{Jor71,Bol89}) along with Likelihood Ratio tests,
though 
Lagrange Multiplier and Wald tests may also be constructed for this
purpose.
Scales may also violate measurement invariance with
respect to a continuous variable such as income.  In this situation,
there are no groups, but the condition in Equation~\eqref{eq:midef}
may still be violated.  The situation involving continuous auxiliary
variables is directly addressed by the tests proposed in this paper.
The tests proposed in this paper also have the ability to isolate
specific model parameters that violate measurement invariance.
Focusing on this point, Millsap's \citeyear{Mil05} review of 
measurement invariance methods cites ``locating the violation of the
invariance'' as one of the major outstanding problems.  The tests
resulting from the proposed project directly address this issue,
leading to an important advance in measurement invariance research.

In the following pages, we first provide a brief review of Likelihood
Ratio Tests, Lagrange Multiplier Tests, and Wald Tests.  We then
proposed tests that generalize the Lagrange Multiplier
tests, providing background and relevant theoretical results.  Next,
we study the 
proposed tests' abilities through example and simulation.  Finally,
we discuss some interesting extensions of the tests.
Throughout the manuscript, we use the term {\em{test}} to refer to a
statistical test and the term {\em{scale}} to refer to a psychometric 
test or scale.

\section{Existing Tests of Measurement Invariance}

Measurement invariance is often studied in a factor-analytic context.
For a $p \times 1$ vector of scales  
${\bm x}$ and $m$ factors, the model is
specified as:
% Define data model via equation
\begin{equation}
    \label{eq:famod}
    {\bm x}_i = {\bm \mu} + {\bm \lambda}{\bm z}_i + {\bm e}_i,\ \ \ i=1,\ldots,n,
\end{equation}
where ${\bm \mu}$ is a $p \times 1$ vector of scale means, ${\bm
  \lambda}$ is a $p \times m$ matrix of factor loadings, ${\bm z}_i$ is
an $m \times 1$ vector of factor scores for individual $i$, and ${\bm
  e}_i$ is a $p \times 1$ vector of error terms for individual $i$.
The model states that each scale is a
linear combination of the latent factor(s) and error.  
Multivariate normality of the ${\bm z}_i$ and of the ${\bm e}_i$ are
traditionally assumed (e.g., \citeNP{LawMax71}), leading to a model
that is often estimated via Maximum Likelihood methods.  We restrict
ourselves to Maximum Likelihood estimation in this manuscript, leaving
us with at least three statistical approaches to assessing measurement
invariance: the Likelihood Ratio Test, the
Lagrange Multiplier Test, and the Wald Test.  While these three tests
are asymptotically equivalent \cite{Sat89}, each is detailed
individually below.

\subsection{Likelihood Ratio Test}

Measurement invariance is achieved
if sets of factor analysis model parameters are equal across
subgroups of individuals 
(\citeNP{Mer93}; there are many types of measurement invariance based
on different sets of model parameters being held constant, but any of these
can be examined via the proposed tests).  If measurement invariance is
not achieved, 
then the parameters assume different values for different
subgroups.  This leads to statistical tests of measurement
invariance via the comparison of two models: a restricted model where
parameters are equal across subgroups, and a ``full'' model
where parameters are allowed to assume different values for each
subgroup.  Statistically, this comparison is often 
accomplished by a likelihood ratio test (also called a
difference test).  Let the full model parameters be denoted 
${\bm \theta} = ({\bm \theta}_1^{\prime}\ {\bm
  \theta}_2^{\prime})^{\prime}$, where we can obtain the restricted model
by placing a set of equality constraints on ${\bm \theta}_1$ described
by the vector-valued function ${\bm
  h}({\bm \theta}_1) = {\bm 0}$.  A $\chi^2$ statistic testing
measurement invariance may then be calculated via:
\begin{equation}
    \label{eq:chisq}
 \chi^2 = -2(n-1)(\log L({\bm X},
\widehat{{\bm \theta}}_c) - \log L({\bm X},\widehat{{\bm \theta}})),
\end{equation}
where ${\bm{X}}$ is an $n \times p$ data matrix, $\widehat{\bm \theta}_c$
contains Maximum Likelihood parameter estimates satisfying the
constraints ${\bm h}({\bm \theta}_1) = {\bm 0}$, and $\widehat{\bm
  \theta}$ contains Maximum Likelihood parameter estimates from the
full model.  The log-likelihood function in the above equation arises
from the multivariate normal distribution and is defined as:
\begin{equation}
    \label{eq:loglik}
    \log L( \bm{X}, \bm{\theta}) = -\frac{n}{2} \left \{ p
          \log(2\pi) + \log | {\bm{\Sigma}}({\bm{\theta}}) | +
          \text{tr}[{\bm{\Sigma}}^{-1}({\bm{\theta}})({\bm{S}} +
          ({\overline{{\bm{X}}} -
            {\bm{\mu}}({\bm{\theta}})})({\overline{{\bm{X}}} -
            {\bm{\mu}}({\bm{\theta}})})^{\prime})]\right \},
\end{equation}
where $\overline{{\bm{X}}}$ is the observed mean vector, ${\bm S}$ is
the observed covariance matrix, ${\bm{\mu}}({\bm{\theta}})$ is the
model-implied mean vector, and ${\bm{\Sigma}}({\bm{\theta}})$ is the
model-implied covariance matrix.  The test resulting from
Equation~\eqref{eq:chisq} has 
degrees of freedom equal to the difference in number of free
parameters for the two models, and it rejects the hypothesis of measurement
invariance for large values of $\chi^2$.

While the Likelihood Ratio Test is most popular for testing
measurement invariance, Lagrange Multiplier tests and Wald tests
may also be constructed for the same purpose.The Lagrange Multiplier and Wald 
tests, described below, are also advantageous in that they require
estimation of only one 
model (either the full or restricted model), whereas the Likelihood
Ratio Test requires estimation of both models.

\subsection{Lagrange Multiplier Test}
The Lagrange Multiplier test (e.g., \citeNP{Sil59,ChouBent90})
focuses on a restricted model and tests whether the equality
constraints described by ${\bm h}({\bm \theta}_1) = {\bm 0}$ should be
freed.
Estimating the restricted model, we define a vector of Lagrange
multipliers $\widehat{\bm \lambda}$ that satisfy:
\begin{equation}
    \label{eq:lagrange}
    \frac{\partial \log L}{\partial {\bm \theta}} \bigg |_{\widehat{\theta}_c} +
    \frac{\partial {\bm h}}{\partial {\bm \theta}}
    \bigg |_{\widehat{\theta}_c}^{\prime} \widehat{{\bm \lambda}} = {\bm 0}.
\end{equation}
The test itself is then based on the asymptotic distribution of 
$\widehat{{\bm \lambda}}$: if $\widehat{{\bm \lambda}}$ is sufficiently close
to ${\bm 0}$, then the constraints are deemed to hold in the
population.  The resulting test statistic, which is a function of
$\widehat{\bm \lambda}$, follows a $\chi^2$ distribution for large
samples.

\subsection{Wald Test}
In contrast to the Lagrange Multiplier test, the
Wald test (e.g., \citeNP{Wald43})
focuses on the full model and tests the plausibility of the
constraints ${\bm h}({\bm \theta}_1) = {\bm 0}$.
The test is based on the estimates ${\bm h}(\widehat{{\bm \theta}}_1)$
from the full model; if
these are sufficiently close to zero, then the constraints are deemed
to hold in the population.  The resulting test statistic, which is a
function of ${\bm h}(\widehat{{\bm \theta}}_1)$, is also
$\chi^2$-distributed in large 
samples.
Chou and Bentler \citeyear{ChouBent90} compared the 
Wald and Lagrange tests with finite sample sizes via
simulation, and they generally found similar
performance.  They noted, 
however, that the Lagrange test has a tendency toward Type I errors in
some contexts.  They therefore urged caution in the Lagrange test's
use, as it may lead one to discard more constraints than is necessary,
leading to overly-complex models.

\subsection{Relationships to Proposed Tests}
All the tests described above require advance specification of
subgroups that potentially violate measurement invariance.  This is a
major limitation in psychometrics, where researchers
ultimately want to ensure that tests are invariant with respect to all
possible subgroups of students.  For example, there are many possible
ways to define subgroups based on a continuous auxiliary variable like
income.  Assuming two subgroups, we
could define an arbitrary cutoff separating the
poorer individuals from the richer individuals.  Alternatively, we could
consider larger numbers of subgroups with similarly-arbitrary cutoffs
defining the groups.  The ultimate subgroup definitions that we
choose could potentially impact our
conclusions about whether or not a scale is measurement invariant, 
in the same way that dichotomization of continuous variables impacts
general statistical analyses \cite{MacZha02}.
Heuristically, we would like to conduct a separate Lagrange (or
Likelihood Ratio or Wald) test for each possible definition of
subgroups, then declare the scales to be measurement invariant if they
``pass'' all of these tests.  Given the large number of possible
subgroups, however, this would generally result in problems with Type
I errors.\footnote{An alternative to this procedure involves inserting the
continuous income variable directly into the factor analysis model,
which we do not pursue due 
to the potential impact on model fit and on other parameter estimates.}

  % Instead of this, can employ parameter instability tests from
  % theoretical statistics to simultaneously examine all possible
  % groupings.
Instead of testing measurement invariance separately for each possible 
subgroup definition, we can use a single test from the family of tests
proposed here.
The proposed tests can be viewed as generalizations of the
Lagrange test \cite{And93}, 
as they focus on a restricted model and 
simultaneously test for measurement invariance across all
possible subgroup definitions.  This removes the problems associated
with error rates and generally promotes a more flexible framework
for studying measurement invariance.  As previously stated, the proposed
tests also have the ability to isolate specific model parameters violating
measurement invariance.

\section{Stochastic Processes for Measurement Invariance}
The proposed tests of measurement invariance implicitly rely on an
estimated factor analysis model (one that restricts parameters to be
invariant across subgroups).
The traditional factor analysis likelihood function 
(Equation~\eqref{eq:loglik}) 
utilizes the observed covariance matrix
${\bm{S}}$ and observed mean vector $\overline{\bm{X}}$
instead of individuals' data, but the proposed tests
require the likelihood to be evaluated for each individual.  Thus, 
we utilize the equivalent casewise likelihood:
\begin{equation}
    \label{eq:caselik}
    \log \text{L}^{\ast}({\bm{X}}, {\bm{\theta}}) = \sum_{i=1}^{N} \log \mid 
\bm{\Sigma}(\bm{\theta}) \mid + \sum_{i=1}^{N} (\bm{x}_i -
\bm{\mu}(\bm{\theta}))^{\prime} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
\bm{\mu}(\bm{\theta})).
\end{equation}
This equation is often employed for factor analysis and structural
equation models with missing data (e.g., \citeNP{Wot00}), 
in which case we would take subsets of
$\bm{\Sigma}(\bm{\theta})$ and $\bm{\mu}(\bm{\theta})$ corresponding
to the observed data on each individual $i$.

Maximization of Equation~\eqref{eq:caselik} typically involves iterative
algorithms that seek values $\widehat{{\bm{\theta}}}$ such that
$\frac{\partial}{\partial \theta_j} \log \text{L}^{\ast}({\bm{X}},
{\bm{\theta}}) \big |_{\theta_j = \widehat{\theta}_j} = 0\ \ \forall\ j$.
% insert derivatives of casewise evaluated at theta-hat = 0
Such derivatives are taken across individuals, but, using the casewise
likelihood described above, we can also 
evaluate the derivatives with respect to each individual.
If a unique value of $\widehat{{\bm{\theta}}}$ maximized
each individual likelihood, then we would have:
% Equation: \psi(y_i, \theta) = d/d theta casewise_i = 0 \forall i.
\begin{equation}
    \label{eq:casederiv}
    \psi({\bm{x}}_i, \theta_j) = \frac{\partial}{\partial\theta_j}
\left[ \log \mid 
\bm{\Sigma}(\bm{\theta}) \mid + (\bm{x}_i -
\bm{\mu}(\bm{\theta}))^{\text{T}} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
\bm{\mu}(\bm{\theta}))\right] \bigg |_{\theta_j = \widehat{\theta}_j} = 0\ \ \ \forall\ i, j.
\end{equation}
In practice, no value of $\widehat{{\bm{\theta}}}$ will maximize all
individuals' likelihoods, but, if parameter invariance is satisfied, 
we can expect the $\psi(\cdot, \cdot)$ to fluctuate around zero.  This
is the idea underlying the tests described here.

\subsection{Theory}
Theoretical work in statistics (e.g., \citeNP{HjoKon02}) has shown
that functions of the 
$\psi(\cdot, \cdot)$ follow specific stochastic processes, allowing
us to use them to generally test measurement
invariance.  
That is, letting ${\bm \theta}_i$ reflect parameter values
specifically 
associated with individual $i$, measurement invariance can be 
described by the null hypothesis:
\begin{equation}
    \label{eq:h0}
    H_0: {\bm \theta}_i = {\bm \theta}_0,\ i=1,\ldots,n,
\end{equation}
where the $i$'s are typically ordered with respect to the auxiliary
variable of interest (i.e., the variable defining subgroups for tests
of measurement invariance).  For example, if parental income was the
auxiliary variable, we would order students according to parental
income and then examine whether parameters were constant across
students.  The ordering of individuals is crucial to
the test development below, but it is relaxed later in the proposal.

Despite addressing measurement
invariance, the above $H_0$ focuses no longer on subgroups of
individuals but rather on individuals themselves.  The $H_0$ is
generally applicable to situations where parameters may vary across
observations.  For example, it has most often been examined in
longitudinal models where observations are 
ordered according to time.
%In this context, {\em{change-point
%    analyses}} have received much attention from the psychometric
%community \cite{Coh08}. % CITE BOOK
Observations may also be ordered with respect to 
continuous variables other than time, a fact that allows for the
measurement invariance tests described here.  Thus, the proposed tests
stem from seminal work on
detecting turning points in longitudinal models (e.g.,
\citeNP{BroDur75,Nyb89}).

The alternative hypothesis corresponding to the above $H_0$ is that not
all the ${\bm{\theta}}_i$ are constant (i.e., measurement invariance is
violated), though specific local alternatives may also be specified to
make the tests more powerful.  We now review some of the main results
for testing $H_0$ via stochastic processes of first derivatives; more
detailed accounts include Hjort and Koning \citeyear{HjoKon02} and
Zeileis and Hornik \citeyear{ZeiHor07}.

% New subsection starts here?

For application to measurement invariance, the most important
theoretical result involves the fact that, under $H_0$, functions of
the $\psi(\cdot, \cdot)$ result in specific stochastic
processes.  One such function
is the {\em{cumulative score process}}, given as:
\begin{equation}
    \label{eq:cumscore}
    W_n(t,\theta_j) = \frac{1}{\sqrt{n}} \sum_{i=1}^{\lfloor nt
      \rfloor} \psi ({\bm{x}}_i, \theta_j),
\end{equation}
where $t$ is a continuous variable in $(1/n,1)$ and $\lfloor nt
\rfloor$ is the floor operator, defined as the integer part of
$nt$.
As the equation shows, the cumulative score process adds subsets of
casewise first derivatives across individuals who have been ordered
based on an auxiliary variable.  Starting at $t=1/n$, only the
first individual's derivative enters into the summation.  Once
$t=2/n$, the first two individuals' derivatives enter into the
summation.  This procedure continues until we reach $t=n$, at which
point all individuals' first derivatives have entered into the
summation.  There are $n$ cumulative sums obtained by sequentially
inserting individuals' derivatives into the equation, and these are
calculated 
for each of the $p$ model parameters.  This results
in $n\times p$
empirical cumulative scores.

Elements of the summation in Equation~\eqref{eq:cumscore}
can be positive or 
negative, so that the cumulative scores do not necessarily increase
with $t$.  In fact, if measurement invariance (i.e., $H_0$) holds,
then the cumulative scores should 
follow a stochastic process that fluctuates around zero.  
Zeileis and Hornik \citeyear{ZeiHor07} showed specifically that, under
$H_0$:
%which generalizes earlier results from %CITE bigger journals:
% Z+H Theorem 2/Corollary 2
\begin{equation}
    \label{eq:browbridge}
{\bm{B}}(\cdot, \widehat{{\bm{\theta}}}) = {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})
\overset{d}{\rightarrow} {\bm W}^{0}(\cdot), % put d over rightarrow
\end{equation}
where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed information
matrix (e.g., of the factor analysis model), ${\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})$ reflects the $p$-dimensional
cumulative sums (one for each parameter), ${\bm W}^{0}(\cdot)$ is a
$p$-dimensional Brownian bridge, and 
$\overset{d}{\rightarrow}$ reflects convergence in distribution. %?
In words, there are $p$ cumulative score processes, one for each
model parameter.  This collection of processes follows a
multidimensional Brownian bridge as derivatives accumulate in the 
summation from individual 1 to individual $n$.
Multiplication by ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}$ 
``decorrelates'' the $p$
cumulative score processes, such that each univariate process (i.e., each
process for a single model parameter) is independent of all other
processes.

The cumulative scores resulting from Equation~\eqref{eq:browbridge}
can be arranged in an $n \times p$ matrix called an {\em{empirical
    cumulative score process}}, which we will denote as
${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$.
Each column of this matrix follows a univariate Brownian bridge and is
based on a single factor analysis parameter. 
To carry out a test of $H_0$, we need functions that collapse across
rows (individuals) and/or columns (parameters) of the matrix.  These
functions then lead to scalar test statistics and p-values.
The specific choice of function depends on the goals of the analyst:
the row-collapsing functions are useful for isolating subsets of
parameters that violate measurement invariance, and the
column-collapsing functions are useful for isolating subgroups of
individuals who violate measurement invariance across all parameters.
Some functions result in test
statistics whose critical values can be derived analytically, while
others require critical values to be derived via the
simulation of a Brownian bridge.

For example, if we wish to test a specific parameter for
measurement invariance, ``row-collapsing''
functions include the absolute maximum or mean of the corresponding 
column in ${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$
\cite{ZeiHor07}. % (see ZH07 p 7)
Conversely, if we wish to generally test 
measurement invariance across sets of model parameters,
``column-collapsing'' functions 
include the maximum norm or the squared Euclidean norm of the
corresponding row in ${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$
\cite{HjoKon02}.

A general, two-step strategy for testing measurement
invariance involves first collapsing over columns for each individual,
yielding a single statistic for each individual. 
These $n$ statistics (ordered according to an auxiliary variable)
can then be graphed to visually define groups that violate measurement
invariance.  Functions of these $n$ statistics can also yield scalar
test statistics.

As a concrete example of the above strategy, we could
first collapse over columns for each individual $i$ via the Euclidean
norm:
\begin{equation}
    \label{eq:meanl2}
    \lambda_i = \lVert {\bm B}^{\text{emp}}_{i\cdot}(\widehat{\bm \theta}) \rVert,
\end{equation}
where the $i\cdot$ subscript signifies row $i$ of the empirical
fluctuation process matrix.  A univariate test statistic may then be
obtained by taking the mean of the $\lambda_i$.  This is a
Cram\'{e}r-von Mises statistic described by, e.g., Nyblom
\citeyear{Nyb89} and Hansen \citeyear{Han92} in the contexts of
longitudinal models and linear regression.  In general contexts,
this test is recommended
when there are more than two groups exhibiting parameter instability 
and when multiple parameters change across groups \cite{ZeiSha10}.  
We demonstrate these recommendations' implications for
measurement invariance later in the paper.

% We still obtain a unidimensional process; the norm is applied
% to each row vector.
As an alternative to the two-step approach described in the previous
paragraph (collapse over columns, then take the absolute maximum), the
{\em{double-max}} statistic allows for simultaneous isolation of 
individuals violating invariance along with offending model
parameters (that are unspecified in advance).  These functions take
the maximum absolute cumulative sum across all entries:
\begin{equation}
    \label{eq:doubmax}
    DM = \max_{i=1,\ldots,n} \max_{j=1,\ldots,p} |
    B^{\text{emp}}_{ij}(\widehat{\bm{\theta}}) |,
\end{equation}
where $B^{\text{emp}}_{ij}(\widehat{\bm{\theta}})$ represents the
entry in row $i$ and column $j$ of
${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$.  This is different from
the two-step approach because, in the two-step approach, we collapse
over parameters (columns) first.  We then obtain a test statistic
from an $n \times 1$ vector.  In contrast, the
$DM$ statistic takes an absolute maximum with respect to the $n \times
p$ empirical cumulative score process (i.e., with respect to
cumulative sums associated with individual parameters).  This is
especially useful for visualization, as the fluctuation process for
each individual parameter can be displayed along with the appropriate
critical value.  An example of this visualization appears in the
Example section.

There exist other functions of first derivatives and of the
empirical fluctuation process that
result in other test statistics; these are described by, e.g.,
Zeileis, Shah, and Patnaik \citeyear{ZeiSha10}.  
In the analyses described later, we focus on the abilities
of the Cram\'{e}r-von Mises test and of the double-max test.  However,
the many 
row- and column-collapsing functions offer a large family of tests
that can be applied to the study of measurement invariance.

\subsection{Critical Values \& $p$-values.}
To obtain $p$-values associated with $H_0$, we can rely on analytic
work concerning functions of Gaussian processes.
As an example of this, we focus on use of the 
double-max function (Equation~\eqref{eq:doubmax}) as a test statistic.
There exist 
analytic results for Brownian bridges that allow us to calculate
% fix double-max notation on following line:
$P(DM > \beta | H_0)$, where $\beta$ is a specific value
that the test statistic may achieve.

We first focus on a univariate
Brownian bridge, which could test a single parameter (say, $\theta_j$) for
measurement invariance using the test statistic $DM_j$.
Under $H_0$, it can be shown that (e.g., \citeNP{ShoWel86}):
 % see Beghin & Orsingher Eq (4.12).  They cite Shorack &
            % Wellner 1986 book
\begin{eqnarray}
    P(DM_j < \beta | H_0) &
    = & \sum_{h=-\infty}^{\infty} (-1)^h \exp(-2h^2\beta^2) \\
   & = & 1 + 2 \sum_{h=1}^{\infty} (-1)^h \exp(-2h^2\beta^2).
\end{eqnarray}
The associated $p$-value is therefore given by:
\begin{equation}
    \label{eq:absmax_p}
        p = P(DM_j > \beta | H_0)
    = 1 - (1 + 2 \sum_{h=1}^{\infty} (-1)^h \exp(-2h^2\beta^2)).
\end{equation}
The terms in the summation quickly go to zero as $h$ goes to infinity,
so, in practice, we only need to compute the first 100 terms (say).

For multidimensional Brownian bridges whose dimensions are independent
(i.e., those that 
have been ``decorrelated'' by the information matrix), $p$-values can
be obtained similarly.  Assuming a Brownian bridge with $p$
independent dimensions (i.e., simultaneously testing $p$ model
parameters for measurement invariance), we have:
\begin{eqnarray}
    \label{eq:multidim}
    p & = & P(DM_1 >
    \beta\ \cup\ DM_2 > \beta\ \cup\ \ldots \cup\ DM_p > \beta | H_0)
      \nonumber \\
   & = & 1 - P(DM_1 <
    \beta\ \cap\ DM_2 < \beta\ \cap\ \ldots \cap\ DM_p < \beta | H_0)
     \nonumber \\
   & = & 1 - \left ( 1 + 2 \sum_{h=1}^{\infty} (-1)^h
       \exp(-2h^2\beta^2) \right )^p.
\end{eqnarray}
These expressions are used to compute $p$-values for the double-max
test statistic, with associated critical values being
obtained, e.g., through a simple grid search using
Equation~\eqref{eq:multidim}.
For other test statistics, we can either obtain similar expressions
for $p$-values or obtain $p$-values via simulation of Brownian
bridges.

\subsection{Locating the Invariance.}
If a test described above implies a measurement invariance violation,
then the researcher may want to: (1) define subgroups of individuals that
the scales measure differently; or (2) isolated model
parameters affected by the invariance.  
In the context of a Cram\'{e}r-von Mises test, we can accomplish the
first task by graphing the $\lambda_i$ (Equation~\eqref{eq:meanl2})
and noting the location of the peaks and valleys.  We can accomplish
the second task by taking the norm in Equation~\eqref{eq:meanl2}
across specific columns of $B^{\text{emp}}(\widehat{\bm \theta})$.

In the context of a double-max test, we can accomplish the first task
by defining subgroups
based on the individual who achieves the absolute
maximum.  That is, assuming the existence of two subgroups
violating measurement invariance, the absolute maximum cumulative
score should be achieved close to 
the ``cut point'' between subgroups.  For more than two subgroups, we
could heuristically define groups based on the number of peaks and
valleys in the graph.
We can accomplish the second task by conducting posthoc
tests on specific model parameters.  These ideas are illustrated in
the example below, which focuses on the double-max test.

\begin{figure}
\caption{Path diagram representing the factor analysis model used for
  the Example and Simulations.}
\label{fig:famod}
\includegraphics[height=5in]{famod.pdf}
%\input{figs/path}
\end{figure}

\section{Example with Artificial Data}

As an example of the ideas discussed above, consider a
hypothetical battery of six scales administered to students aged 14 to 17
years.  Three of the scales are intended to measure verbal ability, and
three of the scales are intended to measure mathematical ability.  We
may observe a maturation effect in the resulting data, whereby the scales
measure older students better than younger students (i.e., the scales
are too difficult for younger students).  The researcher's goal is to
study whether the scales are measurement invariant with
respect to the full range of ages.

\subsection{Method}

To formally represent these ideas, we specify that the data arise from
a factor analysis model with two factors.  As displayed in
Figure~\ref{fig:famod},
three scales arise from the verbal factor and three 
scales arise from the mathematical second factor.
A seventh observed variable (not pictured in Figure~\ref{fig:famod}),
student age, determines the values of 
verbal factor loadings in the model: 
if students are 16 or 17 years of age, then the factor
loadings corresponding to the first 
factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
Figure~\ref{fig:famod}.  If students are between 14 or 15 years of age,
however, then the factor loadings corresponding to the first factor
are three standard errors lower than those in Figure~\ref{fig:famod}
(the standard errors were obtained by fitting
  the model to a sample of size 200, where the sample consisted of all
  ``older'' subjects).
This violation states that the verbal ability scales
lack measurement invariance with respect to age.  For simplicity, we
assume that the mathematical scales are invariant.

A sample of size 200 was generated from the model described above, and
a test was conducted to examine measurement invariance of the three
verbal scales.  To carry out
the test, a confirmatory factor analysis model (with the paths
displayed in Figure~\ref{fig:famod}) was fit to the
data.  Casewise derivatives and 
the observed information matrix were then obtained, and they were used
to calculate the cumulative score process via
Equations~\eqref{eq:cumscore} and~\eqref{eq:browbridge}.  Finally, we
obtained a 
double-max statistic via Equation~\eqref{eq:doubmax} and associated
$p$-value via Equation~\eqref{eq:multidim}.  Note that, because we
were examining only the three verbal factor loading parameters, the
specific null hypothesis being tested is:
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{11}\ \lambda_{21}\ \lambda_{31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$.

\subsection{Results}
In the results section, we first describe overall results and
describe isolation of groups and model parameters violating
measurement invariance.

\subsubsection{Overall Results}
Results related to the cumulative score process are displayed in
Figure~\ref{fig:cusumex}.  In this figure, the processes
are displayed individually 
for each factor loading, with student number (ordered by age and
numbered from 1 to 200, but divided by 200 to correspond 
to a Brownian bridge) on the x-axis and
absolute values of the cumulative scores on the y-axis.  The dashed horizontal
lines represent the critical value for rejecting $H_0$ at $\alpha=.05$
based on the double-max statistic; if any of the three processes
crosses the horizontal line, then we reject $H_0$.  It is seen that
the processes for both the second and third factor loadings
($\lambda_{21}$ and $\lambda_{31}$) cross the dashed line, indeed
implying rejection of $H_0$.  The maximum across all three processes,
corresponding to the double-max test statistic, is 1.72.  This maximum
occurs in the third graph and is represented as a solid circle, 
achieving a $p$-value of .016.

\subsubsection{Group and Parameter Isolation}
The results of the test imply that the 
verbal scales lack measurement invariance, and the age at which the
maximum is achieved allows us to define subgroups violating
measurement invariance.  The student corresponding to the maximum
absolute cumulative sum (the solid circle in the third graph) is 15.8
years of age, implying one subgroup of 
students below 15.8 years and one above 15.8 years.  This agrees well
with the true subgroup cut point of 16.0 years.

The results of the test also yield information about specific
parameters violating measurement invariance.  We simultaneously tested
three factor loading parameters for measurement invariance.  Given
rejection of the hypothesis of measurement invariance, we can 
do posthoc 
tests to study specific parameters violating measurement invariance.
This involves individual examination of each cumulative score process
(i.e., of each of the three graphs in Figure~\ref{fig:cusumex}),
along with a procedure that controls for Type I error.  Taking a
Bonferroni approach, we can examine whether each process's maximum
surpasses the critical value associated with an $\alpha$ of
$.05/3=.017$.  This amounts to the same critical value as that used in
the overall test (as shown by Equation~\eqref{eq:multidim}),
meaning that we simply examine the individual processes that cross
the horizontal dashed lines in Figure~\ref{fig:cusumex}.  The Figure
shows that the second and third parameters (i.e., $\lambda_{21}$ and
$\lambda_{31}$) cross the dashed line, so we would conclude a measurement
invariance violation with respect to age in the second and third
verbal tests.  The fact that the cumulative score process for
$\lambda_{11}$ did not achieve the critical value represents a Type II
error, which implicitly brings into question the posthoc tests' power.  We
generally address the issue of power later in the simulations below.

\begin{figure}
\caption{Cumulative score processes for the verbal factor loadings,
  based on the example involving measurement invariance with respect
  to student age.  The dashed lines represent critical values for the
  test of measurement invariance at $\alpha=.05$.}
\label{fig:cusumex}
\includegraphics[height=5in]{example.pdf}
\end{figure}

\section{Simulation}
% Simulations demonstrating power of test(s) at various
% sample sizes, types of invariance violations, ?
In this section, we conduct a simulation designed to
examine the tests' power and Type I error rates in the context
described previously (testing three factor loadings for measurement
invariance).  We 
examine the power and error rates of three tests: the double-max test, the
Cram\'{e}r-von Mises test, and the traditional $\chi^2$ likelihood
ratio (difference) test involving nested 
multiple-group models.  The $\chi^2$ test has an advantage over the
other two tests here, in that it requires advance definition of
groups violating measurement invariance (we supply it with the known
subgroups here).  This test is useful for comparative purposes,
however, as it helps indicate the extent to which the other tests
``lose'' power due to lack of subgroup information.

\subsection{Method}
Simulated data were generated in a similar way to that of 
the Example section.
These data included verbal scales that violated measurement invariance
with respect to students' ages, with students 14--15 years of age
having smaller factor loadings than students 16--17 years of age.  
Simulation data were generated from the same general factor analysis
model (see Figure~\ref{fig:famod}), but sample size and magnitude
of measurement 
invariance violation was manipulated to examine power:
we examined power to detect 
invariance violations across three sample sizes ($n=100,
200, 500$) and three magnitudes of violations (factor loadings
for the younger students being 0, 1, 2, or 3 standard
errors\footnote{The standard errors used here were
  obtained by fitting
  the model to a sample of size 100, where the sample consisted of all
  ``older'' subjects.} below 
those for the older students).  The 0-standard error condition was
used to study Type I error rate.

For each sample size $\times$
magnitude combination, 500 datasets were generated and tested for
measurement invariance.  For the 0-standard error condition, we
generated 5,000 datasets to get a more precise measure of Type I
error.  In each dataset, half the students were 14--15
years of age and half were 16--17 years of age.  As stated previously,
group
designations (below 16.0 years, above 16.0 years) were assumed to be
known for the likelihood ratio test.

\subsection{Results}
% MODIFY RESULTS TO DISCUSS ALL THREE TESTS
% ALSO NEED 0-SE CONDITION
For each test statistic, power and Type I error rates for the twelve
sample size $\times$ magnitude combinations are displayed in
Table~\ref{tab:sim1}.  The 0-SE column lists Type I error rates; this
is the situation where the tests reject the hypothesis of measurement
invariance at $\alpha=.05$ when none exist.  This column generally
shows that the Cram\'{e}r-von Mises test has accurate Type I error
rates at smaller sample sizes, but it becomes slightly conservative at
$n=500$.  The double-max test, on the other hand, is conservative at
all sample sizes examined.  Finally, the likelihood ratio test is
slightly liberal at smaller sample sizes, with Type I error rates of
.07 and .06 at $n=100$ and $200$, respectively.  Type I error for the
likelihood ratio test at $n=500$ is accurate, however.

Focusing on power, the table includes many interesting results.
First, the likelihood ratio test has superior power over the stochastic
process-based tests at all sample sizes and invariance violations.
This is likely due to the fact that the likelihood ratio test had
extra information (the subgroup definitions), whereas the other tests
were essentially testing measurement invariance over all possible
subgroups.  While power was lower for the tests developed in this
paper, it was not terribly low.  At $n=200$ and above, all tests were
on the same side of the traditional power cutoff of 0.8.  The tests
were able to detect large measurement invariance violations (of three
and sometimes two standard errors) at the larger sample sizes, and
they were generally unable to detect the one-standard-error violation
at any of the sample sizes studied.

Comparing only the two proposed test statistics, we see that the
Cram\'{e}r-von Mises test has better power in all conditions.  This is
similar to the general recommendation described by Zeileis et al.\
\citeyear{ZeiSha10}, stating that the Cram\'{e}r-von Mises test
is generally superior when multiple parameters change across
subgroups.  Because the three verbal factor loadings all changed
across the two subgroups in the simulation, the Cram\'{e}r-von Mises
test excels.

In summary, we found that the proposed tests have less power than a
likelihood ratio test with known subgroups.  However, the power levels
remain adequate after considering the fact that the proposed tests are
testing measurement invariance across all possible subgroups.  For the
two-factor model examined here, all tests had trouble detecting small
measurement invariance violations at small sample size.  Larger
violations were more easily detected at larger sample sizes, with the
Cram\'{e}r-von Mises test having better power due to multiple
parameters being involved in the invariance violation.

% 0-SE column is fixed for n=100 (using 5000 reps instead of 500)
\begin{table*}
\caption{Simulated power and Type I error rates for three test statistics
  across three sample sizes and four magnitudes of measurement
  invariance violations. Abbreviations: CvM = Cram\'{e}r-von Mises
  test; DM = Double-max test; $\chi^2$ = Chi-square likelihood ratio
  test.}
\label{tab:sim1}
\begin{tabular}{llcccc} \thickline
 & & \multicolumn{4}{c}{Violation Magnitude (SEs)} \\
$n$ & Test Stat & 0 & 1 & 2 & 3 \\\hline
100 & CvM       & 0.05 & 0.10 & 0.33 & 0.63 \\ 
    & DM        & 0.03 & 0.06 & 0.19 & 0.41 \\ 
    & $\chi^2$  & 0.07 & 0.16 & 0.49 & 0.83 \\ \hline
200 & CvM       & 0.05 & 0.19 & 0.64 & 0.96 \\ 
    & DM        & 0.04 & 0.12 & 0.44 & 0.88 \\ 
    & $\chi^2$  & 0.06 & 0.27 & 0.76 & 0.99 \\ \hline
500 & CvM       & 0.03 & 0.44 & 0.98 & 1.00 \\ 
    & DM        & 0.03 & 0.35 & 0.94 & 1.00 \\ 
    & $\chi^2$  & 0.05 & 0.58 & 1.00 & 1.00 \\\thickline
\end{tabular}
\end{table*}

% Need to re-do simulation to monitor specific parameters in the
% posthoc test.

% Cramer-von Mises now in Sim 1.  May skip Sim 2 for the time being.
%\section{Simulation 2: Improving Power}
%Simulation 2 examines whether we can improve power using a Cramer-von
%Mises statistic (summing instead of taking maximum).

% Table 2 of Nyblom89 contains critical values that may prove useful.
% Hansen 1992 p. 524 has table of critical values that matches 
%  Nyblom89 within reason.
% See Zeileis05 p. 450-451

% Use model that is similar to the example.

\section{Discussion}
In this paper, we have presented a new class of statistical tests for
the study of measurement invariance in psychometrics.  The tests,
based on stochastic processes, have reasonable power, can isolate
groups of individuals violating measurement invariance, and can
isolate specific model parameters affected by the violation.  In this
section, we consider the tests' use in practice and their 
extension to more complex scenarios.

\subsection{Use in Practice}
% Simultaneous test of Meredith's various types of invariance
The proposed tests give researchers a new set of tools for studying 
measurement invariance.  First, they give researchers the
opportunity to simultaneously test for many types of measurement
invariance (see, e.g., Meredith \citeNP{Mer93}).  The traditional
steps have involved hypothesizing a specific type of invariance and
then testing for it via a Likelihood Ratio Test, but the tests
proposed here do not require one to hypothesize a specific type of
invariance a priori.  Instead, one could define a type of measurement
invariance after seeing the proposed tests' results.

% Researchers interpreting subgroups violating MI
In addition to simultaneously testing different types of measurement
invariance, the proposed tests allow researchers to easily interpret
the nature of the invariance violation.  This is made possible through
the tests' abilities to identify subgroups of individuals violating
measurement invariance.  While subgroup identification was
accomplished in this paper via visualization, it is also possible to
define 
formal rules for identifying subgroups.  This idea leads us to the
next section, which describes further issues associated with the
tests' application to the study of measurement invariance.

\subsection{Further Issues}
Further issues associated with the proposed tests include: (1) formal
definition of subgroups violating measurement invariance based on the
results of the statistical test; and (2) dealing
with ties on the auxiliary variable.  We address each in turn below.

% More than two groups violating MI
\subsubsection{Subgroup Definition}
Given rejection of the hypothesis of measurement invariance, the most
straightforward way to formally identify offending groups is through
the location of the maximum of the cumulative score process.  Assuming two
subgroups, the maximum should be achieved close to the split point
between the the groups.  Referring to the Example (see
Figure~\ref{fig:cusumex}), we would split students at the
age corresponding to the solid dot in the third panel.

To study this method for group identification in more detail, we used
results of our simulation to examine the age at which the maximum of
the cumulative score process occurs.
If the test accurately recovers the violating subgroups,
then the maximum should be close to 16 years
of age.\footnote{The data generated for these
  simulations 
  contained ages measured in days, as opposed to years.  Thus,
  the tests could split students at continuous values like 15.6 or
  16.3 years of age.}
  For each simulation condition, Table~\ref{tab:age1}
  displays confidence intervals for the location of the maximum
  absolute cumulative score.  These confidence intervals were
  calculated across generated 
  datasets, with each dataset contributing one maximum absolute cumulative
  score
  (regardless of whether the hypothesis of
  measurement invariance was rejected).
It is observed that these confidence intervals slightly underestimate
the true age dividing the subgroups, with the degree of bias
decreasing as sample size increases.  The interval widths
are impacted by both the magnitude of the invariance and the sample
size.  In future work, we plan to study alternative ways of
identifying subgroups.

\begin{table*}
\caption{Confidence intervals for the location of the maximum absolute
  cumulative score across the conditions from the simulation.}
\label{tab:age1}
  \begin{tabular}{lccc} \thickline
  & \multicolumn{3}{c}{Violation Magnitude} \\
  $n$ & 1SE & 2SE & 3SE \\\hline
  100 & (15.61; 15.73) & (15.55; 15.65) & (15.50; 15.59) \\
  200 & (15.67; 15.77) & (15.64; 15.72) & (15.67; 15.73) \\
  500 & (15.71; 15.81) & (15.77; 15.83) & (15.80; 15.83) \\\thickline
  \end{tabular}
\end{table*}

% Differing proportions of people in each group
%\subsubsection{Subgroup Proportions}
%The power of the proposed tests is impacted by the proportion of
%individuals within each subgroup.

% Ties on auxiliary variable (mention use in categorical contexts)
\subsubsection{Ties on the Auxiliary Variable}
There remain unaddressed theoretical issues specific to the 
tests' application to measurement invariance.  These
issues generally stem from the types of auxiliary variables on which
researchers may study measurement invariance.  In
the example described earlier, researchers wanted to know whether a
scale was measurement invariant with respect to student age.
Students could then be ordered based on age, and a test could be 
constructed via Equation~\eqref{eq:cumscore}.  However, an issue arises if there
are ties on age; that is, if two students' ages are exactly
the same.  The proposed tests assume complete ordering of the
students, but these two students cannot be ordered.  In this case, we
may randomly order the two students, but there is an
arbitrariness associated with this strategy.  Alternatively, to obtain
tests that liberally reject the hypothesis of measurement invariance,
we may choose the ordering that results in the largest test
statistic.  If there are many tied students, then there are many
possible orderings that must be examined.  However, examination of all
possible orderings remains computationally feasible because the
casewise first derivatives need to be calculated only once; the
derivatives simply need to be reordered for each ordering of tied
students.

A related issue involves the use of categorical demographic
variables to study measurement invariance; this may be conceptualized
as a situation when many students are tied.  In this case, groups are
already specified in advance, and so traditional methods (i.e.,
Likelihood Ratio tests) may suffice.
However, we could use the framework developed here to obtain an
alternative $\chi^2$ statistic testing measurement invariance.
Assume the observations are divided into $K$ categories $I_1, I_2,
\ldots, I_K$.
First, we can sum first derivatives within each category of the
demographic variable via:
\begin{equation}
    \label{eq:catsum}
    \Delta W_n(I_k) = \frac{1}{n_k\sqrt{n}} \sum_{i \in I_k}
    {\bm{I}}(\widehat{\bm{\theta}})^{-1/2} {\bm{\psi}}({\bm{y}}_i, \widehat{{\bm
      \theta}}),
\end{equation}
where $n_k$ is the sample size in category $k$ and $n$ is the overall
sample size.  This results in a $K \times p$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $K \times p$ matrix, summing over the squared entries in
the column to obtain a $\chi^2$ statistic with
$(K-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.

\subsection{Extensions}
The proposed family of tests can be extended to the simultaneous study
of measurement invariance with 
respect to multiple auxiliary variables and to other psychometric
models.  We consider each extension below.

% Recursive method
\subsubsection{Recursive Multiple-Groups Models}
The proposed tests have the ability to automatically define groups of
individuals violating measurement invariance with respect to a single
auxiliary (often demographic) variable.  If a researcher has multiple
auxiliary variables that he/she wishes to assess for measurement
invariance, then the proposed tests can be inserted into a recursive
algorithm 
that isolates groups of individuals violating measurement
invariance.  Such an algorithm is related to 
classification and regression trees \cite{BreFri84,MerSha10,StrMal09} and
C4.5 \cite{Qui93}.  The algorithm would proceed via the following
steps: 
\begin{enumerate}
  \item Fit a confirmatory factor analysis model to the data, with all
  parameters constrained to be equal across observations.
  \item Conduct a measurement invariance test with respect to each
    auxiliary variable (auxiliary variables are determined in
    advance by the researcher).
  \item If violations of measurement invariance are detected,
  split the data into two (or more) groups based on the auxiliary variable
    corresponding to the largest test statistic.
  \item Fit a multiple-groups factor analysis model based on step 3.
  \item Repeat steps 2--5.
\end{enumerate}
The algorithm would continue until it reaches a
predefined stopping rule.  For example, we may stop 
the algorithm when the tests fail to indicate measurement invariance
violations.  Alternatively, the algorithm may be stopped when the
sample size within any subgroup reaches a minimum level.

The algorithm described in the previous paragraph is related to one
developed by Zeileis et al.\ \citeyear{ZeiHot08} for representing
parameter invariance violations in generalized linear models.  Zeileis
et al.\ present the results in a tree structure, which is
suboptimal for factor analysis models.  Instead, the subgroups
discovered by the algorithm are used here to adaptively define a
multiple-groups factor analysis model.  The
simultaneous study of measurement invariance with respect to multiple
auxiliary variables is an important future direction.

\subsubsection{Other Latent Variable Models}
Aside from recursive measurement invariance detection, the proposed tests
readily extend to other popular psychometric models.  
Instead of studying measurement invariance in factor
analysis, the tests can be used to generally 
study the stability of structural equation model parameters across
observations.  For 
example, it would be possible to assess whether relationships between
latent variables are stronger for some students than for
others.  Secondly, 
the tests may be extended to study differential item functioning (DIF)
in item response models (e.g., \citeNP{KopZei10}, who focused on an
application to Rasch models).  
DIF methods are similar to those for factor analysis in
that subgroups must be specified in advance.  The tests proposed here
can detect subgroups automatically, however, offering a unified way of
studying both 
measurement invariance in factor analysis and differential item
functioning in item response.  While these
two methods have developed largely independently of one another,
some treatments \cite{Mcd99} and recent research \cite{StaChe06} have
sought to unify the methods.  Extensions of the proposed tests can
support this endeavor.

\subsection{Summary}
We have outlined a family of 
stochastic process-based parameter instability tests from theoretical
statistics to the issue of measurement invariance in psychometrics.
The paper included both theoretical development and 
study of the tests' performance.  The tests were found via simulation
to have good properties, making them useful for many psychometric
applications.  More generally, the tests help to
solve standing problems in
measurement invariance research and provide many avenues for
future research, both through extensions of the tests within a
factor-analytic context and through application of the tests to new
models.

\bibliography{list}

\end{document}

### 2010-10-04 MIpaper/par_instable.tex
Updated Type I error rates in simulation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@11 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2010-09-24 par_instable.tex
@@
 measurement invariance.  For the 0-standard error condition, we
-generated 5,000 datasets to get a more precise measure of Type I
-error.  In each dataset, half the students were 14--15
+generated 5,000 datasets because Type I
+error is more difficult to estimate precisely.  In each dataset, half the
+students were 14--15 
 years of age and half were 16--17 years of age.  As stated previously,
@@
 is the situation where the tests reject the hypothesis of measurement
-invariance at $\alpha=.05$ when none exist.  This column generally
+invariance at $\alpha=.05$ when none exist.  This column 
 shows that the Cram\'{e}r-von Mises test has accurate Type I error
-rates at smaller sample sizes, but it becomes slightly conservative at
-$n=500$.  The double-max test, on the other hand, is conservative at
+rates at all sample sizes examined.
+The double-max test, on the other hand, is conservative at
 all sample sizes examined.  Finally, the likelihood ratio test is
@@
 200 & CvM       & 0.05 & 0.19 & 0.64 & 0.96 \\ 
-    & DM        & 0.04 & 0.12 & 0.44 & 0.88 \\ 
+    & DM        & 0.03 & 0.12 & 0.44 & 0.88 \\ 
     & $\chi^2$  & 0.06 & 0.27 & 0.76 & 0.99 \\ \hline
-500 & CvM       & 0.03 & 0.44 & 0.98 & 1.00 \\ 
-    & DM        & 0.03 & 0.35 & 0.94 & 1.00 \\ 
+500 & CvM       & 0.05 & 0.44 & 0.98 & 1.00 \\ 
+    & DM        & 0.04 & 0.35 & 0.94 & 1.00 \\ 
     & $\chi^2$  & 0.05 & 0.58 & 1.00 & 1.00 \\\thickline

### 2010-11-05 MIpaper/par_instable.tex
Updated based on Z feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@12 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\abstract{Common methods for the study of measurement invariance in a
-  factor-analytic context include
+\abstract{The issue of measurement invariance commonly arises in
+  factor-analytic contexts, with methods for assessment
+  including 
   likelihood ratio tests, Lagrange multiplier tests, and
-  Wald tests. 
-  These methods require advance definition of the number of groups, group
+  Wald tests.  These tests all require advance definition of the number of
+  groups, group 
   membership, and offending model parameters.
@@
 general design and use of psychometric scales.
-A set of scales are defined to be measurement invariant with respect to
-an auxiliary variable $V$ if:
-\begin{equation}
-    \label{eq:midef}
-      f({\bf X} | T, V) = f({\bf X} | T),
-\end{equation}
-where ${\bf X}$ is a matrix of scale scores, $T$ is the latent
-variable that the scales purport to measure, and $f$ is some assumed
-distributional form.  This concept is 
+This concept is 
 particularly important because violations can render the scales
 useless.  
-That is, if a set of scales violates the condition described by
-Equation~\eqref{eq:midef}, then individuals with the same
-score on $T$ can systematically receive 
+That is, if a set of scales violates measurement invariance,
+then individuals with the same ``amount'' of a latent variable 
+may systematically receive 
 different scale scores.
@@
 
-Focusing on applications, measurement
-invariance is commonly assessed with respect to categorical variables
-such as ethnicity. 
-These assessments are often carried out using nested Multiple Groups
-Models (e.g., \citeNP{Jor71,Bol89}) along with Likelihood Ratio tests,
-though 
-Lagrange Multiplier and Wald tests may also be constructed for this
-purpose.
-Scales may also violate measurement invariance with
-respect to a continuous variable such as income.  In this situation,
-there are no groups, but the condition in Equation~\eqref{eq:midef}
-may still be violated.  The situation involving continuous auxiliary
-variables is directly addressed by the tests proposed in this paper.
-The tests proposed in this paper also have the ability to isolate
-specific model parameters that violate measurement invariance.
-Focusing on this point, Millsap's \citeyear{Mil05} review of 
-measurement invariance methods cites ``locating the violation of the
-invariance'' as one of the major outstanding problems.  The tests
-resulting from the proposed project directly address this issue,
-leading to an important advance in measurement invariance research.
-
-In the following pages, we first provide a brief review of Likelihood
-Ratio Tests, Lagrange Multiplier Tests, and Wald Tests.  We then
-proposed tests that generalize the Lagrange Multiplier
-tests, providing background and relevant theoretical results.  Next,
+In this paper, we consider a new family of tests for assessing
+measurement invariance that have important advantages over existing
+tests.  We begin by developing a general framework for the tests.
+This leads to a discussion of theoretical results relevant to the
+proposed tests, as well as a comparison of the proposed tests with
+existing tests.  Next,
 we study the 
@@
 
-\section{Existing Tests of Measurement Invariance}
-
-Measurement invariance is often studied in a factor-analytic context.
-For a $p \times 1$ vector of scales  
-${\bm x}$ and $m$ factors, the model is
-specified as:
-% Define data model via equation
-\begin{equation}
-    \label{eq:famod}
-    {\bm x}_i = {\bm \mu} + {\bm \lambda}{\bm z}_i + {\bm e}_i,\ \ \ i=1,\ldots,n,
-\end{equation}
-where ${\bm \mu}$ is a $p \times 1$ vector of scale means, ${\bm
-  \lambda}$ is a $p \times m$ matrix of factor loadings, ${\bm z}_i$ is
-an $m \times 1$ vector of factor scores for individual $i$, and ${\bm
-  e}_i$ is a $p \times 1$ vector of error terms for individual $i$.
-The model states that each scale is a
-linear combination of the latent factor(s) and error.  
-Multivariate normality of the ${\bm z}_i$ and of the ${\bm e}_i$ are
-traditionally assumed (e.g., \citeNP{LawMax71}), leading to a model
-that is often estimated via Maximum Likelihood methods.  We restrict
-ourselves to Maximum Likelihood estimation in this manuscript, leaving
-us with at least three statistical approaches to assessing measurement
-invariance: the Likelihood Ratio Test, the
-Lagrange Multiplier Test, and the Wald Test.  While these three tests
+\section{Framework}
+
+The tests proposed here are generally relevant to situations where the
+$p$-dimensional 
+data vector $\bm{x}_i, i=1,\cdots,n$ is specified to arise from a model with
+density $f(\bm{x}_i | \bm{\theta}_i)$ and log-likelihood 
+$\log L(\bm{\theta}_i | \bm{x}_i)$.  
+For the factor-analytic situations
+considered in this paper, the $\bm{x}_i$ consist of individuals' scale
+scores, and 
+$f()$ and $\log L()$ arise from the multivariate normal distribution.
+The density is defined as:
+\begin{equation}
+    \label{eq:mvndensity}
+    f(\bm{x}_i | \theta_i) = \frac{1}{(2\pi)^{p/2} |
+      \bm{\Sigma}(\bm{\theta}_i) |^{1/2}}\exp(-\frac{1}{2}(\bm{x}_i -
+    \bm{\mu}(\bm{\theta}_i))^{\prime} \bm{\Sigma}(\bm{\theta}_i)^{-1} (\bm{x}_i -
+    \bm{\mu}(\bm{\theta}_i)),
+\end{equation}
+where ${\bm{\mu}}({\bm{\theta}}_i)$ is the
+model-implied mean vector, and ${\bm{\Sigma}}({\bm{\theta}}_i)$ is the
+model-implied covariance matrix.
+The casewise log-likelihood is defined as:
+\begin{equation}
+    \label{eq:caselik}
+    \log \text{L}({\bm{x}}_i, {\bm{\theta}}_i) = %\sum_{i=1}^{N} 
+\log \mid 
+\bm{\Sigma}(\bm{\theta}) \mid + %\sum_{i=1}^{N} 
+(\bm{x}_i -
+\bm{\mu}(\bm{\theta}))^{\prime} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
+\bm{\mu}(\bm{\theta})).
+\end{equation}
+This equation is often summed over $n$ observations, and it is also
+often employed for factor analysis and structural
+equation models with missing data (e.g., \citeNP{Wot00}).  We use
+the likelihood here to facilitate the proposed tests.
+
+Within the general framework outlined above, model estimation
+often proceeds by assuming that $\bm{\theta}_i = \bm{\theta}\
+\forall\ i$, then choosing $\hat{\bm{\theta}}$ to satisfy the
+criterion:
+\begin{equation}
+    \label{eq:score}
+  \sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}})] = 0,    
+\end{equation}
+ where $\psi(\bm{x}_i, \bm{\theta})$ 
+is a score function based on the model.  
+For example, in factor analysis models estimated via Maximum
+Likelihood, the score function is the first derivative of $\log
+\text{L}(\bm{x}_i, \bm{\theta})$ with respect to $\bm{\theta}$.
+General M-estimation methods may also be handled within this
+framework, though we do not discuss them here.
+
+\section{Tests of Measurement Invariance}
+
+Measurement invariance specifically implies that factor analysis model
+parameters are invariant across individuals.  It can be framed as the
+hypothesis:
+\begin{equation}
+    \label{eq:h0}
+    H_0: {\bm \theta}_i = {\bm \theta}_0,\ i=1,\ldots,n,
+\end{equation}
+signifying that the parameter vector is the same for all individuals
+$i$.  
+This general hypothesis is very difficult to test because $n$ is large, and
+there are many ways by which the $\bm{\theta}_i$ may stray from one
+another.  To circumvent this problem, researchers often measure
+individuals on an auxiliary variable $V$ and divide them into groups
+based on that variable.  
+Assuming a continuous $V$, we might divide 
+individuals into two groups based on some threshold $\eta$.
+Equation~\eqref{eq:h0} then becomes:
+\begin{equation}
+    \label{eq:h0.cut}
+    H_0: \bm{\theta}_{V < \eta} = \bm{\theta}_{V > \eta},
+\end{equation}
+where, i.e., $\bm{\theta}_{V < \eta}$ reflects the factor analysis
+parameters for all individuals with $V < \eta$.
+
+Once we have grouped individuals based on $V$,
+factor-analytic assessments
+of measurement invariance are well developed.
+These assessments are often carried out using nested Multiple Groups
+Models (e.g., \citeNP{Jor71,Bol89}) coupled with Likelihood Ratio tests,
+though 
+Lagrange Multiplier and Wald tests may also be constructed for this
+purpose.    While these three tests
 are asymptotically equivalent \cite{Sat89}, each is detailed
 individually below.
+
+% MAY BE USEFUL SOMEWHERE
+% Focusing on the latter point, Millsap's \citeyear{Mil05} review of 
+% measurement invariance methods cites ``locating the violation of the
+% invariance'' as one of the major outstanding problems.  The tests
+% resulting from the proposed project directly address this issue,
+% leading to an important advance in measurement invariance research.
 
@@
 from the multivariate normal distribution and is defined as:
-\begin{equation}
-    \label{eq:loglik}
-    \log L( \bm{X}, \bm{\theta}) = -\frac{n}{2} \left \{ p
-          \log(2\pi) + \log | {\bm{\Sigma}}({\bm{\theta}}) | +
-          \text{tr}[{\bm{\Sigma}}^{-1}({\bm{\theta}})({\bm{S}} +
-          ({\overline{{\bm{X}}} -
-            {\bm{\mu}}({\bm{\theta}})})({\overline{{\bm{X}}} -
-            {\bm{\mu}}({\bm{\theta}})})^{\prime})]\right \},
-\end{equation}
-where $\overline{{\bm{X}}}$ is the observed mean vector, ${\bm S}$ is
-the observed covariance matrix, ${\bm{\mu}}({\bm{\theta}})$ is the
-model-implied mean vector, and ${\bm{\Sigma}}({\bm{\theta}})$ is the
-model-implied covariance matrix.  The test resulting from
+  The test resulting from
 Equation~\eqref{eq:chisq} has 
@@
 Estimating the restricted model, we define a vector of Lagrange
-multipliers $\widehat{\bm \lambda}$ that satisfy:
+multipliers $\widehat{\bm \kappa}$ that satisfy:
 \begin{equation}
@@
     \frac{\partial {\bm h}}{\partial {\bm \theta}}
-    \bigg |_{\widehat{\theta}_c}^{\prime} \widehat{{\bm \lambda}} = {\bm 0}.
+    \bigg |_{\widehat{\theta}_c}^{\prime} \widehat{{\bm \kappa}} = {\bm 0}.
 \end{equation}
 The test itself is then based on the asymptotic distribution of 
-$\widehat{{\bm \lambda}}$: if $\widehat{{\bm \lambda}}$ is sufficiently close
+$\widehat{{\bm \kappa}}$: if $\widehat{{\bm \kappa}}$ is sufficiently close
 to ${\bm 0}$, then the constraints are deemed to hold in the
 population.  The resulting test statistic, which is a function of
-$\widehat{\bm \lambda}$, follows a $\chi^2$ distribution for large
+$\widehat{\bm \kappa}$, follows a $\chi^2$ distribution for large
 samples.
@@
 
-\subsection{Relationships to Proposed Tests}
-All the tests described above require advance specification of
-subgroups that potentially violate measurement invariance.  This is a
-major limitation in psychometrics, where researchers
-ultimately want to ensure that tests are invariant with respect to all
-possible subgroups of students.  For example, there are many possible
-ways to define subgroups based on a continuous auxiliary variable like
-income.  Assuming two subgroups, we
-could define an arbitrary cutoff separating the
-poorer individuals from the richer individuals.  Alternatively, we could
-consider larger numbers of subgroups with similarly-arbitrary cutoffs
-defining the groups.  The ultimate subgroup definitions that we
+\subsection{Extensions to Proposed Tests}
+For continuous $V$, the three tests described above require the
+threshold $\eta$ to be known.
+  This is a
+major limitation in psychometrics, where $\eta$ is often unknown or
+undefined.
+For example, if $V$ represents yearly income, there are many possible
+values of 
+$\eta$ that could be used to divide individuals into poorer and richer
+groups.  The ultimate $\eta$ that we
 choose could potentially impact our
@@
 general statistical analyses \cite{MacZha02}.
-Heuristically, we would like to conduct a separate Lagrange (or
-Likelihood Ratio or Wald) test for each possible definition of
-subgroups, then declare the scales to be measurement invariant if they
-``pass'' all of these tests.  Given the large number of possible
-subgroups, however, this would generally result in problems with Type
-I errors.\footnote{An alternative to this procedure involves inserting the
-continuous income variable directly into the factor analysis model,
-which we do not pursue due 
-to the potential impact on model fit and on other parameter estimates.}
-
-  % Instead of this, can employ parameter instability tests from
-  % theoretical statistics to simultaneously examine all possible
-  % groupings.
-Instead of testing measurement invariance separately for each possible 
-subgroup definition, we can use a single test from the family of tests
-proposed here.
-The proposed tests can be viewed as generalizations of the
-Lagrange test \cite{And93}, 
-as they focus on a restricted model and 
-simultaneously test for measurement invariance across all
-possible subgroup definitions.  This removes the problems associated
-with error rates and generally promotes a more flexible framework
-for studying measurement invariance.  As previously stated, the proposed
-tests also have the ability to isolate specific model parameters violating
-measurement invariance.
+
+Instead of choosing a specific $\eta$, we may elect to test
+individuals across all possible values of $\eta$.  For example, we
+could conduct a Likelihood Ratio Test for every possible
+dichotomization of $V$, yielding a test statistic for each
+dichotomization.  The distribution of the maximum of these test
+statistics is derived in Andrews \citeyear{And93}, providing a test of
+measurement invariance over all values of $\eta$.  
+This is a special case of the family of tests proposed here.  The
+tests are designed for situations where
+$\eta$ is unknown or where a dichotomization of $V$ is insufficient to
+detect measurement invariance violations, and they can be applied both
+to individual model parameters and sets of model parameters.
+
+We now describe some general ideas for capturing measurement
+invariance violations with respect to a continuous auxiliary variable
+$V$.
 
@@
 estimated factor analysis model (one that restricts parameters to be
-invariant across subgroups).
-The traditional factor analysis likelihood function 
-(Equation~\eqref{eq:loglik}) 
-utilizes the observed covariance matrix
-${\bm{S}}$ and observed mean vector $\overline{\bm{X}}$
-instead of individuals' data, but the proposed tests
-require the likelihood to be evaluated for each individual.  Thus, 
-we utilize the equivalent casewise likelihood:
-\begin{equation}
-    \label{eq:caselik}
-    \log \text{L}^{\ast}({\bm{X}}, {\bm{\theta}}) = \sum_{i=1}^{N} \log \mid 
-\bm{\Sigma}(\bm{\theta}) \mid + \sum_{i=1}^{N} (\bm{x}_i -
-\bm{\mu}(\bm{\theta}))^{\prime} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
-\bm{\mu}(\bm{\theta})).
-\end{equation}
-This equation is often employed for factor analysis and structural
-equation models with missing data (e.g., \citeNP{Wot00}), 
-in which case we would take subsets of
-$\bm{\Sigma}(\bm{\theta})$ and $\bm{\mu}(\bm{\theta})$ corresponding
-to the observed data on each individual $i$.
-
-Maximization of Equation~\eqref{eq:caselik} typically involves iterative
-algorithms that seek values $\widehat{{\bm{\theta}}}$ such that
-$\frac{\partial}{\partial \theta_j} \log \text{L}^{\ast}({\bm{X}},
-{\bm{\theta}}) \big |_{\theta_j = \widehat{\theta}_j} = 0\ \ \forall\ j$.
+invariant across subgroups).  For the purposes of this paper, we
+assume that the model is estimated via Maximum Likelihood.  
+The factor analysis log-likelihood may be obtained by summing 
+Equation~\eqref{eq:caselik} across individuals.  
+Maximization of the likelihood then involves iterative 
+algorithms that seek values $\widehat{{\bm{\theta}}}$ to satisfy
+Equation~\eqref{eq:score}, where 
+$\psi(\bm{x}_i, \theta_j) = \frac{\partial}{\partial \bm{\theta}} \log
+\text{L}({\bm{x}}_i, 
+{\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}$.
 % insert derivatives of casewise evaluated at theta-hat = 0
-Such derivatives are taken across individuals, but, using the casewise
-likelihood described above, we can also 
+Such derivatives are taken across individuals, but, using individual
+terms of the log-likelihood (i.e., Equation~\eqref{eq:caselik}),
+we can also 
 evaluate the derivatives with respect to each individual.
@@
     \label{eq:casederiv}
-    \psi({\bm{x}}_i, \theta_j) = \frac{\partial}{\partial\theta_j}
-\left[ \log \mid 
-\bm{\Sigma}(\bm{\theta}) \mid + (\bm{x}_i -
-\bm{\mu}(\bm{\theta}))^{\text{T}} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
-\bm{\mu}(\bm{\theta}))\right] \bigg |_{\theta_j = \widehat{\theta}_j} = 0\ \ \ \forall\ i, j.
+    \psi({\bm{x}}_i, \bm{\theta}) = \frac{\partial}{\partial \bm{\theta}} \log
+\text{L}({\bm{x}}_i, 
+{\bm{\theta}}) \bigg |_{\bm{\theta} = \widehat{\bm{\theta}}} = \bm{0}\ \ \ \forall\ i.
 \end{equation}
 In practice, no value of $\widehat{{\bm{\theta}}}$ will maximize all
-individuals' likelihoods, but, if parameter invariance is satisfied, 
+individuals' likelihoods, but, if measurement invariance is satisfied, 
 we can expect the $\psi(\cdot, \cdot)$ to fluctuate around zero.  This
@@
 \subsection{Theory}
-Theoretical work in statistics (e.g., \citeNP{HjoKon02}) has shown
+Theoretical work in statistics (e.g.,
+\citeNP{BroDur75,HjoKon02,Nyb89}) has shown 
 that functions of the 
@@
 invariance.  
-That is, letting ${\bm \theta}_i$ reflect parameter values
-specifically 
-associated with individual $i$, measurement invariance can be 
-described by the null hypothesis:
-\begin{equation}
-    \label{eq:h0}
-    H_0: {\bm \theta}_i = {\bm \theta}_0,\ i=1,\ldots,n,
-\end{equation}
-where the $i$'s are typically ordered with respect to the auxiliary
-variable of interest (i.e., the variable defining subgroups for tests
-of measurement invariance).  For example, if parental income was the
-auxiliary variable, we would order students according to parental
-income and then examine whether parameters were constant across
-students.  The ordering of individuals is crucial to
-the test development below, but it is relaxed later in the proposal.
-
-Despite addressing measurement
-invariance, the above $H_0$ focuses no longer on subgroups of
-individuals but rather on individuals themselves.  The $H_0$ is
-generally applicable to situations where parameters may vary across
-observations.  For example, it has most often been examined in
-longitudinal models where observations are 
-ordered according to time.
-%In this context, {\em{change-point
-%    analyses}} have received much attention from the psychometric
-%community \cite{Coh08}. % CITE BOOK
-Observations may also be ordered with respect to 
-continuous variables other than time, a fact that allows for the
-measurement invariance tests described here.  Thus, the proposed tests
-stem from seminal work on
-detecting turning points in longitudinal models (e.g.,
-\citeNP{BroDur75,Nyb89}).
-
-The alternative hypothesis corresponding to the above $H_0$ is that not
-all the ${\bm{\theta}}_i$ are constant (i.e., measurement invariance is
-violated), though specific local alternatives may also be specified to
-make the tests more powerful.  We now review some of the main results
-for testing $H_0$ via stochastic processes of first derivatives; more
+We rely on the null hypothesis specified in Equation~\eqref{eq:h0},
+which focuses on individuals instead of groups of individuals.
+We now review some of the main results
+for testing this $H_0$; more
 detailed accounts include Hjort and Koning \citeyear{HjoKon02} and
@@
 $H_0$:
-%which generalizes earlier results from %CITE bigger journals:
-% Z+H Theorem 2/Corollary 2
 \begin{equation}
@@
 $p$-dimensional Brownian bridge, and 
-$\overset{d}{\rightarrow}$ reflects convergence in distribution. %?
+$\overset{d}{\rightarrow}$ reflects convergence in distribution.
 In words, there are $p$ cumulative score processes, one for each
@@
 cumulative score processes, such that each univariate process (i.e., each
-process for a single model parameter) is independent of all other
-processes.
+process for a single model parameter) is unrelated to all other
+processes (and is asymptotically independent of the other processes).
 
@@
     cumulative score process}}, which we will denote as
-${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$.
+${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$.
 Each column of this matrix follows a univariate Brownian bridge and is
@@
 To carry out a test of $H_0$, we need functions that collapse across
-rows (individuals) and/or columns (parameters) of the matrix.  These
+rows (individuals) and/or columns (parameters) of the matrix.  
+These
 functions then lead to scalar test statistics and p-values.
@@
 functions include the absolute maximum or mean of the corresponding 
-column in ${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$
+column in ${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$
 \cite{ZeiHor07}. % (see ZH07 p 7)
@@
 include the maximum norm or the squared Euclidean norm of the
-corresponding row in ${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$
+corresponding row in ${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$
 \cite{HjoKon02}.
 
-A general, two-step strategy for testing measurement
-invariance involves first collapsing over columns for each individual,
-yielding a single statistic for each individual. 
-These $n$ statistics (ordered according to an auxiliary variable)
-can then be graphed to visually define groups that violate measurement
-invariance.  Functions of these $n$ statistics can also yield scalar
-test statistics.
-
-As a concrete example of the above strategy, we could
-first collapse over columns for each individual $i$ via the Euclidean
-norm:
+% A general, two-step strategy for testing measurement
+% invariance involves first collapsing over columns for each individual,
+% yielding a single statistic for each individual. 
+% These $n$ statistics (ordered according to an auxiliary variable)
+% can then be graphed to visually define groups that violate measurement
+% invariance.  Functions of these $n$ statistics can also yield scalar
+% test statistics.
+
+As a concrete example of these ideas, consider the situation
+where we carry out a Likelihood Ratio test for all possible values of
+$\eta$, then take the maximum of these test statistics.  A related
+idea here involves taking the maximum of the empirical cumulative
+score process:
+\begin{equation}
+    \label{eq:doubmax}
+    DM = \max_{i=1,\ldots,n} \max_{j=1,\ldots,p} |
+    B^{\ast}_{ij}(\widehat{\bm{\theta}}) |,
+\end{equation}
+where $B^{\ast}_{ij}(\widehat{\bm{\theta}})$ represents the
+entry in row $i$ and column $j$ of
+${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$.
+This is a 
+{\em{double-max}} statistic that allows for simultaneous isolation of 
+$\eta$ along with model parameters that change over $\eta$.
+This statistic is
+especially useful for visualization, as the fluctuation process for
+each individual parameter can be displayed along with the appropriate
+critical value.  An example of this visualization appears in the
+Example section.
+
+As a second example of ``collapsing'' functions,
+we could first collapse over columns for each individual $i$ via the
+Euclidean norm:
 \begin{equation}
     \label{eq:meanl2}
-    \lambda_i = \lVert {\bm B}^{\text{emp}}_{i\cdot}(\widehat{\bm \theta}) \rVert,
+    \nu_i = \lVert {\bm B}^{\ast}_{i\cdot}(\widehat{\bm \theta}) \rVert,
 \end{equation}
@@
 fluctuation process matrix.  A univariate test statistic may then be
-obtained by taking the mean of the $\lambda_i$.  This is a
+obtained by taking the mean of the $\nu_i$.  This is a
 Cram\'{e}r-von Mises statistic described by, e.g., Nyblom
@@
 % to each row vector.
-As an alternative to the two-step approach described in the previous
-paragraph (collapse over columns, then take the absolute maximum), the
-{\em{double-max}} statistic allows for simultaneous isolation of 
-individuals violating invariance along with offending model
-parameters (that are unspecified in advance).  These functions take
-the maximum absolute cumulative sum across all entries:
-\begin{equation}
-    \label{eq:doubmax}
-    DM = \max_{i=1,\ldots,n} \max_{j=1,\ldots,p} |
-    B^{\text{emp}}_{ij}(\widehat{\bm{\theta}}) |,
-\end{equation}
-where $B^{\text{emp}}_{ij}(\widehat{\bm{\theta}})$ represents the
-entry in row $i$ and column $j$ of
-${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$.  This is different from
-the two-step approach because, in the two-step approach, we collapse
-over parameters (columns) first.  We then obtain a test statistic
-from an $n \times 1$ vector.  In contrast, the
-$DM$ statistic takes an absolute maximum with respect to the $n \times
-p$ empirical cumulative score process (i.e., with respect to
-cumulative sums associated with individual parameters).  This is
-especially useful for visualization, as the fluctuation process for
-each individual parameter can be displayed along with the appropriate
-critical value.  An example of this visualization appears in the
-Example section.
-
 There exist other functions of first derivatives and of the
@@
 
-We first focus on a univariate
-Brownian bridge, which could test a single parameter (say, $\theta_j$) for
-measurement invariance using the test statistic $DM_j$.
-Under $H_0$, it can be shown that (e.g., \citeNP{ShoWel86}):
+Focusing on a univariate
+Brownian bridge under $H_0$, it can be shown that (e.g., \citeNP{ShoWel86}):
  % see Beghin & Orsingher Eq (4.12).  They cite Shorack &
@@
 so, in practice, we only need to compute the first 100 terms (say).
-
-For multidimensional Brownian bridges whose dimensions are independent
-(i.e., those that 
-have been ``decorrelated'' by the information matrix), $p$-values can
-be obtained similarly.  Assuming a Brownian bridge with $p$
-independent dimensions (i.e., simultaneously testing $p$ model
-parameters for measurement invariance), we have:
-\begin{eqnarray}
-    \label{eq:multidim}
-    p & = & P(DM_1 >
-    \beta\ \cup\ DM_2 > \beta\ \cup\ \ldots \cup\ DM_p > \beta | H_0)
-      \nonumber \\
-   & = & 1 - P(DM_1 <
-    \beta\ \cap\ DM_2 < \beta\ \cap\ \ldots \cap\ DM_p < \beta | H_0)
-     \nonumber \\
-   & = & 1 - \left ( 1 + 2 \sum_{h=1}^{\infty} (-1)^h
-       \exp(-2h^2\beta^2) \right )^p.
-\end{eqnarray}
-These expressions are used to compute $p$-values for the double-max
-test statistic, with associated critical values being
-obtained, e.g., through a simple grid search using
-Equation~\eqref{eq:multidim}.
-For other test statistics, we can either obtain similar expressions
-for $p$-values or obtain $p$-values via simulation of Brownian
+For multidimensional Brownian bridges whose dimensions have been
+``decorrelated'', $p$-values can
+be obtained similarly with a Bonferroni correction.
+
+For the Cram\'{e}r-von Mises test statistic, analytic expressions for
+$p$-values are more cumbersome.  Nyblom \citeyear{Nyb89} and Hansen
+\citeyear{Han92} provide small tables of critical values, and other
+critical values are easily approximated via simulation of Brownian
 bridges.
@@
 If a test described above implies a measurement invariance violation,
-then the researcher may want to: (1) define subgroups of individuals that
-the scales measure differently; or (2) isolated model
+then the researcher may want to: (1) estimate $\eta$, so that she may
+define subgroups of individuals impacted by measurement invariance;
+or (2) isolate model
 parameters affected by the invariance.  
 In the context of a Cram\'{e}r-von Mises test, we can accomplish the
-first task by graphing the $\lambda_i$ (Equation~\eqref{eq:meanl2})
+first task by graphing the $\nu_i$ (Equation~\eqref{eq:meanl2})
 and noting the location of the peaks and valleys.  We can accomplish
 the second task by taking the norm in Equation~\eqref{eq:meanl2}
-across specific columns of $B^{\text{emp}}(\widehat{\bm \theta})$.
+across specific columns of $B^{\ast}(\widehat{\bm \theta})$.
 
 In the context of a double-max test, we can accomplish the first task
-by defining subgroups
-based on the individual who achieves the absolute
-maximum.  That is, assuming the existence of two subgroups
-violating measurement invariance, the absolute maximum cumulative
-score should be achieved close to 
-the ``cut point'' between subgroups.  For more than two subgroups, we
+by defining $\eta$ to be the value
+of $V$ at which the absolute
+maximum cumulative score is achieved.  This value maximizes the
+segmented likelihood, defined as:
+\begin{equation}
+    \label{eq:seglik}
+    \sum_{i; V_i \leq \eta} L(\bm{x}_i, \bm{\theta}_{V \leq \eta}) + 
+     \sum_{i; V_i > \eta} L(\bm{x}_i, \bm{\theta}_{V > \eta}).
+\end{equation}
+where $\bm{\theta}_{V \leq \eta}$ and $\bm{\theta}_{V > \eta}$ are ML estimates
+for individuals with 
+$V_i \leq \eta$ and with $V_i > \eta$, respectively.
+  For multiple thresholds, we
 could heuristically define groups based on the number of peaks and
-valleys in the graph.
-We can accomplish the second task by conducting posthoc
-tests on specific model parameters.  These ideas are illustrated in
-the example below, which focuses on the double-max test.
+valleys in the graph, or we could maximize a segmented likelihood
+similar to that in Equation~\eqref{eq:seglik}.
+We can accomplish the second task (isolating model parameters) by
+conducting posthoc 
+tests on specific model parameters.  
+%Some of these ideas are illustrated in
+%the example below.
 
 \begin{figure}
-\caption{Path diagram representing the factor analysis model used for
-  the Example and Simulations.}
+\caption{Path diagram representing the base factor analysis model used for
+  the Example and Simulations.  To induce measurement invariance
+  violations, a seventh observed variable (student age) determines the
+values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
+$\lambda_{31}$).}
 \label{fig:famod}
@@
 study whether the scales are measurement invariant with
-respect to the full range of ages.
+respect to age, which is taken to be the auxiliary variable $V$.
 
@@
 To formally represent these ideas, we specify that the data arise from
-a factor analysis model with two factors.  As displayed in
-Figure~\ref{fig:famod},
-three scales arise from the verbal factor and three 
-scales arise from the mathematical second factor.
-A seventh observed variable (not pictured in Figure~\ref{fig:famod}),
-student age, determines the values of 
+a factor analysis model with two factors.  The base model, displayed in
+Figure~\ref{fig:famod}, specifies that measurement invariance holds,
+with three scales arising from the verbal factor and three 
+scales arising from the mathematical second factor.
+For measurement invariance violations, we specify that $V$
+(student age) impacts the values of 
 verbal factor loadings in the model: 
@@
 factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
-Figure~\ref{fig:famod}.  If students are between 14 or 15 years of age,
+Figure~\ref{fig:famod}.  If students are 14 or 15 years of age,
 however, then the factor loadings corresponding to the first factor
@@
 double-max statistic via Equation~\eqref{eq:doubmax} and associated
-$p$-value via Equation~\eqref{eq:multidim}.  Note that, because we
+$p$-value via Equation~\eqref{eq:absmax_p}.  Note that, because we
 were examining only the three verbal factor loading parameters, the
@@
 \subsection{Results}
-In the results section, we first describe overall results and
-describe isolation of groups and model parameters violating
+In the results section, we first describe overall results and then 
+describe estimation of $\eta$ and 
+isolation of model parameters violating
 measurement invariance.
@@
 for each factor loading, with student number (ordered by age and
-numbered from 1 to 200, but divided by 200 to correspond 
-to a Brownian bridge) on the x-axis and
+rescaled to (0,1)) on the x-axis and
 absolute values of the cumulative scores on the y-axis.  The dashed horizontal
@@
 
-\subsubsection{Group and Parameter Isolation}
+\subsubsection{Estimation of $\eta$ and Parameter Isolation}
 The results of the test imply that the 
 verbal scales lack measurement invariance, and the age at which the
-maximum is achieved allows us to define subgroups violating
-measurement invariance.  The student corresponding to the maximum
+maximum is achieved provides the estimate $\hat{\eta}$.
+The student corresponding to the maximum
 absolute cumulative sum (the solid circle in the third graph) is 15.8
@@
 students below 15.8 years and one above 15.8 years.  This agrees well
-with the true subgroup cut point of 16.0 years.
+with the true $\eta$ of 16.0 years.
 
@@
 $.05/3=.017$.  This amounts to the same critical value as that used in
-the overall test (as shown by Equation~\eqref{eq:multidim}),
+the overall test, 
 meaning that we simply examine the individual processes that cross
-the horizontal dashed lines in Figure~\ref{fig:cusumex}.  The Figure
+the horizontal dashed lines in Figure~\ref{fig:cusumex}.  The figure
 shows that the second and third parameters (i.e., $\lambda_{21}$ and
@@
 error, which implicitly brings into question the posthoc tests' power.  We
-generally address the issue of power later in the simulations below.
+generally address the issue of power in the simulations below.
 
@@
 examine the power and error rates of three tests: the double-max test, the
-Cram\'{e}r-von Mises test, and the traditional $\chi^2$ likelihood
+Cram\'{e}r-von Mises test, and the traditional likelihood
 ratio (difference) test involving nested 
-multiple-group models.  The $\chi^2$ test has an advantage over the
+multiple-group models.  The likelihood ratio test has an advantage over the
 other two tests here, in that it requires advance definition of
-groups violating measurement invariance (we supply it with the known
-subgroups here).  This test is useful for comparative purposes,
+$\eta$.  This test is useful for comparative purposes,
 however, as it helps indicate the extent to which the other tests
-``lose'' power due to lack of subgroup information.
+``lose'' power due to $\eta$ being unknown.
 
@@
 the Example section.
-These data included verbal scales that violated measurement invariance
+The data included verbal scales that violated measurement invariance
 with respect to students' ages, with students 14--15 years of age
@@
 measurement invariance.  For the 0-standard error condition, we
-generated 5,000 datasets because Type I
-error is more difficult to estimate precisely.  In each dataset, half the
+generated 5,000 datasets because small proportions are more
+ difficult to estimate precisely.  In each dataset, half the
 students were 14--15 
@@
 group
-designations (below 16.0 years, above 16.0 years) were assumed to be
+designations ($\eta$) were assumed to be
 known for the likelihood ratio test.
@@
 This is likely due to the fact that the likelihood ratio test had
-extra information (the subgroup definitions), whereas the other tests
-were essentially testing measurement invariance over all possible
-subgroups.  While power was lower for the tests developed in this
+extra information ($\eta$), whereas the other tests
+were testing measurement invariance over all values
+of $\eta$.  While power was lower for the tests developed in this
 paper, it was not terribly low.  At $n=200$ and above, all tests were
@@
 \citeyear{ZeiSha10}, stating that the Cram\'{e}r-von Mises test
-is generally superior when multiple parameters change across
-subgroups.  Because the three verbal factor loadings all changed
+is generally superior when multiple parameters change simultaneously.
+Because the three verbal factor loadings all changed
 across the two subgroups in the simulation, the Cram\'{e}r-von Mises
@@
 \section{Discussion}
-In this paper, we have presented a new class of statistical tests for
+In this paper, we have presented a new family of statistical tests for
 the study of measurement invariance in psychometrics.  The tests,
 based on stochastic processes, have reasonable power, can isolate
-groups of individuals violating measurement invariance, and can
+groups of individuals violating measurement invariance based on a
+continuous, auxiliary variable, and can
 isolate specific model parameters affected by the violation.  In this
@@
 The proposed tests give researchers a new set of tools for studying 
-measurement invariance.  First, they give researchers the
+measurement invariance.  For example, they give researchers the
 opportunity to simultaneously test for many types of measurement
-invariance (see, e.g., Meredith \citeNP{Mer93}).  The traditional
+invariance (see, e.g., \citeNP{Mer93}).  The traditional
 steps have involved hypothesizing a specific type of invariance and
@@
 the nature of the invariance violation.  This is made possible through
-the tests' abilities to identify subgroups of individuals violating
-measurement invariance.  While subgroup identification was
-accomplished in this paper via visualization, it is also possible to
-define 
-formal rules for identifying subgroups.  This idea leads us to the
-next section, which describes further issues associated with the
-tests' application to the study of measurement invariance.
-
-\subsection{Further Issues}
-Further issues associated with the proposed tests include: (1) formal
-definition of subgroups violating measurement invariance based on the
-results of the statistical test; and (2) dealing
-with ties on the auxiliary variable.  We address each in turn below.
-
-% More than two groups violating MI
-\subsubsection{Subgroup Definition}
-Given rejection of the hypothesis of measurement invariance, the most
-straightforward way to formally identify offending groups is through
-the location of the maximum of the cumulative score process.  Assuming two
-subgroups, the maximum should be achieved close to the split point
-between the the groups.  Referring to the Example (see
-Figure~\ref{fig:cusumex}), we would split students at the
-age corresponding to the solid dot in the third panel.
-
-To study this method for group identification in more detail, we used
-results of our simulation to examine the age at which the maximum of
-the cumulative score process occurs.
-If the test accurately recovers the violating subgroups,
-then the maximum should be close to 16 years
-of age.\footnote{The data generated for these
-  simulations 
-  contained ages measured in days, as opposed to years.  Thus,
-  the tests could split students at continuous values like 15.6 or
-  16.3 years of age.}
-  For each simulation condition, Table~\ref{tab:age1}
-  displays confidence intervals for the location of the maximum
-  absolute cumulative score.  These confidence intervals were
-  calculated across generated 
-  datasets, with each dataset contributing one maximum absolute cumulative
-  score
-  (regardless of whether the hypothesis of
-  measurement invariance was rejected).
-It is observed that these confidence intervals slightly underestimate
-the true age dividing the subgroups, with the degree of bias
-decreasing as sample size increases.  The interval widths
-are impacted by both the magnitude of the invariance and the sample
-size.  In future work, we plan to study alternative ways of
-identifying subgroups.
-
-\begin{table*}
-\caption{Confidence intervals for the location of the maximum absolute
-  cumulative score across the conditions from the simulation.}
-\label{tab:age1}
-  \begin{tabular}{lccc} \thickline
-  & \multicolumn{3}{c}{Violation Magnitude} \\
-  $n$ & 1SE & 2SE & 3SE \\\hline
-  100 & (15.61; 15.73) & (15.55; 15.65) & (15.50; 15.59) \\
-  200 & (15.67; 15.77) & (15.64; 15.72) & (15.67; 15.73) \\
-  500 & (15.71; 15.81) & (15.77; 15.83) & (15.80; 15.83) \\\thickline
-  \end{tabular}
-\end{table*}
+the tests' abilities to estimate $\eta$, the threshold dividing
+individuals into subgroups that violate
+measurement invariance.  While a single $\eta$ was estimate in this
+paper, it is also possible to define 
+formal rules for estimating multiple $\eta$ parameters.  We plan to
+explore these ideas in the future.
+
+% \subsection{Further Issues}
+% Further issues associated with the proposed tests include: (1)
+% alternative methods for estimating $\eta$;
+% and (2) dealing
+% with categorical $V$.  We address each in turn below.
+
+% % More than two groups violating MI
+% \subsubsection{Estimation of $\eta$}
+% Given rejection of the hypothesis of measurement invariance, the most
+% straightforward way to formally estimate $\eta$ is through
+% the location of the maximum of the cumulative score process.  Assuming
+% a single threshold, the maximum should be achieved close to the true
+% $\eta$.  Referring to the Example (see
+% Figure~\ref{fig:cusumex}), we defined $\eta$ as the 
+% age corresponding to the solid dot in the third panel.
+
+% To study this method in more detail, we used
+% results of our simulation to examine the age at which the maximum of
+% the cumulative score process occurs.
+% If the test accurately recovers the violating subgroups,
+% then the maximum should be close to 16 years
+% of age.\footnote{The data generated for these
+%   simulations 
+%   contained ages measured in days, as opposed to years.  Thus,
+%   the tests could split students at continuous values like 15.6 or
+%   16.3 years of age.}
+%   For each simulation condition, Table~\ref{tab:age1}
+%   displays confidence intervals for the location of the maximum
+%   absolute cumulative score.  These confidence intervals were
+%   calculated across generated 
+%   datasets, with each dataset contributing one maximum absolute cumulative
+%   score
+%   (regardless of whether the hypothesis of
+%   measurement invariance was rejected).
+% It is observed that these confidence intervals slightly underestimate
+% the true age dividing the subgroups, with the degree of bias
+% decreasing as sample size increases.  The interval widths
+% are impacted by both the magnitude of the invariance and the sample
+% size.  In future work, we plan to study alternative ways of
+% identifying subgroups.
+
+% \begin{table*}
+% \caption{Confidence intervals for the location of the maximum absolute
+%   cumulative score across the conditions from the simulation.}
+% \label{tab:age1}
+%   \begin{tabular}{lccc} \thickline
+%   & \multicolumn{3}{c}{Violation Magnitude} \\
+%   $n$ & 1SE & 2SE & 3SE \\\hline
+%   100 & (15.61; 15.73) & (15.55; 15.65) & (15.50; 15.59) \\
+%   200 & (15.67; 15.77) & (15.64; 15.72) & (15.67; 15.73) \\
+%   500 & (15.71; 15.81) & (15.77; 15.83) & (15.80; 15.83) \\\thickline
+%   \end{tabular}
+% \end{table*}
 
@@
 % Ties on auxiliary variable (mention use in categorical contexts)
-\subsubsection{Ties on the Auxiliary Variable}
-There remain unaddressed theoretical issues specific to the 
-tests' application to measurement invariance.  These
-issues generally stem from the types of auxiliary variables on which
-researchers may study measurement invariance.  In
-the example described earlier, researchers wanted to know whether a
-scale was measurement invariant with respect to student age.
-Students could then be ordered based on age, and a test could be 
-constructed via Equation~\eqref{eq:cumscore}.  However, an issue arises if there
-are ties on age; that is, if two students' ages are exactly
-the same.  The proposed tests assume complete ordering of the
-students, but these two students cannot be ordered.  In this case, we
-may randomly order the two students, but there is an
-arbitrariness associated with this strategy.  Alternatively, to obtain
-tests that liberally reject the hypothesis of measurement invariance,
-we may choose the ordering that results in the largest test
-statistic.  If there are many tied students, then there are many
-possible orderings that must be examined.  However, examination of all
-possible orderings remains computationally feasible because the
-casewise first derivatives need to be calculated only once; the
-derivatives simply need to be reordered for each ordering of tied
-students.
-
-A related issue involves the use of categorical demographic
-variables to study measurement invariance; this may be conceptualized
-as a situation when many students are tied.  In this case, groups are
+\subsection{Categorical Auxiliary Variable}
+% There remain unaddressed theoretical issues specific to the 
+% tests' application to measurement invariance.  These
+% issues generally stem from the types of auxiliary variables on which
+% researchers may study measurement invariance.  In
+% the example described earlier, researchers wanted to know whether a
+% scale was measurement invariant with respect to student age.
+% Students could then be ordered based on age, and a test could be 
+% constructed via Equation~\eqref{eq:cumscore}.  However, an issue arises if there
+% are ties on age; that is, if two students' ages are exactly
+% the same.  The proposed tests assume complete ordering of the
+% students, but these two students cannot be ordered.  In this case, we
+% may randomly order the two students, but there is an
+% arbitrariness associated with this strategy.  Alternatively, to obtain
+% tests that liberally reject the hypothesis of measurement invariance,
+% we may choose the ordering that results in the largest test
+% statistic.  If there are many tied students, then there are many
+% possible orderings that must be examined.  However, examination of all
+% possible orderings remains computationally feasible because the
+% casewise first derivatives need to be calculated only once; the
+% derivatives simply need to be reordered for each ordering of tied
+% students.
+
+One issue that was largely unaddressed in this paper involved the use
+of categorical $V$ to study measurement invariance.  In this case,
+groups are 
 already specified in advance, and so traditional methods (i.e.,
@@
 \ldots, I_K$.
-First, we can sum first derivatives within each category of the
+First, we can sum scores within each category of the
 demographic variable via:
@@
 \subsection{Extensions}
-The proposed family of tests can be extended to the simultaneous study
-of measurement invariance with 
-respect to multiple auxiliary variables and to other psychometric
-models.  We consider each extension below.
+The proposed family of tests can be extended in various ways.  First,
+it is possible to construct an algorithm that recursively defines
+groups of individuals violating measurement invariance with respect to
+multiple auxiliary variables.  Such an algorithm is related to 
+classification and regression trees \cite{BreFri84,MerSha10,StrMal09},
+and related algorithms have been developed for generalized linear
+models \cite{ZeiHot08} and Rasch models \cite{KopZei10}.
 
 % Recursive method
-\subsubsection{Recursive Multiple-Groups Models}
-The proposed tests have the ability to automatically define groups of
-individuals violating measurement invariance with respect to a single
-auxiliary (often demographic) variable.  If a researcher has multiple
-auxiliary variables that he/she wishes to assess for measurement
-invariance, then the proposed tests can be inserted into a recursive
-algorithm 
-that isolates groups of individuals violating measurement
-invariance.  Such an algorithm is related to 
-classification and regression trees \cite{BreFri84,MerSha10,StrMal09} and
-C4.5 \cite{Qui93}.  The algorithm would proceed via the following
-steps: 
-\begin{enumerate}
-  \item Fit a confirmatory factor analysis model to the data, with all
-  parameters constrained to be equal across observations.
-  \item Conduct a measurement invariance test with respect to each
-    auxiliary variable (auxiliary variables are determined in
-    advance by the researcher).
-  \item If violations of measurement invariance are detected,
-  split the data into two (or more) groups based on the auxiliary variable
-    corresponding to the largest test statistic.
-  \item Fit a multiple-groups factor analysis model based on step 3.
-  \item Repeat steps 2--5.
-\end{enumerate}
-The algorithm would continue until it reaches a
-predefined stopping rule.  For example, we may stop 
-the algorithm when the tests fail to indicate measurement invariance
-violations.  Alternatively, the algorithm may be stopped when the
-sample size within any subgroup reaches a minimum level.
-
-The algorithm described in the previous paragraph is related to one
-developed by Zeileis et al.\ \citeyear{ZeiHot08} for representing
-parameter invariance violations in generalized linear models.  Zeileis
-et al.\ present the results in a tree structure, which is
-suboptimal for factor analysis models.  Instead, the subgroups
-discovered by the algorithm are used here to adaptively define a
-multiple-groups factor analysis model.  The
-simultaneous study of measurement invariance with respect to multiple
-auxiliary variables is an important future direction.
-
-\subsubsection{Other Latent Variable Models}
+% \subsubsection{Recursive Multiple-Groups Models}
+% The proposed tests have the ability to automatically define groups of
+% individuals violating measurement invariance with respect to a single
+% auxiliary (often demographic) variable.  If a researcher has multiple
+% auxiliary variables that he/she wishes to assess for measurement
+% invariance, then the proposed tests can be inserted into a recursive
+% algorithm 
+% that isolates groups of individuals violating measurement
+% invariance.   and
+% C4.5 \cite{Qui93}.  The algorithm would proceed via the following
+% steps: 
+% \begin{enumerate}
+%   \item Fit a confirmatory factor analysis model to the data, with all
+%   parameters constrained to be equal across observations.
+%   \item Conduct a measurement invariance test with respect to each
+%     auxiliary variable (auxiliary variables are determined in
+%     advance by the researcher).
+%   \item If violations of measurement invariance are detected,
+%   split the data into two (or more) groups based on the auxiliary variable
+%     corresponding to the largest test statistic.
+%   \item Fit a multiple-groups factor analysis model based on step 3.
+%   \item Repeat steps 2--5.
+% \end{enumerate}
+% The algorithm would continue until it reaches a
+% predefined stopping rule.  For example, we may stop 
+% the algorithm when the tests fail to indicate measurement invariance
+% violations.  Alternatively, the algorithm may be stopped when the
+% sample size within any subgroup reaches a minimum level.
+
+% The algorithm described in the previous paragraph is related to one
+% developed by Zeileis et al.\ \citeyear{ZeiHot08} for representing
+% parameter invariance violations in generalized linear models.  Zeileis
+% et al.\ present the results in a tree structure, which is
+% suboptimal for factor analysis models.  Instead, the subgroups
+% discovered by the algorithm are used here to adaptively define a
+% multiple-groups factor analysis model.  The
+% simultaneous study of measurement invariance with respect to multiple
+% auxiliary variables is an important future direction.
+
+%\subsubsection{Other Latent Variable Models}
 Aside from recursive measurement invariance detection, the proposed tests

### 2010-11-17 www/def.model.R
Added code files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@14 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,98 @@
+def.model <- function (fadat, age)
+{
+    dat1 <- fadat[age < 16, ]
+    dat2 <- fadat[age > 16, ]
+    TwoFA <- mxModel("Two Factor Model",
+                     mxModel("group1", 
+                             type = "RAM",
+                             mxData(observed = dat1, type = "raw"),
+                             manifestVars = c("x1", "x2", "x3",
+                               "y1", "y2", "y3"), 
+                             latentVars = c("F1", "F2"),
+                             mxPath(from = c("x1", "x2", "x3",
+                                             "y1", "y2", "y3"),
+                                    arrows = 2,
+                                    free = TRUE, 
+                                    values = c(1, 1, 1, 1, 1, 1),
+                                    labels = c("e1", "e2", "e3",
+                                               "e4", "e5", "e6"),
+                                    lbound=0),
+                             mxPath(from = c("F1", "F2"),
+                                    arrows = 2,
+                                    all = TRUE,
+                                    free = c(FALSE, TRUE, TRUE, FALSE),
+                                    values = c(1, 0.5, 0.5, 1),
+                                    labels = c("varF1", "cov", "cov", "varF2"),
+                                    lbound=c(NA,0,0,NA)),
+                             mxPath(from = "F1",
+                                    to = c("x1", "x2", "x3"),
+                                    arrows = 1,
+                                    free = TRUE,
+                                    values = c(1, 1, 1),
+                                    labels = c("l1", "l2", "l3"),
+                                    lbound=0),
+                             mxPath(from = "F2", 
+                                    to = c("y1", "y2", "y3"),
+                                    arrows = 1,
+                                    free = TRUE, 
+                                    values = c(1, 1, 1),
+                                    labels = c("l4", "l5", "l6"),
+                                    lbound=0), 
+                             mxPath(from = "one",
+                                    to = c("x1", "x2", "x3", "y1", "y2", "y3",
+                                           "F1", "F2"),
+                                    arrows = 1,
+                                    free = c(TRUE, TRUE, TRUE, TRUE,
+                                             TRUE, TRUE, FALSE, FALSE),
+                                    values = c(1, 1, 1, 1, 1, 1, 0, 0),
+                                    labels = c("meanx1", "meanx2", "meanx3",
+                                      "meany1", "meany2", "meany3", NA, NA))), 
+                     mxModel("group2",
+                             type = "RAM",
+                             mxData(observed = dat2, type = "raw"),
[…]

### 2010-11-17 www/fa_score_extraction.R

@@ -1,0 +1,52 @@
+"score.fa" <- function(res.list)
+  {
+    # Code to extract scores from a 2-factor model that
+    # is fit by OpenMx.  res is a list containing
+    # the model fit by mxRun() and the observed data matrix.
+
+    require(numDeriv)
+
+    res <- res.list$res
+    dat <- res.list$dat
+    
+    # Define casewise likelihood function
+    likefun <- function(x,ind.dat){
+      mu <- x[14:19]
+      lam <- matrix(c(x[1:3],rep(0,6),x[4:6]),6,2) # loadings
+      phi <- matrix(c(1,x[13],x[13],1),2,2) # factor covs
+      psi <- diag(x[7:12]) # error vars
+
+      impl.cov <- lam %*% phi %*% t(lam) + psi
+
+      fval <- log(det(impl.cov)) + t(ind.dat - mu) %*% solve(impl.cov) %*%
+                                   (ind.dat - mu)
+      fval
+    }
+
+    # Results from OpenMx
+    ests <- res@output$estimate
+    # Need n from OpenMx output
+    n <- nrow(dat)
+
+    scores <- matrix(NA,n,length(ests))
+    for (j in 1:n){
+      scores[j,] <- grad(likefun, ests,
+                         ind.dat=as.matrix(dat)[j,],
+                         method="Richardson")
+    }
+
+    scores
+  }
+
+
+"inf.fa" <- function(res.list, order.by, data)
+  {
+    # Code to extract information matrix of model parameters,
+    # for strucchange.
+    res <- res.list$res
+    
+    n <- nrow(res.list$dat)
+    inf.mat <- (2/n)*res@output$calculatedHessian
+
+    inf.mat
+  }

### 2010-11-17 www/script.R

@@ -1,0 +1,40 @@
+# Main R script file to carry out tests of measurement invariance
+# described in Merkle & Zeileis.
+
+# Load requisite packages
+library(OpenMx)
+library(strucchange)
+
+# Read data
+obs.data <- dget("data.dat")
+dat <- obs.data$datmat
+V <- obs.data$age
+
+# Obtain functions for score and information matrix extraction
+source("fa_score_extraction.R")
+
+# Obtain OpenMx model specification
+# (Two-group model with all parameters constrained to be equal,
+#  though this could also be estimated as a single-group model.)
+source("def.model.R")
+fa.mod <- def.model(dat, V)
+
+# Fit model via OpenMx
+mx.res <- mxRun(fa.mod)
+
+# Organize results for strucchange:
+fa.res <- list(res=mx.res, dat=dat, V=V)
+
+# Calculate cumulative scores for first three factor loadings:
+fa.cums <- gefp(fa.res, fit=NULL, scores=score.fa, vcov=inf.fa, 
+                order.by=fa.res$V, sandwich=FALSE, parm=1:3)
+
+# Plot individual cumulative score processes:
+plot(fa.cums, agg=FALSE, ylim=c(-1.8,0.3),
+     ylab="Cumulative scores", xlab="Age")
+
+# Obtain results of double-max test:
+sctest(fa.cums, functional="max")
+
+# Obtain results of Cramer-von Mises test:
+sctest(fa.cums, functional="meanL2")

### 2010-11-19 MIpaper/par_instable.tex
Modified example in manuscript, modified bib file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@15 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
     \label{eq:score}
-  \sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}})] = 0,    
+  \sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) = 0,    
 \end{equation}
@@
 
+These attributes give the proposed tests important advantages over
+existing tests, as existing measurement invariance methods cannot
+accomplish either of the tasks outlined above.
+In particular, Millsap \citeyear{Mil05} cites
+``locating the invariance violation'' as a major outstanding problem
+in the field.
+
 \begin{figure}
@@
 isolation of model parameters violating
-measurement invariance.
+measurement invariance.\footnote{R code for obtaining these results is
+  available at \url{http://semtools.r-forge.r-project.org}.}
 
@@
 Results related to the cumulative score process are displayed in
-Figure~\ref{fig:cusumex}.  In this figure, the processes
-are displayed individually 
-for each factor loading, with student number (ordered by age and
-rescaled to (0,1)) on the x-axis and
-absolute values of the cumulative scores on the y-axis.  The dashed horizontal
+Figure~\ref{fig:cusumex}.  In this figure, each panel represents the
+process for a single factor loading, with student age on the x-axis and
+cumulative scores on the y-axis.  The dashed horizontal
 lines represent the critical value for rejecting $H_0$ at $\alpha=.05$
@@
 ($\lambda_{21}$ and $\lambda_{31}$) cross the dashed line, indeed
-implying rejection of $H_0$.  The maximum across all three processes,
-corresponding to the double-max test statistic, is 1.72.  This maximum
-occurs in the third graph and is represented as a solid circle, 
-achieving a $p$-value of .016.
+implying rejection of $H_0$.  The absolute maximum across all three processes,
+corresponding to the double-max test statistic, is 1.77.  This
+statistic is obtained from the minimum of the third
+panel (whose absolute value is the maximum), close to an
+age of 16.  It achieves a $p$-value of .012.
 
@@
 The student corresponding to the maximum
-absolute cumulative sum (the solid circle in the third graph) is 15.8
+absolute cumulative sum (the solid circle in the third graph) is 15.7
 years of age, implying one subgroup of 
-students below 15.8 years and one above 15.8 years.  This agrees well
+students below 15.7 years and one above 15.7 years.  This agrees well
 with the true $\eta$ of 16.0 years.
@@
 
+% NEED TO MODIFY
 \begin{figure}
@@
   based on the example involving measurement invariance with respect
-  to student age.  The dashed lines represent critical values for the
-  test of measurement invariance at $\alpha=.05$.}
+  to student age.  Each panel represents the cumulative score process
+  for a single factor loading; the red line represents the critical values for 
+  the absolute maximum at $\alpha=.05$.}
 \label{fig:cusumex}

### 2011-01-12 MIpaper/par_instable.tex
merged theory revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@18 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \rightheader{Parameter instability tests}
+
+\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+
+%% for internal use
+\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
+\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}
 
@@
 
-The tests proposed here are generally relevant to situations where the
-$p$-dimensional 
-data vector $\bm{x}_i, i=1,\cdots,n$ is specified to arise from a model with
-density $f(\bm{x}_i | \bm{\theta}_i)$ and log-likelihood 
-$\log L(\bm{\theta}_i | \bm{x}_i)$.  
-For the factor-analytic situations
-considered in this paper, the $\bm{x}_i$ consist of individuals' scale
-scores, and 
-$f()$ and $\log L()$ arise from the multivariate normal distribution.
-The density is defined as:
+The methods proposed here are generally relevant to situations where the
+$p$-dimensional random variable $X$ with associated observations $\bm{x}_i, i=1,\cdots,n$
+is specified to arise from a model with density $f(\bm{x}_i; \bm{\theta})$ and
+associated joint log-likelihood
+\begin{equation} \label{eq:loglik}
+  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
+    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
+    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
+\end{equation}
+where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
+the distribution. 
+The methods are applicable under very general conditions, essentially
+whenever standard assumptions for maximum likelihood inference hold (for more
+details see below). For empirical applications, we employ
+a specific factor analysis model where the data follow a multivariate normal
+distribution:
+\begin{eqnarray}
+    \label{eq:mvndensity}
+    f(\bm{x}_i; \bm{\theta}) & = & \frac{1}{(2\pi)^{p/2} |
+      \bm{\Sigma}(\bm{\theta}) |^{1/2}} ~ \exp \left\{ -\frac{1}{2}(\bm{x}_i -
+    \bm{\mu}(\bm{\theta}))^{\top} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
+    \bm{\mu}(\bm{\theta}) \right\}, \\
+    \label{eq:caselik}
+    \ell(\bm{\theta}; \bm{x}_i) & = & -\frac{1}{2} \left\{
+      (\bm{x}_i - \bm{\mu}(\bm{\theta}))^{\top} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i - \bm{\mu}(\bm{\theta}))
+      ~+~ \log | \bm{\Sigma}(\bm{\theta}) | ~+~ p \log(2 \pi) \right\}.
+\end{eqnarray}
+with model-implied mean vector ${\bm{\mu}}({\bm{\theta}})$ and
+covariance matrix ${\bm{\Sigma}}({\bm{\theta}})$. As pointed out above,
+the assumptions for the tests introduced below do not require this specific
+form of the likelihood but it is presented for illustration due to
+its importance in practice.  Many expositions of factor analysis
+utilize the likelihood for the sample covariance matrix, which is
+based on a Wishart distribution when $\bm{x}_i$ are assumed to be
+multivariate normal.
+However, the techniques presented below require the casewise contributions
+to the likelihood; this situation is also generally
+encountered in structural equation models with missing data (e.g.,
+\citeNP{Wot00}).
+
+Within the general framework outlined above, and under the usual
+regularity conditions, the model parameters $\bm{\theta}$ can
+be estimated by maximum likelihood (ML), i.e.,
+\begin{equation} \label{eq:ml}
+  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \ell(\bm{\theta}; x_1, \dots, x_n),
+\end{equation}
+or equivalently by solving the first order conditions
 \begin{equation}
-    \label{eq:mvndensity}
-    f(\bm{x}_i | \theta_i) = \frac{1}{(2\pi)^{p/2} |
-      \bm{\Sigma}(\bm{\theta}_i) |^{1/2}}\exp(-\frac{1}{2}(\bm{x}_i -
-    \bm{\mu}(\bm{\theta}_i))^{\prime} \bm{\Sigma}(\bm{\theta}_i)^{-1} (\bm{x}_i -
-    \bm{\mu}(\bm{\theta}_i)),
-\end{equation}
-where ${\bm{\mu}}({\bm{\theta}}_i)$ is the
-model-implied mean vector, and ${\bm{\Sigma}}({\bm{\theta}}_i)$ is the
-model-implied covariance matrix.
-The casewise log-likelihood is defined as:
-\begin{equation}
-    \label{eq:caselik}
-    \log \text{L}({\bm{x}}_i, {\bm{\theta}}_i) = %\sum_{i=1}^{N} 
-\log \mid 
-\bm{\Sigma}(\bm{\theta}) \mid + %\sum_{i=1}^{N} 
-(\bm{x}_i -
-\bm{\mu}(\bm{\theta}))^{\prime} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
-\bm{\mu}(\bm{\theta})).
-\end{equation}
-This equation is often summed over $n$ observations, and it is also
-often employed for factor analysis and structural
-equation models with missing data (e.g., \citeNP{Wot00}).  We use
-the likelihood here to facilitate the proposed tests.
-
-Within the general framework outlined above, model estimation
-often proceeds by assuming that $\bm{\theta}_i = \bm{\theta}\
-\forall\ i$, then choosing $\hat{\bm{\theta}}$ to satisfy the
-criterion:
+    \label{eq:ml1}
+  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
+\end{equation}
+where 
 \begin{equation}
   \label{eq:score}
-  \sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) = 0,    
-\end{equation}
- where $\psi(\bm{x}_i, \bm{\theta})$ 
-is a score function based on the model.  
-For example, in factor analysis models estimated via Maximum
-Likelihood, the score function is the first derivative of $\log
-\text{L}(\bm{x}_i, \bm{\theta})$ with respect to $\bm{\theta}$.
-General M-estimation methods may also be handled within this
-framework, though we do not discuss them here.
+  {\bm s}({\bm \theta}; x_i) ~=~ \left(
+    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_1},
+    \dots,
+    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_k}
+  \right)^\top,
+\end{equation}
+is the score function of the model, i.e., the partial derivative of the casewise
+likelihood contributions w.r.t.\ the parameters $\bm{\theta}$.
+
+% \fixme{I have deferred individual-specific parameters for ease of exposition.
+% My feeling is that it should be briefly mentioned at the end of this
+% section and then connected with measurement invariance in the next section.}
+
+One central assumption -- sometimes made implicitly -- is that
+the same model parameters $\bm{\theta}$ hold for all individuals $i = 1, \dots, n$.
+If this is not satisfied, the estimates $\hat{\bm{\theta}}$ are typically
+not meaningful and cannot easily be interpreted.
+One potential source of deviation from this assumption is lack of
+measurement invariance, investigated in the following section.
 
@@
 
-Measurement invariance specifically implies that factor analysis model
-parameters are invariant across individuals.  It can be framed as the
-hypothesis:
+In general terms, a set of scales is defined to be measurement
+invariant with respect to an auxiliary variable $V$ if:
+\begin{equation}
+    \label{eq:midef}
+      f({\bf X} | T, V, \dots) = f({\bf X} | T, \dots),
+\end{equation}
+where ${\bf X}$ is a data matrix, $T$ is the latent
+variable that the scales purport to measure, and $f$ is the model's
+distributional form.  In the parametric framework
+adopted here, this means that the parameter vector ${\bm \theta}$ (or
+some subset of ${\bm \theta}$; see \citeNP{Mer93}) is equal
+across subgroups of individuals and thus does not vary
+with any variable $V$.
+
+To frame this as a formal hypothesis, we assume that -- in principle --
+Model~(\ref{eq:loglik}) holds for all individuals but with a
+potentially individual-specific 
+parameter vector ${\bm \theta}_i$. The null hypothesis of measurement invariance
+is then equivalent to the null hypothesis of parameter constancy
 \begin{equation}
     \label{eq:h0}
-    H_0: {\bm \theta}_i = {\bm \theta}_0,\ i=1,\ldots,n,
-\end{equation}
-signifying that the parameter vector is the same for all individuals
-$i$.  
-This general hypothesis is very difficult to test because $n$ is large, and
-there are many ways by which the $\bm{\theta}_i$ may stray from one
-another.  To circumvent this problem, researchers often measure
-individuals on an auxiliary variable $V$ and divide them into groups
-based on that variable.  
-Assuming a continuous $V$, we might divide 
-individuals into two groups based on some threshold $\eta$.
-Equation~\eqref{eq:h0} then becomes:
+    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
+\end{equation}
+which should be tested against the alternative that the parameters
+are some nonconstant function ${\bm \theta}(\cdot)$ of the variable $V$ with observations
+$v_1, \dots, v_n$, i.e.,
 \begin{equation}
-    \label{eq:h0.cut}
-    H_0: \bm{\theta}_{V < \eta} = \bm{\theta}_{V > \eta},
-\end{equation}
-where, i.e., $\bm{\theta}_{V < \eta}$ reflects the factor analysis
-parameters for all individuals with $V < \eta$.
-
-Once we have grouped individuals based on $V$,
-factor-analytic assessments
-of measurement invariance are well developed.
-These assessments are often carried out using nested Multiple Groups
-Models (e.g., \citeNP{Jor71,Bol89}) coupled with Likelihood Ratio tests,
-though 
-Lagrange Multiplier and Wald tests may also be constructed for this
-purpose.    While these three tests
-are asymptotically equivalent \cite{Sat89}, each is detailed
-individually below.
+    \label{eq:h1}
+    H_1:~ {\bm \theta}_i = {\bm \theta}(v_i),\quad (i=1,\ldots,n).
+\end{equation}
+where the pattern ${\bm \theta}(V)$ of deviation from measurement invariance is
+typically not known (exactly) in practice. If it were (see below for
+some concrete examples), then standard inference methods (such as likelihood
+ratio, Wald, or Lagrange multiplier tests) could be employed. However,
+if the pattern is unknown, it is difficult to develop a single test
+that is well-suited for all conceivable patterns. But it is possible to
+derive a family of tests so that representatives from this family are well-suited
+for a wide range of possible patterns. One pattern of particular interest
+involves $V$ dividing the individuals into two subgroups with different
+parameter vectors
+\begin{equation}
+  \label{eq:h1*}
+  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
+    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
+    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
+  \end{array} \right.
+\end{equation}
+where ${\bm \theta}^{(A)} \neq {\bm \theta}^{(B)}$. This could pertain to 
+two different age groups, income groups, genders, etc.
+
+Note that even when adopting $H_1^*$ as the relevant alternative, the pattern ${\bm \theta}(V)$
+is not completely specified unless the cutpoint $\nu$ is known in
+advance. 
+In this situation, all individuals can be grouped based on
+$V$, and we can apply standard
+theory: nested multiple group
+models (e.g., \citeNP{Jor71,Bol89}) coupled with likelihood ratio (LR)
+tests are most common,
+though Lagrange multiplier (LM) and Wald tests may also be constructed for this
+purpose.\footnote{All three of these approaches are asymptotically
+  equivalent \cite{Sat89}.}  However, if $\nu$ is unknown (as is often
+the case for continuous $V$), then the standard theory does not apply
+and we can instead rely on the tests proposed in this paper.
+
+In the following section, we describe the standard approaches to
+testing measurement invariance with $\nu$ known.  We then contrast
+these approaches with the tests proposed in this paper.
+We assume throughout that the observations $i = 1, \dots, n$ are
+ordered with respect to the random variable $V$ of interest such that
+$v_1 \le v_2 \le \dots \le v_n$.
 
@@
 
-\subsection{Likelihood Ratio Test}
-
-Measurement invariance is achieved
-if sets of factor analysis model parameters are equal across
-subgroups of individuals 
-(\citeNP{Mer93}; there are many types of measurement invariance based
-on different sets of model parameters being held constant, but any of these
-can be examined via the proposed tests).  If measurement invariance is
-not achieved, 
-then the parameters assume different values for different
-subgroups.  This leads to statistical tests of measurement
-invariance via the comparison of two models: a restricted model where
-parameters are equal across subgroups, and a ``full'' model
-where parameters are allowed to assume different values for each
-subgroup.  Statistically, this comparison is often 
-accomplished by a likelihood ratio test (also called a
-difference test).  Let the full model parameters be denoted 
-${\bm \theta} = ({\bm \theta}_1^{\prime}\ {\bm
-  \theta}_2^{\prime})^{\prime}$, where we can obtain the restricted model
-by placing a set of equality constraints on ${\bm \theta}_1$ described
-by the vector-valued function ${\bm
-  h}({\bm \theta}_1) = {\bm 0}$.  A $\chi^2$ statistic testing
-measurement invariance may then be calculated via:
-\begin{equation}
-    \label{eq:chisq}
- \chi^2 = -2(n-1)(\log L({\bm X},
-\widehat{{\bm \theta}}_c) - \log L({\bm X},\widehat{{\bm \theta}})),
-\end{equation}
-where ${\bm{X}}$ is an $n \times p$ data matrix, $\widehat{\bm \theta}_c$
-contains Maximum Likelihood parameter estimates satisfying the
-constraints ${\bm h}({\bm \theta}_1) = {\bm 0}$, and $\widehat{\bm
-  \theta}$ contains Maximum Likelihood parameter estimates from the
-full model.  The log-likelihood function in the above equation arises
-from the multivariate normal distribution and is defined as:
-  The test resulting from
-Equation~\eqref{eq:chisq} has 
-degrees of freedom equal to the difference in number of free
-parameters for the two models, and it rejects the hypothesis of measurement
-invariance for large values of $\chi^2$.
-
-While the Likelihood Ratio Test is most popular for testing
-measurement invariance, Lagrange Multiplier tests and Wald tests
-may also be constructed for the same purpose.The Lagrange Multiplier and Wald 
-tests, described below, are also advantageous in that they require
-estimation of only one 
-model (either the full or restricted model), whereas the Likelihood
-Ratio Test requires estimation of both models.
-
-\subsection{Lagrange Multiplier Test}
-The Lagrange Multiplier test (e.g., \citeNP{Sil59,ChouBent90})
-focuses on a restricted model and tests whether the equality
-constraints described by ${\bm h}({\bm \theta}_1) = {\bm 0}$ should be
-freed.
-Estimating the restricted model, we define a vector of Lagrange
-multipliers $\widehat{\bm \kappa}$ that satisfy:
-\begin{equation}
-    \label{eq:lagrange}
-    \frac{\partial \log L}{\partial {\bm \theta}} \bigg |_{\widehat{\theta}_c} +
-    \frac{\partial {\bm h}}{\partial {\bm \theta}}
-    \bigg |_{\widehat{\theta}_c}^{\prime} \widehat{{\bm \kappa}} = {\bm 0}.
-\end{equation}
-The test itself is then based on the asymptotic distribution of 
-$\widehat{{\bm \kappa}}$: if $\widehat{{\bm \kappa}}$ is sufficiently close
-to ${\bm 0}$, then the constraints are deemed to hold in the
-population.  The resulting test statistic, which is a function of
-$\widehat{\bm \kappa}$, follows a $\chi^2$ distribution for large
-samples.
-
-\subsection{Wald Test}
-In contrast to the Lagrange Multiplier test, the
-Wald test (e.g., \citeNP{Wald43})
-focuses on the full model and tests the plausibility of the
-constraints ${\bm h}({\bm \theta}_1) = {\bm 0}$.
-The test is based on the estimates ${\bm h}(\widehat{{\bm \theta}}_1)$
-from the full model; if
-these are sufficiently close to zero, then the constraints are deemed
-to hold in the population.  The resulting test statistic, which is a
-function of ${\bm h}(\widehat{{\bm \theta}}_1)$, is also
-$\chi^2$-distributed in large 
-samples.
-Chou and Bentler \citeyear{ChouBent90} compared the 
-Wald and Lagrange tests with finite sample sizes via
-simulation, and they generally found similar
-performance.  They noted, 
-however, that the Lagrange test has a tendency toward Type I errors in
-some contexts.  They therefore urged caution in the Lagrange test's
-use, as it may lead one to discard more constraints than is necessary,
-leading to overly-complex models.
-
-\subsection{Extensions to Proposed Tests}
-For continuous $V$, the three tests described above require the
-threshold $\eta$ to be known.
-  This is a
-major limitation in psychometrics, where $\eta$ is often unknown or
-undefined.
+\subsection{Likelihood Ratio, Wald, and Lagrange Multiplier Test for Fixed Subgroups}
+
+To employ the LR test for assessing measurement invariance, the data
+are typically separated into a certain number of given subgroups where
+the parameters are estimated separately (for ease of
+  exposition, we assume no parameters are restricted to be equal
+  across subgroups). Then, the sum of maximized likelihoods 
+from the subgroups are compared with the original maximized full-sample likelihood
+in a $\chi^2$~test. More specifically, for the special case of two subgroups,
+the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
+and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
+The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
+observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
+$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
+observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
+for the given threshold $\nu$ is then
+\begin{equation} \label{eq:lr}
+  \mathit{LR}(\nu) ~=~ -2 \left[
+         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
+   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
+    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
+    \right],
+\end{equation}
+which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
+of parameters in~${\bm \theta}$.
+\readme{You previously modified the above section so that $\theta$
+  consists only of parameters that are ``doubled'' under the
+  alternative.  This appears to not allow for situations where
+  parameters are restricted to be equal across subgroups.  I am ok
+  with this, but most applications of LR tests with these models have
+  some parameters restricted to be equal across subgroups.  So I've
+  added some detail in parentheses at the beginning of the above
+  paragraph.}
+
+Analogously to the LR test, the Wald test and LM test (also known as score test)
+can be employed to test the null hypothesis $H_1^*$ for a fixed threshold $\nu$.
+For the Wald test, the idea is to compute the Wald statistic $W(\nu)$ as a quadratic
+form in $\hat {\bm \theta}^{(A)} - \hat {\bm \theta}^{(B)}$, utilizing its estimated
+covariance matrix for standardization. For the LM test, the LM statistic
+$\mathit{LM}(\nu)$ is a quadratic form in ${\bm s}(\hat {\bm \theta}; x_1, \dots, x_m)$
+and ${\bm s}(\hat {\bm \theta}; x_{m+1}, \dots, x_n)$. Thus, the three tests all assess
+differences that should be zero under $H_0$: for the LR test the difference of maximized
+likelihoods; for the Wald test, the difference of parameter estimates; and for the
+LM test, the differences of likelihood scores from zero. In the LR case, the parameters
+have to be estimated both under null hypothesis and alternative while in the Wald case
+only the estimates under the alternative are required and in the LM case only under the
+null hypothesis.
+
+\subsection{Extensions for Unknown Subgroups}
+
+For assessing measurement invariance in psychometric models, the major limitation of
+the three tests is that the potential subgroups have to be known in advance. 
+Even if the variable $V$ w.r.t.\ which the violation of invariance occurs is
+known, the threshold $\nu$ from (\ref{eq:h1*}) is often unknown in practice.
 For example, if $V$ represents yearly income, there are many possible
-values of 
-$\eta$ that could be used to divide individuals into poorer and richer
-groups.  The ultimate $\eta$ that we
+values of $\nu$ that could be used to divide individuals into poorer and richer
+groups.  The ultimate $\nu$ that we
 choose could potentially impact our
@@
 in the same way that dichotomization of continuous variables impacts
-general statistical analyses \cite{MacZha02}.
-
-Instead of choosing a specific $\eta$, we may elect to test
-individuals across all possible values of $\eta$.  For example, we
-could conduct a Likelihood Ratio Test for every possible
-dichotomization of $V$, yielding a test statistic for each
-dichotomization.  The distribution of the maximum of these test
-statistics is derived in Andrews \citeyear{And93}, providing a test of
-measurement invariance over all values of $\eta$.  
-This is a special case of the family of tests proposed here.  The
-tests are designed for situations where
-$\eta$ is unknown or where a dichotomization of $V$ is insufficient to
-detect measurement invariance violations, and they can be applied both
-to individual model parameters and sets of model parameters.
-
-We now describe some general ideas for capturing measurement
-invariance violations with respect to a continuous auxiliary variable
-$V$.
+general psychometric analyses \cite{MacZha02}.
+
+Instead of choosing a specific $\nu$, a natural idea is to compute $\mathit{LR}(\nu)$
+for each possible value in some interval $[\underline{\nu}, \overline{\nu}]$ and
+reject if their maximum
+\begin{equation} \label{eq:maxlr}
+  \max_{\nu \in [\underline{\nu}, \overline{\nu}]} \mathit{LR}(\nu)
+\end{equation}
+becomes large. Note that this corresponds to maximizing the likelihood w.r.t.\
+an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
+maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. However, Andrews \citeyear{And93}
+showed that the asymptotic distribution is in fact tractable, but nonstandard.
+Specifically, the asymptotic distribution of the maximum is the maximum of a
+certain tied-down Bessel process whose specifics also depend on the minimal
+and maximal thresholds $\underline{\nu}$ and $\overline{\nu}$, respectively.
+See below for more details.
+
+Analogously, one can consider $\max W(\nu)$ and
+$\max \mathit{LM}(\nu)$, respectively, which both have the same asymptotic properties
+as $\max \mathit{LR}(\nu)$ and are asymptotically equivalent \cite{And93}.
+From a computational perspective, the latter test is particularly convenient
+because it requires just a single set of estimated parameters $\hat {\bm \theta}$
+which is employed for all thresholds $\nu$ in $[\underline{\nu}, \overline{\nu}]$
+while the other two techniques require reestimation of the subgroup model for each
+$\nu$.
+
+So far, the discussion focused on the alternative $H_1^*$: The maximum LR, Wald, and LM
+tests are designed for a situation where there is a single threshold at which the full
+parameter vector ${\bm \theta}$ changes. While this is plausible and intuitive
+in many applications, it would also be desirable to obtain tests that direct their power
+against other types of alternatives, i.e., against $H_1$ with other patterns ${\bm \theta}(V)$.
+For example, there might be multiple thresholds at which the parameters change or parameters
+fluctuate randomly. Alternatively, only one (or just a few) of the parameters in
+the vector ${\bm \theta}$ change while the remaining parameters are
+constant (a common occurrence in psychometric models).
+To address such situations in a unified way, the next section contains 
+a general framework for testing measurement invariance along a
+(continuous) variable $V$ that
+includes the maximum LM test as a special case.
 
 \section{Stochastic Processes for Measurement Invariance}
-The proposed tests of measurement invariance implicitly rely on an
-estimated factor analysis model (one that restricts parameters to be
-invariant across subgroups).  For the purposes of this paper, we
-assume that the model is estimated via Maximum Likelihood.  
-The factor analysis log-likelihood may be obtained by summing 
-Equation~\eqref{eq:caselik} across individuals.  
-Maximization of the likelihood then involves iterative 
-algorithms that seek values $\widehat{{\bm{\theta}}}$ to satisfy
-Equation~\eqref{eq:score}, where 
-$\psi(\bm{x}_i, \theta_j) = \frac{\partial}{\partial \bm{\theta}} \log
-\text{L}({\bm{x}}_i, 
-{\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}$.
-% insert derivatives of casewise evaluated at theta-hat = 0
-Such derivatives are taken across individuals, but, using individual
-terms of the log-likelihood (i.e., Equation~\eqref{eq:caselik}),
-we can also 
-evaluate the derivatives with respect to each individual.
-If a unique value of $\widehat{{\bm{\theta}}}$ maximized
-each individual likelihood, then we would have:
-% Equation: \psi(y_i, \theta) = d/d theta casewise_i = 0 \forall i.
+
+As discussed above, factor analysis models are typically
+estimated by fitting the model to all $i = 1, \dots, n$ individuals, assuming
+that the parameter vector ${\bm \theta}$ is constant across individuals.
+Having estimated the parameters $\hat {\bm \theta}$, the goal is to check that
+there are no subgroups of individuals that fail to conform with the model (for any
+of the parameters). Hence, some measure of model deviation or residual is required
+that captures the lack of fit for the $i$-th individual at the $j$-th parameter
+($i = 1, \dots, n$, $j = 1, \dots, k$). A natural measure -- that employs the ideas
+of the LM test -- is ${\bm s}(\hat {\bm \theta}; x_i)_j$: the $j$-th component of the
+contribution of the $i$-th observation to the score function. By construction the sum
+of the score contributions over all individuals is zero for each component, see
+(\ref{eq:ml1}). Moreover, if there are no systematic deviations the score contributions
+should fluctuate randomly around zero while they should be shifted away from zero for
+subgroups where the model does not fit.
+
+Therefore, to employ this quantity for tests of measurement invariance against alternatives
+of type (\ref{eq:h1}), we need to overcome two obstacles: (1)~make use of the ordering of
+the observations w.r.t.\ $V$ because we want to test for changes ``along'' $V$; (2)~account
+for potential correlations between the $k$~components of the parameters to be able to detect
+which parameter(s) change (if any).
+
+\subsection{Theory}
+
+The test problem of the null hypothesis (\ref{eq:h0}) against the alternatives
+(\ref{eq:h1}) and (\ref{eq:h1*}), respectively, has been studied extensively
+in the statistics and econometrics literature under the label ``structural change
+tests'' (see e.g., \citeNP{BroDur75,And93}) where the focus of interest is the
+detection of parameter instabilities of time series models ``along'' time.
+Specifically, it has been shown (e.g., \citeNP{Nyb89,Han92,HjoKon02,ZeiHor07})
+that cumulative sums of the empirical scores follow specific stochastic processes, allowing
+us to use them to generally test measurement invariance. Here, we review
+some of the main results from that literature and adapt it to the specific challenges
+of factor analysis models. More detailed accounts of the underlying structural
+change methods include Hjort and Koning \citeyear{HjoKon02} and
+Zeileis and Hornik \citeyear{ZeiHor07}.
+
+For application to measurement invariance, the most important
+theoretical result involves the fact that, under $H_0$, the
+\emph{cumulative score process} converges to a specific asymptotic
+process. The $k$-dimensional cumulative score process is defined as
+\begin{equation} \label{eq:cumscore}
+  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
+    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_i)
+  \qquad (0 \le t \le 1)
+\end{equation}
+where $\lfloor nt \rfloor$ is the integer part of $nt$ and
+$\hat I$ is some consistent estimate of the covariance matrix of the scores,
+e.g., their outer product or the observed information matrix.
+As the equation shows, the cumulative score process adds subsets of
+casewise score contributions across individuals along the ordering
+w.r.t.\ the variable $V$ of interest. At $t=1/n$, only the
+first individual's contribution enters into the summation; at
+$t=2/n$, the first two individuals' contributions etc.\ until $t = n/n$
+where all contributions enter the sum. Thus, due to (\ref{eq:ml1}),
+the cumulative score process always equals zero at $t = 0$ and
+returns to zero at $t = 1$. Furthermore, multiplication by
+$\hat {\bm I}^{-1/2}$ ``decorrelates'' the $k$ cumulative score
+processes, such that each univariate process (i.e., each
+process for a single model parameter) is unrelated to (and asymptotically independent of) all other
+processes. Therefore, this cumulative process ${\bm B}(t; \hat {\bm \theta})$
+accomplishes the challenges discussed at the beginning of this section:
+it makes use of the ordering of the observations by taking cumulative sums
+and it decorrelates the contributions of the $k$ different parameters.
+
+Inference can then be based on an extension of the usual central limit
+theorem. Under the assumption of independence of individuals (implicit
+already in Equation~\ref{eq:loglik}) and under
+the usual ML regularity conditions (assuring asymptotic normality
+of $\hat {\bm \theta}$), Hjort \& Koning \citeyear{HjoKon02} show
+that 
+\begin{equation} \label{eq:fclt}
+  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
+\end{equation}
+where $\overset{d}{\rightarrow}$ denotes convergence in distribution
+and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
+In words, there are $k$ cumulative score processes, one for each
+model parameter.  This collection of processes follows a
+multidimensional Brownian bridge as score contributions accumulate in the 
+summation from individual 1 (with lowest value of $V$) to individual $n$
+(with highest value of $V$).
+
+The empirical cumulative score process from (\ref{eq:cumscore}) can also be viewed
+as an $n \times k$ matrix with elements ${\bm B}(i/n; \hat {\bm \theta})_j$ that
+we also denote ${\bm B}(\hat {\bm \theta})_{ij}$ below for brevity.
+Each column of this matrix converges to a univariate Brownian bridge and pertains
+to a single factor analysis parameter. To carry out a test of $H_0$, this process/matrix
+needs to be aggregated to a scalar test statistic by collapsing across
+rows (individuals) and columns (parameters) of the matrix. The asymptotic distribution
+of this test statistic is then easily obtained by applying the same
+aggregation to the asymptotic process ${\bm B}^0$ \cite{HjoKon02,ZeiHor07},
+so that corresponding $p$-values can be derived.
+
+As argued above, no single aggregation function will have high power for any
+conceivable pattern of measurement invariance ${\bm \theta}(V)$ while any
+(reasonable) aggregation function will have non-trivial power under $H_1$.
+Thus, various aggregation strategies should be employed depending on which
+pattern ${\bm \theta}(V)$ is most plausible (because the exact is typically
+unknown). A particularly agnostic aggregation strategy is to reject $H_0$
+if any component of the the cumulative score process ${\bm B}(t; \hat {\bm \theta})$
+departs ``too far'' from its zero at any time, i.e., if
 \begin{equation}
-    \label{eq:casederiv}
-    \psi({\bm{x}}_i, \bm{\theta}) = \frac{\partial}{\partial \bm{\theta}} \log
-\text{L}({\bm{x}}_i, 
-{\bm{\theta}}) \bigg |_{\bm{\theta} = \widehat{\bm{\theta}}} = \bm{0}\ \ \ \forall\ i.
-\end{equation}
-In practice, no value of $\widehat{{\bm{\theta}}}$ will maximize all
-individuals' likelihoods, but, if measurement invariance is satisfied, 
-we can expect the $\psi(\cdot, \cdot)$ to fluctuate around zero.  This
-is the idea underlying the tests described here.
-
-\subsection{Theory}
-Theoretical work in statistics (e.g.,
-\citeNP{BroDur75,HjoKon02,Nyb89}) has shown 
-that functions of the 
-$\psi(\cdot, \cdot)$ follow specific stochastic processes, allowing
-us to use them to generally test measurement
-invariance.  
-We rely on the null hypothesis specified in Equation~\eqref{eq:h0},
-which focuses on individuals instead of groups of individuals.
-We now review some of the main results
-for testing this $H_0$; more
-detailed accounts include Hjort and Koning \citeyear{HjoKon02} and
-Zeileis and Hornik \citeyear{ZeiHor07}.
-
-% New subsection starts here?
-
-For application to measurement invariance, the most important
-theoretical result involves the fact that, under $H_0$, functions of
-the $\psi(\cdot, \cdot)$ result in specific stochastic
-processes.  One such function
-is the {\em{cumulative score process}}, given as:
-\begin{equation}
-    \label{eq:cumscore}
-    W_n(t,\theta_j) = \frac{1}{\sqrt{n}} \sum_{i=1}^{\lfloor nt
-      \rfloor} \psi ({\bm{x}}_i, \theta_j),
-\end{equation}
-where $t$ is a continuous variable in $(1/n,1)$ and $\lfloor nt
-\rfloor$ is the floor operator, defined as the integer part of
-$nt$.
-As the equation shows, the cumulative score process adds subsets of
-casewise first derivatives across individuals who have been ordered
-based on an auxiliary variable.  Starting at $t=1/n$, only the
-first individual's derivative enters into the summation.  Once
-$t=2/n$, the first two individuals' derivatives enter into the
-summation.  This procedure continues until we reach $t=n$, at which
-point all individuals' first derivatives have entered into the
-summation.  There are $n$ cumulative sums obtained by sequentially
-inserting individuals' derivatives into the equation, and these are
-calculated 
-for each of the $p$ model parameters.  This results
-in $n\times p$
-empirical cumulative scores.
-
-Elements of the summation in Equation~\eqref{eq:cumscore}
-can be positive or 
-negative, so that the cumulative scores do not necessarily increase
-with $t$.  In fact, if measurement invariance (i.e., $H_0$) holds,
-then the cumulative scores should 
-follow a stochastic process that fluctuates around zero.  
-Zeileis and Hornik \citeyear{ZeiHor07} showed specifically that, under
-$H_0$:
-\begin{equation}
-    \label{eq:browbridge}
-{\bm{B}}(\cdot, \widehat{{\bm{\theta}}}) = {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})
-\overset{d}{\rightarrow} {\bm W}^{0}(\cdot), % put d over rightarrow
-\end{equation}
-where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed information
-matrix (e.g., of the factor analysis model), ${\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})$ reflects the $p$-dimensional
-cumulative sums (one for each parameter), ${\bm W}^{0}(\cdot)$ is a
-$p$-dimensional Brownian bridge, and 
-$\overset{d}{\rightarrow}$ reflects convergence in distribution.
-In words, there are $p$ cumulative score processes, one for each
-model parameter.  This collection of processes follows a
-multidimensional Brownian bridge as derivatives accumulate in the 
-summation from individual 1 to individual $n$.
-Multiplication by ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}$ 
-``decorrelates'' the $p$
-cumulative score processes, such that each univariate process (i.e., each
-process for a single model parameter) is unrelated to all other
-processes (and is asymptotically independent of the other processes).
-
-The cumulative scores resulting from Equation~\eqref{eq:browbridge}
-can be arranged in an $n \times p$ matrix called an {\em{empirical
-    cumulative score process}}, which we will denote as
-${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$.
-Each column of this matrix follows a univariate Brownian bridge and is
-based on a single factor analysis parameter. 
-To carry out a test of $H_0$, we need functions that collapse across
-rows (individuals) and/or columns (parameters) of the matrix.  
-These
-functions then lead to scalar test statistics and p-values.
-The specific choice of function depends on the goals of the analyst:
-the row-collapsing functions are useful for isolating subsets of
-parameters that violate measurement invariance, and the
-column-collapsing functions are useful for isolating subgroups of
-individuals who violate measurement invariance across all parameters.
-Some functions result in test
-statistics whose critical values can be derived analytically, while
-others require critical values to be derived via the
-simulation of a Brownian bridge.
-
-For example, if we wish to test a specific parameter for
-measurement invariance, ``row-collapsing''
-functions include the absolute maximum or mean of the corresponding 
-column in ${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$
-\cite{ZeiHor07}. % (see ZH07 p 7)
-Conversely, if we wish to generally test 
-measurement invariance across sets of model parameters,
-``column-collapsing'' functions 
-include the maximum norm or the squared Euclidean norm of the
-corresponding row in ${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$
-\cite{HjoKon02}.
-
-% A general, two-step strategy for testing measurement
-% invariance involves first collapsing over columns for each individual,
-% yielding a single statistic for each individual. 
-% These $n$ statistics (ordered according to an auxiliary variable)
-% can then be graphed to visually define groups that violate measurement
-% invariance.  Functions of these $n$ statistics can also yield scalar
-% test statistics.
-
-As a concrete example of these ideas, consider the situation
-where we carry out a Likelihood Ratio test for all possible values of
-$\eta$, then take the maximum of these test statistics.  A related
-idea here involves taking the maximum of the empirical cumulative
-score process:
-\begin{equation}
-    \label{eq:doubmax}
-    DM = \max_{i=1,\ldots,n} \max_{j=1,\ldots,p} |
-    B^{\ast}_{ij}(\widehat{\bm{\theta}}) |,
-\end{equation}
-where $B^{\ast}_{ij}(\widehat{\bm{\theta}})$ represents the
-entry in row $i$ and column $j$ of
-${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$.
-This is a 
-{\em{double-max}} statistic that allows for simultaneous isolation of 
-$\eta$ along with model parameters that change over $\eta$.
-This statistic is
-especially useful for visualization, as the fluctuation process for
+    \label{eq:dmax}
+    \mathit{DM} = \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,
+\end{equation}
+becomes large. Consquently, this double maximum statistic allows for simultaneous
+isolation of the threshold(s) of parameter change (over the individuals $i = 1, \dots, n$)
+and the parameter(s) affected by it (over $j = 1, \dots, k$). This test is
+especially useful for visualization, as the cumulative score process for
 each individual parameter can be displayed along with the appropriate
@@
 
-As a second example of ``collapsing'' functions,
-we could first collapse over columns for each individual $i$ via the
-Euclidean norm:
-\begin{equation}
-    \label{eq:meanl2}
-    \nu_i = \lVert {\bm B}^{\ast}_{i\cdot}(\widehat{\bm \theta}) \rVert,
-\end{equation}
-where the $i\cdot$ subscript signifies row $i$ of the empirical
-fluctuation process matrix.  A univariate test statistic may then be
-obtained by taking the mean of the $\nu_i$.  This is a
-Cram\'{e}r-von Mises statistic described by, e.g., Nyblom
-\citeyear{Nyb89} and Hansen \citeyear{Han92} in the contexts of
-longitudinal models and linear regression.  In general contexts,
-this test is recommended
-when there are more than two groups exhibiting parameter instability 
-and when multiple parameters change across groups \cite{ZeiSha10}.  
-We demonstrate these recommendations' implications for
-measurement invariance later in the paper.
-
-% We still obtain a unidimensional process; the norm is applied
-% to each row vector.
-There exist other functions of first derivatives and of the
-empirical fluctuation process that
-result in other test statistics; these are described by, e.g.,
-Zeileis, Shah, and Patnaik \citeyear{ZeiSha10}.  
-In the analyses described later, we focus on the abilities
-of the Cram\'{e}r-von Mises test and of the double-max test.  However,
-the many 
-row- and column-collapsing functions offer a large family of tests
-that can be applied to the study of measurement invariance.
+However, taking maximums ``wastes'' power if many of the $k$~parameters change
+at the same threshold or if the score process takes large values for
+many of the $n$~individuals (and just a single threshold). In such
+cases, sums insteads of maximums are more suitable for collapsing across
+parameters and/or individuals because they combine the deviations instead of
+picking out only the single largest deviation. Thus, if the parameter instability
+${\bm \theta}(V)$ affects many parameters and leads to many subgroups, sums
+of (absolute or squared) values should be used for collapsing both across
+parameters and individuals. On the other hand, if there is just a single
+threshold that affects multiple parameters, then the natural aggregation
+is by sums over parameters and then by the maximum over individuals.
+More precisely, the former idea leads to a Cram{\'e}r-von Mises type
+statistic and the latter to the maximum LM statistic from the previous section:
+\begin{eqnarray}
+    \label{eq:cvm}
+    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
+    \label{eq:maxlm}
+    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
+      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
+\end{eqnarray}
+where the $\max \mathit{LM}$ statistic is additionally scaled by the
+asymptotic variance $t (1 - t)$ of the process ${\bm B}(t, \hat {\bm \theta})$.
+It is equivalent to the $\max_\nu \mathit{LM}(\nu)$ from the previous
+section (provided that the boundaries for the subgroups sizes $\underline{i}$/$\underline{\nu}$
+and $\overline{\imath}$/$\overline{\nu}$ are chosen analogously).
+
+Further aggregation functions have been suggested in the structural
+change literature (see e.g., \citeNP{Zei05,ZeiSha10}) but the three
+tests above are probably most likely to be useful in psychometric settings.
+
+\fixme{I have explicitly written down all three test statistics: DM because it is
+intuitive, max LM because we need it to connect to the previous section, and CvM
+because it nicely complements the picture. For the simulations, I think we should
+also consider all three tests, maybe also the max LR test. Of course, we shouldn't
+make things too complex but we can then still decide which results to choose.}
 
 \subsection{Critical Values \& $p$-values.}
-To obtain $p$-values associated with $H_0$, we can rely on analytic
-work concerning functions of Gaussian processes.
-As an example of this, we focus on use of the 
-double-max function (Equation~\eqref{eq:doubmax}) as a test statistic.
-There exist 
-analytic results for Brownian bridges that allow us to calculate
-% fix double-max notation on following line:
-$P(DM > \beta | H_0)$, where $\beta$ is a specific value
-that the test statistic may achieve.
-
-Focusing on a univariate
-Brownian bridge under $H_0$, it can be shown that (e.g., \citeNP{ShoWel86}):
- % see Beghin & Orsingher Eq (4.12).  They cite Shorack &
-            % Wellner 1986 book
-\begin{eqnarray}
-    P(DM_j < \beta | H_0) &
-    = & \sum_{h=-\infty}^{\infty} (-1)^h \exp(-2h^2\beta^2) \\
-   & = & 1 + 2 \sum_{h=1}^{\infty} (-1)^h \exp(-2h^2\beta^2).
-\end{eqnarray}
-The associated $p$-value is therefore given by:
+
+As pointed out above, specifying the asymptotic distribution und $H_0$ for the test
+statistics from the previous section is straightforward: It is simply the aggregation
+of the asympotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
+from (\ref{eq:dmax}) converges to $\sup_t || {\bm B}^0(t) ||_{\infty}$, where
+$||\cdot||_\infty$ denotes the maximum norm. Similarly, $\mathit{CvM}$
+from (\ref{eq:cvm}) converges to $\int_0^1 || {\bm B}^0(t) ||_2^2 d t$, where $||\cdot||^2$
+denotes the Euclidean norm. Finally, $\max \mathit{LM}$ from (\ref{eq:maxlm}) -- and analogously
+the maximum Wald and LR tests -- converges to $\sup_t (t (1-t))^{-1} || {\bm B}^0(t) ||_2^2$
+(which can also be interpreted as the maximum of a tied-down Bessel process, as pointed out
+previously).
+
+While it is easy to formulate these asymptotic distributions theoretically,
+it is not always easy to find closed-form solutions for computing critical
+values and $p$-values from them. In some cases -- in particular for the double maximum test --
+such a closed-form solution is available from analytic results for Gaussian
+processes (see e.g., \citeNP{ShoWel86}). For all other cases, tables of critical
+values can be obtained from direct simulation \cite{Zei06} or in combination with
+more refined techniques such as response surface regression \cite{Han97}.
+
+The analytic solution for the asymptotic $p$-value of a $\mathit{DM}$ statistic
+$d$ is
 \begin{equation}
     \label{eq:absmax_p}
-        p = P(DM_j > \beta | H_0)
-    = 1 - (1 + 2 \sum_{h=1}^{\infty} (-1)^h \exp(-2h^2\beta^2)).
-\end{equation}
-The terms in the summation quickly go to zero as $h$ goes to infinity,
-so, in practice, we only need to compute the first 100 terms (say).
-For multidimensional Brownian bridges whose dimensions have been
-``decorrelated'', $p$-values can
-be obtained similarly with a Bonferroni correction.
-
-For the Cram\'{e}r-von Mises test statistic, analytic expressions for
-$p$-values are more cumbersome.  Nyblom \citeyear{Nyb89} and Hansen
-\citeyear{Han92} provide small tables of critical values, and other
-critical values are easily approximated via simulation of Brownian
-bridges.
+        P(\mathit{DM} > d ~|~ H_0) ~\overset{asy}{=}~ 1 - \left\{1 ~+~ 2 \sum_{h = 1}^{\infty} (-1)^h \exp(-2 h^2 d^2) \right\}^k.
+\end{equation}
+This combines the crossing probability of a univariate Brownian bridge
+(see e.g., \citeNP{ShoWel86,PloKra92}) with a straightforward Bonferroni
+correction to obtain the $k$-dimensional case. As the terms in the summation
+quickly go to zero as $h$ goes to infinity, only some large finite
+number (say, 100 terms) need to be evaluated in practice.
+
+For the Cram\'{e}r-von Mises test statistic $\mathit{CvM}$, Nyblom \citeyear{Nyb89}
+and Hansen \citeyear{Han92} provide small tables of critical values which
+have been extended in the software provided by Zeileis \citeyear{Zei06}.
+
+Critical values for the distribution of the maximum LR/Wald/LM tests
+are provided by Hansen \citeyear{Han97}. Note that the distribution depends on
+the minimal and maximal thresholds employed in the test.
 
 \subsection{Locating the Invariance.}
-If a test described above implies a measurement invariance violation,
-then the researcher may want to: (1) estimate $\eta$, so that she may
-define subgroups of individuals impacted by measurement invariance;
-or (2) isolate model
-parameters affected by the invariance.  
-In the context of a Cram\'{e}r-von Mises test, we can accomplish the
-first task by graphing the $\nu_i$ (Equation~\eqref{eq:meanl2})
-and noting the location of the peaks and valleys.  We can accomplish
-the second task by taking the norm in Equation~\eqref{eq:meanl2}
-across specific columns of $B^{\ast}(\widehat{\bm \theta})$.
-
-In the context of a double-max test, we can accomplish the first task
-by defining $\eta$ to be the value
-of $V$ at which the absolute
-maximum cumulative score is achieved.  This value maximizes the
-segmented likelihood, defined as:
-\begin{equation}
-    \label{eq:seglik}
-    \sum_{i; V_i \leq \eta} L(\bm{x}_i, \bm{\theta}_{V \leq \eta}) + 
-     \sum_{i; V_i > \eta} L(\bm{x}_i, \bm{\theta}_{V > \eta}).
-\end{equation}
-where $\bm{\theta}_{V \leq \eta}$ and $\bm{\theta}_{V > \eta}$ are ML estimates
-for individuals with 
-$V_i \leq \eta$ and with $V_i > \eta$, respectively.
-  For multiple thresholds, we
-could heuristically define groups based on the number of peaks and
-valleys in the graph, or we could maximize a segmented likelihood
-similar to that in Equation~\eqref{eq:seglik}.
-We can accomplish the second task (isolating model parameters) by
-conducting posthoc 
-tests on specific model parameters.  
-%Some of these ideas are illustrated in
-%the example below.
+
+If the employed parameter instability test detects a measurement invariance violation,
+the researcher is typically interested in identification of the parameter(s) affected
+by it and/or the associated threshold(s). As argued above, the double maximum test
+is particularly appealing for this because the $k$-dimensional empirical cumulative
+score process can be graphed along with boundaries for the associated critical values.
+Boundary crossing then implies a violation of measurement invariance, and the location of
+the most extreme deviation(s) in the process convey threshold(s) in the underlying
+ordering $V$.
+
+For the maximum LR/Wald/LM tests, it is natural to graph the sequence of LR/Wald/LM
+statistics along $V$ along with a boundary corresponding to the critical value.
+Again, a boundary crossing signals a significant violation, and the peak(s) in the
+sequence of statistics conveys threshold(s). Note that due to summing over all
+parameters, no specific parameter can be identified that is responsible for the
+violation. Similarly, for the Cram\'{e}r-von Mises test neither component(s) nor threshold(s)
+can be formally identified. However, graphing of (transformations of) the cumulative
+score process may still be valuable for gaining some insights.
+
+If a measurement invariance violation is detected by any of the tests,
+one may want to incorporate it into the model to account for it. How this should
+be done typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
+discussed above often prove helpful in determining a suitable parametrization.
+In particular, one approach that is often employed in practice is to adopt a model
+with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
+single threshold case). In the multiple threshold case, their location can
+be determined by maximizing the corresponding segmented log-likelihood over all
+possible combinations of thresholds (\citeNP{ZeiSha10} adapt a dynamic programming
+algorithm to this task). For the single threshold case, this reduced to
+maximizing the segmented log-likelihood
+\begin{equation} \label{eq:seglik}
+  \ell(\hat {\bm \theta}^{({A)}}; x_1, \dots, x_m) ~+~ \ell(\hat {\bm \theta}^{({B)}}; x_{m+1}, \dots, x_n)
+\end{equation}
+over all values of $m$ corresponding to possible thresholds $\nu$ (such that $v_m \le \nu$ and
+$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR statistic
+from (\ref{eq:lr}) and some minimal subgroup size is typically imposed.
+
+Formally speaking, the maximization of (\ref{eq:seglik}) -- or equivalently (\ref{eq:lr}) --
+yields an estimate $\hat \nu$ of the threshold in $H_1^*$. If $H_1^*$ is in fact the true model,
+the peaks in the Wald/LM sequences and the cumulative score process, respectively, will occur at the
+same threshold asymptotically. However, in empirical samples, their location may differ somewhat
+(although often not much).
 
 These attributes give the proposed tests important advantages over
-existing tests, as existing measurement invariance methods cannot
-accomplish either of the tasks outlined above.
+existing tests, as existing measurement invariance methods cannot:
+(1) isolate specific parameters violating measurement invariance, or
+(2) test measurement invariance for unknown $\nu$. 
 In particular, Millsap \citeyear{Mil05} cites
@@
 to calculate the cumulative score process via
-Equations~\eqref{eq:cumscore} and~\eqref{eq:browbridge}.  Finally, we
+\eqref{eq:cumscore}.  Finally, we
 obtained a 
-double-max statistic via Equation~\eqref{eq:doubmax} and associated
+double-max statistic via Equation~\eqref{eq:dmax} and associated
 $p$-value via Equation~\eqref{eq:absmax_p}.  Note that, because we

### 2011-01-12 MIpaper/par_instable.tex
added to general discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@19 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+Relatedly,
+S\'{a}nchez \citeyear{San09} describes a general
+method for partitioning/segmenting structural equation models within a 
+partial least squares framework.  This method involves direct maximization of
+the likelihood ratio (i.e., fitting the model for various subgroups based on
+$V$ and choosing the subgroups with the largest likelihood ratio),
+however, instead of the tests described in this paper.  The tests
+described here are highly advantageous because they only require the
+model to be estimated once.
+
 % Recursive method

### 2011-01-18 www/script.R
some additions to strucchange application
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@21 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -2,8 +2,8 @@
 # described in Merkle & Zeileis.
 
 # Load requisite packages
-library(OpenMx)
-library(strucchange)
+library("OpenMx")
+library("strucchange")
 
 # Read data
 obs.data <- dget("data.dat")
@@ -23,18 +23,42 @@ fa.mod <- def.model(dat, V)
 mx.res <- mxRun(fa.mod)
 
 # Organize results for strucchange:
-fa.res <- list(res=mx.res, dat=dat, V=V)
+fa.res <- list(res = mx.res, dat = dat, V = V)
+
+## store scores in list as well to avoid slow re-computation
+fa.res$estfun <- score.fa(fa.res)
+colnames(fa.res$estfun) <- summary(mx.res)$parameters$name
 
 # Calculate cumulative scores for first three factor loadings:
-fa.cums <- gefp(fa.res, fit=NULL, scores=score.fa, vcov=inf.fa, 
-                order.by=fa.res$V, sandwich=FALSE, parm=1:3)
+fa.cums1 <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
+  vcov = inf.fa, order.by = fa.res$V, sandwich = FALSE, parm = 1:3)
+
+## the same for all parameters
+fa.cums2 <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
+  vcov = inf.fa, order.by = fa.res$V, sandwich = FALSE)
+
+## the same with OPG (outer product of gradients) vcov estimator
+fa.cums1opg <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
+  order.by = fa.res$V, parm = 1:3)
+fa.cums2opg <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
+  order.by = fa.res$V)
+
 
 # Plot individual cumulative score processes:
-plot(fa.cums, agg=FALSE, ylim=c(-1.8,0.3),
-     ylab="Cumulative scores", xlab="Age")
+plot(fa.cums1, aggregate = FALSE, ylim = c(-1.8, 1.8), xlab = "Age")
 
-# Obtain results of double-max test:
-sctest(fa.cums, functional="max")
+## Cramer-von Mises test
+## -> all lead to significant results but with all 19 parameters
+##    the peak is even clearer at closer to age = 16
+par(mfrow = c(2, 2))
[…]

### 2011-01-18 www/def.model.R
added supLR computations by hand
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@22 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -96,3 +96,51 @@ def.model <- function (fadat, age)
                      mxAlgebraObjective("h12"))
     TwoFA
 }
+
+def.model1 <- function(dat)
+{
+	mxModel("Factor Model", 
+		type = "RAM",
+		mxData(observed = dat, type = "raw"),
+		manifestVars = c("x1", "x2", "x3",
+		  "y1", "y2", "y3"), 
+		latentVars = c("F1", "F2"),
+		mxPath(from = c("x1", "x2", "x3",
+				"y1", "y2", "y3"),
+		       arrows = 2,
+		       free = TRUE, 
+		       values = c(1, 1, 1, 1, 1, 1),
+		       labels = c("e1", "e2", "e3",
+				  "e4", "e5", "e6"),
+		       lbound=0),
+		mxPath(from = c("F1", "F2"),
+		       arrows = 2,
+		       all = TRUE,
+		       free = c(FALSE, TRUE, TRUE, FALSE),
+		       values = c(1, 0.5, 0.5, 1),
+		       labels = c("varF1", "cov", "cov", "varF2"),
+		       lbound=c(NA,0,0,NA)),
+		mxPath(from = "F1",
+		       to = c("x1", "x2", "x3"),
+		       arrows = 1,
+		       free = TRUE,
+		       values = c(1, 1, 1),
+		       labels = c("l1", "l2", "l3"),
+		       lbound=0),
+		mxPath(from = "F2", 
+		       to = c("y1", "y2", "y3"),
+		       arrows = 1,
+		       free = TRUE, 
+		       values = c(1, 1, 1),
+		       labels = c("l4", "l5", "l6"),
+		       lbound=0), 
+		mxPath(from = "one",
+		       to = c("x1", "x2", "x3", "y1", "y2", "y3",
+			      "F1", "F2"),
+		       arrows = 1,
+		       free = c(TRUE, TRUE, TRUE, TRUE,
+				TRUE, TRUE, FALSE, FALSE),
+		       values = c(1, 1, 1, 1, 1, 1, 0, 0),
+		       labels = c("meanx1", "meanx2", "meanx3",
+                                  "meany1", "meany2", "meany3", NA, NA)))
+}

### 2011-01-18 www/script.R

@@ -22,12 +22,16 @@ fa.mod <- def.model(dat, V)
 # Fit model via OpenMx
 mx.res <- mxRun(fa.mod)
 
+## the same with a single model (not split/constrained by age)
+fa.mod1 <- def.model1(dat)
+mx.res1 <- mxRun(fa.mod1)
+
 # Organize results for strucchange:
-fa.res <- list(res = mx.res, dat = dat, V = V)
+fa.res <- list(res = mx.res1, dat = dat, V = V)
 
 ## store scores in list as well to avoid slow re-computation
 fa.res$estfun <- score.fa(fa.res)
-colnames(fa.res$estfun) <- summary(mx.res)$parameters$name
+colnames(fa.res$estfun) <- summary(mx.res1)$parameters$name
 
 # Calculate cumulative scores for first three factor loadings:
 fa.cums1 <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
@@ -62,3 +66,25 @@ plot(fa.cums1,    functional = supLM(0.1), main = "supLM: 3 pars, Info")
 plot(fa.cums2,    functional = supLM(0.1), main = "supLM: 19 pars, Info")
 plot(fa.cums1opg, functional = supLM(0.1), main = "supLM: 3 pars, OPG")
 plot(fa.cums2opg, functional = supLM(0.1), main = "supLM: 19 pars, OPG")
+
+## compute supLR "by hand"
+lrstat <- function(age) {
+  summary(mx.res1)$Minus2LogLikelihood -
+  summary(mxRun(def.model1(dat[V <= age,])))$Minus2LogLikelihood -
+  summary(mxRun(def.model1(dat[V >  age,])))$Minus2LogLikelihood
+}
+fa.age <- sort(V[V >= quantile(V, 0.1) & V <= quantile(V, 0.9)])
+fa.lrstat <- sapply(fa.age, lrstat)
+
+## compare supLM and supLR test
+## (the critical value is the same)
+par(mfrow = c(1, 1))
+plot(fa.cums2, functional = supLM(0.1),
+  xlab = "Age", ylab = "LM and LR statistics", main = "",
+  ylim = c(0, 65))
+lines(fa.age, fa.lrstat, col = 4)
+abline(v = fa.age[which.max(fa.lrstat)], lty = 2)
+## -> in this case, the conclusions are exactly the same, the
+##    maximum is even assumed for the same value (15.96474),
+##    the remaining differences may possibly be reduced when
+##    the numerical computations are improved

### 2011-01-19 MIpaper/par_instable.tex
Added maxLM to simulation + minor edits
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@23 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ordered with respect to the random variable $V$ of interest such that
-$v_1 \le v_2 \le \dots \le v_n$.
+$v_1 \le v_2 \le \dots \le v_n$.  We also assume that the measurement
+model is correctly specified, as is implicitly assumed under traditional
+measurement invariance approaches.
 
@@
 invariance).  We 
-examine the power and error rates of three tests: the double-max test, the
-Cram\'{e}r-von Mises test, and the traditional likelihood
+examine the power and error rates of four tests: the double-max test, the
+Cram\'{e}r-von Mises test, the max-LM test, and the traditional likelihood
 ratio (difference) test involving nested 
 multiple-group models.  The likelihood ratio test has an advantage over the
-other two tests here, in that it requires advance definition of
+other three tests here, in that it requires advance definition of
 $\eta$.  This test is useful for comparative purposes,
@@
 \subsection{Results}
-% MODIFY RESULTS TO DISCUSS ALL THREE TESTS
+% MODIFY RESULTS TO DISCUSS ALL FOUR TESTS
 % ALSO NEED 0-SE CONDITION
@@
 
-Comparing only the two proposed test statistics, we see that the
-Cram\'{e}r-von Mises test has better power in all conditions.  This is
+Comparing the three proposed test statistics, we see that the
+Cram\'{e}r-von Mises test and max-LM test are comparable, and they are
+both better than the double-max test in all conditions.  This is
 similar to the general recommendation described by Zeileis et al.\
-\citeyear{ZeiSha10}, stating that the Cram\'{e}r-von Mises test
-is generally superior when multiple parameters change simultaneously.
+\citeyear{ZeiSha10}, stating that tests involving sums of scores
+are generally superior when multiple parameters change simultaneously.
 Because the three verbal factor loadings all changed
 across the two subgroups in the simulation, the Cram\'{e}r-von Mises
-test excels.
+and max-LM tests excel.
 
@@
 violations were more easily detected at larger sample sizes, with the
-Cram\'{e}r-von Mises test having better power due to multiple
+Cram\'{e}r-von Mises and max-LM tests having better power due to multiple
 parameters being involved in the invariance violation.
@@
 \begin{table*}
-\caption{Simulated power and Type I error rates for three test statistics
+\caption{Simulated power and Type I error rates for four test statistics
   across three sample sizes and four magnitudes of measurement
   invariance violations. Abbreviations: CvM = Cram\'{e}r-von Mises
-  test; DM = Double-max test; $\chi^2$ = Chi-square likelihood ratio
-  test.}
+  test; MLM = Maximum Lagrange multiplier test; DM = Double-max test; 
+  $\chi^2$ = Chi-square likelihood ratio test.}
 \label{tab:sim1}
@@
 100 & CvM       & 0.05 & 0.10 & 0.33 & 0.63 \\ 
+    & MLM       &      & 0.09 & 0.28 & 0.64 \\
     & DM        & 0.03 & 0.06 & 0.19 & 0.41 \\ 
@@
 200 & CvM       & 0.05 & 0.19 & 0.64 & 0.96 \\
+    & MLM       &      & 0.19 & 0.63 & 0.96 \\ 
     & DM        & 0.03 & 0.12 & 0.44 & 0.88 \\ 
@@
 500 & CvM       & 0.05 & 0.44 & 0.98 & 1.00 \\ 
+    & MLM       &      & 0.41 & 0.99 & 1.00 \\
     & DM        & 0.04 & 0.35 & 0.94 & 1.00 \\ 
@@
 
-% Need to re-do simulation to monitor specific parameters in the
-% posthoc test.
-
-% Cramer-von Mises now in Sim 1.  May skip Sim 2 for the time being.
-%\section{Simulation 2: Improving Power}
-%Simulation 2 examines whether we can improve power using a Cramer-von
-%Mises statistic (summing instead of taking maximum).
-
 % Table 2 of Nyblom89 contains critical values that may prove useful.
@@
 % See Zeileis05 p. 450-451
-
-% Use model that is similar to the example.
 

### 2011-01-20 www/def.model.R
Minor changes to def.model to make model estimation more stable
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@24 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -46,7 +46,8 @@ def.model <- function (fadat, age)
                                              TRUE, TRUE, FALSE, FALSE),
                                     values = c(1, 1, 1, 1, 1, 1, 0, 0),
                                     labels = c("meanx1", "meanx2", "meanx3",
-                                      "meany1", "meany2", "meany3", NA, NA))), 
+                                      "meany1", "meany2", "meany3", NA, NA),
+                                    lbound=0)), 
                      mxModel("group2",
                              type = "RAM",
                              mxData(observed = dat2, type = "raw"),
@@ -90,7 +91,8 @@ def.model <- function (fadat, age)
                                              TRUE, TRUE, FALSE, FALSE), 
                                     values = c(1, 1, 1, 1, 1, 1, 0, 0),
                                     labels = c("meanx1", "meanx2", "meanx3",
-                                      "meany1", "meany2", "meany3", NA, NA))),
+                                      "meany1", "meany2", "meany3", NA, NA),
+                                    lbound=0)),
                      mxAlgebra(expression=group1.objective + group2.objective, 
                                name = "h12"),
                      mxAlgebraObjective("h12"))
@@ -142,5 +144,6 @@ def.model1 <- function(dat)
 				TRUE, TRUE, FALSE, FALSE),
 		       values = c(1, 1, 1, 1, 1, 1, 0, 0),
 		       labels = c("meanx1", "meanx2", "meanx3",
-                                  "meany1", "meany2", "meany3", NA, NA)))
+                                  "meany1", "meany2", "meany3", NA, NA),
+                       lbound=0))
 }

### 2011-01-20 www/script.R

@@ -4,6 +4,7 @@
 # Load requisite packages
 library("OpenMx")
 library("strucchange")
+library("numDeriv")
 
 # Read data
 obs.data <- dget("data.dat")

### 2011-01-20 www/gen.data.R
Added code for data generation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@25 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,54 @@
+"gen.data" <- function(sampsize,ses)
+  {
+    # Generates data from a factor analysis model that violates
+    # measurement invariance.  Also generates a continuous auxiliary
+    # variable related to the invariance (loosely called "age").
+    
+    # ses is number of SEs between group 1's parameters and group 2's
+    # parameters.
+    
+    require("mvtnorm")
+
+    # Could eventually manipulate this so break point is not directly
+    # in the middle
+    num.y <- round(sampsize/2,0)
+    num.o <- sampsize - num.y
+
+    age <- c(runif(num.y,13,16),runif(num.o,16,18))
+    
+    # Define parameter vectors/matrices:
+    mu <- c(29.32,24.70,14.84,10.59,19.30,18.01)
+
+    # Loadings for "young" individuals:
+    lambda.y <- matrix(0,6,2)
+    lambda.y[,1] <- c(4.92,2.96,5.96,0,0,0)
+    lambda.y[,2] <- c(0,0,0,3.24,4.32,7.21)
+
+    # Loadings for "old" individuals:
+    lambda.o <- matrix(0,6,2)
+    lambda.o[,1] <- lambda.y[,1] - (ses * c(.87,.6,1.32,rep(0,3)))
+    lambda.o[,2] <- c(0,0,0,3.24,4.32,7.21)
+    
+    phi <- matrix(.48,2,2)
+    diag(phi) <- 1
+
+    psi <- matrix(0,6,6)
+    diag(psi) <- c(26.77,13.01,30.93,3.17,8.82,22.5)
+
+    # Initialize data matrix:
+    datmat <- matrix(0,sampsize,6)
+
+    # Generate z and u vectors:
+    z <- t(rmvnorm(sampsize,rep(0,2),phi))
+    u <- t(rmvnorm(sampsize,rep(0,6),psi))
+
+    # Based on z and u, generate data:
+    for (i in 1:num.y){
+      datmat[i,] <- mu + lambda.y%*%z[,i] + u[,i]
+    }
+    for (i in (num.y+1):sampsize){
+      datmat[i,] <- mu + lambda.o%*%z[,i] + u[,i]
+    }
+    
+    list(datmat=datmat, age=age)
+  }

### 2011-01-21 www/gen.data.R
added correct column names in DGP
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@26 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -37,6 +37,7 @@
 
     # Initialize data matrix:
     datmat <- matrix(0,sampsize,6)
+    colnames(datmat) <- c("x1", "x2", "x3", "y1", "y2", "y3")
 
     # Generate z and u vectors:
     z <- t(rmvnorm(sampsize,rep(0,2),phi))

### 2011-01-21 www/script2.R
added new script analogous to script.R which generates new artificial data on the fly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@27 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,93 @@
+## packages
+library("OpenMx")
+library("strucchange")
+library("numDeriv")
+library("mvtnorm")
+
+## additional code
+source("gen.data.R")
+source("fa_score_extraction.R")
+source("def.model.R")
+
+## generate artificial data
+set.seed(1090)
+obs.data <- gen.data(200, 2)
+dat <- obs.data$datmat
+V <- obs.data$age
+
+## full sample and "true" subsample models 
+mx.res0 <- mxRun(def.model1(dat))
+mx.res1 <- mxRun(def.model1(dat[1:100,]))
+mx.res2 <- mxRun(def.model1(dat[101:200,]))
+
+## check: Which model parameters differ pre- and post-break?
+round(structure( 
+  (summary(mx.res1)$parameters[,5] - summary(mx.res2)$parameters[,5]) /
+    sqrt(summary(mx.res1)$parameters[,6]^2 + summary(mx.res2)$parameters[,6]^2),
+  names = summary(mx.res1)$parameters[,1]
+), digits = 2)
+
+## organize results for strucchange
+fa.res <- list(res = mx.res0, dat = dat, V = V)
+fa.res$estfun <- score.fa(fa.res)
+colnames(fa.res$estfun) <- summary(mx.res0)$parameters$name
+
+## empirical fluctuation processes (3 vs 19 pars, info vs OPG vcov)
+gefp_3_info <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
+  vcov = inf.fa, order.by = fa.res$V, sandwich = FALSE, parm = 1:3)
+gefp_19_info <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
+  vcov = inf.fa, order.by = fa.res$V, sandwich = FALSE)
+gefp_3_opg <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
+  order.by = fa.res$V, parm = 1:3)
+gefp_19_opg <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
+  order.by = fa.res$V)
+
+## Double-maximum test
+par(mfrow = c(2, 2))
+plot(gefp_3_info,  functional = maxBB, main = "dmax: 3 pars, Info")
+plot(gefp_19_info, functional = maxBB, main = "dmax: 19 pars, Info")
+plot(gefp_3_opg,   functional = maxBB, main = "dmax: 3 pars, OPG")
[…]

### 2011-01-21 www/fa_score_extraction.R
Added option for score calculation via analytic derivatives
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@28 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,43 +1,80 @@
-"score.fa" <- function(res.list)
+"score.fa" <- function(res.list,derivs="a")
   {
     # Code to extract scores from a 2-factor model that
     # is fit by OpenMx.  res is a list containing
     # the model fit by mxRun() and the observed data matrix.
 
-    require(numDeriv)
+    # Possible options for derivs:
+    #   a: Analytic derivatives (fast)
+    #   n: Numerical derivatives (slow)
+
+    derivs <- tolower(derivs)
+    if (derivs=="n"){
+      require("numDeriv")
+    }
 
     res <- res.list$res
     dat <- res.list$dat
-    
-    # Define casewise likelihood function
-    likefun <- function(x,ind.dat){
-      mu <- x[14:19]
-      lam <- matrix(c(x[1:3],rep(0,6),x[4:6]),6,2) # loadings
-      phi <- matrix(c(1,x[13],x[13],1),2,2) # factor covs
-      psi <- diag(x[7:12]) # error vars
-
-      impl.cov <- lam %*% phi %*% t(lam) + psi
-
-      fval <- log(det(impl.cov)) + t(ind.dat - mu) %*% solve(impl.cov) %*%
-                                   (ind.dat - mu)
-      fval
-    }
 
     # Results from OpenMx
-    ests <- res@output$estimate
-    # Need n from OpenMx output
-    n <- nrow(dat)
+    theta <- res@output$estimate
 
-    scores <- matrix(NA,n,length(ests))
-    for (j in 1:n){
-      scores[j,] <- grad(likefun, ests,
-                         ind.dat=as.matrix(dat)[j,],
-                         method="Richardson")
-    }
+    # Get n and k from OpenMx output
+    n <- nrow(dat)
+    k <- length(theta)
+
+    # Score matrix
+    scores <- matrix(NA,n,k)
+    
+    switch(derivs, `n` = {
+    
+      # Define casewise likelihood function
+      likefun <- function(x,ind.dat){
+        mu <- x[14:19]
+        lam <- matrix(c(x[1:3],rep(0,6),x[4:6]),6,2) # loadings
+        phi <- matrix(c(1,x[13],x[13],1),2,2) # factor covs
+        psi <- diag(x[7:12]) # error vars
+
[…]

### 2011-01-22 www/mz.R
Provided new file mz.R that collects and streamlines all code:
  o dgp() generates an artificial data.frame with 7 columns:
    6 data columns and 1 "age" column.

  o mzfit() takes a data.frame/matrix as input as generated
    by dgp() and returns the fitted "MxModel" object.

  o Methods for "MxModel" objects are provided for the generics
    coef(), vcov(), logLik(), bread(), and estfun().

  o This simplifies application of gefp() as well as many other
    fitted model computations.

script2.R has been adapted to the new code in mz.R.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/

@@ -1,0 +1,178 @@
+mzfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
+{
+  ## data checks
+  data <- as.data.frame(data)
+  stopifnot(c("x1", "x2", "x3", "y1", "y2", "y3") %in% names(data))
+  data <- data[, c("x1", "x2", "x3", "y1", "y2", "y3")]
+
+  ## need OpenMx for model fitting
+  stopifnot(require("OpenMx"))
+
+  ## model from Merkle & Zeileis
+  rval <- mxModel("Factor Model", 
+    type = "RAM",
+    mxData(observed = data, type = "raw"),
+    manifestVars = c("x1", "x2", "x3", "y1", "y2", "y3"), 
+    latentVars = c("F1", "F2"),
+    mxPath(from = c("x1", "x2", "x3", "y1", "y2", "y3"),
+      arrows = 2, free = TRUE, values = c(1, 1, 1, 1, 1, 1),
+      labels = c("e1", "e2", "e3", "e4", "e5", "e6"), lbound=0),
+    mxPath(from = c("F1", "F2"),
+      arrows = 2, all = TRUE, free = c(FALSE, TRUE, TRUE, FALSE),
+      values = c(1, 0.5, 0.5, 1), labels = c("varF1", "cov", "cov", "varF2"),
+      lbound=c(NA, 0, 0, NA)),
+    mxPath(from = "F1", to = c("x1", "x2", "x3"),
+      arrows = 1, free = TRUE, values = c(1, 1, 1),
+      labels = c("l1", "l2", "l3"), lbound=0),
+    mxPath(from = "F2", to = c("y1", "y2", "y3"),
+      arrows = 1, free = TRUE, values = c(1, 1, 1),
+      labels = c("l4", "l5", "l6"), lbound=0), 
+    mxPath(from = "one",
+      to = c("x1", "x2", "x3", "y1", "y2", "y3", "F1", "F2"),
+      arrows = 1, free = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE),
+      values = c(1, 1, 1, 1, 1, 1, 0, 0), labels = c("meanx1", "meanx2",
+        "meanx3", "meany1", "meany2", "meany3", NA, NA),
+      lbound = 0)
+  )
+
+  ## fit model
+  rval <- mxRun(rval, silent = silent, suppressWarnings = suppressWarnings, ...)
+  return(rval)
+}
+
[…]

### 2011-01-22 www/script2.R

@@ -1,46 +1,28 @@
-## packages
+## packages and code
 library("OpenMx")
 library("strucchange")
-library("numDeriv")
 library("mvtnorm")
-
-## additional code
-source("gen.data.R")
-source("fa_score_extraction.R")
-source("def.model.R")
+source("mz.R")
 
 ## generate artificial data
 set.seed(1090)
-obs.data <- gen.data(200, 2)
-dat <- obs.data$datmat
-V <- obs.data$age
+d <- dgp(200, 2)
 
 ## full sample and "true" subsample models 
-mx.res0 <- mxRun(def.model1(dat))
-mx.res1 <- mxRun(def.model1(dat[1:100,]))
-mx.res2 <- mxRun(def.model1(dat[101:200,]))
+mz0 <- mzfit(d)
+mz1 <- mzfit(d[1:100,])
+mz2 <- mzfit(d[101:200,])
 
 ## check: Which model parameters differ pre- and post-break?
-round(structure( 
-  (summary(mx.res1)$parameters[,5] - summary(mx.res2)$parameters[,5]) /
-    sqrt(summary(mx.res1)$parameters[,6]^2 + summary(mx.res2)$parameters[,6]^2),
-  names = summary(mx.res1)$parameters[,1]
-), digits = 2)
-
-## organize results for strucchange
-fa.res <- list(res = mx.res0, dat = dat, V = V)
-fa.res$estfun <- score.fa(fa.res)
-colnames(fa.res$estfun) <- summary(mx.res0)$parameters$name
+(coef(mz1) - coef(mz2)) / sqrt(diag(vcov(mz1)) + diag(vcov(mz2)))
 
 ## empirical fluctuation processes (3 vs 19 pars, info vs OPG vcov)
-gefp_3_info <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
-  vcov = inf.fa, order.by = fa.res$V, sandwich = FALSE, parm = 1:3)
-gefp_19_info <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
-  vcov = inf.fa, order.by = fa.res$V, sandwich = FALSE)
-gefp_3_opg <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
-  order.by = fa.res$V, parm = 1:3)
-gefp_19_opg <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
[…]

### 2011-01-22 www/mz.R
wrote new simulation suite in sim.R (employing the same simulation strategy as Kleiber & Zeileis, 2010)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@30 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -53,6 +53,9 @@ vcov.MxModel <- function(object, ...)
 bread.MxModel <- function(x, ...)
   2 * solve(x@output$calculatedHessian) * x@data@numObs
 
+info.MxModel <- function(x, ...)
+  (1/(2 * x@data@numObs)) * x@output$calculatedHessian
+
 estfun.MxModel <- function(x, type = c("analytic", "numeric"), ...)
 {
   ## numeric or analytic derivatives
@@ -116,63 +119,4 @@ estfun.MxModel <- function(x, type = c("analytic", "numeric"), ...)
   })
 
   return(scores)
-}
-
-
-dgp <- function(n = 200, diff = 3)
-{
-  # Generates data from a factor analysis model that violates
-  # measurement invariance.  Also generates a continuous auxiliary
-  # variable related to the invariance (loosely called "age").
-  
-  # ses is number of SEs between group 1's parameters and group 2's
-  # parameters.
-  sampsize <- n
-  ses <- diff
-  
-  stopifnot(require("mvtnorm"))
-
-  # Could eventually manipulate this so break point is not directly
-  # in the middle
-  num.y <- round(sampsize/2, 0)
-  num.o <- sampsize - num.y
-
-  age <- c(runif(num.y,13,16),runif(num.o,16,18))
-  
-  # Define parameter vectors/matrices:
-  mu <- c(29.32,24.70,14.84,10.59,19.30,18.01)
-
-  # Loadings for "young" individuals:
-  lambda.y <- matrix(0,6,2)
-  lambda.y[,1] <- c(4.92,2.96,5.96,0,0,0)
-  lambda.y[,2] <- c(0,0,0,3.24,4.32,7.21)
-
-  # Loadings for "old" individuals:
-  lambda.o <- matrix(0,6,2)
-  lambda.o[,1] <- lambda.y[,1] - (ses * c(.87,.6,1.32,rep(0,3)))
-  lambda.o[,2] <- c(0,0,0,3.24,4.32,7.21)
-  
-  phi <- matrix(.48,2,2)
-  diag(phi) <- 1
-
-  psi <- matrix(0,6,6)
-  diag(psi) <- c(26.77,13.01,30.93,3.17,8.82,22.5)
-
-  # Initialize data matrix:
-  datmat <- matrix(0,sampsize,6)
[…]

### 2011-01-22 www/script2.R

@@ -3,6 +3,7 @@ library("OpenMx")
 library("strucchange")
 library("mvtnorm")
 source("mz.R")
+source("sim.R")
 
 ## generate artificial data
 set.seed(1090)
@@ -19,9 +20,9 @@ mz2 <- mzfit(d[101:200,])
 ## empirical fluctuation processes (3 vs 19 pars, info vs OPG vcov)
 gefp_3_opg <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3)
 gefp_19_opg <- gefp(mz0, fit = NULL, order.by = d$age)
-gefp_3_info <- gefp(mz0, fit = NULL, vcov = function(x, ...) solve(bread(x)),
+gefp_3_info <- gefp(mz0, fit = NULL, vcov = info.MxModel,
   order.by = d$age, sandwich = FALSE, parm = 1:3)
-gefp_19_info <- gefp(mz0, fit = NULL, vcov = function(x, ...) solve(bread(x)),
+gefp_19_info <- gefp(mz0, fit = NULL, vcov = info.MxModel,
   order.by = d$age, sandwich = FALSE)
 
 ## Double-maximum test
@@ -69,3 +70,13 @@ abline(v = mz.age[which.max(mz.lrstat)], lty = 2)
 ##     is close to supLM
 ##   - Tests with information matrix seem to be more powerful
 ##     than OPG.
+
+
+library("strucchange")
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1090)
+sc_sim <- simulation()
+
+## Results in tabular form (replication of Table 6)
+tab <- xtabs(power ~ intensity + test + angle + timing, data = sc_sim)
+ftable(tab, row.vars = c("angle", "timing", "test"), col.vars = "intensity")

### 2011-01-22 www/sim.R

@@ -1,0 +1,125 @@
+## Data-generating process
+dgp <- function(nobs = 200, diff = 3)
+{
+  # Generates data from a factor analysis model that violates
+  # measurement invariance.  Also generates a continuous auxiliary
+  # variable related to the invariance (loosely called "age").
+  
+  # ses is number of SEs between group 1's parameters and group 2's
+  # parameters.
+  sampsize <- nobs
+  ses <- diff
+  
+  # Could eventually manipulate this so break point is not directly
+  # in the middle
+  num.y <- round(sampsize/2, 0)
+  num.o <- sampsize - num.y
+
+  age <- c(runif(num.y,13,16),runif(num.o,16,18))
+  
+  # Define parameter vectors/matrices:
+  mu <- c(29.32,24.70,14.84,10.59,19.30,18.01)
+
+  # Loadings for "young" individuals:
+  lambda.y <- matrix(0,6,2)
+  lambda.y[,1] <- c(4.92,2.96,5.96,0,0,0)
+  lambda.y[,2] <- c(0,0,0,3.24,4.32,7.21)
+
+  # Loadings for "old" individuals:
+  lambda.o <- matrix(0,6,2)
+  lambda.o[,1] <- lambda.y[,1] - (ses * c(.87,.6,1.32,rep(0,3)))
+  lambda.o[,2] <- c(0,0,0,3.24,4.32,7.21)
+  
+  phi <- matrix(.48,2,2)
+  diag(phi) <- 1
+
+  psi <- matrix(0,6,6)
+  diag(psi) <- c(26.77,13.01,30.93,3.17,8.82,22.5)
+
+  # Initialize data matrix:
+  datmat <- matrix(0,sampsize,6)
+  colnames(datmat) <- c("x1", "x2", "x3", "y1", "y2", "y3")
+
+  # Generate z and u vectors:
+  z <- t(rmvnorm(sampsize,rep(0,2),phi))
+  u <- t(rmvnorm(sampsize,rep(0,6),psi))
+
+  # Based on z and u, generate data:
+  for (i in 1:num.y){
+    datmat[i,] <- mu + lambda.y%*%z[,i] + u[,i]
+  }
+  for (i in (num.y+1):sampsize){
+    datmat[i,] <- mu + lambda.o%*%z[,i] + u[,i]
+  }
+  
+  cbind(as.data.frame(datmat), age)
+}
+
+## Evaluate power simulation on a single dgp() scenario
[…]

### 2011-01-31 www/sim.R
More extensive simulation script, along with resulting .rda.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@31 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -56,7 +56,7 @@ dgp <- function(nobs = 200, diff = 3)
 }
 
 ## Evaluate power simulation on a single dgp() scenario
-testpower <- function(nrep = 1000, size = 0.05, ...)
+testpower <- function(nrep = 5000, size = 0.05, verbose = TRUE, ...)
 {
   pval <- matrix(rep(NA, 6 * nrep), ncol = 6)
   colnames(pval) <- c("dmax3", "CvM3", "supLM3", "dmax19", "CvM19", "supLM19")
@@ -78,25 +78,33 @@ testpower <- function(nrep = 1000, size = 0.05, ...)
       pval[i, 6] <- sctest(mz_gefp19, functional = supLM(0.1))$p.value
     }
   }
-  return(colMeans(pval < size, na.rm = TRUE))
+  rval <- colMeans(pval < size, na.rm = TRUE)
+  if(verbose) print(rval)
+  
+  return(rval)
 }
 
 ## Loop over scenarios
-simulation <- function(diff = seq(0, 5, by = 0.5), ...)
+simulation <- function(diff = seq(0, 5, by = 0.25),
+  nobs = c(100, 200, 500), verbose = TRUE, ...)
 {
-  prs <- expand.grid(diff = diff) ## somewhat complex setup if we want to add further vars
+  prs <- expand.grid(diff = diff, nobs = nobs)
   nprs <- nrow(prs)
   
   test <- c("dmax3", "CvM3", "supLM3", "dmax19", "CvM19", "supLM19")
   ntest <- length(test)
 
   pow <- matrix(rep(NA, ntest * nprs), ncol = ntest)
-  for(i in 1:nprs) pow[i,] <- testpower(diff = prs$diff[i], ...)
+  for(i in 1:nprs) {
+    if(verbose) print(prs[i,])
+    pow[i,] <- testpower(diff = prs$diff[i], nobs = prs$nobs[i], verbose = verbose, ...)
+  }
 
   rval <- data.frame()
   for(i in 1:ntest) rval <- rbind(rval, prs)
   rval$test <- factor(rep(c("dmax", "CvM", "supLM", "dmax", "CvM", "supLM"), each = nprs), levels = c("dmax", "CvM", "supLM"))
   rval$pars <- factor(rep(c("3", "3", "3", "19", "19", "19"), each = nprs), levels = c("3", "19"))
+  rval$nobs <- factor(rval$nobs)
   rval$power <- as.vector(pow)
   return(rval)
 }
[…]

### 2011-01-31 MIpaper/par_instable.tex
Small wording changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@32 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 with three scales arising from the verbal factor and three 
-scales arising from the mathematical second factor.
+scales arising from the mathematical factor.
 For measurement invariance violations, we specify that $V$

### 2011-02-04 www/script2.R
Added code for manuscript example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@33 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -53,12 +53,55 @@ mz.lrstat <- sapply(mz.age, lrstat)
 
 ## compare supLM and supLR test
 ## (the critical value is the same)
+#pdf("lmlr_19.pdf")
+
 par(mfrow = c(1, 1))
 plot(gefp_19_info, functional = supLM(0.1),
   xlab = "Age", ylab = "LM and LR statistics", main = "",
   ylim = c(0, 65))
 lines(mz.age, mz.lrstat, col = 4)
 abline(v = mz.age[which.max(mz.lrstat)], lty = 2)
+
+#dev.off()
+
+## Plot for example in the paper (3 parameters and 19 parameters)
+#pdf("gefp_3_19.pdf",height=9,width=6)
+par(mfcol = c(3, 2))
+
+## 3 parameters:
+plot(gefp_3_info,  functional = maxBB, main = "Three Parameters\n dmax", xlab="Age")
+plot(gefp_3_info,  functional = meanL2BB, main = "CvM", xlab="Age")
+plot(gefp_3_info,  functional = supLM(0.1), main = "supLM", xlab="Age")
+
+## 19 parameters:
+plot(gefp_19_info,  functional = maxBB, main = "Nineteen Parameters\n dmax", xlab="Age")
+plot(gefp_19_info,  functional = meanL2BB, main = "CvM", xlab="Age")
+plot(gefp_19_info,  functional = supLM(0.1), main = "supLM", xlab="Age")
+
+#dev.off()
+
+
+## Plot for posthoc tests in the example:
+#pdf("gefp_posthoc.pdf")
+
+plot(gefp_3_info,  functional = maxBB, main = "", xlab="Age", ylim=c(0,2.25),
+     aggregate=FALSE)
+
+#dev.off()
+
+
+## Find points at which test statistics achieve their maxima:
+
+# L2 statistics:
+L2_process <- zoo(apply(as.matrix(gefp_3_info$process)^2, 1, sum),
+                    time(gefp_3_info))
+L2_max <- L2_process[which.max(L2_process)]
+
+# Double max statistic:
+dmax_process <- zoo(apply(abs(as.matrix(gefp_3_info$process)), 1, max),
+                    time(gefp_3_info))
+dmax <- dmax_process[which.max(dmax_process)]
+
 
 ## EMPIRICAL RESULTS (all conforming with theory!):
 ##   - Only first 3 of all 19 parameters change.

### 2011-02-04 MIpaper/par_instable.tex
Revised example section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@34 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\fixme{I have explicitly written down all three test statistics: DM because it is
-intuitive, max LM because we need it to connect to the previous section, and CvM
-because it nicely complements the picture. For the simulations, I think we should
-also consider all three tests, maybe also the max LR test. Of course, we shouldn't
-make things too complex but we can then still decide which results to choose.}
-
 \subsection{Critical Values \& $p$-values.}
@@
 As an example of the ideas discussed above, consider a
-hypothetical battery of six scales administered to students aged 14 to 17
+hypothetical battery of six scales administered to students aged 13 to 18
 years.  Three of the scales are intended to measure verbal ability, and
@@
 may observe a maturation effect in the resulting data, whereby the scales
-measure older students better than younger students (i.e., the scales
-are too difficult for younger students).  The researcher's goal is to
+measure older students better than younger students (i.e., younger
+students have not yet learned all the material).  The researcher's goal is to
 study whether the scales are measurement invariant with
@@
 verbal factor loadings in the model: 
-if students are 16 or 17 years of age, then the factor
+if students are 16 through 18 years of age, then the factor
 loadings corresponding to the first 
 factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
-Figure~\ref{fig:famod}.  If students are 14 or 15 years of age,
+Figure~\ref{fig:famod}.  If students are 13 through 15 years of age,
 however, then the factor loadings corresponding to the first factor
@@
 \eqref{eq:cumscore}.  Finally, we
-obtained a 
-double-max statistic via Equation~\eqref{eq:dmax} and associated
-$p$-value via Equation~\eqref{eq:absmax_p}.  Note that, because we
-were examining only the three verbal factor loading parameters, the
-specific null hypothesis being tested is:
+obtained various test statistics and $p$-values from the cumulative score
+process.  These include the double-max statistic from \eqref{eq:dmax},
+the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the max-LM
+statistic from \eqref{eq:maxlm}.
+
+Note that the tests give us the flexibility to test various subsets of
+parameters.  For example, were we only interested in the 
+three verbal factor loading parameters, we can test
 \begin{equation}
@@
 where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
-verbal factor loading parameters for student $i$.
+verbal factor loading parameters for student $i$.  Alternatively, we
+can consider all model parameters (including means), leading to a test
+of \eqref{eq:h0}.  We consider both of these tests below.
 
 \subsection{Results}
-In the results section, we first describe overall results and then 
-describe estimation of $\eta$ and 
+In the results section, we first describe overall results.  We then 
+describe estimation of $\nu$, and 
 isolation of model parameters violating
-measurement invariance.\footnote{R code for obtaining these results is
-  available at \url{http://semtools.r-forge.r-project.org}.}
+measurement invariance.  R code for obtaining these results is
+  available at \url{http://semtools.r-forge.r-project.org}.
 
 \subsubsection{Overall Results}
-Results related to the cumulative score process are displayed in
-Figure~\ref{fig:cusumex}.  In this figure, each panel represents the
-process for a single factor loading, with student age on the x-axis and
-cumulative scores on the y-axis.  The dashed horizontal
-lines represent the critical value for rejecting $H_0$ at $\alpha=.05$
-based on the double-max statistic; if any of the three processes
-crosses the horizontal line, then we reject $H_0$.  It is seen that
-the processes for both the second and third factor loadings
-($\lambda_{21}$ and $\lambda_{31}$) cross the dashed line, indeed
-implying rejection of $H_0$.  The absolute maximum across all three processes,
-corresponding to the double-max test statistic, is 1.77.  This
-statistic is obtained from the minimum of the third
-panel (whose absolute value is the maximum), close to an
-age of 16.  It achieves a $p$-value of .012.
-
-\subsubsection{Estimation of $\eta$ and Parameter Isolation}
-The results of the test imply that the 
-verbal scales lack measurement invariance, and the age at which the
-maximum is achieved provides the estimate $\hat{\eta}$.
-The student corresponding to the maximum
-absolute cumulative sum (the solid circle in the third graph) is 15.7
-years of age, implying one subgroup of 
-students below 15.7 years and one above 15.7 years.  This agrees well
-with the true $\eta$ of 16.0 years.
-
-The results of the test also yield information about specific
-parameters violating measurement invariance.  We simultaneously tested
-three factor loading parameters for measurement invariance.  Given
-rejection of the hypothesis of measurement invariance, we can 
-do posthoc 
+Test statistics for the hypotheses \eqref{eq:exh0} and~\eqref{eq:h0}
+are displayed in  
+Figure~\ref{fig:cusumex}.
+Each panel displays a test 
+statistic's fluctuation across values of student age, with the first
+column containing tests of \eqref{eq:exh0} and the second column
+containing tests of \eqref{eq:h0}.  The solid
+horizontal 
+lines represent critical values for $\alpha=.05$, and
+the Cram\'{e}r-von Mises panels 
+also contain a dashed line depicting the value of the test
+statistic (test statistics for the others are simply 
+the maxima of the processes).
+In other words, for panels in the first and third rows, \eqref{eq:exh0} is
+rejected if the process crosses the horizontal line.  For panels in
+the second row, \eqref{eq:h0} is rejected if the dashed horizontal line is
+higher than the solid horizontal line.
+
+The figures display many test attributes, all of which generally
+conform with the underlying theory.  First, when we test
+only the three verbal factor loadings, all three tests can detect the
+change.  Conversely, when all nineteen model parameters are tested,
+only the double-max test detects a change.  This implies that 
+power increases as larger proportions of tested parameters
+change.  The results also imply that the double-max test is more
+sensitive to fluctuations among a small subset of parameters.
+
+Figure~\ref{fig:lrlm} 
+compares the max-LM statistic (black line) to the max-LR
+statistic (blue line) from \eqref{eq:maxlr}, as applied to testing
+\eqref{eq:h0}.  The critical values for these two
+tests are identical, hence the single horizontal line.
+The figure
+shows that the two statistics are very similar to one another, with
+both maxima at the dotted vertical line.  This
+is generally to be expected, because the two tests are asymptotically
+equivalent.  For this specific situation, however, the max-LR
+statistic rejects \eqref{eq:h0} while the max-LM statistic does not.
+More generally, the 
+max-LR statistic cannot be obtained from the empirical fluctuation
+process, so it must be computed ``by hand'' (i.e., by computing
+the LR statistic over all values of the threshold $\nu$, re-fitting the
+factor analysis model for each value).
+
+\subsubsection{Estimation of $\nu$ and Parameter Isolation}
+As described above, the tests of \eqref{eq:exh0} imply that the 
+verbal scales lack measurement invariance.  
+We can also use the tests to: (1) estimate the 
+threshold $\nu$, and (2) isolate specific parameters that violate
+measurement invariance.  
+For example, as described previously, estimates of $\nu$ can be obtained by
+examining the peaks in Figure~\ref{fig:cusumex}.  For all six panels
+in the figure, the peaks occur near an age of 16.1.  This agrees well
+with the true threshold of 16.0.
+
+The double-max test yields specific information about 
+parameters violating measurement invariance.  Given
+rejection of the hypothesis of measurement invariance \eqref{eq:exh0},
+we can do posthoc 
 tests to study specific parameters violating measurement invariance.
-This involves individual examination of each cumulative score process
-(i.e., of each of the three graphs in Figure~\ref{fig:cusumex}),
-along with a procedure that controls for Type I error.  Taking a
-Bonferroni approach, we can examine whether each process's maximum
-surpasses the critical value associated with an $\alpha$ of
-$.05/3=.017$.  This amounts to the same critical value as that used in
-the overall test, 
-meaning that we simply examine the individual processes that cross
-the horizontal dashed lines in Figure~\ref{fig:cusumex}.  The figure
-shows that the second and third parameters (i.e., $\lambda_{21}$ and
-$\lambda_{31}$) cross the dashed line, so we would conclude a measurement
-invariance violation with respect to age in the second and third
-verbal tests.  The fact that the cumulative score process for
-$\lambda_{11}$ did not achieve the critical value represents a Type II
+This involves individual examination of each parameter's cumulative
+score process,
+along with a procedure that controls for Type I error.  
+Figure~\ref{fig:posthoc} shows the individual cumulative score processes for the
+verbal factor loadings, with the horizontal lines reflecting the
+critical value at $\alpha=.05$.
+  The figure
+shows that the third parameter (i.e., $\lambda_{31}$) crosses the
+dashed line, so we would conclude a measurement 
+invariance violation with respect to age for the third
+verbal test.  The fact that the cumulative score process for the first
+and second loadings did not achieve the critical value represents a Type II
 error, which implicitly brings into question the posthoc tests' power.  We
@@
 
-% NEED TO MODIFY
 \begin{figure}
-\caption{Cumulative score processes for the verbal factor loadings,
+\caption{Three test statistics of \eqref{eq:exh0} and \eqref{eq:h0}, 
   based on the example involving measurement invariance with respect
-  to student age.  Each panel represents the cumulative score process
-  for a single factor loading; the red line represents the critical values for 
-  the absolute maximum at $\alpha=.05$.}
+  to student age.  Solid, horizontal lines represent critical values
+  at $\alpha=.05$, and the dotted, horizontal lines (second row)
+  represent values of the Cram\'{e}r-von Mises test statistic.}
 \label{fig:cusumex}
-\includegraphics[height=5in]{example.pdf}
+\includegraphics[height=9in]{gefp_3_19.pdf}
+\end{figure}
+
+\begin{figure}
+\caption{A comparison of the max-LM (black line) and max-LR (blue line) test
+  statistics for \eqref{eq:h0}.  The solid, horizontal line
+  corresponds to the critical value at $\alpha=.05$.}
+\label{fig:lrlm}
+\includegraphics[height=5in]{lmlr_19.pdf}
+\end{figure}
+
+\begin{figure}
+\caption{Posthoc tests of individual verbal factor loadings, based on
+  the absolute maximum of each cumulative score process.  The top
+  panel depicts the process for $\lambda_{11}$, the middle panel
+  depicts the process for $\lambda_{21}$, and the bottom panel depicts
+  the process for $\lambda_{31}$.  The solid, horizontal lines
+  correspond to the critical value at $\alpha=.05$.}
+\label{fig:posthoc}
+\includegraphics[height=5in]{gefp_posthoc.pdf}
 \end{figure}

### 2011-02-04 MIpaper/par_instable.tex
Partial revision to simulation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@35 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 of parameters in~${\bm \theta}$.
-\readme{You previously modified the above section so that $\theta$
+\readme{Achim: You previously modified the above section so that $\theta$
   consists only of parameters that are ``doubled'' under the
@@
   parameters are restricted to be equal across subgroups.  I am ok
-  with this, but most applications of LR tests with these models have
+  with this, but many applications of LR tests with these models have
   some parameters restricted to be equal across subgroups.  So I've
@@
 In this section, we conduct a simulation designed to
-examine the tests' power and Type I error rates in the context
-described previously (testing three factor loadings for measurement
-invariance).  We 
-examine the power and error rates of four tests: the double-max test, the
-Cram\'{e}r-von Mises test, the max-LM test, and the traditional likelihood
-ratio (difference) test involving nested 
-multiple-group models.  The likelihood ratio test has an advantage over the
-other three tests here, in that it requires advance definition of
-$\eta$.  This test is useful for comparative purposes,
-however, as it helps indicate the extent to which the other tests
-``lose'' power due to $\eta$ being unknown.
+examine the tests' power and Type I error rates in a context
+similar to the example.
+We examine the power and error rates of three tests: the double-max test, the
+Cram\'{e}r-von Mises test, and the max-LM test.  We also compare 
+tests involving only the parameters that changed with tests of all model
+parameters.  The example implied that the former tests had more
+power, and the simulations provide more detail on the extent to which
+power increases.  Finally, we also examine the tests' power across
+various magnitudes of measurement invariance violations.
 
 \subsection{Method}
-Simulated data were generated in a similar way to that of 
+Data were generated in a similar way to that of 
 the Example section.
 The data included verbal scales that violated measurement invariance
-with respect to students' ages, with students 14--15 years of age
-having smaller factor loadings than students 16--17 years of age.  
+with respect to students' ages, with students 13--15 years of age
+having smaller factor loadings than students 16--18 years of age.  
 Simulation data were generated from the same general factor analysis
@@
 of measurement 
-invariance violation was manipulated to examine power:
+invariance violation were manipulated to examine power:
 we examined power to detect 
@@
 200, 500$) and three magnitudes of violations (factor loadings
-for the younger students being 0, 1, 2, or 3 standard
+for the younger students being $(0, 0.25, 0.5, \dots, 5)$ standard
 errors\footnote{The standard errors used here were
@@
 
-For each sample size $\times$
-magnitude combination, 500 datasets were generated and tested for
-measurement invariance.  For the 0-standard error condition, we
-generated 5,000 datasets because small proportions are more
- difficult to estimate precisely.  In each dataset, half the
-students were 14--15 
-years of age and half were 16--17 years of age.  As stated previously,
-group
-designations ($\eta$) were assumed to be
-known for the likelihood ratio test.
+\readme{Achim: Is it true that you used 5000 reps per cell?}
+For each combination of sample size $\times$
+violation magnitude $\times$ number of parameters being tested,
+5,000 datasets were generated and tested.  In each dataset, half the
+students were 13--15 
+years of age and half were 16--18 years of age.
 
@@
 % ALSO NEED 0-SE CONDITION
-For each test statistic, power and Type I error rates for the twelve
-sample size $\times$ magnitude combinations are displayed in
-Table~\ref{tab:sim1}.  The 0-SE column lists Type I error rates; this
-is the situation where the tests reject the hypothesis of measurement
-invariance at $\alpha=.05$ when none exist.  This column 
-shows that the Cram\'{e}r-von Mises test has accurate Type I error
-rates at all sample sizes examined.
-The double-max test, on the other hand, is conservative at
-all sample sizes examined.  Finally, the likelihood ratio test is
-slightly liberal at smaller sample sizes, with Type I error rates of
-.07 and .06 at $n=100$ and $200$, respectively.  Type I error for the
-likelihood ratio test at $n=500$ is accurate, however.
-
-Focusing on power, the table includes many interesting results.
-First, the likelihood ratio test has superior power over the stochastic
-process-based tests at all sample sizes and invariance violations.
-This is likely due to the fact that the likelihood ratio test had
-extra information ($\eta$), whereas the other tests
-were testing measurement invariance over all values
-of $\eta$.  While power was lower for the tests developed in this
-paper, it was not terribly low.  At $n=200$ and above, all tests were
-on the same side of the traditional power cutoff of 0.8.  The tests
-were able to detect large measurement invariance violations (of three
-and sometimes two standard errors) at the larger sample sizes, and
-they were generally unable to detect the one-standard-error violation
-at any of the sample sizes studied.
-
-Comparing the three proposed test statistics, we see that the
-Cram\'{e}r-von Mises test and max-LM test are comparable, and they are
-both better than the double-max test in all conditions.  This is
-similar to the general recommendation described by Zeileis et al.\
-\citeyear{ZeiSha10}, stating that tests involving sums of scores
-are generally superior when multiple parameters change simultaneously.
-Because the three verbal factor loadings all changed
-across the two subgroups in the simulation, the Cram\'{e}r-von Mises
-and max-LM tests excel.
-
-In summary, we found that the proposed tests have less power than a
-likelihood ratio test with known subgroups.  However, the power levels
-remain adequate after considering the fact that the proposed tests are
-testing measurement invariance across all possible subgroups.  For the
-two-factor model examined here, all tests had trouble detecting small
-measurement invariance violations at small sample size.  Larger
-violations were more easily detected at larger sample sizes, with the
-Cram\'{e}r-von Mises and max-LM tests having better power due to multiple
-parameters being involved in the invariance violation.
-
-% 0-SE column is fixed for n=100 (using 5000 reps instead of 500)
+Full simulation results are presented in Figure~\ref{fig:simres}, and
+a subset of the results (for finer distinctions) is displayed in
+Table~\ref{tab:simres}.  In describing the results, we largely refer
+to the figure.
+
+Figure~\ref{fig:simres} displays power as a function of violation
+magnitude, with panels for each combination of sample size $\times$
+number of parameters being tested.  
+Circles reflect the double-max test, triangles reflect the max-LM
+test, and plus signs reflect the Cram\`{e}r-von Mises test.  One can
+generally observe that simultaneous tests of all 19 parameters
+result in decreased power, with the tests performing very similarly at
+the larger sample sizes.  The tests distinguish themselves from one
+another when only the three factor loadings are tested, with the
+Cram\`{e}r-von Mises test having the most power, followed by max-LM,
+followed by the double-max test.  This advantage decreases with
+increases in sample size.  
+Absolute differences in power can be found in Table~\ref{tab:simres},
+which contains a subset of the results.  For example, the table shows that the
+power advantage of the Cram\`{e}r-von
+Mises test can be as large as .1 (at $n=100$, 3 parameters being
+tested).  It also shows that the Cram\`{e}r-von Mises test generally
+has true Type-I error rates, with the double-max test being slightly
+conservative and the max-LM test being slightly liberal at small
+sample sizes.
+
+In summary, we found that the proposed tests have adequate power to detect
+measurement invariance violations in applied contexts.  The specific
+statstics' are not equivalent to one another, and we found that the
+Cram\`{e}r-von Mises statistic exhibited the best performance for the
+data generated here.  In the discussion, we consider the extent to
+which these findings may extend to other factor-analytic situations.
+
+% Need to modify table and add figure
 \begin{table*}

### 2011-02-08 MIpaper/par_instable.tex
New figure of simulation results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@36 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 In summary, we found that the proposed tests have adequate power to detect
-measurement invariance violations in applied contexts.  The specific
-statstics' are not equivalent to one another, and we found that the
+measurement invariance violations in applied contexts.  The 
 Cram\`{e}r-von Mises statistic exhibited the best performance for the
-data generated here.  In the discussion, we consider the extent to
+data generated here, though this result may not hold for all
+situations.  In the discussion, we specifically consider the extent to 
 which these findings may extend to other factor-analytic situations.
 
-% Need to modify table and add figure
+\begin{figure}
+\caption{Simulated power for three test statistics across three sample
+  sizes, two subsets of tested parameters, and measurement invariance
+  violations of 0 to 5 standard 
+  errors.  Circles reflect the double-max test, triangles reflect the
+  max-LM test, and plus signs reflect the Cram\`{e}r-von Mises test.
+  Numbers above each panel reflect the sample size and number of
+  parameters being tested, respectively.}
+\label{fig:simres}
+\includegraphics[height=5in]{simres.pdf}
+\end{figure}
+
+% Need to modify table
 \begin{table*}

### 2011-02-09 MIpaper/par_instable.tex
Revised simulation/discussion sections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@37 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   consists only of parameters that are ``doubled'' under the
-  alternative.  This appears to not allow for situations where
-  parameters are restricted to be equal across subgroups.  I am ok
-  with this, but many applications of LR tests with these models have
+  alternative.  This appears to not allow for situations where some 
+  parameters are restricted to be equal across subgroups.  This should
+  work for the current presentation, 
+  but many applications of LR tests with these models have
   some parameters restricted to be equal across subgroups.  So I've
@@
 \subsection{Results}
-% MODIFY RESULTS TO DISCUSS ALL FOUR TESTS
-% ALSO NEED 0-SE CONDITION
 Full simulation results are presented in Figure~\ref{fig:simres}, and
@@
 
-% Need to modify table
 \begin{table*}
-\caption{Simulated power and Type I error rates for four test statistics
-  across three sample sizes and four magnitudes of measurement
-  invariance violations. Abbreviations: CvM = Cram\'{e}r-von Mises
-  test; MLM = Maximum Lagrange multiplier test; DM = Double-max test; 
-  $\chi^2$ = Chi-square likelihood ratio test.}
-\label{tab:sim1}
-\begin{tabular}{llcccc} \thickline
- & & \multicolumn{4}{c}{Violation Magnitude (SEs)} \\
-$n$ & Test Stat & 0 & 1 & 2 & 3 \\\hline
-100 & CvM       & 0.05 & 0.10 & 0.33 & 0.63 \\ 
-    & MLM       &      & 0.09 & 0.28 & 0.64 \\
-    & DM        & 0.03 & 0.06 & 0.19 & 0.41 \\ 
-    & $\chi^2$  & 0.07 & 0.16 & 0.49 & 0.83 \\ \hline
-200 & CvM       & 0.05 & 0.19 & 0.64 & 0.96 \\
-    & MLM       &      & 0.19 & 0.63 & 0.96 \\ 
-    & DM        & 0.03 & 0.12 & 0.44 & 0.88 \\ 
-    & $\chi^2$  & 0.06 & 0.27 & 0.76 & 0.99 \\ \hline
-500 & CvM       & 0.05 & 0.44 & 0.98 & 1.00 \\ 
-    & MLM       &      & 0.41 & 0.99 & 1.00 \\
-    & DM        & 0.04 & 0.35 & 0.94 & 1.00 \\ 
-    & $\chi^2$  & 0.05 & 0.58 & 1.00 & 1.00 \\\thickline
+\caption{Simulated power for three test statistics
+  across three sample sizes, eight magnitudes of measurement
+  invariance violations, and two subsets of tested
+  parameters. Abbreviations: CvM = Cram\'{e}r-von Mises 
+  test; Max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
+\label{tab:simres}
+\begin{center}
+\begin{tabular}{lllllllllll}
+  \hline
+   & & & \multicolumn{8}{l}{Violation Magnitude (SE)} \\
+  $n$ &  npars &  Statistic    &  0 & 0.5 &    1 &  1.5
+       &    2 &  2.5  &   3 &  3.5 \\ 
+  \hline
+  100  &  3    &  DM           &  3.1 &    3.6 &    6.6 &   12.4 &   20.6 &   32.1  &  46.1 &   57.8 \\ 
+         &         &  CvM            &  5.0 &    5.7 &   10.6 &   19.1 &   32.1 &   49.2  &  67.2 &   78.8 \\ 
+         &         &  max-LM          &  6.2 &    6.2 &    9.8 &   15.0 &   24.7 &   39.7  &  57.0 &   69.2 \\ 
+         &  19   &  DM           &  2.5 &    2.8 &    3.1 &    4.9 &    7.1 &   12.4  &  19.6 &   25.3 \\ 
+         &         &  CvM            &  4.2 &    4.8 &    5.8 &    8.6 &   13.0 &   18.9  &  28.2 &   37.4 \\ 
+         &         &  max-LM          &  6.7 &    6.9 &    7.7 &    9.6 &   13.3 &   17.3  &  23.6 &   30.6 \\ 
+  200  &  3    &  DM           &  3.7 &    5.2 &   12.5 &   26.3 &   47.0 &   70.6  &  86.8 &   95.0 \\ 
+         &         &  CvM            &  5.3 &    7.9 &   18.6 &   39.1 &   64.3 &   85.8  &  96.2 &   99.0 \\ 
+         &         &  max-LM          &  6.0 &    8.0 &   14.5 &   30.2 &   54.7 &   79.4  &  93.2 &   98.5 \\ 
+         &  19   &  DM           &  3.2 &    3.6 &    5.7 &   10.8 &   21.9 &   40.2  &  61.1 &   76.5 \\ 
+         &         &  CvM            &  4.6 &    5.8 &    8.5 &   15.0 &   29.1 &   46.4  &  65.4 &   80.7 \\ 
+         &         &  max-LM          &  5.6 &    6.4 &    8.3 &   13.2 &   22.3 &   38.2  &  57.5 &   74.8 \\ 
+  500  &  3    &  DM           &  4.1 &    8.3 &   32.2 &   69.0 &   93.4 &   99.5  & 100.0 &  100.0 \\ 
+         &         &  CvM            &  4.7 &   11.8 &   42.2 &   81.3 &   98.0 &  100.0  & 100.0 &  100.0 \\ 
+         &         &  max-LM          &  4.6 &    9.8 &   33.9 &   75.5 &   97.1 &  100.0  & 100.0 &  100.0 \\ 
+         &  19   &  DM           &  3.6 &    4.4 &   14.0 &   40.1 &   75.3 &   96.0  &  99.7 &  100.0 \\ 
+         &         &  CvM            &  5.0 &    6.7 &   17.9 &   43.6 &   77.4 &   95.9  &  99.7 &   99.9 \\ 
+         &         &  max-LM          &  4.9 &    7.0 &   14.2 &   36.0 &   71.0 &   94.9  &  99.6 &  100.0 \\ 
+   \hline
 \end{tabular}
+\end{center}
 \end{table*}
-
-% Table 2 of Nyblom89 contains critical values that may prove useful.
-% Hansen 1992 p. 524 has table of critical values that matches 
-%  Nyblom89 within reason.
-% See Zeileis05 p. 450-451
 
@@
 measurement invariance.  For example, they give researchers the
-opportunity to simultaneously test for many types of measurement
-invariance (see, e.g., \citeNP{Mer93}).  The traditional
+flexibility to: (1) simultaneously test all model parameters, yielding
+results relevant to many types of measurement
+invariance (see, e.g., \citeNP{Mer93}), or (2) test a single subset of
+model parameters, potentially leading to improved power to detect a
+single type of measurement invariance.  The traditional
 steps have involved hypothesizing a specific type of invariance and
-then testing for it via a Likelihood Ratio Test, but the tests
-proposed here do not require one to hypothesize a specific type of
-invariance a priori.  Instead, one could define a type of measurement
-invariance after seeing the proposed tests' results.
+then testing for it via a Likelihood Ratio Test, but this is
+unnecessary under the proposed framework.
 
@@
 the nature of the invariance violation.  This is made possible through
-the tests' abilities to estimate $\eta$, the threshold dividing
+the tests' abilities to estimate $\nu$, the threshold dividing
 individuals into subgroups that violate
-measurement invariance.  While a single $\eta$ was estimate in this
+measurement invariance.  While a single $\nu$ was assumed in this
 paper, it is also possible to define 
-formal rules for estimating multiple $\eta$ parameters.  We plan to
+formal rules for estimating multiple $\nu$ parameters.  We plan to
 explore these ideas in the future.
@@
 column of the $K \times p$ matrix, summing over the squared entries in
-the column to obtain a $\chi^2$ statistic with
+the column to obtain a $\chi^2$-distributed statistic with
 $(K-1)$ degrees of freedom \cite{HjoKon02}.
@@
 classification and regression trees \cite{BreFri84,MerSha10,StrMal09},
-and related algorithms have been developed for generalized linear
+with related algorithms being developed for generalized linear
 models \cite{ZeiHot08} and Rasch models \cite{KopZei10}.
@@
 Relatedly,
-S\'{a}nchez \citeyear{San09} describes a general
+S\'{a}nchez \citeyear{San09} described a general
 method for partitioning/segmenting structural equation models within a 
 partial least squares framework.  This method involves direct maximization of
-the likelihood ratio (i.e., fitting the model for various subgroups based on
+the likelihood ratio (i.e., fitting the model for various subgroups
+defined by 
 $V$ and choosing the subgroups with the largest likelihood ratio),

### 2011-02-09 www/sim.R
Slight changes to xyplot and ftable commands
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@38 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -135,12 +135,15 @@ ftable(xtabs(power ~ nobs + pars + test + diff, data = mz_sim), col.vars = "diff
 round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
   data = mz_sim, subset = diff %in% c(seq(0, 3.5, by = 0.5))),
   col.vars = "diff"), digits = 1)
+# To export to Latex, I wrote ftable contents to a text file, then read it
+# back in with read.fwf, then used xtable.  There must be an easier way?
 
 ## display result in graphic
 library("lattice")
 trellis.par.set(theme = canonical.theme(color = FALSE))
 ## all
-xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim, type = "b")
+xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim, type = "b",
+       xlab="Violation Magnitude", ylab="Power")
 ## subset
 xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim,
   subset = diff <= 3, type = "b")

### 2011-02-10 MIpaper/par_instable.tex
added computational details section, including additional references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@40 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\section*{Computational Details}
+
+All results were obtained using the {R}~system for statistical computing \cite{R2010},
+version~2.11.1, employing the add-on packages {OpenMx}~0.9.1-1421 \cite{BokNea11} for
+fitting of the factor analysis models and {strucchange}~1.4.2 \cite{ZeiLei02,Zei06}
+for evaluating the parameter instability tests. {R} and {strucchange} are
+freely available under the General Public License~2 from the
+Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/}
+while {OpenMx} is available under the
+Apache License~2.0 from \url{http://openmx.psyc.virginia.edu/}.
+R code for replication of our results is
+available at \url{http://semtools.R-Forge.R-project.org/}.
+
 \bibliography{list}

### 2011-02-10 MIpaper/par_instable.tex
revised the paper: mostly smaller modifications and some readme/fixme comments
revised the code, including the artificial example: especially the DGP can
be improved using a suitably scaled version of the violation magnitude



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@41 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\abstract{The issue of measurement invariance commonly arises in
-  factor-analytic contexts, with methods for assessment
-  including 
-  likelihood ratio tests, Lagrange multiplier tests, and
-  Wald tests.  These tests all require advance definition of the number of
-  groups, group 
-  membership, and offending model parameters.
-In this paper, we construct tests of parameter
-invariance based on stochastic processes of casewise
-derivatives of the likelihood function.  These tests can be viewed as
-generalizations of the Lagrange multiplier test, and they are especially
-useful for: (1) isolating specific parameters affected by measurement
-invariance violations, and (2) identifying subgroups of individuals
-that violated measurement invariance based on a continuous auxiliary
-variable.  The tests are presented and illustrated in detail, along
-with simulations examining the tests' abilities in controlled conditions.}
-
-\acknowledgements{Correspondence to Edgar C.\
-  Merkle, Department of 
+\abstract{
+  The issue of measurement invariance commonly arises in factor-analytic
+  contexts, with methods for assessment including likelihood ratio tests,
+  Lagrange multiplier tests, and Wald tests. These tests all require advance
+  definition of the number of groups, group membership, and offending model
+  parameters. In this paper, we construct tests of parameter invariance based on
+  stochastic processes of casewise derivatives of the likelihood function. These
+  tests can be viewed as generalizations of the Lagrange multiplier test, and they
+  are especially useful for: (1)~isolating specific parameters affected by
+  measurement invariance violations, and (2)~identifying subgroups of individuals
+  that violated measurement invariance based on a continuous auxiliary variable. 
+  The tests are presented and illustrated in detail, along with simulations
+  examining the tests' abilities in controlled conditions.
+}
+
+\acknowledgements{Correspondence to Edgar C.\ Merkle, Department of 
   Psychology, Wichita State University, Wichita, KS 67260-0034.
@@
 \newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}
+
+\spacing{1}
 
@@
 invariance equivocates conclusions and casts doubt on theory in the
-behavioral sciences'' (p.\ 141).  Borsboom \citeyear{Bor06a} also
+behavioral sciences'' (p.~141).  Borsboom \citeyear{Bor06a} also
 notes that researchers often fail to assess whether measurement
@@
 typically not known (exactly) in practice. If it were (see below for
-some concrete examples), then standard inference methods (such as likelihood
-ratio, Wald, or Lagrange multiplier tests) could be employed. However,
+some concrete examples), then standard inference methods -- such as likelihood
+ratio, Wald, or Lagrange multiplier tests -- could be employed. However,
 if the pattern is unknown, it is difficult to develop a single test
@@
 
-Note that even when adopting $H_1^*$ as the relevant alternative, the pattern ${\bm \theta}(V)$
-is not completely specified unless the cutpoint $\nu$ is known in
-advance. 
-In this situation, all individuals can be grouped based on
-$V$, and we can apply standard
-theory: nested multiple group
-models (e.g., \citeNP{Jor71,Bol89}) coupled with likelihood ratio (LR)
-tests are most common,
-though Lagrange multiplier (LM) and Wald tests may also be constructed for this
-purpose.\footnote{All three of these approaches are asymptotically
-  equivalent \cite{Sat89}.}  However, if $\nu$ is unknown (as is often
-the case for continuous $V$), then the standard theory does not apply
-and we can instead rely on the tests proposed in this paper.
+Note that even when adopting $H_1^*$ as the relevant alternative, the pattern
+${\bm \theta}(V)$ is not completely specified unless the cutpoint $\nu$ is known
+in advance.  In this situation, all individuals can be grouped based on $V$, and
+we can apply standard theory: nested multiple group models (e.g.,
+\citeNP{Jor71,Bol89}) coupled with likelihood ratio (LR) tests are most common,
+although the asymptotically equivalent (see \citeNP{Sat89}) Lagrange multiplier
+(LM) and Wald tests may also be constructed for this purpose. However, if $\nu$
+is unknown (as is often the case for continuous $V$), then the standard theory
+cannot be easily applied and instead nonstandard inference methods are required
+such as the tests proposed in this paper. \readme{Slightly rephrase but probably
+still not optimal.}
 
@@
 
-% MAY BE USEFUL SOMEWHERE
-% Focusing on the latter point, Millsap's \citeyear{Mil05} review of 
-% measurement invariance methods cites ``locating the violation of the
-% invariance'' as one of the major outstanding problems.  The tests
-% resulting from the proposed project directly address this issue,
-% leading to an important advance in measurement invariance research.
-
 \subsection{Likelihood Ratio, Wald, and Lagrange Multiplier Test for Fixed Subgroups}
@@
 are typically separated into a certain number of given subgroups where
-the parameters are estimated separately (for ease of
-  exposition, we assume no parameters are restricted to be equal
-  across subgroups). Then, the sum of maximized likelihoods 
+the parameters are estimated separately, potentially with some parameters
+restricted to be equal across subgroups. For ease of exposition we present
+the results for the case where there are no such restrictions (although it is
+straightforward to extend all methods to the more general case).
+Then, the sum of maximized likelihoods 
 from the subgroups are compared with the original maximized full-sample likelihood
@@
 of parameters in~${\bm \theta}$.
-\readme{Achim: You previously modified the above section so that $\theta$
-  consists only of parameters that are ``doubled'' under the
-  alternative.  This appears to not allow for situations where some 
-  parameters are restricted to be equal across subgroups.  This should
-  work for the current presentation, 
-  but many applications of LR tests with these models have
-  some parameters restricted to be equal across subgroups.  So I've
-  added some detail in parentheses at the beginning of the above
-  paragraph.}
 
@@
 showed that the asymptotic distribution is in fact tractable, but nonstandard.
-Specifically, the asymptotic distribution of the maximum is the maximum of a
+Specifically, the asymptotic distribution of (\ref{eq:maxlr}) is the maximum of a
 certain tied-down Bessel process whose specifics also depend on the minimal
@@
 So far, the discussion focused on the alternative $H_1^*$: The maximum LR, Wald, and LM
-tests are designed for a situation where there is a single threshold at which the full
-parameter vector ${\bm \theta}$ changes. While this is plausible and intuitive
+tests are designed for a situation where there is a single threshold at which all
+parameters in the vector ${\bm \theta}$ change. While this is plausible and intuitive
 in many applications, it would also be desirable to obtain tests that direct their power
 against other types of alternatives, i.e., against $H_1$ with other patterns ${\bm \theta}(V)$.
-For example, there might be multiple thresholds at which the parameters change or parameters
-fluctuate randomly. Alternatively, only one (or just a few) of the parameters in
+For example, the parameters may fluctuate randomly or there might be multiple thresholds
+at which the parameters change. Alternatively, only one (or just a few) of the parameters in
 the vector ${\bm \theta}$ change while the remaining parameters are
@@
 critical value.  An example of this visualization appears in the
-Example section.
+example section. \readme{Maybe give the explicit figure reference?}
 
@@
 at the same threshold or if the score process takes large values for
-many of the $n$~individuals (and just a single threshold). In such
+many of the $n$~individuals (and not just a single threshold). In such
 cases, sums insteads of maximums are more suitable for collapsing across
@@
 
+Finally, all tests can be easily modified to address the situation of
+so-called ``partial structural changes'' \cite{And93}. This refers to the case of some
+parameters being known to be stable, i.e., restricted to be constant across
+potential subgroups. Tests for potential changes/instabilities only in the
+remaining parameters are easily constructed by omitting those columns from
+${\bm B}(\hat {\bm \theta})_{ij}$ that are restricted/stable and only retaining those
+that are potentially instable. \readme{I guess we should say something like
+this in the theory section. Is this intelligible enough?}
+
 \subsection{Critical Values \& $p$-values.}
@@
 can be formally identified. However, graphing of (transformations of) the cumulative
-score process may still be valuable for gaining some insights.
+score process may still be valuable for gaining some insights.  \readme{Maybe give figure
+references for the example?}
 
@@
 possible combinations of thresholds (\citeNP{ZeiSha10} adapt a dynamic programming
-algorithm to this task). For the single threshold case, this reduced to
+algorithm to this task). For the single threshold case, this reduces to
 maximizing the segmented log-likelihood
@@
 \caption{Path diagram representing the base factor analysis model used for
-  the Example and Simulations.  To induce measurement invariance
+  the example and simulations.  To induce measurement invariance
   violations, a seventh observed variable (student age) determines the
@@
 respect to age, which is taken to be the auxiliary variable $V$.
+
+\fixme{I think it would be useful to say early on that a single static
+data set is drawn from the model first for illustration -- and subsequently
+a full simulation study using the same scenario is considered.}
 
@@
   ``older'' subjects).
+\fixme{This wasn't fully clear to me before and I probably should
+have asked earlier. I think we should employ the asymptotic standard
+errors divided by $\sqrt{n}$.}
 This violation states that the verbal ability scales
@@
 H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
-(\lambda_{11}\ \lambda_{21}\ \lambda_{31}),\ i=1,\ldots,n,
+(\lambda_{0, 11}\ \lambda_{0, 21}\ \lambda_{0, 31}),\ i=1,\ldots,n,
 \end{equation}
@@
 
+\fixme{I think the paragraph above should say something like ``If we
+knew or suspected which coefficients are subject to change\dots'' or
+something else along those lines. Just to make clear that we assume
+additional knowledge here. Also it may help to refer back to the
+paragraph that I added in the theory section about partial changes.}
+
 \subsection{Results}
@@
 isolation of model parameters violating
-measurement invariance.  R code for obtaining these results is
-  available at \url{http://semtools.r-forge.r-project.org}.
+measurement invariance.
 
@@
 
-The figures display many test attributes, all of which generally
-conform with the underlying theory.  First, when we test
-only the three verbal factor loadings, all three tests can detect the
-change.  Conversely, when all nineteen model parameters are tested,
-only the double-max test detects a change.  This implies that 
-power increases as larger proportions of tested parameters
-change.  The results also imply that the double-max test is more
+The figures convey information about several properties
+of the tests. First, all three tests are more powerful (and in
+this example significant) if only those parameters are tested
+that are subject to instabilities. Conversely, if all 19~model
+parameters are assessed (including those that are in fact invariant),
+the power is decreased. This decrease, however, is less pronounced
+for the double-max test as it is more 
 sensitive to fluctuations among a small subset of parameters.
@@
 Figure~\ref{fig:lrlm} 
-compares the max-LM statistic (black line) to the max-LR
-statistic (blue line) from \eqref{eq:maxlr}, as applied to testing
+compares the max-LM statistic (solid line) to the max-LR
+statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
 \eqref{eq:h0}.  The critical values for these two
@@
 is generally to be expected, because the two tests are asymptotically
-equivalent.  For this specific situation, however, the max-LR
-statistic rejects \eqref{eq:h0} while the max-LM statistic does not.
-More generally, the 
+equivalent.  More generally, the 
 max-LR statistic cannot be obtained from the empirical fluctuation
-process, so it must be computed ``by hand'' (i.e., by computing
-the LR statistic over all values of the threshold $\nu$, re-fitting the
-factor analysis model for each value).
+process, so the factor analysis model must be refitted before and after
+each of the possible threshold values $\nu$.
+\readme{With the modified DGP code, both max-LR and max-LM, are
+nonsignificant in the example.}
 
@@
 verbal scales lack measurement invariance.  
-We can also use the tests to: (1) estimate the 
-threshold $\nu$, and (2) isolate specific parameters that violate
+We can also use the tests to: (1)~estimate the 
+threshold $\nu$, and (2)~isolate specific parameters that violate
 measurement invariance.  
@@
 with the true threshold of 16.0.
+
+\fixme{I'm not really happy with the following paragraph. I think
+I wouldn't call this a posthoc test. It is really a test of the same
+hypothesis as before, but now using the double-max
+statistic. Would it be possible to simply show this and say that
+it is an advantage of the double-max test that both the threshold
+and the parameter affected by the invariance can be ``seen'' in the
+visualization. Then, we can still go on and talk about the power issue:
+The double-max test potentially wastes a lot of power, namely
+in the case where many of the parameters change at the same time. Note
+that in the case where all parameters change, we wouldn't have to ask
+the question ``Which parameter did change?''. Thus, luckily, the double-max test
+is more powerful when we need it. But, unfortunately, we don't know that
+in advance.}
 
@@
 \begin{figure}
-\caption{A comparison of the max-LM (black line) and max-LR (blue line) test
-  statistics for \eqref{eq:h0}.  The solid, horizontal line
-  corresponds to the critical value at $\alpha=.05$.}
+\caption{A comparison of the max-LM (solid line) and max-LR (dashed line) test
+  statistics for \eqref{eq:h0}.  The solid horizontal line
+  corresponds to the critical value at $\alpha=.05$ while the dotted vertical
+  line highlights the threshold at which the test statistics are maximized.}
 \label{fig:lrlm}
@@
 \section{Simulation}
+
+\readme{I didn't look at this section in detail because I think we
+should rerun the simulations first with the modified DGP.}
+
 % Simulations demonstrating power of test(s) at various
@@
 Data were generated in a similar way to that of 
-the Example section.
+the example section.
 The data included verbal scales that violated measurement invariance
@@
 
-\readme{Achim: Is it true that you used 5000 reps per cell?}
 For each combination of sample size $\times$
@@
 Circles reflect the double-max test, triangles reflect the max-LM
-test, and plus signs reflect the Cram\`{e}r-von Mises test.  One can
+test, and plus signs reflect the Cram{\'e}r-von Mises test.  One can
 generally observe that simultaneous tests of all 19 parameters
@@
 another when only the three factor loadings are tested, with the
-Cram\`{e}r-von Mises test having the most power, followed by max-LM,
+Cram\'{e}r-von Mises test having the most power, followed by max-LM,
 followed by the double-max test.  This advantage decreases with
@@
 which contains a subset of the results.  For example, the table shows that the
-power advantage of the Cram\`{e}r-von
+power advantage of the Cram\'{e}r-von
 Mises test can be as large as .1 (at $n=100$, 3 parameters being
-tested).  It also shows that the Cram\`{e}r-von Mises test generally
+tested).  It also shows that the Cram\'{e}r-von Mises test generally
 has true Type-I error rates, with the double-max test being slightly
@@
 measurement invariance violations in applied contexts.  The 
-Cram\`{e}r-von Mises statistic exhibited the best performance for the
+Cram\'{e}r-von Mises statistic exhibited the best performance for the
 data generated here, though this result may not hold for all
@@
   errors.  Circles reflect the double-max test, triangles reflect the
-  max-LM test, and plus signs reflect the Cram\`{e}r-von Mises test.
+  max-LM test, and plus signs reflect the Cram\'{e}r-von Mises test.
   Numbers above each panel reflect the sample size and number of
@@
 groups of individuals violating measurement invariance based on a
-continuous, auxiliary variable, and can
+continuous auxiliary variable, and can
 isolate specific model parameters affected by the violation.  In this
@@
 measurement invariance.  For example, they give researchers the
-flexibility to: (1) simultaneously test all model parameters, yielding
+flexibility to: (1)~simultaneously test all model parameters, yielding
 results relevant to many types of measurement
-invariance (see, e.g., \citeNP{Mer93}), or (2) test a single subset of
+invariance (see, e.g., \citeNP{Mer93}), or (2)~test a single subset of
 model parameters, potentially leading to improved power to detect a
@@
 steps have involved hypothesizing a specific type of invariance and
-then testing for it via a Likelihood Ratio Test, but this is
+then testing for it via a LR test, but this is
 unnecessary under the proposed framework.
@@
 % a single threshold, the maximum should be achieved close to the true
-% $\eta$.  Referring to the Example (see
+% $\eta$.  Referring to the example (see
 % Figure~\ref{fig:cusumex}), we defined $\eta$ as the 
@@
 groups are 
-already specified in advance, and so traditional methods (i.e.,
-Likelihood Ratio tests) may suffice.
-However, we could use the framework developed here to obtain an
-alternative $\chi^2$ statistic testing measurement invariance.
-Assume the observations are divided into $K$ categories $I_1, I_2,
-\ldots, I_K$.
-First, we can sum scores within each category of the
-demographic variable via:
+already specified in advance, and so traditional methods for fixed
+subgroups (i.e., LR, Wald, and LM tests) may suffice.
+Furthermore, we can also obtain an LM-type statistic from the
+framework developed here. Assume the observations are divided into
+$C$ categories $I_1, I_2, \ldots, I_C$. Then, the increment of
+the cumulative score process $\Delta_{I_c} {\bm B}(\hat {\bm \theta})$
+within in each category is just the sum of the corresponding scores.
+In somewhat sloppy notation:
 \begin{equation}
     \label{eq:catsum}
-    \Delta W_n(I_k) = \frac{1}{n_k\sqrt{n}} \sum_{i \in I_k}
-    {\bm{I}}(\widehat{\bm{\theta}})^{-1/2} {\bm{\psi}}({\bm{y}}_i, \widehat{{\bm
-      \theta}}),
-\end{equation}
-where $n_k$ is the sample size in category $k$ and $n$ is the overall
-sample size.  This results in a $K \times p$ matrix, with one entry for
+    \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
+    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i)
+%% \\
+%%    X^2 & = & \sum_{c = 1}^C n_c^{-1} n^{-1} || \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ||^2
+\end{equation}
+This results in a $C \times k$ matrix, with one entry for
 each category-by-model parameter combination.  We can test
 a specific model parameter for invariance by focusing on the associated
-column of the $K \times p$ matrix, summing over the squared entries in
-the column to obtain a $\chi^2$-distributed statistic with
-$(K-1)$ degrees of freedom \cite{HjoKon02}.
+column of the $C \times k$ matrix and employing a weighted squared sum of
+the entries in the column to obtain a $\chi^2$-distributed statistic with
+$(C-1)$ degrees of freedom \cite{HjoKon02}.
 Alternatively, to simultaneously test multiple parameters, we can sum
@@
 parameters.
+
+\readme{Should we also say something about ordinal variables and
+variables with a lot of ties? These may also be better treated as categorical
+variables if the number of potential thresholds is very low -- because
+the asymptotics go in the number of thresholds.}
 
@@
 Relatedly,
-S\'{a}nchez \citeyear{San09} described a general
+S\'{a}nchez \citeyear{San09} describes a general
 method for partitioning/segmenting structural equation models within a 
@@
 defined by 
-$V$ and choosing the subgroups with the largest likelihood ratio),
-however, instead of the tests described in this paper.  The tests
-described here are highly advantageous because they only require the
-model to be estimated once.
+$V$ and choosing the subgroups with the largest likelihood ratio).
+Thus, unlike the tests described in this paper, this approach does not
+provide a formal significance test with a controlled level of Type~I errors.
+
+\readme{For PATHMOX, I would just use the argument that it is not a formal
+test. Whether or not it is computationally more expensive would only be
+a second argument, but I would probably avoid to use it.}
 
@@
 the tests may be extended to study differential item functioning (DIF)
-in item response models (e.g., \citeNP{KopZei10}, who focused on an
-application to Rasch models).  
-DIF methods are similar to those for factor analysis in
-that subgroups must be specified in advance.  The tests proposed here
-can detect subgroups automatically, however, offering a unified way of
-studying both 
-measurement invariance in factor analysis and differential item
-functioning in item response.  While these
-two methods have developed largely independently of one another,
+in item response models (e.g., \citeNP{KopZei10}, who focused on recursive
+partitioning of Rasch models).  
+Traditional DIF methods are similar to those for factor analysis in
+that subgroups must be specified in advance.
+%% The tests proposed here can detect subgroups automatically, however, offering a unified way of
+%% studying both  measurement invariance in factor analysis and differential item
+%% functioning in item response.
+While these two methods have developed largely independently of one another,
 some treatments \cite{Mcd99} and recent research \cite{StaChe06} have
@@
 support this endeavor.
+\readme{I have omitted (= commented) one sentence in the last paragraph.
+I feel the paragraph may be somewhat more intelligible like this.}
 
@@
 
+\readme{Instead of just giving a single URL, I prefer to give some more details
+about the computational methods used. Especially in this case this makes sense,
+because an OpenMx paper is forthcoming in Psychometrika.}
+
 All results were obtained using the {R}~system for statistical computing \cite{R2010},
@@
 freely available under the General Public License~2 from the
-Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/}
+Comprehensive {R} Archive Network at \url{http://CRAN.R-project.org/}
 while {OpenMx} is available under the

### 2011-02-10 www/script2.R

@@ -1,3 +1,7 @@
+###################
+## Preliminaries ##
+###################
+
 ## packages and code
 library("OpenMx")
 library("strucchange")
@@ -5,9 +9,13 @@ library("mvtnorm")
 source("mz.R")
 source("sim.R")
 
-## generate artificial data
+##################################
+## Example with Artificial Data ##
+##################################
+
+## generate single artificial data set
 set.seed(1090)
-d <- dgp(200, 2)
+d <- dgp(200, 3)
 
 ## full sample and "true" subsample models 
 mz0 <- mzfit(d)
@@ -15,111 +23,81 @@ mz1 <- mzfit(d[1:100,])
 mz2 <- mzfit(d[101:200,])
 
 ## check: Which model parameters differ pre- and post-break?
-(coef(mz1) - coef(mz2)) / sqrt(diag(vcov(mz1)) + diag(vcov(mz2)))
+round((coef(mz1) - coef(mz2)) / sqrt(diag(vcov(mz1)) + diag(vcov(mz2))), digits = 2)
 
-## empirical fluctuation processes (3 vs 19 pars, info vs OPG vcov)
-gefp_3_opg <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3)
-gefp_19_opg <- gefp(mz0, fit = NULL, order.by = d$age)
-gefp_3_info <- gefp(mz0, fit = NULL, vcov = info.MxModel,
-  order.by = d$age, sandwich = FALSE, parm = 1:3)
-gefp_19_info <- gefp(mz0, fit = NULL, vcov = info.MxModel,
-  order.by = d$age, sandwich = FALSE)
+## empirical fluctuation processes (3 vs all 19 pars)
+gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
+  vcov = info.MxModel, sandwich = FALSE)
+gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
+  vcov = info.MxModel, sandwich = FALSE)
 
-## Double-maximum test
-par(mfrow = c(2, 2))
-plot(gefp_3_info,  functional = maxBB, main = "dmax: 3 pars, Info")
-plot(gefp_19_info, functional = maxBB, main = "dmax: 19 pars, Info")
-plot(gefp_3_opg,   functional = maxBB, main = "dmax: 3 pars, OPG")
-plot(gefp_19_opg,  functional = maxBB, main = "dmax: 19 pars, OPG")
+## Figure 2
[…]

### 2011-02-10 www/sim.R

@@ -27,7 +27,8 @@ dgp <- function(nobs = 200, diff = 3)
 
   # Loadings for "old" individuals:
   lambda.o <- matrix(0,6,2)
-  lambda.o[,1] <- lambda.y[,1] - (ses * c(.87,.6,1.32,rep(0,3)))
+  #old# lambda.o[,1] <- lambda.y[,1] - (ses * c(.87,.6,1.32,rep(0,3)))
+  lambda.o[,1] <- lambda.y[,1] - (ses * c(8.66, 5.52, 9.19, rep(0, 3))/sqrt(nobs))
   lambda.o[,2] <- c(0,0,0,3.24,4.32,7.21)
   
   phi <- matrix(.48,2,2)
@@ -102,7 +103,8 @@ simulation <- function(diff = seq(0, 5, by = 0.25),
 
   rval <- data.frame()
   for(i in 1:ntest) rval <- rbind(rval, prs)
-  rval$test <- factor(rep(c("dmax", "CvM", "supLM", "dmax", "CvM", "supLM"), each = nprs), levels = c("dmax", "CvM", "supLM"))
+  rval$test <- factor(rep(c("dmax", "CvM", "supLM", "dmax", "CvM", "supLM"), each = nprs),
+    levels = c("dmax", "CvM", "supLM"))
   rval$pars <- factor(rep(c("3", "3", "3", "19", "19", "19"), each = nprs), levels = c("3", "19"))
   rval$nobs <- factor(rval$nobs)
   rval$power <- as.vector(pow)
@@ -117,14 +119,15 @@ library("mvtnorm")
 source("mz.R")
 
 ## seed for replication
-set.seed(6020)
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1090)
 
 ## run simulation
 mz_sim <- simulation()
 save(mz_sim, file = "mz_sim.rda")
 }
 
-if(TRUE) {
+if(FALSE) {
 ## load simulation results
 load("mz_sim.rda")
 
@@ -135,8 +138,6 @@ ftable(xtabs(power ~ nobs + pars + test + diff, data = mz_sim), col.vars = "diff
 round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
   data = mz_sim, subset = diff %in% c(seq(0, 3.5, by = 0.5))),
   col.vars = "diff"), digits = 1)
-# To export to Latex, I wrote ftable contents to a text file, then read it
[…]

### 2011-02-10 www/sim.R
reduced diff to 3.5 max in simulation, should be sufficient due to new scaling
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@42 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -86,7 +86,7 @@ testpower <- function(nrep = 5000, size = 0.05, verbose = TRUE, ...)
 }
 
 ## Loop over scenarios
-simulation <- function(diff = seq(0, 5, by = 0.25),
+simulation <- function(diff = seq(0, 3.5, by = 0.25),
   nobs = c(100, 200, 500), verbose = TRUE, ...)
 {
   prs <- expand.grid(diff = diff, nobs = nobs)

### 2011-02-10 www/script2.R
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@43 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -15,7 +15,7 @@ source("sim.R")
 
 ## generate single artificial data set
 set.seed(1090)
-d <- dgp(200, 3)
+d <- dgp(200, 2)
 
 ## full sample and "true" subsample models 
 mz0 <- mzfit(d)

### 2011-02-10 www/sim.R

@@ -28,7 +28,7 @@ dgp <- function(nobs = 200, diff = 3)
   # Loadings for "old" individuals:
   lambda.o <- matrix(0,6,2)
   #old# lambda.o[,1] <- lambda.y[,1] - (ses * c(.87,.6,1.32,rep(0,3)))
-  lambda.o[,1] <- lambda.y[,1] - (ses * c(8.66, 5.52, 9.19, rep(0, 3))/sqrt(nobs))
+  lambda.o[,1] <- lambda.y[,1] - (ses * c(8.66, 5.52, 9.19, rep(0, 3))/sqrt(nobs/2))
   lambda.o[,2] <- c(0,0,0,3.24,4.32,7.21)
   
   phi <- matrix(.48,2,2)

### 2011-02-11 MIpaper/par_instable.tex
Revisions for first half of paper, modified figures
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@44 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 violating measurement invariance into structural equation
-models, where relationships between constructs are hypothesized.  Horn
+models, where relationships between latent variables are hypothesized.  Horn
 and McArdle \citeyear{HorMca92} concisely summarize the impact of
@@
 invariance equivocates conclusions and casts doubt on theory in the
-behavioral sciences'' (p.~141).  Borsboom \citeyear{Bor06a} also
+behavioral sciences'' (p.~141).  Borsboom \citeyear{Bor06a} further 
 notes that researchers often fail to assess whether measurement
@@
 In this paper, we consider a new family of tests for assessing
-measurement invariance that have important advantages over existing
+measurement invariance that has important advantages over existing
 tests.  We begin by developing a general framework for the tests.
@@
 whenever standard assumptions for maximum likelihood inference hold (for more
-details see below). For empirical applications, we employ
-a specific factor analysis model where the data follow a multivariate normal
-distribution:
+details see below). For the measurement invariance applications
+considered in this paper, we employ
+a factor analysis model with assumed multivariate normality:
 \begin{eqnarray}
@@
       (\bm{x}_i - \bm{\mu}(\bm{\theta}))^{\top} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i - \bm{\mu}(\bm{\theta}))
-      ~+~ \log | \bm{\Sigma}(\bm{\theta}) | ~+~ p \log(2 \pi) \right\}.
+      ~+~ \log | \bm{\Sigma}(\bm{\theta}) | ~+~ p \log(2 \pi) \right\},
 \end{eqnarray}
@@
 the assumptions for the tests introduced below do not require this specific
-form of the likelihood but it is presented for illustration due to
+form of the likelihood, but it is presented for illustration due to
 its importance in practice.  Many expositions of factor analysis
 utilize the likelihood for the sample covariance matrix, which is
-based on a Wishart distribution when $\bm{x}_i$ are assumed to be
+based on a Wishart distribution when the $\bm{x}_i$ are assumed to be
 multivariate normal.
@@
 likelihood contributions w.r.t.\ the parameters $\bm{\theta}$.
-
-% \fixme{I have deferred individual-specific parameters for ease of exposition.
-% My feeling is that it should be briefly mentioned at the end of this
-% section and then connected with measurement invariance in the next section.}
+Evaluation of the score function at $\hat{\bm{\theta}}$ for $i=1,
+\dots, n$ essentially measures the extent to which each individual's
+likelihood is maximized.
+
+\readme{I added a sentence above about score functions because I don't think
+  they are well known in psychometrics.  You might check to see whether my
+  interpretation is reasonable.}
 
@@
 \citeNP{Jor71,Bol89}) coupled with likelihood ratio (LR) tests are most common,
-although the asymptotically equivalent (see \citeNP{Sat89}) Lagrange multiplier
-(LM) and Wald tests may also be constructed for this purpose. However, if $\nu$
-is unknown (as is often the case for continuous $V$), then the standard theory
-cannot be easily applied and instead nonstandard inference methods are required
-such as the tests proposed in this paper. \readme{Slightly rephrase but probably
-still not optimal.}
+although the asymptotically equivalent  Lagrange multiplier
+(LM) and Wald tests may also be constructed for this purpose (see
+\citeNP{Sat89}). If $\nu$
+is unknown (as is often the case for continuous $V$), however, then
+standard theory 
+is not easily applied.   Nonstandard inference methods, such as those
+proposed in this paper, are then required.
 
@@
 
-To employ the LR test for assessing measurement invariance, the data
-are typically separated into a certain number of given subgroups where
-the parameters are estimated separately, potentially with some parameters
-restricted to be equal across subgroups. For ease of exposition we present
-the results for the case where there are no such restrictions (although it is
-straightforward to extend all methods to the more general case).
+To employ the LR test for assessing measurement invariance, model
+parameters are estimated separately for a certain number of subgroups
+of the data (with some parameters potentially restricted to be equal across
+subgroups).\footnote{For ease of exposition we present
+the results for the case where there are no such restrictions, although it is
+straightforward to extend all methods to the more general case.}
 Then, the sum of maximized likelihoods 
@@
 LM test, the differences of likelihood scores from zero. In the LR case, the parameters
-have to be estimated both under null hypothesis and alternative while in the Wald case
-only the estimates under the alternative are required and in the LM case only under the
-null hypothesis.
+have to be estimated under both the null hypothesis and alternative.
+Conversely, the Wald case requires  
+only the estimates under the alternative, while the LM case requires
+only the estimates under the null hypothesis.
 
@@
 as $\max \mathit{LR}(\nu)$ and are asymptotically equivalent \cite{And93}.
-From a computational perspective, the latter test is particularly convenient
+From a computational perspective, the $\max \mathit{LM}(\nu)$ test is
+particularly convenient 
 because it requires just a single set of estimated parameters $\hat {\bm \theta}$
-which is employed for all thresholds $\nu$ in $[\underline{\nu}, \overline{\nu}]$
-while the other two techniques require reestimation of the subgroup model for each
+which is employed for all thresholds $\nu$ in $[\underline{\nu},
+\overline{\nu}]$.  
+The other two tests require reestimation of the subgroup model for each
 $\nu$.
@@
 of the LM test -- is ${\bm s}(\hat {\bm \theta}; x_i)_j$: the $j$-th component of the
-contribution of the $i$-th observation to the score function. By construction the sum
-of the score contributions over all individuals is zero for each component, see
-(\ref{eq:ml1}). Moreover, if there are no systematic deviations the score contributions
-should fluctuate randomly around zero while they should be shifted away from zero for
+contribution of the $i$-th observation to the score function. By construction, the sum
+of the score contributions over all individuals is zero for each
+component, see
+(\ref{eq:ml1}). Moreover, if there are no systematic deviations, the score contributions
+should fluctuate randomly around zero.  Conversely, the score
+contributions should be shifted away from zero for
 subgroups where the model does not fit.
@@
 accomplishes the challenges discussed at the beginning of this section:
-it makes use of the ordering of the observations by taking cumulative sums
+it makes use of the ordering of the observations by taking cumulative sums,
 and it decorrelates the contributions of the $k$ different parameters.
@@
 As argued above, no single aggregation function will have high power for any
-conceivable pattern of measurement invariance ${\bm \theta}(V)$ while any
+conceivable pattern of measurement invariance ${\bm \theta}(V)$, while any
 (reasonable) aggregation function will have non-trivial power under $H_1$.
 Thus, various aggregation strategies should be employed depending on which
-pattern ${\bm \theta}(V)$ is most plausible (because the exact is typically
+pattern ${\bm \theta}(V)$ is most plausible (because the exact pattern
+is typically
 unknown). A particularly agnostic aggregation strategy is to reject $H_0$
 if any component of the the cumulative score process ${\bm B}(t; \hat {\bm \theta})$
-departs ``too far'' from its zero at any time, i.e., if
+departs ``too far'' from zero at any time, i.e., if
 \begin{equation}
@@
 critical value.  An example of this visualization appears in the
-example section. \readme{Maybe give the explicit figure reference?}
+example section (Figure~\ref{fig:lambdas}).
 
 However, taking maximums ``wastes'' power if many of the $k$~parameters change
-at the same threshold or if the score process takes large values for
+at the same threshold, or if the score process takes large values for
 many of the $n$~individuals (and not just a single threshold). In such
@@
 remaining parameters are easily constructed by omitting those columns from
-${\bm B}(\hat {\bm \theta})_{ij}$ that are restricted/stable and only retaining those
-that are potentially instable. \readme{I guess we should say something like
-this in the theory section. Is this intelligible enough?}
+${\bm B}(\hat {\bm \theta})_{ij}$ that are restricted/stable,
+retaining only those
+that are potentially instable. This may be of special interest to
+those wishing to test specific types of measurement invariance, where
+subsets of model parameters are assumed to be stable.
 
@@
 
-As pointed out above, specifying the asymptotic distribution und $H_0$ for the test
+As pointed out above, specification of the asymptotic distribution under $H_0$ for the test
 statistics from the previous section is straightforward: It is simply the aggregation
@@
 (see e.g., \citeNP{ShoWel86,PloKra92}) with a straightforward Bonferroni
-correction to obtain the $k$-dimensional case. As the terms in the summation
-quickly go to zero as $h$ goes to infinity, only some large finite
-number (say, 100 terms) need to be evaluated in practice.
+correction to obtain the $k$-dimensional case. The terms in the summation
+quickly go to zero as $h$ goes to infinity, so only some large finite
+number of terms (say, 100) need to be evaluated in practice.
 
@@
 have been extended in the software provided by Zeileis \citeyear{Zei06}.
-
 Critical values for the distribution of the maximum LR/Wald/LM tests
@@
 Again, a boundary crossing signals a significant violation, and the peak(s) in the
-sequence of statistics conveys threshold(s). Note that due to summing over all
+sequence of statistics conveys threshold(s). Note that, due to summing over all
 parameters, no specific parameter can be identified that is responsible for the
-violation. Similarly, for the Cram\'{e}r-von Mises test neither component(s) nor threshold(s)
-can be formally identified. However, graphing of (transformations of) the cumulative
-score process may still be valuable for gaining some insights.  \readme{Maybe give figure
-references for the example?}
+violation. Similarly,  neither component(s) nor threshold(s)
+can be formally identified for the Cram\'{e}r-von Mises test. However, graphing of (transformations of) the cumulative
+score process may still be valuable for gaining some insights (see,
+e.g., Figure~\ref{fig:gefp_3_19}).
 
 If a measurement invariance violation is detected by any of the tests,
-one may want to incorporate it into the model to account for it. How this should
-be done typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
-discussed above often prove helpful in determining a suitable parametrization.
-In particular, one approach that is often employed in practice is to adopt a model
+one may want to incorporate it into the model to account for it. The
+procedure for doing this  typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
+discussed above often prove helpful in determining a suitable parameterization.
+In particular, one approach that is often employed in practice
+involves adoptions of a model
 with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
@@
 $v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR statistic
-from (\ref{eq:lr}) and some minimal subgroup size is typically imposed.
+from (\ref{eq:lr}) (with some minimal subgroup size typically imposed).
 
@@
 in the field.
+
+In the next section, we present an example involving a single data set
+drawn from a 
+known factor analysis model.  Following the example, we consider a
+full simulation study involving the same factor analysis model.
 
@@
 
-As an example of the ideas discussed above, consider a
+Consider a 
 hypothetical battery of six scales administered to students aged 13 to 18
@@
 three of the scales are intended to measure mathematical ability.  We
-may observe a maturation effect in the resulting data, whereby the scales
-measure older students better than younger students (i.e., younger
-students have not yet learned all the material).  The researcher's goal is to
+may observe a maturation effect in the resulting data, whereby the
+factor loadings for older students are larger than those for younger
+students.  The researcher's goal is to
 study whether the scales are measurement invariant with
 respect to age, which is taken to be the auxiliary variable $V$.
-
-\fixme{I think it would be useful to say early on that a single static
-data set is drawn from the model first for illustration -- and subsequently
-a full simulation study using the same scenario is considered.}
 
@@
 are three standard errors lower than those in Figure~\ref{fig:famod}
-(the standard errors were obtained by fitting
-  the model to a sample of size 200, where the sample consisted of all
-  ``older'' subjects).
-\fixme{This wasn't fully clear to me before and I probably should
-have asked earlier. I think we should employ the asymptotic standard
-errors divided by $\sqrt{n}$.}
+(the standard errors are based on the asymptotic standard errors
+divided by $\sqrt{n}$).
 This violation states that the verbal ability scales
@@
 
-Note that the tests give us the flexibility to test various subsets of
-parameters.  For example, were we only interested in the 
-three verbal factor loading parameters, we can test
+Note that, as mentioned in the theory section, the 
+tests give us the flexibility to study hypotheses of partial change.
+That is, we have the ability to 
+test various subsets of 
+parameters.  For example, if we suspected that the verbal factor loadings
+lacked measurement invariance, we could test
 \begin{equation}
@@
 
-\fixme{I think the paragraph above should say something like ``If we
-knew or suspected which coefficients are subject to change\dots'' or
-something else along those lines. Just to make clear that we assume
-additional knowledge here. Also it may help to refer back to the
-paragraph that I added in the theory section about partial changes.}
+% stopped here Feb 10 2011
 
@@
 In the results section, we first describe overall results.  We then 
-describe estimation of $\nu$, and 
+describe estimation of $\nu$ and 
 isolation of model parameters violating
@@
 along with a procedure that controls for Type I error.  
-Figure~\ref{fig:posthoc} shows the individual cumulative score processes for the
+Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
 verbal factor loadings, with the horizontal lines reflecting the
@@
   correspond to the critical value at $\alpha=.05$.}
-\label{fig:posthoc}
-\includegraphics[height=5in]{gefp_posthoc.pdf}
+\label{fig:lambdas}
+\includegraphics[height=5in]{gefp_lambdas.pdf}
 \end{figure}

### 2011-02-11 MIpaper/par_instable.tex
Revised manuscript (except simulation section)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@45 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 of the data (with some parameters potentially restricted to be equal across
-subgroups).\footnote{For ease of exposition we present
-the results for the case where there are no such restrictions, although it is
+subgroups).\footnote{For ease of exposition, we describe
+the case where there are no such parameter restrictions.  As shown in
+the example and simulation below, it is
 straightforward to extend all methods to the more general case.}
@@
 score process may still be valuable for gaining some insights (see,
-e.g., Figure~\ref{fig:gefp_3_19}).
+e.g., Figure~\ref{fig:cusumex}).
 
@@
 
-% stopped here Feb 10 2011
-
 \subsection{Results}
@@
 of the tests. First, all three tests are more powerful (and in
-this example significant) if only those parameters are tested
+this example significant) if we test only those parameters 
 that are subject to instabilities. Conversely, if all 19~model
@@
 is generally to be expected, because the two tests are asymptotically
-equivalent.  More generally, the 
+equivalent.  The 
 max-LR statistic cannot be obtained from the empirical fluctuation
-process, so the factor analysis model must be refitted before and after
-each of the possible threshold values $\nu$.
-\readme{With the modified DGP code, both max-LR and max-LM, are
-nonsignificant in the example.}
+process, however, so the factor analysis model must be refitted before
+and after each of the possible threshold values $\nu$.
 
@@
 
-\fixme{I'm not really happy with the following paragraph. I think
+\readme{Achim previous comment: I'm not really happy with the following
+  paragraph. I think 
 I wouldn't call this a posthoc test. It is really a test of the same
 hypothesis as before, but now using the double-max
-statistic. Would it be possible to simply show this and say that
-it is an advantage of the double-max test that both the threshold
-and the parameter affected by the invariance can be ``seen'' in the
-visualization. Then, we can still go on and talk about the power issue:
-The double-max test potentially wastes a lot of power, namely
-in the case where many of the parameters change at the same time. Note
-that in the case where all parameters change, we wouldn't have to ask
-the question ``Which parameter did change?''. Thus, luckily, the double-max test
-is more powerful when we need it. But, unfortunately, we don't know that
-in advance.}
-
-The double-max test yields specific information about 
-parameters violating measurement invariance.  Given
-rejection of the hypothesis of measurement invariance \eqref{eq:exh0},
-we can do posthoc 
-tests to study specific parameters violating measurement invariance.
-This involves individual examination of each parameter's cumulative
-score process,
-along with a procedure that controls for Type I error.  
+statistic. \\ \ \\
+Ed response: I am fine with not calling it a posthoc test.  However,
+it could sometimes be advantageous to first test the hypothesis using the CvM
+statistic, then, if CvM rejects the hypothesis, employ the double-max
+statistic to say something about individual model parameters.  This
+would be useless if all parameters change, but, for measurement
+invariance problems, one often expects many model parameters to be stable.}
+
+As mentioned previously, the double-max test is advantageous because
+it yields information about individual 
+parameters violating measurement invariance.  
+That is, it allows us to examine whether or not individual parameters' 
+cumulative score processes lead one to reject the hypothesis of
+measurement invariance.
 Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
@@
 and second loadings did not achieve the critical value represents a Type II
-error, which implicitly brings into question the posthoc tests' power.  We
+error, which implicitly brings into question the posthoc tests' power.
+This is an example of a situation where the double-max test ``wastes''
+power, as it cannot make use of the fact that multiple model
+parameters change simultaneously.  We
 generally address the issue of power in the simulations below.
@@
 \begin{figure}
-\caption{Posthoc tests of individual verbal factor loadings, based on
-  the absolute maximum of each cumulative score process.  The top
-  panel depicts the process for $\lambda_{11}$, the middle panel
-  depicts the process for $\lambda_{21}$, and the bottom panel depicts
-  the process for $\lambda_{31}$.  The solid, horizontal lines
+\caption{Cumulative score processes for each verbal factor
+  loadings, with critical values stemming from the double-max test. 
+  The solid, horizontal lines
   correspond to the critical value at $\alpha=.05$.}
@@
 based on stochastic processes, have reasonable power, can isolate
-groups of individuals violating measurement invariance based on a
+subgroups of individuals violating measurement invariance based on a
 continuous auxiliary variable, and can
@@
 formal rules for estimating multiple $\nu$ parameters.  We plan to
-explore these ideas in the future.
+explore these issues in the future.
 
@@
 the $\chi^2$ statistics and degrees of freedom for the individual
-parameters.
-
-\readme{Should we also say something about ordinal variables and
-variables with a lot of ties? These may also be better treated as categorical
-variables if the number of potential thresholds is very low -- because
-the asymptotics go in the number of thresholds.}
+parameters.  In addition to categorical $V$, this framework may also
+be useful for continuous $V$ with many ties and ordinal $V$.  
+The number of potential thresholds may be very low in these
+situations, which impacts the extent to which asymptotic results hold
+for the main test statistics described in this paper.
 
@@
 provide a formal significance test with a controlled level of Type~I errors.
-
-\readme{For PATHMOX, I would just use the argument that it is not a formal
-test. Whether or not it is computationally more expensive would only be
-a second argument, but I would probably avoid to use it.}
 
@@
 %\subsubsection{Other Latent Variable Models}
-Aside from recursive measurement invariance detection, the proposed tests
+The proposed tests also 
 readily extend to other popular psychometric models.  
@@
 example, it would be possible to assess whether relationships between
-latent variables are stronger for some students than for
+latent variables are stronger for some individuals than for
 others.  Secondly, 
@@
 %% functioning in item response.
-While these two methods have developed largely independently of one another,
+While factor-analytic measurement invariance methods and DIF methods
+have developed 
+largely independently of one another, 
 some treatments \cite{Mcd99} and recent research \cite{StaChe06} have
@@
 support this endeavor.
-\readme{I have omitted (= commented) one sentence in the last paragraph.
-I feel the paragraph may be somewhat more intelligible like this.}
 
@@
 
-\readme{Instead of just giving a single URL, I prefer to give some more details
-about the computational methods used. Especially in this case this makes sense,
-because an OpenMx paper is forthcoming in Psychometrika.}
-
 All results were obtained using the {R}~system for statistical computing \cite{R2010},

### 2011-02-11 www/script2.R
Modified web page content
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@46 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -60,7 +60,7 @@ abline(v = mz.age[which.max(mz.lrstat)], lty = 3)
 
 
 ## Figure 4
- # pdf("gefp_posthoc.pdf", height = 7, width = 6)
+ # pdf("gefp_lambdas.pdf", height = 7, width = 6)
 plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
   main = "", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
  # dev.off()

### 2011-02-14 www/script2.R
prettified (somewhat) trellis panel labels
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@48 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -99,5 +99,7 @@ round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
 ## display result in graphic
 library("lattice")
 trellis.par.set(theme = canonical.theme(color = FALSE))
+levels(mz_sim$pars) <- paste("k =", levels(mz_sim$pars))
+levels(mz_sim$nobs) <- paste("n =", levels(mz_sim$nobs))
 xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim, type = "b",
-       xlab = "Violation Magnitude", ylab = "Power")
+       xlab = "Violation Magnitude", ylab = "Power", ylim = c(0, 1))

### 2011-02-17 MIpaper/par_instable.tex
Revised simulation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@49 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-Note that, as mentioned in the theory section, the 
+As mentioned in the theory section, the 
 tests give us the flexibility to study hypotheses of partial change.
@@
 
-\readme{I didn't look at this section in detail because I think we
-should rerun the simulations first with the modified DGP.}
-
 % Simulations demonstrating power of test(s) at various
@@
 power, and the simulations provide more detail on the extent to which
-power increases.  Finally, we also examine the tests' power across
+the tests' power levels differ.
+Finally, we also examine the tests' power across
 various magnitudes of measurement invariance violations.
@@
 invariance violations across three sample sizes ($n=100,
-200, 500$) and three magnitudes of violations (factor loadings
-for the younger students being $(0, 0.25, 0.5, \dots, 5)$ standard
-errors\footnote{The standard errors used here were
-  obtained by fitting
-  the model to a sample of size 100, where the sample consisted of all
-  ``older'' subjects.} below 
-those for the older students).  The 0-standard error condition was
+200, 500$) and fifteen magnitudes of violations.  These violations
+involved the younger students' values of $\{\lambda_{11},
+\lambda_{21}, \lambda_{31}\}$ deviating from the older students'
+values by $d$ 
+times the parameters' asymptotic standard errors, with $d = 0, 0.25,
+0.5, \dots, 3.5$.  The 0-standard error condition was
 used to study Type I error rate.
 
-For each combination of sample size $\times$
-violation magnitude $\times$ number of parameters being tested,
+For each combination of sample size ($n$) $\times$
+violation magnitude ($d$) $\times$ number of parameters being tested
+($k^{\ast}$),
 5,000 datasets were generated and tested.  In each dataset, half the
@@
 Figure~\ref{fig:simres} displays power as a function of violation
-magnitude, with panels for each combination of sample size $\times$
-number of parameters being tested.  
+magnitude, with panels for each combination of sample size ($n$) $\times$
+number of parameters being tested ($k^{\ast}$).  
 Circles reflect the double-max test, triangles reflect the max-LM
@@
 generally observe that simultaneous tests of all 19 parameters
-result in decreased power, with the tests performing very similarly at
+result in decreased power, with the tests performing more similarly at
 the larger sample sizes.  The tests distinguish themselves from one
@@
 increases in sample size.  
-Absolute differences in power can be found in Table~\ref{tab:simres},
-which contains a subset of the results.  For example, the table shows that the
+Table~\ref{tab:simres} presents the same results as
+Figure~\ref{fig:simres}, but it is easier to see exact power
+magnitudes in the table.
+The table shows that the
 power advantage of the Cram\'{e}r-von
-Mises test can be as large as .1 (at $n=100$, 3 parameters being
-tested).  It also shows that the Cram\'{e}r-von Mises test generally
+Mises test can be as large as .1, most notably when three parameters
+are being tested.
+  It also shows that the Cram\'{e}r-von Mises test generally
 has true Type-I error rates, with the double-max test being slightly
-conservative and the max-LM test being slightly liberal at small
-sample sizes.
+conservative and the max-LM test being slightly liberal.
 
@@
 Cram\'{e}r-von Mises statistic exhibited the best performance for the
-data generated here, though this result may not hold for all
-situations.  In the discussion, we specifically consider the extent to 
-which these findings may extend to other factor-analytic situations.
+data generated here, though more simulations are warranted to examine
+the generality of this finding.
+In the discussion, we describe extensions of the tests in factor
+analysis and beyond.
 
@@
   sizes, two subsets of tested parameters, and measurement invariance
-  violations of 0 to 5 standard 
-  errors.  Circles reflect the double-max test, triangles reflect the
+  violations of 0--3.5 standard 
+  errors (scaled by $n$).  Circles reflect the double-max test,
+  triangles reflect the 
   max-LM test, and plus signs reflect the Cram\'{e}r-von Mises test.
@@
 \label{fig:simres}
-\includegraphics[height=5in]{simres.pdf}
+\includegraphics[height=9in]{simres.pdf}
 \end{figure}
@@
 \caption{Simulated power for three test statistics
-  across three sample sizes, eight magnitudes of measurement
+  across three sample sizes, fifteen magnitudes of measurement
   invariance violations, and two subsets of tested
@@
   \hline
-  100  &  3    &  DM           &  3.1 &    3.6 &    6.6 &   12.4 &   20.6 &   32.1  &  46.1 &   57.8 \\ 
-         &         &  CvM            &  5.0 &    5.7 &   10.6 &   19.1 &   32.1 &   49.2  &  67.2 &   78.8 \\ 
-         &         &  max-LM          &  6.2 &    6.2 &    9.8 &   15.0 &   24.7 &   39.7  &  57.0 &   69.2 \\ 
-         &  19   &  DM           &  2.5 &    2.8 &    3.1 &    4.9 &    7.1 &   12.4  &  19.6 &   25.3 \\ 
-         &         &  CvM            &  4.2 &    4.8 &    5.8 &    8.6 &   13.0 &   18.9  &  28.2 &   37.4 \\ 
-         &         &  max-LM          &  6.7 &    6.9 &    7.7 &    9.6 &   13.3 &   17.3  &  23.6 &   30.6 \\ 
-  200  &  3    &  DM           &  3.7 &    5.2 &   12.5 &   26.3 &   47.0 &   70.6  &  86.8 &   95.0 \\ 
-         &         &  CvM            &  5.3 &    7.9 &   18.6 &   39.1 &   64.3 &   85.8  &  96.2 &   99.0 \\ 
-         &         &  max-LM          &  6.0 &    8.0 &   14.5 &   30.2 &   54.7 &   79.4  &  93.2 &   98.5 \\ 
-         &  19   &  DM           &  3.2 &    3.6 &    5.7 &   10.8 &   21.9 &   40.2  &  61.1 &   76.5 \\ 
-         &         &  CvM            &  4.6 &    5.8 &    8.5 &   15.0 &   29.1 &   46.4  &  65.4 &   80.7 \\ 
-         &         &  max-LM          &  5.6 &    6.4 &    8.3 &   13.2 &   22.3 &   38.2  &  57.5 &   74.8 \\ 
-  500  &  3    &  DM           &  4.1 &    8.3 &   32.2 &   69.0 &   93.4 &   99.5  & 100.0 &  100.0 \\ 
-         &         &  CvM            &  4.7 &   11.8 &   42.2 &   81.3 &   98.0 &  100.0  & 100.0 &  100.0 \\ 
-         &         &  max-LM          &  4.6 &    9.8 &   33.9 &   75.5 &   97.1 &  100.0  & 100.0 &  100.0 \\ 
-         &  19   &  DM           &  3.6 &    4.4 &   14.0 &   40.1 &   75.3 &   96.0  &  99.7 &  100.0 \\ 
-         &         &  CvM            &  5.0 &    6.7 &   17.9 &   43.6 &   77.4 &   95.9  &  99.7 &   99.9 \\ 
-         &         &  max-LM          &  4.9 &    7.0 &   14.2 &   36.0 &   71.0 &   94.9  &  99.6 &  100.0 \\ 
+100 & 3 & DM & 3.1 & 4.1 & 8.5 & 16.1 & 27.2 & 43.2 & 57.1 & 67.4 \\ 
+   &  & CvM & 4.9 & 6.6 & 13.1 & 26.4 & 45.0 & 65.9 & 80.4 & 89.2 \\ 
+   &  & Max-LM & 4.9 & 6.8 & 11.4 & 20.6 & 34.8 & 54.8 & 71.4 & 82.9 \\ 
+   & 19 & DM & 2.4 & 3.0 & 4.3 & 5.9 & 9.5 & 15.9 & 23.8 & 31.5 \\ 
+   &  & CvM & 4.6 & 4.4 & 7.0 & 11.1 & 17.1 & 27.1 & 38.1 & 49.2 \\ 
+   &  & Max-LM & 7.0 & 7.1 & 9.0 & 11.2 & 16.2 & 21.4 & 31.0 & 39.7 \\ 
+  200 & 3 & DM & 3.7 & 5.0 & 9.0 & 17.8 & 32.0 & 49.3 & 67.1 & 84.0 \\ 
+   &  & CvM & 4.9 & 6.5 & 13.9 & 26.2 & 48.2 & 68.1 & 85.5 & 95.1 \\ 
+   &  & Max-LM & 5.5 & 6.1 & 11.1 & 20.9 & 39.1 & 58.6 & 78.5 & 92.0 \\ 
+   & 19 & DM & 3.8 & 3.4 & 4.6 & 7.1 & 13.1 & 22.0 & 34.7 & 52.5 \\ 
+   &  & CvM & 4.6 & 4.9 & 7.3 & 11.8 & 20.1 & 31.7 & 45.8 & 64.0 \\ 
+   &  & Max-LM & 6.0 & 6.1 & 8.2 & 11.3 & 16.3 & 24.4 & 37.6 & 54.3 \\ 
+  500 & 3 & DM & 3.9 & 5.6 & 9.7 & 19.5 & 34.3 & 52.3 & 71.9 & 87.0 \\ 
+   &  & CvM & 4.8 & 7.3 & 12.9 & 26.9 & 47.3 & 69.1 & 86.7 & 94.9 \\ 
+   &  & Max-LM & 5.5 & 6.7 & 11.4 & 21.4 & 38.5 & 60.1 & 81.4 & 93.2 \\ 
+   & 19 & DM & 3.7 & 4.4 & 5.3 & 8.7 & 14.6 & 25.6 & 41.6 & 60.9 \\ 
+   &  & CvM & 4.8 & 5.3 & 7.1 & 11.9 & 20.6 & 32.3 & 50.3 & 68.8 \\ 
+   &  & Max-LM & 5.6 & 6.2 & 7.5 & 10.4 & 15.9 & 25.8 & 41.6 & 60.8 \\ 
    \hline

### 2011-02-17 www/script2.R
Modified xyplot command
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@50 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -99,7 +99,11 @@ round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
 ## display result in graphic
 library("lattice")
 trellis.par.set(theme = canonical.theme(color = FALSE))
-levels(mz_sim$pars) <- paste("k =", levels(mz_sim$pars))
-levels(mz_sim$nobs) <- paste("n =", levels(mz_sim$nobs))
+
 xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim, type = "b",
-       xlab = "Violation Magnitude", ylab = "Power", ylim = c(0, 1))
+       xlab = "Violation Magnitude", ylab = "Power", ylim = c(0, 1),
+       strip=function(which.given, ..., factor.levels){
+         levs <- if (which.given == 1) c(expression(k^symbol("\052") == 3),
+                                         expression(k^symbol("\052") == 19))
+                 else paste("n =", levels(mz_sim$nobs))
+         strip.default(which.given, ..., factor.levels=levs)})

### 2011-03-23 www/script2.R
fully adapted all scripts to new OpenMx/lavaan infrastructure
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@54 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -89,22 +89,18 @@ set.seed(1090)
 ## run simulation
 mz_sim <- simulation()
 
-## display result in table
-## all without rounding
-ftable(xtabs(power ~ nobs + pars + test + diff, data = mz_sim), col.vars = "diff")
-## subset with rounding, in percent
+## display subset of results in table
+## (with rounding, in percent)
+mz_sim_sub <- subset(mz_sim, pars %in% c(3, 19))
+mz_sim_sub$pars <- factor(mz_sim_sub$pars)
+levels(mz_sim_sub$pars) <- paste("k* =", levels(mz_sim_sub$pars))
+levels(mz_sim_sub$nobs) <- paste("n =", levels(mz_sim_sub$nobs))
 round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
-  data = mz_sim, subset = diff %in% c(seq(0, 3.5, by = 0.5))),
+  data = mz_sim_sub, subset = diff %in% c(seq(0, 3.5, by = 0.5))),
   col.vars = "diff"), digits = 1)
 
 ## display result in graphic
 library("lattice")
 trellis.par.set(theme = canonical.theme(color = FALSE))
-
-xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim, type = "b",
-       xlab = "Violation Magnitude", ylab = "Power", ylim = c(0, 1),
-       strip=function(which.given, ..., factor.levels){
-         levs <- if (which.given == 1) c(expression(k^symbol("\052") == 3),
-                                         expression(k^symbol("\052") == 19))
-                 else paste("n =", levels(mz_sim$nobs))
-         strip.default(which.given, ..., factor.levels=levs)})
+xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub, type = "b",
+  xlab = "Violation Magnitude", ylab = "Power", ylim = c(0, 1))

### 2011-03-23 www/sim.R

@@ -117,7 +117,7 @@ simulation <- function(diff = seq(0, 4, by = 0.25),
   return(rval)
 }
 
-if(TRUE) {
+if(FALSE) {
 ## code and packages
 library("lavaan")
 library("strucchange")

### 2011-03-23 www/def.model.R
Cleanup of 'www' directory:   - old .R and data files have been removed from current revision   - script2.R has been renamed to replication.R   - index.php has been updated accordingly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@55 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,149 +1,0 @@
-def.model <- function (fadat, age)
-{
-    dat1 <- fadat[age < 16, ]
-    dat2 <- fadat[age > 16, ]
-    TwoFA <- mxModel("Two Factor Model",
-                     mxModel("group1", 
-                             type = "RAM",
-                             mxData(observed = dat1, type = "raw"),
-                             manifestVars = c("x1", "x2", "x3",
-                               "y1", "y2", "y3"), 
-                             latentVars = c("F1", "F2"),
-                             mxPath(from = c("x1", "x2", "x3",
-                                             "y1", "y2", "y3"),
-                                    arrows = 2,
-                                    free = TRUE, 
-                                    values = c(1, 1, 1, 1, 1, 1),
-                                    labels = c("e1", "e2", "e3",
-                                               "e4", "e5", "e6"),
-                                    lbound=0),
-                             mxPath(from = c("F1", "F2"),
-                                    arrows = 2,
-                                    all = TRUE,
-                                    free = c(FALSE, TRUE, TRUE, FALSE),
-                                    values = c(1, 0.5, 0.5, 1),
-                                    labels = c("varF1", "cov", "cov", "varF2"),
-                                    lbound=c(NA,0,0,NA)),
-                             mxPath(from = "F1",
-                                    to = c("x1", "x2", "x3"),
-                                    arrows = 1,
-                                    free = TRUE,
-                                    values = c(1, 1, 1),
-                                    labels = c("l1", "l2", "l3"),
-                                    lbound=0),
-                             mxPath(from = "F2", 
-                                    to = c("y1", "y2", "y3"),
-                                    arrows = 1,
-                                    free = TRUE, 
-                                    values = c(1, 1, 1),
-                                    labels = c("l4", "l5", "l6"),
-                                    lbound=0), 
-                             mxPath(from = "one",
-                                    to = c("x1", "x2", "x3", "y1", "y2", "y3",
-                                           "F1", "F2"),
-                                    arrows = 1,
-                                    free = c(TRUE, TRUE, TRUE, TRUE,
-                                             TRUE, TRUE, FALSE, FALSE),
[…]

### 2011-03-23 www/fa_score_extraction.R

@@ -1,89 +1,0 @@
-"score.fa" <- function(res.list,derivs="a")
-  {
-    # Code to extract scores from a 2-factor model that
-    # is fit by OpenMx.  res is a list containing
-    # the model fit by mxRun() and the observed data matrix.
-
-    # Possible options for derivs:
-    #   a: Analytic derivatives (fast)
-    #   n: Numerical derivatives (slow)
-
-    derivs <- tolower(derivs)
-    if (derivs=="n"){
-      require("numDeriv")
-    }
-
-    res <- res.list$res
-    dat <- res.list$dat
-
-    # Results from OpenMx
-    theta <- res@output$estimate
-
-    # Get n and k from OpenMx output
-    n <- nrow(dat)
-    k <- length(theta)
-
-    # Score matrix
-    scores <- matrix(NA,n,k)
-    
-    switch(derivs, `n` = {
-    
-      # Define casewise likelihood function
-      likefun <- function(x,ind.dat){
-        mu <- x[14:19]
-        lam <- matrix(c(x[1:3],rep(0,6),x[4:6]),6,2) # loadings
-        phi <- matrix(c(1,x[13],x[13],1),2,2) # factor covs
-        psi <- diag(x[7:12]) # error vars
-
-        impl.cov <- lam %*% phi %*% t(lam) + psi
-
-        # Added (-1/2) term:
-        fval <- (-1/2)*(log(det(impl.cov)) +
-                        t(ind.dat - mu) %*% solve(impl.cov) %*% (ind.dat - mu))
-        fval
-      }
-
-
-      for (j in 1:n){
-        scores[j,] <- grad(likefun, theta,
-                           ind.dat=as.matrix(dat)[j,],
-                           method="Richardson")
-      }
-    }, `a` = {
-      mu <- theta[14:19]
-      Lambda <- matrix(c(theta[1:3],rep(0,6),theta[4:6]),6,2) # loadings
-      Phi <- matrix(c(1,theta[13],theta[13],1),2,2) # factor covs
-      Psi <- matrix(0,6,6)
-      diag(Psi) <- theta[7:12] # error vars
-
-      Sigma <- Lambda %*% Phi %*% t(Lambda) + Psi
-  
-      Sig.inv <- solve(Sigma)
-      for (i in 1:n){
-        Xmu <- (dat[i,]-mu)
[…]

### 2011-03-23 www/gen.data.R

@@ -1,55 +1,0 @@
-"gen.data" <- function(sampsize,ses)
-  {
-    # Generates data from a factor analysis model that violates
-    # measurement invariance.  Also generates a continuous auxiliary
-    # variable related to the invariance (loosely called "age").
-    
-    # ses is number of SEs between group 1's parameters and group 2's
-    # parameters.
-    
-    require("mvtnorm")
-
-    # Could eventually manipulate this so break point is not directly
-    # in the middle
-    num.y <- round(sampsize/2,0)
-    num.o <- sampsize - num.y
-
-    age <- c(runif(num.y,13,16),runif(num.o,16,18))
-    
-    # Define parameter vectors/matrices:
-    mu <- c(29.32,24.70,14.84,10.59,19.30,18.01)
-
-    # Loadings for "young" individuals:
-    lambda.y <- matrix(0,6,2)
-    lambda.y[,1] <- c(4.92,2.96,5.96,0,0,0)
-    lambda.y[,2] <- c(0,0,0,3.24,4.32,7.21)
-
-    # Loadings for "old" individuals:
-    lambda.o <- matrix(0,6,2)
-    lambda.o[,1] <- lambda.y[,1] - (ses * c(.87,.6,1.32,rep(0,3)))
-    lambda.o[,2] <- c(0,0,0,3.24,4.32,7.21)
-    
-    phi <- matrix(.48,2,2)
-    diag(phi) <- 1
-
-    psi <- matrix(0,6,6)
-    diag(psi) <- c(26.77,13.01,30.93,3.17,8.82,22.5)
-
-    # Initialize data matrix:
-    datmat <- matrix(0,sampsize,6)
-    colnames(datmat) <- c("x1", "x2", "x3", "y1", "y2", "y3")
-
-    # Generate z and u vectors:
-    z <- t(rmvnorm(sampsize,rep(0,2),phi))
-    u <- t(rmvnorm(sampsize,rep(0,6),psi))
-
-    # Based on z and u, generate data:
-    for (i in 1:num.y){
-      datmat[i,] <- mu + lambda.y%*%z[,i] + u[,i]
-    }
-    for (i in (num.y+1):sampsize){
-      datmat[i,] <- mu + lambda.o%*%z[,i] + u[,i]
-    }
-    
-    list(datmat=datmat, age=age)
-  }

### 2011-03-23 www/script.R

@@ -1,91 +1,0 @@
-# Main R script file to carry out tests of measurement invariance
-# described in Merkle & Zeileis.
-
-# Load requisite packages
-library("OpenMx")
-library("strucchange")
-library("numDeriv")
-
-# Read data
-obs.data <- dget("data.dat")
-dat <- obs.data$datmat
-V <- obs.data$age
-
-# Obtain functions for score and information matrix extraction
-source("fa_score_extraction.R")
-
-# Obtain OpenMx model specification
-# (Two-group model with all parameters constrained to be equal,
-#  though this could also be estimated as a single-group model.)
-source("def.model.R")
-fa.mod <- def.model(dat, V)
-
-# Fit model via OpenMx
-mx.res <- mxRun(fa.mod)
-
-## the same with a single model (not split/constrained by age)
-fa.mod1 <- def.model1(dat)
-mx.res1 <- mxRun(fa.mod1)
-
-# Organize results for strucchange:
-fa.res <- list(res = mx.res1, dat = dat, V = V)
-
-## store scores in list as well to avoid slow re-computation
-fa.res$estfun <- score.fa(fa.res)
-colnames(fa.res$estfun) <- summary(mx.res1)$parameters$name
-
-# Calculate cumulative scores for first three factor loadings:
-fa.cums1 <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
-  vcov = inf.fa, order.by = fa.res$V, sandwich = FALSE, parm = 1:3)
-
-## the same for all parameters
-fa.cums2 <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
-  vcov = inf.fa, order.by = fa.res$V, sandwich = FALSE)
-
-## the same with OPG (outer product of gradients) vcov estimator
-fa.cums1opg <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
-  order.by = fa.res$V, parm = 1:3)
-fa.cums2opg <- gefp(fa.res, fit=NULL, scores = function(x) x$estfun,
-  order.by = fa.res$V)
-
-
-# Plot individual cumulative score processes:
-plot(fa.cums1, aggregate = FALSE, ylim = c(-1.8, 1.8), xlab = "Age")
-
-## Cramer-von Mises test
-## -> all lead to significant results but with all 19 parameters
[…]

### 2011-03-23 www/paper.Rnw
Paper moved to "www" directory and turned into .Rnw file.
All results in the manuscript can now be generated dynamically using Sweave().
Hence, no static PDF graphics (except the model diagram) are required
and these will be generated on the fly.

Also renamed .bib and added lavaan reference.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@56 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2011-02-17 MIpaper/par_instable.tex
@@
 \usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
+%% need no \usepackage{Sweave}
+\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}
+
 \title{Studying measurement invariance in factor analysis models
@@
 \maketitle
+
+<<preliminaries>>=
+library("OpenMx")
+library("lavaan")
+library("strucchange")
+library("mvtnorm")
+library("lattice")
+source("mz.R")
+source("sim.R")
+@
 
@@
 
-To employ the LR test for assessing measurement invariance, model
+To employ the LR~test for assessing measurement invariance, model
 parameters are estimated separately for a certain number of subgroups
@@
 $\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
-observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
+observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR~test statistic
 for the given threshold $\nu$ is then
@@
 
-Analogously to the LR test, the Wald test and LM test (also known as score test)
+Analogously to the LR~test, the Wald test and LM test (also known as score test)
 can be employed to test the null hypothesis $H_1^*$ for a fixed threshold $\nu$.
@@
 and ${\bm s}(\hat {\bm \theta}; x_{m+1}, \dots, x_n)$. Thus, the three tests all assess
-differences that should be zero under $H_0$: for the LR test the difference of maximized
+differences that should be zero under $H_0$: for the LR~test the difference of maximized
 likelihoods; for the Wald test, the difference of parameter estimates; and for the
-LM test, the differences of likelihood scores from zero. In the LR case, the parameters
+LM test, the differences of likelihood scores from zero. In the LR~case, the parameters
 have to be estimated under both the null hypothesis and alternative.
@@
 denotes the Euclidean norm. Finally, $\max \mathit{LM}$ from (\ref{eq:maxlm}) -- and analogously
-the maximum Wald and LR tests -- converges to $\sup_t (t (1-t))^{-1} || {\bm B}^0(t) ||_2^2$
+the maximum Wald and LR~tests -- converges to $\sup_t (t (1-t))^{-1} || {\bm B}^0(t) ||_2^2$
 (which can also be interpreted as the maximum of a tied-down Bessel process, as pointed out
@@
 over all values of $m$ corresponding to possible thresholds $\nu$ (such that $v_m \le \nu$ and
-$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR statistic
+$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR~statistic
 from (\ref{eq:lr}) (with some minimal subgroup size typically imposed).
@@
 \includegraphics[height=5in]{famod.pdf}
-%\input{figs/path}
 \end{figure}
@@
 
+<<dgp>>=
+set.seed(1090)
+d <- dgp(200, 2)
+@
+
 A sample of size 200 was generated from the model described above, and
@@
 measurement invariance.
+
+<<mz_mod>>=
+if(file.exists("mz_mod.rda")) {
+  load("mz_mod.rda")
+} else {
+  ## full sample model
+  mz0 <- mzfit(d)
+  ## subsample models
+  lrstat <- function(age) 2 * as.vector(logLik(mzfit(d[d$age <= age,])) + logLik(mzfit(d[d$age > age,])) - logLik(mz0))
+  mz.age <- sort(d$age[d$age >= quantile(d$age, 0.1) & d$age <= quantile(d$age, 0.9)])
+  mz.lrstat <- sapply(mz.age, lrstat)
+  save(mz0, mz.age, mz.lrstat, file = "mz_mod.rda")
+}
+@
 
@@
 equivalent.  The 
-max-LR statistic cannot be obtained from the empirical fluctuation
+max-LR~statistic cannot be obtained from the empirical fluctuation
 process, however, so the factor analysis model must be refitted before
@@
   to student age.  Solid, horizontal lines represent critical values
-  at $\alpha=.05$, and the dotted, horizontal lines (second row)
+  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
   represent values of the Cram\'{e}r-von Mises test statistic.}
 \label{fig:cusumex}
-\includegraphics[height=9in]{gefp_3_19.pdf}
+\setkeys{Gin}{width=\textwidth}
+<<gefp-3-19, fig=TRUE, height=8, width=6>>=
+## empirical fluctuation processes (3 vs all 19 pars)
+gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
+  vcov = info.mzfit, sandwich = FALSE)
+gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
+  vcov = info.mzfit, sandwich = FALSE)
+
+## visualization
+par(mfcol = c(3, 2))
+plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = supLM(0.1), main = "max-LM, k* = 3", xlab = "Age")
+plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = supLM(0.1), main = "max-LM, k* = 19", xlab = "Age")
+@
 \end{figure}
@@
 \label{fig:lrlm}
-\includegraphics[height=5in]{lmlr_19.pdf}
+\setkeys{Gin}{width=\textwidth}
+<<lm-lr-19, fig=TRUE, height=5, width=6>>=
+plot(gefp_19_info, functional = supLM(0.1),
+  xlab = "Age", ylab = "LR and LM statistics", main = "", ylim = c(0, 50))
+legend("topleft", c("LR", "LM"), lty = c(2, 1), bty = "n")
+lines(mz.age, mz.lrstat, lty = 2)
+abline(v = mz.age[which.max(mz.lrstat)], lty = 3)
+@
 \end{figure}
@@
 \label{fig:lambdas}
-\includegraphics[height=5in]{gefp_lambdas.pdf}
+\setkeys{Gin}{width=\textwidth}
+<<gefp-lambdas, fig=TRUE, height=8, width=6>>=
+plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
+  main = "", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
+@
 \end{figure}
@@
 
-% Simulations demonstrating power of test(s) at various
-% sample sizes, types of invariance violations, ?
 In this section, we conduct a simulation designed to
@@
 times the parameters' asymptotic standard errors, with $d = 0, 0.25,
-0.5, \dots, 3.5$.  The 0-standard error condition was
+0.5, \dots, 4$.  The 0-standard error condition was
 used to study Type I error rate.
@@
 years of age and half were 16--18 years of age.
+
+<<mz_sim>>=
+if(file.exists("mz_sim.rda")) {
+  load("mz_sim.rda")
+} else {
+  ## seed for replication
+  RNGkind(kind = "default", normal.kind = "default")
+  set.seed(1090)
+  ## run simulation
+  mz_sim <- simulation()
+  save(mz_sim, file = "mz_sim.rda")
+}
+
+## select only subset
+mz_sim_sub <- subset(mz_sim, pars %in% c(3, 19))
+mz_sim_sub$pars <- factor(mz_sim_sub$pars)
+## labeling
+levels(mz_sim_sub$pars) <- paste("k* =", levels(mz_sim_sub$pars))
+levels(mz_sim_sub$nobs) <- paste("n =", levels(mz_sim_sub$nobs))
+@
 
@@
 \label{fig:simres}
-\includegraphics[height=9in]{simres.pdf}
+\setkeys{Gin}{width=\textwidth}
+<<mzsim-xyplot, fig=TRUE, height=7, width=7>>=
+## visualization
+trellis.par.set(theme = canonical.theme(color = FALSE))
+print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub, type = "b",
+  xlab = "Violation Magnitude", ylab = "Power", ylim = c(0, 1)))
+@
 \end{figure}
@@
   parameters. Abbreviations: CvM = Cram\'{e}r-von Mises 
-  test; Max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
+  test; max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
 \label{tab:simres}
 \begin{center}
-\begin{tabular}{lllllllllll}
+\begin{tabular}{lrlrrrrrrrrr}
   \hline
-   & & & \multicolumn{8}{l}{Violation Magnitude (SE)} \\
-  $n$ &  npars &  Statistic    &  0 & 0.5 &    1 &  1.5
-       &    2 &  2.5  &   3 &  3.5 \\ 
-  \hline
-100 & 3 & DM & 3.1 & 4.1 & 8.5 & 16.1 & 27.2 & 43.2 & 57.1 & 67.4 \\ 
-   &  & CvM & 4.9 & 6.6 & 13.1 & 26.4 & 45.0 & 65.9 & 80.4 & 89.2 \\ 
-   &  & Max-LM & 4.9 & 6.8 & 11.4 & 20.6 & 34.8 & 54.8 & 71.4 & 82.9 \\ 
-   & 19 & DM & 2.4 & 3.0 & 4.3 & 5.9 & 9.5 & 15.9 & 23.8 & 31.5 \\ 
-   &  & CvM & 4.6 & 4.4 & 7.0 & 11.1 & 17.1 & 27.1 & 38.1 & 49.2 \\ 
-   &  & Max-LM & 7.0 & 7.1 & 9.0 & 11.2 & 16.2 & 21.4 & 31.0 & 39.7 \\ 
-  200 & 3 & DM & 3.7 & 5.0 & 9.0 & 17.8 & 32.0 & 49.3 & 67.1 & 84.0 \\ 
-   &  & CvM & 4.9 & 6.5 & 13.9 & 26.2 & 48.2 & 68.1 & 85.5 & 95.1 \\ 
-   &  & Max-LM & 5.5 & 6.1 & 11.1 & 20.9 & 39.1 & 58.6 & 78.5 & 92.0 \\ 
-   & 19 & DM & 3.8 & 3.4 & 4.6 & 7.1 & 13.1 & 22.0 & 34.7 & 52.5 \\ 
-   &  & CvM & 4.6 & 4.9 & 7.3 & 11.8 & 20.1 & 31.7 & 45.8 & 64.0 \\ 
-   &  & Max-LM & 6.0 & 6.1 & 8.2 & 11.3 & 16.3 & 24.4 & 37.6 & 54.3 \\ 
-  500 & 3 & DM & 3.9 & 5.6 & 9.7 & 19.5 & 34.3 & 52.3 & 71.9 & 87.0 \\ 
-   &  & CvM & 4.8 & 7.3 & 12.9 & 26.9 & 47.3 & 69.1 & 86.7 & 94.9 \\ 
-   &  & Max-LM & 5.5 & 6.7 & 11.4 & 21.4 & 38.5 & 60.1 & 81.4 & 93.2 \\ 
-   & 19 & DM & 3.7 & 4.4 & 5.3 & 8.7 & 14.6 & 25.6 & 41.6 & 60.9 \\ 
-   &  & CvM & 4.8 & 5.3 & 7.1 & 11.9 & 20.6 & 32.3 & 50.3 & 68.8 \\ 
-   &  & Max-LM & 5.6 & 6.2 & 7.5 & 10.4 & 15.9 & 25.8 & 41.6 & 60.8 \\ 
-   \hline
+   & & & \multicolumn{9}{l}{Violation Magnitude (SE)} \\
+<<mz_sim-table, results=tex>>=
+## labeling
+levels(mz_sim_sub$pars) <- gsub("k* = ", "", levels(mz_sim_sub$pars), fixed = TRUE)
+levels(mz_sim_sub$nobs) <- gsub( "n = ", "", levels(mz_sim_sub$nobs), fixed = TRUE)
+levels(mz_sim_sub$test) <- gsub("supLM", "max-LM", levels(mz_sim_sub$test), fixed = TRUE)
+
+## table
+mz_tab <- round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
+  data = mz_sim_sub, subset = diff %in% c(seq(0, 4, by = 0.5))),
+  col.vars = "diff"), digits = 1)
+mz_tab <- format(mz_tab, quote = FALSE)[-2, -4]
+mz_tab[1,] <- c("$n$", "$k^*$", "Statistic", format(seq(0, 4, by = 0.5)))
+mz_tab <- paste(apply(mz_tab, 1, paste, collapse = " & "), "\\\\")
+mz_tab[c(1, length(mz_tab))] <- paste(mz_tab[c(1, length(mz_tab))], "\\hline")
+writeLines(mz_tab)
+@
+
 \end{tabular}
@@
 steps have involved hypothesizing a specific type of invariance and
-then testing for it via a LR test, but this is
+then testing for it via a LR~test, but this is
 unnecessary under the proposed framework.
@@
 
-All results were obtained using the {R}~system for statistical computing \cite{R2010},
-version~2.11.1, employing the add-on packages {OpenMx}~0.9.1-1421 \cite{BokNea11} for
-fitting of the factor analysis models and {strucchange}~1.4.2 \cite{ZeiLei02,Zei06}
-for evaluating the parameter instability tests. {R} and {strucchange} are
+All results were obtained using the {R}~system for statistical computing \cite{R11},
+version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on packages
+{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} and
+{OpenMx}~\Sexpr{packageDescription("OpenMx")["Version"]} \cite{BokNea11} for fitting of the factor analysis models and
+{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
+for evaluating the parameter instability tests. {R}~and the packages {lavaan} and {strucchange} are
 freely available under the General Public License~2 from the
@@
 Apache License~2.0 from \url{http://openmx.psyc.virginia.edu/}.
-R code for replication of our results is
+{R}~code for replication of our results is
 available at \url{http://semtools.R-Forge.R-project.org/}.
 
-\bibliography{list}
+\bibliography{refs}
 

### 2011-03-25 www/mz.R
replaced estfun() code for lavaan models by Yves' more general code...now in principle applicable to more general SEMs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@58 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -119,58 +119,66 @@ bread.mzfit <- function(x, ...) vcov(x) * nrow(x$data)
 
 info.mzfit <- function(x, ...) solve(vcov(x) * nrow(x$data))
 
-estfun.mzfit <- function(x, type = c("analytic", "numeric"), ...)
+estfun.mzfit <- function(x, ...)
 {
-  ## numeric or analytic derivatives
-  type <- match.arg(type)
-  if(type == "numeric") stopifnot(require("numDeriv"))
-
   ## estimated coefficients
-  theta <- coef(x)
+  cf <- coef(x)
 
-  ## extract observed data
-  dat <- x$data
-  dat <- as.matrix(dat)
-  n <- nrow(dat)
+  ## observed data
+  X <- as.matrix(x$data)
+  nvar <- ncol(X)
+  nobs <- nrow(X)
 
-  # score matrix
-  scores <- matrix(NA, n, length(theta))
-  colnames(scores) <- names(theta)
-  rownames(scores) <- rownames(dat)
+  rval <- switch(x$engine,
   
-  switch(type,
-  
-  "numeric" = {    
-    # casewise likelihood function
-    likefun <- function(x, ind.dat){
-      mu <- x[14:19]
-      lam <- matrix(c(x[1:3], rep(0,6), x[4:6]), 6, 2) # loadings
-      phi <- matrix(c(1,x[13], x[13], 1), 2, 2)        # factor covs
-      psi <- diag(x[7:12])			       # error vars
+  ## for lavaan we could in principle also handle
+  ## other models:
+  "lavaan" = {
+    # fitted moments
+    moments <- fitted(x$model)
+    Sigma.hat <- moments$cov
+    Mu.hat <- moments$mean
+    Sigma.inv <- solve(Sigma.hat)
 
-      impl.cov <- lam %*% phi %*% t(lam) + psi
+    # scores.H1 (H1 = saturated model)
+    pstar <- nvar * (nvar + 1)/2 + nvar
+    scores.H1 <- matrix(NA, nobs, pstar)
 
-      fval <- (-1/2) * (log(det(impl.cov)) + t(ind.dat - mu) %*% solve(impl.cov) %*% (ind.dat - mu))
-      fval
+    # compute H1 scores per case
+    for(i in 1:nrow(X)) {
+      diff.i <- t(X[i,] - Mu.hat)
[…]

### 2011-03-26 www/paper.Rnw
Some touch-ups in paper   - removed last \readme comments   - used k^* consistently throughout the manuscript   - enhanced the figures/tables and their captions   - slightly changed the wording in a few places
and in the references:
  - updated BRP reference
  - changed internal usage of "title" field to title style
    (coerced to sentence style by some BibTeX styles as in Psychometrika
    but not by others such as jss.bib often employed by Z)



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@59 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\readme{I added a sentence above about score functions because I don't think
-  they are well known in psychometrics.  You might check to see whether my
-  interpretation is reasonable.}
-
 One central assumption -- sometimes made implicitly -- is that
@@
 potential subgroups. Tests for potential changes/instabilities only in the
-remaining parameters are easily constructed by omitting those columns from
-${\bm B}(\hat {\bm \theta})_{ij}$ that are restricted/stable,
-retaining only those
-that are potentially instable. This may be of special interest to
-those wishing to test specific types of measurement invariance, where
-subsets of model parameters are assumed to be stable.
+$k^*$ remaining parameters (from overall $k$ parameters) are easily constructed
+by omitting those $k - k^*$ columns from ${\bm B}(\hat {\bm \theta})_{ij}$ that are
+restricted/stable, retaining only those $k^*$ columns that are potentially instable.
+This may be of special interest to those wishing to test specific types of measurement
+invariance, where subsets of model parameters are assumed to be stable.
 
@@
 scales arising from the mathematical factor.
-For measurement invariance violations, we specify that $V$
+For the measurement invariance violation, we specify that $V$
 (student age) impacts the values of 
@@
 however, then the factor loadings corresponding to the first factor
-are three standard errors lower than those in Figure~\ref{fig:famod}
-(the standard errors are based on the asymptotic standard errors
-divided by $\sqrt{n}$).
+are three standard errors ($=$ asymptotic standard errors divided by $\sqrt{n}$)
+lower than those in Figure~\ref{fig:famod}.
 This violation states that the verbal ability scales
@@
 process.  These include the double-max statistic from \eqref{eq:dmax},
-the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the max-LM
+the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
 statistic from \eqref{eq:maxlm}.
@@
 where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
-verbal factor loading parameters for student $i$.  Alternatively, we
-can consider all model parameters (including means), leading to a test
-of \eqref{eq:h0}.  We consider both of these tests below.
+verbal factor loading parameters for student $i$. Thus, here only $k^* = 3$
+from the overall $k = 19$~model parameters (including means) are assessed.
+Alternatively, we can consider all $k^* = k = 19$~parameters, leading to a test
+of~\eqref{eq:h0}.  We consider both of these tests below.
 
@@
 statistic's fluctuation across values of student age, with the first
-column containing tests of \eqref{eq:exh0} and the second column
-containing tests of \eqref{eq:h0}.  The solid
+column containing tests of~\eqref{eq:exh0} and the second column
+containing tests of~\eqref{eq:h0}.  The solid
 horizontal 
-lines represent critical values for $\alpha=.05$, and
+lines represent critical values for $\alpha= 0.05$, and
 the Cram\'{e}r-von Mises panels 
@@
 Figure~\ref{fig:lrlm} 
-compares the max-LM statistic (solid line) to the max-LR
+compares the $\max \mathit{LM}$ statistic (solid line) to the $\max \mathit{LR}$
 statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
@@
 equivalent.  The 
-max-LR~statistic cannot be obtained from the empirical fluctuation
+$\max \mathit{LR}$~statistic cannot be obtained from the empirical fluctuation
 process, however, so the factor analysis model must be refitted before
-and after each of the possible threshold values $\nu$.
+and after each of the possible threshold values $\nu$ (i.e., here
+$\Sexpr{2 * length(mz.lrstat)}$~model fits for $\Sexpr{length(mz.lrstat)}$~thresholds).
 
 \subsubsection{Estimation of $\nu$ and Parameter Isolation}
-As described above, the tests of \eqref{eq:exh0} imply that the 
+As described above, the tests of~\eqref{eq:exh0} imply that the 
 verbal scales lack measurement invariance.  
@@
 
-\readme{Achim previous comment: I'm not really happy with the following
-  paragraph. I think 
-I wouldn't call this a posthoc test. It is really a test of the same
-hypothesis as before, but now using the double-max
-statistic. \\ \ \\
-Ed response: I am fine with not calling it a posthoc test.  However,
-it could sometimes be advantageous to first test the hypothesis using the CvM
-statistic, then, if CvM rejects the hypothesis, employ the double-max
-statistic to say something about individual model parameters.  This
-would be useless if all parameters change, but, for measurement
-invariance problems, one often expects many model parameters to be stable.}
-
 As mentioned previously, the double-max test is advantageous because
@@
 verbal factor loadings, with the horizontal lines reflecting the
-critical value at $\alpha=.05$.
+critical value at $\alpha= 0.05$.
   The figure
@@
 and second loadings did not achieve the critical value represents a Type II
-error, which implicitly brings into question the posthoc tests' power.
+error, which implicitly brings into question the tests' power.
 This is an example of a situation where the double-max test ``wastes''
@@
 \begin{figure}
-\caption{Three test statistics of \eqref{eq:exh0} and \eqref{eq:h0}, 
+\caption{Three test statistics of~\eqref{eq:exh0} (with $k^* = 3$)
+  and~\eqref{eq:h0} (with $k^* = 19$), 
   based on the example involving measurement invariance with respect
@@
 plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = supLM(0.1), main = "max-LM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
 plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
 plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = supLM(0.1), main = "max-LM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
 @
@@
 \begin{figure}
-\caption{A comparison of the max-LM (solid line) and max-LR (dashed line) test
-  statistics for \eqref{eq:h0}.  The solid horizontal line
-  corresponds to the critical value at $\alpha=.05$ while the dotted vertical
-  line highlights the threshold at which the test statistics are maximized.}
+\caption{A comparison of the $\max \mathit{LM}$ (solid line) and $\max \mathit{LR}$ (dashed line) test
+  statistics for~\eqref{eq:h0} (i.e., $k^* = 19$).  The solid horizontal line
+  corresponds to the critical value at $\alpha= 0.05$ while the dotted vertical
+  line highlights the threshold at which both test statistics assume their maximum.}
 \label{fig:lrlm}
@@
 plot(gefp_19_info, functional = supLM(0.1),
-  xlab = "Age", ylab = "LR and LM statistics", main = "", ylim = c(0, 50))
+  xlab = "Age", ylab = "LR and LM statistics (k* = 19)", main = "", ylim = c(0, 50))
 legend("topleft", c("LR", "LM"), lty = c(2, 1), bty = "n")
@@
   The solid, horizontal lines
-  correspond to the critical value at $\alpha=.05$.}
+  correspond to the critical value at $\alpha= 0.05$.}
 \label{fig:lambdas}
@@
 In this section, we conduct a simulation designed to
-examine the tests' power and Type I error rates in a context
-similar to the example.
+examine the tests' power and Type I error rates in the context also
+employed for the previous example.
 We examine the power and error rates of three tests: the double-max test, the
-Cram\'{e}r-von Mises test, and the max-LM test.  We also compare 
-tests involving only the parameters that changed with tests of all model
-parameters.  The example implied that the former tests had more
+Cram\'{e}r-von Mises test, and the $\max \mathit{LM}$ test.  We also compare 
+tests involving only the $k^* = 3$ parameters that changed with tests of all
+$k^* = k = 19$ model parameters.  The example implied that the former tests had more
 power, and the simulations provide more detail on the extent to which
@@
 \subsection{Method}
-Data were generated in a similar way to that of 
-the example section.
+Data were generated analogously to the example section.
 The data included verbal scales that violated measurement invariance
@@
 invariance violations across three sample sizes ($n=100,
-200, 500$) and fifteen magnitudes of violations.  These violations
+200, 500$) and 17~magnitudes of violations.  These violations
 involved the younger students' values of $\{\lambda_{11},
@@
 values by $d$ 
-times the parameters' asymptotic standard errors, with $d = 0, 0.25,
-0.5, \dots, 4$.  The 0-standard error condition was
+times the parameters' asymptotic standard errors (scaled by $\sqrt{n}$),
+with $d = 0, 0.25, 0.5, \dots, 4$.  The 0-standard error condition was
 used to study Type I error rate.
@@
 Full simulation results are presented in Figure~\ref{fig:simres}, and
-a subset of the results (for finer distinctions) is displayed in
+the underlying numeric values for a subset of the results is additionally displayed in
 Table~\ref{tab:simres}.  In describing the results, we largely refer
@@
 
-Figure~\ref{fig:simres} displays power as a function of violation
+Figure~\ref{fig:simres} displays power curves as a function of violation
 magnitude, with panels for each combination of sample size ($n$) $\times$
 number of parameters being tested ($k^{\ast}$).  
-Circles reflect the double-max test, triangles reflect the max-LM
-test, and plus signs reflect the Cram{\'e}r-von Mises test.  One can
-generally observe that simultaneous tests of all 19 parameters
+Separate curves are drawn for the double-max test (solid lines), 
+the Cram{\'e}r-von Mises test (dashed lines), and the $\max \mathit{LM}$ (dotted lines).
+One can generally observe that simultaneous tests of all 19~parameters
 result in decreased power, with the tests performing more similarly at
@@
 another when only the three factor loadings are tested, with the
-Cram\'{e}r-von Mises test having the most power, followed by max-LM,
+Cram\'{e}r-von Mises test having the most power, followed by $\max \mathit{LM}$,
 followed by the double-max test.  This advantage decreases with
@@
 power advantage of the Cram\'{e}r-von
-Mises test can be as large as .1, most notably when three parameters
+Mises test can be as large as 0.1, most notably when three parameters
 are being tested.
   It also shows that the Cram\'{e}r-von Mises test generally
-has true Type-I error rates, with the double-max test being slightly
-conservative and the max-LM test being slightly liberal.
+has true Type-I error rates, with the double-max test being somewhat
+conservative and the $\max \mathit{LM}$ test being slightly liberal.
 
@@
 data generated here, though more simulations are warranted to examine
-the generality of this finding.
+the generality of this finding in other models or other parameter constellations.
 In the discussion, we describe extensions of the tests in factor
@@
 \begin{figure}
-\caption{Simulated power for three test statistics across three sample
-  sizes, two subsets of tested parameters, and measurement invariance
-  violations of 0--3.5 standard 
-  errors (scaled by $n$).  Circles reflect the double-max test,
-  triangles reflect the 
-  max-LM test, and plus signs reflect the Cram\'{e}r-von Mises test.
-  Numbers above each panel reflect the sample size and number of
-  parameters being tested, respectively.}
+\caption{Simulated power curves for the double-max test (solid), 
+  Cram\'{e}r-von Mises test (dashed), and $\max \mathit{LM}$ test (dotted)
+  across three sample sizes $n$, two subsets of tested parameters $k^*$, and
+  measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$). 
+  See Table~\ref{tab:simres} for the underlying numeric values (using a
+  subset of nine violation magnitudes).}
 \label{fig:simres}
 \setkeys{Gin}{width=\textwidth}
-<<mzsim-xyplot, fig=TRUE, height=7, width=7>>=
+<<mzsim-xyplot, fig=TRUE, height=7.5, width=7>>=
 ## visualization
 trellis.par.set(theme = canonical.theme(color = FALSE))
-print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub, type = "b",
-  xlab = "Violation Magnitude", ylab = "Power", ylim = c(0, 1)))
+mykey <- simpleKey(c("CvM", "max LM", "DM"), points = FALSE, lines = TRUE)
+mykey$lines$pch <- c(2, 3, 1)
+mykey$lines$lty <- c(1.2, 2, 1.2)
+print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub,
+  type = "l", lwd = c(1, 1, 2), ylim = c(0, 1), key = mykey,
+  xlab = "Violation Magnitude", ylab = "Power"))
 @
@@
 \caption{Simulated power for three test statistics
-  across three sample sizes, fifteen magnitudes of measurement
+  across three sample sizes $n$, nine magnitudes of measurement
   invariance violations, and two subsets of tested
-  parameters. Abbreviations: CvM = Cram\'{e}r-von Mises 
-  test; max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
+  parameters $k^*$. Abbreviations: CvM = Cram\'{e}r-von Mises 
+  test; $\max \mathit{LM}$ = Maximum Lagrange multiplier test; DM = Double-max test.
+  See Figure~\ref{fig:simres} for a visualization (using all 17~violation magnitudes).}
 \label{tab:simres}
@@
 levels(mz_sim_sub$nobs) <- gsub( "n = ", "", levels(mz_sim_sub$nobs), fixed = TRUE)
-levels(mz_sim_sub$test) <- gsub("supLM", "max-LM", levels(mz_sim_sub$test), fixed = TRUE)
+levels(mz_sim_sub$test) <- c("$\\mathit{DM}$", "$\\mathit{CvM}$", "$\\max \\mathit{LM}$")
 
@@
 paper, it is also possible to define 
-formal rules for estimating multiple $\nu$ parameters.  We plan to
-explore these issues in the future.
+formal rules for estimating multiple $\nu$ parameters
+(see \citeNP{ZeiSha10}).  We plan to explore these issues in the future.
 
@@
 classification and regression trees \cite{BreFri84,MerSha10,StrMal09},
-with related algorithms being developed for generalized linear
-models \cite{ZeiHot08} and Rasch models \cite{KopZei10}.
+with related algorithms being developed for general parametric models
+\cite{ZeiHot08} and Rasch models in particular \cite{KopZei10}.
 
@@
 while {OpenMx} is available under the
-Apache License~2.0 from \url{http://openmx.psyc.virginia.edu/}.
+Apache License~2.0 from \url{http://OpenMx.psyc.virginia.edu/}.
 {R}~code for replication of our results is

### 2011-03-30 www/report.Rnw
Copy of the current paper.Rnw, modified to use JSS styles for release as a working paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@60 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2011-03-26 www/paper.Rnw
@@
-\documentclass[man]{apa}
-\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
+\documentclass[nojss]{jss}
+
+\usepackage{amsmath,bm,thumbpdf}
 %% need no \usepackage{Sweave}
@@
 
-\title{Studying measurement invariance in factor analysis models
-  via stochastic processes}
-\twoauthors{Edgar C. Merkle}{Achim Zeileis}
-\twoaffiliations{Wichita State University}{Universit\"{a}t Innsbruck}
-
-\abstract{
+\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+
+\title{\vspace*{-0.4cm} A Generalization of Classical Measurement Invariance Tests for Factor Analysis}
+\Plaintitle{A Generalization of Classical Measurement Invariance Tests for Factor Analysis}
+\Shorttitle{Generalized Measurement Invariance Tests for Factor Analysis}
+
+\author{Edgar C. Merkle\\Wichita State University
+   \And Achim Zeileis\\Universit\"{a}t Innsbruck}
+\Plainauthor{Edgar C. Merkle, Achim Zeileis}
+
+\Abstract{
   The issue of measurement invariance commonly arises in factor-analytic
@@
 
-\acknowledgements{Correspondence to Edgar C.\ Merkle, Department of 
-  Psychology, Wichita State University, Wichita, KS 67260-0034.
-  Email: {\texttt{edgar.merkle@wichita.edu}}.}
-\shorttitle{Parameter instability tests}
-\rightheader{Parameter instability tests}
-
-\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+\Keywords{measurement invariance, parameter stability, factor analysis, structural equation models}
+
+\Address{
+  Edgar C. Merkle\\
+  Department of Psychology\\
+  Wichita State University\\
+  Wichita, KS 67260-0034, United States of America\\
+  E-mail: \email{Edgar.Merkle@wichita.edu}\\
+  URL: \url{http://psychology.wichita.edu/merkle/}\\
+
+  Achim Zeileis\\
+  Department of Statistics\\
+  Universit\"at Innsbruck\\
+  Universit\"atsstr.~15\\
+  A-6020 Innsbruck, Austria\\
+  E-mail: \email{Achim.Zeileis@R-project.org}\\
+  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
+}
 
@@
 
-\spacing{1}
-
 \begin{document}
-\maketitle
 
@@
 @
+
+\vspace*{-0.6cm}
+
+\section{Introduction}
 
@@
 violating measurement invariance into structural equation
-models, where relationships between latent variables are hypothesized.  Horn
-and McArdle \citeyear{HorMca92} concisely summarize the impact of
+models, where relationships between latent variables are hypothesized.
+\cite{HorMca92} concisely summarize the impact of
 these issues, stating ``Lack of evidence of measurement 
 invariance equivocates conclusions and casts doubt on theory in the
-behavioral sciences'' (p.~141).  Borsboom \citeyear{Bor06a} further 
+behavioral sciences'' (p.~141).  \cite{Bor06a} further 
 notes that researchers often fail to assess whether measurement
@@
 encountered in structural equation models with missing data (e.g.,
-\citeNP{Wot00}).
+\citealp{Wot00}).
 
@@
 
-\section{Tests of Measurement Invariance}
+\section{Tests of measurement invariance}
 
@@
 adopted here, this means that the parameter vector ${\bm \theta}$ (or
-some subset of ${\bm \theta}$; see \citeNP{Mer93}) is equal
+some subset of ${\bm \theta}$; see \citealp{Mer93}) is equal
 across subgroups of individuals and thus does not vary
@@
 we can apply standard theory: nested multiple group models (e.g.,
-\citeNP{Jor71,Bol89}) coupled with likelihood ratio (LR) tests are most common,
+\citealp{Jor71,Bol89}) coupled with likelihood ratio (LR) tests are most common,
 although the asymptotically equivalent  Lagrange multiplier
 (LM) and Wald tests may also be constructed for this purpose (see
-\citeNP{Sat89}). If $\nu$
+\citealp{Sat89}). If $\nu$
 is unknown (as is often the case for continuous $V$), however, then
@@
 
-\subsection{Likelihood Ratio, Wald, and Lagrange Multiplier Test for Fixed Subgroups}
+\subsection{Likelihood ratio, Wald, and Lagrange multiplier test for fixed subgroups}
+
+\vspace*{-0.6cm}
 
@@
 
-\subsection{Extensions for Unknown Subgroups}
+\subsection{Extensions for unknown subgroups}
 
@@
 in the same way that dichotomization of continuous variables impacts
-general psychometric analyses \cite{MacZha02}.
+general psychometric analyses \citep{MacZha02}.
 
@@
 an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
-maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. However, Andrews \citeyear{And93}
+maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. However, \cite{And93}
 showed that the asymptotic distribution is in fact tractable, but nonstandard.
@@
 $\max \mathit{LM}(\nu)$, respectively, which both have the same asymptotic properties
-as $\max \mathit{LR}(\nu)$ and are asymptotically equivalent \cite{And93}.
+as $\max \mathit{LR}(\nu)$ and are asymptotically equivalent \citep{And93}.
 From a computational perspective, the $\max \mathit{LM}(\nu)$ test is
@@
 
-\section{Stochastic Processes for Measurement Invariance}
+\section{Stochastic processes for measurement invariance}
 
@@
 in the statistics and econometrics literature under the label ``structural change
-tests'' (see e.g., \citeNP{BroDur75,And93}) where the focus of interest is the
+tests'' (see e.g., \citealp{BroDur75,And93}) where the focus of interest is the
 detection of parameter instabilities of time series models ``along'' time.
-Specifically, it has been shown (e.g., \citeNP{Nyb89,Han92,HjoKon02,ZeiHor07})
+Specifically, it has been shown (e.g., \citealp{Nyb89,Han92,HjoKon02,ZeiHor07})
 that cumulative sums of the empirical scores follow specific stochastic processes, allowing
@@
 of factor analysis models. More detailed accounts of the underlying structural
-change methods include Hjort and Koning \citeyear{HjoKon02} and
-Zeileis and Hornik \citeyear{ZeiHor07}.
+change methods include \cite{HjoKon02} and \cite{ZeiHor07}.
 
@@
 the usual ML regularity conditions (assuring asymptotic normality
-of $\hat {\bm \theta}$), Hjort \& Koning \citeyear{HjoKon02} show
+of $\hat {\bm \theta}$), \cite{HjoKon02} show
 that 
@@
 of this test statistic is then easily obtained by applying the same
-aggregation to the asymptotic process ${\bm B}^0$ \cite{HjoKon02,ZeiHor07},
-so that corresponding $p$-values can be derived.
+aggregation to the asymptotic process ${\bm B}^0$ \citep{HjoKon02,ZeiHor07},
+so that corresponding $p$~values can be derived.
 
@@
 Further aggregation functions have been suggested in the structural
-change literature (see e.g., \citeNP{Zei05,ZeiSha10}) but the three
+change literature (see e.g., \citealp{Zei05,ZeiSha10}) but the three
 tests above are probably most likely to be useful in psychometric settings.
@@
 Finally, all tests can be easily modified to address the situation of
-so-called ``partial structural changes'' \cite{And93}. This refers to the case of some
+so-called ``partial structural changes'' \citep{And93}. This refers to the case of some
 parameters being known to be stable, i.e., restricted to be constant across
@@
 
-\subsection{Critical Values \& $p$-values.}
+\subsection[Critical values and p values]{Critical values and $p$~values}
 
@@
 statistics from the previous section is straightforward: It is simply the aggregation
-of the asympotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
+of the asympotic process ${\bm B}^0(t)$ \citep{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
 from (\ref{eq:dmax}) converges to $\sup_t || {\bm B}^0(t) ||_{\infty}$, where
@@
 it is not always easy to find closed-form solutions for computing critical
-values and $p$-values from them. In some cases -- in particular for the double maximum test --
+values and $p$~values from them. In some cases -- in particular for the double maximum test --
 such a closed-form solution is available from analytic results for Gaussian
-processes (see e.g., \citeNP{ShoWel86}). For all other cases, tables of critical
-values can be obtained from direct simulation \cite{Zei06} or in combination with
-more refined techniques such as response surface regression \cite{Han97}.
-
-The analytic solution for the asymptotic $p$-value of a $\mathit{DM}$ statistic
+processes (see e.g., \citealp{ShoWel86}). For all other cases, tables of critical
+values can be obtained from direct simulation \citep{Zei06} or in combination with
+more refined techniques such as response surface regression \citep{Han97}.
+
+The analytic solution for the asymptotic $p$~value of a $\mathit{DM}$ statistic
 $d$ is
@@
 This combines the crossing probability of a univariate Brownian bridge
-(see e.g., \citeNP{ShoWel86,PloKra92}) with a straightforward Bonferroni
+(see e.g., \citealp{ShoWel86,PloKra92}) with a straightforward Bonferroni
 correction to obtain the $k$-dimensional case. The terms in the summation
@@
 
-For the Cram\'{e}r-von Mises test statistic $\mathit{CvM}$, Nyblom \citeyear{Nyb89}
-and Hansen \citeyear{Han92} provide small tables of critical values which
-have been extended in the software provided by Zeileis \citeyear{Zei06}.
+For the Cram\'{e}r-von Mises test statistic $\mathit{CvM}$, \cite{Nyb89}
+and \cite{Han92} provide small tables of critical values which
+have been extended in the software provided by \cite{Zei06}.
 Critical values for the distribution of the maximum LR/Wald/LM tests
-are provided by Hansen \citeyear{Han97}. Note that the distribution depends on
+are provided by \cite{Han97}. Note that the distribution depends on
 the minimal and maximal thresholds employed in the test.
 
-\subsection{Locating the Invariance.}
+\subsection{Locating the invariance}
 
@@
 be determined by maximizing the corresponding segmented log-likelihood over all
-possible combinations of thresholds (\citeNP{ZeiSha10} adapt a dynamic programming
+possible combinations of thresholds (\citealp{ZeiSha10} adapt a dynamic programming
 algorithm to this task). For the single threshold case, this reduces to
@@
 (2) test measurement invariance for unknown $\nu$. 
-In particular, Millsap \citeyear{Mil05} cites
+In particular, \cite{Mil05} cites
 ``locating the invariance violation'' as a major outstanding problem
@@
 
-\begin{figure}
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.78\textwidth]{famod.pdf}
 \caption{Path diagram representing the base factor analysis model used for
@@
 values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
-$\lambda_{31}$).}
-\label{fig:famod}
-\includegraphics[height=5in]{famod.pdf}
+$\lambda_{31}$).
+\label{fig:famod}}
+\end{center}
 \end{figure}
 
-\section{Example with Artificial Data}
+\section{Example with artificial data}
 
@@
 \eqref{eq:cumscore}.  Finally, we
-obtained various test statistics and $p$-values from the cumulative score
+obtained various test statistics and $p$~values from the cumulative score
 process.  These include the double-max statistic from \eqref{eq:dmax},
@@
 of~\eqref{eq:h0}.  We consider both of these tests below.
-
-\subsection{Results}
-In the results section, we first describe overall results.  We then 
-describe estimation of $\nu$ and 
-isolation of model parameters violating
-measurement invariance.
 
@@
 
+\begin{figure}[p!]
+\begin{center}
+\setkeys{Gin}{width=0.95\textwidth}
+<<gefp-3-19, fig=TRUE, height=8, width=6>>=
+## empirical fluctuation processes (3 vs all 19 pars)
+gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
+  vcov = info.mzfit, sandwich = FALSE)
+gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
+  vcov = info.mzfit, sandwich = FALSE)
+
+## visualization
+par(mfcol = c(3, 2))
+plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
+plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
+@
+\caption{Three test statistics of~\eqref{eq:exh0} (with $k^* = 3$)
+  and~\eqref{eq:h0} (with $k^* = 19$), 
+  based on the example involving measurement invariance with respect
+  to student age.  Solid, horizontal lines represent critical values
+  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
+  represent values of the Cram\'{e}r-von Mises test statistic.
+\label{fig:cusumex}}
+\end{center}
+\end{figure}
+
+\subsection{Results}
+
+In the results section, we first describe overall results.  We then 
+describe estimation of $\nu$ and 
+isolation of model parameters violating
+measurement invariance.
+
 \subsubsection{Overall Results}
+
 Test statistics for the hypotheses \eqref{eq:exh0} and~\eqref{eq:h0}
@@
 
+\begin{figure}[t!]
+\begin{center}
+\setkeys{Gin}{width=0.8\textwidth}
+<<lm-lr-19, fig=TRUE, height=5, width=6>>=
+plot(gefp_19_info, functional = supLM(0.1),
+  xlab = "Age", ylab = "LR and LM statistics (k* = 19)", main = "", ylim = c(0, 50))
+legend("topleft", c("LR", "LM"), lty = c(2, 1), bty = "n")
+lines(mz.age, mz.lrstat, lty = 2)
+abline(v = mz.age[which.max(mz.lrstat)], lty = 3)
+@
+\caption{A comparison of the $\max \mathit{LM}$ (solid line) and $\max \mathit{LR}$ (dashed line) test
+  statistics for~\eqref{eq:h0} (i.e., $k^* = 19$).  The solid horizontal line
+  corresponds to the critical value at $\alpha= 0.05$ while the dotted vertical
+  line highlights the threshold at which both test statistics assume their maximum.
+\label{fig:lrlm}}
+\end{center}
+\end{figure}
+
 The figures convey information about several properties
@@
 sensitive to fluctuations among a small subset of parameters.
+
+\begin{figure}[t!]
+\begin{center}
+\setkeys{Gin}{width=0.8\textwidth}
+<<gefp-lambdas, fig=TRUE, height=8, width=6>>=
+plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
+  main = "", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
+@
+\caption{Cumulative score processes for each verbal factor
+  loadings, with critical values stemming from the double-max test. 
+  The solid, horizontal lines
+  correspond to the critical value at $\alpha= 0.05$.
+\label{fig:lambdas}}
+\end{center}
+\end{figure}
 
@@
 
-\subsubsection{Estimation of $\nu$ and Parameter Isolation}
+\subsubsection[Estimation of nu and parameter isolation]{Estimation of $\nu$ and parameter isolation}
+
 As described above, the tests of~\eqref{eq:exh0} imply that the 
@@
 
-\begin{figure}
-\caption{Three test statistics of~\eqref{eq:exh0} (with $k^* = 3$)
-  and~\eqref{eq:h0} (with $k^* = 19$), 
-  based on the example involving measurement invariance with respect
-  to student age.  Solid, horizontal lines represent critical values
-  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
-  represent values of the Cram\'{e}r-von Mises test statistic.}
-\label{fig:cusumex}
-\setkeys{Gin}{width=\textwidth}
-<<gefp-3-19, fig=TRUE, height=8, width=6>>=
-## empirical fluctuation processes (3 vs all 19 pars)
-gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
-  vcov = info.mzfit, sandwich = FALSE)
-gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
-  vcov = info.mzfit, sandwich = FALSE)
-
-## visualization
-par(mfcol = c(3, 2))
-plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
-plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
-@
-\end{figure}
-
-\begin{figure}
-\caption{A comparison of the $\max \mathit{LM}$ (solid line) and $\max \mathit{LR}$ (dashed line) test
-  statistics for~\eqref{eq:h0} (i.e., $k^* = 19$).  The solid horizontal line
-  corresponds to the critical value at $\alpha= 0.05$ while the dotted vertical
-  line highlights the threshold at which both test statistics assume their maximum.}
-\label{fig:lrlm}
-\setkeys{Gin}{width=\textwidth}
-<<lm-lr-19, fig=TRUE, height=5, width=6>>=
-plot(gefp_19_info, functional = supLM(0.1),
-  xlab = "Age", ylab = "LR and LM statistics (k* = 19)", main = "", ylim = c(0, 50))
-legend("topleft", c("LR", "LM"), lty = c(2, 1), bty = "n")
-lines(mz.age, mz.lrstat, lty = 2)
-abline(v = mz.age[which.max(mz.lrstat)], lty = 3)
-@
-\end{figure}
-
-\begin{figure}
-\caption{Cumulative score processes for each verbal factor
-  loadings, with critical values stemming from the double-max test. 
-  The solid, horizontal lines
-  correspond to the critical value at $\alpha= 0.05$.}
-\label{fig:lambdas}
-\setkeys{Gin}{width=\textwidth}
-<<gefp-lambdas, fig=TRUE, height=8, width=6>>=
-plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
-  main = "", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
-@
-\end{figure}
-
 \section{Simulation}
@@
 
+\begin{figure}[t!]
+\begin{center}
+\setkeys{Gin}{width=\textwidth}
+<<mzsim-xyplot, fig=TRUE, height=7.5, width=7>>=
+## visualization
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(c("CvM", "max LM", "DM"), points = FALSE, lines = TRUE)
+mykey$lines$pch <- c(2, 3, 1)
+mykey$lines$lty <- c(1.2, 2, 1.2)
+print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub,
+  type = "l", lwd = c(1, 1, 2), ylim = c(0, 1), key = mykey,
+  xlab = "Violation Magnitude", ylab = "Power"))
+@
+\caption{Simulated power curves for the double-max test (solid), 
+  Cram\'{e}r-von Mises test (dashed), and $\max \mathit{LM}$ test (dotted)
+  across three sample sizes $n$, two subsets of tested parameters $k^*$, and
+  measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$). 
+  See Table~\ref{tab:simres} for the underlying numeric values (using a
+  subset of nine violation magnitudes).
+\label{fig:simres}}
+\end{center}
+\end{figure}
+
 \subsection{Results}
@@
 to the figure.
+
+\begin{table}[t!]
+\begin{center}
+\begin{tabular}{lrlrrrrrrrrr}
+  \hline
+   & & & \multicolumn{9}{l}{Violation Magnitude (SE)} \\
+<<mz_sim-table, results=tex>>=
+## labeling
+levels(mz_sim_sub$pars) <- gsub("k* = ", "", levels(mz_sim_sub$pars), fixed = TRUE)
+levels(mz_sim_sub$nobs) <- gsub( "n = ", "", levels(mz_sim_sub$nobs), fixed = TRUE)
+levels(mz_sim_sub$test) <- c("$\\mathit{DM}$", "$\\mathit{CvM}$", "$\\max \\mathit{LM}$")
+
+## table
+mz_tab <- round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
+  data = mz_sim_sub, subset = diff %in% c(seq(0, 4, by = 0.5))),
+  col.vars = "diff"), digits = 1)
+mz_tab <- format(mz_tab, quote = FALSE)[-2, -4]
+mz_tab[1,] <- c("$n$", "$k^*$", "Statistic", format(seq(0, 4, by = 0.5)))
+mz_tab <- paste(apply(mz_tab, 1, paste, collapse = " & "), "\\\\")
+mz_tab[c(1, length(mz_tab))] <- paste(mz_tab[c(1, length(mz_tab))], "\\hline")
+writeLines(mz_tab)
+@
+
+\end{tabular}
+\caption{Simulated power for three test statistics
+  across three sample sizes $n$, nine magnitudes of measurement
+  invariance violations, and two subsets of tested
+  parameters $k^*$. Abbreviations: CvM = Cram\'{e}r-von Mises 
+  test; $\max \mathit{LM}$ = Maximum Lagrange multiplier test; DM = Double-max test.
+  See Figure~\ref{fig:simres} for a visualization (using all 17~violation magnitudes).
+\label{tab:simres}}
+\end{center}
+\end{table}
 
@@
 
-\begin{figure}
-\caption{Simulated power curves for the double-max test (solid), 
-  Cram\'{e}r-von Mises test (dashed), and $\max \mathit{LM}$ test (dotted)
-  across three sample sizes $n$, two subsets of tested parameters $k^*$, and
-  measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$). 
-  See Table~\ref{tab:simres} for the underlying numeric values (using a
-  subset of nine violation magnitudes).}
-\label{fig:simres}
-\setkeys{Gin}{width=\textwidth}
-<<mzsim-xyplot, fig=TRUE, height=7.5, width=7>>=
-## visualization
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(c("CvM", "max LM", "DM"), points = FALSE, lines = TRUE)
-mykey$lines$pch <- c(2, 3, 1)
-mykey$lines$lty <- c(1.2, 2, 1.2)
-print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub,
-  type = "l", lwd = c(1, 1, 2), ylim = c(0, 1), key = mykey,
-  xlab = "Violation Magnitude", ylab = "Power"))
-@
-\end{figure}
-
-\begin{table*}
-\caption{Simulated power for three test statistics
-  across three sample sizes $n$, nine magnitudes of measurement
-  invariance violations, and two subsets of tested
-  parameters $k^*$. Abbreviations: CvM = Cram\'{e}r-von Mises 
-  test; $\max \mathit{LM}$ = Maximum Lagrange multiplier test; DM = Double-max test.
-  See Figure~\ref{fig:simres} for a visualization (using all 17~violation magnitudes).}
-\label{tab:simres}
-\begin{center}
-\begin{tabular}{lrlrrrrrrrrr}
-  \hline
-   & & & \multicolumn{9}{l}{Violation Magnitude (SE)} \\
-<<mz_sim-table, results=tex>>=
-## labeling
-levels(mz_sim_sub$pars) <- gsub("k* = ", "", levels(mz_sim_sub$pars), fixed = TRUE)
-levels(mz_sim_sub$nobs) <- gsub( "n = ", "", levels(mz_sim_sub$nobs), fixed = TRUE)
-levels(mz_sim_sub$test) <- c("$\\mathit{DM}$", "$\\mathit{CvM}$", "$\\max \\mathit{LM}$")
-
-## table
-mz_tab <- round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
-  data = mz_sim_sub, subset = diff %in% c(seq(0, 4, by = 0.5))),
-  col.vars = "diff"), digits = 1)
-mz_tab <- format(mz_tab, quote = FALSE)[-2, -4]
-mz_tab[1,] <- c("$n$", "$k^*$", "Statistic", format(seq(0, 4, by = 0.5)))
-mz_tab <- paste(apply(mz_tab, 1, paste, collapse = " & "), "\\\\")
-mz_tab[c(1, length(mz_tab))] <- paste(mz_tab[c(1, length(mz_tab))], "\\hline")
-writeLines(mz_tab)
-@
-
-\end{tabular}
-\end{center}
-\end{table*}
-
 \section{Discussion}
@@
 
-\subsection{Use in Practice}
+\subsection{Use in practice}
 % Simultaneous test of Meredith's various types of invariance
@@
 results relevant to many types of measurement
-invariance (see, e.g., \citeNP{Mer93}), or (2)~test a single subset of
+invariance (see, e.g., \citealp{Mer93}), or (2)~test a single subset of
 model parameters, potentially leading to improved power to detect a
@@
 formal rules for estimating multiple $\nu$ parameters
-(see \citeNP{ZeiSha10}).  We plan to explore these issues in the future.
-
-% \subsection{Further Issues}
+(see \citealp{ZeiSha10}).  We plan to explore these issues in the future.
+
+% \subsection{Further issues}
 % Further issues associated with the proposed tests include: (1)
@@
 % Differing proportions of people in each group
-%\subsubsection{Subgroup Proportions}
+%\subsubsection{Subgroup proportions}
 %The power of the proposed tests is impacted by the proportion of
@@
 % Ties on auxiliary variable (mention use in categorical contexts)
-\subsection{Categorical Auxiliary Variable}
+\subsection{Categorical auxiliary variable}
 % There remain unaddressed theoretical issues specific to the 
@@
 the entries in the column to obtain a $\chi^2$-distributed statistic with
-$(C-1)$ degrees of freedom \cite{HjoKon02}.
+$(C-1)$ degrees of freedom \citep{HjoKon02}.
 Alternatively, to simultaneously test multiple parameters, we can sum
@@
 multiple auxiliary variables.  Such an algorithm is related to 
-classification and regression trees \cite{BreFri84,MerSha10,StrMal09},
+classification and regression trees \citep{BreFri84,MerSha10,StrMal09},
 with related algorithms being developed for general parametric models
-\cite{ZeiHot08} and Rasch models in particular \cite{KopZei10}.
+\citep{ZeiHot08} and Rasch models in particular \citep{KopZei10}.
 
 Relatedly,
-S\'{a}nchez \citeyear{San09} describes a general
+\cite{San09} describes a general
 method for partitioning/segmenting structural equation models within a 
@@
 % invariance.   and
-% C4.5 \cite{Qui93}.  The algorithm would proceed via the following
+% C4.5 \citep{Qui93}.  The algorithm would proceed via the following
 % steps: 
@@
 % The algorithm described in the previous paragraph is related to one
-% developed by Zeileis et al.\ \citeyear{ZeiHot08} for representing
+% developed by \cite{ZeiHot08} for representing
 % parameter invariance violations in generalized linear models.  Zeileis
@@
 the tests may be extended to study differential item functioning (DIF)
-in item response models (e.g., \citeNP{KopZei10}, who focused on recursive
+in item response models (e.g., \citealp{KopZei10}, who focused on recursive
 partitioning of Rasch models).  
@@
 largely independently of one another, 
-some treatments \cite{Mcd99} and recent research \cite{StaChe06} have
+some treatments \citep{Mcd99} and recent research \citep{StaChe06} have
 sought to unify the methods.  Extensions of the proposed tests can
@@
 
-\section*{Computational Details}
-
-All results were obtained using the {R}~system for statistical computing \cite{R11},
+\section*{Computational details}
+
+All results were obtained using the {R}~system for statistical computing \citep{R11},
 version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on packages
-{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} and
-{OpenMx}~\Sexpr{packageDescription("OpenMx")["Version"]} \cite{BokNea11} for fitting of the factor analysis models and
-{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
+{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \citep{lavaan11} and
+{OpenMx}~\Sexpr{packageDescription("OpenMx")["Version"]} \citep{BokNea11} for fitting of the factor analysis models and
+{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \citep{ZeiLei02,Zei06}
 for evaluating the parameter instability tests. {R}~and the packages {lavaan} and {strucchange} are

### 2011-03-31 www/paper.Rnw
Minor wording changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@61 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\title{Studying measurement invariance in factor analysis models
-  via stochastic processes}
+\title{Generalized measurement invariance tests with application to
+  factor analysis}
 \twoauthors{Edgar C. Merkle}{Achim Zeileis}
-\twoaffiliations{Wichita State University}{Universit\"{a}t Innsbruck}
+\twoaffiliations{University of Missouri--Columbia}{Universit\"{a}t Innsbruck}
 
@@
   definition of the number of groups, group membership, and offending model
-  parameters. In this paper, we construct tests of parameter invariance based on
+  parameters. In this paper, we construct tests of measurement invariance based on
   stochastic processes of casewise derivatives of the likelihood function. These
@@
 \acknowledgements{Correspondence to Edgar C.\ Merkle, Department of 
-  Psychology, Wichita State University, Wichita, KS 67260-0034.
-  Email: {\texttt{edgar.merkle@wichita.edu}}.}
-\shorttitle{Parameter instability tests}
-\rightheader{Parameter instability tests}
+  Psychological Sciences, University of Missouri--Columbia, Columbia,
+  MO 65211.
+  Email: {\texttt{ecmerkle@gmail.com}}.}
+\shorttitle{Generalized measurement invariance tests}
+\rightheader{Generalized measurement invariance tests}
 
@@
 
-\spacing{1}
+%\spacing{1}
 
@@
 This leads to a discussion of theoretical results relevant to the
-proposed tests, as well as a comparison of the proposed tests with
+proposed tests, as well as a comparison of the proposed tests to the 
 existing tests.  Next,
@@
 covariance matrix ${\bm{\Sigma}}({\bm{\theta}})$. As pointed out above,
-the assumptions for the tests introduced below do not require this specific
+the assumptions for the tests introduced here do not require this specific
 form of the likelihood, but it is presented for illustration due to
@@
 
-Within the general framework outlined above, and under the usual
+Within the general framework outlined above and under the usual
 regularity conditions, the model parameters $\bm{\theta}$ can
@@
 If this is not satisfied, the estimates $\hat{\bm{\theta}}$ are typically
-not meaningful and cannot easily be interpreted.
+not meaningful and cannot be easily interpreted.
 One potential source of deviation from this assumption is lack of
@@
 of the data (with some parameters potentially restricted to be equal across
-subgroups).\footnote{For ease of exposition, we describe
-the case where there are no such parameter restrictions.  As shown in
-the example and simulation below, it is
-straightforward to extend all methods to the more general case.}
-Then, the sum of maximized likelihoods 
+subgroups).  For ease of exposition, we describe
+the case where there are no such parameter restrictions; as shown in
+the example and simulation below, however, it is
+straightforward to extend all methods to the more general case.
+After fitting the model to each subgroup, the sum of maximized likelihoods 
 from the subgroups are compared with the original maximized full-sample likelihood
-in a $\chi^2$~test. More specifically, for the special case of two subgroups,
+in a $\chi^2$~test. For the special case of two subgroups,
 the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
 and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
-The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
+The parameter estimates $\hat {\bm \theta}^{(A)}$ can then be obtained from the
 observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
@@
 an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
-maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. However, Andrews \citeyear{And93}
-showed that the asymptotic distribution is in fact tractable, but nonstandard.
+maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. Andrews \citeyear{And93}
+showed that the asymptotic distribution is in fact tractable but nonstandard.
 Specifically, the asymptotic distribution of (\ref{eq:maxlr}) is the maximum of a
@@
 Having estimated the parameters $\hat {\bm \theta}$, the goal is to check that
-there are no subgroups of individuals that fail to conform with the model (for any
+all subgroups of individuals conform with the model (for all 
 of the parameters). Hence, some measure of model deviation or residual is required
@@
 of the score contributions over all individuals is zero for each
-component, see
+component; see
 (\ref{eq:ml1}). Moreover, if there are no systematic deviations, the score contributions
@@
 if any component of the the cumulative score process ${\bm B}(t; \hat {\bm \theta})$
-departs ``too far'' from zero at any time, i.e., if
+strays ``too far'' from zero at any time, i.e., if
 \begin{equation}
@@
 change literature (see e.g., \citeNP{Zei05,ZeiSha10}) but the three
-tests above are probably most likely to be useful in psychometric settings.
+tests above are most likely to be useful in psychometric settings.
 
@@
 correction to obtain the $k$-dimensional case. The terms in the summation
-quickly go to zero as $h$ goes to infinity, so only some large finite
+quickly go to zero as $h$ goes to infinity, so that only some large finite
 number of terms (say, 100) need to be evaluated in practice.
@@
 For the maximum LR/Wald/LM tests, it is natural to graph the sequence of LR/Wald/LM
-statistics along $V$ along with a boundary corresponding to the critical value.
+statistics along $V$, with a boundary corresponding to the critical value.
 Again, a boundary crossing signals a significant violation, and the peak(s) in the
@@
 same threshold asymptotically. However, in empirical samples, their location may differ somewhat
-(although often not much).
+(although often not by much).
 
@@
 
-Consider a 
-hypothetical battery of six scales administered to students aged 13 to 18
-years.  Three of the scales are intended to measure verbal ability, and
-three of the scales are intended to measure mathematical ability.  We
-may observe a maturation effect in the resulting data, whereby the
+Consider the factor analysis model in Figure~\ref{fig:famod}, which
+includes six manifest variables, two 
+correlated factors, and a single auxiliary variable $V$.  Though an
+applied framework is unnecessary for the example, we may imagine that
+six scales have been administered to students aged 13 to 18
+years, with three of the scales intended to measure verbal ability and
+three of the scales intended to measure mathematical ability.  We
+assume a measurement invariance violation for the factor loading
+parameters, whereby the factor loadings for individuals with low
+values of $V$ are smaller than the factor loadings for individuals
+with high values of $V$ (e.g.,  
 factor loadings for older students are larger than those for younger
-students.  The researcher's goal is to
-study whether the scales are measurement invariant with
-respect to age, which is taken to be the auxiliary variable $V$.
+students).  We generally wish to assess measurement invariance with
+respect to $V$ (student age).
 
@@
 
-To formally represent these ideas, we specify that the data arise from
-a factor analysis model with two factors.  The base model, displayed in
-Figure~\ref{fig:famod}, specifies that measurement invariance holds,
-with three scales arising from the verbal factor and three 
-scales arising from the mathematical factor.
+The base factor analysis model, displayed in
+Figure~\ref{fig:famod} with set parameter values, specifies that
+measurement invariance holds.
 For the measurement invariance violation, we specify that $V$
@@
 lower than those in Figure~\ref{fig:famod}.
-This violation states that the verbal ability scales
+This violation states, e.g., that the verbal ability scales
 lack measurement invariance with respect to age.  For simplicity, we
@@
 dashed line, so we would conclude a measurement 
-invariance violation with respect to age for the third
+invariance violation for the third
 verbal test.  The fact that the cumulative score process for the first
@@
 Data were generated analogously to the example section.
-The data included verbal scales that violated measurement invariance
-with respect to students' ages, with students 13--15 years of age
-having smaller factor loadings than students 16--18 years of age.  
-Simulation data were generated from the same general factor analysis
-model (see Figure~\ref{fig:famod}), but sample size and magnitude
+The data were generated from a factor analysis model with two
+correlated factor and six manifest variables, with individuals with
+low $V$ (13--15) having smaller factor loadings than individuals with
+large $V$ (16--18).
+Sample size and magnitude
 of measurement 
@@
 5,000 datasets were generated and tested.  In each dataset, half the
-students were 13--15 
-years of age and half were 16--18 years of age.
+individuals had ``low $V$'' (e.g., 13--15 
+years of age) and half had ``high $V$'' (e.g., 16--18 years of age).
 
@@
 
-% \subsection{Further Issues}
-% Further issues associated with the proposed tests include: (1)
-% alternative methods for estimating $\eta$;
-% and (2) dealing
-% with categorical $V$.  We address each in turn below.
-
-% % More than two groups violating MI
-% \subsubsection{Estimation of $\eta$}
-% Given rejection of the hypothesis of measurement invariance, the most
-% straightforward way to formally estimate $\eta$ is through
-% the location of the maximum of the cumulative score process.  Assuming
-% a single threshold, the maximum should be achieved close to the true
-% $\eta$.  Referring to the example (see
-% Figure~\ref{fig:cusumex}), we defined $\eta$ as the 
-% age corresponding to the solid dot in the third panel.
-
-% To study this method in more detail, we used
-% results of our simulation to examine the age at which the maximum of
-% the cumulative score process occurs.
-% If the test accurately recovers the violating subgroups,
-% then the maximum should be close to 16 years
-% of age.\footnote{The data generated for these
-%   simulations 
-%   contained ages measured in days, as opposed to years.  Thus,
-%   the tests could split students at continuous values like 15.6 or
-%   16.3 years of age.}
-%   For each simulation condition, Table~\ref{tab:age1}
-%   displays confidence intervals for the location of the maximum
-%   absolute cumulative score.  These confidence intervals were
-%   calculated across generated 
-%   datasets, with each dataset contributing one maximum absolute cumulative
-%   score
-%   (regardless of whether the hypothesis of
-%   measurement invariance was rejected).
-% It is observed that these confidence intervals slightly underestimate
-% the true age dividing the subgroups, with the degree of bias
-% decreasing as sample size increases.  The interval widths
-% are impacted by both the magnitude of the invariance and the sample
-% size.  In future work, we plan to study alternative ways of
-% identifying subgroups.
-
-% \begin{table*}
-% \caption{Confidence intervals for the location of the maximum absolute
-%   cumulative score across the conditions from the simulation.}
-% \label{tab:age1}
-%   \begin{tabular}{lccc} \thickline
-%   & \multicolumn{3}{c}{Violation Magnitude} \\
-%   $n$ & 1SE & 2SE & 3SE \\\hline
-%   100 & (15.61; 15.73) & (15.55; 15.65) & (15.50; 15.59) \\
-%   200 & (15.67; 15.77) & (15.64; 15.72) & (15.67; 15.73) \\
-%   500 & (15.71; 15.81) & (15.77; 15.83) & (15.80; 15.83) \\\thickline
-%   \end{tabular}
-% \end{table*}
-
-% Differing proportions of people in each group
-%\subsubsection{Subgroup Proportions}
-%The power of the proposed tests is impacted by the proportion of
-%individuals within each subgroup.
-
-% Ties on auxiliary variable (mention use in categorical contexts)
 \subsection{Categorical Auxiliary Variable}
-% There remain unaddressed theoretical issues specific to the 
-% tests' application to measurement invariance.  These
-% issues generally stem from the types of auxiliary variables on which
-% researchers may study measurement invariance.  In
-% the example described earlier, researchers wanted to know whether a
-% scale was measurement invariant with respect to student age.
-% Students could then be ordered based on age, and a test could be 
-% constructed via Equation~\eqref{eq:cumscore}.  However, an issue arises if there
-% are ties on age; that is, if two students' ages are exactly
-% the same.  The proposed tests assume complete ordering of the
-% students, but these two students cannot be ordered.  In this case, we
-% may randomly order the two students, but there is an
-% arbitrariness associated with this strategy.  Alternatively, to obtain
-% tests that liberally reject the hypothesis of measurement invariance,
-% we may choose the ordering that results in the largest test
-% statistic.  If there are many tied students, then there are many
-% possible orderings that must be examined.  However, examination of all
-% possible orderings remains computationally feasible because the
-% casewise first derivatives need to be calculated only once; the
-% derivatives simply need to be reordered for each ordering of tied
-% students.
-
 One issue that was largely unaddressed in this paper involved the use
@@
 the cumulative score process $\Delta_{I_c} {\bm B}(\hat {\bm \theta})$
-within in each category is just the sum of the corresponding scores.
+within each category is just the sum of the corresponding scores.
 In somewhat sloppy notation:
@@
 parameters.  In addition to categorical $V$, this framework may also
-be useful for continuous $V$ with many ties and ordinal $V$.  
+be useful for both continuous $V$ with many ties and ordinal $V$.  
 The number of potential thresholds may be very low in these
@@
 
-% Recursive method
-% \subsubsection{Recursive Multiple-Groups Models}
-% The proposed tests have the ability to automatically define groups of
-% individuals violating measurement invariance with respect to a single
-% auxiliary (often demographic) variable.  If a researcher has multiple
-% auxiliary variables that he/she wishes to assess for measurement
-% invariance, then the proposed tests can be inserted into a recursive
-% algorithm 
-% that isolates groups of individuals violating measurement
-% invariance.   and
-% C4.5 \cite{Qui93}.  The algorithm would proceed via the following
-% steps: 
-% \begin{enumerate}
-%   \item Fit a confirmatory factor analysis model to the data, with all
-%   parameters constrained to be equal across observations.
-%   \item Conduct a measurement invariance test with respect to each
-%     auxiliary variable (auxiliary variables are determined in
-%     advance by the researcher).
-%   \item If violations of measurement invariance are detected,
-%   split the data into two (or more) groups based on the auxiliary variable
-%     corresponding to the largest test statistic.
-%   \item Fit a multiple-groups factor analysis model based on step 3.
-%   \item Repeat steps 2--5.
-% \end{enumerate}
-% The algorithm would continue until it reaches a
-% predefined stopping rule.  For example, we may stop 
-% the algorithm when the tests fail to indicate measurement invariance
-% violations.  Alternatively, the algorithm may be stopped when the
-% sample size within any subgroup reaches a minimum level.
-
-% The algorithm described in the previous paragraph is related to one
-% developed by Zeileis et al.\ \citeyear{ZeiHot08} for representing
-% parameter invariance violations in generalized linear models.  Zeileis
-% et al.\ present the results in a tree structure, which is
-% suboptimal for factor analysis models.  Instead, the subgroups
-% discovered by the algorithm are used here to adaptively define a
-% multiple-groups factor analysis model.  The
-% simultaneous study of measurement invariance with respect to multiple
-% auxiliary variables is an important future direction.
-
-%\subsubsection{Other Latent Variable Models}
 The proposed tests also 
@@
 observations.  For 
-example, it would be possible to assess whether relationships between
+example, it would be possible to assess whether paths between
 latent variables are stronger for some individuals than for
@@
 stochastic process-based parameter instability tests from theoretical
-statistics to the issue of measurement invariance in psychometrics.
+statistics and applied them to the issue of measurement invariance in psychometrics.
 The paper included both theoretical development and 
-study of the tests' performance.  The tests were found via simulation
-to have good properties, making them useful for many psychometric
-applications.  More generally, the tests help to
+study of the tests' performance.  The tests were found to have good
+properties via simulation, making them useful for many psychometric
+applications.  More generally, the tests help 
 solve standing problems in

### 2011-03-31 www/report.Rnw
Minor wording changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@61 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\title{\vspace*{-0.4cm} A Generalization of Classical Measurement Invariance Tests for Factor Analysis}
-\Plaintitle{A Generalization of Classical Measurement Invariance Tests for Factor Analysis}
+\title{\vspace*{-0.4cm} Generalized Measurement Invariance Tests with
+  Application to Factor Analysis}
+\Plaintitle{Generalized Measurement Invariance Tests with Application
+  to Factor Analysis}
 \Shorttitle{Generalized Measurement Invariance Tests for Factor Analysis}
@@
 an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
-maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. However, \cite{And93}
-showed that the asymptotic distribution is in fact tractable, but nonstandard.
+maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. \cite{And93}
+showed that the asymptotic distribution is in fact tractable but nonstandard.
 Specifically, the asymptotic distribution of (\ref{eq:maxlr}) is the maximum of a
@@
 Having estimated the parameters $\hat {\bm \theta}$, the goal is to check that
-there are no subgroups of individuals that fail to conform with the model (for any
+all subgroups of individuals conform with the model (for all
 of the parameters). Hence, some measure of model deviation or residual is required
@@
 change literature (see e.g., \citealp{Zei05,ZeiSha10}) but the three
-tests above are probably most likely to be useful in psychometric settings.
+tests above are most likely to be useful in psychometric settings.
 
@@
 
-% \subsection{Further issues}
-% Further issues associated with the proposed tests include: (1)
-% alternative methods for estimating $\eta$;
-% and (2) dealing
-% with categorical $V$.  We address each in turn below.
-
-% % More than two groups violating MI
-% \subsubsection{Estimation of $\eta$}
-% Given rejection of the hypothesis of measurement invariance, the most
-% straightforward way to formally estimate $\eta$ is through
-% the location of the maximum of the cumulative score process.  Assuming
-% a single threshold, the maximum should be achieved close to the true
-% $\eta$.  Referring to the example (see
-% Figure~\ref{fig:cusumex}), we defined $\eta$ as the 
-% age corresponding to the solid dot in the third panel.
-
-% To study this method in more detail, we used
-% results of our simulation to examine the age at which the maximum of
-% the cumulative score process occurs.
-% If the test accurately recovers the violating subgroups,
-% then the maximum should be close to 16 years
-% of age.\footnote{The data generated for these
-%   simulations 
-%   contained ages measured in days, as opposed to years.  Thus,
-%   the tests could split students at continuous values like 15.6 or
-%   16.3 years of age.}
-%   For each simulation condition, Table~\ref{tab:age1}
-%   displays confidence intervals for the location of the maximum
-%   absolute cumulative score.  These confidence intervals were
-%   calculated across generated 
-%   datasets, with each dataset contributing one maximum absolute cumulative
-%   score
-%   (regardless of whether the hypothesis of
-%   measurement invariance was rejected).
-% It is observed that these confidence intervals slightly underestimate
-% the true age dividing the subgroups, with the degree of bias
-% decreasing as sample size increases.  The interval widths
-% are impacted by both the magnitude of the invariance and the sample
-% size.  In future work, we plan to study alternative ways of
-% identifying subgroups.
-
-% \begin{table*}
-% \caption{Confidence intervals for the location of the maximum absolute
-%   cumulative score across the conditions from the simulation.}
-% \label{tab:age1}
-%   \begin{tabular}{lccc} \thickline
-%   & \multicolumn{3}{c}{Violation Magnitude} \\
-%   $n$ & 1SE & 2SE & 3SE \\\hline
-%   100 & (15.61; 15.73) & (15.55; 15.65) & (15.50; 15.59) \\
-%   200 & (15.67; 15.77) & (15.64; 15.72) & (15.67; 15.73) \\
-%   500 & (15.71; 15.81) & (15.77; 15.83) & (15.80; 15.83) \\\thickline
-%   \end{tabular}
-% \end{table*}
-
-% Differing proportions of people in each group
-%\subsubsection{Subgroup proportions}
-%The power of the proposed tests is impacted by the proportion of
-%individuals within each subgroup.
-
-% Ties on auxiliary variable (mention use in categorical contexts)
 \subsection{Categorical auxiliary variable}
-% There remain unaddressed theoretical issues specific to the 
-% tests' application to measurement invariance.  These
-% issues generally stem from the types of auxiliary variables on which
-% researchers may study measurement invariance.  In
-% the example described earlier, researchers wanted to know whether a
-% scale was measurement invariant with respect to student age.
-% Students could then be ordered based on age, and a test could be 
-% constructed via Equation~\eqref{eq:cumscore}.  However, an issue arises if there
-% are ties on age; that is, if two students' ages are exactly
-% the same.  The proposed tests assume complete ordering of the
-% students, but these two students cannot be ordered.  In this case, we
-% may randomly order the two students, but there is an
-% arbitrariness associated with this strategy.  Alternatively, to obtain
-% tests that liberally reject the hypothesis of measurement invariance,
-% we may choose the ordering that results in the largest test
-% statistic.  If there are many tied students, then there are many
-% possible orderings that must be examined.  However, examination of all
-% possible orderings remains computationally feasible because the
-% casewise first derivatives need to be calculated only once; the
-% derivatives simply need to be reordered for each ordering of tied
-% students.
-
 One issue that was largely unaddressed in this paper involved the use
@@
 
-% Recursive method
-% \subsubsection{Recursive Multiple-Groups Models}
-% The proposed tests have the ability to automatically define groups of
-% individuals violating measurement invariance with respect to a single
-% auxiliary (often demographic) variable.  If a researcher has multiple
-% auxiliary variables that he/she wishes to assess for measurement
-% invariance, then the proposed tests can be inserted into a recursive
-% algorithm 
-% that isolates groups of individuals violating measurement
-% invariance.   and
-% C4.5 \citep{Qui93}.  The algorithm would proceed via the following
-% steps: 
-% \begin{enumerate}
-%   \item Fit a confirmatory factor analysis model to the data, with all
-%   parameters constrained to be equal across observations.
-%   \item Conduct a measurement invariance test with respect to each
-%     auxiliary variable (auxiliary variables are determined in
-%     advance by the researcher).
-%   \item If violations of measurement invariance are detected,
-%   split the data into two (or more) groups based on the auxiliary variable
-%     corresponding to the largest test statistic.
-%   \item Fit a multiple-groups factor analysis model based on step 3.
-%   \item Repeat steps 2--5.
-% \end{enumerate}
-% The algorithm would continue until it reaches a
-% predefined stopping rule.  For example, we may stop 
-% the algorithm when the tests fail to indicate measurement invariance
-% violations.  Alternatively, the algorithm may be stopped when the
-% sample size within any subgroup reaches a minimum level.
-
-% The algorithm described in the previous paragraph is related to one
-% developed by \cite{ZeiHot08} for representing
-% parameter invariance violations in generalized linear models.  Zeileis
-% et al.\ present the results in a tree structure, which is
-% suboptimal for factor analysis models.  Instead, the subgroups
-% discovered by the algorithm are used here to adaptively define a
-% multiple-groups factor analysis model.  The
-% simultaneous study of measurement invariance with respect to multiple
-% auxiliary variables is an important future direction.
-
-%\subsubsection{Other Latent Variable Models}
 The proposed tests also 

### 2011-11-07 www/efpFunctional-cat.R
Added two efpFunctional-generating functions for dealing with categorical covariates.
catL2BB() implements Equation 21 (which lacks squares!)

ordL2BB() implements the analogous statistic to Equation 17 for
ordered categorical variables. The simple idea is to not take the maximum
over all i but only those at the "end" of a category.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@69 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,121 @@
+catL2BB <- function(freq)
+{
+  if(inherits(freq, "gefp")) freq <- as.factor(time(freq)[-1L])
+  if(is.factor(freq)) freq <- prop.table(table(freq))
+  freq <- freq / sum(freq)
+  ncat <- length(freq)
+  
+  catdiffL2 <- function(x) {
+    d <- diff(c(0, x[round(cumsum(freq) * length(x))]))
+    sum(d^2 / freq)
+  }
+  
+  computePval <- function(x, nproc = 1) pchisq(x, df = (ncat - 1) * nproc, lower.tail = FALSE)
+  computeCritval <- function(alpha, nproc = 1) qchisq(alpha, df = (ncat - 1) * nproc, lower.tail = FALSE)
+
+  efpFunctional(lim.process = "Brownian bridge",
+    functional = list(time = catdiffL2, comp = sum),
+    computePval = computePval,
+    computeCritval = computeCritval)
+}
+
+ordL2BB <- function(freq, nobs = NULL, nproc = NULL, ...)
+{
+  if(inherits(freq, "gefp")) {
+    if(is.null(nproc)) nproc <- NCOL(freq$process)
+    freq <- prop.table(table(time(freq)[-1L]))
+  } else if(is.factor(freq)) {
+    if(is.null(nproc)) nproc <- 1:20
+    freq <- prop.table(table(freq))
+  } else {
+    if(is.null(nproc)) nproc <- 1:20
+  }
+  freq <- freq / sum(freq)
+  ncat <- length(freq)
+  tcat <- cumsum(freq[-ncat])
+  if(is.null(nobs)) nobs <- 50 / min(freq)
+
+  catwmax <- function(x) {
+    n <- length(x)
+    tt <- seq(along = x)/n
+    ix <- round(tcat * n)
+    x <- x[ix]
+    tt <- tt[ix]
+    x <- x/(tt * (1-tt))
+    return(max(x))
+  }
+
+  boundary <- function(x) rep(1, length(x))
+
+  plotProcess <- function(x, alpha = 0.05, aggregate = TRUE,
+    xlab = NULL, ylab = NULL, main = x$type.name, ylim = NULL, type = "b", ...)
+  {
+    n <- x$nobs
+    ix <- round(tcat * n)
+    tt <- ix/n
+
+##    bound <- computeCritval(alpha = alpha, nproc = NCOL(x$process)) * boundary(tt)
[…]

### 2011-11-07 www/paper.Rnw
Deleted the file paperR1.Rnw and instead moved all changes to the original file, paper.Rnw.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@70 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}
+\SweaveOpts{engine = R, eps = TRUE, echo = FALSE, results = hide}
 
@@
 \twoauthors{Edgar C. Merkle}{Achim Zeileis}
-\twoaffiliations{University of Missouri--Columbia}{Universit\"{a}t Innsbruck}
+\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}
 
@@
 
-\acknowledgements{Correspondence to Edgar C.\ Merkle, Department of 
-  Psychological Sciences, University of Missouri--Columbia, Columbia,
+\acknowledgements{This work was supported by National Science
+  Foundation grant SES-1061334. Correspondence to Edgar C.\ Merkle,
+  Department of  
+  Psychological Sciences, University of Missouri, Columbia,
   MO 65211.
-  Email: {\texttt{ecmerkle@gmail.com}}.}
+  Email: {\texttt{merklee@missouri.edu}}.}
 \shorttitle{Generalized measurement invariance tests}
@@
 source("mz.R")
+source("wdh.R")
 source("sim.R")
@@
 ${\bm \theta}(V)$ is not completely specified unless the cutpoint $\nu$ is known
-in advance.  In this situation, all individuals can be grouped based on $V$, and
+in advance (i.e., unless there is observed heterogeneity).  In this situation, all individuals can be grouped based on $V$, and
 we can apply standard theory: nested multiple group models (e.g.,
@@
 is unknown (as is often the case for continuous $V$), however, then
-standard theory 
+there is unobserved heterogeneity and standard theory 
 is not easily applied.   Nonstandard inference methods, such as those
@@
 \end{equation}
-which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
-of parameters in~${\bm \theta}$.
+which, when \eqref{eq:h1*} holds, has an asymptotic $\chi^2$ with degrees
+of freedom equal to the number of parameters in~${\bm \theta}$.
 
@@
 For application to measurement invariance, the most important
-theoretical result involves the fact that, under $H_0$, the
+theoretical result involves the fact that, under the $H_0$ in \eqref{eq:h0}, the
 \emph{cumulative score process} converges to a specific asymptotic
@@
 
-In the next section, we present an example involving a single data set
-drawn from a 
-known factor analysis model.  Following the example, we consider a
-full simulation study involving the same factor analysis model.
-
-\begin{figure}
-\caption{Path diagram representing the base factor analysis model used for
-  the example and simulations.  To induce measurement invariance
-  violations, a seventh observed variable (student age) determines the
-values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
-$\lambda_{31}$).}
-\label{fig:famod}
-\includegraphics[height=5in]{famod.pdf}
-\end{figure}
-
-\section{Example with Artificial Data}
-
-Consider the factor analysis model in Figure~\ref{fig:famod}, which
-includes six manifest variables, two 
-correlated factors, and a single auxiliary variable $V$.  Though an
-applied framework is unnecessary for the example, we may imagine that
-six scales have been administered to students aged 13 to 18
-years, with three of the scales intended to measure verbal ability and
-three of the scales intended to measure mathematical ability.  We
-assume a measurement invariance violation for the factor loading
-parameters, whereby the factor loadings for individuals with low
-values of $V$ are smaller than the factor loadings for individuals
-with high values of $V$ (e.g.,  
-factor loadings for older students are larger than those for younger
-students).  We generally wish to assess measurement invariance with
-respect to $V$ (student age).
+In the next section, we employ the proposed tests to analyze existing
+data from Wicherts, Dolan, and Hessen \citeyear{WicDol05}.
+Following the example, we consider a
+full simulation study involving artificial data.
+
+\section{Example: Stereotype Threat}
+
+Wicherts, Dolan, and Hessen \citeyear{WicDol05} utilized confirmatory
+factor models to study measurement invariance in a series of
+general intelligence scales.  They were interested in the
+notion of {\em{stereotype threat}}, whereby stereotypes concerning a
+subgroup's ability adversely impact the subgroup's performance on tests of that
+ability.  The authors specifically focused on stereotypes concerning
+ethnic minorities' performance on tests of general intelligence.  In this section, we used the proposed tests to supplement the authors' original analyses.
+
+\subsection{Background}
+
+To study stereotype threat in a measurement invariance framework,
+Study 1 of Wicherts et al.\ \citeyear{WicDol05} involved 295 high school
+students completing three intelligence tests in two between-subjects
+conditions.  Conditions were defined by
+whether or not students received primes about ethnic stereotypes prior to testing.
+To study the data, Wicherts et al.\ employed a four-group,
+one-factor model with the three intelligence tests as manifest variables.
+The groups were defined by ethnicity (majority/minority) and by the
+experimental manipulation (received/did not receive stereotype prime).
+Results indicated that the intelligence tests lacked measurement
+invariance, with the minority group receiving stereotype primes being
+particularly different from the other three groups on the most
+difficult intelligence test.  Most notably, the factor loading and
+intercept for the "minority, stereotype prime" group were smaller, and the unique variance
+was larger.
+
+As Wicherts et al.
+describe, one explanation for this pattern of results involves the idea that ability is positively related to "domain identification" \cite{Ste97}: individuals who are high on ability may be more prone to stereotype threat because they are more "tied in" to the domain of interest and are more aware of stereotypes.
+To test this idea, we employ a four-group, confirmatory factor model used by Wicherts et al.\ in tandem with the tests proposed in this paper.
+Our $V$ variable is participants' grade point averages (which were unused in the original analyses), which will reflect participants' identification with the academic domain.
 
 \subsection{Method}
-
-The base factor analysis model, displayed in
-Figure~\ref{fig:famod} with set parameter values, specifies that
-measurement invariance holds.
-For the measurement invariance violation, we specify that $V$
-(student age) impacts the values of 
-verbal factor loadings in the model: 
-if students are 16 through 18 years of age, then the factor
-loadings corresponding to the first 
-factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
-Figure~\ref{fig:famod}.  If students are 13 through 15 years of age,
-however, then the factor loadings corresponding to the first factor
-are three standard errors ($=$ asymptotic standard errors divided by $\sqrt{n}$)
-lower than those in Figure~\ref{fig:famod}.
-This violation states, e.g., that the verbal ability scales
-lack measurement invariance with respect to age.  For simplicity, we
-assume that the mathematical scales are invariant.
-
-<<dgp>>=
-set.seed(1090)
-d <- dgp(200, 2)
-@
-
-A sample of size 200 was generated from the model described above, and
-a test was conducted to examine measurement invariance of the three
-verbal scales.  To carry out
-the test, a confirmatory factor analysis model (with the paths
-displayed in Figure~\ref{fig:famod}) was fit to the
-data.  Casewise derivatives and 
+Wicherts et al.\ tested a series of confirmatory factor models for
+various types of measurement invariance.  We focus on the model used
+in their Step 5b (see pp.\ 703--704 of their paper), which involved
+across-group parameter restrictions on the factor loadings, unique
+variances, intercepts, and factor variances.  These parameters were
+typically restricted to be equal across groups, though a subset of the
+parameters were allowed to be group-specific upon examination of
+modification indices.
+Most notably, the most difficult intelligence test's factor loading, unique variance, and intercept differed for the "minority, stereotype threat" group, as compared to the other groups.
+Factor means were also estimated for three of the groups.  The model
+provided a reasonable fit to the data, as judged by examination of
+$\chi^2$, RMSEA, and CFI statistics.  Wicherts et al.\ also had a Step
+6 (the final step), where they examined a model that was simpler than
+described above and that provided a slightly better fit to the data.
+This model restricted factor means to be equal across a subset of the
+groups, however, which was not helpful for our purposes (because we are focusing on a single group).
+
+The stereotype threat effect of interest is most relevant to the
+"minority, stereotype prime" group.  Thus, to test the idea that
+domain identification impacts the magnitude of stereotype threat
+effect, we focus on parameters specific to this group.  We test these
+parameters for measurement invariance across the GPA variable, with
+violations supporting the hypothesis that domain identification
+impacts the magnitude of the stereotype threat effect.  We also test
+the intercept parameter individually.  This is because, if domain
+identification does impact stereotype threat, it could be
+conceptualized as the error term having a nonzero expectation across
+individuals (i.e., individuals with high domain identification have a
+negative error expectation and individuals with low domain
+identification have a positive error expectation).  The expected value
+for the error term cannot be uniquely separated from the intercept, so
+that the estimated intercept would display a lack of measurement
+invariance (see, e.g., \citeNP{Mil11}, p.\ 77).
+
+To carry out
+the tests, the four-group confirmatory factor model described above
+was first fit to the data.  Casewise derivatives and 
 the observed information matrix were then obtained, and they were used
@@
 test various subsets of 
-parameters.  For example, if we suspected that the verbal factor loadings
-lacked measurement invariance, we could test
+parameters.  For the Wicherts et al.\ data, we first test
 \begin{equation}
     \label{eq:exh0}
-H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
-(\lambda_{0, 11}\ \lambda_{0, 21}\ \lambda_{0, 31}),\ i=1,\ldots,n,
-\end{equation}
-where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
-verbal factor loading parameters for student $i$. Thus, here only $k^* = 3$
-from the overall $k = 19$~model parameters (including means) are assessed.
-Alternatively, we can consider all $k^* = k = 19$~parameters, leading to a test
-of~\eqref{eq:h0}.  We consider both of these tests below.
+H_0:\ (\nu_{i,1,\text{MSP}}\ \lambda_{i,1,\text{MSP}}\
+\theta_{i,1,\text{MSP}}\ \alpha_{i,1,\text{MSP}}) = 
+(\nu_{0,1,\text{MSP}}\ \lambda_{0,1,\text{MSP}}\
+\theta_{0,1,\text{MSP}}\ \alpha_{0,1,\text{MSP}}),\ i=1,\ldots,n_{\text{MSP}},
+\end{equation}
+where $\nu_{i,1,\text{MSP}}$ reflects the intercept for student $i$ on intelligence
+test 1 in the ``minority, stereotype prime'' group, with other
+subscripts defined similarly.  Additionally, $\lambda$ is a factor loading, $\theta$
+is a unique variance, and $\alpha$ is a factor mean, and the
+hypothesis is taken only wrt the $n_{\text{MSP}}$ individuals in the
+``minority, stereotype prime'' group.  In words, the hypothesis states
+that the four group-specific model parameters are invariant with
+respect to GPA.
+
+We can construct a similar hypothesis involving only the intercept
+parameter
+\begin{equation}
+    \label{eq:exh0.2}
+H_0:\ \nu_{i,1,\text{MSP}}\ = \nu_{0,1,\text{MSP}},\ i=1,\ldots,n_{\text{MSP}},
+\end{equation}
+which, as described previously, tests a more specific theory of the
+manner in which domain identification impacts stereotype threat.
+We consider tests of both \eqref{eq:exh0} and \eqref{eq:exh0.2}
+below.
 
 \subsection{Results}
-In the results section, we first describe overall results.  We then 
-describe estimation of $\nu$ and 
-isolation of model parameters violating
-measurement invariance.
-
-<<mz_mod>>=
-if(file.exists("mz_mod.rda")) {
-  load("mz_mod.rda")
-} else {
-  ## full sample model
-  mz0 <- mzfit(d)
+<<<mz_mod>>=
+d <- dget("WichertsEtAl2005_Study1.txt")
+
+wdh0 <- wdhfit(d)
+
 ## subsample models
-  lrstat <- function(age) 2 * as.vector(logLik(mzfit(d[d$age <= age,])) + logLik(mzfit(d[d$age > age,])) - logLik(mz0))
-  mz.age <- sort(d$age[d$age >= quantile(d$age, 0.1) & d$age <= quantile(d$age, 0.9)])
-  mz.lrstat <- sapply(mz.age, lrstat)
-  save(mz0, mz.age, mz.lrstat, file = "mz_mod.rda")
-}
+## This could be used to compare sequential LRTs with the LM
+## statistic that tests all parameters simultaneously.
+## (This is not done in the data example.)
+##############################################################
+## lrstat <- function(gpa) 2 * as.vector(logLik(wdhfit(d[d$gpa <= gpa,])) + logLik(wdhfit(d[d$gpa > gpa,])) - logLik(wdh0))
+## wdh.gpa <- sort(d$gpa[d$gpa >= quantile(d$gpa, 0.1) & d$gpa <= quantile(d$gpa, 0.9)])
+## # Results in many "computationally singular" errors:
+## wdh.lrstat <- sapply(wdh.gpa, lrstat)
+## save(wdh0, wdh.gpa, wdh.lrstat, file = "wicherts_mod.rda")
 @
 
-\subsubsection{Overall Results}
-Test statistics for the hypotheses \eqref{eq:exh0} and~\eqref{eq:h0}
-are displayed in  
-Figure~\ref{fig:cusumex}.
+Test statistics for the hypotheses \eqref{eq:exh0}
+and~\eqref{eq:exh0.2} are displayed in Figure~\ref{cusumex}.
 Each panel displays a test 
-statistic's fluctuation across values of student age, with the first
-column containing tests of~\eqref{eq:exh0} and the second column
-containing tests of~\eqref{eq:h0}.  The solid
+statistic's fluctuation across values of student GPA, with the first
+column containing tests of~\eqref{eq:exh0} (testing $k*=4$ parameters)
+and the second column containing tests of~\eqref{eq:exh0.2} (testing
+$k*=1$ parameter).
+The solid
 horizontal 
@@
 
-The figures convey information about several properties
-of the tests. First, all three tests are more powerful (and in
-this example significant) if we test only those parameters 
-that are subject to instabilities. Conversely, if all 19~model
-parameters are assessed (including those that are in fact invariant),
-the power is decreased. This decrease, however, is less pronounced
-for the double-max test as it is more 
-sensitive to fluctuations among a small subset of parameters.
-
-Figure~\ref{fig:lrlm} 
-compares the $\max \mathit{LM}$ statistic (solid line) to the $\max \mathit{LR}$
-statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
-\eqref{eq:h0}.  The critical values for these two
-tests are identical, hence the single horizontal line.
-The figure
-shows that the two statistics are very similar to one another, with
-both maxima at the dotted vertical line.  This
-is generally to be expected, because the two tests are asymptotically
-equivalent.  The 
-$\max \mathit{LR}$~statistic cannot be obtained from the empirical fluctuation
-process, however, so the factor analysis model must be refitted before
-and after each of the possible threshold values $\nu$ (i.e., here
-$\Sexpr{2 * length(mz.lrstat)}$~model fits for $\Sexpr{length(mz.lrstat)}$~thresholds).
-
-\subsubsection{Estimation of $\nu$ and Parameter Isolation}
-As described above, the tests of~\eqref{eq:exh0} imply that the 
-verbal scales lack measurement invariance.  
-We can also use the tests to: (1)~estimate the 
-threshold $\nu$, and (2)~isolate specific parameters that violate
-measurement invariance.  
-For example, as described previously, estimates of $\nu$ can be obtained by
-examining the peaks in Figure~\ref{fig:cusumex}.  For all six panels
-in the figure, the peaks occur near an age of 16.1.  This agrees well
-with the true threshold of 16.0.
-
-As mentioned previously, the double-max test is advantageous because
-it yields information about individual 
-parameters violating measurement invariance.  
-That is, it allows us to examine whether or not individual parameters' 
-cumulative score processes lead one to reject the hypothesis of
-measurement invariance.
-Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
-verbal factor loadings, with the horizontal lines reflecting the
-critical value at $\alpha= 0.05$.
-  The figure
-shows that the third parameter (i.e., $\lambda_{31}$) crosses the
-dashed line, so we would conclude a measurement 
-invariance violation for the third
-verbal test.  The fact that the cumulative score process for the first
-and second loadings did not achieve the critical value represents a Type II
-error, which implicitly brings into question the tests' power.
-This is an example of a situation where the double-max test ``wastes''
-power, as it cannot make use of the fact that multiple model
-parameters change simultaneously.  We
-generally address the issue of power in the simulations below.
+%% Details on test results
+The results displayed in Figure~\ref{cusumex} generally show that the
+``minority, stereotype prime'' group lacks invariance with respect to
+GPA.  All three tests are significant at the .05 level when the four
+group-specific parameters are tested simultaneously (first column).
+When the intercept parameter is tested individually, two of the three
+tests achieve significance (the CvM test and $\max \mathit{LM}$
+statistics).  These results generally support the notion that domain
+identification (as measured by GPA) influences the magnitude of
+stereotype threat.  The $k*=1$ results also imply that the tests
+differ in power and/or in error rates, an issue we address later in the
+paper.
+
+%% add paragraph about peaks of graphs
+We can also use the test results to estimate the threshold $\nu$,
+which in this context can be used to define GPA-based subgroups whose
+measurement parameters differ.  As described previously, estimates of
+$\nu$ can be obtained by 
+examining the peaks in Figure~\ref{fig:cusumex}.  For the $k*=4$
+panels (first column), two peaks occur near GPAs of 6.0 and of 6.3 (6
+corresponds to the American `C' range).  These peaks roughly divide
+individuals into those receiving D's and F's, those receiving solid
+C's, and those receiving high C's, B's, and A's.  Parameter-specific
+results can also be obtained, as shown for the intercept parameter in
+the second column of Figure~\ref{fig:cusumex}.  Here, a single peak
+occurs in the three panels around a GPA of 7.0 (`B' range).
 
 \begin{figure}
-\caption{Three test statistics of~\eqref{eq:exh0} (with $k^* = 3$)
-  and~\eqref{eq:h0} (with $k^* = 19$), 
-  based on the example involving measurement invariance with respect
-  to student age.  Solid, horizontal lines represent critical values
+\caption{Three test statistics of~\eqref{eq:exh0} (set of
+  group-specific parameters)
+  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
+  using data from Study 1 of Wicherts et al.\ (2005).
+  Solid, horizontal lines represent critical values
   at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
@@
 \setkeys{Gin}{width=\textwidth}
-<<gefp-3-19, fig=TRUE, height=8, width=6>>=
-## empirical fluctuation processes (3 vs all 19 pars)
-gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
-  vcov = info.mzfit, sandwich = FALSE)
-gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
-  vcov = info.mzfit, sandwich = FALSE)
+<<gefp-4-1, fig=TRUE, height=8, width=6>>=
+## add small amount of random noise to duplicated GPAs, so that everything
+## works with zoo.  (Without this, one can still obtain DM and CvM test 
+##                   results but not supLM results.)
+set.seed(1090)
+ties <- duplicated(d$gpa)
+d$gpa[ties] <- runif(sum(ties), d$gpa[ties]-.01, d$gpa[ties]+.01)
+## empirical fluctuation processes (4 vs 1 par)
+gefp_4_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
+gefp_1_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)
 
@@
 par(mfcol = c(3, 2))
-plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
-plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
+plot(gefp_4_info,  functional = maxBB, main = "DM, k* = 4", xlab = "GPA")
+plot(gefp_4_info,  functional = meanL2BB, main = "CvM, k* = 4", xlab = "GPA")
+plot(gefp_4_info,  functional = supLM(0.1), main = "max LM, k* = 4", xlab = "GPA")
+plot(gefp_1_info, functional = maxBB, main = "DM, k* = 1", xlab = "GPA")
+plot(gefp_1_info, functional = meanL2BB, main = "CvM, k* = 1", xlab = "GPA")
+plot(gefp_1_info, functional = supLM(0.1), main = "max LM, k* = 1", xlab = "GPA")
 @
@@
 
+\subsection{Discussion}
+Application of the proposed measurement invariance tests to the
+Wicherts et al.\ \citeyear{WicDol05} data allowed us to study the
+domain identification's impact on stereotype threat in a
+straightforward manner.  The tests focused on a set of group-specific
+parameters within a four-group confirmatory factor model, and they
+could be carried out using the results from a single estimated model.
+The latter fact is important, as other approaches to the problem
+described here may adversely impact model fit and degrees of freedom (e.g., if
+GPA is inserted directly into the model) or require multiple models to
+be estimated (e.g., for LRTs).  Further, through examination of both
+test statistics and cumulative score processes, the tests were easily
+interpretable from both a theoretical and applied standpoint.  In the
+next section, we study the proposed tests' power via simulation.
+
 \begin{figure}
-\caption{A comparison of the $\max \mathit{LM}$ (solid line) and $\max \mathit{LR}$ (dashed line) test
-  statistics for~\eqref{eq:h0} (i.e., $k^* = 19$).  The solid horizontal line
-  corresponds to the critical value at $\alpha= 0.05$ while the dotted vertical
-  line highlights the threshold at which both test statistics assume their maximum.}
-\label{fig:lrlm}
-\setkeys{Gin}{width=\textwidth}
-<<lm-lr-19, fig=TRUE, height=5, width=6>>=
-plot(gefp_19_info, functional = supLM(0.1),
-  xlab = "Age", ylab = "LR and LM statistics (k* = 19)", main = "", ylim = c(0, 50))
-legend("topleft", c("LR", "LM"), lty = c(2, 1), bty = "n")
-lines(mz.age, mz.lrstat, lty = 2)
-abline(v = mz.age[which.max(mz.lrstat)], lty = 3)
-@
+\caption{Path diagram representing the base factor analysis model used for
+  the example and simulations.  To induce measurement invariance
+  violations, a seventh observed variable (student age) determines the
+values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
+$\lambda_{31}$).}
+\label{fig:famod}
+\includegraphics[height=5in]{famod.pdf}
 \end{figure}
 
-\begin{figure}
-\caption{Cumulative score processes for each verbal factor
-  loadings, with critical values stemming from the double-max test. 
-  The solid, horizontal lines
-  correspond to the critical value at $\alpha= 0.05$.}
-\label{fig:lambdas}
-\setkeys{Gin}{width=\textwidth}
-<<gefp-lambdas, fig=TRUE, height=8, width=6>>=
-plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
-  main = "", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
-@
-\end{figure}
-
 \section{Simulation}
@@
 In this section, we conduct a simulation designed to
-examine the tests' power and Type I error rates in the context also
-employed for the previous example.
-We examine the power and error rates of three tests: the double-max test, the
+examine the tests' power and Type I error rates in situations where
+measurement invariance violations are known to exist.  Specifically,
+we generated data from the factor analysis model displayed in
+Figure~\ref{fig:famod}, with measurement invariance violations in the
+``verbal'' factor loadings with respect to a continuous auxliary variable.
+For simplicity, we
+assume other model parameters are invariant.  We also employ a
+single-group model to reflect situations where there exist only continuous
+$V$, so that no groups can be predefined.
+
+We examine the power and error rates of the three
+tests used in the Example section: the double-max test, the
 Cram\'{e}r-von Mises test, and the $\max \mathit{LM}$ test.  We also compare 
-tests involving only the $k^* = 3$ parameters that changed with tests of all
-$k^* = k = 19$ model parameters.  The example implied that the former tests had more
-power, and the simulations provide more detail on the extent to which
-the tests' power levels differ.
+tests involving only the $k^* = 3$ factor loadings lacking invariance
+ with tests of all
+$k^* = k = 19$ model parameters.  It is likely that power
+is higher when testing only the affected model parameters, but
+the affected parameters are usually unknown in practice.
 Finally, we also examine the tests' power across
@@
 \subsection{Method}
-Data were generated analogously to the example section.
-The data were generated from a factor analysis model with two
-correlated factor and six manifest variables, with individuals with
-low $V$ (13--15) having smaller factor loadings than individuals with
-large $V$ (16--18).
+Data were generated using the model in Figure~\ref{fig:famod}, with
+$V$ being designed to reflect student ages.
+For the measurement invariance violation, ``young'' students (13--15)
+had smaller factor loadings 
+than ``old'' students (16--18).
 Sample size and magnitude
@@
   invariance violations, and two subsets of tested
-  parameters $k^*$. Abbreviations: CvM = Cram\'{e}r-von Mises 
-  test; $\max \mathit{LM}$ = Maximum Lagrange multiplier test; DM = Double-max test.
+  parameters $k^*$. 
   See Figure~\ref{fig:simres} for a visualization (using all 17~violation magnitudes).}
@@
 mz_tab[c(1, length(mz_tab))] <- paste(mz_tab[c(1, length(mz_tab))], "\\hline")
+mz_tab <- c(mz_tab,"\\multicolumn{12}{l}{{\\scriptsize{Abbreviations: CvM = Cram\\'{e}r-von Mises 
+  test; $\\max \\mathit{LM}$ = Maximum Lagrange multiplier test; DM = Double-max test.}}}")
 writeLines(mz_tab)

### 2011-11-08 www/paper.Rnw
Addressed a number of reviewer comments pertaining to factor mixture models, moderated factor models, and some other miscellaneous issues.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@71 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \subsection{Use in Practice}
-% Simultaneous test of Meredith's various types of invariance
 The proposed tests give researchers a new set of tools for studying 
-measurement invariance.  For example, they give researchers the
-flexibility to: (1)~simultaneously test all model parameters, yielding
+measurement invariance, allowing them the 
+flexibility to: (1)~simultaneously test all model parameters across
+all individuals, yielding
 results relevant to many types of measurement
-invariance (see, e.g., \citeNP{Mer93}), or (2)~test a single subset of
-model parameters, potentially leading to improved power to detect a
-single type of measurement invariance.  The traditional
-steps have involved hypothesizing a specific type of invariance and
-then testing for it via a LR~test, but this is
-unnecessary under the proposed framework.
-
-% Researchers interpreting subgroups violating MI
-In addition to simultaneously testing different types of measurement
-invariance, the proposed tests allow researchers to easily interpret
-the nature of the invariance violation.  This is made possible through
-the tests' abilities to estimate $\nu$, the threshold dividing
+invariance (see, e.g., \citeNP{Mer93}), (2)~test a subset of
+model parameters, either across all individuals or within a
+multiple-group model, and (3)~use the tests as a type of modification
+index following significant LR~tests.
+Traditional steps to studying measurement invariance have involved
+sequential LR~tests for various types of invariance using
+multiple-group models.  As shown in the Example section, it can be
+beneficial to employ the traditional steps in tandem with the proposed
+tests.  Further, when groups are not defined in advance (e.g.,
+continuous $V$), the traditional steps fail unless further assumptions
+about the nature of the groups are made.
+
+In addition to providing general information about whether or not
+measurement 
+invariance holds, the proposed tests allow researchers to interpret
+the nature of the invariance violation.  This is made possible, e.g.,
+through the tests' abilities to estimate $\nu$, the threshold dividing
 individuals into subgroups that violate
-measurement invariance.  While a single $\nu$ was assumed in this
-paper, it is also possible to define 
-formal rules for estimating multiple $\nu$ parameters
-(see \citeNP{ZeiSha10}).  We plan to explore these issues in the future.
+measurement invariance (see Equation~\eqref{eq:h1*}).  It is also
+possible to define formal rules for estimating multiple $\nu$
+parameters (see \citeNP{ZeiSha10}).
+
+\subsection{Comparison to Other Methods}
+The step
+functions from~\eqref{eq:h18} were employed in the current paper to
+highlight connections with nested model
+comparison, but the proposed tests typically have non-trivial power for
+all (non-constant) deviation patterns $\theta(v_i)$ (considered in
+Equation~\eqref{eq:h1}). Thus, the tests will also have power for linear
+deviations from parameter constancy, although other techniques may 
+perform better in this particular case.  Examples include the 
+class of moderated factor models \cite{BauHus09,MolDol10,Pur02}, 
+whereby continuous moderators are allowed to linearly impact model
+parameters (Purcell also considers quadratic effects of the
+moderators).   
+Returning to deviations in parameter constancy, if the change in
+parameters is continuous but not exactly linear - e.g., a sigmoidal
+shift from one set of parameters to another one - then the simple
+single shift model in~\eqref{eq:h1*} may be a useful first
+approximation and have better power than linear
+techniques.
+
+% paragraph on factor mixture models
+The proposed tests are also similar to factor mixture models (e.g.,
+\citeNP{DolMaa98,LubMut05}) in that they can handle unknown subgroups
+violating measurement invariance.  Covariates can also be used to
+predict the unknown subgroups under both approaches.
+For example, for the 
+data used in our simulations, one could employ a factor mixture
+model with two latent classes.  This would typically involve the
+specification that student age has a linear effect on logit(P(Class
+2)), which highlights the main difference between the two techniques:
+the factor mixture model requires a specific assumption about the type
+of deviation from parameter constancy, while the proposed tests do
+not.
+
+More generally, all the approaches described above can be distinguished
+from the proposed tests in that they require estimation of a new model
+with increased complexity (due to the moderator/covariate).
+The proposed tests, on the other hand, are of the
+``posthoc'' variety, relying only on results calculated during the
+original model estimation.  While no method clearly dominates 
+across all situations, use of the proposed tests can at least reduce 
+some of the technical issues associated with estimating and
+interpreting models of greater complexity.  This may not be a
+good reason to use the tests, but it is a consideration that is often
+meaningful in practice.
 

### 2011-11-09 www/paper.Rnw
More reviewer comments addressed in paper.Rnw, including normality violations and local alternatives.  Also added a file containing data for the new data example.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@72 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 particularly important because violations can render the scales
-useless.  
+difficult to interpret.  
 That is, if a set of scales violates measurement invariance,
@@
 differences on a wide variety of interesting constructs 
-when, in reality, the scales are the sole cause of the 
+when, in reality, the scales impact the magnitude of the 
 differences.  Further, it can be inappropriate to incorporate scales
@@
 Within the general framework outlined above and under the usual
-regularity conditions, the model parameters $\bm{\theta}$ can
+regularity conditions (e.g., \citeNP{Fer96}), the model parameters
+$\bm{\theta}$ can 
 be estimated by maximum likelihood (ML), i.e.,
@@
 variable that the scales purport to measure, and $f$ is the model's
-distributional form.  In the parametric framework
-adopted here, this means that the parameter vector ${\bm \theta}$ (or
-some subset of ${\bm \theta}$; see \citeNP{Mer93}) is equal
-across subgroups of individuals and thus does not vary
-with any variable $V$.
+distributional form \cite{Mel89}.  In the parametric,
+factor-analytic framework
+adopted here, this implies that the measurement parameters are equal
+across subgroups of individuals and, thus, do not vary with any
+variable $V$ \cite{Mer93}.
 
@@
 model is correctly specified, as is implicitly assumed under traditional
-measurement invariance approaches.
+measurement invariance approaches.  In particular, violations of
+normality may lead to spurious results in the proposed tests just as
+they do in other approaches (e.g., \citeNP{BauCur04}).
 
@@
 and maximal thresholds $\underline{\nu}$ and $\overline{\nu}$, respectively.
-See below for more details.
+See Andrews for the original results and further references, and see 
+below for more details on the results' application to measurement invariance.
 
@@
 e.g., their outer product or the observed information matrix.
+The $k$ dimensions of the 
+process arise from the fact that a separate cumulative score is
+maintained for each of the $k$ model parameters.  
 As the equation shows, the cumulative score process adds subsets of
@@
 first individual's contribution enters into the summation; at
-$t=2/n$, the first two individuals' contributions etc.\ until $t = n/n$
-where all contributions enter the sum. Thus, due to (\ref{eq:ml1}),
+$t=2/n$, the first two individuals' contributions enter into the
+summation, etc., until $t = n/n$
+where all contributions enter into the summation. Thus, due to (\ref{eq:ml1}),
 the cumulative score process always equals zero at $t = 0$ and
@@
 we also denote ${\bm B}(\hat {\bm \theta})_{ij}$ below for brevity.
-Each column of this matrix converges to a univariate Brownian bridge and pertains
-to a single factor analysis parameter. To carry out a test of $H_0$, this process/matrix
+That is, assuming that individuals are ordered by $V$, the first row
+of ${\bm B}(\hat {\bm \theta})$ corresponds to the first individual's
+scores.  The second row of ${\bm B}(\hat {\bm \theta})$ corresponds to
+the sum of the first two individuals' scores, etc., until the last row
+of ${\bm B}(\hat {\bm \theta})$ corresponds to the sum of all
+individuals' scores (which will be a row of zeroes).
+Under this setup, each column of ${\bm B}(\hat {\bm \theta})$
+converges to a univariate Brownian bridge and pertains 
+to a single factor analysis parameter. To carry out a test of $H_0$, the process/matrix
 needs to be aggregated to a scalar test statistic by collapsing across
@@
 of this test statistic is then easily obtained by applying the same
-aggregation to the asymptotic process ${\bm B}^0$ \cite{HjoKon02,ZeiHor07},
+aggregation to the asymptotic Brownian bridge \cite{HjoKon02,ZeiHor07},
 so that corresponding $p$-values can be derived.
 
-As argued above, no single aggregation function will have high power for any
-conceivable pattern of measurement invariance ${\bm \theta}(V)$, while any
+As argued above, no single aggregation function will have high power for all
+conceivable patterns of measurement invariance ${\bm \theta}(V)$, while any
 (reasonable) aggregation function will have non-trivial power under $H_1$.
@@
 \end{equation}
-becomes large. Consquently, this double maximum statistic allows for simultaneous
-isolation of the threshold(s) of parameter change (over the individuals $i = 1, \dots, n$)
+becomes large.  In determining where in ${\bm B}(\hat {\bm \theta})$
+this maximum occurred, we are able to determine  
+threshold(s) of parameter change (over the individuals $i = 1, \dots, n$)
 and the parameter(s) affected by it (over $j = 1, \dots, k$). This test is
@@
 critical value.  An example of this visualization appears in the 
-example section (Figure~\ref{fig:lambdas}).
+example section (first row, second column of Figure~\ref{fig:cusumex}).
 
@@
 many of the $n$~individuals (and not just a single threshold). In such
-cases, sums insteads of maximums are more suitable for collapsing across
-parameters and/or individuals because they combine the deviations instead of
+cases, sums insteads of maxima are more suitable for collapsing across
+parameters and/or individuals, because they combine the deviations instead of
 picking out only the single largest deviation. Thus, if the parameter instability
@@
 asymptotic variance $t (1 - t)$ of the process ${\bm B}(t, \hat {\bm \theta})$.
-It is equivalent to the $\max_\nu \mathit{LM}(\nu)$ from the previous
-section (provided that the boundaries for the subgroups sizes $\underline{i}$/$\underline{\nu}$
-and $\overline{\imath}$/$\overline{\nu}$ are chosen analogously).
-
+It is equivalent to the $\max_\nu \mathit{LM}(\nu)$ statistic from the previous
+section, provided that the boundaries for the subgroups sizes $\underline{i}$/$\underline{\nu}$
+and $\overline{\imath}$/$\overline{\nu}$ are chosen analogously.
 Further aggregation functions have been suggested in the structural
@@
 
-\subsection{Critical Values \& $p$-values.}
+\subsection{Critical Values \& $p$-values}
 
 As pointed out above, specification of the asymptotic distribution under $H_0$ for the test
-statistics from the previous section is straightforward: It is simply the aggregation
+statistics from the previous section is straightforward: it is simply the aggregation
 of the asympotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
@@
 
-\subsection{Locating the Invariance.}
+\subsection{Local Alternatives}
+Using results from Hjort and Koning \citeyear{HjoKon02} and Zeileis
+and Hornik \citeyear{ZeiHor07}, we can also
+capture the behaviour of the process $\bm{B}(t, \hat{\bm{\theta}})$
+under specific alternatives 
+of parameter instability. In particular, we can assume that the pattern
+of deviation $\bm{theta}(v_i)$ can be described as a constant
+parameter plus some non-constant deviation $\bm{g}(i/n)$:
+\begin{equation}
+    \label{eq:local}
+  \bm{\theta}(v_i) = \bm{\theta}_0 + n^{-1/2} \bm{g}(i/n).
+\end{equation}
+In this case, the scores $\bm{s}(\bm{\theta}_0; x_i)$ from
+Equation~\eqref{eq:score} do not have zero expectation but rather
+\begin{equation}
+    \label{eq:localE}
+  E[\bm{s}(\bm{\theta}_0; x_i)] = \bm{0} + n^{-1/2}
+  \bm{C}(\bm{\theta}_0) \bm{g}(i/n).
+\end{equation}
+In this case, the covariance matrix $\bm{C}(\bm{\theta})$ from the expected
+outer product of gradients is
+\begin{equation}
+    \label{eq:localS}
+  \bm{C}(\bm{\theta}_0) = E[\bm{s}(\bm{\theta}_0; x_i) \bm{s}(\bm{\theta}; x_i)']    
+\end{equation}
+which, at $\bm{\theta}_0$, coincides with the expected information matrix.
+
+Under the local alternative described above, Hjort and Koning
+\citeyear{HjoKon02} show that the process
+$\bm{B}(t, \hat{\bm{\theta}})$ behaves asymptotically like
+\begin{equation}
+    \label{eq:localB}
+    \bm{B}^0(t) + \hat{\bm{I}}^{-1/2} \bm{C}(\hat{\bm{\theta}}) \bm{G}^0(t),
+\end{equation}
+i.e., a zero-mean Brownian bridge plus a term with non-zero mean driven by
+$\bm{G}^0(t) = \bm{G}(t) - t \bm{G}(1)$, where $\bm{G}(t) = \int_0^t
+\bm{g}(y) dy$.  Hence, unless the local
+alternative has $\bm{g}(t) \equiv \bm{0}$, the empirical process $\bm{B}(t,
+\hat{\bm{theta}})$ will 
+have a non-zero mean.  Hence, the corresponding tests will have non-trivial
+power (asymptotically).  We use these results in the Simulation
+section to describe the expected behavior of the Brownian bridge.
+
+\subsection{Locating the Invariance}
 
@@
 assume other model parameters are invariant.  We also employ a
-single-group model to reflect situations where there exist only continuous
-$V$, so that no groups can be predefined.
+single-group model, reflecting situations where there exist only continuous
+$V$ (so that no groups can be predefined).
 
@@
 years of age) and half had ``high $V$'' (e.g., 16--18 years of age).
+
+Using results discussed previously in the ``Local Alternatives''
+section, we can derive the expected behavior of the univariate 
+Brownian bridges for the parameters $\{\lambda_{11},
+\lambda_{21}, \lambda_{31}\}$.  For these parameters, 
+we use a simple step function $g(t) = \text{I}(t > 0.5)$ multiplied
+by the violation magnitude.  Thus, $G(t) = 
+\text{I}(t > 0.5) \times (t - 0.5)$ and $G^0(t) = \{\text{I}(t > 0.5) - 0.5 } \times t - \text{I}(t > 0.5) \times 0.5$,
+both again multiplied by the violation magnitude. Thus, the mean under the
+alternative is driven by a function that is triangle-shaped, with a peak at the
+changepoint $t = 0.5$ for the parameters affected by the change (and
+equal to zero for all remaining parameters).
 
@@
 The step
-functions from~\eqref{eq:h18} were employed in the current paper to
+functions from~\eqref{eq:h1*} were employed in the current paper to
 highlight connections with nested model

### 2011-11-10 www/paper.Rnw
Added an n=50 condition to the simulations:   - File mz_sim.rda now contains results from n=50   - Table 1/figure 3 in paper.Rnw were modified to include the     n=50 condition.
Also, there was an issue with the Figure 3 lines that I fixed
(involved pch and lty arguments for the xyplot key).



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@73 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 by the violation magnitude.  Thus, $G(t) = 
-\text{I}(t > 0.5) \times (t - 0.5)$ and $G^0(t) = \{\text{I}(t > 0.5) - 0.5 } \times t - \text{I}(t > 0.5) \times 0.5$,
+\text{I}(t > 0.5) \times (t - 0.5)$ and $G^0(t) = \{\text{I}(t > 0.5) - 0.5 \} \times t - \text{I}(t > 0.5) \times 0.5$,
 both again multiplied by the violation magnitude. Thus, the mean under the
@@
 mykey <- simpleKey(c("CvM", "max LM", "DM"), points = FALSE, lines = TRUE)
-mykey$lines$pch <- c(2, 3, 1)
-mykey$lines$lty <- c(1.2, 2, 1.2)
+mykey$lines$lty <- c(2, 3, 1)
+mykey$lines$lwd <- c(1.2, 2, 1.2)
 print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub,
@@
 mz_tab[c(1, length(mz_tab))] <- paste(mz_tab[c(1, length(mz_tab))], "\\hline")
-mz_tab <- c(mz_tab,"\\multicolumn{12}{l}{{\\scriptsize{Abbreviations: CvM = Cram\\'{e}r-von Mises 
-  test; $\\max \\mathit{LM}$ = Maximum Lagrange multiplier test; DM = Double-max test.}}}")
+mz_tab <- c(mz_tab,"\\multicolumn{2}{l}{{\\scriptsize{Abbreviations:}}} &
+\\multicolumn{10}{l}{{\\scriptsize{CvM = Cram\\'{e}r-von Mises 
+  test; $\\max \\mathit{LM}$ = Maximum Lagrange multiplier test;}}} \\\\")
+mz_tab <- c(mz_tab," & & \\multicolumn{10}{l}{{\\scriptsize{DM = Double-max test.}}}")
 writeLines(mz_tab)

### 2011-11-11 www/rev1/response.txt
Addressed more reviewer comments, including:   - Added commentary on the n=50 simulation condition   - Addressed the issue of identification conditions impacting     the test results
Also added a draft set of responses to reviewers in the rev1 directory.


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@74 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Editor

Comment E.1. While I share the reviewers generally positive view of your paper, I find myself in the unusual position of being less enthusiastic about the present version of the paper than the reviewers seem to be.  One reason for this is that I think that your method is much less practical than the reviewers do, chiefly because in practical applications of invariance work, it is unusual to encounter a case in which V is continuous. The one exception to this situation is when V is age, but even then, in the context of employment discrimination for example, legal considerations often demand an arbitrary split on 40 years, or some other such threshold (which is admittedly arbitrary, but is demanded by the law).

Response E.1. Though we did not highlight it in the paper, the methods can be used with categorical V.  We have also developed a new strategy for handling ordinal V and briefly describe it in the Discussion.  Additionally, we would argue that applications of invariance do not use continuous V because methods for handling continuous V are still emerging (e.g., the moderated factor models discussed by Reviewer 1).  We believe that our proposed methods will encourage invariance research involving continuous V.

Comment E.2. The second point is related to issues raised by Reviewers one and four.  In any invariance analyses, some elements of theta are allowed to vary across groups freely because they are not part of the invariance constraints dictated by equation 7.  In the factor model, the factor means and factor covariance elements are not constrained.  How is this fact represented in your method?  While I saw your treatment of partial invariance (i.e., hold some elements to invariance while letting other elements go free), you seem in that context to be testing invariance in all of the free elements.  Does your method allow one to permit some elements to be unconstrained, while others are
invariant, without looking for violations of invariance in the free elements?  If it does, we need to see how this works.

Response E.2. This comment focuses on the fact that our methods generally work from a "reduced model" standpoint, whereby all model parameters are constrained to be equal and we examine the stability of parameter estimates across individuals (or across groups).  When groups are not known in advance (e.g., when V is continuous), this appears to be the only way to proceed.

However, if groups are known in advance, it is possible to employ the proposed methods in tandem with a multiple-group model.  The multiple-group model can then handle both unconstrained and invariant parameters, because scores can be computed for all model parameters.  We now illustrate this approach in the data example. DONE

Comment E.3. You also did not mention how model identification constraints, which in the invariance case are sometimes invariance constraints, are handled.

Response E.3. The proposed tests are based on scores associated with all free parameters.  Thus, it is straightforward to handle parameters that are constrained to be equal to one another.  However, this comment may also refer to the fact that one's choice of identification constraint will influence tests of other free parameters.  We now briefly describe this issue in the "Use in Practice" section of the Discussion.  DONE

Comment E.4. I think that we need to have a real example, as suggested by Reviewer one.  This would allow you to answer my second point above, and also show that your method can give answers that are sensible in context.  With a real example, you could drop the artificial one (but keep the simulations). Having a real example would also help to illustrate that there are real applications in which your method would prove useful.

Response E.4. We have dropped the artificial example and added a real example. DONE

Reviewer #1

Comment 1.1. The section on the asymptotic properties of the Beta-function on pages 11-14 was still quite hard to follow. The authors may want to expand upon some of the more technical aspects of these models to allow more readers to grasp how they would do the job in their given application.

Response 1.1. We have revised the theory development and added extra detail to the text.  If some specific concepts are still unclear, we are happy to revisit this section.  DONE

Comment 1.2. There are clear links between the current approach based on an observed continuous grouping variable V and the factor mixture approach. In factor mixture modeling, the grouping variable that characterizes the subgroups that differ in terms of particular model parameters are also typically unknown (i.e., latent classes), but class membership can be predicted on the basis of observed predictor(s). So the model in Figure 1 and the scenario of 2 subgroups that the authors describe on pages 17-20 is essentially a factor mixture model with two latent classes in which age predicts class membership. Therefore, the authors should take steps to relate their modeling approach to factor mixture modeling.

Response 1.2. We agree and have now described our methods' relationship to mixture models in the Discussion.  Some advantages of our approach include the facts that it does not require a priori specification of the number of latent classes or a priori specification of the way by which age impacts class membership.  DONE

Comment 1.3. Similarly, there exist a family of moderation-type factor models that may serve the same purpose of studying failures of invariance in common factor models with respect to (unknown) "grouping" variables. For instance, Dylan Molenaar, Conor Dolan and colleagues have recently worked out CFA models that were based on Shaun Purcell's work in which certain parameters are moderated by either observed or unobserved variables. These models too could be applied to the important problems related to invariance with respect to unknown groups although they may not work as well as do mixture models and the current approach when indeed there is some step function that describes the violations of invariance. The existence of these alternative approaches by no means diminishes the potential value of the current developments, but it does mean that the authors should discuss the relation between their approach and other existing methods.

Response 1.3. We now describe factor analysis models employing moderation in the "Comparison to other methods" section of the discussion.  DONE

Comment 1.4. The authors chose to simulate data only under models that did not include the mean structure. Since violations of measurement invariance with real data are most often related to the mean structure (to be sure due to larger power associated with modeling first-order moments), it may be more realistic and interesting to have scenarios in which such types of failures of invariance (called uniform measurement bias) occur.

Response 1.4. Mean parameters were estimated and tested in the simulation models, though they were unstructured.  Mean structures are often made possible by constraining parameters across groups, and our simulation focuses on a case where groups cannot be predefined.  Thus, we have not modified the model in the simulation.  In the real data example that we have added, however, the mean structure is utilized. DONE

Comment 1.5. Also, many "nuisance" variable that are hypothesized to affect indicators, say, in cognitive testing (e.g., see Wicherts et al.'s work on measurement invariance and stereotype threat) in a given multi-group CFA are expected to linearly affect these indicators. Now I can come up with a host of scenarios in which the measurement bias is approximately linear and so can be modeled in simple terms because the regression on V results in an additional factor in the model.  Surely I don't expect the authors to deal with all of such potential scenarios' of failures of measurement invariance in CFA. However, I do wish to point out that underlying the current approach is a specific kind of measurement bias that is expected to be discontinuous while many failures of invariance are expected to be continuous and approximately linear. Therefore, the authors should go into more detail on the nature of the measurement bias that the approach is trying to uncover. Their approach may be generic, but it is not really general in the sense that many failures of strict factorial invariance can be modeled in more straightforward and traditional ways. The step function in Eq. 10 should be discussed in more detail and may be supplemented by a hypothetical example of why it occurs in linear factor models.

Response 1.5. The step function from Equation 10 is employed as an illustrative example because it highlights the connection between nested model comparisons and the proposed techniques. Also, the maxLM statistic from Equation 17 has particularly high power for this scenario. However, the proposed tests typically have non-trivial power for all (non-constant) deviation patterns theta(v_i) (considered in Equation 9).  We have now provided more details on these issues in the Discussion section (and tied the details to Comment/Response 1.3).  DONE

Comment 1.6. On page 3, the authors state that failures of measurement invariance render scales useless, but this statement may be overly strong. Specifically, scales that are only partially invariant  may still be useful. In addition, the statement that "the scales are the sole cause of the differences" appears to suggest that failures of invariance can explain all group differences in observed test scores, which is seldom the case. Failures of measurement invariance do not preclude the existence of group differences in the distribution of latent factors scores.

Response 1.6. We agree that these statements were too strong and have revised accordingly.  DONE

Comment 1.7. When presenting the definition of invariance on page 5, the authors may wish to indicate that the definition (due originally to Mellenbergh, 1989) implies (as shown by Meredith, 1993) that not all of the parameters in the CFA are invariant but only the measurement parameters. So factor means, factor covariances, and (in structural models) factor regression weights are not expected to be invariant under this definition of invariance.
Another issue that could be discussed is that non-normality in the data may lead to spurious findings in the current approach just like they do in regular multi-group confirmatory analyses and in factor mixture modeling (e.g., see work by Gitta Lubke and Michael Neale).

Response 1.7. We now describe the invariance definitions in more detail, and we also clarify that our methods are not necessarily robust to violations of normality.  DONE

Comment 1.8. Finally, it may be interesting to apply the method to real data and to compare the outcomes from more traditional modeling approaches. After all, in real life the exact nature of the measurement bias with respect to V is not know. One interesting approach may be to use data from experiments like in the stereotype threat framework that was used by Wicherts et al. (who I am sure will be happy to share raw data for this purpose).

Response 1.8. See Response E.4.  DONE

Reviewer #3

Comment 3.1. The result from Hjort and Koning is a nice trick to rewrite the asymptotic representation of the likelihood estimator as an empirical process. This then relates the cumulative sums of the likelihood estimator to a Brownian bridge. The weak convergence of the monitoring process is of course asymptotic. The properties of the tests in the current manuscript are investigated decently enough for medium to large sample sizes (n = 100 to 500). However, for small samples no results are obtained. Given the relatively small model in the simulations, I would expect a relatively small sample size of n = 50 to be included in the simulations. The main reason is that this may show a point where the power functions show a "breakdown."

Response 3.1. We now include the n=50 condition in the simulations, which indeed exhibits a breakdown in the sense that power is not monotonic with respect to violation magnitude.  DONE

Comment 3.2. The weak convergence of the monitoring process is proved in the paper by Hjort and Koning in two cases: (i) for theta_0 the true parameter, and (ii)   for local alternatives O(1/sqrt(n)) from theta_0. In the latter case the monitoring process is shown to be a Brownian bridge including a constant (bias) even when using -E{u'u}, Fisher's information (which is consistent up to O(1/n)). I would imagine this is important since model misspecification (measurement invariance) is the topic of the manuscript. This result allows the authors to to determine exactly what the result of misspecification is. It fits well with their line of reasoning, which I agree with. The misspecification is determined by a deviation h/sqrt(n) of the true parameter. This corresponds exactly to the misspecification in the simulations. Measurement invariance can be imagined as a latent variable acting on the factor loadings, and can therefore be incorporated in the parameter theta. If there were no
measurement invariance, theta_v woold be 0, but in the case of misspecification, it is exactly the same as that of Hjort and Koning. I would recommend that the authors use this result by Hjort and Koning in the manuscript to show analytically what happens to the distribution and tests.

Response 3.2. We have now incorporated these results in a new section, titled "Local alternatives".  We also discuss these results in the context of the simulation, showing the impact of our imposed measurement invariance violation on the Brownian bridge.  DONE

We addressed all minor issues identified by Reviewer 3.  We have a response to one of these issues:

Comment 3.3. p. 7 second paragraph: Assuming that the model is correctly specified is bonus when the second point above is incorporated. I think this would improve the manuscript.

Response 3.3. Violations of distributional assumptions may still cause problems with the proposed tests, even if the "structure" of the model (e.g., parameter constraints, included variables, etc) is correct.  We have modified this section of text accordingly.  DONE

Reviewer #4

Comment 4.1. Page 6/7:	It could be emphasized that a known cutpoint ν means “observed heterogeneity” and an unknown cutpoint ν means “unobserved heterogeneity”.

Response 4.1. We have now added this detail.  DONE

Comment 4.2. Page 9:	The authors should give a reference or a short explanation of the “tied-down Bessel process”. Readers might be unfamiliar with the process.

Response 4.2. We now note that Andrews (1993) provides results and references relevant to Bessel processes.  DONE

Comment 4.3. Page 10:	To test for changes “along” V means that V is always continuous (also mentioned in the first paragraph). What happens if V is an ordered categorical variable? Would that be a special case? Adding a reference to the discussion section would be helpful.

Response 4.3. We have added some discussion on the use of categorical V.  Also see Response E.1.

Comment 4.4. Page 19:	The overall k=19 model parameter consists of six factor loadings, six measurement errors and one covariance between the factors. The authors refer to the six means of the manifest variables. Usually, in factor models mean differences of the factors are of research interest. The authors should give some information why they considered the manifest means instead of the latent mean differences.

Response 4.4. Factor means can be estimated if one is constraining other model parameters to be equal across groups.  In our new data example, we do estimate factor means and use them in the proposed tests.  In our simulation (and in the example that was removed), we use a scenario where no grouping variables exist (i.e., there is only a continuous V).  In this scenario, it seems difficult to estimate factor means prior to model estimation because we cannot constrain parameters across groups.  We have added some extra detail to the simulation section on our use of a single-group model.  DONE

Comment 4.5.	The simulations later on focus on invariance of factor loadings. Basicly, the covariance structure in the simulations are analyzed with the factor model of Figure 1. So, means are not included in the analyses (as far as I read the simulation results correctly).

Response 4.5. Mean parameters are included in the casewise likelihood function, though they were unmodeled in the simulation and thus omitted from Figure 1.  The casewise likelihood function is needed to make the proposed tests possible.  Also see Comment/Response 1.5.  DONE

Comment 4.6. Page 25: 	Implications of the study of measurement invariance in factor analysis to general structural equation models could be added with comments to mean differences (see comment for page 19).

Response 4.6. We have added some brief discussion surrounding extensions to structural equation models.  DONE

Comment 4.7. Table 1:	Table captions should be moved to the bottom of the table. (Abbreviations: …)

Response 4.7. We have moved the Abbreviations part of the captions.
DONE

### 2011-11-11 www/paper.Rnw
Addressed more reviewer comments, including:   - Added commentary on the n=50 simulation condition   - Addressed the issue of identification conditions impacting     the test results
Also added a draft set of responses to reviewers in the rev1 directory.


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@74 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+# Refit model with new identification constraints
+wdh0.1 <- wdhfit2(d)
+
 ## subsample models
@@
 <<gefp-4-1, fig=TRUE, height=8, width=6>>=
-## add small amount of random noise to duplicated GPAs, so that everything
-## works with zoo.  (Without this, one can still obtain DM and CvM test 
+## add a small amount of random noise to duplicated GPAs, so that everything
+## works with gefp().  (Without this, one can still obtain DM and CvM test 
 ##                      results but not supLM results.)
@@
 gefp_1_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)
+
+# processes for second identification constraints
+gefp_4_info_1 <- gefp(wdh0.1, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
+gefp_1_info_1 <- gefp(wdh0.1, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)
 
@@
 we examined power to detect 
-invariance violations across three sample sizes ($n=100,
+invariance violations across four sample sizes ($n=50, 100,
 200, 500$) and 17~magnitudes of violations.  These violations
@@
 Cram\'{e}r-von Mises test having the most power, followed by $\max \mathit{LM}$,
-followed by the double-max test.  This advantage decreases with
-increases in sample size.  
-Table~\ref{tab:simres} presents the same results as
+followed by the double-max test.  
+This advantage decreases at larger sample sizes.  
+At $n=50$, generally regarded as a
+small sample size for factor analysis, the power functions break down
+and are not monotonic with respect to violation magnitude.
+
+Table~\ref{tab:simres} presents a subset of the results displayed in 
 Figure~\ref{fig:simres}, but it is easier to see exact power
@@
   Cram\'{e}r-von Mises test (dashed), and $\max \mathit{LM}$ test (dotted)
-  across three sample sizes $n$, two subsets of tested parameters $k^*$, and
+  across four sample sizes $n$, two subsets of tested parameters $k^*$, and
   measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$). 
@@
 \caption{Simulated power for three test statistics
-  across three sample sizes $n$, nine magnitudes of measurement
+  across four sample sizes $n$, nine magnitudes of measurement
   invariance violations, and two subsets of tested
@@
 extension to more complex scenarios.
+
+<<<constraint_study>>=
+tests <- c("maxBB", "meanL2BB", "supLM(0.1)")
+
+p.diffs <- matrix(NA, 2*length(tests), 2)
+
+for (i in 1:length(tests)){
+  p.diffs[i,1] <- unlist(sctest(gefp_4_info, functional=
+                                eval(parse(text=tests[i])))[2])
+  p.diffs[i,2] <- unlist(sctest(gefp_4_info_1, functional=
+                                     eval(parse(text=tests[i])))[2])
+  }
+for (i in (length(tests)+1):(2*length(tests))){
+  p.diffs[i,1] <- unlist(sctest(gefp_1_info, functional=
+                                     eval(parse(text=tests[i-length(tests)])))[2])
+  p.diffs[i,2] <- unlist(sctest(gefp_1_info_1, functional=
+                                   eval(parse(text=tests[i-length(tests)])))[2])
+}
+@
 
@@
 about the nature of the groups are made.
+
+Another difference between the proposed tests and traditional tests
+involves the fact that, for the proposed tests, one's choice of
+identification constraints can influence tests of the remaining
+parameters.  This essentially reflects the fact that, while the model
+fits equally well regardless of identification constraint, the
+free model parameters adjust themselves to the specific identification
+constraint chosen.  To examine this issue in more detail, we refit the model
+used in the Example section, using an alternative identification
+constraint (i.e., fixing a different factor loading at 1).  Across the 6
+tests presented in the example section (see, e.g.,
+Figure~\ref{fig:cusumex}), the largest difference in p-values for the
+two identification constraints 
+was~$\Sexpr{round(max(abs(p.diffs[,2] - p.diffs[,1])),3)}$.  This is not a large difference, though further
+study is warranted.
 
@@
 from the proposed tests in that they require estimation of a new model
-with increased complexity (due to the moderator/covariate).
+of increased complexity (due to the moderator/covariate).
 The proposed tests, on the other hand, are of the
@@
 some of the technical issues associated with estimating and
-interpreting models of greater complexity.  This may not be a
+interpreting models of greater complexity.  This alone is not a
 good reason to use the tests, but it is a consideration that is often
@@
 readily extend to other popular psychometric models.  
-Instead of studying measurement invariance in factor
-analysis, the tests can be used to generally 
+For example, the tests can be used to generally 
 study the stability of structural equation model parameters across
-observations.  For 
-example, it would be possible to assess whether paths between
+observations.  Under this framework, it is possible to assess whether
+paths between 
 latent variables are stronger for some individuals than for
-others.  Secondly, 
-the tests may be extended to study differential item functioning (DIF)
-in item response models (e.g., \citeNP{KopZei10}, who focused on recursive
-partitioning of Rasch models).  
+others.  Further, in applying the proposed tests to factor mean
+parameters (as was done in the example), implications of measurement
+invariance violations for structural equation modeling may be better
+understood.
+
+The proposed tests also naturally extend to the study of differential
+item functioning (DIF) 
+in item response models (e.g., \citeNP{KopZei10}, who focused on Rasch
+models).   
 Traditional DIF methods are similar to those for factor analysis in
@@
 have developed 
-largely independently of one another, 
-some treatments \cite{Mcd99} and recent research \cite{StaChe06} have
-sought to unify the methods.  Extensions of the proposed tests can
-support this endeavor.
+largely independently of one another, the methods can certainly be
+treated from a unified perspective (e.g.,
+\citeNP{Mcd99,Mil11,StaChe06}).   
+The tests proposed here were designed with this perspective in mind.
 
@@
 statistics and applied them to the issue of measurement invariance in psychometrics.
-The paper included both theoretical development and 
+The paper included theoretical development, an applied example, and 
 study of the tests' performance.  The tests were found to have good
@@
 measurement invariance research and provide many avenues for
-future research, both through extensions of the tests within a
-factor-analytic context and through application of the tests to new
+future research.  This can happen both through extensions of the tests
+within a factor-analytic context and through application of the tests to new
 models.

### 2011-11-11 www/wdh.R

@@ -37,19 +37,45 @@ wdhfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
   return(rval)
 }
 
-#coef.wdhfit <- function(object, ...) {
-#  switch(object$engine,  
-#  "OpenMx" = {
-#    n <- object$names
-#    cf <- object$model@output$estimate
-#    cf <- cf[n[,1]]
-#    names(cf) <- n[,2]
-#    cf
-#  },
-#  "lavaan" = {
-#    unclass(getMethod("coef", "lavaan")(object$model, ...))
-#  })
-#}
+wdhfit2 <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
+{
+  ## Fit the same model but with different identification constraints.
+  ## data checks
+  data <- as.data.frame(data)
+  stopifnot(c("VERBREA","ABSREA","NUMER","gpa","grouping") %in% names(data))
+  data <- data[, c("VERBREA","ABSREA","NUMER","gpa","grouping")]
+
+  ## require package
+  stopifnot(require("lavaan"))
+  
+  ## model from Step 5b of Wicherts et al., 2005:
+  mod5b <- 'f1 =~ label(rep("l3",4))*VERBREA + 1*ABSREA + label(c("l1","l1","l1","l1expmin"))*NUMER 
+                  f1 ~ label(c("a","faccontmin","facexpmaj","facexpmin"))*1 + c(0,NA,NA,NA)*1
+                  f1 ~~ label(c("vmaj","vmin","vmaj","vmin"))*f1
+                  NUMER ~~ label(c("e1","e1","e1","e1expmin"))*NUMER
+                  ABSREA ~~ label(rep("e2",4))*ABSREA
+                  VERBREA ~~ label(rep("e3",4))*VERBREA
+
+                  NUMER ~ label(c("meanNA","meanNA","meanNA","meanNAexpmin"))*1
+                  ABSREA ~ label(c("meanARcontmaj","meanARcontmin","meanAR","meanAR"))*1
+                  VERBREA ~ label(rep("meanVR",4))*1'
+
+  ## fit model
+  rval <- lavaan(mod5b, data=data, meanstructure=TRUE, group="grouping")
+
+  ## store (1) data, (2) fitted model object, (3) grouping variable,
+  ## (4) focal group nubmers (group ordering is defined by lavaan and
+  ##                          can be found in rval@Sample@group.label)
+  rval <- list(
[…]

### 2011-11-11 www/paper.Rnw
Fixed some broken citations in the paper and added to refs.bib.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@75 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Test statistics for the hypotheses \eqref{eq:exh0}
-and~\eqref{eq:exh0.2} are displayed in Figure~\ref{cusumex}.
+and~\eqref{eq:exh0.2} are displayed in Figure~\ref{fig:cusumex}.
 Each panel displays a test 
@@
 %% Details on test results
-The results displayed in Figure~\ref{cusumex} generally show that the
+The results displayed in Figure~\ref{fig:cusumex} generally show that the
 ``minority, stereotype prime'' group lacks invariance with respect to

### 2011-11-17 www/paper.Rnw
touch-ups in references
set line-spacing to 1 (for the moment)



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@76 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
some LaTeX touch-ups:
{theta} -> {\theta}

replaced a few undirected double quotes by directed double quotes

\citeyear constructs replaced by \citeA



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@77 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-%\spacing{1}
+\spacing{1}
 
@@
 violating measurement invariance into structural equation
-models, where relationships between latent variables are hypothesized.  Horn
-and McArdle \citeyear{HorMca92} concisely summarize the impact of
+models, where relationships between latent variables are hypothesized.
+\citeA{HorMca92} concisely summarize the impact of
 these issues, stating ``Lack of evidence of measurement 
 invariance equivocates conclusions and casts doubt on theory in the
-behavioral sciences'' (p.~141).  Borsboom \citeyear{Bor06a} further 
+behavioral sciences'' (p.~141).  \citeA{Bor06a} further 
 notes that researchers often fail to assess whether measurement
@@
 an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
-maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. Andrews \citeyear{And93}
+maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. \citeA{And93}
 showed that the asymptotic distribution is in fact tractable but nonstandard.
@@
 of factor analysis models. More detailed accounts of the underlying structural
-change methods include Hjort and Koning \citeyear{HjoKon02} and
-Zeileis and Hornik \citeyear{ZeiHor07}.
+change methods include \citeA{HjoKon02} and \citeA{ZeiHor07}.
 
@@
 the usual ML regularity conditions (assuring asymptotic normality
-of $\hat {\bm \theta}$), Hjort \& Koning \citeyear{HjoKon02} show
+of $\hat {\bm \theta}$), \citeA{HjoKon02} show
 that 
@@
 
-For the Cram\'{e}r-von Mises test statistic $\mathit{CvM}$, Nyblom \citeyear{Nyb89}
-and Hansen \citeyear{Han92} provide small tables of critical values which
-have been extended in the software provided by Zeileis \citeyear{Zei06}.
+For the Cram\'{e}r-von Mises test statistic $\mathit{CvM}$, \citeA{Nyb89}
+and \citeA{Han92} provide small tables of critical values which
+have been extended in the software provided by \citeA{Zei06}.
 Critical values for the distribution of the maximum LR/Wald/LM tests
-are provided by Hansen \citeyear{Han97}. Note that the distribution depends on
+are provided by \citeA{Han97}. Note that the distribution depends on
 the minimal and maximal thresholds employed in the test.
@@
 \subsection{Local Alternatives}
-Using results from Hjort and Koning \citeyear{HjoKon02} and Zeileis
-and Hornik \citeyear{ZeiHor07}, we can also
+Using results from \citeA{HjoKon02} and \citeA{ZeiHor07}, we can also
 capture the behaviour of the process $\bm{B}(t, \hat{\bm{\theta}})$
@@
 of parameter instability. In particular, we can assume that the pattern
-of deviation $\bm{theta}(v_i)$ can be described as a constant
+of deviation $\bm{\theta}(v_i)$ can be described as a constant
 parameter plus some non-constant deviation $\bm{g}(i/n)$:
@@
 
-Under the local alternative described above, Hjort and Koning
-\citeyear{HjoKon02} show that the process
+Under the local alternative described above, \citeA{HjoKon02} show that the process
 $\bm{B}(t, \hat{\bm{\theta}})$ behaves asymptotically like
@@
 alternative has $\bm{g}(t) \equiv \bm{0}$, the empirical process $\bm{B}(t,
-\hat{\bm{theta}})$ will 
+\hat{\bm{\theta}})$ will 
 have a non-zero mean.  Hence, the corresponding tests will have non-trivial
@@
 (2) test measurement invariance for unknown $\nu$. 
-In particular, Millsap \citeyear{Mil05} cites
+In particular, \citeA{Mil05} cites
 ``locating the invariance violation'' as a major outstanding problem
@@
 In the next section, we employ the proposed tests to analyze existing
-data from Wicherts, Dolan, and Hessen \citeyear{WicDol05}.
+data from \citeA{WicDol05}.
 Following the example, we consider a
@@
 
-Wicherts, Dolan, and Hessen \citeyear{WicDol05} utilized confirmatory
+\citeA{WicDol05} utilized confirmatory
 factor models to study measurement invariance in a series of
@@
 
-To study stereotype threat in a measurement invariance framework,
-Study 1 of Wicherts et al.\ \citeyear{WicDol05} involved 295 high school
-students completing three intelligence tests in two between-subjects
-conditions.  Conditions were defined by
-whether or not students received primes about ethnic stereotypes prior to testing.
-To study the data, Wicherts et al.\ employed a four-group,
-one-factor model with the three intelligence tests as manifest variables.
-The groups were defined by ethnicity (majority/minority) and by the
-experimental manipulation (received/did not receive stereotype prime).
-Results indicated that the intelligence tests lacked measurement
-invariance, with the minority group receiving stereotype primes being
-particularly different from the other three groups on the most
-difficult intelligence test.  Most notably, the factor loading and
-intercept for the "minority, stereotype prime" group were smaller, and the unique variance
-was larger.
-
-As Wicherts et al.
-describe, one explanation for this pattern of results involves the idea that ability is positively related to "domain identification" \cite{Ste97}: individuals who are high on ability may be more prone to stereotype threat because they are more "tied in" to the domain of interest and are more aware of stereotypes.
-To test this idea, we employ a four-group, confirmatory factor model used by Wicherts et al.\ in tandem with the tests proposed in this paper.
-Our $V$ variable is participants' grade point averages (which were unused in the original analyses), which will reflect participants' identification with the academic domain.
+To study stereotype threat in a measurement invariance framework, Study 1 of
+\citeA{WicDol05} involved 295 high school students
+completing three intelligence tests in two between-subjects conditions. 
+Conditions were defined by whether or not students received primes about ethnic
+stereotypes prior to testing. To study the data, Wicherts et al.\ employed a
+four-group, one-factor model with the three intelligence tests as manifest
+variables. The groups were defined by ethnicity (majority/minority) and by the
+experimental manipulation (received/did not receive stereotype prime). Results
+indicated that the intelligence tests lacked measurement invariance, with the
+minority group receiving stereotype primes being particularly different from the
+other three groups on the most difficult intelligence test.  Most notably, the
+factor loading and intercept for the ``minority, stereotype prime'' group were
+smaller, and the unique variance was larger.
+
+As Wicherts et al. describe, one explanation for this pattern of results involves the idea that ability is positively related to ``domain identification''
+\cite{Ste97}: individuals who are high on ability may be more prone to stereotype threat because they are more ``tied in'' to the domain of interest and are
+more aware of stereotypes.  To test this idea, we employ a four-group, confirmatory factor model used by Wicherts et al.\ in tandem with the tests proposed
+in this paper.  Our $V$ variable is participants' grade point averages (GPA, which were unused in the original analyses), which will reflect participants' identification with the academic domain.
 
@@
 parameters were allowed to be group-specific upon examination of
-modification indices.
-Most notably, the most difficult intelligence test's factor loading, unique variance, and intercept differed for the "minority, stereotype threat" group, as compared to the other groups.
+modification indices.  Most notably, the most difficult intelligence test's factor loading, unique variance, and intercept differed for the ``minority,
+stereotype threat'' group, as compared to the other groups.
 Factor means were also estimated for three of the groups.  The model
@@
 \subsection{Results}
-<<<mz_mod>>=
+<<mz_mod>>=
 d <- dget("WichertsEtAl2005_Study1.txt")
@@
 Application of the proposed measurement invariance tests to the
-Wicherts et al.\ \citeyear{WicDol05} data allowed us to study the
+\citeA{WicDol05} data allowed us to study the
 domain identification's impact on stereotype threat in a
@@
 
-<<<constraint_study>>=
+<<constraint_study>>=
 tests <- c("maxBB", "meanL2BB", "supLM(0.1)")
@@
 
-Relatedly,
-S\'{a}nchez \citeyear{San09} describes a general
+Relatedly, \citeA{San09} describes a general
 method for partitioning/segmenting structural equation models within a 

### 2011-11-17 www/paper.Rnw
further small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@78 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 various types of measurement invariance.  We focus on the model used
-in their Step 5b (see pp.\ 703--704 of their paper), which involved
+in their Step~5b (see pp.~703--704 of their paper), which involved
 across-group parameter restrictions on the factor loadings, unique
@@
 provided a reasonable fit to the data, as judged by examination of
-$\chi^2$, RMSEA, and CFI statistics.  Wicherts et al.\ also had a Step
-6 (the final step), where they examined a model that was simpler than
+$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
+(comparative fit index) statistics.  Wicherts et al.\ also had a
+Step~6 (the final step), where they examined a model that was simpler than
 described above and that provided a slightly better fit to the data.
@@
 statistic's fluctuation across values of student GPA, with the first
-column containing tests of~\eqref{eq:exh0} (testing $k*=4$ parameters)
+column containing tests of~\eqref{eq:exh0} (testing $k^*=4$ parameters)
 and the second column containing tests of~\eqref{eq:exh0.2} (testing
-$k*=1$ parameter).
+$k^*=1$ parameter).
 The solid
@@
 identification (as measured by GPA) influences the magnitude of
-stereotype threat.  The $k*=1$ results also imply that the tests
+stereotype threat.  The $k^*=1$ results also imply that the tests
 differ in power and/or in error rates, an issue we address later in the
@@
 $\nu$ can be obtained by 
-examining the peaks in Figure~\ref{fig:cusumex}.  For the $k*=4$
+examining the peaks in Figure~\ref{fig:cusumex}.  For the $k^*=4$
 panels (first column), two peaks occur near GPAs of 6.0 and of 6.3 (6

### 2011-12-13 www/paper.Rnw
Modified example section:   - Revised text for simplicity   - Modified R code to work with lavaan-S3.R, as opposed to wdhfit objects.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@84 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ability.  The authors specifically focused on stereotypes concerning
-ethnic minorities' performance on tests of general intelligence.  In this section, we used the proposed tests to supplement the authors' original analyses.
+ethnic minorities' performance on tests of general intelligence.  In this section, we use the proposed tests to supplement the authors' original analyses.
 
@@
 minority group receiving stereotype primes being particularly different from the
-other three groups on the most difficult intelligence test.  Most notably, the
-factor loading and intercept for the ``minority, stereotype prime'' group were
-smaller, and the unique variance was larger.
-
-As Wicherts et al. describe, one explanation for this pattern of results involves the idea that ability is positively related to ``domain identification''
-\cite{Ste97}: individuals who are high on ability may be more prone to stereotype threat because they are more ``tied in'' to the domain of interest and are
-more aware of stereotypes.  To test this idea, we employ a four-group, confirmatory factor model used by Wicherts et al.\ in tandem with the tests proposed
-in this paper.  Our $V$ variable is participants' grade point averages (GPA, which were unused in the original analyses), which will reflect participants' identification with the academic domain.
+other three groups on the most difficult intelligence test.  In the
+example below, we employ one of the models used by Wicherts et al.
+Our $V$ variable is participants' grade point averages, which were unused in the original analyses.
 
@@
 parameters were allowed to be group-specific upon examination of
-modification indices.  Most notably, the most difficult intelligence test's factor loading, unique variance, and intercept differed for the ``minority,
-stereotype threat'' group, as compared to the other groups.
-Factor means were also estimated for three of the groups.  The model
+modification indices.    The model
 provided a reasonable fit to the data, as judged by examination of
 $\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
-(comparative fit index) statistics.  Wicherts et al.\ also had a
-Step~6 (the final step), where they examined a model that was simpler than
-described above and that provided a slightly better fit to the data.
-This model restricted factor means to be equal across a subset of the
-groups, however, which was not helpful for our purposes (because we are focusing on a single group).
-
-The stereotype threat effect of interest is most relevant to the
-"minority, stereotype prime" group.  Thus, to test the idea that
-domain identification impacts the magnitude of stereotype threat
-effect, we focus on parameters specific to this group.  We test these
-parameters for measurement invariance across the GPA variable, with
-violations supporting the hypothesis that domain identification
-impacts the magnitude of the stereotype threat effect.  We also test
-the intercept parameter individually.  This is because, if domain
-identification does impact stereotype threat, it could be
-conceptualized as the error term having a nonzero expectation across
-individuals (i.e., individuals with high domain identification have a
-negative error expectation and individuals with low domain
-identification have a positive error expectation).  The expected value
-for the error term cannot be uniquely separated from the intercept, so
-that the estimated intercept would display a lack of measurement
-invariance (see, e.g., \citeNP{Mil11}, p.\ 77).
+(comparative fit index) statistics.  While the model included four
+groups as described above, we focus only on the submodel for the minority
+group that received stereotype primes.
 
 To carry out
-the tests, the four-group confirmatory factor model described above
-was first fit to the data.  Casewise derivatives and 
-the observed information matrix were then obtained, and they were used
-to calculate the cumulative score process via
-\eqref{eq:cumscore}.  Finally, we
-obtained various test statistics and $p$-values from the cumulative score
-process.  These include the double-max statistic from \eqref{eq:dmax},
+the tests, we first fit the four-group model to the data,
+calculating casewise derivatives and 
+the observed information matrix.  We then
+calculated the cumulative score process wrt student GPA via
+\eqref{eq:cumscore}, allowing us to
+obtain various test statistics and $p$-values from this 
+process.  Test statistics include the double-max statistic from \eqref{eq:dmax},
 the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
@@
     \label{eq:exh0}
-H_0:\ (\nu_{i,1,\text{MSP}}\ \lambda_{i,1,\text{MSP}}\
-\theta_{i,1,\text{MSP}}\ \alpha_{i,1,\text{MSP}}) = 
-(\nu_{0,1,\text{MSP}}\ \lambda_{0,1,\text{MSP}}\
-\theta_{0,1,\text{MSP}}\ \alpha_{0,1,\text{MSP}}),\ i=1,\ldots,n_{\text{MSP}},
-\end{equation}
-where $\nu_{i,1,\text{MSP}}$ reflects the intercept for student $i$ on intelligence
-test 1 in the ``minority, stereotype prime'' group, with other
-subscripts defined similarly.  Additionally, $\lambda$ is a factor loading, $\theta$
-is a unique variance, and $\alpha$ is a factor mean, and the
+H_0:\ (\nu_{i}\ \lambda_{i}\
+\theta_{i}\ \alpha_{i}) = 
+(\nu_{0}\ \lambda_{0}\
+\theta_{0}\ \alpha_{0}),\ i=1,\ldots,n_{\text{MSP}},
+\end{equation}
+where $\nu_{i}$ is the intercept for student $i$ on intelligence
+test 1, $\lambda_{i}$ the factor loading, $\theta_{i}$ the unique
+variance, and $\alpha_i$ the factor mean.
+The individuals are ordered according to GPA, and the 
 hypothesis is taken only wrt the $n_{\text{MSP}}$ individuals in the
-``minority, stereotype prime'' group.  In words, the hypothesis states
-that the four group-specific model parameters are invariant with
+``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states 
+that these four group-specific model parameters are invariant with
 respect to GPA.
@@
     \label{eq:exh0.2}
-H_0:\ \nu_{i,1,\text{MSP}}\ = \nu_{0,1,\text{MSP}},\ i=1,\ldots,n_{\text{MSP}},
-\end{equation}
-which, as described previously, tests a more specific theory of the
-manner in which domain identification impacts stereotype threat.
+H_0:\ \nu_{i}\ = \nu_{0},\ i=1,\ldots,n_{\text{MSP}},
+\end{equation}
+which, e.g., may be used to isolate the impact of stereotype threat.
 We consider tests of both \eqref{eq:exh0} and \eqref{eq:exh0.2}
@@
 <<mz_mod>>=
-d <- dget("WichertsEtAl2005_Study1.txt")
-
-wdh0 <- wdhfit(d)
-
-# Refit model with new identification constraints
-wdh0.1 <- wdhfit2(d)
-
-## subsample models
-## This could be used to compare sequential LRTs with the LM
-## statistic that tests all parameters simultaneously.
-## (This is not done in the data example.)
-##############################################################
-## lrstat <- function(gpa) 2 * as.vector(logLik(wdhfit(d[d$gpa <= gpa,])) + logLik(wdhfit(d[d$gpa > gpa,])) - logLik(wdh0))
-## wdh.gpa <- sort(d$gpa[d$gpa >= quantile(d$gpa, 0.1) & d$gpa <= quantile(d$gpa, 0.9)])
-## # Results in many "computationally singular" errors:
-## wdh.lrstat <- sapply(wdh.gpa, lrstat)
-## save(wdh0, wdh.gpa, wdh.lrstat, file = "wicherts_mod.rda")
+load("wicherts05_study1.rda")
+
+d <- subset(d, !is.na(gpa))
+
+## fit model 5b
+wdh5b <- lavaan(
+   'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical + label(rep("load_v", 4)) * verbal + 1 * abstract
+    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
+    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
+    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
+    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
+    verbal    ~  label(rep("mean_v", 4)) * 1
+    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
+    abstract  ~~ label(rep("var_a", 4)) * abstract
+    verbal    ~~ label(rep("var_v", 4)) * verbal',
+  data = d, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")
+
+## Refit model with new identification constraints
+wdh5b.2 <- lavaan(
+   'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
+    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
+    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
+    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
+    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
+    verbal    ~  label(rep("mean_v", 4)) * 1
+    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
+    abstract  ~~ label(rep("var_a", 4)) * abstract
+    verbal    ~~ label(rep("var_v", 4)) * verbal',
+  data = d, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
 @
@@
 ``minority, stereotype prime'' group lacks invariance with respect to
-GPA.  All three tests are significant at the .05 level when the four
-group-specific parameters are tested simultaneously (first column).
-When the intercept parameter is tested individually, two of the three
-tests achieve significance (the CvM test and $\max \mathit{LM}$
-statistics).  These results generally support the notion that domain
-identification (as measured by GPA) influences the magnitude of
-stereotype threat.  The $k^*=1$ results also imply that the tests
-differ in power and/or in error rates, an issue we address later in the
-paper.
-
-%% add paragraph about peaks of graphs
+GPA.  All three tests are significant at the .05 level, both when the four
+group-specific parameters are tested simultaneously (first column) and
+when the intercept parameter is tested individually (second column).
+
 We can also use the test results to estimate the threshold $\nu$,
@@
 examining the peaks in Figure~\ref{fig:cusumex}.  For the $k^*=4$
-panels (first column), two peaks occur near GPAs of 6.0 and of 6.3 (6
+panels (first column), two peaks occur near GPAs of 5.9 and of 6.3 (6
 corresponds to the American `C' range).  These peaks roughly divide
@@
 results can also be obtained, as shown for the intercept parameter in
-the second column of Figure~\ref{fig:cusumex}.  Here, a single peak
-occurs in the three panels around a GPA of 7.0 (`B' range).
+the second column of Figure~\ref{fig:cusumex}.  The peak locations are
+generally similar to those of all four parameters, though this finding
+will not hold in general.
 
@@
 ties <- duplicated(d$gpa)
-d$gpa[ties] <- runif(sum(ties), d$gpa[ties]-.01, d$gpa[ties]+.01)
+d$gpa[ties] <- runif(sum(ties), min = d$gpa[ties] - .001, max = d$gpa[ties] + .001)
+
+## FIXME It seems like info_full should go in lavaan_S3.R.
+info_full <- function(x, ...) solve(vcov(x) * nobs(x))
+scores_min_t <- function(x, ...) {
+  ef <- estfun(x, ...)
+  ef[d$group == "exp:minority", ]
+}
+gpa_min_t <- subset(d, group == "exp:minority")$gpa
+
 ## empirical fluctuation processes (4 vs 1 par)
-gefp_4_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
-gefp_1_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)
+gefp_4_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
+gefp_1_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
 
 # processes for second identification constraints
-gefp_4_info_1 <- gefp(wdh0.1, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
-gefp_1_info_1 <- gefp(wdh0.1, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)
+gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
+gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
 
@@
 Application of the proposed measurement invariance tests to the
-\citeA{WicDol05} data allowed us to study the
-domain identification's impact on stereotype threat in a
-straightforward manner.  The tests focused on a set of group-specific
+\citeA{WicDol05} data allowed us to study the extent to which model
+parameters are invariant wrt GPA in a straightforward manner.
+The tests focused on a set of group-specific
 parameters within a four-group confirmatory factor model, and they
@@
 The latter fact is important, as other approaches to the problem
-described here may adversely impact model fit and degrees of freedom (e.g., if
-GPA is inserted directly into the model) or require multiple models to
-be estimated (e.g., for LRTs).  Further, through examination of both
-test statistics and cumulative score processes, the tests were easily
+described here may require multiple models to
+be estimated (e.g., for LRTs) or adversely impact model fit and
+degrees of freedom (e.g., if GPA is inserted directly into the model).
+Further, through examination of both 
+test statistics and cumulative score processes, the tests were
 interpretable from both a theoretical and applied standpoint.  In the

### 2011-12-13 www/paper.Rnw
Revised example sections of paper   - Re-introduced the artificial example   - Ordered the sections so that artificial stuff comes first, then the real example.   - Revised the text so that these sections flow smoothly.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@85 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \acknowledgements{This work was supported by National Science
-  Foundation grant SES-1061334. Correspondence to Edgar C.\ Merkle,
+  Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
+  generously shared data for the stereotype threat example.
+  Correspondence to Edgar C.\ Merkle, 
   Department of  
@@
 source("mz.R")
-source("wdh.R")
+source("lavaan-S3.R")
 source("sim.R")
@@
 ``locating the invariance violation'' as a major outstanding problem
-in the field.
-
-In the next section, we employ the proposed tests to analyze existing
-data from \citeA{WicDol05}.
-Following the example, we consider a
-full simulation study involving artificial data.
-
-\section{Example: Stereotype Threat}
-
-\citeA{WicDol05} utilized confirmatory
-factor models to study measurement invariance in a series of
-general intelligence scales.  They were interested in the
-notion of {\em{stereotype threat}}, whereby stereotypes concerning a
-subgroup's ability adversely impact the subgroup's performance on tests of that
-ability.  The authors specifically focused on stereotypes concerning
-ethnic minorities' performance on tests of general intelligence.  In this section, we use the proposed tests to supplement the authors' original analyses.
-
-\subsection{Background}
-
-To study stereotype threat in a measurement invariance framework, Study 1 of
-\citeA{WicDol05} involved 295 high school students
-completing three intelligence tests in two between-subjects conditions. 
-Conditions were defined by whether or not students received primes about ethnic
-stereotypes prior to testing. To study the data, Wicherts et al.\ employed a
-four-group, one-factor model with the three intelligence tests as manifest
-variables. The groups were defined by ethnicity (majority/minority) and by the
-experimental manipulation (received/did not receive stereotype prime). Results
-indicated that the intelligence tests lacked measurement invariance, with the
-minority group receiving stereotype primes being particularly different from the
-other three groups on the most difficult intelligence test.  In the
-example below, we employ one of the models used by Wicherts et al.
-Our $V$ variable is participants' grade point averages, which were unused in the original analyses.
+in the field.  In the following sections, we demonstrate the proposed
+tests' uses and properties via example and simulation.  We first
+describe an example and simulation with artificial data, and we then
+provide an illustrative example with real data.  Finally, we conclude
+the paper with a discussion of test extensions and general summary.
+
+\begin{figure}
+\caption{Path diagram representing the base factor analysis model used for
+  the example and simulations.  To induce measurement invariance
+  violations, a seventh observed variable (student age) determines the
+values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
+$\lambda_{31}$).}
+\label{fig:famod}
+\includegraphics[height=5in]{famod.pdf}
+\end{figure}
+
+\section{Example with Artificial Data}
+
+Consider a 
+hypothetical battery of six scales administered to students aged 13 to 18
+years.  Three of the scales are intended to measure verbal ability, and
+three of the scales are intended to measure mathematical ability.  We
+may observe a maturation effect in the resulting data, whereby the
+factor loadings for older students are larger than those for younger
+students.  The researcher's goal is to
+study whether the scales are measurement invariant with
+respect to age, which is taken to be the auxiliary variable $V$.
 
 \subsection{Method}
-Wicherts et al.\ tested a series of confirmatory factor models for
-various types of measurement invariance.  We focus on the model used
-in their Step~5b (see pp.~703--704 of their paper), which involved
-across-group parameter restrictions on the factor loadings, unique
-variances, intercepts, and factor variances.  These parameters were
-typically restricted to be equal across groups, though a subset of the
-parameters were allowed to be group-specific upon examination of
-modification indices.    The model
-provided a reasonable fit to the data, as judged by examination of
-$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
-(comparative fit index) statistics.  While the model included four
-groups as described above, we focus only on the submodel for the minority
-group that received stereotype primes.
-
-To carry out
-the tests, we first fit the four-group model to the data,
-calculating casewise derivatives and 
-the observed information matrix.  We then
-calculated the cumulative score process wrt student GPA via
-\eqref{eq:cumscore}, allowing us to
-obtain various test statistics and $p$-values from this 
-process.  Test statistics include the double-max statistic from \eqref{eq:dmax},
+
+To formally represent these ideas, we specify that the data arise from
+a factor analysis model with two factors.  The base model, displayed in
+Figure~\ref{fig:famod}, specifies that measurement invariance holds,
+with three scales arising from the verbal factor and three 
+scales arising from the mathematical factor.
+For the measurement invariance violation, we specify that $V$
+(student age) impacts the values of 
+verbal factor loadings in the model: 
+if students are 16 through 18 years of age, then the factor
+loadings corresponding to the first 
+factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
+Figure~\ref{fig:famod}.  If students are 13 through 15 years of age,
+however, then the factor loadings corresponding to the first factor
+are three standard errors ($=$ asymptotic standard errors divided by $\sqrt{n}$)
+lower than those in Figure~\ref{fig:famod}.
+This violation states that the verbal ability scales
+lack measurement invariance with respect to age.  For simplicity, we
+assume that the mathematical scales are invariant.
+
+<<dgp>>=
+set.seed(1090)
+d <- dgp(200, 2)
+@
+
+A sample of size 200 was generated from the model described above, and
+a test was conducted to examine measurement invariance of the three
+verbal scales.  To carry out
+the test, a confirmatory factor analysis model (with the paths
+displayed in Figure~\ref{fig:famod}) was fit to the
+data.  Casewise derivatives and 
+the observed information matrix were then obtained, and they were used
+to calculate the cumulative score process via
+\eqref{eq:cumscore}.  Finally, we
+obtained various test statistics and $p$-values from the cumulative score
+process.  These include the double-max statistic from \eqref{eq:dmax},
 the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
@@
 test various subsets of 
-parameters.  For the Wicherts et al.\ data, we first test
+parameters.  For example, if we suspected that the verbal factor loadings
+lacked measurement invariance, we could test
 \begin{equation}
     \label{eq:exh0}
-H_0:\ (\nu_{i}\ \lambda_{i}\
-\theta_{i}\ \alpha_{i}) = 
-(\nu_{0}\ \lambda_{0}\
-\theta_{0}\ \alpha_{0}),\ i=1,\ldots,n_{\text{MSP}},
-\end{equation}
-where $\nu_{i}$ is the intercept for student $i$ on intelligence
-test 1, $\lambda_{i}$ the factor loading, $\theta_{i}$ the unique
-variance, and $\alpha_i$ the factor mean.
-The individuals are ordered according to GPA, and the 
-hypothesis is taken only wrt the $n_{\text{MSP}}$ individuals in the
-``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states 
-that these four group-specific model parameters are invariant with
-respect to GPA.
-
-We can construct a similar hypothesis involving only the intercept
-parameter
-\begin{equation}
-    \label{eq:exh0.2}
-H_0:\ \nu_{i}\ = \nu_{0},\ i=1,\ldots,n_{\text{MSP}},
-\end{equation}
-which, e.g., may be used to isolate the impact of stereotype threat.
-We consider tests of both \eqref{eq:exh0} and \eqref{eq:exh0.2}
-below.
+H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
+(\lambda_{0, 11}\ \lambda_{0, 21}\ \lambda_{0, 31}),\ i=1,\ldots,n,
+\end{equation}
+where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
+verbal factor loading parameters for student $i$. Thus, here only $k^* = 3$
+from the overall $k = 19$~model parameters (including means) are assessed.
+Alternatively, we can consider all $k^* = k = 19$~parameters, leading to a test
+of~\eqref{eq:h0}.  We consider both of these tests below.
 
 \subsection{Results}
+In the results section, we first describe overall results.  We then 
+describe estimation of $\nu$ and 
+isolation of model parameters violating
+measurement invariance.
+
 <<mz_mod>>=
-load("wicherts05_study1.rda")
-
-d <- subset(d, !is.na(gpa))
-
-## fit model 5b
-wdh5b <- lavaan(
-   'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical + label(rep("load_v", 4)) * verbal + 1 * abstract
-    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
-    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
-    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
-    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
-    verbal    ~  label(rep("mean_v", 4)) * 1
-    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
-    abstract  ~~ label(rep("var_a", 4)) * abstract
-    verbal    ~~ label(rep("var_v", 4)) * verbal',
-  data = d, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")
-
-## Refit model with new identification constraints
-wdh5b.2 <- lavaan(
-   'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
-    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
-    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
-    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
-    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
-    verbal    ~  label(rep("mean_v", 4)) * 1
-    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
-    abstract  ~~ label(rep("var_a", 4)) * abstract
-    verbal    ~~ label(rep("var_v", 4)) * verbal',
-  data = d, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
+if(file.exists("mz_mod.rda")) {
+  load("mz_mod.rda")
+} else {
+  ## full sample model
+  mz0 <- mzfit(d)
+  ## subsample models
+  lrstat <- function(age) 2 * as.vector(logLik(mzfit(d[d$age <= age,])) + logLik(mzfit(d[d$age > age,])) - logLik(mz0))
+  mz.age <- sort(d$age[d$age >= quantile(d$age, 0.1) & d$age <= quantile(d$age, 0.9)])
+  mz.lrstat <- sapply(mz.age, lrstat)
+  save(mz0, mz.age, mz.lrstat, file = "mz_mod.rda")
+}
 @
 
-Test statistics for the hypotheses \eqref{eq:exh0}
-and~\eqref{eq:exh0.2} are displayed in Figure~\ref{fig:cusumex}.
+\subsubsection{Overall Results}
+Test statistics for the hypotheses \eqref{eq:exh0} and~\eqref{eq:h0}
+are displayed in  
+Figure~\ref{fig:cusumex}.
 Each panel displays a test 
-statistic's fluctuation across values of student GPA, with the first
-column containing tests of~\eqref{eq:exh0} (testing $k^*=4$ parameters)
-and the second column containing tests of~\eqref{eq:exh0.2} (testing
-$k^*=1$ parameter).
-The solid
+statistic's fluctuation across values of student age, with the first
+column containing tests of~\eqref{eq:exh0} and the second column
+containing tests of~\eqref{eq:h0}.  The solid
 horizontal 
@@
 
-%% Details on test results
-The results displayed in Figure~\ref{fig:cusumex} generally show that the
-``minority, stereotype prime'' group lacks invariance with respect to
-GPA.  All three tests are significant at the .05 level, both when the four
-group-specific parameters are tested simultaneously (first column) and
-when the intercept parameter is tested individually (second column).
-
-We can also use the test results to estimate the threshold $\nu$,
-which in this context can be used to define GPA-based subgroups whose
-measurement parameters differ.  As described previously, estimates of
-$\nu$ can be obtained by 
-examining the peaks in Figure~\ref{fig:cusumex}.  For the $k^*=4$
-panels (first column), two peaks occur near GPAs of 5.9 and of 6.3 (6
-corresponds to the American `C' range).  These peaks roughly divide
-individuals into those receiving D's and F's, those receiving solid
-C's, and those receiving high C's, B's, and A's.  Parameter-specific
-results can also be obtained, as shown for the intercept parameter in
-the second column of Figure~\ref{fig:cusumex}.  The peak locations are
-generally similar to those of all four parameters, though this finding
-will not hold in general.
+The figures convey information about several properties
+of the tests. First, all three tests are more powerful (and in
+this example significant) if we test only those parameters 
+that are subject to instabilities. Conversely, if all 19~model
+parameters are assessed (including those that are in fact invariant),
+the power is decreased. This decrease, however, is less pronounced
+for the double-max test as it is more 
+sensitive to fluctuations among a small subset of parameters.
+
+Figure~\ref{fig:lrlm} 
+compares the $\max \mathit{LM}$ statistic (solid line) to the $\max \mathit{LR}$
+statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
+\eqref{eq:h0}.  The critical values for these two
+tests are identical, hence the single horizontal line.
+The figure
+shows that the two statistics are very similar to one another, with
+both maxima at the dotted vertical line.  This
+is generally to be expected, because the two tests are asymptotically
+equivalent.  The 
+$\max \mathit{LR}$~statistic cannot be obtained from the empirical fluctuation
+process, however, so the factor analysis model must be refitted before
+and after each of the possible threshold values $\nu$ (i.e., here
+$\Sexpr{2 * length(mz.lrstat)}$~model fits for $\Sexpr{length(mz.lrstat)}$~thresholds).
+
+\subsubsection{Estimation of $\nu$ and Parameter Isolation}
+As described above, the tests of~\eqref{eq:exh0} imply that the 
+verbal scales lack measurement invariance.  
+We can also use the tests to: (1)~estimate the 
+threshold $\nu$, and (2)~isolate specific parameters that violate
+measurement invariance.  
+For example, as described previously, estimates of $\nu$ can be obtained by
+examining the peaks in Figure~\ref{fig:cusumex}.  For all six panels
+in the figure, the peaks occur near an age of 16.1.  This agrees well
+with the true threshold of 16.0.
+
+As mentioned previously, the double-max test is advantageous because
+it yields information about individual 
+parameters violating measurement invariance.  
+That is, it allows us to examine whether or not individual parameters' 
+cumulative score processes lead one to reject the hypothesis of
+measurement invariance.
+Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
+verbal factor loadings, with the horizontal lines reflecting the
+critical value at $\alpha= 0.05$.
+  The figure
+shows that the third parameter (i.e., $\lambda_{31}$) crosses the
+dashed line, so we would conclude a measurement 
+invariance violation with respect to age for the third
+verbal test.  The fact that the cumulative score process for the first
+and second loadings did not achieve the critical value represents a Type II
+error, which implicitly brings into question the tests' power.
+This is an example of a situation where the double-max test ``wastes''
+power, as it cannot make use of the fact that multiple model
+parameters change simultaneously.  We
+generally address the issue of power in the simulations below.
 
 \begin{figure}
-\caption{Three test statistics of~\eqref{eq:exh0} (set of
-  group-specific parameters)
-  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
-  using data from Study 1 of Wicherts et al.\ (2005).
-  Solid, horizontal lines represent critical values
+\caption{Three test statistics of~\eqref{eq:exh0} (with $k^* = 3$)
+  and~\eqref{eq:h0} (with $k^* = 19$), 
+  based on the example involving measurement invariance with respect
+  to student age.  Solid, horizontal lines represent critical values
   at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
@@
 \setkeys{Gin}{width=\textwidth}
-<<gefp-4-1, fig=TRUE, height=8, width=6>>=
-## add a small amount of random noise to duplicated GPAs, so that everything
-## works with gefp().  (Without this, one can still obtain DM and CvM test 
-##                      results but not supLM results.)
-set.seed(1090)
-ties <- duplicated(d$gpa)
-d$gpa[ties] <- runif(sum(ties), min = d$gpa[ties] - .001, max = d$gpa[ties] + .001)
-
-## FIXME It seems like info_full should go in lavaan_S3.R.
-info_full <- function(x, ...) solve(vcov(x) * nobs(x))
-scores_min_t <- function(x, ...) {
-  ef <- estfun(x, ...)
-  ef[d$group == "exp:minority", ]
-}
-gpa_min_t <- subset(d, group == "exp:minority")$gpa
-
-## empirical fluctuation processes (4 vs 1 par)
-gefp_4_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
-gefp_1_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
-
-# processes for second identification constraints
-gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
-gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
+<<gefp-3-19, fig=TRUE, height=8, width=6>>=
+## empirical fluctuation processes (3 vs all 19 pars)
+gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
+  vcov = info.mzfit, sandwich = FALSE)
+gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
+  vcov = info.mzfit, sandwich = FALSE)
 
@@
 par(mfcol = c(3, 2))
-plot(gefp_4_info,  functional = maxBB, main = "DM, k* = 4", xlab = "GPA")
-plot(gefp_4_info,  functional = meanL2BB, main = "CvM, k* = 4", xlab = "GPA")
-plot(gefp_4_info,  functional = supLM(0.1), main = "max LM, k* = 4", xlab = "GPA")
-plot(gefp_1_info, functional = maxBB, main = "DM, k* = 1", xlab = "GPA")
-plot(gefp_1_info, functional = meanL2BB, main = "CvM, k* = 1", xlab = "GPA")
-plot(gefp_1_info, functional = supLM(0.1), main = "max LM, k* = 1", xlab = "GPA")
+plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
+plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
 @
@@
 
-\subsection{Discussion}
-Application of the proposed measurement invariance tests to the
-\citeA{WicDol05} data allowed us to study the extent to which model
-parameters are invariant wrt GPA in a straightforward manner.
-The tests focused on a set of group-specific
-parameters within a four-group confirmatory factor model, and they
-could be carried out using the results from a single estimated model.
-The latter fact is important, as other approaches to the problem
-described here may require multiple models to
-be estimated (e.g., for LRTs) or adversely impact model fit and
-degrees of freedom (e.g., if GPA is inserted directly into the model).
-Further, through examination of both 
-test statistics and cumulative score processes, the tests were
-interpretable from both a theoretical and applied standpoint.  In the
-next section, we study the proposed tests' power via simulation.
-
 \begin{figure}
-\caption{Path diagram representing the base factor analysis model used for
-  the example and simulations.  To induce measurement invariance
-  violations, a seventh observed variable (student age) determines the
-values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
-$\lambda_{31}$).}
-\label{fig:famod}
-\includegraphics[height=5in]{famod.pdf}
+\caption{A comparison of the $\max \mathit{LM}$ (solid line) and $\max \mathit{LR}$ (dashed line) test
+  statistics for~\eqref{eq:h0} (i.e., $k^* = 19$).  The solid horizontal line
+  corresponds to the critical value at $\alpha= 0.05$ while the dotted vertical
+  line highlights the threshold at which both test statistics assume their maximum.}
+\label{fig:lrlm}
+\setkeys{Gin}{width=\textwidth}
+<<lm-lr-19, fig=TRUE, height=5, width=6>>=
+plot(gefp_19_info, functional = supLM(0.1),
+  xlab = "Age", ylab = "LR and LM statistics (k* = 19)", main = "", ylim = c(0, 50))
+legend("topleft", c("LR", "LM"), lty = c(2, 1), bty = "n")
+lines(mz.age, mz.lrstat, lty = 2)
+abline(v = mz.age[which.max(mz.lrstat)], lty = 3)
+@
+\end{figure}
+
+\begin{figure}
+\caption{Cumulative score processes for each verbal factor
+  loadings, with critical values stemming from the double-max test. 
+  The solid, horizontal lines
+  correspond to the critical value at $\alpha= 0.05$.}
+\label{fig:lambdas}
+\setkeys{Gin}{width=\textwidth}
+<<gefp-lambdas, fig=TRUE, height=8, width=6>>=
+plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
+  main = "", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
+@
 \end{figure}
@@
 measurement invariance violations are known to exist.  Specifically,
-we generated data from the factor analysis model displayed in
-Figure~\ref{fig:famod}, with measurement invariance violations in the
-``verbal'' factor loadings with respect to a continuous auxliary variable.
-For simplicity, we
-assume other model parameters are invariant.  We also employ a
-single-group model, reflecting situations where there exist only continuous
-$V$ (so that no groups can be predefined).
+we generated data from the factor analysis model used in the previous
+section, with measurement invariance violations in the
+``verbal'' factor loadings with respect to a continuous auxiliary variable.
 
 We examine the power and error rates of the three
-tests used in the Example section: the double-max test, the
+tests used in the previous section: the double-max test, the
 Cram\'{e}r-von Mises test, and the $\max \mathit{LM}$ test.  We also compare 
@@
 \subsection{Method}
-Data were generated using the model in Figure~\ref{fig:famod}, with
-$V$ being designed to reflect student ages.
-For the measurement invariance violation, ``young'' students (13--15)
-had smaller factor loadings 
-than ``old'' students (16--18).
+Data were generated from the same model and with the same type of
+measurement invariance as was described in the Artificial Example section.
 Sample size and magnitude
@@
 \text{I}(t > 0.5) \times (t - 0.5)$ and $G^0(t) = \{\text{I}(t > 0.5) - 0.5 \} \times t - \text{I}(t > 0.5) \times 0.5$,
-both again multiplied by the violation magnitude. Thus, the mean under the
+both again multiplied by the violation magnitude. This implies that
+the mean under the 
 alternative is driven by a function that is triangle-shaped, with a peak at the
@@
 
-\section{Discussion}
+\section{Example: Stereotype Threat}
+
+\citeA{WicDol05} utilized confirmatory
+factor models to study measurement invariance in a series of
+general intelligence scales.  They were interested in the
+notion of {\em{stereotype threat}}, whereby stereotypes concerning a
+subgroup's ability adversely impact the subgroup's performance on tests of that
+ability.  The authors specifically focused on stereotypes concerning
+ethnic minorities' performance on tests of general intelligence.  In this section, we use the proposed tests to supplement the authors' original analyses.
+
+\subsection{Background}
+
+To study stereotype threat in a measurement invariance framework, Study 1 of
+\citeA{WicDol05} involved 295 high school students
+completing three intelligence tests in two between-subjects conditions. 
+Conditions were defined by whether or not students received primes about ethnic
+stereotypes prior to testing. To study the data, Wicherts et al.\ employed a
+four-group, one-factor model with the three intelligence tests as manifest
+variables. The groups were defined by ethnicity (majority/minority) and by the
+experimental manipulation (received/did not receive stereotype prime). Results
+indicated that the intelligence tests lacked measurement invariance, with the
+minority group receiving stereotype primes being particularly different from the
+other three groups on the most difficult intelligence test.  In the
+example below, we employ one of the models used by Wicherts et al.
+Our $V$ variable is participants' grade point averages, which were unused in the original analyses.
+
+\subsection{Method}
+Wicherts et al.\ tested a series of confirmatory factor models for
+various types of measurement invariance.  We focus on the model used
+in their Step~5b (see pp.~703--704 of their paper), which involved
+across-group parameter restrictions on the factor loadings, unique
+variances, intercepts, and factor variances.  These parameters were
+typically restricted to be equal across groups, though a subset of the
+parameters were allowed to be group-specific upon examination of
+modification indices.    The model
+provided a reasonable fit to the data, as judged by examination of
+$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
+(comparative fit index) statistics.  While the model included four
+groups as described above, we focus only on the submodel for the minority
+group that received stereotype primes.
+
+To carry out
+the tests, we first fit the four-group model to the data,
+calculating casewise derivatives and 
+the observed information matrix.  We then
+calculated the cumulative score process wrt student GPA via
+\eqref{eq:cumscore}, allowing us to
+obtain various test statistics and $p$-values from this 
+process.  Test statistics include the double-max statistic from \eqref{eq:dmax},
+the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
+statistic from \eqref{eq:maxlm}.
+
+As mentioned in the theory section, the 
+tests give us the flexibility to study hypotheses of partial change.
+That is, we have the ability to 
+test various subsets of 
+parameters.  For the Wicherts et al.\ data, we first test
+\begin{equation}
+    \label{eq:exh0}
+H_0:\ (\nu_{i}\ \lambda_{i}\
+\theta_{i}\ \alpha_{i}) = 
+(\nu_{0}\ \lambda_{0}\
+\theta_{0}\ \alpha_{0}),\ i=1,\ldots,n_{\text{MSP}},
+\end{equation}
+where $\nu_{i}$ is the intercept for student $i$ on intelligence
+test 1, $\lambda_{i}$ the factor loading, $\theta_{i}$ the unique
+variance, and $\alpha_i$ the factor mean.
+The individuals are ordered according to GPA, and the 
+hypothesis is taken only wrt the $n_{\text{MSP}}$ individuals in the
+``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states 
+that these four group-specific model parameters are invariant with
+respect to GPA.
+
+We can construct a similar hypothesis involving only the intercept
+parameter
+\begin{equation}
+    \label{eq:exh0.2}
+H_0:\ \nu_{i}\ = \nu_{0},\ i=1,\ldots,n_{\text{MSP}},
+\end{equation}
+which, e.g., may be used to isolate the impact of stereotype threat.
+We consider tests of both \eqref{eq:exh0} and \eqref{eq:exh0.2}
+below.
+
+\subsection{Results}
+<<mz_mod>>=
+load("wicherts05_study1.rda")
+
+d <- subset(d, !is.na(gpa))
+
+## fit model 5b
+wdh5b <- lavaan(
+   'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical + label(rep("load_v", 4)) * verbal + 1 * abstract
+    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
+    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
+    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
+    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
+    verbal    ~  label(rep("mean_v", 4)) * 1
+    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
+    abstract  ~~ label(rep("var_a", 4)) * abstract
+    verbal    ~~ label(rep("var_v", 4)) * verbal',
+  data = d, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")
+
+## Refit model with new identification constraints
+wdh5b.2 <- lavaan(
+   'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
+    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
+    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
+    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
+    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
+    verbal    ~  label(rep("mean_v", 4)) * 1
+    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
+    abstract  ~~ label(rep("var_a", 4)) * abstract
+    verbal    ~~ label(rep("var_v", 4)) * verbal',
+  data = d, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
+@
+
+Test statistics for the hypotheses \eqref{eq:exh0}
+and~\eqref{eq:exh0.2} are displayed in Figure~\ref{fig:cusumex}.
+Each panel displays a test 
+statistic's fluctuation across values of student GPA, with the first
+column containing tests of~\eqref{eq:exh0} (testing $k^*=4$ parameters)
+and the second column containing tests of~\eqref{eq:exh0.2} (testing
+$k^*=1$ parameter).
+The solid
+horizontal 
+lines represent critical values for $\alpha= 0.05$, and
+the Cram\'{e}r-von Mises panels 
+also contain a dashed line depicting the value of the test
+statistic (test statistics for the others are simply 
+the maxima of the processes).
+In other words, for panels in the first and third rows, \eqref{eq:exh0} is
+rejected if the process crosses the horizontal line.  For panels in
+the second row, \eqref{eq:h0} is rejected if the dashed horizontal line is
+higher than the solid horizontal line.
+
+%% Details on test results
+The results displayed in Figure~\ref{fig:cusumex} generally show that the
+``minority, stereotype prime'' group lacks invariance with respect to
+GPA.  All three tests are significant at the .05 level, both when the four
+group-specific parameters are tested simultaneously (first column) and
+when the intercept parameter is tested individually (second column).
+
+We can also use the test results to estimate the threshold $\nu$,
+which in this context can be used to define GPA-based subgroups whose
+measurement parameters differ.  As described previously, estimates of
+$\nu$ can be obtained by 
+examining the peaks in Figure~\ref{fig:cusumex}.  For the $k^*=4$
+panels (first column), two peaks occur near GPAs of 5.9 and of 6.3 (6
+corresponds to the American `C' range).  These peaks roughly divide
+individuals into those receiving D's and F's, those receiving solid
+C's, and those receiving high C's, B's, and A's.  Parameter-specific
+results can also be obtained, as shown for the intercept parameter in
+the second column of Figure~\ref{fig:cusumex}.  The peak locations are
+generally similar to those of all four parameters, though this finding
+will not hold in general.
+
+\begin{figure}
+\caption{Three test statistics of~\eqref{eq:exh0} (set of
+  group-specific parameters)
+  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
+  using data from Study 1 of Wicherts et al.\ (2005).
+  Solid, horizontal lines represent critical values
+  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
+  represent values of the Cram\'{e}r-von Mises test statistic.}
+\label{fig:cusumex}
+\setkeys{Gin}{width=\textwidth}
+<<gefp-4-1, fig=TRUE, height=8, width=6>>=
+## add a small amount of random noise to duplicated GPAs, so that everything
+## works with gefp().  (Without this, one can still obtain DM and CvM test 
+##                      results but not supLM results.)
+set.seed(1090)
+ties <- duplicated(d$gpa)
+d$gpa[ties] <- runif(sum(ties), min = d$gpa[ties] - .001, max = d$gpa[ties] + .001)
+
+## FIXME It seems like info_full should go in lavaan_S3.R.
+info_full <- function(x, ...) solve(vcov(x) * nobs(x))
+scores_min_t <- function(x, ...) {
+  ef <- estfun(x, ...)
+  ef[d$group == "exp:minority", ]
+}
+gpa_min_t <- subset(d, group == "exp:minority")$gpa
+
+## empirical fluctuation processes (4 vs 1 par)
+gefp_4_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
+gefp_1_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
+
+# processes for second identification constraints
+gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
+gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
+
+## visualization
+par(mfcol = c(3, 2))
+plot(gefp_4_info,  functional = maxBB, main = "DM, k* = 4", xlab = "GPA")
+plot(gefp_4_info,  functional = meanL2BB, main = "CvM, k* = 4", xlab = "GPA")
+plot(gefp_4_info,  functional = supLM(0.1), main = "max LM, k* = 4", xlab = "GPA")
+plot(gefp_1_info, functional = maxBB, main = "DM, k* = 1", xlab = "GPA")
+plot(gefp_1_info, functional = meanL2BB, main = "CvM, k* = 1", xlab = "GPA")
+plot(gefp_1_info, functional = supLM(0.1), main = "max LM, k* = 1", xlab = "GPA")
+@
+\end{figure}
+
+\subsection{Summary}
+Application of the proposed measurement invariance tests to the
+\citeA{WicDol05} data allowed us to study the extent to which model
+parameters are invariant wrt GPA in a straightforward manner.
+The tests focused on a set of group-specific
+parameters within a four-group confirmatory factor model, and they
+could be carried out using the results from a single estimated model.
+The latter fact is important, as other approaches to the problem
+described here may require multiple models to
+be estimated (e.g., for LRTs) or adversely impact model fit and
+degrees of freedom (e.g., if GPA is inserted directly into the model).
+Further, through examination of both 
+test statistics and cumulative score processes, the tests were
+interpretable from both a theoretical and applied standpoint.
+
+\section{General Discussion}
 In this paper, we have presented a new family of statistical tests for

### 2011-12-14 www/paper.Rnw
small touch-ups to R code
(data columns were not correctly formatted, now use StereotypeThreat data
from psychotools-devel)



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@86 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine = R, eps = TRUE, echo = FALSE, results = hide}
+\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}
 
@@
 \subsection{Results}
-<<mz_mod>>=
-load("wicherts05_study1.rda")
-
-d <- subset(d, !is.na(gpa))
-
+<<wdh_data>>=
+## data and transformations
+data("StereotypeThreat", package = "psychotools")
+## include group variable
+StereotypeThreat <- transform(StereotypeThreat, group = interaction(ethnicity, condition))
+StereotypeThreat <- StereotypeThreat[order(StereotypeThreat$group),]
+## omit NAs in GPA and break ties randomly
+StereotypeThreat <- subset(StereotypeThreat, !is.na(gpa))
+ties <- duplicated(StereotypeThreat$gpa)
+set.seed(1090)
+StereotypeThreat$gpa[ties] <- runif(sum(ties),
+  min = StereotypeThreat$gpa[ties] - 0.001, max = StereotypeThreat$gpa[ties] + 0.01)
+## keep only data with NAs in GPA
+wdh <- subset(StereotypeThreat, !is.na(gpa))
+@
+
+<<wdh_model>>=
 ## fit model 5b
@@
     verbal    ~~ label(rep("var_v", 4)) * verbal',
-  data = d, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")
-
-## Refit model with new identification constraints
+  data = wdh, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")
+
+## refit model with new identification constraints
 wdh5b.2 <- lavaan(
@@
     verbal    ~~ label(rep("var_v", 4)) * verbal',
-  data = d, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
+  data = wdh, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
 @
@@
 
+<<gefp-4-1>>=
+## FIXME It seems like info_full should go in lavaan_S3.R.
+info_full <- function(x, ...) solve(vcov(x) * nobs(x))
+scores_min_t <- function(x, ...) {
+  ef <- estfun(x, ...)
+  ef[wdh$group == "minority.threat", ]
+}
+gpa_min_t <- subset(wdh, group == "minority.threat")$gpa
+
+## empirical fluctuation processes (4 vs 1 par)
+gefp_4_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
+gefp_1_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
+
+# processes for second identification constraints
+gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
+gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
+@
+
 \begin{figure}
@@
 \setkeys{Gin}{width=\textwidth}
-<<gefp-4-1, fig=TRUE, height=8, width=6>>=
-## add a small amount of random noise to duplicated GPAs, so that everything
-## works with gefp().  (Without this, one can still obtain DM and CvM test 
-##                      results but not supLM results.)
-set.seed(1090)
-ties <- duplicated(d$gpa)
-d$gpa[ties] <- runif(sum(ties), min = d$gpa[ties] - .001, max = d$gpa[ties] + .001)
-
-## FIXME It seems like info_full should go in lavaan_S3.R.
-info_full <- function(x, ...) solve(vcov(x) * nobs(x))
-scores_min_t <- function(x, ...) {
-  ef <- estfun(x, ...)
-  ef[d$group == "exp:minority", ]
-}
-gpa_min_t <- subset(d, group == "exp:minority")$gpa
-
-## empirical fluctuation processes (4 vs 1 par)
-gefp_4_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
-gefp_1_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
-
-# processes for second identification constraints
-gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
-gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
-
-## visualization
+<<gefp-4-1-plot, fig=TRUE, height=8, width=6>>=
 par(mfcol = c(3, 2))

### 2012-01-16 www/psychoco2012/invariance.tex
Added files for Psychoco 2012 talk in directory psychoco2012.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@87 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{beamer}

\usepackage{beamerthemeHannover,graphics,amsmath,bm}

\title[Measurement Invariance]{Generalized Measurement Invariance
  Tests for Factor Analysis}
\author{Ed Merkle\inst{1} \and Achim Zeileis\inst{2}}
\institute{\inst{1} University of Missouri \and \inst{2} Universit\"{a}t Innsbruck}
\date{Supported by grant SES-1061334 from the U.S.\ National
  Science Foundation}

\begin{document}

\frame{\titlepage}

\frame{
  \frametitle{}
  \tableofcontents}

\section{Background}
% Define measurement invariance
\frame{
  \frametitle{Measurement Invariance}
  \begin{itemize}
    \item Measurement invariance: Sets of tests/items consistently
      assigning scores across diverse groups of individuals.\\ \ \\
    \item Notable violations of measurement invariance:
      \begin{itemize}
        \item SAT for different ethnic groups (Atkinson, 2001)
        \item Intelligence tests \& the Flynn effect
          (Wicherts et al., 2004)
      \end{itemize}
  \end{itemize}}

% FA path diagram, types of MI
\frame{
  \frametitle{Studying Measurement Invariance}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{famod.pdf}
  \end{center}}

% Define as theta_i = theta_0, i=1,..,n
\frame{
  \frametitle{Studying Measurement Invariance}
  \begin{itemize}
    % \item There exist many types of measurement invariance,
    %   based on specific subsets of parameters being allowed
    %   to vary across individuals (e.g., Meredith, 1993).\\ \ \\
    \item Hypothesis of ``full'' measurement invariance:\\
      \begin{center}
          $H_0: \bm{\theta}_i = \bm{\theta}_0, i=1,\ldots,n$\\
          $H_1: \text{Not all the }\bm{\theta}_i = \bm{\theta}_0$
      \end{center}
  \end{itemize}}

% Too difficult to assess, so use an auxiliary variable to place
% people in meaningful groups
  % continuous vs categorical auxiliary variable
\frame{
  \frametitle{Studying Measurement Invariance}
  \begin{itemize}
    \item $H_0$ from the previous slide is difficult to fully assess
      due to all the ways by which individuals may differ.\\ \ \\
    \item We typically place people into groups based on a meaningful
      auxiliary variable, then study measurement invariance across
      those groups (via Likelihood Ratio tests, Lagrange multiplier
      tests, Wald tests).
  \end{itemize}}

\section{Proposed Tests}
\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item In contrast to existing tests of measurement invariance, the
      proposed tests offer the abilities to:
      \begin{itemize}
        \item Test for measurement invariance when groups are
          ill-defined (e.g., when the grouping variable is continuous).
        \item Test for measurement invariance in any subset of model
          parameters.
        \item Interpret the nature of measurement invariance violations.
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item The proposed family of tests rely on first derivatives of
      the model's log-likelihood function.
  % Equation of gradient vs scores for individuals
  \end{itemize}
\begin{eqnarray*}
  \displaystyle\sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) =
  {\bm 0},\text{ where} \\
 \\
  \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
\end{eqnarray*}}

\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item We can also consider individual terms ({\em{scores}}) of the
      gradient.  These scores tell us how well a particular
      parameter describes a particular individual.  If your score is
      0 for some parameter, that parameter describes you well.  If your score
      is far from 0, that parameter describes you poorly. \\ \ \\
  \end{itemize}
  \begin{equation*}
  \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
  \text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}      
  \end{equation*}}

\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item Under measurement invariance, parameter estimates should
      roughly describe everyone equally well.  So people's scores
      should fluctuate around zero.\\ \ \\
    \item If measurement invariance is violated, the scores should
      stray from zero.
  \end{itemize}}

\frame{
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by an auxiliary variable.\\ \ \\
        \item Define $t \in (1/n, n)$.
          The {\em empirical cumulative score process} is defined by:\\
      \end{itemize}
  \end{itemize}
  \begin{center}
    $\bm{W}_n(t,\bm{\theta}) = \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} \psi ({\bm{x}}_i, \bm{\theta}).$\\ \ \\
  where $\lfloor nt \rfloor$ is the integer part of $nt$.
\end{center}}

% EXAMPLE OF CUMULATIVE SCORE
% \frame{
%   \frametitle{Cumulative Scores}
% \begin{center}
% \begin{tabular}{rrrr}
%   \hline
%  & Age & Score & Cumul Sc \\ 
%   \hline
% 1 & 13.0 & -0.24 & -0.24 \\ 
%   2 & 13.5 & 0.17 & -0.08 \\ 
%   3 & 14.0 & -0.38 & -0.45 \\ 
%   4 & 14.5 & 0.27 & -0.19 \\ 
%   5 & 15.0 & -0.27 & -0.46 \\ 
%    \hline
% \end{tabular}
% \end{center}}

\frame{
  \frametitle{Theorem}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Theorem (Zeileis \& Hornik, 2007): Under the hypothesis of
      measurement invariance, the following functional central limit
      theorem holds:
  \end{itemize}
  \begin{center}
  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}}) 
\overset{d}{\rightarrow} {\bm W}^{0}(\cdot),$\\ \ \\
 where ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}$ is the observed
 information matrix and ${\bm W}^{0}(\cdot)$ is a
 $p$-dimensional Brownian bridge.
\end{center}}

% \frame{
%   \frametitle{Brownian Bridge}
%   \begin{itemize}
%     \item Brownian bridge: A sequence of normal random variables
%       $X(t)$ satisfying:
%     % notation
% \begin{eqnarray*}
%     \mu(t) & = & 0\ \forall\ t\\
%     \Gamma(t_1, t_2) & = & \text{min}(t_1, t_2)\\
%     X(0) = & X(1) & = 0.
% \end{eqnarray*}
%   \end{itemize}}

% Picture of Brownian bridge
\frame{
  \frametitle{Brownian Bridge}
  \begin{center}
      \includegraphics[height=3in]{brownian.pdf}
  \end{center}}

% \frame{
%   \frametitle{Theorem}
%   \begin{itemize}
%     \item Use of the theorem:
%       \begin{itemize}
%         \item The theorem gives us a way to construct statistical
%           tests of measurement invariance using the scores.
%         \item Various test statistics can be obtained depending on our
%           goals.
%         \item Ordering individuals based on an auxiliary variable
%           allows us to interpret measurement invariance violations in
%           the context of that variable.
%       \end{itemize}
%   \end{itemize}}

\frame{
  \frametitle{Software}
  \begin{itemize}
    \item To carry out the tests, we use
      \begin{itemize}
        \item \texttt{lavaan} for model estimation.
        \item A combination of \texttt{lavaan} and our own code for calculating scores of fitted models.
        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
      \end{itemize}
  \end{itemize}}

% INSERT WICHERTS EXAMPLE HERE
\section{Illustration}
\frame{
  \frametitle{Example}
  \begin{itemize}
    \item Example: Studying stereotype threat via factor analysis (Wicherts et al., 2005)
      \begin{itemize}
        \item Stereotype threat: Knowledge of stereotypes about one's social group might cause one to fulfill the stereotypes.
        \item Wicherts et al.\ study: 295 students were administered three intelligence tests.  Stereotypes were primed for half of the students.
        \item Groups defined by: Ethnicity (majority/minority) and whether or not stereotypes were primed.
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Model}
  \begin{itemize}
    \item To study the data, Wicherts et al.\ employed a series of four-group, one-factor models.
      \begin{itemize}
        \item General finding: Minorities with stereotype primes have different measurement parameters than other groups.
        \item Current example: Is measurement further impacted by academic performance (as measured by student GPA)?
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Model}
  \begin{itemize}
    \item We utilize a model employed by Wicherts et al., where four model parameters are specific to the ``minority, stereotype prime'' group.
      \begin{itemize}
        \item Test for measurement invariance in these parameters wrt the student GPA variable (either all four together or only the factor mean).
        \item Violations of measurement invariance imply that stereotype threat is more problematic for students of low (or high) GPA.
      \end{itemize}
  \end{itemize}}

% Insert new results
\frame{
  \frametitle{Results}
  \begin{center}
      \includegraphics[height=3in]{paper-gefp-4-1.pdf}
  \end{center}}

\frame{
  \frametitle{Simulation}
  \begin{itemize}
    \item Simulation: What is the power of the proposed tests?\\ \ \\
      \begin{itemize}
        \item Two-factor model, with three indicators each.
        \item Measurement invariance violation in three factor loading parameters, with magnitude from 0--4 standard errors.
        \item Sample size in $\{100, 200, 500\}$
        \item Model parameters tested in $\{3, 19\}$
        \item Three test statistics
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Simulation}
  \begin{center}
      \includegraphics[height=3in]{paper-mzsim-xyplot.pdf}
  \end{center}
}

\section{Conclusions}
\frame{
  \frametitle{Conclusions}
  \begin{itemize}
    \item Measurement invariance tests utilizing stochastic processes have 
      important advantages over existing tests:
      \begin{itemize}
        \item Isolating specific parameters that violate
          measurement invariance, allowing the researcher to define
          specific types of measurement invariance ``post hoc''
          instead of ``a priori''.
        \item Isolating
          groups of individuals whose parameter values differ.\\ \ \\
        \item Studying the impact of continuous variables on model estimates, without ``ruining'' the rest of the model.
      \end{itemize}
    \item Power is reasonable, with specific tests being better in
      specific circumstances.
  \end{itemize}}

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Detailed examination of test properties via simulation.\\ \ \\
    \item Extension to related psychometric issues (modification
      indices, mediation).
  \end{itemize}}

\frame{
  \frametitle{}
  \begin{itemize}
  \item Questions?
  \end{itemize}}

\end{document}

### 2012-01-23 www/psychoco2012/invariance.tex
Minor revisions to Psychoco 2012 slides, along with TODO's for revisions that will take longer than 2min.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@88 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+% TODO: Add something like Fig 3 from paper to show test stat construction?
+
 \section{Background}
@@
 % FA path diagram, types of MI
+% TODO: Overlays with different parameter values on graphic
 \frame{
@@
     %   to vary across individuals (e.g., Meredith, 1993).\\ \ \\
-    \item Hypothesis of ``full'' measurement invariance:\\
+    \item Hypothesis of ``full'' measurement invariance ($\bm{\theta}_i$ is the full parameter vector for individual $i$):\\
       \begin{center}
@@
 
+% TODO Consistent notation with paper
 \frame{
@@
       the model's log-likelihood function.
+    \item We can also consider individual terms ({\em scores}) of the gradient.  These scores tell us how well a particular parameter describes a particular individual.
   % Equation of gradient vs scores for individuals
@@
  \\
-  \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
+  \psi(\bm{x}_i, \hat{{\bm \theta}}) = \frac{\partial}{\partial \bm{\theta}} \log
 \text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
@@
 
-\frame{
-  \frametitle{Proposed Tests}
-  \begin{itemize}
-    \item We can also consider individual terms ({\em{scores}}) of the
-      gradient.  These scores tell us how well a particular
-      parameter describes a particular individual.  If your score is
-      0 for some parameter, that parameter describes you well.  If your score
-      is far from 0, that parameter describes you poorly. \\ \ \\
-  \end{itemize}
-  \begin{equation*}
-  \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
-  \text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}      
-  \end{equation*}}
+% \frame{
+%   \frametitle{Proposed Tests}
+%   \begin{itemize}
+%     \item We can also consider individual terms ({\em{scores}}) of the
+%       gradient.  These scores tell us how well a particular
+%       parameter describes a particular individual.  If your score is
+%       0 for some parameter, that parameter describes you well.  If your score
+%       is far from 0, that parameter describes you poorly. \\ \ \\
+%   \end{itemize}
+%   \begin{equation*}
+%   \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
+%   \text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}      
+%   \end{equation*}}
 
@@
       theorem holds:
+  \begin{equation}
+  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}}) 
+\overset{d}{\rightarrow} {\bm W}^{0}(\cdot),$
+  \end{equation}
+ where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
+ information matrix and ${\bm W}^{0}(\cdot)$ is a
+ $p$-dimensional Brownian bridge (where $p$ is the number of parameters).
   \end{itemize}
-  \begin{center}
-  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}}) 
-\overset{d}{\rightarrow} {\bm W}^{0}(\cdot),$\\ \ \\
- where ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}$ is the observed
- information matrix and ${\bm W}^{0}(\cdot)$ is a
- $p$-dimensional Brownian bridge.
-\end{center}}
+}
 
@@
 % Picture of Brownian bridge
-\frame{
-  \frametitle{Brownian Bridge}
-  \begin{center}
-      \includegraphics[height=3in]{brownian.pdf}
-  \end{center}}
+% \frame{
+%   \frametitle{Brownian Bridge}
+%   \begin{center}
+%       \includegraphics[height=3in]{brownian.pdf}
+%   \end{center}}
 
@@
 %   \end{itemize}}
-
-\frame{
-  \frametitle{Software}
-  \begin{itemize}
-    \item To carry out the tests, we use
-      \begin{itemize}
-        \item \texttt{lavaan} for model estimation.
-        \item A combination of \texttt{lavaan} and our own code for calculating scores of fitted models.
-        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
-      \end{itemize}
-  \end{itemize}}
 
@@
 
-% Insert new results
+% TODO: Break up results over multiple slides
 \frame{
@@
 
+% TODO: Add some detail on software
+\frame{
+  \frametitle{Software}
+  \begin{itemize}
+    \item To carry out the tests, we use
+      \begin{itemize}
+        \item \texttt{lavaan} for model estimation.
+        \item A combination of \texttt{lavaan} and our own code for calculating scores of fitted models.
+        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
+      \end{itemize}
+  \end{itemize}}
+
 \frame{
@@
       indices, mediation).
+   % TODO: URL for working paper
   \end{itemize}}

### 2012-01-24 www/psychoco2012/invariance.tex
Revised psychoco 2012 slides: - Added "overlay" figures to demonstrate invariance - Modified notation to match paper - Added some detail on software - Added some detail on test statistic construction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@89 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-% TODO: Add something like Fig 3 from paper to show test stat construction?
-
 \section{Background}
@@
 % FA path diagram, types of MI
-% TODO: Overlays with different parameter values on graphic
 \frame{
-  \frametitle{Studying Measurement Invariance}
+  \frametitle{Studying Measurement Invariance: Age <= 16}
   % path diagram illustrating parameters being the same for everyone
   \begin{center}
-      \includegraphics[height=3in]{famod.pdf}
+      \includegraphics[height=3in]{model_under16.pdf}
+  \end{center}}
+
+\frame{
+  \frametitle{Studying Measurement Invariance: Age > 16}
+  % path diagram illustrating parameters being the same for everyone
+  \begin{center}
+      \includegraphics[height=3in]{model_over16.pdf}
   \end{center}}
@@
 \frame{
-  \frametitle{Studying Measurement Invariance}
+  \frametitle{Hypotheses}
   \begin{itemize}
@@
   \end{itemize}}
+
+\frame{
+  \frametitle{Lack of Grouping}
+  \begin{itemize}
+    \item If we did not know the groups in advance, we could conduct a LRT wrt to all possible groups, taking the maximum.  The proposed tests are related to this idea.
+  \end{itemize}
+ % TODO Insert Fig 3
+}
 
@@
 
-% TODO Consistent notation with paper
 \frame{
@@
 \begin{eqnarray*}
-  \displaystyle\sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) =
+  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
   {\bm 0},\text{ where} \\
  \\
-  \psi(\bm{x}_i, \hat{{\bm \theta}}) = \frac{\partial}{\partial \bm{\theta}} \log
+  s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} \log
 \text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
@@
   \begin{center}
-    $\bm{W}_n(t,\bm{\theta}) = \frac{1}{\sqrt{n}}
+    $\bm{B}(\hat{\bm{\theta}}; t) = \frac{1}{\sqrt{n}}
     \displaystyle\sum_{i=1}^{\lfloor nt 
@@
   \begin{equation}
-  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}}) 
-\overset{d}{\rightarrow} {\bm W}^{0}(\cdot),$
+  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) 
+\overset{d}{\rightarrow} {\bm B}^{0}(\cdot),$
   \end{equation}
  where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
- information matrix and ${\bm W}^{0}(\cdot)$ is a
- $p$-dimensional Brownian bridge (where $p$ is the number of parameters).
+ information matrix and ${\bm B}^{0}(\cdot)$ is a
+ $p$-dimensional Brownian bridge ($p$ is the number of parameters).
   \end{itemize}
@@
 
-% TODO: Add some detail on software
 \frame{
@@
   \begin{itemize}
-    \item To carry out the tests, we use
+    \item To carry out the tests, we utilize
       \begin{itemize}
         \item \texttt{lavaan} for model estimation.
-        \item A combination of \texttt{lavaan} and our own code for calculating scores of fitted models.
+        \item \texttt{estfun()} for score extraction, which is (currently) a combination of our own code and \texttt{lavaan} code.
         \item \texttt{strucchange} for carrying out the proposed tests with the scores.
+          \begin{itemize}
+            \item Required input: Fitted model, function for score extraction, and information matrix (optional)
+            \item \texttt{gefp()} constructs the process.
+            \item \texttt{sctest()} and \texttt{plot()} calculate and visualize test statistics.
+          \end{itemize}
       \end{itemize}
@@
       indices, mediation).
-   % TODO: URL for working paper
+    \item Working paper: \url{http://econpapers.repec.org/RePEc:inn:wpaper:2011-09}
   \end{itemize}}

### 2012-01-26 www/psychoco2012/invariance.tex
Made small changes to lavaan-S3.R so that it works with the most recent devel version of lavaan (0.4-12): two slots in lavaan objects were moved.
Revised Psychoco 2012 slides:
  - Divided stereotype threat results into multiple graphs
  - Other minor changes to equations


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@90 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \frame{
-  \frametitle{Studying Measurement Invariance: Age <= 16}
+  \frametitle{Studying Measurement Invariance: Age $\leq$ 16}
   % path diagram illustrating parameters being the same for everyone
@@
 \frame{
-  \frametitle{Studying Measurement Invariance: Age > 16}
+  \frametitle{Studying Measurement Invariance: Age $>$ 16}
   % path diagram illustrating parameters being the same for everyone
@@
   \begin{itemize}
-    \item If we did not know the groups in advance, we could conduct a LRT wrt to all possible groups, taking the maximum.  The proposed tests are related to this idea.
-  \end{itemize}
- % TODO Insert Fig 3
-}
+    \item If we did not know the groups in advance, we could conduct a LRT wrt for each possible grouping, then take the maximum.  The proposed tests are related to this idea.
+  \begin{center}
+      \includegraphics[height=2.65in]{paper-lm-lr-19.pdf}
+  \end{center}
+  \end{itemize}}
 
@@
 
-\frame{
+\begin{frame}[fragile]
   \frametitle{Aggregating Scores}
@@
           The {\em empirical cumulative score process} is defined by:\\
+  \begin{equation*}
+    \bm{B}(\hat{\bm{\theta}}; t) = \frac{1}{\sqrt{n}}
+    \displaystyle\sum_{i=1}^{\lfloor nt 
+      \rfloor} \psi ({\bm{x}}_i, \bm{\theta}).
+    \end{equation*}
+  where $\lfloor nt \rfloor$ is the integer part of $nt$.
       \end{itemize}
   \end{itemize}
-  \begin{center}
-    $\bm{B}(\hat{\bm{\theta}}; t) = \frac{1}{\sqrt{n}}
-    \displaystyle\sum_{i=1}^{\lfloor nt 
-      \rfloor} \psi ({\bm{x}}_i, \bm{\theta}).$\\ \ \\
-  where $\lfloor nt \rfloor$ is the integer part of $nt$.
-\end{center}}
+\end{frame}
 
@@
 
-\frame{
+\begin{frame}[fragile]
   \frametitle{Theorem}
@@
       theorem holds:
-  \begin{equation}
-  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) 
-\overset{d}{\rightarrow} {\bm B}^{0}(\cdot),$
-  \end{equation}
+  \begin{equation*}
+  {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) 
+\overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
+  \end{equation*}
  where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
@@
   \end{itemize}
-}
+\end{frame}
 
@@
 
-% TODO: Break up results over multiple slides
 \frame{
-  \frametitle{Results}
+  \frametitle{Results for Single Parameters}
   \begin{center}
-      \includegraphics[height=3in]{paper-gefp-4-1.pdf}
+      \includegraphics[height=3in]{gefp_4_ind.pdf}
+  \end{center}}
+
+\frame{
+  \frametitle{Aggregated Results}
+  \begin{center}
+      \includegraphics[height=3in]{gefp_4_agg.pdf}
   \end{center}}
@@
         \item \texttt{lavaan} for model estimation.
-        \item \texttt{estfun()} for score extraction, which is (currently) a combination of our own code and \texttt{lavaan} code.
+        \item \texttt{estfun()} for score extraction, which is currently a combination of our own code and \texttt{lavaan} code.
         \item \texttt{strucchange} for carrying out the proposed tests with the scores.
@@
     \item Extension to related psychometric issues (modification
-      indices, mediation).
+      indices, mediation).\\ \ \\
     \item Working paper: \url{http://econpapers.repec.org/RePEc:inn:wpaper:2011-09}

### 2012-01-26 www/lavaan-S3.R

@@ -13,9 +13,9 @@ df.residual.lavaan <- function(object, ...) NULL
 estfun.lavaan <- function(x, group = NULL, ...)
 {
   ## observed data
-  X <- x@Sample@data.obs
+  X <- x@Data
   ## number variables/sample size
-  nvar <- x@Sample@nvar
+  nvar <- x@Model@nvar
   nobs <- x@Sample@ntotal
   ntab <- unlist(x@Sample@nobs)
 
@@ -53,7 +53,7 @@ estfun.lavaan <- function(x, group = NULL, ...)
 
     ## Junk matrices for multiplication
     J <- matrix(1, 1L, ntab[i]) ## FIXME: needed? better maybe rowSums/colSums?
-    J2 <- matrix(1, nvar, nvar)
+    J2 <- matrix(1, nvar[i], nvar[i])
     diag(J2) <- 0.5
 
     ## scores.H1 (H1 = saturated model)

### 2012-01-29 www/psychoco2012/invariance.tex
Some suggestions for improvement   - full parameter vector more explicit   - some brief discussion of test statistics   - simulation before empirical example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@92 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \frame{
-  \frametitle{Studying Measurement Invariance: Age $\leq$ 16}
+  \frametitle{Example (Age $\leq$ 16)}
   % path diagram illustrating parameters being the same for everyone
@@
 \frame{
-  \frametitle{Studying Measurement Invariance: Age $>$ 16}
+  \frametitle{Example (Age~$>$~16)}
   % path diagram illustrating parameters being the same for everyone
@@
     %   to vary across individuals (e.g., Meredith, 1993).\\ \ \\
-    \item Hypothesis of ``full'' measurement invariance ($\bm{\theta}_i$ is the full parameter vector for individual $i$):\\
+    \item Hypothesis of ``full'' measurement invariance:\\
       \begin{center}
@@
       \end{center}
+      where $\bm{\theta}_i = (\lambda_{i, 1, 1}, \dots, \psi_{i, 1, 1}, \dots, \varphi_{i, 1, 2}, \dots)^\top$
+      is the full $p$-dimensional parameter vector for individual $i$.
   \end{itemize}}
@@
       those groups (via Likelihood Ratio tests, Lagrange multiplier
-      tests, Wald tests).
+      tests, Wald tests).\\ \ \\
+    \item If we did not know the groups in advance, we could conduct
+      a LR or LM test for each possible grouping, then take the maximum.
+      Requires different critical values! (Can be obtained from proposed
+      tests.)
   \end{itemize}}
@@
   \frametitle{Lack of Grouping}
-  \begin{itemize}
-    \item If we did not know the groups in advance, we could conduct a LRT wrt for each possible grouping, then take the maximum.  The proposed tests are related to this idea.
   \begin{center}
-      \includegraphics[height=2.65in]{paper-lm-lr-19.pdf}
+      \includegraphics[height=3in]{paper-lm-lr-19.pdf}
   \end{center}
-  \end{itemize}}
+}
 
@@
 \begin{frame}[fragile]
-  \frametitle{Theorem}
+  \frametitle{Tests}
  % show that cumulative scores follow a Brownian bridge under H_0
   \begin{itemize}
-    \item Theorem (Zeileis \& Hornik, 2007): Under the hypothesis of
-      measurement invariance, the following functional central limit
+    \item Theorem: Under the hypothesis of
+      measurement invariance, a functional central limit
       theorem holds:
@@
  information matrix and ${\bm B}^{0}(\cdot)$ is a
- $p$-dimensional Brownian bridge ($p$ is the number of parameters).
+ $p$-dimensional Brownian bridge.\\ \ \\
+
+   \item Testing procedure: Compute an aggregated statistic of empirical
+     score process and compare with corresponding quantile of aggregated
+     Brownian motion.\\ \ \\
+
+    \item Test statistics: Special cases include double maximum (DM), Cram\'er-von Mises
+      (CvM), maximum of LM statistics.
   \end{itemize}
 \end{frame}
+
+\frame{
+  \frametitle{Simulation}
+  \begin{itemize}
+    \item Simulation: What is the power of the proposed tests?\\ \ \\
+      \begin{itemize}
+        \item Two-factor model, with three indicators each.
+        \item Measurement invariance violation in three factor loading parameters, with magnitude from 0--4 standard errors.
+        \item Sample size in $\{100, 200, 500\}$.
+        \item Model parameters tested in $\{3, 19\}$.
+        \item Three test statistics.
+      \end{itemize}
+  \end{itemize}}
+
+\frame{
+  \frametitle{Simulation}
+  \begin{center}
+      \includegraphics[height=3in]{paper-mzsim-xyplot.pdf}
+  \end{center}
+}
 
@@
 
-\frame{
-  \frametitle{Simulation}
-  \begin{itemize}
-    \item Simulation: What is the power of the proposed tests?\\ \ \\
-      \begin{itemize}
-        \item Two-factor model, with three indicators each.
-        \item Measurement invariance violation in three factor loading parameters, with magnitude from 0--4 standard errors.
-        \item Sample size in $\{100, 200, 500\}$
-        \item Model parameters tested in $\{3, 19\}$
-        \item Three test statistics
-      \end{itemize}
-  \end{itemize}}
-
-\frame{
-  \frametitle{Simulation}
-  \begin{center}
-      \includegraphics[height=3in]{paper-mzsim-xyplot.pdf}
-  \end{center}
-}
-
 \section{Conclusions}
@@
         \item Isolating
-          groups of individuals whose parameter values differ.\\ \ \\
+          groups of individuals whose parameter values differ.
         \item Studying the impact of continuous variables on model estimates, without ``ruining'' the rest of the model.
@@
           \begin{itemize}
-            \item Required input: Fitted model, function for score extraction, and information matrix (optional)
+            \item Required input: Fitted model, function for score extraction, and information matrix (optional).
             \item \texttt{gefp()} constructs the process.

### 2012-02-02 www/psychoco2012/invariance.tex
Minor changes to psychoco talk:   - Fixed some notation typos   - Modified labels of path diagrams   - A few wording changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@93 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor wording changes to psychoco slides.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@94 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
     \displaystyle\sum_{i=1}^{\lfloor nt 
-      \rfloor} \psi ({\bm{x}}_i, \bm{\theta}).
+      \rfloor} s(\hat{\bm{\theta}} ; \bm{x}_i).
     \end{equation*}
@@
         \item Test for measurement invariance in these parameters wrt the student GPA variable (either all four together or only the factor mean).
-        \item Violations of measurement invariance imply that stereotype threat is more problematic for students of low (or high) GPA.
+        \item Violations of measurement invariance imply that stereotype threat is more problematic for students of low or high GPA.
       \end{itemize}
@@
     \item Detailed examination of test properties via simulation.\\ \ \\
-    \item Extension to related psychometric issues (modification
-      indices, mediation).\\ \ \\
+    \item Extension to related psychometric issues.\\ \ \\
     \item Working paper: \url{http://econpapers.repec.org/RePEc:inn:wpaper:2011-09}

### 2012-02-08 www/rev1/response.txt
Added introduction to response letter, along with minor revisions.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@96 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
-Editor
+Accompanying this letter is our revised manuscript, titled "Generalized measurement invariance tests with application to factor analysis."  We found the reviews to be very helpful and believe the paper has improved as a result.
+We have added details on the use of our methods with categorical V, the ability of our methods to handle multiple groups models and mean structure, and the relationship of our methods to mixture models and moderated factor models.
+Importantly, we have also included an example with real data that involves a four group model with mean structure.
+We elected to use this real example to supplement, vs replace, the artificial example, however.
+The artificial example allows us to compare test statistics to one another in a situation where the true model is known, which is obviously impossible with real data, and the artificial example also serves as a gateway to the simulation.
+We are willing to reconsider this decision if deemed necessary, however.
+
+We look forward to receiving feedback on the manuscript.
+
+Sincerely,
+Ed Merkle
+
+Editor Comments
 
@@
 We now illustrate this approach in the data example.
-DONE
 
@@
 We now briefly describe this issue in the "Use in Practice" section of the Discussion.
-DONE
 
@@
 
-Response E.4. We have dropped the artificial example and added a real example. DONE
+Response E.4. We have dropped the artificial example and added a real example.
 
@@
 
-Response 1.1. We have revised the theory development and added extra detail to the text.  If some specific concepts are still unclear, we are happy to revisit this section.  DONE
+Response 1.1. We have revised the theory development and added extra detail to the text.  If some specific concepts are still unclear, we are happy to revisit this section.
 
@@
 Some advantages of our approach include the facts that it does not require a priori specification of the number of latent classes or a priori specification of the way by which age impacts class membership.
-DONE
 
@@
 
-Response 1.3. We now describe factor analysis models employing moderation in the "Comparison to other methods" section of the discussion.  DONE
+Response 1.3. We now describe factor analysis models employing moderation in the "Comparison to other methods" section of the discussion.
 
@@
 In the real data example that we have added, however, the mean structure is utilized.
-DONE
 
@@
 We have now provided more details on these issues in the Discussion section (and tied the details to Comment/Response 1.3).
-DONE
 
@@
 
-Response 1.6. We agree that these statements were too strong and have revised accordingly.  DONE
+Response 1.6. We agree that these statements were too strong and have revised accordingly.
 
@@
 
-Response 1.7. We now describe the invariance definitions in more detail, and we also clarify that our methods are not necessarily robust to violations of normality.  DONE
+Response 1.7. We now describe the invariance definitions in more detail, and we also clarify that our methods are not necessarily robust to violations of normality.
 
@@
 
-Response 1.8. See Response E.4.  DONE
+Response 1.8. See Response E.4.
 
@@
 
-Response 3.1. We now include the n=50 condition in the simulations, which indeed exhibits a breakdown in the sense that power is not monotonic with respect to violation magnitude.  DONE
+Response 3.1. We now include the n=50 condition in the simulations, which indeed exhibits a breakdown in the sense that power is not monotonic with respect to violation magnitude.
 
@@
 We also discuss these results in the context of the simulation, showing the impact of our imposed measurement invariance violation on the Brownian bridge.
-DONE
 
@@
 We have modified this section of text accordingly.
-DONE
 
@@
 
-Response 4.1. We have now added this detail.  DONE
+Response 4.1. We have now added this detail.
 
@@
 
-Response 4.2. We now note that Andrews (1993) provides results and references relevant to Bessel processes.  DONE
+Response 4.2. We now note that Andrews (1993) provides results and references relevant to Bessel processes.
 
@@
 We have added some extra detail to the simulation section on our use of a single-group model.
-DONE
 
@@
 Also see Comment/Response 1.5.
-DONE
 
@@
 
-Response 4.6. We have added some brief discussion surrounding extensions to structural equation models.  DONE
+Response 4.6. We have added some brief discussion surrounding extensions to structural equation models.
 
@@
 Response 4.7. We have moved the Abbreviations part of the captions.
-DONE

### 2012-02-08 www/psychoco2012/invariance.tex
Final revisions to psychoco talk: - Modified individual score processes to have correct critical values - Added a path diagram for the data example - Minor wording changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@97 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
       \end{center}
-      where $\bm{\theta}_i = (\lambda_{i, 1, 1}, \dots, \psi_{i, 1, 1}, \dots, \varphi_{i, 1, 2}, \dots)^\top$
+      where $\bm{\theta}_i = (\lambda_{i, 1, 1}, \dots, \psi_{i, 1, 1}, \dots, \varphi_{i, 1, 2})^\top$
       is the full $p$-dimensional parameter vector for individual $i$.
@@
       the model's log-likelihood function.
-    \item We can also consider individual terms ({\em scores}) of the gradient.  These scores tell us how well a particular parameter describes a particular individual.
+    \item We consider individual terms ({\em scores}) of the gradient.  These scores tell us how well a particular parameter describes a particular individual.
   % Equation of gradient vs scores for individuals
@@
   \begin{itemize}
-    \item Theorem: Under the hypothesis of
+    \item Under the hypothesis of
       measurement invariance, a functional central limit
@@
 
-   \item Testing procedure: Compute an aggregated statistic of empirical
+   \item Testing procedure: Compute an aggregated statistic of the empirical
      score process and compare with corresponding quantile of aggregated
@@
       \begin{itemize}
-        \item Test for measurement invariance in these parameters wrt the student GPA variable (either all four together or only the factor mean).
+        \item Test for measurement invariance in these parameters wrt the student GPA variable (either all four together or individually).
         \item Violations of measurement invariance imply that stereotype threat is more problematic for students of low or high GPA.
@@
   \end{itemize}}
+
+\frame{
+  \frametitle{Model}
+  \begin{center}
+      \includegraphics[height=3in]{wicherts_msp.pdf}
+  \end{center}}
 
@@
     \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
-    \item Detailed examination of test properties via simulation.\\ \ \\
+    \item Detailed examination of test properties.\\ \ \\
     \item Extension to related psychometric issues.\\ \ \\

### 2012-02-08 www/paper.Rnw
Z revision after discussion ECM, only a few FIXMEs remain plus some READMEs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@98 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   that violated measurement invariance based on a continuous auxiliary variable. 
-  The tests are presented and illustrated in detail, along with simulations
-  examining the tests' abilities in controlled conditions.
+  The tests are presented and illustrated in detail, along with an application
+  to a study of stereotype threat and simulations examining the tests'
+  abilities in controlled conditions.
 }
@@
 The methods proposed here are generally relevant to situations where the
-$p$-dimensional random variable $X$ with associated observations $\bm{x}_i, i=1,\cdots,n$
+$p$-dimensional random variable $X$ with associated observations $\bm{x}_i, i=1,\dots,n$
 is specified to arise from a model with density $f(\bm{x}_i; \bm{\theta})$ and
@@
     \bm{\mu}(\bm{\theta}))^{\top} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
-    \bm{\mu}(\bm{\theta}) \right\}, \\
+    \bm{\mu}(\bm{\theta})) \right\}, \\
     \label{eq:caselik}
@@
 form of the likelihood, but it is presented for illustration due to
-its importance in practice.  Many expositions of factor analysis
-utilize the likelihood for the sample covariance matrix, which is
-based on a Wishart distribution when the $\bm{x}_i$ are assumed to be
-multivariate normal.
-However, the techniques presented below require the casewise contributions
-to the likelihood; this situation is also generally
-encountered in structural equation models with missing data (e.g.,
-\citeNP{Wot00}).
+its importance in practice.
+
+Many expositions of factor analysis utilize the likelihood for the sample
+covariance matrix (instead of the likelihood of the individual observations $\bm{x}_i$),
+employing a Wishart distribution. However, in that approach the sample covariance
+matrix can also be disaggregated to the sum of its individual-specific contributions
+leading essentially to a likelihood like \eqref{eq:mvndensity} with multivariate normal
+observations $\bm{x}_i$. This situation is also generally encountered in structural
+equation models with missing data (e.g., \citeNP{Wot00}).
+
+\readme{maybe like this?}
 
@@
 although the asymptotically equivalent  Lagrange multiplier
-(LM) and Wald tests may also be constructed for this purpose (see
+(LM) tests (also known as score tests) and Wald tests may also be constructed for this purpose (see
 \citeNP{Sat89}). If $\nu$
@@
 
-Analogously to the LR~test, the Wald test and LM test (also known as score test)
+Analogously to the LR~test, the Wald test and LM test
 can be employed to test the null hypothesis $H_1^*$ for a fixed threshold $\nu$.
@@
 \overline{\nu}]$.  
-The other two tests require reestimation of the subgroup model for each
+The other two tests require reestimation of the subgroup models for each
 $\nu$.
@@
 at which the parameters change. Alternatively, only one (or just a few) of the parameters in
-the vector ${\bm \theta}$ change while the remaining parameters are
+the vector ${\bm \theta}$ may change while the remaining parameters are
 constant (a common occurrence in psychometric models).
@@
 e.g., their outer product or the observed information matrix.
-The $k$ dimensions of the 
+The $k$~dimensions of the 
 process arise from the fact that a separate cumulative score is
@@
 returns to zero at $t = 1$. Furthermore, multiplication by
-$\hat {\bm I}^{-1/2}$ ``decorrelates'' the $k$ cumulative score
+$\hat {\bm I}^{-1/2}$ ``decorrelates'' the $k$~cumulative score
 processes, such that each univariate process (i.e., each
@@
 each individual parameter can be displayed along with the appropriate
-critical value.  An example of this visualization appears in the 
-example section (first row, second column of Figure~\ref{fig:cusumex}).
+critical value. For an example of this see Figure~\ref{fig:lambdas}
+(in the ``Example'' section) which shows the the cumulative score processes for three
+factor loadings along with the critical values at 5\% level.
 
@@
 have a non-zero mean.  Hence, the corresponding tests will have non-trivial
-power (asymptotically).  We use these results in the Simulation
+power (asymptotically).  We use these results in the ``Simulation''
 section to describe the expected behavior of the Brownian bridge.
@@
 is particularly appealing for this because the $k$-dimensional empirical cumulative
-score process can be graphed along with boundaries for the associated critical values.
+score process can be graphed along with boundaries for the associated critical values
+(see Figure~\ref{fig:lambdas}).
 Boundary crossing then implies a violation of measurement invariance, and the location of
@@
 For the maximum LR/Wald/LM tests, it is natural to graph the sequence of LR/Wald/LM
-statistics along $V$, with a boundary corresponding to the critical value.
+statistics along $V$, with a boundary corresponding to the critical value
+(see Figure~\ref{fig:lrlm} and the third row of Figure~\ref{fig:cusumex}).
 Again, a boundary crossing signals a significant violation, and the peak(s) in the
@@
 score process may still be valuable for gaining some insights (see,
-e.g., Figure~\ref{fig:cusumex}).
+e.g., the second row of Figure~\ref{fig:cusumex}).
 
@@
 be determined by maximizing the corresponding segmented log-likelihood over all
-possible combinations of thresholds (\citeNP{ZeiSha10} adapt a dynamic programming
+possible combinations of thresholds (\citeNP{ZeiSha10}, adapt a dynamic programming
 algorithm to this task). For the single threshold case, this reduces to
@@
 the peaks in the Wald/LM sequences and the cumulative score process, respectively, will occur at the
-same threshold asymptotically. However, in empirical samples, their location may differ somewhat
+same threshold asymptotically. However, in empirical samples, their location may differ
 (although often not by much).
@@
 existing tests, as existing measurement invariance methods cannot:
-(1) isolate specific parameters violating measurement invariance, or
-(2) test measurement invariance for unknown $\nu$. 
+(1)~test measurement invariance for unknown $\nu$, or
+(2)~isolate specific parameters violating measurement invariance.
 In particular, \citeA{Mil05} cites
@@
 
+\section{Example with Artificial Data}
+
 \begin{figure}
 \caption{Path diagram representing the base factor analysis model used for
-  the example and simulations.  To induce measurement invariance
+  the artificial example and simulations.  To induce measurement invariance
   violations, a seventh observed variable (student age) determines the
@@
 
-\section{Example with Artificial Data}
-
-Consider a 
-hypothetical battery of six scales administered to students aged 13 to 18
+Consider a hypothetical battery of six scales administered to students aged 13 to 18
 years.  Three of the scales are intended to measure verbal ability, and
@@
 \begin{equation}
-    \label{eq:exh0}
+    \label{eq:simh0}
 H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
@@
 from the overall $k = 19$~model parameters (including means) are assessed.
+\fixme{Explain what the 19 parameters are.}
 Alternatively, we can consider all $k^* = k = 19$~parameters, leading to a test
@@
 \subsection{Results}
-In the results section, we first describe overall results.  We then 
-describe estimation of $\nu$ and 
-isolation of model parameters violating
+
+Here, we first describe overall results and subsequently
+the identification of $\nu$ and isolation of model parameters violating
 measurement invariance.
@@
 
+<<gefp-3-19>>=
+## empirical fluctuation processes (3 vs all 19 pars)
+gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
+  vcov = info.mzfit, sandwich = FALSE)
+gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
+  vcov = info.mzfit, sandwich = FALSE)
+@
+
 \subsubsection{Overall Results}
-Test statistics for the hypotheses \eqref{eq:exh0} and~\eqref{eq:h0}
-are displayed in  
-Figure~\ref{fig:cusumex}.
-Each panel displays a test 
+Test statistics for the hypotheses \eqref{eq:simh0} and~\eqref{eq:h0}
+are displayed in Figure~\ref{fig:cusumex}. Each panel displays a test 
 statistic's fluctuation across values of student age, with the first
-column containing tests of~\eqref{eq:exh0} and the second column
+column containing tests of~\eqref{eq:simh0} and the second column
 containing tests of~\eqref{eq:h0}.  The solid
@@
 the maxima of the processes).
-In other words, for panels in the first and third rows, \eqref{eq:exh0} is
+In other words, for panels in the first and third rows, \eqref{eq:simh0} is
 rejected if the process crosses the horizontal line.  For panels in
@@
 for the double-max test as it is more 
-sensitive to fluctuations among a small subset of parameters.
-
-Figure~\ref{fig:lrlm} 
-compares the $\max \mathit{LM}$ statistic (solid line) to the $\max \mathit{LR}$
-statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
-\eqref{eq:h0}.  The critical values for these two
-tests are identical, hence the single horizontal line.
-The figure
-shows that the two statistics are very similar to one another, with
-both maxima at the dotted vertical line.  This
-is generally to be expected, because the two tests are asymptotically
-equivalent.  The 
-$\max \mathit{LR}$~statistic cannot be obtained from the empirical fluctuation
-process, however, so the factor analysis model must be refitted before
-and after each of the possible threshold values $\nu$ (i.e., here
-$\Sexpr{2 * length(mz.lrstat)}$~model fits for $\Sexpr{length(mz.lrstat)}$~thresholds).
-
-\subsubsection{Estimation of $\nu$ and Parameter Isolation}
-As described above, the tests of~\eqref{eq:exh0} imply that the 
-verbal scales lack measurement invariance.  
-We can also use the tests to: (1)~estimate the 
-threshold $\nu$, and (2)~isolate specific parameters that violate
-measurement invariance.  
-For example, as described previously, estimates of $\nu$ can be obtained by
-examining the peaks in Figure~\ref{fig:cusumex}.  For all six panels
-in the figure, the peaks occur near an age of 16.1.  This agrees well
-with the true threshold of 16.0.
-
-As mentioned previously, the double-max test is advantageous because
-it yields information about individual 
-parameters violating measurement invariance.  
-That is, it allows us to examine whether or not individual parameters' 
-cumulative score processes lead one to reject the hypothesis of
-measurement invariance.
-Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
-verbal factor loadings, with the horizontal lines reflecting the
-critical value at $\alpha= 0.05$.
-  The figure
-shows that the third parameter (i.e., $\lambda_{31}$) crosses the
-dashed line, so we would conclude a measurement 
-invariance violation with respect to age for the third
-verbal test.  The fact that the cumulative score process for the first
-and second loadings did not achieve the critical value represents a Type II
-error, which implicitly brings into question the tests' power.
-This is an example of a situation where the double-max test ``wastes''
-power, as it cannot make use of the fact that multiple model
-parameters change simultaneously.  We
-generally address the issue of power in the simulations below.
-
-\begin{figure}
-\caption{Three test statistics of~\eqref{eq:exh0} (with $k^* = 3$)
-  and~\eqref{eq:h0} (with $k^* = 19$), 
-  based on the example involving measurement invariance with respect
-  to student age.  Solid, horizontal lines represent critical values
-  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
-  represent values of the Cram\'{e}r-von Mises test statistic.}
-\label{fig:cusumex}
-\setkeys{Gin}{width=\textwidth}
-<<gefp-3-19, fig=TRUE, height=8, width=6>>=
-## empirical fluctuation processes (3 vs all 19 pars)
-gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
-  vcov = info.mzfit, sandwich = FALSE)
-gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
-  vcov = info.mzfit, sandwich = FALSE)
-
-## visualization
-par(mfcol = c(3, 2))
-plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
-plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
-@
-\end{figure}
+sensitive to fluctuations among a small subset of parameters
+(3 out of 19 here).
 
@@
 
+Figure~\ref{fig:lrlm} 
+compares the $\max \mathit{LM}$ statistic (solid line) to the $\max \mathit{LR}$
+statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
+\eqref{eq:h0}. (Note also that the visualization of the $\max \mathit{LM}$ in
+Figure~\ref{fig:lrlm} is identical to the bottom right panel of Figure~\ref{fig:cusumex}.)
+The critical values for these two
+tests are identical, hence the single horizontal line.
+The figure
+shows that the two statistics are very similar to one another, with
+both maxima at the dotted vertical line.  This
+is generally to be expected, because the two tests are asymptotically
+equivalent.  The 
+$\max \mathit{LR}$~statistic cannot be obtained from the empirical fluctuation
+process, however, so the factor analysis model must be refitted before
+and after each of the possible threshold values $\nu$ (i.e., here
+$\Sexpr{2 * length(mz.lrstat)}$~model fits for $\Sexpr{length(mz.lrstat)}$~thresholds).
+
+\subsubsection{Interpretation of Test Results}
+As described above, the tests of~\eqref{eq:simh0} imply that the 
+verbal scales lack measurement invariance.  
+We can also use the tests to: (1)~locate the threshold $\nu$, and
+(2)~isolate specific parameters that violate measurement invariance.  
+For example, as described previously, information about the location of $\nu$
+can be obtained by examining the peaks in Figure~\ref{fig:cusumex}.  For all six panels
+in the figure, the peaks occur near an age of 16.1.  This agrees well
+with the true threshold of 16.0.
+
 \begin{figure}
-\caption{Cumulative score processes for each verbal factor
-  loadings, with critical values stemming from the double-max test. 
-  The solid, horizontal lines
-  correspond to the critical value at $\alpha= 0.05$.}
+\centering
+\caption{Cumulative score processes for each verbal factor loading. 
+  The horizontal lines correspond to the critical value of the
+  double-max test at $\alpha= 0.05$.}
 \label{fig:lambdas}
-\setkeys{Gin}{width=\textwidth}
-<<gefp-lambdas, fig=TRUE, height=8, width=6>>=
+\setkeys{Gin}{width=0.8\textwidth}
+<<gefp-3-visualization, fig=TRUE, height=8, width=5.5>>=
 plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
-  main = "", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
+  main = "DM, k* = 3", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
 @
@@
 
+As mentioned previously, the double-max test is advantageous because
+it yields information about individual parameters violating measurement invariance.  
+That is, it allows us to examine whether or not individual parameters' 
+cumulative score processes lead one to reject the hypothesis of
+measurement invariance.
+Figure~\ref{fig:lambdas} shows the individual cumulative score processes for the
+three verbal factor loadings and the top left panel of Figure~\ref{fig:cusumex} shows
+the same processes aggregated over the three parameters. In both graphics, horizontal lines reflecting the
+same critical value at $\alpha= 0.05$. The figure
+shows that the third parameter (i.e., $\lambda_{31}$) crosses the
+dashed line, so we would conclude a measurement 
+invariance violation with respect to age for the third
+verbal test.  The fact that the cumulative score process for the first
+and second loadings did not achieve the critical value represents a type~II
+error, which implicitly brings into question the tests' power.
+This is an example of a situation where the double-max test ``wastes''
+power, as it cannot make use of the fact that multiple model
+parameters change simultaneously.  We
+generally address the issue of power in the simulations below.
+
+\begin{figure}
+\caption{Three test statistics of~\eqref{eq:simh0} (with $k^* = 3$)
+  and~\eqref{eq:h0} (with $k^* = 19$), 
+  based on the example involving measurement invariance with respect
+  to student age.  Solid, horizontal lines represent critical values
+  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
+  represent values of the Cram\'{e}r-von Mises test statistic.}
+\label{fig:cusumex}
+\setkeys{Gin}{width=\textwidth}
+<<gefp-3-19-visualization, fig=TRUE, height=8, width=6>>=
+par(mfcol = c(3, 2))
+plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
+plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
+@
+\end{figure}
+
 \section{Simulation}
@@
 In this section, we conduct a simulation designed to
-examine the tests' power and Type I error rates in situations where
+examine the tests' power and type~I error rates in situations where
 measurement invariance violations are known to exist.  Specifically,
@@
 Data were generated from the same model and with the same type of
-measurement invariance as was described in the Artificial Example section.
+measurement invariance as was described in the ``Artificial Example'' section.
 Sample size and magnitude
@@
 with $d = 0, 0.25, 0.5, \dots, 4$.  The 0-standard error condition was
-used to study Type I error rate.
+used to study type~I error rate.
 
@@
 Mises test can be as large as 0.1, most notably when three parameters
-are being tested.
-  It also shows that the Cram\'{e}r-von Mises test generally
-has true Type-I error rates, with the double-max test being somewhat
+are being tested. It also shows that the Cram\'{e}r-von Mises test generally
+is very close to its nominal type~I error rate, with the double-max test being somewhat
 conservative and the $\max \mathit{LM}$ test being slightly liberal.
@@
 
-\section{Example: Stereotype Threat}
-
-\citeA{WicDol05} utilized confirmatory
+\section{Application: Stereotype Threat}
+
+\readme{now called ``Application'' as opposed to the artificial ``Example''}
+
+\citeA{WicDol05}, henceforth WDH, utilized confirmatory
 factor models to study measurement invariance in a series of
@@
 
-To study stereotype threat in a measurement invariance framework, Study 1 of
+\fixme{Include path diagram and a bit more background so that the paper is more
+accessible for readers that are not familiar with the WDH paper.}
+
+To study stereotype threat in a measurement invariance framework, Study~1 of
 \citeA{WicDol05} involved 295 high school students
@@
 Conditions were defined by whether or not students received primes about ethnic
-stereotypes prior to testing. To study the data, Wicherts et al.\ employed a
+stereotypes prior to testing. To study the data, WDH employed a
 four-group, one-factor model with the three intelligence tests as manifest
@@
 other three groups on the most difficult intelligence test.  In the
-example below, we employ one of the models used by Wicherts et al.
-Our $V$ variable is participants' grade point averages, which were unused in the original analyses.
+example below, we employ one of the models used by WDH.
+Our $V$ variable is participants' grade point averages (GPAs), which were unused in the original analyses.
+\fixme{Maybe add: ``\dots as a measure of aptitude.'' Or something along these lines? I'm not
+sure what the appropriate psychometric terminology is.}
 
 \subsection{Method}
-Wicherts et al.\ tested a series of confirmatory factor models for
+WDH tested a series of confirmatory factor models for
 various types of measurement invariance.  We focus on the model used
@@
 the observed information matrix.  We then
-calculated the cumulative score process wrt student GPA via
+calculated the cumulative score process with respect to student GPA via
 \eqref{eq:cumscore}, allowing us to
@@
 As mentioned in the theory section, the 
-tests give us the flexibility to study hypotheses of partial change.
-That is, we have the ability to 
-test various subsets of 
-parameters.  For the Wicherts et al.\ data, we first test
+tests give us the flexibility to study hypotheses of partial change. That is, we
+have the ability to  test various subsets of  parameters.  For the WDH data, we
+first test
 \begin{equation}
     \label{eq:exh0}
-H_0:\ (\nu_{i}\ \lambda_{i}\
-\theta_{i}\ \alpha_{i}) = 
-(\nu_{0}\ \lambda_{0}\
-\theta_{0}\ \alpha_{0}),\ i=1,\ldots,n_{\text{MSP}},
-\end{equation}
-where $\nu_{i}$ is the intercept for student $i$ on intelligence
-test 1, $\lambda_{i}$ the factor loading, $\theta_{i}$ the unique
-variance, and $\alpha_i$ the factor mean.
-The individuals are ordered according to GPA, and the 
-hypothesis is taken only wrt the $n_{\text{MSP}}$ individuals in the
-``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states 
-that these four group-specific model parameters are invariant with
-respect to GPA.
-
-We can construct a similar hypothesis involving only the intercept
-parameter
-\begin{equation}
-    \label{eq:exh0.2}
-H_0:\ \nu_{i}\ = \nu_{0},\ i=1,\ldots,n_{\text{MSP}},
-\end{equation}
-which, e.g., may be used to isolate the impact of stereotype threat.
-We consider tests of both \eqref{eq:exh0} and \eqref{eq:exh0.2}
-below.
+H_0:\ (\lambda_{i, \text{num}}\ \eta_{i}\ \mu_{i, \text{num}}\ \psi_{i, \text{num}}) = 
+      (\lambda_{0, \text{num}}\ \eta_{0}\ \mu_{0, \text{num}}\ \psi_{0, \text{num}}),\
+      i = 1,\ldots, n_{\text{MSP}},
+\end{equation}
+where
+$\lambda_{i, \text{num}}$ is the factor loading of student $i$'s intelligence on the numerical scale,
+$\eta_{i}$ is student $i$'s factor mean,
+$\mu_{i, \text{num}}$ is student $i$'s intercept on the numerical scale, and
+$\psi_{i, \text{num}}$ is student $i$'s unique variance on the numerical scale.
+The students are ordered according to GPA (i.e.,  in this application the variable $V$),
+and the  hypothesis is taken only with respect to the $n_{\text{MSP}}$ individuals in the
+``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states  that these four group-specific
+model parameters are invariant with respect to GPA.
+
+\readme{I have omitted the single parameter hypothesis. Still it would be good to
+double-check the estfun method once more, though.}
 
 \subsection{Results}
+
 <<wdh_data>>=
@@
   min = StereotypeThreat$gpa[ties] - 0.001, max = StereotypeThreat$gpa[ties] + 0.01)
-## keep only data with NAs in GPA
+## keep only data without NAs in GPA
 wdh <- subset(StereotypeThreat, !is.na(gpa))
@@
 ## refit model with new identification constraints
-wdh5b.2 <- lavaan(
+wdh5b_1 <- lavaan(
    'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
@@
 
-Test statistics for the hypotheses \eqref{eq:exh0}
-and~\eqref{eq:exh0.2} are displayed in Figure~\ref{fig:cusumex}.
-Each panel displays a test 
-statistic's fluctuation across values of student GPA, with the first
-column containing tests of~\eqref{eq:exh0} (testing $k^*=4$ parameters)
-and the second column containing tests of~\eqref{eq:exh0.2} (testing
-$k^*=1$ parameter).
-The solid
-horizontal 
-lines represent critical values for $\alpha= 0.05$, and
-the Cram\'{e}r-von Mises panels 
-also contain a dashed line depicting the value of the test
-statistic (test statistics for the others are simply 
-the maxima of the processes).
-In other words, for panels in the first and third rows, \eqref{eq:exh0} is
-rejected if the process crosses the horizontal line.  For panels in
-the second row, \eqref{eq:h0} is rejected if the dashed horizontal line is
-higher than the solid horizontal line.
-
-%% Details on test results
-The results displayed in Figure~\ref{fig:cusumex} generally show that the
-``minority, stereotype prime'' group lacks invariance with respect to
-GPA.  All three tests are significant at the .05 level, both when the four
-group-specific parameters are tested simultaneously (first column) and
-when the intercept parameter is tested individually (second column).
-
-We can also use the test results to estimate the threshold $\nu$,
-which in this context can be used to define GPA-based subgroups whose
-measurement parameters differ.  As described previously, estimates of
-$\nu$ can be obtained by 
-examining the peaks in Figure~\ref{fig:cusumex}.  For the $k^*=4$
-panels (first column), two peaks occur near GPAs of 5.9 and of 6.3 (6
-corresponds to the American `C' range).  These peaks roughly divide
-individuals into those receiving D's and F's, those receiving solid
-C's, and those receiving high C's, B's, and A's.  Parameter-specific
-results can also be obtained, as shown for the intercept parameter in
-the second column of Figure~\ref{fig:cusumex}.  The peak locations are
-generally similar to those of all four parameters, though this finding
-will not hold in general.
+The double-max test for the hypothesis~\eqref{eq:exh0} is shown in
+Figure~\ref{fig:cusumgpa}, displaying the cumulative score processes
+for each of the four group-specific parameters across GPA along
+with horizontal lines representing the critical value for $\alpha= 0.05$.
+This shows that the ``minority, stereotype prime'' group lacks invariance
+with respect to GPA,
+because one of the processes crosses its boundaries. Furthermore, as this
+process pertains to the variance $\psi_\text{num}$ of the numerical scale
+(bottom panel) it can be concluded that the invariance violation is associated
+with this parameter. However, also the factor loading for the numerical scale
+$\lambda_\text{num}$ and the factor mean $\eta$ exhibit increased fluctuation
+albeit they are just non-significant at level $\alpha = 0.05$. Only the process
+associated with the intercept $\mu_\text{num}$ just shows moderate (and hence
+clearly non-significant) fluctuation across GPA.
 
 <<gefp-4-1>>=
-## FIXME It seems like info_full should go in lavaan_S3.R.
 info_full <- function(x, ...) solve(vcov(x) * nobs(x))
@@
 # processes for second identification constraints
-gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
-gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
+gefp_4_info_1 <- gefp(wdh5b_1, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
+gefp_1_info_1 <- gefp(wdh5b_1, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
 @
@@
 \begin{figure}
-\caption{Three test statistics of~\eqref{eq:exh0} (set of
-  group-specific parameters)
-  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
-  using data from Study 1 of Wicherts et al.\ (2005).
-  Solid, horizontal lines represent critical values
-  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
-  represent values of the Cram\'{e}r-von Mises test statistic.}
-\label{fig:cusumex}
-\setkeys{Gin}{width=\textwidth}
-<<gefp-4-1-plot, fig=TRUE, height=8, width=6>>=
-par(mfcol = c(3, 2))
-plot(gefp_4_info,  functional = maxBB, main = "DM, k* = 4", xlab = "GPA")
-plot(gefp_4_info,  functional = meanL2BB, main = "CvM, k* = 4", xlab = "GPA")
-plot(gefp_4_info,  functional = supLM(0.1), main = "max LM, k* = 4", xlab = "GPA")
-plot(gefp_1_info, functional = maxBB, main = "DM, k* = 1", xlab = "GPA")
-plot(gefp_1_info, functional = meanL2BB, main = "CvM, k* = 1", xlab = "GPA")
-plot(gefp_1_info, functional = supLM(0.1), main = "max LM, k* = 1", xlab = "GPA")
+\centering
+\caption{Cumulative score processes for the four group-specific parameters across GPA
+  in the ``minority, stereotype prime'' group
+  using data from Study 1 of Wicherts et al.\ (2005). %% \citeA{WicDol05} does not work in caption
+  The horizontal lines correspond to the critical value of
+  the doule-max test at $\alpha = 0.05$.}
+\label{fig:cusumgpa}
+\setkeys{Gin}{width=0.8\textwidth}
+<<gefp-4-plot, fig=TRUE, height=8, width=5.5>>=
+plot(gefp_4_info,  functional = maxBB, aggregate = FALSE,
+  ylim = c(-2.15, 2.15), main = "DM, k* = 4", xlab = "GPA",
+  ylab = expression(lambda[num], eta, mu[num], psi[num]))
 @
 \end{figure}
+
+We can also use the test results to locate the threshold $\nu$,
+which in this context can be used to define GPA-based subgroups whose
+measurement parameters differ.  As described previously, information about
+the location of $\nu$ can be obtained by 
+examining the peaks in Figure~\ref{fig:cusumgpa}. The main peak occurs
+around a GPA of~6.3 and a second one near~5.9
+(6~corresponds to the American `C' range).  These peaks roughly divide
+individuals into those receiving D's and F's, those receiving solid
+C's, and those receiving high C's, B's, and A's.
+
+In addition to the double-max test, the Cram\'er-von Mises or $\max \mathit{LM}$
+statistics could also be employed. They also lead to clearly significant
+violations of measurement invariance ($p < 0.005$ for both tests) and
+the corresponding visualizations bring out similar
+peaks as the double-max test in Figure~\ref{fig:cusumgpa} (and hence
+are omitted here).
 
@@
 \citeA{WicDol05} data allowed us to study the extent to which model
-parameters are invariant wrt GPA in a straightforward manner.
+parameters are invariant with respect to GPA in a straightforward manner.
 The tests focused on a set of group-specific
@@
 described here may require multiple models to
-be estimated (e.g., for LRTs) or adversely impact model fit and
+be estimated (e.g., for LR tests) or adversely impact model fit and
 degrees of freedom (e.g., if GPA is inserted directly into the model).
@@
 <<constraint_study>>=
-tests <- c("maxBB", "meanL2BB", "supLM(0.1)")
-
-p.diffs <- matrix(NA, 2*length(tests), 2)
-
+tests <- list(maxBB, meanL2BB, supLM(0.1))
+p.diffs <- matrix(NA, nrow = 2 * length(tests), ncol = 2)
 for (i in 1:length(tests)){
-  p.diffs[i,1] <- unlist(sctest(gefp_4_info, functional=
-                                eval(parse(text=tests[i])))[2])
-  p.diffs[i,2] <- unlist(sctest(gefp_4_info_1, functional=
-                                     eval(parse(text=tests[i])))[2])
-  }
-for (i in (length(tests)+1):(2*length(tests))){
-  p.diffs[i,1] <- unlist(sctest(gefp_1_info, functional=
-                                     eval(parse(text=tests[i-length(tests)])))[2])
-  p.diffs[i,2] <- unlist(sctest(gefp_1_info_1, functional=
-                                   eval(parse(text=tests[i-length(tests)])))[2])
+  p.diffs[                i, 1] <- sctest(gefp_4_info,   functional = tests[[i]])$p.value
+  p.diffs[length(tests) + i, 1] <- sctest(gefp_1_info,   functional = tests[[i]])$p.value
+  p.diffs[                i, 2] <- sctest(gefp_4_info_1, functional = tests[[i]])$p.value
+  p.diffs[length(tests) + i, 2] <- sctest(gefp_1_info_1, functional = tests[[i]])$p.value
 }
@@
 sequential LR~tests for various types of invariance using
-multiple-group models.  As shown in the Example section, it can be
+multiple-group models.  As shown in the ``Application'' section, it can be
 beneficial to employ the traditional steps in tandem with the proposed
-tests.  Further, when groups are not defined in advance (e.g.,
+tests.  Furthermore, when groups are not defined in advance (e.g.,
 continuous $V$), the traditional steps fail unless further assumptions
@@
 
-Another difference between the proposed tests and traditional tests
-involves the fact that, for the proposed tests, one's choice of
-identification constraints can influence tests of the remaining
-parameters.  This essentially reflects the fact that, while the model
-fits equally well regardless of identification constraint, the
-free model parameters adjust themselves to the specific identification
-constraint chosen.  To examine this issue in more detail, we refit the model
-used in the Example section, using an alternative identification
-constraint (i.e., fixing a different factor loading at 1).  Across the 6
-tests presented in the example section (see, e.g.,
-Figure~\ref{fig:cusumex}), the largest difference in p-values for the
-two identification constraints 
-was~$\Sexpr{round(max(abs(p.diffs[,2] - p.diffs[,1])),3)}$.  This is not a large difference, though further
-study is warranted.
-
-In addition to providing general information about whether or not
-measurement 
+It is worth mentioning that the proposed tests are not invariant
+with respect to the choice of identification constraints. They share this
+property with both LM and Wald tests (while LR tests are invariant to
+the choice of identification constraints). However, asymptotically
+the influence of the parametrization disappears and also in finite
+samples it typically does not change the results significantly.
+For example, in the stereotype threat application, we assessed the p-values
+of all three tests under the constraint that the abstract reasoning
+loading is fixed to 1 \fixme{figure ref} and compared them to the p-values
+under the constraint that the verbal loading is fixed to 1. All p-values
+remained clearly significant and the changes were all smaller than
+\Sexpr{format(round(max(abs(p.diffs[1:3,2] - p.diffs[1:3,1])), digits = 5), scientific = FALSE)}.
+
+In addition to providing general information about whether or not measurement 
 invariance holds, the proposed tests allow researchers to interpret
 the nature of the invariance violation.  This is made possible, e.g.,
-through the tests' abilities to estimate $\nu$, the threshold dividing
+through the tests' abilities to locate $\nu$, the threshold dividing
 individuals into subgroups that violate
 measurement invariance (see Equation~\eqref{eq:h1*}).  It is also
-possible to define formal rules for estimating multiple $\nu$
-parameters (see \citeNP{ZeiSha10}).
+possible to formally estimate one or more $\nu$ thresholds by
+adopting a (partially) segmented model (see e.g., \citeNP{ZeiSha10}).
 
 \subsection{Comparison to Other Methods}
-The step
-functions from~\eqref{eq:h1*} were employed in the current paper to
+
+The step functions from~\eqref{eq:h1*} were employed in the current paper to
 highlight connections with nested model
@@
 parameters (Purcell also considers quadratic effects of the
-moderators).   
-Returning to deviations in parameter constancy, if the change in
-parameters is continuous but not exactly linear - e.g., a sigmoidal
-shift from one set of parameters to another one - then the simple
+moderators). Moreover, if the change in
+parameters is continuous but not exactly linear -- e.g., a sigmoidal
+shift from one set of parameters to another one -- then the simple
 single shift model in~\eqref{eq:h1*} may be a useful first
-approximation and have better power than linear
-techniques.
-
-% paragraph on factor mixture models
+approximation and may have better power than linear techniques
+(depending on how clearly the two regimes are separated).
+
 The proposed tests are also similar to factor mixture models (e.g.,
@@
 violating measurement invariance. Covariates can also be used to
-predict the unknown subgroups under both approaches.
-For example, for the 
+predict the unknown subgroups under both approaches. For example, for the 
 data used in our simulations, one could employ a factor mixture
-model with two latent classes.  This would typically involve the
-specification that student age has a linear effect on logit(P(Class
-2)), which highlights the main difference between the two techniques:
-the factor mixture model requires a specific assumption about the type
-of deviation from parameter constancy, while the proposed tests do
-not.
+model with two latent classes where the class probabilities depend
+on student's age through a logit link. If the assumptions of such
+a mixture hold, i.e., that each observation is a weighted combination
+from a (known) number of classes, the fitted model has more power
+uncovering this structure and makes it easy to interpret. However,
+the proposed tests are easier to use with fewer assumptions about
+the type of deviations (as long as they occur along the selected
+variable $V$).
+
+\readme{slightly rephrased}
 
@@
 parameters.  In addition to categorical $V$, this framework may also
-be useful for both continuous $V$ with many ties and ordinal $V$.  
+be useful for ordinal $V$ (or continuous $V$ with many ties).
+Furthermore, in this situation, one could also adapt the statistics
+\eqref{eq:dmax}, \eqref{eq:cvm}, and \eqref{eq:maxlm} computed from the same cumulative score
+process as usual. The only modification required should be that the statistics
+should not depend on the process's values ``within'' a category (or group
+of ties). This is easily achieved by taking the maximum (or sum) not over
+all observations $i = 1, \dots, n$ but only those $i$ at the ``end'' of a
+category.
+%% If $i_j$ is the cumulative frequency of category $j$ or lower
+%% (with $j = 1, \dots, C$), then one could take the outer maximum in
+%% \eqref{} over $i = i_1, i_2, \dots, i_{C-1}$.
+
+\readme{Is this intelligible or do we need more information?}
+
 The number of potential thresholds may be very low in these
@@
 Thus, unlike the tests described in this paper, this approach does not
-provide a formal significance test with a controlled level of Type~I errors.
+provide a formal significance test with a controlled level of type~I errors.
 
@@
 \subsection{Summary}
-We have outlined a family of 
+We outline a family of 
 stochastic process-based parameter instability tests from theoretical
-statistics and applied them to the issue of measurement invariance in psychometrics.
-The paper included theoretical development, an applied example, and 
-study of the tests' performance.  The tests were found to have good
+statistics and apply them to the issue of measurement invariance in psychometrics.
+The paper includes theoretical development, an applied example, and 
+study of the tests' performance under controlled conditions.  The tests are found to have good
 properties via simulation, making them useful for many psychometric
-applications.  More generally, the tests help 
-solve standing problems in
+applications.  More generally, the tests help solve standing problems in
 measurement invariance research and provide many avenues for

### 2012-02-09 www/efpFunctional-cat.R
improved ordinal data example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@100 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -116,6 +116,18 @@ sctest(scus, functional = ordfun)
 
 ## corresponding visualization
 plot(scus, functional = ordfun)
+abline(h = ordfun$computeCritval(alpha = 0.05, nproc = 2), col = 4)
+
+## comparison with supLM(0.1) test with broken ties
+d$zcont <- d$z + rep(-29:0/116, 10)
+scus2 <- gefp(y ~ x, order.by = ~ zcont, data = d)
+plot(scus2, functional = supLM(0.1),
+  col = "gray", lwd = 2, alpha = 0.0001,
+  xlim = c(0, 10), ylim = c(0, 20), xlab = "z")
+par(new = TRUE)
+plot(scus, functional = ordfun,
+  xlim = c(0, 10), ylim = c(0, 20), xlab = "", ylab = "")
+abline(h = supLM(0.1)$computeCritval(alpha = 0.05, nproc = 2), col = 2, lty = 2)
 abline(h = ordfun$computeCritval(alpha = 0.05, nproc = 2), col = 2)
 
 }

### 2012-02-09 www/rev1/response.txt
some improvements from Z, see README
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@101 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2011-11-11 www/rev1/response.txt
@@
-Editor
+Accompanying this letter is our revised manuscript, titled "Generalized
+measurement invariance tests with application to factor analysis."  We found the
+reviews to be very helpful and believe the paper has improved considerably as a
+result.
 
-Comment E.1.
-While I share the reviewers generally positive view of your paper, I find myself in the unusual position of being less enthusiastic about the present version of the paper than the reviewers seem to be.
-One reason for this is that I think that your method is much less practical than the reviewers do, chiefly because in practical applications of invariance work, it is unusual to encounter a case in which V is continuous.
-The one exception to this situation is when V is age, but even then, in the context of employment discrimination for example, legal considerations often demand an arbitrary split on 40 years, or some other such threshold (which is admittedly arbitrary, but is demanded by the law).
+The main addition to the manuscript is an application of our proposed tests to a
+four-group model with mean structure in a study of stereotype threat by
+Wicherts, Dolan, and Hessen (2005), as suggested by Reviewer 1.  We elected to
+use this real example to supplement, vs replace, the artificial example. While
+the real application allows us to illustrate how the tests can be adapted to
+specific challenges, the artificial example is more flexible in the illustration
+of other general aspects of the tests. Also, it serves as a gateway to the
+simulation.  We are willing to streamline this further if you or the reviewers
+feel that there is too much redundancy between the artificial and the real
+example, however.
+README
 
-Response E.1.
-Though we did not highlight it in the paper, the methods can be used with categorical V.
-We have also developed a new strategy for handling ordinal V and briefly describe it in the Discussion.
-Additionally, we would argue that applications of invariance do not use continuous V because methods for handling continuous V are still emerging (e.g., the moderated factor models discussed by Reviewer 1).
-We believe that our proposed methods will encourage invariance research involving continuous V.
+Furthermore, we have added some theory for the study of local alternatives as
+well as details on the use of our methods with categorical V, the ability of our
+methods to handle multiple groups models and mean structure, and the
+relationship of our methods to mixture models and moderated factor models.
 
-Comment E.2.
-The second point is related to issues raised by Reviewers one and four.
-In any invariance analyses, some elements of theta are allowed to vary across groups freely because they are not part of the invariance constraints dictated by equation 7.
-In the factor model, the factor means and factor covariance elements are not constrained.
-How is this fact represented in your method?
-While I saw your treatment of partial invariance (i.e., hold some elements to invariance while letting other elements go free), you seem in that context to be testing invariance in all of the free elements.
-Does your method allow one to permit some elements to be unconstrained, while others are
-invariant, without looking for violations of invariance in the free elements?  If it does, we need to see how this works.
+We look forward to receiving feedback on the manuscript.
 
-Response E.2.
-This comment focuses on the fact that our methods generally work from a "reduced model" standpoint, whereby all model parameters are constrained to be equal and we examine the stability of parameter estimates across individuals (or across groups).
-When groups are not known in advance (e.g., when V is continuous), this appears to be the only way to proceed.
+Sincerely,
+Ed Merkle
 
-However, if groups are known in advance, it is possible to employ the proposed methods in tandem with a multiple-group model.
-The multiple-group model can then handle both unconstrained and invariant parameters, because scores can be computed for all model parameters.
-We now illustrate this approach in the data example.
-DONE
+Editor Comments
 
-Comment E.3. You also did not mention how model identification constraints, which in the invariance case are sometimes invariance constraints, are handled.
+Comment E.1. While I share the reviewers generally positive view of your paper,
+I find myself in the unusual position of being less enthusiastic about the
+present version of the paper than the reviewers seem to be.  One reason for this
+is that I think that your method is much less practical than the reviewers do,
+chiefly because in practical applications of invariance work, it is unusual to
+encounter a case in which V is continuous. The one exception to this situation
+is when V is age, but even then, in the context of employment discrimination for
+example, legal considerations often demand an arbitrary split on 40 years, or
+some other such threshold (which is admittedly arbitrary, but is demanded by the
+law).
 
-Response E.3.
-The proposed tests are based on scores associated with all free parameters.
-Thus, it is straightforward to handle parameters that are constrained to be equal to one another.
-However, this comment may also refer to the fact that one's choice of identification constraint will influence tests of other free parameters.
-We now briefly describe this issue in the "Use in Practice" section of the Discussion.
-DONE
+Response E.1. Though we did not highlight it in the paper, the methods can be
+used with categorical V.  We now briefly describe several strategies for
+handling ordinal data in the Discussion.  Additionally, we would argue that many
+applications of invariance do not use continuous V because methods for handling
+continuous V are still emerging (e.g., the moderated factor models discussed by
+Reviewer 1).  We believe that our proposed methods will encourage further
+invariance research involving continuous V.
 
-Comment E.4.
-I think that we need to have a real example, as suggested by Reviewer one.
-This would allow you to answer my second point above, and also show that your method can give answers that are sensible in context.
-With a real example, you could drop the artificial one (but keep the simulations).
-Having a real example would also help to illustrate that there are real applications in which your method would prove useful.
+Comment E.2. The second point is related to issues raised by Reviewers one and
+four.  In any invariance analyses, some elements of theta are allowed to vary
+across groups freely because they are not part of the invariance constraints
+dictated by equation 7.  In the factor model, the factor means and factor
+covariance elements are not constrained.  How is this fact represented in your
+method?  While I saw your treatment of partial invariance (i.e., hold some
+elements to invariance while letting other elements go free), you seem in that
+context to be testing invariance in all of the free elements.  Does your method
+allow one to permit some elements to be unconstrained, while others are
+invariant, without looking for violations of invariance in the free elements? 
+If it does, we need to see how this works.
 
-Response E.4. We have dropped the artificial example and added a real example. DONE
+Response E.2. This comment focuses on the fact that our methods generally work
+from a "reduced model" standpoint, whereby all model parameters are constrained
+to be equal and we examine the stability of parameter estimates across
+individuals (or across groups).  When groups are not known in advance (e.g.,
+when V is continuous), this appears to be the only way to proceed.
+
+However, if groups are known in advance, it is possible to employ the proposed
+methods in tandem with a multiple-group model.  The multiple-group model can
+then handle both unconstrained and invariant parameters, because scores can be
+computed for all model parameters.  We now illustrate this approach in the data
+example.
+
+Comment E.3. You also did not mention how model identification constraints,
+which in the invariance case are sometimes invariance constraints, are
+handled.
+
+Response E.3. The proposed tests are based on scores associated with all free
+parameters.  Thus, it is straightforward to handle parameters that are
+constrained to be equal to one another.  However, this comment may also refer to
+the fact that one's choice of identification constraint will influence tests of
+other free parameters.  We now briefly describe this issue in the "Use in
+Practice" section of the Discussion.
+
+Comment E.4. I think that we need to have a real example, as suggested by
+Reviewer one.  This would allow you to answer my second point above, and also
+show that your method can give answers that are sensible in context.  With a
+real example, you could drop the artificial one (but keep the simulations).
+Having a real example would also help to illustrate that there are real
+applications in which your method would prove useful.
+
+Response E.4. As discussed in the introduction above, we have added a real
+example and hope that it brings out more clearly how the proposed tests can
+be adapted to some of the challenges you and the reviewers raised. However,
+we have also kept the artificial example because we feel that some of the basic
+properties of the tests are easier to illustrate under controlled conditions.
+README
 
@@
 
-Comment 1.1.
-The section on the asymptotic properties of the Beta-function on pages 11-14 was still quite hard to follow.
-The authors may want to expand upon some of the more technical aspects of these models to allow more readers to grasp how they would do the job in their given application.
+Comment 1.1. The section on the asymptotic properties of the Beta-function on
+pages 11-14 was still quite hard to follow. The authors may want to expand upon
+some of the more technical aspects of these models to allow more readers to
+grasp how they would do the job in their given application.
 
-Response 1.1. We have revised the theory development and added extra detail to the text.  If some specific concepts are still unclear, we are happy to revisit this section.  DONE
+Response 1.1. We have revised the theory development and added extra detail to
+the text.  If some specific concepts are still unclear, we are happy to revisit
+this section.
 
-Comment 1.2.
-There are clear links between the current approach based on an observed continuous grouping variable V and the factor mixture approach.
-In factor mixture modeling, the grouping variable that characterizes the subgroups that differ in terms of particular model parameters are also typically unknown (i.e., latent classes), but class membership can be predicted on the basis of observed predictor(s).
-So the model in Figure 1 and the scenario of 2 subgroups that the authors describe on pages 17-20 is essentially a factor mixture model with two latent classes in which age predicts class membership.
-Therefore, the authors should take steps to relate their modeling approach to factor mixture modeling.
+Comment 1.2. There are clear links between the current approach based on an
+observed continuous grouping variable V and the factor mixture approach. In
+factor mixture modeling, the grouping variable that characterizes the subgroups
+that differ in terms of particular model parameters are also typically unknown
+(i.e., latent classes), but class membership can be predicted on the basis of
+observed predictor(s). So the model in Figure 1 and the scenario of 2 subgroups
+that the authors describe on pages 17-20 is essentially a factor mixture model
+with two latent classes in which age predicts class membership. Therefore, the
+authors should take steps to relate their modeling approach to factor mixture
+modeling.
 
-Response 1.2.
-We agree and have now described our methods' relationship to mixture models in the Discussion.
-Some advantages of our approach include the facts that it does not require a priori specification of the number of latent classes or a priori specification of the way by which age impacts class membership.
-DONE
+Response 1.2. We agree and have now described our methods' relationship to
+mixture models in the Discussion.  Some advantages of our approach include the
+facts that it does not require a priori specification of the number of latent
+classes or a priori specification of the way by which age impacts class
+membership.
 
-Comment 1.3.
-Similarly, there exist a family of moderation-type factor models that may serve the same purpose of studying failures of invariance in common factor models with respect to (unknown) "grouping" variables.
-For instance, Dylan Molenaar, Conor Dolan and colleagues have recently worked out CFA models that were based on Shaun Purcell's work in which certain parameters are moderated by either observed or unobserved variables.
-These models too could be applied to the important problems related to invariance with respect to unknown groups although they may not work as well as do mixture models and the current approach when indeed there is some step function that describes the violations of invariance.
-The existence of these alternative approaches by no means diminishes the potential value of the current developments, but it does mean that the authors should discuss the relation between their approach and other existing methods.
+Comment 1.3. Similarly, there exist a family of moderation-type factor models
+that may serve the same purpose of studying failures of invariance in common
+factor models with respect to (unknown) "grouping" variables. For instance,
+Dylan Molenaar, Conor Dolan and colleagues have recently worked out CFA models
+that were based on Shaun Purcell's work in which certain parameters are
+moderated by either observed or unobserved variables. These models too could be
+applied to the important problems related to invariance with respect to unknown
+groups although they may not work as well as do mixture models and the current
+approach when indeed there is some step function that describes the violations
+of invariance. The existence of these alternative approaches by no means
+diminishes the potential value of the current developments, but it does mean
+that the authors should discuss the relation between their approach and other
+existing methods.
 
-Response 1.3. We now describe factor analysis models employing moderation in the "Comparison to other methods" section of the discussion.  DONE
+Response 1.3. We now describe factor analysis models employing moderation in the
+"Comparison to other methods" section of the discussion.
 
-Comment 1.4.
-The authors chose to simulate data only under models that did not include the mean structure.
-Since violations of measurement invariance with real data are most often related to the mean structure (to be sure due to larger power associated with modeling first-order moments), it may be more realistic and interesting to have scenarios in which such types of failures of invariance (called uniform measurement bias) occur.
+Comment 1.4. The authors chose to simulate data only under models that did not
+include the mean structure. Since violations of measurement invariance with real
+data are most often related to the mean structure (to be sure due to larger
+power associated with modeling first-order moments), it may be more realistic
+and interesting to have scenarios in which such types of failures of invariance
+(called uniform measurement bias) occur.
 
-Response 1.4.
-Mean parameters were estimated and tested in the simulation models, though they were unstructured.
-Mean structures are often made possible by constraining parameters across groups, and our simulation focuses on a case where groups cannot be predefined.
-Thus, we have not modified the model in the simulation.
-In the real data example that we have added, however, the mean structure is utilized.
-DONE
+Response 1.4. Mean parameters were estimated and tested in the simulation
+models, though they were unstructured.  Mean structures are often made possible
+by constraining parameters across groups, and our simulation focuses on a case
+where groups cannot be predefined.  Thus, we have not modified the model in the
+simulation but point out more clearly what the 19 parameters in the model are.
+In the real data example that we have added, however, the mean structure is
+utilized. 
+README
 
-Comment 1.5.
-Also, many "nuisance" variable that are hypothesized to affect indicators, say, in cognitive testing (e.g., see Wicherts et al.'s work on measurement invariance and stereotype threat) in a given multi-group CFA are expected to linearly affect these indicators.
-Now I can come up with a host of scenarios in which the measurement bias is approximately linear and so can be modeled in simple terms because the regression on V results in an additional factor in the model.
-Surely I don't expect the authors to deal with all of such potential scenarios' of failures of measurement invariance in CFA.
-However, I do wish to point out that underlying the current approach is a specific kind of measurement bias that is expected to be discontinuous while many failures of invariance are expected to be continuous and approximately linear.
-Therefore, the authors should go into more detail on the nature of the measurement bias that the approach is trying to uncover.
-Their approach may be generic, but it is not really general in the sense that many failures of strict factorial invariance can be modeled in more straightforward and traditional ways.
-The step function in Eq.
-10 should be discussed in more detail and may be supplemented by a hypothetical example of why it occurs in linear factor models.
+Comment 1.5. Also, many "nuisance" variable that are hypothesized to affect
+indicators, say, in cognitive testing (e.g., see Wicherts et al.'s work on
+measurement invariance and stereotype threat) in a given multi-group CFA are
+expected to linearly affect these indicators. Now I can come up with a host of
+scenarios in which the measurement bias is approximately linear and so can be
+modeled in simple terms because the regression on V results in an additional
+factor in the model.  Surely I don't expect the authors to deal with all of such
+potential scenarios' of failures of measurement invariance in CFA. However, I do
+wish to point out that underlying the current approach is a specific kind of
+measurement bias that is expected to be discontinuous while many failures of
+invariance are expected to be continuous and approximately linear. Therefore,
+the authors should go into more detail on the nature of the measurement bias
+that the approach is trying to uncover. Their approach may be generic, but it is
+not really general in the sense that many failures of strict factorial
+invariance can be modeled in more straightforward and traditional ways. The step
+function in Eq. 10 should be discussed in more detail and may be supplemented by
+a hypothetical example of why it occurs in linear factor models.
 
-Response 1.5.
-The step function from Equation 10 is employed as an illustrative example because it highlights the connection between nested model comparisons and the proposed techniques.
-Also, the maxLM statistic from Equation 17 has particularly high power for this scenario.
-However, the proposed tests typically have non-trivial power for all (non-constant) deviation patterns theta(v_i) (considered in Equation 9).
-We have now provided more details on these issues in the Discussion section (and tied the details to Comment/Response 1.3).
-DONE
+Response 1.5. The step function from Equation 10 is employed as an illustrative
+example because it highlights the connection between nested model comparisons
+and the proposed techniques. Also, the maxLM statistic from Equation 17 has
+particularly high power for this scenario. However, the proposed tests typically
+have non-trivial power for all (non-constant) deviation patterns theta(v_i)
+(considered in Equation 9).  We have now provided more details on these issues
+in the Discussion section (and tied the details to Comment/Response 1.3).
 
-Comment 1.6.
-On page 3, the authors state that failures of measurement invariance render scales useless, but this statement may be overly strong.
-Specifically, scales that are only partially invariant  may still be useful.
-In addition, the statement that "the scales are the sole cause of the differences" appears to suggest that failures of invariance can explain all group differences in observed test scores, which is seldom the case.
-Failures of measurement invariance do not preclude the existence of group differences in the distribution of latent factors scores.
+Comment 1.6. On page 3, the authors state that failures of measurement
+invariance render scales useless, but this statement may be overly strong.
+Specifically, scales that are only partially invariant  may still be useful. In
+addition, the statement that "the scales are the sole cause of the differences"
+appears to suggest that failures of invariance can explain all group differences
+in observed test scores, which is seldom the case. Failures of measurement
+invariance do not preclude the existence of group differences in the
+distribution of latent factors scores.
 
-Response 1.6. We agree that these statements were too strong and have revised accordingly.  DONE
+Response 1.6. We agree that these statements were too strong and have revised
+accordingly.
 
-Comment 1.7.
-When presenting the definition of invariance on page 5, the authors may wish to indicate that the definition (due originally to Mellenbergh, 1989) implies (as shown by Meredith, 1993) that not all of the parameters in the CFA are invariant but only the measurement parameters.
-So factor means, factor covariances, and (in structural models) factor regression weights are not expected to be invariant under this definition of invariance.
-Another issue that could be discussed is that non-normality in the data may lead to spurious findings in the current approach just like they do in regular multi-group confirmatory analyses and in factor mixture modeling (e.g., see work by Gitta Lubke and Michael Neale).
+Comment 1.7. When presenting the definition of invariance on page 5, the authors
+may wish to indicate that the definition (due originally to Mellenbergh, 1989)
+implies (as shown by Meredith, 1993) that not all of the parameters in the CFA
+are invariant but only the measurement parameters. So factor means, factor
+covariances, and (in structural models) factor regression weights are not
+expected to be invariant under this definition of invariance. Another issue that
+could be discussed is that non-normality in the data may lead to spurious
+findings in the current approach just like they do in regular multi-group
+confirmatory analyses and in factor mixture modeling (e.g., see work by Gitta
+Lubke and Michael Neale).
 
-Response 1.7. We now describe the invariance definitions in more detail, and we also clarify that our methods are not necessarily robust to violations of normality.  DONE
+Response 1.7. We now describe the invariance definitions in more detail, and we
+also clarify that our methods are not necessarily robust to violations of
+normality.
 
-Comment 1.8.
-Finally, it may be interesting to apply the method to real data and to compare the outcomes from more traditional modeling approaches.
-After all, in real life the exact nature of the measurement bias with respect to V is not know.
-One interesting approach may be to use data from experiments like in the stereotype threat framework that was used by Wicherts et al.
-(who I am sure will be happy to share raw data for this purpose).
+Comment 1.8. Finally, it may be interesting to apply the method to real data and
+to compare the outcomes from more traditional modeling approaches. After all, in
+real life the exact nature of the measurement bias with respect to V is not
+know. One interesting approach may be to use data from experiments like in the
+stereotype threat framework that was used by Wicherts et al. (who I am sure will
+be happy to share raw data for this purpose).
 
-Response 1.8. See Response E.4.  DONE
+Response 1.8. See Response E.4.
 
@@
 
-Comment 3.1.
-The result from Hjort and Koning is a nice trick to rewrite the asymptotic representation of the likelihood estimator as an empirical process.
-This then relates the cumulative sums of the likelihood estimator to a Brownian bridge.
-The weak convergence of the monitoring process is of course asymptotic.
-The properties of the tests in the current manuscript are investigated decently enough for medium to large sample sizes (n = 100 to 500).
-However, for small samples no results are obtained.
-Given the relatively small model in the simulations, I would expect a relatively small sample size of n = 50 to be included in the simulations.
-The main reason is that this may show a point where the power functions show a "breakdown."
+Comment 3.1. The result from Hjort and Koning is a nice trick to rewrite the
+asymptotic representation of the likelihood estimator as an empirical process.
+This then relates the cumulative sums of the likelihood estimator to a Brownian
+bridge. The weak convergence of the monitoring process is of course asymptotic.
+The properties of the tests in the current manuscript are investigated decently
+enough for medium to large sample sizes (n = 100 to 500). However, for small
+samples no results are obtained. Given the relatively small model in the
+simulations, I would expect a relatively small sample size of n = 50 to be
+included in the simulations. The main reason is that this may show a point where
+the power functions show a "breakdown."
 
-Response 3.1. We now include the n=50 condition in the simulations, which indeed exhibits a breakdown in the sense that power is not monotonic with respect to violation magnitude.  DONE
+Response 3.1. We now include the n=50 condition in the simulations, which indeed
+exhibits a breakdown in the sense that power is not monotonic with respect to
+violation magnitude.
 
-Comment 3.2.
-The weak convergence of the monitoring process is proved in the paper by Hjort and Koning in two cases: (i) for theta_0 the true parameter, and (ii)   for local alternatives O(1/sqrt(n)) from theta_0.
-In the latter case the monitoring process is shown to be a Brownian bridge including a constant (bias) even when using -E{u'u}, Fisher's information (which is consistent up to O(1/n)).
-I would imagine this is important since model misspecification (measurement invariance) is the topic of the manuscript.
-This result allows the authors to to determine exactly what the result of misspecification is.
-It fits well with their line of reasoning, which I agree with.
-The misspecification is determined by a deviation h/sqrt(n) of the true parameter.
-This corresponds exactly to the misspecification in the simulations.
-Measurement invariance can be imagined as a latent variable acting on the factor loadings, and can therefore be incorporated in the parameter theta.
-If there were no
-measurement invariance, theta_v woold be 0, but in the case of misspecification, it is exactly the same as that of Hjort and Koning.
-I would recommend that the authors use this result by Hjort and Koning in the manuscript to show analytically what happens to the distribution and tests.
+Comment 3.2. The weak convergence of the monitoring process is proved in the
+paper by Hjort and Koning in two cases: (i) for theta_0 the true parameter, and
+(ii)   for local alternatives O(1/sqrt(n)) from theta_0. In the latter case the
+monitoring process is shown to be a Brownian bridge including a constant (bias)
+even when using -E{u'u}, Fisher's information (which is consistent up to
+O(1/n)). I would imagine this is important since model misspecification
+(measurement invariance) is the topic of the manuscript. This result allows the
+authors to to determine exactly what the result of misspecification is. It fits
+well with their line of reasoning, which I agree with. The misspecification is
+determined by a deviation h/sqrt(n) of the true parameter. This corresponds
+exactly to the misspecification in the simulations. Measurement invariance can
+be imagined as a latent variable acting on the factor loadings, and can
+therefore be incorporated in the parameter theta. If there were no measurement
+invariance, theta_v woold be 0, but in the case of misspecification, it is
+exactly the same as that of Hjort and Koning. I would recommend that the authors
+use this result by Hjort and Koning in the manuscript to show analytically what
+happens to the distribution and tests.
 
-Response 3.2.
-We have now incorporated these results in a new section, titled "Local alternatives".
-We also discuss these results in the context of the simulation, showing the impact of our imposed measurement invariance violation on the Brownian bridge.
-DONE
+Response 3.2. We have now incorporated these results in a new section, titled
+"Local alternatives".  We also discuss these results in the context of the
+simulation, showing the impact of our imposed measurement invariance violation
+on the Brownian bridge.
 
-We addressed all minor issues identified by Reviewer 3.  We have a response to one of these issues:
+We addressed all minor issues identified by Reviewer 3.  We have a response to
+one of these issues:
 
-Comment 3.3. p. 7 second paragraph: Assuming that the model is correctly specified is bonus when the second point above is incorporated. I think this would improve the manuscript.
+Comment 3.3. p. 7 second paragraph: Assuming that the model is correctly
+specified is bonus when the second point above is incorporated. I think this
+would improve the manuscript.
 
-Response 3.3.
-Violations of distributional assumptions may still cause problems with the proposed tests, even if the "structure" of the model (e.g., parameter constraints, included variables, etc) is correct.
-We have modified this section of text accordingly.
-DONE
+Response 3.3. Violations of distributional assumptions may still cause problems
+with the proposed tests, even if the "structure" of the model (e.g., parameter
+constraints, included variables, etc) is correct.  We have modified this section
+of text accordingly.
 
@@
 
-Comment 4.1. Page 6/7:	It could be emphasized that a known cutpoint ν means “observed heterogeneity” and an unknown cutpoint ν means “unobserved heterogeneity”.
+Comment 4.1. Page 6/7:	It could be emphasized that a known cutpoint nu means
+"observed heterogeneity" and an unknown cutpoint nu means "unobserved
+heterogeneity".
 
-Response 4.1. We have now added this detail.  DONE
+Response 4.1. We have now added this detail.
 
-Comment 4.2. Page 9:	The authors should give a reference or a short explanation of the “tied-down Bessel process”. Readers might be unfamiliar with the process.
+Comment 4.2. Page 9:	The authors should give a reference or a short
+explanation of the "tied-down Bessel process". Readers might be unfamiliar
+with the process.
 
-Response 4.2. We now note that Andrews (1993) provides results and references relevant to Bessel processes.  DONE
+Response 4.2. We now note that Andrews (1993) provides results and references
+relevant to Bessel processes and also included a forward reference to later
+sections. Then, in the "Critical Values & p-values" section, we then state a
+brief formal definition of the process (sup_t (t * (1 - t))^{-1} ||B^0(t)||^2)
+and point to Hansen (1997) where simulations for critical values are discussed
+in detail.
 
-Comment 4.3.
-Page 10:	To test for changes “along” V means that V is always continuous (also mentioned in the first paragraph).
-What happens if V is an ordered categorical variable?
-Would that be a special case?
-Adding a reference to the discussion section would be helpful.
+Comment 4.3. Page 10:	To test for changes "along" V means that V is always
+continuous (also mentioned in the first paragraph). What happens if V is an
+ordered categorical variable? Would that be a special case? Adding a reference
+to the discussion section would be helpful.
 
-Response 4.3. We have added some discussion on the use of categorical V.  Also see Response E.1.
+Response 4.3. We have added some discussion on the use of categorical V.  Also
+see Response E.1.
 
-Comment 4.4.
-Page 19:	The overall k=19 model parameter consists of six factor loadings, six measurement errors and one covariance between the factors.
-The authors refer to the six means of the manifest variables.
-Usually, in factor models mean differences of the factors are of research interest.
-The authors should give some information why they considered the manifest means instead of the latent mean differences.
+Comment 4.4. Page 19:	The overall k=19 model parameter consists of six factor
+loadings, six measurement errors and one covariance between the factors. The
+authors refer to the six means of the manifest variables. Usually, in factor
+models mean differences of the factors are of research interest. The authors
+should give some information why they considered the manifest means instead of
+the latent mean differences.
 
-Response 4.4.
-Factor means can be estimated if one is constraining other model parameters to be equal across groups.
-In our new data example, we do estimate factor means and use them in the proposed tests.
-In our simulation (and in the example that was removed), we use a scenario where no grouping variables exist (i.e., there is only a continuous V).
-In this scenario, it seems difficult to estimate factor means prior to model estimation because we cannot constrain parameters across groups.
-We have added some extra detail to the simulation section on our use of a single-group model.
-DONE
+Response 4.4. Factor means can be estimated if one is constraining other model
+parameters to be equal across groups.  In our new data example, we do estimate
+factor means and use them in the proposed tests.  In our simulation we use a
+scenario where no grouping variables exist (i.e., there is only a continuous
+V).  In this scenario, it seems difficult to estimate factor means prior to
+model estimation because we cannot constrain parameters across groups.  We have
+added some extra detail to the simulation section on our use of a single-group
+model.
 
 Comment 4.5.	The simulations later on focus on invariance of factor loadings.
-Basicly, the covariance structure in the simulations are analyzed with the factor model of Figure 1.
-So, means are not included in the analyses (as far as I read the simulation results correctly).
+Basicly, the covariance structure in the simulations are analyzed with the
+factor model of Figure 1. So, means are not included in the analyses (as far as
+I read the simulation results correctly).
 
-Response 4.5.
-Mean parameters are included in the casewise likelihood function, though they were unmodeled in the simulation and thus omitted from Figure 1.
+Response 4.5. Mean parameters are included in the casewise likelihood function,
+though they were unmodeled in the simulation and thus omitted from Figure 1. 
 The casewise likelihood function is needed to make the proposed tests possible. 
 Also see Comment/Response 1.5.
-DONE
 
-Comment 4.6.
-Page 25: 	Implications of the study of measurement invariance in factor analysis to general structural equation models could be added with comments to mean differences (see comment for page 19).
+Comment 4.6. Page 25: 	Implications of the study of measurement invariance in
+factor analysis to general structural equation models could be added with
+comments to mean differences (see comment for page 19).
 
-Response 4.6. We have added some brief discussion surrounding extensions to structural equation models.  DONE
+Response 4.6. We have added some brief discussion surrounding extensions to
+structural equation models.
 
-Comment 4.7. Table 1:	Table captions should be moved to the bottom of the table. (Abbreviations: …)
+Comment 4.7. Table 1:	Table captions should be moved to the bottom of the
+table. (Abbreviations: ...)
 
 Response 4.7. We have moved the Abbreviations part of the captions.
-DONE

### 2012-02-09 www/paper.Rnw
Addressed all readmes and fixmes in the paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@102 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
-  generously shared data for the stereotype threat example.
+  generously shared data for the stereotype threat application.
   Correspondence to Edgar C.\ Merkle, 
@@
 employing a Wishart distribution. However, in that approach the sample covariance
-matrix can also be disaggregated to the sum of its individual-specific contributions
+matrix can also be disaggregated to the sum of its individual-specific contributions, 
 leading essentially to a likelihood like \eqref{eq:mvndensity} with multivariate normal
@@
 equation models with missing data (e.g., \citeNP{Wot00}).
-
-\readme{maybe like this?}
 
@@
 verbal factor loading parameters for student $i$. Thus, here only $k^* = 3$
-from the overall $k = 19$~model parameters (including means) are assessed.
-\fixme{Explain what the 19 parameters are.}
+from the overall $k = 19$~model parameters are assessed (where the 19 parameters include six factor loadings, six unique variances, six intercepts, and one factor correlation).
 Alternatively, we can consider all $k^* = k = 19$~parameters, leading to a test
@@
 
-\readme{now called ``Application'' as opposed to the artificial ``Example''}
-
 \citeA{WicDol05}, henceforth WDH, utilized confirmatory
@@
 \subsection{Background}
-
-\fixme{Include path diagram and a bit more background so that the paper is more
-accessible for readers that are not familiar with the WDH paper.}
 
@@
 example below, we employ one of the models used by WDH.
-Our $V$ variable is participants' grade point averages (GPAs), which were unused in the original analyses.
-\fixme{Maybe add: ``\dots as a measure of aptitude.'' Or something along these lines? I'm not
-sure what the appropriate psychometric terminology is.}
+Our $V$ is participants' aptitudes, as measured by their grade point averages (GPAs).  The GPAs were unused in the original analyses.
 
@@
 in their Step~5b (see pp.~703--704 of their paper), which involved
-across-group parameter restrictions on the factor loadings, unique
+across-group restrictions on the factor loadings, unique
 variances, intercepts, and factor variances.  These parameters were
@@
 groups as described above, we focus only on the submodel for the minority
-group that received stereotype primes.
+group that received stereotype primes, which is pictured in Figure~\ref{fig:wpath}.
+In the figure, paths with dashed lines and red parameters signify group-specific parameters.
+These include the factor mean $\eta$, numerical factor loading $\lambda_{\text{num}}$, numerical intercept $\mu_{\text{num}}$, and numerical unique variance $\psi_{\text{num}}$.
+Other parameters (not displayed) were restricted to be equal across all four groups, with the exception of the factor variance.
+This parameter was restricted to be equal within both minority groups, and separately within both majority groups.
+
+\begin{figure}
+\caption{Path diagram representing the submodel for the ``minority, stereotype'' group in model 5b of WDH.
+Dashed paths and red parameters are group specific.
+Other parameters (not displayed) were restricted to be equal across all four groups, with the factor variance only being restricted across the two minority groups.}
+\label{fig:wpath}
+\includegraphics[height=3.5in]{wicherts_msp.pdf}
+\end{figure}
 
@@
 where
-$\lambda_{i, \text{num}}$ is the factor loading of student $i$'s intelligence on the numerical scale,
+$\lambda_{i, \text{num}}$ is the factor loading on the numerical scale,
 $\eta_{i}$ is student $i$'s factor mean,
@@
 model parameters are invariant with respect to GPA.
-
-\readme{I have omitted the single parameter hypothesis. Still it would be good to
-double-check the estfun method once more, though.}
 
@@
 of all three tests under the constraint that the abstract reasoning
-loading is fixed to 1 \fixme{figure ref} and compared them to the p-values
+loading is fixed to 1 (see Figure~\ref{fig:wpath}) and compared them to the p-values
 under the constraint that the verbal loading is fixed to 1. All p-values
@@
 variable $V$).
-
-\readme{slightly rephrased}
 
@@
 the $\chi^2$ statistics and degrees of freedom for the individual
-parameters.  In addition to categorical $V$, this framework may also
+parameters.  In addition to categorical $V$, this framework may 
 be useful for ordinal $V$ (or continuous $V$ with many ties).
-Furthermore, in this situation, one could also adapt the statistics
+In this situation, one could also adapt the statistics
 \eqref{eq:dmax}, \eqref{eq:cvm}, and \eqref{eq:maxlm} computed from the same cumulative score
-process as usual. The only modification required should be that the statistics
+process as usual. The only required modification is that the statistics
 should not depend on the process's values ``within'' a category (or group
 of ties). This is easily achieved by taking the maximum (or sum) not over
-all observations $i = 1, \dots, n$ but only those $i$ at the ``end'' of a
+all observations $i = 1, \dots, n$, but over only those $i$ at the ``end'' of a
 category.
@@
 %% \eqref{} over $i = i_1, i_2, \dots, i_{C-1}$.
-
-\readme{Is this intelligible or do we need more information?}
 

### 2012-02-11 www/paper.Rnw
Minor wording changes to the revised paper, based on final read-through.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@103 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   tests can be viewed as generalizations of the Lagrange multiplier test, and they
-  are especially useful for: (1)~isolating specific parameters affected by
+  are especially useful for: (1)~identifying specific parameters impacted by
   measurement invariance violations, and (2)~identifying subgroups of individuals
   that violated measurement invariance based on a continuous auxiliary variable. 
-  The tests are presented and illustrated in detail, along with an application
+  The tests are presented and illustrated in detail, including an application
   to a study of stereotype threat and simulations examining the tests'
@@
 leading essentially to a likelihood like \eqref{eq:mvndensity} with multivariate normal
-observations $\bm{x}_i$. This situation is also generally encountered in structural
+observations $\bm{x}_i$. This situation is similar to that encountered in structural
 equation models with missing data (e.g., \citeNP{Wot00}).
@@
 the three tests is that the potential subgroups have to be known in advance. 
-Even if the variable $V$ w.r.t.\ which the violation of invariance occurs is
-known, the threshold $\nu$ from (\ref{eq:h1*}) is often unknown in practice.
+Even if the scales are known to violate measurement invariance w.r.t.\ the variable $V$, the threshold $\nu$ from (\ref{eq:h1*}) is often unknown in practice.
 For example, if $V$ represents yearly income, there are many possible
@@
 In particular, one approach that is often employed in practice
-involves adoptions of a model
+involves adoption of a model
 with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
@@
 
-As mentioned previously, the double-max test is advantageous because
+We previously noted that the double-max test is advantageous because
 it yields information about individual parameters violating measurement invariance.  
@@
 three verbal factor loadings and the top left panel of Figure~\ref{fig:cusumex} shows
-the same processes aggregated over the three parameters. In both graphics, horizontal lines reflecting the
+the same processes aggregated over the three parameters. In both graphics, horizontal lines reflect the
 same critical value at $\alpha= 0.05$. The figure
@@
 error, which implicitly brings into question the tests' power.
-This is an example of a situation where the double-max test ``wastes''
-power, as it cannot make use of the fact that multiple model
-parameters change simultaneously.  We
+% This is an example of a situation where the double-max test ``wastes''
+% power, as it cannot make use of the fact that multiple model
+% parameters change simultaneously.
+We
 generally address the issue of power in the simulations below.
@@
 Other parameters (not displayed) were restricted to be equal across all four groups, with the exception of the factor variance.
-This parameter was restricted to be equal within both minority groups, and separately within both majority groups.
+The factor variance was restricted to be equal within both minority groups, and separately within both majority groups.
 
@@
 As mentioned in the theory section, the 
-tests give us the flexibility to study hypotheses of partial change. That is, we
+tests give us the flexibility to study hypotheses of partial change: we
 have the ability to  test various subsets of  parameters.  For the WDH data, we
@@
 (bottom panel) it can be concluded that the invariance violation is associated
-with this parameter. However, also the factor loading for the numerical scale
-$\lambda_\text{num}$ and the factor mean $\eta$ exhibit increased fluctuation
-albeit they are just non-significant at level $\alpha = 0.05$. Only the process
-associated with the intercept $\mu_\text{num}$ just shows moderate (and hence
+with this parameter.  Both the factor loading for the numerical scale
+$\lambda_\text{num}$ and the factor mean $\eta$ also exhibit increased fluctuation, 
+although they are just non-significant at level $\alpha = 0.05$. Only the process
+associated with the intercept $\mu_\text{num}$ shows moderate (and hence
 clearly non-significant) fluctuation across GPA.
@@
 subgroups (i.e., LR, Wald, and LM tests) may suffice.
-Furthermore, we can also obtain an LM-type statistic from the
+However, we can also obtain an LM-type statistic from the
 framework developed here. Assume the observations are divided into
@@
     \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
-    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i)
+    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i).
 %% \\

### 2012-02-11 www/rev1/response.txt
Minor revisions to response letter.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@104 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 example, however.
-README
 
@@
 properties of the tests are easier to illustrate under controlled conditions.
-README
 
@@
 Response 1.3. We now describe factor analysis models employing moderation in the
-"Comparison to other methods" section of the discussion.
+Discussion.
 
@@
 utilized.
-README
 
@@
 Response 3.3. Violations of distributional assumptions may still cause problems
-with the proposed tests, even if the "structure" of the model (e.g., parameter
-constraints, included variables, etc) is correct.  We have modified this section
-of text accordingly.
+with the proposed tests, even if we need not worry about the "structure" of the model (e.g., parameter constraints, included variables, etc).  We have modified this section of text accordingly.
 
@@
 scenario where no grouping variables exist (i.e., there is only a continuous
-V).  In this scenario, it seems difficult to estimate factor means prior to
-model estimation because we cannot constrain parameters across groups.  We have
-added some extra detail to the simulation section on our use of a single-group
-model.
+V).
+In this scenario, it seems difficult to estimate factor means prior to model estimation because we cannot constrain parameters across groups.
+We have added some extra detail to the simulation section on our use of a single-group model.
 
@@
 though they were unmodeled in the simulation and thus omitted from Figure 1. 
-The casewise likelihood function is needed to make the proposed tests possible. 
-Also see Comment/Response 1.5.
+The proposed tests are most easily carried out using the casewise likelihood function.  Also see Comment/Response 1.5.
 

### 2012-02-11 www/paper.Rnw
changed paper title to better reflect that there are no predefined subgroups and that the tests extend classical approaches
updated acknowledgements (added Yves and Psychoco)

further small improvements



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@105 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\title{Generalized measurement invariance tests with application to
-  factor analysis}
+\title{Tests of measurement invariance without subgroups:
+  A~generalization of classical methods}
 \twoauthors{Edgar C. Merkle}{Achim Zeileis}
@@
   tests can be viewed as generalizations of the Lagrange multiplier test, and they
-  are especially useful for: (1)~identifying specific parameters impacted by
-  measurement invariance violations, and (2)~identifying subgroups of individuals
-  that violated measurement invariance based on a continuous auxiliary variable. 
+  are especially useful for: (1)~identifying subgroups of individuals
+  that violate measurement invariance along a continuous auxiliary variable
+  without prespecified thresholds, and (2)~identifying specific parameters impacted by
+  measurement invariance violations.
   The tests are presented and illustrated in detail, including an application
@@
   Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
-  generously shared data for the stereotype threat application.
-  Correspondence to Edgar C.\ Merkle, 
-  Department of  
+  generously shared data for the stereotype threat application,
+  Yves Rosseel, who provided feedback and code for performing the
+  tests with the lavaan package, and the participants of the
+  Psychoco 2012 workshop on psychometric computing for
+  discussions.
+  Correspondence to Edgar C.\ Merkle, Department of  
   Psychological Sciences, University of Missouri, Columbia,
@@
   Email: {\texttt{merklee@missouri.edu}}.}
-\shorttitle{Generalized measurement invariance tests}
-\rightheader{Generalized measurement invariance tests}
+\shorttitle{Tests of measurement invariance without subgroups}
+\rightheader{Tests of measurement invariance without subgroups}
 
@@
 
-To carry out
-the tests, we first fit the four-group model to the data,
-calculating casewise derivatives and 
-the observed information matrix.  We then
-calculated the cumulative score process with respect to student GPA via
-\eqref{eq:cumscore}, allowing us to
-obtain various test statistics and $p$-values from this 
-process.  Test statistics include the double-max statistic from \eqref{eq:dmax},
-the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
+To carry out the tests, we first fit the four-group model to the data,
+calculating casewise derivatives and  the observed information matrix.   Second,
+to assess measurement invariance within the ``minority, stereotype prime'' group
+only, the scores of the $n_{\text{MSP}}$ individuals from that group are ordered
+and aggregated along GPA (i.e.,  in this application the variable $V$). The
+cumulative score process \eqref{eq:cumscore} with respect to GPA then allows us
+to obtain various test statistics and $p$-values from this  process.  Test
+statistics include the double-max statistic from \eqref{eq:dmax}, the
+Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
 statistic from \eqref{eq:maxlm}.
@@
 $\psi_{i, \text{num}}$ is student $i$'s unique variance on the numerical scale.
-The students are ordered according to GPA (i.e.,  in this application the variable $V$),
-and the  hypothesis is taken only with respect to the $n_{\text{MSP}}$ individuals in the
-``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states  that these four group-specific
+Thus, \eqref{eq:exh0} states  that these four group-specific
 model parameters are invariant with respect to GPA.

### 2012-02-11 www/wdh.R
removed wdh.R script as the wdhfit() interface is not needed anymore
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@107 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,157 +1,0 @@
-wdhfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
-{
-  ## data checks
-  data <- as.data.frame(data)
-  stopifnot(c("VERBREA","ABSREA","NUMER","gpa","grouping") %in% names(data))
-  data <- data[, c("VERBREA","ABSREA","NUMER","gpa","grouping")]
-
-  ## require package
-  stopifnot(require("lavaan"))
-  
-  ## model from Step 5b of Wicherts et al., 2005:
-  mod5b <- 'f1 =~ 1*VERBREA + label(rep("l2",4))*ABSREA + label(c("l1","l1","l1","l1expmin"))*NUMER 
-                  f1 ~ label(c("a","faccontmin","facexpmaj","facexpmin"))*1 + c(0,NA,NA,NA)*1
-                  f1 ~~ label(c("vmaj","vmin","vmaj","vmin"))*f1
-                  NUMER ~~ label(c("e1","e1","e1","e1expmin"))*NUMER
-                  ABSREA ~~ label(rep("e2",4))*ABSREA
-                  VERBREA ~~ label(rep("e3",4))*VERBREA
-
-                  NUMER ~ label(c("meanNA","meanNA","meanNA","meanNAexpmin"))*1
-                  ABSREA ~ label(c("meanARcontmaj","meanARcontmin","meanAR","meanAR"))*1
-                  VERBREA ~ label(rep("meanVR",4))*1'
-
-  ## fit model
-  rval <- lavaan(mod5b, data=data, meanstructure=TRUE, group="grouping")
-
-  ## store (1) data, (2) fitted model object, (3) grouping variable,
-  ## (4) focal group nubmers (group ordering is defined by lavaan and
-  ##                          can be found in rval@Sample@group.label)
-  rval <- list(
-    data = data[,1:3],
-    model = rval,
-    groupvar = data$grouping,
-    groupfoc = 4
-  )
-  class(rval) <- "wdhfit"
-
-  return(rval)
-}
-
-wdhfit2 <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
-{
-  ## Fit the same model but with different identification constraints.
-  ## data checks
-  data <- as.data.frame(data)
[…]

### 2012-02-11 www/wdh-replication.R
removed wdh-*.R scripts: replication code is now in psychotools and the 5b model is contained in paper.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@108 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,87 +1,0 @@
-## Package and data
-library("lavaan")
-data("StereotypeThreat", package = "psychotools")
-
-## Include group variable and order correspondingly
-StereotypeThreat <- transform(StereotypeThreat, group = interaction(ethnicity, condition))
-StereotypeThreat <- StereotypeThreat[order(StereotypeThreat$group),]
-
-## Step 2: Fix loadings across groups
-m2  <- c(
-'ability   =~ label(rep("load_n", 4)) * numerical  + 1 * abstract + label(rep("load_v", 4)) * verbal',
-'ability   ~  0 * 1',
-'ability   ~~ label(c("lvar:maj_c", "lvar:min_c", "lvar:maj_t", "lvar:min_t")) * ability',
-
-'numerical ~  label(c("mean_n:maj_c", "mean_n:min_c", "mean_n:maj_t", "mean_n:min_t")) * 1',
-'abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", "mean_a:maj_t", "mean_a:min_t")) * 1',
-'verbal    ~  label(c("mean_v:maj_c", "mean_v:min_c", "mean_v:maj_t", "mean_v:min_t")) * 1',
-
-'numerical ~~ label(c("var_n:maj_c", "var_n:min_c", "var_n:maj_t", "var_n:min_t")) * numerical',
-'abstract  ~~ label(c("var_a:maj_c", "var_a:min_c", "var_a:maj_t", "var_a:min_t")) * abstract',
-'verbal    ~~ label(c("var_v:maj_c", "var_v:min_c", "var_v:maj_t", "var_v:min_t")) * verbal')
-
-## Step 2a: Free numerical loading in group min_t (minority.threat)
-m2a <- m2
-m2a[1] <- 'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + 1 * abstract + label(rep("load_v", 4)) * verbal'
-
-## Step 3: Fix variances across groups
-m3 <- m2a
-m3[7:9] <- c(
-'numerical ~~ label(rep("var_n", 4)) * numerical',
-'abstract  ~~ label(rep("var_a", 4)) * abstract',
-'verbal    ~~ label(rep("var_v", 4)) * verbal')
-
-## Step 3a: Free numerical variance in group min_t
[…]

### 2012-02-11 www/wdh-simplification.R

@@ -1,30 +1,0 @@
-## Package and data
-library("lavaan")
-library(strucchange)
-source("lavaan-S3.R")
-data("StereotypeThreat", package = "psychotools")
-
-## Include group variable and order correspondingly
-StereotypeThreat <- transform(StereotypeThreat, group = interaction(ethnicity, condition))
-StereotypeThreat <- StereotypeThreat[order(StereotypeThreat$group),]
-
-## Disregard groups:
-m2.simp  <- c(
-'ability   =~ label("load_n") * numerical  + 1 * abstract + label("load_v") * verbal',
-'ability   ~  0 * 1',
-'ability   ~~ label("lvar") * ability',
-
-'numerical ~  label("mean_n:min_t") * 1',
-'abstract  ~  label("mean_a:min_t") * 1',
-'verbal    ~  label("mean_v:min_t") * 1',
-
-'numerical ~~ label("var_n:min_t") * numerical',
-'abstract  ~~ label("var_a:min_t") * abstract',
-'verbal    ~~ label("var_v:min_t") * verbal')
-
-m2.simp.fit <- lavaan(m2.simp, data = StereotypeThreat, meanstructure = TRUE)
-
-gefp_m2.simp <- gefp(m2.simp.fit, fit = NULL, order.by = StereotypeThreat$gpa, sandwich = FALSE) #, parm=c(1,4,7))  To test only parameters related to numerical variable
-
-plot(gefp_m2.simp, functional=maxBB)
-plot(gefp_m2.simp, functional=meanL2BB)

### 2012-02-11 www/replication.R
updated replication script:
- small touch ups (mostly labeling) for artificial data example

- added stereotype threat application code

- simulation study still lacks n = 50 case => resimulate all
  results when paper is accepted



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@113 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -7,8 +7,11 @@ library("OpenMx")
 library("lavaan")
 library("strucchange")
 library("mvtnorm")
+library("lattice")
 source("mz.R")
+source("lavaan-S3.R")
 source("sim.R")
+
 
 ##################################
 ## Example with Artificial Data ##
@@ -33,15 +36,13 @@ gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
   vcov = info.mzfit, sandwich = FALSE)
 
 ## Figure 2
- # pdf("gefp_3_19.pdf",height = 9, width = 6)
 par(mfcol = c(3, 2))
-plot(gefp_3_info,  functional = maxBB, main = "Three Parameters\nDM", xlab = "Age")
-plot(gefp_3_info,  functional = meanL2BB, main = "CvM", xlab = "Age")
-plot(gefp_3_info,  functional = supLM(0.1), main = "maxLM", xlab = "Age")
-plot(gefp_19_info, functional = maxBB, main = "Nineteen Parameters\nDM", xlab = "Age")
-plot(gefp_19_info, functional = meanL2BB, main = "CvM", xlab = "Age")
-plot(gefp_19_info, functional = supLM(0.1), main = "maxLM", xlab = "Age")
- # dev.off()
+plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
+plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
+plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
+plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
 
 
 ## Figure 3
@@ -50,31 +51,27 @@ lrstat <- function(age) 2 * as.vector(logLik(mzfit(d[d$age <= age,])) + logLik(m
 mz.age <- sort(d$age[d$age >= quantile(d$age, 0.1) & d$age <= quantile(d$age, 0.9)])
 mz.lrstat <- sapply(mz.age, lrstat)
 
- # pdf("lmlr_19.pdf", height = 5, width = 6)
 par(mfrow = c(1, 1))
 plot(gefp_19_info, functional = supLM(0.1),
[…]

### 2012-04-27 www/paper.Rnw
Minor revisions to the manuscript, as suggested by reviewers.
Also minor revision to sim.R, to include the n=50 condition.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@126 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+%% TODO 1. discussion of measurement vs structural parameters
+%%      2. discussion of longitudinal MI
+%%      3. re-run simulations
 \title{Tests of measurement invariance without subgroups:
@@
   definition of the number of groups, group membership, and offending model
-  parameters. In this paper, we construct tests of measurement invariance based on
+  parameters. In this paper, we study tests of measurement invariance based on
   stochastic processes of casewise derivatives of the likelihood function. These
@@
   tests with the lavaan package, and the participants of the
-  Psychoco 2012 workshop on psychometric computing for
-  discussions.
+  Psychoco 2012 workshop on psychometric computing for helpful 
+  discussion.
   Correspondence to Edgar C.\ Merkle, Department of  
@@
 
-In this paper, we consider a new family of tests for assessing
-measurement invariance that has important advantages over existing
-tests.  We begin by developing a general framework for the tests.
-This leads to a discussion of theoretical results relevant to the
-proposed tests, as well as a comparison of the proposed tests to the 
-existing tests.  Next,
-we study the 
+In this paper, we apply a family of statistical tests based on
+stochastic processes to the assessment of measurement invariance.
+The tests are shown to have useful advantages over existing
+tests.  We begin by developing a general framework for the tests,
+including discussion of theoretical results relevant to the
+proposed tests and comparison of the proposed tests to the 
+existing tests.  Next, we study the 
 proposed tests' abilities through example and simulation.  Finally,
@@
 although the asymptotically equivalent  Lagrange multiplier
-(LM) tests (also known as score tests) and Wald tests may also be constructed for this purpose (see
+(LM) tests (also known as score tests) and Wald tests may also be used for this purpose (see
 \citeNP{Sat89}). If $\nu$
@@
 potential subgroups. Tests for potential changes/instabilities only in the
-$k^*$ remaining parameters (from overall $k$ parameters) are easily constructed
+$k^*$ remaining parameters (from overall $k$ parameters) are easily obtained
 by omitting those $k - k^*$ columns from ${\bm B}(\hat {\bm \theta})_{ij}$ that are
@@
 is higher when testing only the affected model parameters, but
-the affected parameters are usually unknown in practice.
+the affected parameters are usually unknown in practice.  Thus, the
+$k^*=3$ and $k^*=19$ conditions reflect boundary cases, with the
+performance of tests involving other subsets of model parameters
+(e.g., all factor loadings) falling in between the boundaries.  
 Finally, we also examine the tests' power across
@@
 Separate curves are drawn for the double-max test (solid lines), 
-the Cram{\'e}r-von Mises test (dashed lines), and the $\max \mathit{LM}$ (dotted lines).
+the Cram{\'e}r-von Mises test (dashed lines), and the $\max
+\mathit{LM}$ test (dotted lines).
 One can generally observe that simultaneous tests of all 19~parameters
@@
 followed by the double-max test.  
-This advantage decreases at larger sample sizes.  
+This advantage decreases at larger sample sizes, and we also surmise
+that it decreases as extra parameters satisfying measurement
+invariance are included in the tests.  
 At $n=50$, generally regarded as a
@@
 are being tested. It also shows that the Cram\'{e}r-von Mises test generally
-is very close to its nominal type~I error rate, with the double-max test being somewhat
+is very close to its nominal type~I error rate, with the double-max
+test being somewhat 
 conservative and the $\max \mathit{LM}$ test being slightly liberal.
@@
 Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
-statistic from \eqref{eq:maxlm}.
+statistic from \eqref{eq:maxlm}.  We focus on the double-max statistic
+due to its ease of interpretation and intuitive visual display.
 
@@
 \section{General Discussion}
-In this paper, we have presented a new family of statistical tests for
-the study of measurement invariance in psychometrics.  The tests,
-based on stochastic processes, have reasonable power, can isolate
+In this paper, we have applied a family of stochastic process-based
+statistical tests to  
+the study of measurement invariance in psychometrics.  The tests
+have reasonable power, can isolate
 subgroups of individuals violating measurement invariance based on a
@@
 \subsection{Extensions}
-The proposed family of tests can be extended in various ways.  First,
+The family of tests described in this paper can be extended in various
+ways.  First, 
 it is possible to construct an algorithm that recursively defines

### 2012-04-27 www/sim.R

@@ -93,7 +93,7 @@ testpower <- function(nrep = 5000, size = 0.05, verbose = TRUE, ...)
 
 ## Loop over scenarios
 simulation <- function(diff = seq(0, 4, by = 0.25),
-  nobs = c(100, 200, 500), verbose = TRUE, ...)
+  nobs = c(50, 100, 200, 500), verbose = TRUE, ...)
 {
   prs <- expand.grid(diff = diff, nobs = nobs)
   nprs <- nrow(prs)

### 2012-04-30 www/paper.Rnw
Revisions: o Addressed reviewers' final comments o Re-ran simulation with n=50 condition o Added two new references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@127 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-%% TODO 1. discussion of measurement vs structural parameters
-%%      2. discussion of longitudinal MI
-%%      3. re-run simulations
+%%      4. deal with color in figures
 \title{Tests of measurement invariance without subgroups:
@@
     \label{eq:midef}
-      f({\bf X} | T, V, \dots) = f({\bf X} | T, \dots),
-\end{equation}
-where ${\bf X}$ is a data matrix, $T$ is the latent
-variable that the scales purport to measure, and $f$ is the model's
-distributional form \cite{Mel89}.  In the parametric,
-factor-analytic framework
-adopted here, this implies that the measurement parameters are equal
-across subgroups of individuals and, thus, do not vary with any
-variable $V$ \cite{Mer93}.
+      f(x_i | t_i, v_i, \dots) = f({\bf x}_i | t_i, \dots),
+\end{equation}
+where $x_i$ is the data vector for individual $i$, $t_i$ is the latent
+variable for individual $i$ that the scales purport to measure, and
+$f$ is the model's 
+distributional form \cite{Mel89}.  
+We adopt a parametric,
+factor-analytic framework here, so that the above equation being true
+for all $V$ implies that the measurement
+parameters are equal 
+across individuals and, thus, do not vary with any
+$V$ \cite{Mer93}.
 
@@
 \end{equation}
-In this case, the covariance matrix $\bm{C}(\bm{\theta})$ from the expected
-outer product of gradients is
+The covariance matrix $\bm{C}(\bm{\theta})$ from the expected
+outer product of gradients is then 
 \begin{equation}
@@
 
+We included a factor mean in the application in order to
+demonstrate the tests' generality.  To deal with 
+measurement invariance specifically, however, we may elect to focus on
+only the 
+measurement parameters within the model (omitting the factor mean).
+That is, it would 
+not have been notable or surprising if we had observed that
+the factor mean lacked measurement invariance with respect to GPA:
+this result would have implied that individuals' latent
+intelligence fluctuates with GPA.  We speculate that such a result
+would often be obtained when $V$ is related to the latent variable of
+interest.
+
 \section{General Discussion}
@@
 The proposed tests also 
-readily extend to other popular psychometric models.  
+readily extend to other psychometric models.  
 For example, the tests can be used to generally 
 study the stability of structural equation model parameters across
-observations.  Under this framework, it is possible to assess whether
-paths between 
-latent variables are stronger for some individuals than for
-others.  Further, in applying the proposed tests to factor mean
-parameters (as was done in the example), implications of measurement
-invariance violations for structural equation modeling may be better
-understood.
-
-The proposed tests also naturally extend to the study of differential
+observations.  This includes the study of measurement invariance
+in second-order growth models and related models for longitudinal
+data (e.g., \citeNP{FerBal08,McA09}).
+The issue of measurement invariance is important in these models in
+order to establish that a ``true'' change has occurred in the 
+individuals to which the model has been fitted.  The tests described
+here can be used to help establish measurement invariance with respect
+to both continuous and categorical auxiliary variables, using only
+output from the estimated model of interest.
+
+Finally, the tests can be used for the study of differential
 item functioning (DIF) 
@@
 that subgroups must be specified in advance.
-%% The tests proposed here can detect subgroups automatically, however, offering a unified way of
-%% studying both  measurement invariance in factor analysis and differential item
-%% functioning in item response.
+The tests proposed here can detect subgroups automatically, however, offering a unified way of
+studying both  measurement invariance in factor analysis and differential item
+functioning in item response.
 While factor-analytic measurement invariance methods and DIF methods

### 2012-05-09 m3_2012/invariance.tex
Added draft slides for the m3 conference.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@128 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2012-02-08 www/psychoco2012/invariance.tex
@@
 
-\title[Measurement Invariance]{Generalized Measurement Invariance
-  Tests for Factor Analysis}
+\title[Measurement Invariance]{Tests of Measurement Invariance without Subgroups}
 \author{Ed Merkle\inst{1} \and Achim Zeileis\inst{2}}
 \institute{\inst{1} University of Missouri \and \inst{2} Universit\"{a}t Innsbruck}
-\date{Supported by grant SES-1061334 from the U.S.\ National
+\date{Supported by grant SES-1061334 from the National
   Science Foundation}
@@
   \begin{center}
-      \includegraphics[height=3in]{model_under16.pdf}
+      \includegraphics[height=3in]{../psychoco2012/model_under16.pdf}
   \end{center}}
@@
   \begin{center}
-      \includegraphics[height=3in]{model_over16.pdf}
+      \includegraphics[height=3in]{../psychoco2012/model_over16.pdf}
   \end{center}}
@@
   \begin{center}
-      \includegraphics[height=3in]{paper-lm-lr-19.pdf}
+      \includegraphics[height=3in]{../psychoco2012/paper-lm-lr-19.pdf}
   \end{center}
@@
   \begin{center}
-      \includegraphics[height=3in]{paper-mzsim-xyplot.pdf}
+      \includegraphics[height=3in]{../psychoco2012/paper-mzsim-xyplot.pdf}
   \end{center}
@@
   \frametitle{Model}
-  \begin{center}
-      \includegraphics[height=3in]{wicherts_msp.pdf}
-  \end{center}}
+  %\begin{center}
+      \includegraphics[height=2.3in]{../psychoco2012/wicherts_msp.pdf}
+  %\end{center}
+}
 
@@
   \begin{center}
-      \includegraphics[height=3in]{gefp_4_ind.pdf}
+      \includegraphics[height=3in]{../psychoco2012/gefp_4_ind.pdf}
   \end{center}}
@@
   \begin{center}
-      \includegraphics[height=3in]{gefp_4_agg.pdf}
+      \includegraphics[height=3in]{../psychoco2012/gefp_4_agg.pdf}
   \end{center}}
@@
         \item Isolating specific parameters that violate
-          measurement invariance, allowing the researcher to define
-          specific types of measurement invariance ``post hoc''
-          instead of ``a priori''.
+          measurement invariance, allowing the researcher to interpret the nature of measurement invariance violations.
         \item Isolating
@@
   \begin{itemize}
-    \item To carry out the tests, we utilize
+    \item We can carry out the tests for general SEMs in R, using
       \begin{itemize}
@@
         \item \texttt{strucchange} for carrying out the proposed tests with the scores.
-          \begin{itemize}
-            \item Required input: Fitted model, function for score extraction, and information matrix (optional).
-            \item \texttt{gefp()} constructs the process.
-            \item \texttt{sctest()} and \texttt{plot()} calculate and visualize test statistics.
-          \end{itemize}
+          % \begin{itemize}
+          %   \item Required input: Fitted model, function for score extraction, and information matrix (optional).
+          %   \item \texttt{gefp()} constructs the process.
+          %   \item \texttt{sctest()} and \texttt{plot()} calculate and visualize test statistics.
+          % \end{itemize}
       \end{itemize}
@@
     \item Extension to related psychometric issues.\\ \ \\
-    \item Working paper: \url{http://econpapers.repec.org/RePEc:inn:wpaper:2011-09}
+    \item Paper: Merkle, E.\ C.\ \& Zeileis, A.\ (accepted).  Tests of measurement invariance without subgroups:\ {A} generalization of classical methods.  {\em Psychometrika}.
   \end{itemize}}

### 2012-05-31 www/estfun-lavaan.R
lavaan@Sample is now called lavaan@SampleStats (lavaan >= 0.4-14).
lmtest (>= 0.9-30) can now also deal with S4 objects, so the S3 glue in
lavaan-S3.R is not needed anymore.

estfun-lavaan.R thus only contains the updated estfun.lavaan() function.

Small improvements in replication script.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@129 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,23 +1,11 @@
-## provide S3 counterparts to S4 methods in lavaan
-## (this is really a bad hack and lmtest/strucchange probably need improvement)
-coef.lavaan <- function(object, ...) getMethod("coef", "lavaan")(object, ...)
-vcov.lavaan <- function(object, ...) getMethod("vcov", "lavaan")(object, ...)
-coef.lavaan <- function(object, ...) getMethod("coef", "lavaan")(object, ...)
-residuals.lavaan <- function(object, ...) getMethod("residuals", "lavaan")(object, ...)
-logLik.lavaan <- function(object, ...) getMethod("logLik", "lavaan")(object, ...)
-
-## new methods
-formula.lavaan <- function(x, ...) x@Options$model.syntax
-df.residual.lavaan <- function(object, ...) NULL
-
 estfun.lavaan <- function(x)
 {
   ## observed data
   X <- x@Data@X
   ## number variables/sample size
   nvar <- x@Model@nvar
-  nobs <- x@Sample@ntotal
-  ntab <- unlist(x@Sample@nobs)
+  nobs <- x@SampleStats@ntotal
+  ntab <- unlist(x@SampleStats@nobs)
 
   ## number of groups
   ngrp <- x@Model@ngroups

### 2012-05-31 www/replication.R

@@ -9,7 +9,7 @@ library("strucchange")
 library("mvtnorm")
 library("lattice")
 source("mz.R")
-source("lavaan-S3.R")
+source("estfun-lavaan.R")
 source("sim.R")
 
 
@@ -172,7 +172,7 @@ plot(gefp_4_info,  functional = maxBB, aggregate = FALSE,
 
 ## p-values from all three tests under both sets of constraints
 tests <- list(maxBB, meanL2BB, supLM(0.1))
-pvals <- matrix(NA, nrow = 2 * length(tests), ncol = 2)
+pvals <- matrix(NA, nrow = length(tests), ncol = 2)
 for (i in 1:length(tests)){
   pvals[i, 1] <- sctest(gefp_4_info,   functional = tests[[i]])$p.value
   pvals[i, 2] <- sctest(gefp_4_info_1, functional = tests[[i]])$p.value

### 2012-05-31 www/paper.Rnw
Improved paper.Rnw so that instead of the default red boundaries, gray boundaries are plotted. Requires strucchange >= 1.4-7.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@130 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<preliminaries>>=
+## packages
 library("OpenMx")
@@
 library("lattice")
+
+## auxiliary code
 source("mz.R")
-source("lavaan-S3.R")
+source("estfun-lavaan.R")
 source("sim.R")
+
+## convenience function for plotting boundaries without color
+get_boundary <- function(obj, fun, trim = 0) {
+  bound <- fun$boundary(0:obj$nobs/obj$nobs)
+  bound <- fun$computeCritval(0.05, ncol(obj$process)) * bound
+  bound <- zoo(bound, time(obj$process))
+  if(trim > 0) bound <- head(tail(bound, - floor(trim * obj$nobs)), - floor(trim * obj$nobs))
+  return(bound)
+}
 @
@@
 \caption{A comparison of the $\max \mathit{LM}$ (solid line) and $\max \mathit{LR}$ (dashed line) test
-  statistics for~\eqref{eq:h0} (i.e., $k^* = 19$).  The solid horizontal line
+  statistics for~\eqref{eq:h0} (i.e., $k^* = 19$).  The gray horizontal line
   corresponds to the critical value at $\alpha= 0.05$ while the dotted vertical
@@
 plot(gefp_19_info, functional = supLM(0.1),
-  xlab = "Age", ylab = "LR and LM statistics (k* = 19)", main = "", ylim = c(0, 50))
+  xlab = "Age", ylab = "LR and LM statistics (k* = 19)",
+  main = "", ylim = c(0, 50), boundary = FALSE)
 legend("topleft", c("LR", "LM"), lty = c(2, 1), bty = "n")
 lines(mz.age, mz.lrstat, lty = 2)
+lines(get_boundary(gefp_19_info, supLM(0.1), 0.1), col = "slategray", lwd = 1.7)
 abline(v = mz.age[which.max(mz.lrstat)], lty = 3)
@@
 \caption{Cumulative score processes for each verbal factor loading. 
-  The horizontal lines correspond to the critical value of the
+  The solid gray horizontal lines correspond to the critical value of the
   double-max test at $\alpha= 0.05$.}
@@
 <<gefp-3-visualization, fig=TRUE, height=8, width=5.5>>=
-plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
-  main = "DM, k* = 3", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
+gefp_3_bound <- get_boundary(gefp_3_info, maxBB)
+plot(gefp_3_info$process,  ylim = c(-2, 2),
+  main = "DM, k* = 3", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]),
+  panel = function(x, y, ...) {
+    abline(h = 0)
+    lines(gefp_3_bound, col = "slategray", lwd = 1.7)
+    lines(-gefp_3_bound, col = "slategray", lwd = 1.7)
+    lines(x, y, ...)
+  })
 @
@@
   based on the example involving measurement invariance with respect
-  to student age.  Solid, horizontal lines represent critical values
+  to student age.  Solid gray horizontal lines represent critical values
   at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
@@
 par(mfcol = c(3, 2))
-plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
-plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
-plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
-plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
+plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age", boundary = FALSE)
+lines(get_boundary(gefp_3_info, maxBB), col = "slategray", lwd = 1.7)
+plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age", boundary = FALSE)
+lines(get_boundary(gefp_3_info, meanL2BB), col = "slategray", lwd = 1.7)
+plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age", boundary = FALSE)
+lines(get_boundary(gefp_3_info, supLM(0.1), 0.1), col = "slategray", lwd = 1.7)
+plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age", boundary = FALSE)
+lines(get_boundary(gefp_19_info, maxBB), col = "slategray", lwd = 1.7)
+plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age", boundary = FALSE)
+lines(get_boundary(gefp_19_info, meanL2BB), col = "slategray", lwd = 1.7)
+plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age", boundary = FALSE)
+lines(get_boundary(gefp_19_info, supLM(0.1), 0.1), col = "slategray", lwd = 1.7)
 @
@@
   using data from Study 1 of Wicherts et al.\ (2005). %% \citeA{WicDol05} does not work in caption
-  The horizontal lines correspond to the critical value of
+  The solid gray horizontal lines correspond to the critical value of
   the doule-max test at $\alpha = 0.05$.}
@@
 <<gefp-4-plot, fig=TRUE, height=8, width=5.5>>=
-plot(gefp_4_info,  functional = maxBB, aggregate = FALSE,
+gefp_4_bound <- get_boundary(gefp_4_info, maxBB)
+plot(gefp_4_info$process,
   ylim = c(-2.15, 2.15), main = "DM, k* = 4", xlab = "GPA",
-  ylab = expression(lambda[num], eta, mu[num], psi[num]))
+  ylab = expression(lambda[num], eta, mu[num], psi[num]),
+  panel = function(x, y, ...) {
+    abline(h = 0)
+    lines(gefp_4_bound, col = "slategray", lwd = 1.7)
+    lines(-gefp_4_bound, col = "slategray", lwd = 1.7)
+    lines(x, y, ...)
+  })
+
 @

### 2012-06-07 prelim_code/ordinal_warning.R
Added code to replicate a warning coming from ordL2BB().
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@132 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,26 @@
+library("lavaan")
+library("strucchange")
+source("../www/estfun-lavaan.R")
+source("efpFunctional-cat.R")
+
+## Define 15-category grouping variable
+grp <- c(rep(1:15, each=20),15)
+
+## Fit a model
+HS.model <- ' visual  =~ x1 + x2 + x3
+              textual =~ x4 + x5 + x6
+              speed   =~ x7 + x8 + x9 '
+     
+fit <- cfa(HS.model, data=HolzingerSwineford1939, meanstructure=TRUE)
+
+fit.gefp <- gefp(fit, fit = NULL, order.by = grp, sandwich = FALSE,
+                 parm=7:15)
+
+## Takes about 5min on my machine, with this warning:
+#### $`data length [6772] is not a sub-multiple or multiple of the number of rows [753]`
+#### matrix(rnorm(nproc * nobs), ncol = nproc)
+critval <- ordL2BB(fit.gefp)
+
+## No warnings with this, so I think the issue lies on
+## line 36 of efpFunctional-cat.R
+critval2 <- ordL2BB(fit.gefp, nobs = 301)

### 2012-06-08 www/paper.Rnw
Various small touch-ups and improvements. Should be ready to go now.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@134 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \end{equation}
-which, when \eqref{eq:h1*} holds, has an asymptotic $\chi^2$ with degrees
+which, when the null hypothesis holds, has an asymptotic $\chi^2$ with degrees
 of freedom equal to the number of parameters in~${\bm \theta}$.
@@
 For application to measurement invariance, the most important
-theoretical result involves the fact that, under the $H_0$ in \eqref{eq:h0}, the
+theoretical result involves the fact that, under $H_0$ in \eqref{eq:h0}, the
 \emph{cumulative score process} converges to a specific asymptotic
@@
 of ${\bm B}(\hat {\bm \theta})$ corresponds to the first individual's
-scores.  The second row of ${\bm B}(\hat {\bm \theta})$ corresponds to
+decorrelated scores.  The second row of ${\bm B}(\hat {\bm \theta})$ corresponds to
 the sum of the first two individuals' scores, etc., until the last row
@@
 unknown). A particularly agnostic aggregation strategy is to reject $H_0$
-if any component of the the cumulative score process ${\bm B}(t; \hat {\bm \theta})$
-strays ``too far'' from zero at any time, i.e., if
+if any component of the cumulative score process ${\bm B}(t; \hat {\bm \theta})$
+strays ``too far'' from zero at any time, i.e., if the double maximum statistic
 \begin{equation}
@@
 critical value. For an example of this see Figure~\ref{fig:lambdas}
-(in the ``Example'' section) which shows the the cumulative score processes for three
+(in the ``Example'' section) which shows the cumulative score processes for three
 factor loadings along with the critical values at 5\% level.
@@
 $||\cdot||_\infty$ denotes the maximum norm. Similarly, $\mathit{CvM}$
-from (\ref{eq:cvm}) converges to $\int_0^1 || {\bm B}^0(t) ||_2^2 d t$, where $||\cdot||^2$
+from (\ref{eq:cvm}) converges to $\int_0^1 || {\bm B}^0(t) ||_2^2 d t$, where $||\cdot||_2$
 denotes the Euclidean norm. Finally, $\max \mathit{LM}$ from (\ref{eq:maxlm}) -- and analogously
@@
 also contain a dashed line depicting the value of the test
-statistic (test statistics for the others are simply 
+statistic $\mathit{CvM}$ (test statistics for the others are simply 
 the maxima of the processes).
@@
 Data were generated from the same model and with the same type of
-measurement invariance as was described in the ``Artificial Example'' section.
+measurement invariance violation as was described in the ``Artificial Example'' section.
 Sample size and magnitude
@@
 \lambda_{21}, \lambda_{31}\}$ deviating from the older students'
-values by $d$ 
-times the parameters' asymptotic standard errors (scaled by $\sqrt{n}$),
-with $d = 0, 0.25, 0.5, \dots, 4$.  The 0-standard error condition was
+values by $d$~times the parameters' asymptotic standard errors (scaled by $\sqrt{n}$),
+with $d = 0, 0.25, 0.5, \dots, 4$.  The $d=0$ standard error condition was
 used to study type~I error rate.
@@
 Full simulation results are presented in Figure~\ref{fig:simres}, and
-the underlying numeric values for a subset of the results is additionally displayed in
+the underlying numeric values for a subset of the results are additionally displayed in
 Table~\ref{tab:simres}.  In describing the results, we largely refer
@@
 example below, we employ one of the models used by WDH.
-Our $V$ is participants' aptitudes, as measured by their grade point averages (GPAs).  The GPAs were unused in the original analyses.
+Our $V$ is participants' aptitudes, as measured by their grade point averages (GPAs) which were unused in the original analyses.
 
@@
 group that received stereotype primes, which is pictured in Figure~\ref{fig:wpath}.
-In the figure, paths with dashed lines and red parameters signify group-specific parameters.
+In the figure, paths with dashed lines and \fixme{red} parameters signify group-specific parameters.
 These include the factor mean $\eta$, numerical factor loading $\lambda_{\text{num}}$, numerical intercept $\mu_{\text{num}}$, and numerical unique variance $\psi_{\text{num}}$.
@@
 \begin{figure}
-\caption{Path diagram representing the submodel for the ``minority, stereotype'' group in model 5b of WDH.
-Dashed paths and red parameters are group specific.
+\caption{Path diagram representing the submodel for the ``minority, stereotype prime'' group in model 5b of WDH.
+Dashed paths and red (FIXME) parameters are group specific.
 Other parameters (not displayed) were restricted to be equal across all four groups, with the factor variance only being restricted across the two minority groups.}
@@
   in the ``minority, stereotype prime'' group
-  using data from Study 1 of Wicherts et al.\ (2005). %% \citeA{WicDol05} does not work in caption
+  using data from Study 1 of WDH. %% \citeA{WicDol05} does not work in caption
   The solid gray horizontal lines correspond to the critical value of
@@
 whereby continuous moderators are allowed to linearly impact model
-parameters (Purcell also considers quadratic effects of the
+parameters (\citeNP{Pur02}, also considers quadratic effects of the
 moderators). Moreover, if the change in

### 2012-06-08 prelim_code/efpFunctional-cat.R
assure that nobs is integer
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@135 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -33,7 +33,7 @@ ordL2BB <- function(freq, nobs = NULL, nproc = NULL, ...)
   freq <- freq / sum(freq)
   ncat <- length(freq)
   tcat <- cumsum(freq[-ncat])
-  if(is.null(nobs)) nobs <- 50 / min(freq)
+  if(is.null(nobs)) nobs <- ceiling(50 / min(freq))
 
   catwmax <- function(x) {
     n <- length(x)

### 2012-06-08 www/paper.Rnw
Final paper revisions: o Removed color from remaining figures, fixed text. o Added a citation based on some recent comments. o Spell check
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@136 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-%%      4. deal with color in figures
 \title{Tests of measurement invariance without subgroups:
@@
   Yves Rosseel, who provided feedback and code for performing the
-  tests with the lavaan package, and the participants of the
+  tests with the lavaan package, Kris Preacher, who provided helpful
+  comments 
+  on the manuscript, and the participants of the
   Psychoco 2012 workshop on psychometric computing for helpful 
@@
 many of the $n$~individuals (and not just a single threshold). In such
-cases, sums insteads of maxima are more suitable for collapsing across
+cases, sums instead of maxima are more suitable for collapsing across
 parameters and/or individuals, because they combine the deviations instead of
@@
 by omitting those $k - k^*$ columns from ${\bm B}(\hat {\bm \theta})_{ij}$ that are
-restricted/stable, retaining only those $k^*$ columns that are potentially instable.
+restricted/stable, retaining only those $k^*$ columns that are potentially unstable.
 This may be of special interest to those wishing to test specific types of measurement
@@
 statistics from the previous section is straightforward: it is simply the aggregation
-of the asympotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
+of the asymptotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
 from (\ref{eq:dmax}) converges to $\sup_t || {\bm B}^0(t) ||_{\infty}$, where
@@
 Using results from \citeA{HjoKon02} and \citeA{ZeiHor07}, we can also
-capture the behaviour of the process $\bm{B}(t, \hat{\bm{\theta}})$
+capture the behavior of the process $\bm{B}(t, \hat{\bm{\theta}})$
 under specific alternatives 
@@
 group that received stereotype primes, which is pictured in Figure~\ref{fig:wpath}.
-In the figure, paths with dashed lines and \fixme{red} parameters signify group-specific parameters.
+In the figure, paths with dashed lines and parameters in bold/italics signify group-specific parameters.
 These include the factor mean $\eta$, numerical factor loading $\lambda_{\text{num}}$, numerical intercept $\mu_{\text{num}}$, and numerical unique variance $\psi_{\text{num}}$.
@@
 \caption{Path diagram representing the submodel for the ``minority, stereotype prime'' group in model 5b of WDH.
-Dashed paths and red (FIXME) parameters are group specific.
+Dashed paths and parameters in bold/italics are group specific.
 Other parameters (not displayed) were restricted to be equal across all four groups, with the factor variance only being restricted across the two minority groups.}
@@
 wdh5b <- lavaan(
-   'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical + label(rep("load_v", 4)) * verbal + 1 * abstract
-    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
-    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
-    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
-    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
-    verbal    ~  label(rep("mean_v", 4)) * 1
-    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
-    abstract  ~~ label(rep("var_a", 4)) * abstract
-    verbal    ~~ label(rep("var_v", 4)) * verbal',
+   'ability   =~ c(rep("load_n", 3), "load_n:min_t") * numerical + rep("load_v", 4) * verbal + 1 * abstract
+    ability   ~  c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t") * 1 + c(0, NA, NA, NA) * 1
+    ability   ~~ c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min") * ability
+    numerical ~  c(rep("mean_n", 3), "mean_n:min_t") * 1
+    abstract  ~  c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2)) * 1
+    verbal    ~  rep("mean_v", 4) * 1
+    numerical ~~ c(rep("var_n", 3), "var_n:min_t") * numerical
+    abstract  ~~ rep("var_a", 4) * abstract
+    verbal    ~~ rep("var_v", 4) * verbal',
   data = wdh, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")
@@
 wdh5b_1 <- lavaan(
-   'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
-    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
-    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
-    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
-    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
-    verbal    ~  label(rep("mean_v", 4)) * 1
-    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
-    abstract  ~~ label(rep("var_a", 4)) * abstract
-    verbal    ~~ label(rep("var_v", 4)) * verbal',
+   'ability   =~ 1 * verbal + c(rep("load_n", 3), "load_n:min_t") * numerical  + rep("load_a", 4) * abstract
+    ability   ~  c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t") * 1 + c(0, NA, NA, NA) * 1
+    ability   ~~ c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min") * ability
+    numerical ~  c(rep("mean_n", 3), "mean_n:min_t") * 1
+    abstract  ~  c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2)) * 1
+    verbal    ~  rep("mean_v", 4) * 1
+    numerical ~~ c(rep("var_n", 3), "var_n:min_t") * numerical
+    abstract  ~~ rep("var_a", 4) * abstract
+    verbal    ~~ rep("var_v", 4) * verbal',
   data = wdh, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
@@
   The solid gray horizontal lines correspond to the critical value of
-  the doule-max test at $\alpha = 0.05$.}
+  the double-max test at $\alpha = 0.05$.}
 \label{fig:cusumgpa}
@@
 perform better in this particular case.  Examples include the 
-class of moderated factor models \cite{BauHus09,MolDol10,Pur02}, 
+class of moderated factor models \cite{BauHus09,MolDol10,Nea06,Pur02}, 
 whereby continuous moderators are allowed to linearly impact model

### 2012-06-08 ordinal/mz.R
Added 'ordinal' directory containing code for ordL2BB simulation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@137 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,74 @@
+ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
+{
+  ## data checks
+  data <- as.data.frame(data)
+  stopifnot(c("x1", "x2", "x3", "y1", "y2", "y3", "age") %in% names(data))
+  age <- data$age
+
+  ## require package
+  stopifnot(require("lavaan"))
+
+  ## Fit multi-group models for LRT; similar to catL2BB
+  rval <- 'verbal =~ x1 + x2 + x3; math =~ y1 + y2 + y3;
+           verbal ~ 0*1; math ~ 0*1'
+
+  m2 <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
+            group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances")))
+  m3 <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
+            group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances","residuals","means")))
+
+  ## model from Merkle & Zeileis
+  rval <- 'verbal =~ x1 + x2 + x3; math =~ y1 + y2 + y3'
+
+  data <- data[, c("x1", "x2", "x3", "y1", "y2", "y3")]
+  
+  ## fit model
+  rval <- cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE)
+
+  ## p-value for LRT Stat
+  if(!inherits(m2, "try-error") & !inherits(m3, "try-error")) {
+    lrt.p <- anova(m3,m2)[[7]][2]
+  } else {
+    lrt.p <- NA
+  }
+  
+  ## store (1) fitted model object, (2) lavaan coefficient
+  ## labeling and ordering
+  rval <- list(
+    data = data,
+    model = rval,
+    lrt.p = lrt.p,
+    names = c("verbal=~x1", "verbal=~x2", "verbal=~x3", "math=~y1", "math=~y2", "math=~y3", "x1~~x1", "x2~~x2", "x3~~x3",
+        "y1~~y1", "y2~~y2", "y3~~y3", "verbal~~math", "x1~1", "x2~1", "x3~1", "y1~1", "y2~1", "y3~1")
+            )
+  
+  class(rval) <- "mzfit"
+
+  return(rval)
+}
+
+coef.mzfit <- function(object, ...) {
[…]

### 2012-06-08 ordinal/sim_ordinal.R

@@ -1,0 +1,174 @@
+
+## Data-generating process
+dgp <- function(nobs = 200, diff = 3, nlevels=10)
+{
+  # Generates data from a factor analysis model that violates
+  # measurement invariance.  Also generates an ordinal auxiliary
+  # variable related to the invariance (loosely called "age").
+  stopifnot(require("mvtnorm"))
+  
+  # ses is SE increase in parameters for each increasing value
+  # of "age" past the halfway point.
+  sampsize <- nobs
+  ses <- diff
+  sampsize.per.nobs <- nobs %/% nlevels
+
+  ## Deal with cases where nobs/nlevels is not an integer
+  age <- rep(NA, sampsize)
+  age[1:(nlevels*sampsize.per.nobs)] <- rep(1:nlevels, each=sampsize.per.nobs)
+  age[is.na(age)] <- sample((1:nlevels),sum(is.na(age)))
+  half.level <- ifelse(nlevels %% 2 == 1, (nlevels %/% 2)+1, nlevels/2)
+  
+  # Define parameter vectors/matrices:
+  mu <- c(29.32,24.70,14.84,10.59,19.30,18.01)
+
+  # Loadings:
+  lambda <- matrix(0,6,2)
+  lambda[,1] <- c(4.92,2.96,5.96,0,0,0)
+  lambda[,2] <- c(0,0,0,3.24,4.32,7.21)
+
+  phi <- matrix(.48,2,2)
+  diag(phi) <- 1
+
+  ## Base variances
+  psi <- matrix(0,6,6)
+  diag(psi) <- c(26.77,13.01,30.93,3.17,8.82,22.5)
+
+  ## Initialize data matrix:
+  datmat <- matrix(NA,sampsize,6)
+  colnames(datmat) <- c("x1", "x2", "x3", "y1", "y2", "y3")
+
+  ## Generate z and u vectors for invariant individuals:
+  n.invariant <- sum(age <= half.level)
+  z <- t(rmvnorm(n.invariant,rep(0,2),phi))
+  u <- t(rmvnorm(n.invariant,rep(0,6),psi))
+
+  ## Based on z and u, generate data:
+  tmp.ind <- which(age <= half.level)
+  for (i in 1:n.invariant){
+    datmat[tmp.ind[i],] <- mu + lambda%*%z[,i] + u[,i]
+  }
+
+  ## MODEL CODE FOR FINDING APPROPRIATE SES
[…]

### 2012-06-28 ordinal/sim_ordinal.R
Added functionality to ordinal simulation code: o Added ordwmax() statistic. o Added use of mclapply() to speed up the simulation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@141 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -72,13 +72,13 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10)
 }
 
 ## Evaluate power simulation on a single dgp() scenario
-testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, test = NULL, verbose = TRUE, ...)
+testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, test = NULL, verbose = TRUE, nobs, diff, nlevels)
 {
   pval <- matrix(rep(NA, length(test) * nrep), ncol = length(test))
   colnames(pval) <- test
   
   for(i in 1:nrep) {
-    d <- dgp(...)
+    d <- dgp(nobs, diff, nlevels)
 
     mz <- ordfit(d)
 
@@ -87,9 +87,10 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, test = NULL, verb
 
     if(!inherits(ord_gefp, "try-error")) {
       pval[i, 1] <- sctest(ord_gefp,  functional = ordfun)$p.value
-      pval[i, 2] <- sctest(ord_gefp,  functional = catL2BB(ord_gefp))$p.value
-      pval[i, 3] <- sctest(ord_gefp,  functional = supLM(0.1))$p.value
-      pval[i, 4] <- mz$lrt.p
+      pval[i, 2] <- sctest(ord_gefp,  functional = ordwmax(ord_gefp))$p.value
+      pval[i, 3] <- sctest(ord_gefp,  functional = catL2BB(ord_gefp))$p.value
+      pval[i, 4] <- sctest(ord_gefp,  functional = supLM(0.1))$p.value
+      pval[i, 5] <- mz$lrt.p
     }
   }
   rval <- colMeans(pval < size, na.rm = TRUE)
@@ -99,13 +100,13 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, test = NULL, verb
 }
 
 ## Loop over scenarios
-simulation <- function(diff = seq(0, 3, by = 0.5),
-  nobs = c(120, 480, 960), nlevels = c(4, 8, 12, 20), verbose = TRUE, ...)
+simulation <- function(diff = seq(0, 1.5, by = 0.25),
+  nobs = c(120, 480, 960), nlevels = c(4, 8, 12), verbose = TRUE, ...)
 {
   prs <- expand.grid(diff = diff, nlevels = nlevels, nobs = nobs)
[…]

### 2012-06-28 ordinal/sim_ordinal.R
Further touch-ups to sim_ordinal.R
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@142 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -72,13 +72,13 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10)
 }
 
 ## Evaluate power simulation on a single dgp() scenario
-testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, test = NULL, verbose = TRUE, nobs, diff, nlevels)
+testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, test = NULL, verbose = TRUE, ...)
 {
   pval <- matrix(rep(NA, length(test) * nrep), ncol = length(test))
   colnames(pval) <- test
   
   for(i in 1:nrep) {
-    d <- dgp(nobs, diff, nlevels)
+    d <- dgp(...)
 
     mz <- ordfit(d)
 
@@ -123,12 +123,12 @@ simulation <- function(diff = seq(0, 1.5, by = 0.25),
   do.parallel <- require("parallel")
   if (do.parallel){
     pow <- mclapply(1:nprs, function(i){
-      testpower(nrep = 10, diff = prs$diff[i], nobs = prs$nobs[i],
+      testpower(diff = prs$diff[i], nobs = prs$nobs[i],
                 nlevels = prs$nlevels[i], test = test,
                 ordfun = cval[[which(cval.conds == prs$nlevels[i])]],
                 verbose = verbose)},
-                    mc.cores = round(.75*detectCores())
-    pow <- unlist(pow)
+                    mc.cores = round(.75*detectCores()))
+    pow <- t(matrix(unlist(pow), ntest, nprs))
   } else {
     pow <- matrix(rep(NA, ntest * nprs), ncol = ntest)
     for(i in 1:nprs) {
@@ -137,7 +137,7 @@ simulation <- function(diff = seq(0, 1.5, by = 0.25),
       ## Find critical values we need
       ordfun <- cval[[which(cval.conds == prs$nlevels[i])]]
 
-      pow[i,] <- testpower(nrep = 10, diff = prs$diff[i], nobs = prs$nobs[i], nlevels = prs$nlevels[i], ordfun = ordfun, test = test, verbose = verbose, ...)
+      pow[i,] <- testpower(diff = prs$diff[i], nobs = prs$nobs[i], nlevels = prs$nlevels[i], ordfun = ordfun, test = test, verbose = verbose, ...)
     }
   }
 

### 2012-06-28 imps2012/invariance.tex
Added draft of slides for the Psychometric conference.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@143 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2012-05-09 m3_2012/invariance.tex
@@
 
-\title[Measurement Invariance]{Tests of Measurement Invariance without Subgroups}
+\title[Measurement Invariance]{Tests for Measurement Invariance when Subgroups are Unknown}
 \author{Ed Merkle\inst{1} \and Achim Zeileis\inst{2}}
@@
       assigning scores across diverse groups of individuals.\\ \ \\
-    \item Notable violations of measurement invariance:
-      \begin{itemize}
-        \item SAT for different ethnic groups (Atkinson, 2001)
-        \item Intelligence tests \& the Flynn effect
-          (Wicherts et al., 2004)
-      \end{itemize}
+    \item Commonly studied via factor analysis (today's focus) and
+      item response models.
   \end{itemize}}
@@
 
-\section{Proposed Tests}
-\frame{
-  \frametitle{Proposed Tests}
-  \begin{itemize}
-    \item In contrast to existing tests of measurement invariance, the
-      proposed tests offer the abilities to:
-      \begin{itemize}
-        \item Test for measurement invariance when groups are
-          ill-defined (e.g., when the grouping variable is continuous).
-        \item Test for measurement invariance in any subset of model
-          parameters.
-        \item Interpret the nature of measurement invariance violations.
-      \end{itemize}
-  \end{itemize}}
+\section{Tests Along a Continuum}
+% \frame{
+%   \frametitle{Proposed Tests}
+%   \begin{itemize}
+%     \item In contrast to existing tests of measurement invariance, the
+%       proposed tests offer the abilities to:
+%       \begin{itemize}
+%         \item Test for measurement invariance when groups are
+%           ill-defined (e.g., when the grouping variable is continuous).
+%         \item Test for measurement invariance in any subset of model
+%           parameters.
+%         \item Interpret the nature of measurement invariance violations.
+%       \end{itemize}
+%   \end{itemize}}
 
@@
   \begin{itemize}
-    \item The proposed family of tests rely on first derivatives of
-      the model's log-likelihood function.
-    \item We consider individual terms ({\em scores}) of the gradient.  These scores tell us how well a particular parameter describes a particular individual.
+    \item We propose a number of tests for measurement invariance
+      along a continuous auxiliary variable.
+    \item Tests are based on individual terms ({\em scores}) of the gradient.  The scores tell us how well a particular parameter describes a particular individual.
   % Equation of gradient vs scores for individuals
@@
 
-% \frame{
-%   \frametitle{Proposed Tests}
-%   \begin{itemize}
-%     \item We can also consider individual terms ({\em{scores}}) of the
-%       gradient.  These scores tell us how well a particular
-%       parameter describes a particular individual.  If your score is
-%       0 for some parameter, that parameter describes you well.  If your score
-%       is far from 0, that parameter describes you poorly. \\ \ \\
-%   \end{itemize}
-%   \begin{equation*}
-%   \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
-%   \text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}      
-%   \end{equation*}}
-
 \frame{
@@
     \item Under measurement invariance, parameter estimates should
-      roughly describe everyone equally well.  So people's scores
-      should fluctuate around zero.\\ \ \\
+      roughly describe everyone equally well.  So, people's scores
+      should fluctuate around zero as we move up the auxiliary variable.\\ \ \\
     \item If measurement invariance is violated, the scores should
-      stray from zero.
+      stray from zero around specific points of the auxiliary variable.
   \end{itemize}}
@@
       \begin{itemize}
-        \item Order individuals by an auxiliary variable.\\ \ \\
+        \item Order individuals by the auxiliary variable.\\ \ \\
         \item Define $t \in (1/n, n)$.
@@
 \end{frame}
-
-% EXAMPLE OF CUMULATIVE SCORE
-% \frame{
-%   \frametitle{Cumulative Scores}
-% \begin{center}
-% \begin{tabular}{rrrr}
-%   \hline
-%  & Age & Score & Cumul Sc \\ 
-%   \hline
-% 1 & 13.0 & -0.24 & -0.24 \\ 
-%   2 & 13.5 & 0.17 & -0.08 \\ 
-%   3 & 14.0 & -0.38 & -0.45 \\ 
-%   4 & 14.5 & 0.27 & -0.19 \\ 
-%   5 & 15.0 & -0.27 & -0.46 \\ 
-%    \hline
-% \end{tabular}
-% \end{center}}
 
@@
 
-% \frame{
-%   \frametitle{Brownian Bridge}
-%   \begin{itemize}
-%     \item Brownian bridge: A sequence of normal random variables
-%       $X(t)$ satisfying:
-%     % notation
-% \begin{eqnarray*}
-%     \mu(t) & = & 0\ \forall\ t\\
-%     \Gamma(t_1, t_2) & = & \text{min}(t_1, t_2)\\
-%     X(0) = & X(1) & = 0.
-% \end{eqnarray*}
-%   \end{itemize}}
+\section{Tests Along an Ordinal Variable}
+\frame{
+  \frametitle{Tests Along Ordinal Variables}
+  \begin{itemize}
+    \item The statistics described above do not immediately translate
+      to ordinal auxiliary variables, where there is no unique ordering of
+      individuals.
+    \item To obtain a test statistic, we allow all individuals with the
+      same value of the 
+      auxiliary variable to simultaneously enter into the cumulative
+      sum.  We then take the maximum of the cumulative sum.
+    \item Critical values are obtained by summing bins of a Brownian
+      bridge, where bin sizes match the observed bin sizes of the 
+      ordinal variable.
+  \end{itemize}}
 
-% Picture of Brownian bridge
-% \frame{
-%   \frametitle{Brownian Bridge}
-%   \begin{center}
-%       \includegraphics[height=3in]{brownian.pdf}
-%   \end{center}}
+% Achim: Can you outline ordmax and possibly ordwmax here?
+\frame{
+  \frametitle{Test Statistics}
+}
 
-% \frame{
-%   \frametitle{Theorem}
-%   \begin{itemize}
-%     \item Use of the theorem:
-%       \begin{itemize}
-%         \item The theorem gives us a way to construct statistical
-%           tests of measurement invariance using the scores.
-%         \item Various test statistics can be obtained depending on our
-%           goals.
-%         \item Ordering individuals based on an auxiliary variable
-%           allows us to interpret measurement invariance violations in
-%           the context of that variable.
-%       \end{itemize}
-%   \end{itemize}}
-
-% INSERT WICHERTS EXAMPLE HERE
-\section{Illustration}
 \frame{
-  \frametitle{Example}
+  \frametitle{Simulation}
   \begin{itemize}
-    \item Example: Studying stereotype threat via factor analysis (Wicherts et al., 2005)
+    \item Simulation:
       \begin{itemize}
-        \item Stereotype threat: Knowledge of stereotypes about one's social group might cause one to fulfill the stereotypes.
-        \item Wicherts et al.\ study: 295 students were administered three intelligence tests.  Stereotypes were primed for half of the students.
-        \item Groups defined by: Ethnicity (majority/minority) and whether or not stereotypes were primed.
+        \item Same two-factor model from before.
+        \item Measurement invariance violation in three unique
+          variance parameters, growing as we move up the ordinal
+          variable.
+        \item Sample size in $\{120, 480, 960\}$.
+        \item Levels of the ordinal variable in $\{4, 8, 12\}$.
+        \item Five test statistics.
       \end{itemize}
@@
 \frame{
-  \frametitle{Model}
-  \begin{itemize}
-    \item To study the data, Wicherts et al.\ employed a series of four-group, one-factor models.
-      \begin{itemize}
-        \item General finding: Minorities with stereotype primes have different measurement parameters than other groups.
-        \item Current example: Is measurement further impacted by academic performance (as measured by student GPA)?
-      \end{itemize}
-  \end{itemize}}
-
-\frame{
-  \frametitle{Model}
-  \begin{itemize}
-    \item We utilize a model employed by Wicherts et al., where four model parameters are specific to the ``minority, stereotype prime'' group.
-      \begin{itemize}
-        \item Test for measurement invariance in these parameters wrt the student GPA variable (either all four together or individually).
-        \item Violations of measurement invariance imply that stereotype threat is more problematic for students of low or high GPA.
-      \end{itemize}
-  \end{itemize}}
-
-\frame{
-  \frametitle{Model}
-  %\begin{center}
-      \includegraphics[height=2.3in]{../psychoco2012/wicherts_msp.pdf}
-  %\end{center}
+  \frametitle{Simulation}
+  % Graph of results
 }
 
-\frame{
-  \frametitle{Results for Single Parameters}
-  \begin{center}
-      \includegraphics[height=3in]{../psychoco2012/gefp_4_ind.pdf}
-  \end{center}}
-
-\frame{
-  \frametitle{Aggregated Results}
-  \begin{center}
-      \includegraphics[height=3in]{../psychoco2012/gefp_4_agg.pdf}
-  \end{center}}
+% Wicherts example removed due to time constraints (about 15min)
 
@@
       \begin{itemize}
-        \item Isolating specific parameters that violate
-          measurement invariance, allowing the researcher to interpret the nature of measurement invariance violations.
-        \item Isolating
-          groups of individuals whose parameter values differ.
-        \item Studying the impact of continuous variables on model estimates, without ``ruining'' the rest of the model.
+        \item Testing for measurement invariance along a continuum,
+          defining subgroups based on the test.
+        \item Testing for measurement invariance wrt an ordinal
+          variable, with sensitivity to ``ordinal'' violations.
+        \item Testing for measurement invariance at low $n$, where
+          multiple-group models are not feasible.
       \end{itemize}
-    \item Power is reasonable, with specific tests being better in
-      specific circumstances.
   \end{itemize}}
@@
         \item \texttt{lavaan} for model estimation.
-        \item \texttt{estfun()} for score extraction, which is currently a combination of our own code and \texttt{lavaan} code.
-        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
-          % \begin{itemize}
-          %   \item Required input: Fitted model, function for score extraction, and information matrix (optional).
-          %   \item \texttt{gefp()} constructs the process.
-          %   \item \texttt{sctest()} and \texttt{plot()} calculate and visualize test statistics.
-          % \end{itemize}
+        \item \texttt{estfun()} for score extraction, which is
+          currently a combination of our own code and \texttt{lavaan}
+          code.
+        \item \texttt{strucchange} for carrying out the proposed tests
+          with the scores from \texttt{estfun()}.
       \end{itemize}
@@
   \begin{itemize}
-    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
     \item Detailed examination of test properties.\\ \ \\
     \item Extension to related psychometric issues.\\ \ \\
+    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
     \item Paper: Merkle, E.\ C.\ \& Zeileis, A.\ (accepted).  Tests of measurement invariance without subgroups:\ {A} generalization of classical methods.  {\em Psychometrika}.

### 2012-07-03 imps2012/invariance.tex
some small touch-ups, changed definition of efp to match the Psychometrika paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@144 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\title[Measurement Invariance]{Tests for Measurement Invariance when Subgroups are Unknown}
+\title[Measurement Invariance]{Tests for Measurement Invariance When Subgroups Are Unknown}
 \author{Ed Merkle\inst{1} \and Achim Zeileis\inst{2}}
@@
 
-\section{Tests Along a Continuum}
+\section{Tests along a Continuum}
 % \frame{
@@
       along a continuous auxiliary variable.
-    \item Tests are based on individual terms ({\em scores}) of the gradient.  The scores tell us how well a particular parameter describes a particular individual.
+    \item Tests are based on individual terms of the gradient ({\em scores}).  The scores tell us how well a particular parameter describes a particular individual.
   % Equation of gradient vs scores for individuals
@@
         \item Order individuals by the auxiliary variable.\\ \ \\
-        \item Define $t \in (1/n, n)$.
+        \item Define $t \in (0, 1)$.
           The {\em empirical cumulative score process} is defined by:\\
   \begin{equation*}
-    \bm{B}(\hat{\bm{\theta}}; t) = \frac{1}{\sqrt{n}}
+    \bm{B}(\hat{\bm{\theta}}; t) = 
+      {\bm{\widehat I}}^{-1/2} \frac{1}{\sqrt{n}}
     \displaystyle\sum_{i=1}^{\lfloor nt 
@@
     \end{equation*}
-  where $\lfloor nt \rfloor$ is the integer part of $nt$.
+  where $\lfloor nt \rfloor$ is the integer part of $nt$
+  and $\bm{\widehat I}$ is some consistent covariance matrix estimate, e.g.,
+  the observed information matrix ${\bm{I}}(\widehat{{\bm{\theta}}})$.
+
       \end{itemize}
@@
   \begin{equation*}
-  {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) 
-\overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
+  {\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) \overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
   \end{equation*}
- where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
- information matrix and ${\bm B}^{0}(\cdot)$ is a
- $p$-dimensional Brownian bridge.\\ \ \\
+ where ${\bm B}^{0}(\cdot)$ is a $p$-dimensional Brownian bridge.\\ \ \\
 
@@
 
-\section{Tests Along an Ordinal Variable}
+\section{Tests along an Ordinal Variable}
 \frame{
-  \frametitle{Tests Along Ordinal Variables}
+  \frametitle{Tests along Ordinal Variables}
   \begin{itemize}
@@
       to ordinal auxiliary variables, where there is no unique ordering of
-      individuals.
+      all individuals.
     \item To obtain a test statistic, we allow all individuals with the

### 2012-07-05 imps2012/invariance.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@145 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor changes to IMPS slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@146 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
added tests for complete ordering
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@147 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Added ordinal simulation results to imps slides.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@148 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   \begin{equation*}
-    \bm{B}(\hat{\bm{\theta}}; t) = 
+    \bm{B}(t; \hat{\bm{\theta}}) = 
       {\bm{\widehat I}}^{-1/2} \frac{1}{\sqrt{n}}
@@
   \begin{equation*}
-  {\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) \overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
+  {\bm{B}}(\cdot; \widehat{{\bm{\theta}}}) \overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
   \end{equation*}
@@
   \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Tests}
+
+\begin{itemize}
+  \item Test statistics aggregate over parameters $j = 1, \dots, k$ and observations $i = 1, \dots, n$,
+    employing different norms.
+  \item Special cases:
+    \begin{eqnarray*}
+      \mathit{DM}      & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
+      \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
+      \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
+                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
+    \end{eqnarray*}
+  \item ${\bm B}(\hat {\bm \theta})_{ij}$ is short for $\bm{B}(i/n; \hat{\bm{\theta}})_j$.
+\end{itemize}
+
 \end{frame}
@@
 
-% Achim: Can you outline ordmax and possibly ordwmax here?
-\frame{
-  \frametitle{Test Statistics}
-}
-
 \frame{
@@
         \item Levels of the ordinal variable in $\{4, 8, 12\}$.
-        \item Five test statistics.
+        \item Three test statistics (two new statistics, plus the
+          multiple group LRT).
       \end{itemize}
@@
   % Graph of results
+  \begin{center}
+      \includegraphics[height=3in]{../ordinal/ordsim-xyplot.pdf}
+  \end{center}
 }
 
-% Wicherts example removed due to time constraints (about 15min)
+% library("lattice")
+% load("simres2.rda")
+% res$nlevels <- factor(paste("nlev=",res$nlevels,sep=""),
+%   levels=c("nlev=4","nlev=8","nlev=12"))
+% res$nobs <- paste("n=",res$nobs,sep="")
+
+% pdf("ordsim-xyplot.pdf")
+% trellis.par.set(theme = canonical.theme(color = FALSE))
+% mykey <- simpleKey(c("LRT", "ordmaxLM", "orddwmax"), points = FALSE, lines = TRUE)
+% mykey$lines$lty <- c(2, 1, 3)
+% mykey$lines$lwd <- c(1.2, 1.2, 2)
+% print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data=
+% res[res$test %in% c("lrt","ordmax","ordwmax"),], type="l", xlab="Violation Magnitude", ylab="Power",key=mykey, ylim=c(0,1), lty=c(3,2,1), lwd=c(1.5,1.3,1)))
+
+% dev.off()
 
@@
     \item Extension to related psychometric issues.\\ \ \\
-    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
+    %\item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
     \item Paper: Merkle, E.\ C.\ \& Zeileis, A.\ (accepted).  Tests of measurement invariance without subgroups:\ {A} generalization of classical methods.  {\em Psychometrika}.

### 2012-07-07 imps2012/invariance.tex
added ordinal test statistics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@149 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \frame{
+  \frametitle{Tests along Ordinal Variables}
+  \begin{itemize}
+    \item Instead of aggregating over all $i = 1, \dots, n$, aggregate
+      only over $i_\ell = \lfloor n \cdot t_\ell\rfloor$ for the cumulative proportions
+      $t_\ell$ ($\ell = 1, \dots, m$).
+    \item Tests: Obtain ordinal versions of the max LM test and (weighted) double maximum test.
+      Additionally use LM-type test that does not exploit ordering.
+    \begin{eqnarray*}
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}      
+                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~
+                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,\\
+      \mathit{LM}_\mathit{uo} & = & \sum_{\ell = 1, \dots, m+1} \sum_{j = 1, \dots, k}
+        \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2.
+    \end{eqnarray*}
+  \end{itemize}}
+
+\frame{
   \frametitle{Simulation}

### 2012-07-10 ordinal/ordinal.Rnw
Added frame of an ordinal manuscript.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@151 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}

\title{Testing for measurement invariance with respect to an ordinal variable}
\threeauthors{Edgar C. Merkle}{Achim Zeileis}{Jinyan Fan}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}{Auburn University}

\abstract{Researchers are often interested in testing for 
  measurement invariance with respect to an ordinal auxiliary variable such as
  age group, income class, or school grade.  In a factor-analytic
  context, these tests are traditionally carried out via a likelihood
  ratio statistic that compares a model where parameters differ across groups
  to a model where parameters are equal across groups.  This test
  neglects the fact that the auxiliary variable is ordinal, and it
  is also known to be overly sensitive at large sample sizes.  In this
  paper, we propose a new measurement invariance test statistic that
  explicitly accounts for the ordinality of the auxiliary variable.
  The statistic is derived from a family of tests based on stochastic
  processes that have recently received attention in the psychometric
  literature.  The statistic is illustrated via an application
  involving real data, and its performance is studied via simulation.
}

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. 
  Correspondence to Edgar C.\ Merkle, Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Ordinal measurement invariance}
\rightheader{Ordinal measurement invariance}

\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

%% for internal use
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}

\spacing{1}

\begin{document}
\maketitle

<<preliminaries>>=
## packages
library("OpenMx")
library("lavaan")
library("strucchange")
library("mvtnorm")
library("lattice")

## auxiliary code
source("mz.R")
source("estfun-lavaan.R")
source("sim.R")

## convenience function for plotting boundaries without color
get_boundary <- function(obj, fun, trim = 0) {
  bound <- fun$boundary(0:obj$nobs/obj$nobs)
  bound <- fun$computeCritval(0.05, ncol(obj$process)) * bound
  bound <- zoo(bound, time(obj$process))
  if(trim > 0) bound <- head(tail(bound, - floor(trim * obj$nobs)), - floor(trim * obj$nobs))
  return(bound)
}
@

The test statistic proposed here is derived from a family of tests
that were recently applied to the study of measurement invariance in
psychometric models \cite{MerZei12}.  % TODO Add kopf
In the following section, we provide an overview of the family and
describe the proposed statistic in detail.  Next, we report on the
results of two simulation studies designed to compare the proposed
test statistic to existing statistics for testing measurement
invariance.  Next, we illustrate the proposed statistic using
psychometric data on scales purported to measure youth gratitude.  
Finally, we provide some detail on the test's use in practice.

\section{Theoretical Detail}

It is of interest to test a hypothesis of 
measurement invariance with respect to subsets of model parameters.
The hypothesis can be written as
%% TODO equation
which reflects invariance in the full set of model parameters
$\theta$.  It is straightforward to test a subset of model parameters,
and we do so later in the paper.  In this section, we first provide
background on the family of tests from which the proposed test
statistic is obtained.  We then describe the proposed statistic for
testing measurement invariance with respect to an ordinal variable.

\subsection{Background}

The ordinal test statistic proposed here, along with
related ``non-ordinal'' statistics reported elsewhere, rely on the
casewise contributions to the model likelihood function (i.e., the
{\em scores}).  These contributions are defined as
%% TODO equation
the sum of which equals $\bm{0}$ when maximum likelihood estimates
$\hat{\bm{\theta}}$ are substituted for $\bm{\theta}$.  
Under the hypothesis of measurement invariance, the
individual scores are expected to fluctuate around zero.  Conversely,
when measurement invariance is violated, the scores should
systematically stray from zero for some individuals.
%% TODO add citations:
As described in Merkle and Zeileis (\citeyearNP{MerZei12}; see XX for
further detail), we can create formal test statistics based on these
ideas.  These statistics rely on ordering of individuals with
respect to an auxiliary variable, then taking a cumulative sum of the
scores:
%% TODO insert cumsum
Under %eqref
terms in this cumulative sum combine to follow a Brownian bridge.
Thus, we can summarize the empirical cumulative scores in various
manners, obtaining critical values and $p$-values from the Brownian
bridge.

%% Paragraph on test statistics
Specific test statistics that can be obtained under this framework
include

\subsection{An Ordinal Test Statistic}

The above framework was originally designed for situations where
the auxiliary variable is continuous, so that there is a unique
ordering of individuals with respect to the auxiliary variable.  In
situations where the auxiliary variable is categorical, it is possible
to obtain a test statistic by first summing scores within each level of the
auxiliary variable, then summing the sums to obtain a test statistic.
%% TODO make sure the above is true.
However, these sums discard the ordinal nature of the auxiliary
variable.  A similar issue is observed in testing for measurement
invariance via multiple groups models and Likelihood Ratio tests: we
can allow $\bm{\theta}$ to be unique to each level of the ordinal
variable, but the ordinality of the auxiliary variable is then lost.
In contrast, the proposed statistic explicitly accounts for the fact
that the auxiliary variable is ordinal.

The statistic that we proposed here is defined as
%% TODO insert equation
so that all individuals with the same value of the ordinal variable
simultaneously enter into the cumulative sum.  The test statistic is
then the maximum of this cumulative sum.  Closed-form solutions for
this particular functional of a Brownian bridge are unavailable, but
critical values and $p$-values can be obtained through repeated
simulation of a Brownian bridge.  This functionality is built in to
R's strucchange package \cite{strucchange}, which we use to carry out
the tests for general SEMs.

As we demonstrate in the simulations below, the proposed test
statistic is sensitive to the measurement invariance violations that
an analyst would typically expect from an ordinal auxiliary variable.
These include violations that increase or decrease as we move up
levels of the auxiliary variable, as well as violations that abruptly
begin at a single point on the auxiliary variable and continue after
that.  The test statistic is insensitive to anomolies in a small
number of categories of the auxiliary variable that are
unrelated to its ordering.  This is especially relevant to situations
where the analyst has a large sample size.  In these cases, the usual
Likelihood Ratio test is notoriously sensitive to minute parameter
instabilities.

\section{Simulation 1: Detecting Ordinal Invariance Violations}

In this simulation, we demonstrate that the proposed test statistic is
sensitive to ordinal measurement invariance violations, moreso than
other competing statistics.

\section{Simulation 2: Minor Anomolies and Large $n$}

In this simulation, we demonstrate that the proposed statistic is
relatively insensitive to minor parameter violations that are
unrelated to the ordering of the auxiliary variable.  As noted
earlier, this feature is especially applicable to situations where
one's sample size is very large.  Analysts often resort to informal
fit measures in this case, because the traditional LRT is nearly
guaranteed to result in significance.  This simulation is intended to
show that the proposed ordinal test remains viable for very large $n$.

\section{Application: Youth Gratitude}

\section{General Discussion}

%% application to other psychometric models

\subsection{Summary}

\section*{Computational Details}

All results were obtained using the {R}~system for statistical computing \cite{R11},
version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on package
{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} for fitting of the factor analysis models and
{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
for evaluating the parameter instability tests. {R}~and the packages {lavaan} and {strucchange} are
freely available under the General Public License~2 from the
Comprehensive {R} Archive Network at \url{http://CRAN.R-project.org/}.
{R}~code for replication of our results is
available at \url{http://semtools.R-Forge.R-project.org/}.

\bibliography{refs}

\end{document}

### 2012-07-10 ordinal/ordinal.Rnw
Additions to ordinal paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@152 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   context, these tests are traditionally carried out via a likelihood
-  ratio statistic that compares a model where parameters differ across groups
+  ratio test statistic comparing a model where parameters differ across groups
   to a model where parameters are equal across groups.  This test
@@
   is also known to be overly sensitive at large sample sizes.  In this
-  paper, we propose a new measurement invariance test statistic that
-  explicitly accounts for the ordinality of the auxiliary variable.
-  The statistic is derived from a family of tests based on stochastic
+  paper, we propose test statistics that
+  explicitly account for the ordinality of the auxiliary variable.
+  The statistics are derived from a family of tests based on stochastic
   processes that have recently received attention in the psychometric
-  literature.  The statistic is illustrated via an application
-  involving real data, and its performance is studied via simulation.
+  literature.  The statistics are illustrated via an application
+  involving real data, and their performance is studied via simulation.
 }
@@
 that were recently applied to the study of measurement invariance in
-psychometric models \cite{MerZei12}.  % TODO Add kopf
+psychometric models \cite{KopZei10,MerZei12}.
 In the following section, we provide an overview of the family and
-describe the proposed statistic in detail.  Next, we report on the
+describe the proposed statistics in detail.  Next, we report on the
 results of two simulation studies designed to compare the proposed
-test statistic to existing statistics for testing measurement
-invariance.  Next, we illustrate the proposed statistic using
+test statistics to existing tests of measurement
+invariance.  Next, we illustrate the proposed statistics using
 psychometric data on scales purported to measure youth gratitude.  
-Finally, we provide some detail on the test's use in practice.
+Finally, we provide some detail on the tests' use in practice.
 
@@
 
-It is of interest to test a hypothesis of 
-measurement invariance with respect to subsets of model parameters.
-The hypothesis can be written as
-%% TODO equation
-which reflects invariance in the full set of model parameters
-$\theta$.  It is straightforward to test a subset of model parameters,
-and we do so later in the paper.  In this section, we first provide
-background on the family of tests from which the proposed test
-statistic is obtained.  We then describe the proposed statistic for
-testing measurement invariance with respect to an ordinal variable.
+We will consider situations in which a $p$-dimensional variable $X$
+with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
+model with density $f(\bm{x}_i; \bm{\theta})$ and
+associated joint log-likelihood
+\begin{equation} \label{eq:loglik}
+  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
+    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
+    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
+\end{equation}
+where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
+the distribution.
 
-\subsection{Background}
+We focus on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises
+from a structural equation model with assumed multivariate normality,
+though the proposed tests extend beyond this family of models.  Under
+the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
+$\bm{\theta}$ can 
+be estimated by maximum likelihood (ML), i.e.,
+\begin{equation} \label{eq:ml}
+  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \ell(\bm{\theta}; x_1, \dots, x_n),
+\end{equation}
+or equivalently by solving the first order conditions
+\begin{equation}
+    \label{eq:ml1}
+  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
+\end{equation}
+where 
+\begin{equation}
+  \label{eq:score}
+  {\bm s}({\bm \theta}; x_i) ~=~ \left(
+    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_1},
+    \dots,
+    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_k}
+  \right)^\top,
+\end{equation}
+is the score function of the model (the partial derivative of the casewise
+likelihood contributions w.r.t.\ the parameters $\bm{\theta}$).
+Evaluation of the score function at $\hat{\bm{\theta}}$ for $i=1,
+\dots, n$ essentially measures the extent to which the model maximizes
+each individual's likelihood: as an individual's scores go further
+from zero, the model provides a poorer description of that individual.
 
-The ordinal test statistic proposed here, along with
-related ``non-ordinal'' statistics reported elsewhere, rely on the
-casewise contributions to the model likelihood function (i.e., the
-{\em scores}).  These contributions are defined as
-%% TODO equation
-the sum of which equals $\bm{0}$ when maximum likelihood estimates
-$\hat{\bm{\theta}}$ are substituted for $\bm{\theta}$.  
-Under the hypothesis of measurement invariance, the
-individual scores are expected to fluctuate around zero.  Conversely,
-when measurement invariance is violated, the scores should
-systematically stray from zero for some individuals.
-%% TODO add citations:
+Measurement invariance tests essentially test the assumption that all
+individuals arise from the same 
+parameter vector ${\bm \theta}$.
+Thus, a hypothesis of invariance can be written as
+\begin{equation}
+    \label{eq:h0}
+    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
+\end{equation}
+where ${\bm \theta}_i$ reflects the parameter vector for individual
+$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  This
+hypothesis is typically tested against an alternative
+\begin{equation}
+  \label{eq:h1*}
+  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
+    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
+    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
+  \end{array} \right.
+\end{equation}
+where $v_i$ reflects individual $i$'s value on an auxiliary variable
+$V$ and $\nu$ is a threshold dividing individuals into groups based on
+$V$.  In this paper, we generally focus on situations where $V$ is
+ordinal and 
+where $\nu$ is unknown.  In the section below, we review tests where
+$V$ is continuous (and $\nu$ is unknown) as background for our
+proposed tests for ordinal $V$.
+
+\subsection{Tests for Continuous $V$}
+% stopped here
+
 As described in Merkle and Zeileis (\citeyearNP{MerZei12}; see XX for
@@
 
-\subsection{An Ordinal Test Statistic}
+\subsection{Proposed Tests for Ordinal $V$}
 
@@
 
-The statistic that we proposed here is defined as
+In contrast to the likelihood ratio test, the statistics that we
+propose here explicitly take into 
+account the fact that the auxiliary variable is ordinal.  They are
+similar to the statistics from % eqref
+above, except that we focus on ``bins'' of individuals at each level of the
+ordinal variable:
 %% TODO insert equation
-so that all individuals with the same value of the ordinal variable
-simultaneously enter into the cumulative sum.  The test statistic is
-then the maximum of this cumulative sum.  Closed-form solutions for
+Critical values associated with these test statistics can be obtained
+by applying these functionals to bins of a Brownian bridge, where the
+bin sizes equal the empirical proportion of individuals at each level
+of the ordinal variable.
+Closed-form solutions for
 this particular functional of a Brownian bridge are unavailable, but
@@
 simulation of a Brownian bridge.  This functionality is built in to
-R's strucchange package \cite{strucchange}, which we use to carry out
+R's strucchange package \cite{strucchange}, which can be used to carry out
 the tests for general SEMs.
@@
 
-In this simulation, we demonstrate that the proposed test statistic is
+In this simulation, we demonstrate that the proposed test statistics are
 sensitive to ordinal measurement invariance violations, moreso than
-other competing statistics.
+other competing statistics.  We generate data from a two-factor,
+six-indicator model, with a measurement invariance violation occurring
+in the three unique variances associated with the first factor.  We
+use the proposed test statistics to test for measurement invariance
+simultaneously across all six unique variances, mimicking the first
+step of the ``prescribed'' course of measurement invariance tests. % cite
+
+For comparison, we also compute two statistics that treat the ordinal
+auxiliary variable as categorical.  These include the likelihood ratio
+test of measurement 
+invariance in the six unique variance parameters, along with .
+
+\subsection{Method}
+As described above, data were generated from a two-factor model
+lacking measurement invariance in three of six unique variance
+parameters.  Magnitude of measurement invariance violation, sample
+size, and number of categories of the ordinal variable were manipulated.
+We examined three sample sizes ($n=120, 480, 960$), three numbers of
+categories ($l=4, 8, 12$), and seven magnitudes of invariance
+violations.  These violations started at level $1+l/2$ and continued
+for higher levels of the ordinal variable.  The unique variances for
+the ``violating'' levels deviated from the lower levels' unique
+variances by $d$~times the parameters' asymptotic standard
+errors (scaled by $\sqrt{n}$), with $d=0, 0.25, 0.5, \dots, 1.5$.
+
+For each combination of $n \times l \times d$, 5,000 datasets were
+generated and tested via the XX statistics described above.  In all
+conditions, we maintained equal sample sizes at each level of the
+ordinal variable.
+
+%% sweave code here
+
+\subsection{Results}
 
@@
 
+\subsection{Method}
+
+\subsection{Results}
+
 \section{Application: Youth Gratitude}
+
+\subsection{Background}
+
+\subsection{Method}
+
+\subsection{Results}
+
+\subsection{Summary}
 
@@
 %% application to other psychometric models
+
+%% should we care about non-ordinal violations when using an ordinal variable?
 

### 2012-07-26 ordinal/ordinal.Rnw
Completed a draft of the intro to the ordinal paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@153 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \title{Testing for measurement invariance with respect to an ordinal variable}
-\threeauthors{Edgar C. Merkle}{Achim Zeileis}{Jinyan Fan}
-\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}{Auburn University}
+\threeauthors{Edgar C. Merkle}{Jinyan Fan}{Achim Zeileis}
+\threeaffiliations{University of Missouri}{Auburn University}{Universit\"{a}t Innsbruck}
 
@@
 ## packages
-library("OpenMx")
 library("lavaan")
@@
 ## auxiliary code
-source("mz.R")
-source("estfun-lavaan.R")
-source("sim.R")
+source("mzo.R")
+source("../www/estfun-lavaan.R")
+source("../prelim_code/efpFunctional-cat.R")
+source("sim_ordinal.R")
 
@@
 
-The test statistic proposed here is derived from a family of tests
+%% TODO add some more citations to these initial paragraphs
+The study of measurement invariance and differential item functioning
+(DIF) 
+has received considerable attention in the psychometric literature
+(see, e.g., \citeNP{Mill11} for a thorough review).  A set of
+psychometric scales ${\bm X}$ are defined to be measurement invariant
+with respect to an auxiliary variable $V$ if \cite{Mel89}
+\begin{equation}
+    \label{eq:midef}
+      f(x_i | t_i, v_i, \dots) = f({\bf x}_i | t_i, \dots),
+\end{equation}
+where $T$ is the latent variable that the scales measure, $f$ is the
+model's distributional form, and the $i$ subscript refers to
+individual cases.  If the above equation does not hold, then there
+exists a measurement invariance violation.  We focus here on
+situations where $f()$ is parametric, and the measurement invariance
+violation occurs because the model parameters are
+unequal across individuals (and related to $V$).
+
+As a concrete example of the study of measurement invariance, consider
+a situation where ${\bm X}$ contains some ``high stakes'' tests of
+ability and $V$ is ethnicity.
+One's ethnicity should be unrelated to the measurement parameters
+within $f()$, and this expectation can be studied within a
+psychometric model by fitting the model examining whether or not
+measurement parameters 
+vary across different ethnicities.  Statistical tools that can be used
+to carry out this study include Likelihood Ratio Tests, Lagrange
+Multiplier tests, and Wald tests \cite{Sat89}.
+These tools have greatly aided in
+the development of improved, ``fairer'' psychometric tests and
+scales.
+
+Along with ethnicity, researchers are often interested in studying
+measurement invariance with respect to ordinal $V$.
+Such variables can arise from multiple choice surveys, where
+continuous variables such as age or income are binned into a small number
+of categories.  Alternatively, the variables may arise from gross,
+qualitative assessments of a particular measure of interest, where
+individuals may be categorized as having a
+``low,'' ``medium,'' or ``high'' level of the variable of interest.
+While these variables are relatively easy to find in the literature,
+there exist very few measurement invariance methods that specifically
+account for the fact that $V$ is ordinal.  More often, $V$ is simply
+treated as categorical so that the traditional tests can be applied.
+It is the intent of this paper to propose
+two test statistics that fill this gap and to show that the statistics
+possess good properties to be used in practice.
+
+The test statistics proposed here are derived from a family of tests
 that were recently applied to the study of measurement invariance in
@@
 \dots, n$ essentially measures the extent to which the model maximizes
-each individual's likelihood: as an individual's scores go further
+each individual's likelihood: as an individual's scores stray further
 from zero, the model provides a poorer description of that individual.
 
-Measurement invariance tests essentially test the assumption that all
+In studying measurement invariance, we are essentially testing the
+assumption that all  
 individuals arise from the same 
@@
 where $\nu$ is unknown.  In the section below, we review tests where
-$V$ is continuous (and $\nu$ is unknown) as background for our
+$V$ is continuous (and $\nu$ is unknown) before proceeding to the 
 proposed tests for ordinal $V$.
@@
 \subsection{Tests for Continuous $V$}
-% stopped here
 
-As described in Merkle and Zeileis (\citeyearNP{MerZei12}; see XX for
-further detail), we can create formal test statistics based on these
-ideas.  These statistics rely on ordering of individuals with
-respect to an auxiliary variable, then taking a cumulative sum of the
-scores:
-%% TODO insert cumsum
-Under %eqref
-terms in this cumulative sum combine to follow a Brownian bridge.
-Thus, we can summarize the empirical cumulative scores in various
-manners, obtaining critical values and $p$-values from the Brownian
+When $V$ is categorical with a relatively-small number of categories,
+tests of measurement invariance typically proceed via multiple-group
+models.  In this situation, we use likelihood ratio tests to compare a
+model whose parameters differ across groups to a model whose
+parameters are constrained to be equal across groups.  When $V$ is
+continuous, however, multiple-group models obviously cannot be used 
+because there are no existing groups.  Instead, we can fit a model
+whose parameters are restricted to be equal across all individuals, then
+examine individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
+fluctuate with their values of $V$.  If measurement invariance holds
+with respect to $V$, then the scores should randomly fluctuate around
+zero.  Conversely, if measurement invariance does not hold, then the
+scores should systematically depart from zero.  These ideas are
+related to those underlying the Lagrange multiplier test; see
+\citeNP{MerZei12} for more details.
+
+To formalize these ideas, we define the $k$-dimensional {\em
+  cumulative score process} as 
+\begin{equation} \label{eq:cumscore}
+  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
+    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_i)
+  \qquad (0 \le t \le 1)
+\end{equation}
+where $\lfloor nt \rfloor$ is the integer part of $nt$ and
+$\hat I$ is some consistent estimate of the covariance matrix of the
+scores.  Equation~\eqref{eq:cumscore} simultaneously accounts for the
+ordering of individuals w.r.t.\ $V$ and decorrelates the scores
+associated with each of the $k$ model parameters (which allows us to
+potentially make inferences separately for each individual model
+parameter).  Using ideas similar to those that were outlined in the previous
+paragraph, the cumulative score process associated with each
+model parameter should randomly fluctuate around zero under
+measurement invariance.  Further, there exists a functional central
+limit theorem that allows us to make formal inference with this
+cumulative score process.  Assuming that individuals are independent
+and the usual ML regularity conditions hold, it is possible to show
+that \citeA{HjoKon02}
+\begin{equation} \label{eq:fclt}
+  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
+\end{equation}
+where $\overset{d}{\rightarrow}$ denotes convergence in distribution
+and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
+Thus, we can construct tests of measurement invariance by comparing
+the cumulative score process's behavior to that of a Brownian bridge.
+This is accomplished by comparing a scalar statistic associated with
+the cumulative score process to the analogous statistic of a Brownian
 bridge.
 
+In practice, we have a finite sample size $n$ and so the
+{\em empirical} cumulative score can be represented within an $n
+\times k$ matrix with elements ${\bm B}(i/n; \hat {\bm \theta})_j$ that
+we also denote ${\bm B}(\hat {\bm \theta})_{ij}$ below for brevity.
+Each row of the matrix contains cumulative sums of the scores of
+individuals who were at the $i/n$ percentile of $V$ or below.  Scalar
+test statistics are then obtained by collapsing over rows
+(individuals) and columns (parameters) of the matrix, with asymptotic
+distributions of the test statistics under~\eqref{eq:h0}  being
+obtained by applying the 
+same functional to the Brownian bridge \cite{HjoKon02,ZeiHor07}.
+
 %% Paragraph on test statistics
-Specific test statistics that can be obtained under this framework
-include
+Specific test statistics commonly obtained under this framework
+include the double maximum statistic
+\begin{equation}
+    \label{eq:dmax}
+    \mathit{DM} = \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,
+\end{equation}
+which essentially tests whether any component of the cumulative score
+process strays too far from zero.  This test discards information
+related to multiple parameters fluctuating simultaneously, and Merkle
+and Zeileis \citeyear{MerZei12} found it to have relatively-low power
+for simultaneously testing multiple factor analysis parameters for
+measurement invariance.
+
+Test statistics with higher power
+aggregate information across parameters and possibly also across
+individuals.  These test statistics include
+\begin{eqnarray}
+    \label{eq:cvm}
+    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
+    \label{eq:maxlm}
+    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
+      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
+\end{eqnarray}
+with the former being a Cram{\'e}r-von Mises statistic and the latter
+corresponding to a ``maximum'' Lagrange multiplier test, where the
+maximum is taken across all possible divisions of individuals into two
+groups w.r.t.\ $V$.  Additionally, the $\max \mathit{LM}$ statistic is
+scaled by the asymptotic variance $t (1 - t)$ of the process ${\bm
+  B}(t, \hat {\bm \theta})$.  In simulations, Merkle and Zeileis found that the
+$\mathit{CvM}$ test exhibited the highest power for testing multiple
+factor analysis parameters for measurement invariance w.r.t.\
+continuous $V$.  These simulations included situations in which
+subsets of model parameters were tested; such situations are handled
+by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
+that correspond to the parameters of interest.
 
@@
 
-The above framework was originally designed for situations where
-the auxiliary variable is continuous, so that there is a unique
-ordering of individuals with respect to the auxiliary variable.  In
-situations where the auxiliary variable is categorical, it is possible
-to obtain a test statistic by first summing scores within each level of the
-auxiliary variable, then summing the sums to obtain a test statistic.
-%% TODO make sure the above is true.
-However, these sums discard the ordinal nature of the auxiliary
+The theory described above was designed for situations where
+$V$ is continuous, so that there is a unique
+ordering of individuals with respect to $V$.  In
+situations where $V$ is categorical, it is possible
+to obtain a test statistic by first summing scores within each of the
+$m$ levels of the
+auxiliary variable, then ``summing the sums'' to obtain a test
+statistic \cite{HjoKon02}.  This test statistic can be formally
+written as
+\begin{equation}
+    \label{eq:lmuo}
+     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
+\end{equation}
+where, again, tests of subsets of model parameters can be obtained by
+taking the inner sum over only the $k^* < k$ parameters of interest.
+This test statistic discards the ordinal nature of the auxiliary
 variable.  A similar issue is observed in testing for measurement
-invariance via multiple groups models and Likelihood Ratio tests: we
-can allow $\bm{\theta}$ to be unique to each level of the ordinal
-variable, but the ordinality of the auxiliary variable is then lost.
-In contrast, the proposed statistic explicitly accounts for the fact
-that the auxiliary variable is ordinal.
+invariance via multiple groups models and Likelihood Ratio tests (or,
+equivalently, via Wald tests or Lagrange multiplier tests): we
+can allow $\bm{\theta}$ to be unique at each level of the ordinal
+variable, but the ordinality of the auxiliary variable is lost.
+In contrast, the statistics proposed below explicitly account for the fact
+that $V$ is ordinal.
 
-In contrast to the likelihood ratio test, the statistics that we
-propose here explicitly take into 
-account the fact that the auxiliary variable is ordinal.  They are
-similar to the statistics from % eqref
+The statistics proposed here are
+similar to those described in
+Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}
 above, except that we focus on ``bins'' of individuals at each level of the
-ordinal variable:
+ordinal variable.  That is, instead of aggregating over all $i = 1,
+\dots, n$ individuals, we first compute cumulative proportions
+$t_\ell$ ($\ell = 1, \dots, m-1$) associated with the first $m-1$
+levels of $V$.  We then aggregate the 
+cumulative scores only over $i_\ell = \lfloor n
+\cdot t_\ell\rfloor$.  Test statistics related to~\eqref{eq:dmax}
+and~\eqref{eq:maxlm} above can then be written as
 %% TODO insert equation
+\begin{eqnarray}
+    \label{eq:wdm}
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}      
+                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
+    \label{eq:maxlmo}
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~
+                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.\\
+\end{eqnarray}
 Critical values associated with these test statistics can be obtained
-by applying these functionals to bins of a Brownian bridge, where the
-bin sizes equal the empirical proportion of individuals at each level
-of the ordinal variable.
+by applying the same functionals to bins of a Brownian bridge, where the
+bin sizes result in the cumulative proportions $t_\ell$ ($\ell=1, \dots,
+m-1$) associated with the observed $V$.
 Closed-form solutions for
-this particular functional of a Brownian bridge are unavailable, but
+these particular functionals of a Brownian bridge are unavailable, but
 critical values and $p$-values can be obtained through repeated
 simulation of a Brownian bridge.  This functionality is built in to
-R's strucchange package \cite{strucchange}, which can be used to carry out
-the tests for general SEMs.
+R's strucchange package \cite{strucchange}, which can be used to
+generally carry out the tests.
 
-As we demonstrate in the simulations below, the proposed test
-statistic is sensitive to the measurement invariance violations that
-an analyst would typically expect from an ordinal auxiliary variable.
+As we demonstrate in the simulations below, the test
+statistics proposed above are sensitive to the measurement invariance
+violations that 
+an analyst would typically expect from an ordinal $V$.
 These include violations that increase or decrease as we move up
-levels of the auxiliary variable, as well as violations that abruptly
-begin at a single point on the auxiliary variable and continue after
-that.  The test statistic is insensitive to anomolies in a small
-number of categories of the auxiliary variable that are
-unrelated to its ordering.  This is especially relevant to situations
-where the analyst has a large sample size.  In these cases, the usual
-Likelihood Ratio test is notoriously sensitive to minute parameter
-instabilities.
+levels of $V$, as well as violations that abruptly
+begin at a single point on $V$ and continue after
+that.  The test statistics are insensitive to anomolies in a small
+number of categories of $V$ that are
+unrelated to the ordering of $V$.  This is especially relevant to situations
+where the analyst has a large sample size, where the usual
+Likelihood Ratio Test is notoriously sensitive to minute parameter
+instabilities (e.g., \citeNP{Ben80}).
 
@@
 sensitive to ordinal measurement invariance violations, moreso than
-other competing statistics.  We generate data from a two-factor,
+traditional statistics.  We generate data from a two-factor,
 six-indicator model, with a measurement invariance violation occurring
@@
 test of measurement 
-invariance in the six unique variance parameters, along with .
+invariance in the six unique variance parameters, along with the
+unordered LM test from~\eqref{eq:lmuo}.
 
@@
 lacking measurement invariance in three of six unique variance
-parameters.  Magnitude of measurement invariance violation, sample
+parameters.  Magnitude of measurement invariance violation, type of
+measurement invariance violation, sample
 size, and number of categories of the ordinal variable were manipulated.
 We examined three sample sizes ($n=120, 480, 960$), three numbers of
-categories ($l=4, 8, 12$), and seven magnitudes of invariance
-violations.  These violations started at level $1+l/2$ and continued
-for higher levels of the ordinal variable.  The unique variances for
+categories ($m=4, 8, 12$), and
+seven magnitudes of invariance 
+violations.  The measurement invariance violations began
+at level $1+l/2$ of $V$ and were constant thereafter
+  The unique variances for
 the ``violating'' levels deviated from the lower levels' unique
@@
 
-For each combination of $n \times l \times d$, 5,000 datasets were
+For each combination of $n \times m \times d$, 5,000 datasets were
 generated and tested via the XX statistics described above.  In all

### 2012-07-27 ordinal/ordinal.Rnw
Additions to ordinal manuscript: o Modified dgp of simulation 1 to include 'constant' instability. o Started to add R code to the Sweave document. o Worked on data example for the manuscript.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@154 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 violations.  The measurement invariance violations began
-at level $1+l/2$ of $V$ and were constant thereafter
+at level $1+l/2$ of $V$ and were constant thereafter.
   The unique variances for
@@
 For each combination of $n \times m \times d$, 5,000 datasets were
-generated and tested via the XX statistics described above.  In all
+generated and tested via the 4 statistics described above.  In all
 conditions, we maintained equal sample sizes at each level of the
@@
 
-%% sweave code here
+<<sim1>>=
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim1.rda")) {
+  load("sim1.rda")
+} else {
+  sim1 <- simulation()
+  save(sim1, file="sim1.rda")
+}
+@
 
@@
 
+Froh et al.\ \citeyear{FroFan11} used confirmatory factor analysis
+models to study the invariance of three youth gratitude scales across
+students aged 10 to 19 years.  Each of the scales were individually
+factor-analyzed using the items comprising the scale.
+For each model, the authors 
+first fit a congeneric model (all parameters free for each age group),
+followed by a tau-equivalent model (factor loadings restricted to be
+equal across groups) and a parallel model (all parameters restricted
+to be equal across groups).  Because their sample size was large ($n
+\approx 1400$), they could not rely solely on likelihood ratio tests
+(i.e., chi-square difference tests) for model comparison because the
+tests were overly sensitive at their sample size.  To supplement these
+tests, Froh et al examined a set of alternative fit indices, including
+the non-normed fit index, the comparative fit index, and the
+incremental fit index. %% TODO cite
+The authors generally found support for the tau-equivalent models
+through these alternative fit indices; the likelihood ratio test often
+resulted in significance event when the alternative indices indicated
+good fit.
+
+In the analyses described below, we re-analyze the Froh et al.\ data using
+the ordinal test statistics proposed in this paper.  This results in a
+series of tests that are less sensitive to sample size than the
+likelihood ratio test, while being
+more sensitive to ``ordinal'' violations of measurement invariance.
+We focus on two analyses from Froh et al.\ where the likelihood
+ratio test resulted in significance (indicating that the restricted
+model did not fit as well as the less-restricted model), but the
+alternative fit measures indicated the opposite (using the rules of
+thumb employed by Froh et al).  These include
+comparison of a one-factor congeneric model to a one-factor
+tau-equivalent model using the 
+GQ6 and comparison of a one-factor tau-equivalent model to a one-factor
+parallel model 
+using the GAC.  To conduct equivalent analyses via the proposed tests, we
+fit the restricted model and test for instability in the focal
+model parameters.
+
+Of the 1401 cases originally collected by Froh et al., we use here all
+subjects with complete data (resulting in $n=1327$).
+
+<<frohdata>>=
+yg <- read.spss("grat data combined_9.29.09.sav", to.data.frame=TRUE)
+names(yg) <- tolower(names(yg))
+
+## Remove cases with 'imputed' values
+yg <- yg[yg$ao_4 <= 9,]
+decimals <- apply(yg[,4:28]%%1 > 0, 1, sum)
+yg <- yg[decimals==0,]
+
+## TODO Simulate critical values for ordL2BB
+@
+
 \subsection{Results}
+
+The results section is divided into two subsections, one for each of
+the analyses described previously.
+
+<<gq6>>=
+m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+          data=yg, group="agegroup", group.equal="loadings",
+          meanstructure=TRUE)
+gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
+              vcov = info_full, sandwich = FALSE, parm=1:4)
+omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
+ol2bb1 <- sctest(gefp1, functional=ordL2BB(gefp1))
+
+## Full model
+m1.f <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+            data=yg, group="agegroup", meanstructure=TRUE)
+lrt1 <- anova(m1.f, m1)
+@
+
+\paragraph{GQ6.} Ordinal tests do not support LRT.
+
+<<gac>>=
+m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
+           f1 ~ 0*1', data=yg, group="agegroup",
+           group.equal=c("loadings","residuals","lv.variances"), 
+           meanstructure=TRUE)
+gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
+              vcov = info_full, sandwich = FALSE, parm=3:6)
+omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
+ol2bb2 <- sctest(gefp2, functional=ordL2BB(gefp2))
+
+## Full model
+m2.f <- cfa('f1 =~ gac_1 + gac_2 + gac_3
+           f1 ~ 0*1', data=yg, group="agegroup",
+           group.equal="loadings", meanstructure=TRUE)
+lrt2 <- anova(m2.f, m2)
+@
+
+\paragraph{GAC.} Ordinal tests do support LRT.
 

### 2012-07-27 ordinal/sim_ordinal.R

@@ -1,10 +1,13 @@
 
 ## Data-generating process
-dgp <- function(nobs = 200, diff = 3, nlevels=10)
+dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE)
 {
   # Generates data from a factor analysis model that violates
   # measurement invariance.  Also generates an ordinal auxiliary
   # variable related to the invariance (loosely called "age").
+  # If gradual==TRUE, mild instability starts halfway up age
+  # and gets larger.  If gradual==FALSE, the instability starts
+  # halfway up age and stays constant.
   stopifnot(require("mvtnorm"))
   
   # ses is SE increase in parameters for each increasing value
@@ -61,8 +64,14 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10)
     ## observed SES by sqrt(n)
     ## FIXME: Should ses refer to maximum difference in 'end' category, or to se increase
     ##        with each category (currently coded as difference in end category)?
-    diag(psi) <- diag(psi) + (each.vary*c(62.1,26.6,82.1,9.05,19.35,51.59))/sqrt(tmp.n)
-    u <- t(rmvnorm(tmp.n,rep(0,6),psi))
+    if (gradual){
+      diag(psi) <- diag(psi) + (each.vary*c(62.1,26.6,82.1,9.05,19.35,51.59))/sqrt(tmp.n)
+      tmp.psi <- psi
+    } else {
+      tmp.psi <- psi
+      diag(tmp.psi) <- diag(tmp.psi) + (ses*c(62.1,26.6,82.1,9.05,19.35,51.59))/sqrt(tmp.n)
+    }
+    u <- t(rmvnorm(tmp.n,rep(0,6),tmp.psi))
     for (j in 1:tmp.n){
       datmat[tmp.ind[j],] <- mu + lambda%*%z[,j] + u[,j]
     }
@@ -89,8 +98,8 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, test = NULL, verb
       pval[i, 1] <- sctest(ord_gefp,  functional = ordfun)$p.value
       pval[i, 2] <- sctest(ord_gefp,  functional = ordwmax(ord_gefp))$p.value
       pval[i, 3] <- sctest(ord_gefp,  functional = catL2BB(ord_gefp))$p.value
-      pval[i, 4] <- sctest(ord_gefp,  functional = supLM(0.1))$p.value
-      pval[i, 5] <- mz$lrt.p
[…]

### 2012-07-31 ordinal/ordinal.Rnw
Work on ordinal paper: o Wrote up results of simulation 1. o Modified simulation code for simulation 2.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@155 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 As described above, data were generated from a two-factor model
-lacking measurement invariance in three of six unique variance
-parameters.  Magnitude of measurement invariance violation, type of
-measurement invariance violation, sample
+lacking measurement invariance in the six unique variance
+parameters.  Magnitude of measurement invariance violation, 
+sample
 size, and number of categories of the ordinal variable were manipulated.
@@
 violations.  The measurement invariance violations began
-at level $1+l/2$ of $V$ and were constant thereafter.
+at level $1+m/2$ of $V$ and were constant thereafter.
   The unique variances for
@@
 
+Simulation results are presented in Figure XX.  Rows of the figure
+correspond to $n$, columns of the figure correspond to $m$, the x-axis
+of each panel corresponds to $d$, and the y-axis of each panel
+corresponds to power.  It is seen that one of the proposed test
+statistics, the $\max \mathit{LM}_O$ statistic from~\eqref{eq:maxlmo},
+generally has the largest power to detect the ordinal measurement
+invariance violations.  The other three tests are considerably closer
+in power, with the second proposed ordinal statistic
+(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
+at large violation magnitudes.  This is because the double-max
+test discards information about multiple parameters changing together
+at specific levels of the ordinal variable
+(see Merkle and Zeileis, 2012, for related discussion), while the
+three other tests under consideration make use of this information.
+Finally, it is seen that, in the small $n$ and large $m$ conditions,
+the likelihood ratio test exhibits large Type-I error rates (i.e.,
+power greater than .05 at $d=0$).  This is because the likelihood
+ratio test requires 
+estimation of a multiple groups model, which is very unstable with
+large numbers of groups and small sample sizes.  The proposed
+statistics, which require estimation of only the single-group model,
+exhibit a clear advantage in these conditions.
+
+\begin{figure}
+\caption{Simulated power curves for the ordered and unordered
+      $\max \mathit{LM}$ tests, the ordered double-max test, and the
+      likelihood ratio test across three sample sizes $n$, three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
+\label{fig:sim1res}
+\setkeys{Gin}{width=\textwidth}
+<<sim1res, fig=TRUE, height=7, width=7>>=
+mykey <- simpleKey(levels(sim1$test), points = FALSE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+\end{figure}
+
+To summarize, we found the $\max \mathit{LM}_O$ statistic to be
+advantageous for detecting measurement invariance violations that are
+related to an ordinal auxiliary variable.  In particular, power is
+generally higher, and the test does not require estimation of a
+multiple group model.  Thus, the statistic allows reasonable
+measurement invariance tests to be carried out at small $n$/large $m$
+combinations.  To further illustrate that the proposed statistics are
+useful for testing violations related to an ordinal variable, we
+now compare their performance to the likelihood ratio test at large
+$n$ and small $d$.
+
 \section{Simulation 2: Minor Anomolies and Large $n$}
@@
 \subsection{Method}
+
+<<sim2>>=
+if(file.exists("sim2.rda")) {
+  load("sim2.rda")
+} else {
+  sim2 <- simulation(diff = seq(0, 0.5, by = 0.1), nobs = c(1200, 4800, 9600),
+                     anomaly=TRUE)
+  save(sim2, file="sim2.rda")
+}
+@
 

### 2012-07-31 ordinal/sim_ordinal.R

@@ -1,19 +1,22 @@
 
 ## Data-generating process
-dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE)
+dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE)
 {
-  # Generates data from a factor analysis model that violates
-  # measurement invariance.  Also generates an ordinal auxiliary
-  # variable related to the invariance (loosely called "age").
-  # If gradual==TRUE, mild instability starts halfway up age
-  # and gets larger.  If gradual==FALSE, the instability starts
-  # halfway up age and stays constant.
+  ## Generates data from a factor analysis model that violates
+  ## measurement invariance.  Also generates an ordinal auxiliary
+  ## variable related to the invariance (loosely called "age").
+
+  ## If gradual==TRUE, mild instability starts halfway up age
+  ## and gets larger.  If gradual==FALSE, the instability starts
+  ## halfway up age and stays constant.
+  ## If anomaly==TRUE, instability only occurs at level (1+nlevels/2).
+  ## Otherwise, anomaly continues up the ordinal variable.
   stopifnot(require("mvtnorm"))
   
-  # ses is SE increase in parameters for each increasing value
-  # of "age" past the halfway point.
-  sampsize <- nobs
+  # ses is SE increase in parameters at last ordinal category.
+  # (This is constant if gradual==FALSE but not if gradual==TRUE.)
   ses <- diff
+  sampsize <- nobs
   sampsize.per.nobs <- nobs %/% nlevels
 
   ## Deal with cases where nobs/nlevels is not an integer
@@ -60,10 +63,10 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE)
     tmp.ind <- which(age==i)
     z <- t(rmvnorm(tmp.n,rep(0,2),phi))
 
+    if (anomaly) ifelse(i==(half.level+1), ses, 1)
+    
     ## SES found by fitting model to population S, multiplying
     ## observed SES by sqrt(n)
-    ## FIXME: Should ses refer to maximum difference in 'end' category, or to se increase
[…]

### 2012-08-07 ordinal/sim_ordinal.R
Minor bugs fixed in dgp.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@156 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -56,25 +56,26 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE)
   }
 
   ## Now do the same for invariant-violating levels
-  n.vary <- nlevels - (half.level + 1)
+  n.vary <- nlevels - half.level
   each.vary <- ses/n.vary
   for (i in (half.level+1):nlevels){
     tmp.n <- sum(age==i)
     tmp.ind <- which(age==i)
     z <- t(rmvnorm(tmp.n,rep(0,2),phi))
 
-    if (anomaly) ifelse(i==(half.level+1), ses, 1)
+    mult <- ses
+    if (anomaly) mult <- ifelse(i==(half.level+1), ses, 0)
     
     ## SES found by fitting model to population S, multiplying
     ## observed SES by sqrt(n)
     if (gradual){
+      if (anomaly) stop("gradual=TRUE and anomaly=TRUE do not work together.")
       diag(psi) <- diag(psi) + (each.vary*c(62.1,26.6,82.1,9.05,19.35,51.59))/sqrt(tmp.n)
       tmp.psi <- psi
     } else {
       tmp.psi <- psi
-      diag(tmp.psi) <- diag(tmp.psi) + (ses*c(62.1,26.6,82.1,9.05,19.35,51.59))/sqrt(tmp.n)
+      diag(tmp.psi) <- diag(tmp.psi) + (mult*c(62.1,26.6,82.1,9.05,19.35,51.59))/sqrt(tmp.n)
     }
-
 
     u <- t(rmvnorm(tmp.n,rep(0,6),tmp.psi))
     for (j in 1:tmp.n){

### 2012-08-09 ordinal/ordinal.Rnw
Added Simulation 2 method, results to ordinal paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@157 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 library("lattice")
+library("foreign")
 
@@
 }
+
+## convenience function for applying tests to real data
+info_full <- function(x, ...) solve(vcov(x) * nobs(x))
 @
@@
   sim1 <- simulation()
+  sim1$nlevels <- factor(sim1$nlevels)
+  levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
+  sim1$nobs <- factor(sim1$nobs)
+  levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
+  levels(sim1$test) <- c("LM_uo", "LRT", "LM_o", "WDM_o")
   save(sim1, file="sim1.rda")
@@
 
-Simulation results are presented in Figure XX.  Rows of the figure
+Simulation results are presented in Figure~\ref{fig:sim1res}.  Rows of
+the figure 
 correspond to $n$, columns of the figure correspond to $m$, the x-axis
@@
 guaranteed to result in significance.  This simulation is intended to
-show that the proposed ordinal test remains viable for very large $n$.
+show that the proposed ordinal test remains viable for large $n$.
 
 \subsection{Method}
+Data were generated from the same factor analysis model that was used
+in Simulation 1, with measurement invariance violations in the unique
+variance parameters.  To implement a minor measurement invariance
+violation, the unique variances were equal across all levels of the
+ordinal variable except one (level $1+m/2$).  At this particular
+level, the unique variances were greater by a factor of $d$~times the
+parameters' asymptotic standard errors (scaled by $\sqrt{n}$), with
+$d=0, 0.5, 1.0, \dots, 3.0$.  The number of levels of the ordinal
+variable were the same as those in Simulation 1 ($m=4,8,12$), and
+sample sizes were set at $n=1200,4800,9600$.  All other simulation
+features matches those from Simulation 1.
 
@@
 } else {
-  sim2 <- simulation(diff = seq(0, 0.5, by = 0.1), nobs = c(1200, 4800, 9600),
+  sim2 <- simulation(diff = seq(0, 3, by = 0.5), nobs = c(1200, 4800, 9600),
                      anomaly=TRUE)
+  sim2$nlevels <- factor(sim2$nlevels)
+  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
+  sim2$nobs <- factor(sim2$nobs)
+  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
+  levels(sim2$test) <- c("LM_uo", "LRT", "LM_o", "WDM_o")
   save(sim2, file="sim2.rda")
@@
 \subsection{Results}
+Simulation results are presented in Figure~\ref{fig:sim2res}.  It is
+observed that results are very consistent across the sample sizes
+tested, implying that "practical infinity" is reached for this model
+by $n=1200$. \readme{Also implying that we don't need all three sample
+  sizes or could try values smaller than 1200.}  We also observe
+a negative 
+relationship between power and $m$; this is because the measurement
+invariance violation occurred at only one level of the auxiliary
+variable.  As $m$ increases (and $n$ is held
+constant), the number of individuals violating measurement invariance
+therefore decreases.  As a result, power to detect the violation decreases with
+increasing $m$.
+
+The more interesting result of this simulation lies in the comparison
+of the four test statistics within each panel of the figure.  The
+two ``unordered'' test statistics both have relatively-high power to
+detect the measurement invariance violation, illustrating the finding
+of Bentler and Bonett \citeyear{BenBon80} and others that the
+likelihood ratio test statistic picks out minor 
+parameter discrepancies at large $n$.  In contrast, the two ordinal
+test statistics that we proposed have considerably lower power, with
+the $\text{WDM}_O$ statistic being the lowest and the $\text{LM}_O$
+statistic being higher at larger values of $d$.  
+This demonstrates that the proposed ordinal test
+statistics can be especially useful at large sample sizes, where
+traditional test statistics result in frequent significance.
+Both statistics exhibited much lower power to detect a measurement
+invariance violation occurring at a single level of the auxiliary
+variable.
+
+\begin{figure}
+\caption{Simulated power curves for the ordered and unordered
+      $\max \mathit{LM}$ tests, the ordered double-max test, and the
+      likelihood ratio test across three sample sizes $n$, three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
+      occurring at a single level (the $(1 +
+      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
+\label{fig:sim2res}
+\setkeys{Gin}{width=\textwidth}
+<<sim2res, fig=TRUE, height=7, width=7>>=
+mykey <- simpleKey(levels(sim2$test), points = FALSE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+\end{figure}
+
+Taken together, the results from Simulation 1 and Simulation
+2 provide evidence that the $\text{LM}_O$ statistic should be
+preferred to the $\text{WDM}_O$ statistic for testing measurement
+invariance.  The $\text{LM}_O$ statistic has higher power to detect
+ordinal violations, and its power to detect non-ordinal violations was
+similar to that of $\text{WDM}_O$ when the violation magnitude was
+small (e.g., for $d \leq 1.5$).  The $\text{LM}_O$ statistic is
+advantageous because it can make use of invariance violations that
+simultaneously 
+occur in multiple parameters, whereas the $\text{WDM}_O$ focuses only
+on the parameter with the largest invariance violation.  
+Thus, the statistics are likely to exhibit
+similar performance if only a single model parameter violated
+measurement invariance.
+
+In the next section, we compare the proposed statistics to the
+likelihood ratio test with real data.
 
@@
               vcov = info_full, sandwich = FALSE, parm=1:4)
+if(file.exists("ol2bb1.rda")){
+  load("ol2bb1.rda")
+} else {
+  ol2bb1.fun <- ordL2BB(gefp1)
+  save(ol2bb1.fun, file="ol2bb1.rda")
+}
 omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
-ol2bb1 <- sctest(gefp1, functional=ordL2BB(gefp1))
+ol2bb1 <- sctest(gefp1, functional=ol2bb1.fun)
 
@@
               vcov = info_full, sandwich = FALSE, parm=3:6)
+if(file.exists("ol2bb2.rda")){
+  load("ol2bb2.rda")
+} else {
+  ol2bb2.fun <- ordL2BB(gefp2)
+  save(ol2bb2.fun, file="ol2bb2.rda")
+}
 omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
-ol2bb2 <- sctest(gefp2, functional=ordL2BB(gefp2))
+ol2bb2 <- sctest(gefp2, functional=ol2bb2.fun)
 
@@
 
-%% application to other psychometric models
-
-%% should we care about non-ordinal violations when using an ordinal variable?
+\begin{enumerate}
+  \item The only proposed statistics that treat ordinal auxiliary
+    variables as truly ordinal?  LRT treats them as categorical,
+    inclusion in the model typically treats tham as continuous.
+  \item The first point is important because, if we test wrt an
+    ordinal auxiliary variable, we are generally looking for ordinal
+    violations.  We don't care if a middle level differs from
+    all others because we would not usually have an intuitive
+    explanation for it.
+  \item The proposed statistics should generally transfer to IRT and
+    other psychometric models.
+\end{enumerate}
 

### 2012-08-10 ordinal/ordinal.Rnw
Additions to manuscript: o Added application section using youth gratitude data. o Expanded outline for general discussion. o General revisions to paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@158 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 has received considerable attention in the psychometric literature
-(see, e.g., \citeNP{Mill11} for a thorough review).  A set of
+(see, e.g., \citeNP{Mil11} for a thorough review).  A set of
 psychometric scales ${\bm X}$ are defined to be measurement invariant
@@
 within $f()$, and this expectation can be studied within a
-psychometric model by fitting the model examining whether or not
+psychometric model by fitting the model and examining whether or not
 measurement parameters 
@@
 to carry out this study include Likelihood Ratio Tests, Lagrange
-Multiplier tests, and Wald tests \cite{Sat89}.
+Multiplier tests, and Wald tests (e.g., \citeNP{Sat89}).
 These tools have greatly aided in
@@
 \section{Theoretical Detail}
+
+This section contains background on the theory underlying the
+proposed statistics; for a more detailed
+account, see Merkle and Zeileis \citeyear{MerZei12}.
 
@@
 $i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  This
-hypothesis is typically tested against an alternative
+hypothesis is often tested against an alternative
 \begin{equation}
@@
 $V$ and $\nu$ is a threshold dividing individuals into groups based on
-$V$.  In this paper, we generally focus on situations where $V$ is
-ordinal and 
-where $\nu$ is unknown.  In the section below, we review tests where
+$V$.  Note that this alternative hypothesis can also be used when $V$
+is ordinal, testing whether the model parameters shift at some point
+on $V$.  
+Alternatively, if $V$ is categorical with a relatively-small number of
+groups, a frequent alternative of interest is
+\begin{equation}
+  \label{eq:h2*}
+  H_2^*:~ {\bm \theta}_i = {\bm \theta}_{v_i}.
+\end{equation}
+In this paper, we generally focus on situations where $V$ is
+ordinal and where the measurement invariance violation is related to the ordinal
+variable (e.g., the violation is of the type from~\eqref{eq:h1*} or
+the violation grows/shrinks with $V$).  Researchers typically test for
+measurement invariance w.r.t.\ ordinal $V$ by employing the alternative
+from~\eqref{eq:h2*}, which implicitly treats $V$ as categorical.
+Thus, test statistics that explicitly treat $V$ as ordinal should have
+higher power to detect the ``ordinal'' types of violations described above.
+
+In the section below, we review tests where
 $V$ is continuous (and $\nu$ is unknown) before proceeding to the 
@@
 whose parameters are restricted to be equal across all individuals, then
-examine individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
+examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
 fluctuate with their values of $V$.  If measurement invariance holds
@@
 and the usual ML regularity conditions hold, it is possible to show
-that \citeA{HjoKon02}
+that \cite{HjoKon02}
 \begin{equation} \label{eq:fclt}
@@
 
-Test statistics with higher power
+Test statistics that exhibit better performance in such situations 
 aggregate information across parameters and possibly also across
@@
 This test statistic discards the ordinal nature of the auxiliary
-variable.  A similar issue is observed in testing for measurement
+variable, essentially employing the alternative hypothesis
+from~\eqref{eq:h2*}.  A similar issue is observed in testing for
+measurement 
 invariance via multiple groups models and Likelihood Ratio tests (or,
@@
 similar to those described in
-Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}
+Equations~\eqref{eq:dmax} and~\eqref{eq:maxlm}
 above, except that we focus on ``bins'' of individuals at each level of the
@@
                              \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.\\
+                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
 \end{eqnarray}
@@
 simulation of a Brownian bridge.  This functionality is built in to
-R's strucchange package \cite{strucchange}, which can be used to
+R's strucchange package \cite{ZeiLei02}, which can be used to
 generally carry out the tests.
 
-As we demonstrate in the simulations below, the test
+As demonstrated in the simulations below, the test
 statistics proposed above are sensitive to the measurement invariance
@@
 detect the measurement invariance violation, illustrating the finding
-of Bentler and Bonett \citeyear{BenBon80} and others that the
+of Bentler and Bonett \citeyear{Ben80} and others that the
 likelihood ratio test statistic picks out minor 
@@
 
+\readme{Fan: Can you provide a short
+  background on the study of youth gratitude and the specific intent
+  of Froh et al?}
+
 \subsection{Method}
@@
 models to study the invariance of three youth gratitude scales across
-students aged 10 to 19 years.  Each of the scales were individually
-factor-analyzed using the items comprising the scale.
+students aged 10 to 19 years.  Due to sample size constraints, the age
+variable included six categories: 10-11 years, 12-13 years, 14 years,
+15 years, 16 years, and 17-19 years.  Thus, age is an ordinal variable
+with which the proposed tests can be applied.
+
+To test for measurement invariance w.r.t.\ age, each of the three scales
+was individually 
+factor-analyzed using the items that comprised the scale.
 For each model, the authors 
-first fit a congeneric model (all parameters free for each age group),
+first fit a congeneric model (all parameters free for each level of age),
 followed by a tau-equivalent model (factor loadings restricted to be
-equal across groups) and a parallel model (all parameters restricted
-to be equal across groups).  Because their sample size was large ($n
+equal across each level of age) and a parallel model (all parameters restricted
+to be equal across levels of age).  Because their sample size was large ($n
 \approx 1400$), they could not rely solely on likelihood ratio tests
@@
 the ordinal test statistics proposed in this paper.  This results in a
-series of tests that are less sensitive to sample size than the
-likelihood ratio test, while being
-more sensitive to ``ordinal'' violations of measurement invariance.
+series of tests that are less sensitive than the likelihood ratio test
+to minor parameter discrepancies,
+while being more sensitive to ordinal violations of measurement invariance.
 We focus on two analyses from Froh et al.\ where the likelihood
@@
 model did not fit as well as the less-restricted model), but the
-alternative fit measures indicated the opposite (using the rules of
-thumb employed by Froh et al).  These include
+alternative fit measures employed by Froh et al.\ indicated the opposite.
+These include
 comparison of a one-factor congeneric model to a one-factor
@@
 using the GAC.  To conduct equivalent analyses via the proposed tests, we
-fit the restricted model and test for instability in the focal
-model parameters.
+fit the more-restricted model in each case and test for instability in
+the focal model parameters.
 
@@
 The results section is divided into two subsections, one for each of
-the analyses described previously.
+the analyses described previously.  The first section contains an
+example of the ordinal statistics disagreeing with the likelihood
+ratio tests, while the second section contains the opposite.
 
@@
 
-\paragraph{GQ6.} Ordinal tests do not support LRT.
+\subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
+model to the GQ-6 data, Froh et al.\ used alternative fit indices to
+conclude that the tau-equivalent model was as good as the congeneric
+model.  However, the likelihood ratio test comparing these two models
+was significant ($\chi^2_{20} = \Sexpr{round(lrt1[[5]][2],1)}, p=\Sexpr{round(lrt1[[7]][2],2)}$ for the data considered
+here).
+
+We can use the proposed ordinal statistics to compare the
+tau-equivalent and congeneric models,
+fitting the tau-equivalent model and testing whether or
+not the factor loadings fluctuate with respect to age group.  In doing
+so, we obtain~$\text{WDM}_O=\Sexpr{round(omax1$statistic,1)}, p=\Sexpr{round(omax1$p.value,2)}$
+ and~$\text{LM}_O=\Sexpr{round(ol2bb1$statistic,1)}, p=\Sexpr{round(ol2bb1$p.value,2)}$.
+Neither statistic is significant at an $\alpha$ of .05,
+which supports the conclusions that Froh et al.\ obtained from
+alternative fit statistics.  This provides further evidence that the
+likelihood ratio statistic is overly sensitive for this particular
+analysis.
 
@@
 
-\paragraph{GAC.} Ordinal tests do support LRT.
+\subsubsection{GAC} In fitting tau-equivalent and parallel models to the
+GAC data, Froh et al.\ obtained mixed results.  The alternative fit
+indices did not all agree with one another, and the likelihood ratio test
+indicated that the parallel model fit worse than the tau-equivalent
+model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2],1)}, p<.01$
+for the data considered here).  Froh et al.\ ultimately 
+concluded that the tau-equivalent model provided a better fit than did
+the parallel model.
+
+To apply the ordinal statistics proposed in this paper, we fit the
+parallel model and test for instability in the variance parameters
+(unique variance and factor variance) w.r.t.\ age.
+This results in~$\text{WDM}_O=\Sexpr{round(omax2$statistic,1)}, p<.01$
+and~$\text{LM}_O=\Sexpr{round(ol2bb2$statistic,1)}, p<.01$.
+Both of these statistics agree with the likelihood ratio test and with
+the general conclusion that the tau-equivalent model is better than
+the parallel model.  This provides further evidence that the
+significant likelihood ratio test is not simply an artifact of the
+large sample size.
 
 \subsection{Summary}
+The application considered above shows that the ordinal test
+statistics can provide useful information in situations where one
+might question significant likelihood ratio test statistics.  While
+researchers use rules of thumb to obtain decisions from other
+alternative fit measures, the proposed statistics are proper tests of
+the hypothesis of interest.  They can be used to either supplement or
+replace the likelihood ratio test, depending upon the measurement
+invariance violations in which the researcher is interested a priori.
+We further describe the issue of supplementing vs.\ replacing the
+likelihood ratio test in the general discussion.
 
@@
 \begin{enumerate}
-  \item The only proposed statistics that treat ordinal auxiliary
-    variables as truly ordinal?  LRT treats them as categorical,
+  \item To my knowledge, these are the only proposed statistics that
+    treat ordinal auxiliary 
+    variables as truly ordinal.  LRT treats them as categorical,
     inclusion in the model typically treats tham as continuous.
-  \item The first point is important because, if we test wrt an
+  \item The first point is important because, if we test w.r.t.\ an
     ordinal auxiliary variable, we are generally looking for ordinal
@@
     explanation for it.
+  \item If we know our sample size is large and have an ordinal
+    auxiliary variable, should we even use the LRT in
+    the first place?  If the measurement invariance violation is truly
+    ordinal, the proposed statistics should be better.  If it is not
+    ordinal, the violation will be difficult to explain as anything
+    but an anomaly.  The stakes might also play a role here.  In
+    high-stakes testing, you might be more interested in treating the
+    ordinal variable as categorical to detect any possible
+    instability, even those that don't make sense.
   \item The proposed statistics should generally transfer to IRT and

### 2012-08-14 ordinal/ordinal.Rnw
Added background to application section.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@159 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\readme{Fan: Can you provide a short
-  background on the study of youth gratitude and the specific intent
-  of Froh et al?}
+With the positive psychology movement, the construct of gratitude has
+received much research attention (for a review, see \citeNP{EmmMcc04}).
+Recently, scholars have begun to explore gratitude
+in youth. However, one potential problem with this research is that
+scholars, with no exception, have used adult gratitude inventories to
+measure youth gratitude, thus raising the question of whether the
+existing gratitude scales used with adults are valid in research with
+youth. Addressing this issue, Froh et al.\ \citeyear{FroFan11} had a
+large sample of 
+youth ($n$ = 1401, ranging from late childhood (10 years old) to
+late adolescent (19 years old)) complete the three most widely used
+adult gratitude inventories, including Gratitude Questionnaire – 6
+(GQ-6; \citeNP{MccEmm02}), Gratitude Adjective
+Checklist (GAC; \citeNP{MccEmm02}), and Gratitude, Resentment,
+Appreciation Test-Short Form (GRAT-Short Form;
+\citeNP{ThoWat03}). The authors were interested in whether the
+youth factor structure 
+for the gratitude scales resembles that of 
+adults, and whether the gratitude scales are invariant across the
+youth age groups.
+
+% references
+% Emmons, R. A., & McCullough, M. E. (Eds.). (2004). The psychology of gratitude. New York, NY: Oxford University Press. 
+% McCullough, M.
+E., Emmons, R.
+A., & Tsang, J.-A.
+(2002).
+The grateful disposition: A conceptual and empirical topography.
+Journal of Personality and Social Psychology, 82, 112-127.
+doi: 10.1037/0022-3514.82.1.112
+% Thomas, M., & Watkins, P.
+(2003, April).
+Measuring the grateful trait: Development of the revised GRAT.
+Poster session presented at the Annual Convention of the Western Psychological Association, Vancouver, British Columbia, Canada.
 

### 2012-08-16 ordinal/ordinal.Rnw
started to go through manuscript (up to p.5), more changes coming up
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@160 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 unequal across individuals (and related to $V$).
+\readme{Maybe we should explicitly mention that in parametric settings, f is
+the probability density function of x? Also, do we need to point out
+the relation between X and x, V and v, T and t?}
 
@@
 vary across different ethnicities.  Statistical tools that can be used
-to carry out this study include Likelihood Ratio Tests, Lagrange
+to carry out this study include Likelihood Ratio tests, Lagrange
 Multiplier tests, and Wald tests (e.g., \citeNP{Sat89}).
@@
 In the following section, we provide an overview of the family and
-describe the proposed statistics in detail.  Next, we report on the
+describe the proposed statistics in detail.  Subsequently, we report on the
 results of two simulation studies designed to compare the proposed
 test statistics to existing tests of measurement
-invariance.  Next, we illustrate the proposed statistics using
+invariance.  Moreover, we illustrate the proposed statistics using
 psychometric data on scales purported to measure youth gratitude.  
@@
 
+\readme{My feeling is that we still need to explain better what type
+of patterns are covered by which alternative. Maybe we should do it
+the other way round and start with $H_2^*$ as the most general type
+of deviation from $H_0$. Then, we could explain that these also
+covers changes that are not `monotonic' in V, i.e., where parameter
+values change in one direction as V increases and then change into
+another direction again. After that we could continue and argue that
+such monotonic changes are of particular interest for categorical V
+and that the simplest form of such a change is $H_1^*$. And as a final
+note one could add that this one-shift alternative is also often
+employed in practice where practitioners often use the median V
+as the cutoff whereas we want to assess all possible cutoffs.}
+
 In the section below, we review tests where
@@
 because there are no existing groups.  Instead, we can fit a model
-whose parameters are restricted to be equal across all individuals, then
+whose parameters are restricted to be equal across all individuals and then
 examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
@@
 scores should systematically depart from zero.  These ideas are
-related to those underlying the Lagrange multiplier test; see
-\citeNP{MerZei12} for more details.
+related to those underlying the Lagrange multiplier test and are
+discussed by \citeNP{MerZei12} in detail. Here, we review the tests'
+most important properties that are relevant for extending the tests
+to the ordinal case.
 
-To formalize these ideas, we define the $k$-dimensional {\em
-  cumulative score process} as 
+To formalize these ideas, we assume that the observations are ordered
+w.r.t.\ $V$ (so that $v_i \le v_{i + 1}$) and define the $k$-dimensional
+{\em cumulative score process} as 
 \begin{equation} \label{eq:cumscore}
@@
 $\hat I$ is some consistent estimate of the covariance matrix of the
-scores.  Equation~\eqref{eq:cumscore} simultaneously accounts for the
+scores, e.g., based on the information matrix or an outer product of the scores.
+Equation~\eqref{eq:cumscore} simultaneously accounts for the
 ordering of individuals w.r.t.\ $V$ and decorrelates the scores
@@
 
+\readme{I changed/extended the comments about the power of the tests
+and their relative advantages somewhat.}
+
 %% Paragraph on test statistics
@@
 process strays too far from zero.  This test discards information
-related to multiple parameters fluctuating simultaneously, and Merkle
-and Zeileis \citeyear{MerZei12} found it to have relatively-low power
-for simultaneously testing multiple factor analysis parameters for
-measurement invariance.
+related to multiple parameters fluctuating simultaneously, and
+\citeNP{MerZei12} found it to be easy to visualize and
+interpret but to have relatively-low power for assessing measurement
+invariance when multiple factor analysis parameters change
+simultaneously.
 
@@
 groups w.r.t.\ $V$.  Additionally, the $\max \mathit{LM}$ statistic is
-scaled by the asymptotic variance $t (1 - t)$ of the process ${\bm
-  B}(t, \hat {\bm \theta})$.  In simulations, Merkle and Zeileis found that the
-$\mathit{CvM}$ test exhibited the highest power for testing multiple
-factor analysis parameters for measurement invariance w.r.t.\
-continuous $V$.  These simulations included situations in which
-subsets of model parameters were tested; such situations are handled
+scaled by the asymptotic variance $t (1 - t)$ of the process
+${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeNP{MerZei12} found
+that both tests perform well when assessing simultaneous changes
+in multiple factor analysis parameters with some advantages for the
+$\mathit{CvM}$ test in their simulation setup. These simulations included
+situations in which subsets of model parameters were tested; such situations are handled
 by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
@@
 corresponds to power.  It is seen that one of the proposed test
-statistics, the $\max \mathit{LM}_O$ statistic from~\eqref{eq:maxlmo},
+statistics, the $\max \mathit{LM}_o$ statistic from~\eqref{eq:maxlmo},
 generally has the largest power to detect the ordinal measurement
@@
 
-To summarize, we found the $\max \mathit{LM}_O$ statistic to be
+To summarize, we found the $\max \mathit{LM}_o$ statistic to be
 advantageous for detecting measurement invariance violations that are
@@
 test statistics that we proposed have considerably lower power, with
-the $\text{WDM}_O$ statistic being the lowest and the $\text{LM}_O$
+the $\text{WDM}_o$ statistic being the lowest and the $\text{LM}_o$
 statistic being higher at larger values of $d$.  
@@
 Taken together, the results from Simulation 1 and Simulation
-2 provide evidence that the $\text{LM}_O$ statistic should be
-preferred to the $\text{WDM}_O$ statistic for testing measurement
-invariance.  The $\text{LM}_O$ statistic has higher power to detect
+2 provide evidence that the $\text{LM}_o$ statistic should be
+preferred to the $\text{WDM}_o$ statistic for testing measurement
+invariance.  The $\text{LM}_o$ statistic has higher power to detect
 ordinal violations, and its power to detect non-ordinal violations was
-similar to that of $\text{WDM}_O$ when the violation magnitude was
-small (e.g., for $d \leq 1.5$).  The $\text{LM}_O$ statistic is
+similar to that of $\text{WDM}_o$ when the violation magnitude was
+small (e.g., for $d \leq 1.5$).  The $\text{LM}_o$ statistic is
 advantageous because it can make use of invariance violations that
 simultaneously 
-occur in multiple parameters, whereas the $\text{WDM}_O$ focuses only
+occur in multiple parameters, whereas the $\text{WDM}_o$ focuses only
 on the parameter with the largest invariance violation.  
@@
 not the factor loadings fluctuate with respect to age group.  In doing
-so, we obtain~$\text{WDM}_O=\Sexpr{round(omax1$statistic,1)}, p=\Sexpr{round(omax1$p.value,2)}$
- and~$\text{LM}_O=\Sexpr{round(ol2bb1$statistic,1)}, p=\Sexpr{round(ol2bb1$p.value,2)}$.
+so, we obtain~$\text{WDM}_o=\Sexpr{round(omax1$statistic,1)}, p=\Sexpr{round(omax1$p.value,2)}$
+ and~$\text{LM}_o=\Sexpr{round(ol2bb1$statistic,1)}, p=\Sexpr{round(ol2bb1$p.value,2)}$.
 Neither statistic is significant at an $\alpha$ of .05,
@@
 (unique variance and factor variance) w.r.t.\ age.
-This results in~$\text{WDM}_O=\Sexpr{round(omax2$statistic,1)}, p<.01$
-and~$\text{LM}_O=\Sexpr{round(ol2bb2$statistic,1)}, p<.01$.
+This results in~$\text{WDM}_o=\Sexpr{round(omax2$statistic,1)}, p<.01$
+and~$\text{LM}_o=\Sexpr{round(ol2bb2$statistic,1)}, p<.01$.
 Both of these statistics agree with the likelihood ratio test and with

### 2012-08-20 ordinal/ordinal.Rnw
some more comments/refs for ordinal V case
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@161 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 The theory described above was designed for situations where
-$V$ is continuous, so that there is a unique
-ordering of individuals with respect to $V$.  In
-situations where $V$ is categorical, it is possible
-to obtain a test statistic by first summing scores within each of the
-$m$ levels of the
-auxiliary variable, then ``summing the sums'' to obtain a test
-statistic \cite{HjoKon02}.  This test statistic can be formally
-written as
+$V$ is continuous, so that there is a unique ordering of individuals with respect to $V$.
+However, in situations where $V$ is ordinal, there is only a partial ordering
+of all individuals, i.e., observations with the same level of $V$ have no unique
+ordering. (Note that the same also applies if $V$ is metric in nature but is only
+discretely measured leading to many ties.) \readme{Extended the previous sentence somewhat.
+My feeling is that we now should first comment on the ordinal test statistics because these
+directly build on the previous `continuous V' section. Then we can comment later on
+on the categorical case.}
+
+\readme{I have also extended the comments about the nominal/categorical case.}
+Furthermore, if $V$ is only nominal/categorical,
+there is not even a partial ordering, i.e., measurement invariance tests should neither
+exploit the ordering of $V$'s levels nor of the observations within the level.
+In this situation, it is possible to obtain a test statistic by first summing scores within
+each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
+statistic \cite{HjoKon02}.  This test statistic can be formally written as
 \begin{equation}
@@
 taking the inner sum over only the $k^* < k$ parameters of interest.
+\fixme{The $i_\ell$ still need to be introduced. This would happen automatically
+if we move the nominal statistics behind the ordinal statistics.}
 This test statistic discards the ordinal nature of the auxiliary
@@
 measurement 
-invariance via multiple groups models and Likelihood Ratio tests (or,
+invariance via multiple groups models and likelihood ratio tests (or,
 equivalently, via Wald tests or Lagrange multiplier tests): we
@@
 and~\eqref{eq:maxlm} above can then be written as
-%% TODO insert equation
 \begin{eqnarray}
@@
 bin sizes result in the cumulative proportions $t_\ell$ ($\ell=1, \dots,
-m-1$) associated with the observed $V$.
-Closed-form solutions for
-these particular functionals of a Brownian bridge are unavailable, but
+m-1$) associated with the observed $V$. For $\mathit{WDM}_o$ this
+just requires computations of quantiles from a multivariate normal
+distribution with standard normal marginals and a certain correlation
+structure \cite{hotzei08} which can be easily obtained using R's
+mvtnorm package \cite{mvtnorm}. For $\max \mathit{LM}_o$, no
+closed-form solution for this particular functional of a Brownian bridge is unavailable, but
 critical values and $p$-values can be obtained through repeated
-simulation of a Brownian bridge.  This functionality is built in to
-R's strucchange package \cite{ZeiLei02}, which can be used to
+simulation of Brownian bridges.  This functionality is built in to
+R's strucchange package \cite{Zei06}, which can be used to
 generally carry out the tests.

### 2012-08-21 ordinal/ordinal.Rnw
included more technical detail about ordinal statistics and their limiting counterparts
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@162 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
     \label{eq:wdm}
-      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}      
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
                              \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
@@
 bin sizes result in the cumulative proportions $t_\ell$ ($\ell=1, \dots,
-m-1$) associated with the observed $V$. For $\mathit{WDM}_o$ this
-just requires computations of quantiles from a multivariate normal
-distribution with standard normal marginals and a certain correlation
-structure \cite{hotzei08} which can be easily obtained using R's
-mvtnorm package \cite{mvtnorm}. For $\max \mathit{LM}_o$, no
-closed-form solution for this particular functional of a Brownian bridge is unavailable, but
+m-1$) associated with the observed $V$.
+
+\readme{Several details/extensions added to the following three paragraphs.}
+
+For $\mathit{WDM}_o$ the resulting asymptotic distribution is
+$\max_{j = 1, \dots, k} \max_{\ell = 1, \dots, m-1} {\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$.
+Note that the effect outer maximum can be easily captured by a Bonferroni
+correction as the $k$~components of the Brownian bridge are asymptotically independent.
+Moreover, the inner maximum is taken over $m-1$ variables
+${\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$ which are standard normal
+(due to the scaling with the standard deviation of a Brownian bridge) and have
+a simple correlation structure: $\sqrt{s (1 - t)}/\sqrt{t (1 - s)}$ for
+$s \le t$ and both $\in \{t_1, \dots, t_{m-1}\}$. Therefore, critical
+values and $p$-values can be easily computed from a multivariate normal
+distribution with standard normal marginals and this particular correlation
+matrix; see also \citeNP{hotzei08} for more details. In R, this can be accomplished
+using the mvtnorm package \cite{mvtnorm}.
+
+For $\max \mathit{LM}_o$ the resulting asymptotic distribution is
+$\max{\ell = 1, \dots, m-1} || {\bm B}^{0}(t_\ell) ||_2^2 / (t_\ell (1 - t_\ell))$
+for which no simple closed-form solution is available. However,
 critical values and $p$-values can be obtained through repeated
@@
 R's strucchange package \cite{Zei06}, which can be used to
-generally carry out the tests.
+generally carry out the tests. Note that for models with only a single
+parameter to be tested (i.e., $k = 1$) both test statistics are
+equivalent because then $\max \mathit{LM}_o = \mathit{WDM}_o^2$.
 
@@
 statistics proposed above are sensitive to the measurement invariance
-violations that 
-an analyst would typically expect from an ordinal $V$.
-These include violations that increase or decrease as we move up
-levels of $V$, as well as violations that abruptly
-begin at a single point on $V$ and continue after
-that.  The test statistics are insensitive to anomolies in a small
+violations that an analyst would typically expect from an ordinal $V$.
+In particular, due to computing cumulative sums in ${\bm B}(\hat {\bm \theta})$,
+violations that occur as we move along the levels of $V$ can be captured
+wll. This includes abrupt shifts in the parameters $\bm \theta$ at a
+certain level of $V$ as well as smooth increases decreases in the parameters.
+Taking a maximum over the $k$~parameters as in $\mathit{WDM}_o$ will be
+more sensitive to changes that occur only in one out of many parameters
+while $\max \mathit{LM}_o$ will be more sensitive to changes occuring
+in several (or even all of the) parameters simultaneously.
+Moreover, the test statistics are rather insensitive to anomolies in a small
 number of categories of $V$ that are
@@
 where the analyst has a large sample size, where the usual
-Likelihood Ratio Test is notoriously sensitive to minute parameter
+likelihood ratio test is notoriously sensitive to minor parameter
 instabilities (e.g., \citeNP{Ben80}).
@@
 conditions, we maintained equal sample sizes at each level of the
-ordinal variable.
+ordinal variable (i.e., $t_\ell = \ell/m$).
 
@@
 the likelihood ratio test exhibits large Type-I error rates (i.e.,
-power greater than .05 at $d=0$).  This is because the likelihood
-ratio test requires 
-estimation of a multiple groups model, which is very unstable with
-large numbers of groups and small sample sizes.  The proposed
-statistics, which require estimation of only the single-group model,
-exhibit a clear advantage in these conditions.
+power greater than 0.05 at $d=0$).  This is because the likelihood
+ratio test requires estimation of a multiple-groups model, which is very
+unstable with large numbers of groups and small sample sizes as only
+$n/m$ observations are available in each subsample.  The proposed
+statistics are all of the LM-type and just require estimation of the single-group model,
+leading to a clear advantage in these conditions.
 
@@
 observed that results are very consistent across the sample sizes
-tested, implying that "practical infinity" is reached for this model
+tested, implying that ``practical infinity'' is reached for this model
 by $n=1200$. \readme{Also implying that we don't need all three sample
@@
 preferred to the $\text{WDM}_o$ statistic for testing measurement
-invariance.  The $\text{LM}_o$ statistic has higher power to detect
+invariance. \readme{This is not surprising given that the models
+have quite a few parameters and we change many of them. We could 
+try to set up a simulation where only a single out of many parameters
+changes where the WDM should perform better. Also,
+in models that are simpler than most CFA models, the WDM may be more useful.
+I'm not sure how much we want to comment on this or illustrate this.}
+The $\text{LM}_o$ statistic has higher power to detect
 ordinal violations, and its power to detect non-ordinal violations was
@@
 existing gratitude scales used with adults are valid in research with
-youth. Addressing this issue, Froh et al.\ \citeyear{FroFan11} had a
+youth. Addressing this issue, \citeNP{FroFan11} had a
 large sample of 
@@
 late adolescent (19 years old)) complete the three most widely used
-adult gratitude inventories, including Gratitude Questionnaire – 6
+adult gratitude inventories, including Gratitude Questionnaire~6
 (GQ-6; \citeNP{MccEmm02}), Gratitude Adjective
@@
 
-Froh et al.\ \citeyear{FroFan11} used confirmatory factor analysis
+\citeNP{FroFan11} used confirmatory factor analysis
 models to study the invariance of three youth gratitude scales across
 students aged 10 to 19 years.  Due to sample size constraints, the age
-variable included six categories: 10-11 years, 12-13 years, 14 years,
-15 years, 16 years, and 17-19 years.  Thus, age is an ordinal variable
+variable included six categories: 10--11 years, 12--13 years, 14 years,
+15 years, 16 years, and 17--19 years.  Thus, age is an ordinal variable
 with which the proposed tests can be applied.
@@
 \approx 1400$), they could not rely solely on likelihood ratio tests
-(i.e., chi-square difference tests) for model comparison because the
+(i.e., $\chi^2$ difference tests) for model comparison because the
 tests were overly sensitive at their sample size.  To supplement these
-tests, Froh et al examined a set of alternative fit indices, including
+tests, \citeNP{FroFan11} examined a set of alternative fit indices, including
 the non-normed fit index, the comparative fit index, and the
@@
 
-In the analyses described below, we re-analyze the Froh et al.\ data using
+In the analyses described below, we re-analyze the \citeNP{FroFan11} data using
 the ordinal test statistics proposed in this paper.  This results in a
@@
 while being more sensitive to ordinal violations of measurement invariance.
-We focus on two analyses from Froh et al.\ where the likelihood
+We focus on two analyses from \citeNP{FroFan11} where the likelihood
 ratio test resulted in significance (indicating that the restricted
 model did not fit as well as the less-restricted model), but the
-alternative fit measures employed by Froh et al.\ indicated the opposite.
+alternative fit measures employed by \citeNP{FroFan11} indicated the opposite.
 These include
@@
 
-Of the 1401 cases originally collected by Froh et al., we use here all
+Of the 1401 cases originally collected by \citeNP{FroFan11}, we use here all
 subjects with complete data (resulting in $n=1327$).
@@
 \subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
-model to the GQ-6 data, Froh et al.\ used alternative fit indices to
+model to the GQ-6 data, \citeNP{FroFan11} used alternative fit indices to
 conclude that the tau-equivalent model was as good as the congeneric
@@
  and~$\text{LM}_o=\Sexpr{round(ol2bb1$statistic,1)}, p=\Sexpr{round(ol2bb1$p.value,2)}$.
-Neither statistic is significant at an $\alpha$ of .05,
-which supports the conclusions that Froh et al.\ obtained from
+Neither statistic is significant at an $\alpha$ of 0.05,
+which supports the conclusions that \citeNP{FroFan11} obtained from
 alternative fit statistics.  This provides further evidence that the
@@
 \subsubsection{GAC} In fitting tau-equivalent and parallel models to the
-GAC data, Froh et al.\ obtained mixed results.  The alternative fit
+GAC data, \citeNP{FroFan11} obtained mixed results.  The alternative fit
 indices did not all agree with one another, and the likelihood ratio test
 indicated that the parallel model fit worse than the tau-equivalent
-model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2],1)}, p<.01$
-for the data considered here).  Froh et al.\ ultimately 
+model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2],1)}, p<0.01$
+for the data considered here).  \citeNP{FroFan11} ultimately 
 concluded that the tau-equivalent model provided a better fit than did
@@
 (unique variance and factor variance) w.r.t.\ age.
-This results in~$\text{WDM}_o=\Sexpr{round(omax2$statistic,1)}, p<.01$
-and~$\text{LM}_o=\Sexpr{round(ol2bb2$statistic,1)}, p<.01$.
+This results in~$\text{WDM}_o=\Sexpr{round(omax2$statistic,1)}, p<0.01$
+and~$\text{LM}_o=\Sexpr{round(ol2bb2$statistic,1)}, p<0.01$.
 Both of these statistics agree with the likelihood ratio test and with

### 2012-08-21 ordinal/ordinal.Rnw
improved some comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@163 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 unequal across individuals (and related to $V$).
-\readme{Maybe we should explicitly mention that in parametric settings, f is
+\fixme{Maybe we should explicitly mention that in parametric settings, f is
 the probability density function of x? Also, do we need to point out
@@
 
-\readme{My feeling is that we still need to explain better what type
+\fixme{My feeling is that we still need to explain better what type
 of patterns are covered by which alternative. Maybe we should do it

### 2012-08-21 ordinal/ordinal.Rnw
Edited intro in light of Achim's comments.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@164 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 (see, e.g., \citeNP{Mil11} for a thorough review).  A set of
-psychometric scales ${\bm X}$ are defined to be measurement invariant
+psychometric scales $X$ are defined to be measurement invariant
 with respect to an auxiliary variable $V$ if \cite{Mel89}
@@
     \label{eq:midef}
-      f(x_i | t_i, v_i, \dots) = f({\bf x}_i | t_i, \dots),
+      f(x_i | t_i, v_i, \dots) = f(x_i | t_i, \dots),
 \end{equation}
 where $T$ is the latent variable that the scales measure, $f$ is the
-model's distributional form, and the $i$ subscript refers to
-individual cases.  If the above equation does not hold, then there
+model's distributional form, the $i$ subscript refers to
+individual cases, capital letters signify random variables, and lowercase
+letters signify realizations of the variables.  If the above equation
+does not hold, then there 
 exists a measurement invariance violation.  We focus here on
-situations where $f()$ is parametric, and the measurement invariance
+situations where $f()$ is the probability density function of $X$,
+and the measurement invariance
 violation occurs because the model parameters are
 unequal across individuals (and related to $V$).
-\fixme{Maybe we should explicitly mention that in parametric settings, f is
-the probability density function of x? Also, do we need to point out
-the relation between X and x, V and v, T and t?}
 
 As a concrete example of the study of measurement invariance, consider
-a situation where ${\bm X}$ contains some ``high stakes'' tests of
+a situation where $X$ includes ``high stakes'' tests of
 ability and $V$ is ethnicity.
 One's ethnicity should be unrelated to the measurement parameters
-within $f()$, and this expectation can be studied within a
-psychometric model by fitting the model and examining whether or not
+within $f()$, and this expectation can be studied by fitting the model
+and examining whether or not 
 measurement parameters 
@@
 
-Along with ethnicity, researchers are often interested in studying
+Along with categorical variables such as ethnicity, researchers are
+often interested in studying 
 measurement invariance with respect to ordinal $V$.
@@
 two test statistics that fill this gap and to show that the statistics
-possess good properties to be used in practice.
+possess good properties for use in practice.
 
@@
 where ${\bm \theta}_i$ reflects the parameter vector for individual
-$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  This
-hypothesis is often tested against an alternative
+$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  
+The most general alternative hypothesis related to $V$ may then be
+written as
+\begin{equation}
+  \label{eq:h2*}
+  H_1^*:~ {\bm \theta}_i = {\bm \theta}_{v_i},
+\end{equation}
+stating that the parameter vector differs for every unique realization
+of $V$.  This alternative is commonly employed when $V$ is
+categorical.  In these situations, the likelihood ratio test compares
+a model where 
+parameters are restricted across groups (i.e., across values of $V$)
+to a model where parameters are free across groups; the exact
+parameter values within each group are completely unrestricted.
+However, in situations where $V$ is categorical or
+continuous,~\eqref{eq:h2*} includes non-monotonic violations of
+measurement invariance.   This allows for instances where, e.g., the
+parameter values initially increase with $V$ and then decrease.
+Researchers typically do not expect such a result when testing
+measurement invariance wrt continuous or ordinal $V$, and 
+researchers often cannot interpret such violations.  Monotonic
+parameter changes wrt $V$ are of much more interest in these
+situations, with the simplest type of change given by the alternative
+hypothesis
 \begin{equation}
   \label{eq:h1*}
-  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
+  H_2^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
     {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
@@
 $V$ and $\nu$ is a threshold dividing individuals into groups based on
-$V$.  Note that this alternative hypothesis can also be used when $V$
-is ordinal, testing whether the model parameters shift at some point
-on $V$.  
-Alternatively, if $V$ is categorical with a relatively-small number of
-groups, a frequent alternative of interest is
-\begin{equation}
-  \label{eq:h2*}
-  H_2^*:~ {\bm \theta}_i = {\bm \theta}_{v_i}.
-\end{equation}
+$V$.  This alternative is implicitly employed in ``median split''
+analyses, % TODO cite preacher
+where $\nu$ is given as the sample median of $V$.  The threshold $\nu$
+is usually unknown, however, so it is generally of interest to
+test~\eqref{eq:h1*} across all possible values of $\nu$.  The tests
+proposed below generally allow for this.
+
 In this paper, we generally focus on situations where $V$ is
@@
 Thus, test statistics that explicitly treat $V$ as ordinal should have
-higher power to detect the ``ordinal'' types of violations described above.
-
-\fixme{My feeling is that we still need to explain better what type
-of patterns are covered by which alternative. Maybe we should do it
-the other way round and start with $H_2^*$ as the most general type
-of deviation from $H_0$. Then, we could explain that these also
-covers changes that are not `monotonic' in V, i.e., where parameter
-values change in one direction as V increases and then change into
-another direction again. After that we could continue and argue that
-such monotonic changes are of particular interest for categorical V
-and that the simplest form of such a change is $H_1^*$. And as a final
-note one could add that this one-shift alternative is also often
-employed in practice where practitioners often use the median V
-as the cutoff whereas we want to assess all possible cutoffs.}
+higher power to detect measurement invariance violations that are
+monotonic with $V$.
 
@@
 
-When $V$ is categorical with a relatively-small number of categories,
+As mentioned previously, when $V$ is categorical with a
+relatively-small number of categories, 
 tests of measurement invariance typically proceed via multiple-group
@@
 related to those underlying the Lagrange multiplier test and are
-discussed by \citeNP{MerZei12} in detail. Here, we review the tests'
-most important properties that are relevant for extending the tests
+discussed by Merkle and Zeileis \citeyear{MerZei12} in detail. Here,
+we review the tests' properties that are relevant for extending them
 to the ordinal case.
 
-To formalize these ideas, we assume that the observations are ordered
+To formalize the ideas discussed in the previous paragraph, we assume
+that the observations are ordered 
 w.r.t.\ $V$ (so that $v_i \le v_{i + 1}$) and define the $k$-dimensional
@@
 Thus, we can construct tests of measurement invariance by comparing
-the cumulative score process's behavior to that of a Brownian bridge.
+the behavior of the cumulative score process to that of a Brownian bridge.
 This is accomplished by comparing a scalar statistic associated with
@@
 
-\readme{I changed/extended the comments about the power of the tests
-and their relative advantages somewhat.}
-
 %% Paragraph on test statistics
@@
 which essentially tests whether any component of the cumulative score
-process strays too far from zero.  This test discards information
-related to multiple parameters fluctuating simultaneously, and
-\citeNP{MerZei12} found it to be easy to visualize and
-interpret but to have relatively-low power for assessing measurement
+process strays too far from zero and is easily visualized.  This test
+discards information 
+related to multiple parameters fluctuating simultaneously, resulting
+in it having relatively-low power for assessing measurement
 invariance when multiple factor analysis parameters change
-simultaneously.
+simultaneously \cite{MerZei12}.
 
@@
 scaled by the asymptotic variance $t (1 - t)$ of the process
-${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeNP{MerZei12} found
+${\bm B}(t, \hat {\bm \theta})$.  In simulations, Merkle and Zeileis
+\citeyear{MerZei12} found 
 that both tests perform well when assessing simultaneous changes
-in multiple factor analysis parameters with some advantages for the
-$\mathit{CvM}$ test in their simulation setup. These simulations included
-situations in which subsets of model parameters were tested; such situations are handled
+in multiple factor analysis parameters, with the
+$\mathit{CvM}$ test being somewhat advantageous in their particular
+simulation setup. These simulations included 
+situations in which subsets of model parameters were tested; such
+situations are handled 
 by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
@@
 ordering. (Note that the same also applies if $V$ is metric in nature but is only
-discretely measured leading to many ties.) \readme{Extended the previous sentence somewhat.
-My feeling is that we now should first comment on the ordinal test statistics because these
-directly build on the previous `continuous V' section. Then we can comment later on
-on the categorical case.}
+discretely measured leading to many ties.)
 
-\readme{I have also extended the comments about the nominal/categorical case.}
-Furthermore, if $V$ is only nominal/categorical,
-there is not even a partial ordering, i.e., measurement invariance tests should neither
-exploit the ordering of $V$'s levels nor of the observations within the level.
-In this situation, it is possible to obtain a test statistic by first summing scores within
-each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
-statistic \cite{HjoKon02}.  This test statistic can be formally written as
-\begin{equation}
-    \label{eq:lmuo}
-     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
-       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
-\end{equation}
-where, again, tests of subsets of model parameters can be obtained by
-taking the inner sum over only the $k^* < k$ parameters of interest.
-\fixme{The $i_\ell$ still need to be introduced. This would happen automatically
-if we move the nominal statistics behind the ordinal statistics.}
-This test statistic discards the ordinal nature of the auxiliary
-variable, essentially employing the alternative hypothesis
-from~\eqref{eq:h2*}.  A similar issue is observed in testing for
-measurement 
-invariance via multiple groups models and likelihood ratio tests (or,
-equivalently, via Wald tests or Lagrange multiplier tests): we
-can allow $\bm{\theta}$ to be unique at each level of the ordinal
-variable, but the ordinality of the auxiliary variable is lost.
-In contrast, the statistics proposed below explicitly account for the fact
-that $V$ is ordinal.
-
-The statistics proposed here are
+The ordinal statistics proposed here are
 similar to those described in
@@
 
-\readme{Several details/extensions added to the following three paragraphs.}
-
 For $\mathit{WDM}_o$ the resulting asymptotic distribution is
 $\max_{j = 1, \dots, k} \max_{\ell = 1, \dots, m-1} {\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$.
-Note that the effect outer maximum can be easily captured by a Bonferroni
+Note that the effect of the outer maximum can be easily captured by a Bonferroni
 correction as the $k$~components of the Brownian bridge are asymptotically independent.
@@
 distribution with standard normal marginals and this particular correlation
-matrix; see also \citeNP{hotzei08} for more details. In R, this can be accomplished
-using the mvtnorm package \cite{mvtnorm}.
+matrix; see also \citeNP{hotzei08} for more details. In R, this can be
+accomplished using the mvtnorm package \cite{mvtnorm}.
 
@@
 
-As demonstrated in the simulations below, the test
+If $V$ is only nominal/categorical,
+there is not even a partial ordering, i.e., measurement invariance tests should neither
+exploit the ordering of $V$'s levels nor of the observations within the level.
+In this situation, it is possible to obtain a test statistic by first summing scores within
+each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
+statistic \cite{HjoKon02}.  This test statistic can be formally written as
+\begin{equation}
+    \label{eq:lmuo}
+     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
+\end{equation}
+where, again, tests of subsets of model parameters can be obtained by
+taking the inner sum over only the $k^* < k$ parameters of interest.
+This test statistic discards the ordinal nature of the auxiliary
+variable, essentially employing the alternative hypothesis
+from~\eqref{eq:h2*}.  A similar issue is observed in testing for
+measurement 
+invariance via multiple groups models and likelihood ratio tests (or,
+equivalently, via Wald tests or Lagrange multiplier tests): we
+can allow $\bm{\theta}$ to be unique at each level of the ordinal
+variable, but the ordinality of the auxiliary variable is lost.
+In contrast, the statistics proposed below explicitly account for the fact
+that $V$ is ordinal.
+
+As demonstrated in the simulations below, the ordinal test
 statistics proposed above are sensitive to the measurement invariance
@@
 wll. This includes abrupt shifts in the parameters $\bm \theta$ at a
-certain level of $V$ as well as smooth increases decreases in the parameters.
+certain level of $V$ as well as smooth increases/decreases in the parameters.
 Taking a maximum over the $k$~parameters as in $\mathit{WDM}_o$ will be
-more sensitive to changes that occur only in one out of many parameters
+more sensitive to changes that occur only in one out of many
+parameters, 
 while $\max \mathit{LM}_o$ will be more sensitive to changes occuring
@@
 use the proposed test statistics to test for measurement invariance
-simultaneously across all six unique variances, mimicking the first
-step of the ``prescribed'' course of measurement invariance tests. % cite
+simultaneously across all six unique variances, which mimics the
+comparison of a tau-equivalent model to a congeneric model (e.g.,
+\citeNP{VanLan00}).
 

### 2012-08-22 ordinal/ordinal.Rnw
Revised and added to the general discussion of the ordinal paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@165 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\begin{enumerate}
-  \item To my knowledge, these are the only proposed statistics that
-    treat ordinal auxiliary 
-    variables as truly ordinal.  LRT treats them as categorical,
-    inclusion in the model typically treats tham as continuous.
-  \item The first point is important because, if we test w.r.t.\ an
-    ordinal auxiliary variable, we are generally looking for ordinal
-    violations.  We don't care if a middle level differs from
-    all others because we would not usually have an intuitive
-    explanation for it.
-  \item If we know our sample size is large and have an ordinal
-    auxiliary variable, should we even use the LRT in
-    the first place?  If the measurement invariance violation is truly
-    ordinal, the proposed statistics should be better.  If it is not
-    ordinal, the violation will be difficult to explain as anything
-    but an anomaly.  The stakes might also play a role here.  In
-    high-stakes testing, you might be more interested in treating the
-    ordinal variable as categorical to detect any possible
-    instability, even those that don't make sense.
-  \item The proposed statistics should generally transfer to IRT and
-    other psychometric models.
-\end{enumerate}
+In this paper, we proposed two statistics for studying measurement
+invariance that can be used when one has an auxiliary ordinal
+variable.    To our knowledge,
+the ordinal measurement invariance statistics proposed here are the
+only ones that treat auxiliary variables as ordinal.  Other methods
+treat the auxiliary variable as either continuous or categorical, in a
+manner similar to the treatment of ordinal predictor variables in
+linear regression.  In the remainder of the paper, we provide 
+detail on test choice and on the tests' applicability to other
+models.
+
+\paragraph{Choice of Test}
+The results presented in this paper imply that the proposed
+ordinal statistics may ``miss'' measurement invariance violations that
+are not monotonic w.r.t.\ $V$, moreso than the likelihood ratio test.  We
+speculate that, in most applications, this will not be a major issue
+because the researcher's a priori hypotheses exclusively focus on
+monotonic measurement 
+invariance violations.  For example, in the youth gratitude
+application, we tested for measurement invariance across six age
+groups.  If we observed a measurement invariance violation whereby
+factor loadings were equal at all age groups except 14 years,
+we would
+have a hard time explaining the violation as anything but an anomaly
+in the 14-year-olds.  Further, if $n$ is large, we are likely to
+suspect that the result arises from the large sample size.
+We may still be interested in why the
+14-year-olds differed, but the analysis is purely exploratory at this
+point because this type of violation was unexpected.  
+However, there is generally a tradeoff between the ordinal statistics
+and the likelihood ratio statistic.  The
+proposed ordinal statistics usually provide sharper tests of one's a
+priori hypothesis regarding measurement invariance w.r.t.\ ordinal
+$V$, while the likelihood ratio statistic provides a sharper test of
+general (non-monotonic) measurement invariance w.r.t.\ $V$.  While the
+latter feature may be important in some high-stakes applications, many
+researchers are likely to find the former feature appealing for their
+work.
+
+\paragraph{Extension to Other Models}
+We focused on testing for measurement invariance in factor
+analysis models here, but the proposed test statistics are applicable to
+other psychometric models that are estimated via ML.  The only
+requirement for carrying out the tests is that the casewise scores
+(Equation~\eqref{eq:score}) be available following model estimation.
+As a result, applications to studying DIF in IRT are immediate,
+as are general psychometric applications to studying parameter
+stability w.r.t.\ ordinal auxiliary variables.  We expect the
+same general results to hold for these applications, whereby the
+proposed test statistics are better than the LRT for detecting
+monotonic instabilities.   The strucchange package \cite{Zei06} noted
+previously can be used for these more-general applications.
 
 \subsection{Summary}
+The test statistics proposed in this paper have high power for
+detecting measurement 
+invariance violations that are monotonic with the ordinal variable,
+and they have low power for detecting minor violations that are not
+monotonic.  The former feature implies that the statistics are good at
+detecting measurement invariance violations that are interpretable to
+the researcher, while the latter feature implies that the statistics
+are feasible in situations where the likelihood ratio test commonly
+rejects $H_0$ in practice (e.g., \citeNP{Bent80}).  Furthermore, the
+focal psychometric model does not have to be modified in any way,
+which differs from approaches that may treat the ordinal variable as
+continuous 
+(e.g., \citeNP{BauHus09,MolDol10,NeaAgg06}).  In all, the tests have
+advantageous properties that should be useful in practice.
 

### 2012-08-29 ordinal/ordinal.Rnw
Added fluctuation plots to application.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@166 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 yg <- yg[decimals==0,]
-
-## TODO Simulate critical values for ordL2BB
 @
@@
           meanstructure=TRUE)
-gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
+gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, 
+              order.by = as.numeric(yg$agegroup), 
               vcov = info_full, sandwich = FALSE, parm=1:4)
@@
 @
+
+\begin{figure}
+\caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
+  the $\text{LM}_o$ statistic (right panel).}
+\label{fig:gq6res}
+\setkeys{Gin}{width=\textwidth}
+<<gq6res, fig=TRUE, height=3, width=6>>=
+par(mfrow=c(1,2), cex=.65)
+
+plot(gefp1, functional=ordwmax(gefp1), axes=FALSE, main="",
+     xlab="Age group", ylab=expression("WDM"[o]))
+axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
+axis(2)
+
+plot(gefp1, functional=ol2bb1.fun, axes=FALSE, ylim=c(0,14), main="",
+     xlab="Age group", ylab=expression("LM"[o]))
+axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
+axis(2)
+@
+\end{figure}
 
@@
 
+Plots representing the statistics' fluctuations across levels of age
+group are displayed in Figure~\ref{fig:gq6res}.  The left panel
+displays the process associated with $\text{WDM}_o$, while the right
+panel displays the process associated with $\text{LM}_o$.  The final
+age group (17-19 years) is not displayed, because the statistics
+associated with this final age group are always zero (scores are
+summed across all individuals in the sample at this point).
+%% more detail
+
 <<gac>>=
@@
            meanstructure=TRUE)
-gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
+gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, 
+              order.by = as.numeric(yg$agegroup), 
               vcov = info_full, sandwich = FALSE, parm=3:6)
@@
 
-\paragraph{Choice of Test}
+\subsection{Choice of Test}
 The results presented in this paper imply that the proposed
@@
 
-\paragraph{Extension to Other Models}
+\subsection{Extension to Other Models}
 We focused on testing for measurement invariance in factor
@@
 \subsection{Summary}
-The test statistics proposed in this paper have high power for
+The test statistics proposed in this paper have relatively-high power for
 detecting measurement 
 invariance violations that are monotonic with the ordinal variable,
-and they have low power for detecting minor violations that are not
+and they have relatively-low power for detecting minor violations that are not
 monotonic.  The former feature implies that the statistics are good at
@@
 are feasible in situations where the likelihood ratio test commonly
-rejects $H_0$ in practice (e.g., \citeNP{Bent80}).  Furthermore, the
+rejects $H_0$ in practice (e.g., \citeNP{Ben80}).  Furthermore, the
 focal psychometric model does not have to be modified in any way,

### 2012-08-30 ordinal/ordinal.Rnw
Finished full draft of ordinal manuscript: o Added figures to application section. o Added to general discussion. o Added references to bib file. o Added saved simulation results for fast compiling.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@167 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \title{Testing for measurement invariance with respect to an ordinal variable}
-\threeauthors{Edgar C. Merkle}{Jinyan Fan}{Achim Zeileis}
+\threeauthors{Edgar C.\ Merkle}{Jinyan Fan}{Achim Zeileis}
 \threeaffiliations{University of Missouri}{Auburn University}{Universit\"{a}t Innsbruck}
@@
 violations that occur as we move along the levels of $V$ can be captured
-wll. This includes abrupt shifts in the parameters $\bm \theta$ at a
+well. This includes abrupt shifts in the parameters $\bm \theta$ at a
 certain level of $V$ as well as smooth increases/decreases in the parameters.
@@
 parameters, 
-while $\max \mathit{LM}_o$ will be more sensitive to changes occuring
+while $\max \mathit{LM}_o$ will be more sensitive to changes occurring
 in several (or even all of the) parameters simultaneously.
-Moreover, the test statistics are rather insensitive to anomolies in a small
+Moreover, the test statistics are rather insensitive to anomalies in a small
 number of categories of $V$ that are
@@
 six-indicator model, with a measurement invariance violation occurring
-in the three unique variances associated with the first factor.  We
+in the unique variance parameters.  We
 use the proposed test statistics to test for measurement invariance
-simultaneously across all six unique variances, which mimics the
-comparison of a tau-equivalent model to a congeneric model (e.g.,
+simultaneously across the unique variances, which is similar to
+comparing a tau-equivalent model to a congeneric model (e.g.,
 \citeNP{VanLan00}).
 
-For comparison, we also compute two statistics that treat the ordinal
+We also compute two statistics that treat the ordinal
 auxiliary variable as categorical.  These include the likelihood ratio
@@
 
-\section{Simulation 2: Minor Anomolies and Large $n$}
+\section{Simulation 2: Minor Anomalies and Large $n$}
 
@@
 tested, implying that ``practical infinity'' is reached for this model
-by $n=1200$. \readme{Also implying that we don't need all three sample
-  sizes or could try values smaller than 1200.}  We also observe
-a negative 
+by $n=1200$.   We also observe a negative 
 relationship between power and $m$; this is because the measurement
@@
 2 provide evidence that the $\text{LM}_o$ statistic should be
-preferred to the $\text{WDM}_o$ statistic for testing measurement
-invariance. \readme{This is not surprising given that the models
+preferred to the $\text{WDM}_o$ statistic for simultaneously assessing
+measurement invariance across multiple model parameters.  
+\readme{This is not surprising given that the models
 have quite a few parameters and we change many of them. We could 
@@
 I'm not sure how much we want to comment on this or illustrate this.}
+\readme{Achim: I decided not to set up a case where a single parameter
+changes, because measurement invariance researchers usually test
+multiple parameters simultaneously.  For now, I added the phrase "across
+multiple model parameters" above as a brief reference to this issue.}
 The $\text{LM}_o$ statistic has higher power to detect
@@
 \caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
-  the $\text{LM}_o$ statistic (right panel).}
+  the $\text{LM}_o$ statistic (right panel), arising from the GQ-6 data.}
 \label{fig:gq6res}
@@
 summed across all individuals in the sample at this point).
-%% more detail
+Critical values are displayed as horizontal red lines, so that the
+hypothesis of invariance is rejected if the process ever crosses the
+red line.
+It is observed that both statistics generally increase with age, with
+$\text{WDM}_o$ being largest in the 15-year group and $\text{LM}_o$
+being largest in the 16-year group.  The $\text{WDM}_o$ panel displays
+(a function of) the largest
+cumulative sum associated with a single parameter, while the
+$\text{LM}_o$ panel aggregates across the tested parameters.  The
+differing pattern of values for the 15- and 16-year-olds can be taken
+as evidence that a single factor loading is relatively unstable for the
+15-year-olds, but this is not a general trend across all the tested
+loadings.
 
@@
 
+\begin{figure}
+\caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
+  the $\text{LM}_o$ statistic (right panel), arising from the GAC data.}
+\label{fig:gacres}
+\setkeys{Gin}{width=\textwidth}
+<<gacres, fig=TRUE, height=3, width=6>>=
+par(mfrow=c(1,2), cex=.65)
+
+plot(gefp2, functional=ordwmax(gefp2), axes=FALSE, main="",
+     xlab="Age group", ylab=expression("WDM"[o]))
+axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
+axis(2)
+
+plot(gefp2, functional=ol2bb2.fun, axes=FALSE, ylim=c(0,115), main="",
+     xlab="Age group", ylab=expression("LM"[o]))
+axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
+axis(2)
+@
+\end{figure}
+
 \subsubsection{GAC} In fitting tau-equivalent and parallel models to the
@@
 
+Plots representing the statistics' fluctuations across levels of age
+group are displayed in Figure~\ref{fig:gacres}.  The left panel
+displays the process associated with $\text{WDM}_o$, while the right
+panel displays the process associated with $\text{LM}_o$.  It is
+observed that both processes are fully above the critical value,
+implying the measurement invariance violation.  Additionally, both
+processes peak at the 12-13 age group.  This implies that fluctuation 
+in the variance parameters is largest within this age group, with the 
+parameters differing between individuals younger and
+older than 12 years.
+
 \subsection{Summary}
@@
 the hypothesis of interest.  They can be used to either supplement or
-replace the likelihood ratio test, depending upon the measurement
-invariance violations in which the researcher is interested a priori.
+replace the likelihood ratio test, depending upon the types of
+measurement
+invariance violations in which the researcher has a priori interest.
 We further describe the issue of supplementing vs.\ replacing the
@@
 
+Along with using likelihood ratio tests to study measurement
+invariance, researchers may wish to treat ordinal $V$ as continuous
+(especially if $V$ has many levels).  This treatment would allow the
+researcher to employ moderated factor models
+\cite{BauHus09,MolDol10,Pur02} or factor mixture models
+\cite{DolMaa98,LubMut05}.  Under the moderated factor model approach,
+$V$ is inserted directly into the factor analysis model and allowed to
+have a linear relationship with model parameters.  Under the factor
+mixture model approach, individuals are typically assumed to arise
+from a small number of distinct factor analysis models.  The ordinal
+variable $V$ could then be used to predict the probability that an
+individual arises from each model.  These treatments of ordinal $V$ as
+continuous are likely to be advantageous, especially if the levels of
+$V$ are approximately equally-spaced and the relationship between $V$
+and the measurement invariance violation is linear.  The approaches do
+require models of greater complexity and may not be suitable for all
+ordinal $V$, however, while that the methods we propose here are
+generally suitable for ordinal $V$.
+
 \subsection{Extension to Other Models}
@@
 which differs from approaches that may treat the ordinal variable as
-continuous 
-(e.g., \citeNP{BauHus09,MolDol10,NeaAgg06}).  In all, the tests have
+continuous.  In all, the tests have
 advantageous properties that should be useful in practice.

### 2012-08-30 ordinal/sim_ordinal.R

@@ -10,7 +10,7 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE)
   ## and gets larger.  If gradual==FALSE, the instability starts
   ## halfway up age and stays constant.
   ## If anomaly==TRUE, instability only occurs at level (1+nlevels/2).
-  ## Otherwise, anomaly continues up the ordinal variable.
+  ## Otherwise, instability continues up the ordinal variable.
   stopifnot(require("mvtnorm"))
   
   # ses is SE increase in parameters at last ordinal category.

### 2012-09-12 ordinal/ordinal.Rnw
changed path in read.spss call to include version-controlled .sav file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@168 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
extensions/improvements/suggestions, especially in application and general discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@169 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\abstract{Researchers are often interested in testing for 
+\abstract{
+  Researchers are often interested in testing for 
   measurement invariance with respect to an ordinal auxiliary variable such as
@@
   is also known to be overly sensitive at large sample sizes.  In this
-  paper, we propose test statistics that
-  explicitly account for the ordinality of the auxiliary variable.
+  paper, we propose test statistics that explicitly account for the ordinality
+  of the auxiliary variable and hence have higher power against ``monotonic''
+  violations of measurement invariance and less power against ``non-monotonic''
+  ones. \readme{I thought a comment like this might help. But I'm not sure
+  whether ``(non-)monotonic'' is intuitive to all readers.}  
   The statistics are derived from a family of tests based on stochastic
@@
 of $V$.  This alternative is commonly employed when $V$ is
-categorical.  In these situations, the likelihood ratio test compares
+categorical.  In these situations, the likelihood ratio test (LRT) compares
 a model where 
@@
 measurement invariance.   This allows for instances where, e.g., the
-parameter values initially increase with $V$ and then decrease.
+parameter values initially increase with $V$ and then decrease
+or where just one or two ``middle'' levels of $V$ differ from the rest.
 Researchers typically do not expect such a result when testing
-measurement invariance wrt continuous or ordinal $V$, and 
+measurement invariance w.r.t.\ continuous or ordinal $V$, and 
 researchers often cannot interpret such violations.  Monotonic
-parameter changes wrt $V$ are of much more interest in these
+parameter changes w.r.t.\ $V$ are of much more interest in these
 situations, with the simplest type of change given by the alternative
@@
 where $v_i$ reflects individual $i$'s value on an auxiliary variable
-$V$ and $\nu$ is a threshold dividing individuals into groups based on
+$V$ and $\nu$ is a threshold dividing individuals into two groups based on
 $V$.  This alternative is implicitly employed in ``median split''
@@
     \label{eq:wdm}
-      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
                              \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
     \label{eq:maxlmo}
-      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
                              \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
@@
 }
+sim1$test <- factor(as.character(sim1$test),
+  levels = c("LM_o", "WDM_o", "LRT", "LM_uo"),
+  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo"))
 @
@@
 corresponds to power.  It is seen that one of the proposed test
-statistics, the $\max \mathit{LM}_o$ statistic from~\eqref{eq:maxlmo},
+statistics, the $\max \mathit{LM}_o$ statistic
+\readme{$\max \mathit{LM}_o$ is now consistently employed as the name
+for this statistic, also in the graphics. First, this is more consistent
+with the notation in the theory section. Second, it should remind the
+reader that this is not an ordinary LM test but an aggregation of
+LM statistics.}
+from~\eqref{eq:maxlmo},
 generally has the largest power to detect the ordinal measurement
@@
 <<sim1res, fig=TRUE, height=7, width=7>>=
-mykey <- simpleKey(levels(sim1$test), points = FALSE, lines = TRUE)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
 print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1, 
@@
 }
+sim2$test <- factor(as.character(sim2$test),
+  levels = c("LM_o", "WDM_o", "LRT", "LM_uo"),
+  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo"))
 @
@@
 test statistics that we proposed have considerably lower power, with
-the $\text{WDM}_o$ statistic being the lowest and the $\text{LM}_o$
+the $\mathit{WDM}_o$ statistic being the lowest and the $\max\mathit{LM}_o$
 statistic being higher at larger values of $d$.  
@@
 <<sim2res, fig=TRUE, height=7, width=7>>=
-mykey <- simpleKey(levels(sim2$test), points = FALSE, lines = TRUE)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
 print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2, 
@@
 
-Taken together, the results from Simulation 1 and Simulation
-2 provide evidence that the $\text{LM}_o$ statistic should be
-preferred to the $\text{WDM}_o$ statistic for simultaneously assessing
-measurement invariance across multiple model parameters.  
-\readme{This is not surprising given that the models
-have quite a few parameters and we change many of them. We could 
-try to set up a simulation where only a single out of many parameters
-changes where the WDM should perform better. Also,
-in models that are simpler than most CFA models, the WDM may be more useful.
-I'm not sure how much we want to comment on this or illustrate this.}
-\readme{Achim: I decided not to set up a case where a single parameter
-changes, because measurement invariance researchers usually test
-multiple parameters simultaneously.  For now, I added the phrase "across
-multiple model parameters" above as a brief reference to this issue.}
-The $\text{LM}_o$ statistic has higher power to detect
+Taken together, the results from Simulation~1 and Simulation~2
+provide evidence that the $\max\mathit{LM}_o$ statistic should be
+preferred to the $\mathit{WDM}_o$ statistic for simultaneously assessing
+measurement invariance across multiple parameters in factor analysis models.
+\readme{Ed: For now, I added the phrase ``across multiple model parameters'' above as a brief reference to this issue.}
+\readme{Z: OK, I appended ``in factor analysis models'' to make the statement somewhat less general and more specific to our results.}
+The $\max\mathit{LM}_o$ statistic has higher power to detect
 ordinal violations, and its power to detect non-ordinal violations was
-similar to that of $\text{WDM}_o$ when the violation magnitude was
-small (e.g., for $d \leq 1.5$).  The $\text{LM}_o$ statistic is
+similar to that of $\mathit{WDM}_o$ when the violation magnitude was
+small (e.g., for $d \leq 1.5$).  The $\max\mathit{LM}_o$ statistic is
 advantageous because it can make use of invariance violations that
 simultaneously 
-occur in multiple parameters, whereas the $\text{WDM}_o$ focuses only
+occur in multiple parameters, whereas the $\mathit{WDM}_o$ focuses only
 on the parameter with the largest invariance violation.  
@@
 through these alternative fit indices; the likelihood ratio test often
-resulted in significance event when the alternative indices indicated
+resulted in significance even when the alternative indices indicated
 good fit.
@@
 <<frohdata>>=
-yg <- read.spss("grat data combined_9.29.09.sav", to.data.frame=TRUE)
+yg <- read.spss("../prelim_code/grat_data_combined_2009-09-29.sav", to.data.frame=TRUE)
 names(yg) <- tolower(names(yg))
@@
 \begin{figure}
-\caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
-  the $\text{LM}_o$ statistic (right panel), arising from the GQ-6 data.}
+\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
+  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GQ-6 data.}
 \label{fig:gq6res}
@@
 <<gq6res, fig=TRUE, height=3, width=6>>=
-par(mfrow=c(1,2), cex=.65)
+par(mfrow=c(1, 2), cex = 0.65)
 
 plot(gefp1, functional=ordwmax(gefp1), axes=FALSE, main="",
-     xlab="Age group", ylab=expression("WDM"[o]))
+     xlab="Age group", ylab="Weighted max statistics")
 axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
@@
 plot(gefp1, functional=ol2bb1.fun, axes=FALSE, ylim=c(0,14), main="",
-     xlab="Age group", ylab=expression("LM"[o]))
+     xlab="Age group", ylab="LM statistics")
 axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
@@
 model.  However, the likelihood ratio test comparing these two models
-was significant ($\chi^2_{20} = \Sexpr{round(lrt1[[5]][2],1)}, p=\Sexpr{round(lrt1[[7]][2],2)}$ for the data considered
+was significant ($\chi^2_{20} = \Sexpr{round(lrt1[[5]][2], digits = 2)}, p=\Sexpr{round(lrt1[[7]][2], digits = 3)}$ for the data considered
 here).
 
+\fixme{Here and also subsequently, formulations like ``we compare model~A
+and model~B'' are used. I'm not sure whether this sends the correct message.
+If I want to compare exactly these two models, then the LRT is the correct
+thing to do. But we are looking for other types of deviations, especially if
+a split into two (rather than all) groups might be sufficient to capture
+the effect of $V$. Hence, maybe we should explain in some more detail what
+we are testing and also explore other models when rejecting the model? For
+more comments in this vein see below.}
 We can use the proposed ordinal statistics to compare the
@@
 fitting the tau-equivalent model and testing whether or
-not the factor loadings fluctuate with respect to age group.  In doing
-so, we obtain~$\text{WDM}_o=\Sexpr{round(omax1$statistic,1)}, p=\Sexpr{round(omax1$p.value,2)}$
- and~$\text{LM}_o=\Sexpr{round(ol2bb1$statistic,1)}, p=\Sexpr{round(ol2bb1$p.value,2)}$.
-Neither statistic is significant at an $\alpha$ of 0.05,
+not the factor loadings fluctuate with respect to age group.
+\fixme{Here, I would prefer something like: We can use the proposed
+ordinal statistics to assess whether or not the factor loadings in the
+tau-equivalent model are in fact stable or fluctuate with respect to age.}
+In doing so, we obtain~$\mathit{WDM}_o=\Sexpr{round(omax1$statistic, digits = 2)}, p=\Sexpr{round(omax1$p.value, digits = 3)}$
+ and~$\max\mathit{LM}_o=\Sexpr{round(ol2bb1$statistic, digits = 2)}, p=\Sexpr{round(ol2bb1$p.value, digits = 3)}$.
+Both $p$-values are clearly larger than that of the likelihood ratio test
+and neither is significant at an $\alpha$ of 0.05,
 which supports the conclusions that \citeNP{FroFan11} obtained from
-alternative fit statistics.  This provides further evidence that the
-likelihood ratio statistic is overly sensitive for this particular
-analysis.
+alternative fit statistics.  This provides further evidence that 
+there is no systematic deviation of the factor loadings along age
+and that the likelihood ratio statistic is overly sensitive, picking
+up some non-systematic dependence on age. \readme{I have changed the
+previous sentence somewhat to make clear that the LRT really picks up
+something but that it's not what we are looking for.}
 
@@
 group are displayed in Figure~\ref{fig:gq6res}.  The left panel
-displays the process associated with $\text{WDM}_o$, while the right
-panel displays the process associated with $\text{LM}_o$.  The final
-age group (17-19 years) is not displayed, because the statistics
-associated with this final age group are always zero (scores are
-summed across all individuals in the sample at this point).
-Critical values are displayed as horizontal red lines, so that the
-hypothesis of invariance is rejected if the process ever crosses the
-red line.
+displays the process associated with $\mathit{WDM}_o$ from \eqref{eq:wdm},
+i.e., the sequence of weighted maximum (over $j$) statistics for each
+potential threshold~$i$.
+The right panel displays the process associated with $\max\mathit{LM}_o$
+from \eqref{eq:maxlmo}, i.e., the sequence of LM statistics for each
+potential threshold~$i$.
+In both cases, the test statistics in the sequence assess a split of
+the observations up to age group~$i$ vs.\ greater than $i$ and the
+null hypothesis is rejected if the maximum of the statistics is
+larger than its 5\% critical value (visualized by the horizontal red line).
+Therefore, the final age group (17--19 years) is not displayed, because the statistics
+associated with this final age group would encompass all observations
+in a single group and hence are always zero.
 It is observed that both statistics generally increase with age, with
-$\text{WDM}_o$ being largest in the 15-year group and $\text{LM}_o$
-being largest in the 16-year group.  The $\text{WDM}_o$ panel displays
-(a function of) the largest
-cumulative sum associated with a single parameter, while the
-$\text{LM}_o$ panel aggregates across the tested parameters.  The
+$\mathit{WDM}_o$ being largest for a threshold of 15~years and $\max\mathit{LM}_o$
+for a threshold of 16~years. The
 differing pattern of values for the 15- and 16-year-olds can be taken
-as evidence that a single factor loading is relatively unstable for the
-15-year-olds, but this is not a general trend across all the tested
-loadings.
+as an indication that some factor loading is somewhat unstable at an
+age of 15 or 16, but this is not a clear and general trend across all the tested
+loadings and age groups.
+\readme{Added some more detail about how to read the visualizations of
+the test statistics. Modified the interpretations somewhat.}
 
@@
 \begin{figure}
-\caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
-  the $\text{LM}_o$ statistic (right panel), arising from the GAC data.}
+\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
+  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GAC data.}
 \label{fig:gacres}
@@
 <<gacres, fig=TRUE, height=3, width=6>>=
-par(mfrow=c(1,2), cex=.65)
+par(mfrow=c(1, 2), cex = 0.65)
 
 plot(gefp2, functional=ordwmax(gefp2), axes=FALSE, main="",
-     xlab="Age group", ylab=expression("WDM"[o]))
+     xlab="Age group", ylab="Weighted max statistics")
 axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
@@
 plot(gefp2, functional=ol2bb2.fun, axes=FALSE, ylim=c(0,115), main="",
-     xlab="Age group", ylab=expression("LM"[o]))
+     xlab="Age group", ylab="LM statistics")
 axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
@@
 indicated that the parallel model fit worse than the tau-equivalent
-model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2],1)}, p<0.01$
+model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2], digits = 2)}, p<0.01$
 for the data considered here).  \citeNP{FroFan11} ultimately 
@@
 (unique variance and factor variance) w.r.t.\ age.
-This results in~$\text{WDM}_o=\Sexpr{round(omax2$statistic,1)}, p<0.01$
-and~$\text{LM}_o=\Sexpr{round(ol2bb2$statistic,1)}, p<0.01$.
-Both of these statistics agree with the likelihood ratio test and with
-the general conclusion that the tau-equivalent model is better than
-the parallel model.  This provides further evidence that the
+This results in~$\mathit{WDM}_o=\Sexpr{round(omax2$statistic, digits = 2)}, p<0.01$
+and~$\max\mathit{LM}_o=\Sexpr{round(ol2bb2$statistic, digits = 2)}, p<0.01$.
+Both of these statistics agree with the general conclusion that the
+parallel model is not sufficient, providing further evidence that the
 significant likelihood ratio test is not simply an artifact of the
-large sample size.
-
-Plots representing the statistics' fluctuations across levels of age
-group are displayed in Figure~\ref{fig:gacres}.  The left panel
-displays the process associated with $\text{WDM}_o$, while the right
-panel displays the process associated with $\text{LM}_o$.  It is
+large sample size. \readme{Sentence somewhat modified.}
+
+Plots representing the statistics' fluctuations across age
+groups are displayed in Figure~\ref{fig:gacres}.  The left panel
+displays the process associated with $\mathit{WDM}_o$, while the right
+panel displays the process associated with $\max\mathit{LM}_o$.  It is
 observed that both processes are fully above the critical value,
 implying the measurement invariance violation.  Additionally, both
-processes peak at the 12-13 age group.  This implies that fluctuation 
-in the variance parameters is largest within this age group, with the 
-parameters differing between individuals younger and
-older than 12 years.
+processes peak at the 12--13 age group, suggesting that
+parameters differ between individuals up to 13 years of age and older
+than 13 years. \fixme{Maybe we could additionally fit a tau-equivalent
+model for only two age groups: $\le 13$ vs.\ $> 13$. Then, we could
+re-run tests and check whether this already fits sufficiently well?
+This could help making the split into two subgroups clearer, rather
+than always splitting into all possible age groups.}
 
@@
 the ordinal measurement invariance statistics proposed here are the
-only ones that treat auxiliary variables as ordinal.  Other methods
+only ones that treat auxiliary variables as ordinal and thus direct power against
+alternatives that are typically of interest to practitioners
+\readme{added half-sentence}.  Other methods
 treat the auxiliary variable as either continuous or categorical, in a
@@
 \subsection{Choice of Test}
+
 The results presented in this paper imply that the proposed
 ordinal statistics may ``miss'' measurement invariance violations that
-are not monotonic w.r.t.\ $V$, moreso than the likelihood ratio test.  We
-speculate that, in most applications, this will not be a major issue
+are not monotonic w.r.t.\ $V$. More precisely, while the suggested tests
+are also consistent for such non-monotonic violations, they seem to be
+clearly less powerful than the likelihood ratio test \readme{added some
+more ``technical'' detail}.
+We speculate that, in most applications, this will not be a major issue
 because the researcher's a priori hypotheses exclusively focus on
@@
 and the likelihood ratio statistic.  The
-proposed ordinal statistics usually provide sharper tests of one's a
+proposed ordinal statistics usually provide more powerful tests of one's a
 priori hypothesis regarding measurement invariance w.r.t.\ ordinal
-$V$, while the likelihood ratio statistic provides a sharper test of
-general (non-monotonic) measurement invariance w.r.t.\ $V$.  While the
+$V$, while the likelihood ratio statistic provides a more powerful test of
+general (non-monotonic) measurement invariance w.r.t.\ $V$. \readme{I
+replace ``sharper'' by ``more powerful'' which would be the correct
+statistical jargon. Or is sharper the terminus technicus in psychometrics?}
+While the
 latter feature may be important in some high-stakes applications, many
@@
 invariance, researchers may wish to treat ordinal $V$ as continuous
-(especially if $V$ has many levels).  This treatment would allow the
-researcher to employ moderated factor models
+(especially if $V$ has many levels).  
+\fixme{We should also point to the supLM test of our first paper here.
+The maxLM and supLM will generally be very close if the number of
+potential thresholds is large. Maybe we could also re-iterate that
+the forming of age groups a priori might not even be necessary and
+more studies could collect data about age as a continuous rather than
+just an ordinal variable.}
+This treatment would allow the researcher to employ moderated factor models
 \cite{BauHus09,MolDol10,Pur02} or factor mixture models
@@
 analysis models here, but the proposed test statistics are applicable to
-other psychometric models that are estimated via ML.  The only
+other psychometric models that are estimated via ML (or similar estimation
+techniques for independent observations that are governed by a central limit
+theorem). \readme{Added the comment in brackets. Useful or too detailed and
+potentially confusing?} The only
 requirement for carrying out the tests is that the casewise scores
@@
 previously can be used for these more-general applications.
+
+\fixme{Should we make some practical remarks about the computation of
+$p$-values somewhere? The issue with the maxLM test is that it is hard
+to tabulate critical values in advance and hence simulation has to be
+done for each case and is potentially rather time-consuming. Not sure
+whether we should do this now or wait until the revision or so\dots}
 

### 2012-10-15 ordinal/ordinal.Rnw
Addressed Achim's comments on the ordinal ms.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@170 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   paper, we propose test statistics that explicitly account for the ordinality
-  of the auxiliary variable and hence have higher power against ``monotonic''
-  violations of measurement invariance and less power against ``non-monotonic''
-  ones. \readme{I thought a comment like this might help. But I'm not sure
-  whether ``(non-)monotonic'' is intuitive to all readers.}  
+  of the auxiliary variable, resulting in  higher power against ``monotonic''
+  violations of measurement invariance and lower power against ``non-monotonic''
+  ones. 
   The statistics are derived from a family of tests based on stochastic
@@
 letters signify realizations of the variables.  If the above equation
-does not hold, then there 
-exists a measurement invariance violation.  We focus here on
+does not hold, then  
+a measurement invariance violation is said to exist.  We focus here on
 situations where $f()$ is the probability density function of $X$,
@@
 While these variables are relatively easy to find in the literature,
-there exist very few measurement invariance methods that specifically
-account for the fact that $V$ is ordinal.  More often, $V$ is simply
-treated as categorical so that the traditional tests can be applied.
+there exist very few psychometric methods that specifically
+account for the fact that $V$ is ordinal.  More often, $V$ is 
+treated as categorical so that the traditional tests can be applied.  Alternatively, if there are many levels of $V$, $V$ may also be treated as continuous. 
 It is the intent of this paper to propose
-two test statistics that fill this gap and to show that the statistics
+two test statistics that explicitly treat $V$ as ordinal and to show
+that the statistics 
 possess good properties for use in practice.
@@
 
-We will consider situations in which a $p$-dimensional variable $X$
+We  consider situations in which a $p$-dimensional variable $X$
 with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
@@
 measurement invariance.   This allows for instances where, e.g., the
-parameter values initially increase with $V$ and then decrease
+parameter values initially increase with $V$ and then decrease, 
 or where just one or two ``middle'' levels of $V$ differ from the rest.
@@
 \end{equation}
-where $v_i$ reflects individual $i$'s value on an auxiliary variable
-$V$ and $\nu$ is a threshold dividing individuals into two groups based on
+where $\nu$ is a threshold dividing individuals into two groups based on
 $V$.  This alternative is implicitly employed in ``median split''
@@
 
-In this paper, we generally focus on situations where $V$ is
+As stated previously, we focus here on situations where $V$ is
 ordinal and where the measurement invariance violation is related to the ordinal
@@
 parameters are constrained to be equal across groups.  When $V$ is
-continuous, however, multiple-group models obviously cannot be used 
+continuous, however, multiple-group models usually cannot be used 
 because there are no existing groups.  Instead, we can fit a model
@@
 related to those underlying the Lagrange multiplier test and are
-discussed by Merkle and Zeileis \citeyear{MerZei12} in detail. Here,
+discussed in detail by Merkle and Zeileis \citeyear{MerZei12}. Here,
 we review the tests' properties that are relevant for extending them
@@
 of all individuals, i.e., observations with the same level of $V$ have no unique
-ordering. (Note that the same also applies if $V$ is metric in nature but is only
+ordering. (Note that the same also applies if $V$ is continuous in nature but is only
 discretely measured leading to many ties.)
@@
 
-For $\mathit{WDM}_o$ the resulting asymptotic distribution is
+For the $\mathit{WDM}_o$ statistic, the resulting asymptotic distribution is
 $\max_{j = 1, \dots, k} \max_{\ell = 1, \dots, m-1} {\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$.
@@
 For $\max \mathit{LM}_o$ the resulting asymptotic distribution is
-$\max{\ell = 1, \dots, m-1} || {\bm B}^{0}(t_\ell) ||_2^2 / (t_\ell (1 - t_\ell))$
+$\max_{\ell = 1, \dots, m-1} || {\bm B}^{0}(t_\ell) ||_2^2 / (t_\ell (1 - t_\ell))$
 for which no simple closed-form solution is available. However,
@@
 variable, but the ordinality of the auxiliary variable is lost.
-In contrast, the statistics proposed below explicitly account for the fact
+In contrast, the statistics proposed above explicitly account for the fact
 that $V$ is ordinal.
 
-As demonstrated in the simulations below, the ordinal test
-statistics proposed above are sensitive to the measurement invariance
+As demonstrated in the simulations below, the proposed ordinal test
+statistics are sensitive to the measurement invariance
 violations that an analyst would typically expect from an ordinal $V$.
@@
 unrelated to the ordering of $V$.  This is especially relevant to situations
-where the analyst has a large sample size, where the usual
+in which the analyst has a large sample size, so that the usual
 likelihood ratio test is notoriously sensitive to minor parameter
@@
 \subsection{Method}
-As described above, data were generated from a two-factor model
+Data were generated from a two-factor model
 lacking measurement invariance in the six unique variance
@@
 statistics, the $\max \mathit{LM}_o$ statistic
-\readme{$\max \mathit{LM}_o$ is now consistently employed as the name
-for this statistic, also in the graphics. First, this is more consistent
-with the notation in the theory section. Second, it should remind the
-reader that this is not an ordinary LM test but an aggregation of
-LM statistics.}
 from~\eqref{eq:maxlmo},
@@
 
-In this simulation, we demonstrate that the proposed statistic is
+In this simulation, we demonstrate that the proposed statistics are 
 relatively insensitive to minor parameter violations that are
@@
 guaranteed to result in significance.  This simulation is intended to
-show that the proposed ordinal test remains viable for large $n$.
+show that the proposed ordinal tests remain viable for large $n$.
 
@@
 sample sizes were set at $n=1200,4800,9600$.  All other simulation
-features matches those from Simulation 1.
+features match those from Simulation 1.
 
@@
 two ``unordered'' test statistics both have relatively-high power to
-detect the measurement invariance violation, illustrating the finding
+detect the measurement invariance violation, illustrating the result 
 of Bentler and Bonett \citeyear{Ben80} and others that the
@@
 Both statistics exhibited much lower power to detect a measurement
-invariance violation occurring at a single level of the auxiliary
-variable.
+invariance violation that occurs only at a single level of $V$.
 
@@
 measurement invariance across multiple parameters in factor analysis models.
-\readme{Ed: For now, I added the phrase ``across multiple model parameters'' above as a brief reference to this issue.}
-\readme{Z: OK, I appended ``in factor analysis models'' to make the statement somewhat less general and more specific to our results.}
 The $\max\mathit{LM}_o$ statistic has higher power to detect
@@
 Recently, scholars have begun to explore gratitude
-in youth. However, one potential problem with this research is that
+in youth. One potential problem with this research is that
 scholars, with no exception, have used adult gratitude inventories to
@@
 
-\citeNP{FroFan11} used confirmatory factor analysis
+\citeNP{FroFan11} used confirmatory factor 
 models to study the invariance of three youth gratitude scales across
@@
 while being more sensitive to ordinal violations of measurement invariance.
-We focus on two analyses from \citeNP{FroFan11} where the likelihood
+We focus on two analyses from Froh et al.\ where the likelihood
 ratio test resulted in significance (indicating that the restricted
-model did not fit as well as the less-restricted model), but the
-alternative fit measures employed by \citeNP{FroFan11} indicated the opposite.
+model did not fit as well as the less-restricted model) but the
+alternative fit measures indicated the opposite.
 These include
@@
 
-The results section is divided into two subsections, one for each of
-the analyses described previously.  The first section contains an
+The results section is divided into two subsections, one for each
+analysis described above.  The first subsection contains an
 example of the ordinal statistics disagreeing with the likelihood
-ratio tests, while the second section contains the opposite.
+ratio tests, while the second subsection contains the opposite.
 
@@
 \subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
-model to the GQ-6 data, \citeNP{FroFan11} used alternative fit indices to
+model to the GQ-6 data, Froh et al.\ used alternative fit indices to
 conclude that the tau-equivalent model was as good as the congeneric
@@
 more comments in this vein see below.}
-We can use the proposed ordinal statistics to compare the
-tau-equivalent and congeneric models,
-fitting the tau-equivalent model and testing whether or
-not the factor loadings fluctuate with respect to age group.
-\fixme{Here, I would prefer something like: We can use the proposed
-ordinal statistics to assess whether or not the factor loadings in the
-tau-equivalent model are in fact stable or fluctuate with respect to age.}
-In doing so, we obtain~$\mathit{WDM}_o=\Sexpr{round(omax1$statistic, digits = 2)}, p=\Sexpr{round(omax1$p.value, digits = 3)}$
+\readme{Achim: Added two sentences here that hopefully address the issue.}
+We can use the proposed ordinal statistics to assess whether or not
+the factor loadings in the tau-equivalent model fluctuate with respect
+to age.  Unlike the LRT, the test does not require parameters to
+differ within each subgroup.  Instead, we test for deviations such
+that a split into two subgroups is sufficient to capture the effect of
+$V$.
+In employing the ordinal tests, we obtain~$\mathit{WDM}_o=\Sexpr{round(omax1$statistic, digits = 2)}, p=\Sexpr{round(omax1$p.value, digits = 3)}$
  and~$\max\mathit{LM}_o=\Sexpr{round(ol2bb1$statistic, digits = 2)}, p=\Sexpr{round(ol2bb1$p.value, digits = 3)}$.
@@
 and neither is significant at an $\alpha$ of 0.05,
-which supports the conclusions that \citeNP{FroFan11} obtained from
+which supports the conclusions that Froh et al.\ obtained from
 alternative fit statistics.  This provides further evidence that 
@@
 and that the likelihood ratio statistic is overly sensitive, picking
-up some non-systematic dependence on age. \readme{I have changed the
-previous sentence somewhat to make clear that the LRT really picks up
-something but that it's not what we are looking for.}
+up some non-systematic dependence on age.
 
@@
 In both cases, the test statistics in the sequence assess a split of
-the observations up to age group~$i$ vs.\ greater than $i$ and the
+the observations up to age group~$i$ vs.\ greater than $i$, and the
 null hypothesis is rejected if the maximum of the statistics is
@@
 associated with this final age group would encompass all observations
-in a single group and hence are always zero.
+in a single group and hence always equal zero.
 It is observed that both statistics generally increase with age, with
@@
 loadings and age groups.
-\readme{Added some more detail about how to read the visualizations of
-the test statistics. Modified the interpretations somewhat.}
 
@@
 model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2], digits = 2)}, p<0.01$
-for the data considered here).  \citeNP{FroFan11} ultimately 
+for the data considered here).  Froh et al.\ ultimately 
 concluded that the tau-equivalent model provided a better fit than did
@@
 significant likelihood ratio test is not simply an artifact of the
-large sample size. \readme{Sentence somewhat modified.}
+large sample size.
 
@@
 than always splitting into all possible age groups.}
+\readme{Achim: See code starting on line 963 of the Rnw file. Even if we
+  split around age 13, the test statistics are all significant.}
+
+<<gac.2grp>>=
+yg <- transform(yg, under14=as.numeric(agegroup)<=2)
+
+m3 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
+           f1 ~ 0*1', data=yg, group="under14",
+           group.equal=c("loadings","residuals","lv.variances"), 
+           meanstructure=TRUE)
+gefp3 <- gefp(m3, fit = NULL, scores=estfun.lavaan, 
+              order.by = as.numeric(yg$under14), 
+              vcov = info_full, sandwich = FALSE, parm=3:6)
+if(file.exists("ol2bb3.rda")){
+  load("ol2bb3.rda")
+} else {
+  ol2bb3.fun <- ordL2BB(gefp3)
+  save(ol2bb3.fun, file="ol2bb3.rda")
+}
+omax3 <- sctest(gefp3, functional=ordwmax(gefp3))
+ol2bb3 <- sctest(gefp3, functional=ol2bb3.fun)
+
+## Full model
+m3.f <- cfa('f1 =~ gac_1 + gac_2 + gac_3
+           f1 ~ 0*1', data=yg, group="under14",
+           group.equal="loadings", meanstructure=TRUE)
+lrt3 <- anova(m3.f, m3)
+@
 
@@
 
-In this paper, we proposed two statistics for studying measurement
-invariance that can be used when one has an auxiliary ordinal
-variable.    To our knowledge,
+In this paper, we proposed two statistics that can be used when one
+has an auxiliary ordinal variable and wishes to study measurement
+invariance.    To our knowledge,
 the ordinal measurement invariance statistics proposed here are the
-only ones that treat auxiliary variables as ordinal and thus direct power against
-alternatives that are typically of interest to practitioners
-\readme{added half-sentence}.  Other methods
+only ones that treat auxiliary variables as ordinal and thus direct
+power against 
+alternatives that are typically of interest to practitioners.  Other methods
 treat the auxiliary variable as either continuous or categorical, in a
@@
 are also consistent for such non-monotonic violations, they seem to be
-clearly less powerful than the likelihood ratio test \readme{added some
-more ``technical'' detail}.
+ less powerful than the likelihood ratio test.
 We speculate that, in most applications, this will not be a major issue
@@
 statistical jargon. Or is sharper the terminus technicus in psychometrics?}
+\readme{Achim: No, ``sharper'' was simply my terminus technicus when I
+  was writing this section.}
 While the
@@
 just an ordinal variable.}
-This treatment would allow the researcher to employ moderated factor models
-\cite{BauHus09,MolDol10,Pur02} or factor mixture models
+\readme{Achim: You say both ``maxLM'' and ``supLM'' above, but I assume
+  you are generally referring to maxLM for ordinal $V$ vs maxLM for
+  continuous $V$.  See sentences below.}
+As described in detail by Merkle and Zeileis \citeyear{MerZei12}, we
+can also use cumulative score processes with continuous $V$,
+resulting in, e.g., maximum LM statistics and Cram\'{e}r-von Mises
+statistics.  In fact, when the number of potential thresholds is
+large, the proposed~$\max\mathit{LM}_o$ statistic will be very close
+to the $\max\mathit{LM}$ statistic described in Merkle and Zeileis.
+Thus, the formation of ordinal age groups (or other variables) is not
+necessary for testing measurement invariance, and it may be beneficial
+to collect continuous age data (e.g., measured by days instead of by 
+years).
+
+There also exist alternative methods for testing measurement
+invariance w.r.t.\ continuous $V$, including  moderated factor models
+\cite{BauHus09,MolDol10,Pur02} and factor mixture models
 \cite{DolMaa98,LubMut05}.  Under the moderated factor model approach,
@@
 techniques for independent observations that are governed by a central limit
-theorem). \readme{Added the comment in brackets. Useful or too detailed and
-potentially confusing?} The only
+theorem). The only
 requirement for carrying out the tests is that the casewise scores
@@
 whether we should do this now or wait until the revision or so\dots}
+\readme{Achim: I am happy to wait for the revision, and for now have
+  inserted a TODO comment into the Rnw file.}
+
+% TODO Practical remarks about the computation of p-values
 

### 2012-10-16 ordinal/ordinal.Rnw
small touch-ups, removing most readme/fixme improved cite/citeNP/citeA usage, avoiding spelling out author names by hand consistency changes in refs.bib (especially title case)
to do: described m2.2 instead of m3



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@171 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 proposed statistics; for a more detailed
-account, see Merkle and Zeileis \citeyear{MerZei12}.
+account, see \citeA{MerZei12}.
 
@@
 related to those underlying the Lagrange multiplier test and are
-discussed in detail by Merkle and Zeileis \citeyear{MerZei12}. Here,
+discussed in detail by \citeA{MerZei12}. Here,
 we review the tests' properties that are relevant for extending them
@@
 scaled by the asymptotic variance $t (1 - t)$ of the process
-${\bm B}(t, \hat {\bm \theta})$.  In simulations, Merkle and Zeileis
-\citeyear{MerZei12} found 
+${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeA{MerZei12} found 
 that both tests perform well when assessing simultaneous changes
@@
 distribution with standard normal marginals and this particular correlation
-matrix; see also \citeNP{hotzei08} for more details. In R, this can be
+matrix; see also \citeA{hotzei08} for more details. In R, this can be
 accomplished using the mvtnorm package \cite{mvtnorm}.
@@
 at specific levels of the ordinal variable
-(see Merkle and Zeileis, 2012, for related discussion), while the
+(see \citeNP{MerZei12}, for related discussion), while the
 three other tests under consideration make use of this information.
@@
 detect the measurement invariance violation, illustrating the result 
-of Bentler and Bonett \citeyear{Ben80} and others that the
+of \citeA{Ben80} and others that the
 likelihood ratio test statistic picks out minor 
@@
 existing gratitude scales used with adults are valid in research with
-youth. Addressing this issue, \citeNP{FroFan11} had a
+youth. Addressing this issue, \citeA{FroFan11} had a
 large sample of 
@@
 
-\citeNP{FroFan11} used confirmatory factor 
+\citeA{FroFan11} used confirmatory factor 
 models to study the invariance of three youth gratitude scales across
@@
 tests were overly sensitive at their sample size.  To supplement these
-tests, \citeNP{FroFan11} examined a set of alternative fit indices, including
+tests, \citeA{FroFan11} examined a set of alternative fit indices, including
 the non-normed fit index, the comparative fit index, and the
@@
 
-In the analyses described below, we re-analyze the \citeNP{FroFan11} data using
+In the analyses described below, we re-analyze the \citeA{FroFan11} data using
 the ordinal test statistics proposed in this paper.  This results in a
@@
 while being more sensitive to ordinal violations of measurement invariance.
-We focus on two analyses from Froh et al.\ where the likelihood
+We focus on two analyses from \citeA{FroFan11} where the likelihood
 ratio test resulted in significance (indicating that the restricted
@@
 
-Of the 1401 cases originally collected by \citeNP{FroFan11}, we use here all
+Of the 1401 cases originally collected by \citeA{FroFan11}, we use here all
 subjects with complete data (resulting in $n=1327$).
@@
           meanstructure=TRUE)
-gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, 
-              order.by = as.numeric(yg$agegroup), 
+gefp1 <- gefp(m1, fit = NULL, order.by = as.numeric(yg$agegroup), 
               vcov = info_full, sandwich = FALSE, parm=1:4)
@@
 \subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
-model to the GQ-6 data, Froh et al.\ used alternative fit indices to
+model to the GQ-6 data, \citeA{FroFan11} used alternative fit indices to
 conclude that the tau-equivalent model was as good as the congeneric
@@
 
-\fixme{Here and also subsequently, formulations like ``we compare model~A
-and model~B'' are used. I'm not sure whether this sends the correct message.
-If I want to compare exactly these two models, then the LRT is the correct
-thing to do. But we are looking for other types of deviations, especially if
-a split into two (rather than all) groups might be sufficient to capture
-the effect of $V$. Hence, maybe we should explain in some more detail what
-we are testing and also explore other models when rejecting the model? For
-more comments in this vein see below.}
-\readme{Achim: Added two sentences here that hopefully address the issue.}
 We can use the proposed ordinal statistics to assess whether or not
@@
 to age.  Unlike the LRT, the test does not require parameters to
-differ within each subgroup.  Instead, we test for deviations such
-that a split into two subgroups is sufficient to capture the effect of
-$V$.
-In employing the ordinal tests, we obtain~$\mathit{WDM}_o=\Sexpr{round(omax1$statistic, digits = 2)}, p=\Sexpr{round(omax1$p.value, digits = 3)}$
- and~$\max\mathit{LM}_o=\Sexpr{round(ol2bb1$statistic, digits = 2)}, p=\Sexpr{round(ol2bb1$p.value, digits = 3)}$.
+differ across \emph{all} subgroups.  Instead, we test for deviations such
+that a split into two subgroups is sufficient to capture the effect of $V$.
+In employing the ordinal tests, we obtain~$\mathit{WDM}_o=\Sexpr{round(omax1$statistic, digits = 2)},
+p = \Sexpr{format(round(omax1$p.value, digits = 3), nsmall = 3)}$
+and~$\max\mathit{LM}_o=\Sexpr{round(ol2bb1$statistic, digits = 2)},
+p=\Sexpr{round(ol2bb1$p.value, digits = 3)}$.
 Both $p$-values are clearly larger than that of the likelihood ratio test
-and neither is significant at an $\alpha$ of 0.05,
-which supports the conclusions that Froh et al.\ obtained from
+and neither is significant at $\alpha = 0.05$,
+which supports the conclusions that \citeA{FroFan11} obtained from
 alternative fit statistics.  This provides further evidence that 
@@
            meanstructure=TRUE)
-gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, 
-              order.by = as.numeric(yg$agegroup), 
+gefp2 <- gefp(m2, fit = NULL, order.by = as.numeric(yg$agegroup), 
               vcov = info_full, sandwich = FALSE, parm=3:6)
@@
 \subsubsection{GAC} In fitting tau-equivalent and parallel models to the
-GAC data, \citeNP{FroFan11} obtained mixed results.  The alternative fit
+GAC data, \citeA{FroFan11} obtained mixed results.  The alternative fit
 indices did not all agree with one another, and the likelihood ratio test
@@
 model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2], digits = 2)}, p<0.01$
-for the data considered here).  Froh et al.\ ultimately 
+for the data considered here).  \citeA{FroFan11} ultimately 
 concluded that the tau-equivalent model provided a better fit than did
@@
 parameters differ between individuals up to 13 years of age and older
-than 13 years. \fixme{Maybe we could additionally fit a tau-equivalent
+than 13 years. \fixme{Z: Maybe we could additionally fit a tau-equivalent
 model for only two age groups: $\le 13$ vs.\ $> 13$. Then, we could
@@
 than always splitting into all possible age groups.}
-\readme{Achim: See code starting on line 963 of the Rnw file. Even if we
+\readme{ECM: See code starting on line 963 of the Rnw file. Even if we
   split around age 13, the test statistics are all significant.}
+\readme{Z: My formulation above was misleading. We started out from
+model m2 with 24 parameters where variances are restricted to be equal
+across all groups. The LRT compares this to model m2.f with 44 parameters,
+where the group-specific variances for all 6 age groups are estimated.
+This ``costs'' us $(6 - 1) \cdot 4 = 20$ additional parameters. Instead,
+we could fit a model m2.2 with 4 additional parameters -- i.e., 28 in total --
+where the variances are just estimated separately for two age groups (split at age 14).
+This should improve significantly upon m2 but if we are lucky it is not
+significantly worse than m2.f.}
 
@@
 $V$, while the likelihood ratio statistic provides a more powerful test of
-general (non-monotonic) measurement invariance w.r.t.\ $V$. \readme{I
-replace ``sharper'' by ``more powerful'' which would be the correct
-statistical jargon. Or is sharper the terminus technicus in psychometrics?}
-\readme{Achim: No, ``sharper'' was simply my terminus technicus when I
-  was writing this section.}
-While the
+general (non-monotonic) measurement invariance w.r.t.\ $V$. While the
 latter feature may be important in some high-stakes applications, many
@@
 invariance, researchers may wish to treat ordinal $V$ as continuous
-(especially if $V$ has many levels).  
-\fixme{We should also point to the supLM test of our first paper here.
-The maxLM and supLM will generally be very close if the number of
-potential thresholds is large. Maybe we could also re-iterate that
-the forming of age groups a priori might not even be necessary and
-more studies could collect data about age as a continuous rather than
-just an ordinal variable.}
-\readme{Achim: You say both ``maxLM'' and ``supLM'' above, but I assume
-  you are generally referring to maxLM for ordinal $V$ vs maxLM for
-  continuous $V$.  See sentences below.}
-As described in detail by Merkle and Zeileis \citeyear{MerZei12}, we
+(especially if $V$ has very many levels).  
+As described in detail by \citeA{MerZei12}, we
 can also use cumulative score processes with continuous $V$,
@@
 large, the proposed~$\max\mathit{LM}_o$ statistic will be very close
-to the $\max\mathit{LM}$ statistic described in Merkle and Zeileis.
+to the $\max\mathit{LM}$ statistic described in \citeA{MerZei12}.
 Thus, the formation of ordinal age groups (or other variables) is not
 necessary for testing measurement invariance, and it may be beneficial
-to collect continuous age data (e.g., measured by days instead of by 
+to collect continuous age data (e.g., age measured in days rather than in
 years).
@@
 
-\fixme{Should we make some practical remarks about the computation of
-$p$-values somewhere? The issue with the maxLM test is that it is hard
-to tabulate critical values in advance and hence simulation has to be
-done for each case and is potentially rather time-consuming. Not sure
-whether we should do this now or wait until the revision or so\dots}
-\readme{Achim: I am happy to wait for the revision, and for now have
-  inserted a TODO comment into the Rnw file.}
-
-% TODO Practical remarks about the computation of p-values
+% TODO % \fixme{Z: Should we make some practical remarks about the computation of
+% TODO % $p$-values somewhere? The issue with the maxLM test is that it is hard
+% TODO % to tabulate critical values in advance and hence simulation has to be
+% TODO % done for each case and is potentially rather time-consuming. Not sure
+% TODO % whether we should do this now or wait until the revision or so\dots}
+% TODO % \readme{ECM: I am happy to wait for the revision, and for now have
+% TODO %   inserted a TODO comment into the Rnw file.}
 

### 2012-10-21 ordinal/ordinal.Rnw
Revisions to ordinal paper: o Added an intermediate, tau-equivalent model to application section o Added a reference
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@172 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-%% TODO add some more citations to these initial paragraphs
 The study of measurement invariance and differential item functioning
@@
 $V$.  This alternative is implicitly employed in ``median split''
-analyses, % TODO cite preacher
+analyses, 
 where $\nu$ is given as the sample median of $V$.  The threshold $\nu$
@@
 the non-normed fit index, the comparative fit index, and the
-incremental fit index. %% TODO cite
+incremental fit index (e.g., \citeNP{BroCud93}).
 The authors generally found support for the tau-equivalent models
@@
 processes peak at the 12--13 age group, suggesting that
-parameters differ between individuals up to 13 years of age and older
-than 13 years. \fixme{Z: Maybe we could additionally fit a tau-equivalent
-model for only two age groups: $\le 13$ vs.\ $> 13$. Then, we could
-re-run tests and check whether this already fits sufficiently well?
-This could help making the split into two subgroups clearer, rather
-than always splitting into all possible age groups.}
-\readme{ECM: See code starting on line 963 of the Rnw file. Even if we
-  split around age 13, the test statistics are all significant.}
-\readme{Z: My formulation above was misleading. We started out from
-model m2 with 24 parameters where variances are restricted to be equal
-across all groups. The LRT compares this to model m2.f with 44 parameters,
-where the group-specific variances for all 6 age groups are estimated.
-This ``costs'' us $(6 - 1) \cdot 4 = 20$ additional parameters. Instead,
-we could fit a model m2.2 with 4 additional parameters -- i.e., 28 in total --
-where the variances are just estimated separately for two age groups (split at age 14).
-This should improve significantly upon m2 but if we are lucky it is not
-significantly worse than m2.f.}
-
-<<gac.2grp>>=
-yg <- transform(yg, under14=as.numeric(agegroup)<=2)
-
-m3 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
-           f1 ~ 0*1', data=yg, group="under14",
+parameters differ between individuals up to 13 years of age and
+individuals older 
+than 13 years of age.
+
+<<gac.mod>>=
+m2.2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
+             f1 ~ 0*1
+             gac_1 ~~ c("v1.1","v1.1","v1.2","v1.2","v1.2","v1.2")*gac_1
+             gac_2 ~~ c("v2.1","v2.1","v2.2","v2.2","v2.2","v2.2")*gac_2
+             gac_3 ~~ c("v3.1","v3.1","v3.2","v3.2","v3.2","v3.2")*gac_3
+             f1 ~~ c("v4.1","v4.1","v4.2","v4.2","v4.2","v4.2")*f1', 
+             data=yg, group="agegroup",
              group.equal=c("loadings","residuals","lv.variances"), 
              meanstructure=TRUE)
-gefp3 <- gefp(m3, fit = NULL, scores=estfun.lavaan, 
-              order.by = as.numeric(yg$under14), 
-              vcov = info_full, sandwich = FALSE, parm=3:6)
-if(file.exists("ol2bb3.rda")){
-  load("ol2bb3.rda")
+
+gefp2.2 <- gefp(m2.2, fit = NULL, scores=estfun.lavaan, 
+                order.by = as.numeric(yg$agegroup), 
+                vcov = info_full, sandwich = FALSE, parm=c(3:6,13:16))
+
+if(file.exists("ol2bb22.rda")){
+  load("ol2bb22.rda")
 } else {
-  ol2bb3.fun <- ordL2BB(gefp3)
-  save(ol2bb3.fun, file="ol2bb3.rda")
+  ol2bb22.fun <- ordL2BB(gefp2.2)
+  save(ol2bb22.fun, file="ol2bb22.rda")
 }
-omax3 <- sctest(gefp3, functional=ordwmax(gefp3))
-ol2bb3 <- sctest(gefp3, functional=ol2bb3.fun)
-
-## Full model
-m3.f <- cfa('f1 =~ gac_1 + gac_2 + gac_3
-           f1 ~ 0*1', data=yg, group="under14",
-           group.equal="loadings", meanstructure=TRUE)
-lrt3 <- anova(m3.f, m3)
+
+omax2.2 <- sctest(gefp2.2, functional=ordwmax(gefp2.2))
+ol2bb2.2 <- sctest(gefp2.2, functional=ol2bb22.fun)
+lrt2.2par <- anova(m2.2, m2)
+lrt2.2te <- anova(m2.f, m2.2)
 @
+
+The finding that variance parameters differ between individuals up to 13 years and individuals over 13 years is reinforced by comparing the
+tau-equivalent and parallel models to an intermediate model.  This intermediate model is tau-equivalent in nature, but there exist only two groups: individuals up to 13 years
+and individuals older than 13 years.  A likelihood ratio test implies that
+this intermediate model fits as well as the original tau-equivalent model~($\chi^2_{16} = \Sexpr{round(lrt2.2te[[5]][2], digits = 2)}$, $p=\Sexpr{round(lrt2.2te[[7]][2], digits=2)}$), with 16 fewer parameters.
+The intermediate model also fits
+better than the parallel model, as judged by a second likelihood ratio test~($\chi^2_{4} = \Sexpr{round(lrt2.2par[[5]][2], digits = 2)}, p<.01$).
+Finally, using the proposed ordinal test statistics with the intermediate model, we no 
+longer observe instability in the variance parameters
+($\mathit{WDM}_o=\Sexpr{round(omax2.2$statistic, digits = 2)}, p=\Sexpr{round(omax2.2$p.value, digits = 2)}$; 
+$\max\mathit{LM}_o=\Sexpr{round(ol2bb2.2$statistic, digits = 2)}, p=\Sexpr{round(ol2bb2.2$p.value, digits = 2)}$).
 

### 2012-10-21 ordinal/ordinal.Rnw
added few details
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@174 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 and individuals older than 13 years.  A likelihood ratio test implies that
-this intermediate model fits as well as the original tau-equivalent model~($\chi^2_{16} = \Sexpr{round(lrt2.2te[[5]][2], digits = 2)}$, $p=\Sexpr{round(lrt2.2te[[7]][2], digits=2)}$), with 16 fewer parameters.
-The intermediate model also fits
-better than the parallel model, as judged by a second likelihood ratio test~($\chi^2_{4} = \Sexpr{round(lrt2.2par[[5]][2], digits = 2)}, p<.01$).
+this intermediate model fits as well as the original tau-equivalent model~($\chi^2_{16} = \Sexpr{round(lrt2.2te[[5]][2], digits = 2)}$, $p=\Sexpr{round(lrt2.2te[[7]][2], digits=2)}$),
+with 16 fewer parameters ($= 6 \cdot 4 - 2 \cdot 4$ because the four variances have to be estimated only in two rather than in six age groups).  The intermediate model also fits 
+better than the parallel model, as judged by a second likelihood ratio test~($\chi^2_{4} = \Sexpr{format(round(lrt2.2par[[5]][2], digits = 2), nsmall = 2)},
+p<0.01$).
 Finally, using the proposed ordinal test statistics with the intermediate model, we no 
-longer observe instability in the variance parameters
+longer observe further instability in the variance parameters
 ($\mathit{WDM}_o=\Sexpr{round(omax2.2$statistic, digits = 2)}, p=\Sexpr{round(omax2.2$p.value, digits = 2)}$; 

### 2012-10-23 ordinal/ordinal.Rnw
touch-ups in code to - use YouthGratitude from psychotools package - set seeds for all ordL2BB objects - prettify code for subsequent extraction to replication-ordinal.R
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@178 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 library("lattice")
-library("foreign")
 
 ## auxiliary code
-source("mzo.R")
+source("../www/mz-ordinal.R")
 source("../www/estfun-lavaan.R")
-source("../prelim_code/efpFunctional-cat.R")
-source("sim_ordinal.R")
+source("../www/efpFunctional-cat.R")
+source("../www/sim-ordinal.R")
 
@@
 } else {
-  sim2 <- simulation(diff = seq(0, 3, by = 0.5), nobs = c(1200, 4800, 9600),
-                     anomaly=TRUE)
+  sim2 <- simulation(diff = seq(0, 3, by = 0.5), nobs = c(1200, 4800, 9600), anomaly = TRUE)
   sim2$nlevels <- factor(sim2$nlevels)
@@
 
-<<frohdata>>=
-yg <- read.spss("../prelim_code/grat_data_combined_2009-09-29.sav", to.data.frame=TRUE)
-names(yg) <- tolower(names(yg))
-
-## Remove cases with 'imputed' values
-yg <- yg[yg$ao_4 <= 9,]
-decimals <- apply(yg[,4:28]%%1 > 0, 1, sum)
-yg <- yg[decimals==0,]
+<<frohetal-data>>=
+data("YouthGratitude", package = "psychotools")
+## remove cases with 'imputed' values (not in 1, ..., 9)
+yg <- YouthGratitude[apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9)), ]
 @
@@
 <<gq6>>=
-m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
-          data=yg, group="agegroup", group.equal="loadings",
-          meanstructure=TRUE)
+m1 <- cfa(
+  'f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+  data = yg, group = "agegroup", meanstructure=TRUE,
+  group.equal = "loadings")
 gefp1 <- gefp(m1, fit = NULL, order.by = as.numeric(yg$agegroup), 
@@
 } else {
+  set.seed(1163)
   ol2bb1.fun <- ordL2BB(gefp1)
-  save(ol2bb1.fun, file="ol2bb1.rda")
+  save(ol2bb1.fun, file = "ol2bb1.rda")
 }
-omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
-ol2bb1 <- sctest(gefp1, functional=ol2bb1.fun)
+omax1 <- sctest(gefp1, functional = ordwmax(gefp1))
+ol2bb1 <- sctest(gefp1, functional = ol2bb1.fun)
 
 ## Full model
-m1.f <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
-            data=yg, group="agegroup", meanstructure=TRUE)
+m1.f <- cfa(
+  'f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+  data = yg, group = "agegroup", meanstructure = TRUE)
 lrt1 <- anova(m1.f, m1)
@@
 <<gq6res, fig=TRUE, height=3, width=6>>=
-par(mfrow=c(1, 2), cex = 0.65)
-
-plot(gefp1, functional=ordwmax(gefp1), axes=FALSE, main="",
-     xlab="Age group", ylab="Weighted max statistics")
-axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
+par(mfrow = c(1, 2), cex = 0.65)
+
+plot(gefp1, functional = ordwmax(gefp1), axes = FALSE, main = "",
+     xlab = "Age group", ylab = "Weighted max statistics")
+axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
 axis(2)
 
-plot(gefp1, functional=ol2bb1.fun, axes=FALSE, ylim=c(0,14), main="",
-     xlab="Age group", ylab="LM statistics")
-axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
+plot(gefp1, functional = ol2bb1.fun, axes = FALSE, ylim = c(0, 14), main = "",
+     xlab = "Age group", ylab = "LM statistics")
+axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
 axis(2)
@@
 <<gac>>=
-m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
-           f1 ~ 0*1', data=yg, group="agegroup",
-           group.equal=c("loadings","residuals","lv.variances"), 
-           meanstructure=TRUE)
+m2 <- cfa(
+  'f1 =~ gac_1 + gac_2 + gac_3
+   f1 ~ 0*1',
+  data = yg, group = "agegroup", meanstructure = TRUE,
+  group.equal = c("loadings", "residuals", "lv.variances"))
+
 gefp2 <- gefp(m2, fit = NULL, order.by = as.numeric(yg$agegroup), 
-              vcov = info_full, sandwich = FALSE, parm=3:6)
+  vcov = info_full, sandwich = FALSE, parm = 3:6)
 if(file.exists("ol2bb2.rda")){
@@
 } else {
+  set.seed(1163)
   ol2bb2.fun <- ordL2BB(gefp2)
-  save(ol2bb2.fun, file="ol2bb2.rda")
+  save(ol2bb2.fun, file = "ol2bb2.rda")
 }
-omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
-ol2bb2 <- sctest(gefp2, functional=ol2bb2.fun)
+omax2 <- sctest(gefp2, functional = ordwmax(gefp2))
+ol2bb2 <- sctest(gefp2, functional = ol2bb2.fun)
 
 ## Full model
-m2.f <- cfa('f1 =~ gac_1 + gac_2 + gac_3
-           f1 ~ 0*1', data=yg, group="agegroup",
-           group.equal="loadings", meanstructure=TRUE)
+m2.f <- cfa(
+  'f1 =~ gac_1 + gac_2 + gac_3
+   f1 ~ 0*1',
+  data = yg, group = "agegroup", meanstructure = TRUE,
+  group.equal = "loadings")
 lrt2 <- anova(m2.f, m2)
@@
 <<gacres, fig=TRUE, height=3, width=6>>=
-par(mfrow=c(1, 2), cex = 0.65)
-
-plot(gefp2, functional=ordwmax(gefp2), axes=FALSE, main="",
-     xlab="Age group", ylab="Weighted max statistics")
-axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
+par(mfrow = c(1, 2), cex = 0.65)
+
+plot(gefp2, functional = ordwmax(gefp2), axes = FALSE, main = "",
+     xlab = "Age group", ylab = "Weighted max statistics")
+axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
 axis(2)
 
-plot(gefp2, functional=ol2bb2.fun, axes=FALSE, ylim=c(0,115), main="",
-     xlab="Age group", ylab="LM statistics")
-axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
+plot(gefp2, functional = ol2bb2.fun, axes = FALSE, ylim = c(0, 115), main = "",
+     xlab = "Age group", ylab = "LM statistics")
+axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
 axis(2)
@@
 <<gac.mod>>=
-m2.2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
+m2.2 <- cfa(
+  'f1 =~ gac_1 + gac_2 + gac_3
    f1 ~ 0*1
@@
    f1 ~~ c("v4.1","v4.1","v4.2","v4.2","v4.2","v4.2")*f1', 
-             data=yg, group="agegroup",
-             group.equal=c("loadings","residuals","lv.variances"), 
-             meanstructure=TRUE)
-
-gefp2.2 <- gefp(m2.2, fit = NULL, scores=estfun.lavaan, 
-                order.by = as.numeric(yg$agegroup), 
-                vcov = info_full, sandwich = FALSE, parm=c(3:6,13:16))
+  data = yg, group = "agegroup", meanstructure = TRUE,
+  group.equal = c("loadings", "residuals", "lv.variances"))
+
+gefp2.2 <- gefp(m2.2, fit = NULL, order.by = as.numeric(yg$agegroup), 
+  vcov = info_full, sandwich = FALSE, parm = c(3:6, 13:16))
 
@@
 } else {
+  set.seed(1163)
   ol2bb22.fun <- ordL2BB(gefp2.2)
@@
 
-omax2.2 <- sctest(gefp2.2, functional=ordwmax(gefp2.2))
-ol2bb2.2 <- sctest(gefp2.2, functional=ol2bb22.fun)
+omax2.2 <- sctest(gefp2.2, functional = ordwmax(gefp2.2))
+ol2bb2.2 <- sctest(gefp2.2, functional = ol2bb22.fun)
 lrt2.2par <- anova(m2.2, m2)

### 2012-10-23 www/sim-ordinal.R
updated web page -> moved all .R files necessary for replication to www/ directory
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@179 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -179,51 +179,3 @@ critvals <- function(nlevels = c(4, 8, 12), verbose = TRUE, ...)
 
   save(cval, file="critvals.rda")
 }
-
-
-if(FALSE) {
-## code and packages
-library("lavaan")
-library("strucchange")
-library("mvtnorm")
-source("mzo.R")
-source("../www/estfun-lavaan.R")
-source("../prelim_code/efpFunctional-cat.R")
-
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1090)
-
-## run simulation
-sim1 <- simulation()
-save(sim1, file = "sim1.rda")
-}
-
-if(FALSE) {
-## load simulation results
-load("sim1.rda")
-
-sim1$nlevels <- paste("m=",sim1$nlevels,sep="")
-sim1$nlevels <- factor(sim1$nlevels, levels=c("m=4","m=8","m=12"))
-levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
-levels(sim1$test) <- c("LM_uo","LRT","LM_o","WDM_o")
-
-## display result in table
-## all without rounding
-ftable(xtabs(power ~ nobs + pars + test + diff, data = mz_sim), col.vars = "diff")
-## subset with rounding, in percent
-round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
-  data = mz_sim, subset = diff %in% c(seq(0, 3.5, by = 0.5))),
-  col.vars = "diff"), digits = 1)
-
-## display result in graphic
-library("lattice")
-trellis.par.set(theme = canonical.theme(color = FALSE))
-
-mykey <- simpleKey(levels(sim1$test), points = FALSE, lines = TRUE)
-#mykey$lines$lty <- c(2, 3, 1)
-#mykey$lines$lwd <- c(1.2, 2, 1.2)
-## all
-xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1, type = "b",
-       xlab="Violation Magnitude", ylab="Power", key=mykey)
-}

### 2012-10-23 www/replication-ordinal.R
added replication script for ordinal paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@180 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,193 @@
+###################
+## Preliminaries ##
+###################
+
+## packages
+library("lavaan")
+library("strucchange")
+library("mvtnorm")
+library("lattice")
+
+## auxiliary code
+source("mz-ordinal.R")
+source("estfun-lavaan.R")
+source("efpFunctional-cat.R")
+source("sim-ordinal.R")
+
+## convenience function for plotting boundaries without color
+get_boundary <- function(obj, fun, trim = 0) {
+  bound <- fun$boundary(0:obj$nobs/obj$nobs)
+  bound <- fun$computeCritval(0.05, ncol(obj$process)) * bound
+  bound <- zoo(bound, time(obj$process))
+  if(trim > 0) bound <- head(tail(bound, - floor(trim * obj$nobs)), - floor(trim * obj$nobs))
+  return(bound)
+}
+
+## convenience function for computing information matrix
+info_full <- function(x, ...) solve(vcov(x) * nobs(x))
+
+
+##################
+## Simulation 1 ##
+##################
+
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+## run simulation
+sim1 <- simulation()
+
+## change labels for plotting
+sim1$nlevels <- factor(sim1$nlevels)
+levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
+sim1$nobs <- factor(sim1$nobs)
+levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
+levels(sim1$test) <- c("LM_uo", "LRT", "LM_o", "WDM_o")
+sim1$test <- factor(as.character(sim1$test),
+  levels = c("LM_o", "WDM_o", "LRT", "LM_uo"),
+  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo"))
+
+## visualization
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1, 
+  type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey)
+
+
+##################
+## Simulation 2 ##
[…]

### 2012-10-23 ordinal/report.Rnw
prepared report version of ordinal paper (using jss.cls)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@181 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2012-10-23 ordinal/ordinal.Rnw
@@
-\documentclass[man]{apa}
-\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
+\documentclass[nojss]{jss}
+\usepackage{amsmath,alltt,setspace,bm,thumbpdf}
 %% need no \usepackage{Sweave}
 \SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}
-
-\title{Testing for measurement invariance with respect to an ordinal variable}
-\threeauthors{Edgar C.\ Merkle}{Jinyan Fan}{Achim Zeileis}
-\threeaffiliations{University of Missouri}{Auburn University}{Universit\"{a}t Innsbruck}
-
-\abstract{
+\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+
+\title{Testing for Measurement Invariance with Respect to an Ordinal Variable}
+\Shorttitle{Testing for Ordinal Measurement Invariance}
+
+\author{Edgar C. Merkle\\University of Missouri
+   \And Jinyan Fan\\Auburn University
+   \And Achim Zeileis\\Universit\"{a}t Innsbruck}
+\Plainauthor{Edgar C. Merkle, Jinyan Fan, Achim Zeileis}
+
+\Abstract{
   Researchers are often interested in testing for 
@@
 
-\acknowledgements{This work was supported by National Science
-  Foundation grant SES-1061334. 
-  Correspondence to Edgar C.\ Merkle, Department of  
-  Psychological Sciences, University of Missouri, Columbia,
-  MO 65211.
-  Email: {\texttt{merklee@missouri.edu}}.}
-\shorttitle{Ordinal measurement invariance}
-\rightheader{Ordinal measurement invariance}
-
-\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
-
-%% for internal use
-\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
-\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}
-
-\spacing{1}
+\Keywords{measurement invariance, ordinal variable, parameter stability, factor analysis, structural equation models}
+
+\Address{
+  Edgar C. Merkle\\
+  Department of Psychological Sciences\\
+  University of Missouri\\
+  Columbia, MO 65211, United States of America\\
+  E-mail: \email{merklee@missouri.edu}\\
+
+  Jinyan Fan\\
+  Department of Psychology\\
+  Auburn University\\
+  225 Thach\\
+  Auburn, AL 36849-5214, United States of America\\
+  E-mail: \email{jzf0007@auburn.edu}\\
+
+  Achim Zeileis\\
+  Department of Statistics\\
+  Faculty of Economics and Statistics\\
+  Universit\"at Innsbruck\\
+  Universit\"atsstr.~15\\
+  6020 Innsbruck, Austria\\
+  E-mail: \email{Achim.Zeileis@R-project.org}\\
+  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
+}
 
 \begin{document}
-\maketitle
 
@@
 
+\section{Introduction}
+
 The study of measurement invariance and differential item functioning
@@
 has received considerable attention in the psychometric literature
-(see, e.g., \citeNP{Mil11} for a thorough review).  A set of
+(see, e.g., \citealp{Mil11}, for a thorough review).  A set of
 psychometric scales $X$ are defined to be measurement invariant
-with respect to an auxiliary variable $V$ if \cite{Mel89}
+with respect to an auxiliary variable $V$ if \citep{Mel89}
 \begin{equation}
@@
 vary across different ethnicities.  Statistical tools that can be used
-to carry out this study include Likelihood Ratio tests, Lagrange
-Multiplier tests, and Wald tests (e.g., \citeNP{Sat89}).
+to carry out this study include likelihood ratio tests, Lagrange
+multiplier tests, and Wald tests (e.g., \citealp{Sat89}).
 These tools have greatly aided in
@@
 that were recently applied to the study of measurement invariance in
-psychometric models \cite{KopZei10,MerZei12}.
+psychometric models \citep{KopZei10,MerZei12}.
 In the following section, we provide an overview of the family and
@@
 
-\section{Theoretical Detail}
+\section{Theoretical detail}
 
@@
 proposed statistics; for a more detailed
-account, see \citeA{MerZei12}.
+account, see \citet{MerZei12}.
 
@@
 though the proposed tests extend beyond this family of models.  Under
-the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
+the usual regularity conditions (e.g., \citealp{Fer96}), the model parameters
 $\bm{\theta}$ can 
@@
 
-\subsection{Tests for Continuous $V$}
+\subsection{Tests for continuous $V$}
 
@@
 related to those underlying the Lagrange multiplier test and are
-discussed in detail by \citeA{MerZei12}. Here,
+discussed in detail by \citet{MerZei12}. Here,
 we review the tests' properties that are relevant for extending them
@@
 and the usual ML regularity conditions hold, it is possible to show
-that \cite{HjoKon02}
+that \citep{HjoKon02}
 \begin{equation} \label{eq:fclt}
@@
 obtained by applying the 
-same functional to the Brownian bridge \cite{HjoKon02,ZeiHor07}.
+same functional to the Brownian bridge \citep{HjoKon02,ZeiHor07}.
 
@@
 invariance when multiple factor analysis parameters change
-simultaneously \cite{MerZei12}.
+simultaneously \citep{MerZei12}.
 
@@
 scaled by the asymptotic variance $t (1 - t)$ of the process
-${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeA{MerZei12} found 
+${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citet{MerZei12} found 
 that both tests perform well when assessing simultaneous changes
@@
 
-\subsection{Proposed Tests for Ordinal $V$}
+\subsection{Proposed tests for ordinal $V$}
 
@@
 
-For the $\mathit{WDM}_o$ statistic, the resulting asymptotic distribution is
+For the $\mathit{WDM}_o$ statistic, the resulting asymptotic distribution is\linebreak
 $\max_{j = 1, \dots, k} \max_{\ell = 1, \dots, m-1} {\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$.
@@
 distribution with standard normal marginals and this particular correlation
-matrix; see also \citeA{hotzei08} for more details. In R, this can be
-accomplished using the mvtnorm package \cite{mvtnorm}.
+matrix; see also \citet{hotzei08} for more details. In R, this can be
+accomplished using the mvtnorm package \citep{mvtnorm}.
 
@@
 simulation of Brownian bridges.  This functionality is built in to
-R's strucchange package \cite{Zei06}, which can be used to
+R's strucchange package \citep{Zei06}, which can be used to
 generally carry out the tests. Note that for models with only a single
@@
 each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
-statistic \cite{HjoKon02}.  This test statistic can be formally written as
+statistic \citep{HjoKon02}.  This test statistic can be formally written as
 \begin{equation}
@@
 likelihood ratio test is notoriously sensitive to minor parameter
-instabilities (e.g., \citeNP{Ben80}).
-
-\section{Simulation 1: Detecting Ordinal Invariance Violations}
+instabilities (e.g., \citealp{Ben80}).
+
+\section{Simulation 1: Detecting ordinal invariance violations}
 
@@
 comparing a tau-equivalent model to a congeneric model (e.g.,
-\citeNP{VanLan00}).
+\citealp{VanLan00}).
 
@@
 at specific levels of the ordinal variable
-(see \citeNP{MerZei12}, for related discussion), while the
+(see \citealp{MerZei12}, for related discussion), while the
 three other tests under consideration make use of this information.
@@
 
-\begin{figure}
-\caption{Simulated power curves for the ordered and unordered
-      $\max \mathit{LM}$ tests, the ordered double-max test, and the
-      likelihood ratio test across three sample sizes $n$, three
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
-\label{fig:sim1res}
-\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\setkeys{Gin}{width=0.95\textwidth}
 <<sim1res, fig=TRUE, height=7, width=7>>=
@@
 @
+\caption{\label{fig:sim1res}
+      Simulated power curves for the ordered and unordered
+      $\max \mathit{LM}$ tests, the ordered double-max test, and the
+      likelihood ratio test across three sample sizes $n$, three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
 \end{figure}
@@
 
-\section{Simulation 2: Minor Anomalies and Large $n$}
+\section{Simulation 2: Minor anomalies and large $n$}
 
@@
 show that the proposed ordinal tests remain viable for large $n$.
-
-\subsection{Method}
-Data were generated from the same factor analysis model that was used
-in Simulation 1, with measurement invariance violations in the unique
-variance parameters.  To implement a minor measurement invariance
-violation, the unique variances were equal across all levels of the
-ordinal variable except one (level $1+m/2$).  At this particular
-level, the unique variances were greater by a factor of $d$~times the
-parameters' asymptotic standard errors (scaled by $\sqrt{n}$), with
-$d=0, 0.5, 1.0, \dots, 3.0$.  The number of levels of the ordinal
-variable were the same as those in Simulation 1 ($m=4,8,12$), and
-sample sizes were set at $n=1200,4800,9600$.  All other simulation
-features match those from Simulation 1.
 
@@
 
+\begin{figure}[t!]
+\setkeys{Gin}{width=0.95\textwidth}
+<<sim2res, fig=TRUE, height=7, width=7>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+\caption{\label{fig:sim2res}
+      Simulated power curves for the ordered and unordered
+      $\max \mathit{LM}$ tests, the ordered double-max test, and the
+      likelihood ratio test across three sample sizes $n$, three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
+      occurring at a single level (the $(1 +
+      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
+\end{figure}
+
+\subsection{Method}
+Data were generated from the same factor analysis model that was used
+in Simulation 1, with measurement invariance violations in the unique
+variance parameters.  To implement a minor measurement invariance
+violation, the unique variances were equal across all levels of the
+ordinal variable except one (level $1+m/2$).  At this particular
+level, the unique variances were greater by a factor of $d$~times the
+parameters' asymptotic standard errors (scaled by $\sqrt{n}$), with
+$d=0, 0.5, 1.0, \dots, 3.0$.  The number of levels of the ordinal
+variable were the same as those in Simulation 1 ($m=4,8,12$), and
+sample sizes were set at $n=1200,4800,9600$.  All other simulation
+features match those from Simulation 1.
+
 \subsection{Results}
@@
 detect the measurement invariance violation, illustrating the result 
-of \citeA{Ben80} and others that the
+of \citet{Ben80} and others that the
 likelihood ratio test statistic picks out minor 
@@
 invariance violation that occurs only at a single level of $V$.
-
-\begin{figure}
-\caption{Simulated power curves for the ordered and unordered
-      $\max \mathit{LM}$ tests, the ordered double-max test, and the
-      likelihood ratio test across three sample sizes $n$, three
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
-      occurring at a single level (the $(1 +
-      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
-\label{fig:sim2res}
-\setkeys{Gin}{width=\textwidth}
-<<sim2res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-\end{figure}
 
@@
 
-\section{Application: Youth Gratitude}
+\section{Application: Youth gratitude}
 
@@
 With the positive psychology movement, the construct of gratitude has
-received much research attention (for a review, see \citeNP{EmmMcc04}).
+received much research attention (for a review, see \citealp{EmmMcc04}).
 Recently, scholars have begun to explore gratitude
@@
 existing gratitude scales used with adults are valid in research with
-youth. Addressing this issue, \citeA{FroFan11} had a
+youth. Addressing this issue, \citet{FroFan11} had a
 large sample of 
@@
 adult gratitude inventories, including Gratitude Questionnaire~6
-(GQ-6; \citeNP{MccEmm02}), Gratitude Adjective
-Checklist (GAC; \citeNP{MccEmm02}), and Gratitude, Resentment,
+(GQ-6; \citealp{MccEmm02}), Gratitude Adjective
+Checklist (GAC; \citealp{MccEmm02}), and Gratitude, Resentment,
 Appreciation Test-Short Form (GRAT-Short Form;
-\citeNP{ThoWat03}). The authors were interested in whether the
+\citealp{ThoWat03}). The authors were interested in whether the
 youth factor structure 
@@
 
-\citeA{FroFan11} used confirmatory factor 
+\citet{FroFan11} used confirmatory factor 
 models to study the invariance of three youth gratitude scales across
@@
 tests were overly sensitive at their sample size.  To supplement these
-tests, \citeA{FroFan11} examined a set of alternative fit indices, including
+tests, \citet{FroFan11} examined a set of alternative fit indices, including
 the non-normed fit index, the comparative fit index, and the
-incremental fit index (e.g., \citeNP{BroCud93}).
+incremental fit index (e.g., \citealp{BroCud93}).
 The authors generally found support for the tau-equivalent models
@@
 
-In the analyses described below, we re-analyze the \citeA{FroFan11} data using
+In the analyses described below, we re-analyze the \citet{FroFan11} data using
 the ordinal test statistics proposed in this paper.  This results in a
@@
 while being more sensitive to ordinal violations of measurement invariance.
-We focus on two analyses from \citeA{FroFan11} where the likelihood
+We focus on two analyses from \citet{FroFan11} where the likelihood
 ratio test resulted in significance (indicating that the restricted
@@
 
-Of the 1401 cases originally collected by \citeA{FroFan11}, we use here all
+Of the 1401 cases originally collected by \citet{FroFan11}, we use here all
 subjects with complete data (resulting in $n=1327$).
@@
 
-\begin{figure}
-\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
-  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GQ-6 data.}
-\label{fig:gq6res}
+\begin{figure}[t!]
 \setkeys{Gin}{width=\textwidth}
@@
 @
+\caption{\label{fig:gq6res}
+  Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
+  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GQ-6 data.}
 \end{figure}
@@
 \subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
-model to the GQ-6 data, \citeA{FroFan11} used alternative fit indices to
+model to the GQ-6 data, \citet{FroFan11} used alternative fit indices to
 conclude that the tau-equivalent model was as good as the congeneric
@@
 and neither is significant at $\alpha = 0.05$,
-which supports the conclusions that \citeA{FroFan11} obtained from
+which supports the conclusions that \citet{FroFan11} obtained from
 alternative fit statistics.  This provides further evidence that 
@@
 
-\begin{figure}
-\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
-  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GAC data.}
-\label{fig:gacres}
+\begin{figure}[t!]
 \setkeys{Gin}{width=\textwidth}
@@
 @
+\caption{\label{fig:gacres}
+  Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
+  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GAC data.}
 \end{figure}
@@
 \subsubsection{GAC} In fitting tau-equivalent and parallel models to the
-GAC data, \citeA{FroFan11} obtained mixed results.  The alternative fit
+GAC data, \citet{FroFan11} obtained mixed results.  The alternative fit
 indices did not all agree with one another, and the likelihood ratio test
@@
 model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2], digits = 2)}, p<0.01$
-for the data considered here).  \citeA{FroFan11} ultimately 
+for the data considered here).  \citet{FroFan11} ultimately 
 concluded that the tau-equivalent model provided a better fit than did
@@
 
-\section{General Discussion}
+\section{General discussion}
 
@@
 
-\subsection{Choice of Test}
+\subsection{Choice of test}
 
@@
 (especially if $V$ has very many levels).  
-As described in detail by \citeA{MerZei12}, we
+As described in detail by \citet{MerZei12}, we
 can also use cumulative score processes with continuous $V$,
@@
 large, the proposed~$\max\mathit{LM}_o$ statistic will be very close
-to the $\max\mathit{LM}$ statistic described in \citeA{MerZei12}.
+to the $\max\mathit{LM}$ statistic described in \citet{MerZei12}.
 Thus, the formation of ordinal age groups (or other variables) is not
@@
 invariance w.r.t.\ continuous $V$, including  moderated factor models
-\cite{BauHus09,MolDol10,Pur02} and factor mixture models
-\cite{DolMaa98,LubMut05}.  Under the moderated factor model approach,
+\citep{BauHus09,MolDol10,Pur02} and factor mixture models
+\citep{DolMaa98,LubMut05}.  Under the moderated factor model approach,
 $V$ is inserted directly into the factor analysis model and allowed to
@@
 
-\subsection{Extension to Other Models}
+\subsection{Extension to other models}
 We focused on testing for measurement invariance in factor
@@
 proposed test statistics are better than the LRT for detecting
-monotonic instabilities.   The strucchange package \cite{Zei06} noted
+monotonic instabilities.   The strucchange package \citep{Zei06} noted
 previously can be used for these more-general applications.
-
-% TODO % \fixme{Z: Should we make some practical remarks about the computation of
-% TODO % $p$-values somewhere? The issue with the maxLM test is that it is hard
-% TODO % to tabulate critical values in advance and hence simulation has to be
-% TODO % done for each case and is potentially rather time-consuming. Not sure
-% TODO % whether we should do this now or wait until the revision or so\dots}
-% TODO % \readme{ECM: I am happy to wait for the revision, and for now have
-% TODO %   inserted a TODO comment into the Rnw file.}
 
@@
 are feasible in situations where the likelihood ratio test commonly
-rejects $H_0$ in practice (e.g., \citeNP{Ben80}).  Furthermore, the
+rejects $H_0$ in practice (e.g., \citealp{Ben80}).  Furthermore, the
 focal psychometric model does not have to be modified in any way,
@@
 
-\section*{Computational Details}
-
-All results were obtained using the {R}~system for statistical computing \cite{R11},
+\section*{Computational details}
+
+All results were obtained using the {R}~system for statistical computing \citep{R11},
 version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on package
-{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} for fitting of the factor analysis models and
-{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
+{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \citep{lavaan11} for fitting of the factor analysis models and
+{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \citep{ZeiLei02,Zei06}
 for evaluating the parameter instability tests. {R}~and the packages {lavaan} and {strucchange} are
@@
 
+\section*{Acknowledgments}
+
+This work was supported by National Science Foundation grant SES-1061334.
+
 \bibliography{refs}

### 2013-02-19 ordinal/ordinal.Rnw
Modified ordinal simulation code to add new statistics: o Yuan-Bentler adjusted WLS statistic o Satorra-Bentler adjusted statistic o AIC
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@190 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 source("../www/mz-ordinal.R")
-source("../www/estfun-lavaan.R")
+## TODO Change this:
+source("../sandbox/yves/estfun.R")
 source("../www/efpFunctional-cat.R")
@@
 vary across different ethnicities.  Statistical tools that can be used
-to carry out this study include Likelihood Ratio tests, Lagrange
-Multiplier tests, and Wald tests (e.g., \citeNP{Sat89}).
+to carry out this study include likelihood ratio tests, Lagrange
+multiplier tests, and Wald tests (e.g., \citeNP{Sat89}).
 These tools have greatly aided in
@@
   levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
-  levels(sim1$test) <- c("LM_uo", "LRT", "LM_o", "WDM_o")
+  levels(sim1$test) <- c("AIC","LM_uo", "LRT", "LRT_sb", "LM_o", "WDM_o",
+                         "YB97")
   save(sim1, file="sim1.rda")
@@
 sim1$test <- factor(as.character(sim1$test),
-  levels = c("LM_o", "WDM_o", "LRT", "LM_uo"),
-  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo"))
+  levels = c("LM_o", "WDM_o", "LRT", "LM_uo", "AIC", "LRT_sb", "YB97"),
+  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo", "AIC", "LRT_sb", "YB97"))
 @

### 2013-02-19 www/mz-ordinal.R

@@ -13,9 +13,9 @@ ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
            verbal ~ 0*1; math ~ 0*1'
 
   m2 <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
-            group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances")))
+            group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances"), test="Satorra.Bentler"))
   m3 <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
-            group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances","residuals","means")))
+            group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances","residuals","means"), test="Satorra.Bentler"))
 
   ## Get test stats at values of m3, for Satorra-Bentler 2010 correction
   ## for difference tests.  This is currently difficult in lavaan due to
@@ -42,48 +42,61 @@ ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
   ##    the charts.
   ## NB It seems that these models do not converge as often
   ##    at smaller n.  The simulations will gloss over this.
-  m2.wls <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
-            group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances"), estimator="WLS"))
+  zinv <- try(get.zcov(data))
+  if(!inherits(zinv, "try-error")){
+    m2.wls <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
+                      group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances"), estimator="WLS", WLS.V=zinv))
 
-  m3.wls <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
-            group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances","residuals","means"), estimator="WLS"))
[…]

### 2013-02-19 www/sim-ordinal.R

@@ -106,8 +106,9 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, test = NULL, verb
       pval[i, 3] <- sctest(ord_gefp,  functional = catL2BB(ord_gefp))$p.value
       #pval[i, 4] <- sctest(ord_gefp,  functional = supLM(0.1))$p.value
       pval[i, 4] <- mz$lrt.p
-      pval[i, 5] <- mz$yb.p
-      pval[i, 6] <- mz$aic
+      pval[i, 5] <- mz$lrt.sb
+      pval[i, 6] <- mz$yb.p
+      pval[i, 7] <- mz$aic
     }
   }
   rval <- colMeans(pval < size, na.rm = TRUE)
@@ -124,7 +125,7 @@ simulation <- function(diff = seq(0, 1.5, by = 0.25),
   prs <- expand.grid(diff = diff, nlevels = nlevels, nobs = nobs)
   nprs <- nrow(prs)
   
-  test <- c("ordmax","ordwmax","catdiff","lrt","yb97","aic") # had suplm here
+  test <- c("ordmax","ordwmax","catdiff","lrt","lrt.sb","yb97","aic") # had suplm here
   ntest <- length(test)
 
   ## Only simulate critical values if we need it.
@@ -138,13 +139,13 @@ simulation <- function(diff = seq(0, 1.5, by = 0.25),
   }
   cval.conds <- as.numeric(names(cval))
 
-  do.parallel <- F #require("parallel")
+  do.parallel <- require("parallel")
   if (do.parallel){
     pow <- mclapply(1:nprs, function(i){
       testpower(diff = prs$diff[i], nobs = prs$nobs[i],
                 nlevels = prs$nlevels[i], gradual = gradual, test = test,
                 ordfun = cval[[which(cval.conds == prs$nlevels[i])]],
-                anomaly = anomaly, verbose = verbose)},
+                anomaly = anomaly, verbose = verbose, ...)},
                     mc.cores = round(.75*detectCores()))
     pow <- t(matrix(unlist(pow), ntest, nprs))
   } else {

### 2013-02-25 ordinal/ordinal.Rnw
Additions for ordinal simulations: o Code modifications for new test stats o Changes to code in ordinal.Rnw
Also changed estfun-lavaan.R to match the code that will be incorporated into lavaan.


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@191 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ## TODO Change this:
-source("../sandbox/yves/estfun.R")
+source("../www/estfun-lavaan.R")
 source("../www/efpFunctional-cat.R")

### 2013-02-25 www/estfun-lavaan.R

@@ -1,59 +1,104 @@
-estfun.lavaan <- function(x)
+# contributed by Ed Merkle (17 Jan 2013)
+# small changes by YR (12 Feb 2013) to match the results
+# of computeGradient in the multiple group case
+estfun.lavaan <- lavScores <- function(object, scaling=FALSE)
 {
-  ## observed data
-  X <- x@Data@X
   ## number variables/sample size
-  nvar <- x@Model@nvar
-  nobs <- x@SampleStats@ntotal
-  ntab <- unlist(x@SampleStats@nobs)
+  samplestats <- object@SampleStats
+  ntab <- unlist(samplestats@nobs)
+  ntot <- samplestats@ntotal
 
-  ## number of groups
-  ngrp <- x@Model@ngroups
-
-  ## Define matrix that will hold all scores
-  ## Assumes means are estimated ## FIXME: check and otherwise throw error
-  scores.H0 <- matrix(NA, nobs, x@Model@nx.free)
+  Score.mat <- matrix(NA, ntot, length(coef(object)))
   
-  for (i in 1:ngrp) {
-    ## fitted moments
-    if(ngrp > 1){
-      moments <- fitted(x)[[i]]
-    }else{
-      moments <- fitted(x)
+  for(g in 1:samplestats@ngroups) {
+    if (samplestats@ngroups > 1){
+      moments <- fitted(object)[[g]]
+    } else {
+      moments <- fitted(object)
     }
     Sigma.hat <- moments$cov
-    ## deal with wishart likelihood? FIXME: check
-    if (x@Options$likelihood == "wishart"){
-      Sigma.hat <- ((ntab[i] - 1)/(ntab[i])) * Sigma.hat
-    }
-    Mu.hat <- moments$mean
-    Sigma.inv <- solve(Sigma.hat)
+  
+    if(!samplestats@missing.flag) { # complete data
+      if(object@Model@meanstructure) { # mean structure
+        nvar <- ncol(samplestats@cov[[g]])
+        Mu.hat <- moments$mean
+        X <- object@Data@X[[g]]
+        Sigma.inv <- lavaan:::inv.chol(Sigma.hat, logdet=FALSE)
+        group.w <- (unlist(samplestats@nobs)/samplestats@ntotal)
 
-    ## Junk matrices for multiplication
[…]

### 2013-02-25 www/mz-ordinal.R

@@ -42,17 +42,15 @@ ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
   ##    the charts.
   ## NB It seems that these models do not converge as often
   ##    at smaller n.  The simulations will gloss over this.
-  zinv <- try(get.zcov(data))
-  if(!inherits(zinv, "try-error")){
-    m2.wls <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
-                      group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances"), estimator="WLS", WLS.V=zinv))
+  ## NB: Use of zinv results in a stat with terrible Type I error
+  ## zinv <- try(get.zcov(data))
+  
+  m2.wls <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
+                      group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances"), estimator="WLS")) #, WLS.V=zinv))
 
-    m3.wls <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
-                      group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances","residuals","means"), estimator="WLS", WLS.V=zinv))
-  }else{
-    m2.wls <- try(y12, silent=TRUE)
-    m3.wls <- try(y12, silent=TRUE)
-  }
+  m3.wls <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE,
+                      group = "age", group.equal = c("loadings","intercepts","regressions","lv.covariances","residuals","means"), estimator="WLS")) #, WLS.V=zinv))
+
 
   ## p-value for LRT Stat, AIC, Sat-Bent scaled difference
   if(!inherits(m2, "try-error") & !inherits(m3, "try-error") &

### 2013-03-21 ordinal/ordinal.Rnw
Simulation & Results: o Modified code to deal with small bug in lavaan which involves   the Satorra-Bentler test stat. o Obtained new results for both Simulation 1 and Simulation 2.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@193 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 source("../www/mz-ordinal.R")
-## TODO Change this:
 source("../www/estfun-lavaan.R")
@@
   levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
-  levels(sim2$test) <- c("LM_uo", "LRT", "LM_o", "WDM_o")
+  levels(sim2$test) <- c("AIC","LM_uo", "LRT", "LRT_sb","LM_o", "WDM_o",
+                         "YB97")
   save(sim2, file="sim2.rda")
@@
 sim2$test <- factor(as.character(sim2$test),
-  levels = c("LM_o", "WDM_o", "LRT", "LM_uo"),
-  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo"))
+  levels = c("LM_o", "WDM_o", "LRT", "LM_uo", "LRT_sb",
+             "YB97", "AIC"),
+  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo", "LRT_sb", "YB97", "AIC"))
 @

### 2013-03-21 www/mz-ordinal.R

@@ -65,7 +65,12 @@ ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
 
     ## Satorra-Bentler scaled difference test automatically
     ## computed via anova()
-    lrt.sb <- anova(m2,m3,SB.classic=TRUE)[[7]][2]
+    lrt.sb <- try(anova(m2,m3,SB.classic=TRUE)[[7]][2])
+    if(inherits(lrt.sb, "try-error")){
+      lrt.sb <- NA
+      tmp <- list(m2=m2, m3=m3, data=data)
+      save(tmp, file="SBerror.rda")
+    }
   } else {
     lrt.p <- NA
     aic <- NA
@@ -94,7 +99,6 @@ ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
 
     yb.diff <- m3.yb - m2.yb
     yb.p <- pchisq(yb.diff, df=anova(m2,m3)[[6]][2], lower.tail=FALSE)
-    if(!m2.wls@Fit@converged | !m3.wls@Fit@converged) yb.p <- NA
   } else {
     yb.p <- NA
   }
@@ -105,7 +109,7 @@ ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
   data <- data[, c("x1", "x2", "x3", "y1", "y2", "y3")]
   
   ## fit model
-  rval <- cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE)
+  rval <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE))
   
   ## store (1) fitted model object, (2) lavaan coefficient
   ## labeling and ordering

### 2013-03-21 www/sim-ordinal.R

@@ -146,7 +146,8 @@ simulation <- function(diff = seq(0, 1.5, by = 0.25),
                 nlevels = prs$nlevels[i], gradual = gradual, test = test,
                 ordfun = cval[[which(cval.conds == prs$nlevels[i])]],
                 anomaly = anomaly, verbose = verbose, ...)},
-                    mc.cores = round(.75*detectCores()))
+                    mc.cores = round(.75*detectCores()),
+                    mc.preschedule = FALSE)
     pow <- t(matrix(unlist(pow), ntest, nprs))
   } else {
     pow <- matrix(rep(NA, ntest * nprs), ncol = ntest)

### 2013-04-04 ordinal/response.txt
Revisions to ordinal paper: o Added new simulation results to ms o Addressed other reviewer comments in ms o Added draft of response letter o Updated references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@194 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Reviewer 1

Although I regard this work as rigorous, I have some doubts whether it adds sufficiently to the previous paper to warrant publication in its current form. I think it may need some additional extensions or additional work on power to render it publishable in Psychometrika. 
Such extensions may entail the following:

Comment 1.1. A comparison of the current approach with scenarios in which the factor analyst uses Modification Indices (MIs) to free parameters. In practice if one has, say, 5 age groups and a given parameter is found to lead to significant increases in model fit when restricted across groups, it is quite common to use MIs to detect the source(s) of the misfit and then to end up with, say one parameter for age groups 1-3  and another parameter for age groups 4 & 5 in a so-called partially invariant model (see Byrne et al., 1989). It would be interesting to see how that approach fares in comparison to the approaches described in the manuscript. I am also interested in reading the authors' ideas about the use MI in "ad hoc" model fitting.

Response 1.1. MIs are quite related to the proposed tests: the MI described by, e.g., Sorbom 1989 is equivalent to a Lagrange multiplier test, and the maximum Lagrange multiplier test statistic (computed across all potential thresholds of V) is within the family of statistics described by Merkle & Zeileis (2013).  However, as the reviewer states, MIs are typically applied only after a significant likelihood ratio test has been obtained (because the MI focuses on a single model parameter).  Our proposed statistics can be used across multiple parameters, potentially in place of the likelihood ratio test.  If the test is significant, one could use MIs or parameter fluctuation plots such as those in manuscript Figures 5 & 6.  We have added some brief detail on this relationship to the "Tests for Continuous V" section.

Comment 1.2. Given the widespread literature on ordinal factor analyses, it would be worthwhile to link the current modelling approach to ordinal factor analysis. Another major reason is that ordinal factor analysis naturally extends towards more standard IRT models. Therefore, by bridging the gap from the current modelling approach to ordinal factor analysis, the authors may also take steps to integrate it with the literature in measurement invariance in the IRT framework.

Response 1.2. We have added detail in the General Discussion on the extension of the proposed tests to other factor analysis models and to IRT.  In particular, our proposed tests can potentially extend to any psychometric model that assumes independent observations and that uses ML estimation.  Additionally, we cite some work on the application of similar tests to IRT.

Comment 1.3. Study of power under other violations of MI related to factor loadings or measurement intercepts.

Response 1.3. We now acknowledge that the obtained simulation results may not generalize to other measurement invariance violations or to other factor analysis models.  We feel that this suggestion would require multiple simulations that would result in a much longer paper, so we have not added new simulations to the manuscript.

Comment 1.4. Consider issues related to model comparison by means of AICs or BICs and the like. This is particularly relevant in light of the fact that monotonic violations of MI should be detectable in a more parsimonious manner than how it is often done in practice with likelihood ratio tests with many DFs.

Response 1.4. If we understand this comment correctly, the reviewer is suggesting that we compare the proposed statistics to the use of AIC or BIC (in addition to the LRT).  Under this framework, one would, e.g., reject the hypothesis of measurement invariance if the AIC for the "full" model were less than the AIC for the restricted model.  We have added the AIC strategy to our simulation, along with some other adjusted versions of the LRT (see Response 2.2).  The AIC turns out to have particularly low power.

Reviewer 2

Comment 2.1. On page 5 is the scaled sum of scores which is normalized by the square root of n and the covariance matrix of the scores I. This is fine theoretically, but what choice do you make in practice? In fact I think that here the problem is model misspecification, other than the biasing factor V. If the model used in ML is incorrect with respect to the number of factors for instance, then this will result in a inconsistent estimate of I. Why not use the sandwich by the late Halbert White (JASA, 1982 for instance)? This is easy to calculate and will improve the limiting distribution of the test statistics.  The choice of I is also missing in the simulations. Please be more specific what you did there.

Response 2.1. We have added an extended discussion of choices for the covariance matrix of scores, with references.  We have also clarified that, in the simulations, we used the information matrix.  Because our simulation models were correctly specified (except for the measurement invariance violation), we do not expect the choice of I to play a large role there.

Comment 2.2. The simulations show great improvement over the standard LR test on measurement invariance. And on page 10 the authors mention the results by Bentler and Bonett (1980) on LR tests picking up deviations too quickly. I think the alternative tests, adjustments of LR tests, like the one proposed in Yuan and Bentler (JASA, 1997). There the LR is downweighted ate rate O_p(1/n), which brings the results of the LR closer to the results of the authors' maxLM test. I would like to see a comparison with this or maybe some other adjusted LR test.

Response 2.2. We implemented three adjusted LRTs in our simulations: (i) the Yuan-Bentler 1997 adjustment suggested by the reviewer, (ii) the use of AIC (which, for nested models, essentially adjusts the LRT for complexity), and (iii) the Satorra-Bentler adjustment with correction for difference tests.  While these corrections downweight individual LRTs, the impact of the adjustments on difference tests is less clear.  Further, the Yuan-Bentler adjustment is seemingly designed for situations where the model is estimated via WLS, and there are no available corrections for difference tests.  We suspect that these issues are problematic, as our simulation results indicated that the Yuan-Bentler adjustment performs poorly in nearly all conditions.  Because the reason for this poor performance is not obvious (and because this issue is not focal for the paper), we only report results of statistics (ii) and (iii) in the manuscript.

Reviewer 3

Comment 3.1. The section entitled 'Theoretical Detail' (page 4) starts with the background for the score-based test statistics. However, from lines 18 onwards, there is a section on measurement invariance (starting with 'In studying measurement invariance, we are essentially testing...' up to line 19 on page 5, just before the subsection 'Tests for Continuous V'). I would suggest to put the section on measurement invariance testing in a separate (sub)section, and present this first, before providing the background for the test statistics.

Response 3.1.  We have moved these sections with thanks to the reviewer, so that measurement invariance comes before the theoretical discussion.

Comment 3.2. For readers unfamiliar with case-based scores, it may be useful to have a table with artificial values for some B(theta)_ij values; say for 20 subjects, and 4 parameters; in an additional column, there could be ordered continuous values for a V variable; in the next column, the V values could be binned to get an ordinal variable; referring to this table, it would be trivial to illustrate how the values of eg DM, CvM, maxLM, etc are computed...

Response 3.2. We agree that this would be useful for some readers, but we think that it would take up more space than it is worth.  For example, the calculation of ordinal test statistics would require both calculations within each bin and unique scaling (by (i/n)*(1 - i/n)) for each bin.  While these calculations are not difficult, they also consume space to write them out (especially if we included the non-ordinal statistics).  Additionally, besides the calculation of test statistics, it might be equally useful to show how the scores (s(theta; x_i)) can be used to calculate B(theta).  This implies that we would need both a table of scores (s(theta; x_i)) and a table of cumulative scores (B(theta)).  We feel that these ideas are represented more concisely and intuitively via their equations.

Comment 3.3. related to 2), I would suggest adding (perhaps in an appendix) some R code illustrating how (for example, for the values in the aforementioned table) the p-values can be computed for the various test statistics

Response 3.3. We have added some practical discussion of p-value computation to the "Choice of Test" subsection of the General Discussion.  We have also posted all the R code for replicating our results (p-values and otherwise) on our R-Forge page (listed in the "Computational Details" section).  We do not include code within the manuscript because the p-values associated with some of the statistics are computed via simulation (i.e., there is not a closed-form solution for critical values associated with the test statistic), and much of the underlying R code is already contained within R's strucchange package.

All minor points have been addressed, with thanks to the reviewer.

### 2013-04-04 ordinal/ordinal.Rnw
Revisions to ordinal paper: o Added new simulation results to ms o Addressed other reviewer comments in ms o Added draft of response letter o Updated references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@194 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 that were recently applied to the study of measurement invariance in
-psychometric models \cite{KopZei10,MerZei12}.
+psychometric models \cite{KopZei10,MerZei13}.
 In the following section, we provide an overview of the family and
@@
 
-\section{Theoretical Detail}
-
-This section contains background on the theory underlying the
-proposed statistics; for a more detailed
-account, see \citeA{MerZei12}.
-
-We  consider situations in which a $p$-dimensional variable $X$
-with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
-model with density $f(\bm{x}_i; \bm{\theta})$ and
-associated joint log-likelihood
-\begin{equation} \label{eq:loglik}
-  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
-    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
-    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
-\end{equation}
-where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
-the distribution.
-
-We focus on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises
-from a structural equation model with assumed multivariate normality,
-though the proposed tests extend beyond this family of models.  Under
-the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
-$\bm{\theta}$ can 
-be estimated by maximum likelihood (ML), i.e.,
-\begin{equation} \label{eq:ml}
-  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \ell(\bm{\theta}; x_1, \dots, x_n),
-\end{equation}
-or equivalently by solving the first order conditions
-\begin{equation}
-    \label{eq:ml1}
-  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
-\end{equation}
-where 
-\begin{equation}
-  \label{eq:score}
-  {\bm s}({\bm \theta}; x_i) ~=~ \left(
-    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_1},
-    \dots,
-    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_k}
-  \right)^\top,
-\end{equation}
-is the score function of the model (the partial derivative of the casewise
-likelihood contributions w.r.t.\ the parameters $\bm{\theta}$).
-Evaluation of the score function at $\hat{\bm{\theta}}$ for $i=1,
-\dots, n$ essentially measures the extent to which the model maximizes
-each individual's likelihood: as an individual's scores stray further
-from zero, the model provides a poorer description of that individual.
+\section{Measurement Invariance}
 
@@
 
-In the section below, we review tests where
-$V$ is continuous (and $\nu$ is unknown) before proceeding to the 
-proposed tests for ordinal $V$.
+In the section below, we review the theory underlying tests where
+$V$ is continuous (and $\nu$ is unknown).  We then propose novel tests
+for ordinal $V$.
+
+\section{Theoretical Detail}
+
+This section contains background on the theory underlying the
+proposed statistics; for a more detailed
+account, see \citeA{MerZei13}.
+
+We  consider situations in which a $p$-dimensional variable $X$
+with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
+model with density $f(\bm{x}_i; \bm{\theta})$ and
+associated joint log-likelihood
+\begin{equation} \label{eq:loglik}
+  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
+    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
+    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
+\end{equation}
+where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
+the distribution.
+
+We focus on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises
+from a structural equation model with assumed multivariate normality,
+though the proposed tests extend beyond this family of models.  Under
+the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
+$\bm{\theta}$ can 
+be estimated by maximum likelihood (ML), i.e.,
+\begin{equation} \label{eq:ml}
+  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \ell(\bm{\theta}; x_1, \dots, x_n),
+\end{equation}
+or equivalently by solving the first order conditions
+\begin{equation}
+    \label{eq:ml1}
+  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
+\end{equation}
+where 
+\begin{equation}
+  \label{eq:score}
+  {\bm s}({\bm \theta}; x_i) ~=~ \left(
+    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_1},
+    \dots,
+    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_k}
+  \right)^\top,
+\end{equation}
+is the score function of the model (the partial derivative of the casewise
+likelihood contributions w.r.t.\ the parameters $\bm{\theta}$).
+Evaluation of the score function at $\hat{\bm{\theta}}$ for $i=1,
+\dots, n$ essentially measures the extent to which the model maximizes
+each individual's likelihood: as an individual's scores stray further
+from zero, the model provides a poorer description of that individual.
 
@@
 related to those underlying the Lagrange multiplier test and are
-discussed in detail by \citeA{MerZei12}. Here,
-we review the tests' properties that are relevant for extending them
+discussed in detail by \citeA{MerZei13}.  Additionally, these ideas
+are related to those underlying Modification Indexes (e.g.,
+\citeNP{Sor89}): the Modification Index is equivalent to a Lagrange
+multiplier test, and the Lagrange multiplier test is contained in the
+family described by \citeA{MerZei13}.  Here,
+we focus on the tests' properties that are relevant for extending them
 to the ordinal case.
@@
 that the observations are ordered 
-w.r.t.\ $V$ (so that $v_i \le v_{i + 1}$) and define the $k$-dimensional
+w.r.t.\ $V$, with $x_{(i)}$ reflecting
+the data for the individual who has the 
+$i^{\text{th}}$-smallest value of $V$.  We then define the $k$-dimensional
 {\em cumulative score process} as 
@@
   {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
-    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_i)
+    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
   \qquad (0 \le t \le 1)
@@
 $\hat I$ is some consistent estimate of the covariance matrix of the
-scores, e.g., based on the information matrix or an outer product of the scores.
+scores.  Natural choices for $\hat I$ include the information matrix
+(which we use in our applications and simulations below) or
+alternatively some kind of outer product of the scores or sandwich
+estimator to guard the inference against potential misspecification of
+the model (see \citeNP{hub67} for the theoretical foundation and
+\citeNP{zei06b} for a computational framework).  
 Equation~\eqref{eq:cumscore} simultaneously accounts for the
@@
 invariance when multiple factor analysis parameters change
-simultaneously \cite{MerZei12}.
+simultaneously \cite{MerZei13}.
 
@@
 scaled by the asymptotic variance $t (1 - t)$ of the process
-${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeA{MerZei12} found 
+${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeA{MerZei13} found 
 that both tests perform well when assessing simultaneous changes
@@
 ordering. (Note that the same also applies if $V$ is continuous in nature but is only
-discretely measured leading to many ties.)
+discretely measured, leading to many ties.)
 
@@
                              \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
+                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
 \end{eqnarray}
+resulting in a ``weighted'' double maximum statistic (weighted by the
+asymptotic variance of the Brownian bridge) and an ordinal, maximum
+Lagrange multiplier statistic.
 Critical values associated with these test statistics can be obtained
@@
 use the proposed test statistics to test for measurement invariance
-simultaneously across the unique variances, which is similar to
-comparing a tau-equivalent model to a congeneric model (e.g.,
-\citeNP{VanLan00}).
-
-We also compute two statistics that treat the ordinal
-auxiliary variable as categorical.  These include the likelihood ratio
-test of measurement 
-invariance in the six unique variance parameters, along with the
-unordered LM test from~\eqref{eq:lmuo}.
+simultaneously across the unique variances, which is similar to a test
+of ``invariant uniquenesses'' (see \citeNP{VanLan00}). % Or comparing
+                                        % parallel to tau-equivalent
+
+The four ``traditional'' statistics that we consider generally treat
+the ordinal auxiliary variable as categorical.  The first two are the
+likelihood ratio test of measurement 
+invariance in the six unique variance parameters and the
+unordered LM test from~\eqref{eq:lmuo}.  Next, we consider the
+Satorra-Bentler \citeyear{SatBen01} scaled likelihood ratio test
+statistic with correction for difference testing.  Finally, we
+consider the use of AIC \cite{Aka74} for detecting measurement
+invariance.  In this context, we compute AIC for both the ``reduced'' 
+model (all parameters constrained equal across groups) and the
+``full'' model (unique variances allowed to differ between groups).
+We then declare a measurement invariance violation if the AIC for the
+full model is lower than the AIC for the reduced model.
 
@@
 
-Simulation results are presented in Figure~\ref{fig:sim1res}.  Rows of
+Simulation results comparing the ordinal tests to the unordered LM
+test and the LRT are presented in Figure~\ref{fig:sim1res}.  Rows of
 the figure 
@@
 at specific levels of the ordinal variable
-(see \citeNP{MerZei12}, for related discussion), while the
+(see \citeNP{MerZei13}, for related discussion), while the
 three other tests under consideration make use of this information.
@@
 $n/m$ observations are available in each subsample.  The proposed
-statistics are all of the LM-type and just require estimation of the single-group model,
+statistics are all of the LM-type and just require estimation of the
+single-group model, 
 leading to a clear advantage in these conditions.
+
+Simulation results comparing the ordinal tests to the Satorra-Bentler
+scaled test statistic and to the AIC are presented in
+Figure~\ref{fig:sim1bres}.  This figure has the same design as the
+previous figure.  It is again seen that the $\max \mathit{LM}_o$
+statistic exhibits the greatest power across the conditions.
+Additionally, the Satorra-Bentler statistic has very large Type I
+error rates at low-$n$/large-$m$ combinations but is comparable to the
+ordinal double-max test at large $n$.  Finally, the ``power'' of the
+AIC (measured by the proportion of times that the full model has a
+lower AIC than the reduced model) is exceptionally low across all
+conditions.  The full model's gain in fit is seldom worth the extra
+parameters, at least according to AIC.
 
@@
 \caption{Simulated power curves for the ordered and unordered
-      $\max \mathit{LM}$ tests, the ordered double-max test, and the
+      $\max\mathit{LM}$ tests, the ordered double-max test, and the
       likelihood ratio test across three sample sizes $n$, three
@@
 <<sim1res, fig=TRUE, height=7, width=7>>=
+sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
+sim1.tmp$test <- factor(sim1.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1, 
+mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+\end{figure}
+
+\begin{figure}
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the ordered double-max test, the
+      Satorra-Bentler (2001) scaled likelihood ratio
+      test, and the use of AIC 
+      across three sample sizes $n$, three 
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
+\label{fig:sim1bres}
+\setkeys{Gin}{width=\textwidth}
+<<sim1bres, fig=TRUE, height=7, width=7>>=
+sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
+sim1.tmp$test <- factor(sim1.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
              type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@@
 \subsection{Results}
-Simulation results are presented in Figure~\ref{fig:sim2res}.  It is
+Simulation results for the two ordinal test statistics, the unordered LM
+test, and the LRT are presented in Figure~\ref{fig:sim2res}.  It is
 observed that results are very consistent across the sample sizes
@@
 
-The more interesting result of this simulation lies in the comparison
+The more interesting result of Figure~\ref{fig:sim2res} lies in the comparison
 of the four test statistics within each panel of the figure.  The
@@
 statistic being higher at larger values of $d$.
-This demonstrates that the proposed ordinal test
+
+Simulation results for the two ordinal test statistics, the Satorra-Bentler
+scaled test statistic, and the AIC are presented in
+Figure~\ref{fig:sim2bres}.  It is observed that the Satorra-Bentler
+statistic exhibits performance that is similar to the usual LRT,
+with the two ordinal statistics having lower power.  The AIC, which
+generally exhibited low power even for Simulation 1, continues to
+exhibit low power here.  However, if $n$ or $d$ is large enough, the
+power exhibited by AIC is larger than the power of the ordinal test
+statistics.
+
+These results demonstrate that the proposed ordinal test
 statistics can be especially useful at large sample sizes, where
@@
 <<sim2res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2, 
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+\end{figure}
+
+\begin{figure}
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the ordered double-max test, the
+      Satorra-Bentler (2001) scaled likelihood ratio
+      test, and the use of AIC 
+      across three sample sizes $n$, three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
+      occurring at a single level (the $(1 +
+      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
+\label{fig:sim2bres}
+\setkeys{Gin}{width=\textwidth}
+<<sim2bres, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp, 
              type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@@
 similar performance if only a single model parameter violated
-measurement invariance.
+measurement invariance.  The only disadvantage to $\max\mathit{LM}_o$ is
+that its critical- and/or $p$-values must be computed by simulation,
+which can significantly increase computation time.  We return to this
+issue in the General Discussion.
 
@@
 has an auxiliary ordinal variable and wishes to study measurement
-invariance.    To our knowledge,
+invariance.  We demonstrated via simulation that these statistics have
+good properties, though these results necessarily examined a small
+number of models and invariance violations and may not hold in all
+situations.  To our knowledge, 
 the ordinal measurement invariance statistics proposed here are the
@@
 (especially if $V$ has very many levels).  
-As described in detail by \citeA{MerZei12}, we
+As described in detail by \citeA{MerZei13}, we
 can also use cumulative score processes with continuous $V$,
@@
 large, the proposed~$\max\mathit{LM}_o$ statistic will be very close
-to the $\max\mathit{LM}$ statistic described in \citeA{MerZei12}.
+to the $\max\mathit{LM}$ statistic described in \citeA{MerZei13}.
 Thus, the formation of ordinal age groups (or other variables) is not
@@
 
+Finally, a practical issue associated with the proposed
+$\max\mathit{LM}_o$ statistic involves the computation of $p$-values:
+$p$-values and/or critical values must be computed via simulation of a
+Brownian bridge, with the simulation depending on
+the relative proportion of cases at each level of the ordinal
+auxiliary variable.  Hence, a new simulation usually must be conducted
+for each dataset, which can be somewhat time-consuming.
+
 \subsection{Extension to Other Models}
@@
 (Equation~\eqref{eq:score}) be available following model estimation.
-As a result, applications to studying DIF in IRT are immediate,
+As a result, applications to studying DIF in IRT are immediate (for a
+presentation involving non-ordinal variables, see \citeNP{KopZei10}),
 as are general psychometric applications to studying parameter
-stability w.r.t.\ ordinal auxiliary variables.  We expect the
+stability w.r.t.\ ordinal auxiliary variables.  These could 
+include applications where ordinal variables are explicitly included
+in the model, such as ordinal factor analysis.
+We expect the
 same general results to hold for these applications, whereby the
@@
 
-% TODO % \fixme{Z: Should we make some practical remarks about the computation of
-% TODO % $p$-values somewhere? The issue with the maxLM test is that it is hard
-% TODO % to tabulate critical values in advance and hence simulation has to be
-% TODO % done for each case and is potentially rather time-consuming. Not sure
-% TODO % whether we should do this now or wait until the revision or so\dots}
-% TODO % \readme{ECM: I am happy to wait for the revision, and for now have
-% TODO %   inserted a TODO comment into the Rnw file.}
-
 \subsection{Summary}
-The test statistics proposed in this paper have relatively-high power for
-detecting measurement 
+As demonstrated via simulation, the proposed test statistics 
+have relatively-high power for detecting measurement 
 invariance violations that are monotonic with the ordinal variable,

### 2013-04-15 ordinal/ordinal.Rnw
Added babel package to ordinal.Rnw, which combats the tilde problem in texi2pdf
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@195 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
small improvements, update of Rasch tree ref
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@196 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
+%% This keeps the tildes out when you use Rnw2pdf:
+\usepackage[english]{babel}
 %% need no \usepackage{Sweave}
@@
 that were recently applied to the study of measurement invariance in
-psychometric models \cite{KopZei10,MerZei13}.
+psychometric models \cite{MerZei13,StrKop13}.
 In the following section, we provide an overview of the family and
@@
 As mentioned previously, when $V$ is categorical with a
-relatively-small number of categories, 
+relatively small number of categories, 
 tests of measurement invariance typically proceed via multiple-group
@@
 discussed in detail by \citeA{MerZei13}.  Additionally, these ideas
-are related to those underlying Modification Indexes (e.g.,
-\citeNP{Sor89}): the Modification Index is equivalent to a Lagrange
+are related to those underlying modification indexes (e.g.,
+\citeNP{Sor89}): the modification index is equivalent to a Lagrange
 multiplier test, and the Lagrange multiplier test is contained in the
@@
 related to multiple parameters fluctuating simultaneously, resulting
-in it having relatively-low power for assessing measurement
+in it having relatively low power for assessing measurement
 invariance when multiple factor analysis parameters change
@@
 of the four test statistics within each panel of the figure.  The
-two ``unordered'' test statistics both have relatively-high power to
+two ``unordered'' test statistics both have relatively high power to
 detect the measurement invariance violation, illustrating the result 
@@
 As a result, applications to studying DIF in IRT are immediate (for a
-presentation involving non-ordinal variables, see \citeNP{KopZei10}),
+presentation involving non-ordinal variables, see \citeNP{StrKop10}),
 as are general psychometric applications to studying parameter
@@
 As demonstrated via simulation, the proposed test statistics 
-have relatively-high power for detecting measurement 
+have relatively high power for detecting measurement 
 invariance violations that are monotonic with the ordinal variable,
-and they have relatively-low power for detecting minor violations that are not
+and they have relatively low power for detecting minor violations that are not
 monotonic.  The former feature implies that the statistics are good at

### 2013-04-16 ordinal/response.txt
Revisions to ordinal ms: o Added detail to re-ordered introductory sections o Small revisions to response letter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@197 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Response 2.2.
-We implemented three adjusted LRTs in our simulations: (i) the Yuan-Bentler 1997 adjustment suggested by the reviewer, (ii) the use of AIC (which, for nested models, essentially adjusts the LRT for complexity), and (iii) the Satorra-Bentler adjustment with correction for difference tests.
-While these corrections downweight individual LRTs, the impact of the adjustments on difference tests is less clear.
+We implemented three adjusted LRTs in our simulations: (i) the Yuan-Bentler 1997 adjustment suggested by the reviewer, (ii) the use of AIC (which, for nested models, essentially increases the critical value of the LRT), and (iii) the Satorra-Bentler adjustment with correction for difference tests.
+While these adjustments downweight the LRT associated with individual models (e.g., comparing an individual model to a saturated model), the impact of the adjustments on difference tests is less clear.
 Further, the Yuan-Bentler adjustment is seemingly designed for situations where the model is estimated via WLS, and there are no available corrections for difference tests.
-We suspect that these issues are problematic, as our simulation results indicated that the Yuan-Bentler adjustment performs poorly in nearly all conditions.
-Because the reason for this poor performance is not obvious (and because this issue is not focal for the paper), we only report results of statistics (ii) and (iii) in the manuscript.
+We suspect that these issues are problematic, as our simulation results indicated that these statistics are worse than the usual LRT in nearly all conditions.
+We have included graphs of the results with the resubmission but have not included these results in the manuscript.
+We are willing to reconsider the exclusion of these results if the AE or reviewers find them of interest.
 

### 2013-04-16 ordinal/ordinal.Rnw
Revisions to ordinal ms: o Added detail to re-ordered introductory sections o Small revisions to response letter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@197 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-In studying measurement invariance, we are essentially testing the
+In studying measurement invariance, we consider situations in which a $p$-dimensional variable $X$
+with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
+model with density $f(\bm{x}_i; \bm{\theta})$ and
+associated joint log-likelihood
+\begin{equation} \label{eq:loglik}
+  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
+    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
+    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
+\end{equation}
+where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
+the distribution.
+
+Tests of  measurement invariance are essentially tests of the 
 assumption that all  
@@
 parameter vector ${\bm \theta}$.
-Thus, a hypothesis of invariance can be written as
+Thus, a hypothesis of measurement invariance can be written as
 \begin{equation}
@@
 
-We  consider situations in which a $p$-dimensional variable $X$
-with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
-model with density $f(\bm{x}_i; \bm{\theta})$ and
-associated joint log-likelihood
-\begin{equation} \label{eq:loglik}
-  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
-    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
-    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
-\end{equation}
-where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
-the distribution.
-
-We focus on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises
+\subsection{Model Estimation}
+We focus specifically on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises 
 from a structural equation model with assumed multivariate normality,
-though the proposed tests extend beyond this family of models.  Under
+though the proposed tests extend beyond this family of models.
+  Under
 the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
@@
 Evaluation of the score function at $\hat{\bm{\theta}}$ for $i=1,
-\dots, n$ essentially measures the extent to which the model maximizes
+\dots, n$ measures the extent to which the model maximizes
 each individual's likelihood: as an individual's scores stray further
@@
 
-The four ``traditional'' statistics that we consider generally treat
-the ordinal auxiliary variable as categorical.  The first two are the
+The two ``traditional'' statistics that we consider generally treat
+the ordinal auxiliary variable as categorical.  These include the
 likelihood ratio test of measurement 
 invariance in the six unique variance parameters and the
-unordered LM test from~\eqref{eq:lmuo}.  Next, we consider the
+unordered LM test from~\eqref{eq:lmuo}.  At the request of reviewers,
+we also considered the
 Satorra-Bentler \citeyear{SatBen01} scaled likelihood ratio test
-statistic with correction for difference testing.  Finally, we
-consider the use of AIC \cite{Aka74} for detecting measurement
-invariance.  In this context, we compute AIC for both the ``reduced'' 
-model (all parameters constrained equal across groups) and the
-``full'' model (unique variances allowed to differ between groups).
-We then declare a measurement invariance violation if the AIC for the
-full model is lower than the AIC for the reduced model.
+statistic with correction for difference testing, the Yuan-Bentler
+\citeyear{YuaBen97} scaled test statistic, and the use of AIC
+\cite{Aka74} for detecting measurement 
+invariance.  We do not report the latter results, because these test
+statistics performed worse than the usual likelihood ratio test (for
+both Simulation 1 and Simulation 2).
+
+%% In this context, we compute AIC for both the ``reduced'' 
+%% model (all parameters constrained equal across groups) and the
+%% ``full'' model (unique variances allowed to differ between groups).
+%% We then declare a measurement invariance violation if the AIC for the
+%% full model is lower than the AIC for the reduced model.
 
@@
 
-Simulation results comparing the ordinal tests to the Satorra-Bentler
-scaled test statistic and to the AIC are presented in
-Figure~\ref{fig:sim1bres}.  This figure has the same design as the
-previous figure.  It is again seen that the $\max \mathit{LM}_o$
-statistic exhibits the greatest power across the conditions.
-Additionally, the Satorra-Bentler statistic has very large Type I
-error rates at low-$n$/large-$m$ combinations but is comparable to the
-ordinal double-max test at large $n$.  Finally, the ``power'' of the
-AIC (measured by the proportion of times that the full model has a
-lower AIC than the reduced model) is exceptionally low across all
-conditions.  The full model's gain in fit is seldom worth the extra
-parameters, at least according to AIC.
+%% Simulation results comparing the ordinal tests to the Satorra-Bentler
+%% scaled test statistic and to the AIC are presented in
+%% Figure~\ref{fig:sim1bres}.  This figure has the same design as the
+%% previous figure.  It is again seen that the $\max \mathit{LM}_o$
+%% statistic exhibits the greatest power across the conditions.
+%% Additionally, the Satorra-Bentler statistic has very large Type I
+%% error rates at low-$n$/large-$m$ combinations but is comparable to the
+%% ordinal double-max test at large $n$.  Finally, the ``power'' of the
+%% AIC (measured by the proportion of times that the full model has a
+%% lower AIC than the reduced model) is exceptionally low across all
+%% conditions.  The full model's gain in fit is seldom worth the extra
+%% parameters, at least according to AIC.
 
@@
 
-\begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the ordered double-max test, the
-      Satorra-Bentler (2001) scaled likelihood ratio
-      test, and the use of AIC 
-      across three sample sizes $n$, three 
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
-\label{fig:sim1bres}
-\setkeys{Gin}{width=\textwidth}
-<<sim1bres, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
-sim1.tmp$test <- factor(sim1.tmp$test)
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-\end{figure}
+%% \begin{figure}
+%% \caption{Simulated power curves for the ordered 
+%%       $\max\mathit{LM}$ tests, the ordered double-max test, the
+%%       Satorra-Bentler (2001) scaled likelihood ratio
+%%       test, and the use of AIC 
+%%       across three sample sizes $n$, three 
+%%       levels of the ordinal variable $m$, and measurement invariance
+%%       violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
+%% \label{fig:sim1bres}
+%%\setkeys{Gin}{width=\textwidth}
+%%<<sim1bres, fig=TRUE, height=7, width=7>>=
+%%sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
+%%sim1.tmp$test <- factor(sim1.tmp$test)
+%%trellis.par.set(theme = canonical.theme(color = FALSE))
+%%mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
+%%print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
+%%             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+%%@
+%%\end{figure}
 
@@
 
-Simulation results for the two ordinal test statistics, the Satorra-Bentler
-scaled test statistic, and the AIC are presented in
-Figure~\ref{fig:sim2bres}.  It is observed that the Satorra-Bentler
-statistic exhibits performance that is similar to the usual LRT,
-with the two ordinal statistics having lower power.  The AIC, which
-generally exhibited low power even for Simulation 1, continues to
-exhibit low power here.  However, if $n$ or $d$ is large enough, the
-power exhibited by AIC is larger than the power of the ordinal test
-statistics.
+%% Simulation results for the two ordinal test statistics, the Satorra-Bentler
+%% scaled test statistic, and the AIC are presented in
+%% Figure~\ref{fig:sim2bres}.  It is observed that the Satorra-Bentler
+%% statistic exhibits performance that is similar to the usual LRT,
+%% with the two ordinal statistics having lower power.  The AIC, which
+%% generally exhibited low power even for Simulation 1, continues to
+%% exhibit low power here.  However, if $n$ or $d$ is large enough, the
+%% power exhibited by AIC is larger than the power of the ordinal test
+%% statistics.
 
@@
 
-\begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the ordered double-max test, the
-      Satorra-Bentler (2001) scaled likelihood ratio
-      test, and the use of AIC 
-      across three sample sizes $n$, three
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
-      occurring at a single level (the $(1 +
-      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
-\label{fig:sim2bres}
-\setkeys{Gin}{width=\textwidth}
-<<sim2bres, fig=TRUE, height=7, width=7>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-\end{figure}
+%% \begin{figure}
+%% \caption{Simulated power curves for the ordered 
+%%       $\max\mathit{LM}$ tests, the ordered double-max test, the
+%%       Satorra-Bentler (2001) scaled likelihood ratio
+%%       test, and the use of AIC 
+%%       across three sample sizes $n$, three
+%%       levels of the ordinal variable $m$, and measurement invariance
+%%       violations of 0--3 standard errors (scaled by $\sqrt{n}$)
+%%       occurring at a single level (the $(1 +
+%%       m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
+%% \label{fig:sim2bres}
+%%\setkeys{Gin}{width=\textwidth}
+%%<<sim2bres, fig=TRUE, height=7, width=7>>=
+%%sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
+%%sim2.tmp$test <- factor(sim2.tmp$test)
+%%trellis.par.set(theme = canonical.theme(color = FALSE))
+%%mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+%%print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp, 
+%%             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+%%@
+%%\end{figure}
 

### 2013-04-16 ordinal/response.txt
Revisions to ordinal ms: o Further revised ordering of intro section o Added detail to response letter o Added references o Created graphs of simulation results for unused statistics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@198 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Response 1.1.
-MIs are quite related to the proposed tests: the MI described by, e.g., Sorbom 1989 is equivalent to a Lagrange multiplier test, and the maximum Lagrange multiplier test statistic (computed across all potential thresholds of V) is within the family of statistics described by Merkle & Zeileis (2013).
-However, as the reviewer states, MIs are typically applied only after a significant likelihood ratio test has been obtained (because the MI focuses on a single model parameter).
+MIs are quite related to the proposed tests: the MI described by, e.g., Sorbom 1989 is equivalent to a Lagrange multiplier test.
+The maximum of Lagrange multiplier tests (e.g., computed across all potential thresholds of V), on the other hand, is within the family of statistics described by Merkle & Zeileis (2013).
+However, as the reviewer states, MIs are typically applied only after a significant likelihood ratio test has been obtained.
 Our proposed statistics can be used across multiple parameters, potentially in place of the likelihood ratio test.
-If the test is significant, one could use MIs or parameter fluctuation plots such as those in manuscript Figures 5 & 6.
+If the test is significant, one could use MIs or parameter fluctuation plots such as those in manuscript Figures 3 & 4 to determine the model misfit.
 We have added some brief detail on this relationship to the "Tests for Continuous V" section.
@@
 We suspect that these issues are problematic, as our simulation results indicated that these statistics are worse than the usual LRT in nearly all conditions.
-We have included graphs of the results with the resubmission but have not included these results in the manuscript.
-We are willing to reconsider the exclusion of these results if the AE or reviewers find them of interest.
+
+We have included some graphs of the results with this response letter but have not included them in the manuscript.
+Both graphs are organized in the same manner as the graphs in the paper, except that they contain different statistics: the ordinal LM test (LM_o), the Likelihood ratio test (LRT), the Satorra-Bentler LRT (LRT_sb), and the use of AIC (AIC).
+We do not think these results add anything to the paper, but we are willing to reconsider if the AE or reviewers find them worthwhile.
 
@@
 
-Response 3.1.  We have moved these sections with thanks to the reviewer, so that measurement invariance comes before the theoretical discussion.
+Response 3.1.  We have moved around these sections, so that measurement invariance comes before the theoretical discussion.
 
@@
 While these calculations are not difficult, they also consume space to write them out (especially if we included the non-ordinal statistics).
-Additionally, besides the calculation of test statistics, it might be equally useful to show how the scores (s(theta; x_i)) can be used to calculate B(theta).
-This implies that we would need both a table of scores (s(theta; x_i)) and a table of cumulative scores (B(theta)).
+Additionally, besides the calculation of test statistics, it might be equally useful to show how the scores s(theta; x_i) can be used to calculate B(theta).
+This implies that we would need both a table of s(theta; x_i) and a table of cumulative scores B(theta).
 We feel that these ideas are represented more concisely and intuitively via their equations.

### 2013-04-16 ordinal/ordinal.Rnw
Revisions to ordinal ms: o Further revised ordering of intro section o Added detail to response letter o Added references o Created graphs of simulation results for unused statistics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@198 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-In studying measurement invariance, we consider situations in which a $p$-dimensional variable $X$
+In studying measurement invariance, we consider situations where a $p$-dimensional variable $X$
 with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
@@
 Researchers typically do not expect such a result when testing
-measurement invariance w.r.t.\ continuous or ordinal $V$, and 
+measurement invariance w.r.t.\ categorical or continuous $V$, and 
 researchers often cannot interpret such violations.  Monotonic
@@
 
-As stated previously, we focus here on situations where $V$ is
+As stated previously, we specifically focus here on situations where $V$ is
 ordinal and where the measurement invariance violation is related to the ordinal
@@
 Note that the effect of the outer maximum can be easily captured by a Bonferroni
-correction as the $k$~components of the Brownian bridge are asymptotically independent.
+correction, as the $k$~components of the Brownian bridge are asymptotically independent.
 Moreover, the inner maximum is taken over $m-1$ variables
@@
 As a result, applications to studying DIF in IRT are immediate (for a
-presentation involving non-ordinal variables, see \citeNP{StrKop10}),
+presentation involving non-ordinal variables, see \citeNP{StrKop13}),
 as are general psychometric applications to studying parameter

### 2013-04-17 ordinal/response.Rnw
turned response letter into formatted .Rnw with graphics dynamically generated
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@199 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{a4wide}

\begin{document}

\setkeys{Gin}{width=\textwidth}

\section*{Reviewer 1}

\begin{enumerate}
\item \emph{A comparison of the current approach with scenarios in which the
factor analyst uses Modification Indices (MIs) to free parameters. In practice
if one has, say, 5 age groups and a given parameter is found to lead to
significant increases in model fit when restricted across groups, it is quite
common to use MIs to detect the source(s) of the misfit and then to end up with,
say one parameter for age groups 1-3  and another parameter for age groups 4 \& 5
in a so-called partially invariant model (see Byrne et al., 1989). It would be
interesting to see how that approach fares in comparison to the approaches
described in the manuscript. I am also interested in reading the authors' ideas
about the use MI in ``ad hoc" model fitting.}

MIs are quite related to the proposed tests: the MI described by, e.g., Sorbom
1989 is equivalent to a Lagrange multiplier test.  The maximum of Lagrange
multiplier tests (e.g., computed across all potential thresholds of V), on the
other hand, is within the family of statistics described by Merkle \& Zeileis
(2013).  However, as the reviewer states, MIs are typically applied only after a
significant likelihood ratio test has been obtained.  Our proposed statistics
can be used across multiple parameters, potentially in place of the likelihood
ratio test.  If the test is significant, one could use MIs or parameter
fluctuation plots such as those in manuscript Figures 3 \& 4 to determine the
model misfit.  We have added some brief detail on this relationship to the
"Tests for Continuous V" section.

\item \emph{Given the widespread literature on ordinal factor analyses, it would
be worthwhile to link the current modelling approach to ordinal factor analysis.
Another major reason is that ordinal factor analysis naturally extends towards
more standard IRT models. Therefore, by bridging the gap from the current
modelling approach to ordinal factor analysis, the authors may also take steps
to integrate it with the literature in measurement invariance in the IRT
framework.}

We have added detail in the General Discussion on the extension of the proposed
tests to other factor analysis models and to IRT.  In particular, our proposed
tests can potentially extend to any psychometric model that assumes independent
observations and that uses ML estimation.  Additionally, we cite some work on
the application of similar tests to IRT.

\item \emph{Study of power under other violations of MI related to factor
loadings or measurement intercepts.}

We now acknowledge that the obtained simulation results may not generalize to
other measurement invariance violations or to other factor analysis models.  We
feel that this suggestion would require multiple simulations that would result
in a much longer paper, so we have not added new simulations to the manuscript.

\item \emph{Consider issues related to model comparison by means of AICs or BICs
and the like. This is particularly relevant in light of the fact that monotonic
violations of MI should be detectable in a more parsimonious manner than how it
is often done in practice with likelihood ratio tests with many DFs.}

If we understand this comment correctly, the reviewer is suggesting that we
compare the proposed statistics to the use of AIC or BIC (in addition to the
LRT).  Under this framework, one would, e.g., reject the hypothesis of
measurement invariance if the AIC for the ``full" model were less than the AIC
for the restricted model.  We have added the AIC strategy to our simulation,
along with some other adjusted versions of the LRT (see Response 2.2).  The AIC
turns out to have particularly low power.
\end{enumerate}

\section*{Reviewer 2}

\begin{enumerate} \item \emph{On page 5 is the scaled sum of scores which is
normalized by the square root of n and the covariance matrix of the scores I.
This is fine theoretically, but what choice do you make in practice? In fact I
think that here the problem is model misspecification, other than the biasing
factor V. If the model used in ML is incorrect with respect to the number of
factors for instance, then this will result in a inconsistent estimate of I. Why
not use the sandwich by the late Halbert White (JASA, 1982 for instance)? This
is easy to calculate and will improve the limiting distribution of the test
statistics.  The choice of I is also missing in the simulations. Please be more
specific what you did there.}

We have added an extended discussion of choices for the covariance matrix of
scores, with references.  We have also clarified that, in the simulations, we
used the information matrix.  Because our simulation models were correctly
specified (except for the measurement invariance violation), we do not expect
the choice of I to play a large role there.

\item \emph{The simulations show great improvement over the standard LR test on
measurement invariance. And on page 10 the authors mention the results by
Bentler and Bonett (1980) on LR tests picking up deviations too quickly. I think
the alternative tests, adjustments of LR tests, like the one proposed in Yuan
and Bentler (JASA, 1997). There the LR is downweighted at rate $O_p(1/n)$, which
brings the results of the LR closer to the results of the authors' maxLM test. I
would like to see a comparison with this or maybe some other adjusted LR test.}

<<fig1, echo=FALSE, results=hide, fig=TRUE, height=10, width=10>>=
load("sim1.rda")
library("lattice")

lev0 <- c(   "LM_o","LRT","LRT_sb","AIC")
lev  <- c("maxLM_o","LRT","LRT_sb","AIC")
cl <- c(1, 4, 2, 3)
pc <- c(1, 2, 3, 4)
lt <- c(1, 1, 2, 2)

sim1.tmp <- sim1[sim1$test %in% lev0,]
sim1.tmp$test <- factor(sim1.tmp$test, levels = lev0, labels = lev)
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp,
  type = "b", xlab="Violation Magnitude", ylab="Power",
  key=list(text = list(lab = lev), lines = list(col = cl, lty = lt), points = list(col = cl, pch = pc)),
  col = cl, pch = pc, lty = lt))
@

<<fig2, echo=FALSE, results=hide, fig=TRUE, height=10, width=10>>=
load("sim2.rda")
sim2.tmp <- sim2[sim2$test %in% lev0,]
sim2.tmp$test <- factor(sim2.tmp$test, levels = lev0, labels = lev)
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp,
  type = "b", xlab="Violation Magnitude", ylab="Power",
  key=list(text = list(lab = lev), lines = list(col = cl, lty = lt), points = list(col = cl, pch = pc)),
  col = cl, pch = pc, lty = lt))
@

We implemented three adjusted LRTs in our simulations: (i) the Yuan-Bentler 1997
adjustment suggested by the reviewer, (ii) the use of AIC, and (iii) the
Satorra-Bentler adjustment with correction for difference tests.  While these
adjustments downweight the LRT associated with individual models (e.g.,
comparing an individual model to a saturated model), the impact of the
adjustments on difference tests is less clear.  Further, the Yuan-Bentler
adjustment is seemingly designed for situations where the model is estimated via
WLS, and there are no available corrections for difference tests.  We suspect
that these issues are problematic, as our simulation results indicated that
these statistics are worse than the usual LRT in nearly all conditions.
We have included the resulting graphs for Simulation~1 and Simulation~2 above
(but not in the manuscript).
\begin{itemize}
  \item Both graphs are organized in the same manner as the graphs in the paper,
    except that they contain different statistics: the ordinal maxLM test
    ($\mathrm{LM}_o$), the Likelihood ratio test (LRT), the Satorra-Bentler LRT ($\mathrm{LRT}_\mathrm{sb}$),
    and the use of AIC (AIC). The former two are presented in the manuscripts while the latter
    two are only shown here.
  \item It can be seen that $\mathrm{LRT}_\mathrm{sb}$ is virtually identical to
    the plain LRT in situations with small number of groups $m$ and/or large
    sample sizes $n$ (in particular throughout Simulation~2) while for the large $m$/small
    $n$ scenarios , the problems of the LRT are even amplified.
  \item Finally, the AIC has relatively low power throughout Simulation~1. This is
    because using AIC corresponds to using the LRT with an increased critical value
    here (because the models being compared are nested with relatively large differences in degrees
    of freedom).
\end{itemize}
Hence, we feel that the additional discussion of these modifications do not add
substantial new insights and hence have omitted them from the paper (and only
mention them very briefly). But we are willing to reconsider if the AE or
reviewers find the results worthwhile.
\end{enumerate}

\section*{Reviewer 3}

\begin{enumerate} \item \emph{The section entitled `Theoretical Detail' (page 4)
starts with the background for the score-based test statistics. However, from
lines 18 onwards, there is a section on measurement invariance (starting with
`In studying measurement invariance, we are essentially testing...' up to line
19 on page 5, just before the subsection `Tests for Continuous V'). I would
suggest to put the section on measurement invariance testing in a separate
(sub)section, and present this first, before providing the background for the
test statistics.}

We have moved around these sections, so that measurement invariance comes before
the theoretical discussion.

\item \emph{For readers unfamiliar with case-based scores, it may be useful to
have a table with artificial values for some $B(\theta)_{ij}$ values; say for 20
subjects, and 4 parameters; in an additional column, there could be ordered
continuous values for a V variable; in the next column, the V values could be
binned to get an ordinal variable; referring to this table, it would be trivial
to illustrate how the values of eg DM, CvM, maxLM, etc are computed...}

We agree that this would be useful for some readers, but we think that it would
take up more space than it is worth.  For example, the calculation of ordinal
test statistics would require both calculations within each bin and unique
scaling (by $(i/n) \cdot (1 - i/n)$) for each bin.  While these calculations are not
difficult, they also consume space to write them out (especially if we included
the non-ordinal statistics).  Additionally, besides the calculation of test
statistics, it might be equally useful to show how the scores $s(\theta; x_i)$ can
be used to calculate $B(\theta)$.  This implies that we would need both a table of
$s(\theta; x_i)$ and a table of cumulative scores $B(\theta)$.  We feel that these
ideas are represented more concisely and intuitively via their equations.

\item \emph{related to 2), I would suggest adding (perhaps in an appendix) some
R code illustrating how (for example, for the values in the aforementioned
table) the p-values can be computed for the various test statistics}

We have added some practical discussion of p-value computation to the ``Choice of
Test" subsection of the General Discussion.  We have also posted all the R code
for replicating our results (p-values and otherwise) on our R-Forge page (listed
in the ``Computational Details" section).  We do not include code within the
manuscript because the p-values associated with some of the statistics are
computed via simulation (i.e., there is not a closed-form solution for critical
values associated with the test statistic), and much of the underlying R code is
already contained within R's strucchange package.
\end{enumerate}

All minor points have been addressed, with thanks to the reviewer.

\end{document}

### 2013-04-17 ordinal/response.Rnw
Final to ordinal ms prior to resubmission o Added YB97 statistic to response o Used apa6 class instead of apa class for the ms,   allowing for easy masking of authors
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@200 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{a4wide}
+\usepackage[english]{babel} % To avoid tildes
+\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}
 
@@
 described in the manuscript. I am also interested in reading the authors' ideas
-about the use MI in ``ad hoc" model fitting.}
+about the use MI in ``ad hoc'' model fitting.}
 
@@
 model misfit.  We have added some brief detail on this relationship to the
-"Tests for Continuous V" section.
+``Tests for Continuous V'' section.
 
@@
 LRT).  Under this framework, one would, e.g., reject the hypothesis of
-measurement invariance if the AIC for the ``full" model were less than the AIC
+measurement invariance if the AIC for the ``full'' model were less than the AIC
 for the restricted model.  We have added the AIC strategy to our simulation,
@@
 
-lev0 <- c(   "LM_o","LRT","LRT_sb","AIC")
-lev  <- c("maxLM_o","LRT","LRT_sb","AIC")
-cl <- c(1, 4, 2, 3)
-pc <- c(1, 2, 3, 4)
-lt <- c(1, 1, 2, 2)
+lev0 <- c(   "LM_o","LRT","LRT_sb","AIC","YB97")
+lev  <- c("maxLM_o","LRT","LRT_sb","AIC","YB97")
+cl <- c(1, 4, 2, 3, 5)
+pc <- c(1, 2, 3, 4, 5)
+lt <- c(1, 1, 2, 2, 2)
 
@@
     except that they contain different statistics: the ordinal maxLM test
-    ($\mathrm{LM}_o$), the Likelihood ratio test (LRT), the Satorra-Bentler LRT ($\mathrm{LRT}_\mathrm{sb}$),
-    and the use of AIC (AIC). The former two are presented in the manuscripts while the latter
-    two are only shown here.
+    ($\mathrm{maxLM}_o$), the Likelihood ratio test (LRT), the
+    Satorra-Bentler LRT ($\mathrm{LRT}_\mathrm{sb}$), the Yuan-Bentler
+    statistic (YB97), 
+    and the use of AIC (AIC). The first two are presented in the
+    manuscript while the last three are only shown here.
   \item It can be seen that $\mathrm{LRT}_\mathrm{sb}$ is virtually identical to
@@
     sample sizes $n$ (in particular throughout Simulation~2) while for the large $m$/small
-    $n$ scenarios , the problems of the LRT are even amplified.
+    $n$ scenarios, the problems of the LRT are even amplified.
+  \item The YB97 statistic is problematic at smaller $n$ and at large
+    $m$, generally resulting in 
+    very high Type I error rates.  We suspect that this is because the
+    statistic is based on models fits via WLS, which is generally
+    known to require very large $n$ to be reliable.  The extra
+    parameters from large $m$ would then seem to require even larger
+    $n$.  However, 
+    even in the Simulation~2 situations where the Type I error rate is
+    true, the performance of the statistic is similar to that of the plain LRT.
   \item Finally, the AIC has relatively low power throughout Simulation~1. This is
@@
 Hence, we feel that the additional discussion of these modifications do not add
-substantial new insights and hence have omitted them from the paper (and only
+substantial new insights and so have omitted them from the paper (and only
 mention them very briefly). But we are willing to reconsider if the AE or
@@
 
-\item \emph{related to 2), I would suggest adding (perhaps in an appendix) some
+\item \emph{Related to 2), I would suggest adding (perhaps in an appendix) some
 R code illustrating how (for example, for the values in the aforementioned
@@
 We have added some practical discussion of p-value computation to the ``Choice of
-Test" subsection of the General Discussion.  We have also posted all the R code
+Test'' subsection of the General Discussion.  We have also posted all the R code
 for replicating our results (p-values and otherwise) on our R-Forge page (listed
-in the ``Computational Details" section).  We do not include code within the
+in the ``Computational Details'' section).  We do not include code within the
 manuscript because the p-values associated with some of the statistics are

### 2013-04-22 ordinal/ordinal.Rnw
Final to ordinal ms prior to resubmission o Added YB97 statistic to response o Used apa6 class instead of apa class for the ms,   allowing for easy masking of authors
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@200 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Ordinal ms, unmasked the author names
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@201 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Final read of ordinal ms, minor wording changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@202 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
-\documentclass[man]{apa}
+\documentclass[man,apacite]{apa6}
 \usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
@@
 
-\acknowledgements{This work was supported by National Science
+\authornote{This work was supported by National Science
   Foundation grant SES-1061334. 
@@
 (see, e.g., \citeNP{Mil11} for a thorough review).  A set of
-psychometric scales $X$ are defined to be measurement invariant
+psychometric scales $X$ is defined to be measurement invariant
 with respect to an auxiliary variable $V$ if \cite{Mel89}
@@
 account for the fact that $V$ is ordinal.  More often, $V$ is 
-treated as categorical so that the traditional tests can be applied.  Alternatively, if there are many levels of $V$, $V$ may also be treated as continuous. 
-It is the intent of this paper to propose
+treated as categorical so that the traditional tests can be applied.
+Additionally, if there are many levels, then $V$ may also be treated
+as continuous.  
+The goals of this paper are to propose 
 two test statistics that explicitly treat $V$ as ordinal and to show
@@
 parameter values within each group are completely unrestricted.
-However, in situations where $V$ is categorical or
+However, in situations where $V$ is ordinal or
 continuous,~\eqref{eq:h2*} includes non-monotonic violations of
@@
 Researchers typically do not expect such a result when testing
-measurement invariance w.r.t.\ categorical or continuous $V$, and 
+measurement invariance w.r.t.\ ordinal or continuous $V$, and 
 researchers often cannot interpret such violations.  Monotonic
@@
 
-As stated previously, we specifically focus here on situations where $V$ is
+As stated previously, we specifically focus on situations where $V$ is
 ordinal and where the measurement invariance violation is related to the ordinal
@@
   {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
-    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
+    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
   \qquad (0 \le t \le 1)
@@
 \end{equation}
-where, again, tests of subsets of model parameters can be obtained by
+where the first subscript on the two terms in parentheses are $i_\ell$
+and $i_{\ell-1}$, respectively.  Again, tests of subsets of model
+parameters can be obtained by 
 taking the inner sum over only the $k^* < k$ parameters of interest.
@@
 in which the analyst has a large sample size, so that the usual
-likelihood ratio test is notoriously sensitive to minor parameter
+likelihood ratio test is overly sensitive to minor parameter
 instabilities (e.g., \citeNP{Ben80}).
@@
 invariance.  We do not report the latter results, because these test
-statistics performed worse than the usual likelihood ratio test (for
-both Simulation 1 and Simulation 2).
+statistics performed worse than the usual likelihood ratio test (both
+here and in Simulation 2).
 
@@
 ratio test requires estimation of a multiple-groups model, which is very
-unstable with large numbers of groups and small sample sizes as only
-$n/m$ observations are available in each subsample.  The proposed
-statistics are all of the LM-type and just require estimation of the
+unstable with large numbers of groups and small sample sizes (as only
+$n/m$ observations are available in each subsample).  The statistics
+proposed here
+are all of the LM-type and just require estimation of the
 single-group model, 
@@
 one's sample size is very large.  Analysts often resort to informal
-fit measures in this case, because the traditional LRT is nearly
+fit measures in practice, because the traditional LRT is nearly
 guaranteed to result in significance.  This simulation is intended to
@@
 received much research attention (for a review, see \citeNP{EmmMcc04}).
-Recently, scholars have begun to explore gratitude
-in youth. One potential problem with this research is that
-scholars, with no exception, have used adult gratitude inventories to
+Recently, researchers have begun to explore gratitude
+in youth. One potential problem with this is that
+researchers, with no exception, have used adult gratitude inventories to
 measure youth gratitude, thus raising the question of whether the
@@
 15 years, 16 years, and 17--19 years.  Thus, age is an ordinal variable
-with which the proposed tests can be applied.
+to which the proposed tests can be applied.
 
@@
 The authors generally found support for the tau-equivalent models
-through these alternative fit indices; the likelihood ratio test often
+through these alternative fit indices: the likelihood ratio test often
 resulted in significance even when the alternative indices indicated
@@
 differing pattern of values for the 15- and 16-year-olds can be taken
-as an indication that some factor loading is somewhat unstable at an
+as an indication that some factor loading is unstable at an
 age of 15 or 16, but this is not a clear and general trend across all the tested
@@
 this intermediate model fits as well as the original tau-equivalent model~($\chi^2_{16} = \Sexpr{round(lrt2.2te[[5]][2], digits = 2)}$, $p=\Sexpr{round(lrt2.2te[[7]][2], digits=2)}$),
-with 16 fewer parameters ($= 6 \cdot 4 - 2 \cdot 4$ because the four variances have to be estimated only in two rather than in six age groups).  The intermediate model also fits 
+with 16 fewer parameters ($= 6 \cdot 4 - 2 \cdot 4$ because the four
+variances have to be estimated in only two rather than six age groups).  The intermediate model also fits 
 better than the parallel model, as judged by a second likelihood ratio test~($\chi^2_{4} = \Sexpr{format(round(lrt2.2par[[5]][2], digits = 2), nsmall = 2)},
@@
 In this paper, we proposed two statistics that can be used when one
-has an auxiliary ordinal variable and wishes to study measurement
+has an ordinal auxiliary variable and wishes to study measurement
 invariance.  We demonstrated via simulation that these statistics have
@@
 individual arises from each model.  These treatments of ordinal $V$ as
-continuous are likely to be advantageous, especially if the levels of
+continuous will often be advantageous, especially if the levels of
 $V$ are approximately equally-spaced and the relationship between $V$
@@
 require models of greater complexity and may not be suitable for all
-ordinal $V$, however, while that the methods we propose here are
+ordinal $V$, however, while the methods we propose here are
 generally suitable for ordinal $V$.
@@
 auxiliary variable.  Hence, a new simulation usually must be conducted
-for each dataset, which can be somewhat time-consuming.
+for each dataset, which can be somewhat time-consuming (on the order
+of minutes, as opposed to seconds or hours).
 

### 2013-04-24 ordinal/ordinal.Rnw
try to explan i_0
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@203 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 where the first subscript on the two terms in parentheses are $i_\ell$
-and $i_{\ell-1}$, respectively.  Again, tests of subsets of model
-parameters can be obtained by 
+and $i_{\ell-1}$, respectively (and for $i_0 = 0$ the cumulative score is
+${\bm B}(0, \hat {\bm \theta}) = 0$).
+Again, tests of subsets of model parameters can be obtained by 
 taking the inner sum over only the $k^* < k$ parameters of interest.

### 2013-04-25 ordinal/ordinal.Rnw
Minor changes to ordinal ms to remove color from figures.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@204 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2013-04-22 ordinal/ordinal.Rnw
@@
 where the first subscript on the two terms in parentheses are $i_\ell$
-and $i_{\ell-1}$, respectively.  Again, tests of subsets of model
-parameters can be obtained by 
+and $i_{\ell-1}$, respectively (and where we take $i_0 = 0$, so that
+the cumulative score is ${\bm B}(0, \hat {\bm \theta}) = \bm{0}$).
+Again, tests of subsets of model parameters can be obtained by 
 taking the inner sum over only the $k^* < k$ parameters of interest.
@@
 plot(gefp1, functional = ordwmax(gefp1), axes = FALSE, main = "",
-     xlab = "Age group", ylab = "Weighted max statistics")
+     xlab = "Age group", ylab = "Weighted max statistics",
+     boundary = FALSE)
+lines(get_boundary(gefp1, ordwmax(gefp1)), col = "slategray", lwd = 1.7)
 axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
@@
 plot(gefp1, functional = ol2bb1.fun, axes = FALSE, ylim = c(0, 14), main = "",
-     xlab = "Age group", ylab = "LM statistics")
+     xlab = "Age group", ylab = "LM statistics", boundary = FALSE)
+lines(get_boundary(gefp1, ol2bb1.fun), col = "slategray", lwd = 1.7)
 axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
@@
 plot(gefp2, functional = ordwmax(gefp2), axes = FALSE, main = "",
-     xlab = "Age group", ylab = "Weighted max statistics")
+     xlab = "Age group", ylab = "Weighted max statistics",
+     boundary = FALSE)
+lines(get_boundary(gefp2, ordwmax(gefp2)), col = "slategray", lwd = 1.7)
 axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
@@
 plot(gefp2, functional = ol2bb2.fun, axes = FALSE, ylim = c(0, 115), main = "",
-     xlab = "Age group", ylab = "LM statistics")
+     xlab = "Age group", ylab = "LM statistics", boundary = FALSE)
+lines(get_boundary(gefp2, ol2bb2.fun), col = "slategray", lwd = 1.7)
 axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))

### 2013-04-26 ordinal/preprint.Rnw
renamed report.Rnw -> preprint.Rnw ported ordinal.Rnw into preprint.Rnw changed to title case and fixed a few other fields in new .bib entries
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@206 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2013-04-25 ordinal/ordinal.Rnw
@@
-\documentclass[man,apacite]{apa6}
-\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
-%% This keeps the tildes out when you use Rnw2pdf:
-\usepackage[english]{babel}
+\documentclass[nojss,noheadings]{jss}
+\usepackage{amsmath,alltt,setspace,bm,thumbpdf}
 %% need no \usepackage{Sweave}
 \SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}
-
-\title{Testing for measurement invariance with respect to an ordinal variable}
-\threeauthors{Edgar C.\ Merkle}{Jinyan Fan}{Achim Zeileis}
-\threeaffiliations{University of Missouri}{Auburn University}{Universit\"{a}t Innsbruck}
-
-\abstract{
+\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+
+\title{Testing for Measurement Invariance with Respect to an Ordinal Variable}
+\Shorttitle{Testing for Ordinal Measurement Invariance}
+
+\author{Edgar C. Merkle\\University of Missouri
+   \And Jinyan Fan\\Auburn University
+   \And Achim Zeileis\\Universit\"{a}t Innsbruck}
+\Plainauthor{Edgar C. Merkle, Jinyan Fan, Achim Zeileis}
+
+\Abstract{
   Researchers are often interested in testing for 
@@
 
-\authornote{This work was supported by National Science
-  Foundation grant SES-1061334. 
-  Correspondence to Edgar C.\ Merkle, Department of  
-  Psychological Sciences, University of Missouri, Columbia,
-  MO 65211.
-  Email: {\texttt{merklee@missouri.edu}}.}
-\shorttitle{Ordinal measurement invariance}
-\rightheader{Ordinal measurement invariance}
-
-\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
-
-%% for internal use
-\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
-\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}
-
-\spacing{1}
+\Keywords{measurement invariance, ordinal variable, parameter stability, factor analysis, structural equation models}
+
+\Address{
+  Edgar C. Merkle\\
+  Department of Psychological Sciences\\
+  University of Missouri\\
+  Columbia, MO 65211, United States of America\\
+  E-mail: \email{merklee@missouri.edu}\\
+
+  Jinyan Fan\\
+  Department of Psychology\\
+  Auburn University\\
+  225 Thach\\
+  Auburn, AL 36849-5214, United States of America\\
+  E-mail: \email{jzf0007@auburn.edu}\\
+
+  Achim Zeileis\\
+  Department of Statistics\\
+  Faculty of Economics and Statistics\\
+  Universit\"at Innsbruck\\
+  Universit\"atsstr.~15\\
+  6020 Innsbruck, Austria\\
+  E-mail: \email{Achim.Zeileis@R-project.org}\\
+  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
+}
+
+%% publication info and copyright notice
+\usepackage{fancyhdr}
+\setlength{\headheight}{15pt}
+\renewcommand{\headrulewidth}{0pt}
+\pagestyle{fancy}
+\thispagestyle{plain}
+\fancyhf{}
+\fancyhead[RO,LE]{\thepage}
+\fancyhead[CO]{\textit{\normalsize Edgar C. Merkle, Jinyan Fan, Achim Zeileis}}
+\fancyhead[CE]{\textit{\normalsize Testing for Ordinal Measurement Invariance}}
+\fancyfoot[LO,LE]{\small Copyright {\copyright} 2013 The Psychometric Society}
+\fancypagestyle{plain}{
+\fancyhf{}
+\fancyfoot[LO,LE]{\small This is a preprint of an article accepted for publication in \textit{Psychometrika}.\\ %%, \textbf{78}(1), 59--82.\\
+  Copyright~{\copyright} 2013 The Psychometric Society \url{http://www.psychometrika.org/}}
+  %% \doi{10.1007/S11336-012-9302-4}}
+}
 
 \begin{document}
-\maketitle
 
@@
 has received considerable attention in the psychometric literature
-(see, e.g., \citeNP{Mil11} for a thorough review).  A set of
+(see, e.g., \citealp{Mil11} for a thorough review).  A set of
 psychometric scales $X$ is defined to be measurement invariant
-with respect to an auxiliary variable $V$ if \cite{Mel89}
+with respect to an auxiliary variable $V$ if \citep{Mel89}
 \begin{equation}
@@
 to carry out this study include likelihood ratio tests, Lagrange
-multiplier tests, and Wald tests (e.g., \citeNP{Sat89}).
+multiplier tests, and Wald tests (e.g., \citealp{Sat89}).
 These tools have greatly aided in
@@
 that were recently applied to the study of measurement invariance in
-psychometric models \cite{MerZei13,StrKop13}.
+psychometric models \citep{MerZei13,StrKop13}.
 In the following section, we provide an overview of the family and
@@
 
-\section{Measurement Invariance}
+\section{Measurement invariance}
 
@@
 proposed statistics; for a more detailed
-account, see \citeA{MerZei13}.
-
-\subsection{Model Estimation}
+account, see \cite{MerZei13}.
+
+\subsection{Model estimation}
 We focus specifically on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises 
@@
   Under
-the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
+the usual regularity conditions (e.g., \citealp{Fer96}), the model parameters
 $\bm{\theta}$ can 
@@
 related to those underlying the Lagrange multiplier test and are
-discussed in detail by \citeA{MerZei13}.  Additionally, these ideas
+discussed in detail by \cite{MerZei13}.  Additionally, these ideas
 are related to those underlying modification indexes (e.g.,
-\citeNP{Sor89}): the modification index is equivalent to a Lagrange
+\citealp{Sor89}): the modification index is equivalent to a Lagrange
 multiplier test, and the Lagrange multiplier test is contained in the
-family described by \citeA{MerZei13}.  Here,
+family described by \cite{MerZei13}.  Here,
 we focus on the tests' properties that are relevant for extending them
@@
 estimator to guard the inference against potential misspecification of
-the model (see \citeNP{hub67} for the theoretical foundation and
-\citeNP{zei06b} for a computational framework).  
+the model (see \citealp{hub67} for the theoretical foundation and
+\citealp{zei06b} for a computational framework).  
 Equation~\eqref{eq:cumscore} simultaneously accounts for the
@@
 invariance when multiple factor analysis parameters change
-simultaneously \cite{MerZei13}.
+simultaneously \citep{MerZei13}.
 
@@
 scaled by the asymptotic variance $t (1 - t)$ of the process
-${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeA{MerZei13} found 
+${\bm B}(t, \hat {\bm \theta})$.  In simulations, \cite{MerZei13} found 
 that both tests perform well when assessing simultaneous changes
@@
 distribution with standard normal marginals and this particular correlation
-matrix; see also \citeA{hotzei08} for more details. In R, this can be
-accomplished using the mvtnorm package \cite{mvtnorm}.
+matrix; see also \cite{hotzei08} for more details. In {R}, this can be
+accomplished using the {mvtnorm} package \citep{mvtnorm}.
 
@@
 simulation of Brownian bridges.  This functionality is built in to
-R's strucchange package \cite{Zei06}, which can be used to
+{R}'s {strucchange} package \citep{Zei06}, which can be used to
 generally carry out the tests. Note that for models with only a single
@@
 likelihood ratio test is overly sensitive to minor parameter
-instabilities (e.g., \citeNP{Ben80}).
-
-\section{Simulation 1: Detecting Ordinal Invariance Violations}
+instabilities (e.g., \citealp{Ben80}).
+
+\section{Simulation 1: Detecting ordinal invariance violations}
 
@@
 simultaneously across the unique variances, which is similar to a test
-of ``invariant uniquenesses'' (see \citeNP{VanLan00}). % Or comparing
+of ``invariant uniquenesses'' (see \citealp{VanLan00}). % Or comparing
                                         % parallel to tau-equivalent
@@
 we also considered the
-Satorra-Bentler \citeyear{SatBen01} scaled likelihood ratio test
-statistic with correction for difference testing, the Yuan-Bentler
-\citeyear{YuaBen97} scaled test statistic, and the use of AIC
-\cite{Aka74} for detecting measurement 
+\cite{SatBen01} scaled likelihood ratio test
+statistic with correction for difference testing, the
+\cite{YuaBen97} scaled test statistic, and the use of AIC
+\citep{Aka74} for detecting measurement 
 invariance.  We do not report the latter results, because these test
 statistics performed worse than the usual likelihood ratio test (both
-here and in Simulation 2).
+here and in Simulation~2).
 
@@
 
+\pagebreak
+
 \subsection{Results}
+
+\begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.84\textwidth}
+<<sim1res, fig=TRUE, height=7, width=7>>=
+sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
+sim1.tmp$test <- factor(sim1.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+\caption{Simulated power curves for the ordered and unordered
+      $\max\mathit{LM}$ tests, the ordered double-max test, and the
+      likelihood ratio test across three sample sizes $n$, three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).
+\label{fig:sim1res}}
+\end{figure}
 
@@
 at specific levels of the ordinal variable
-(see \citeNP{MerZei13}, for related discussion), while the
+(see \citealp{MerZei13}, for related discussion), while the
 three other tests under consideration make use of this information.
@@
 %% parameters, at least according to AIC.
-
-\begin{figure}
-\caption{Simulated power curves for the ordered and unordered
-      $\max\mathit{LM}$ tests, the ordered double-max test, and the
-      likelihood ratio test across three sample sizes $n$, three
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
-\label{fig:sim1res}
-\setkeys{Gin}{width=\textwidth}
-<<sim1res, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
-sim1.tmp$test <- factor(sim1.tmp$test)
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-\end{figure}
 
@@
 
+\begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.84\textwidth}
+<<sim2res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+\caption{Simulated power curves for the ordered and unordered
+      $\max\mathit{LM}$ tests, the ordered double-max test, and the
+      likelihood ratio test across three sample sizes $n$, three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
+      occurring at a single level (the $(1 +
+      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.
+\label{fig:sim2res}}
+\end{figure}
+
 The more interesting result of Figure~\ref{fig:sim2res} lies in the comparison
@@
 detect the measurement invariance violation, illustrating the result 
-of \citeA{Ben80} and others that the
+of \cite{Ben80} and others that the
 likelihood ratio test statistic picks out minor 
@@
 invariance violation that occurs only at a single level of $V$.
-
-\begin{figure}
-\caption{Simulated power curves for the ordered and unordered
-      $\max\mathit{LM}$ tests, the ordered double-max test, and the
-      likelihood ratio test across three sample sizes $n$, three
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
-      occurring at a single level (the $(1 +
-      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
-\label{fig:sim2res}
-\setkeys{Gin}{width=\textwidth}
-<<sim2res, fig=TRUE, height=7, width=7>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-\end{figure}
 
@@
 which can significantly increase computation time.  We return to this
-issue in the General Discussion.
+issue in the Section~\ref{sec:discussion}.
 
@@
 
-\section{Application: Youth Gratitude}
+\vspace*{-0.2cm}
+
+\section{Application: Youth gratitude}
+
+\vspace*{-0.2cm}
 
@@
 With the positive psychology movement, the construct of gratitude has
-received much research attention (for a review, see \citeNP{EmmMcc04}).
+received much research attention (for a review, see \citealp{EmmMcc04}).
 Recently, researchers have begun to explore gratitude
@@
 existing gratitude scales used with adults are valid in research with
-youth. Addressing this issue, \citeA{FroFan11} had a
+youth. Addressing this issue, \cite{FroFan11} had a
 large sample of 
@@
 
-\citeA{FroFan11} used confirmatory factor 
+\cite{FroFan11} used confirmatory factor 
 models to study the invariance of three youth gratitude scales across
@@
 tests were overly sensitive at their sample size.  To supplement these
-tests, \citeA{FroFan11} examined a set of alternative fit indices, including
+tests, \cite{FroFan11} examined a set of alternative fit indices, including
 the non-normed fit index, the comparative fit index, and the
-incremental fit index (e.g., \citeNP{BroCud93}).
+incremental fit index (e.g., \citealp{BroCud93}).
 The authors generally found support for the tau-equivalent models
@@
 
-In the analyses described below, we re-analyze the \citeA{FroFan11} data using
+In the analyses described below, we re-analyze the \cite{FroFan11} data using
 the ordinal test statistics proposed in this paper.  This results in a
@@
 while being more sensitive to ordinal violations of measurement invariance.
-We focus on two analyses from \citeA{FroFan11} where the likelihood
+We focus on two analyses from \cite{FroFan11} where the likelihood
 ratio test resulted in significance (indicating that the restricted
@@
 
-Of the 1401 cases originally collected by \citeA{FroFan11}, we use here all
+Of the 1401 cases originally collected by \cite{FroFan11}, we use here all
 subjects with complete data (resulting in $n=1327$).
@@
 
-\begin{figure}
-\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
-  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GQ-6 data.}
-\label{fig:gq6res}
-\setkeys{Gin}{width=\textwidth}
-<<gq6res, fig=TRUE, height=3, width=6>>=
-par(mfrow = c(1, 2), cex = 0.65)
-
-plot(gefp1, functional = ordwmax(gefp1), axes = FALSE, main = "",
-     xlab = "Age group", ylab = "Weighted max statistics",
-     boundary = FALSE)
-lines(get_boundary(gefp1, ordwmax(gefp1)), col = "slategray", lwd = 1.7)
-axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
-axis(2)
-
-plot(gefp1, functional = ol2bb1.fun, axes = FALSE, ylim = c(0, 14), main = "",
-     xlab = "Age group", ylab = "LM statistics", boundary = FALSE)
-lines(get_boundary(gefp1, ol2bb1.fun), col = "slategray", lwd = 1.7)
-axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
-axis(2)
-@
-\end{figure}
-
 \subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
-model to the GQ-6 data, \citeA{FroFan11} used alternative fit indices to
+model to the GQ-6 data, \cite{FroFan11} used alternative fit indices to
 conclude that the tau-equivalent model was as good as the congeneric
@@
 and neither is significant at $\alpha = 0.05$,
-which supports the conclusions that \citeA{FroFan11} obtained from
+which supports the conclusions that \cite{FroFan11} obtained from
 alternative fit statistics.  This provides further evidence that 
@@
 up some non-systematic dependence on age.
+
+\begin{figure}[t!]
+\setkeys{Gin}{width=\textwidth}
+<<gq6res, fig=TRUE, height=3, width=6>>=
+par(mfrow = c(1, 2), cex = 0.65)
+
+plot(gefp1, functional = ordwmax(gefp1), axes = FALSE, main = "",
+     xlab = "Age group", ylab = "Weighted max statistics",
+     boundary = FALSE)
+lines(get_boundary(gefp1, ordwmax(gefp1)), col = "slategray", lwd = 1.7)
+axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
+axis(2)
+
+plot(gefp1, functional = ol2bb1.fun, axes = FALSE, ylim = c(0, 14), main = "",
+     xlab = "Age group", ylab = "LM statistics", boundary = FALSE)
+lines(get_boundary(gefp1, ol2bb1.fun), col = "slategray", lwd = 1.7)
+axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
+axis(2)
+@
+\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
+  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GQ-6 data.
+\label{fig:gq6res}}
+\end{figure}
 
@@
 
-\begin{figure}
-\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
-  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GAC data.}
-\label{fig:gacres}
+\subsubsection{GAC} In fitting tau-equivalent and parallel models to the
+GAC data, \cite{FroFan11} obtained mixed results.  The alternative fit
+indices did not all agree with one another, and the likelihood ratio test
+indicated that the parallel model fit worse than the tau-equivalent
+model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2], digits = 2)}, p<0.01$
+for the data considered here).  \cite{FroFan11} ultimately 
+concluded that the tau-equivalent model provided a better fit than did
+the parallel model.
+
+\begin{figure}[t!]
 \setkeys{Gin}{width=\textwidth}
@@
 @
+\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and
+  the $\max\mathit{LM}_o$ statistic (right panel), arising from the GAC data.
+\label{fig:gacres}}
 \end{figure}
-
-\subsubsection{GAC} In fitting tau-equivalent and parallel models to the
-GAC data, \citeA{FroFan11} obtained mixed results.  The alternative fit
-indices did not all agree with one another, and the likelihood ratio test
-indicated that the parallel model fit worse than the tau-equivalent
-model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2], digits = 2)}, p<0.01$
-for the data considered here).  \citeA{FroFan11} ultimately 
-concluded that the tau-equivalent model provided a better fit than did
-the parallel model.
 
@@
 
-\section{General Discussion}
+\section{General discussion} \label{sec:discussion}
 
@@
 (especially if $V$ has very many levels).  
-As described in detail by \citeA{MerZei13}, we
+As described in detail by \cite{MerZei13}, we
 can also use cumulative score processes with continuous $V$,
@@
 large, the proposed~$\max\mathit{LM}_o$ statistic will be very close
-to the $\max\mathit{LM}$ statistic described in \citeA{MerZei13}.
+to the $\max\mathit{LM}$ statistic described in \cite{MerZei13}.
 Thus, the formation of ordinal age groups (or other variables) is not
@@
 As a result, applications to studying DIF in IRT are immediate (for a
-presentation involving non-ordinal variables, see \citeNP{StrKop13}),
+presentation involving non-ordinal variables, see \citealp{StrKop13}),
 as are general psychometric applications to studying parameter
@@
 proposed test statistics are better than the LRT for detecting
-monotonic instabilities.   The strucchange package \cite{Zei06} noted
+monotonic instabilities.   The {strucchange} package \citep{Zei06} noted
 previously can be used for these more-general applications.

### 2013-05-15 m3_2013/ord_invariance.tex
Added presentation files for m3 conference.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@208 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2012-07-07 imps2012/invariance.tex
@@
 \usepackage{beamerthemeHannover,graphics,amsmath,bm}
+\usepackage{Sweave}
 
-\title[Measurement Invariance]{Tests for Measurement Invariance When Subgroups Are Unknown}
-\author{Ed Merkle\inst{1} \and Achim Zeileis\inst{2}}
-\institute{\inst{1} University of Missouri \and \inst{2} Universit\"{a}t Innsbruck}
-\date{Supported by grant SES-1061334 from the National
-  Science Foundation}
+\title[Ordinal Invariance]{Testing for Measurement Invariance with
+  respect to an Ordinal Variable}
+\author{Ed Merkle\inst{1} \and Jinyan Fan\inst{2} \and Achim Zeileis\inst{3}}
+\institute{\inst{1} University of Missouri \and \inst{2} Auburn
+  University \and \inst{3} Universit\"{a}t Innsbruck}
+\date{\ }
+%\date{Supported by NSF grant SES-1061334}
 
@@
 \frame{\titlepage}
-
-\frame{
-  \frametitle{}
-  \tableofcontents}
 
@@
     \item Commonly studied via factor analysis (today's focus) and
-      item response models.
+      item response models.\\ \ \\
+    \item Focal problem: How do we test for measurement invariance
+      with respect to an ordinal auxiliary variable, in a way that
+      ``respects'' the fact that the variable is ordinal?
   \end{itemize}}
@@
 \frame{
-  \frametitle{Example (Age $\leq$ 16)}
-  % path diagram illustrating parameters being the same for everyone
+  \frametitle{Example (13--14 years)}
   \begin{center}
@@
 \frame{
-  \frametitle{Example (Age~$>$~16)}
-  % path diagram illustrating parameters being the same for everyone
+  \frametitle{Example (15--16 years)}
   \begin{center}
@@
 
+\frame{
+  \frametitle{Example (17 years)}
+  \begin{center}
+      \includegraphics[height=3in]{../psychoco2012/model_over16.pdf}
+  \end{center}}
+
+\frame{
+  \frametitle{Outline}
+  \begin{itemize}
+    \item The traditional way to test for measurement invariance here
+      involves multiple-group models and likelihood ratio tests.
+      However, these tests treat the ordinal variable as unordered.
+      In this talk, we:
+      \begin{itemize}
+        \item Review last year's family of measurement invariance
+          tests, which can be 
+          used when the auxiliary variable is continuous.
+        \item Present new test statistics to deal with situations when the
+          auxiliary variable is ordinal.
+        \item Illustrate the statistics and compare them to
+          traditional methods.
+      \end{itemize}
+  \end{itemize}}
+
+\section{Theory}
 % Define as theta_i = theta_0, i=1,..,n
@@
       due to all the ways by which individuals may differ.\\ \ \\
-    \item We typically place people into groups based on a meaningful
-      auxiliary variable, then study measurement invariance across
-      those groups (via Likelihood Ratio tests, Lagrange multiplier
-      tests, Wald tests).\\ \ \\
-    \item If we did not know the groups in advance, we could conduct
-      a LR or LM test for each possible grouping, then take the maximum.
-      Requires different critical values! (Can be obtained from proposed
-      tests.)
+    \item If our auxiliary variable of interest is continuous (i.e.,
+      we do not know the groups in advance), we could conduct 
+      a LR or LM test for each possible pair of groups, then take the maximum.
+      Requires different critical values!
   \end{itemize}}
@@
 
-\section{Tests along a Continuum}
-% \frame{
-%   \frametitle{Proposed Tests}
-%   \begin{itemize}
-%     \item In contrast to existing tests of measurement invariance, the
-%       proposed tests offer the abilities to:
-%       \begin{itemize}
-%         \item Test for measurement invariance when groups are
-%           ill-defined (e.g., when the grouping variable is continuous).
-%         \item Test for measurement invariance in any subset of model
-%           parameters.
-%         \item Interpret the nature of measurement invariance violations.
-%       \end{itemize}
-%   \end{itemize}}
+\frame{
+  \frametitle{Instability Tests}
+  \begin{itemize}
+    \item The tests presented last year expand on this idea, removing
+      the need to fit more than one model.\\ \ \\
+    \item Idea: Fit a ``reduced'' model to the data (assuming, e.g.,
+      that parameters are equal across groups).  Examine how
+      individuals' {\em scores} vary across the continuous auxiliary
+      variable.
+  \end{itemize}}
 
 \frame{
-  \frametitle{Proposed Tests}
+  \frametitle{Scores}
   \begin{itemize}
-    \item We propose a number of tests for measurement invariance
-      along a continuous auxiliary variable.
-    \item Tests are based on individual terms of the gradient ({\em scores}).  The scores tell us how well a particular parameter describes a particular individual.
-  % Equation of gradient vs scores for individuals
+    \item What are scores?
+   \begin{itemize}
+    \item The traditional way to obtain maximum likelihood estimates
+      involves choosing estimates $\widehat{\bm{\theta}}$ so that\\
+  \begin{center}
+  $\frac{\partial}{\partial \bm{\theta}} \displaystyle\sum_{i=1}^{n} \log
+\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}} = 
+  {\bm 0}$
+  \end{center}\ \\ \ \\
+    \item Individual terms of this summation are the scores\\
+  \begin{center}
+  $s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} \log
+\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} =
+  \widehat{\bm{\theta}}}$
+  \end{center}\ \\ \ \\
+    \item $s(\hat{{\bm \theta}}; \bm{x}_i)$ can be conceptualized 
+      as a residual for individual $i$: close to ${\bm 0}$ is good,
+      far from ${\bm 0}$ is not so good. 
   \end{itemize}
-\begin{eqnarray*}
-  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
-  {\bm 0},\text{ where} \\
- \\
-  s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} \log
-\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
-\end{eqnarray*}}
+ \end{itemize}}
 
 \frame{
-  \frametitle{Proposed Tests}
+  \frametitle{Instability Tests}
   \begin{itemize}
@@
   \begin{itemize}
-    \item We need a way to aggregate scores across people so that we
-      can draw some general conclusions.
+    \item We need a way to aggregate scores across individuals so that we
+      can draw some general conclusions.  We define a {\em cumulative score
+      process}, which is advantageous because we know its distribution
+      under $H_0$:
       \begin{itemize}
@@
         \item Define $t \in (0, 1)$.
-          The {\em empirical cumulative score process} is defined by:\\
+          The empirical cumulative score process is defined by:\\
   \begin{equation*}
@@
   the observed information matrix ${\bm{I}}(\widehat{{\bm{\theta}}})$.
-
       \end{itemize}
@@
 
-    \item Test statistics: Special cases include double maximum (DM), Cram\'er-von Mises
-      (CvM), maximum of LM statistics.
+    \item Test statistics: Special cases include double maximum (DM),
+      Cram\'er-von Mises (CvM), maximum of LM statistics.
   \end{itemize}
@@
 \begin{itemize}
-  \item Test statistics aggregate over parameters $j = 1, \dots, k$ and observations $i = 1, \dots, n$,
+  \item Test statistics aggregate over parameters (columns) $j = 1,
+    \dots, k$ and observations (rows) $i = 1, \dots, n$,
     employing different norms.
@@
 \end{itemize}
-
 \end{frame}
@@
 \frame{
-  \frametitle{Simulation}
+  \frametitle{Illustration}
+  \begin{center}
+      \includegraphics[height=3in]{../psychoco2012/paper-lm-lr-19.pdf}
+  \end{center}
+}
+
+\section{Ordinal Tests}
+\frame{
+  \frametitle{Ordinal vs Continuous Tests}
   \begin{itemize}
-    \item Simulation: What is the power of the proposed tests?\\ \ \\
-      \begin{itemize}
-        \item Two-factor model, with three indicators each.
-        \item Measurement invariance violation in three factor loading parameters, with magnitude from 0--4 standard errors.
-        \item Sample size in $\{100, 200, 500\}$.
-        \item Model parameters tested in $\{3, 19\}$.
-        \item Three test statistics.
-      \end{itemize}
+    \item Last year, we showed that the above tests have adequate
+      power to be useful in practice.\\ \ \\
+    \item However, the tests require a unique ordering of individuals
+      w.r.t.\ the auxiliary variable.\\ \ \\
+    \item If the auxiliary variable is ordinal, we have a large number
+      of ties.  What to do?
   \end{itemize}}
@@
 \frame{
-  \frametitle{Simulation}
-  \begin{center}
-      \includegraphics[height=3in]{../psychoco2012/paper-mzsim-xyplot.pdf}
-  \end{center}
-}
-
-\section{Tests along an Ordinal Variable}
-\frame{
   \frametitle{Tests along Ordinal Variables}
   \begin{itemize}
-    \item The statistics described above do not immediately translate
-      to ordinal auxiliary variables, where there is no unique ordering of
-      all individuals.
-    \item To obtain a test statistic, we allow all individuals with the
-      same value of the 
+    \item To obtain a test statistic in the ordinal case, we allow all
+      individuals with the same value of the 
       auxiliary variable to simultaneously enter into the cumulative
-      sum.  We then take the maximum of the cumulative sum.
+      sum.  We then apply the same functionals that were used in the
+      continuous case.
     \item Critical values are obtained by summing bins of a Brownian
       bridge, where bin sizes match the observed bin sizes of the 
-      ordinal variable.
+      ordinal variable.  This allows us to treat ordinal auxiliary
+      variables as truly ordinal, which no existing methods (to our
+      knowledge) can do.
   \end{itemize}}
@@
     \item Tests: Obtain ordinal versions of the max LM test and (weighted) double maximum test.
-      Additionally use LM-type test that does not exploit ordering.
+      % Additionally use LM-type test that does not exploit ordering.
     \begin{eqnarray*}
@@
                              \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,\\
-      \mathit{LM}_\mathit{uo} & = & \sum_{\ell = 1, \dots, m+1} \sum_{j = 1, \dots, k}
-        \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2.
+                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.\\
+%      \mathit{LM}_\mathit{uo} & = & \sum_{\ell = 1, \dots, m+1} \sum_{j = 1, \dots, k}
+%        \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2.
     \end{eqnarray*}
@@
 
+\subsection{Simulations}
+
 \frame{
-  \frametitle{Simulation}
+  \frametitle{Simulation 1}
   \begin{itemize}
-    \item Simulation:
+    \item Simulation 1:
       \begin{itemize}
-        \item Same two-factor model from before.
+        \item Two-factor model, with each factor having three unique
+          indicators.
         \item Measurement invariance violation in three unique
@@
         \item Levels of the ordinal variable in $\{4, 8, 12\}$.
-        \item Three test statistics (two new statistics, plus the
-          multiple group LRT).
+        \item Four test statistics (two new statistics, the
+          multiple group LRT, and the unordered Lagrange multiplier test).
       \end{itemize}
@@
 \frame{
-  \frametitle{Simulation}
+  \frametitle{Simulation 1}
   % Graph of results
   \begin{center}
-      \includegraphics[height=3in]{../ordinal/ordsim-xyplot.pdf}
+      \includegraphics[height=3in]{ordinal-sim1res}
   \end{center}
@@
 
-% library("lattice")
-% load("simres2.rda")
-% res$nlevels <- factor(paste("nlev=",res$nlevels,sep=""),
-%   levels=c("nlev=4","nlev=8","nlev=12"))
-% res$nobs <- paste("n=",res$nobs,sep="")
-
-% pdf("ordsim-xyplot.pdf")
-% trellis.par.set(theme = canonical.theme(color = FALSE))
-% mykey <- simpleKey(c("LRT", "ordmaxLM", "orddwmax"), points = FALSE, lines = TRUE)
-% mykey$lines$lty <- c(2, 1, 3)
-% mykey$lines$lwd <- c(1.2, 1.2, 2)
-% print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data=
-% res[res$test %in% c("lrt","ordmax","ordwmax"),], type="l", xlab="Violation Magnitude", ylab="Power",key=mykey, ylim=c(0,1), lty=c(3,2,1), lwd=c(1.5,1.3,1)))
-
-% dev.off()
-
-\section{Conclusions}
 \frame{
-  \frametitle{Conclusions}
+  \frametitle{Simulation 2}
   \begin{itemize}
-    \item Measurement invariance tests utilizing stochastic processes have 
-      important advantages over existing tests:
+    \item Simulation 2: Are the proposed test statistics less
+      sensitive to large $n$, as compared to the analogous likelihood
+      ratio test (e.g., Bentler and Bonnett, 1980)?
       \begin{itemize}
-        \item Testing for measurement invariance along a continuum,
-          defining subgroups based on the test.
-        \item Testing for measurement invariance wrt an ordinal
-          variable, with sensitivity to ``ordinal'' violations.
-        \item Testing for measurement invariance at low $n$, where
-          multiple-group models are not feasible.
+        \item Same two-factor model from before.
+        \item Small measurement invariance violation (0, 0.5, \dots,
+          3 standard errors apart) in all unique
+          variance parameters at a middle level of the ordinal auxiliary
+          variable.
+        \item Sample size in $\{1200, 4800, 9600\}$.
+        \item Levels of the ordinal variable in $\{4, 8, 12\}$.
+        \item Four test statistics (two new statistics, the
+          multiple group LRT, and the unordered Lagrange multiplier test).
       \end{itemize}
@@
 \frame{
-  \frametitle{Software}
+  \frametitle{Simulation 2}
+  % Graph of results
+  \begin{center}
+      \includegraphics[height=3in]{ordinal-sim2res}
+  \end{center}
+}
+
+\frame{
+  \frametitle{Simulations}
   \begin{itemize}
-    \item We can carry out the tests for general SEMs in R, using
+  \item Conclusions from ordinal simulations
     \begin{itemize}
-        \item \texttt{lavaan} for model estimation.
-        \item \texttt{estfun()} for score extraction, which is
-          currently a combination of our own code and \texttt{lavaan}
-          code.
-        \item \texttt{strucchange} for carrying out the proposed tests
-          with the scores from \texttt{estfun()}.
+    \item In the ordinal scenarios examined, the ordered Lagrange multiplier
+      statistic has higher power to detect measurement invariance
+      violations than does the traditional LRT.
+    \item The ordered Lagrange multiplier statistic is also less
+      sensitive to minor invariance violations at large $n$, as
+      compared to the traditional LRT (the ordered double-max
+      statistic is also less sensitive, but its power is generally
+      poor).
+    \item The ordered Lagrange multiplier statistic has good potential
+      for practical application (though a downside is that critical
+      values must be simulated).
     \end{itemize}
@@
 
+\subsection{Example}
+
 \frame{
-  \frametitle{Current Work}
+  \frametitle{Example}
   \begin{itemize}
-    \item Detailed examination of test properties.\\ \ \\
-    \item Extension to related psychometric issues.\\ \ \\
-    %\item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
-    \item Paper: Merkle, E.\ C.\ \& Zeileis, A.\ (accepted).  Tests of measurement invariance without subgroups:\ {A} generalization of classical methods.  {\em Psychometrika}.
+    \item Example: A small number of scales have been developed to
+      measure gratitude in adults.  Can these scales also be used with
+      youth? (Froh, Fan, et al., 2011)
+      \begin{enumerate}
+        \item $n=1,401$ youth ages 10--19 complete three gratitude
+          scales.  Age groups are \{10--11, 12--13, 14, 15, 16,
+          17--19\}.
+        \item Measurement invariance examined via factor analyses of
+          each individual scale (one-factor models), with
+          increasingly-restricted parameters.
+      \end{enumerate}
   \end{itemize}}
+
+\frame{
+  \frametitle{Example}
+  \begin{itemize}
+    \item The large sample size made the authors suspicious about
+      significant likelihood ratio tests of model fit, making it difficult
+      to draw final conclusions about measurement invariance.
+      \begin{itemize}
+        \item GQ6 scale: Compare congeneric model to tau-equivalent model,
+          obtain $\chi^2_{20}=38.08, p=.009$
+        \item GAC scale: Compare tau-equivalent model to parallel model,
+          obtain $\chi^2_{20}=167.72, p<.001$\\ \ \\
+      \end{itemize}
+    \item To use our proposed statistics in each case, we fit the
+      restricted model and test the parameters that are freed in the
+      less-restricted model.
+  \end{itemize}}
+
+\begin{frame}[fragile]
+  \frametitle{GQ6 Results}
+  \begin{itemize}
+    \item LRT implies the congeneric model is better, whereas our
+      proposed statistics imply the tau-equivalent model is as good
+      ($WDM_{o}=2.91, p=.06$; $\max LM_o = 11.16, p=.10$)
+  \end{itemize}
+  \begin{center}
+      \includegraphics[height=2in]{ordinal-gq6res.pdf}
+  \end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{GAC Results}
+  \begin{itemize}
+    \item LRT implies the tau-equivalent model is better, and our
+      proposed statistics agree ($WDM_{o}=6.55, p<.01$; $\max LM_o =
+      113.13, p<.01$)
+  \end{itemize}
+  \begin{center}
+      \includegraphics[height=2in]{ordinal-gacres.pdf}
+  \end{center}
+\end{frame}
+
+\section{Conclusions}
+\frame{
+  \frametitle{General Conclusions}
+  \begin{itemize}
+    \item The family of measurement invariance tests considered here 
+      usefully extends to the ordinal case:
+      \begin{itemize}
+        \item Increased sensitivity to ``ordinal'' violations.
+        \item Decreased sensitivity to anomalous violations induced by
+          large $n$.
+        \item Ability to test at low $n$, where
+          multiple-group models are not feasible (the proposed tests
+          only require the ``reduced'' model).
+      \end{itemize}
+  \end{itemize}}
+
+\begin{frame}[fragile]
+  \frametitle{Software}
+  \begin{itemize}
+    \item R currently has functionality to carry out the proposed
+      tests for general SEMs:
+      \begin{itemize}
+        \item \texttt{lavaan} for model estimation, with score
+          extraction included in recent versions via \verb+estfun()+.\\ \ \\
+        \item \texttt{strucchange} for carrying out the proposed
+          tests.\\ \ \\
+        \item See \url{http://semtools.r-forge.r-project.org/} for
+          papers and relevant code.
+      \end{itemize}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Acknowledgements}
+  \begin{itemize}
+    \item Support from NSF grant SES-1061334\\ \ \\
+    \item Yves Rosseel, \verb+lavaan+ development\\ \ \\
+  \end{itemize}
+\end{frame}
 

### 2013-07-01 frontiers/abstract.tex
Created new folder for Frontiers paper, added abstract.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@210 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[11pt]{article}
\usepackage{hyperref,amsmath,bm,apacite}

\begin{document}

{\large
\begin{center}
\noindent {\em Title:} Score-based tests of measurment invariance: Use
in practice\\ \ \\

\noindent {\em Authors:} Ting Wang, Edgar Merkle, Achim Zeileis
\end{center}}

\pagestyle{empty}

% Paragraph 1: Background
%% General statement about measurement invariance

%% Summary of Merkle + Zeileis 2013

%% The tests have had only limited application to measurement
%% invariance, and their properties as applied to measurement
%% invariance are not fully known.

As the Research Topic indicates, measurement invariance is a critical
issue in psychometric modeling.  Recently, Merkle and
Zeileis \citeyear{MerZei13} studied a novel family of
score-based tests of 
measurement invariance for factor analysis models.  This family can
be used to test for measurement invariance w.r.t.\ a continuous
auxiliary variable, and the family is advantageous because it can
identify subgroups of observations and model parameters that violate
measurement invariance.  Merkle, Fan, and Zeileis \citeyear{MerFanZei}
extended this framework to testing measurement invariance w.r.t.\ an
ordinal 
auxiliary variable, yielding test statistics that are sensitive to
violations that are monotonically related to the auxiliary variable
(and insensitive to non-monotonic violations).

% Paragraph 2: Tutorial in lavaan
The papers cited in the above paragraph have been relatively technical
and slanted towards psychometricians.  In the proposed paper for the
Research Topic, we plan to focus on the 
test statistics' use in practice.  We will first provide a 
minimally-technical overview of the tests, along with a tutorial on
carrying out the tests using the R packages lavaan (for model
estimation) and strucchange (for measurement invariance testing).
These packages currently have all the functionality necessary for
carrying out the tests on arbitrary structural equation models, so the
tutorial will be especially useful for researchers who wish to apply
the tests to their own data.
%% Ting: I think the following details are too specific for the abstract.
% To carry out these tests, a confirmatory factor analysis model is fit to the data first. Then R package lavaan could
% provide casewise derivatives and observed information matrix, and then they are used to calculate the cumulative
% score process. In the end, by using different aggregation function of cumulative
% score process, we obtain three tests' statistics and \emph{p}-values, which could indicate whether or not measurement invariance is 
% violated.

% Paragraph 3: Simulations to examine issues that are of importance in
% practice.
Following the tutorial, we plan to conduct and report on
some novel simulations that are relevant to the tests' use in practice.  In
Simulation 1, we will study the extent to which the proposed
tests attribute measurement invariance violations to the correct model
parameters.  In Simulation 2, we will study the extent to which the
tests are robust to model misspecification.  Taken together, the
tutorial and simulations will provide researchers with both the means
to carry out the tests and the knowledge of when the tests are likely to
be useful in practice. 
%% Ting: For this proposed abstract, I think we do not need to discuss
%% any results.
% The resutls indicate that the tests
% could correctly locate the violation parameter, as long as the violation parameter is included
% in the model. If not, the tests show violation in more than one parameter associated with the
% particular manifest variable which is unduely modeled.
\bibliographystyle{apacite}
\bibliography{refs}

\end{document}

### 2013-07-08 frontiers/abstract.tex
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@211 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-As the Research Topic indicates, measurement invariance is a critical
-issue in psychometric modeling.  Recently, Merkle and
-Zeileis \citeyear{MerZei13} studied a novel family of
-score-based tests of 
-measurement invariance for factor analysis models.  This family can
-be used to test for measurement invariance w.r.t.\ a continuous
-auxiliary variable, and the family is advantageous because it can
-identify subgroups of observations and model parameters that violate
-measurement invariance.  Merkle, Fan, and Zeileis \citeyear{MerFanZei}
-extended this framework to testing measurement invariance w.r.t.\ an
-ordinal 
-auxiliary variable, yielding test statistics that are sensitive to
-violations that are monotonically related to the auxiliary variable
-(and insensitive to non-monotonic violations).
+As the Research Topic indicates, measurement invariance is a critical issue in
+psychometric modeling. Recently, Merkle and Zeileis \citeyear{MerZei13} studied
+a novel family of score-based tests for measurement invariance in factor
+analysis models. This family can be used to test for measurement invariance
+w.r.t.\ a continuous auxiliary variable without pre-specification of subgroups.
+Moreover, the family of tests can be employed to identify subgroups of
+observations and model parameters that violate measurement invariance. Merkle,
+Fan, and Zeileis \citeyear{MerFanZei} extended this framework to testing
+measurement invariance w.r.t.\ an ordinal auxiliary variable, yielding test
+statistics that are sensitive to violations that are monotonically related to
+the auxiliary variable (and much less sensitive to non-monotonic violations).
 
 % Paragraph 2: Tutorial in lavaan
-The papers cited in the above paragraph have been relatively technical
-and slanted towards psychometricians.  In the proposed paper for the
-Research Topic, we plan to focus on the 
-test statistics' use in practice.  We will first provide a 
-minimally-technical overview of the tests, along with a tutorial on
-carrying out the tests using the R packages lavaan (for model
-estimation) and strucchange (for measurement invariance testing).
-These packages currently have all the functionality necessary for
-carrying out the tests on arbitrary structural equation models, so the
-tutorial will be especially useful for researchers who wish to apply
-the tests to their own data.
+The papers cited in the above paragraph have been relatively technical and
+slanted towards psychometricians. In the proposed paper for the Research Topic,
+we plan to focus on the test statistics' use in practice. We will first
+provide a minimally-technical overview of the tests, along with a tutorial on
+carrying out the tests using the {R} packages \emph{lavaan} (for model estimation) and
+\emph{strucchange} (for measurement invariance testing). These packages currently have
+all the functionality necessary for carrying out the tests on arbitrary
+structural equation models, so the tutorial will be especially useful for
+researchers who wish to apply the tests to their own data.
+
 %% Ting: I think the following details are too specific for the abstract.
@@
 % practice.
-Following the tutorial, we plan to conduct and report on
-some novel simulations that are relevant to the tests' use in practice.  In
-Simulation 1, we will study the extent to which the proposed
-tests attribute measurement invariance violations to the correct model
-parameters.  In Simulation 2, we will study the extent to which the
-tests are robust to model misspecification.  Taken together, the
-tutorial and simulations will provide researchers with both the means
-to carry out the tests and the knowledge of when the tests are likely to
-be useful in practice. 
+Following the tutorial, we plan to conduct and report on some novel simulations
+that are relevant to the tests' use in practice. In Simulation~1, we will study
+the extent to which the proposed tests attribute measurement invariance
+violations to the correct model parameters. In Simulation~2, we will study the
+extent to which the tests are robust to model misspecification. Taken together,
+the tutorial and simulations will provide researchers with both the means to
+carry out the tests and the knowledge of when the tests are likely to be useful
+in practice. 
 %% Ting: For this proposed abstract, I think we do not need to discuss
@@
 % particular manifest variable which is unduely modeled.
+
 \bibliographystyle{apacite}

### 2013-07-08 frontiers/testprops1.Rnw
add writing files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@212 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Editing Rules:                                                            %
%1. Ed's comment and questions:\readme{}                                   %     
%2. Ting's comment and questions:\seeme{}                                  % 
%3. When comments and questions are resovled, the corresponding \readme or %  
%   \seeme will be removed by Ting.                                        %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[doc]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
\usepackage[english]{babel}

\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}

\title{Tests of measurement invariance via stochastic processes:
  {S}ome properties}
\twoauthors{Ting Wang and Edgar C.\ Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

\abstract{
We study a family of 
recently-proposed measurement invariance tests that are derived from
stochastic processes.  As applied to factor analysis and other
psychometric models, the tests are advantageous in that (i) they do
not require advance definition of subgroups of cases that violate
measurement invariance; (ii) they can potentially identify specific
model parameters that violate measurement invariance; and (iii) they
do not require a model of additional complexity to be estimated.  
Because the tests have been applied to only a small number of
psychometric examples, we conduct a series of simulation studies that provide
a detailed examination of test properties.  These studies allow us
to study (details here).  The studies indicate that.}

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. 
  Correspondence to. 
  Email: }
\shorttitle{Measurement invariance test properties}
\rightheader{Measurement invariance test properties}

\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
%% for internal use
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}
\newcommand{\seeme}[1]{\emph{\marginpar{SEEME} (#1)}}

\spacing{1}

\begin{document}
\maketitle

<<preliminaries1>>=
## packages
library("lavaan")
library("strucchange")
library("mvtnorm")
library("lattice")

## auxiliary code
source("simall.R")
source("mz.R")

## convenience function for plotting boundaries without color
get_boundary <- function(obj, fun, trim = 0) {
  bound <- fun$boundary(0:obj$nobs/obj$nobs)
  bound <- fun$computeCritval(0.05, ncol(obj$process)) * bound
  bound <- zoo(bound, time(obj$process))
  if(trim > 0) bound <- head(tail(bound, - floor(trim * obj$nobs)), - floor(trim * obj$nobs))
  return(bound)
}

## convenience function for applying tests to real data
info_full <- function(x, ...) solve(vcov(x) * nobs(x))
@

Parameter invariance is a fundamental assumption for measurement model in 
developing psychometric scales. Violation of it could happen in any estimated parameter.  
In order to fix the psychometric scale, we need to specify the violating parameter.  
Moreover, in practice model misspecification is an inevitable issue to a certain extent 
and may interfere with parameter invariance specificity. A family of tests derived from 
stochastic process has been recently proposed to identify specific parameters impacted 
by measurement invariance violation. However, these tests have only been used to a limited 
type of parameter violation. It is this paper's intent to provide a detailed examination 
of the tests' performance under practical research scenarios.

In the following section, we first briefly review the 
theoretical framework of proposed statistical tests. Subsequently, we study the tests'
ability through simulations that mimic practical research. 
Finally, some suggestion on the tests' use in practice is provided.

\section{Simulation 1}

In Simulation 1, we examined the extent to which the proposed tests
are sensitive to the specific model parameter that violates measurement
invariance.  If, say, a factor loading violates measurement
invariance, it is plausible that this violation impacts other
parameters such as factor covariances or other loadings 
associated with the factor in question.  Thus, the goal of the
simulation was to examine the extent to which the proposed tests
attribute the measurement invariance violation to the parameters that
are truly in violation.

\subsection{Methods}
To examine these specificity issues, we generated data from a
two-factor model with three indicators each (the same model used in
Merkle and Zeileis, \citeyearNP{MerZei13}).  The measurement invariance
violation occurred in one of three parameters: the factor loading
associated with the first indicator $\lambda_{11}$, the unique variance associated
with the first indicator $\epsilon_{11}$, or the factor covariance $\sigma_{12}$.  
We then tested for measurement invariance in seven subsets of parameters: the three
individual parameters noted above, all six factor loadings $\lambda_{11}--\lambda_{62}$, 
all six unique variances $\epsilon_{11}--\epsilon_{66}$. The path diagram is shown in  
Figure~\ref{fig:path}.

\begin{figure}
\caption{Path diagram representing the factor analysis model 
         used for simulations.}
\label{fig:path}
\includegraphics{PathDiagram}
\end{figure}

Power and Type I error were examined across three sample sizes
($n=100,200,500$) and 17 magnitudes of invariance violations. These
violations were based on the auxiliary variable $V$: individuals below
the 50th percentile of $V$ deviated from individuals above the 50th
percentile  by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter, 5000 datasets were
generated and tested.

\subsection{Results}
Full simulation results are presented in Figure~\ref{fig:sim11res} to
Figure~\ref{fig:sim13res}; all three test statistics exhibited very
similar results, so the figures include only results for the CvM statistic.
Figure~\ref{fig:sim11res} displays power curves as a function of 
violation magnitude for the first
factor loading $\lambda_{11}$, with panels reflecting the parameters 
being tested and lines reflecting $n$.
Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
similar power curves when the factor covariance 
$\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
invariance, respectively.

From these figures, one generally observe that the tests isolate 
the parameter violating measurementinvariance.  Additionally, the tests 
have somewhat-higher power to detect measurement invariance violations 
in the factor loading and factor covariance parameters, as opposed to the error variance
parameter.  Finally, simultaneous tests of all factor loadings or all error
parameters result in decreased power, as compared to the situation
where one tests only the violating parameter.  This occurs because, in
testing a subset of parameters (only one of which violates measurement
invariance), we are effectively dampening the signal of a measurement
invariance violation.

In summary, we found that the proposed tests can attribute
measurement invariance violations to the correct parameter.  This
provides evidence that, in practice, one can have confidence in the
tests' abilities to locate the measurement invariance violation.  Of
course, this statement is qualified by the fact that, in this
simulation, the model was correctly specified.  In the following
simulations, we examine the tests' performance when the model is
misspecified.

%% Example of how to call a simulation in R:
<<sim1>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1090)

if(file.exists("sim1.rda")) {
  load("sim1.rda")
} else {
  ## Arguments lead to small-scale simulation:
  sim1 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
  save(sim1, file="sim1.rda")
}

## Add to column names so that plots look nicer
sim1$nobs <- paste("n=", sim1$nobs, sep="")

#Not sure how to express all loadings and all error terms in greek letters.
greekExprList<-expression(lambda[11],epsilon[11],sigma[12],lambda[11]--lambda[62],
                          epsilon[11]--epsilon[66])
greekpar<-as.expression(greekExprList)

@

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 1.Violation Parameter is loading}
\label{fig:sim11res}
\setkeys{Gin}{width=\textwidth}
<<sim11res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey1 <- simpleKey(unique(sim1$nobs), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
             subset = diff <= 4 & parms == "loading" & test=="CvM" &  pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
             as.table=TRUE))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 1.Violation Parameter is covariance}
\label{fig:sim12res}
\setkeys{Gin}{width=\textwidth}
<<sim12res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
             subset = diff <= 4 & parms == "var" & test == "CvM" &  pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
             as.table=TRUE))
@
\end{figure}

\begin{figure}
\caption{Simulated power curves for Simulation 1.Violation Parameter is error terms}
\label{fig:sim13res}
\setkeys{Gin}{width=\textwidth}
<<sim13res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
             subset = diff <= 4 & parms == "error" & test == "CvM" &  pars %in%
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
             as.table=TRUE))
@
\end{figure}

%###########################################################################################
\section{Simulation 2}
In Simulation 2, we examine the extent to which the results of
Simulation 1 are robust to model misspecification. Specifically, we
generate data from the factor analysis model used in the previous
section, but with adding an extra loading from the second factor to the first
indicator (of the first factor).  The estimated model lacks this extra
loading, however. The goal of this simulation is to examine the proposed statitics'
power and spicificity under model misspecification.

\subsection{Method}
Measurement invariance violations could occur in
each of the three parameters from Simulation 1 (factor loading,factor covariance,unique
variance), and they could also occur in the extra,unmodeled loading. 
Sample size and magnitude of measurementinvariance violation were manipulated 
in the same way as they were in Simulation 1.The tested parameters are the same as Simulation
1,too.

\subsection{Results}
Results are presented by the paramter that violated measurement invariance.
Figure~\ref{fig:sim21res} displays results when the unmodeled loading violates 
measurement invariance. One can generally observe that 
the first loading and error variance parameter showed high measurement invariance 
violation simultaneously, while factor covariance did not demonstrate violation.

Figure~\ref{fig:sim22res} displays results when the loading $\lambda_{11}$ violates
measurement invariance. Even though lacking one loading associated with the first indicator, 
CvM statistic showed enough power to detect the first loading violation specifically. 
Similar pattern of specificity and power were observed when the violation parameter 
was $\sigma_{12}$ (Figure~\ref{fig:sim23res}) and $\epsilon_{11}$ (Figure~\ref{fig:sim24res}).

In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
as the umodeled parameter does not violate measurment invariance. If the unmodeled parameter 
dose violate measurement invariance, however, then the tests assign this violation to other 
parameters that do not violate measurement invariance. The impacted parameters include the error
variance and other loadings associated with the manifest variables that had the unmodeled 
loading. Thus, prior to testing measurement invariance, it is important to study the extent to
which the hypothesized model includes all parameters of importance.

%% Example of how to call a simulation in R:
<<sim2>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1090)

if(file.exists("sim2.rda")) {
  load("sim2.rda")
} else {
  ## Arguments lead to small-scale simulation:
  sim2 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("extra",
                    "extra+loading", "extra+var","extra+error"))
  save(sim2, file="sim2.rda")
}

## Add to column names so that plots look nicer
sim2$nobs <- paste("n=", sim2$nobs, sep="")

@

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 2.Violation Parameter is extra loading}
\label{fig:sim21res}
\setkeys{Gin}{width=\textwidth}
<<sim21res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey2 <- simpleKey(unique(sim2$nobs), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff |pars , group = ~ nobs, data = sim2,
             subset = diff <= 4 & parms == "extra" & test == "CvM" & pars %in%
              c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 2.Violation Parameter is loading, 
         but model misspecified}
\label{fig:sim22res}
\setkeys{Gin}{width=\textwidth}
<<sim22res, fig=TRUE, height=7, width=7>>=
print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
             subset = diff <= 4 & parms == "extra+loading" & test == "CvM" & pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 2.Violation Parameter is var, 
         but model misspecified}
\label{fig:sim23res}
\setkeys{Gin}{width=\textwidth}
<<sim23res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
             subset = diff <=4 & parms == "extra+var" & test == "CvM" & pars %in%
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 2.Violation Parameter is var, 
         but model misspecified}
\label{fig:sim24res}
\setkeys{Gin}{width=\textwidth}
<<sim24res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
## 
print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
             subset = diff <=4 & parms == "extra+error" & test == "CvM" & pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
@
\end{figure}

%###########################################################################################

\section{Simulation 3}
In this section, we are aiming to test the extent to which the results of Simulation 1 
are robust to percentage of violation group impact. Specifically, we generate data from 
the factor analysis model used in Simulation 1 but manipulating the percentage of violating 
individuals. The goal of this simulation is to examine the proposed three statistics'
power and spicificity in different percentage violating individuals.

\subsection{Method}
The percentage of violation group is manipulated from 10\% to 90\% by 10\% in 
each simulation. Violation magnitude is fixed to 3. Sample size is fixed to 500. 
The violation parameters are the same as Simulation 1, including factor loading, 
factor correlation and error term. 5000 datasets were generated and tested.

\subsection{Results}
Simulation results were shown in Figure~\ref{fig:sim31res} 
to Figure~\ref{fig:sim33res}. The highest power exhibited 
when the percentage is 0.5 for all three violation paramter, as shown in Simualition 1 when 
group divided by half. As the percentage decrease or increase from 0.5, the power decreases.
However, the decreasing pattern is not symmetric. Specifically, when the violation percentage
is approaching to 0.1, the power decreases more rappidly and down to lower than 0.1 when the 
violation percentage is 10\%. Meanwhile, the power decrease less dramatically as the violation
percentage approaching to 0.9. As for the difference among these three statistics, we noticed 
that supLM which takes the maximum over individuals is the most robust to violation percentage.
The power remains to 0.2 as violation percentage reaches 0.9, while dmax and CvM drop to near 
0.1. This observation is consistent with the statistics' property. dmax or CvM takes the 
maximum or sums over the parameters and individuals, respectively. While supLM sums over 
parameters then take the maximum over individuals. Therefore, when the violation percentage is
small, this statistics is more resistent to the contamination produced by aggregation across 
individuals. 
However, when there are more than one parameter violates the measurement invariance, the 
difference pattern among these three statistics might change.We will explore this problem in 
simulation 4.

\seeme{The asymptotic pattern should be interpreted with the parameter value. I write some 
  in the discussion part. Basically, if most individuals' parameter value is larger, 
  the power will be larger.}
\seeme{I think the natural question would be when more than one paramter violate in different
  individual percentage, how the three statistics would perform. So these two simulations would
  be looked as exploration of the difference among these three statisitcs.}

%% Example of how to call a simulation in R:
<<sim3>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1090)

if(file.exists("sim3.rda")) {
  load("sim3.rda")
} else {
  ## Arguments lead to small-scale simulation:
  sim3 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
  save(sim3, file="sim3.rda")
}

## Add to column names so that plots look nicer
sim3$test <- paste("", sim3$test, sep="")
levels(sim3$perc)<-c(".05",".1",".15",".2",".25",".3",".35",".4",".45",".5",".55",
                     ".6",".65",".7",".75",".8",".85",".9",".95")

@

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 3.Violation Parameter is loading}
\label{fig:sim31res}
\setkeys{Gin}{width=\textwidth}
<<sim31res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey3 <- simpleKey(unique(sim3$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ perc | pars, group = ~ test, data = sim3,
             subset = diff ==3 & parms == "loading"  & nobs == 500 & pars %in%
             c("1","7","13","1:6","7:12") & perc %in% c(".1",".2",
               ".3",".4",".5",".6",".7",".8",".9"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Percentage",ylab="Power", key=mykey3))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 3.Violation Parameter is var}
\label{fig:sim32res}
\setkeys{Gin}{width=\textwidth}
<<sim32res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey3 <- simpleKey(unique(sim3$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ perc | pars, group = ~ test, data = sim3,
             subset = diff ==3 & parms == "var" & nobs == 500 & pars %in% 
             c("1","7","13","1:6","7:12") & perc %in% c(".1",".2",
               ".3",".4",".5",".6",".7",".8",".9"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Percentage",ylab="Power", key=mykey3))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 3.Violation Parameter is loading}
\label{fig:sim33res}
\setkeys{Gin}{width=\textwidth}
<<sim33res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey3 <- simpleKey(unique(sim3$test), points = TRUE, lines = TRUE)

print(xyplot(power ~ perc | pars , group = ~ test, data = sim3,
             subset = diff ==3 & parms == "error" & nobs == 500 & pars %in% 
             c("1","7","13","1:6","7:12") & perc %in% c(".1",".2",
               ".3",".4",".5",".6",".7",".8",".9"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Percentage", ylab="Power", key=mykey3))
@
\end{figure}

%###########################################################################################

\section{Simulation 4}
In Simulation 4, we are aiming to test the extent to which the results of Simulation 1 
are robust to multiple group violation.Specifically, we generate data from the factor 
analysis model used in Simulation 1 but adding one more group which violates the measurement 
invariance by a different degree. The goal of this simulation is to examine the proposed 
statistics'power and spicificity in multiple group violation condition.

\subsection{Method}
The violations were still based on the auxiliary variable $V$, but there are two cutoff 
points for $V$. Therefore,individuals is divided into three groups.For simplicity reason,
individuals were assigned to these three groups equally. Group difference were $d/2$ times 
the parameters' asymptotic standard errors(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation magnitude ($d$) $\times$ violating 
parameter, 5000 datasets were generated and tested.

\subsection{Results}
Simulation results were displayed in Figure~\ref{fig:sim41res} to Figure~\ref{fig:sim43res}. 
The test specificity were retained as in two group setting.However, the powers are generally 
lower than the corresponding two group setting as shown in Simulation 1.

In summary, when there is more than one group violating measurement invariance, the proposed 
test statistic could remain specificity and enough power to test the particular violating 
parameter.

%% Example of how to call a simulation in R:
<<sim4>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1090)

if(file.exists("sim4.rda")) {
  load("sim4.rda")
} else {
  ## Arguments lead to small-scale simulation:
  sim4 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
  save(sim4, file="sim4.rda")
}

## Add to column names so that plots look nicer
sim4$nobs <- paste("n=", sim4$nobs, sep="")
## Not fully happy with this, but haven't figured out how
## to get greek letters in xyplot panel labels.

@

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 4.Violation Parameter is loading}
\label{fig:sim41res}
\setkeys{Gin}{width=\textwidth}
<<sim41res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey4 <- simpleKey(unique(sim4$nobs), points = TRUE, lines = TRUE)

print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim4,
             subset = diff <= 4 & parms == "loading" & test=="CvM" & pars %in%
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey4))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 4. Violation Parameter is covariance}
\label{fig:sim42res}
\setkeys{Gin}{width=\textwidth}
<<sim42res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim4,
             subset = diff <= 4 & parms == "var" & test == "CvM" & pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey4))
@
\end{figure}

\begin{figure}
\caption{Simulated power curves for Simulation 4. Violation Parameter is error terms}
\label{fig:sim43res}
\setkeys{Gin}{width=\textwidth}
<<sim43res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))

print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim4,
             subset = diff <= 4 & parms == "error" & test == "CvM" & pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey4))
@
\end{figure}

%###########################################################################################

\section{Simulation 6}
Follow sim3. The problem need concern is that when fix diff=3, the plot is similar to sim3. However,
if we look at with diffrent diff in perc=0.5, the plot would be similar to psychometric paper. 
CvM's advantageous is depending on percentage.

%% Example of how to call a simulation in R:
<<sim6>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1090)

if(file.exists("sim6.rda")) {
  load("sim6.rda")
} else {
  ## Arguments lead to small-scale simulation:
  sim6 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
  save(sim6, file="sim6.rda")
}

## Add to column names so that plots look nicer
sim6$test <- paste("", sim6$test, sep="")
levels(sim6$perc)<-c(".05",".1",".15",".2",".25",".3",".35",".4",".45",".5",".55",
                     ".6",".65",".7",".75",".8",".85",".9",".95")

@

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 3.Violation Parameter is loading}
\label{fig:sim61res}
\setkeys{Gin}{width=\textwidth}
<<sim61res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey6 <- simpleKey(unique(sim6$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ perc | pars, group = ~ test, data = sim6,
             subset = diff ==3 & parms == "loading"  & nobs == 500 & pars %in%
             c("1","7","13","1:6","7:12","1:3","7:9") & perc %in% c(".1",".2",
               ".3",".4",".5",".6",".7",".8",".9"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,3),
             type = "b", xlab="Violation Percentage",ylab="Power", key=mykey6))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 3.Violation Parameter is loading}
\label{fig:sim62res}
\setkeys{Gin}{width=\textwidth}
<<sim62res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey6 <- simpleKey(unique(sim6$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ perc | pars, group = ~ test, data = sim6,
             subset = diff ==3 & parms == "error"  & nobs == 500 & pars %in%
             c("1","7","13","1:6","7:12","1:3","7:9") & perc %in% c(".1",".2",
               ".3",".4",".5",".6",".7",".8",".9"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,3),
             type = "b", xlab="Violation Percentage",ylab="Power", key=mykey6))
@
\end{figure}

\section{General Discussion}

What do the results mean, future studies, etc.

In Simulation 3, the asymmetric pattern is asscociated with parameter value.
Specifically, given unequal groups, when most(percentage is 0.6,0.7,0.8,0.9) of the 
individuals' parameter value is larger, the measurement variance is easier to be detected.
In the situation that violation group parameter value is smaller, as the percentage increases,
individuals' parameter value is smaller, too, the measurement variance is more difficult to be 
detected. Thus the asymmptotic pattern would look reversed (percentage 0.1,0.2,0.3,0.4 would be
associated with larger power compared with the 0.9,0.8,0.7,0.6).

\subsection{Summary}

Concluding paragraph.

\section*{Computational Details}

All results were obtained using the {R}~system for statistical computing \cite{R11},
version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on package
{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} for fitting 
of the factor analysis models and
{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
for evaluating the parameter instability tests. {R}~and the packages {lavaan} and 
{strucchange} are
freely available under the General Public License~2 from the
Comprehensive {R} Archive Network at \url{http://CRAN.R-project.org/}.
{R}~code for replication of our results is
available at \url{http://semtools.R-Forge.R-project.org/}.

\bibliography{refs}

\end{document}

### 2013-07-08 frontiers/simall.R
add simall file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@213 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,763 @@
+## Data-generating process, different dgp function represent different simulation study.
+
+
+
+#######################################################################################
+#Simualtion1, different parameter voilation############################################
+dgp1 <- function(nobs = 200, diff = 3, parms="loading", perc = 0.5)
+{
+  # Generates data from a factor analysis model that violates
+  # measurement invariance.  Also generates a continuous auxiliary
+  # variable related to the invariance (loosely called "age").
+  
+  # ses is number of SEs between group 1's parameters and group 2's
+  # parameters.
+  # theta[1-6]:loadings, theta[7-12]:residual,theta[13]:covariance, theta[14-19]:mean
+  sampsize <- nobs
+  ses <- diff
+  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5,-0.48,
+           29.32,24.70,14.84,10.59,19.30,18.01)
+  asym<-c( 7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
+            56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
+           1.024326,7.139776,4.666005,8.151785,3.696972,5.242366,
+            8.630417)
+  # Could eventually manipulate this so break point is not directly
+  # in the middle
+  num.y <- round(sampsize*perc, 0)
+  num.o <- sampsize - num.y
+  age <- c(runif(num.y,13,16),runif(num.o,16,18))
+  # Define parameter vectors/matrices:
+  
+  # thetas for "young" individuals:
+  theta.y <- theta
+  
+  # thetas for "old" individuals:
+  #this is the block need to change by naming different parms and assigning different vectors.
+  ###################################################
+  if(parms=="loading"){parmsvector=c(1,rep(0,18))}
+  if(parms=="var"){parmsvector=c(rep(0,12),1,rep(0,6))}
+  if(parms=="error"){parmsvector=c(rep(0,6),1,rep(0,5),rep(0,7))}
+  ####################################################
[…]

### 2013-07-08 frontiers/estfun.working.R
adding supporting files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@214 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,143 @@
+estfun.working <- function(fit)
+{
+  ## Code to extract scores from lavaan
+  ## number variables/sample size
+  samplestats <- fit@SampleStats
+  ntab <- unlist(samplestats@nobs)
+  nobs <- samplestats@ntotal
+
+  Score.mat <- matrix(NA, nobs, length(coef(fit)))
+  
+  for(g in 1:samplestats@ngroups) {
+    if (samplestats@ngroups > 1){
+      moments <- fitted(fit)[[g]]
+    } else {
+      moments <- fitted(fit)
+    }
+    Sigma.hat <- moments$cov
+  
+    if(!samplestats@missing.flag) { # complete data
+      if(fit@Model@meanstructure) { # mean structure
+        nvar <- ncol(samplestats@cov[[g]])
+        Mu.hat <- moments$mean
+        X <- fit@Data@X[[g]]
+        Sigma.inv <- lavaan:::inv.chol(Sigma.hat, logdet=FALSE)
+
+        J <- matrix(1, 1L, ntab[g]) ## FIXME: needed? better maybe rowSums/colSums?
+        J2 <- matrix(1, nvar[g], nvar[g])
+        diag(J2) <- 0.5
+
+        ## scores.H1 (H1 = saturated model)
+        mean.diff <- t(t(X) - Mu.hat %*% J)
+
+        dx.Mu <- -1 * mean.diff %*% Sigma.inv
+
+        dx.Sigma <- t(apply(mean.diff, 1L,
+           function(x) lavaan:::vech(- J2 * (Sigma.inv %*% (tcrossprod(x) - Sigma.hat) %*% Sigma.inv))))
+
+        scores.H1 <- cbind(dx.Mu, dx.Sigma)
+    
+      } else {
+        ## no mean structure
+        stop("Score calculation with no mean structure is not implemented.")
+      }
+    } else { # incomplete data
+      nsub <- ntab[g]
+      M <- samplestats@missing[[g]]
+      MP1 <- fit@Data@Mp[[g]]
+      pat.idx <- match(MP1$id, MP1$order)
+
+      Mu.hat <- moments$mean
+      nvar <- ncol(samplestats@cov[[g]])
+      score.sigma   <- matrix(0, nsub, nvar*(nvar+1)/2)
+      score.mu <- matrix(0, nsub, nvar)
+    
+      for(p in 1:length(M)) {
+        ## Data
+        X <- M[[p]][["X"]]
[…]

### 2013-07-08 frontiers/mz.R

@@ -1,0 +1,54 @@
+mzfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
+{
+  ## data checks
+  data <- as.data.frame(data)
+  stopifnot(c("x1", "x2", "x3", "y1", "y2", "y3") %in% names(data))
+  data <- data[, c("x1", "x2", "x3", "y1", "y2", "y3")]
+
+  ## require package
+  stopifnot(require("lavaan"))
+
+  ## model from Merkle & Zeileis
+  rval <- 'verbal =~ x1 + x2 + x3; math =~ y1 + y2 + y3'
+
+  ## fit model (set likelihood = "wishart" for N-1 correction as in OpenMx)
+  rval <- sem(rval, data = data, meanstructure = TRUE, std.lv = TRUE)
+
+  ## store (1) fitted model object, (2) engine used,
+  ## (3) match between OpenMx and lavaan coefficient labeling and ordering
+  rval <- list(
+    data = data,
+    model = rval,
+    names = c("verbal=~x1", "verbal=~x2", "verbal=~x3", "math=~y1", "math=~y2", "math=~y3", "x1~~x1", "x2~~x2", "x3~~x3",
+        "y1~~y1", "y2~~y2", "y3~~y3", "verbal~~math", "x1~1", "x2~1", "x3~1", "y1~1", "y2~1", "y3~1"))
+
+  class(rval) <- "mzfit"
+
+  return(rval)
+}
+
+coef.mzfit <- function(object, ...) {
+    unclass(getMethod("coef", "lavaan")(object$model, ...))
+}
+
+logLik.mzfit <- function(object, ...) {
+  rval <- as.vector(fitMeasures(object$model, "logl"))
+
+  structure(rval, df = length(coef(object)),
+    nobs = nrow(object$data), class = "logLik")
+}
+
+vcov.mzfit <- function(object, ...) {
+    vc <- getMethod("vcov", "lavaan")(object$model, ...)
+    class(vc) <- "matrix"
+    vc
+}
+
+bread.mzfit <- function(x, ...) vcov(x) * nrow(x$data)
+
+info.mzfit <- function(x, ...) solve(vcov(x) * nrow(x$data))
+
+estfun.mzfit <- function(x, ...)
+{
+  estfun(x$model)
+}

### 2013-07-15 frontiers/frontiers.Rnw
add frontiers.Rnw and delete other unrelated files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@215 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2013-07-08 frontiers/testprops1.Rnw
@@
 
+\section{Theoretical Detail}
+This section contains background on the theory underlying the proposed statistics; for a more detailed account, see Merkle and Zeileis\citeyearNP{MerZei13}. 
+We consider situations in which a $p$-dimensional variable $X$ with observations $x_{i},i=1,...n$ is described by a model with density $f(x_{i};\theta)$ and associated joint log-likelihood
+
+\begin{equation}
+\label{equ:likelihood}
+\ell(\theta;x_{1},...,x_{n})=\overset{n}{\underset{i=1}{\Sigma}}\ell(\theta;x_{i})=\overset{n}{\underset{i=1}{\Sigma}}logf(x_{i};\theta)
+\end{equation}
+
+where $\theta$ is some $k-$dimensional parameter vector that characterizes the distribution. 
+We focus on applications where the density $f(x_{i};\theta)$ arises from a structural equation model with assumed multivariate normality, though the proposed tests extend beyond this family of models.
+Under the usual regularity condition\cite{}, the model parameters $\theta$ can be estimated by maximum likelihood(ML),i.e.,
+
+\begin{equation}
+\label{equ:ml}
+\hat{\theta}=\underset{\theta}{argmax}\ell(\theta;x_{1},...,x_{n})
+\end{equation}
+
+or equivalently by solving the first order conditions
+
+\begin{equation}
+\label{equ:max}
+\overset{n}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i})=0,
+\end{equation}
+
+where
+
+\begin{equation}
+\label{equ:score}
+s(\theta;x_{i})=(\frac{\partial l(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial l(\theta;x_{i})}{\partial(\theta_{k})})^{T}
+\end{equation}
+
+is the score function of the model(the partial derivative of the casewise likelihood contributions w.r.t.
+the parameters $\theta$).
+Evaluation of the score function at $\hat{\theta}$ for $i=1,...n$ essentially measures the extent to which the model maximizes each individual's likelihood: as an individual's score stray further from zero,the model provides a poorer description of that individual.
+
+\section{Tutorial}
 \section{Simulation 1}
@@
 From these figures, one generally observe that the tests isolate 
-the parameter violating measurementinvariance.  Additionally, the tests 
+the parameter violating measurement invariance.  Additionally, the tests 
 have somewhat-higher power to detect measurement invariance violations 
@@
 
-%###########################################################################################
-
-\section{Simulation 3}
-In this section, we are aiming to test the extent to which the results of Simulation 1 
-are robust to percentage of violation group impact. Specifically, we generate data from 
-the factor analysis model used in Simulation 1 but manipulating the percentage of violating 
-individuals. The goal of this simulation is to examine the proposed three statistics'
-power and spicificity in different percentage violating individuals.
-
-\subsection{Method}
-The percentage of violation group is manipulated from 10\% to 90\% by 10\% in 
-each simulation. Violation magnitude is fixed to 3. Sample size is fixed to 500. 
-The violation parameters are the same as Simulation 1, including factor loading, 
-factor correlation and error term. 5000 datasets were generated and tested.
-
-\subsection{Results}
-Simulation results were shown in Figure~\ref{fig:sim31res} 
-to Figure~\ref{fig:sim33res}. The highest power exhibited 
-when the percentage is 0.5 for all three violation paramter, as shown in Simualition 1 when 
-group divided by half. As the percentage decrease or increase from 0.5, the power decreases.
-However, the decreasing pattern is not symmetric. Specifically, when the violation percentage
-is approaching to 0.1, the power decreases more rappidly and down to lower than 0.1 when the 
-violation percentage is 10\%. Meanwhile, the power decrease less dramatically as the violation
-percentage approaching to 0.9. As for the difference among these three statistics, we noticed 
-that supLM which takes the maximum over individuals is the most robust to violation percentage.
-The power remains to 0.2 as violation percentage reaches 0.9, while dmax and CvM drop to near 
-0.1. This observation is consistent with the statistics' property. dmax or CvM takes the 
-maximum or sums over the parameters and individuals, respectively. While supLM sums over 
-parameters then take the maximum over individuals. Therefore, when the violation percentage is
-small, this statistics is more resistent to the contamination produced by aggregation across 
-individuals. 
-However, when there are more than one parameter violates the measurement invariance, the 
-difference pattern among these three statistics might change.We will explore this problem in 
-simulation 4.
-
-\seeme{The asymptotic pattern should be interpreted with the parameter value. I write some 
-  in the discussion part. Basically, if most individuals' parameter value is larger, 
-  the power will be larger.}
-\seeme{I think the natural question would be when more than one paramter violate in different
-  individual percentage, how the three statistics would perform. So these two simulations would
-  be looked as exploration of the difference among these three statisitcs.}
-
-%% Example of how to call a simulation in R:
-<<sim3>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1090)
-
-if(file.exists("sim3.rda")) {
-  load("sim3.rda")
-} else {
-  ## Arguments lead to small-scale simulation:
-  sim3 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
-  save(sim3, file="sim3.rda")
-}
-
-## Add to column names so that plots look nicer
-sim3$test <- paste("", sim3$test, sep="")
-levels(sim3$perc)<-c(".05",".1",".15",".2",".25",".3",".35",".4",".45",".5",".55",
-                     ".6",".65",".7",".75",".8",".85",".9",".95")
-
-@
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 3.Violation Parameter is loading}
-\label{fig:sim31res}
-\setkeys{Gin}{width=\textwidth}
-<<sim31res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey3 <- simpleKey(unique(sim3$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ perc | pars, group = ~ test, data = sim3,
-             subset = diff ==3 & parms == "loading"  & nobs == 500 & pars %in%
-             c("1","7","13","1:6","7:12") & perc %in% c(".1",".2",
-               ".3",".4",".5",".6",".7",".8",".9"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Percentage",ylab="Power", key=mykey3))
-@
-\end{figure}
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 3.Violation Parameter is var}
-\label{fig:sim32res}
-\setkeys{Gin}{width=\textwidth}
-<<sim32res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey3 <- simpleKey(unique(sim3$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ perc | pars, group = ~ test, data = sim3,
-             subset = diff ==3 & parms == "var" & nobs == 500 & pars %in% 
-             c("1","7","13","1:6","7:12") & perc %in% c(".1",".2",
-               ".3",".4",".5",".6",".7",".8",".9"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Percentage",ylab="Power", key=mykey3))
-@
-\end{figure}
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 3.Violation Parameter is loading}
-\label{fig:sim33res}
-\setkeys{Gin}{width=\textwidth}
-<<sim33res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey3 <- simpleKey(unique(sim3$test), points = TRUE, lines = TRUE)
-
-print(xyplot(power ~ perc | pars , group = ~ test, data = sim3,
-             subset = diff ==3 & parms == "error" & nobs == 500 & pars %in% 
-             c("1","7","13","1:6","7:12") & perc %in% c(".1",".2",
-               ".3",".4",".5",".6",".7",".8",".9"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Percentage", ylab="Power", key=mykey3))
-@
-\end{figure}
-
-%###########################################################################################
-
-\section{Simulation 4}
-In Simulation 4, we are aiming to test the extent to which the results of Simulation 1 
-are robust to multiple group violation.Specifically, we generate data from the factor 
-analysis model used in Simulation 1 but adding one more group which violates the measurement 
-invariance by a different degree. The goal of this simulation is to examine the proposed 
-statistics'power and spicificity in multiple group violation condition.
-
-\subsection{Method}
-The violations were still based on the auxiliary variable $V$, but there are two cutoff 
-points for $V$. Therefore,individuals is divided into three groups.For simplicity reason,
-individuals were assigned to these three groups equally. Group difference were $d/2$ times 
-the parameters' asymptotic standard errors(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
-For each combination of sample size ($n$) $\times$ violation magnitude ($d$) $\times$ violating 
-parameter, 5000 datasets were generated and tested.
-
-\subsection{Results}
-Simulation results were displayed in Figure~\ref{fig:sim41res} to Figure~\ref{fig:sim43res}. 
-The test specificity were retained as in two group setting.However, the powers are generally 
-lower than the corresponding two group setting as shown in Simulation 1.
-
-In summary, when there is more than one group violating measurement invariance, the proposed 
-test statistic could remain specificity and enough power to test the particular violating 
-parameter.
-
-%% Example of how to call a simulation in R:
-<<sim4>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1090)
-
-if(file.exists("sim4.rda")) {
-  load("sim4.rda")
-} else {
-  ## Arguments lead to small-scale simulation:
-  sim4 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
-  save(sim4, file="sim4.rda")
-}
-
-## Add to column names so that plots look nicer
-sim4$nobs <- paste("n=", sim4$nobs, sep="")
-## Not fully happy with this, but haven't figured out how
-## to get greek letters in xyplot panel labels.
-
-@
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 4.Violation Parameter is loading}
-\label{fig:sim41res}
-\setkeys{Gin}{width=\textwidth}
-<<sim41res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey4 <- simpleKey(unique(sim4$nobs), points = TRUE, lines = TRUE)
-
-print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim4,
-             subset = diff <= 4 & parms == "loading" & test=="CvM" & pars %in%
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey4))
-@
-\end{figure}
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 4. Violation Parameter is covariance}
-\label{fig:sim42res}
-\setkeys{Gin}{width=\textwidth}
-<<sim42res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim4,
-             subset = diff <= 4 & parms == "var" & test == "CvM" & pars %in% 
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey4))
-@
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for Simulation 4. Violation Parameter is error terms}
-\label{fig:sim43res}
-\setkeys{Gin}{width=\textwidth}
-<<sim43res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-
-print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim4,
-             subset = diff <= 4 & parms == "error" & test == "CvM" & pars %in% 
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey4))
-@
-\end{figure}
-
-%###########################################################################################
-
-\section{Simulation 6}
-Follow sim3. The problem need concern is that when fix diff=3, the plot is similar to sim3. However,
-if we look at with diffrent diff in perc=0.5, the plot would be similar to psychometric paper. 
-CvM's advantageous is depending on percentage.
-
-%% Example of how to call a simulation in R:
-<<sim6>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1090)
-
-if(file.exists("sim6.rda")) {
-  load("sim6.rda")
-} else {
-  ## Arguments lead to small-scale simulation:
-  sim6 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
-  save(sim6, file="sim6.rda")
-}
-
-## Add to column names so that plots look nicer
-sim6$test <- paste("", sim6$test, sep="")
-levels(sim6$perc)<-c(".05",".1",".15",".2",".25",".3",".35",".4",".45",".5",".55",
-                     ".6",".65",".7",".75",".8",".85",".9",".95")
-
-@
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 3.Violation Parameter is loading}
-\label{fig:sim61res}
-\setkeys{Gin}{width=\textwidth}
-<<sim61res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey6 <- simpleKey(unique(sim6$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ perc | pars, group = ~ test, data = sim6,
-             subset = diff ==3 & parms == "loading"  & nobs == 500 & pars %in%
-             c("1","7","13","1:6","7:12","1:3","7:9") & perc %in% c(".1",".2",
-               ".3",".4",".5",".6",".7",".8",".9"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,3),
-             type = "b", xlab="Violation Percentage",ylab="Power", key=mykey6))
-@
-\end{figure}
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 3.Violation Parameter is loading}
-\label{fig:sim62res}
-\setkeys{Gin}{width=\textwidth}
-<<sim62res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey6 <- simpleKey(unique(sim6$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ perc | pars, group = ~ test, data = sim6,
-             subset = diff ==3 & parms == "error"  & nobs == 500 & pars %in%
-             c("1","7","13","1:6","7:12","1:3","7:9") & perc %in% c(".1",".2",
-               ".3",".4",".5",".6",".7",".8",".9"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,3),
-             type = "b", xlab="Violation Percentage",ylab="Power", key=mykey6))
-@
-\end{figure}
-
 \section{General Discussion}
@@
 What do the results mean, future studies, etc.
-
-In Simulation 3, the asymmetric pattern is asscociated with parameter value.
-Specifically, given unequal groups, when most(percentage is 0.6,0.7,0.8,0.9) of the 
-individuals' parameter value is larger, the measurement variance is easier to be detected.
-In the situation that violation group parameter value is smaller, as the percentage increases,
-individuals' parameter value is smaller, too, the measurement variance is more difficult to be 
-detected. Thus the asymmptotic pattern would look reversed (percentage 0.1,0.2,0.3,0.4 would be
-associated with larger power compared with the 0.9,0.8,0.7,0.6).
-
-\subsection{Summary}
-
-Concluding paragraph.
 

### 2013-07-17 frontiers/frontiers.Rnw
update frontiers to show code in Rnw file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@216 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor revisions to frontiers paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@217 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
update frontier with lavaan example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@218 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}
+\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
 
@@
 
+\SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
 \section{Tutorial}
+In general, a set of scales is defined to be measurement invariant with respect to an auxiliary variable $V$ if:
+
+\begin{equation}
+\label{equ:invariance}
+f(x_{i}|t_{i},v_{i},...)=f(x_{i}|t_{i},...),
+\end{equation}
+
+where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable for variable $i$ that the scales purport to measure, and $f$ is the model's distributional form \citep{}.
+If the cut point of $v$ is known in advance, i.e.
+gender, nested multiple group models\citep{} coupled with likelihood ratio test are most commonly applied.
+Suppose there are two groups A and B. The null hypothesis of measurement invariance is
+
+\begin{equation}
+\label{equ:null}
+H_{0}: \theta_{A}=\theta_{B},
+\end{equation}
+
+The alternative hypothesis is
+
+\begin{equation}
+\label{equ:alternative}
+H_{1}:\theta_{i}=
+\begin{cases}
+\theta^{(A)} & v_{i}\leq v\\
+\theta^{(B)} & v_{i}\geq v
+\end{cases}
+\end{equation}
+
+The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ (group A and group B) can be obtained by Maximum Likelihood(ML) from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,respectively.
+The LR test statistic for the given threshold $v$ is then
+\begin{equation}
+\label{equ:LRT}
+LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
+{(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
+}],
+\end{equation}
+
+has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal to the number difference of parameters in $\theta$.
+
+To be more specific, in the following section, we will introduce how to implement multi group analysis in lavaan package. For more details, see\citep{}
+\subsection{Multiple Group Analysis}
+First of all, we need to read in data to R.  
+<<read data>>=
+# Reading in the data
+## (need data here)Data <- read.table("example.dat")
+## names(Data) <- c("y1","y2","y3","y4","y5","y6","g")
+@
+Second, the model need to be specified using lavaan package. 
+<<specify model>>=
+library(lavaan)
+model <- 'f1=~NA*y1+y2+y3
+          f2=~NA*y4+y5+y6
+          f1 ~~1*f1
+          f2 ~~1*f2'
+@
+Third, we want to test which parameter(s) violate measurement invariance.
+In lavaan package, loading, manifest variables' intercept and factor means could be tested together or specifically.
+For more details, please see \citep{}.
+Since in most cases researchers are interested in the factor loading invariance, we will use that as the example in this tutorial.
+To implement LRT as ~\ref{equ:likelihood}, we need to get estimation for \hat{\theta}
+<<theta estimate>>=
+# estimate theta 
+## theta<-cfa(model,data=Data,group="g",group.equal="loadings")
+@
+Then we can get estimate for group A and group B. 
+<<theta A and theta B estimate>>=
+# estimate theta A,B 
+## thetaAB<-cfa(model,data=Data,group="g")
+@
+In the end, we can get the LRT results by calling the measurement invariance function in lavaan. 
+<<LRT>>=
+# LRT
+## measurementInvariance(model,data=Data,group="g",strict=FALSE)
+@
+\subsection{Extensions for Unknown Group}
+
+\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
 \section{Simulation 1}

### 2013-07-17 frontiers/simall.R
update simall with readme
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@220 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,7 +1,20 @@
+
+######################################################################################
+#READ ME##############################################################################
+#                                                                                    #
+#This study aims to initiate different dpg functions to explore measurment invariance#
+#test properties in different scenarios.Specifically,                                # 
+#dgp1:different parameter, var, loading, error violate measurement invariance(MI).   #
+#dgp2:model misspecified by lacking one loading, the other are the same as dgp1.     #
+#dgp3:group percentage change while different parameter violate MI.                  #
+#dgp4:Three groups generated by two groups violating MI with different amptitude.    #
+#dgp5:Observations violate MI according to age continuously.                         # 
+#dgp6 Group percentage change while 3 loading or 3 error term violate MI.            # 
+######################################################################################
+
+
+
 ## Data-generating process, different dgp function represent different simulation study.
-
-
-
 #######################################################################################
 #Simualtion1, different parameter voilation############################################
 dgp1 <- function(nobs = 200, diff = 3, parms="loading", perc = 0.5)
@@ -638,13 +651,13 @@ if(FALSE) {  ## Ignore this line, it just tells R to ignore this code:
                      c("loading","var","error"))
   sim2 <- simulation(sim=c("sim2"), nobs=c(100,200,500), diff=seq(0,4,.25),
                      parms=c("extra","extra+loading","extra+var","extra+error"))
-  sim3 <- simulation(sim=c("sim3"), nobs=500, diff=3, parms=c("loading","var","error"),
+  sim3 <- simulation(sim=c("sim3"), nobs=500, diff=seq(0,4,.25), parms=c("loading","var","error"),
                      perc=seq(0.05,0.95,0.05))
   sim4 <- simulation(sim=c("sim4"), nobs=c(100,200,500), diff=seq(0,4,.25),
                      parms=c("loading","var","error"),perc=1/3)
   sim5 <- simulation(sim=c("sim5"), nobs=c(100,200,500),diff=seq(10,100,5),
                      parms=c("loading","error"))
-  sim6 <- simulation(sim=c("sim6"), nobs=500, diff=3, parms=c("loading","error"),
+  sim6 <- simulation(sim=c("sim6"), nobs=500, diff=seq(0,4,.25), parms=c("loading","error"),
                      perc=seq(0.05,0.95,0.05))
   
   save(sim1, file = "sim1.rda")

### 2013-07-18 frontiers/simall.R
update simall
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@221 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -368,7 +368,7 @@ dgp4 <- function(nobs = 200, diff = 3, parms="loading", perc = 1/3)
  
 
 ######################################################################################
-dgp5 <- function(nobs = 200, diff = 3, parms="loading", perc = 0.5)
+dgp5 <- function(nobs = 200, diff = 3, parms="loading",perc=1)
 {
   # Generates data from a factor analysis model that violates
   # measurement invariance.  Also generates a continuous auxiliary
@@ -381,7 +381,16 @@ dgp5 <- function(nobs = 200, diff = 3, parms="loading", perc = 0.5)
   ses <- diff
   theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5,-0.48,
            29.32,24.70,14.84,10.59,19.30,18.01)
+  asym<-c( 7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
+            56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
+           1.024326,7.139776,4.666005,8.151785,3.696972,5.242366,
+            8.630417)
   age <- c(runif(sampsize,13,18))
+  z.age<-scale(age)
+
+  # Initialize data matrix:
+  datmat <- matrix(0,sampsize,6)
+  colnames(datmat) <- c("x1", "x2", "x3", "y1", "y2", "y3")
  
   
   # Define parameter vectors/matrices:  
@@ -394,45 +403,34 @@ dgp5 <- function(nobs = 200, diff = 3, parms="loading", perc = 0.5)
   ####################################################
   theta.age <- matrix(0,sampsize,length(theta))
   
-  lambda <-list()
-  phi<-list()
-  psi<-list()
-  mu<-list()
-  z<-list()
-  u<-list()
   
   for (i in 1:sampsize){
-  theta.age[i,] <- theta + diff*age[i]*parmsvector/sqrt(nobs/2) 
+  theta.age[i,] <- theta + diff*asym*z.age[i]*parmsvector/sqrt(nobs/2)*perc 
 
-  lambda[[i]] <- matrix(0,6,2)
+  lambda <- matrix(0,6,2)
   
-  lambda[[i]][,1] <- c(theta.age[i,1:3],0,0,0)
-  lambda[[i]][,2] <- c(0,0,0,theta.age[i,4:6])
[…]

### 2013-07-24 pkg/sctest.lavaan.R
Added sctest.lavaan to simplify strucchange calls for lavaan users.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@222 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,34 @@
+sctest.lavaan <- function(x, order.by, parm, stat, fplot = FALSE, ...){
+  ## Carry out a score-based test of a fitted lavaan model x
+
+  require("strucchange")
+  
+  ## TODO Check that the fit uses MVN likelihood fn
+  
+  ## Couldn't gefp() have a "type" argument like efp()?
+  fp <- gefp(x, fit = NULL, order.by = order.by, vcov = info_full,
+             sandwich = FALSE, parm = parm)
+
+  ## switch between stat argument names and corresponding functionals
+  ## ordwmax(fp), ordL2BB(fp), maxBB, meanL2BB, supLM(0.1), rangeBB
+  ## WDMo         maxLMo       DM     CvM       maxLM,      range
+  ## TODO? Allow user to modify the 0.1 of supLM?
+  fnl <- switch(stat,
+                WDMo = ordwmax(fp),
+                maxLMo = ordL2BB(fp),
+                DM = maxBB,
+                CvM = meanL2BB,
+                maxLM = supLM(0.1),
+                range = rangeBB)
+
+  if(is.null(fnl)) stop(paste("Argument stat=",stat," is unknown",sep=""))
+  
+  testres <- sctest(fp, functional = fnl)
+
+  if(fplot) plot(fp, functional = fnl)
+
+  testres
+}
+  
+## convenience function for applying tests to real data
+info_full <- function(x, ...) solve(vcov(x) * nobs(x))

### 2013-07-25 frontiers/frontiers.Rnw
Rolled frontiers.Rnw back to r222
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@224 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable for variable $i$ that the scales purport to measure, and $f$ is the model's distributional form \citep{}.
+where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable for variable $i$ that the scales purport to measure, and $f$ is the model's distributional form {addref}.
 If the cut point of $v$ is known in advance, i.e.
-gender, nested multiple group models\citep{} coupled with likelihood ratio test are most commonly applied.
+gender, nested multiple group models {addref} coupled with likelihood ratio test are most commonly applied.
 Suppose there are two groups A and B. The null hypothesis of measurement invariance is
@@
 
-To be more specific, in the following section, we will introduce how to implement multi group analysis in lavaan package. For more details, see\citep{}
+To be more specific, in the following section, we will introduce how to implement multi group analysis in lavaan package. For more details, see {addref}
 \subsection{Multiple Group Analysis}
@@
 In lavaan package, loading, manifest variables' intercept and factor means could be tested together or specifically.
-For more details, please see \citep{}.
+For more details, please see {addref}.
 Since in most cases researchers are interested in the factor loading invariance, we will use that as the example in this tutorial.
-To implement LRT as ~\ref{equ:likelihood}, we need to get estimation for \hat{\theta}
+To implement LRT as ~\ref{equ:LRT}, we need to get estimation for $\hat{\theta}$, without differing groups.
+
 <<theta estimate>>=

### 2013-07-25 frontiers/frontiers.Rnw
fix the empty frontiers.Rnw.Update the unknown group section.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@225 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \subsection{Extensions for Unknown Group}
+The multiple group test performs typically well when the $V$ is a relatively-small number of categories and the sample size is not too large (LRT is sensitive to large sample and will return "false" significance results, details see next section).
+However, when $V$ (in this case, age) continuous variable, however, multiple-group analysis usually cannot be performed because there are no existing groups.
+Instead, we can follow the procedure proposed by Ed Merkle and Zeileis.
+Specifically, we can fit a model whose parameters are restricted to be equal across across all individuals and then examine how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$.
+To formalize the idea, we assume that the observations are ordered w.r.t.
+$V$ and define the $k$ dimensional cumulative score process as
+\begin{equation}
+\label{equ:B}
+B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
+\end{equation}
+where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ is some consistent estimate of the covariance matrix of the scores,e.g.
+information matrix.
+~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t.
+$V$ and decorrelates the scores between $k$ components of the parameters in the model.
+Further, there exists a functional central limit theorem that allows us to make formal inference with this cumulative score process.
+Assuming that individuals are independent and the usual ML regularity condition hold, it is possible to show that(addref)
+\begin{equation}
+B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
+\end{equation} 
+where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ is a $k$ dimensional Brownian bridge.
+Thus, we can construct tests of measurement invariance by comparing the behavior of cumulative score process to that of a Brownian bridge.
+In practice, there are three test statistics can be used: double maximum, CvM and maxLM.
+\begin{equation}
+DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
+\end{equation}
+\begin{equation}
+CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
+\end{equation}
+\begin{equation}
+maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
+\end{equation}
+
+Specifically, we will use the previous example, except changing categorical agegroup to continuous age.
+Like before, after read in data, we fit the model in lavaan package. 
+<<theta estimate>>=
+# estimate theta 
+## theta<-cfa(model,data=Data,group="g",group.equal="loadings")
+@
+Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
+<<Get B Estimate>>=
+library(strucchange)
+ for(i in 1:1327) {
+  parnum <- c(1,2,3,4,5)
+  tname <- c("dmax","CvM","maxLM")
+ test <- paste(rep(tname,length(parnum)), rep(parnum, each=length(tname)),
+               sep="")
+  nrep=1
+  pval <- matrix(rep(NA,length(test) * nrep), ncol = length(test))
+   colnames(pval) <- test
+    mz_gefp <- vector("list", length(parnum))
+    for(j in 1:length(parnum)){
+      mz_gefp[[j]] <- try(gefp(theta, fit = NULL, vcov = info.mzfit,
+                               order.by = yg$agegroup, sandwich = FALSE,
+                                 parm = eval(parse(text=parnum[j]))),
+                            silent = TRUE)
+     if(!inherits(mz_gefp[[j]], "try-error")) {
+       pval[i, (j-1)*3 + 1] <- sctest(mz_gefp[[j]],  functional = maxBB)$p.value
+      pval[i, (j-1)*3 + 2] <- sctest(mz_gefp[[j]],  functional = meanL2BB)$p.value
+       pval[i, (j-1)*3 + 3] <- sctest(mz_gefp[[j]],  functional = supLM(0.1))$p.value
+         }
+    }
+   }
+  size = 0.05
+  rval <- colMeans(pval < size, na.rm = TRUE)
+  return(pval)
+@
 

### 2013-07-30 frontiers/frontiers.Rnw
read in gratitude data and conduct LRT. Update the continuous one. Update the beginning part of the ordinal one.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@226 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \subsection{Multiple Group Analysis}
-First of all, we need to read in data to R.  
+Froh et al had a large sample of youth(n=1401, ranging from late childhood(10 years old) to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, including Gratitude Questionnaire 6(GQ-6; addref), Gratitude Adjective Checklist(GAC; addref) and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); addref).
+In this specific analysis, the authors were interested in whether the youth factor structure for GQ-6 is tau equivalent model or a congeneric model.
+
+To conduct the analysis, first of all, we need to read in data to R and clean the data.
 <<read data>>=
 # Reading in the data
-## (need data here)Data <- read.table("example.dat")
-## names(Data) <- c("y1","y2","y3","y4","y5","y6","g")
+data("YouthGratitude",package="psychotools")
+# remove cases with 'imputed' values(not in 1,...,9)
+yg<-YouthGratitude[apply(YouthGratitude[,4:28],1,function(x)all(x %in% 1:9)),]
 @
-Second, the model need to be specified using lavaan package. 
-<<specify model>>=
-library(lavaan)
-model <- 'f1=~NA*y1+y2+y3
-          f2=~NA*y4+y5+y6
-          f1 ~~1*f1
-          f2 ~~1*f2'
-@
-Third, we want to test which parameter(s) violate measurement invariance.
+Second, we want to test which parameter(s) violate measurement invariance.
 In lavaan package, loading, manifest variables' intercept and factor means could be tested together or specifically.
@@
 # estimate theta 
-## theta<-cfa(model,data=Data,group="g",group.equal="loadings")
+library(lavaan)
+theta<-cfa(
+  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+  data=yg,group="agegroup",meanstructure=TRUE,
+  group.equal="loadings")
 @
@@
 # estimate theta A,B 
-## thetaAB<-cfa(model,data=Data,group="g")
+thetaAB<-cfa(
+  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+  data=yg,group="agegroup",meanstructure=TRUE)
 @
@@
 # LRT
-## measurementInvariance(model,data=Data,group="g",strict=FALSE)
+LRT<-anova(theta, thetaAB)
+LRT
 @
@@
 <<theta estimate>>=
-# estimate theta 
-## theta<-cfa(model,data=Data,group="g",group.equal="loadings")
+library(lavaan)
+theta<-cfa(
+  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+  data=yg,group="agegroup",meanstructure=TRUE,
+  group.equal="loadings")
 @
 Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
-<<Get B Estimate>>=
-library(strucchange)
- for(i in 1:1327) {
-  parnum <- c(1,2,3,4,5)
-  tname <- c("dmax","CvM","maxLM")
- test <- paste(rep(tname,length(parnum)), rep(parnum, each=length(tname)),
-               sep="")
-  nrep=1
-  pval <- matrix(rep(NA,length(test) * nrep), ncol = length(test))
-   colnames(pval) <- test
-    mz_gefp <- vector("list", length(parnum))
-    for(j in 1:length(parnum)){
-      mz_gefp[[j]] <- try(gefp(theta, fit = NULL, vcov = info.mzfit,
-                               order.by = yg$agegroup, sandwich = FALSE,
-                                 parm = eval(parse(text=parnum[j]))),
-                            silent = TRUE)
-     if(!inherits(mz_gefp[[j]], "try-error")) {
-       pval[i, (j-1)*3 + 1] <- sctest(mz_gefp[[j]],  functional = maxBB)$p.value
-      pval[i, (j-1)*3 + 2] <- sctest(mz_gefp[[j]],  functional = meanL2BB)$p.value
-       pval[i, (j-1)*3 + 3] <- sctest(mz_gefp[[j]],  functional = supLM(0.1))$p.value
-         }
-    }
-   }
-  size = 0.05
-  rval <- colMeans(pval < size, na.rm = TRUE)
-  return(pval)
+<<Get >>=
+gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
+       vcov=info_full,sandwich=FALSE,parm=1:4)
+pval=rep(NA,3)
+pval[1]<-sctest(gefp1,functional=maxBB)$p.value
+pval[2]<-sctest(gefp1,functional=meanL2BB)$p.value
+pval[3]<-sctest(gefp1,functional=supLM(0.1))$p.value
+pval
 @
+
+\subsection{Ordinal Test}
+However, in this dataset, $V$ is actually ordinal instead of continuous.
+There is only a partial ordering of individuals with respect to $V$,i.e.
+observations with the same level of $V$ have no unique ordering.
+The ordinal statistics proposed here are similar to those described in Equation and above.
 

### 2013-07-31 frontiers/frontiers.Rnw
Frontiers paper: o Added to background section o Minor update of refs.bib
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@227 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\section{Theoretical Detail}
-This section contains background on the theory underlying the proposed statistics; for a more detailed account, see Merkle and Zeileis\citeyearNP{MerZei13}. 
-We consider situations in which a $p$-dimensional variable $X$ with observations $x_{i},i=1,...n$ is described by a model with density $f(x_{i};\theta)$ and associated joint log-likelihood
+\section{Background}
+This section contains background and discussion of the 
+proposed statistics as applied to SEM; for a more detailed account, see
+\citeA{MerZei13} and \citeA{MerFanZei}.  For details on the
+statistics' application to general statistical models, see
+\citeA{zeihor07}.
 
+As currently implemented for SEM, the statistical tests described in
+this paper can be applied to models that are estimated via a
+multivariate normal or Wishart likelihood (or discrepancy) function
+(though extension to other discrepancy functions is straightforward).
+The tests are carried out after model estimation, making use of output
+associated with the fitted model.  In general, we fit a model that
+restricts parameters to be equal across observations, then carry out a
+``posthoc'' test to examine whether certain parameters instead vary
+across observations.  This procedure is similar in spirit to the
+calculation of modification indices CITE and to Lagrange multiplier
+tests CITE, and, in fact, these statistics can be viewed as  
+special cases of the family described here.
+
+Following model estimation, the tests primarily work on the {\em
+  scores} of the fitted model; these are defined as
+%% TODO indicate that this is evaluated at theta hat
 \begin{equation}
-\label{equ:likelihood}
-\ell(\theta;x_{1},...,x_{n})=\overset{n}{\underset{i=1}{\Sigma}}\ell(\theta;x_{i})=\overset{n}{\underset{i=1}{\Sigma}}logf(x_{i};\theta)
+\label{eq:score}
+s(\hat{\theta};x_{i}) = \left ( \frac{\partial
+      l(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
+      l(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
 \end{equation} 
+where $l(\theta; x_i)$ is the likelihood associated with individual
+$i$, $\theta$ is a $k$-dimensional parameter vector, and
+$\hat{\theta}$ is the maximum likelihood estimate.  In words, each
+individual has $k$ scores describing the extent to which the fitted
+model describes that particular individual.  Scores close to zero
+indicate a ``good'' description, and scores far from zero indicate a
+``bad'' description.  Each of an individual's $k$ scores represent one
+model parameter, so that the scores provide specific information about
+each parameter's fit to each individual.
 
-where $\theta$ is some $k-$dimensional parameter vector that characterizes the distribution. 
-We focus on applications where the density $f(x_{i};\theta)$ arises from a structural equation model with assumed multivariate normality, though the proposed tests extend beyond this family of models.
-Under the usual regularity condition\cite{}, the model parameters $\theta$ can be estimated by maximum likelihood(ML),i.e.,
-
-\begin{equation}
-\label{equ:ml}
-\hat{\theta}=\underset{\theta}{argmax}\ell(\theta;x_{1},...,x_{n})
-\end{equation}
-
-or equivalently by solving the first order conditions
-
-\begin{equation}
-\label{equ:max}
-\overset{n}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i})=0,
-\end{equation}
-
-where
-
-\begin{equation}
-\label{equ:score}
-s(\theta;x_{i})=(\frac{\partial l(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial l(\theta;x_{i})}{\partial(\theta_{k})})^{T}
-\end{equation}
-
-is the score function of the model(the partial derivative of the casewise likelihood contributions w.r.t.
-the parameters $\theta$).
-Evaluation of the score function at $\hat{\theta}$ for $i=1,...n$ essentially measures the extent to which the model maximizes each individual's likelihood: as an individual's score stray further from zero,the model provides a poorer description of that individual.
+%% Next: How scores can be used for testing
 

### 2013-08-01 frontiers/frontiers.Rnw
Added detail to background of frontiers paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@228 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 this paper can be applied to models that are estimated via a
-multivariate normal or Wishart likelihood (or discrepancy) function
-(though extension to other discrepancy functions is straightforward).
-The tests are carried out after model estimation, making use of output
+multivariate normal or Wishart likelihood (or discrepancy) function,
+with extension to other discrepancy functions being straightforward.
+The tests are carried out following model estimation, making use of output
 associated with the fitted model.  In general, we fit a model that
 restricts parameters to be equal across observations, then carry out a
-``posthoc'' test to examine whether certain parameters instead vary
+posthoc test to examine whether specific parameters vary
 across observations.  This procedure is similar in spirit to the
 calculation of modification indices CITE and to Lagrange multiplier
-tests CITE, and, in fact, these statistics can be viewed as  
+tests CITE, and, in fact, those statistics can be viewed as  
 special cases of the family described here.
@@
 $i$, $\theta$ is a $k$-dimensional parameter vector, and
-$\hat{\theta}$ is the maximum likelihood estimate.  In words, each
+$\hat{\theta}$ is the maximum likelihood estimate.  The cross-product
+of these scores forms the ``meat'' for the calculation of robust
+(Huber-White) standard errors. %TODO cite
+
+To expand on the above equation, each
 individual has $k$ scores describing the extent to which the fitted
@@
 
-%% Next: How scores can be used for testing
+To use these scores for testing, we order individuals according to
+an auxiliary variable (the variable against which we are testing 
+measurement invariance) and look for trends in the
+scores (roughly speaking).  For example, imagine that we are testing
+for measurement 
+invariance w.r.t.\ age.  If the manifest variables violate measurement
+invariance here, then some parameter estimates may be too large for
+young individuals and too small for old individuals (say).  This
+result would be reflected in the scores, where young individuals'
+scores would be greater than zero and old individuals' scores would be
+less than zero.  Conversely, if measurement invariance holds, then all
+individuals' scores will fluctuate around zero.
+
+To formalize the ideas in the previous paragraph, we focus on a cumulative
+sum of the ordered scores.  Specifically, we focus on how the
+cumulative sum fluctuates as more individuals' scores are added to it,
+starting with the youngest and ending with the oldest individual
+(assuming that age is the auxiliary variable of interest).  Under the
+hypothesis of measurement invariance, a central limit theorem can be
+used to show that the fluctuation of the cumulative sum follows a
+Brownian bridge.  This result allows us to calculate $p$-values and
+critical values for test statistics under the hypothesis of
+measurement invariance.  This includes test statistics associated with
+all model parameters and subsets of model parameters.
+
+%% TODO focus on ordinal
+Multiple test statistics are available, depending on how one
+summarizes the behavior of the cumulative sum of scores.  One could
+take the absolute maximum that the cumulative sum attains for any
+parameter of interest, resulting in a {\em double max} statistic.
+Alternatively, one could sum the (squared) cumulative sum across parameters
+of interest and take the maximum across individuals, resulting in a
+{\em maximum Lagrange multiplier} statistic 
+\cite<see>[for further discussion of this idea]{MerZei13}.
 

### 2013-08-05 frontiers/frontiers.Rnw
turn off the signif codes in anova. use sctest.lavaan function to get the continuous and ordinal statistics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@229 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 source("mz.R")
+source("../www/efpFunctional-cat.R")
+source("../pkg/sctest.lavaan.R")
 
@@
 # LRT
-LRT<-anova(theta, thetaAB)
-LRT
+LRT<-as.matrix(anova(theta, thetaAB))
+print(LRT,na.print="", quote=FALSE)
 @
@@
 Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
-<<Get >>=
+<<Get Estimate>>=
 gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
@@
 @
+In applying the proposed continuous test, the $p$ value for CvM and maxBB $p$ value is larger than likelihood test. Besides, CvM $p$ value is not significant.
 
+\SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
+Alternatively, we could use sctest.lavaan function to get the results. 
+<<Get Estimate Alternative Way>>=
+DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
+CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
+#maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
+DM
+CvM
+#maxLM
+@
 \subsection{Ordinal Test}
@@
 observations with the same level of $V$ have no unique ordering.
-The ordinal statistics proposed here are similar to those described in Equation and above.
+The ordinal statistics proposed here are similar to those described in Equation and above, except that we focus on "bins" of individuals at each level of the ordinal variable.
+That is, instead of aggregating over all $i=1,...n$ individuals, we first compute cumulative proportions $t_{l}(l=1,...,m-1)$ associated with the first $m-1$ levels of $V$.
+We then aggregate the cumulative scores only over $i_{l}=\lfloor n \cdot t_{l} \rfloor$.
+Test statistics related can be written as
+\begin{eqnarray}
+    \label{eq:wdm}
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
+    \label{eq:maxlmo}
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
+\end{eqnarray}
 
-\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
+The difference between the continuous test and ordinal test is that we don't force arbitary order on the observations within the same age groups.
+Therefore, this ordinal test should reflect more of the data.
+In terms of code, we still need the gefp function to get the B estimate.
+After that, we can use different functions to aggregate and scale B estimation.
+The function are called ordwmax and ol2bb1.fun, respectively.
+<<Ordinal Test>>=
+gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
+       vcov=info_full,sandwich=FALSE,parm=1:4)
+pval.ord=rep(NA,2)
+pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value
+pval.ord[2]<-sctest(gefp1,functional=ordL2BB(gefp1))$p.value
+pval.ord
+@
+
+<<Get Estimate Alternative Way for Ordinal>>=
+WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE) 
+#maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLMo", fplot=TRUE)
+WDMo
+#maxLMo
+@
+
+In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. Both $p-$values are clearly larger than that of the likelihood ratio test and neigher is significant at $\alpha=0.5$
+
 \section{Simulation 1}

### 2013-08-06 frontiers/frontiers.Rnw
fix the plot problem generated by sctest functions. didn't change anything in simall functions.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@230 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
+\SweaveOpts{engine = R, eps = TRUE, echo = TRUE, pdf=TRUE}
 Alternatively, we could use sctest.lavaan function to get the results.
+
 <<Get Estimate Alternative Way>>=
 DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
+DM
 CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
+CvM
 #maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
-DM
-CvM
 #maxLM
 @
+
+\begin{figure}
+\label{fig:statistics}
+\includegraphics[page=1]{Rplots}
+\end{figure}
+
+\begin{figure}
+\label{fig:statistics}
+\includegraphics[page=2]{Rplots}
+\end{figure}
+
 \subsection{Ordinal Test}
@@
 WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
-#maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLMo", fplot=TRUE)
 WDMo 
-#maxLMo
+maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fplot=TRUE)
+maxLMo
 @
+
+\begin{figure}
+\label{fig:statistics}
+\includegraphics[page=3]{Rplots}
+\end{figure}
+
+\begin{figure}
+\label{fig:statistics}
+\includegraphics[page=4]{Rplots}
+\end{figure}
 
@@
 
-\section{Simulation 1}
-
-In Simulation 1, we examined the extent to which the proposed tests
-are sensitive to the specific model parameter that violates measurement
-invariance.  If, say, a factor loading violates measurement
-invariance, it is plausible that this violation impacts other
-parameters such as factor covariances or other loadings 
-associated with the factor in question.  Thus, the goal of the
-simulation was to examine the extent to which the proposed tests
-attribute the measurement invariance violation to the parameters that
-are truly in violation.
-
-\subsection{Methods}
-To examine these specificity issues, we generated data from a
-two-factor model with three indicators each (the same model used in
-Merkle and Zeileis, \citeyearNP{MerZei13}). The measurement invariance
-violation occurred in one of three parameters: the factor loading
-associated with the first indicator $\lambda_{11}$, the unique variance associated
-with the first indicator $\epsilon_{11}$, or the factor covariance $\sigma_{12}$.  
-We then tested for measurement invariance in seven subsets of parameters: the three
-individual parameters noted above, all six factor loadings $\lambda_{11}--\lambda_{62}$, 
-all six unique variances $\epsilon_{11}--\epsilon_{66}$. The path diagram is shown in  
-Figure~\ref{fig:path}.
-
-\begin{figure}
-\caption{Path diagram representing the factor analysis model 
-         used for simulations.}
-\label{fig:path}
-\includegraphics{PathDiagram}
-\end{figure}
-
-Power and Type I error were examined across three sample sizes
-($n=100,200,500$) and 17 magnitudes of invariance violations. These
-violations were based on the auxiliary variable $V$: individuals below
-the 50th percentile of $V$ deviated from individuals above the 50th
-percentile  by $d$ times the parameters' asymptotic standard errors
-(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
-For each combination of sample size ($n$) $\times$ violation
-magnitude ($d$) $\times$ violating parameter, 5000 datasets were
-generated and tested.
-
-\subsection{Results}
-Full simulation results are presented in Figure~\ref{fig:sim11res} to
-Figure~\ref{fig:sim13res}; all three test statistics exhibited very
-similar results, so the figures include only results for the CvM statistic.
-Figure~\ref{fig:sim11res} displays power curves as a function of 
-violation magnitude for the first
-factor loading $\lambda_{11}$, with panels reflecting the parameters 
-being tested and lines reflecting $n$.
-Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
-similar power curves when the factor covariance 
-$\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
-invariance, respectively.
-
-From these figures, one generally observe that the tests isolate 
-the parameter violating measurement invariance.  Additionally, the tests 
-have somewhat-higher power to detect measurement invariance violations 
-in the factor loading and factor covariance parameters, as opposed to the error variance
-parameter.  Finally, simultaneous tests of all factor loadings or all error
-parameters result in decreased power, as compared to the situation
-where one tests only the violating parameter.  This occurs because, in
-testing a subset of parameters (only one of which violates measurement
-invariance), we are effectively dampening the signal of a measurement
-invariance violation.
-
-In summary, we found that the proposed tests can attribute
-measurement invariance violations to the correct parameter.  This
-provides evidence that, in practice, one can have confidence in the
-tests' abilities to locate the measurement invariance violation.  Of
-course, this statement is qualified by the fact that, in this
-simulation, the model was correctly specified.  In the following
-simulations, we examine the tests' performance when the model is
-misspecified.
-
-%% Example of how to call a simulation in R:
-<<sim1>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1090)
-
-if(file.exists("sim1.rda")) {
-  load("sim1.rda")
-} else {
-  ## Arguments lead to small-scale simulation:
-  sim1 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
-  save(sim1, file="sim1.rda")
-}
-
-## Add to column names so that plots look nicer
-sim1$nobs <- paste("n=", sim1$nobs, sep="")
-
-#Not sure how to express all loadings and all error terms in greek letters.
-greekExprList<-expression(lambda[11],epsilon[11],sigma[12],lambda[11]--lambda[62],
-                          epsilon[11]--epsilon[66])
-greekpar<-as.expression(greekExprList)
-
-@
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 1.Violation Parameter is loading}
-\label{fig:sim11res}
-\setkeys{Gin}{width=\textwidth}
-<<sim11res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey1 <- simpleKey(unique(sim1$nobs), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
-             subset = diff <= 4 & parms == "loading" & test=="CvM" &  pars %in% 
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
-             as.table=TRUE))
-@
-\end{figure}
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 1.Violation Parameter is covariance}
-\label{fig:sim12res}
-\setkeys{Gin}{width=\textwidth}
-<<sim12res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
-             subset = diff <= 4 & parms == "var" & test == "CvM" &  pars %in% 
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
-             as.table=TRUE))
-@
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for Simulation 1.Violation Parameter is error terms}
-\label{fig:sim13res}
-\setkeys{Gin}{width=\textwidth}
-<<sim13res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
-             subset = diff <= 4 & parms == "error" & test == "CvM" &  pars %in%
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
-             as.table=TRUE))
-@
-\end{figure}
-
-%###########################################################################################
-\section{Simulation 2}
-In Simulation 2, we examine the extent to which the results of
-Simulation 1 are robust to model misspecification. Specifically, we
-generate data from the factor analysis model used in the previous
-section, but with adding an extra loading from the second factor to the first
-indicator (of the first factor).  The estimated model lacks this extra
-loading, however. The goal of this simulation is to examine the proposed statitics'
-power and spicificity under model misspecification.
-
-\subsection{Method}
-Measurement invariance violations could occur in
-each of the three parameters from Simulation 1 (factor loading,factor covariance,unique
-variance), and they could also occur in the extra,unmodeled loading. 
-Sample size and magnitude of measurementinvariance violation were manipulated 
-in the same way as they were in Simulation 1.The tested parameters are the same as Simulation
-1,too.
-
-\subsection{Results}
-Results are presented by the paramter that violated measurement invariance.
-Figure~\ref{fig:sim21res} displays results when the unmodeled loading violates 
-measurement invariance. One can generally observe that 
-the first loading and error variance parameter showed high measurement invariance 
-violation simultaneously, while factor covariance did not demonstrate violation.
-
-Figure~\ref{fig:sim22res} displays results when the loading $\lambda_{11}$ violates
-measurement invariance. Even though lacking one loading associated with the first indicator, 
-CvM statistic showed enough power to detect the first loading violation specifically. 
-Similar pattern of specificity and power were observed when the violation parameter 
-was $\sigma_{12}$ (Figure~\ref{fig:sim23res}) and $\epsilon_{11}$ (Figure~\ref{fig:sim24res}).
-
-In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
-as the umodeled parameter does not violate measurment invariance. If the unmodeled parameter 
-dose violate measurement invariance, however, then the tests assign this violation to other 
-parameters that do not violate measurement invariance. The impacted parameters include the error
-variance and other loadings associated with the manifest variables that had the unmodeled 
-loading. Thus, prior to testing measurement invariance, it is important to study the extent to
-which the hypothesized model includes all parameters of importance.
-
-%% Example of how to call a simulation in R:
-<<sim2>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1090)
-
-if(file.exists("sim2.rda")) {
-  load("sim2.rda")
-} else {
-  ## Arguments lead to small-scale simulation:
-  sim2 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("extra",
-                    "extra+loading", "extra+var","extra+error"))
-  save(sim2, file="sim2.rda")
-}
-
-## Add to column names so that plots look nicer
-sim2$nobs <- paste("n=", sim2$nobs, sep="")
-
-@
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 2.Violation Parameter is extra loading}
-\label{fig:sim21res}
-\setkeys{Gin}{width=\textwidth}
-<<sim21res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey2 <- simpleKey(unique(sim2$nobs), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff |pars , group = ~ nobs, data = sim2,
-             subset = diff <= 4 & parms == "extra" & test == "CvM" & pars %in%
-              c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
-@
-\end{figure}
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 2.Violation Parameter is loading, 
-         but model misspecified}
-\label{fig:sim22res}
-\setkeys{Gin}{width=\textwidth}
-<<sim22res, fig=TRUE, height=7, width=7>>=
-print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
-             subset = diff <= 4 & parms == "extra+loading" & test == "CvM" & pars %in% 
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
-@
-\end{figure}
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 2.Violation Parameter is var, 
-         but model misspecified}
-\label{fig:sim23res}
-\setkeys{Gin}{width=\textwidth}
-<<sim23res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
-             subset = diff <=4 & parms == "extra+var" & test == "CvM" & pars %in%
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
-@
-\end{figure}
-
-%% Example of how to create a figure of results
-\begin{figure}
-\caption{Simulated power curves for Simulation 2.Violation Parameter is var, 
-         but model misspecified}
-\label{fig:sim24res}
-\setkeys{Gin}{width=\textwidth}
-<<sim24res, fig=TRUE, height=7, width=7>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-## 
-print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
-             subset = diff <=4 & parms == "extra+error" & test == "CvM" & pars %in% 
-             c("1","7","13","1:6","7:12"),
-              strip = function(which.given, which.panel, var.name,
-                        strip.levels = FALSE,
-                        strip.names = TRUE, ...) {
-       	      strip.default(which.given, which.panel,
-       	               var.name = greekpar[which.panel],
-                       strip.levels = FALSE,
-                       strip.names = TRUE, ...)},
-             layout = c(3,2),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
-@
-\end{figure}
+\section{Simulation}
 

### 2013-08-06 frontiers/simall.R

@@ -770,7 +770,7 @@ if(FALSE) {  ## Ignore this line, it just tells R to ignore this code:
   args.list <- c(xyplot,list(nrow=6,ncol=7))
   dev.off()
 
-
+  
 }
 
  

### 2013-08-07 frontiers/frontiers.Rnw
write simall-ordinal.R file. update frontiers.Rnw simulation section R code
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@231 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \section{Tutorial}
-In general, a set of scales is defined to be measurement invariant with respect to an auxiliary variable $V$ if:
+In general, a set of scales is defined to be measurement invariant with respect
+to an auxiliary variable $V$ if:
 
@@
 
-where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable for variable $i$ that the scales purport to measure, and $f$ is the model's distributional form {addref}.
-If the cut point of $v$ is known in advance, i.e.
-gender, nested multiple group models {addref} coupled with likelihood ratio test are most commonly applied.
-Suppose there are two groups A and B. The null hypothesis of measurement invariance is
+where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
+for variable $i$ that the scales purport to measure, and $f$ is the model's 
+distributional form {addref}. If the cut point of $v$ is known in advance, i.e. 
+gender, nested multiple group models {addref} coupled with likelihood ratio test 
+are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
+of measurement invariance is
 
@@
 
-The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ (group A and group B) can be obtained by Maximum Likelihood(ML) from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,respectively.
-The LR test statistic for the given threshold $v$ is then
+The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ 
+(group A and group B) can be obtained by Maximum Likelihood(ML) 
+from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,
+respectively. The LR test statistic for the given threshold $v$ is then
+
 \begin{equation}
@@
 
-has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal to the number difference of parameters in $\theta$.
+has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
+to the number difference of parameters in $\theta$.
 
-To be more specific, in the following section, we will introduce how to implement multi group analysis in lavaan package. For more details, see {addref}
+To be more specific, in the following section, we will introduce how to implement 
+multi group analysis in lavaan package. For more details, see {addref}
 \subsection{Multiple Group Analysis}
-Froh et al had a large sample of youth(n=1401, ranging from late childhood(10 years old) to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, including Gratitude Questionnaire 6(GQ-6; addref), Gratitude Adjective Checklist(GAC; addref) and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); addref).
-In this specific analysis, the authors were interested in whether the youth factor structure for GQ-6 is tau equivalent model or a congeneric model.
+Froh et al had a large sample of youth(n=1401, ranging from late childhood(10 years old) 
+to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, 
+including Gratitude Questionnaire 6(GQ-6; addref), Gratitude Adjective Checklist(GAC; addref) 
+and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); addref). 
+In this specific analysis, the authors were interested in whether the youth factor 
+structure for GQ-6 is tau equivalent model or a congeneric model.
 
@@
 Second, we want to test which parameter(s) violate measurement invariance. 
-In lavaan package, loading, manifest variables' intercept and factor means could be tested together or specifically.
-For more details, please see {addref}.
-Since in most cases researchers are interested in the factor loading invariance, we will use that as the example in this tutorial.
-To implement LRT as ~\ref{equ:LRT}, we need to get estimation for $\hat{\theta}$, without differing groups.
+In lavaan package, loading, manifest variables' intercept and factor means 
+could be tested together or specifically. For more details, 
+please see {addref}. Since in most cases researchers are 
+interested in the factor loading invariance, we will use that as the 
+example in this tutorial.
+To implement LRT as ~\ref{equ:LRT}, we need to get estimation 
+for $\hat{\theta}$, without differing groups.
 
@@
 @
-In the end, we can get the LRT results by calling the measurement invariance function in lavaan. 
+In the end, we can get the LRT results by conducting 
+log likelihood test, which is done by anova function in R. 
 <<LRT>>=
@@
 \subsection{Extensions for Unknown Group}
-The multiple group test performs typically well when the $V$ is a relatively-small number of categories and the sample size is not too large (LRT is sensitive to large sample and will return "false" significance results, details see next section).
-However, when $V$ (in this case, age) continuous variable, however, multiple-group analysis usually cannot be performed because there are no existing groups.
-Instead, we can follow the procedure proposed by Ed Merkle and Zeileis.
-Specifically, we can fit a model whose parameters are restricted to be equal across across all individuals and then examine how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$.
-To formalize the idea, we assume that the observations are ordered w.r.t.
-$V$ and define the $k$ dimensional cumulative score process as
+The multiple group test performs typically well when the $V$ 
+is a relatively-small number of categories and the sample size 
+is not too large (LRT is sensitive to large sample and will 
+return "false" significance results, details see next section). 
+However, when $V$ (in this case, age) continuous variable, however, 
+multiple-group analysis usually cannot be performed because there are 
+no existing groups. Instead, we can follow the procedure proposed by 
+Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters 
+are restricted to be equal across across all individuals and then examine 
+how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. 
+To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and 
+define the $k$ dimensional cumulative score process as 
 \begin{equation}
 \label{equ:B}
-B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
+B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}
+{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
 \end{equation}
-where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ is some consistent estimate of the covariance matrix of the scores,e.g.
-information matrix.
+where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ 
+is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. 
 ~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. 
 $V$ and decorrelates the scores between $k$ components of the parameters in the model. 
-Further, there exists a functional central limit theorem that allows us to make formal inference with this cumulative score process.
-Assuming that individuals are independent and the usual ML regularity condition hold, it is possible to show that(addref)
+Further, there exists a functional central limit theorem 
+that allows us to make formal inference with this cumulative 
+score process. Assuming that individuals are independent and the usual 
+ML regularity condition hold, it is possible to show that(addref) 
 \begin{equation}
@@
 \end{equation} 
-where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ is a $k$ dimensional Brownian bridge.
-Thus, we can construct tests of measurement invariance by comparing the behavior of cumulative score process to that of a Brownian bridge.
-In practice, there are three test statistics can be used: double maximum, CvM and maxLM.
+where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ 
+is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement 
+invariance by comparing the behavior of cumulative score process to that of 
+a Brownian bridge. In practice, there are three test statistics can be used: 
+double maximum, CvM and maxLM. 
 \begin{equation}
@@
 
-Specifically, we will use the previous example, except changing categorical agegroup to continuous age.
+Specifically, we will use the previous example, except assuming ordinal agegroup to be continuous.
 Like before, after read in data, we fit the model in lavaan package. 
 <<theta estimate>>=
-library(lavaan)
 theta<-cfa(
@@
 @
-Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
+Then we will use the gefp function in strucchange package to 
+get the $B(\hat{\theta}_{ij})$. After that, we can get the 
+three test statistics' significance by using sctest function.
+
 <<Get Estimate>>=
@@
 @
-In applying the proposed continuous test, the $p$ value for CvM and maxBB $p$ value is larger than likelihood test. Besides, CvM $p$ value is not significant.
+In applying the proposed continuous test, the $p$ value for 
+CvM and maxBB $p$ value is larger than likelihood test. 
+Besides, CvM $p$ value is not significant.
 
@@
 #maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
-#maxLM
+#maxLM (Problem: nonconsistent with sctest(gefp1))
 @
@@
 \subsection{Ordinal Test}
-However, in this dataset, $V$ is actually ordinal instead of continuous.
-There is only a partial ordering of individuals with respect to $V$,i.e.
-observations with the same level of $V$ have no unique ordering.
-The ordinal statistics proposed here are similar to those described in Equation and above, except that we focus on "bins" of individuals at each level of the ordinal variable.
-That is, instead of aggregating over all $i=1,...n$ individuals, we first compute cumulative proportions $t_{l}(l=1,...,m-1)$ associated with the first $m-1$ levels of $V$.
-We then aggregate the cumulative scores only over $i_{l}=\lfloor n \cdot t_{l} \rfloor$.
-Test statistics related can be written as
+However, in this dataset, $V$ is actually ordinal instead of 
+continuous. There is only a partial ordering of individuals 
+with respect to $V$,i.e. observations with the same level 
+of $V$ have no unique ordering. The ordinal statistics 
+proposed here are similar to those described in Equation 
+and above, except that we focus on "bins" of individuals 
+at each level of the ordinal variable. That is, instead 
+of aggregating over all $i=1,...n$ individuals, we 
+first compute cumulative proportions $t_{l}(l=1,...,m-1)$ 
+associated with the first $m-1$ levels of $V$. 
+We then aggregate the cumulative scores only over 
+$i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics 
+related can be written as 
 \begin{eqnarray}
     \label{eq:wdm}
-      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ 
+      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
+      \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
     \label{eq:maxlmo}
-      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} 
+      ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
+      \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
 \end{eqnarray}
 
-The difference between the continuous test and ordinal test is that we don't force arbitary order on the observations within the same age groups.
-Therefore, this ordinal test should reflect more of the data.
-In terms of code, we still need the gefp function to get the B estimate.
-After that, we can use different functions to aggregate and scale B estimation.
+The difference between the continuous test and ordinal 
+test is that we don't force arbitary order on the observations 
+within the same age groups. Therefore, this ordinal test should 
+reflect more of the data. In terms of code, we still need the gefp 
+function to get the B estimate. After that, we can use different 
+functions to aggregate and scale B estimation. 
 The function are called ordwmax and ol2bb1.fun, respectively. 
@@
 
-In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. Both $p-$values are clearly larger than that of the likelihood ratio test and neigher is significant at $\alpha=0.5$
+In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. 
+Both $p-$values are clearly larger than that of the 
+likelihood ratio test and neither is significant at $\alpha=0.5$
 
 \section{Simulation}
+\section{Simulation 1}
+
+In Simulation 1, we examined the extent to which the proposed tests
+are sensitive to the specific model parameter that violates measurement
+invariance.  If, say, a factor loading violates measurement
+invariance, it is plausible that this violation impacts other
+parameters such as factor covariances or other loadings 
+associated with the factor in question.  Thus, the goal of the
+simulation was to examine the extent to which the proposed tests
+attribute the measurement invariance violation to the parameters that
+are truly in violation.
+
+\subsection{Methods}
+To examine these specificity issues, we generated data from a
+two-factor model with three indicators each (the same model used in
+Merkle and Zeileis, \citeyearNP{MerZei13}). The measurement invariance
+violation occurred in one of three places: the factor loadings
+ $\lambda_{11}--\lambda_{62}$, the unique variance associated
+ $\epsilon_{11}--\epsilon_{66}$, or the factor covariance $\sigma_{12}$.  
+We then tested for measurement invariance in seven subsets of parameters: the three
+individual parameters noted above, all six factor loadings $\lambda_{11}--\lambda_{62}$, 
+all six unique variances $\epsilon_{11}--\epsilon_{66}$. The path diagram is shown in  
+Figure~\ref{fig:path}.
+
+\begin{figure}
+\caption{Path diagram representing the factor analysis model 
+         used for simulations.}
+\label{fig:path}
+\includegraphics{PathDiagram}
+\end{figure}
+
+Power and Type I error were examined across three sample sizes
+($n=120,480,960$) , three numbers of categories ($m=4,8,12$)and
+7 magnitudes of invariance violations. 
+The meaurement invariance violations began at level $1+m/2$ of $V$
+and were consistant thereafter. These
+violations were based on the auxiliary variable $V$: individuals below
+the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
+percentile  by $d$ times the parameters' asymptotic standard errors
+(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,1.5$. 
+For each combination of sample size ($n$) $\times$ violation
+magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
+generated and tested.Four statistics were examined above. In all conditions, we
+maintained equal sample sizes at each level of the ordinal variable.
+
+\subsection{Results}
+<<sim1>>=
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim1.rda")) {
+  load("sim1.rda")
+} else {
+  sim1 <- simulation()
+  sim1$nlevels <- factor(sim1$nlevels)
+  levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
+  sim1$nobs <- factor(sim1$nobs)
+  levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
+  levels(sim1$test) <- c("AIC","LM_uo", "LRT", "LRT_sb", "LM_o", "WDM_o",
+                         "YB97")
+  save(sim1, file="sim1.rda")
+}
+sim1$test <- factor(as.character(sim1$test),
+  levels = c("LM_o", "WDM_o", "LRT", "LM_uo", "AIC", "LRT_sb", "YB97"),
+  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo", "AIC", "LRT_sb", "YB97"))
+@
+
+\begin{figure}
+\caption{Simulated power curves for the ordered and unordered
+      $\max\mathit{LM}$ tests, the ordered double-max test, and the
+      likelihood ratio test across three sample sizes $n$, three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
+\label{fig:sim1res}
+\setkeys{Gin}{width=\textwidth}
+<<sim1res, fig=TRUE, height=7, width=7>>=
+sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
+sim1.tmp$test <- factor(sim1.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+\end{figure}
 

### 2013-08-07 frontiers/simall-ordinal.R

@@ -1,0 +1,231 @@
+
+## Data-generating process
+dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE, parms="loadings")
+{
+  ## Generates data from a factor analysis model that violates
+  ## measurement invariance.  Also generates an ordinal auxiliary
+  ## variable related to the invariance (loosely called "age").
+
+  ## If gradual==TRUE, mild instability starts halfway up age
+  ## and gets larger.  If gradual==FALSE, the instability starts
+  ## halfway up age and stays constant.
+  ## If anomaly==TRUE, instability only occurs at level (1+nlevels/2).
+  ## Otherwise, instability continues up the ordinal variable.
+  stopifnot(require("mvtnorm"))
+  
+  # ses is SE increase in parameters at last ordinal category.
+  # (This is constant if gradual==FALSE but not if gradual==TRUE.)
+  ses <- diff
+  sampsize <- nobs
+  sampsize.per.nobs <- nobs %/% nlevels
+
+  ## Deal with cases where nobs/nlevels is not an integer
+  age <- rep(NA, sampsize)
+  age[1:(nlevels*sampsize.per.nobs)] <- rep(1:nlevels, each=sampsize.per.nobs)
+  if (any(is.na(age))) age[is.na(age)] <- sample((1:nlevels),sum(is.na(age)))
+  half.level <- ifelse(nlevels %% 2 == 1, (nlevels %/% 2)+1, nlevels/2)
+  
+  # Define parameter vectors/matrices:
+  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5,-0.48,
+           29.32,24.70,14.84,10.59,19.30,18.01)
+  asym<-c( 7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
+            56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
+           1.024326,7.139776,4.666005,8.151785,3.696972,5.242366,
+            8.630417)
+
+  # Loadings:
+  lambda <- matrix(0,6,2)
+  lambda[,1] <- c(theta[1:3],0,0,0)
+  lambda[,2] <- c(0,0,0,theta[4:6])
+
+  phi <- matrix(theta[13],2,2)
[…]

### 2013-08-07 frontiers/frontiers.Rnw
delete sim1 code for now.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@232 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \subsection{Results}
-<<sim1>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1163)
-
-if(file.exists("sim1.rda")) {
-  load("sim1.rda")
-} else {
-  sim1 <- simulation()
-  sim1$nlevels <- factor(sim1$nlevels)
-  levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
-  sim1$nobs <- factor(sim1$nobs)
-  levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
-  levels(sim1$test) <- c("AIC","LM_uo", "LRT", "LRT_sb", "LM_o", "WDM_o",
-                         "YB97")
-  save(sim1, file="sim1.rda")
-}
-sim1$test <- factor(as.character(sim1$test),
-  levels = c("LM_o", "WDM_o", "LRT", "LM_uo", "AIC", "LRT_sb", "YB97"),
-  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo", "AIC", "LRT_sb", "YB97"))
-@
-
-\begin{figure}
-\caption{Simulated power curves for the ordered and unordered
-      $\max\mathit{LM}$ tests, the ordered double-max test, and the
-      likelihood ratio test across three sample sizes $n$, three
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
-\label{fig:sim1res}
-\setkeys{Gin}{width=\textwidth}
-<<sim1res, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
-sim1.tmp$test <- factor(sim1.tmp$test)
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-\end{figure}
 

### 2013-08-08 frontiers/simall-ordinal.R
update simall-ordinal.R file, simulation function to include parms condition
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@234 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -113,7 +113,8 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE, 
 }
 
 ## Evaluate power simulation on a single dgp() scenario
-testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, test = NULL, verbose = TRUE, ...)
+testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, test = NULL,
+                      verbose = TRUE, ...)
 {
   pval <- matrix(rep(NA, length(test) * nrep), ncol = length(test))
   colnames(pval) <- test
@@ -124,7 +125,8 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, 
     mz <- ordfit(d)
 
     ## Requires estfun.lavaan
-    ord_gefp  <- try(gefp(mz, fit = NULL, vcov = info.mzfit, order.by = d$age, sandwich = FALSE, parm = eval(parse(text=parnum[j]))), silent = TRUE)
+    ord_gefp  <- try(gefp(mz, fit = NULL, vcov = info.mzfit, order.by = d$age, sandwich =
+                          FALSE, parm = eval(parse(text=parnum[j]))), silent = TRUE)
 
     if(!inherits(ord_gefp[[j]], "try-error")) {
       pval[i, (j-1)*7+1] <- sctest(ord_gefp,  functional = ordfun)$p.value
@@ -148,7 +150,7 @@ simulation <- function(diff = seq(0, 1.5, by = 0.25),parms="error",
   nobs = c(120, 480, 960), nlevels = c(4, 8, 12), gradual = FALSE,
   anomaly = FALSE, verbose = TRUE, ...)
 {
-  prs <- expand.grid(diff = diff, nlevels = nlevels, nobs = nobs)
+  prs <- expand.grid(diff = diff, nlevels = nlevels, nobs = nobs, parms = parms)
   nprs <- nrow(prs)
 
   parnum <- c("1:6",13,"7:12")
@@ -174,7 +176,8 @@ simulation <- function(diff = seq(0, 1.5, by = 0.25),parms="error",
   if (do.parallel){
     pow <- mclapply(1:nprs, function(i){
       testpower(diff = prs$diff[i], nobs = prs$nobs[i],
-                parms = prs$parms[i],parnum = parnum, nlevels = prs$nlevels[i], gradual = gradual, test = test,
[…]

### 2013-08-08 frontiers/frontiers.Rnw
Frontiers manuscript: Added to background and tutorial sctest.lavaan: Added fnl argument in case critical values                have already been simulated.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@235 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+## turn off stars
+options(show.signif.stars = FALSE)
+
 ## convenience function for plotting boundaries without color
@@
 In the following section, we first briefly review the 
-theoretical framework of proposed statistical tests. Subsequently, we study the tests'
+theoretical framework of proposed statistical tests and
+provide a short tutorial illustrating the use of the tests in R.
+Next, we study the tests'
 ability through simulations that mimic practical research. 
@@
 restricts parameters to be equal across observations, then carry out a
-posthoc test to examine whether specific parameters vary
+posthoc test to examine whether specific parameters appear to vary
 across observations.  This procedure is similar in spirit to the
@@
 scores would be greater than zero and old individuals' scores would be
-less than zero.  Conversely, if measurement invariance holds, then all
+less than zero (though the sign of the scores will also depend on whether a
+function is being minimized or maximized).  Conversely, if measurement
+invariance holds, then all 
 individuals' scores will fluctuate around zero.
@@
 To formalize the ideas in the previous paragraph, we focus on a cumulative
-sum of the ordered scores.  Specifically, we focus on how the
+sum of the ordered scores.  This cumulative sum may be written as
+\begin{equation} \label{eq:cumscore}
+  {\bm B}(t; \hat {\bm \theta}) ~=~ %\hat {\bm I}^{-1/2} n^{-1/2}
+    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
+  \qquad (0 \le t \le 1)
+\end{equation}
+where $\lfloor nt \rfloor$ is the integer part of $nt$ (i.e., a floor
+operator) and $x_{(i)}$ reflects the individual with the
+$i^{\text{th}}$-smallest value of the auxiliary variable.
+We specifically focus on how the
 cumulative sum fluctuates as more individuals' scores are added to it,
 starting with the youngest and ending with the oldest individual
-(assuming that age is the auxiliary variable of interest).  Under the
+(assuming that age is the auxiliary variable of interest).  
+The summation is also typically premultiplied by the inverse square
+root of the information matrix, which serves to decorrelate the
+fluctuation processes associated with individual model parameters.
+
+Under the
 hypothesis of measurement invariance, a central limit theorem can be
-used to show that the fluctuation of the cumulative sum follows a
+used to show that the fluctuation of the above cumulative sum follows a
 Brownian bridge.  This result allows us to calculate $p$-values and
 critical values for test statistics under the hypothesis of
-measurement invariance.  This includes test statistics associated with
-all model parameters and subsets of model parameters.
+measurement invariance.  We can obtain test statistics associated with
+all model parameters, as well as test statistics associated with
+subsets of model parameters.
 
@@
 \cite<see>[for further discussion of this idea]{MerZei13}.
+%% TODO equations of statistics?
+
+Importantly, these statistics were derived for situations where
+individuals are uniquely ordered according to the auxiliary variable.
+This is not always the case for measurement invariance applications,
+where the auxiliary variable is often ordinal.  To remedy this
+situation, \citeA{MerFanZei} extended the framework to situations
+where one has an ordinal auxiliary variable of interest.  Essentially,
+one allows all individuals with the same value of the auxiliary
+variable to enter into the cumulative sum at the same time.  Test
+statistics analogous to those describe above are then computed, with
+modified critical values being adopted to reflect the change in the
+statistics' computation.
 
@@
 # estimate theta 
-library(lavaan)
 theta<-cfa(
@@
 # LRT
-LRT<-as.matrix(anova(theta, thetaAB))
-print(LRT,na.print="", quote=FALSE)
+anova(theta, thetaAB)
 @
+
 \subsection{Extensions for Unknown Group}
@@
 multiple-group analysis usually cannot be performed because there are 
-no existing groups. Instead, we can follow the procedure proposed by 
-Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters 
+no existing groups. Instead, we can follow the procedure described in
+the Background section.  That is, we can fit a model whose parameters 
 are restricted to be equal across across all individuals and then examine 
-how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. 
-To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and 
-define the $k$ dimensional cumulative score process as 
-\begin{equation}
-\label{equ:B}
-B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}
-{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
-\end{equation}
-where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ 
-is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. 
-~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. 
-$V$ and decorrelates the scores between $k$ components of the parameters in the model. 
-Further, there exists a functional central limit theorem 
-that allows us to make formal inference with this cumulative 
-score process. Assuming that individuals are independent and the usual 
-ML regularity condition hold, it is possible to show that(addref) 
-\begin{equation}
-B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
-\end{equation} 
-where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ 
-is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement 
-invariance by comparing the behavior of cumulative score process to that of 
-a Brownian bridge. In practice, there are three test statistics can be used: 
-double maximum, CvM and maxLM. 
-\begin{equation}
-DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
-\end{equation}
-\begin{equation}
-CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
-\end{equation}
-\begin{equation}
-maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
-\end{equation}
+how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
+values of $V$.
 
-Specifically, we will use the previous example, except assuming ordinal agegroup to be continuous.
+%% TO BE MOVED TO BACKGROUND
+%% In practice, there are three test statistics can be used: 
+%% double maximum, CvM and maxLM. 
+%% \begin{equation}
+%% DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
+%% \end{equation}
+%% \begin{equation}
+%% CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
+%% \end{equation}
+%% \begin{equation}
+%% maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
+%% \end{equation}
+
+To demonstrate these tests in R, we will use the youth gratitude data
+and assume that age group is continuous.
 Like before, after read in data, we fit the model in lavaan package. 
@@
 
+\readme{Ting: The problem with the last statistic is having
+  fplot=TRUE.  To get around this, you can add random noise or set
+  fplot=FALSE (I've done the latter).}
 <<Get Estimate Alternative Way>>=
-DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
+DM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
 DM
-CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
+CvM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
 CvM
-#maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
-#maxLM (Problem: nonconsistent with sctest(gefp1))
+maxLM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
+maxLM
 @
@@
 associated with the first $m-1$ levels of $V$. 
+%% TODO Move these details to Background section
 We then aggregate the cumulative scores only over 
@@
 pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value
-pval.ord[2]<-sctest(gefp1,functional=ordL2BB(gefp1))$p.value
+
+if(file.exists("opval1.rda")){
+  load("opval1.rda")
+} else {
+  opval1 <- ordL2BB(gefp1)
+  save(opval1, file="opval1.rda")
+}
+pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
 pval.ord
@@
 WDMo 
-maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fplot=TRUE)
+maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
+               fnl=opval1, fplot=TRUE)
 maxLMo
@@
 Both $p-$values are clearly larger than that of the 
-likelihood ratio test and neither is significant at $\alpha=0.5$
+likelihood ratio test and neither is significant at $\alpha=0.05$
 

### 2013-08-08 pkg/sctest.lavaan.R

@@ -1,4 +1,4 @@
-sctest.lavaan <- function(x, order.by, parm, stat, fplot = FALSE, ...){
+sctest.lavaan <- function(x, order.by, parm, stat, fnl = NULL, fplot = FALSE, ...){
   ## Carry out a score-based test of a fitted lavaan model x
 
   require("strucchange")
@@ -13,14 +13,17 @@ sctest.lavaan <- function(x, order.by, parm, stat, fplot = FALSE, ...){
   ## ordwmax(fp), ordL2BB(fp), maxBB, meanL2BB, supLM(0.1), rangeBB
   ## WDMo         maxLMo       DM     CvM       maxLM,      range
   ## TODO? Allow user to modify the 0.1 of supLM?
-  fnl <- switch(stat,
-                WDMo = ordwmax(fp),
-                maxLMo = ordL2BB(fp),
-                DM = maxBB,
-                CvM = meanL2BB,
-                maxLM = supLM(0.1),
-                range = rangeBB)
+  if(is.null(fnl)){
+    fnl <- switch(stat,
+                  WDMo = ordwmax(fp),
+                  maxLMo = ordL2BB(fp),
+                  DM = maxBB,
+                  CvM = meanL2BB,
+                  maxLM = supLM(0.1),
+                  range = rangeBB)
+  }
 
+  ## If stat argument is messed up
   if(is.null(fnl)) stop(paste("Argument stat=",stat," is unknown",sep=""))
   
   testres <- sctest(fp, functional = fnl)

### 2013-08-12 frontiers/frontiers.Rnw
update figure part, change code to include R code naturally
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@236 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\readme{Ting: The problem with the last statistic is having
-  fplot=TRUE.  To get around this, you can add random noise or set
-  fplot=FALSE (I've done the latter).}
-<<Get Estimate Alternative Way>>=
+\begin{figure}
+\caption{Get Continuous Estimate DM Alternative Way}
+\label{fig:conti1res}
+\setkeys{Gin}{width=\textwidth}
+<<conti1res, fig=TRUE, height=7, width=7>>=
 DM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
 DM
+@
+\end{figure}
+
+\begin{figure}
+\caption{Get Continuous Estimate CvM Alternative Way}
+\label{fig:conti2res}
+\setkeys{Gin}{width=\textwidth}
+<<conti2res, fig=TRUE, height=7, width=7>>=
 CvM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
 CvM
+@
+\end{figure}
+
+<<conti3resmaxLM>>=
 maxLM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
@@
 @
-
-\begin{figure}
-\label{fig:statistics}
-\includegraphics[page=1]{Rplots}
-\end{figure}
-
-\begin{figure}
-\label{fig:statistics}
-\includegraphics[page=2]{Rplots}
-\end{figure}
 
@@
 
-<<Get Estimate Alternative Way for Ordinal>>=
+\begin{figure}
+\caption{Get Ordinal Estimate WDMo Alternative Way}
+\label{fig:ord1res}
+\setkeys{Gin}{width=\textwidth}
+<<ord1res, fig=TRUE, height=7, width=7>>=
 WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
 WDMo 
+@
+\end{figure}
+
+\begin{figure}
+\caption{Get Ordinal Estimate maxLMo Alternative Way}
+\label{fig:ord2res}
+\setkeys{Gin}{width=\textwidth}
+<<ord2res, fig=TRUE, height=7, width=7>>=
 maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
@@
 @
-
-\begin{figure}
-\label{fig:statistics}
-\includegraphics[page=3]{Rplots}
-\end{figure}
-
-\begin{figure}
-\label{fig:statistics}
-\includegraphics[page=4]{Rplots}
 \end{figure}
@@
 
-In Simulation 1, we examined the extent to which the proposed tests
+In Simulation 1, we examined the extent
+
+ to which the proposed tests
 are sensitive to the specific model parameter that violates measurement

### 2013-08-13 frontiers/simall-ordinal.R
update simall-ordinal.R file. dgp function is OK. But in testpower function, ordfun can't accept list mode. Haven't figured out how to solve this.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@237 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,6 +1,6 @@
 
 ## Data-generating process
-dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE, parms="loadings")
+dgp <- function(nobs = 200, diff = 3, nlevels=4, gradual=FALSE, anomaly=FALSE, parms="loadings")
 {
   ## Generates data from a factor analysis model that violates
   ## measurement invariance.  Also generates an ordinal auxiliary
@@ -32,7 +32,7 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE, 
             56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
            1.024326,7.139776,4.666005,8.151785,3.696972,5.242366,
             8.630417)
-
+  mu<-theta[14:19]
   # Loadings:
   lambda <- matrix(0,6,2)
   lambda[,1] <- c(theta[1:3],0,0,0)
@@ -66,7 +66,11 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE, 
   for (i in (half.level+1):nlevels){
     tmp.n <- sum(age==i)
     tmp.ind <- which(age==i)
-    z <- t(rmvnorm(tmp.n,rep(0,2),phi))
+    tmp.lambda <- matrix(0,6,2)
+    tmp.lambda[,1] <- lambda[,1] 
+    tmp.lambda[,2] <- lambda[,2]
+    tmp.psi <- psi
+    tmp.phi <- phi
 
     mult <- ses
     if (anomaly) mult <- ifelse(i==(half.level+1), ses, 0)
@@ -76,36 +80,34 @@ dgp <- function(nobs = 200, diff = 3, nlevels=10, gradual=FALSE, anomaly=FALSE, 
     if (gradual){
       if (anomaly) stop("gradual=TRUE and anomaly=TRUE do not work together.")
       if (parms=="error"){
-      diag(psi) <- diag(psi) + (each.vary*asym[7:12])/sqrt(tmp.n)
-      tmp.psi <- psi
+      diag(tmp.psi) <- diag(psi) + (each.vary*asym[7:12])/sqrt(tmp.n)
     }
       if (parms=="var"){
[…]

### 2013-08-14 frontiers/frontiers.Rnw
update refs.bib for reference. go through the connection part in tutorials.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@238 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 for variable $i$ that the scales purport to measure, and $f$ is the model's 
-distributional form {addref}. If the cut point of $v$ is known in advance, i.e. 
-gender, nested multiple group models {addref} coupled with likelihood ratio test 
+distributional form \cite{mel89}. If the cut point of $v$ is known in advance, i.e. 
+gender, nested multiple group models \cite{bollen98} coupled with likelihood ratio test 
 are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
@@
 
-has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
+,which has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
 to the number difference of parameters in $\theta$.
 
-To be more specific, in the following section, we will introduce how to implement 
-multi group analysis in lavaan package. For more details, see {addref}
+To be more specific, in the following section, we will illustrate with an example to demonstrate how to implement multi group analysis in lavaan package. For more details, see \cite{vande12}
+
 \subsection{Multiple Group Analysis}
-Froh et al had a large sample of youth(n=1401, ranging from late childhood(10 years old) 
+Froh et al\citeA{FroFan11} had a large sample of youth(n=1401, ranging from late childhood(10 years old) 
 to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, 
-including Gratitude Questionnaire 6(GQ-6; addref), Gratitude Adjective Checklist(GAC; addref) 
-and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); addref). 
+including Gratitude Questionnaire 6(GQ-6; \citeNP{MccEmm02}), Gratitude Adjective Checklist(GAC; \citeNP{MccEmm02})  and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); \citeNP{ThoWat03}).
 In this specific analysis, the authors were interested in whether the youth factor
-structure for GQ-6 is tau equivalent model or a congeneric model.
+structure for GQ-6 is a tau equivalent model (factor loadings restricted to be equal across
+levels of age)or a congeneric model(all parameters free for each level of age).
 
@@
 could be tested together or specifically. For more details, 
-please see {addref}. Since in most cases researchers are 
+please see \cite{vande12}. Since in most cases researchers are 
 interested in the factor loading invariance, we will use that as the 
@@
 @
+Likelihood test is sensitive to large observations (CITE), it is not surprising to 
+have a significant result, although Froh et al\citeA{FroFan11} showed that tau equivalent
+model has as good alternative fit indices as congeneric model.
 
@@
 is a relatively-small number of categories and the sample size 
-is not too large (LRT is sensitive to large sample and will 
-return "false" significance results, details see next section). 
-However, when $V$ (in this case, age) continuous variable, however, 
+is not too large. However, when $V$ (in this case, age) continuous variable, however, 
 multiple-group analysis usually cannot be performed because there are 
@@
 \SweaveOpts{engine = R, eps = TRUE, echo = TRUE, pdf=TRUE}
-Alternatively, we could use sctest.lavaan function to get the results.
+Alternatively, we could use sctest.lavaan function to get the statistics value,
+corresponding $p$ value and the plot ordered by age level.
 
@@
 @
+Since there are more than one observations located in each age level. 
+So the observation order within each age level is not unique, thus, 
+maxLM can't produce the plot. Actually, all three continuous statistics 
+were calculated on the arbitrary order that particular dataset entered 
+in each age level. In order to accurately account for the multiple observations 
+at the same age level, it is more appropriate to use the proposed ordinal test.
 
 \subsection{Ordinal Test}
-However, in this dataset, $V$ is actually ordinal instead of 
-continuous. There is only a partial ordering of individuals 
+As a matter of fact, there is only a partial ordering of individuals 
 with respect to $V$,i.e. observations with the same level 
@@
 The difference between the continuous test and ordinal 
-test is that we don't force arbitary order on the observations 
+test is that we don't force arbitrary order on the observations 
 within the same age groups. Therefore, this ordinal test should 
-reflect more of the data. In terms of code, we still need the gefp 
+reflect more real status of the data. In terms of code, we still need the gefp 
 function to get the B estimate. After that, we can use different 
@@
 
+Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
 \begin{figure}
@@
 Both $p-$values are clearly larger than that of the 
-likelihood ratio test and neither is significant at $\alpha=0.05$
+likelihood ratio test and neither is significant at $\alpha=0.05$. 
+This result is consistent with alternative fit indices used in Froh et al paper\citeA{FroFan11}.
 

### 2013-08-16 frontiers/simall-ordinal.R
change simall-ordinal to replicate results of ordinal paper. It worked. Tomorrow to try parameter loading and var to see the specificity
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@239 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,6 +1,7 @@
 
 ## Data-generating process
-dgp <- function(nobs = 200, diff = 3, nlevels=4, gradual=FALSE, anomaly=FALSE, parms="loadings")
+set.seed=1090
+dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALSE, parms="loadings")
 {
   ## Generates data from a factor analysis model that violates
   ## measurement invariance.  Also generates an ordinal auxiliary
@@ -26,7 +27,7 @@ dgp <- function(nobs = 200, diff = 3, nlevels=4, gradual=FALSE, anomaly=FALSE, p
   half.level <- ifelse(nlevels %% 2 == 1, (nlevels %/% 2)+1, nlevels/2)
   
   # Define parameter vectors/matrices:
-  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5,-0.48,
+  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5, 0.48,
            29.32,24.70,14.84,10.59,19.30,18.01)
   asym<-c( 7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
             56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
@@ -105,7 +106,7 @@ dgp <- function(nobs = 200, diff = 3, nlevels=4, gradual=FALSE, anomaly=FALSE, p
     
 
     u <- t(rmvnorm(tmp.n,rep(0,6),tmp.psi))
-    z <- t(rmvnorm(n.invariant,rep(0,2),tmp.phi))
+    z <- t(rmvnorm(tmp.n,rep(0,2),tmp.phi))
     for (j in 1:tmp.n){
       datmat[tmp.ind[j],] <- mu + tmp.lambda%*%z[,j] + u[,j]
     }
@@ -122,7 +123,7 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, 
   colnames(pval) <- test
   
   for(i in 1:nrep) {
-    d <- dgp()
+    d <- dgp(...)
 
     mz <- ordfit(d)
 
@@ -134,14 +135,14 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, 
                           silent = TRUE)
 
[…]

### 2013-08-21 frontiers/frontiers.Rnw
Added to frontiers paper: o Added detail to background o Revised tutorial discussion o Revised code in tutorial
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@240 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\title{Tests of measurement invariance via stochastic processes:
-  {S}ome properties}
+\title{Score-based tests of measurement invariance: Use in practice}
 \twoauthors{Ting Wang and Edgar C.\ Merkle}{Achim Zeileis}
@@
 multivariate normal or Wishart likelihood (or discrepancy) function,
-with extension to other discrepancy functions being straightforward.
+with extension to other discrepancy functions being conceptually straightforward. 
 The tests are carried out following model estimation, making use of output
@@
 restricts parameters to be equal across observations, then carry out a
-posthoc test to examine whether specific parameters appear to vary
+posthoc test to examine whether specific parameters vary
 across observations.  This procedure is similar in spirit to the
@@
 s(\hat{\theta};x_{i}) = \left ( \frac{\partial
-      l(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
-      l(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
+      \ell(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
+      \ell(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
 \end{equation} 
-where $l(\theta; x_i)$ is the likelihood associated with individual
+where $\ell(\theta; x_i)$ is the likelihood associated with individual
 $i$, $\theta$ is a $k$-dimensional parameter vector, and
@@
 
-To expand on the above equation, each
+To verbally describe the above equation, each
 individual has $k$ scores describing the extent to which the fitted
@@
 an auxiliary variable (the variable against which we are testing 
-measurement invariance) and look for trends in the
-scores (roughly speaking).  For example, imagine that we are testing
+measurement invariance) and look for ``trends'' in the
+scores.  For example, imagine that we are testing
 for measurement 
@@
 result would be reflected in the scores, where young individuals'
-scores would be greater than zero and old individuals' scores would be
-less than zero (though the sign of the scores will also depend on whether a
+scores may be greater than zero and old individuals' scores 
+less than zero (though the sign of the scores will depend on whether a
 function is being minimized or maximized).  Conversely, if measurement
@@
 
-To formalize the ideas in the previous paragraph, we focus on a cumulative
+To formalize the ideas in the previous paragraph, we define a cumulative
 sum of the ordered scores.  This cumulative sum may be written as
-\begin{equation} \label{eq:cumscore}
+\begin{equation} 
+    \label{eq:cumscore}
   {\bm B}(t; \hat {\bm \theta}) ~=~ %\hat {\bm I}^{-1/2} n^{-1/2}
@@
 root of the information matrix, which serves to decorrelate the
-fluctuation processes associated with individual model parameters.
+fluctuation processes associated with individual model parameters (i.e., the process associated with one parameter is not correlated with the processes associated with other parameters).
 
@@
 measurement invariance.  We can obtain test statistics associated with
-all model parameters, as well as test statistics associated with
-subsets of model parameters.
+all model parameters and with subsets of model parameters.
 
@@
 Multiple test statistics are available, depending on how one
-summarizes the behavior of the cumulative sum of scores.  One could
+summarizes the behavior of the cumulative sum of scores.  For example, one could
 take the absolute maximum that the cumulative sum attains for any
-parameter of interest, resulting in a {\em double max} statistic.
+parameter of interest, resulting in a {\em double max} statistic (the maximum is taken across parameters and individuals).
 Alternatively, one could sum the (squared) cumulative sum across parameters
-of interest and take the maximum across individuals, resulting in a
-{\em maximum Lagrange multiplier} statistic 
-\cite<see>[for further discussion of this idea]{MerZei13}.
-%% TODO equations of statistics?
+of interest and take the maximum or the average across individuals, resulting in a {\em maximum Lagrange multiplier} statistic and Cram\'{e}r-von Mises statistic, respectively \cite<see>[for further discussion]{MerZei13}.
+These statistics are given by
+\begin{eqnarray}
+    \label{eq:dmax}
+    \mathit{DM} & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} \\
+        \label{eq:cvm}
+    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
+    \label{eq:maxlm}
+    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
+      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
+\end{eqnarray}
 
-Importantly, these statistics were derived for situations where
+Critical values associated with $\mathit{DM}$ can be obtained analytically, while critical values associated with the other statistics can be obtained from direct simulation \cite{Zei06} or from
+more refined techniques \cite{han97}.  This issue should not be important to the user, as critical values are obtained directly from the R implementation described later.
+
+Importantly, the above statistics were derived for situations where
 individuals are uniquely ordered according to the auxiliary variable.
@@
 one allows all individuals with the same value of the auxiliary
-variable to enter into the cumulative sum at the same time.  Test
-statistics analogous to those describe above are then computed, with
+variable to enter into the cumulative sum at the same time.  Analogous test 
+statistics are then computed, with
 modified critical values being adopted to reflect the change in the
@@
 
+For an ordinal auxiliary variable with $m$ levels, these modifications are based on $t_{\ell}$ $(\ell=1,\ldots,m-1)$, which are the empirical, cumulative proportions of individuals observed at the first $m-1$ levels.
+The modified statistics are then given by
+\begin{eqnarray}
+    \label{eq:wdm}
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
+                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
+    \label{eq:maxlmo}
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
+\end{eqnarray}
+where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$ $(\ell=1,\ldots,m-1)$.
+Critical values associated with the $\mathit{WDM}_o$ statistic can be obtained directly from a multivariate normal distribution \cite<see>{hotzei08}, while critical values associated with $\max \mathit{LM}_o$ can be obtained via simulation.
+This simulation is somewhat computationally intensive and, in practice, takes about 10 minutes on the authors' computers.
+However, the wait appears to be worth it, as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic to be considerably better than the $\mathit{WDM}_o$ statistic.
+
 \SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
 \section{Tutorial}
-In general, a set of scales is defined to be measurement invariant with respect
-to an auxiliary variable $V$ if:
 
-\begin{equation}
-\label{equ:invariance}
-f(x_{i}|t_{i},v_{i},...)=f(x_{i}|t_{i},...),
-\end{equation}
+In this section, we demonstrate how the above tests can be carried out in~R, using the package lavaan \cite{lavaan11} for model estimation and strucchange \cite{ZeiLei02,Zei06} for testing.
+We use data from \citeA{FroFan11} concerning the applicability of adult gratitude scales to youth, available in the~R package psychotools CITE.
+The data consist of responses to three adult gratitude scales from $n=1401$ youth aged 10--19 years.
+The original authors were specifically interested in whether the scales were measurement invariant across age.
+Because the sample size at each age was unbalanced, the authors created age groups of approximately equal sample size.
+In the examples below, we test for measurement invariance across these age groups.
 
-where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
-for variable $i$ that the scales purport to measure, and $f$ is the model's 
-distributional form \cite{mel89}. If the cut point of $v$ is known in advance, i.e. 
-gender, nested multiple group models \cite{bollen98} coupled with likelihood ratio test 
-are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
-of measurement invariance is
+We focus on measurement invariance of the factor loadings associated with the GQ-6 scale, using a one-factor model.
+While the ``age group'' variable against which we are testing measurement invariance is best considered ordinal, for demonstration we consider its treatment as categorical, continuous, and ordinal.
+Each of these treatments is described below in a separate section.
 
-\begin{equation}
-\label{equ:null}
-H_{0}: \theta_{A}=\theta_{B},
-\end{equation}
+%% THE FOLLOWING DETAILS ARE REMOVED FOR NOW
+%% In general, a set of scales is defined to be measurement invariant with respect
+%% to an auxiliary variable $V$ if:
 
-The alternative hypothesis is
+%% \begin{equation}
+%% \label{equ:invariance}
+%% f(x_{i}|t_{i},v_{i},...)=f(x_{i}|t_{i},...),
+%% \end{equation}
 
-\begin{equation}
-\label{equ:alternative}
-H_{1}:\theta_{i}=
-\begin{cases}
-\theta^{(A)} & v_{i}\leq v\\
-\theta^{(B)} & v_{i}\geq v
-\end{cases}
-\end{equation}
+%% where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
+%% for variable $i$ that the scales purport to measure, and $f$ is the model's 
+%% distributional form \cite{mel89}. If the cut point of $v$ is known in advance, i.e. 
+%% gender, nested multiple group models \cite{bollen98} coupled with likelihood ratio test 
+%% are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
+%% of measurement invariance is
 
-The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ 
-(group A and group B) can be obtained by Maximum Likelihood(ML) 
-from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,
-respectively. The LR test statistic for the given threshold $v$ is then
+%% \begin{equation}
+%% \label{equ:null}
+%% H_{0}: \theta_{A}=\theta_{B},
+%% \end{equation}
 
-\begin{equation}
-\label{equ:LRT}
-LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
-{(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
-}],
-\end{equation}
+%% The alternative hypothesis is
 
-,which has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
-to the number difference of parameters in $\theta$.
+%% \begin{equation}
+%% \label{equ:alternative}
+%% H_{1}:\theta_{i}=
+%% \begin{cases}
+%% \theta^{(A)} & v_{i}\leq v\\
+%% \theta^{(B)} & v_{i}\geq v
+%% \end{cases}
+%% \end{equation}
 
-To be more specific, in the following section, we will illustrate with an example to demonstrate how to implement multi group analysis in lavaan package. For more details, see \cite{vande12}
+%% The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ 
+%% (group A and group B) can be obtained by Maximum Likelihood(ML) 
+%% from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,
+%% respectively. The LR test statistic for the given threshold $v$ is then
 
-\subsection{Multiple Group Analysis}
-Froh et al\citeA{FroFan11} had a large sample of youth(n=1401, ranging from late childhood(10 years old) 
-to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, 
-including Gratitude Questionnaire 6(GQ-6; \citeNP{MccEmm02}), Gratitude Adjective Checklist(GAC; \citeNP{MccEmm02})  and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); \citeNP{ThoWat03}).
-In this specific analysis, the authors were interested in whether the youth factor
-structure for GQ-6 is a tau equivalent model (factor loadings restricted to be equal across
-levels of age)or a congeneric model(all parameters free for each level of age).
+%% \begin{equation}
+%% \label{equ:LRT}
+%% LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
+%% {(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
+%% }],
+%% \end{equation}
 
-To conduct the analysis, first of all, we need to read in data to R and clean the data.
-<<read data>>=
-# Reading in the data
+%% ,which has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
+%% to the number difference of parameters in $\theta$.
+
+\subsection{Categorical Treatment}
+
+\fixme{Need to update the citation below so that the last name is ``van de Schoot''.}
+Measurement invariance is most often tested using multiple groups models \cite<see, e.g., >{vande12}.
+This amounts to assuming that our auxiliary variable is categorical (i.e., unordered), which is not true for the age groups in the data.
+However, we demonstrate the procedure for completeness.
+
+To conduct the analysis, we first load the data and keep only complete cases for simplicity (though the tests can be applied to incomplete data).
+<<data>>=
+## Reading in the data
 data("YouthGratitude",package="psychotools")
-# remove cases with 'imputed' values(not in 1,...,9)
-yg<-YouthGratitude[apply(YouthGratitude[,4:28],1,function(x)all(x %in% 1:9)),]
+
+## remove cases with 'imputed' values(not in 1,...,9)
+compcases <- apply(YouthGratitude[,4:28], 1, function(x) all(x %in% 1:9))
+yg <- YouthGratitude[compcases,]
 @
-Second, we want to test which parameter(s) violate measurement invariance. 
-In lavaan package, loading, manifest variables' intercept and factor means 
-could be tested together or specifically. For more details, 
-please see \cite{vande12}. Since in most cases researchers are 
-interested in the factor loading invariance, we will use that as the 
-example in this tutorial.
-To implement LRT as ~\ref{equ:LRT}, we need to get estimation 
-for $\hat{\theta}$, without differing groups.
+Next, we fit two models in lavaan: a one-factor model where loadings are restricted to be equal across age groups, and a one-factor model where loadings are free across age groups.
+<<modest>>=
+## Restricted model
+reduced <- cfa('f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+             data=yg, group="agegroup", meanstructure=TRUE,
+             group.equal="loadings")
 
-<<theta estimate>>=
-# estimate theta 
-theta<-cfa(
-  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
-  data=yg,group="agegroup",meanstructure=TRUE,
-  group.equal="loadings")
+## Unrestricted model
+full <- cfa('f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+             data=yg, group="agegroup", meanstructure=TRUE)
 @
-Then we can get estimate for group A and group B. 
-<<theta A and theta B estimate>>=
-# estimate theta A,B 
-thetaAB<-cfa(
-  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
-  data=yg,group="agegroup",meanstructure=TRUE)
+Finally, we can get the results of a likelihood ratio test via the \verb+anova()+ function, which implies that the GQ-6 violates measurement invariance.
+<<LRT>>=
+## LRT
+anova(full, reduced)
 @
-In the end, we can get the LRT results by conducting 
-log likelihood test, which is done by anova function in R. 
-<<LRT>>=
-# LRT
-anova(theta, thetaAB)
-@
-Likelihood test is sensitive to large observations (CITE), it is not surprising to 
-have a significant result, although Froh et al\citeA{FroFan11} showed that tau equivalent
-model has as good alternative fit indices as congeneric model.
+Because our sample size is large, this likelihood ratio test is sensitive to small measurement invariance violations \cite{Ben80}.
+Thus, the above test may be significant solely due to the large sample size.
+Additionally, the above test is sensitive to small measurement invariance violations that are not likely to be of interest to researchers.
+For example, imagine that the 15-year-olds' parameters are slightly different than the other age groups.
+The 15-year-olds are in the middle of the age groups, and there is not likely to be any theoretical justification for 15-year-olds differing from every other age group.
+We need test statistics that are sensitive to measurement invariance violations that are monotonic with age.
+The score-based statistics described previously possess this property.
 
-\subsection{Extensions for Unknown Group}
-The multiple group test performs typically well when the $V$ 
-is a relatively-small number of categories and the sample size 
-is not too large. However, when $V$ (in this case, age) continuous variable, however, 
-multiple-group analysis usually cannot be performed because there are 
-no existing groups. Instead, we can follow the procedure described in
-the Background section.  That is, we can fit a model whose parameters 
+%% TODO Illustrate catL2BB statistic?
+
+\subsection{Continuous Treatment}
+If we are interested in measurement invariance violations that are monotonic with the age groups, it is perhaps simplest to treat the age groups as continuous.
+In doing so, we can use the statistics from Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.
+That is, we can fit a model whose parameters
 are restricted to be equal across across all individuals and then examine 
 how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
-values of $V$.
+age (where age ties are broken arbitrarily).
+This is demonstrated below, with similar code being useful when one is testing measurement invariance w.r.t.\ other continuous variables.
 
-%% TO BE MOVED TO BACKGROUND
-%% In practice, there are three test statistics can be used: 
-%% double maximum, CvM and maxLM. 
-%% \begin{equation}
-%% DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
-%% \end{equation}
-%% \begin{equation}
-%% CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
-%% \end{equation}
-%% \begin{equation}
-%% maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
-%% \end{equation}
+%% The multiple group test performs typically well when the $V$ 
+%% is a relatively-small number of categories and the sample size 
+%% is not too large. However, when $V$ (in this case, age) continuous variable, however, 
+%% multiple-group analysis usually cannot be performed because there are 
+%% no existing groups. Instead, we can follow the procedure described in
+%% the Background section.
 
-To demonstrate these tests in R, we will use the youth gratitude data
-and assume that age group is continuous.
-Like before, after read in data, we fit the model in lavaan package. 
-<<theta estimate>>=
-theta<-cfa(
-  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
-  data=yg,group="agegroup",meanstructure=TRUE,
-  group.equal="loadings")
-@
-Then we will use the gefp function in strucchange package to 
-get the $B(\hat{\theta}_{ij})$. After that, we can get the 
-three test statistics' significance by using sctest function.
+To proceed, we focus only on the restricted model that was fit in the previous section (the object \verb+reduced+).
+We obtain scores associated with the factor loadings from this model, then test the fluctuation of these scores across \verb+agegroup+.
+As demonstrated below, the score computation and testing can be efficiently carried out using \verb+sctest.lavaan()+.
+%% Ting: I removed this because the model was already fit in the
+%%       previous section
+%%<<theta estimate>>=
+%%theta<-cfa(
+%%  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+%%  data=yg,group="agegroup",meanstructure=TRUE,
+%%  group.equal="loadings")
+%%@
 
 <<Get Estimate>>=
-gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
-       vcov=info_full,sandwich=FALSE,parm=1:4)
-pval=rep(NA,3)
-pval[1]<-sctest(gefp1,functional=maxBB)$p.value
-pval[2]<-sctest(gefp1,functional=meanL2BB)$p.value
-pval[3]<-sctest(gefp1,functional=supLM(0.1))$p.value
-pval
+dm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM")
+cvm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM")
+maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM")
+
+c(dm$p.value, cvm$p.value, maxlm$p.value)
 @
-In applying the proposed continuous test, the $p$ value for 
-CvM and maxBB $p$ value is larger than likelihood test. 
-Besides, CvM $p$ value is not significant.
+In the code above, the \verb+parm=1:4+ argument contains parameter numbers reflecting the factor loadings that we wish to test.
+The ordering of parameters can be obtained via the command \verb+coef(reduced)+.
+Two of the three $p$-values output at the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).
 
-\SweaveOpts{engine = R, eps = TRUE, echo = TRUE, pdf=TRUE}
-Alternatively, we could use sctest.lavaan function to get the statistics value,
-corresponding $p$ value and the plot ordered by age level.
+%% Aug 21: Ed stopped here.  TODO: graphs
 
@@
 <<conti1res, fig=TRUE, height=7, width=7>>=
-DM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
+DM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
 DM
@@
 <<conti2res, fig=TRUE, height=7, width=7>>=
-CvM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
+CvM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
 CvM
@@
 <<conti3resmaxLM>>=
-maxLM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
+maxLM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
 maxLM
@@
 and above, except that we focus on "bins" of individuals 
-at each level of the ordinal variable. That is, instead 
-of aggregating over all $i=1,...n$ individuals, we 
-first compute cumulative proportions $t_{l}(l=1,...,m-1)$ 
-associated with the first $m-1$ levels of $V$. 
-%% TODO Move these details to Background section
-We then aggregate the cumulative scores only over 
-$i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics 
-related can be written as 
-\begin{eqnarray}
-    \label{eq:wdm}
-      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ 
-      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
-      \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
-    \label{eq:maxlmo}
-      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} 
-      ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
-      \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
-\end{eqnarray}
+at each level of the ordinal variable.
 
@@
 <<Ordinal Test>>=
-gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
+gefp1<-gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
        vcov=info_full,sandwich=FALSE,parm=1:4)
@@
 <<ord1res, fig=TRUE, height=7, width=7>>=
-WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
+WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
 WDMo 
@@
 <<ord2res, fig=TRUE, height=7, width=7>>=
-maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
+maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
                fnl=opval1, fplot=TRUE)

### 2013-08-23 frontiers/frontiers.Rnw
Frontiers ms, additions to tutorial section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@241 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-c(dm$p.value, cvm$p.value, maxlm$p.value)
+round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
 @
-In the code above, the \verb+parm=1:4+ argument contains parameter numbers reflecting the factor loadings that we wish to test.
+In the code above, the \verb+parm=1:4+ argument means that we wish to test model parameter numbers 1 through 4 (which are the factor loadings), which are supplied by lavaan.
 The ordering of parameters can be obtained via the command \verb+coef(reduced)+.
-Two of the three $p$-values output at the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).
+We see that two of the three $p$-values output at the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).
 
-%% Aug 21: Ed stopped here.  TODO: graphs
+%% TODO Illustrate with graphs?
 
-\begin{figure}
-\caption{Get Continuous Estimate DM Alternative Way}
-\label{fig:conti1res}
-\setkeys{Gin}{width=\textwidth}
-<<conti1res, fig=TRUE, height=7, width=7>>=
-DM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
-DM
-@
-\end{figure}
+%% THE FOLLOWING HAS BEEN MOVED
+%% \begin{figure}
+%% \caption{Get Continuous Estimate DM Alternative Way}
+%% \label{fig:conti1res}
+%% \setkeys{Gin}{width=\textwidth}
+%% <<conti1res, fig=TRUE, height=7, width=7>>=
+%% DM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
+%% DM
+%% @
+%% \end{figure}
 
-\begin{figure}
-\caption{Get Continuous Estimate CvM Alternative Way}
-\label{fig:conti2res}
-\setkeys{Gin}{width=\textwidth}
-<<conti2res, fig=TRUE, height=7, width=7>>=
-CvM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
-CvM
-@
-\end{figure}
+%% \begin{figure}
+%% \caption{Get Continuous Estimate CvM Alternative Way}
+%% \label{fig:conti2res}
+%% \setkeys{Gin}{width=\textwidth}
+%% <<conti2res, fig=TRUE, height=7, width=7>>=
+%% CvM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
+%% CvM
+%% @
+%% \end{figure}
 
-<<conti3resmaxLM>>=
-maxLM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
-maxLM
-@
-Since there are more than one observations located in each age level. 
-So the observation order within each age level is not unique, thus, 
-maxLM can't produce the plot. Actually, all three continuous statistics 
-were calculated on the arbitrary order that particular dataset entered 
-in each age level. In order to accurately account for the multiple observations 
-at the same age level, it is more appropriate to use the proposed ordinal test.
+%% <<conti3resmaxLM>>=
+%% maxLM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
+%% maxLM
+%% @
+The tests carried out here assume a unique ordering of individuals by age, but this is obviously not the case.
+To compute the statistics and $p$-values, the strucchange package implicitly assumed an arbitrary ordering of individuals who are tied on age.
+If we were to change this ordering, the resulting statistics and $p$-values would also change, which is obviously problematic.
+To accurately account for the multiple observations
+at the same age level, we must used the ordinal tests from Equations~\eqref{eq:wdm} and~\eqref{eq:maxlmo}.  These are described next.
 
-\subsection{Ordinal Test}
-As a matter of fact, there is only a partial ordering of individuals 
-with respect to $V$,i.e. observations with the same level 
-of $V$ have no unique ordering. The ordinal statistics 
-proposed here are similar to those described in Equation 
-and above, except that we focus on "bins" of individuals 
-at each level of the ordinal variable.
+\subsection{Ordinal Treatment}
 
-The difference between the continuous test and ordinal 
-test is that we don't force arbitrary order on the observations 
-within the same age groups. Therefore, this ordinal test should 
-reflect more real status of the data. In terms of code, we still need the gefp 
-function to get the B estimate. After that, we can use different 
-functions to aggregate and scale B estimation. 
-The function are called ordwmax and ol2bb1.fun, respectively. 
-<<Ordinal Test>>=
-gefp1<-gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
-       vcov=info_full,sandwich=FALSE,parm=1:4)
-pval.ord=rep(NA,2)
-pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value
+%% As a matter of fact, there is only a partial ordering of individuals 
+%% with respect to $V$,i.e. observations with the same level 
+%% of $V$ have no unique ordering. The ordinal statistics 
+%% proposed here are similar to those described in Equation 
+%% and above, except that we focus on "bins" of individuals 
+%% at each level of the ordinal variable.
 
+The difference between the ordinal test statistics and their
+continuous counterparts is that the ordinal statistics do not require
+an ordering of individuals  
+within the same age group.  To compute the test statistics, we allow
+the scores of all tied individuals to enter the cumulative sum
+(Equation~\eqref{eq:cumscore}) simultaneously.  This results in
+modified critical values and test statistics that are sensitive to
+measurement invariance violations that are monotonic with age group.
+%% Therefore, this ordinal test should 
+%% reflect more real status of the data. In terms of code, we still need the gefp 
+%% function to get the B estimate. After that, we can use different 
+%% functions to aggregate and scale B estimation. 
+%% The function are called ordwmax and ol2bb1.fun, respectively.
+
+To carry out the tests, we can rely on the same function that we used
+for the continuous test statistics.  As mentioned previously,
+calculation of the ``maxLMo'' statistic (Equation~\eqref{eq:maxlmo})
+can be lengthy from the need to simulate critical values.
+<<critvals,echo=FALSE>>=
 if(file.exists("opval1.rda")){
@@
 }
-pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
-pval.ord
+## Get test statistics using fnl=opval1 without showing it in the paper
+wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
+maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo",
+                 fnl=opval1)
 @
 
-Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
+<<ordtest,eval=FALSE>>=
+wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
+maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")
+@ 
+<<pvals>>=
+round(c(wdmo$p.value, maxlmo$p.value), 2)
+@
+
+%% TODO Describe graphs
+
+%% gefp1<-gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
+%%        vcov=info_full,sandwich=FALSE,parm=1:4)
+%% pval.ord=rep(NA,2)
+%% pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value
+
+%% pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
+%% pval.ord
+%% @
+
+%Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
 \begin{figure}
@@
 \setkeys{Gin}{width=\textwidth}
-<<ord1res, fig=TRUE, height=7, width=7>>=
+<<ord1res, fig=TRUE, height=7, width=7, echo=FALSE>>=
 WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
-WDMo 
+#WDMo 
 @
@@
 \setkeys{Gin}{width=\textwidth}
-<<ord2res, fig=TRUE, height=7, width=7>>=
+<<ord2res, fig=TRUE, height=7, width=7, echo=FALSE>>=
 maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
                fnl=opval1, fplot=TRUE)
-maxLMo 
+#maxLMo 
 @
@@
 
-In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. 
+In computing the ordinal test statistics, we
+obtain~$p=\Sexpr{round(wdmo$p.value,2)}$
+and~$p=\Sexpr{round(maxlmo$p.value,2)}$, respectively.
 Both $p-$values are clearly larger than that of the 
 likelihood ratio test and neither is significant at $\alpha=0.05$. 
-This result is consistent with alternative fit indices used in Froh et al paper\citeA{FroFan11}.
+This provides evidence that there is no measurement
+invariance violation that is monotonic with age group.  Instead, given
+the large sample size, the likelihood ratio test may be overly
+sensitive to anomalous, non-monotonic violations at one (or a few) age groups.
+
+In the above sections, we have illustrated the score-based tests'
+computation in R.  We suspect that the ordinal tests will be most
+popular with users, because measurement invariance tests are typically
+carried out across categories (ordered or not), as opposed to
+continuous variables.  Thus, in the sections below, we report
+novel simulations to study the ordinal statistics' expected behavior
+in practice.  In particular, we wish to study (i) the extent to which the
+ordinal statistics attribute measurement invariance violations to the
+correct parameter(s), and (ii) the extent to which the tests are
+robust to model misspecification.
 

### 2013-08-26 frontiers/simall-ordinal.R
have sim1 results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@242 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -27,7 +27,7 @@ dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
   half.level <- ifelse(nlevels %% 2 == 1, (nlevels %/% 2)+1, nlevels/2)
   
   # Define parameter vectors/matrices:
-  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5, 0.48,
+  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5, -0.48,
            29.32,24.70,14.84,10.59,19.30,18.01)
   asym<-c( 7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
             56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
@@ -104,11 +104,10 @@ dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
     }
     }
     
-
-    u <- t(rmvnorm(tmp.n,rep(0,6),tmp.psi))
-    z <- t(rmvnorm(tmp.n,rep(0,2),tmp.phi))
+    u.tmp <- t(rmvnorm(tmp.n,rep(0,6),tmp.psi))
+    z.tmp <- t(rmvnorm(tmp.n,rep(0,2),tmp.phi))
     for (j in 1:tmp.n){
-      datmat[tmp.ind[j],] <- mu + tmp.lambda%*%z[,j] + u[,j]
+      datmat[tmp.ind[j],] <- mu + tmp.lambda%*%z.tmp[,j] + u.tmp[,j]
     }
   }
   
@@ -153,7 +152,7 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, 
 }
 
 ## Loop over scenarios
-simulation <- function(diff = seq(0, 1.5, by = 0.25),parms="error",
+simulation <- function(diff = seq(0, 2, by = 0.25),parms=c("error","loading","var")
   nobs = c(120, 480, 960), nlevels = c(4, 8, 12), gradual = FALSE,
   anomaly = FALSE, verbose = TRUE, ...)
 {
@@ -244,4 +243,6 @@ source("../www/estfun-lavaan.R")
 source("../www/efpFunctional-cat.R")
 source("simall-ordinal.R")
[…]

### 2013-08-27 frontiers/frontiers.Rnw
Try a small simulation in simall-ordinal.R, but when parameter is loading or var, the plots seems wrong. I haven't figured out why.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@243 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 root of the information matrix, which serves to decorrelate the
-fluctuation processes associated with individual model parameters (i.e., the process associated with one parameter is not correlated with the processes associated with other parameters).
+fluctuation processes associated with individual model parameters 
+(i.e., the process associated with one parameter is not correlated 
+with the processes associated with other parameters).
 
@@
 Multiple test statistics are available, depending on how one
-summarizes the behavior of the cumulative sum of scores.  For example, one could
-take the absolute maximum that the cumulative sum attains for any
-parameter of interest, resulting in a {\em double max} statistic (the maximum is taken across parameters and individuals).
+summarizes the behavior of the cumulative sum of scores.  For example, 
+one could take the absolute maximum that the cumulative sum attains for any
+parameter of interest, resulting in a {\em double max} statistic 
+(the maximum is taken across parameters and individuals).
 Alternatively, one could sum the (squared) cumulative sum across parameters
-of interest and take the maximum or the average across individuals, resulting in a {\em maximum Lagrange multiplier} statistic and Cram\'{e}r-von Mises statistic, respectively \cite<see>[for further discussion]{MerZei13}.
-These statistics are given by
+of interest and take the maximum or the average across individuals, resulting in a 
+{\em maximum Lagrange multiplier} statistic and Cram\'{e}r-von Mises statistic, respectively 
+\cite<see>[for further discussion]{MerZei13}.  These statistics are given by
+
 \begin{eqnarray}
@@
 
-Critical values associated with $\mathit{DM}$ can be obtained analytically, while critical values associated with the other statistics can be obtained from direct simulation \cite{Zei06} or from
-more refined techniques \cite{han97}.  This issue should not be important to the user, as critical values are obtained directly from the R implementation described later.
+Critical values associated with $\mathit{DM}$ can be obtained analytically, 
+while critical values associated with the other statistics can be obtained 
+from direct simulation \cite{Zei06} or from more refined techniques \cite{han97}.  
+This issue should not be important to the user, as critical values are obtained 
+directly from the R implementation described later.
 
@@
 
-For an ordinal auxiliary variable with $m$ levels, these modifications are based on $t_{\ell}$ $(\ell=1,\ldots,m-1)$, which are the empirical, cumulative proportions of individuals observed at the first $m-1$ levels.
+For an ordinal auxiliary variable with $m$ levels, these modifications are 
+based on $t_{\ell}$ $(\ell=1,\ldots,m-1)$, which are the empirical, cumulative 
+proportions of individuals observed at the first $m-1$ levels.
 The modified statistics are then given by
@@
 where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$ $(\ell=1,\ldots,m-1)$.
-Critical values associated with the $\mathit{WDM}_o$ statistic can be obtained directly from a multivariate normal distribution \cite<see>{hotzei08}, while critical values associated with $\max \mathit{LM}_o$ can be obtained via simulation.
-This simulation is somewhat computationally intensive and, in practice, takes about 10 minutes on the authors' computers.
-However, the wait appears to be worth it, as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic to be considerably better than the $\mathit{WDM}_o$ statistic.
+Critical values associated with the $\mathit{WDM}_o$ statistic can be obtained 
+directly from a multivariate normal distribution \cite<see>{hotzei08}, while 
+critical values associated with $\max \mathit{LM}_o$ can be obtained via simulation.  
+This simulation is somewhat computationally intensive and, in practice, takes about 
+10 minutes on the authors' computers.  However, the wait appears to be worth it, 
+as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic 
+to be considerably better than the $\mathit{WDM}_o$ statistic.
 
@@
 
-In this section, we demonstrate how the above tests can be carried out in~R, using the package lavaan \cite{lavaan11} for model estimation and strucchange \cite{ZeiLei02,Zei06} for testing.
-We use data from \citeA{FroFan11} concerning the applicability of adult gratitude scales to youth, available in the~R package psychotools CITE.
-The data consist of responses to three adult gratitude scales from $n=1401$ youth aged 10--19 years.
-The original authors were specifically interested in whether the scales were measurement invariant across age.
-Because the sample size at each age was unbalanced, the authors created age groups of approximately equal sample size.
-In the examples below, we test for measurement invariance across these age groups.
+In this section, we demonstrate how the above tests can be carried out in~R, 
+using the package lavaan \cite{lavaan11} for model estimation and strucchange 
+\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
+the applicability of adult gratitude scales to youth, available in the~R 
+package psychotools CITE.  The data consist of responses to three adult 
+gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
+were specifically interested in whether the scales were measurement invariant 
+across age.  Because the sample size at each age was unbalanced, the authors 
+created age groups of approximately equal sample size.  In the examples below, 
+we test for measurement invariance across these age groups.
 
-We focus on measurement invariance of the factor loadings associated with the GQ-6 scale, using a one-factor model.
-While the ``age group'' variable against which we are testing measurement invariance is best considered ordinal, for demonstration we consider its treatment as categorical, continuous, and ordinal.
+We focus on measurement invariance of the factor loadings associated with t
+he GQ-6 scale, using a one-factor model.While the ``age group'' variable against 
+which we are testing measurement invariance is best considered ordinal, 
+for demonstration we consider its treatment as categorical, continuous, and ordinal.  
 Each of these treatments is described below in a separate section.
@@
 Measurement invariance is most often tested using multiple groups models \cite<see, e.g., >{vande12}.  
-This amounts to assuming that our auxiliary variable is categorical (i.e., unordered), which is not true for the age groups in the data.
-However, we demonstrate the procedure for completeness.
+This amounts to assuming that our auxiliary variable is categorical (i.e., unordered), 
+which is not true for the age groups in the data.  However, we demonstrate the procedure for completeness.
 
-To conduct the analysis, we first load the data and keep only complete cases for simplicity (though the tests can be applied to incomplete data).
+To conduct the analysis, we first load the data and keep only complete cases for simplicity 
+(though the tests can be applied to incomplete data).
 <<data>>=
@@
 @
-Next, we fit two models in lavaan: a one-factor model where loadings are restricted to be equal across age groups, and a one-factor model where loadings are free across age groups.
+Next, we fit two models in lavaan: a one-factor model where loadings are restricted 
+to be equal across age groups, and a one-factor model where loadings are free across age groups.
 <<modest>>=
@@
 @
-Finally, we can get the results of a likelihood ratio test via the \verb+anova()+ function, which implies that the GQ-6 violates measurement invariance.
+Finally, we can get the results of a likelihood ratio test via the \verb+anova()+ 
+function, which implies that the GQ-6 violates measurement invariance.
 <<LRT>>=
@@
 @
-Because our sample size is large, this likelihood ratio test is sensitive to small measurement invariance violations \cite{Ben80}.
-Thus, the above test may be significant solely due to the large sample size.
-Additionally, the above test is sensitive to small measurement invariance violations that are not likely to be of interest to researchers.
-For example, imagine that the 15-year-olds' parameters are slightly different than the other age groups.
-The 15-year-olds are in the middle of the age groups, and there is not likely to be any theoretical justification for 15-year-olds differing from every other age group.
-We need test statistics that are sensitive to measurement invariance violations that are monotonic with age.
-The score-based statistics described previously possess this property.
+Because our sample size is large, this likelihood ratio test is sensitive to 
+small measurement invariance violations \cite{Ben80}.  Thus, the above test may 
+be significant solely due to the large sample size.  Additionally, the above test 
+is sensitive to small measurement invariance violations that are not likely to be 
+of interest to researchers.  For example, imagine that the 15-year-olds' 
+parameters are slightly different than the other age groups.  
+The 15-year-olds are in the middle of the age groups, and there 
+is not likely to be any theoretical justification for 15-year-olds 
+differing from every other age group.  We need test statistics that 
+are sensitive to measurement invariance violations that are monotonic 
+with age.  The score-based statistics described previously possess this property.
 
@@
 \subsection{Continuous Treatment}
-If we are interested in measurement invariance violations that are monotonic with the age groups, it is perhaps simplest to treat the age groups as continuous.
-In doing so, we can use the statistics from Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.
+If we are interested in measurement invariance violations that are 
+monotonic with the age groups, it is perhaps simplest to treat the 
+age groups as continuous.  In doing so, we can use the statistics 
+from Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
 That is, we can fit a model whose parameters 
@@
 age (where age ties are broken arbitrarily).
-This is demonstrated below, with similar code being useful when one is testing measurement invariance w.r.t.\ other continuous variables.
+This is demonstrated below, with similar code being useful when one is 
+testing measurement invariance w.r.t.\ other continuous variables.
 
@@
 
-To proceed, we focus only on the restricted model that was fit in the previous section (the object \verb+reduced+).
-We obtain scores associated with the factor loadings from this model, then test the fluctuation of these scores across \verb+agegroup+.
+To proceed, we focus only on the restricted model that was fit in 
+the previous section (the object \verb+reduced+).  We obtain scores 
+associated with the factor loadings from this model, then test the 
+fluctuation of these scores across \verb+agegroup+.  
 As demonstrated below, the score computation and testing can be efficiently carried out using \verb+sctest.lavaan()+.
@@
 @
-In the code above, the \verb+parm=1:4+ argument means that we wish to test model parameter numbers 1 through 4 (which are the factor loadings), which are supplied by lavaan.
-The ordering of parameters can be obtained via the command \verb+coef(reduced)+.
-We see that two of the three $p$-values output at the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).
+In the code above, the \verb+parm=1:4+ argument means that 
+we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
+which are supplied by lavaan.  The ordering of parameters can be obtained via the 
+command \verb+coef(reduced)+.  We see that two of the three $p$-values output at 
+the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).
 
@@
 %% @
-The tests carried out here assume a unique ordering of individuals by age, but this is obviously not the case.
-To compute the statistics and $p$-values, the strucchange package implicitly assumed an arbitrary ordering of individuals who are tied on age.
-If we were to change this ordering, the resulting statistics and $p$-values would also change, which is obviously problematic.
+The tests carried out here assume a unique ordering of individuals by age, but 
+this is obviously not the case.  To compute the statistics and $p$-values, 
+the strucchange package implicitly assumed an arbitrary ordering of 
+individuals who are tied on age.  If we were to change this ordering, 
+the resulting statistics and $p$-values would also change, which is obviously problematic.  
 To accurately account for the multiple observations  
-at the same age level, we must used the ordinal tests from Equations~\eqref{eq:wdm} and~\eqref{eq:maxlmo}.  These are described next.
+at the same age level, we must used the ordinal tests from 
+Equations~\eqref{eq:wdm} and~\eqref{eq:maxlmo}.  These are described next.
 
@@
 can be lengthy from the need to simulate critical values.
+
 <<critvals,echo=FALSE>>=
@@
 
-%% gefp1<-gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
+%% gefp1<gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
 %%        vcov=info_full,sandwich=FALSE,parm=1:4)
@@
 
-%% pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
+%%pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
 %% pval.ord
@@
 \section{Simulation 1}
-
-In Simulation 1, we examined the extent
 

### 2013-08-27 frontiers/simall-ordinal.R

@@ -70,9 +70,9 @@ dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
     tmp.lambda <- matrix(0,6,2)
     tmp.lambda[,1] <- lambda[,1] 
     tmp.lambda[,2] <- lambda[,2]
-    tmp.psi <- psi
     tmp.phi <- phi
-
+    tmp.psi <- psi
+    
     mult <- ses
     if (anomaly) mult <- ifelse(i==(half.level+1), ses, 0)
 
@@ -81,23 +81,29 @@ dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
     if (gradual){
       if (anomaly) stop("gradual=TRUE and anomaly=TRUE do not work together.")
       if (parms=="error"){
-      diag(tmp.psi) <- diag(psi) + (each.vary*asym[7:12])/sqrt(tmp.n)
+      tmp.psi <- psi
+      diag(tmp.psi) <- diag(tmp.psi) + (each.vary*asym[7:12])/sqrt(tmp.n)
     }
       if (parms=="var"){
-      diag(tmp.phi) <- diag(phi) + (each.vary*asym[13])/sqrt(tmp.n)
+      tmp.phi <- phi
+      diag(tmp.phi) <- diag(tmp.phi) + (each.vary*asym[13])/sqrt(tmp.n)
     }
       if (parms=="loading"){
+      tmp.lambda <- matrix(0,6,2)
       tmp.lambda[,1] <- lambda[,1] + (each.vary*asym[1:3])/sqrt(tmp.n)
       tmp.lambda[,2] <- lambda[,2] + (each.vary*asym[4:6])/sqrt(tmp.n)
     }
     } else {
       if (parms=="error"){
+      tmp.psi <- psi
       diag(tmp.psi) <- diag(tmp.psi) + (mult*asym[7:12])/sqrt(tmp.n)
     }
       if (parms=="var"){
+      tmp.phi <- phi
       diag(tmp.phi) <- diag(tmp.phi) + (mult*asym[13])/sqrt(tmp.n)
     }
       if (parms=="loading"){
+      tmp.lambda <- matrix(0,6,2)
       tmp.lambda[,1] <- lambda[,1] + (mult*asym[1:3])/sqrt(tmp.n)
[…]

### 2013-08-27 frontiers/simall-ordinal.R
fix the simulation problem. The problem is that I used the wrong matrix in loading and var.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@244 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -29,7 +29,7 @@ dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
   # Define parameter vectors/matrices:
   theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5, -0.48,
            29.32,24.70,14.84,10.59,19.30,18.01)
-  asym<-c( 7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
+  asym<-c(7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
             56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
            1.024326,7.139776,4.666005,8.151785,3.696972,5.242366,
             8.630417)
@@ -67,12 +67,11 @@ dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
   for (i in (half.level+1):nlevels){
     tmp.n <- sum(age==i)
     tmp.ind <- which(age==i)
-    tmp.lambda <- matrix(0,6,2)
-    tmp.lambda[,1] <- lambda[,1] 
-    tmp.lambda[,2] <- lambda[,2]
-    tmp.phi <- phi
-    tmp.psi <- psi
-    
+
+    if (parms=="loading"){parmsvector=c(1,rep(0,18))}
+    if (parms=="var"){parmsvector=c(rep(0,12),1,rep(0,6))}
+    if (parms=="error"){parmsvector=c(rep(0,6),1,rep(0,5),rep(0,7))}
+   
     mult <- ses
     if (anomaly) mult <- ifelse(i==(half.level+1), ses, 0)
 
@@ -80,34 +79,31 @@ dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
     ## observed SES by sqrt(n)
     if (gradual){
       if (anomaly) stop("gradual=TRUE and anomaly=TRUE do not work together.")
-      if (parms=="error"){
-      tmp.psi <- psi
-      diag(tmp.psi) <- diag(tmp.psi) + (each.vary*asym[7:12])/sqrt(tmp.n)
-    }
-      if (parms=="var"){
-      tmp.phi <- phi
[…]

### 2013-08-28 frontiers/simall-ordinal.R
update simall-ordinal. R file. The code is ready to run full simulation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@246 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,7 +1,7 @@
 
 ## Data-generating process
 set.seed=1090
-dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALSE, parms="loadings")
+dgp1 <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALSE, parms="loadings")
 {
   ## Generates data from a factor analysis model that violates
   ## measurement invariance.  Also generates an ordinal auxiliary
@@ -116,16 +116,133 @@ dgp <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
   cbind(as.data.frame(datmat), age)
 }
 
+dgp2 <-function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALSE,
+                parms="loadings")
+{
+  ## Generates data from a factor analysis model that violates
+  ## measurement invariance.  Also generates an ordinal auxiliary
+  ## variable related to the invariance (loosely called "age").
+
+  ## If gradual==TRUE, mild instability starts halfway up age
+  ## and gets larger.  If gradual==FALSE, the instability starts
+  ## halfway up age and stays constant.
+  ## If anomaly==TRUE, instability only occurs at level (1+nlevels/2).
+  ## Otherwise, instability continues up the ordinal variable.
+  stopifnot(require("mvtnorm"))
+  
+  # ses is SE increase in parameters at last ordinal category.
+  # (This is constant if gradual==FALSE but not if gradual==TRUE.)
+  ses <- diff
+  sampsize <- nobs
+  sampsize.per.nobs <- nobs %/% nlevels
+
+  ## Deal with cases where nobs/nlevels is not an integer
+  age <- rep(NA, sampsize)
+  age[1:(nlevels*sampsize.per.nobs)] <- rep(1:nlevels, each=sampsize.per.nobs)
+  if (any(is.na(age))) age[is.na(age)] <- sample((1:nlevels),sum(is.na(age)))
+  half.level <- ifelse(nlevels %% 2 == 1, (nlevels %/% 2)+1, nlevels/2)
+  
+  # Define parameter vectors/matrices:
[…]

### 2013-08-30 frontiers/frontiers.Rnw
set up the full simulation 1 and simulation 2 in frontiers.Rnw file.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@247 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ## auxiliary code
-source("simall.R")
+source("simall-ordinal.R")
 source("mz.R")
@@
 source("../pkg/sctest.lavaan.R")
+source("../www/mz-ordinal.R")
+source("../www/estfun-lavaan.R")
 
@@
 \section{Simulation 1}
-
- to which the proposed tests
+In Simulation 1, we examined the extent to which the proposed tests
 are sensitive to the specific model parameter that violates measurement
@@
 
-\begin{figure}
-\caption{Path diagram representing the factor analysis model 
-         used for simulations.}
-\label{fig:path}
-\includegraphics{PathDiagram}
-\end{figure}
-
 Power and Type I error were examined across three sample sizes
@@
 
+<<sim1>>=
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim1.rda")) {
+  load("sim1.rda")
+} else {
+  sim1 <- simulation(sim=c("sim1"), nobs=c(480),parms=c("loading","error","var"))
+  sim1$nlevels <- factor(sim1$nlevels)
+  levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
+  sim1$nobs <- factor(sim1$nobs)
+  levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
+  levels(sim1$test) <- c("ordmax","ordwmax","catdiff")
+  save(sim1, file="sim1.rda")
+}
+sim1$test <- factor(as.character(sim1$test),
+  levels = c("ordmax","ordwmax","catdiff"),
+  labels = c("ordmax","ordwmax","catdiff"))
+@
+
 \subsection{Results}
+
+\section{Simulation 2}
+In Simulation 2, we examine the extent to which the results of
+Simulation 1 are robust to model misspecification. Specifically, we
+generate data from the factor analysis model used in the previous
+section, but with adding an extra loading from the second factor to the first
+indicator (of the first factor).  The estimated model lacks this extra
+loading, however. The goal of this simulation is to examine the proposed statitics'
+power and spicificity under model misspecification.
+
+\subsection{Method}
+Measurement invariance violations could occur in
+each of the three parameters from Simulation 1 (factor loading,factor covariance,unique
+variance), and they could also occur in the extra,unmodeled loading. 
+Sample size and magnitude of measurementinvariance violation were manipulated 
+in the same way as they were in Simulation 1.The tested parameters are the same as Simulation
+1,too.
+
+\subsection{Results}
+
+<<sim2>>=
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim2.rda")) {
+  load("sim2.rda")
+} else {
+  sim2 <- simulation(sim=c("sim2"), nobs=c(480),parms=c("extra","extra+loading","extra+var","extra+error"))
+  sim2$nlevels <- factor(sim2$nlevels)
+  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
+  sim2$nobs <- factor(sim2$nobs)
+  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
+  levels(sim2$test) <- c("ordmax","ordwmax","catdiff")
+  save(sim2, file="sim2.rda")
+}
+sim2$test <- factor(as.character(sim2$test),
+  levels = c("ordmax","ordwmax","catdiff"),
+  labels = c("ordmax","ordwmax","catdiff"))
+@
 

### 2013-08-30 frontiers/simall-ordinal.R

@@ -30,9 +30,10 @@ dgp1 <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FAL
   theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5, -0.48,
            29.32,24.70,14.84,10.59,19.30,18.01)
   asym<-c(7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
-            56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
+           62.1,26.6,82.1,9.05,19.35,51.59,
            1.024326,7.139776,4.666005,8.151785,3.696972,5.242366,
             8.630417)
+    # 56.672724,24.255196,74.917170,8.264380,17.664706,47.093002,
   mu<-theta[14:19]
   # Loadings:
   lambda <- matrix(0,6,2)

### 2013-09-05 frontiers/frontiers.Rnw
go through the intro,tutorial and sim method sections, change some typos and write some small questions.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@248 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage[english]{babel}
-
-\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
 
@@
 to study (details here).  The studies indicate that.}
+
+\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
 
@@
 (Huber-White) standard errors. %TODO cite
+\seeme{I don't understand this sentence. cross-product of each individual for each parameter? Why this 
+is the standard error?}
 
@@
 
-We focus on measurement invariance of the factor loadings associated with t
-he GQ-6 scale, using a one-factor model.While the ``age group'' variable against 
+We focus on measurement invariance of the factor loadings associated with 
+the GQ-6 scale, using a one-factor model. While the ``age group'' variable against 
 which we are testing measurement invariance is best considered ordinal, 
@@
 \fixme{Need to update the citation below so that the last name is ``van de Schoot''.}
+\seeme{Is it a different paper from vande12?}
 Measurement invariance is most often tested using multiple groups models \cite<see, e.g., >{vande12}.  
@@
 That is, we can fit a model whose parameters 
-are restricted to be equal across across all individuals and then examine 
+are restricted to be equal across all individuals and then examine 
 how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
@@
 ($n=120,480,960$) , three numbers of categories ($m=4,8,12$)and
-7 magnitudes of invariance violations. 
-The meaurement invariance violations began at level $1+m/2$ of $V$
-and were consistant thereafter. These
+17 magnitudes of invariance violations. 
+The measurement invariance violations began at level $1+m/2$ of $V$
+and were consistent thereafter. These
 violations were based on the auxiliary variable $V$: individuals below
@@
 percentile  by $d$ times the parameters' asymptotic standard errors
-(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,1.5$. 
+(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
 For each combination of sample size ($n$) $\times$ violation
 magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
-generated and tested.Four statistics were examined above. In all conditions, we
+generated and tested. Four statistics were examined above. In all conditions, we
 maintained equal sample sizes at each level of the ordinal variable.
@@
 indicator (of the first factor).  The estimated model lacks this extra
-loading, however. The goal of this simulation is to examine the proposed statitics'
-power and spicificity under model misspecification.
+loading, however. The goal of this simulation is to examine the proposed statistics'
+power and specificity under model misspecification.
 
@@
 each of the three parameters from Simulation 1 (factor loading,factor covariance,unique
-variance), and they could also occur in the extra,unmodeled loading. 
+variance), and they could also occur in the extra, unmodeled loading. 
 Sample size and magnitude of measurementinvariance violation were manipulated 

### 2013-09-05 frontiers/frontiers.Rnw
o Minor changes to frontiers manuscript o Added sim1 results file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@249 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 } else {
-  sim1 <- simulation(sim=c("sim1"), nobs=c(480),parms=c("loading","error","var"))
+  sim1 <- simulation(sim="sim1", nobs=480, parms=c("loading","error","var"))
   sim1$nlevels <- factor(sim1$nlevels)
@@
 } else {
-  sim2 <- simulation(sim=c("sim2"), nobs=c(480),parms=c("extra","extra+loading","extra+var","extra+error"))
+  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
   sim2$nlevels <- factor(sim2$nlevels)

### 2013-09-08 frontiers/frontiers.Rnw
add sim1 plots in the file and edit results section roughly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@250 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass[doc]{apa}
-\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
+\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm,float}
 \usepackage[english]{babel}
@@
 %Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
-\begin{figure}
+\begin{figure}[H]
 \caption{Get Ordinal Estimate WDMo Alternative Way}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \caption{Get Ordinal Estimate maxLMo Alternative Way}
@@
 
+\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
 \section{Simulation}
@@
 
+\subsection{Results}
+Full simulation results are presented in Figure~\ref{fig:sim11res} to
+Figure~\ref{fig:sim13res}; all three test statistics exhibited 
+similar results, with the unordered $\max\mathit{LM}$ 
+Figure~\ref{fig:sim11res} displays power curves as a function of 
+violation magnitude for the first
+factor loading $\lambda_{11}$, with panels reflecting the parameters 
+being tested and lines reflecting $n$.
+Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
+similar power curves when the factor covariance 
+$\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
+invariance, respectively.
+
+From these figures, one generally observe that the tests isolate 
+the parameter violating measurement invariance.  Additionally, the tests 
+have somewhat-higher power to detect measurement invariance violations 
+in the factor loading and factor covariance parameters, as opposed to the error variance
+parameter.  Finally, simultaneous tests of all factor loadings or all error
+parameters result in decreased power, as compared to the situation
+where one tests only the violating parameter.  This occurs because, in
+testing a subset of parameters (only one of which violates measurement
+invariance), we are effectively dampening the signal of a measurement
+invariance violation.
+
+In summary, we found that the proposed tests can attribute
+measurement invariance violations to the correct parameter.  This
+provides evidence that, in practice, one can have confidence in the
+tests' abilities to locate the measurement invariance violation.  Of
+course, this statement is qualified by the fact that, in this
+simulation, the model was correctly specified.  In the following
+simulations, we examine the tests' performance when the model is
+misspecified.
+
 <<sim1>>=
@@
   levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
+
   sim1$nobs <- factor(sim1$nobs)
@@
   labels = c("ordmax","ordwmax","catdiff"))
+  levels(sim1$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors")
+
 @
 
-\subsection{Results}
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Loading1.}
+\label{fig:sim11res}
+\setkeys{Gin}{width=\textwidth}
+<<sim11res, fig=TRUE, height=7, width=7>>=
+sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("loading"),]
+sim1.tmp$test <- factor(sim1.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the ordered double-max test the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Covariance}
+\label{fig:sim12res}
+\setkeys{Gin}{width=\textwidth}
+<<sim12res, fig=TRUE, height=7, width=7>>=
+sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("var"),]
+sim1.tmp$test <- factor(sim1.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the ordered double-max test the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Error1}
+\label{fig:sim13res}
+\setkeys{Gin}{width=\textwidth}
+<<sim13res, fig=TRUE, height=7, width=7>>=
+sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("error"),]
+sim1.tmp$test <- factor(sim1.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
 
@@
 
-<<sim2>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1163)
+%<<sim2>>=
+%## seed for replication
+%RNGkind(kind = "default", normal.kind = "default")
+%set.seed(1163)
 
-if(file.exists("sim2.rda")) {
-  load("sim2.rda")
-} else {
-  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
-  sim2$nlevels <- factor(sim2$nlevels)
-  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
-  sim2$nobs <- factor(sim2$nobs)
-  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
-  levels(sim2$test) <- c("ordmax","ordwmax","catdiff")
-  save(sim2, file="sim2.rda")
-}
-sim2$test <- factor(as.character(sim2$test),
-  levels = c("ordmax","ordwmax","catdiff"),
-  labels = c("ordmax","ordwmax","catdiff"))
-@
+%if(file.exists("sim2.rda")) {
+%  load("sim2.rda")
+%} else {
+%  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
+%  sim2$nlevels <- factor(sim2$nlevels)
+%  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
+%  sim2$nobs <- factor(sim2$nobs)
+%  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
+%  levels(sim2$test) <- c("ordmax","ordwmax","catdiff")
+%  save(sim2, file="sim2.rda")
+%}
+%sim2$test <- factor(as.character(sim2$test),
+%  levels = c("ordmax","ordwmax","catdiff"),
+%  labels = c("ordmax","ordwmax","catdiff"))
+%@
 

### 2013-09-09 frontiers/frontiers.Rnw
use mz-ordinal-proptest.R in simall-ordinal file. Update sim1 results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@251 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Figure~\ref{fig:sim13res}; all three test statistics exhibited 
-similar results, with the unordered $\max\mathit{LM}$ 
+similar results, with unordered  $\max\mathit{LM}$ demonstrated lower powers across all situations.
+ordered $\max\mathit{LM}$ and double-max exhibited equivalent power curves since there is only one parameter violating measurement invariance.
+Thus, different ways of aggregating $\bm B(\hat{\theta})$ would be the same.
 Figure~\ref{fig:sim11res} displays power curves as a function of 

### 2013-09-09 frontiers/simall-ordinal.R

@@ -359,7 +359,8 @@ if(FALSE){
 library("lavaan")
 library("strucchange")
 library("mvtnorm")
-source("../www/mz-ordinal.R")
+#source("../www/mz-ordinal.R")
+source("mz-ordinal-proptest.R")
 source("../www/estfun-lavaan.R")
 source("../www/efpFunctional-cat.R")
 source("simall-ordinal.R")
@@ -381,6 +382,24 @@ save(simtry,file ="simtry.rda")
 sim1 <- simulation(sim=c("sim1"), nobs=c(480),parms=c("loading","error","var"))
 sim2 <- simulation(sim=c("sim2"), nobs=c(480),
                    parms=c("extra","extra+loading","extra+var","extra+error"))
+
+#load data file and give names for different levels in each variable.
+if(file.exists("simtry.rda")) {
+  load("simtry.rda")
+} else {
+  simtry <- simulation()
+  simtry$nlevels <- factor(simtry$nlevels)
+  levels(simtry$nlevels) <- paste("m=",levels(simtry$nlevels),sep="")
+  simtry$nobs <- factor(simtry$nobs)
+  levels(simtry$nobs) <- paste("n=",levels(simtry$nobs),sep="")
+  levels(simtry$test) <- c("ordmax","ordwmax","catdiff")
+  save(simtry, file="simtry.rda")
+}
+simtry$test <- factor(as.character(simtry$test),
+  levels = c("ordmax","ordwmax","catdiff"),
+  labels = c("ordmax","ordwmax","catdiff"))
+   simtry$nlevels <- factor(simtry$nlevels)
+ levels(simtry$nlevels) <- paste("m=",levels(simtry$nlevels),sep="")
 
 library(lattice)
 #parm=loading
@@ -418,24 +437,6 @@ xyplot(power ~ diff | nlevels + pars + parms, group = ~ test, data = simtry.tmp,
 dev.off()
 
 
-
-#load data file and give names for different levels in each variable.
-if(file.exists("simtry.rda")) {
-  load("simtry.rda")
-} else {
-  simtry <- simulation()
-  simtry$nlevels <- factor(simtry$nlevels)
-  levels(simtry$nlevels) <- paste("m=",levels(simtry$nlevels),sep="")
-  simtry$nobs <- factor(simtry$nobs)
-  levels(simtry$nobs) <- paste("n=",levels(simtry$nobs),sep="")
[…]

### 2013-09-09 frontiers/mz-ordinal-proptest.R
add mz-ordinal-proptest.R, this file only included proposed test and statisics, so it saves a lot of time running simulation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@252 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,80 @@
+ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
+{
+ ## data checks
+  data <- as.data.frame(data)
+  stopifnot(c("x1", "x2", "x3", "y1", "y2", "y3", "age") %in% names(data))
+  age <- data$age
+
+  ## require package
+  stopifnot(require("lavaan"))
+
+  ## model from Merkle & Zeileis (for proposed tests)
+  rval <- 'verbal =~ x1 + x2 + x3; math =~ y1 + y2 + y3'
+
+  data <- data[, c("x1", "x2", "x3", "y1", "y2", "y3")]
+  
+  ## fit model
+  rval <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE))
+  
+  ## store (1) fitted model object, (2) lavaan coefficient
+  ## labeling and ordering
+  rval <- list(
+    data = data,
+    model = rval,
+    names = c("verbal=~x1", "verbal=~x2", "verbal=~x3", "math=~y1", "math=~y2", "math=~y3", "x1~~x1", "x2~~x2", "x3~~x3",
+        "y1~~y1", "y2~~y2", "y3~~y3", "verbal~~math", "x1~1", "x2~1", "x3~1", "y1~1", "y2~1", "y3~1")
+            )
+  
+  class(rval) <- "mzfit"
+
+  return(rval)
+}
+
+coef.mzfit <- function(object, ...) {
+    unclass(getMethod("coef", "lavaan")(object$model, ...))
+}
+
+logLik.mzfit <- function(object, ...) {
+  rval <- as.vector(fitMeasures(object$model, "logl"))
+
+  structure(rval, df = length(coef(object)),
+    nobs = nrow(object$data), class = "logLik")
+}
+
+vcov.mzfit <- function(object, ...) {
+    vc <- getMethod("vcov", "lavaan")(object$model, ...)
+    class(vc) <- "matrix"
+    vc
+}
+
+bread.mzfit <- function(x, ...) vcov(x) * nrow(x$data)
+
+info.mzfit <- function(x, ...) solve(vcov(x) * nrow(x$data))
+
+estfun.mzfit <- function(x, ...)
+{
+  estfun.lavaan(x$model)
+}
+
+get.zcov <- function(data){
+  age <- data$age
[…]

### 2013-09-09 frontiers/frontiers.Rnw
describe the plot for the ordinal tests. Write the outline of general discussion.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@253 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-%% TODO Describe graphs
-
 %% gefp1<gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
@@
 \begin{figure}[H]
-\caption{Get Ordinal Estimate WDMo Alternative Way}
-\label{fig:ord1res}
+\caption{Fluctuation processes for the WDM statistic(left panel) and the maxLM statistic(right panel)}
+\label{fig:ordres}
 \setkeys{Gin}{width=\textwidth}
-<<ord1res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+<<ordres, fig=TRUE, height=3, width=6, echo=FALSE>>=
+par(mfrow = c(1, 2), cex = 0.65)
 WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
 #WDMo 
-@
-\end{figure}
-
-\begin{figure}[H]
-\caption{Get Ordinal Estimate maxLMo Alternative Way}
-\label{fig:ord2res}
-\setkeys{Gin}{width=\textwidth}
-<<ord2res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
-               fnl=opval1, fplot=TRUE)
+maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fnl=opval1, fplot=TRUE)
 #maxLMo 
@@
 sensitive to anomalous, non-monotonic violations at one (or a few) age groups.
+
+%% TODO Describe graphs
+%% Is there some argument for fplot=TRUE so user can change  axis-name and critical value line
+Plots representing the statistics' fluctuations across levels of age group are displayed in Figure~\ref{fig:ordres}.
+The left panel displays the process associated with WDM.
+The right panel displays the process associated with maxLM.
+In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs.
+greater than $i$, and the null hypothesis is rejected if the maximum of the statistics is larger than its 5\% critical value(visualized by the horizontal red line).
+The final group is not displayed because it encompass all observations and always equals to zero.
+It is observed that both statistics generally increase with age.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-What do the results mean, future studies, etc.
+In this paper, we examined the recently proposed ordinal statistics specificity property and robustness to model specification to study measurement invariance.
+As demonstrated in the results, the proposed statistics could select the parameter violating measurement invariance even though the model is misspecified in general.
+However, when an unmodeled parameter is violating measurement invariance, the statistics will demonstrating these variances in the related parameters, for example, associated loading and error terms.
+
+These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context.
+It is very common to observe several individuals with the same age, IQ, income rank, etc.
+Therefore, the ordinal tests which dealing with multiple ties in auxiliary variable is more suitable than the continuous test.
+
+Moreover, even though researchers might be originally interested in comparing categorical differences.
+For example, researchers might want to test whether there is gender difference or cultural difference for the scale.
+Using traditional likelihood test or wald test, there might existing some differences regarding to the categorical variable.
+Then the next natural question would be why there is gender or culture differences?
+At this point, researchers might need to examine education level, social economic status, etc, which were often recorded in the ordinal format.
+Thus these test statistics offer an convenient assessing tool to further exploring research hypothesis without making complicated assumptions or fitting multiple different models.
+
+The core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity.
 

### 2013-09-10 frontiers/frontiers.Rnw
update citations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@254 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 across observations.  This procedure is similar in spirit to the
-calculation of modification indices CITE and to Lagrange multiplier
-tests CITE, and, in fact, those statistics can be viewed as  
+calculation of modification indices \cite{bentler90} and to Lagrange multiplier
+tests \cite{sat89}, and, in fact, those statistics can be viewed as  
 special cases of the family described here.
@@
 of these scores forms the ``meat'' for the calculation of robust
-(Huber-White) standard errors. %TODO cite
-\seeme{I don't understand this sentence. cross-product of each individual for each parameter? Why this 
-is the standard error?}
+(Huber-White) standard errors \cite{freed06}. \seeme{not sure about this citation}
 
@@
 the applicability of adult gratitude scales to youth, available in the~R 
-package psychotools CITE.  The data consist of responses to three adult 
+package psychotools \cite{zeilei11}.  The data consist of responses to three adult 
 gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 

### 2013-09-12 frontiers/frontiers.Rnw
include sim2 results section.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@256 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 variance), and they could also occur in the extra, unmodeled loading. 
-Sample size and magnitude of measurementinvariance violation were manipulated 
+Sample size and magnitude of measurement invariance violation were manipulated 
 in the same way as they were in Simulation 1.The tested parameters are the same as Simulation
@@
 \subsection{Results}
+Results are presented by the parameter that violated measurement invariance.
+Figure~\ref{fig:sim21res} displays results when the unmodeled loading violates 
+measurement invariance. One can generally observe that 
+the first loading and error variance parameter showed high measurement invariance 
+violation simultaneously, while factor covariance did not demonstrate violation.
 
-%<<sim2>>=
-%## seed for replication
-%RNGkind(kind = "default", normal.kind = "default")
-%set.seed(1163)
+Figure~\ref{fig:sim22res} displays results when the loading $\lambda_{11}$ violates
+measurement invariance. Even though lacking one loading associated with the first indicator, 
+statistic showed enough power to detect the first loading violation specifically. 
+Similar pattern of specificity and power were observed when the violation parameter 
+was $\sigma_{12}$ (Figure~\ref{fig:sim23res}) and $\epsilon_{11}$ (Figure~\ref{fig:sim24res}).
 
-%if(file.exists("sim2.rda")) {
-%  load("sim2.rda")
-%} else {
-%  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
-%  sim2$nlevels <- factor(sim2$nlevels)
-%  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
-%  sim2$nobs <- factor(sim2$nobs)
-%  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
-%  levels(sim2$test) <- c("ordmax","ordwmax","catdiff")
-%  save(sim2, file="sim2.rda")
-%}
-%sim2$test <- factor(as.character(sim2$test),
-%  levels = c("ordmax","ordwmax","catdiff"),
-%  labels = c("ordmax","ordwmax","catdiff"))
-%@
+In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
+as the umodeled parameter does not violate measurment invariance. If the unmodeled parameter 
+dose violate measurement invariance, however, then the tests assign this violation to other 
+parameters that do not violate measurement invariance. The impacted parameters include the error
+variance and other loadings associated with the manifest variables that had the unmodeled 
+loading. Thus, prior to testing measurement invariance, it is important to study the extent to
+which the hypothesized model includes all parameters of importance.
 
+<<sim2>>=
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim2.rda")) {
+  load("sim2.rda")
+} else {
+  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
+  sim2$nlevels <- factor(sim2$nlevels)
+  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
+  sim2$nobs <- factor(sim2$nobs)
+  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
+  levels(sim2$test) <- c("ordmax","ordwmax","catdiff")
+  save(sim2, file="sim2.rda")
+}
+sim2$test <- factor(as.character(sim2$test),
+  levels = c("ordmax","ordwmax","catdiff"),
+  labels = c("ordmax","ordwmax","catdiff"))
+  levels(sim2$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors") 
+@
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim21res}
+\setkeys{Gin}{width=\textwidth}
+<<sim21res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim22res}
+\setkeys{Gin}{width=\textwidth}
+<<sim22res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra+loading"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim23res}
+\setkeys{Gin}{width=\textwidth}
+<<sim23res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra+var"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim24res}
+\setkeys{Gin}{width=\textwidth}
+<<sim24res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra+error"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
 \section{General Discussion}

### 2013-09-12 pkg/sctest.lavaan.R
Small modification to sctest.lavaan for plot customization.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@257 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -28,7 +28,7 @@ sctest.lavaan <- function(x, order.by, parm, stat, fnl = NULL, fplot = FALSE, ..
   
   testres <- sctest(fp, functional = fnl)
 
-  if(fplot) plot(fp, functional = fnl)
+  if(fplot) plot(fp, functional = fnl, ...)
 
   testres
 }

### 2013-09-13 frontiers/frontiers.Rnw
update cat2bb, paragraph, change fplot option, change sim1 plots and move sim2 plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@258 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+%% TODO Illustrate catL2BB statistic
+
+If $V$ is only nominal/categorical,
+there is not even a partial ordering, i.e., measurement invariance tests should neither
+exploit the ordering of $V$'s levels nor of the observations within the level.
+In this situation, it is possible to obtain a test statistic by first summing scores within
+each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
+statistic \cite{HjoKon02}.  This test statistic can be formally written as
+\begin{equation}
+    \label{eq:lmuo}
+     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
+\end{equation}
+where the first subscript on the two terms in parentheses are $i_\ell$
+and $i_{\ell-1}$, respectively (and where we take $i_0 = 0$, so that
+the cumulative score is ${\bm B}(0, \hat {\bm \theta}) = \bm{0}$).
+Again, tests of subsets of model parameters can be obtained by 
+taking the inner sum over only the $k^* < k$ parameters of interest.
+This test statistic discards the ordinal nature of the auxiliary
+variable, essentially employing the alternative hypothesis
+from~\eqref{eq:h2*}.  A similar issue is observed in testing for
+measurement 
+invariance via multiple groups models and likelihood ratio tests (or,
+equivalently, via Wald tests or Lagrange multiplier tests): we
+can allow $\bm{\theta}$ to be unique at each level of the ordinal
+variable, but the ordinality of the auxiliary variable is lost.
+
 \SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
@@
 @
+
+Another similar statistics is unordered $\mathit{LM}_\mathit{uo}$ as shown in Equ. We can obtain this statistic by using the following function. 
+<<Categorical>>=
+## unordered LM
+catdiff <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="range")
+round(c(catdiff$p.value), 2)
+@
 Because our sample size is large, this likelihood ratio test is sensitive to 
@@
 with age.  The score-based statistics described previously possess this property.
-
-%% TODO Illustrate catL2BB statistic?
 
@@
 maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM")
-
 round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
@@
 maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")
+
 @ 
@@
 par(mfrow = c(1, 2), cex = 0.65)
-WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
+WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE,ylim=c(0,3.5),xlab="Age group",ylab = "Weighted max statistics")
+
 #WDMo 
-maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fnl=opval1, fplot=TRUE)
+maxLMo<-sctest(reduced,as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fnl=opval1, fplot=TRUE,ylim=c(0,14),xlab="Age group",ylab = "LM statistics")
 #maxLMo 
@@
 \end{figure}
+
+\seeme{can't change age group axis, since order.by argument needs agegroup as numeric}
 
@@
 The final group is not displayed because it encompass all observations and always equals to zero.
+The age group 1:5 in the x-axis corresponds to 10-11, 12-13, 14,15,16 respectively.
 It is observed that both statistics generally increase with age.
@@
 <<sim11res, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("loading"),]
+sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("loading")&sim1$pars %in% c("Loading1","All Loadings","All Errors"),]
 sim1.tmp$test <- factor(sim1.tmp$test)
@@
 <<sim12res, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("var"),]
+sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("var") &sim1$pars %in% c("All Loadings", "Covariance","All Errors") ,]
 sim1.tmp$test <- factor(sim1.tmp$test)
@@
 <<sim13res, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("error"),]
+sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("error") &sim1$pars %in% c("Error1","All Errors","All Loadings"),]
 sim1.tmp$test <- factor(sim1.tmp$test)
@@
   levels(sim2$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors") 
+  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
 @
@@
 <<sim21res, fig=TRUE, height=7, width=7>>=
-sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra"),]
+sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra")&sim2$pars %in% c("Loading1","Error1","Covariance"),]
 sim2.tmp$test <- factor(sim2.tmp$test)
@@
 
+\section{General Discussion}
+
+In this paper, we examined the recently proposed ordinal statistics specificity property and robustness to model specification to study measurement invariance.
+As demonstrated in the results, the proposed statistics could select the parameter violating measurement invariance even though the model is misspecified in general.
+However, when an unmodeled parameter is violating measurement invariance, the statistics will demonstrating these variances in the related parameters, for example, associated loading and error terms.
+
+These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context.
+It is very common to observe several individuals with the same age, IQ, income rank, etc.
+Therefore, the ordinal tests which dealing with multiple ties in auxiliary variable is more suitable than the continuous test.
+
+Moreover, even though researchers might be originally interested in comparing categorical differences.
+For example, researchers might want to test whether there is gender difference or cultural difference for the scale.
+Using traditional likelihood test or wald test, there might existing some differences regarding to the categorical variable.
+Then the next natural question would be why there is gender or culture differences?
+At this point, researchers might need to examine education level, social economic status, etc, which were often recorded in the ordinal format.
+Thus these test statistics offer an convenient assessing tool to further exploring research hypothesis without making complicated assumptions or fitting multiple different models.
+
+The core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity.
+
+\section{Appendix}
 \begin{figure}[H]
@@
 \end{figure}
-\section{General Discussion}
-
-In this paper, we examined the recently proposed ordinal statistics specificity property and robustness to model specification to study measurement invariance.
-As demonstrated in the results, the proposed statistics could select the parameter violating measurement invariance even though the model is misspecified in general.
-However, when an unmodeled parameter is violating measurement invariance, the statistics will demonstrating these variances in the related parameters, for example, associated loading and error terms.
-
-These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context.
-It is very common to observe several individuals with the same age, IQ, income rank, etc.
-Therefore, the ordinal tests which dealing with multiple ties in auxiliary variable is more suitable than the continuous test.
-
-Moreover, even though researchers might be originally interested in comparing categorical differences.
-For example, researchers might want to test whether there is gender difference or cultural difference for the scale.
-Using traditional likelihood test or wald test, there might existing some differences regarding to the categorical variable.
-Then the next natural question would be why there is gender or culture differences?
-At this point, researchers might need to examine education level, social economic status, etc, which were often recorded in the ordinal format.
-Thus these test statistics offer an convenient assessing tool to further exploring research hypothesis without making complicated assumptions or fitting multiple different models.
-
-The core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity.
 

### 2013-09-15 frontiers/frontiers.Rnw
organize cat2bb, organize path diagram, organize general discussion, organize sim2 results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@259 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2013-09-10 frontiers/frontiers.Rnw
@@
 
+%% TODO Illustrate catL2BB statistic
+
+If $V$ is only nominal/categorical,
+there is not even a partial ordering, i.e., measurement invariance tests should neither
+exploit the ordering of $V$'s levels nor of the observations within the level.
+In this situation, unordered $\mathit{LM}$ is proposed.  This test statistic can be formally written as
+\begin{equation}
+    \label{eq:lmuo}
+     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
+\end{equation}
+This test statistic discards the ordinal nature of the auxiliary
+variable, essentially similar to the nature of likelihood ratio tests (or,
+equivalently, via Wald tests or Lagrange multiplier tests): the ordinality of the auxiliary variable is
+ignored. By contrast, this statistics is easier to get compared to likelihood ratio test statistics since it doesn't involve specifying different sets of competitive models.
+
 \SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
@@
 
-\fixme{Need to update the citation below so that the last name is ``van de Schoot''.}
-\seeme{Is it a different paper from vande12?}
 Measurement invariance is most often tested using multiple groups models \cite<see, e.g., >{vande12}.  
@@
 @
+
+Another similar statistics is $\mathit{LM}_\mathit{uo}$ as shown in Equation~ \eqref{eq:lmuo}. We can obtain this statistic by using the following function. 
+<<LM_UO>>=
+## unordered LM
+lm_u0 <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="range")
+round(c(lm_u0$p.value), 2)
+@
+we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
+which are supplied by lavaan.  The ordering of parameters can be obtained via the 
+command \verb+coef(reduced)+.
+
 Because our sample size is large, this likelihood ratio test is sensitive to 
@@
 maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM")
-
 round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
 @
-In the code above, the \verb+parm=1:4+ argument means that 
-we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
-which are supplied by lavaan.  The ordering of parameters can be obtained via the 
-command \verb+coef(reduced)+.  We see that two of the three $p$-values output at 
+%% move this code description to the cat2BB part. 
+%In the code above, the \verb+parm=1:4+ argument means that 
+%we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
+%which are supplied by lavaan.  The ordering of parameters can be obtained via the 
+%command \verb+coef(reduced)+.
+
+We see that two of the three $p$-values output at 
 the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).
@@
 @
+
+In computing the ordinal test statistics, we
+obtain~$p=\Sexpr{round(wdmo$p.value,2)}$
+and~$p=\Sexpr{round(maxlmo$p.value,2)}$, respectively.
+Both $p-$values are clearly larger than that of the 
+likelihood ratio test and neither is significant at $\alpha=0.05$. 
+This provides evidence that there is no measurement
+invariance violation that is monotonic with age group.  Instead, given
+the large sample size, the likelihood ratio test may be overly
+sensitive to anomalous, non-monotonic violations at one (or a few) age groups.
+
+Besides, plots can be generated by using \verb+fplot=TRUE+ argument inside of \verb+sctest.lavaan()+ function.
+Plots representing the statistics' fluctuations across levels of age group are displayed in Figure~\ref{fig:ordres}.
+The left panel displays the process associated with $\mathit{WDM}$.
+The right panel displays the process associated with $\mathit{maxLM}$.
+In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs.
+greater than $i$, and the null hypothesis is rejected if the maximum of the statistics is larger than its 5\% critical value(visualized by the horizontal red line).
+The final group is not displayed because it encompass all observations and always equals to zero.
+The age group 1:5 in the x-axis corresponds to 10-11, 12-13, 14,15,16 respectively.
+It is observed that both statistics generally increase with age.
 
@@
 
-In computing the ordinal test statistics, we
-obtain~$p=\Sexpr{round(wdmo$p.value,2)}$
-and~$p=\Sexpr{round(maxlmo$p.value,2)}$, respectively.
-Both $p-$values are clearly larger than that of the 
-likelihood ratio test and neither is significant at $\alpha=0.05$. 
-This provides evidence that there is no measurement
-invariance violation that is monotonic with age group.  Instead, given
-the large sample size, the likelihood ratio test may be overly
-sensitive to anomalous, non-monotonic violations at one (or a few) age groups.
+\seeme{can't change age group axis, since order.by argument needs agegroup as numeric}
 
@@
 %% Is there some argument for fplot=TRUE so user can change  axis-name and critical value line
-Plots representing the statistics' fluctuations across levels of age group are displayed in Figure~\ref{fig:ordres}.
-The left panel displays the process associated with WDM.
-The right panel displays the process associated with maxLM.
-In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs.
-greater than $i$, and the null hypothesis is rejected if the maximum of the statistics is larger than its 5\% critical value(visualized by the horizontal red line).
-The final group is not displayed because it encompass all observations and always equals to zero.
-It is observed that both statistics generally increase with age.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{figure}[H]
+\caption{Path diagram representing the factor analysis model 
+         used for simulations.}
+\label{fig:path}
+\includegraphics{PathDiagram}
+\end{figure}
+
 Power and Type I error were examined across three sample sizes
-($n=120,480,960$) , three numbers of categories ($m=4,8,12$)and
+($n=120,480,960$), three numbers of categories ($m=4,8,12$) and
 17 magnitudes of invariance violations. 
@@
   levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
-  levels(sim1$test) <- c("ordmax","ordwmax","catdiff")
+  levels(sim1$test) <- c("maxLM_O","WDM_O","LM_uO")
   save(sim1, file="sim1.rda")
@@
   levels = c("ordmax","ordwmax","catdiff"),
-  labels = c("ordmax","ordwmax","catdiff"))
+  labels = c("maxLM_o","WDM_o","LM_uo"))
   levels(sim1$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors")
@@
 <<sim11res, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("loading"),]
+sim1.tmp <- sim1[sim1$parms %in% c("loading")& sim1$pars %in% c("Loading1","All Loadings","All Errors"),]
 sim1.tmp$test <- factor(sim1.tmp$test)
@@
 <<sim12res, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("var"),]
+sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LM_uo")&sim1$parms %in% c("var") &sim1$pars %in% c("All Loadings", "Covariance","All Errors") ,]
 sim1.tmp$test <- factor(sim1.tmp$test)
@@
 <<sim13res, fig=TRUE, height=7, width=7>>=
-sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("error"),]
+sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LM_uo")&sim1$parms %in% c("error") &sim1$pars %in% c("Error1","All Errors","All Loadings"),]
 sim1.tmp$test <- factor(sim1.tmp$test)
@@
 \subsection{Results}
+Results are presented by the parameter that violated measurement invariance.
+Figure~\ref{fig:sim21res} displays results when the unmodeled loading violates 
+measurement invariance. One can generally observe that 
+the first loading and error variance parameter showed high measurement invariance 
+violation simultaneously, while factor covariance did not demonstrate violation.
 
-%<<sim2>>=
-%## seed for replication
-%RNGkind(kind = "default", normal.kind = "default")
-%set.seed(1163)
+When the loading $\lambda_{11}$ violates
+measurement invariance, even though lacking one loading associated with the first indicator, 
+statistic showed enough power to detect the first loading violation specifically. 
+Similar pattern of specificity and power were observed when the violation parameter 
+was $\sigma_{12}$ and $\epsilon_{11}$. Figures are shown in Appendix for simplicity.
 
-%if(file.exists("sim2.rda")) {
-%  load("sim2.rda")
-%} else {
-%  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
-%  sim2$nlevels <- factor(sim2$nlevels)
-%  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
-%  sim2$nobs <- factor(sim2$nobs)
-%  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
-%  levels(sim2$test) <- c("ordmax","ordwmax","catdiff")
-%  save(sim2, file="sim2.rda")
-%}
-%sim2$test <- factor(as.character(sim2$test),
-%  levels = c("ordmax","ordwmax","catdiff"),
-%  labels = c("ordmax","ordwmax","catdiff"))
-%@
+In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
+as the umodeled parameter does not violate measurment invariance. If the unmodeled parameter 
+dose violate measurement invariance, however, then the tests assign this violation to other 
+parameters that do not violate measurement invariance. The impacted parameters include the error
+variance and other loadings associated with the manifest variables that had the unmodeled 
+loading. Thus, prior to testing measurement invariance, it is important to study the extent to
+which the hypothesized model includes all parameters of importance.
+
+<<sim2>>=
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim2.rda")) {
+  load("sim2.rda")
+} else {
+  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
+
+  save(sim2, file="sim2.rda")
+}
+sim2$test <- factor(as.character(sim2$test),
+  levels = c("ordmax","ordwmax","catdiff"),
+  labels = c("maxLM_o","WDM_o","LM_uo"))
+  sim2$nlevels <- factor(sim2$nlevels)
+  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
+  sim2$nobs <- factor(sim2$nobs)
+  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
+  levels(sim2$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors")
+
+@
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim21res}
+\setkeys{Gin}{width=\textwidth}
+<<sim21res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra")&sim2$pars %in% c("Loading1","Error1","Covariance"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
 
@@
 
-The core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity.
+We focused on testing for measurement invariance in factor analysis model in this paper, however, the core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity.
+For example, studying DIF in IRT is immediate.
+
+In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable.
+This
+feature is robust as long as the violation parameter is included in the testing model.
+Furthermore, as illustrated in tutorial part, the use of these tests is convenient without changing the focal psychometric model.
+In all, the tests have advantageous properties that should be implemented in practice.
+
+\section{Appendix}
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim22res}
+\setkeys{Gin}{width=\textwidth}
+<<sim22res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+loading"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim23res}
+\setkeys{Gin}{width=\textwidth}
+<<sim23res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+var"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}[H]
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim24res}
+\setkeys{Gin}{width=\textwidth}
+<<sim24res, fig=TRUE, height=7, width=7>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+error"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
 

### 2013-09-15 frontiers/simall-ordinal.R

@@ -484,4 +484,6 @@ mykey <- simpleKey(levels(simtry.tmp$test), points = TRUE, lines = TRUE)
 xyplot(power ~ diff | nlevels + pars + parms, group = ~ test, data = simtry.tmp, 
              type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey)
 dev.off()
+
+ 
 }

### 2013-09-17 frontiers/frontiers.Rnw
update refenrence and general discussion section.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@260 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 the applicability of adult gratitude scales to youth, available in the~R 
-package psychotools \cite{zeilei11}.  The data consist of responses to three adult 
+package psychotools \cite{zeilei12}.  The data consist of responses to three adult 
 gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
@@
 maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")
-
 @ 
@@
 @
+
+\seeme{I don't know how to adjust the line interval space for the code part.}
 
@@
 
+\seeme{When I make the picture bigger, the symbols became blur. I think I might need to work on the plot tex file.}
 Power and Type I error were examined across three sample sizes
@@
 
+\seeme{I think it might be nicer to keep these plots all have three rows?(Not sure)}
 \section{Simulation 2}
@@
 In this paper, we examined the recently proposed ordinal statistics specificity property and robustness to model specification to study measurement invariance.
-As demonstrated in the results, the proposed statistics could select the parameter violating measurement invariance even though the model is misspecified in general.
-However, when an unmodeled parameter is violating measurement invariance, the statistics will demonstrating these variances in the related parameters, for example, associated loading and error terms.
+As demonstrated in the results, the proposed statistics could select the modeled parameter violating measurement invariance even though the model is misspecified in general.
 
+\subsection{Application}
 These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context.
@@
 
+\subsection{Extension}
 We focused on testing for measurement invariance in factor analysis model in this paper, however, the core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity.
@@
 
+\subsection{Conclusion}
 In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable.

### 2013-09-17 frontiers/frontiers.Rnw
Work on frontiers paper: o Small revisions to paper o Small revisions/additions to references o Added LMuo to sctest.lavaan
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@261 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \abstract{
-We study a family of 
-recently-proposed measurement invariance tests that are derived from
-stochastic processes.  As applied to factor analysis and other
-psychometric models, the tests are advantageous in that (i) they do
-not require advance definition of subgroups of cases that violate
-measurement invariance; (ii) they can potentially identify specific
-model parameters that violate measurement invariance; and (iii) they
-do not require a model of additional complexity to be estimated.  
-Because the tests have been applied to only a small number of
-psychometric examples, we conduct a series of simulation studies that provide
-a detailed examination of test properties.  These studies allow us
-to study (details here).  The studies indicate that.}
-
-\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
+In this paper, we consider a family of 
+recently-proposed measurement invariance tests that are based on the {\em scores} of a fitted model.  This family can be used to test for measurement invariance
+w.r.t.\ a continuous auxiliary variable, without pre-specification of subgroups.
+Moreover, the family can be used when one wishes to test for 
+measurement invariance w.r.t.\ an ordinal auxiliary variable, yielding test
+statistics that are sensitive to violations that are monotonically related to
+the ordinal variable (and less sensitive to non-monotonic violations).
+The paper is specifically aimed at potential users of the tests who may wish to know (i) how the tests can be employed for their data, and (ii) whether the tests can accurately identify measurement invariance violations in the presence of model misspecification.
+After providing an overview of the tests, we illustrate their general use via the R packages lavaan and strucchange.
+We then describe two novel simulations that provide evidence of the tests' practical abilities.
+As a whole, the paper provides researchers with the tools and knowledge needed to apply these tests to general measurement invariance scenarios.}
 
@@
   Foundation grant SES-1061334. 
-  Correspondence to. 
-  Email: }
+  Correspondence to Edgar Merkle. 
+  Email: merklee@missouri.edu.}
 \shorttitle{Measurement invariance test properties}
@@
 \newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
-\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}
-\newcommand{\seeme}[1]{\emph{\marginpar{SEEME} (#1)}}
+\newcommand{\readme}[1]{\emph{\marginpar{Ed} (#1)}}
+\newcommand{\seeme}[1]{\emph{\marginpar{Ting} (#1)}}
 
@@
 
+\readme{We need a better intro here.  The special topic is about measurement invariance, so most of this paragraph is obvious to everyone.  Maybe start with how so many other papers are about BSEM.}
 Parameter invariance is a fundamental assumption for measurement model in 
@@
 of these scores forms the ``meat'' for the calculation of robust
-(Huber-White) standard errors \cite{freed06}. \seeme{not sure about this citation}
+(Huber-White) standard errors \cite<e.g.,>{zei06b}.
 
@@
 This simulation is somewhat computationally intensive and, in practice, takes about 
-10 minutes on the authors' computers.  However, the wait appears to be worth it, 
+10 minutes on the authors' computers.  However, the wait is often worth it, 
 as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic 
@@
 
-If $V$ is only nominal/categorical,
-there is not even a partial ordering, i.e., measurement invariance tests should neither
-exploit the ordering of $V$'s levels nor of the observations within the level.
-In this situation, unordered $\mathit{LM}$ is proposed.  This test statistic can be formally written as
+Finally, if $V$ is only nominal/categorical, the cumulative sums of scores can be used to obtain the a Lagrange multiplier statistic.
+This test statistic can be formally written as
 \begin{equation}
@@
 \end{equation}
-This test statistic discards the ordinal nature of the auxiliary
-variable, essentially similar to the nature of likelihood ratio tests (or,
-equivalently, via Wald tests or Lagrange multiplier tests): the ordinality of the auxiliary variable is
-ignored. By contrast, this statistics is easier to get compared to likelihood ratio test statistics since it doesn't involve specifying different sets of competitive models.
+and it is asymptotically equivalent to the usual, likelihood ratio test statistic.
+It is advantageous over the likelihood ratio test because it requires estimation of only one model (the restricted model).
+We make use of this advantage in the simulations, described later.
 
-\SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
 \section{Tutorial}
@@
 
+\readme{I just added LMuo to the sctest command; the range statistic is different.}
 Another similar statistics is $\mathit{LM}_\mathit{uo}$ as shown in Equation~ \eqref{eq:lmuo}. We can obtain this statistic by using the following function. 
@@
 ## unordered LM
-lm_u0 <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="range")
-round(c(lm_u0$p.value), 2)
+lm_uo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="LMuo")
+round(lm_uo$p.value, 4)
 @
-we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
-which are supplied by lavaan.  The ordering of parameters can be obtained via the 
-command \verb+coef(reduced)+.
+The above command specifies that we wish to test model parameter numbers 1 through 4, which are the factor loadings supplied by lavaan.
+The ordering of parameters can be obtained via the command \verb+coef(reduced)+.
 
-Because our sample size is large, this likelihood ratio test is sensitive to 
-small measurement invariance violations \cite{Ben80}.  Thus, the above test may 
-be significant solely due to the large sample size.  Additionally, the above test 
-is sensitive to small measurement invariance violations that are not likely to be 
+Because our sample size is large, this likelihood ratio test is known to be sensitive to 
+small measurement invariance violations \cite{Ben80}.  That is, the above tests 
+are sensitive to small measurement invariance violations that are not likely to be 
 of interest to researchers.  For example, imagine that the 15-year-olds' 
@@
 is not likely to be any theoretical justification for 15-year-olds 
-differing from every other age group.  We need test statistics that 
-are sensitive to measurement invariance violations that are monotonic 
-with age.  The score-based statistics described previously possess this property.
+differing from every other age group.
+One solution to this problem would be the {\em approximate invariance} methods proposed by Muth\'{e}n and others.\readme{Ting: Can you get some approximate invariance references and put them here?}
+Alternatively, we can use the proposed family of score-based statistics to obtain tests that are sensitive to the ordering of age.
 
@@
 are restricted to be equal across all individuals and then examine 
-how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
+how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
 age (where age ties are broken arbitrarily).
 This is demonstrated below, with similar code being useful when one is 
-testing measurement invariance w.r.t.\ other continuous variables.
+testing measurement invariance w.r.t.\ truly continuous variables.
 
@@
 To proceed, we focus only on the restricted model that was fit in 
-the previous section (the object \verb+reduced+).  We obtain scores 
+the previous section (stored in the object \verb+reduced+).  We obtain scores 
 associated with the factor loadings from this model, then test the 
@@
 %command \verb+coef(reduced)+.  
-
 We see that two of the three $p$-values output at 
@@
 To accurately account for the multiple observations  
-at the same age level, we must used the ordinal tests from 
+at the same age level, we must use the ordinal tests from 
 Equations~\eqref{eq:wdm} and~\eqref{eq:maxlmo}.  These are described next.
@@
 %% at each level of the ordinal variable.  
-
-The difference between the ordinal test statistics and their
+The main difference between the ordinal test statistics and their
 continuous counterparts is that the ordinal statistics do not require
@@
 can be lengthy from the need to simulate critical values.
-
-<<critvals,echo=FALSE>>=
+%% This is more complicated than necessary, but it is written like
+%% this so that we can load the ordinal critical values behind the
+%% scenes without confusing the reader.
+<<critvals,echo=FALSE,strip.white=all>>=
 if(file.exists("opval1.rda")){
@@
 @ 
-
-<<ordtest,eval=FALSE>>=
+<<ordtest,eval=FALSE,strip.white=all>>=
 wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
@@
 @ 
-<<pvals>>=
+<<pval,strip.white=all>>=
 round(c(wdmo$p.value, maxlmo$p.value), 2)
 @ 
-
-\seeme{I don't know how to adjust the line interval space for the code part.}
-
 In computing the ordinal test statistics, we
@@
 
-Besides, plots can be generated by using \verb+fplot=TRUE+ argument inside of \verb+sctest.lavaan()+ function.
+In addition to test statistics, ``instability plots'' can be generated by using the \verb+fplot=TRUE+ argument within \verb+sctest.lavaan()+.
 Plots representing the statistics' fluctuations across levels of age group are displayed in Figure~\ref{fig:ordres}.
-The left panel displays the process associated with $\mathit{WDM}$.
-The right panel displays the process associated with $\mathit{maxLM}$.
-In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs.
+The left panel displays the process associated with $\mathit{WDM}_o$, and the right panel displays the process associated with $\max\mathit{LM}_o$.
+%The following sentence is not correct for all statistics.  
+%In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs.
 greater than $i$, and the null hypothesis is rejected if the maximum of the statistics is larger than its 5\% critical value(visualized by the horizontal red line).
-The final group is not displayed because it encompass all observations and always equals to zero.
-The age group 1:5 in the x-axis corresponds to 10-11, 12-13, 14,15,16 respectively.
-It is observed that both statistics generally increase with age.
+These plots display the loadings' fluctuation across age groups, with the x-axis reflecting age group and the y-axis reflecting test statistic values (larger values reflect more instability).
+The red lines reflect critical values, so that the hypothesis of measurement invariance is rejected if a black line crosses the red line.
+While the measurement invariance tests are non-significant, the plots imply some instability in the older age groups (15 and 16).
 
@@
 \begin{figure}[H]
-\caption{Fluctuation processes for the WDM statistic(left panel) and the maxLM statistic(right panel)}
+\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and the $\max\mathit{LM}_o$ statistic (right panel).}
 \label{fig:ordres}
 \setkeys{Gin}{width=\textwidth}
-<<ordres, fig=TRUE, height=3, width=6, echo=FALSE>>=
+<<ordres, fig=TRUE, height=3, width=6, echo=FALSE, results=hide>>=
 par(mfrow = c(1, 2), cex = 0.65)
-WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE,ylim=c(0,3.5),xlab="Age group",ylab = "Weighted max statistics")
+sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo", 
+       fplot=TRUE, ylim=c(0,3.5), xlab="Age group",
+       ylab = "Weighted max statistics", axes=FALSE)
+axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
+box()
 
-#WDMo 
-maxLMo<-sctest(reduced,as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fnl=opval1, fplot=TRUE,ylim=c(0,14),xlab="Age group",ylab = "LM statistics")
-#maxLMo 
+sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo", 
+       fnl=opval1, fplot=TRUE, ylim=c(0,14), xlab="Age group",
+       ylab = "LM statistics", axes = FALSE)
+axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
+box()
 @
 \end{figure}
-
-\seeme{can't change age group axis, since order.by argument needs agegroup as numeric}
 
@@
 
-\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
-\section{Simulation}
 \section{Simulation 1}
@@
 
-<<sim1>>=
+<<sim1,echo=FALSE>>=
 ## seed for replication
@@
   levels(sim1$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors")  
-
 @

### 2013-09-17 pkg/sctest.lavaan.R

@@ -20,7 +20,8 @@ sctest.lavaan <- function(x, order.by, parm, stat, fnl = NULL, fplot = FALSE, ..
                   DM = maxBB,
                   CvM = meanL2BB,
                   maxLM = supLM(0.1),
-                  range = rangeBB)
+                  range = rangeBB,
+                  LMuo = catL2BB(fp))
   }
 
   ## If stat argument is messed up

### 2013-09-18 frontiers/frontiers.Rnw
o Revisions to frontiers ms o Added saved critical values to speed up frontiers compilation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@262 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-<<preliminaries1>>=
+<<preliminaries1,echo=FALSE>>=
 ## packages
@@
 
-\readme{We need a better intro here.  The special topic is about measurement invariance, so most of this paragraph is obvious to everyone.  Maybe start with how so many other papers are about BSEM.}
+\readme{We need a better intro here.
+The special topic is about measurement invariance, so most of this paragraph is obvious to everyone.
+Maybe start with a remark about approximate invariance and try to tie our topic to that somehow.}
 Parameter invariance is a fundamental assumption for measurement model in 
@@
 
-%% TODO Illustrate catL2BB statistic
-
 Finally, if $V$ is only nominal/categorical, the cumulative sums of scores can be used to obtain the a Lagrange multiplier statistic.
@@
 
-\readme{I just added LMuo to the sctest command; the range statistic is different.}
+\readme{I just added LMuo to the sctest command and corrected the command below; the range statistic is different.}
 Another similar statistics is $\mathit{LM}_\mathit{uo}$ as shown in Equation~ \eqref{eq:lmuo}. We can obtain this statistic by using the following function. 
@@
 } else {
+  gefp1 <- gefp(reduced, fit=NULL, order.by=as.numeric(yg$agegroup),
+                vcov=info_full, sandwich=FALSE, parm=1:4)
   opval1 <- ordL2BB(gefp1)
@@
 and~$p=\Sexpr{round(maxlmo$p.value,2)}$, respectively.
-Both $p-$values are clearly larger than that of the 
+Both $p$-values are clearly larger than that of the 
 likelihood ratio test and neither is significant at $\alpha=0.05$. 
@@
 The left panel displays the process associated with $\mathit{WDM}_o$, and the right panel displays the process associated with $\max\mathit{LM}_o$.
-%The following sentence is not correct for all statistics.  
+%The following sentence is not correct for all statistics:  
 %In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs.
@@
 %Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
-\begin{figure}[H]
+\begin{figure}
 \caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and the $\max\mathit{LM}_o$ statistic (right panel).}
@@
 \section{Simulation 1}
+
 In Simulation 1, we examined the extent to which the proposed tests
@@
 \subsection{Methods}
+
 To examine these specificity issues, we generated data from a
@@
 
-\begin{figure}[H]
+\begin{figure}
 \caption{Path diagram representing the factor analysis model 
@@
 
-\begin{figure}[H]
+\begin{figure}
 \caption{Simulated power curves for the ordered 
@@
 \setkeys{Gin}{width=\textwidth}
-<<sim11res, fig=TRUE, height=7, width=7>>=
+<<sim11res, fig=TRUE, height=7, width=7, echo=FALSE>>=
 sim1.tmp <- sim1[sim1$parms %in% c("loading")& sim1$pars %in% c("Loading1","All Loadings","All Errors"),]
@@
 \setkeys{Gin}{width=\textwidth}
-<<sim12res, fig=TRUE, height=7, width=7>>=
+<<sim12res, fig=TRUE, height=7, width=7, echo=FALSE>>=
 sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LM_uo")&sim1$parms %in% c("var") &sim1$pars %in% c("All Loadings", "Covariance","All Errors") ,]
@@
 \setkeys{Gin}{width=\textwidth}
-<<sim13res, fig=TRUE, height=7, width=7>>=
+<<sim13res, fig=TRUE, height=7, width=7, echo=FALSE>>=
 sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LM_uo")&sim1$parms %in% c("error") &sim1$pars %in% c("Error1","All Errors","All Loadings"),]
@@
 
-<<sim2>>=
+<<sim2,echo=FALSE>>=
 ## seed for replication
@@
 \setkeys{Gin}{width=\textwidth}
-<<sim21res, fig=TRUE, height=7, width=7>>=
+<<sim21res, fig=TRUE, height=7, width=7, echo=FALSE>>=
 sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra")&sim2$pars %in% c("Loading1","Error1","Covariance"),]
@@
 
-\section{Appendix}
-\begin{figure}[H]
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance 
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
-\label{fig:sim22res}
-\setkeys{Gin}{width=\textwidth}
-<<sim22res, fig=TRUE, height=7, width=7>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+loading"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-
-\end{figure}
-
-\begin{figure}[H]
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance 
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
-\label{fig:sim23res}
-\setkeys{Gin}{width=\textwidth}
-<<sim23res, fig=TRUE, height=7, width=7>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+var"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-
-\end{figure}
-
-\begin{figure}[H]
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance 
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
-\label{fig:sim24res}
-\setkeys{Gin}{width=\textwidth}
-<<sim24res, fig=TRUE, height=7, width=7>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+error"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
-@
-
-\end{figure}
-
 \section*{Computational Details}
@@
 
+\section{Appendix}
+
+\readme{Need some text describing what this appendix contains.}
+
+\begin{figure}
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim22res}
+\setkeys{Gin}{width=\textwidth}
+<<sim22res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+loading"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim23res}
+\setkeys{Gin}{width=\textwidth}
+<<sim23res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+var"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
+\begin{figure}
+\caption{Simulated power curves for the ordered 
+      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
+      levels of the ordinal variable $m$, and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\label{fig:sim24res}
+\setkeys{Gin}{width=\textwidth}
+<<sim24res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+error"),]
+sim2.tmp$test <- factor(sim2.tmp$test)
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+@
+
+\end{figure}
+
 \end{document}

### 2013-09-23 frontiers/frontiers.Rnw
include plot code in frontiers.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@263 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage[english]{babel}
+\usepackage{tikz}
 
@@
 Figure~\ref{fig:path}.
+\seeme{How to refer this figure?}
 
-\begin{figure}
-\caption{Path diagram representing the factor analysis model 
-         used for simulations.}
-\label{fig:path}
-\includegraphics{PathDiagram}
-\end{figure}
+%\begin{figure}
+%\caption{Path diagram representing the factor analysis model 
+%         used for simulations.}
+%\label{fig:path}
+%\includegraphics{PathDiagram}
+%\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Plot Code
 
-\seeme{When I make the picture bigger, the symbols became blur. I think I might need to work on the plot tex file.}
+\begin{tikzpicture}
+[latent/.style={circle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt, minimum width=20mm, minimum height=10mm},
+ manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt,minimum width=20mm, minimum height=10mm},
+ error/.style={circle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt, minimum width=10mm},
+ post/.style={->,shorten >=1pt,>=stealth',semithick},
+ sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=20mm,semithick},
+ errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=10mm,semithick}]
+
+%position objects.
+
+\node[latent] (VERBAL) at (0,4) {VERBAL};
+\node[latent] (MATH)   at (0,0) {MATH};
+
+\node[manifest] (SCALE1) at(5,6) {SCALE1};
+\node[manifest] (SCALE2) at (5,4.5) {SCALE2};
+\node[manifest] (SCALE3) at (5,3) {SCALE3};
+\node[manifest] (SCALE4) at(5,1) {SCALE4};
+\node[manifest] (SCALE5) at (5,-0.5) {SCALE5};
+\node[manifest] (SCALE6) at (5,-2) {SCALE6};
+
+\node[error] (E1) at (7,6) {E1};
+\node[error] (E2) at (7,4.5) {E2};
+\node[error] (E3) at (7,3) {E3};
+\node[error] (E4) at (7,1) {E4};
+\node[error] (E5) at (7,-0.5) {E5};
+\node[error] (E6) at (7,-2) {E6};
+
+%arrows, slings and texts
+\node[latent] (VERBAL) at (0,4) {VERBAL}
+   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{11}=4.92$}} (SCALE1)
+   edge [post] node[above=0.001mm,text width=2cm]{\scriptsize{$\lambda_{21}=2.96$}} (SCALE2)
+   edge [post] node[above=0.1cm,text width=2cm]{\scriptsize{$\lambda_{31}=5.96$}} (SCALE3)
+   edge [sling,bend right=100] node[auto,swap]{1}(VERBAL.-180)
+   edge [sling,bend right=70] node[auto,swap]{$\phi_{12}=-0.48$}(MATH)  ;
+
+\node[latent] (MATH) at (0,0) {MATH}
+   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{42}=3.24$}}(SCALE4)
+   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{52}=4.32$}}(SCALE5)
+   edge [post] node[above=0.25cm,text width=2cm]{\scriptsize{$\lambda_{62}=7.21$}}(SCALE6)
+   edge [sling,bend right=100] node[auto,swap]{1}(MATH.270);
+
+\node[error] (E1) at (7,6) {E1}
+   edge [post] node[auto,swap]{\small{1}}(SCALE1)
+   edge [errorsling, bend right=100]node[right=0.1,text width=2cm]{\scriptsize{$\psi_{11}=26.77$}} (E1.45);
+\node[error] (E2) at (7,4.5) {E2}
+   edge [post] node[auto,swap]{\small{1}} (SCALE2)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{22}=13.01$}}(E2.45);
+\node[error] (E3) at (7,3) {E3}
+   edge [post] node[auto,swap]{\small{1}}(SCALE3)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{33}=30.93$}}(E3.45);
+\node[error] (E4) at (7,1) {E4}
+   edge [post] node[auto,swap]{\small{1}}(SCALE4)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{44}=3.17$}}(E4.45);
+\node[error] (E5) at (7,-0.5) {E5}
+   edge [post] node[auto,swap]{\small{1}}(SCALE5)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{55}=8.82$}} (E5.45);
+\node[error] (E6) at (7,-2) {E6}
+   edge [post] node[auto,swap]{\small{1}}(SCALE6)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{11}=22.5$}}(E6.45);
+
+\end{tikzpicture}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Power and Type I error were examined across three sample sizes

### 2013-09-24 frontiers/frontiers.Rnw
Added comments and revisions to frontiers simulations.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@264 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{tikz}
+
+% Seems necessary for texlive 2012?
+\usetikzlibrary{arrows}
 
@@
     \label{eq:lmuo}
-     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+     \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
        \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
@@
 invariance, it is plausible that this violation impacts other
-parameters such as factor covariances or other loadings 
+parameter estimates, including factor covariances or other loadings 
 associated with the factor in question.  Thus, the goal of the
-simulation was to examine the extent to which the proposed tests
+Simulation 1 is to examine the extent to which the proposed tests
 attribute the measurement invariance violation to the parameters that
@@
 To examine these specificity issues, we generated data from a
-two-factor model with three indicators each (the same model used in
-Merkle and Zeileis, \citeyearNP{MerZei13}). The measurement invariance
-violation occurred in one of three places: the factor loadings
- $\lambda_{11}--\lambda_{62}$, the unique variance associated
- $\epsilon_{11}--\epsilon_{66}$, or the factor covariance $\sigma_{12}$.  
+two-factor model with three indicators each (see Figure~\ref{fig:path}).
+\readme{Please make sure the text below is correct.  The previous text stated that the violation occurs in all loadings and all unique variances.}
+The measurement invariance
+violation occurred in one of three places: the factor loading associated with Scale 1 
+ ($\lambda_{11}$), the unique variance associated with Scale 1
+ ($\epsilon_{11}$), or the factor covariance $\sigma_{12}$.  
 We then tested for measurement invariance in seven subsets of parameters: the three
-individual parameters noted above, all six factor loadings $\lambda_{11}--\lambda_{62}$, 
-all six unique variances $\epsilon_{11}--\epsilon_{66}$. The path diagram is shown in  
-Figure~\ref{fig:path}.
-\seeme{How to refer this figure?}
+individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
+and all six unique variances. % $\epsilon_{11}--\epsilon_{66}$.
 
@@
 
+\begin{figure}
+    \caption{General model used for the simulations.}
+    \label{fig:path}
 \begin{tikzpicture}
@@
 \end{tikzpicture}
+\end{figure}
 
@@
 magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
-generated and tested. Four statistics were examined above. In all conditions, we
+generated and tested. Four statistics were examined above. \readme{What four statistics?  Equation numbers would be helpful.} 
+In all conditions, we
 maintained equal sample sizes at each level of the ordinal variable.
@@
 \subsection{Results}
+\readme{When referring to statistics, it would be good to use the notation from Equations 6--8.  For example, instead of ``unordered $\max\mathit{LM}$'', you would have $\max\mathit{LM}_{uo}$.}
 Full simulation results are presented in Figure~\ref{fig:sim11res} to 
-Figure~\ref{fig:sim13res}; all three test statistics exhibited 
-similar results, with unordered  $\max\mathit{LM}$ demonstrated lower powers across all situations.
-ordered $\max\mathit{LM}$ and double-max exhibited equivalent power curves since there is only one parameter violating measurement invariance.
-Thus, different ways of aggregating $\bm B(\hat{\theta})$ would be the same.
+Figure~\ref{fig:sim13res}; 
 Figure~\ref{fig:sim11res} displays power curves as a function of 
@@
 invariance, respectively. 
+All three test statistics exhibited 
+similar results, with unordered  $\max\mathit{LM}$ demonstrated lower powers across all situations.
+ordered $\max\mathit{LM}$ and double-max exhibited equivalent power curves because, when only one parameter is tested, the statistics are equivalent.
 
@@
 
+\readme{We need to compare the specific tests: the unordered test, which mimics the LRT, has lower power.}
+
 In summary, we found that the proposed tests can attribute
@@
 simulations, we examine the tests' performance when the model is
-misspecified.
+misspecified. \readme{The plots have too many points; maybe we should only plot  every other diff value.  In the xyplot command, you could add the argument \verb+subset = diff %% .5 == 0+.}
 
@@
 \seeme{I think it might be nicer to keep these plots all have three rows?(Not sure)}
+
 \section{Simulation 2}
@@
 generate data from the factor analysis model used in the previous
-section, but with adding an extra loading from the second factor to the first
-indicator (of the first factor).  The estimated model lacks this extra
-loading, however. The goal of this simulation is to examine the proposed statistics'
-power and specificity under model misspecification.
+section, except that the model contains an extra loading from the second factor to Scale 1.
+The estimated model matches that displayed in Figure~\ref{fig:path}, however, resulting in model misspecification.  The goal of this simulation is to examine the proposed statistics'
+power and specificity under this misspecification.
 
 \subsection{Method}
+\readme{Make sure there is space after each period and comma, the beginning of each sentence is capitalized, etc.}
 Measurement invariance violations could occur in
-each of the three parameters from Simulation 1 (factor loading,factor covariance,unique
-variance), and they could also occur in the extra, unmodeled loading. 
+each of the three parameters from Simulation 1 (factor loading, factor covariance, and unique variance), and they could also occur in the extra, unmodeled loading.
+In each condition, a single parameter exhibited the violation.
 Sample size and magnitude of measurement invariance violation were manipulated 
-in the same way as they were in Simulation 1.The tested parameters are the same as Simulation
-1,too.
+in the same way as they were in Simulation 1.The tested parameters were also the same as Simulation
+1.
 
@@
 Results are presented by the parameter that violated measurement invariance.
-Figure~\ref{fig:sim21res} displays results when the unmodeled loading violates 
+Of primary interest is Figure~\ref{fig:sim21res}, which displays results when the unmodeled loading violates 
 measurement invariance. One can generally observe that 
-the first loading and error variance parameter showed high measurement invariance 
-violation simultaneously, while factor covariance did not demonstrate violation.
+tests of the first loading and error variance parameter exhibited high power, reflecting a high Type I error rate.  Tests associated with the factor covariance did not demonstrate this error, however.
 
-When the loading $\lambda_{11}$ violates
-measurement invariance, even though lacking one loading associated with the first indicator, 
-statistic showed enough power to detect the first loading violation specifically. 
-Similar pattern of specificity and power were observed when the violation parameter 
-was $\sigma_{12}$ and $\epsilon_{11}$. Figures are shown in Appendix for simplicity.
+\readme{Need some comparison between the unordered and ordered tests.}
 
+When the modeled factor loading, $\lambda_{11}$, violated
+measurement invariance, the statistics were generally able to pick up the violation despite the misspecification.
+A similar pattern of specificity and power were observed when the unique variance and factor covariance paraameters violated measurement invariance; the figures associated with these results are shown in the Appendix.
+
+\readme{We should say something about how the unordered test looks better when the unmodeled parameter violates MI.}
 In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long

### 2013-09-25 frontiers/frontiers.Rnw
Revisions to frontiers paper, simulation sections.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@265 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 All three test statistics exhibited 
-similar results, with unordered  $\max\mathit{LM}$ demonstrated lower powers across all situations.
+similar results, with unordered  $\max\mathit{LM}$ demonstrating lower power across all situations.
 ordered $\max\mathit{LM}$ and double-max exhibited equivalent power curves because, when only one parameter is tested, the statistics are equivalent.
-
-From these figures, one generally observe that the tests isolate 
+\readme{Should say that the unordered statistic is asymptotically equivalent to the LRT.}
+
+From these figures, one generally observes that the tests isolate 
 the parameter violating measurement invariance.  Additionally, the tests 
@@
 
-\readme{We need to compare the specific tests: the unordered test, which mimics the LRT, has lower power.}
-
 In summary, we found that the proposed tests can attribute
@@
 simulations, we examine the tests' performance when the model is
-misspecified. \readme{The plots have too many points; maybe we should only plot  every other diff value.  In the xyplot command, you could add the argument \verb+subset = diff %% .5 == 0+.}
+misspecified. \readme{The plots have too many points; maybe we should only plot  every other diff value.  In the xyplot command, you could add the argument ``subset = diff \%\% .5 == 0''.}
 
@@
   labels = c("maxLM_o","WDM_o","LM_uo"))
+  parlabs <- c(expression(lambda[11]), expression(phi[12]), 
+               expression(psi[11]), expression(lambda[11] - lambda[62]), 
+               expression(psi[11] - psi[66]))
   levels(sim1$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors")  
@@
 \caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance 
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Loading1.}
+      $\max\mathit{LM}$ test, the double-max test, and the unordered $\max\mathit{LM}$ test across three
+      levels of the ordinal variable $m$ and measurement invariance 
+      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement invariance is $\lambda_{11}$.}
 \label{fig:sim11res}
@@
 <<sim11res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-sim1.tmp <- sim1[sim1$parms %in% c("loading")& sim1$pars %in% c("Loading1","All Loadings","All Errors"),]
-sim1.tmp$test <- factor(sim1.tmp$test)
+#sim1.tmp <- sim1[sim1$parms %in% c("loading")& sim1$pars %in% c("Loading1","All Loadings","All Errors"),]
+#sim1.tmp$test <- factor(sim1.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1,
+             subset = (parms=="loading" & 
+                      pars %in% c("Loading1","All Loadings","All Errors")),
+             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey,
+             strip = function(..., which.given, factor.levels){
+                 if(which.given==2){
+                     strip.default(which.given, factor.levels = parlabs[c(1,4:5)], ...)
+                 } else {
+                     strip.default(which.given, factor.levels = factor.levels, ...)
+                 }
+             }))
 @

### 2013-09-25 frontiers/frontiers.Rnw
Minor updates to frontiers.Rnw: o Revised subscripts in path diagram o Removed single quotes in Sweave chunks due to issues with texlive
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@266 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\include{Sweave}
 <<preliminaries1,echo=FALSE>>=
@@
 
-## remove cases with 'imputed' values(not in 1,...,9)
+## remove cases with imputed values(not in 1,...,9)
 compcases <- apply(YouthGratitude[,4:28], 1, function(x) all(x %in% 1:9))
@@
 ## Restricted model
-reduced <- cfa('f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+reduced <- cfa("f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
              data=yg, group="agegroup", meanstructure=TRUE,
@@
 ## Unrestricted model
-full <- cfa('f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
+full <- cfa("f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
              data=yg, group="agegroup", meanstructure=TRUE)
@@
    edge [post] node[auto,swap]{\small{1}}(SCALE6)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{11}=22.5$}}(E6.45);
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{66}=22.5$}}(E6.45);
 

### 2013-09-26 frontiers/frontiers.Rnw
make changes according to Ed's comments.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@267 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\readme{We need a better intro here.
-The special topic is about measurement invariance, so most of this paragraph is obvious to everyone.
-Maybe start with a remark about approximate invariance and try to tie our topic to that somehow.}
 Parameter invariance is a fundamental assumption for measurement model in 
-developing psychometric scales. Violation of it could happen in any estimated parameter.  
-In order to fix the psychometric scale, we need to specify the violating parameter.  
-Moreover, in practice model misspecification is an inevitable issue to a certain extent 
-and may interfere with parameter invariance specificity. A family of tests derived from 
-stochastic process has been recently proposed to identify specific parameters impacted 
-by measurement invariance violation. However, these tests have only been used to a limited 
-type of parameter violation. It is this paper's intent to provide a detailed examination 
-of the tests' performance under practical research scenarios.
+developing psychometric scales. Violation of it could happen in any degree. 
+Generally, researchers want to ignore minor violation parameters, e.g. random 
+normally distributed deviations from a parameter value across groups. BSEM
+(Bayesian Structural Equation Modelling) is proposed not to pick up this kind of 
+violating parameters but only "real" violating ones by assigning priors instead of strictly 
+constraining equality across groups. After identifying the violation parameters, 
+those non-invariant parameters are freely estimated to ensure final correct estimation.
+
+Instead of using this two-step Bayesian analysis procedure, we are applying a family of tests derived
+from stochastic process.
+These tests have been recently proposed to identify specific parameters impacted by measurement invariance.
+However, they have only been used to a limited type of parameter violation.
+It is this paper's intent to provide a detailed examination
+of the tests' performance under practical research scenarios. Specifically, we will examine these tests
+specificity in identifying the violation parameter without giving false positive results.
+Moreover, in practice model misspecification is an inevitable issue to a certain extent, we will demonstrate these tests
+results under model misspecification condition. 
+\seeme{What about this intro? too much BSEM?}
+%In order to fix the psychometric scale, we need to specify the violating parameter.  
+%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
+%and may interfere with parameter invariance specificity. A family of tests derived from 
+%stochastic process has been recently proposed to identify specific parameters impacted 
+%by measurement invariance violation. However, these tests have only been used to a limited 
+%type of parameter violation. It is this paper's intent to provide a detailed examination 
+%of the tests' performance under practical research scenarios.
 
@@
 ## Restricted model
-reduced <- cfa("f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-             data=yg, group="agegroup", meanstructure=TRUE,
-             group.equal="loadings")
+reduced <- cfa("f1 = ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+             data = yg, group = "agegroup", meanstructure = TRUE,
+             group.equal = "loadings")
 
 ## Unrestricted model
-full <- cfa("f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-             data=yg, group="agegroup", meanstructure=TRUE)
+full <- cfa("f1= ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+             data = yg, group = "agegroup", meanstructure = TRUE)
 @
@@
 
-\readme{I just added LMuo to the sctest command and corrected the command below; the range statistic is different.}
 Another similar statistics is $\mathit{LM}_\mathit{uo}$ as shown in Equation~ \eqref{eq:lmuo}. We can obtain this statistic by using the following function. 
@@
 ## unordered LM
-lm_uo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="LMuo")
+lm_uo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "LMuo")
 round(lm_uo$p.value, 4)
@@
 differing from every other age group.
-One solution to this problem would be the {\em approximate invariance} methods proposed by Muth\'{e}n and others.\readme{Ting: Can you get some approximate invariance references and put them here?}
+One solution to this problem would be the {\em approximate invariance} methods proposed by Muth\'{e}n \cite{muthen13}\seeme{How to handle muthen's name in refs.bib file?}.
 Alternatively, we can use the proposed family of score-based statistics to obtain tests that are sensitive to the ordering of age.
@@
 <<Get Estimate>>=
-dm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM")
-cvm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM")
-maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM")
+dm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "DM")
+cvm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "CvM")
+maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLM")
 round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
@@
 } else {
-  gefp1 <- gefp(reduced, fit=NULL, order.by=as.numeric(yg$agegroup),
-                vcov=info_full, sandwich=FALSE, parm=1:4)
+  gefp1 <- gefp(reduced, fit = NULL, order.by = as.numeric(yg$agegroup),
+                vcov = info_full, sandwich = FALSE, parm = 1:4)
   opval1 <- ordL2BB(gefp1)
@@
 ## Get test statistics using fnl=opval1 without showing it in the paper
-wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
-maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo",
-                 fnl=opval1)
+wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "WDMo")
+maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLMo",
+                 fnl = opval1)
 @ 
 <<ordtest,eval=FALSE,strip.white=all>>=
-wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
-maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")
+wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "WDMo")
+maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLMo")
 @ 
@@
 par(mfrow = c(1, 2), cex = 0.65)
-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo", 
-       fplot=TRUE, ylim=c(0,3.5), xlab="Age group",
-       ylab = "Weighted max statistics", axes=FALSE)
+sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "WDMo", 
+       fplot = TRUE, ylim = c(0,3.5), xlab = "Age group",
+       ylab = "Weighted max statistics", axes = FALSE)
 axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
@@
 
-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo", 
-       fnl=opval1, fplot=TRUE, ylim=c(0,14), xlab="Age group",
+sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLMo", 
+       fnl = opval1, fplot = TRUE, ylim = c(0,14), xlab = "Age group",
        ylab = "LM statistics", axes = FALSE)
@@
 two-factor model with three indicators each (see Figure~\ref{fig:path}).
-\readme{Please make sure the text below is correct.  The previous text stated that the violation occurs in all loadings and all unique variances.}
 The measurement invariance
@@
 magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
-generated and tested. Four statistics were examined above. \readme{What four statistics?  Equation numbers would be helpful.} 
+generated and tested. Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo} and~\eqref{eq:lmuo} were examined. 
 In all conditions, we
@@
 \subsection{Results}
-\readme{When referring to statistics, it would be good to use the notation from Equations 6--8.  For example, instead of ``unordered $\max\mathit{LM}$'', you would have $\max\mathit{LM}_{uo}$.}
+
 Full simulation results are presented in Figure~\ref{fig:sim11res} to 
@@
 All three test statistics exhibited 
-similar results, with unordered  $\max\mathit{LM}$ demonstrating lower power across all situations.
-ordered $\max\mathit{LM}$ and double-max exhibited equivalent power curves because, when only one parameter is tested, the statistics are equivalent.
-\readme{Should say that the unordered statistic is asymptotically equivalent to the LRT.}
+similar results, with $\max\mathit{LM}_{uo}$ demonstrating lower power across all situations. This statistic 
+is asymptotically equivalent to the LRT as what we describe in introduction.
+$\mathit{WDM}_o$  and $\max \mathit{LM}_o$ exhibited equivalent power curves because, when only one parameter is tested, the statistics are equivalent.
 
@@
 simulations, we examine the tests' performance when the model is
-misspecified. \readme{The plots have too many points; maybe we should only plot  every other diff value.  In the xyplot command, you could add the argument ``subset = diff \%\% .5 == 0''.}
+misspecified.
 
@@
 } else {
-  sim1 <- simulation(sim="sim1", nobs=480, parms=c("loading","error","var"))
+  sim1 <- simulation(sim = "sim1", nobs = 480, parms = c("loading", "error", "var"))
   sim1$nlevels <- factor(sim1$nlevels)
-  levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
+  levels(sim1$nlevels) <- paste("m=", levels(sim1$nlevels), sep="")
 
   sim1$nobs <- factor(sim1$nobs)
-  levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
-  levels(sim1$test) <- c("maxLM_O","WDM_O","LM_uO")
+  levels(sim1$nobs) <- paste("n=", levels(sim1$nobs), sep="")
+  levels(sim1$test) <- c("maxLM_O", "WDM_O", "LM_uO")
   save(sim1, file="sim1.rda")
@@
 sim1$test <- factor(as.character(sim1$test),
-  levels = c("ordmax","ordwmax","catdiff"),
-  labels = c("maxLM_o","WDM_o","LM_uo"))
+  levels = c("ordmax", "ordwmax", "catdiff"),
+  labels = c("maxLM_o", "WDM_o", "LM_uo"))
   parlabs <- c(expression(lambda[11]), expression(phi[12]), 
@@
                expression(psi[11] - psi[66]))
-  levels(sim1$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors")  
+  levels(sim1$pars) <- c("Loading1", "Covariance", "Error1", "All Loadings", "All Errors")  
 @
@@
 \begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ test, the double-max test, and the unordered $\max\mathit{LM}$ test across three
-      levels of the ordinal variable $m$ and measurement invariance 
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement invariance is $\lambda_{11}$.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement 
+invariance is $\lambda_{11}$.}
 \label{fig:sim11res}
@@
 <<sim11res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-#sim1.tmp <- sim1[sim1$parms %in% c("loading")& sim1$pars %in% c("Loading1","All Loadings","All Errors"),]
-#sim1.tmp$test <- factor(sim1.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1,
-             subset = (parms=="loading" & 
-                      pars %in% c("Loading1","All Loadings","All Errors")),
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey,
+print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+             subset = (parms == "loading" & 
+                       pars %in% c("Loading1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+             type = "b", xlab = "Violation Magnitude", ylab = "Power", key=mykey,
              strip = function(..., which.given, factor.levels){
-                 if(which.given==2){
-                     strip.default(which.given, factor.levels = parlabs[c(1,4:5)], ...)
+                 if(which.given == 2){
+                     strip.default(which.given, factor.levels = parlabs[c(1, 4:5)], ...)
                  } else {
@@
 \begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the ordered double-max test the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Covariance}
+\caption{Simulated power curves for $\max \mathit{LM}_o$ tests, $\mathit{WDM}_o$ test $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$, and measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$).
+The violation parameter is $\sigma_{12}$}
 \label{fig:sim12res}
@@
 <<sim12res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LM_uo")&sim1$parms %in% c("var") &sim1$pars %in% c("All Loadings", "Covariance","All Errors") ,]
-sim1.tmp$test <- factor(sim1.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+             subset = (parms == "var" & 
+                       pars %in% c("Covariance", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+             type = "b", xlab = "Violation Magnitude", ylab="Power", key = mykey,
+             strip = function(..., which.given, factor.levels){
+                 if(which.given == 2){
+                     strip.default(which.given, factor.levels = parlabs[c(2, 4:5)], ...)
+                 } else {
+                     strip.default(which.given, factor.levels = factor.levels, ...)
+                 }
+             }))
 @
@@
 \begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the ordered double-max test the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Error1}
+\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\psi{11}$}
 \label{fig:sim13res}
@@
 <<sim13res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LM_uo")&sim1$parms %in% c("error") &sim1$pars %in% c("Error1","All Errors","All Loadings"),]
-sim1.tmp$test <- factor(sim1.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+             subset = (parms == "error" & 
+                       pars %in% c("Error1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+             type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+             strip = function(..., which.given, factor.levels){
+                 if(which.given == 2){
+                     strip.default(which.given, factor.levels = parlabs[c(3,4:5)], ...)
+                 } else {
+                     strip.default(which.given, factor.levels = factor.levels, ...)
+                 }
+             }))
 @
@@
 \end{figure}
-
-\seeme{I think it might be nicer to keep these plots all have three rows?(Not sure)}
 
@@
 \subsection{Method}
-\readme{Make sure there is space after each period and comma, the beginning of each sentence is capitalized, etc.}
 Measurement invariance violations could occur in
@@
 Sample size and magnitude of measurement invariance violation were manipulated 
-in the same way as they were in Simulation 1.The tested parameters were also the same as Simulation
+in the same way as they were in Simulation 1. The tested parameters were also the same as Simulation
 1.
@@
 measurement invariance. One can generally observe that 
-tests of the first loading and error variance parameter exhibited high power, reflecting a high Type I error rate.  Tests associated with the factor covariance did not demonstrate this error, however.
-
-\readme{Need some comparison between the unordered and ordered tests.}
+tests of the first loading and error variance parameter exhibited high power, reflecting a high Type I error rate.
+Tests associated with the factor covariance did not demonstrate this error, however.
+In terms of specific statistic performance, $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test demonstrated higher power than $\mathit{LM}_{uo}$ in each panel, especially with increasing levels.
+It is indicating ordered statistics is more inclined to pick up the unmodelled parameter violation in the manifest variables' associated loading and error terms.
+\seeme{Is this sentence enough to compare ordered and unordered statistics?}
 
@@
 measurement invariance, the statistics were generally able to pick up the violation despite the misspecification.
-A similar pattern of specificity and power were observed when the unique variance and factor covariance paraameters violated measurement invariance; the figures associated with these results are shown in the Appendix.
-
-\readme{We should say something about how the unordered test looks better when the unmodeled parameter violates MI.}
+A similar pattern of specificity and power were observed when the unique variance and factor covariance parameters violated measurement invariance; the figures associated with these results are shown in the Appendix.
+Also, the power of ordered statistics $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in each panel.
+
 In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
-as the umodeled parameter does not violate measurment invariance. If the unmodeled parameter 
+as the umodeled parameter does not violate measurement invariance. If the unmodeled parameter 
 dose violate measurement invariance, however, then the tests assign this violation to other 
@@
 } else {
-  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
+  sim2 <- simulation(sim = "sim2", nobs = 480, 
+  parms = c("extra", "extra+loading", "extra+var", "extra+error"))
 
@@
 sim2$test <- factor(as.character(sim2$test),
-  levels = c("ordmax","ordwmax","catdiff"),
-  labels = c("maxLM_o","WDM_o","LM_uo"))
+  levels = c("ordmax", "ordwmax", "catdiff"),
+  labels = c("maxLM_o", "WDM_o", "LM_uo"))
   sim2$nlevels <- factor(sim2$nlevels)
-  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
+  levels(sim2$nlevels) <- paste("m=", levels(sim2$nlevels), sep="")
   sim2$nobs <- factor(sim2$nobs)
-  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
-  levels(sim2$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors")
+  levels(sim2$nobs) <- paste("n=", levels(sim2$nobs), sep="")
+  parlabs <- c(expression(lambda[11]), expression(phi[12]), 
+               expression(psi[11]), expression(lambda[11] - lambda[62]), 
+               expression(psi[11] - psi[66]))
+  levels(sim2$pars) <- c("Loading1", "Covariance", "Error1", "All Loadings", "All Errors")
 
@@
 \begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance 
+\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
 violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
@@
 <<sim21res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra")&sim2$pars %in% c("Loading1","Error1","Covariance"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+             subset = (parms == "extra" & 
+                       pars %in% c("Loading1", "Error1", "Covariance") & diff %% 0.5 == 0),
+             type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+             strip = function(..., which.given, factor.levels){
+                 if(which.given == 2){
+                     strip.default(which.given, factor.levels = parlabs[c(1, 2:3)], ...)
+                 } else {
+                     strip.default(which.given, factor.levels = factor.levels, ...)
+                 }
+             }))
 @
@@
 
-\readme{Need some text describing what this appendix contains.}
+This Appendix demonstrated the results for Simulation 2 when there exists model misspecification (lacking one path from Scale 1 to Math).
+Figure ~\ref{fig:sim22res} to Figure ~\ref{fig:sim24res} showed the violation parameter is first loading $\lambda_{11}$, covariance $\sigma{12}$ and
+first error term $\psi_{11}$, respectively.
 
 \begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance 
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\lambda_{11}$.}
 \label{fig:sim22res}
@@
 <<sim22res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+loading"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+             subset = (parms=="extra+loading" & 
+                       pars %in% c("Loading1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+             type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+             strip = function(..., which.given, factor.levels){
+                 if(which.given == 2){
+                     strip.default(which.given, factor.levels = parlabs[c(1, 4:5)], ...)
+                 } else {
+                     strip.default(which.given, factor.levels = factor.levels, ...)
+                 }
+             }))
 @
@@
 \begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance 
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\sigma_{12}$.}
 \label{fig:sim23res}
@@
 <<sim23res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+var"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+             subset = (parms=="extra+var" & 
+                       pars %in% c("Covariance", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+             type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+             strip = function(..., which.given, factor.levels){
+                 if(which.given == 2){
+                     strip.default(which.given, factor.levels = parlabs[c(2, 4:5)], ...)
+                 } else {
+                     strip.default(which.given, factor.levels = factor.levels, ...)
+                 }
+             }))
 @
@@
 \begin{figure}
-\caption{Simulated power curves for the ordered 
-      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
-      levels of the ordinal variable $m$, and measurement invariance 
-      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\psi_{11}$.}
 \label{fig:sim24res}
@@
 <<sim24res, fig=TRUE, height=7, width=7, echo=FALSE>>=
-sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LM_uo")&sim2$parms %in% c("extra+error"),]
-sim2.tmp$test <- factor(sim2.tmp$test)
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
-             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+             subset = (parms == "extra" & 
+                       pars %in% c("Error1", "All Errors", "All Loadings") & diff %% 0.5 == 0),
+             type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+             strip = function(..., which.given, factor.levels){
+                 if(which.given == 2){
+                     strip.default(which.given, factor.levels = parlabs[c(3, 4:5)], ...)
+                 } else {
+                     strip.default(which.given, factor.levels = factor.levels, ...)
+                 }
+             }))
 @

### 2013-09-26 frontiers/frontiers.Rnw
Frontiers paper: o Revisions to General Discussion o Addition to refs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@268 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$ tests, $\mathit{WDM}_o$ test $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$, and measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$).
-The violation parameter is $\sigma_{12}$}
+The parameter violating measurement invariance is $\sigma_{12}$}
 \label{fig:sim12res}
@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\psi{11}$}
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement invariance is $\psi{11}$}
 \label{fig:sim13res}
@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement invariance is the unmodeled loading.}
 \label{fig:sim21res}
@@
 
+\readme{First, summarize each step of the paper: In this paper, we first described a novel family of test statistics for measurement invariance and illustrated their use via the R package lavaan.
+Next, we examined these statistics' ability to -.
+We found -.}
 In this paper, we examined the recently proposed ordinal statistics specificity property and robustness to model specification to study measurement invariance.
@@
 
-\subsection{Application}
-These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context.
+\readme{Now we need another short paragraph here describing what we will say in the sections below.}
+
+\subsection{Applications}
+
+Many of the applications in this volume, along with many applications in general, focus on testing for measurement invariance across unordered categories such as nations or gender.
+As discussed earlier in this paper, the score-based tests for unordered categories are essentially equivalent to the usual likelihood ratio test.
+Given a measurement invariance violation across these unordered categories, however, researchers typically wish to know why the violation occurred.
+At this point, researchers may examine education level, socioeconomic status, income levels, and so on across the unordered categories.
+These variables are often ordinal or continuous in nature, so that the family of tests described in this paper are applicable.
+This is a first step towards describing why measurement invariance violations occur, as opposed to simply detecting measurement invariance violations.
+\readme{In the frontiers abstracts, one paper was proposed along these lines.
+I will get a draft of that paper and then we can cite it here.}
+The tests are convenient for this purpose, as they do not require a new model to be estimated for each ordinal variable.
+Instead, each ordinal variable defines an ordering of observations, which in turn yields a test statistic that is specific to that ordinal variable.
+
+%% I'm not sure we want to make a big deal of this.  If there are only a small number of ties, we are probably better off with the continuous tests:
+%These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context.
 It is very common to observe several individuals with the same age, IQ, income rank, etc.
@@
 
-Moreover, even though researchers might be originally interested in comparing categorical differences.
+%Moreover, even though researchers might be originally interested in comparing categorical differences.
 For example, researchers might want to test whether there is gender difference or cultural difference for the scale.
@@
 Then the next natural question would be why there is gender or culture differences?
-At this point, researchers might need to examine education level, social economic status, etc, which were often recorded in the ordinal format.
+.
 Thus these test statistics offer an convenient assessing tool to further exploring research hypothesis without making complicated assumptions or fitting multiple different models.
@@
 \subsection{Extension}
-We focused on testing for measurement invariance in factor analysis model in this paper, however, the core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity.
-For example, studying DIF in IRT is immediate.
+In this paper, We focused on testing for measurement invariance in factor analysis models that assume multivariate normality and that are estimated via ML.
+The family of tests described here generally apply to estimation methods that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are potentially applicable to alternative SEM discrepancy functions such as Weighted Least Squares CITE.
+Score calculation for these alternative discrepancy functions has not been implemented (to our knowledge), though the calculation certainly could be implemented.
+Test statistic calculation and inference would then proceed in exactly the same manner as the calculation and inference illustrated in this paper.
+
+In addition to alternative fit functions, the tests can be extended to other models estimated via ML.
+Of primary interest to the topic of measurement invariance, the tests can be extended to item response models to examine differential item functioning.
+In particular, \citeA{strkop13} studied application of these tests to the Rasch model, using them as the basis of a recursive partioning procedure that segments subgroups of individuals who exhibit DIF.
+Further study and extension of these tests for IRT seem warranted.
 
 \subsection{Conclusion}
+\readme{Not sure whether we need this paragraph, but I still need to think about it.}
 In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable.
@@
 This Appendix demonstrated the results for Simulation 2 when there exists model misspecification (lacking one path from Scale 1 to Math).
-Figure ~\ref{fig:sim22res} to Figure ~\ref{fig:sim24res} showed the violation parameter is first loading $\lambda_{11}$, covariance $\sigma{12}$ and
+Figure ~\ref{fig:sim22res} to Figure ~\ref{fig:sim24res} showed the violation parameter is first loading $\lambda_{11}$, covariance $\sigma_{12}$ and
 first error term $\psi_{11}$, respectively.

### 2013-09-27 frontiers/frontiers.Rnw
update discussion part according to Ed's comment.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@269 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 is not likely to be any theoretical justification for 15-year-olds 
-differing from every other age group.
-One solution to this problem would be the {\em approximate invariance} methods proposed by Muth\'{e}n \cite{muthen13}\seeme{How to handle muthen's name in refs.bib file?}.
+differing from every other age group.  One solution to this problem would be the {\em approximate invariance} methods proposed by Muth\'{e}n \cite{muthen13}.
 Alternatively, we can use the proposed family of score-based statistics to obtain tests that are sensitive to the ordering of age.
@@
 
-\readme{First, summarize each step of the paper: In this paper, we first described a novel family of test statistics for measurement invariance and illustrated their use via the R package lavaan.
-Next, we examined these statistics' ability to -.
-We found -.}
-In this paper, we examined the recently proposed ordinal statistics specificity property and robustness to model specification to study measurement invariance.
-As demonstrated in the results, the proposed statistics could select the modeled parameter violating measurement invariance even though the model is misspecified in general.
-
-\readme{Now we need another short paragraph here describing what we will say in the sections below.}
+In this paper, we first described a novel family of test statistics for measurement invariance and illustrated their use via the R package lavaan.
+Next, we examined these statistics' ability to specifically pick up the violation parameter and specificity robustness to model specification.
+We found the proposed statistics could select the modeled parameter violating measurement invariance even though the model is misspecified in general.
+
+In the reminder of the paper, we will first consider the applicability of these tests in exploring the reason of measurement invariance.
+Next, the extension of these tests to other fit functions and more specialized models will be discussed.
+\seeme{Are these two paragraphs OK?}
 
@@
 In this paper, We focused on testing for measurement invariance in factor analysis models that assume multivariate normality and that are estimated via ML.
-The family of tests described here generally apply to estimation methods that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are potentially applicable to alternative SEM discrepancy functions such as Weighted Least Squares CITE.
+The family of tests described here generally apply to estimation methods that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are potentially applicable to alternative SEM discrepancy functions such as Weighted Least Squares \cite{bollen98}.
 Score calculation for these alternative discrepancy functions has not been implemented (to our knowledge), though the calculation certainly could be implemented.

### 2013-10-01 frontiers/frontiers.Rnw
Frontiers revisions: o Revised intro o Updated refs.bib and corrected some errors o Added a few extra readmes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@270 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 In this paper, we consider a family of 
-recently-proposed measurement invariance tests that are based on the {\em scores} of a fitted model.  This family can be used to test for measurement invariance
+recently-proposed measurement invariance tests that are based on the scores of a fitted model.  This family can be used to test for measurement invariance
 w.r.t.\ a continuous auxiliary variable, without pre-specification of subgroups.
@@
 
-Parameter invariance is a fundamental assumption for measurement model in 
-developing psychometric scales. Violation of it could happen in any degree. 
-Generally, researchers want to ignore minor violation parameters, e.g. random 
-normally distributed deviations from a parameter value across groups. BSEM
-(Bayesian Structural Equation Modelling) is proposed not to pick up this kind of 
-violating parameters but only "real" violating ones by assigning priors instead of strictly 
-constraining equality across groups. After identifying the violation parameters, 
-those non-invariant parameters are freely estimated to ensure final correct estimation.
-
-Instead of using this two-step Bayesian analysis procedure, we are applying a family of tests derived
-from stochastic process.
-These tests have been recently proposed to identify specific parameters impacted by measurement invariance.
-However, they have only been used to a limited type of parameter violation.
-It is this paper's intent to provide a detailed examination
-of the tests' performance under practical research scenarios. Specifically, we will examine these tests
-specificity in identifying the violation parameter without giving false positive results.
-Moreover, in practice model misspecification is an inevitable issue to a certain extent, we will demonstrate these tests
-results under model misspecification condition. 
-\seeme{What about this intro? too much BSEM?}
+Many of the papers in this special issue focus on the topic of approximate measurement invariance: we know that strict hypotheses of measurement invariance do not hold exactly across different groups, and this should be reflected in tests of measurement invariance.
+BSEM
+(Bayesian Structural Equation Modeling) generally handles these ideas by replacing across-group equality constraints with informative prior distributions.
+In this paper, we describe an alternative approach: the development of test statistics that are sensitive to invariance violations of interest and insensitive to ``anomalous'' invariance violations.
+The test statistics are specifically applicable to situations where one wishes to test for measurement invariance with respect to an ordinal variable, and they are special cases of a family of tests that may generally be used to study measurement invariance w.r.t.\ continuous, categorical, and ordinal variables.
+
+The family of tests described in this paper have recently been studied in the context of measurement invariance \cite{MerZei13,MerFanZei}, though their practical application has been limited to a small set of simulations and datasets.
+In this paper, we provide a detailed illustration of the tests' use in practice and a detailed examination
+of the tests' performance under scenarios likely to be encountered in practice. 
 %In order to fix the psychometric scale, we need to specify the violating parameter.  
@@
 %of the tests' performance under practical research scenarios. 
-
 In the following section, we first briefly review the 
-theoretical framework of proposed statistical tests and
+theoretical framework of the proposed tests and
 provide a short tutorial illustrating the use of the tests in R.
-Next, we study the tests'
-ability through simulations that mimic practical research. 
-Finally, some suggestion on the tests' use in practice is provided.
+Next, we study the tests' performance in simulations that mimic practical research scenarios. 
+Finally, we provide some further discussion on the tests' use in practice.
 
@@
     \label{eq:dmax}
-    \mathit{DM} & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} \\
+    \mathit{DM} & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} | \\
         \label{eq:cvm}
@@
 
-## remove cases with imputed values(not in 1,...,9)
+## Remove cases with imputed values (not in 1,...,9)
 compcases <- apply(YouthGratitude[,4:28], 1, function(x) all(x %in% 1:9))
@@
 
-Another similar statistics is $\mathit{LM}_\mathit{uo}$ as shown in Equation~ \eqref{eq:lmuo}. We can obtain this statistic by using the following function. 
+To obtain the asymptotically equivalent $\mathit{LM}_\mathit{uo}$ (Equation~\eqref{eq:lmuo}), we can use the following command.
 <<LM_UO>>=
@@
 @
-The above command specifies that we wish to test model parameter numbers 1 through 4, which are the factor loadings supplied by lavaan.
+This command specifies that we wish to test model parameter numbers 1 through 4, which are the factor loadings supplied by lavaan.
 The ordering of parameters can be obtained via the command \verb+coef(reduced)+.
 
-Because our sample size is large, this likelihood ratio test is known to be sensitive to 
+Because our sample size is large, the likelihood ratio test is known to be sensitive to 
 small measurement invariance violations \cite{Ben80}.  That is, the above tests 
@@
 This is demonstrated below, with similar code being useful when one is 
-testing measurement invariance w.r.t.\ truly continuous variables.
+testing for measurement invariance w.r.t.\ truly continuous variables.
 
@@
 for the continuous test statistics.  As mentioned previously,
-calculation of the ``maxLMo'' statistic (Equation~\eqref{eq:maxlmo})
+calculation of the $\max\mathit{LM}_o$ statistic (Equation~\eqref{eq:maxlmo})
 can be lengthy from the need to simulate critical values.
@@
 carried out across categories (ordered or not), as opposed to
-continuous variables.  Thus, in the sections below, we report
+continuous variables.  Thus, in the sections below, we conduct
 novel simulations to study the ordinal statistics' expected behavior
@@
  ($\lambda_{11}$), the unique variance associated with Scale 1
- ($\epsilon_{11}$), or the factor covariance $\sigma_{12}$.  
+ ($\psi_{11}$), or the factor covariance $\sigma_{12}$.  
+\readme{Isn't this five subsets of parameters, not seven?}
  We then tested for measurement invariance in seven subsets of parameters: the three
@@
 The measurement invariance violations began at level $1+m/2$ of $V$
-and were consistent thereafter. These
-violations were based on the auxiliary variable $V$: individuals below
+and were consistent thereafter: individuals below
 the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
@@
 All three test statistics exhibited 
-similar results, with $\max\mathit{LM}_{uo}$ demonstrating lower power across all situations. This statistic 
-is asymptotically equivalent to the LRT as what we describe in introduction.
-$\mathit{WDM}_o$  and $\max \mathit{LM}_o$ exhibited equivalent power curves because, when only one parameter is tested, the statistics are equivalent.
+similar results, with $\max\mathit{LM}_{uo}$ demonstrating lower power across all situations. As described earlier, this statistic 
+is asymptotically equivalent to the LRT. $\mathit{WDM}_o$  and $\max \mathit{LM}_o$ exhibited equivalent power curves because, when only one parameter is tested, the statistics are equivalent.
 
@@
 simulation, the model was correctly specified.  In the following
-simulations, we examine the tests' performance when the model is
+simulation, we examine the tests' performance when the model is
 misspecified.
@@
 Tests associated with the factor covariance did not demonstrate this error, however.
-In terms of specific statistic performance, $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test demonstrated higher power than $\mathit{LM}_{uo}$ in each panel, especially with increasing levels.
-It is indicating ordered statistics is more inclined to pick up the unmodelled parameter violation in the manifest variables' associated loading and error terms.
-\seeme{Is this sentence enough to compare ordered and unordered statistics?}
+In terms of specific statistic performance, $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test demonstrated higher Type I error than $\mathit{LM}_{uo}$ in each panel, especially with increasing levels.
+This is likely because the unmodeled loading's noninvariance was monotonic with $V$; if it were not monotonic, we would expect $\mathit{LM}_{uo}$ to have higher Type I error.
 
@@
 A similar pattern of specificity and power were observed when the unique variance and factor covariance parameters violated measurement invariance; the figures associated with these results are shown in the Appendix.
-Also, the power of ordered statistics $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in each panel.
-
+Also, the power of ordered statistics $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in each of these panels.
+
+\readme{Make sure there is a space after each comma and period:}
 In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
@@
 variance and other loadings associated with the manifest variables that had the unmodeled 
-loading. Thus, prior to testing measurement invariance, it is important to study the extent to
+loading. Thus, as it is with other tests of measurement invariance, it is important to study the extent to
 which the hypothesized model includes all parameters of importance.
@@
 In this paper, we first described a novel family of test statistics for measurement invariance and illustrated their use via the R package lavaan.
-Next, we examined these statistics' ability to specifically pick up the violation parameter and specificity robustness to model specification.
-We found the proposed statistics could select the modeled parameter violating measurement invariance even though the model is misspecified in general.
-
-In the reminder of the paper, we will first consider the applicability of these tests in exploring the reason of measurement invariance.
-Next, the extension of these tests to other fit functions and more specialized models will be discussed.
-\seeme{Are these two paragraphs OK?}
+Next, we examined these statistics' abilities to identify the parameter violating measurement invariance under well-specified and misspecified models.
+We found the proposed statistics could generally isolate the model parameter violating measurement invariance, so long as the violating parameter is included in the model.
+
+In the reminder of the paper, we will first consider the applicability of these tests as compared to traditional tests.
+Next, we discuss test extension to other fit functions and to other specialized models.
 
@@
 In this paper, We focused on testing for measurement invariance in factor analysis models that assume multivariate normality and that are estimated via ML.
-The family of tests described here generally apply to estimation methods that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are potentially applicable to alternative SEM discrepancy functions such as Weighted Least Squares \cite{bollen98}.
+The family of tests described here generally apply to estimation methods that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are potentially applicable to alternative SEM discrepancy functions such as Weighted Least Squares \cite<e.g.,>{broarm95}.
 Score calculation for these alternative discrepancy functions has not been implemented (to our knowledge), though the calculation certainly could be implemented.
@@
 
-\subsection{Conclusion}
-\readme{Not sure whether we need this paragraph, but I still need to think about it.}
-In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable.
+% Removed for now:
+%% \subsection{Conclusion}
+%% \readme{Not sure whether we need this paragraph, but I still need to think about it.}
+%% In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable.
 This
-feature is robust as long as the violation parameter is included in the testing model.
+%% feature is robust as long as the violation parameter is included in the testing model.
 Furthermore, as illustrated in tutorial part, the use of these tests is convenient without changing the focal psychometric model.
@@
 
-This Appendix demonstrated the results for Simulation 2 when there exists model misspecification (lacking one path from Scale 1 to Math).
-Figure ~\ref{fig:sim22res} to Figure ~\ref{fig:sim24res} showed the violation parameter is first loading $\lambda_{11}$, covariance $\sigma_{12}$ and
-first error term $\psi_{11}$, respectively.
+This Appendix includes the results for Simulation 2 when there exists model misspecification (lacking one path from Scale 1 to Math).
+Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim24res} display results when the loading $\lambda_{11}$, covariance $\sigma_{12}$, and
+error term $\psi_{11}$ violate invariance, respectively.
 

### 2013-10-02 frontiers/frontiers.Rnw
update frontier.Rnw file according to Ed's comment. Change formate etc.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@271 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
  ($\psi_{11}$), or the factor covariance $\sigma_{12}$.  
-\readme{Isn't this five subsets of parameters, not seven?}
- We then tested for measurement invariance in seven subsets of parameters: the three
+ We then tested for measurement invariance in five subsets of parameters: the three
 individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
@@
 
-\readme{Make sure there is a space after each comma and period:}
-In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
+In summary, the proposed test statistics appear robust to unmodeled loading parameters, so long
 as the umodeled parameter does not violate measurement invariance.  If the unmodeled parameter 

### 2013-10-02 frontiers/frontiers.Rnw
put spaces after commas.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@272 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor revisions to frontiers ms
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@273 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   Email: merklee@missouri.edu.}
-\shorttitle{Measurement invariance test properties}
-\rightheader{Measurement invariance test properties}
+\shorttitle{Score-based tests in practice}
+\rightheader{Score-based tests in practice}
 
@@
 
-Many of the papers in this special issue focus on the topic of approximate measurement invariance: we know that strict hypotheses of measurement invariance do not hold exactly across different groups, and this should be reflected in tests of measurement invariance.
+Many of the papers in this special issue focus on the topic of approximate measurement invariance: we know that strict hypotheses of measurement invariance do not hold exactly across different groups, and this should be reflected in corresponding tests of measurement invariance.
 BSEM
@@
 %of the tests' performance under practical research scenarios. 
-In the following section, we first briefly review the 
+In the following sections, we first briefly review the 
 theoretical framework of the proposed tests and
@@
 age groups as continuous.  In doing so, we can use the statistics 
-from Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
+from Equations~\eqref{eq:dmax}, ~\eqref{eq:cvm} and~\eqref{eq:maxlm}.  
 That is, we can fit a model whose parameters 
@@
 
-<<Get Estimate>>=
+<<getestimate>>=
 dm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "DM")
@@
 magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
-generated and tested.  Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo} and~\eqref{eq:lmuo} were examined. 
-In all conditions, we
-maintained equal sample sizes at each level of the ordinal variable.
+generated and tested.
+Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo} and~\eqref{eq:lmuo} were examined.
+In all conditions, we maintained equal sample sizes at each level of the ordinal variable.
 
@@
 This is a first step towards describing why measurement invariance violations occur, as opposed to simply detecting measurement invariance violations.
-\readme{In the frontiers abstracts, one paper was proposed along these lines.
+\readme{In the abstracts for the special issue (which the special editors sent us), one paper was proposed along these lines.
 I will get a draft of that paper and then we can cite it here.}

### 2013-10-20 frontiers/frontiers.Rnw
Adapted .Rnw to new strucchange devel-package (1.5-0) waiving the need for specialized code in the tutorial section.
Made results exactly reproducible by setting seeds everywhere where needed.

Modified a few comments/descriptions. Especially Equation 2 really needs
to be scaled properly (was commented out), otherwise the subsequent
equations are not correct.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@275 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \abstract{
-In this paper, we consider a family of 
-recently-proposed measurement invariance tests that are based on the scores of a fitted model.  This family can be used to test for measurement invariance
-w.r.t.\ a continuous auxiliary variable, without pre-specification of subgroups.
-Moreover, the family can be used when one wishes to test for 
-measurement invariance w.r.t.\ an ordinal auxiliary variable, yielding test
-statistics that are sensitive to violations that are monotonically related to
-the ordinal variable (and less sensitive to non-monotonic violations).
-The paper is specifically aimed at potential users of the tests who may wish to know (i) how the tests can be employed for their data, and (ii) whether the tests can accurately identify measurement invariance violations in the presence of model misspecification.
-After providing an overview of the tests, we illustrate their general use via the R packages lavaan and strucchange.
-We then describe two novel simulations that provide evidence of the tests' practical abilities.
-As a whole, the paper provides researchers with the tools and knowledge needed to apply these tests to general measurement invariance scenarios.}
+In this paper, we consider a family of  recently-proposed measurement invariance
+tests that are based on the scores of a fitted model.  This family can be used
+to test for measurement invariance w.r.t.\ a continuous auxiliary variable,
+without pre-specification of subgroups. Moreover, the family can be used when
+one wishes to test for  measurement invariance w.r.t.\ an ordinal auxiliary
+variable, yielding test statistics that are sensitive to violations that are
+monotonically related to the ordinal variable (and less sensitive to
+non-monotonic violations).  The paper is specifically aimed at potential users
+of the tests who may wish to know (i) how the tests can be employed for their
+data, and (ii) whether the tests can accurately identify measurement invariance
+violations in the presence of model misspecification.  After providing an
+overview of the tests, we illustrate their general use via the \proglang{R} packages \pkg{lavaan}
+and \pkg{strucchange}.  We then describe two novel simulations that provide evidence
+of the tests' practical abilities.  As a whole, the paper provides researchers
+with the tools and knowledge needed to apply these tests to general measurement
+invariance scenarios.
+}
 
@@
 \newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+\let\proglang=\textsf
+\let\pkg=\emph
+\let\code=\texttt
+
 %% for internal use
@@
 \include{Sweave}
-<<preliminaries1,echo=FALSE>>=
+\setkeys{Gin}{width=\textwidth}
+<<preliminaries, echo=FALSE>>=
 ## packages
 library("lavaan")
-library("strucchange")
+library("strucchange") ## at least 1.5-0
 library("mvtnorm")
@@
 source("mz.R")
-source("../www/efpFunctional-cat.R")
-source("../pkg/sctest.lavaan.R")
 source("../www/mz-ordinal.R")
-source("../www/estfun-lavaan.R")
-
-## turn off stars
-options(show.signif.stars = FALSE)
-
-## convenience function for plotting boundaries without color
-get_boundary <- function(obj, fun, trim = 0) {
-  bound <- fun$boundary(0:obj$nobs/obj$nobs)
-  bound <- fun$computeCritval(0.05, ncol(obj$process)) * bound
-  bound <- zoo(bound, time(obj$process))
-  if(trim > 0) bound <- head(tail(bound, - floor(trim * obj$nobs)), - floor(trim * obj$nobs))
-  return(bound)
-}
-
-## convenience function for applying tests to real data
-info_full <- function(x, ...) solve(vcov(x) * nobs(x))
-@
-
-Many of the papers in this special issue focus on the topic of approximate measurement invariance: we know that strict hypotheses of measurement invariance do not hold exactly across different groups, and this should be reflected in corresponding tests of measurement invariance.
-BSEM
-(Bayesian Structural Equation Modeling) generally handles these ideas by replacing across-group equality constraints with informative prior distributions.
-In this paper, we describe an alternative approach: the development of test statistics that are sensitive to invariance violations of interest and insensitive to ``anomalous'' invariance violations.
-The test statistics are specifically applicable to situations where one wishes to test for measurement invariance with respect to an ordinal variable, and they are special cases of a family of tests that may generally be used to study measurement invariance w.r.t.\ continuous, categorical, and ordinal variables.
-
-The family of tests described in this paper have recently been studied in the context of measurement invariance \cite{MerZei13,MerFanZei}, though their practical application has been limited to a small set of simulations and datasets.
-In this paper, we provide a detailed illustration of the tests' use in practice and a detailed examination
-of the tests' performance under scenarios likely to be encountered in practice. 
+
+## some options
+options(prompt = "R> ", continue = "+  ", digits = 3,
+  useFancyQuotes = FALSE, show.signif.stars = FALSE)
+@
+
+Many of the papers in this special issue focus on the topic of approximate
+measurement invariance: we know that strict hypotheses of measurement invariance
+do not hold exactly across different groups, and this should be reflected in
+corresponding tests of measurement invariance.  BSEM (Bayesian structural
+equation modeling) generally handles these ideas by replacing across-group
+equality constraints with informative prior distributions.  In this paper, we
+describe an alternative approach: the development of test statistics that are
+sensitive to invariance violations of interest and insensitive to ``anomalous''
+invariance violations.  The test statistics are specifically applicable to
+situations where one wishes to test for measurement invariance with respect to
+an ordinal variable, and they are special cases of a family of tests that may
+generally be used to study measurement invariance w.r.t.\ continuous,
+categorical, and ordinal variables.
+
+The family of tests described in this paper have recently been studied in the
+context of measurement invariance \cite{MerZei13,MerFanZei}, though their
+practical application has been limited to a small set of simulations and
+datasets.  In this paper, we provide a detailed illustration of the tests' use
+in practice and a detailed examination  of the tests' performance under
+scenarios likely to be encountered in practice. 
 %In order to fix the psychometric scale, we need to specify the violating parameter.  
@@
 %of the tests' performance under practical research scenarios. 
-In the following sections, we first briefly review the 
-theoretical framework of the proposed tests and
-provide a short tutorial illustrating the use of the tests in R.
-Next, we study the tests' performance in simulations that mimic practical research scenarios. 
-Finally, we provide some further discussion on the tests' use in practice.
+In the following sections, we first briefly review the  theoretical framework of
+the proposed tests and provide a short tutorial illustrating the use of the
+tests in \proglang{R} \cite{R11}. Next, we study the tests' performance in simulations that mimic
+practical research scenarios.  Finally, we provide some further discussion on
+the tests' use in practice.
 
 \section{Background}
-This section contains background and discussion of the 
-proposed statistics as applied to SEM; for a more detailed account, see
-\citeA{MerZei13} and \citeA{MerFanZei}.  For details on the
-statistics' application to general statistical models, see
-\citeA{zeihor07}.
+
+This section contains background and discussion of the  proposed statistics as
+applied to SEM; for a more detailed account, see \citeA{MerZei13} and
+\citeA{MerFanZei}.  For details on the statistics' application to general
+statistical models, see \citeA{zeihor07}.
 
@@
 
-Following model estimation, the tests primarily work on the {\em
-  scores} of the fitted model; these are defined as
+Following model estimation, the tests primarily work on the \emph{scores}
+of the fitted model; these are defined as
 %% TODO indicate that this is evaluated at theta hat
@@
 \label{eq:score}
-s(\hat{\theta};x_{i}) = \left ( \frac{\partial
-      \ell(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
-      \ell(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
+s(\theta;x_{i}) = \left (
+  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{1}}, \dots,
+  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right )^\top,
 \end{equation} 
 where $\ell(\theta; x_i)$ is the likelihood associated with individual
-$i$, $\theta$ is a $k$-dimensional parameter vector, and
-$\hat{\theta}$ is the maximum likelihood estimate.  The cross-product
+$i$, $\theta$ is a $k$-dimensional parameter vector. The corresponding
+maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
+$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$. The cross-product
 of these scores forms the ``meat'' for the calculation of robust
@@
 invariance holds, then all 
-individuals' scores will fluctuate around zero.
-
-To formalize the ideas in the previous paragraph, we define a cumulative
-sum of the ordered scores.  This cumulative sum may be written as
+individuals' scores will fluctuate randomly around zero.
+
+To formalize the ideas in the previous paragraph, we define a
+suitably scaled cumulative sum of the ordered scores.  This may be written as
 \begin{equation} 
     \label{eq:cumscore}
-  {\bm B}(t; \hat {\bm \theta}) ~=~ %\hat {\bm I}^{-1/2} n^{-1/2}
+  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
     \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
@@
 operator) and $x_{(i)}$ reflects the individual with the
-$i^{\text{th}}$-smallest value of the auxiliary variable.
+$i$-th smallest value of the auxiliary variable.
 We specifically focus on how the
-cumulative sum fluctuates as more individuals' scores are added to it,
+cumulative sum fluctuates as more individuals' scores are added to it, e.g.,
 starting with the youngest and ending with the oldest individual
-(assuming that age is the auxiliary variable of interest).  
-The summation is also typically premultiplied by the inverse square
+if age is the auxiliary variable of interest.
+The summation is premultiplied by an estimate of the inverse square
 root of the information matrix, which serves to decorrelate the
-fluctuation processes associated with individual model parameters 
-(i.e., the process associated with one parameter is not correlated 
-with the processes associated with other parameters).
+fluctuation processes associated with individual model parameters.
+Thus, the process associated with one parameter is not correlated 
+with the processes associated with other parameters.
 
@@
 of interest and take the maximum or the average across individuals, resulting in a 
-{\em maximum Lagrange multiplier} statistic and Cram\'{e}r-von Mises statistic, respectively 
+{\em maximum Lagrange multiplier} statistic and \emph{Cram\'{e}r-von Mises} statistic, respectively 
 \cite<see>[for further discussion]{MerZei13}.  These statistics are given by
-
 \begin{eqnarray}
@@
 \end{eqnarray}
-
 Critical values associated with $\mathit{DM}$ can be obtained analytically, 
@@
 This issue should not be important to the user, as critical values are obtained 
-directly from the R implementation described later.
+directly from the \proglang{R} implementation described later.
 
@@
 This simulation is somewhat computationally intensive and, in practice, takes about 
-10 minutes on the authors' computers.  However, the wait is often worth it, 
+10 minutes on the authors' computers when 50,000 replications are sampled from the 
+approximate asymptotic distribution.  However, the wait is often worth it, 
 as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic 
@@
 
-Finally, if $V$ is only nominal/categorical, the cumulative sums of scores can be used to obtain the a Lagrange multiplier statistic.
+Finally, if $V$ is only nominal/categorical, the cumulative sums of scores can be used to obtain a Lagrange multiplier statistic.
 This test statistic can be formally written as
@@
 and it is asymptotically equivalent to the usual, likelihood ratio test statistic.
-It is advantageous over the likelihood ratio test because it requires estimation of only one model (the restricted model).
-We make use of this advantage in the simulations, described later.
+It is advantageous over the likelihood ratio test because it requires estimation of
+only one model (the restricted model).  We make use of this advantage in the
+simulations, described later.
 
@@
 In this section, we demonstrate how the above tests can be carried out in~R, 
-using the package lavaan \cite{lavaan11} for model estimation and strucchange 
+using the package \pkg{lavaan} \cite{lavaan11} for model estimation and \pkg{strucchange} 
 \cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
-the applicability of adult gratitude scales to youth, available in the~R 
-package psychotools \cite{zeilei12}.  The data consist of responses to three adult 
+the applicability of adult gratitude scales to youth, available in the~\proglang{R} 
+package \emph{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
 gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
@@
 (though the tests can be applied to incomplete data).
+%
 <<data>>=
-## Reading in the data
-data("YouthGratitude",package="psychotools")
-
-## Remove cases with imputed values (not in 1,...,9)
-compcases <- apply(YouthGratitude[,4:28], 1, function(x) all(x %in% 1:9))
-yg <- YouthGratitude[compcases,]
-@
-Next, we fit two models in lavaan: a one-factor model where loadings are restricted 
+data("YouthGratitude", package = "psychotools")
+compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
+yg <- YouthGratitude[compcases, ]
+@
+%
+Next, we fit two models in \pkg{lavaan}: a one-factor model where loadings are restricted 
 to be equal across age groups, and a one-factor model where loadings are free across age groups.
-<<modest>>=
-## Restricted model
-reduced <- cfa("f1 = ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+%
+<<models>>=
+restr <- cfa("f1 = ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
   data = yg, group = "agegroup", meanstructure = TRUE,
   group.equal = "loadings")
-
-## Unrestricted model
 full <- cfa("f1= ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
@@
 @
-Finally, we can get the results of a likelihood ratio test via the \verb+anova()+ 
+%
+Finally, we can get the results of a likelihood ratio test via the \code{anova()}
 function, which implies that the GQ-6 violates measurement invariance.
 <<LRT>>=
-## LRT
-anova(full, reduced)
-@
-
-To obtain the asymptotically equivalent $\mathit{LM}_\mathit{uo}$ (Equation~\eqref{eq:lmuo}), we can use the following command.
+anova(full, restr)
+@
+To obtain the asymptotically equivalent $\mathit{LM}_\mathit{uo}$ (Equation~\eqref{eq:lmuo}), we
+can use the \code{sctest()} function from \pkg{strucchange}:
 <<LM_UO>>=
-## unordered LM
-lm_uo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "LMuo")
-round(lm_uo$p.value, 4)
-@
-This command specifies that we wish to test model parameter numbers 1 through 4, which are the factor loadings supplied by lavaan.
-The ordering of parameters can be obtained via the command \verb+coef(reduced)+.
+sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "LMuo")
+@
+This command specifies that we assess the parameter 1--4 of
+model \code{restr} after ordering the observations along the \code{agegroup} and
+the observed information matrix is used as the variance-covariance matrix. Note
+that the model parameters 1--4 are the factor loadings supplied by \pkg{lavaan}
+as can be seen by inspecting \code{coef(restr)}. This also leads to somewhat
+smaller test statistics that is very close to being significant at the 5\% level.
 
@@
 is not likely to be any theoretical justification for 15-year-olds 
-differing from every other age group.  One solution to this problem would be the {\em approximate invariance} methods proposed by Muth\'{e}n \cite{muthen13}.
-Alternatively, we can use the proposed family of score-based statistics to obtain tests that are sensitive to the ordering of age.
+differing from every other age group.  One solution to this problem would be the
+\emph{approximate invariance} methods proposed by Muth\'{e}n \cite{muthen13}.
+Alternatively, we can use the proposed family of score-based statistics to obtain
+tests that are sensitive to the ordering of age.
 
 \subsection{Continuous Treatment}
+
 If we are interested in measurement invariance violations that are 
@@
 how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
-age (where age ties are broken arbitrarily).
+age where age ties are broken arbitrarily (using the original order
+of the observations within each age group).
 This is demonstrated below, with similar code being useful when one is 
@@
 
-To proceed, we focus only on the restricted model that was fit in 
-the previous section (stored in the object \verb+reduced+).  We obtain scores 
-associated with the factor loadings from this model, then test the 
-fluctuation of these scores across \verb+agegroup+.  
-As demonstrated below, the score computation and testing can be efficiently carried out using \verb+sctest.lavaan()+.
-%% Ting: I removed this because the model was already fit in the
-%%       previous section
-%%<<theta estimate>>=
-%%theta<-cfa(
-%%  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
-%%  data=yg,group="agegroup",meanstructure=TRUE,
-%%  group.equal="loadings")
-%%@
-
-<<getestimate>>=
-dm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "DM")
-cvm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "CvM")
-maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLM")
-round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
-@
-%% move this code description to the cat2BB part. 
-%In the code above, the \verb+parm=1:4+ argument means that 
-%we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
-%which are supplied by lavaan.  The ordering of parameters can be obtained via the 
-%command \verb+coef(reduced)+.  
-We see that two of the three $p$-values output at 
-the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).
-
-%% TODO Illustrate with graphs?
-
-%% THE FOLLOWING HAS BEEN MOVED
-%% \begin{figure}
-%% \caption{Get Continuous Estimate DM Alternative Way}
-%% \label{fig:conti1res}
-%% \setkeys{Gin}{width=\textwidth}
-%% <<conti1res, fig=TRUE, height=7, width=7>>=
-%% DM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
-%% DM
-%% @
-%% \end{figure}
-
-%% \begin{figure}
-%% \caption{Get Continuous Estimate CvM Alternative Way}
-%% \label{fig:conti2res}
-%% \setkeys{Gin}{width=\textwidth}
-%% <<conti2res, fig=TRUE, height=7, width=7>>=
-%% CvM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
-%% CvM
-%% @
-%% \end{figure}
-
-%% <<conti3resmaxLM>>=
-%% maxLM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
-%% maxLM
-%% @
+Again, we employ the \code{sctest()} function to assess the parameters~1--4
+from the restricted model \code{restr} after ordering w.r.t.\ \code{agegroup}:
+%
+<<sctest-cont>>=
+dm <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "DM")
+cvm <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "CvM")
+maxlm <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "maxLM")
+c(dm$p.value, cvm$p.value, maxlm$p.value)
+@
+%
+We see that two of the three $p$-values output at the end of the code are
+larger than that associated with the LRT (with the CvM statistic being non-significant).
+
 The tests carried out here assume a unique ordering of individuals by age, but 
 this is obviously not the case.  To compute the statistics and $p$-values, 
-the strucchange package implicitly assumed an arbitrary ordering of 
+the \pkg{strucchange} package implicitly employed the (arbitrary) ordering of 
 individuals who are tied on age.  If we were to change this ordering, 
-the resulting statistics and $p$-values would also change, which is obviously problematic.  
+the resulting statistics and $p$-values would also change, potentially
+switching significant results to being non-significant and vice versa.
+Clearly, this is problematic.  
 To accurately account for the multiple observations  
@@
 The main difference between the ordinal test statistics and their
-continuous counterparts is that the ordinal statistics do not require
-an ordering of individuals  
-within the same age group.  To compute the test statistics, we allow
+continuous counterparts is that the ordinal statistics are unchanged
+when re-ordering individuals within the same age group.  To compute the test statistics, we allow
 the scores of all tied individuals to enter the cumulative sum
@@
 modified critical values and test statistics that are sensitive to
-measurement invariance violations that are monotonic with age group.
+measurement invariance violations that are monotonic w.r.t.\ age group.
 %% Therefore, this ordinal test should 
@@
 %% scenes without confusing the reader.
-<<critvals,echo=FALSE,strip.white=all>>=
-if(file.exists("opval1.rda")){
-  load("opval1.rda")
+<<sctest-ord1, echo=FALSE, results=hide>>=
+set.seed(1090)
+wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "WDMo")
+
+if(file.exists("maxLMo.rda")){
+  load("maxLMo.rda")
 } else {
-  gefp1 <- gefp(reduced, fit = NULL, order.by = as.numeric(yg$agegroup),
-                vcov = info_full, sandwich = FALSE, parm = 1:4)
-  opval1 <- ordL2BB(gefp1)
-  save(opval1, file="opval1.rda")
+  RNGkind(kind = "default", normal.kind = "default")
+  set.seed(1090)
+  maxLMo <- ordL2BB(yg$agegroup, nproc = 4)
+  save(maxLMo, file = "maxLMo.rda")
 }
-## Get test statistics using fnl=opval1 without showing it in the paper
-wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "WDMo")
-maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLMo",
-                 fnl = opval1)
-@ 
-<<ordtest,eval=FALSE,strip.white=all>>=
-wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "WDMo")
-maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLMo")
-@ 
-<<pval,strip.white=all>>=
-round(c(wdmo$p.value, maxlmo$p.value), 2)
+maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = maxLMo)
+@ 
+<<sctest-ord2, eval=FALSE,strip.white=all>>=
+wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "WDMo")
+maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "maxLMo")
+@ 
+<<sctest-ord3,strip.white=all>>=
+c(wdmo$p.value, maxlmo$p.value)
 @ 
 In computing the ordinal test statistics, we
-obtain~$p=\Sexpr{round(wdmo$p.value,2)}$
-and~$p=\Sexpr{round(maxlmo$p.value,2)}$, respectively.
+obtain~$p=\Sexpr{round(wdmo$p.value, 3)}$
+and~$p=\Sexpr{round(maxlmo$p.value, 3)}$, respectively.\footnote{To replicate
+both $p$-values exactly, the \proglang{R}'s random seed needs to be
+set by \code{set.seed(1090)} prior to each \code{sctest()} call.}
 Both $p$-values are clearly larger than that of the 
@@
 
-In addition to test statistics, ``instability plots'' can be generated by using the \verb+fplot=TRUE+ argument within \verb+sctest.lavaan()+.
-Plots representing the statistics' fluctuations across levels of age group are displayed in Figure~\ref{fig:ordres}.
-The left panel displays the process associated with $\mathit{WDM}_o$, and the right panel displays the process associated with $\max\mathit{LM}_o$.
-%The following sentence is not correct for all statistics:  
-%In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs.
-greater than $i$, and the null hypothesis is rejected if the maximum of the statistics is larger than its 5\% critical value(visualized by the horizontal red line).
-These plots display the loadings' fluctuation across age groups, with the x-axis reflecting age group and the y-axis reflecting test statistic values (larger values reflect more instability).
-The red lines reflect critical values, so that the hypothesis of measurement invariance is rejected if a black line crosses the red line.
-While the measurement invariance tests are non-significant, the plots imply some instability in the older age groups (15 and 16).
-
-%% gefp1<gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
-%%        vcov=info_full,sandwich=FALSE,parm=1:4)
-%% pval.ord=rep(NA,2)
-%% pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value
-
-%%pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
-%% pval.ord
-%% @
-
-%Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
+In addition to test statistics, ``instability plots'' can be generated by additionaly setting
+\code{plot = TRUE} in the \code{sctest()} calls above. The resulting plots representing
+the statistics' fluctuations across levels of age group are displayed in
+Figure~\ref{fig:ordres}.
+These plots display the loadings' fluctuation across age groups, with the x-axis
+reflecting age group and the y-axis reflecting test statistic values (larger
+values reflect more instability).  The dashed horizontal lines reflect critical values, so
+that the hypothesis of measurement invariance is rejected if the sequence of
+test statistics crosses the critical value.  While the measurement invariance tests are
+non-significant, the plots imply some instability in the older age groups (15 and 16).
+
 \begin{figure}
@@
 \label{fig:ordres}
-\setkeys{Gin}{width=\textwidth}
-<<ordres, fig=TRUE, height=3, width=6, echo=FALSE, results=hide>>=
-par(mfrow = c(1, 2), cex = 0.65)
-sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "WDMo", 
-       fplot = TRUE, ylim = c(0,3.5), xlab = "Age group",
-       ylab = "Weighted max statistics", axes = FALSE)
-axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
-box()
-
-sctest(reduced, as.numeric(yg$agegroup), parm = 1:4, stat = "maxLMo", 
-       fnl = opval1, fplot = TRUE, ylim = c(0,14), xlab = "Age group",
-       ylab = "LM statistics", axes = FALSE)
-axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
-box()
+<<sctest-ord-graphics, fig=TRUE, height=4.5, width=9, echo=FALSE, results=hide>>=
+par(mfrow = c(1, 2))
+set.seed(1090)
+sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "WDMo", plot = TRUE, ylim = c(0, 3.2),
+  xlab = "Age group", ylab = "Weighted max statistics",
+  main = expression(WDM[o]), boundary = list(col = 1, lty = 2))
+sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = maxLMo, plot = TRUE, ylim = c(0, 14),
+  xlab = "Age group", ylab = "LM statistics",
+  main = expression(maxLM[o]), boundary = list(col = 1, lty = 2))
 @
@@
 
-%% TODO Describe graphs
-%% Is there some argument for fplot=TRUE so user can change  axis-name and critical value line
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 In the above sections, we have illustrated the score-based tests'
-computation in R.  We suspect that the ordinal tests will be most
+computation in \proglang{R}.  We suspect that the ordinal tests will be most
 popular with users, because measurement invariance tests are typically
@@
 \label{fig:sim11res}
-\setkeys{Gin}{width=\textwidth}
-<<sim11res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+<<sim11res, fig=TRUE, height=9, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
        subset = (parms == "loading" & 
@@
 	   }
-             }))
+       })
 @
@@
 \label{fig:sim12res}
-\setkeys{Gin}{width=\textwidth}
-<<sim12res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+<<sim12res, fig=TRUE, height=9, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
        subset = (parms == "var" & 
@@
 \label{fig:sim13res}
-\setkeys{Gin}{width=\textwidth}
-<<sim13res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+<<sim13res, fig=TRUE, height=9, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
        subset = (parms == "error" & 
@@
 \label{fig:sim21res}
-\setkeys{Gin}{width=\textwidth}
-<<sim21res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+<<sim21res, fig=TRUE, height=9, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
        subset = (parms == "extra" & 
@@
 
-In this paper, we first described a novel family of test statistics for measurement invariance and illustrated their use via the R package lavaan.
+In this paper, we first described a novel family of test statistics for measurement invariance and illustrated their use via the \proglang{R} package \pkg{lavaan}.
 Next, we examined these statistics' abilities to identify the parameter violating measurement invariance under well-specified and misspecified models.
@@
 
-All results were obtained using the {R}~system for statistical computing \cite{R11},
+All results were obtained using the \proglang{R}~system for statistical computing \cite{R11},
 version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on package
-{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} for fitting 
+\pkg{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} for fitting 
 of the factor analysis models and
-{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
-for evaluating the parameter instability tests. {R}~and the packages {lavaan} and 
-{strucchange} are
+\pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
+for evaluating the parameter instability tests. {R}~and the packages \pkg{lavaan} and 
+\pkg{strucchange} are
 freely available under the General Public License~2 from the
-Comprehensive {R} Archive Network at \url{http://CRAN.R-project.org/}.
-{R}~code for replication of our results is
+Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/}.
+\proglang{R}~code for replication of our results is
 available at \url{http://semtools.R-Forge.R-project.org/}.
@@
 \label{fig:sim22res}
-\setkeys{Gin}{width=\textwidth}
-<<sim22res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+<<sim22res, fig=TRUE, height=9, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
        subset = (parms=="extra+loading" & 
@@
 \label{fig:sim23res}
-\setkeys{Gin}{width=\textwidth}
-<<sim23res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+<<sim23res, fig=TRUE, height=9, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
        subset = (parms=="extra+var" & 
@@
 \label{fig:sim24res}
-\setkeys{Gin}{width=\textwidth}
-<<sim24res, fig=TRUE, height=7, width=7, echo=FALSE>>=
+<<sim24res, fig=TRUE, height=9, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-print(xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
        subset = (parms == "extra" & 

### 2013-10-20 frontiers/frontiers.Rnw
small improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@276 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 To use these scores for testing, we order individuals according to
-an auxiliary variable (the variable against which we are testing 
+an auxiliary variable $V$ (the variable against which we are testing 
 measurement invariance) and look for ``trends'' in the
@@
 operator) and $x_{(i)}$ reflects the individual with the
-$i$-th smallest value of the auxiliary variable.
+$i$-th smallest value of the auxiliary variable $V$.
 We specifically focus on how the
@@
 
-Finally, if $V$ is only nominal/categorical, the cumulative sums of scores can be used to obtain a Lagrange multiplier statistic.
+Finally, if the auxiliary variable $V$ is only nominal/categorical, the cumulative
+sums of scores can be used to obtain a Lagrange multiplier statistic.
 This test statistic can be formally written as
@@
 17 magnitudes of invariance violations. 
-The measurement invariance violations began at level $1+m/2$ of $V$
-and were consistent thereafter: individuals below
+The measurement invariance violations began at level $1+m/2$ of the
+auxiliary variable $V$ and were consistent thereafter: individuals below
 the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
@@
 magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
-generated and tested.
-Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo} and~\eqref{eq:lmuo} were examined.
-In all conditions, we maintained equal sample sizes at each level of the ordinal variable.
+generated and tested.  Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo}
+and~\eqref{eq:lmuo} were examined. In all conditions, we maintained equal sample
+sizes in each subgroup of the ordinal variable.
 
@@
 invariance, respectively. 
-All three test statistics exhibited 
+All three test statistics exhibit
 similar results, with $\max\mathit{LM}_{uo}$ demonstrating lower power across all situations.  As described earlier, this statistic 
@@
 the parameter violating measurement invariance.  Additionally, the tests 
-have somewhat-higher power to detect measurement invariance violations 
+have somewhat higher power to detect measurement invariance violations 
 in the factor loading and factor covariance parameters, as opposed to the error variance
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
+\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and
+$\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
 violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement 
@@
 		 pars %in% c("Loading1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key=mykey,
+       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ tests, $\mathit{WDM}_o$ test $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$, and measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$).
-The parameter violating measurement invariance is $\sigma_{12}$}
+\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and
+$\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$, and measurement invariance
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement
+invariance is $\sigma_{12}$.}
 \label{fig:sim12res}
@@
 		 pars %in% c("Covariance", "All Loadings", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab="Power", key = mykey,
+       type = "b", xlab = "Violation Magnitude", ylab="Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 		 pars %in% c("Error1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 \subsection{Results}
-Results are presented by the parameter that violated measurement invariance.
-Of primary interest is Figure~\ref{fig:sim21res}, which displays results when the unmodeled loading violates 
-measurement invariance.  One can generally observe that 
-tests of the first loading and error variance parameter exhibited high power, reflecting a high Type I error rate.
-Tests associated with the factor covariance did not demonstrate this error, however.
-In terms of specific statistic performance, $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test demonstrated higher Type I error than $\mathit{LM}_{uo}$ in each panel, especially with increasing levels.
-This is likely because the unmodeled loading's noninvariance was monotonic with $V$; if it were not monotonic, we would expect $\mathit{LM}_{uo}$ to have higher Type I error.
-
-When the modeled factor loading, $\lambda_{11}$, violated
-measurement invariance, the statistics were generally able to pick up the violation despite the misspecification.
-A similar pattern of specificity and power were observed when the unique variance and factor covariance parameters violated measurement invariance; the figures associated with these results are shown in the Appendix.
-Also, the power of ordered statistics $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in each of these panels.
+Results are presented by the parameter that violated measurement invariance. Of
+primary interest is Figure~\ref{fig:sim21res}, which displays results when the
+unmodeled loading violates  measurement invariance.  One can generally observe
+that  tests of the first loading and error variance parameter exhibited high
+power, reflecting a high Type I error rate.  Tests associated with the factor
+covariance did not demonstrate this error, however. In terms of specific
+statistic performance, $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test
+demonstrated higher Type I error than $\mathit{LM}_{uo}$ in each panel,
+especially with increasing levels.  This is likely because the unmodeled
+loading's non-invariance was monotonic with $V$; if it were not monotonic, we
+would expect $\mathit{LM}_{uo}$ to have higher Type I error.
+
+When the modeled factor loading, $\lambda_{11}$, violated measurement
+invariance, the statistics were generally able to pick up the violation despite
+the misspecification. A similar pattern of specificity and power were observed
+when the unique variance and factor covariance parameters violated measurement
+invariance; the figures associated with these results are shown in the Appendix.
+Also, the power of ordered statistics $\max \mathit{LM}_o$ test and
+$\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in
+each of these panels.
 
@@
 		 pars %in% c("Loading1", "Error1", "Covariance") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 
-Many of the applications in this volume, along with many applications in general, focus on testing for measurement invariance across unordered categories such as nations or gender.
-As discussed earlier in this paper, the score-based tests for unordered categories are essentially equivalent to the usual likelihood ratio test.
-Given a measurement invariance violation across these unordered categories, however, researchers typically wish to know why the violation occurred.
-At this point, researchers may examine education level, socioeconomic status, income levels, and so on across the unordered categories.
-These variables are often ordinal or continuous in nature, so that the family of tests described in this paper are applicable.
-This is a first step towards describing why measurement invariance violations occur, as opposed to simply detecting measurement invariance violations.
-\readme{In the abstracts for the special issue (which the special editors sent us), one paper was proposed along these lines.
-I will get a draft of that paper and then we can cite it here.}
-The tests are convenient for this purpose, as they do not require a new model to be estimated for each ordinal variable.
-Instead, each ordinal variable defines an ordering of observations, which in turn yields a test statistic that is specific to that ordinal variable.
+Many of the applications in this volume, along with many applications in
+general, focus on testing for measurement invariance across unordered categories
+such as nations or gender.  As discussed earlier in this paper, the score-based
+tests for unordered categories are essentially equivalent to the usual
+likelihood ratio test.  Given a measurement invariance violation across these
+unordered categories, however, researchers typically wish to know why the
+violation occurred.  At this point, researchers may examine education level,
+socioeconomic status, income levels, and so on across the unordered categories. 
+These variables are often ordinal or continuous in nature, so that the family of
+tests described in this paper are applicable.  This is a first step towards
+describing why measurement invariance violations occur, as opposed to simply
+detecting measurement invariance violations. \readme{In the abstracts for the
+special issue (which the special editors sent us), one paper was proposed along
+these lines.  I will get a draft of that paper and then we can cite it here.}
+The tests are convenient for this purpose, as they do not require a new model to
+be estimated for each ordinal variable.  Instead, each ordinal variable defines
+an ordering of observations, which in turn yields a test statistic that is
+specific to that ordinal variable.
 
@@
 \subsection{Extension}
-In this paper, We focused on testing for measurement invariance in factor analysis models that assume multivariate normality and that are estimated via ML.
-The family of tests described here generally apply to estimation methods that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are potentially applicable to alternative SEM discrepancy functions such as Weighted Least Squares \cite<e.g.,>{broarm95}.
-Score calculation for these alternative discrepancy functions has not been implemented (to our knowledge), though the calculation certainly could be implemented.
-Test statistic calculation and inference would then proceed in exactly the same manner as the calculation and inference illustrated in this paper.
-
-In addition to alternative fit functions, the tests can be extended to other models estimated via ML.
-Of primary interest to the topic of measurement invariance, the tests can be extended to item response models to examine differential item functioning.
-In particular, \citeA{strkop13} studied application of these tests to the Rasch model, using them as the basis of a recursive partioning procedure that segments subgroups of individuals who exhibit DIF.
-Further study and extension of these tests for IRT seem warranted.
+
+In this paper, We focused on testing for measurement invariance in factor
+analysis models that assume multivariate normality and that are estimated via
+maximum likelihood (ML).  The family of tests described here generally apply to estimation methods
+that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are
+potentially applicable to alternative SEM discrepancy functions such as weighted
+least squares \cite<e.g.,>{broarm95}.  Score calculation for these alternative
+discrepancy functions has not been implemented (to our knowledge), though the
+calculation certainly could be implemented.  Test statistic calculation and
+inference would then proceed in exactly the same manner as the calculation and
+inference illustrated in this paper.
+
+In addition to alternative fit functions, the tests can be extended to other
+models estimated via ML.  Of primary interest to the topic of measurement
+invariance, the tests can be extended to item response models to examine
+differential item functioning.  In particular, \citeA{strkop13} studied
+application of these tests to the Rasch model, using them as the basis of a
+recursive partioning procedure that segments subgroups of individuals who
+exhibit DIF.  Further study and extension of these tests for IRT seem warranted.
 
@@
 		 pars %in% c("Loading1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 		 pars %in% c("Covariance", "All Loadings", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 		 pars %in% c("Error1", "All Errors", "All Loadings") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey,
+       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){

### 2013-10-21 frontiers/frontiers.Rnw
change according to Achim comments about results section n and sigma parameter confusion.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@278 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
  ($\lambda_{11}$), the unique variance associated with Scale 1
- ($\psi_{11}$), or the factor covariance $\sigma_{12}$.  
+ ($\psi_{11}$), or the factor covariance $\phi_{12}$.  
  We then tested for measurement invariance in five subsets of parameters: the three
@@
 factor loading $\lambda_{11}$, with panels reflecting the parameters 
-being tested and lines reflecting $n$.
+being tested and lines reflecting different statistics.
 Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
@@
 violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement
-invariance is $\sigma_{12}$.}
+invariance is $\phi_{12}$.}
 \label{fig:sim12res}
@@
 This Appendix includes the results for Simulation 2 when there exists model misspecification (lacking one path from Scale 1 to Math).
-Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim24res} display results when the loading $\lambda_{11}$, covariance $\sigma_{12}$, and
+Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim24res} display results when the loading $\lambda_{11}$, covariance $\phi_{12}$, and
 error term $\psi_{11}$ violate invariance, respectively.
@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\sigma_{12}$.}
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\phi_{12}$.}
 \label{fig:sim23res}

### 2013-10-21 frontiers/simall-ordinal.R

@@ -234,6 +234,8 @@ dgp2 <-function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
   cbind(as.data.frame(datmat), age)
 }
 
+
+
 ## Evaluate power simulation on a single dgp() scenario
 testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, test = test,
                       sim="sim1",verbose = TRUE, ...)
@@ -271,6 +273,8 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, 
   return(rval)
 }
 
+
+
 ## Loop over scenarios
 simulation <- function(diff = seq(0, 4, by = 0.25),parms=c("error","loading","var"),
   nobs = c(120, 480, 960), nlevels = c(4, 8, 12), gradual = FALSE,
@@ -298,8 +302,8 @@ simulation <- function(diff = seq(0, 4, by = 0.25),parms=c("error","loading","va
   cval.conds <- as.numeric(names(cval))
 
   pow <- matrix(rep(NA,ntest*nprs),ncol=ntest)
-  do.parallel <-require("parallel")
-  #do.parallel=FALSE
+#  do.parallel <-require("parallel")
+  do.parallel=FALSE
   if (do.parallel){
     pow <- mclapply(1:nprs, function(i){
       testpower(diff = prs$diff[i], nobs = prs$nobs[i],
@@ -365,6 +369,7 @@ source("../www/estfun-lavaan.R")
 source("../www/efpFunctional-cat.R")
 source("simall-ordinal.R")
 
+
 ## seed for replication
   RNGkind(kind = "default", normal.kind = "default")
   set.seed(1090)
@@ -372,16 +377,18 @@ source("simall-ordinal.R")
 
   
 ## To get a shorter version (say, 20 minutes)to see what's going on, you could do:
-simtry <- simulation(sim=c("sim1"),nobs=c(480),nrep=300, diff=seq(0, 4, by = 1),
-                     parms=c("loading","error","var"))
-simtry <- simulation(sim=c("sim2"), nobs=c(480),nrep=300, diff=seq(0, 4, by = 1),
-                     parms=c("extra","extra+loading","extra+var","extra+error"))
+simtry <- simulation(sim = c("sim1"), nobs = c(480),nrep = 300, diff = seq(0, 4, by = 1),
+                     parms = c("loading", "error", "var"))
[…]

### 2013-10-22 frontiers/frontiers.Rnw
update differentiation for weighted and max statistics. Add motivation for simulation. Delete figure panels showing zero powers.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@279 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \section{Simulation 1}
-
-In Simulation 1, we examined the extent to which the proposed tests
+Factor analysis models(and SEMs) are usually the first step for social scientists to determine the latent construct.
+These models are fairly complex with many inter-related parameters.
+Previous applications of the proposed tests have typically involved simpler models.
+In this paper we intended to study whether the tests can still identify the specific parameters that violate measurement invariance when parameter correlation is present in the model.
+\seeme{the above part is added motivation.
+Please check.} Particularly,
+in Simulation 1, we examined the extent to which the proposed tests
 are sensitive to the specific model parameter that violates measurement
@@
 $\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
-invariance, respectively. 
+invariance, respectively. In those Figures, we only show test parameters demonstrating non-zero powers for clarity. \seeme{Add this sentence to explaining there is no zero power curves.}
 All three test statistics exhibit
@@
 invariance violation.
+This ''dampening'' effect is more apparent for $\max \mathit{LM}_o$ statistic, because it sums across all parameters tested and then take the maximum.
+Whereas $\mathit{WDM}_o$ takes the maximum over parameters being tested directly.\seeme{Add from ''This'' to ''directly''.Please check.}
 
@@
        subset = (parms == "loading" & 
-		 pars %in% c("Loading1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+		 pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 4:5)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(1, 4)], ...)
 	   } else {
@@
        subset = (parms == "var" & 
-		 pars %in% c("Covariance", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+		 pars %in% c("Covariance") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude", ylab="Power", key = mykey, as.table = TRUE,
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(2, 4:5)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(2)], ...)
 	   } else {
@@
        subset = (parms == "error" & 
-		 pars %in% c("Error1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+		 pars %in% c("Error1", "All Errors") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(3,4:5)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(3,5)], ...)
 	   } else {
@@
        subset = (parms == "extra" & 
-		 pars %in% c("Loading1", "Error1", "Covariance") & diff %% 0.5 == 0),
+		 pars %in% c("Loading1", "Error1") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,

### 2013-10-23 frontiers/frontiers.Rnw
correct sim2 plot panel lable.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@280 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 2:3)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(1, 3)], ...)
 	   } else {

### 2013-10-24 frontiers/frontiers.Rnw
change covariance plot for sim1 and get rid of zero power curves in appendix.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@281 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \label{fig:sim12res}
-<<sim12res, fig=TRUE, height=9, width=9, echo=FALSE>>=
+<<sim12res, fig=TRUE, height=5.5, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+xyplot(power ~ diff | nlevels, group = ~ test, data = sim1,
        subset = (parms == "var" & 
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(2)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(2)],...)
 	   } else {
@@
        subset = (parms=="extra+loading" & 
-		 pars %in% c("Loading1", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+		 pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
@@
 \label{fig:sim23res}
-<<sim23res, fig=TRUE, height=9, width=9, echo=FALSE>>=
+<<sim23res, fig=TRUE, height=5.5, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
 mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+xyplot(power ~ diff | nlevels, group = ~ test, data = sim2,
        subset = (parms=="extra+var" & 
-		 pars %in% c("Covariance", "All Loadings", "All Errors") & diff %% 0.5 == 0),
+		 pars %in% c("Covariance") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,

### 2013-10-25 frontiers/frontiers.Rnw
Minor revisions to frontiers paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@282 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 In this paper, we consider a family of  recently-proposed measurement invariance
-tests that are based on the scores of a fitted model.  This family can be used
+tests that are based on the {\em scores} of a fitted model.  This family can be used
 to test for measurement invariance w.r.t.\ a continuous auxiliary variable,
@@
 of the tests who may wish to know (i) how the tests can be employed for their
-data, and (ii) whether the tests can accurately identify measurement invariance
-violations in the presence of model misspecification.  After providing an
+data, and (ii) whether the tests can accurately identify specific models parameters that violate measurement invariance (possibly in the presence of model misspecification).  After providing an
 overview of the tests, we illustrate their general use via the \proglang{R} packages \pkg{lavaan}
@@
 equation modeling) generally handles these ideas by replacing across-group
-equality constraints with informative prior distributions.  In this paper, we
+equality constraints with informative prior distributions \cite<e.g.,>{muthen13}.  In this paper, we
 describe an alternative approach: the development of test statistics that are
@@
 an ordinal variable, and they are special cases of a family of tests that may
-generally be used to study measurement invariance w.r.t.\ continuous,
+be used to study measurement invariance w.r.t.\ continuous,
 categorical, and ordinal variables.
 
-The family of tests described in this paper have recently been studied in the
-context of measurement invariance \cite{MerZei13,MerFanZei}, though their
+The family of tests described in this paper have recently been applied
+to the study of measurement invariance \cite{MerZei13,MerFanZei}, though their
 practical application has been limited to a small set of simulations and
 datasets.  In this paper, we provide a detailed illustration of the tests' use
-in practice and a detailed examination  of the tests' performance under
+and performance under
 scenarios likely to be encountered in practice. 
@@
   \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{1}}, \dots,
-  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right )^\top,
+  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right
+  )^\top,\ i=1,\ldots,n,
 \end{equation} 
 where $\ell(\theta; x_i)$ is the likelihood associated with individual
-$i$, $\theta$ is a $k$-dimensional parameter vector. The corresponding
+$i$ and $\theta$ is a $k$-dimensional parameter vector. The corresponding
 maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
@@
 
-To formalize the ideas in the previous paragraph, we define a
+To formalize these ideas, we define a
 suitably scaled cumulative sum of the ordered scores.  This may be written as
@@
 \end{equation}
-where $\lfloor nt \rfloor$ is the integer part of $nt$ (i.e., a floor
-operator) and $x_{(i)}$ reflects the individual with the
+where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
+nt \rfloor$ is 
+the integer part of $nt$ (i.e., a floor 
+operator), and $x_{(i)}$ reflects the individual with the
 $i$-th smallest value of the auxiliary variable $V$.
@@
 as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic 
-to be considerably better than the $\mathit{WDM}_o$ statistic.
+to have more power than the $\mathit{WDM}_o$ statistic and the traditional
+likelihood ratio test statistic when the measurement invariance
+violation is monotonic with the ordinal variable.
 
@@
 This test statistic can be formally written as
+\readme{Achim: Should $\ell$ be summed from 2 to $m$?  The second term
+  in parentheses is undefined for $\ell=1$.}
 \begin{equation}
@@
 This command specifies that we assess the parameter 1--4 of
-model \code{restr} after ordering the observations along the \code{agegroup} and
+model \code{restr} after ordering the observations according to
+\code{agegroup}.  Additionally, 
 the observed information matrix is used as the variance-covariance matrix. Note
-that the model parameters 1--4 are the factor loadings supplied by \pkg{lavaan}
-as can be seen by inspecting \code{coef(restr)}. This also leads to somewhat
-smaller test statistics that is very close to being significant at the 5\% level.
+that the model parameters 1--4 are the factor loadings supplied by
+\pkg{lavaan}, which can be seen by inspecting \code{coef(restr)}. This
+also leads to somewhat 
+smaller test statistics that are very close to being significant at the 5\% level.
 
@@
 differing from every other age group.  One solution to this problem would be the
-\emph{approximate invariance} methods proposed by Muth\'{e}n \cite{muthen13}.
+Bayesian, \emph{approximate invariance} methods described in the
+introduction \cite{muthen13}.
 Alternatively, we can use the proposed family of score-based statistics to obtain
@@
 age groups as continuous.  In doing so, we can use the statistics 
-from Equations~\eqref{eq:dmax}, ~\eqref{eq:cvm} and~\eqref{eq:maxlm}.  
+from Equations~\eqref{eq:dmax}, ~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
 That is, we can fit a model whose parameters 
@@
 how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
-age where age ties are broken arbitrarily (using the original order
+age (where age ties are broken arbitrarily, using the original order
 of the observations within each age group).
@@
 
-Again, we employ the \code{sctest()} function to assess the parameters~1--4
+Again, we employ the \code{sctest()} function to assess parameters~1--4
 from the restricted model \code{restr} after ordering w.r.t.\ \code{agegroup}:
@@
 and~$p=\Sexpr{round(maxlmo$p.value, 3)}$, respectively.\footnote{To replicate
-both $p$-values exactly, the \proglang{R}'s random seed needs to be
+both $p$-values exactly, \proglang{R}'s random seed needs to be
 set by \code{set.seed(1090)} prior to each \code{sctest()} call.}
@@
 
-In addition to test statistics, ``instability plots'' can be generated by additionaly setting
+In addition to test statistics, ``instability plots'' can be generated
+by setting 
 \code{plot = TRUE} in the \code{sctest()} calls above. The resulting plots representing
-the statistics' fluctuations across levels of age group are displayed in
+the ordinal statistics' fluctuations across levels of age group are displayed in
 Figure~\ref{fig:ordres}.
@@
 novel simulations to study the ordinal statistics' expected behavior
-in practice.  In particular, we wish to study (i) the extent to which the
+in practice. 
+In particular, we wish to study (i) the extent to which the
 ordinal statistics attribute measurement invariance violations to the
 correct parameter(s), and (ii) the extent to which the tests are
-robust to model misspecification.
+robust to model misspecification.  These issues are especially
+important to examine because SEMs are typically complex, with many
+inter-related parameters that may exhibit measurement invariance.
+Previous applications of score-based tests have typically focused on
+regression-like models with only a small number of parameters that
+may exhibit instability \cite<e.g.,>{zeihor07}.  Thus, the simulations
+here provide general evidence about the extent to which the tests accurately
+capture instabilities in complex models.
 
 \section{Simulation 1}
-Factor analysis models(and SEMs) are usually the first step for social scientists to determine the latent construct.
-These models are fairly complex with many inter-related parameters.
-Previous applications of the proposed tests have typically involved simpler models.
-In this paper we intended to study whether the tests can still identify the specific parameters that violate measurement invariance when parameter correlation is present in the model.
-\seeme{the above part is added motivation.
-Please check.} Particularly,
-in Simulation 1, we examined the extent to which the proposed tests
+In Simulation 1, we examined the extent to which the proposed tests
 are sensitive to the specific model parameter that violates measurement
@@
 invariance, it is plausible that this violation impacts other
-parameter estimates, including factor covariances or other loadings 
-associated with the factor in question.  Thus, the goal of the
+parameter estimates, including factor covariances or the unique variance
+associated with the manifest variable in question.  Thus, the goal of the
 Simulation 1 is to examine the extent to which the proposed tests
@@
 Power and Type I error were examined across three sample sizes
-($n=120,480,960$), three numbers of categories ($m=4,8,12$) and
-17 magnitudes of invariance violations. 
+($n=120,480,960$), three numbers of categories ($m=4,8,12$), and
+17 magnitudes of invariance violations (described in the following sentences). 
 The measurement invariance violations began at level $1+m/2$ of the
 auxiliary variable $V$ and were consistent thereafter: individuals below
-the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
-percentile  by $d$ times the parameters' asymptotic standard errors
+level $1+m/2$ of $V$ deviated from individuals at or above level $1+m/2$ 
+by $d$ times the parameters' asymptotic standard errors
 (scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
@@
 generated and tested.  Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo}
-and~\eqref{eq:lmuo} were examined. In all conditions, we maintained equal sample
+and~\eqref{eq:lmuo} were examined. As mentioned previously,
+~\eqref{eq:lmuo} is asymptotically equivalent to the usual likelihood
+ratio test.  Thus, this statistic provides information about the
+relative performance of the ordinal statistics vs the LRT.
+
+In all conditions, we maintained equal sample 
 sizes in each subgroup of the ordinal variable.
@@
 Figure~\ref{fig:sim11res} displays power curves as a function of 
-violation magnitude for the first
+violation magnitude in the 
 factor loading $\lambda_{11}$, with panels reflecting the parameters 
-being tested and lines reflecting different statistics.
+being tested and $m$ and lines reflecting different statistics.
 Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
@@
 $\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
-invariance, respectively. In those Figures, we only show test parameters demonstrating non-zero powers for clarity. \seeme{Add this sentence to explaining there is no zero power curves.}
-All three test statistics exhibit
-similar results, with $\max\mathit{LM}_{uo}$ demonstrating lower power across all situations.  As described earlier, this statistic 
-is asymptotically equivalent to the LRT.  $\mathit{WDM}_o$  and $\max \mathit{LM}_o$ exhibited equivalent power curves because, when only one parameter is tested, the statistics are equivalent.
+invariance, respectively. In these figures, we only show tests
+associated with parameters that exhibited nonzero power curves.  For
+example, in Figure~\ref{fig:sim11res}, we do not show tests associated
+with unique variance parameters or the factor covariance.  This is
+because these tests exhibited near-zero power for all values of $d$.
+
+Within each panel of Figure~\ref{fig:sim11res} to 
+Figure~\ref{fig:sim13res}, the three lines reflect the three test
+statistics.  It is seen that the two ordinal statistics exhibit 
+similar results, with $\max\mathit{LM}_{uo}$ demonstrating lower power
+across all situations.  This demonstrates the sensitivity of the
+ordinal statistics to invariance violations that are monotonic with
+$V$.  In situations where only one parameter is tested,
+$\mathit{WDM}_o$  and $\max \mathit{LM}_o$ exhibit equivalent power
+curves.  This is because, when only one parameter is tested, the
+statistics are equivalent.
 
@@
 invariance), we are effectively dampening the signal of a measurement
-invariance violation.
-This ''dampening'' effect is more apparent for $\max \mathit{LM}_o$ statistic, because it sums across all parameters tested and then take the maximum.
-Whereas $\mathit{WDM}_o$ takes the maximum over parameters being tested directly.\seeme{Add from ''This'' to ''directly''.Please check.}
+invariance violation. This ``dampening'' effect is more apparent for
+the $\max \mathit{LM}_o$ statistic, because it involves a sum across
+all tested parameters (see Equation~\eqref{eq:maxlmo}).
+Conversely, $\mathit{WDM}_o$ takes the maximum over parameters
+(Equation~\eqref{eq:wdm}), so that invariant parameters have no
+impact on this statistic.
 
@@
 simulation, the model was correctly specified.  In the following
-simulation, we examine the tests' performance when the model is
-misspecified.
+simulation, we examine the tests' performance in the likely situation
+of model misspecification.
 
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and
-$\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and
+$\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
 violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement 
-invariance is $\lambda_{11}$.}
+invariance is $\lambda_{11}$.  Panel labels denote the parameter(s)
+being tested (top) and levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim11res}
-<<sim11res, fig=TRUE, height=9, width=9, echo=FALSE>>=
+<<sim11res, fig=TRUE, height=7, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and
-$\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$, and measurement invariance
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement
-invariance is $\phi_{12}$.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and
+$\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$,
+  and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+parameter violating measurement invariance is $\phi_{12}$.}
 \label{fig:sim12res}
-<<sim12res, fig=TRUE, height=5.5, width=9, echo=FALSE>>=
+<<sim12res, fig=TRUE, height=4, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement invariance is $\psi{11}$}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+parameter violating measurement invariance is $\psi_{11}$. Panel labels
+denote the parameter(s) 
+being tested (top) and levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim13res}
-<<sim13res, fig=TRUE, height=9, width=9, echo=FALSE>>=
+<<sim13res, fig=TRUE, height=7, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 generate data from the factor analysis model used in the previous
-section, except that the model contains an extra loading from the second factor to Scale 1.
-The estimated model matches that displayed in Figure~\ref{fig:path}, however, resulting in model misspecification.  The goal of this simulation is to examine the proposed statistics'
-power and specificity under this misspecification.
+section, except that the model contains an extra loading from the
+second factor to Scale 1.
+The estimated model matches that displayed in Figure~\ref{fig:path},
+however, resulting in model misspecification.  The goal of this
+simulation is to examine the proposed statistics' 
+power to detect measurement invariance violations (and to attribute
+the violation to the correct parameter) under this misspecification.
 
 \subsection{Method}
-Measurement invariance violations could occur in
-each of the three parameters from Simulation 1 (factor loading, factor covariance, and unique variance), and they could also occur in the extra, unmodeled loading.
-In each condition, a single parameter exhibited the violation.
+A measurement invariance violation could occur in
+each of the three parameters from Simulation 1 (factor loading, factor
+covariance, and unique variance), and a violation could also occur in
+the extra, unmodeled loading. In each condition, a single parameter
+exhibited the violation. 
 Sample size and magnitude of measurement invariance violation were manipulated 
-in the same way as they were in Simulation 1.  The tested parameters were also the same as Simulation
-1.
+in the same way as they were in Simulation 1.  The tested parameters
+were also the same as Simulation 1.
 
 \subsection{Results}
-Results are presented by the parameter that violated measurement invariance. Of
-primary interest is Figure~\ref{fig:sim21res}, which displays results when the
-unmodeled loading violates  measurement invariance.  One can generally observe
-that  tests of the first loading and error variance parameter exhibited high
-power, reflecting a high Type I error rate.  Tests associated with the factor
-covariance did not demonstrate this error, however. In terms of specific
-statistic performance, $\max \mathit{LM}_o$ test and $\mathit{WDM}_o$ test
+Results of
+primary interest are conditions where the unmodeled loading violates
+measurement invariance.  These results are displayed in Figure~\ref{fig:sim21res}.  One can generally observe
+that tests of the first loading and unique variance exhibited high
+``power,'' which is actually a high Type I error rate here.  Tests
+associated with the factor 
+covariance (not shown) did not demonstrate this error, however. In
+terms of specific 
+statistic performance, $\max \mathit{LM}_o$ and $\mathit{WDM}_o$ 
 demonstrated higher Type I error than $\mathit{LM}_{uo}$ in each panel,
@@
 
+When the parameter violating measurement invariance was modeled,
+results were generally the same as Simulation 1.  
 When the modeled factor loading, $\lambda_{11}$, violated measurement
 invariance, the statistics were generally able to pick up the violation despite
-the misspecification. A similar pattern of specificity and power were observed
+the misspecification. Similar results were observed
 when the unique variance and factor covariance parameters violated measurement
-invariance; the figures associated with these results are shown in the Appendix.
-Also, the power of ordered statistics $\max \mathit{LM}_o$ test and
+invariance; these results are all shown in the Appendix.
+In particular, the power of ordered statistics $\max \mathit{LM}_o$ test and
 $\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in
@@
 parameters that do not violate measurement invariance.  The impacted parameters include the error
-variance and other loadings associated with the manifest variables that had the unmodeled 
-loading.  Thus, as it is with other tests of measurement invariance, it is important to study the extent to
+variance and other loadings associated with the manifest variable
+associated with the unmodeled 
+loading.  Thus, as for other tests of measurement invariance, it is important to study the extent to
 which the hypothesized model includes all parameters of importance.
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement invariance is the unmodeled loading.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+parameter violating measurement invariance is the unmodeled
+loading. Panel labels denote the parameter(s) 
+being tested (top) and levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim21res}
-<<sim21res, fig=TRUE, height=9, width=9, echo=FALSE>>=
+<<sim21res, fig=TRUE, height=7, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 
-In this paper, we first described a novel family of test statistics for measurement invariance and illustrated their use via the \proglang{R} package \pkg{lavaan}.
-Next, we examined these statistics' abilities to identify the parameter violating measurement invariance under well-specified and misspecified models.
-We found the proposed statistics could generally isolate the model parameter violating measurement invariance, so long as the violating parameter is included in the model.
-
-In the reminder of the paper, we will first consider the applicability of these tests as compared to traditional tests.
-Next, we discuss test extension to other fit functions and to other specialized models.
+In this paper, we first described a novel family of test statistics
+for measurement invariance and illustrated their use via the
+\proglang{R} package \pkg{lavaan}.  Next, we examined these
+statistics' abilities to identify the parameter violating measurement
+invariance under well-specified and misspecified models.  We found
+that the
+proposed statistics could generally isolate the model parameter
+violating measurement invariance, so long as the violating parameter
+is included in the model.
+
+In the reminder of the paper, we will first compare
+these tests to traditional tests of measurement invariance.  We also
+discuss test extension to other fit functions and to other specialized
+models.
 
@@
 
-Many of the applications in this volume, along with many applications in
-general, focus on testing for measurement invariance across unordered categories
+Many of the applications in this volume, along with many measurement
+invariance applications in
+general, focus on testing across unordered categories
 such as nations or gender.  As discussed earlier in this paper, the score-based
@@
 describing why measurement invariance violations occur, as opposed to simply
-detecting measurement invariance violations. \readme{In the abstracts for the
-special issue (which the special editors sent us), one paper was proposed along
-these lines.  I will get a draft of that paper and then we can cite it here.}
+detecting measurement invariance violations. 
+%% TODO!!
+%\readme{In the abstracts for the
+%special issue (which the special editors sent us), one paper was proposed along
+%these lines.  I will get a draft of that paper and then we can cite it here.}
 The tests are convenient for this purpose, as they do not require a new model to
@@
 
-In this paper, We focused on testing for measurement invariance in factor
+In this paper, we focused on testing for measurement invariance in factor
 analysis models that assume multivariate normality and that are estimated via
@@
 that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are
-potentially applicable to alternative SEM discrepancy functions such as weighted
+potentially applicable to alternative SEM discrepancy functions such
+as generalized 
 least squares \cite<e.g.,>{broarm95}.  Score calculation for these alternative
 discrepancy functions has not been implemented (to our knowledge), though the
-calculation certainly could be implemented.  Test statistic calculation and
+calculation could be implemented.  Test statistic calculation and
 inference would then proceed in exactly the same manner as the calculation and
@@
 
-This Appendix includes the results for Simulation 2 when there exists model misspecification (lacking one path from Scale 1 to Math).
-Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim24res} display results when the loading $\lambda_{11}$, covariance $\phi_{12}$, and
+This Appendix contains results for Simulation 2, where there existed
+model misspecification (lacking one loading from Scale 1 to
+Math). Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim24res} display
+results when the loading $\lambda_{11}$, covariance $\phi_{12}$, and  
 error term $\psi_{11}$ violate invariance, respectively.
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\lambda_{11}$.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+parameter violating measurement invariance is $\lambda_{11}$.  Panel
+labels denote the parameter(s)  
+being tested (top) and levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim22res}
-<<sim22res, fig=TRUE, height=9, width=9, echo=FALSE>>=
+<<sim22res, fig=TRUE, height=7, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\phi_{12}$.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+parameter violating measurement invariance is $\phi_{12}$.  Panel
+labels denote the parameter(s)  
+being tested (top) and levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim23res}
-<<sim23res, fig=TRUE, height=5.5, width=9, echo=FALSE>>=
+<<sim23res, fig=TRUE, height=4, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 \begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$ test, $\mathit{WDM}_o$ test, and $\mathit{LM}_{uo}$ test across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is $\psi_{11}$.}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+parameter violating measurement invariance is $\psi_{11}$.  Panel
+labels denote the parameter(s)  
+being tested (top) and levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim24res}

### 2013-10-25 frontiers/frontiers.Rnw
fix Fig9(right now Fig10) problem and add one figure in Appendix to show all loading and all errors.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@283 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
-parameter violating measurement invariance is $\phi_{12}$.}
+parameter violating measurement invariance is $\phi_{12}$. Panel
+labels denote the parameter(s)  
+being tested (top) and levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim12res}
@@
 that tests of the first loading and unique variance exhibited high
-``power,'' which is actually a high Type I error rate here.  Tests
+``power,'' which is actually a high Type I error rate here.
+This Type I error also demonstrated when testing all loadings and all unique variances.
+The result is shown in Appendix.\seeme{From ''This type I'' to ''Appendix'' are added to refer to the added Appendix Figure.
+}  Tests
 associated with the factor 
@@
 model misspecification (lacking one loading from Scale 1 to
-Math). Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim24res} display
-results when the loading $\lambda_{11}$, covariance $\phi_{12}$, and  
+Math). Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim25res} display
+results when the unmodeled loading\seeme{add this to refer to the added figure}, loading $\lambda_{11}$, covariance $\phi_{12}$, and  
 error term $\psi_{11}$ violate invariance, respectively.
+
+\begin{figure}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+parameter violating measurement invariance is the unmodeled
+loading. Panel labels denote the parameter(s) 
+being tested (top) and levels of the ordinal variable $m$ (bottom).}
+\label{fig:sim22res}
+<<sim22res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+       subset = (parms == "extra" & 
+		 pars %in% c("All Loadings", "All Errors") & diff %% 0.5 == 0),
+       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
+       strip = function(..., which.given, factor.levels){
+	   if(which.given == 2){
+	       strip.default(which.given, factor.levels = parlabs[c(4, 5)], ...)
+	   } else {
+	       strip.default(which.given, factor.levels = factor.levels, ...)
+	   }
+       })
+@
+
+\end{figure}
 
@@
 being tested (top) and levels of the ordinal variable $m$ (bottom).}
-\label{fig:sim22res}
-<<sim22res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+\label{fig:sim23res}
+<<sim23res, fig=TRUE, height=7, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 being tested (top) and levels of the ordinal variable $m$ (bottom).}
-\label{fig:sim23res}
-<<sim23res, fig=TRUE, height=4, width=9, echo=FALSE>>=
+\label{fig:sim24res}
+<<sim24res, fig=TRUE, height=4, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 being tested (top) and levels of the ordinal variable $m$ (bottom).}
-\label{fig:sim24res}
-<<sim24res, fig=TRUE, height=9, width=9, echo=FALSE>>=
+\label{fig:sim25res}
+<<sim25res, fig=TRUE, height=7, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra" & 
-		 pars %in% c("Error1", "All Errors", "All Loadings") & diff %% 0.5 == 0),
+       subset = (parms == "extra+error" & 
+		 pars %in% c("Error1", "All Errors") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(3, 4:5)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(3,5)], ...)
 	   } else {

### 2013-10-25 frontiers/frontiers.Rnw
Minor formatting revisions to frontiers paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@284 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 do not hold exactly across different groups, and this should be reflected in
-corresponding tests of measurement invariance.  BSEM (Bayesian structural
-equation modeling) generally handles these ideas by replacing across-group
-equality constraints with informative prior distributions \cite<e.g.,>{muthen13}.  In this paper, we
+corresponding tests of measurement invariance.  Under a Bayesian
+approach, we may utilize the idea of approximate invariance
+\cite<e.g.,>{muthen13}, whereby across-group equality constraints on
+parameters are replaced with informative prior distributions.  In this paper, we
 describe an alternative approach: the development of test statistics that are
@@
 This test statistic can be formally written as
-\readme{Achim: Should $\ell$ be summed from 2 to $m$?  The second term
-  in parentheses is undefined for $\ell=1$.}
 \begin{equation}
@@
 \end{equation}
-and it is asymptotically equivalent to the usual, likelihood ratio test statistic.
-It is advantageous over the likelihood ratio test because it requires estimation of
+where ${\bm B}(\hat {\bm \theta})_{i_{0}j} = 0$ for all $j$.
+\readme{Achim: I added a definition of ${\bm B}(\hat {\bm
+    \theta})_{i_{0}j}$.  Please double-check.} 
+This statistic is asymptotically equivalent to the usual, likelihood
+ratio test statistic, and it is advantageous over the likelihood ratio
+test because it requires estimation of 
 only one model (the restricted model).  We make use of this advantage in the
@@
 differing from every other age group.  One solution to this problem would be the
-Bayesian, \emph{approximate invariance} methods described in the
+Bayesian, approximate invariance methods described in the
 introduction \cite{muthen13}.
@@
 mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels, group = ~ test, data = sim1,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
        subset = (parms == "var" & 
@@
 primary interest are conditions where the unmodeled loading violates
-measurement invariance.  These results are displayed in Figure~\ref{fig:sim21res}.  One can generally observe
+measurement invariance.  A subset of results is displayed in Figure~\ref{fig:sim21res}.  One can generally observe
 that tests of the first loading and unique variance exhibited high
-``power,'' which is actually a high Type I error rate here.
-This Type I error also demonstrated when testing all loadings and all unique variances.
-The result is shown in Appendix.\seeme{From ''This type I'' to ''Appendix'' are added to refer to the added Appendix Figure.
-}  Tests
+``power,'' which is actually a high Type I error rate here.  This Type
+I error also demonstrated when testing all loadings and all unique
+variances, as shown in the Appendix.  Tests
 associated with the factor 
@@
 
-In the reminder of the paper, we will first compare
-these tests to traditional tests of measurement invariance.  We also
+In the reminder of the paper, we first compare
+these tests to traditional tests of measurement invariance.  We then 
 discuss test extension to other fit functions and to other specialized
@@
 %these lines.  I will get a draft of that paper and then we can cite it here.}
-The tests are convenient for this purpose, as they do not require a new model to
+The tests described here are convenient for this purpose, as they do
+not require a new model to 
 be estimated for each ordinal variable.  Instead, each ordinal variable defines
@@
 \pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
-for evaluating the parameter instability tests. {R}~and the packages \pkg{lavaan} and 
-\pkg{strucchange} are
+for evaluating the parameter instability tests. \proglang{R}~and the
+packages \pkg{lavaan} and \pkg{strucchange} are
 freely available under the General Public License~2 from the
@@
 
-This Appendix contains results for Simulation 2, where there existed
+This Appendix contains additional results from Simulation 2, where there existed
 model misspecification (lacking one loading from Scale 1 to
 Math). Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim25res} display
-results when the unmodeled loading\seeme{add this to refer to the added figure}, loading $\lambda_{11}$, covariance $\phi_{12}$, and  
-error term $\psi_{11}$ violate invariance, respectively.
+results when the unmodeled loading, the loading $\lambda_{11}$, the
+covariance $\phi_{12}$, and the error term $\psi_{11}$ violate
+invariance, respectively.  Results are only shown for simulation
+conditions exhibiting power curves that increased from zero.  These
+results were generally the same as the Simulation 1 results.
 
@@
 parameter violating measurement invariance is $\phi_{12}$.  Panel
-labels denote the parameter(s)  
+labels denote the parameter  
 being tested (top) and levels of the ordinal variable $m$ (bottom).}
@@
 mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels, group = ~ test, data = sim2,
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
        subset = (parms=="extra+var" & 

### 2013-10-25 frontiers/frontiers.Rnw
improved annotation of graphics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@285 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 where ${\bm B}(\hat {\bm \theta})_{i_{0}j} = 0$ for all $j$.
-\readme{Achim: I added a definition of ${\bm B}(\hat {\bm
-    \theta})_{i_{0}j}$.  Please double-check.} 
 This statistic is asymptotically equivalent to the usual, likelihood
@@
  ($\psi_{11}$), or the factor covariance $\phi_{12}$.  
- We then tested for measurement invariance in five subsets of parameters: the three
+ We then tested for measurement invariance in five subsets of parameters: each of the three
 individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
@@
 For each combination of sample size ($n$) $\times$ violation
-magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
+magnitude ($d$) $\times$ violating parameter $\times$ categories ($m$), 5000 datasets were
 generated and tested.  Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo}
@@
 ratio test.  Thus, this statistic provides information about the
-relative performance of the ordinal statistics vs the LRT.
+relative performance of the ordinal statistics vs.\ the LRT.
 
@@
   parlabs <- c(expression(lambda[11]), expression(phi[12]), 
-               expression(psi[11]), expression(lambda[11] - lambda[62]), 
-               expression(psi[11] - psi[66]))
+               expression(psi[11]),
+	       expression(list(lambda[11], ldots, lambda[62])), 
+               expression(list(psi[11], ldots, psi[66])))
   levels(sim1$pars) <- c("Loading1", "Covariance", "Error1", "All Loadings", "All Errors")  
@@
 invariance is $\lambda_{11}$.  Panel labels denote the parameter(s)
-being tested (top) and levels of the ordinal variable $m$ (bottom).}
+being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim11res}
@@
 		 pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
+       type = "b", xlab = expression(paste("Violation Magnitude (", lambda[11], ")")),
+       ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 labels denote the parameter(s)  
-being tested (top) and levels of the ordinal variable $m$ (bottom).}
+being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim12res}
@@
 		 pars %in% c("Covariance") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab="Power", key = mykey, as.table = TRUE,
+       type = "b", xlab = expression(paste("Violation Magnitude (", phi[12], ")")),
+       ylab="Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 denote the parameter(s) 
-being tested (top) and levels of the ordinal variable $m$ (bottom).}
+being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim13res}
@@
 		 pars %in% c("Error1", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
+       type = "b", xlab = expression(paste("Violation Magnitude (", psi[11], ")")),
+       ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
   parlabs <- c(expression(lambda[11]), expression(phi[12]), 
-               expression(psi[11]), expression(lambda[11] - lambda[62]), 
-               expression(psi[11] - psi[66]))
+               expression(psi[11]),
+	       expression(list(lambda[11], ldots, lambda[62])), 
+               expression(list(psi[11], ldots, psi[66])))
   levels(sim2$pars) <- c("Loading1", "Covariance", "Error1", "All Loadings", "All Errors")
@@
 loading. Panel labels denote the parameter(s) 
-being tested (top) and levels of the ordinal variable $m$ (bottom).}
+being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim21res}
@@
 		 pars %in% c("Loading1", "Error1") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
+       type = "b", xlab = "Violation Magnitude (Unmodeled Loading)",
+       ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 loading. Panel labels denote the parameter(s) 
-being tested (top) and levels of the ordinal variable $m$ (bottom).}
+being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim22res}
@@
 		 pars %in% c("All Loadings", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
+       type = "b", xlab = "Violation Magnitude (Unmodeled Loading)", ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 labels denote the parameter(s)  
-being tested (top) and levels of the ordinal variable $m$ (bottom).}
+being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim23res}
@@
 		 pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
+       type = "b", xlab = expression(paste("Violation Magnitude (", lambda[11], ")")), ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 labels denote the parameter  
-being tested (top) and levels of the ordinal variable $m$ (bottom).}
+being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim24res}
@@
 		 pars %in% c("Covariance") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude", ylab = "Power", key = mykey, as.table = TRUE,
+       type = "b", xlab = expression(paste("Violation Magnitude (", phi[12], ")")),
+       ylab = "Power", key = mykey, as.table = TRUE,
        strip = function(..., which.given, factor.levels){
@@
 labels denote the parameter(s)  
-being tested (top) and levels of the ordinal variable $m$ (bottom).}
+being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
 \label{fig:sim25res}

### 2013-10-27 frontiers/frontiers.Rnw
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@286 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 test statistics crosses the critical value.  While the measurement invariance tests are
-non-significant, the plots imply some instability in the older age groups (15 and 16).
+non-significant, the plots imply some instability in the older age groups (15 and~16).
 
@@
   functional = "WDMo", plot = TRUE, ylim = c(0, 3.2),
-  xlab = "Age group", ylab = "Weighted max statistics",
+  xlab = "Age group", ylab = "Weighted test statistics",
   main = expression(WDM[o]), boundary = list(col = 1, lty = 2))
@@
 Figure~\ref{fig:sim11res} displays power curves as a function of 
-violation magnitude in the 
-factor loading $\lambda_{11}$, with panels reflecting the parameters 
-being tested and $m$ and lines reflecting different statistics.
+violation magnitude in the factor loading $\lambda_{11}$,
+with the parameters being tested changing across rows,
+the number of levels $m$ of the ordinal variable $V$ across columns,
+and lines reflecting different test statistics.
 Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
@@
 invariance is $\lambda_{11}$.  Panel labels denote the parameter(s)
-being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
+being tested and the number of levels of the ordinal variable $m$.}
 \label{fig:sim11res}
@@
 labels denote the parameter(s)  
-being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
+being tested and the number of levels of the ordinal variable $m$.}
 \label{fig:sim12res}
-<<sim12res, fig=TRUE, height=4, width=9, echo=FALSE>>=
+<<sim12res, fig=TRUE, height=4.5, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 denote the parameter(s) 
-being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
+being tested and the number of levels of the ordinal variable $m$.}
 \label{fig:sim13res}
@@
 I error also demonstrated when testing all loadings and all unique
-variances, as shown in the Appendix.  Tests
+variances, as shown in the appendix.  Tests
 associated with the factor 
@@
 loading. Panel labels denote the parameter(s) 
-being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
+being tested and the number of levels of the ordinal variable $m$.}
 \label{fig:sim21res}
@@
 for measurement invariance and illustrated their use via the
-\proglang{R} package \pkg{lavaan}.  Next, we examined these
+\proglang{R} packages \pkg{lavaan} and \pkg{strucchange}.  Next, we examined these
 statistics' abilities to identify the parameter violating measurement
@@
 \pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
-for evaluating the parameter instability tests. \proglang{R}~and the
-packages \pkg{lavaan} and \pkg{strucchange} are
-freely available under the General Public License~2 from the
+for evaluating the parameter instability tests. \proglang{R}~and both
+packages are freely available under the General Public License~2 from the
 Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/}.
@@
 loading. Panel labels denote the parameter(s) 
-being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
+being tested and the number of levels of the ordinal variable $m$.}
 \label{fig:sim22res}
@@
 labels denote the parameter(s)  
-being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
+being tested and the number of levels of the ordinal variable $m$.}
 \label{fig:sim23res}
@@
 labels denote the parameter  
-being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
+being tested and the number of levels of the ordinal variable $m$.}
 \label{fig:sim24res}
-<<sim24res, fig=TRUE, height=4, width=9, echo=FALSE>>=
+<<sim24res, fig=TRUE, height=4.5, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 labels denote the parameter(s)  
-being tested (top) and the number of levels of the ordinal variable $m$ (bottom).}
+being tested and the number of levels of the ordinal variable $m$.}
 \label{fig:sim25res}

### 2013-10-28 frontiers/frontiers.Rnw
change simulation 2 according to Achim's 2-round comments.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@287 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 associated with the factor 
-covariance (not shown) did not demonstrate this error, however. In
+covariance \seeme{delete the ''not shown''} did not demonstrate this error, however.  In
 terms of specific 
@@
 as the umodeled parameter does not violate measurement invariance.  If the unmodeled parameter 
-dose violate measurement invariance, however, then the tests assign this violation to other 
-parameters that do not violate measurement invariance.  The impacted parameters include the error
+dose violate measurement invariance, however, the test could detect there existing measurement invariance in the model but assign this violation to other 
+parameters that do not violate measurement invariance.  \seeme{add the point that the tests could detect measurement invariance}.  The impacted parameters include the error
 variance and other loadings associated with the manifest variable
@@
 loading.  Thus, as for other tests of measurement invariance, it is important to study the extent to
-which the hypothesized model includes all parameters of importance.
+which the hypothesized model includes all parameters of importance.  The tests are model-based and could not inform the researcher model misspecification. \seeme{The last sentence.}
 
@@
 \label{fig:sim21res}
-<<sim21res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+<<sim21res, fig=TRUE, height=9, width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
        subset = (parms == "extra" & 
-		 pars %in% c("Loading1", "Error1") & diff %% 0.5 == 0),
+		 pars %in% c("Loading1", "Covariance", "Error1") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude (Unmodeled Loading)",
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 3)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(1, 2, 3)], ...)
 	   } else {

### 2013-10-28 frontiers/frontiers.Rnw
Added detail on saving critical values via ordL2BB Added extra "zero-power" rows to some plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@288 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 calculation of the $\max\mathit{LM}_o$ statistic (Equation~\eqref{eq:maxlmo})
-can be lengthy from the need to simulate critical values.
+can be lengthy from the need to simulate critical values (though see
+the end of this section, which provides a partial speed-up).
 %% This is more complicated than necessary, but it is written like
@@
 \end{figure}
+
+Finally, if the user anticipates multiple calculations of the
+$\max\mathit{LM}_o$ statistic for a specific dataset, it is possible
+to save time by simulating critical values once and re-using them for
+multiple tests.  We can use the \code{ordL2BB()} command to generate
+critical values, then obtain the test statistic in the usual manner.
+<<ordsave, eval=FALSE>>=
+## Simulate critical values
+mLMo <- ordL2BB(yg$agegroup)
+## Use these critical values for the test
+maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = mLMo)
+@ 
+The \code{ordL2BB()} command automatically generates critical values
+for testing 1 to 20 parameters at a time, with critical values for
+greater numbers of parameters being obtained through the optional
+argument \code{nproc}.  One can re-use \code{mLMo} in this manner for
+further tests of the youth gratitude data.  Critical values must be
+resimulated for new data, however, because they depend on the
+proportion of individuals observed at each level of the ordinal
+variable (see~\eqref{eq:maxlmo}).
 
@@
 
-To examine these specificity issues, we generated data from a
+To examine these issues, we generated data from a
 two-factor model with three indicators each (see Figure~\ref{fig:path}).
@@
 by $d$ times the parameters' asymptotic standard errors
-(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
+(scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$. 
 For each combination of sample size ($n$) $\times$ violation
@@
 
-Full simulation results are presented in Figure~\ref{fig:sim11res} to 
-Figure~\ref{fig:sim13res}; 
+Full simulation results are presented in Figures~\ref{fig:sim11res}
+to~\ref{fig:sim13res}.
 Figure~\ref{fig:sim11res} displays power curves as a function of 
@@
 $\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
-invariance, respectively. In these figures, we only show tests
+invariance, respectively. In these figures, we generally show tests
 associated with parameters that exhibited nonzero power curves.  For
-example, in Figure~\ref{fig:sim11res}, we do not show tests associated
-with unique variance parameters or the factor covariance.  This is
-because these tests exhibited near-zero power for all values of $d$.
+example, in Figure~\ref{fig:sim11res}, the middle row shows that power
+for tests of $\psi_{11}$ stays near zero for all values of $m$ and $d$.
+Similar rows have been omitted from this figure and other figures.
 
@@
 $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The parameter violating measurement 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The parameter violating measurement 
 invariance is $\lambda_{11}$.  Panel labels denote the parameter(s)
@@
        subset = (parms == "loading" & 
-		 pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
+		 pars %in% c("Loading1", "Error1", "All Loadings") & diff %% 0.5 == 0),
        type = "b", xlab = expression(paste("Violation Magnitude (", lambda[11], ")")),
@@
            if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 4)], ...)
+             strip.default(which.given, factor.levels = parlabs[c(1, 3, 4)], ...)
 	   } else {
@@
   and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
 parameter violating measurement invariance is $\phi_{12}$. Panel
@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
 parameter violating measurement invariance is $\psi_{11}$. Panel labels
@@
 associated with the factor 
-covariance \seeme{delete the ''not shown''} did not demonstrate this error, however.  In
+covariance did not demonstrate this error, however.  In
 terms of specific 
@@
 
-In summary, the proposed test statistics appear robust to unmodeled loading parameters, so long
-as the umodeled parameter does not violate measurement invariance.  If the unmodeled parameter 
-dose violate measurement invariance, however, the test could detect there existing measurement invariance in the model but assign this violation to other 
-parameters that do not violate measurement invariance.  \seeme{add the point that the tests could detect measurement invariance}.  The impacted parameters include the error
-variance and other loadings associated with the manifest variable
-associated with the unmodeled 
-loading.  Thus, as for other tests of measurement invariance, it is important to study the extent to
-which the hypothesized model includes all parameters of importance.  The tests are model-based and could not inform the researcher model misspecification. \seeme{The last sentence.}
+In summary, the proposed test statistics appear robust to unmodeled
+loading parameters, so long 
+as the unmodeled loading does not violate measurement invariance.  If
+the unmodeled loading  
+does violate measurement invariance, the tests can still detect
+measurement invariance violations.  The violations are assigned to 
+modeled parameters that do not violate measurement invariance, however.
+The impacted parameters include the error 
+variance and other loadings associated with the manifest variable that
+has an unmodeled loading.
+Thus, as for other tests of measurement invariance, it is important to
+study the extent to 
+which the hypothesized model includes all parameters of importance.
+None of these tests (score-based or otherwise) inform the researcher of
+model misspecification.
 
@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
 parameter violating measurement invariance is the unmodeled
@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
 parameter violating measurement invariance is $\lambda_{11}$.  Panel
@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
 parameter violating measurement invariance is $\phi_{12}$.  Panel
@@
 \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$). The
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
 parameter violating measurement invariance is $\psi_{11}$.  Panel

### 2013-10-29 frontiers/frontiers.Rnw
slightly rephrased discussion of direction usage of ordL2BB
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@289 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 multiple tests.  We can use the \code{ordL2BB()} command to generate
-critical values, then obtain the test statistic in the usual manner.
+critical values and store them in an object \code{mLMo}, say.
+Then, this object can be employed to obtain the test statistic in the usual manner.
 <<ordsave, eval=FALSE>>=
-## Simulate critical values
 mLMo <- ordL2BB(yg$agegroup)
-## Use these critical values for the test
 maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
@@
 The \code{ordL2BB()} command automatically generates critical values
-for testing 1 to 20 parameters at a time, with critical values for
-greater numbers of parameters being obtained through the optional
-argument \code{nproc}.  One can re-use \code{mLMo} in this manner for
-further tests of the youth gratitude data.  Critical values must be
-resimulated for new data, however, because they depend on the
+for testing 1 to 20 parameters at a time. If only a smaller number
+of parameters (e.g., only up to 6) is to be tested, some computation time can be
+saved by setting the \code{nproc} argument accordingly (e.g., \code{nproc = 1:6}).
+In the same way, \code{nproc} can be employed to simulate higher-dimensional
+fluctuation processes suitable for testing more parameters. One can re-use
+\code{mLMo} in this manner for further tests of the youth gratitude data.
+Critical values must be resimulated for new data, however, because they depend on the
 proportion of individuals observed at each level of the ordinal
-variable (see~\eqref{eq:maxlmo}).
+variable (denoted $t_\ell$ for Equation~\eqref{eq:maxlmo}).
 

### 2013-10-29 frontiers/frontiers.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@290 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
small improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@291 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 This section contains background and discussion of the  proposed statistics as
-applied to SEM; for a more detailed account, see \citeA{MerZei13} and
+applied to structural equation models (SEMs); for a more detailed account, see \citeA{MerZei13} and
 \citeA{MerFanZei}.  For details on the statistics' application to general
@@
 
-In this section, we demonstrate how the above tests can be carried out in~R, 
+In this section, we demonstrate how the above tests can be carried out in \proglang{R}, 
 using the package \pkg{lavaan} \cite{lavaan11} for model estimation and \pkg{strucchange} 
@@
 the applicability of adult gratitude scales to youth, available in the~\proglang{R} 
-package \emph{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
+package \pkg{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
 gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
@@
 age groups as continuous.  In doing so, we can use the statistics 
-from Equations~\eqref{eq:dmax}, ~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
+from Equations~\eqref{eq:dmax}, \eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
 That is, we can fit a model whose parameters 
@@
 to save time by simulating critical values once and re-using them for
-multiple tests.  We can use the \code{ordL2BB()} command to generate
+multiple tests.  We can use the \code{ordL2BB()} function to generate
 critical values and store them in an object \code{mLMo}, say.
@@
 magnitude ($d$) $\times$ violating parameter $\times$ categories ($m$), 5000 datasets were
-generated and tested.  Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo}
+generated and tested.  Statistics from Equations~\eqref{eq:wdm}, \eqref{eq:maxlmo}
 and~\eqref{eq:lmuo} were examined. As mentioned previously,
-~\eqref{eq:lmuo} is asymptotically equivalent to the usual likelihood
+\eqref{eq:lmuo} is asymptotically equivalent to the usual likelihood
 ratio test.  Thus, this statistic provides information about the
@@
 
-In the reminder of the paper, we first compare
+In the remainder of the paper, we first compare
 these tests to traditional tests of measurement invariance.  We then 

### 2013-10-30 frontiers/mz-ordinal-proptest.R
Edited index.php to add materials for frontiers paper Related to above, moved files around and deleted some unused files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@292 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,80 +1,0 @@
-ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
-{
- ## data checks
-  data <- as.data.frame(data)
-  stopifnot(c("x1", "x2", "x3", "y1", "y2", "y3", "age") %in% names(data))
-  age <- data$age
-
-  ## require package
-  stopifnot(require("lavaan"))
-
-  ## model from Merkle & Zeileis (for proposed tests)
-  rval <- 'verbal =~ x1 + x2 + x3; math =~ y1 + y2 + y3'
-
-  data <- data[, c("x1", "x2", "x3", "y1", "y2", "y3")]
-  
-  ## fit model
-  rval <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE))
-  
-  ## store (1) fitted model object, (2) lavaan coefficient
-  ## labeling and ordering
-  rval <- list(
-    data = data,
-    model = rval,
-    names = c("verbal=~x1", "verbal=~x2", "verbal=~x3", "math=~y1", "math=~y2", "math=~y3", "x1~~x1", "x2~~x2", "x3~~x3",
-        "y1~~y1", "y2~~y2", "y3~~y3", "verbal~~math", "x1~1", "x2~1", "x3~1", "y1~1", "y2~1", "y3~1")
-            )
-  
-  class(rval) <- "mzfit"
-
-  return(rval)
-}
-
-coef.mzfit <- function(object, ...) {
-    unclass(getMethod("coef", "lavaan")(object$model, ...))
-}
-
-logLik.mzfit <- function(object, ...) {
-  rval <- as.vector(fitMeasures(object$model, "logl"))
-
-  structure(rval, df = length(coef(object)),
-    nobs = nrow(object$data), class = "logLik")
-}
-
-vcov.mzfit <- function(object, ...) {
-    vc <- getMethod("vcov", "lavaan")(object$model, ...)
-    class(vc) <- "matrix"
-    vc
-}
-
-bread.mzfit <- function(x, ...) vcov(x) * nrow(x$data)
-
-info.mzfit <- function(x, ...) solve(vcov(x) * nrow(x$data))
-
-estfun.mzfit <- function(x, ...)
-{
-  estfun.lavaan(x$model)
-}
-
-get.zcov <- function(data){
-  age <- data$age
[…]

### 2013-10-30 frontiers/simall.R

@@ -1,778 +1,0 @@
-
-######################################################################################
-#READ ME##############################################################################
-#                                                                                    #
-#This study aims to initiate different dpg functions to explore measurment invariance#
-#test properties in different scenarios.Specifically,                                # 
-#dgp1:different parameter, var, loading, error violate measurement invariance(MI).   #
-#dgp2:model misspecified by lacking one loading, the other are the same as dgp1.     #
-#dgp3:group percentage change while different parameter violate MI.                  #
-#dgp4:Three groups generated by two groups violating MI with different amptitude.    #
-#dgp5:Observations violate MI according to age continuously.                         # 
-#dgp6 Group percentage change while 3 loading or 3 error term violate MI.            # 
-######################################################################################
-
-
-
-## Data-generating process, different dgp function represent different simulation study.
-#######################################################################################
-#Simualtion1, different parameter voilation############################################
-dgp1 <- function(nobs = 200, diff = 3, parms="loading", perc = 0.5)
-{
-  # Generates data from a factor analysis model that violates
-  # measurement invariance.  Also generates a continuous auxiliary
-  # variable related to the invariance (loosely called "age").
-  
-  # ses is number of SEs between group 1's parameters and group 2's
-  # parameters.
-  # theta[1-6]:loadings, theta[7-12]:residual,theta[13]:covariance, theta[14-19]:mean
-  sampsize <- nobs
-  ses <- diff
-  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5,-0.48,
-           29.32,24.70,14.84,10.59,19.30,18.01)
[…]

### 2013-10-30 www/mz-frontiers.R

@@ -1,27 +1,30 @@
-mzfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
+ordfit <- function(data, silent = TRUE, suppressWarnings = TRUE, ...)
 {
-  ## data checks
+ ## data checks
   data <- as.data.frame(data)
-  stopifnot(c("x1", "x2", "x3", "y1", "y2", "y3") %in% names(data))
-  data <- data[, c("x1", "x2", "x3", "y1", "y2", "y3")]
+  stopifnot(c("x1", "x2", "x3", "y1", "y2", "y3", "age") %in% names(data))
+  age <- data$age
 
   ## require package
   stopifnot(require("lavaan"))
 
-  ## model from Merkle & Zeileis
+  ## model from Merkle & Zeileis (for proposed tests)
   rval <- 'verbal =~ x1 + x2 + x3; math =~ y1 + y2 + y3'
 
-  ## fit model (set likelihood = "wishart" for N-1 correction as in OpenMx)
-  rval <- sem(rval, data = data, meanstructure = TRUE, std.lv = TRUE)
-
-  ## store (1) fitted model object, (2) engine used,
-  ## (3) match between OpenMx and lavaan coefficient labeling and ordering
+  data <- data[, c("x1", "x2", "x3", "y1", "y2", "y3")]
+  
+  ## fit model
+  rval <- try(cfa(rval, data = data, meanstructure = TRUE, std.lv = TRUE))
+  
+  ## store (1) fitted model object, (2) lavaan coefficient
+  ## labeling and ordering
   rval <- list(
     data = data,
     model = rval,
     names = c("verbal=~x1", "verbal=~x2", "verbal=~x3", "math=~y1", "math=~y2", "math=~y3", "x1~~x1", "x2~~x2", "x3~~x3",
-        "y1~~y1", "y2~~y2", "y3~~y3", "verbal~~math", "x1~1", "x2~1", "x3~1", "y1~1", "y2~1", "y3~1"))
-
+        "y1~~y1", "y2~~y2", "y3~~y3", "verbal~~math", "x1~1", "x2~1", "x3~1", "y1~1", "y2~1", "y3~1")
+            )
+  
   class(rval) <- "mzfit"
 
   return(rval)

### 2013-10-30 www/sim-frontiers.R

@@ -1,6 +1,5 @@
 
 ## Data-generating process
-set.seed=1090
 dgp1 <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALSE, parms="loadings")
 {
   ## Generates data from a factor analysis model that violates
@@ -357,7 +356,7 @@ critvals <- function(nlevels = c(4, 8, 12), verbose = TRUE, ...)
 
 
 
-
+## Examples of running the simulation
 if(FALSE){
 
 library("lavaan")
@@ -377,120 +376,14 @@ source("simall-ordinal.R")
 
   
 ## To get a shorter version (say, 20 minutes)to see what's going on, you could do:
-simtry <- simulation(sim = c("sim1"), nobs = c(480),nrep = 300, diff = seq(0, 4, by = 1),
+simtry <- simulation(sim = c("sim1"), nobs = c(480), nrep = 300,
+                     diff = seq(0, 4, by = 1), 
                      parms = c("loading", "error", "var"))
-simtry <- simulation(sim = c("sim2"), nobs = c(480),nrep=300, diff = seq(0, 4, by = 1),
-                     parms = c("extra", "extra+loading", "extra+var", "extra+error"))
+simtry <- simulation(sim = c("sim2"), nobs = c(480), nrep=300,
+                     diff = seq(0, 4, by = 1),
+                     parms = c("extra", "extra+loading", "extra+var",
+                       "extra+error"))
 
 save(simtry,file ="simtry.rda")
-
-## run full simulation
-sim1 <- simulation(sim=c("sim1"), nobs=c(480),parms=c("loading","error","var"))
-sim2 <- simulation(sim=c("sim2"), nobs=c(480),
-                   parms=c("extra","extra+loading","extra+var","extra+error"))
-
-
-#load data file and give names for different levels in each variable.
-if(file.exists("simtry.rda")) {
-  load("simtry.rda")
-} else {
-  simtry <- simulation()
-  simtry$nlevels <- factor(simtry$nlevels)
-  levels(simtry$nlevels) <- paste("m=",levels(simtry$nlevels),sep="")
-  simtry$nobs <- factor(simtry$nobs)
-  levels(simtry$nobs) <- paste("n=",levels(simtry$nobs),sep="")
[…]

### 2013-10-31 frontiers/frontiers.Rnw
Edited index.php to add materials for frontiers paper Related to above, moved files around and deleted some unused files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@292 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor changes to make the code more readable, and to ensure that simulations replicate.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@293 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ## auxiliary code
-source("simall-ordinal.R")
-source("mz.R")
-source("../www/mz-ordinal.R")
+source("../www/sim-frontiers.R")
+source("../www/mz-frontiers.R")
 
@@
   levels(sim1$nobs) <- paste("n=", levels(sim1$nobs), sep="")
-  levels(sim1$test) <- c("maxLM_O", "WDM_O", "LM_uO")
   save(sim1, file="sim1.rda")
 }
-sim1$test <- factor(as.character(sim1$test),
-  levels = c("ordmax", "ordwmax", "catdiff"),
+sim1$test <- factor(as.character(sim1$test), levels = c("ordmax","ordwmax","catdiff"),
                     labels = c("maxLM_o", "WDM_o", "LM_uo"))
 parlabs <- c(expression(lambda[11]), expression(phi[12]), 
-               expression(psi[11]),
-	       expression(list(lambda[11], ldots, lambda[62])), 
+             expression(psi[11]), expression(list(lambda[11], ldots, lambda[62])), 
              expression(list(psi[11], ldots, psi[66])))

### 2013-10-31 www/sim-frontiers.R

@@ -116,8 +116,8 @@ dgp1 <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FAL
   cbind(as.data.frame(datmat), age)
 }
 
-dgp2 <-function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALSE,
-                parms="loadings")
+dgp2 <-function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE,
+                anomaly=FALSE, parms="loadings")
 {
   ## Generates data from a factor analysis model that violates
   ## measurement invariance.  Also generates an ordinal auxiliary
@@ -236,8 +236,8 @@ dgp2 <-function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FALS
 
 
 ## Evaluate power simulation on a single dgp() scenario
-testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, test = test,
-                      sim="sim1",verbose = TRUE, ...)
+testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum,
+                      test = test, sim="sim1", verbose = TRUE, ...)
 {
   pval <- matrix(rep(NA, length(test) * nrep), ncol = length(test))
   colnames(pval) <- test
@@ -250,19 +250,15 @@ testpower <- function(nrep = 5000, size = 0.05, ordfun = NULL, parnum = parnum, 
     ## Requires estfun.lavaan
     ord_gefp <-vector("list",length(parnum))
     for (j in 1: length(parnum)){
-    ord_gefp[[j]]  <- try(gefp(mz, fit = NULL, vcov = info.mzfit, order.by = d$age,
-                               sandwich =FALSE, parm = eval(parse(text=parnum[j]))),
+    ord_gefp[[j]]  <- try(gefp(mz, fit = NULL, vcov = info.mzfit,
+                               order.by = d$age, sandwich =FALSE,
+                               parm = eval(parse(text=parnum[j]))),
                           silent = TRUE)
 
     if(!inherits(ord_gefp[[j]], "try-error")) {
-      pval[i, (j-1)*3+1] <- sctest(ord_gefp[[j]],  functional = ordfun)$p.value
[…]

### 2013-10-31 www/replication-frontiers.R
Frontiers paper, added replication script and modified index.php
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@294 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,281 @@
+####################
+### Preliminaries ##
+####################
+
+## packages
+library("lavaan")
+library("strucchange") ## at least 1.5-0
+library("mvtnorm")
+library("lattice")
+
+## auxiliary code
+source("../www/sim-frontiers.R")
+source("../www/mz-frontiers.R")
+
+## some options
+options(prompt = "R> ", continue = "+  ", digits = 3,
+  useFancyQuotes = FALSE, show.signif.stars = FALSE)
+
+
+#######################
+### Tutorial Section ##
+#######################
+
+data("YouthGratitude", package = "psychotools")
+compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
+yg <- YouthGratitude[compcases, ]
+
+restr <- cfa("f1 = ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+  data = yg, group = "agegroup", meanstructure = TRUE,
+  group.equal = "loadings")
+full <- cfa("f1= ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+  data = yg, group = "agegroup", meanstructure = TRUE)
+
+anova(full, restr)
+
+sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "LMuo")
+
+dm <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "DM")
+cvm <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "CvM")
+maxlm <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "maxLM")
+c(dm$p.value, cvm$p.value, maxlm$p.value)
+
+set.seed(1090)
+wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "WDMo")
+
+## Use of saved critical values:
+if(file.exists("maxLMo.rda")){
+  load("maxLMo.rda")
+} else {
+  RNGkind(kind = "default", normal.kind = "default")
+  set.seed(1090)
+  maxLMo <- ordL2BB(yg$agegroup, nproc = 4)
+  save(maxLMo, file = "maxLMo.rda")
+}
+maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
[…]

### 2013-11-01 www/sim-frontiers.R
Minor changes to frontiers simulation code.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@295 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -361,11 +361,8 @@ if(FALSE){
 library("lavaan")
 library("strucchange")
 library("mvtnorm")
-#source("../www/mz-ordinal.R")
-source("mz-ordinal-proptest.R")
-source("../www/estfun-lavaan.R")
-source("../www/efpFunctional-cat.R")
-source("simall-ordinal.R")
+source("sim-frontiers.R")
+source("mz-frontiers.R")
 
 
 ## seed for replication
@@ -374,15 +371,18 @@ source("simall-ordinal.R")
   
 
   
-## To get a shorter version (say, 20 minutes)to see what's going on, you could do:
-simtry <- simulation(sim = c("sim1"), nobs = c(480), nrep = 300,
+## Short versions:
+## Simulation 1:
+simtry <- simulation(sim = "sim1", nobs = 480, nrep = 300,
                      diff = seq(0, 4, by = 1), 
                      parms = c("loading", "error", "var"))
-simtry <- simulation(sim = c("sim2"), nobs = c(480), nrep=300,
+## Simulation 2:
+simtry <- simulation(sim = "sim2", nobs = 480, nrep=300,
                      diff = seq(0, 4, by = 1),
                      parms = c("extra", "extra+loading", "extra+var",
-                       "extra+error"))
+                               "extra+error"))
 
+## Save some results to file
 save(simtry,file ="simtry.rda")
  
 }

### 2014-01-24 frontiers/frontiers.Rnw
deal with reviewer 1.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@300 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 parameters are replaced with informative prior distributions.  In this paper, we
-describe an alternative approach: the development of test statistics that are
-sensitive to invariance violations of interest and insensitive to ``anomalous''
-invariance violations.  The test statistics are specifically applicable to
-situations where one wishes to test for measurement invariance with respect to
-an ordinal variable, and they are special cases of a family of tests that may
-be used to study measurement invariance w.r.t.\ continuous,
-categorical, and ordinal variables.
+describe an alternative approach: the development of test statistics that 
+are sensitive to monotonic w.r.t. variable of research interest and 
+less sensitive to non-monotonic violations\seeme{original sentence:
+the development of test statistics that are sensitive to invariance 
+violations of interest and insensitive to "anomalous" invariance violations}.  
+The test statistics are specifically applicable to situations where one 
+wishes to test for measurement invariance with respect to an ordinal variable, 
+and they are special cases of a family of tests that may be used to study 
+measurement invariance w.r.t.\ continuous, categorical, and ordinal variables
+\cite{mer93,DolMaa98,lubmut05}\seeme{add these three refs.}.
 
@@
 maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
-$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$. The cross-product
+$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$. \seeme{delete this
+sentence: The cross-product
 of these scores forms the ``meat'' for the calculation of robust
-(Huber-White) standard errors \cite<e.g.,>{zei06b}.
+(Huber-White) standard errors \cite<e.g.,>{zei06b}}.
 
@@
 for measurement 
-invariance w.r.t.\ age.  If the manifest variables violate measurement
-invariance here, then some parameter estimates may be too large for
+invariance w.r.t.\ age.  If there exists measurement invariance w.r.t. age
+\seeme{original sentence:
+If the manifest variables violate measurement
+invariance here}, then some parameter estimates may be too large for
 young individuals and too small for old individuals (say).  This
@@
 the integer part of $nt$ (i.e., a floor 
-operator), and $x_{(i)}$ reflects the individual with the
-$i$-th smallest value of the auxiliary variable $V$.
+operator), thus t could be interpreted as $1/n$, $2/n$, $3/n$, ...$n/n$.
+\seeme{add from thus to n/n}. 
+$x_{(i)}$ reflects the individual with the $i$-th smallest value of the 
+auxiliary variable $V$.  So we replace the score component shown in 
+Equation~\eqref{eq:score} by cumulative score component shown in 
+Equation~\eqref{eq:cumscore} without changing dimension
+\seeme{add from So to dimension}. 
 We specifically focus on how the
@@
 used to show that the fluctuation of the above cumulative sum follows a
-Brownian bridge.  This result allows us to calculate $p$-values and
+Brownian bridge \cite{hjokon02}\seeme{add this ref}.  
+This result allows us to calculate $p$-values and
 critical values for test statistics under the hypothesis of
@@
 package \pkg{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
-gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
+gratitude scales from $n=1401$ youth aged 10--19 years. 
+The response scale is 1 to 9 \seeme{add the scale}. 
+The original authors 
 were specifically interested in whether the scales were measurement invariant 
@@
 %
-Next, we fit two models in \pkg{lavaan}: a one-factor model where loadings are restricted 
-to be equal across age groups, and a one-factor model where loadings are free across age groups.
+Next, we fit two models in \pkg{lavaan}: a one-factor model 
+where loadings are restricted to be equal across age groups, 
+and a one-factor model where loadings are free across age groups
+(Note, the researcher could fit any kind of reduced model, 
+e.g. restrict the residuals to be equal.  For model 
+specification code, please refer to \pkg{lavaan} package)
+\seeme{Add this note}.
 %

### 2014-01-24 frontiers/frontiers.Rnw
make changes according to reviewer 2. --This line, and those below, will be ignored--
M    frontiers/frontiers_comments.txt
M    frontiers/frontiers.Rnw


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@301 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Because our sample size is large, the likelihood ratio test is known to be sensitive to 
-small measurement invariance violations \cite{Ben80}.  That is, the above tests 
+small measurement invariance violations \cite{Ben80}.  That is, the above tests (LRT and  $\mathit{LM}_\mathit{uo}$ test) \seeme{add the LRT to test content}
 are sensitive to small measurement invariance violations that are not likely to be 

### 2014-01-24 www/sim-frontiers.R
add intercept argument value and testing parameter 14 and 14:19 to dgp1 and dpg2.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@302 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -26,7 +26,8 @@ dgp1 <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FAL
   half.level <- ifelse(nlevels %% 2 == 1, (nlevels %/% 2)+1, nlevels/2)
   
   # Define parameter vectors/matrices:
-  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5, -0.48,
+  theta<-c(4.92,2.96,5.96,3.24,4.32,7.21,26.77,13.01,30.93,3.17,8.82,22.5,
+           -0.48,
            29.32,24.70,14.84,10.59,19.30,18.01)
   asym<-c(7.565818,4.951599,8.644422,3.093014,4.499676,7.368059,
            62.1,26.6,82.1,9.05,19.35,51.59,
@@ -71,6 +72,7 @@ dgp1 <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FAL
     if (parms=="loading"){parmsvector=c(1,rep(0,18))}
     if (parms=="var"){parmsvector=c(rep(0,12),1,rep(0,6))}
     if (parms=="error"){parmsvector=c(rep(0,6),1,rep(0,5),rep(0,7))}
+    if (parms=="intercept"){parmsvector=c(rep(0,13),1,rep(0,5))}
    
     mult <- ses
     if (anomaly) mult <- ifelse(i==(half.level+1), ses, 0)
@@ -91,6 +93,9 @@ dgp1 <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FAL
     # variances
     tmp.psi <- matrix(0,6,6)
     diag(tmp.psi) <- theta.tmp[7:12]
+
+    # intercepts
+    tmp.mu <- theta.tmp[14:19]
     } else {
     theta.tmp <- theta + (mult*asym*parmsvector/sqrt(tmp.n))
     # Loadings:
@@ -105,11 +110,14 @@ dgp1 <- function(nobs = 240, diff = 1.5, nlevels = 3, gradual=FALSE, anomaly=FAL
     tmp.psi <- matrix(0,6,6)
     diag(tmp.psi) <- theta.tmp[7:12]  
     }
+
+    # intercepts
+    tmp.mu <- theta.tmp[14:19]
     
     u.tmp <- t(rmvnorm(tmp.n,rep(0,6),tmp.psi))
[…]

### 2014-01-24 frontiers/frontiers.Rnw
Edited comments and frontiers manuscript.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@303 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2013-10-31 frontiers/frontiers.Rnw
@@
 parameters are replaced with informative prior distributions.  In this paper, we
-describe an alternative approach: the development of test statistics that are
-sensitive to invariance violations of interest and insensitive to ``anomalous''
-invariance violations.  The test statistics are specifically applicable to
-situations where one wishes to test for measurement invariance with respect to
-an ordinal variable, and they are special cases of a family of tests that may
-be used to study measurement invariance w.r.t.\ continuous,
-categorical, and ordinal variables.
+describe an alternative approach: the development of test statistics that 
+are especially sensitive to violations that are monotonic w.r.t. the
+variable of interest (and 
+less sensitive to violations that are non monotonic w.r.t. the
+variable of interest).  Because monotonic violations are more likely to be
+interpretable and interesting to the researcher, we can gain more
+power to detect these violations by de-emphasizing other types of violations.
+The resulting test statistics are specifically applicable to
+situations where one  
+wishes to test for measurement invariance with respect to an ordinal variable, 
+and they are special cases of a family of tests that may be used to study 
+measurement invariance w.r.t.\ continuous, categorical, and ordinal variables
+\cite{mer93,DolMaa98,lubmut05}\readme{The new references do not work
+  here (because they are unrelated to tests based on stochastic
+  processes).  I will work on these references.}.
 
@@
 maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
-$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$. The cross-product
-of these scores forms the ``meat'' for the calculation of robust
-(Huber-White) standard errors \cite<e.g.,>{zei06b}.
+$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$.
 
@@
 individual has $k$ scores describing the extent to which the fitted
-model describes that particular individual.  Scores close to zero
-indicate a ``good'' description, and scores far from zero indicate a
-``bad'' description.  Each of an individual's $k$ scores represent one
-model parameter, so that the scores provide specific information about
-each parameter's fit to each individual.
-
-To use these scores for testing, we order individuals according to
+model describes that particular individual.  We can roughly interpret
+scores near zero as providing a ``good'' description of an individual,
+with scores far from zero providing a ``bad'' description of an
+individual.  This is only a rough interpretation as, even when measurement
+invariance holds, some individuals' scores will be further from zero
+than others.  Each of the $k$ scores represents one model parameter,
+which, as further described below, allows us to test subsets of model
+parameters for invariance.
+
+To use the scores for testing, we order individuals according to
 an auxiliary variable $V$ (the variable against which we are testing 
@@
 for measurement 
-invariance w.r.t.\ age.  If the manifest variables violate measurement
-invariance here, then some parameter estimates may be too large for
+invariance w.r.t.\ age.  If there exists measurement invariance
+w.r.t. age, then some parameter estimates may be too large for 
 young individuals and too small for old individuals (say).  This
@@
 the integer part of $nt$ (i.e., a floor 
-operator), and $x_{(i)}$ reflects the individual with the
-$i$-th smallest value of the auxiliary variable $V$.
-We specifically focus on how the
+operator), and $x_{(i)}$ reflects the individual with the $i$-th
+smallest value of the auxiliary variable $V$.  While the above
+equation is written in general form, we can
+restrict the value of $t$ in finite samples to the set $\{0,
+1/n, 2/n, 3/n, \ldots, n/n\}$.
+We focus on how the
 cumulative sum fluctuates as more individuals' scores are added to it, e.g.,
@@
 root of the information matrix, which serves to decorrelate the
-fluctuation processes associated with individual model parameters.
-Thus, the process associated with one parameter is not correlated 
-with the processes associated with other parameters.
+fluctuation processes associated with individual model parameters
+while preserving the behavior of individual parameters' fluctuations..
 
@@
 used to show that the fluctuation of the above cumulative sum follows a
-Brownian bridge.  This result allows us to calculate $p$-values and
+Brownian bridge \cite{hjokon02}.  % TODO add ref to Brownian bridge  
+This result allows us to calculate $p$-values and
 critical values for test statistics under the hypothesis of
@@
 package \pkg{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
-gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
+gratitude scales from $n=1401$ youth aged 10--19 years. 
+The original authors 
 were specifically interested in whether the scales were measurement invariant 
@@
 We focus on measurement invariance of the factor loadings associated with 
-the GQ-6 scale, using a one-factor model. While the ``age group'' variable against 
+the GQ-6 scale, using a one-factor model.  This scale consists of five
+items scored from 1--9.  While the ``age group'' variable against 
 which we are testing measurement invariance is best considered ordinal, 
@@
 %
-Next, we fit two models in \pkg{lavaan}: a one-factor model where loadings are restricted 
-to be equal across age groups, and a one-factor model where loadings are free across age groups.
+Next, we fit two models in \pkg{lavaan}: a one-factor model 
+where loadings are restricted to be equal across age groups, 
+and a one-factor model where loadings are free across age groups.
+This allows us to test a hypothesis of partial measurement invariance
+that was of interest to the original researchers (though, for ordinal
+variables, all types of
+measurement invariance can be examined via the tests described
+previously).
+By default, the code below sets the scale by fixing the first loading
+to 1.
 %
 <<models>>=
-restr <- cfa("f1 = ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
   data = yg, group = "agegroup", meanstructure = TRUE,
   group.equal = "loadings")
-full <- cfa("f1= ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+full <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
   data = yg, group = "agegroup", meanstructure = TRUE)
@@
 Because our sample size is large, the likelihood ratio test is known to be sensitive to 
-small measurement invariance violations \cite{Ben80}.  That is, the above tests 
-are sensitive to small measurement invariance violations that are not likely to be 
+small measurement invariance violations \cite{Ben80}.  That is, the
+LRT and  $\mathit{LM}_\mathit{uo}$ test from~\eqref{eq:lmuo}
+are sensitive to small measurement invariance violations that are not
+likely to be  
 of interest to researchers.  For example, imagine that the 15-year-olds' 
@@
 introduction \cite{muthen13}.
-Alternatively, we can use the proposed family of score-based statistics to obtain
+Alternatively, we can use the ``ordinal'' score-based
+statistics (from~\eqref{eq:wdm} and~\eqref{eq:maxlmo}) to obtain 
 tests that are sensitive to the ordering of age.
@@
  ($\lambda_{11}$), the unique variance associated with Scale 1
- ($\psi_{11}$), or the factor covariance $\phi_{12}$.  
- We then tested for measurement invariance in five subsets of parameters: each of the three
+ ($\psi_{11}$), or the factor covariance $\phi_{12}$.  Note that the
+ latter violation is not necessarily a measurement invariance
+ violation, but it is still a parameter instability that can occur in
+ this type of model.
+ We then tested for measurement invariance (parameter instability) in
+ five subsets of parameters: each of the three 
 individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 

### 2014-01-25 frontiers/frontiers.Rnw
set up new simulations including intercept.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@304 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   data = yg, group = "agegroup", meanstructure = TRUE,
-  group.equal = "loadings")
+  group.equal = "intercepts")
+
 full <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
@@
 invariance, it is plausible that this violation impacts other
-parameter estimates, including factor covariances or the unique variance
-associated with the manifest variable in question.  Thus, the goal of the
-Simulation 1 is to examine the extent to which the proposed tests
-attribute the measurement invariance violation to the parameters that
-are truly in violation.
+parameter estimates, including factor covariances, intercepts or 
+the unique variance associated with the manifest variable in question.  
+Thus, the goal of the Simulation 1 is to examine the extent to which 
+the proposed tests attribute the measurement invariance violation 
+to the parameters that are truly in violation.
 
@@
 The measurement invariance
-violation occurred in one of three places: the factor loading associated with Scale 1 
- ($\lambda_{11}$), the unique variance associated with Scale 1
- ($\psi_{11}$), or the factor covariance $\phi_{12}$.  Note that the
+violation occurred in one of four places: the factor loading associated with Scale 1 
+ ($\lambda_{11}$), the unique variance  ($\psi_{11}$)
+ or intercept associated with Scale 1 ($\mu_{11}$)
+, the factor covariance ($\phi_{12}$).  Note that the
  latter violation is not necessarily a measurement invariance
@@
  We then tested for measurement invariance (parameter instability) in
- five subsets of parameters: each of the three 
+ seven subsets of parameters: each of the four
 individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
-and all six unique variances. % $\epsilon_{11}--\epsilon_{66}$.
+all six unique variances and all six intercepts. % $\epsilon_{11}--\epsilon_{66}$.
 
@@
 Full simulation results are presented in Figures~\ref{fig:sim11res}
-to~\ref{fig:sim13res}.
+to~\ref{fig:sim14res}.
 Figure~\ref{fig:sim11res} displays power curves as a function of 
@@
 and lines reflecting different test statistics.
-Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
+Figure~\ref{fig:sim12res}, Figure~\ref{fig:sim13res} and 
+Figure~\ref{fig:sim14res} display 
 similar power curves when the factor covariance 
-$\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
+$\phi_{12}$, error variance $\epsilon_{11}$ and intercept
+$\mu_{11}$ violate measurement
 invariance, respectively. In these figures, we generally show tests
@@
 Within each panel of Figure~\ref{fig:sim11res} to 
-Figure~\ref{fig:sim13res}, the three lines reflect the three test
+Figure~\ref{fig:sim14res}, the three lines reflect the three test
 statistics.  It is seen that the two ordinal statistics exhibit 
@@
 in the factor loading and factor covariance parameters, as opposed to the error variance
-parameter.  Finally, simultaneous tests of all factor loadings or all error
-parameters result in decreased power, as compared to the situation
+parameter\seeme{TODO:intercept discription}.  
+Finally, simultaneous tests of all factor loadings, all intercepts or 
+all error parameters result in decreased power, as compared to the situation
 where one tests only the violating parameter.  This occurs because, in
@@
 } else {
-  sim1 <- simulation(sim = "sim1", nobs = 480, parms = c("loading", "error", "var"))
+  sim1 <- simulation(sim = "sim1", nobs = 480, parms = c("loading", "error", "var", "intercept"))
   sim1$nlevels <- factor(sim1$nlevels)
@@
 parlabs <- c(expression(lambda[11]), expression(phi[12]), 
-             expression(psi[11]), expression(list(lambda[11], ldots, lambda[62])), 
-             expression(list(psi[11], ldots, psi[66])))
-levels(sim1$pars) <- c("Loading1", "Covariance", "Error1", "All Loadings", "All Errors")  
+             expression(psi[11]), expression(mu[11]), expression(list(lambda[11], ldots, lambda[62])), 
+             expression(list(psi[11], ldots, psi[66])), 
+             expression(list(mu[11], ldots, mu[66])))
+levels(sim1$pars) <- c("Loading1", "Covariance", "Error1", "Intercept1", 
+                       "All Loadings", "All Errors", "All Intercepts")  
 @
@@
            if(which.given == 2){
-             strip.default(which.given, factor.levels = parlabs[c(1, 3, 4)], ...)
+             strip.default(which.given, factor.levels = parlabs[c(1, 3, 5)], ...)
 	   } else {
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(3,5)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(3,6)], ...)
+	   } else {
+	       strip.default(which.given, factor.levels = factor.levels, ...)
+	   }
+       })
+@
+
+\end{figure}
+
+\begin{figure}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
+parameter violating measurement invariance is $\mu_{11}$. Panel labels
+denote the parameter(s) 
+being tested and the number of levels of the ordinal variable $m$.}
+\label{fig:sim14res}
+<<sim14res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+       subset = (parms == "intercept" & 
+		 pars %in% c("Intercept1", "All Intercepts") & diff %% 0.5 == 0),
+       type = "b", xlab = expression(paste("Violation Magnitude (", mu[11], ")")),
+       ylab = "Power", key = mykey, as.table = TRUE,
+       strip = function(..., which.given, factor.levels){
+	   if(which.given == 2){
+	       strip.default(which.given, factor.levels = parlabs[c(4,7)], ...)
 	   } else {
@@
 A measurement invariance violation could occur in
-each of the three parameters from Simulation 1 (factor loading, factor
-covariance, and unique variance), and a violation could also occur in
+each of the four parameters from Simulation 1 (factor loading, factor
+covariance, unique variance and intercept), and a violation could also occur in
 the extra, unmodeled loading. In each condition, a single parameter
@@
 associated with the factor 
-covariance did not demonstrate this error, however.  In
+covariance did not demonstrate this error, however\seeme(TODO:intercepts descrip).  In
 terms of specific 
@@
 the misspecification. Similar results were observed
-when the unique variance and factor covariance parameters violated measurement
+when the unique variance, intercept and factor covariance parameters violated measurement
 invariance; these results are all shown in the Appendix.
@@
   sim2 <- simulation(sim = "sim2", nobs = 480, 
-  parms = c("extra", "extra+loading", "extra+var", "extra+error"))
+  parms = c("extra", "extra+loading", "extra+var", "extra+error", "extra+intercept"))
 
@@
   parlabs <- c(expression(lambda[11]), expression(phi[12]), 
-               expression(psi[11]),
+               expression(psi[11]), expression(mu[11]),
 	       expression(list(lambda[11], ldots, lambda[62])), 
-               expression(list(psi[11], ldots, psi[66])))
-  levels(sim2$pars) <- c("Loading1", "Covariance", "Error1", "All Loadings", "All Errors")
+               expression(list(psi[11], ldots, psi[66])),
+               expression(list(mu[11], ldots, mu[66])))
+  levels(sim2$pars) <- c("Loading1", "Covariance", "Error1", "Intercept1", 
+                         "All Loadings", "All Errors", "All Intercepts")
 
@@
        subset = (parms == "extra" & 
-		 pars %in% c("Loading1", "Covariance", "Error1") & diff %% 0.5 == 0),
+		 pars %in% c("Loading1", "Covariance", "Error1", "Intercept1") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude (Unmodeled Loading)",
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 2, 3)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(1, 2, 3, 4)], ...)
 	   } else {
@@
 model misspecification (lacking one loading from Scale 1 to
-Math). Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim25res} display
+Math). Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim26res} display
 results when the unmodeled loading, the loading $\lambda_{11}$, the
-covariance $\phi_{12}$, and the error term $\psi_{11}$ violate
-invariance, respectively.  Results are only shown for simulation
+covariance $\phi_{12}$, the error term $\psi_{11}$ and intercept $\mu_{11}$
+violate invariance, respectively.  Results are only shown for simulation
 conditions exhibiting power curves that increased from zero.  These
@@
        subset = (parms == "extra" & 
-		 pars %in% c("All Loadings", "All Errors") & diff %% 0.5 == 0),
+		 pars %in% c("All Loadings", "All Errors", "All Intercepts") & diff %% 0.5 == 0),
        type = "b", xlab = "Violation Magnitude (Unmodeled Loading)", ylab = "Power", key = mykey, as.table = TRUE,
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(4, 5)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(5, 6, 7)], ...)
 	   } else {
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 4)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(1, 5)], ...)
 	   } else {
@@
 	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(3,5)], ...)
+	       strip.default(which.given, factor.levels = parlabs[c(3,6)], ...)
 	   } else {
@@
 
+\begin{figure}
+\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
+parameter violating measurement invariance is $\mu_{11}$.  Panel
+labels denote the parameter(s)  
+being tested and the number of levels of the ordinal variable $m$.}
+\label{fig:sim26res}
+<<sim26res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+       subset = (parms == "extra+intercept" & 
+		 pars %in% c("Intercept1", "All Intercepts") & diff %% 0.5 == 0),
+       type = "b", xlab = expression(paste("Violation Magnitude (", mu[11], ")")),
+       ylab = "Power", key = mykey, as.table = TRUE,
+       strip = function(..., which.given, factor.levels){
+	   if(which.given == 2){
+	       strip.default(which.given, factor.levels = parlabs[c(4,7)], ...)
+	   } else {
+	       strip.default(which.given, factor.levels = factor.levels, ...)
+	   }
+       })
+@
+
+\end{figure}
+
 \end{document}

### 2014-01-25 frontiers/frontiers.Rnw
run a small simulation to make sure the frontiers.Rnw is set up correct.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@305 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   data = yg, group = "agegroup", meanstructure = TRUE,
-  group.equal = "intercepts")
+  group.equal = "loadings")
 
@@
 have somewhat higher power to detect measurement invariance violations 
-in the factor loading and factor covariance parameters, as opposed to the error variance
-parameter\seeme{TODO:intercept discription}.  
+in the factor loading, factor covariance and intercept parameters, 
+as opposed to the error variance parameter.  
 Finally, simultaneous tests of all factor loadings, all intercepts or 
@@
 associated with the factor 
-covariance did not demonstrate this error, however\seeme(TODO:intercepts descrip).  In
+covariance and intercept did not demonstrate this error, however.  In
 terms of specific 

### 2014-01-26 frontiers/frontiers.Rnw
work on the comments and Rnw file.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@306 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 and performance under
-scenarios likely to be encountered in practice. 
+scenarios likely to be encountered in practice.  To our knowledge, this 
+is the first detailed examination of the tests' properties. \seeme{add
+this sentence to point the unique contribution.}
 %In order to fix the psychometric scale, we need to specify the violating parameter.  
@@
 and a one-factor model where loadings are free across age groups.
-This allows us to test a hypothesis of partial measurement invariance
+This allows us to test a hypothesis of partial\seeme{I
+think partial should be weak} measurement invariance
 that was of interest to the original researchers (though, for ordinal

### 2014-01-30 frontiers/frontiers.Rnw
update one question about the motivation for intercept.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@310 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
  latter violation is not necessarily a measurement invariance
- violation, but it is still a parameter instability that can occur in
+ violation\seeme{need Meridith reference?}, but it is still a parameter instability that can occur in
  this type of model.

### 2014-01-30 frontiers/frontiers.Rnw
Edited half of the frontiers paper and comments.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@311 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2014-01-25 frontiers/frontiers.Rnw
@@
 and they are special cases of a family of tests that may be used to study 
-measurement invariance w.r.t.\ continuous, categorical, and ordinal variables
-\cite{mer93,DolMaa98,lubmut05}\readme{The new references do not work
-  here (because they are unrelated to tests based on stochastic
-  processes).  I will work on these references.}.
+measurement invariance w.r.t.\ continuous, categorical, and ordinal
+variables.  Other methods for studying measurement
+invariance w.r.t.\ continuous variables include moderated factor
+models \cite{bauhus09,moldol10,pur02} and factor mixture models
+\cite{dolmaa98,lubmut05}.  These methods require a model of greater
+complexity to be estimated, while the tests described here work solely
+on the output of a traditional factor model.  See \citeA{MerZei13} for
+further comparison of these methods.
 
@@
 and performance under
-scenarios likely to be encountered in practice. 
+scenarios likely to be encountered in practice.  While the previous
+papers have described and studied the tests under ideal conditions, we
+focus here on two topics of interest to the applied researcher:
+software considerations for carrying out the tests, and test
+performance under model misspecification.
 %In order to fix the psychometric scale, we need to specify the violating parameter.  
@@
 invariance holds, some individuals' scores will be further from zero
-than others.  Each of the $k$ scores represents one model parameter,
+than others.  However, under measurement noninvariance, the scores
+will differ for different subgroups of individuals (say, scores in
+subgroup A tend to be negative and scores in subgroup B tend to be
+positive).  Each of the $k$ scores represents one model parameter,
 which, as further described below, allows us to test subsets of model
@@
 fluctuation processes associated with individual model parameters
-while preserving the behavior of individual parameters' fluctuations..
+while preserving the behavior of individual parameters' fluctuations.
 
@@
 and a one-factor model where loadings are free across age groups.
-This allows us to test a hypothesis of partial measurement invariance
+This allows us to test a hypothesis of weak measurement invariance
 that was of interest to the original researchers (though, for ordinal
@@
  latter violation is not necessarily a measurement invariance
- violation, but it is still a parameter instability that can occur in
+ \cite<e.g.,>{mer93}, but it is still a parameter instability that can occur in
  this type of model.

### 2014-02-03 frontiers/frontiers.Rnw
Frontiers paper: revisions to paper, response letter, and refs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@312 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 variable of interest (and 
-less sensitive to violations that are non monotonic w.r.t. the
+less sensitive to violations that are non-monotonic w.r.t. the
 variable of interest).  Because monotonic violations are more likely to be
@@
 measurement invariance w.r.t.\ continuous, categorical, and ordinal
-variables.  Other methods for studying measurement
-invariance w.r.t.\ continuous variables include moderated factor
+variables.
+
+The study of measurement invariance w.r.t.\ categorical auxiliary 
+variables (via, e.g., likelihood ratio tests) is popular and well
+known, and ordinal auxiliary variables are typically treated as
+categorical in measurement invariance contexts.
+The study of measurement invariance w.r.t.\ continuous
+variables is newer and less known: along with the family
+described here, other methods include moderated factor
 models \cite{bauhus09,moldol10,pur02} and factor mixture models
-\cite{dolmaa98,lubmut05}.  These methods require a model of greater
-complexity to be estimated, while the tests described here work solely
-on the output of a traditional factor model.  See \citeA{MerZei13} for
-further comparison of these methods.
-
-The family of tests described in this paper have recently been applied
+\cite{dolmaa98,lubmut05}.  These methods require estimation of a model
+of greater complexity, while the tests described in this paper work solely
+on the output of a traditional factor model \cite<see>[for
+further comparison of these methods]{MerZei13}.
+
+The family of tests described here have recently been applied
 to the study of measurement invariance \cite{MerZei13,MerFanZei}, though their
 practical application has been limited to a small set of simulations and
-datasets.  In this paper, we provide a detailed illustration of the tests' use
-and performance under
+data examples.  In this paper, we provide a detailed illustration of
+the tests' use and performance under
 scenarios likely to be encountered in practice.  While the previous
@@
 practical research scenarios.  Finally, we provide some further discussion on
-the tests' use in practice.
+the tests' use in practice.  We acknowledge 
+problems associated with $p$-values (e.g., corrections for multiple
+tests, researcher degrees of freedom, interpretations, etc) but
+ignore them throughout the paper for expository purposes.
 
@@
 
-To verbally describe the above equation, each
+To verbally describe Equation~\eqref{eq:score}, each
 individual has $k$ scores describing the extent to which the fitted
@@
 which, as further described below, allows us to test subsets of model
-parameters for invariance.
+parameters for invariance.  While scores can be obtained
+under the multivariate normal likelihood (discrepancy) function and
+alternatives such as generalized least squares, most SEM software
+fails to supply the scores to the user.
 
@@
 for measurement 
-invariance w.r.t.\ age.  If there exists measurement invariance
+invariance w.r.t.\ age.  If there exists measurement non-invariance
 w.r.t. age, then some parameter estimates may be too large for 
@@
 by setting 
-\code{plot = TRUE} in the \code{sctest()} calls above. The resulting plots representing
-the ordinal statistics' fluctuations across levels of age group are displayed in
-Figure~\ref{fig:ordres}.
-These plots display the loadings' fluctuation across age groups, with the x-axis
-reflecting age group and the y-axis reflecting test statistic values (larger
-values reflect more instability).  The dashed horizontal lines reflect critical values, so
-that the hypothesis of measurement invariance is rejected if the sequence of
+\code{plot = TRUE} in the \code{sctest()} calls
+above. Figure~\ref{fig:ordres} displays the resulting plots, which represent
+the ordinal statistics' fluctuations across levels of age group.
+The x-axis
+reflects age group and the y-axis reflects test statistic values (larger
+values reflect more instability), with the dashed horizontal lines
+reflecting critical values.  The
+hypothesis of measurement invariance is rejected if the sequence of
 test statistics crosses the critical value.  While the measurement invariance tests are
@@
 In Simulation 1, we examined the extent to which the proposed tests
-are sensitive to the specific model parameter that violates measurement
-invariance.  If, say, a factor loading violates measurement
+can ``localize'' a measurement invariance violation.
+If, say, a factor loading violates measurement
 invariance, it is plausible that this violation impacts other
@@
 violation occurred in one of four places: the factor loading associated with Scale 1 
- ($\lambda_{11}$), the unique variance  ($\psi_{11}$)
- or intercept associated with Scale 1 ($\mu_{11}$)
-, the factor covariance ($\phi_{12}$).  Note that the
+ ($\lambda_{11}$), the intercept ($\mu_{11}$),
+ the unique variance  ($\psi_{11}$), or the factor covariance
+ ($\phi_{12}$).  Note that the 
  latter violation is not necessarily a measurement invariance
@@
 and lines reflecting different test statistics.
-Figure~\ref{fig:sim12res}, Figure~\ref{fig:sim13res} and 
+Figure~\ref{fig:sim12res}, Figure~\ref{fig:sim13res}, and 
 Figure~\ref{fig:sim14res} display 
 similar power curves when the factor covariance 
-$\phi_{12}$, error variance $\epsilon_{11}$ and intercept
+$\phi_{12}$, error variance $\epsilon_{11}$, and intercept
 $\mu_{11}$ violate measurement
@@
 have somewhat higher power to detect measurement invariance violations 
-in the factor loading, factor covariance and intercept parameters, 
+in the factor loading, factor covariance, and intercept parameters, 
 as opposed to the error variance parameter.  
-Finally, simultaneous tests of all factor loadings, all intercepts or 
+Finally, simultaneous tests of all factor loadings, all intercepts, or 
 all error parameters result in decreased power, as compared to the situation
@@
 testing a subset of parameters (only one of which violates measurement
-invariance), we are effectively dampening the signal of a measurement
+invariance), we are dampening the signal of a measurement
 invariance violation. This ``dampening'' effect is more apparent for
@@
 each of the four parameters from Simulation 1 (factor loading, factor
-covariance, unique variance and intercept), and a violation could also occur in
+covariance, unique variance, and intercept), and a violation could also occur in
 the extra, unmodeled loading. In each condition, a single parameter
@@
 ``power,'' which is actually a high Type I error rate here.  This Type
-I error also demonstrated when testing all loadings and all unique
-variances, as shown in the appendix.  Tests
+I error is also observed when testing all loadings and all unique
+variances (see the Appendix).  Tests
 associated with the factor 
@@
 invariance; these results are all shown in the Appendix.
-In particular, the power of ordered statistics $\max \mathit{LM}_o$ test and
-$\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in
-each of these panels.
+In particular, power of the ordered statistics was higher than power
+of the unordered statistic in 
+each panel.
 
@@
 study the extent to 
-which the hypothesized model includes all parameters of importance.
-None of these tests (score-based or otherwise) inform the researcher of
-model misspecification.
+which the hypothesized model includes all parameters of importance
+(i.e., the extent to which the model is well specified).
+
+One could begin to study model misspecification by fitting models with
+different discrepancy functions (say, a multivariate normal function and a
+generalized least squares function).  If parameter estimates differ
+greatly across the functions, then this implies model
+misspecification.  Additionally, if one has a large sample size, one
+could split the data into subgroups and examine consistency of results
+across subgroups.  These issues are important for all the tests
+discussed here (score-based or otherwise).
 
@@
 
-In the remainder of the paper, we first compare
-these tests to traditional tests of measurement invariance.  We then 
+In the remainder of the paper, we first compare the use
+these tests to the use of traditional tests in practice.  We then 
 discuss test extension to other fit functions and to other specialized
@@
 such as nations or gender.  As discussed earlier in this paper, the score-based
-tests for unordered categories are essentially equivalent to the usual
+tests for unordered categories are equivalent to the usual
 likelihood ratio test.  Given a measurement invariance violation across these
@@
 application of these tests to the Rasch model, using them as the basis of a
-recursive partioning procedure that segments subgroups of individuals who
-exhibit DIF.  Further study and extension of these tests for IRT seem warranted.
+recursive partitioning procedure that segments subgroups of individuals who
+exhibit DIF.  Further study and extension of these tests for IRT are warranted.
 
@@
 model misspecification (lacking one loading from Scale 1 to
-Math). Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim26res} display
+Math). Figures~\ref{fig:sim22res} to~\ref{fig:sim26res} display
 results when the unmodeled loading, the loading $\lambda_{11}$, the
-covariance $\phi_{12}$, the error term $\psi_{11}$ and intercept $\mu_{11}$
+covariance $\phi_{12}$, the error term $\psi_{11}$, and the intercept $\mu_{11}$
 violate invariance, respectively.  Results are only shown for simulation

### 2014-02-10 frontiers/frontiers.Rnw
small modifications and proposed change in response 2.6
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@313 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor changes to ms in response to comment 2.6.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@314 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 practical research scenarios.  Finally, we provide some further discussion on
-the tests' use in practice.  We acknowledge 
-problems associated with $p$-values (e.g., corrections for multiple
-tests, researcher degrees of freedom, interpretations, etc) but
-ignore them throughout the paper for expository purposes.
+the tests' use in practice.
+% We acknowledge problems associated with $p$-values (e.g., corrections for multiple
+% tests, researcher degrees of freedom, interpretations, etc) but
+% ignore them throughout the paper for expository purposes.
 
@@
 we test for measurement invariance across these age groups.
+For illustrative purposes, we conduct multiple tests and
+compare them to the traditional significance level of .05.  In
+practice, however, one should generally adjust the significance level
+for the number of tests carried out.
 

### 2014-02-10 frontiers/Supple_Material.Rnw
add Supple_Material.Rnw file.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@315 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2014-01-30 frontiers/frontiers.Rnw
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%Editing Rules:                                                            %
-%1. Ed's comment and questions:\readme{}                                   %     
-%2. Ting's comment and questions:\seeme{}                                  % 
-%3. When comments and questions are resovled, the corresponding \readme or %  
-%   \seeme will be removed by Ting.                                        %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \documentclass[doc]{apa}
@@
 
-% Seems necessary for texlive 2012?
-\usetikzlibrary{arrows}
-
 \title{Score-based tests of measurement invariance: Use in practice}
@@
 \twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}
-
-\abstract{
-In this paper, we consider a family of  recently-proposed measurement invariance
-tests that are based on the {\em scores} of a fitted model.  This family can be used
-to test for measurement invariance w.r.t.\ a continuous auxiliary variable,
-without pre-specification of subgroups. Moreover, the family can be used when
-one wishes to test for  measurement invariance w.r.t.\ an ordinal auxiliary
-variable, yielding test statistics that are sensitive to violations that are
-monotonically related to the ordinal variable (and less sensitive to
-non-monotonic violations).  The paper is specifically aimed at potential users
-of the tests who may wish to know (i) how the tests can be employed for their
-data, and (ii) whether the tests can accurately identify specific models parameters that violate measurement invariance (possibly in the presence of model misspecification).  After providing an
-overview of the tests, we illustrate their general use via the \proglang{R} packages \pkg{lavaan}
-and \pkg{strucchange}.  We then describe two novel simulations that provide evidence
-of the tests' practical abilities.  As a whole, the paper provides researchers
-with the tools and knowledge needed to apply these tests to general measurement
-invariance scenarios.
-}
 
@@
   Email: merklee@missouri.edu.}
+
 \shorttitle{Score-based tests in practice}
@@
 
-%% for internal use
-\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
-\newcommand{\readme}[1]{\emph{\marginpar{Ed} (#1)}}
-\newcommand{\seeme}[1]{\emph{\marginpar{Ting} (#1)}}
-
 \spacing{1}
@@
 \begin{document}
+
 \maketitle
@@
 @
-
-Many of the papers in this special issue focus on the topic of approximate
-measurement invariance: we know that strict hypotheses of measurement invariance
-do not hold exactly across different groups, and this should be reflected in
-corresponding tests of measurement invariance.  Under a Bayesian
-approach, we may utilize the idea of approximate invariance
-\cite<e.g.,>{muthen13}, whereby across-group equality constraints on
-parameters are replaced with informative prior distributions.  In this paper, we
-describe an alternative approach: the development of test statistics that 
-are especially sensitive to violations that are monotonic w.r.t. the
-variable of interest (and 
-less sensitive to violations that are non monotonic w.r.t. the
-variable of interest).  Because monotonic violations are more likely to be
-interpretable and interesting to the researcher, we can gain more
-power to detect these violations by de-emphasizing other types of violations.
-The resulting test statistics are specifically applicable to
-situations where one  
-wishes to test for measurement invariance with respect to an ordinal variable, 
-and they are special cases of a family of tests that may be used to study 
-measurement invariance w.r.t.\ continuous, categorical, and ordinal
-variables.  Other methods for studying measurement
-invariance w.r.t.\ continuous variables include moderated factor
-models \cite{bauhus09,moldol10,pur02} and factor mixture models
-\cite{dolmaa98,lubmut05}.  These methods require a model of greater
-complexity to be estimated, while the tests described here work solely
-on the output of a traditional factor model.  See \citeA{MerZei13} for
-further comparison of these methods.
-
-The family of tests described in this paper have recently been applied
-to the study of measurement invariance \cite{MerZei13,MerFanZei}, though their
-practical application has been limited to a small set of simulations and
-datasets.  In this paper, we provide a detailed illustration of the tests' use
-and performance under
-scenarios likely to be encountered in practice.  While the previous
-papers have described and studied the tests under ideal conditions, we
-focus here on two topics of interest to the applied researcher:
-software considerations for carrying out the tests, and test
-performance under model misspecification.
-%In order to fix the psychometric scale, we need to specify the violating parameter.  
-%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
-%and may interfere with parameter invariance specificity. A family of tests derived from 
-%stochastic process has been recently proposed to identify specific parameters impacted 
-%by measurement invariance violation. However, these tests have only been used to a limited 
-%type of parameter violation. It is this paper's intent to provide a detailed examination 
-%of the tests' performance under practical research scenarios. 
-In the following sections, we first briefly review the  theoretical framework of
-the proposed tests and provide a short tutorial illustrating the use of the
-tests in \proglang{R} \cite{R11}. Next, we study the tests' performance in simulations that mimic
-practical research scenarios.  Finally, we provide some further discussion on
-the tests' use in practice.
-
-\section{Background}
-
-This section contains background and discussion of the  proposed statistics as
-applied to structural equation models (SEMs); for a more detailed account, see \citeA{MerZei13} and
-\citeA{MerFanZei}.  For details on the statistics' application to general
-statistical models, see \citeA{zeihor07}.
-
-As currently implemented for SEM, the statistical tests described in
-this paper can be applied to models that are estimated via a
-multivariate normal or Wishart likelihood (or discrepancy) function,
-with extension to other discrepancy functions being conceptually straightforward. 
-The tests are carried out following model estimation, making use of output
-associated with the fitted model.  In general, we fit a model that
-restricts parameters to be equal across observations, then carry out a
-posthoc test to examine whether specific parameters vary
-across observations.  This procedure is similar in spirit to the
-calculation of modification indices \cite{bentler90} and to Lagrange multiplier
-tests \cite{sat89}, and, in fact, those statistics can be viewed as  
-special cases of the family described here.
-
-Following model estimation, the tests primarily work on the \emph{scores}
-of the fitted model; these are defined as
-%% TODO indicate that this is evaluated at theta hat
-\begin{equation}
-\label{eq:score}
-s(\theta;x_{i}) = \left (
-  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{1}}, \dots,
-  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right
-  )^\top,\ i=1,\ldots,n,
-\end{equation} 
-where $\ell(\theta; x_i)$ is the likelihood associated with individual
-$i$ and $\theta$ is a $k$-dimensional parameter vector. The corresponding
-maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
-$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$.
-
-To verbally describe the above equation, each
-individual has $k$ scores describing the extent to which the fitted
-model describes that particular individual.  We can roughly interpret
-scores near zero as providing a ``good'' description of an individual,
-with scores far from zero providing a ``bad'' description of an
-individual.  This is only a rough interpretation as, even when measurement
-invariance holds, some individuals' scores will be further from zero
-than others.  However, under measurement noninvariance, the scores
-will differ for different subgroups of individuals (say, scores in
-subgroup A tend to be negative and scores in subgroup B tend to be
-positive).  Each of the $k$ scores represents one model parameter,
-which, as further described below, allows us to test subsets of model
-parameters for invariance.
-
-To use the scores for testing, we order individuals according to
-an auxiliary variable $V$ (the variable against which we are testing 
-measurement invariance) and look for ``trends'' in the
-scores.  For example, imagine that we are testing
-for measurement 
-invariance w.r.t.\ age.  If there exists measurement invariance
-w.r.t. age, then some parameter estimates may be too large for 
-young individuals and too small for old individuals (say).  This
-result would be reflected in the scores, where young individuals'
-scores may be greater than zero and old individuals' scores 
-less than zero (though the sign of the scores will depend on whether a
-function is being minimized or maximized).  Conversely, if measurement
-invariance holds, then all 
-individuals' scores will fluctuate randomly around zero.
-
-To formalize these ideas, we define a
-suitably scaled cumulative sum of the ordered scores.  This may be written as
-\begin{equation} 
-    \label{eq:cumscore}
-  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
-    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
-  \qquad (0 \le t \le 1)
-\end{equation}
-where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
-nt \rfloor$ is 
-the integer part of $nt$ (i.e., a floor 
-operator), and $x_{(i)}$ reflects the individual with the $i$-th
-smallest value of the auxiliary variable $V$.  While the above
-equation is written in general form, we can
-restrict the value of $t$ in finite samples to the set $\{0,
-1/n, 2/n, 3/n, \ldots, n/n\}$.
-We focus on how the
-cumulative sum fluctuates as more individuals' scores are added to it, e.g.,
-starting with the youngest and ending with the oldest individual
-if age is the auxiliary variable of interest.
-The summation is premultiplied by an estimate of the inverse square
-root of the information matrix, which serves to decorrelate the
-fluctuation processes associated with individual model parameters
-while preserving the behavior of individual parameters' fluctuations.
-
-Under the
-hypothesis of measurement invariance, a central limit theorem can be
-used to show that the fluctuation of the above cumulative sum follows a
-Brownian bridge \cite{hjokon02}.  % TODO add ref to Brownian bridge  
-This result allows us to calculate $p$-values and
-critical values for test statistics under the hypothesis of
-measurement invariance.  We can obtain test statistics associated with
-all model parameters and with subsets of model parameters.
-
-%% TODO focus on ordinal
-Multiple test statistics are available, depending on how one
-summarizes the behavior of the cumulative sum of scores.  For example, 
-one could take the absolute maximum that the cumulative sum attains for any
-parameter of interest, resulting in a {\em double max} statistic 
-(the maximum is taken across parameters and individuals).
-Alternatively, one could sum the (squared) cumulative sum across parameters
-of interest and take the maximum or the average across individuals, resulting in a 
-{\em maximum Lagrange multiplier} statistic and \emph{Cram\'{e}r-von Mises} statistic, respectively 
-\cite<see>[for further discussion]{MerZei13}.  These statistics are given by
-\begin{eqnarray}
-    \label{eq:dmax}
-    \mathit{DM} & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} | \\
-        \label{eq:cvm}
-    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
-    \label{eq:maxlm}
-    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
-      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
-\end{eqnarray}
-Critical values associated with $\mathit{DM}$ can be obtained analytically, 
-while critical values associated with the other statistics can be obtained 
-from direct simulation \cite{Zei06} or from more refined techniques \cite{han97}.  
-This issue should not be important to the user, as critical values are obtained 
-directly from the \proglang{R} implementation described later.
-
-Importantly, the above statistics were derived for situations where
-individuals are uniquely ordered according to the auxiliary variable.
-This is not always the case for measurement invariance applications,
-where the auxiliary variable is often ordinal.  To remedy this
-situation, \citeA{MerFanZei} extended the framework to situations
-where one has an ordinal auxiliary variable of interest.  Essentially,
-one allows all individuals with the same value of the auxiliary
-variable to enter into the cumulative sum at the same time.  Analogous test 
-statistics are then computed, with
-modified critical values being adopted to reflect the change in the
-statistics' computation.
-
-For an ordinal auxiliary variable with $m$ levels, these modifications are 
-based on $t_{\ell}$ $(\ell=1,\ldots,m-1)$, which are the empirical, cumulative 
-proportions of individuals observed at the first $m-1$ levels.
-The modified statistics are then given by
-\begin{eqnarray}
-    \label{eq:wdm}
-      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
-                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
-    \label{eq:maxlmo}
-      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
-                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
-\end{eqnarray}
-where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$ $(\ell=1,\ldots,m-1)$.
-Critical values associated with the $\mathit{WDM}_o$ statistic can be obtained 
-directly from a multivariate normal distribution \cite<see>{hotzei08}, while 
-critical values associated with $\max \mathit{LM}_o$ can be obtained via simulation.  
-This simulation is somewhat computationally intensive and, in practice, takes about 
-10 minutes on the authors' computers when 50,000 replications are sampled from the 
-approximate asymptotic distribution.  However, the wait is often worth it, 
-as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic 
-to have more power than the $\mathit{WDM}_o$ statistic and the traditional
-likelihood ratio test statistic when the measurement invariance
-violation is monotonic with the ordinal variable.
-
-Finally, if the auxiliary variable $V$ is only nominal/categorical, the cumulative
-sums of scores can be used to obtain a Lagrange multiplier statistic.
-This test statistic can be formally written as
-\begin{equation}
-    \label{eq:lmuo}
-     \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
-       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
-\end{equation}
-where ${\bm B}(\hat {\bm \theta})_{i_{0}j} = 0$ for all $j$.
-This statistic is asymptotically equivalent to the usual, likelihood
-ratio test statistic, and it is advantageous over the likelihood ratio
-test because it requires estimation of 
-only one model (the restricted model).  We make use of this advantage in the
-simulations, described later.
-
-\section{Tutorial}
-
-In this section, we demonstrate how the above tests can be carried out in \proglang{R}, 
-using the package \pkg{lavaan} \cite{lavaan11} for model estimation and \pkg{strucchange} 
-\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
-the applicability of adult gratitude scales to youth, available in the~\proglang{R} 
-package \pkg{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
-gratitude scales from $n=1401$ youth aged 10--19 years. 
-The original authors 
-were specifically interested in whether the scales were measurement invariant 
-across age.  Because the sample size at each age was unbalanced, the authors 
-created age groups of approximately equal sample size.  In the examples below, 
-we test for measurement invariance across these age groups.
-
-We focus on measurement invariance of the factor loadings associated with 
-the GQ-6 scale, using a one-factor model.  This scale consists of five
-items scored from 1--9.  While the ``age group'' variable against 
-which we are testing measurement invariance is best considered ordinal, 
-for demonstration we consider its treatment as categorical, continuous, and ordinal.  
-Each of these treatments is described below in a separate section.
-
-%% THE FOLLOWING DETAILS ARE REMOVED FOR NOW
-%% In general, a set of scales is defined to be measurement invariant with respect
-%% to an auxiliary variable $V$ if:
-
-%% \begin{equation}
-%% \label{equ:invariance}
-%% f(x_{i}|t_{i},v_{i},...)=f(x_{i}|t_{i},...),
-%% \end{equation}
-
-%% where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
-%% for variable $i$ that the scales purport to measure, and $f$ is the model's 
-%% distributional form \cite{mel89}. If the cut point of $v$ is known in advance, i.e. 
-%% gender, nested multiple group models \cite{bollen98} coupled with likelihood ratio test 
-%% are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
-%% of measurement invariance is
-
-%% \begin{equation}
-%% \label{equ:null}
-%% H_{0}: \theta_{A}=\theta_{B},
-%% \end{equation}
-
-%% The alternative hypothesis is
-
-%% \begin{equation}
-%% \label{equ:alternative}
-%% H_{1}:\theta_{i}=
-%% \begin{cases}
-%% \theta^{(A)} & v_{i}\leq v\\
-%% \theta^{(B)} & v_{i}\geq v
-%% \end{cases}
-%% \end{equation}
-
-%% The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ 
-%% (group A and group B) can be obtained by Maximum Likelihood(ML) 
-%% from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,
-%% respectively. The LR test statistic for the given threshold $v$ is then
-
-%% \begin{equation}
-%% \label{equ:LRT}
-%% LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
-%% {(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
-%% }],
-%% \end{equation}
-
-%% ,which has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
-%% to the number difference of parameters in $\theta$.
-
-\subsection{Categorical Treatment}
-
-Measurement invariance is most often tested using multiple groups models \cite<see, e.g., >{vande12}.  
-This amounts to assuming that our auxiliary variable is categorical (i.e., unordered), 
-which is not true for the age groups in the data.  However, we demonstrate the procedure for completeness.
-
-To conduct the analysis, we first load the data and keep only complete cases for simplicity 
-(though the tests can be applied to incomplete data).
-%
-<<data>>=
-data("YouthGratitude", package = "psychotools")
-compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
-yg <- YouthGratitude[compcases, ]
-@
-%
-Next, we fit two models in \pkg{lavaan}: a one-factor model 
-where loadings are restricted to be equal across age groups, 
-and a one-factor model where loadings are free across age groups.
-This allows us to test a hypothesis of weak measurement invariance
-that was of interest to the original researchers (though, for ordinal
-variables, all types of
-measurement invariance can be examined via the tests described
-previously).
-By default, the code below sets the scale by fixing the first loading
-to 1.
-%
-<<models>>=
-restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-  data = yg, group = "agegroup", meanstructure = TRUE,
-  group.equal = "loadings")
-
-full <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-  data = yg, group = "agegroup", meanstructure = TRUE)
-@
-%
-Finally, we can get the results of a likelihood ratio test via the \code{anova()}
-function, which implies that the GQ-6 violates measurement invariance.
-<<LRT>>=
-anova(full, restr)
-@
-To obtain the asymptotically equivalent $\mathit{LM}_\mathit{uo}$ (Equation~\eqref{eq:lmuo}), we
-can use the \code{sctest()} function from \pkg{strucchange}:
-<<LM_UO>>=
-sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "LMuo")
-@
-This command specifies that we assess the parameter 1--4 of
-model \code{restr} after ordering the observations according to
-\code{agegroup}.  Additionally, 
-the observed information matrix is used as the variance-covariance matrix. Note
-that the model parameters 1--4 are the factor loadings supplied by
-\pkg{lavaan}, which can be seen by inspecting \code{coef(restr)}. This
-also leads to somewhat 
-smaller test statistics that are very close to being significant at the 5\% level.
-
-Because our sample size is large, the likelihood ratio test is known to be sensitive to 
-small measurement invariance violations \cite{Ben80}.  That is, the
-LRT and  $\mathit{LM}_\mathit{uo}$ test from~\eqref{eq:lmuo}
-are sensitive to small measurement invariance violations that are not
-likely to be  
-of interest to researchers.  For example, imagine that the 15-year-olds' 
-parameters are slightly different than the other age groups.  
-The 15-year-olds are in the middle of the age groups, and there 
-is not likely to be any theoretical justification for 15-year-olds 
-differing from every other age group.  One solution to this problem would be the
-Bayesian, approximate invariance methods described in the
-introduction \cite{muthen13}.
-Alternatively, we can use the ``ordinal'' score-based
-statistics (from~\eqref{eq:wdm} and~\eqref{eq:maxlmo}) to obtain 
-tests that are sensitive to the ordering of age.
-
-\subsection{Continuous Treatment}
-
-If we are interested in measurement invariance violations that are 
-monotonic with the age groups, it is perhaps simplest to treat the 
-age groups as continuous.  In doing so, we can use the statistics 
-from Equations~\eqref{eq:dmax}, \eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
-That is, we can fit a model whose parameters 
-are restricted to be equal across all individuals and then examine 
-how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
-age (where age ties are broken arbitrarily, using the original order
-of the observations within each age group).
-This is demonstrated below, with similar code being useful when one is 
-testing for measurement invariance w.r.t.\ truly continuous variables.
-
-%% The multiple group test performs typically well when the $V$ 
-%% is a relatively-small number of categories and the sample size 
-%% is not too large. However, when $V$ (in this case, age) continuous variable, however, 
-%% multiple-group analysis usually cannot be performed because there are 
-%% no existing groups. Instead, we can follow the procedure described in
-%% the Background section.
-
-Again, we employ the \code{sctest()} function to assess parameters~1--4
-from the restricted model \code{restr} after ordering w.r.t.\ \code{agegroup}:
-%
-<<sctest-cont>>=
-dm <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "DM")
-cvm <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "CvM")
-maxlm <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "maxLM")
-c(dm$p.value, cvm$p.value, maxlm$p.value)
-@
-%
-We see that two of the three $p$-values output at the end of the code are
-larger than that associated with the LRT (with the CvM statistic being non-significant).
-
-The tests carried out here assume a unique ordering of individuals by age, but 
-this is obviously not the case.  To compute the statistics and $p$-values, 
-the \pkg{strucchange} package implicitly employed the (arbitrary) ordering of 
-individuals who are tied on age.  If we were to change this ordering, 
-the resulting statistics and $p$-values would also change, potentially
-switching significant results to being non-significant and vice versa.
-Clearly, this is problematic.  
-To accurately account for the multiple observations  
-at the same age level, we must use the ordinal tests from 
-Equations~\eqref{eq:wdm} and~\eqref{eq:maxlmo}.  These are described next.
-
-\subsection{Ordinal Treatment}
-
-%% As a matter of fact, there is only a partial ordering of individuals 
-%% with respect to $V$,i.e. observations with the same level 
-%% of $V$ have no unique ordering. The ordinal statistics 
-%% proposed here are similar to those described in Equation 
-%% and above, except that we focus on "bins" of individuals 
-%% at each level of the ordinal variable.  
-The main difference between the ordinal test statistics and their
-continuous counterparts is that the ordinal statistics are unchanged
-when re-ordering individuals within the same age group.  To compute the test statistics, we allow
-the scores of all tied individuals to enter the cumulative sum
-(Equation~\eqref{eq:cumscore}) simultaneously.  This results in
-modified critical values and test statistics that are sensitive to
-measurement invariance violations that are monotonic w.r.t.\ age group.
-%% Therefore, this ordinal test should 
-%% reflect more real status of the data. In terms of code, we still need the gefp 
-%% function to get the B estimate. After that, we can use different 
-%% functions to aggregate and scale B estimation. 
-%% The function are called ordwmax and ol2bb1.fun, respectively.
-
-To carry out the tests, we can rely on the same function that we used
-for the continuous test statistics.  As mentioned previously,
-calculation of the $\max\mathit{LM}_o$ statistic (Equation~\eqref{eq:maxlmo})
-can be lengthy from the need to simulate critical values (though see
-the end of this section, which provides a partial speed-up).
-%% This is more complicated than necessary, but it is written like
-%% this so that we can load the ordinal critical values behind the
-%% scenes without confusing the reader.
-<<sctest-ord1, echo=FALSE, results=hide>>=
-set.seed(1090)
-wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "WDMo")
-
-if(file.exists("maxLMo.rda")){
-  load("maxLMo.rda")
-} else {
-  RNGkind(kind = "default", normal.kind = "default")
-  set.seed(1090)
-  maxLMo <- ordL2BB(yg$agegroup, nproc = 4)
-  save(maxLMo, file = "maxLMo.rda")
-}
-maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = maxLMo)
-@ 
-<<sctest-ord2, eval=FALSE,strip.white=all>>=
-wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "WDMo")
-maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "maxLMo")
-@ 
-<<sctest-ord3,strip.white=all>>=
-c(wdmo$p.value, maxlmo$p.value)
-@ 
-In computing the ordinal test statistics, we
-obtain~$p=\Sexpr{round(wdmo$p.value, 3)}$
-and~$p=\Sexpr{round(maxlmo$p.value, 3)}$, respectively.\footnote{To replicate
-both $p$-values exactly, \proglang{R}'s random seed needs to be
-set by \code{set.seed(1090)} prior to each \code{sctest()} call.}
-Both $p$-values are clearly larger than that of the 
-likelihood ratio test and neither is significant at $\alpha=0.05$. 
-This provides evidence that there is no measurement
-invariance violation that is monotonic with age group.  Instead, given
-the large sample size, the likelihood ratio test may be overly
-sensitive to anomalous, non-monotonic violations at one (or a few) age groups.
-
-In addition to test statistics, ``instability plots'' can be generated
-by setting 
-\code{plot = TRUE} in the \code{sctest()} calls above. The resulting plots representing
-the ordinal statistics' fluctuations across levels of age group are displayed in
-Figure~\ref{fig:ordres}.
-These plots display the loadings' fluctuation across age groups, with the x-axis
-reflecting age group and the y-axis reflecting test statistic values (larger
-values reflect more instability).  The dashed horizontal lines reflect critical values, so
-that the hypothesis of measurement invariance is rejected if the sequence of
-test statistics crosses the critical value.  While the measurement invariance tests are
-non-significant, the plots imply some instability in the older age groups (15 and~16).
-
-\begin{figure}
-\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and the $\max\mathit{LM}_o$ statistic (right panel).}
-\label{fig:ordres}
-<<sctest-ord-graphics, fig=TRUE, height=4.5, width=9, echo=FALSE, results=hide>>=
-par(mfrow = c(1, 2))
-set.seed(1090)
-sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "WDMo", plot = TRUE, ylim = c(0, 3.2),
-  xlab = "Age group", ylab = "Weighted test statistics",
-  main = expression(WDM[o]), boundary = list(col = 1, lty = 2))
-sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = maxLMo, plot = TRUE, ylim = c(0, 14),
-  xlab = "Age group", ylab = "LM statistics",
-  main = expression(maxLM[o]), boundary = list(col = 1, lty = 2))
-@
-\end{figure}
-
-Finally, if the user anticipates multiple calculations of the
-$\max\mathit{LM}_o$ statistic for a specific dataset, it is possible
-to save time by simulating critical values once and re-using them for
-multiple tests.  We can use the \code{ordL2BB()} function to generate
-critical values and store them in an object \code{mLMo}, say.
-Then, this object can be employed to obtain the test statistic in the usual manner.
-<<ordsave, eval=FALSE>>=
-mLMo <- ordL2BB(yg$agegroup)
-maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = mLMo)
-@ 
-The \code{ordL2BB()} command automatically generates critical values
-for testing 1 to 20 parameters at a time. If only a smaller number
-of parameters (e.g., only up to 6) is to be tested, some computation time can be
-saved by setting the \code{nproc} argument accordingly (e.g., \code{nproc = 1:6}).
-In the same way, \code{nproc} can be employed to simulate higher-dimensional
-fluctuation processes suitable for testing more parameters. One can re-use
-\code{mLMo} in this manner for further tests of the youth gratitude data.
-Critical values must be resimulated for new data, however, because they depend on the
-proportion of individuals observed at each level of the ordinal
-variable (denoted $t_\ell$ for Equation~\eqref{eq:maxlmo}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-In the above sections, we have illustrated the score-based tests'
-computation in \proglang{R}.  We suspect that the ordinal tests will be most
-popular with users, because measurement invariance tests are typically
-carried out across categories (ordered or not), as opposed to
-continuous variables.  Thus, in the sections below, we conduct
-novel simulations to study the ordinal statistics' expected behavior
-in practice. 
-In particular, we wish to study (i) the extent to which the
-ordinal statistics attribute measurement invariance violations to the
-correct parameter(s), and (ii) the extent to which the tests are
-robust to model misspecification.  These issues are especially
-important to examine because SEMs are typically complex, with many
-inter-related parameters that may exhibit measurement invariance.
-Previous applications of score-based tests have typically focused on
-regression-like models with only a small number of parameters that
-may exhibit instability \cite<e.g.,>{zeihor07}.  Thus, the simulations
-here provide general evidence about the extent to which the tests accurately
-capture instabilities in complex models.
-
-\section{Simulation 1}
-In Simulation 1, we examined the extent to which the proposed tests
-are sensitive to the specific model parameter that violates measurement
-invariance.  If, say, a factor loading violates measurement
-invariance, it is plausible that this violation impacts other
-parameter estimates, including factor covariances, intercepts or 
-the unique variance associated with the manifest variable in question.  
-Thus, the goal of the Simulation 1 is to examine the extent to which 
-the proposed tests attribute the measurement invariance violation 
-to the parameters that are truly in violation.
-
-\subsection{Methods}
-
-To examine these issues, we generated data from a
-two-factor model with three indicators each (see Figure~\ref{fig:path}).
-The measurement invariance
-violation occurred in one of four places: the factor loading associated with Scale 1 
- ($\lambda_{11}$), the unique variance  ($\psi_{11}$)
- or intercept associated with Scale 1 ($\mu_{11}$)
-, the factor covariance ($\phi_{12}$).  Note that the
- latter violation is not necessarily a measurement invariance
- \cite<e.g.,>{mer93}, but it is still a parameter instability that can occur in
- this type of model.
- We then tested for measurement invariance (parameter instability) in
- seven subsets of parameters: each of the four
-individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
-all six unique variances and all six intercepts. % $\epsilon_{11}--\epsilon_{66}$.
-
-%\begin{figure}
-%\caption{Path diagram representing the factor analysis model 
-%         used for simulations.}
-%\label{fig:path}
-%\includegraphics{PathDiagram}
-%\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%Plot Code
-
-\begin{figure}
-    \caption{General model used for the simulations.}
-    \label{fig:path}
-\begin{tikzpicture}
-[latent/.style={circle,draw=black!50,fill=white!20,thick,
-                inner sep=0pt, minimum width=20mm, minimum height=10mm},
- manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
-                inner sep=0pt,minimum width=20mm, minimum height=10mm},
- error/.style={circle,draw=black!50,fill=white!20,thick,
-                inner sep=0pt, minimum width=10mm},
- post/.style={->,shorten >=1pt,>=stealth',semithick},
- sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=20mm,semithick},
- errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=10mm,semithick}]
-
-%position objects.
-
-\node[latent] (VERBAL) at (0,4) {VERBAL};
-\node[latent] (MATH)   at (0,0) {MATH};
-
-\node[manifest] (SCALE1) at(5,6) {SCALE1};
-\node[manifest] (SCALE2) at (5,4.5) {SCALE2};
-\node[manifest] (SCALE3) at (5,3) {SCALE3};
-\node[manifest] (SCALE4) at(5,1) {SCALE4};
-\node[manifest] (SCALE5) at (5,-0.5) {SCALE5};
-\node[manifest] (SCALE6) at (5,-2) {SCALE6};
-
-\node[error] (E1) at (7,6) {E1};
-\node[error] (E2) at (7,4.5) {E2};
-\node[error] (E3) at (7,3) {E3};
-\node[error] (E4) at (7,1) {E4};
-\node[error] (E5) at (7,-0.5) {E5};
-\node[error] (E6) at (7,-2) {E6};
-
-%arrows, slings and texts
-\node[latent] (VERBAL) at (0,4) {VERBAL}
-   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{11}=4.92$}} (SCALE1)
-   edge [post] node[above=0.001mm,text width=2cm]{\scriptsize{$\lambda_{21}=2.96$}} (SCALE2)
-   edge [post] node[above=0.1cm,text width=2cm]{\scriptsize{$\lambda_{31}=5.96$}} (SCALE3)
-   edge [sling,bend right=100] node[auto,swap]{1}(VERBAL.-180)
-   edge [sling,bend right=70] node[auto,swap]{$\phi_{12}=-0.48$}(MATH)  ;
-
-\node[latent] (MATH) at (0,0) {MATH}
-   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{42}=3.24$}}(SCALE4)
-   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{52}=4.32$}}(SCALE5)
-   edge [post] node[above=0.25cm,text width=2cm]{\scriptsize{$\lambda_{62}=7.21$}}(SCALE6)
-   edge [sling,bend right=100] node[auto,swap]{1}(MATH.270);
-
-\node[error] (E1) at (7,6) {E1}
-   edge [post] node[auto,swap]{\small{1}}(SCALE1)
-   edge [errorsling, bend right=100]node[right=0.1,text width=2cm]{\scriptsize{$\psi_{11}=26.77$}} (E1.45);
-\node[error] (E2) at (7,4.5) {E2}
-   edge [post] node[auto,swap]{\small{1}} (SCALE2)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{22}=13.01$}}(E2.45);
-\node[error] (E3) at (7,3) {E3}
-   edge [post] node[auto,swap]{\small{1}}(SCALE3)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{33}=30.93$}}(E3.45);
-\node[error] (E4) at (7,1) {E4}
-   edge [post] node[auto,swap]{\small{1}}(SCALE4)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{44}=3.17$}}(E4.45);
-\node[error] (E5) at (7,-0.5) {E5}
-   edge [post] node[auto,swap]{\small{1}}(SCALE5)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{55}=8.82$}} (E5.45);
-\node[error] (E6) at (7,-2) {E6}
-   edge [post] node[auto,swap]{\small{1}}(SCALE6)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{66}=22.5$}}(E6.45);
-
-\end{tikzpicture}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Power and Type I error were examined across three sample sizes
-($n=120,480,960$), three numbers of categories ($m=4,8,12$), and
-17 magnitudes of invariance violations (described in the following sentences). 
-The measurement invariance violations began at level $1+m/2$ of the
-auxiliary variable $V$ and were consistent thereafter: individuals below
-level $1+m/2$ of $V$ deviated from individuals at or above level $1+m/2$ 
-by $d$ times the parameters' asymptotic standard errors
-(scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$. 
-For each combination of sample size ($n$) $\times$ violation
-magnitude ($d$) $\times$ violating parameter $\times$ categories ($m$), 5000 datasets were
-generated and tested.  Statistics from Equations~\eqref{eq:wdm}, \eqref{eq:maxlmo}
-and~\eqref{eq:lmuo} were examined. As mentioned previously,
-\eqref{eq:lmuo} is asymptotically equivalent to the usual likelihood
-ratio test.  Thus, this statistic provides information about the
-relative performance of the ordinal statistics vs.\ the LRT.
-
-In all conditions, we maintained equal sample 
-sizes in each subgroup of the ordinal variable.
-
-\subsection{Results}
-
-Full simulation results are presented in Figures~\ref{fig:sim11res}
-to~\ref{fig:sim14res}.
-Figure~\ref{fig:sim11res} displays power curves as a function of 
-violation magnitude in the factor loading $\lambda_{11}$,
-with the parameters being tested changing across rows,
-the number of levels $m$ of the ordinal variable $V$ across columns,
-and lines reflecting different test statistics.
-Figure~\ref{fig:sim12res}, Figure~\ref{fig:sim13res} and 
-Figure~\ref{fig:sim14res} display 
-similar power curves when the factor covariance 
-$\phi_{12}$, error variance $\epsilon_{11}$ and intercept
-$\mu_{11}$ violate measurement
-invariance, respectively. In these figures, we generally show tests
-associated with parameters that exhibited nonzero power curves.  For
-example, in Figure~\ref{fig:sim11res}, the middle row shows that power
-for tests of $\psi_{11}$ stays near zero for all values of $m$ and $d$.
-Similar rows have been omitted from this figure and other figures.
-
-Within each panel of Figure~\ref{fig:sim11res} to 
-Figure~\ref{fig:sim14res}, the three lines reflect the three test
-statistics.  It is seen that the two ordinal statistics exhibit 
-similar results, with $\max\mathit{LM}_{uo}$ demonstrating lower power
-across all situations.  This demonstrates the sensitivity of the
-ordinal statistics to invariance violations that are monotonic with
-$V$.  In situations where only one parameter is tested,
-$\mathit{WDM}_o$  and $\max \mathit{LM}_o$ exhibit equivalent power
-curves.  This is because, when only one parameter is tested, the
-statistics are equivalent.
-
-From these figures, one generally observes that the tests isolate 
-the parameter violating measurement invariance.  Additionally, the tests 
-have somewhat higher power to detect measurement invariance violations 
-in the factor loading, factor covariance and intercept parameters, 
-as opposed to the error variance parameter.  
-Finally, simultaneous tests of all factor loadings, all intercepts or 
-all error parameters result in decreased power, as compared to the situation
-where one tests only the violating parameter.  This occurs because, in
-testing a subset of parameters (only one of which violates measurement
-invariance), we are effectively dampening the signal of a measurement
-invariance violation. This ``dampening'' effect is more apparent for
-the $\max \mathit{LM}_o$ statistic, because it involves a sum across
-all tested parameters (see Equation~\eqref{eq:maxlmo}).
-Conversely, $\mathit{WDM}_o$ takes the maximum over parameters
-(Equation~\eqref{eq:wdm}), so that invariant parameters have no
-impact on this statistic.
-
-In summary, we found that the proposed tests can attribute
-measurement invariance violations to the correct parameter.  This
-provides evidence that, in practice, one can have confidence in the
-tests' abilities to locate the measurement invariance violation.  Of
-course, this statement is qualified by the fact that, in this
-simulation, the model was correctly specified.  In the following
-simulation, we examine the tests' performance in the likely situation
-of model misspecification.
-
-<<sim1,echo=FALSE>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1163)
-
-if(file.exists("sim1.rda")) {
-  load("sim1.rda")
-} else {
-  sim1 <- simulation(sim = "sim1", nobs = 480, parms = c("loading", "error", "var", "intercept"))
-  sim1$nlevels <- factor(sim1$nlevels)
-  levels(sim1$nlevels) <- paste("m=", levels(sim1$nlevels), sep="")
-
-  sim1$nobs <- factor(sim1$nobs)
-  levels(sim1$nobs) <- paste("n=", levels(sim1$nobs), sep="")
-  save(sim1, file="sim1.rda")
-}
-sim1$test <- factor(as.character(sim1$test), levels = c("ordmax","ordwmax","catdiff"),
-                    labels = c("maxLM_o", "WDM_o", "LM_uo"))
-parlabs <- c(expression(lambda[11]), expression(phi[12]), 
-             expression(psi[11]), expression(mu[11]), expression(list(lambda[11], ldots, lambda[62])), 
-             expression(list(psi[11], ldots, psi[66])), 
-             expression(list(mu[11], ldots, mu[66])))
-levels(sim1$pars) <- c("Loading1", "Covariance", "Error1", "Intercept1", 
-                       "All Loadings", "All Errors", "All Intercepts")  
-@
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and
-$\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The parameter violating measurement 
-invariance is $\lambda_{11}$.  Panel labels denote the parameter(s)
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim11res}
-<<sim11res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "loading" & 
-		 pars %in% c("Loading1", "Error1", "All Loadings") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", lambda[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-           if(which.given == 2){
-             strip.default(which.given, factor.levels = parlabs[c(1, 3, 5)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and
-$\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$,
-  and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
-parameter violating measurement invariance is $\phi_{12}$. Panel
-labels denote the parameter(s)  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim12res}
-<<sim12res, fig=TRUE, height=4.5, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "var" & 
-		 pars %in% c("Covariance") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", phi[12], ")")),
-       ylab="Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(2)],...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
-parameter violating measurement invariance is $\psi_{11}$. Panel labels
-denote the parameter(s) 
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim13res}
-<<sim13res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "error" & 
-		 pars %in% c("Error1", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", psi[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(3,6)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
-parameter violating measurement invariance is $\mu_{11}$. Panel labels
-denote the parameter(s) 
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim14res}
-<<sim14res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "intercept" & 
-		 pars %in% c("Intercept1", "All Intercepts") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", mu[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(4,7)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\section{Simulation 2}
-In Simulation 2, we examine the extent to which the results of
-Simulation 1 are robust to model misspecification.  Specifically, we
-generate data from the factor analysis model used in the previous
-section, except that the model contains an extra loading from the
-second factor to Scale 1.
-The estimated model matches that displayed in Figure~\ref{fig:path},
-however, resulting in model misspecification.  The goal of this
-simulation is to examine the proposed statistics' 
-power to detect measurement invariance violations (and to attribute
-the violation to the correct parameter) under this misspecification.
-
-\subsection{Method}
-A measurement invariance violation could occur in
-each of the four parameters from Simulation 1 (factor loading, factor
-covariance, unique variance and intercept), and a violation could also occur in
-the extra, unmodeled loading. In each condition, a single parameter
-exhibited the violation. 
-Sample size and magnitude of measurement invariance violation were manipulated 
-in the same way as they were in Simulation 1.  The tested parameters
-were also the same as Simulation 1.
-
-\subsection{Results}
-Results of
-primary interest are conditions where the unmodeled loading violates
-measurement invariance.  A subset of results is displayed in Figure~\ref{fig:sim21res}.  One can generally observe
-that tests of the first loading and unique variance exhibited high
-``power,'' which is actually a high Type I error rate here.  This Type
-I error also demonstrated when testing all loadings and all unique
-variances, as shown in the appendix.  Tests
-associated with the factor 
-covariance and intercept did not demonstrate this error, however.  In
-terms of specific 
-statistic performance, $\max \mathit{LM}_o$ and $\mathit{WDM}_o$ 
-demonstrated higher Type I error than $\mathit{LM}_{uo}$ in each panel,
-especially with increasing levels.  This is likely because the unmodeled
-loading's non-invariance was monotonic with $V$; if it were not monotonic, we
-would expect $\mathit{LM}_{uo}$ to have higher Type I error.
-
-When the parameter violating measurement invariance was modeled,
-results were generally the same as Simulation 1.  
-When the modeled factor loading, $\lambda_{11}$, violated measurement
-invariance, the statistics were generally able to pick up the violation despite
-the misspecification. Similar results were observed
-when the unique variance, intercept and factor covariance parameters violated measurement
-invariance; these results are all shown in the Appendix.
-In particular, the power of ordered statistics $\max \mathit{LM}_o$ test and
-$\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in
-each of these panels.
-
-In summary, the proposed test statistics appear robust to unmodeled
-loading parameters, so long 
-as the unmodeled loading does not violate measurement invariance.  If
-the unmodeled loading  
-does violate measurement invariance, the tests can still detect
-measurement invariance violations.  The violations are assigned to 
-modeled parameters that do not violate measurement invariance, however.
-The impacted parameters include the error 
-variance and other loadings associated with the manifest variable that
-has an unmodeled loading.
-Thus, as for other tests of measurement invariance, it is important to
-study the extent to 
-which the hypothesized model includes all parameters of importance.
-None of these tests (score-based or otherwise) inform the researcher of
-model misspecification.
 
@@
 
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is the unmodeled
-loading. Panel labels denote the parameter(s) 
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim21res}
-<<sim21res, fig=TRUE, height=9, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra" & 
-		 pars %in% c("Loading1", "Covariance", "Error1", "Intercept1") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude (Unmodeled Loading)",
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 2, 3, 4)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\section{General Discussion}
-
-In this paper, we first described a novel family of test statistics
-for measurement invariance and illustrated their use via the
-\proglang{R} packages \pkg{lavaan} and \pkg{strucchange}.  Next, we examined these
-statistics' abilities to identify the parameter violating measurement
-invariance under well-specified and misspecified models.  We found
-that the
-proposed statistics could generally isolate the model parameter
-violating measurement invariance, so long as the violating parameter
-is included in the model.
-
-In the remainder of the paper, we first compare
-these tests to traditional tests of measurement invariance.  We then 
-discuss test extension to other fit functions and to other specialized
-models.
-
-\subsection{Applications}
-
-Many of the applications in this volume, along with many measurement
-invariance applications in
-general, focus on testing across unordered categories
-such as nations or gender.  As discussed earlier in this paper, the score-based
-tests for unordered categories are essentially equivalent to the usual
-likelihood ratio test.  Given a measurement invariance violation across these
-unordered categories, however, researchers typically wish to know why the
-violation occurred.  At this point, researchers may examine education level,
-socioeconomic status, income levels, and so on across the unordered categories. 
-These variables are often ordinal or continuous in nature, so that the family of
-tests described in this paper are applicable.  This is a first step towards
-describing why measurement invariance violations occur, as opposed to simply
-detecting measurement invariance violations. 
-%% TODO!!
-%\readme{In the abstracts for the
-%special issue (which the special editors sent us), one paper was proposed along
-%these lines.  I will get a draft of that paper and then we can cite it here.}
-The tests described here are convenient for this purpose, as they do
-not require a new model to 
-be estimated for each ordinal variable.  Instead, each ordinal variable defines
-an ordering of observations, which in turn yields a test statistic that is
-specific to that ordinal variable.
-
-%% I'm not sure we want to make a big deal of this.  If there are only a small number of ties, we are probably better off with the continuous tests:
-%These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context.
-It is very common to observe several individuals with the same age, IQ, income rank, etc.
-Therefore, the ordinal tests which dealing with multiple ties in auxiliary variable is more suitable than the continuous test.
-
-%Moreover, even though researchers might be originally interested in comparing categorical differences.
-For example, researchers might want to test whether there is gender difference or cultural difference for the scale.
-Using traditional likelihood test or wald test, there might existing some differences regarding to the categorical variable.
-Then the next natural question would be why there is gender or culture differences?
-.
-Thus these test statistics offer an convenient assessing tool to further exploring research hypothesis without making complicated assumptions or fitting multiple different models.
-
-\subsection{Extension}
-
-In this paper, we focused on testing for measurement invariance in factor
-analysis models that assume multivariate normality and that are estimated via
-maximum likelihood (ML).  The family of tests described here generally apply to estimation methods
-that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are
-potentially applicable to alternative SEM discrepancy functions such
-as generalized 
-least squares \cite<e.g.,>{broarm95}.  Score calculation for these alternative
-discrepancy functions has not been implemented (to our knowledge), though the
-calculation could be implemented.  Test statistic calculation and
-inference would then proceed in exactly the same manner as the calculation and
-inference illustrated in this paper.
-
-In addition to alternative fit functions, the tests can be extended to other
-models estimated via ML.  Of primary interest to the topic of measurement
-invariance, the tests can be extended to item response models to examine
-differential item functioning.  In particular, \citeA{strkop13} studied
-application of these tests to the Rasch model, using them as the basis of a
-recursive partioning procedure that segments subgroups of individuals who
-exhibit DIF.  Further study and extension of these tests for IRT seem warranted.
-
-% Removed for now:
-%% \subsection{Conclusion}
-%% \readme{Not sure whether we need this paragraph, but I still need to think about it.}
-%% In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable.
-This
-%% feature is robust as long as the violation parameter is included in the testing model.
-Furthermore, as illustrated in tutorial part, the use of these tests is convenient without changing the focal psychometric model.
-In all, the tests have advantageous properties that should be implemented in practice.
-
-\section*{Computational Details}
-
-All results were obtained using the \proglang{R}~system for statistical computing \cite{R11},
-version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on package
-\pkg{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} for fitting 
-of the factor analysis models and
-\pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
-for evaluating the parameter instability tests. \proglang{R}~and both
-packages are freely available under the General Public License~2 from the
-Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/}.
-\proglang{R}~code for replication of our results is
-available at \url{http://semtools.R-Forge.R-project.org/}.
-
-\bibliography{refs}
-
 \section{Appendix}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 

### 2014-02-18 frontiers/IMPS.Rnw
add IMPS.Rnw file.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@317 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2014-01-30 frontiers/frontiers.Rnw
@@
 
-\include{Sweave}
-\setkeys{Gin}{width=\textwidth}
-<<preliminaries, echo=FALSE>>=
-## packages
-library("lavaan")
-library("strucchange") ## at least 1.5-0
-library("mvtnorm")
-library("lattice")
-
-## auxiliary code
-source("../www/sim-frontiers.R")
-source("../www/mz-frontiers.R")
-
-## some options
-options(prompt = "R> ", continue = "+  ", digits = 3,
-  useFancyQuotes = FALSE, show.signif.stars = FALSE)
-@
-
-Many of the papers in this special issue focus on the topic of approximate
-measurement invariance: we know that strict hypotheses of measurement invariance
-do not hold exactly across different groups, and this should be reflected in
-corresponding tests of measurement invariance.  Under a Bayesian
-approach, we may utilize the idea of approximate invariance
-\cite<e.g.,>{muthen13}, whereby across-group equality constraints on
-parameters are replaced with informative prior distributions.  In this paper, we
-describe an alternative approach: the development of test statistics that 
-are especially sensitive to violations that are monotonic w.r.t. the
-variable of interest (and 
-less sensitive to violations that are non monotonic w.r.t. the
-variable of interest).  Because monotonic violations are more likely to be
-interpretable and interesting to the researcher, we can gain more
-power to detect these violations by de-emphasizing other types of violations.
-The resulting test statistics are specifically applicable to
-situations where one  
-wishes to test for measurement invariance with respect to an ordinal variable, 
-and they are special cases of a family of tests that may be used to study 
-measurement invariance w.r.t.\ continuous, categorical, and ordinal
-variables.  Other methods for studying measurement
-invariance w.r.t.\ continuous variables include moderated factor
-models \cite{bauhus09,moldol10,pur02} and factor mixture models
-\cite{dolmaa98,lubmut05}.  These methods require a model of greater
-complexity to be estimated, while the tests described here work solely
-on the output of a traditional factor model.  See \citeA{MerZei13} for
-further comparison of these methods.
-
-The family of tests described in this paper have recently been applied
-to the study of measurement invariance \cite{MerZei13,MerFanZei}, though their
-practical application has been limited to a small set of simulations and
-datasets.  In this paper, we provide a detailed illustration of the tests' use
-and performance under
-scenarios likely to be encountered in practice.  While the previous
-papers have described and studied the tests under ideal conditions, we
-focus here on two topics of interest to the applied researcher:
-software considerations for carrying out the tests, and test
-performance under model misspecification.
-%In order to fix the psychometric scale, we need to specify the violating parameter.  
-%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
-%and may interfere with parameter invariance specificity. A family of tests derived from 
-%stochastic process has been recently proposed to identify specific parameters impacted 
-%by measurement invariance violation. However, these tests have only been used to a limited 
-%type of parameter violation. It is this paper's intent to provide a detailed examination 
-%of the tests' performance under practical research scenarios. 
-In the following sections, we first briefly review the  theoretical framework of
-the proposed tests and provide a short tutorial illustrating the use of the
-tests in \proglang{R} \cite{R11}. Next, we study the tests' performance in simulations that mimic
-practical research scenarios.  Finally, we provide some further discussion on
-the tests' use in practice.
-
-\section{Background}
-
-This section contains background and discussion of the  proposed statistics as
-applied to structural equation models (SEMs); for a more detailed account, see \citeA{MerZei13} and
-\citeA{MerFanZei}.  For details on the statistics' application to general
-statistical models, see \citeA{zeihor07}.
-
-As currently implemented for SEM, the statistical tests described in
-this paper can be applied to models that are estimated via a
-multivariate normal or Wishart likelihood (or discrepancy) function,
-with extension to other discrepancy functions being conceptually straightforward. 
-The tests are carried out following model estimation, making use of output
-associated with the fitted model.  In general, we fit a model that
-restricts parameters to be equal across observations, then carry out a
-posthoc test to examine whether specific parameters vary
-across observations.  This procedure is similar in spirit to the
-calculation of modification indices \cite{bentler90} and to Lagrange multiplier
-tests \cite{sat89}, and, in fact, those statistics can be viewed as  
-special cases of the family described here.
-
-Following model estimation, the tests primarily work on the \emph{scores}
-of the fitted model; these are defined as
-%% TODO indicate that this is evaluated at theta hat
-\begin{equation}
-\label{eq:score}
-s(\theta;x_{i}) = \left (
-  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{1}}, \dots,
-  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right
-  )^\top,\ i=1,\ldots,n,
-\end{equation} 
-where $\ell(\theta; x_i)$ is the likelihood associated with individual
-$i$ and $\theta$ is a $k$-dimensional parameter vector. The corresponding
-maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
-$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$.
-
-To verbally describe the above equation, each
-individual has $k$ scores describing the extent to which the fitted
-model describes that particular individual.  We can roughly interpret
-scores near zero as providing a ``good'' description of an individual,
-with scores far from zero providing a ``bad'' description of an
-individual.  This is only a rough interpretation as, even when measurement
-invariance holds, some individuals' scores will be further from zero
-than others.  However, under measurement noninvariance, the scores
-will differ for different subgroups of individuals (say, scores in
-subgroup A tend to be negative and scores in subgroup B tend to be
-positive).  Each of the $k$ scores represents one model parameter,
-which, as further described below, allows us to test subsets of model
-parameters for invariance.
-
-To use the scores for testing, we order individuals according to
-an auxiliary variable $V$ (the variable against which we are testing 
-measurement invariance) and look for ``trends'' in the
-scores.  For example, imagine that we are testing
-for measurement 
-invariance w.r.t.\ age.  If there exists measurement invariance
-w.r.t. age, then some parameter estimates may be too large for 
-young individuals and too small for old individuals (say).  This
-result would be reflected in the scores, where young individuals'
-scores may be greater than zero and old individuals' scores 
-less than zero (though the sign of the scores will depend on whether a
-function is being minimized or maximized).  Conversely, if measurement
-invariance holds, then all 
-individuals' scores will fluctuate randomly around zero.
-
-To formalize these ideas, we define a
-suitably scaled cumulative sum of the ordered scores.  This may be written as
-\begin{equation} 
-    \label{eq:cumscore}
-  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
-    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
-  \qquad (0 \le t \le 1)
-\end{equation}
-where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
-nt \rfloor$ is 
-the integer part of $nt$ (i.e., a floor 
-operator), and $x_{(i)}$ reflects the individual with the $i$-th
-smallest value of the auxiliary variable $V$.  While the above
-equation is written in general form, we can
-restrict the value of $t$ in finite samples to the set $\{0,
-1/n, 2/n, 3/n, \ldots, n/n\}$.
-We focus on how the
-cumulative sum fluctuates as more individuals' scores are added to it, e.g.,
-starting with the youngest and ending with the oldest individual
-if age is the auxiliary variable of interest.
-The summation is premultiplied by an estimate of the inverse square
-root of the information matrix, which serves to decorrelate the
-fluctuation processes associated with individual model parameters
-while preserving the behavior of individual parameters' fluctuations.
-
-Under the
-hypothesis of measurement invariance, a central limit theorem can be
-used to show that the fluctuation of the above cumulative sum follows a
-Brownian bridge \cite{hjokon02}.  % TODO add ref to Brownian bridge  
-This result allows us to calculate $p$-values and
-critical values for test statistics under the hypothesis of
-measurement invariance.  We can obtain test statistics associated with
-all model parameters and with subsets of model parameters.
-
-%% TODO focus on ordinal
-Multiple test statistics are available, depending on how one
-summarizes the behavior of the cumulative sum of scores.  For example, 
-one could take the absolute maximum that the cumulative sum attains for any
-parameter of interest, resulting in a {\em double max} statistic 
-(the maximum is taken across parameters and individuals).
-Alternatively, one could sum the (squared) cumulative sum across parameters
-of interest and take the maximum or the average across individuals, resulting in a 
-{\em maximum Lagrange multiplier} statistic and \emph{Cram\'{e}r-von Mises} statistic, respectively 
-\cite<see>[for further discussion]{MerZei13}.  These statistics are given by
-\begin{eqnarray}
-    \label{eq:dmax}
-    \mathit{DM} & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} | \\
-        \label{eq:cvm}
-    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
-    \label{eq:maxlm}
-    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
-      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
-\end{eqnarray}
-Critical values associated with $\mathit{DM}$ can be obtained analytically, 
-while critical values associated with the other statistics can be obtained 
-from direct simulation \cite{Zei06} or from more refined techniques \cite{han97}.  
-This issue should not be important to the user, as critical values are obtained 
-directly from the \proglang{R} implementation described later.
-
-Importantly, the above statistics were derived for situations where
-individuals are uniquely ordered according to the auxiliary variable.
-This is not always the case for measurement invariance applications,
-where the auxiliary variable is often ordinal.  To remedy this
-situation, \citeA{MerFanZei} extended the framework to situations
-where one has an ordinal auxiliary variable of interest.  Essentially,
-one allows all individuals with the same value of the auxiliary
-variable to enter into the cumulative sum at the same time.  Analogous test 
-statistics are then computed, with
-modified critical values being adopted to reflect the change in the
-statistics' computation.
-
-For an ordinal auxiliary variable with $m$ levels, these modifications are 
-based on $t_{\ell}$ $(\ell=1,\ldots,m-1)$, which are the empirical, cumulative 
-proportions of individuals observed at the first $m-1$ levels.
-The modified statistics are then given by
-\begin{eqnarray}
-    \label{eq:wdm}
-      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
-                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
-    \label{eq:maxlmo}
-      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
-                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
-\end{eqnarray}
-where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$ $(\ell=1,\ldots,m-1)$.
-Critical values associated with the $\mathit{WDM}_o$ statistic can be obtained 
-directly from a multivariate normal distribution \cite<see>{hotzei08}, while 
-critical values associated with $\max \mathit{LM}_o$ can be obtained via simulation.  
-This simulation is somewhat computationally intensive and, in practice, takes about 
-10 minutes on the authors' computers when 50,000 replications are sampled from the 
-approximate asymptotic distribution.  However, the wait is often worth it, 
-as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic 
-to have more power than the $\mathit{WDM}_o$ statistic and the traditional
-likelihood ratio test statistic when the measurement invariance
-violation is monotonic with the ordinal variable.
-
-Finally, if the auxiliary variable $V$ is only nominal/categorical, the cumulative
-sums of scores can be used to obtain a Lagrange multiplier statistic.
-This test statistic can be formally written as
-\begin{equation}
-    \label{eq:lmuo}
-     \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
-       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
-\end{equation}
-where ${\bm B}(\hat {\bm \theta})_{i_{0}j} = 0$ for all $j$.
-This statistic is asymptotically equivalent to the usual, likelihood
-ratio test statistic, and it is advantageous over the likelihood ratio
-test because it requires estimation of 
-only one model (the restricted model).  We make use of this advantage in the
-simulations, described later.
-
-\section{Tutorial}
-
-In this section, we demonstrate how the above tests can be carried out in \proglang{R}, 
-using the package \pkg{lavaan} \cite{lavaan11} for model estimation and \pkg{strucchange} 
-\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
-the applicability of adult gratitude scales to youth, available in the~\proglang{R} 
-package \pkg{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
-gratitude scales from $n=1401$ youth aged 10--19 years. 
-The original authors 
-were specifically interested in whether the scales were measurement invariant 
-across age.  Because the sample size at each age was unbalanced, the authors 
-created age groups of approximately equal sample size.  In the examples below, 
-we test for measurement invariance across these age groups.
-
-We focus on measurement invariance of the factor loadings associated with 
-the GQ-6 scale, using a one-factor model.  This scale consists of five
-items scored from 1--9.  While the ``age group'' variable against 
-which we are testing measurement invariance is best considered ordinal, 
-for demonstration we consider its treatment as categorical, continuous, and ordinal.  
-Each of these treatments is described below in a separate section.
-
-%% THE FOLLOWING DETAILS ARE REMOVED FOR NOW
-%% In general, a set of scales is defined to be measurement invariant with respect
-%% to an auxiliary variable $V$ if:
-
-%% \begin{equation}
-%% \label{equ:invariance}
-%% f(x_{i}|t_{i},v_{i},...)=f(x_{i}|t_{i},...),
-%% \end{equation}
-
-%% where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
-%% for variable $i$ that the scales purport to measure, and $f$ is the model's 
-%% distributional form \cite{mel89}. If the cut point of $v$ is known in advance, i.e. 
-%% gender, nested multiple group models \cite{bollen98} coupled with likelihood ratio test 
-%% are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
-%% of measurement invariance is
-
-%% \begin{equation}
-%% \label{equ:null}
-%% H_{0}: \theta_{A}=\theta_{B},
-%% \end{equation}
-
-%% The alternative hypothesis is
-
-%% \begin{equation}
-%% \label{equ:alternative}
-%% H_{1}:\theta_{i}=
-%% \begin{cases}
-%% \theta^{(A)} & v_{i}\leq v\\
-%% \theta^{(B)} & v_{i}\geq v
-%% \end{cases}
-%% \end{equation}
-
-%% The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ 
-%% (group A and group B) can be obtained by Maximum Likelihood(ML) 
-%% from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,
-%% respectively. The LR test statistic for the given threshold $v$ is then
-
-%% \begin{equation}
-%% \label{equ:LRT}
-%% LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
-%% {(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
-%% }],
-%% \end{equation}
-
-%% ,which has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
-%% to the number difference of parameters in $\theta$.
-
-\subsection{Categorical Treatment}
-
-Measurement invariance is most often tested using multiple groups models \cite<see, e.g., >{vande12}.  
-This amounts to assuming that our auxiliary variable is categorical (i.e., unordered), 
-which is not true for the age groups in the data.  However, we demonstrate the procedure for completeness.
-
-To conduct the analysis, we first load the data and keep only complete cases for simplicity 
-(though the tests can be applied to incomplete data).
-%
-<<data>>=
-data("YouthGratitude", package = "psychotools")
-compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
-yg <- YouthGratitude[compcases, ]
-@
-%
-Next, we fit two models in \pkg{lavaan}: a one-factor model 
-where loadings are restricted to be equal across age groups, 
-and a one-factor model where loadings are free across age groups.
-This allows us to test a hypothesis of weak measurement invariance
-that was of interest to the original researchers (though, for ordinal
-variables, all types of
-measurement invariance can be examined via the tests described
-previously).
-By default, the code below sets the scale by fixing the first loading
-to 1.
-%
-<<models>>=
-restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-  data = yg, group = "agegroup", meanstructure = TRUE,
-  group.equal = "loadings")
-
-full <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-  data = yg, group = "agegroup", meanstructure = TRUE)
-@
-%
-Finally, we can get the results of a likelihood ratio test via the \code{anova()}
-function, which implies that the GQ-6 violates measurement invariance.
-<<LRT>>=
-anova(full, restr)
-@
-To obtain the asymptotically equivalent $\mathit{LM}_\mathit{uo}$ (Equation~\eqref{eq:lmuo}), we
-can use the \code{sctest()} function from \pkg{strucchange}:
-<<LM_UO>>=
-sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "LMuo")
-@
-This command specifies that we assess the parameter 1--4 of
-model \code{restr} after ordering the observations according to
-\code{agegroup}.  Additionally, 
-the observed information matrix is used as the variance-covariance matrix. Note
-that the model parameters 1--4 are the factor loadings supplied by
-\pkg{lavaan}, which can be seen by inspecting \code{coef(restr)}. This
-also leads to somewhat 
-smaller test statistics that are very close to being significant at the 5\% level.
-
-Because our sample size is large, the likelihood ratio test is known to be sensitive to 
-small measurement invariance violations \cite{Ben80}.  That is, the
-LRT and  $\mathit{LM}_\mathit{uo}$ test from~\eqref{eq:lmuo}
-are sensitive to small measurement invariance violations that are not
-likely to be  
-of interest to researchers.  For example, imagine that the 15-year-olds' 
-parameters are slightly different than the other age groups.  
-The 15-year-olds are in the middle of the age groups, and there 
-is not likely to be any theoretical justification for 15-year-olds 
-differing from every other age group.  One solution to this problem would be the
-Bayesian, approximate invariance methods described in the
-introduction \cite{muthen13}.
-Alternatively, we can use the ``ordinal'' score-based
-statistics (from~\eqref{eq:wdm} and~\eqref{eq:maxlmo}) to obtain 
-tests that are sensitive to the ordering of age.
-
-\subsection{Continuous Treatment}
-
-If we are interested in measurement invariance violations that are 
-monotonic with the age groups, it is perhaps simplest to treat the 
-age groups as continuous.  In doing so, we can use the statistics 
-from Equations~\eqref{eq:dmax}, \eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
-That is, we can fit a model whose parameters 
-are restricted to be equal across all individuals and then examine 
-how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
-age (where age ties are broken arbitrarily, using the original order
-of the observations within each age group).
-This is demonstrated below, with similar code being useful when one is 
-testing for measurement invariance w.r.t.\ truly continuous variables.
-
-%% The multiple group test performs typically well when the $V$ 
-%% is a relatively-small number of categories and the sample size 
-%% is not too large. However, when $V$ (in this case, age) continuous variable, however, 
-%% multiple-group analysis usually cannot be performed because there are 
-%% no existing groups. Instead, we can follow the procedure described in
-%% the Background section.
-
-Again, we employ the \code{sctest()} function to assess parameters~1--4
-from the restricted model \code{restr} after ordering w.r.t.\ \code{agegroup}:
-%
-<<sctest-cont>>=
-dm <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "DM")
-cvm <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "CvM")
-maxlm <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "maxLM")
-c(dm$p.value, cvm$p.value, maxlm$p.value)
-@
-%
-We see that two of the three $p$-values output at the end of the code are
-larger than that associated with the LRT (with the CvM statistic being non-significant).
-
-The tests carried out here assume a unique ordering of individuals by age, but 
-this is obviously not the case.  To compute the statistics and $p$-values, 
-the \pkg{strucchange} package implicitly employed the (arbitrary) ordering of 
-individuals who are tied on age.  If we were to change this ordering, 
-the resulting statistics and $p$-values would also change, potentially
-switching significant results to being non-significant and vice versa.
-Clearly, this is problematic.  
-To accurately account for the multiple observations  
-at the same age level, we must use the ordinal tests from 
-Equations~\eqref{eq:wdm} and~\eqref{eq:maxlmo}.  These are described next.
-
-\subsection{Ordinal Treatment}
-
-%% As a matter of fact, there is only a partial ordering of individuals 
-%% with respect to $V$,i.e. observations with the same level 
-%% of $V$ have no unique ordering. The ordinal statistics 
-%% proposed here are similar to those described in Equation 
-%% and above, except that we focus on "bins" of individuals 
-%% at each level of the ordinal variable.  
-The main difference between the ordinal test statistics and their
-continuous counterparts is that the ordinal statistics are unchanged
-when re-ordering individuals within the same age group.  To compute the test statistics, we allow
-the scores of all tied individuals to enter the cumulative sum
-(Equation~\eqref{eq:cumscore}) simultaneously.  This results in
-modified critical values and test statistics that are sensitive to
-measurement invariance violations that are monotonic w.r.t.\ age group.
-%% Therefore, this ordinal test should 
-%% reflect more real status of the data. In terms of code, we still need the gefp 
-%% function to get the B estimate. After that, we can use different 
-%% functions to aggregate and scale B estimation. 
-%% The function are called ordwmax and ol2bb1.fun, respectively.
-
-To carry out the tests, we can rely on the same function that we used
-for the continuous test statistics.  As mentioned previously,
-calculation of the $\max\mathit{LM}_o$ statistic (Equation~\eqref{eq:maxlmo})
-can be lengthy from the need to simulate critical values (though see
-the end of this section, which provides a partial speed-up).
-%% This is more complicated than necessary, but it is written like
-%% this so that we can load the ordinal critical values behind the
-%% scenes without confusing the reader.
-<<sctest-ord1, echo=FALSE, results=hide>>=
-set.seed(1090)
-wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "WDMo")
-
-if(file.exists("maxLMo.rda")){
-  load("maxLMo.rda")
-} else {
-  RNGkind(kind = "default", normal.kind = "default")
-  set.seed(1090)
-  maxLMo <- ordL2BB(yg$agegroup, nproc = 4)
-  save(maxLMo, file = "maxLMo.rda")
-}
-maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = maxLMo)
-@ 
-<<sctest-ord2, eval=FALSE,strip.white=all>>=
-wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "WDMo")
-maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "maxLMo")
-@ 
-<<sctest-ord3,strip.white=all>>=
-c(wdmo$p.value, maxlmo$p.value)
-@ 
-In computing the ordinal test statistics, we
-obtain~$p=\Sexpr{round(wdmo$p.value, 3)}$
-and~$p=\Sexpr{round(maxlmo$p.value, 3)}$, respectively.\footnote{To replicate
-both $p$-values exactly, \proglang{R}'s random seed needs to be
-set by \code{set.seed(1090)} prior to each \code{sctest()} call.}
-Both $p$-values are clearly larger than that of the 
-likelihood ratio test and neither is significant at $\alpha=0.05$. 
-This provides evidence that there is no measurement
-invariance violation that is monotonic with age group.  Instead, given
-the large sample size, the likelihood ratio test may be overly
-sensitive to anomalous, non-monotonic violations at one (or a few) age groups.
-
-In addition to test statistics, ``instability plots'' can be generated
-by setting 
-\code{plot = TRUE} in the \code{sctest()} calls above. The resulting plots representing
-the ordinal statistics' fluctuations across levels of age group are displayed in
-Figure~\ref{fig:ordres}.
-These plots display the loadings' fluctuation across age groups, with the x-axis
-reflecting age group and the y-axis reflecting test statistic values (larger
-values reflect more instability).  The dashed horizontal lines reflect critical values, so
-that the hypothesis of measurement invariance is rejected if the sequence of
-test statistics crosses the critical value.  While the measurement invariance tests are
-non-significant, the plots imply some instability in the older age groups (15 and~16).
-
-\begin{figure}
-\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic (left panel) and the $\max\mathit{LM}_o$ statistic (right panel).}
-\label{fig:ordres}
-<<sctest-ord-graphics, fig=TRUE, height=4.5, width=9, echo=FALSE, results=hide>>=
-par(mfrow = c(1, 2))
-set.seed(1090)
-sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "WDMo", plot = TRUE, ylim = c(0, 3.2),
-  xlab = "Age group", ylab = "Weighted test statistics",
-  main = expression(WDM[o]), boundary = list(col = 1, lty = 2))
-sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = maxLMo, plot = TRUE, ylim = c(0, 14),
-  xlab = "Age group", ylab = "LM statistics",
-  main = expression(maxLM[o]), boundary = list(col = 1, lty = 2))
-@
-\end{figure}
-
-Finally, if the user anticipates multiple calculations of the
-$\max\mathit{LM}_o$ statistic for a specific dataset, it is possible
-to save time by simulating critical values once and re-using them for
-multiple tests.  We can use the \code{ordL2BB()} function to generate
-critical values and store them in an object \code{mLMo}, say.
-Then, this object can be employed to obtain the test statistic in the usual manner.
-<<ordsave, eval=FALSE>>=
-mLMo <- ordL2BB(yg$agegroup)
-maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = mLMo)
-@ 
-The \code{ordL2BB()} command automatically generates critical values
-for testing 1 to 20 parameters at a time. If only a smaller number
-of parameters (e.g., only up to 6) is to be tested, some computation time can be
-saved by setting the \code{nproc} argument accordingly (e.g., \code{nproc = 1:6}).
-In the same way, \code{nproc} can be employed to simulate higher-dimensional
-fluctuation processes suitable for testing more parameters. One can re-use
-\code{mLMo} in this manner for further tests of the youth gratitude data.
-Critical values must be resimulated for new data, however, because they depend on the
-proportion of individuals observed at each level of the ordinal
-variable (denoted $t_\ell$ for Equation~\eqref{eq:maxlmo}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-In the above sections, we have illustrated the score-based tests'
-computation in \proglang{R}.  We suspect that the ordinal tests will be most
-popular with users, because measurement invariance tests are typically
-carried out across categories (ordered or not), as opposed to
-continuous variables.  Thus, in the sections below, we conduct
-novel simulations to study the ordinal statistics' expected behavior
-in practice. 
-In particular, we wish to study (i) the extent to which the
-ordinal statistics attribute measurement invariance violations to the
-correct parameter(s), and (ii) the extent to which the tests are
-robust to model misspecification.  These issues are especially
-important to examine because SEMs are typically complex, with many
-inter-related parameters that may exhibit measurement invariance.
-Previous applications of score-based tests have typically focused on
-regression-like models with only a small number of parameters that
-may exhibit instability \cite<e.g.,>{zeihor07}.  Thus, the simulations
-here provide general evidence about the extent to which the tests accurately
-capture instabilities in complex models.
-
-\section{Simulation 1}
-In Simulation 1, we examined the extent to which the proposed tests
-are sensitive to the specific model parameter that violates measurement
-invariance.  If, say, a factor loading violates measurement
-invariance, it is plausible that this violation impacts other
-parameter estimates, including factor covariances, intercepts or 
-the unique variance associated with the manifest variable in question.  
-Thus, the goal of the Simulation 1 is to examine the extent to which 
-the proposed tests attribute the measurement invariance violation 
-to the parameters that are truly in violation.
-
-\subsection{Methods}
-
-To examine these issues, we generated data from a
-two-factor model with three indicators each (see Figure~\ref{fig:path}).
-The measurement invariance
-violation occurred in one of four places: the factor loading associated with Scale 1 
- ($\lambda_{11}$), the unique variance  ($\psi_{11}$)
- or intercept associated with Scale 1 ($\mu_{11}$)
-, the factor covariance ($\phi_{12}$).  Note that the
- latter violation is not necessarily a measurement invariance
- \cite<e.g.,>{mer93}, but it is still a parameter instability that can occur in
- this type of model.
- We then tested for measurement invariance (parameter instability) in
- seven subsets of parameters: each of the four
-individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
-all six unique variances and all six intercepts. % $\epsilon_{11}--\epsilon_{66}$.
-
-%\begin{figure}
-%\caption{Path diagram representing the factor analysis model 
-%         used for simulations.}
-%\label{fig:path}
-%\includegraphics{PathDiagram}
-%\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%Plot Code
-
-\begin{figure}
-    \caption{General model used for the simulations.}
-    \label{fig:path}
-\begin{tikzpicture}
-[latent/.style={circle,draw=black!50,fill=white!20,thick,
-                inner sep=0pt, minimum width=20mm, minimum height=10mm},
- manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
-                inner sep=0pt,minimum width=20mm, minimum height=10mm},
- error/.style={circle,draw=black!50,fill=white!20,thick,
-                inner sep=0pt, minimum width=10mm},
- post/.style={->,shorten >=1pt,>=stealth',semithick},
- sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=20mm,semithick},
- errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',auto,distance=10mm,semithick}]
-
-%position objects.
-
-\node[latent] (VERBAL) at (0,4) {VERBAL};
-\node[latent] (MATH)   at (0,0) {MATH};
-
-\node[manifest] (SCALE1) at(5,6) {SCALE1};
-\node[manifest] (SCALE2) at (5,4.5) {SCALE2};
-\node[manifest] (SCALE3) at (5,3) {SCALE3};
-\node[manifest] (SCALE4) at(5,1) {SCALE4};
-\node[manifest] (SCALE5) at (5,-0.5) {SCALE5};
-\node[manifest] (SCALE6) at (5,-2) {SCALE6};
-
-\node[error] (E1) at (7,6) {E1};
-\node[error] (E2) at (7,4.5) {E2};
-\node[error] (E3) at (7,3) {E3};
-\node[error] (E4) at (7,1) {E4};
-\node[error] (E5) at (7,-0.5) {E5};
-\node[error] (E6) at (7,-2) {E6};
-
-%arrows, slings and texts
-\node[latent] (VERBAL) at (0,4) {VERBAL}
-   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{11}=4.92$}} (SCALE1)
-   edge [post] node[above=0.001mm,text width=2cm]{\scriptsize{$\lambda_{21}=2.96$}} (SCALE2)
-   edge [post] node[above=0.1cm,text width=2cm]{\scriptsize{$\lambda_{31}=5.96$}} (SCALE3)
-   edge [sling,bend right=100] node[auto,swap]{1}(VERBAL.-180)
-   edge [sling,bend right=70] node[auto,swap]{$\phi_{12}=-0.48$}(MATH)  ;
-
-\node[latent] (MATH) at (0,0) {MATH}
-   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{42}=3.24$}}(SCALE4)
-   edge [post] node[above=0.01cm,text width=2cm]{\scriptsize{$\lambda_{52}=4.32$}}(SCALE5)
-   edge [post] node[above=0.25cm,text width=2cm]{\scriptsize{$\lambda_{62}=7.21$}}(SCALE6)
-   edge [sling,bend right=100] node[auto,swap]{1}(MATH.270);
-
-\node[error] (E1) at (7,6) {E1}
-   edge [post] node[auto,swap]{\small{1}}(SCALE1)
-   edge [errorsling, bend right=100]node[right=0.1,text width=2cm]{\scriptsize{$\psi_{11}=26.77$}} (E1.45);
-\node[error] (E2) at (7,4.5) {E2}
-   edge [post] node[auto,swap]{\small{1}} (SCALE2)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{22}=13.01$}}(E2.45);
-\node[error] (E3) at (7,3) {E3}
-   edge [post] node[auto,swap]{\small{1}}(SCALE3)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{33}=30.93$}}(E3.45);
-\node[error] (E4) at (7,1) {E4}
-   edge [post] node[auto,swap]{\small{1}}(SCALE4)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{44}=3.17$}}(E4.45);
-\node[error] (E5) at (7,-0.5) {E5}
-   edge [post] node[auto,swap]{\small{1}}(SCALE5)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{55}=8.82$}} (E5.45);
-\node[error] (E6) at (7,-2) {E6}
-   edge [post] node[auto,swap]{\small{1}}(SCALE6)
-   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]{\scriptsize{$\psi_{66}=22.5$}}(E6.45);
-
-\end{tikzpicture}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Power and Type I error were examined across three sample sizes
-($n=120,480,960$), three numbers of categories ($m=4,8,12$), and
-17 magnitudes of invariance violations (described in the following sentences). 
-The measurement invariance violations began at level $1+m/2$ of the
-auxiliary variable $V$ and were consistent thereafter: individuals below
-level $1+m/2$ of $V$ deviated from individuals at or above level $1+m/2$ 
-by $d$ times the parameters' asymptotic standard errors
-(scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$. 
-For each combination of sample size ($n$) $\times$ violation
-magnitude ($d$) $\times$ violating parameter $\times$ categories ($m$), 5000 datasets were
-generated and tested.  Statistics from Equations~\eqref{eq:wdm}, \eqref{eq:maxlmo}
-and~\eqref{eq:lmuo} were examined. As mentioned previously,
-\eqref{eq:lmuo} is asymptotically equivalent to the usual likelihood
-ratio test.  Thus, this statistic provides information about the
-relative performance of the ordinal statistics vs.\ the LRT.
-
-In all conditions, we maintained equal sample 
-sizes in each subgroup of the ordinal variable.
-
-\subsection{Results}
-
-Full simulation results are presented in Figures~\ref{fig:sim11res}
-to~\ref{fig:sim14res}.
-Figure~\ref{fig:sim11res} displays power curves as a function of 
-violation magnitude in the factor loading $\lambda_{11}$,
-with the parameters being tested changing across rows,
-the number of levels $m$ of the ordinal variable $V$ across columns,
-and lines reflecting different test statistics.
-Figure~\ref{fig:sim12res}, Figure~\ref{fig:sim13res} and 
-Figure~\ref{fig:sim14res} display 
-similar power curves when the factor covariance 
-$\phi_{12}$, error variance $\epsilon_{11}$ and intercept
-$\mu_{11}$ violate measurement
-invariance, respectively. In these figures, we generally show tests
-associated with parameters that exhibited nonzero power curves.  For
-example, in Figure~\ref{fig:sim11res}, the middle row shows that power
-for tests of $\psi_{11}$ stays near zero for all values of $m$ and $d$.
-Similar rows have been omitted from this figure and other figures.
-
-Within each panel of Figure~\ref{fig:sim11res} to 
-Figure~\ref{fig:sim14res}, the three lines reflect the three test
-statistics.  It is seen that the two ordinal statistics exhibit 
-similar results, with $\max\mathit{LM}_{uo}$ demonstrating lower power
-across all situations.  This demonstrates the sensitivity of the
-ordinal statistics to invariance violations that are monotonic with
-$V$.  In situations where only one parameter is tested,
-$\mathit{WDM}_o$  and $\max \mathit{LM}_o$ exhibit equivalent power
-curves.  This is because, when only one parameter is tested, the
-statistics are equivalent.
-
-From these figures, one generally observes that the tests isolate 
-the parameter violating measurement invariance.  Additionally, the tests 
-have somewhat higher power to detect measurement invariance violations 
-in the factor loading, factor covariance and intercept parameters, 
-as opposed to the error variance parameter.  
-Finally, simultaneous tests of all factor loadings, all intercepts or 
-all error parameters result in decreased power, as compared to the situation
-where one tests only the violating parameter.  This occurs because, in
-testing a subset of parameters (only one of which violates measurement
-invariance), we are effectively dampening the signal of a measurement
-invariance violation. This ``dampening'' effect is more apparent for
-the $\max \mathit{LM}_o$ statistic, because it involves a sum across
-all tested parameters (see Equation~\eqref{eq:maxlmo}).
-Conversely, $\mathit{WDM}_o$ takes the maximum over parameters
-(Equation~\eqref{eq:wdm}), so that invariant parameters have no
-impact on this statistic.
-
-In summary, we found that the proposed tests can attribute
-measurement invariance violations to the correct parameter.  This
-provides evidence that, in practice, one can have confidence in the
-tests' abilities to locate the measurement invariance violation.  Of
-course, this statement is qualified by the fact that, in this
-simulation, the model was correctly specified.  In the following
-simulation, we examine the tests' performance in the likely situation
-of model misspecification.
-
-<<sim1,echo=FALSE>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1163)
-
-if(file.exists("sim1.rda")) {
-  load("sim1.rda")
-} else {
-  sim1 <- simulation(sim = "sim1", nobs = 480, parms = c("loading", "error", "var", "intercept"))
-  sim1$nlevels <- factor(sim1$nlevels)
-  levels(sim1$nlevels) <- paste("m=", levels(sim1$nlevels), sep="")
-
-  sim1$nobs <- factor(sim1$nobs)
-  levels(sim1$nobs) <- paste("n=", levels(sim1$nobs), sep="")
-  save(sim1, file="sim1.rda")
-}
-sim1$test <- factor(as.character(sim1$test), levels = c("ordmax","ordwmax","catdiff"),
-                    labels = c("maxLM_o", "WDM_o", "LM_uo"))
-parlabs <- c(expression(lambda[11]), expression(phi[12]), 
-             expression(psi[11]), expression(mu[11]), expression(list(lambda[11], ldots, lambda[62])), 
-             expression(list(psi[11], ldots, psi[66])), 
-             expression(list(mu[11], ldots, mu[66])))
-levels(sim1$pars) <- c("Loading1", "Covariance", "Error1", "Intercept1", 
-                       "All Loadings", "All Errors", "All Intercepts")  
-@
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and
-$\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The parameter violating measurement 
-invariance is $\lambda_{11}$.  Panel labels denote the parameter(s)
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim11res}
-<<sim11res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "loading" & 
-		 pars %in% c("Loading1", "Error1", "All Loadings") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", lambda[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-           if(which.given == 2){
-             strip.default(which.given, factor.levels = parlabs[c(1, 3, 5)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and
-$\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$,
-  and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
-parameter violating measurement invariance is $\phi_{12}$. Panel
-labels denote the parameter(s)  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim12res}
-<<sim12res, fig=TRUE, height=4.5, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "var" & 
-		 pars %in% c("Covariance") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", phi[12], ")")),
-       ylab="Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(2)],...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
-parameter violating measurement invariance is $\psi_{11}$. Panel labels
-denote the parameter(s) 
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim13res}
-<<sim13res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "error" & 
-		 pars %in% c("Error1", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", psi[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(3,6)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 1. The
-parameter violating measurement invariance is $\mu_{11}$. Panel labels
-denote the parameter(s) 
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim14res}
-<<sim14res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "intercept" & 
-		 pars %in% c("Intercept1", "All Intercepts") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", mu[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(4,7)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\section{Simulation 2}
-In Simulation 2, we examine the extent to which the results of
-Simulation 1 are robust to model misspecification.  Specifically, we
-generate data from the factor analysis model used in the previous
-section, except that the model contains an extra loading from the
-second factor to Scale 1.
-The estimated model matches that displayed in Figure~\ref{fig:path},
-however, resulting in model misspecification.  The goal of this
-simulation is to examine the proposed statistics' 
-power to detect measurement invariance violations (and to attribute
-the violation to the correct parameter) under this misspecification.
-
-\subsection{Method}
-A measurement invariance violation could occur in
-each of the four parameters from Simulation 1 (factor loading, factor
-covariance, unique variance and intercept), and a violation could also occur in
-the extra, unmodeled loading. In each condition, a single parameter
-exhibited the violation. 
-Sample size and magnitude of measurement invariance violation were manipulated 
-in the same way as they were in Simulation 1.  The tested parameters
-were also the same as Simulation 1.
-
-\subsection{Results}
-Results of
-primary interest are conditions where the unmodeled loading violates
-measurement invariance.  A subset of results is displayed in Figure~\ref{fig:sim21res}.  One can generally observe
-that tests of the first loading and unique variance exhibited high
-``power,'' which is actually a high Type I error rate here.  This Type
-I error also demonstrated when testing all loadings and all unique
-variances, as shown in the appendix.  Tests
-associated with the factor 
-covariance and intercept did not demonstrate this error, however.  In
-terms of specific 
-statistic performance, $\max \mathit{LM}_o$ and $\mathit{WDM}_o$ 
-demonstrated higher Type I error than $\mathit{LM}_{uo}$ in each panel,
-especially with increasing levels.  This is likely because the unmodeled
-loading's non-invariance was monotonic with $V$; if it were not monotonic, we
-would expect $\mathit{LM}_{uo}$ to have higher Type I error.
-
-When the parameter violating measurement invariance was modeled,
-results were generally the same as Simulation 1.  
-When the modeled factor loading, $\lambda_{11}$, violated measurement
-invariance, the statistics were generally able to pick up the violation despite
-the misspecification. Similar results were observed
-when the unique variance, intercept and factor covariance parameters violated measurement
-invariance; these results are all shown in the Appendix.
-In particular, the power of ordered statistics $\max \mathit{LM}_o$ test and
-$\mathit{WDM}_o$ test showed higher power than unordered $\mathit{LM}_{uo}$ in
-each of these panels.
-
-In summary, the proposed test statistics appear robust to unmodeled
-loading parameters, so long 
-as the unmodeled loading does not violate measurement invariance.  If
-the unmodeled loading  
-does violate measurement invariance, the tests can still detect
-measurement invariance violations.  The violations are assigned to 
-modeled parameters that do not violate measurement invariance, however.
-The impacted parameters include the error 
-variance and other loadings associated with the manifest variable that
-has an unmodeled loading.
-Thus, as for other tests of measurement invariance, it is important to
-study the extent to 
-which the hypothesized model includes all parameters of importance.
-None of these tests (score-based or otherwise) inform the researcher of
-model misspecification.
-
-<<sim2,echo=FALSE>>=
-## seed for replication
-RNGkind(kind = "default", normal.kind = "default")
-set.seed(1163)
-
-if(file.exists("sim2.rda")) {
-  load("sim2.rda")
-} else {
-  sim2 <- simulation(sim = "sim2", nobs = 480, 
-  parms = c("extra", "extra+loading", "extra+var", "extra+error", "extra+intercept"))
-
-  save(sim2, file="sim2.rda")
-}
-sim2$test <- factor(as.character(sim2$test),
-  levels = c("ordmax", "ordwmax", "catdiff"),
-  labels = c("maxLM_o", "WDM_o", "LM_uo"))
-  sim2$nlevels <- factor(sim2$nlevels)
-  levels(sim2$nlevels) <- paste("m=", levels(sim2$nlevels), sep="")
-  sim2$nobs <- factor(sim2$nobs)
-  levels(sim2$nobs) <- paste("n=", levels(sim2$nobs), sep="")
-  parlabs <- c(expression(lambda[11]), expression(phi[12]), 
-               expression(psi[11]), expression(mu[11]),
-	       expression(list(lambda[11], ldots, lambda[62])), 
-               expression(list(psi[11], ldots, psi[66])),
-               expression(list(mu[11], ldots, mu[66])))
-  levels(sim2$pars) <- c("Loading1", "Covariance", "Error1", "Intercept1", 
-                         "All Loadings", "All Errors", "All Intercepts")
-
-@
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is the unmodeled
-loading. Panel labels denote the parameter(s) 
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim21res}
-<<sim21res, fig=TRUE, height=9, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra" & 
-		 pars %in% c("Loading1", "Covariance", "Error1", "Intercept1") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude (Unmodeled Loading)",
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 2, 3, 4)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\section{General Discussion}
-
-In this paper, we first described a novel family of test statistics
-for measurement invariance and illustrated their use via the
-\proglang{R} packages \pkg{lavaan} and \pkg{strucchange}.  Next, we examined these
-statistics' abilities to identify the parameter violating measurement
-invariance under well-specified and misspecified models.  We found
-that the
-proposed statistics could generally isolate the model parameter
-violating measurement invariance, so long as the violating parameter
-is included in the model.
-
-In the remainder of the paper, we first compare
-these tests to traditional tests of measurement invariance.  We then 
-discuss test extension to other fit functions and to other specialized
-models.
-
-\subsection{Applications}
-
-Many of the applications in this volume, along with many measurement
-invariance applications in
-general, focus on testing across unordered categories
-such as nations or gender.  As discussed earlier in this paper, the score-based
-tests for unordered categories are essentially equivalent to the usual
-likelihood ratio test.  Given a measurement invariance violation across these
-unordered categories, however, researchers typically wish to know why the
-violation occurred.  At this point, researchers may examine education level,
-socioeconomic status, income levels, and so on across the unordered categories. 
-These variables are often ordinal or continuous in nature, so that the family of
-tests described in this paper are applicable.  This is a first step towards
-describing why measurement invariance violations occur, as opposed to simply
-detecting measurement invariance violations. 
-%% TODO!!
-%\readme{In the abstracts for the
-%special issue (which the special editors sent us), one paper was proposed along
-%these lines.  I will get a draft of that paper and then we can cite it here.}
-The tests described here are convenient for this purpose, as they do
-not require a new model to 
-be estimated for each ordinal variable.  Instead, each ordinal variable defines
-an ordering of observations, which in turn yields a test statistic that is
-specific to that ordinal variable.
-
-%% I'm not sure we want to make a big deal of this.  If there are only a small number of ties, we are probably better off with the continuous tests:
-%These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context.
-It is very common to observe several individuals with the same age, IQ, income rank, etc.
-Therefore, the ordinal tests which dealing with multiple ties in auxiliary variable is more suitable than the continuous test.
-
-%Moreover, even though researchers might be originally interested in comparing categorical differences.
-For example, researchers might want to test whether there is gender difference or cultural difference for the scale.
-Using traditional likelihood test or wald test, there might existing some differences regarding to the categorical variable.
-Then the next natural question would be why there is gender or culture differences?
-.
-Thus these test statistics offer an convenient assessing tool to further exploring research hypothesis without making complicated assumptions or fitting multiple different models.
-
-\subsection{Extension}
-
-In this paper, we focused on testing for measurement invariance in factor
-analysis models that assume multivariate normality and that are estimated via
-maximum likelihood (ML).  The family of tests described here generally apply to estimation methods
-that maximize/minimize a fit function, however \cite<see>{zeihor07}, so they are
-potentially applicable to alternative SEM discrepancy functions such
-as generalized 
-least squares \cite<e.g.,>{broarm95}.  Score calculation for these alternative
-discrepancy functions has not been implemented (to our knowledge), though the
-calculation could be implemented.  Test statistic calculation and
-inference would then proceed in exactly the same manner as the calculation and
-inference illustrated in this paper.
-
-In addition to alternative fit functions, the tests can be extended to other
-models estimated via ML.  Of primary interest to the topic of measurement
-invariance, the tests can be extended to item response models to examine
-differential item functioning.  In particular, \citeA{strkop13} studied
-application of these tests to the Rasch model, using them as the basis of a
-recursive partioning procedure that segments subgroups of individuals who
-exhibit DIF.  Further study and extension of these tests for IRT seem warranted.
-
-% Removed for now:
-%% \subsection{Conclusion}
-%% \readme{Not sure whether we need this paragraph, but I still need to think about it.}
-%% In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable.
-This
-%% feature is robust as long as the violation parameter is included in the testing model.
-Furthermore, as illustrated in tutorial part, the use of these tests is convenient without changing the focal psychometric model.
-In all, the tests have advantageous properties that should be implemented in practice.
-
-\section*{Computational Details}
-
-All results were obtained using the \proglang{R}~system for statistical computing \cite{R11},
-version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on package
-\pkg{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} for fitting 
-of the factor analysis models and
-\pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
-for evaluating the parameter instability tests. \proglang{R}~and both
-packages are freely available under the General Public License~2 from the
-Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/}.
-\proglang{R}~code for replication of our results is
-available at \url{http://semtools.R-Forge.R-project.org/}.
-
-\bibliography{refs}
-
-\section{Appendix}
-
-This Appendix contains additional results from Simulation 2, where there existed
-model misspecification (lacking one loading from Scale 1 to
-Math). Figure~\ref{fig:sim22res} to Figure~\ref{fig:sim26res} display
-results when the unmodeled loading, the loading $\lambda_{11}$, the
-covariance $\phi_{12}$, the error term $\psi_{11}$ and intercept $\mu_{11}$
-violate invariance, respectively.  Results are only shown for simulation
-conditions exhibiting power curves that increased from zero.  These
-results were generally the same as the Simulation 1 results.
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is the unmodeled
-loading. Panel labels denote the parameter(s) 
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim22res}
-<<sim22res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra" & 
-		 pars %in% c("All Loadings", "All Errors", "All Intercepts") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude (Unmodeled Loading)", ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(5, 6, 7)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is $\lambda_{11}$.  Panel
-labels denote the parameter(s)  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim23res}
-<<sim23res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms=="extra+loading" & 
-		 pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", lambda[11], ")")), ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 5)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is $\phi_{12}$.  Panel
-labels denote the parameter  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim24res}
-<<sim24res, fig=TRUE, height=4.5, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms=="extra+var" & 
-		 pars %in% c("Covariance") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", phi[12], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(2, 4:5)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is $\psi_{11}$.  Panel
-labels denote the parameter(s)  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim25res}
-<<sim25res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra+error" & 
-		 pars %in% c("Error1", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", psi[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(3,6)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is $\mu_{11}$.  Panel
-labels denote the parameter(s)  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim26res}
-<<sim26res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra+intercept" & 
-		 pars %in% c("Intercept1", "All Intercepts") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", mu[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(4,7)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
 \end{document}

### 2014-02-19 frontiers/IMPS.Rnw
Final modifications to frontiers prior to resubmission, and addition of IMPS abstract.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@318 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \abstract{
-In this paper, we consider a family of  recently-proposed measurement invariance
+In this presentation, we study a family of  recently-proposed measurement invariance
 tests that are based on the {\em scores} of a fitted model.  This family can be used
@@
 monotonically related to the ordinal variable (and less sensitive to
-non-monotonic violations).  The paper is specifically aimed at potential users
-of the tests who may wish to know (i) how the tests can be employed for their
-data, and (ii) whether the tests can accurately identify specific models parameters that violate measurement invariance (possibly in the presence of model misspecification).  After providing an
-overview of the tests, we illustrate their general use via the \proglang{R} packages \pkg{lavaan}
-and \pkg{strucchange}.  We then describe two novel simulations that provide evidence
-of the tests' practical abilities.  As a whole, the paper provides researchers
-with the tools and knowledge needed to apply these tests to general measurement
-invariance scenarios.
+non-monotonic violations).  After providing an overview of the family
+of tests, we present new simulation-based results examining (i) the extent
+to which the tests can identify specific parameters violating
+measurement invariance, and (ii) the extent to which the tests are
+impacted by model misspecification.
+We also describe an R implementation that allows users to generally
+apply the tests to SEMs estimated under multivariate normality.
 }
 
-\acknowledgements{This work was supported by National Science
-  Foundation grant SES-1061334. 
-  Correspondence to Edgar Merkle. 
-  Email: merklee@missouri.edu.}
-\shorttitle{Score-based tests in practice}
-\rightheader{Score-based tests in practice}
+% \acknowledgements{This work was supported by National Science
+%   Foundation grant SES-1061334. 
+%   Correspondence to Edgar Merkle. 
+%   Email: merklee@missouri.edu.}
+% \shorttitle{Score-based tests in practice}
+% \rightheader{Score-based tests in practice}
 
-\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
 \let\proglang=\textsf
@@
 \let\code=\texttt
-
-%% for internal use
-\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
-\newcommand{\readme}[1]{\emph{\marginpar{Ed} (#1)}}
-\newcommand{\seeme}[1]{\emph{\marginpar{Ting} (#1)}}
 

### 2014-02-19 frontiers/Supple_Material.Rnw
Final modifications to frontiers prior to resubmission, and addition of IMPS abstract.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@318 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\maketitle
+%\maketitle
 
@@
 
-\section{Appendix}
+\section{Supplementary Material to Wang, Merkle, \& Zeileis}
 
-This Appendix contains additional results from Simulation 2, where there existed
+The Supplementary Material contains additional results from Simulation
+2, where there existed 
 model misspecification (lacking one loading from Scale 1 to

### 2014-02-19 frontiers/frontiers.Rnw
minimal touch-ups in the wording
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@316 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Final modifications to frontiers prior to resubmission, and addition of IMPS abstract.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@318 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 For illustrative purposes, we conduct multiple tests and
-compare them to the traditional significance level of .05.  In
+compare them to the traditional significance level of 0.05.  In
 practice, however, one should generally adjust the significance level
@@
 by $d$ times the parameters' asymptotic standard errors
-(scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$. 
+(scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$ (see replication code
+for specific values of the standard errors).
 For each combination of sample size ($n$) $\times$ violation
@@
 In all conditions, we maintained equal sample 
-sizes in each subgroup of the ordinal variable.
+sizes in each subgroup of the ordinal variable.  Aside from the
+parameter changes that reflect measurement invariance, the fitted
+models matched the data generating model.
 
@@
 I error is also observed when testing all loadings and all unique
-variances (see the Appendix).  Tests
+variances (see the Supplementary Material).  Tests
 associated with the factor 
@@
 when the unique variance, intercept and factor covariance parameters violated measurement
-invariance; these results are all shown in the Appendix.
+invariance; these results are all shown in the Supplementary Material.
 In particular, power of the ordered statistics was higher than power
@@
 In summary, the proposed test statistics appear robust to unmodeled
-loading parameters, so long 
-as the unmodeled loading does not violate measurement invariance.  If
+loading parameters, when
+the unmodeled loading does not violate measurement invariance and when
+the rest of the model is correctly specified (save for the measurement
+invariance violation).  If
 the unmodeled loading  
@@
 
-\section{Appendix}
-
-This Appendix contains additional results from Simulation 2, where there existed
-model misspecification (lacking one loading from Scale 1 to
-Math). Figures~\ref{fig:sim22res} to~\ref{fig:sim26res} display
-results when the unmodeled loading, the loading $\lambda_{11}$, the
-covariance $\phi_{12}$, the error term $\psi_{11}$, and the intercept $\mu_{11}$
-violate invariance, respectively.  Results are only shown for simulation
-conditions exhibiting power curves that increased from zero.  These
-results were generally the same as the Simulation 1 results.
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is the unmodeled
-loading. Panel labels denote the parameter(s) 
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim22res}
-<<sim22res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra" & 
-		 pars %in% c("All Loadings", "All Errors", "All Intercepts") & diff %% 0.5 == 0),
-       type = "b", xlab = "Violation Magnitude (Unmodeled Loading)", ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(5, 6, 7)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is $\lambda_{11}$.  Panel
-labels denote the parameter(s)  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim23res}
-<<sim23res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms=="extra+loading" & 
-		 pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", lambda[11], ")")), ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(1, 5)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is $\phi_{12}$.  Panel
-labels denote the parameter  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim24res}
-<<sim24res, fig=TRUE, height=4.5, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms=="extra+var" & 
-		 pars %in% c("Covariance") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", phi[12], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(2, 4:5)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is $\psi_{11}$.  Panel
-labels denote the parameter(s)  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim25res}
-<<sim25res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra+error" & 
-		 pars %in% c("Error1", "All Errors") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", psi[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(3,6)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
-
-\begin{figure}
-\caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
-violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
-parameter violating measurement invariance is $\mu_{11}$.  Panel
-labels denote the parameter(s)  
-being tested and the number of levels of the ordinal variable $m$.}
-\label{fig:sim26res}
-<<sim26res, fig=TRUE, height=7, width=9, echo=FALSE>>=
-trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
-       subset = (parms == "extra+intercept" & 
-		 pars %in% c("Intercept1", "All Intercepts") & diff %% 0.5 == 0),
-       type = "b", xlab = expression(paste("Violation Magnitude (", mu[11], ")")),
-       ylab = "Power", key = mykey, as.table = TRUE,
-       strip = function(..., which.given, factor.levels){
-	   if(which.given == 2){
-	       strip.default(which.given, factor.levels = parlabs[c(4,7)], ...)
-	   } else {
-	       strip.default(which.given, factor.levels = factor.levels, ...)
-	   }
-       })
-@
-
-\end{figure}
+%% \section{Appendix}
+
+%% This Appendix contains additional results from Simulation 2, where there existed
+%% model misspecification (lacking one loading from Scale 1 to
+%% Math). Figures~\ref{fig:sim22res} to~\ref{fig:sim26res} display
+%% results when the unmodeled loading, the loading $\lambda_{11}$, the
+%% covariance $\phi_{12}$, the error term $\psi_{11}$, and the intercept $\mu_{11}$
+%% violate invariance, respectively.  Results are only shown for simulation
+%% conditions exhibiting power curves that increased from zero.  These
+%% results were generally the same as the Simulation 1 results.
+
+%% \begin{figure}
+%% \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+%% violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
+%% parameter violating measurement invariance is the unmodeled
+%% loading. Panel labels denote the parameter(s) 
+%% being tested and the number of levels of the ordinal variable $m$.}
+%% \label{fig:sim22res}
+%% <<sim22res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+%% trellis.par.set(theme = canonical.theme(color = FALSE))
+%% mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+%% xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+%%        subset = (parms == "extra" & 
+%% 		 pars %in% c("All Loadings", "All Errors", "All Intercepts") & diff %% 0.5 == 0),
+%%        type = "b", xlab = "Violation Magnitude (Unmodeled Loading)", ylab = "Power", key = mykey, as.table = TRUE,
+%%        strip = function(..., which.given, factor.levels){
+%% 	   if(which.given == 2){
+%% 	       strip.default(which.given, factor.levels = parlabs[c(5, 6, 7)], ...)
+%% 	   } else {
+%% 	       strip.default(which.given, factor.levels = factor.levels, ...)
+%% 	   }
+%%        })
+%% @
+
+%% \end{figure}
+
+%% \begin{figure}
+%% \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+%% violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
+%% parameter violating measurement invariance is $\lambda_{11}$.  Panel
+%% labels denote the parameter(s)  
+%% being tested and the number of levels of the ordinal variable $m$.}
+%% \label{fig:sim23res}
+%% <<sim23res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+%% trellis.par.set(theme = canonical.theme(color = FALSE))
+%% mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+%% xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+%%        subset = (parms=="extra+loading" & 
+%% 		 pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
+%%        type = "b", xlab = expression(paste("Violation Magnitude (", lambda[11], ")")), ylab = "Power", key = mykey, as.table = TRUE,
+%%        strip = function(..., which.given, factor.levels){
+%% 	   if(which.given == 2){
+%% 	       strip.default(which.given, factor.levels = parlabs[c(1, 5)], ...)
+%% 	   } else {
+%% 	       strip.default(which.given, factor.levels = factor.levels, ...)
+%% 	   }
+%%        })
+%% @
+
+%% \end{figure}
+
+%% \begin{figure}
+%% \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+%% violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
+%% parameter violating measurement invariance is $\phi_{12}$.  Panel
+%% labels denote the parameter  
+%% being tested and the number of levels of the ordinal variable $m$.}
+%% \label{fig:sim24res}
+%% <<sim24res, fig=TRUE, height=4.5, width=9, echo=FALSE>>=
+%% trellis.par.set(theme = canonical.theme(color = FALSE))
+%% mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+%% xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+%%        subset = (parms=="extra+var" & 
+%% 		 pars %in% c("Covariance") & diff %% 0.5 == 0),
+%%        type = "b", xlab = expression(paste("Violation Magnitude (", phi[12], ")")),
+%%        ylab = "Power", key = mykey, as.table = TRUE,
+%%        strip = function(..., which.given, factor.levels){
+%% 	   if(which.given == 2){
+%% 	       strip.default(which.given, factor.levels = parlabs[c(2, 4:5)], ...)
+%% 	   } else {
+%% 	       strip.default(which.given, factor.levels = factor.levels, ...)
+%% 	   }
+%%        })
+%% @
+
+%% \end{figure}
+
+%% \begin{figure}
+%% \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+%% violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
+%% parameter violating measurement invariance is $\psi_{11}$.  Panel
+%% labels denote the parameter(s)  
+%% being tested and the number of levels of the ordinal variable $m$.}
+%% \label{fig:sim25res}
+%% <<sim25res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+%% trellis.par.set(theme = canonical.theme(color = FALSE))
+%% mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+%% xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+%%        subset = (parms == "extra+error" & 
+%% 		 pars %in% c("Error1", "All Errors") & diff %% 0.5 == 0),
+%%        type = "b", xlab = expression(paste("Violation Magnitude (", psi[11], ")")),
+%%        ylab = "Power", key = mykey, as.table = TRUE,
+%%        strip = function(..., which.given, factor.levels){
+%% 	   if(which.given == 2){
+%% 	       strip.default(which.given, factor.levels = parlabs[c(3,6)], ...)
+%% 	   } else {
+%% 	       strip.default(which.given, factor.levels = factor.levels, ...)
+%% 	   }
+%%        })
+%% @
+
+%% \end{figure}
+
+%% \begin{figure}
+%% \caption{Simulated power curves for $\max \mathit{LM}_o$, $\mathit{WDM}_o$, and $\mathit{LM}_{uo}$ across three levels of the ordinal variable $m$ and measurement invariance 
+%% violations of 0--4 standard errors (scaled by $\sqrt{n}$), Simulation 2. The
+%% parameter violating measurement invariance is $\mu_{11}$.  Panel
+%% labels denote the parameter(s)  
+%% being tested and the number of levels of the ordinal variable $m$.}
+%% \label{fig:sim26res}
+%% <<sim26res, fig=TRUE, height=7, width=9, echo=FALSE>>=
+%% trellis.par.set(theme = canonical.theme(color = FALSE))
+%% mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+%% xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+%%        subset = (parms == "extra+intercept" & 
+%% 		 pars %in% c("Intercept1", "All Intercepts") & diff %% 0.5 == 0),
+%%        type = "b", xlab = expression(paste("Violation Magnitude (", mu[11], ")")),
+%%        ylab = "Power", key = mykey, as.table = TRUE,
+%%        strip = function(..., which.given, factor.levels){
+%% 	   if(which.given == 2){
+%% 	       strip.default(which.given, factor.levels = parlabs[c(4,7)], ...)
+%% 	   } else {
+%% 	       strip.default(which.given, factor.levels = factor.levels, ...)
+%% 	   }
+%%        })
+%% @
+
+%% \end{figure}
 

### 2014-03-03 frontiers/IMPSpres.Rnw
set up presentation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@322 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{beamer}
\setbeamertemplate{caption}[numbered]
\usepackage{beamerthemeSzeged,graphics,amsmath,bm,booktabs}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usefonttheme{serif}

\title{Score-based tests of measurement invariance: Use in practice}
\subtitle{2014 Annual Meeting of Psychometric Society}
\author[shortname]{Ting Wang\inst{1} \and Edgar C. Merkle\inst{1} \and 
  Achim Zeileis \inst{2}}
\institute[shortinst]{\inst{1} University of Missouri \and 
                      \inst{2} Universit\"{a}t Innsbruck}

\renewcommand{\baselinestretch}{1.1}

\begin{document}

<<setup, include=FALSE>>=
library("knitr")
opts_chunk$set(size='footnotesize',tidy.opts=list(width.cutoff=50,
               reindent.spaces=6))

## Color unevaluted commands using Sexpr
knit_hooks$set(inline = function(x) { 
  if (is.numeric(x)) return(knitr:::format_sci(x, 'latex'))
  highr:::hi_latex(x)
}) 
## packages
library("lavaan")
library("strucchange") ## at least 1.5-0
library("mvtnorm")
library("lattice")

## auxiliary code
source("../www/sim-frontiers.R")
source("../www/mz-frontiers.R")

## some options
options(prompt = "R> ", continue = "+  ", digits = 3,
  useFancyQuotes = FALSE, show.signif.stars = FALSE)
@

\frame{\titlepage}

\begin{frame}
\frametitle{Problem Scenario}

\end{frame}
%--------------------------------------------------------------------------

\begin{frame}
\frametitle{Overview} 
% Table of contents slide, comment this block out to remove it
\tableofcontents 
% Throughout your presentation, if you choose to use \section{} and 
%\subsection{} commands, these will automatically be printed on this slide 
%as an overview of your presentation
\end{frame}

\section{Theoretical Background}
\section{Simulation: Locating M.I. Violation Parameter Property}
\subsection{Violation Parameter is IN the Model}
\subsection{Violation Parameter is OUT the Model}
\section{Conclusion from Simulation Studies}
\section{Application}

\end{document}

### 2014-03-03 frontiers/knit.R
knit command to produce latex file.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@323 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,3 @@
+library("knitr")
+Sweave2knitr("IMPSpres.Rnw")
+knit("IMPSpres-knitr.Rnw")

### 2014-03-04 frontiers/IMPSpres.Rnw
update sim figure part.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@324 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   useFancyQuotes = FALSE, show.signif.stars = FALSE)
+
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim1.rda")) {
+  load("sim1.rda")
+} else {
+  sim1 <- simulation(sim = "sim1", nobs = 480, parms = c("loading", "error", 
+                                                   "var", "intercept"))
+  sim1$nlevels <- factor(sim1$nlevels)
+  levels(sim1$nlevels) <- paste("m=", levels(sim1$nlevels), sep="")
+
+  sim1$nobs <- factor(sim1$nobs)
+  levels(sim1$nobs) <- paste("n=", levels(sim1$nobs), sep="")
+  save(sim1, file="sim1.rda")
+}
+sim1$test <- factor(as.character(sim1$test), levels = c("ordmax","ordwmax",
+                                                 "catdiff"),
+                    labels = c("maxLM_o", "WDM_o", "LM_uo"))
+parlabs <- c(expression(lambda[11]), expression(phi[12]), 
+             expression(psi[11]), expression(mu[11]), expression
+             (list(lambda[11], ldots, lambda[62])), 
+             expression(list(psi[11], ldots, psi[66])), 
+             expression(list(mu[11], ldots, mu[66])))
+levels(sim1$pars) <- c("Loading1", "Covariance", "Error1", "Intercept1", 
+                       "All Loadings", "All Errors", "All Intercepts")  
 @
@@
 \subsection{Violation Parameter is IN the Model}
+
+\begin{frame}
+\begin{figure}
+\tiny{\caption{M.I.violation parameter $\lambda_{11}$}}
+\label{fig:sim11res}
+<<sim11res, fig=TRUE, height=5.5, width=9, echo=FALSE>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+       subset = (parms == "loading" & 
+          pars %in% c("Loading1", "Error1", "All Loadings") & diff %% 0.5 == 0),
+       type = "b", xlab = 
+          expression(paste("Violation Magnitude (", lambda[11], ")")),
+       ylab = "Power", key = mykey, as.table = TRUE,
+       strip = function(..., which.given, factor.levels){
+           if(which.given == 2){
+             strip.default(which.given, factor.levels = 
+                           parlabs[c(1, 3, 5)], ...)
+	   } else {
+	       strip.default(which.given, factor.levels = factor.levels, ...)
+	   }
+       })
+@
+
+\end{figure}
+\end{frame}
+
 \subsection{Violation Parameter is OUT the Model}

### 2014-03-04 frontiers/IMPSpres.Rnw
update presentation. need to work on application part.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@325 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2014-03-03 frontiers/IMPSpres.Rnw
@@
 \usepackage{beamerthemeSzeged,graphics,amsmath,bm,booktabs}
+\usepackage{tikz}
 \usepackage[T1]{fontenc}
@@
 \usefonttheme{serif}
+% Seems necessary for texlive 2012?
+\usetikzlibrary{arrows}
 
@@
   Achim Zeileis \inst{2}}
-\institute[shortinst]{\inst{1} University of Missouri \and 
+\institute[]{\inst{1} University of Missouri \and 
                       \inst{2} Universit\"{a}t Innsbruck}
@@
 library("knitr")
-opts_chunk$set(size='footnotesize',tidy.opts=list(width.cutoff=50,
+opts_chunk$set(size='footnotesize',tidy.opts=list(width.cutoff=45,
                reindent.spaces=6))
@@
   useFancyQuotes = FALSE, show.signif.stars = FALSE)
+
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim1.rda")) {
+  load("sim1.rda")
+} else {
+  sim1 <- simulation(sim = "sim1", nobs = 480, parms = c("loading", "error", 
+                                                   "var", "intercept"))
+  sim1$nlevels <- factor(sim1$nlevels)
+  levels(sim1$nlevels) <- paste("m=", levels(sim1$nlevels), sep="")
+
+  sim1$nobs <- factor(sim1$nobs)
+  levels(sim1$nobs) <- paste("n=", levels(sim1$nobs), sep="")
+  save(sim1, file="sim1.rda")
+}
+sim1$test <- factor(as.character(sim1$test), levels = c("ordmax","ordwmax",
+                                                 "catdiff"),
+                    labels = c("maxLM_o", "WDM_o", "LM_uo"))
+parlabs <- c(expression(lambda[11]), expression(phi[12]), 
+             expression(psi[11]), expression(mu[11]), expression
+             (list(lambda[11], ldots, lambda[62])), 
+             expression(list(psi[11], ldots, psi[66])), 
+             expression(list(mu[11], ldots, mu[66])))
+levels(sim1$pars) <- c("Loading1", "Covariance", "Error1", "Intercept1", 
+                       "All Loadings", "All Errors", "All Intercepts")
+
+## seed for replication
+RNGkind(kind = "default", normal.kind = "default")
+set.seed(1163)
+
+if(file.exists("sim2.rda")) {
+  load("sim2.rda")
+} else {
+  sim2 <- simulation(sim = "sim2", nobs = 480, 
+  parms = c("extra", "extra+loading", "extra+var", "extra+error", 
+           "extra+intercept"))
+
+  save(sim2, file="sim2.rda")
+}
+sim2$test <- factor(as.character(sim2$test),
+  levels = c("ordmax", "ordwmax", "catdiff"),
+  labels = c("maxLM_o", "WDM_o", "LM_uo"))
+  sim2$nlevels <- factor(sim2$nlevels)
+  levels(sim2$nlevels) <- paste("m=", levels(sim2$nlevels), sep="")
+  sim2$nobs <- factor(sim2$nobs)
+  levels(sim2$nobs) <- paste("n=", levels(sim2$nobs), sep="")
+  parlabs <- c(expression(lambda[11]), expression(phi[12]), 
+               expression(psi[11]), expression(mu[11]),
+	       expression(list(lambda[11], ldots, lambda[62])), 
+               expression(list(psi[11], ldots, psi[66])),
+               expression(list(mu[11], ldots, mu[66])))
+  levels(sim2$pars) <- c("Loading1", "Covariance", "Error1", "Intercept1", 
+                         "All Loadings", "All Errors", "All Intercepts")
 @
@@
 \end{frame}
+%--------------------------------------------------------------------------
+\section{Theoretical Background}
+\section{Simulation}
+\subsection{Violation Parameter is IN the Model}
+\begin{frame}
+ \begin{figure}
+    \caption{Model used for the simulation.}
+    \label{fig:pathinmodel}
+\resizebox{8.0cm}{!}{
+\begin{tikzpicture}
+[latent/.style={circle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt, minimum width=20mm, minimum height=10mm},
+ manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt,minimum width=20mm, minimum height=10mm},
+ error/.style={circle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt, minimum width=10mm},
+ post/.style={->,shorten >=1pt,>=stealth',semithick},
+ postred/.style={->,shorten >=1pt,>=stealth',semithick,red},
+ postdashed/.style={->,shorten >=1pt,>=stealth',semithick,red,dashed},
+ sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
+   auto,distance=20mm,semithick},
+ errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
+   auto,distance=10mm,semithick}]
 
-\section{Theoretical Background}
-\section{Simulation: Locating M.I. Violation Parameter Property}
-\subsection{Violation Parameter is IN the Model}
+%position objects.
+
+\node[latent] (VERBAL) at (0,4) {VERBAL};
+\node[latent] (MATH)   at (0,0) {MATH};
+
+\node[manifest] (SCALE1) at(5,6) {SCALE1};
+\node[manifest] (SCALE2) at (5,4.5) {SCALE2};
+\node[manifest] (SCALE3) at (5,3) {SCALE3};
+\node[manifest] (SCALE4) at(5,1) {SCALE4};
+\node[manifest] (SCALE5) at (5,-0.5) {SCALE5};
+\node[manifest] (SCALE6) at (5,-2) {SCALE6};
+
+\node[error] (E1) at (7,6) {E1};
+\node[error] (E2) at (7,4.5) {E2};
+\node[error] (E3) at (7,3) {E3};
+\node[error] (E4) at (7,1) {E4};
+\node[error] (E5) at (7,-0.5) {E5};
+\node[error] (E6) at (7,-2) {E6};
+
+%arrows, slings and texts
+\node[latent] (VERBAL) at (0,4) {VERBAL}
+   edge [postred] node[above=0.01cm,text width=2cm]
+   {\scriptsize{$\lambda_{11}=4.92$}} (SCALE1)
+   edge [post] node[above=0.001mm,text width=2cm]
+   {\scriptsize{$\lambda_{21}=2.96$}} (SCALE2)
+   edge [post] node[above=0.1cm,text width=2cm]
+   {\scriptsize{$\lambda_{31}=5.96$}} (SCALE3)
+   edge [sling,bend right=100] node[auto,swap]{1}(VERBAL.-180)
+   edge [sling,bend right=70] node[auto,swap]
+   {$\phi_{12}=-0.48$}(MATH)  ;
+
+\node[latent] (MATH) at (0,0) {MATH}
+   edge [post] node[above=0.01cm,text width=2cm]
+   {\scriptsize{$\lambda_{42}=3.24$}}(SCALE4)
+   edge [post] node[above=0.01cm,text width=2cm]
+   {\scriptsize{$\lambda_{52}=4.32$}}(SCALE5)
+   edge [post] node[above=0.25cm,text width=2cm]
+   {\scriptsize{$\lambda_{62}=7.21$}}(SCALE6)
+   edge [sling,bend right=100] node[auto,swap]{1}(MATH.270);
+
+\node[error] (E1) at (7,6) {E1}
+   edge [post] node[auto,swap]{\small{1}}(SCALE1)
+   edge [errorsling, bend right=100]node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{11}=26.77$}} (E1.45);
+\node[error] (E2) at (7,4.5) {E2}
+   edge [post] node[auto,swap]{\small{1}} (SCALE2)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{22}=13.01$}}(E2.45);
+\node[error] (E3) at (7,3) {E3}
+   edge [post] node[auto,swap]{\small{1}}(SCALE3)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{33}=30.93$}}(E3.45);
+\node[error] (E4) at (7,1) {E4}
+   edge [post] node[auto,swap]{\small{1}}(SCALE4)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{44}=3.17$}}(E4.45);
+\node[error] (E5) at (7,-0.5) {E5}
+   edge [post] node[auto,swap]{\small{1}}(SCALE5)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{55}=8.82$}} (E5.45);
+\node[error] (E6) at (7,-2) {E6}
+   edge [post] node[auto,swap]{\small{1}}(SCALE6)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{66}=22.5$}}(E6.45);
+\end{tikzpicture}}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\begin{figure}
+\tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
+    parameters are $\lambda_{11}$ and $\lambda_{11} \ldots \lambda_{62}$.}}
+\label{fig:sim11res}
+<<sim11res, fig=TRUE, height=5, width=9, echo=FALSE>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+       subset = (parms == "loading" & 
+          pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
+       type = "b", xlab = 
+          expression(paste("Violation Magnitude (", lambda[11], ")")),
+       ylab = "Power", key = mykey, as.table = TRUE,
+       strip = function(..., which.given, factor.levels){
+           if(which.given == 2){
+             strip.default(which.given, factor.levels = 
+                           parlabs[c(1, 5)], ...)
+	   } else {
+	       strip.default(which.given, factor.levels = factor.levels, ...)
+	   }
+       })
+@
+
+\end{figure}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+\begin{figure}
+\tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
+    parmaeter is $\psi_{11} \ldots \psi_{66}, \sigma_{12}$ and
+    $\mu_{11} \ldots \mu_{66}$.}}
+\label{fig:sim12res}
+<<sim12res, fig=TRUE, height=5, width=9, echo=FALSE>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
+       subset = (parms == "loading" & 
+          pars %in% c("All Errors", "Covariance", "All Intercepts") 
+                 & diff %% 0.5 == 0),
+       type = "b", xlab = 
+          expression(paste("Violation Magnitude (", lambda[11], ")")),
+       ylab = "Power", ylim = c(0,1), key = mykey, as.table = TRUE,
+       strip = function(..., which.given, factor.levels){
+           if(which.given == 2){
+             strip.default(which.given, factor.levels = 
+                           parlabs[c(2, 6, 7)], ...)
+	   } else {
+	       strip.default(which.given, factor.levels = factor.levels, ...)
+	   }
+       })
+@
+
+\end{figure}
+\end{frame}
+%--------------------------------------------------------------------------
+
 \subsection{Violation Parameter is OUT the Model}
-\section{Conclusion from Simulation Studies}
+\begin{frame}
+\begin{figure}
+    \caption{Model used for the simulation.}
+    \label{fig:pathinmodel}
+\resizebox{8.0cm}{!}{
+\begin{tikzpicture}
+[latent/.style={circle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt, minimum width=20mm, minimum height=10mm},
+ manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt,minimum width=20mm, minimum height=10mm},
+ error/.style={circle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt, minimum width=10mm},
+ post/.style={->,shorten >=1pt,>=stealth',semithick},
+ postred/.style={->,shorten >=1pt,>=stealth',semithick,red},
+ postdashed/.style={->,shorten >=1pt,>=stealth',semithick,red,dashed},
+ sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
+   auto,distance=20mm,semithick},
+ errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
+   auto,distance=10mm,semithick}]
+
+%position objects.
+
+\node[latent] (VERBAL) at (0,4) {VERBAL};
+\node[latent] (MATH)   at (0,0) {MATH};
+
+\node[manifest] (SCALE1) at(5,6) {SCALE1};
+\node[manifest] (SCALE2) at (5,4.5) {SCALE2};
+\node[manifest] (SCALE3) at (5,3) {SCALE3};
+\node[manifest] (SCALE4) at(5,1) {SCALE4};
+\node[manifest] (SCALE5) at (5,-0.5) {SCALE5};
+\node[manifest] (SCALE6) at (5,-2) {SCALE6};
+
+\node[error] (E1) at (7,6) {E1};
+\node[error] (E2) at (7,4.5) {E2};
+\node[error] (E3) at (7,3) {E3};
+\node[error] (E4) at (7,1) {E4};
+\node[error] (E5) at (7,-0.5) {E5};
+\node[error] (E6) at (7,-2) {E6};
+
+%arrows, slings and texts
+\node[latent] (VERBAL) at (0,4) {VERBAL}
+   edge [post] node[above=0.01cm,text width=2cm]
+   {\scriptsize{$\lambda_{11}=4.92$}} (SCALE1)
+   edge [post] node[above=0.001mm,text width=2cm]
+   {\scriptsize{$\lambda_{21}=2.96$}} (SCALE2)
+   edge [post] node[above=0.1cm,text width=2cm]
+   {\scriptsize{$\lambda_{31}=5.96$}} (SCALE3)
+   edge [sling,bend right=100] node[auto,swap]{1}(VERBAL.-180)
+   edge [sling,bend right=70] node[auto,swap]
+   {$\phi_{12}=-0.48$}(MATH)  ;
+
+\node[latent] (MATH) at (0,0) {MATH}
+   edge [post] node[above=0.01cm,text width=2cm]
+   {\scriptsize{$\lambda_{42}=3.24$}}(SCALE4)
+   edge [postdashed] node [below=0.7cm,text width=2cm, left=0.001cm]
+   {\scriptsize{unmodeled loading=$3$}}(SCALE1.210)
+   edge [post] node[above=0.01cm,text width=2cm]
+   {\scriptsize{$\lambda_{52}=4.32$}}(SCALE5)
+   edge [post] node[above=0.25cm,text width=2cm]
+   {\scriptsize{$\lambda_{62}=7.21$}}(SCALE6)
+   edge [sling,bend right=100] node[auto,swap]{1}(MATH.270);
+
+\node[error] (E1) at (7,6) {E1}
+   edge [post] node[auto,swap]{\small{1}}(SCALE1)
+   edge [errorsling, bend right=100]node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{11}=26.77$}} (E1.45);
+\node[error] (E2) at (7,4.5) {E2}
+   edge [post] node[auto,swap]{\small{1}} (SCALE2)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{22}=13.01$}}(E2.45);
+\node[error] (E3) at (7,3) {E3}
+   edge [post] node[auto,swap]{\small{1}}(SCALE3)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{33}=30.93$}}(E3.45);
+\node[error] (E4) at (7,1) {E4}
+   edge [post] node[auto,swap]{\small{1}}(SCALE4)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{44}=3.17$}}(E4.45);
+\node[error] (E5) at (7,-0.5) {E5}
+   edge [post] node[auto,swap]{\small{1}}(SCALE5)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{55}=8.82$}} (E5.45);
+\node[error] (E6) at (7,-2) {E6}
+   edge [post] node[auto,swap]{\small{1}}(SCALE6)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{66}=22.5$}}(E6.45);
+\end{tikzpicture}}
+\end{figure}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+\begin{figure}
+\tiny{\caption{The parameter violating measurement invariance is the unmodeled
+loading. Testing parameters are $\lambda_{11}$ and $\psi_{11}$.}}
+\label{fig:sim21res}
+<<sim21res, fig=TRUE, height=5, width=9, echo=FALSE>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+       subset = (parms == "extra" & 
+	 pars %in% c("Loading1", "Error1") 
+                 & diff %% 0.5 == 0),
+       type = "b", xlab = "Violation Magnitude (Unmodeled Loading)",
+       ylab = "Power", key = mykey, as.table = TRUE,
+       strip = function(..., which.given, factor.levels){
+	   if(which.given == 2){
+	 strip.default(which.given, factor.levels = parlabs[c(1, 3)], ...)
+	   } else {
+	       strip.default(which.given, factor.levels = factor.levels, ...)
+	   }
+       })
+@
+\end{figure}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+\begin{figure}
+\tiny{\caption{The parameter violating measurement invariance is the unmodeled
+loading. Testing parameters are $\sigma_{12}$ and $\mu_{11}$.}}
+\label{fig:sim22res}
+<<sim22res, fig=TRUE, height=5, width=9, echo=FALSE>>=
+trellis.par.set(theme = canonical.theme(color = FALSE))
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+       subset = (parms == "extra" & 
+	 pars %in% c("Covariance", "Intercept1") 
+                 & diff %% 0.5 == 0),
+       type = "b", xlab = "Violation Magnitude (Unmodeled Loading)",
+       ylab = "Power", ylim = c(0,1), key = mykey, as.table = TRUE,
+       strip = function(..., which.given, factor.levels){
+	   if(which.given == 2){
+	 strip.default(which.given, factor.levels = parlabs[c(2, 4)], ...)
+	   } else {
+	       strip.default(which.given, factor.levels = factor.levels, ...)
+	   }
+       })
+@
+\end{figure}
+\end{frame}
+
+\section{Conclusion}
 \section{Application}
+\begin{frame}[fragile]
+\frametitle{Input data and Fit model}
+\begin{enumerate}
+ \item Input data
+<<data,eval=FALSE>>=
+data("YouthGratitude", package = "psychotools")
+compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
+yg <- YouthGratitude[compcases, ]
+@ 
+\item Fit restricted model and full model
+<<modeest,eval=FALSE>>=
+restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+  data = yg, group = "agegroup", meanstructure = TRUE,
+  group.equal = "loadings")
 
+full <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+  data = yg, group = "agegroup", meanstructure = TRUE)
+@ 
+\end{enumerate}
+\end{frame}
+
+%--------------------------------------------------------------------------
+%TODO: reset counter for enumerate.
+\begin{frame}
+  \item fff
+\end{frame}
 \end{document}

### 2014-03-05 frontiers/IMPSpres.Rnw
update the application part.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@326 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage[english]{babel}
+\usepackage{enumitem}
 \usefonttheme{serif}
@@
 \renewcommand{\baselinestretch}{1.1}
+\newcounter{tmpc}
 
@@
                          "All Loadings", "All Errors", "All Intercepts")
+
+data("YouthGratitude", package = "psychotools")
+compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
+yg <- YouthGratitude[compcases, ]
+
+restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+  data = yg, group = "agegroup", meanstructure = TRUE,
+  group.equal = "loadings")
+
+set.seed(1090)
+wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "WDMo")
+
+if(file.exists("maxLMo.rda")){
+  load("maxLMo.rda")
+} else {
+  RNGkind(kind = "default", normal.kind = "default")
+  set.seed(1090)
+  maxLMo <- ordL2BB(yg$agegroup, nproc = 4)
+  save(maxLMo, file = "maxLMo.rda")
+}
+maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = maxLMo)
+
 @
@@
 \begin{frame}[fragile]
-\frametitle{Input data and Fit model}
+\frametitle{Procedures of Application}
 \begin{enumerate}
  \item Input data
+\asuivre
+\end{enumerate}
 <<data,eval=FALSE>>=
@@
 @
-\item Fit restricted model and full model
+
+\end{frame}
+
+%--------------------------------------------------------------------------
+\begin{frame}[fragile]
+\frametitle{Procedures of Application}
+\begin{enumerate}
+\item Fit your model
+\end{enumerate}
  <<modeest,eval=FALSE>>=
@@
   group.equal = "loadings")
+@
 
-full <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-  data = yg, group = "agegroup", meanstructure = TRUE)
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}[fragile]
+\frametitle{Procedures of Application}
+\begin{enumerate}[resume]
+  \item Test if M.I. is violated.
+\end{enumerate}
+<<sctest-ord2, eval=FALSE>>=
+wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "WDMo",  plot = TRUE, ylim = c(0, 3.2),
+  xlab = "Age group", ylab = "Weighted test statistics")
+maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "maxLMo",  plot = TRUE, ylim = c(0, 14),
+  xlab = "Age group", ylab = "LM statistics",
+  main = expression(maxLM[o]), boundary = list(col = 1, lty = 2))
 @
-\end{enumerate}
+
 \end{frame}
 
-%--------------------------------------------------------------------------
-%TODO: reset counter for enumerate.
 \begin{frame}
-  \item fff
+  \frametitle
+\begin{figure}
+\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic 
+  (left panel) and the $\max\mathit{LM}_o$ statistic (right panel).}
+\label{fig:ordres}
+<<sctest-ord-graphics, fig=TRUE, height=4.5, width=9, echo=FALSE, results=hide>>=
+
+par(mfrow = c(1, 2))
+set.seed(1090)
+sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = "WDMo", plot = TRUE, ylim = c(0, 3.2),
+  xlab = "Age group", ylab = "Weighted test statistics",
+  main = expression(WDM[o]), boundary = list(col = 1, lty = 2))
+sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
+  functional = maxLMo, plot = TRUE, ylim = c(0, 14),
+  xlab = "Age group", ylab = "LM statistics",
+  main = expression(maxLM[o]), boundary = list(col = 1, lty = 2))
+@
+\end{figure}
 \end{frame}

### 2014-03-06 frontiers/IMPSpres.Rnw
update application part. need to work on theory part.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@327 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage[english]{babel}
-\usepackage{enumitem}
 \usefonttheme{serif}
@@
 \renewcommand{\baselinestretch}{1.1}
-\newcounter{tmpc}
+\newcounter{sauvegardeenumi}
+\newcommand{\asuivre}{\setcounter{sauvegardeenumi}{\theenumi}}
+\newcommand{\suite}{\setcounter{enumi}{\thesauvegardeenumi}}
 
@@
 \frametitle{Procedures of Application}
-\begin{enumerate}
+\begin{enumerate}[1.]
  \item Input data
-\asuivre
 \end{enumerate}
@@
 \frametitle{Procedures of Application}
-\begin{enumerate}
+\begin{enumerate}[2.]
+  \suite
 \item Fit your model
@@
  <<modeest,eval=FALSE>>=
+ library(lavaan)
   restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
@@
 \frametitle{Procedures of Application}
-\begin{enumerate}[resume]
+\begin{enumerate}[3.]
+  \suite
   \item Test if M.I. is violated.
@@
 <<sctest-ord2, eval=FALSE>>=
+ library(strucchange)
 wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
@@
   xlab = "Age group", ylab = "Weighted test statistics")
-maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "maxLMo",  plot = TRUE, ylim = c(0, 14),
-  xlab = "Age group", ylab = "LM statistics",
-  main = expression(maxLM[o]), boundary = list(col = 1, lty = 2))
 @ 
-
 \end{frame}
+%--------------------------------------------------------------------------
 
 \begin{frame}
-  \frametitle
 \begin{figure}
-\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic 
-  (left panel) and the $\max\mathit{LM}_o$ statistic (right panel).}
+\tiny{\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic}}
 \label{fig:ordres}
-<<sctest-ord-graphics, fig=TRUE, height=4.5, width=9, echo=FALSE, results=hide>>=
-
-par(mfrow = c(1, 2))
-set.seed(1090)
+<<sctest-ord-graphics,fig=TRUE, height=3.3, width=5, echo=FALSE, results=hide>>=
 sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
@@
   main = expression(WDM[o]), boundary = list(col = 1, lty = 2))
-sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = maxLMo, plot = TRUE, ylim = c(0, 14),
-  xlab = "Age group", ylab = "LM statistics",
-  main = expression(maxLM[o]), boundary = list(col = 1, lty = 2))
 @
@@
 \end{frame}
+
 \end{document}

### 2014-03-13 frontiers/IMPSpres.Rnw
update theory part, need to work on minor details.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@328 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \frametitle{Problem Scenario}
+\begin{itemize}
+  \item
+ \emph{Data:} The data consist of responses to a gratitude scale from $n=1401$ youth 
+ aged 10--19 years, dividing into six age groups: 10-11, 12-13, 14,
+ 15, 16, 17-19. 
+  \item
+\emph{Question:} The researcher is interested in whether the scales 
+were measurement invariant across these age groups.  
+\end{itemize}
 
@@
 %--------------------------------------------------------------------------
-
+\begin{frame}
+  \frametitle{Similar Problem Scenarios and Generalization}
+  \begin{itemize}
+    \item Whether model parameter holds for people with different
+      income level, grade level, motivation level, etc.  
+    \item In general, the problem is whether measurement invariance
+      hypothesis holds with an auxilliary ordinal variable. 
+    \end{itemize}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+\frametitle{Possible Solution}
+\begin{itemize}
+\item Likelihood Ratio Test(LRT) for full model and reduced model(s).
+  \begin{itemize}
+    \item Ignore the nature of ordinal variable, treat it as
+      categorical varaible. 
+    \item Sensitive to sample size. Rejecte the reduced model(s) too often.
+    \item Rather tedious and time consuming to fit and compare all possible
+      models. e.g, for testing one parameter, you need to fit $*2^{m-1}$ models,
+      $m$ is the level of ordinal variable.
+    \end{itemize}
+\item Score-based Tests w.r.t. ordinal variable. 
+\end{itemize}
+  \end{frame}
 \begin{frame}
@@
 \section{Theoretical Background}
+\begin{frame}
+  \frametitle{Score for Each Observation}
+\begin{equation}
+\label{eq:score}
+s(\theta;x_{i}) = \left (
+  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{1}}, \dots,
+  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right
+  )^\top,\ i=1,\ldots,n,
+\end{equation} 
+\scriptsize{where $\ell(\theta; x_i)$ is the likelihood associated with individual
+$i$ and $\theta$ is a $k$-dimensional parameter vector. The corresponding
+maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
+$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$.}  
+\end{frame}
+%--------------------------------------------------------------------------
+
+\begin{frame}
+  \frametitle{Decorrelate and Order Cumulative Score}
+To formalize these ideas, we define a suitably scaled cumulative sum of the 
+ordered scores.  This can be written as
+\begin{equation} 
+    \label{eq:cumscore}
+  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
+    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
+  \qquad (0 \le t \le 1)
+\end{equation}
+\scriptsize{where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
+nt \rfloor$ is the integer part of $nt$, and $x_{(i)}$ reflects the individual 
+with the $i$-thsmallest value of the auxiliary variable $V$.} 
+  \end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{}
+  \begin{itemize}
+   \item  Under the hypothesis of measurement invariance, a central limit theorem can be
+   used to show that the fluctuation of the above cumulative sum follows a
+   Brownian bridge.
+   \item This result allows us to calculate $p$-values and
+   critical values for test statistics under the hypothesis of
+   measurement invariance.
+  \end{itemize}
+ \end{frame}
+%--------------------------------------------------------------------------  
+ \begin{frame}
+   \frametitle{Statistics Based on Decorrelated and Ordered Socre}
+   Multiple test statistics are available, depending on how one
+summarizes the behavior of the cumulative sum of scores. Below shows
+ two statistics for ordinal auxiliary variable with $m$ levels,
+
+\scriptsize{\begin{eqnarray}
+    \label{eq:wdm}
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
+                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |\\
+    \label{eq:maxlmo}
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2
+\end{eqnarray}}
+where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$ $(\ell=1,\ldots,m-1)$.
+   \end{frame}
+%--------------------------------------------------------------------------  
+\begin{frame}
+  \frametitle{Visual Explanation of the Score-based Test for ordinal
+    auxillary varaible.}
+$$\bordermatrix{ \text{$B$}& \theta_{1} & \theta_{2} & \ldots &\theta_{k}\cr
+                x_{1}&  1.32  & -4.32 & \ldots & -2.34\cr
+                x_{1}&  3.37 & 8.32 & \ldots & \textcolor{red}{63.45}\cr
+                x_{..}& \vdots & \vdots & \vdots & \vdots\cr
+                x_{m}&  0 & 0 & \ldots & 0\cr
+                x_{m}& 0     &   0     & \ldots & 0}$$
+
+  \end{frame}
+
 \section{Simulation}
@@
 
+%--------------------------------------------------------------------------
 \section{Conclusion}
+
+\begin{frame}
+  \frametitle{Conclusion based on Simulation Studies}
+ \begin{itemize}
+   \item
+      The ordinal test statistic proposed have relatively-high power for detecting measurement 
+      invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable. 
+   \item This specificity feature is robust as long as the violation parameter is 
+     included in the testing model.
+   \item The use of these tests is convenient without changing the focal psychometric model.
+   \end{itemize}
+\end{frame}
+%--------------------------------------------------------------------------
 \section{Application}
@@
   \suite
-\item Fit your model
+\item Fit your model:no need to comparing various models.
 \end{enumerate}

### 2014-03-14 frontiers/IMPSpres.Rnw
update presentation. almost finished. need to practice orally.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@329 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2014-03-06 frontiers/IMPSpres.Rnw
@@
 \frametitle{Problem Scenario}
+\begin{itemize}
+  \item
+ \emph{Data:} \\
+ The data consist of responses to a gratitude scale from $n=1401$ youth 
+ aged 10--19 years, dividing into six age groups: 10-11, 12-13, 14,
+ 15, 16, 17-19. 
+  \item
+\emph{Question:} \\
+The researcher is interested in whether the scale
+is measurement invariant across these age groups.  
+\end{itemize}
 
@@
 %--------------------------------------------------------------------------
-
+\begin{frame}
+  \frametitle{Similar Problem Scenarios and Generalization}
+  \begin{itemize}
+    \item Whether model parameter(s) hold(s) for people with different
+      income level, grade level, motivation level, etc.  
+    \item In general, the problem is whether measurement invariance(M.I.)
+      hypothesis holds with respect to an auxiliary ordinal variable. 
+    \end{itemize}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+\frametitle{Possible Solution}
+\begin{itemize}
+\item Likelihood Ratio Test(LRT) for full model and reduced model(s).
+  \begin{itemize}
+    \item Ignore the nature of ordinal variable, treat it as
+      categorical varaible. 
+    \item Sensitive to sample size. Rejecte the reduced model(s) too often.
+    \item Rather tedious and time consuming to fit and compare all possible
+      models. e.g, for testing one parameter, you need to fit $2^{m-1}$ 
+      models, $m$ is the level of ordinal variable.
+    \end{itemize}
+\item Score-based Tests w.r.t. ordinal variable. 
+\end{itemize}
+  \end{frame}
 \begin{frame}
@@
 \section{Theoretical Background}
+\begin{frame}
+  \frametitle{Score for Each Observation}
+\begin{equation}
+\label{eq:score}
+s(\theta;x_{i}) = \left (
+  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{1}}, \dots,
+  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right
+  ),\ i=1,\ldots,n,
+\end{equation}
+
+\vspace{0.3in}
+where $\ell(\theta; x_i)$ is the likelihood associated with individual
+$i$ and $\theta$ is a $k$-dimensional parameter vector. The corresponding
+maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
+$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$. 
+\end{frame}
+%--------------------------------------------------------------------------
+
+\begin{frame}
+  \frametitle{Cumulative Sum Score: Decorrelate and Order}
+\begin{equation} 
+    \label{eq:cumscore}
+  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
+    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
+  \qquad (0 \le t \le 1)
+\end{equation}
+
+\vspace{0.3in}
+where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
+nt \rfloor$ is the integer part of $nt$, $t$ is the cumulative proportion 
+of $n$, and $x_{(i)}$ reflects the 
+individual with the $i$-th smallest value of the auxiliary variable $V$. 
+  \end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{Convergence Distribution}
+  \begin{itemize}
+   \item  Under the hypothesis of measurement invariance, a central limit 
+     theorem can be
+   used to show that the fluctuation of the above cumulative sum score
+   matrix ${\bm B}(\cdot; \hat {\bm \theta})$ converge in distribution to a 
+   k-dimensional Brownian bridge ${\bm B}^{0}(\cdot)$.
+   \begin{equation}
+     \label{eq:converge}
+    {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}
+        {\rightarrow}~ {\bm B}^{0}(\cdot),
+   \end{equation}
+   \item This result allows us to calculate $p$-values and
+   critical values for test statistics under the hypothesis of
+   measurement invariance.
+  \end{itemize}
+ \end{frame}
+%--------------------------------------------------------------------------  
+ \begin{frame}
+   \frametitle{Statistics Based on Cumulative Sum Socre}
+   Multiple test statistics are available, depending on how one
+summarizes the behavior of the cumulative sum of scores. Below shows
+ two statistics for ordinal auxiliary variable with $m$ levels,
+
+\scriptsize{\begin{eqnarray}
+    \label{eq:wdm}
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
+                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |\\
+    \label{eq:maxlmo}
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+        \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+          \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2
+\end{eqnarray}}
+where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$,
+$(t_{ell}$ is the cumulative proportions associated with level $\ell$, 
+($\ell=1,\ldots,m-1)$.
+   \end{frame}
+%-------------------------------------------------------------------------- 
+\begin{frame}
+\frametitle{What If the Auxiliary Variable is Indeed Categorical}
+\begin{equation}
+    \label{eq:lmuo}
+     \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+    \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}
+      (\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
+\end{equation}
+
+\vspace{0.3in}
+This statistic is asymptotically equivalent to the likelihood
+ratio test statistic but only requires estimation of only one model 
+(the reduced model).
+ \end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{Visual Explanation of the Score-based Test w.r.t. ordinal
+    auxiliary variable.}
+$$\bordermatrix{\text{}& \theta_{1} & \theta_{2} & \ldots &\theta_{k}\cr
+                x_{1}&  1.32  & -4.32 & \ldots & -2.34\cr
+                x_{2}&  3.37 & 8.32 & \ldots & \textcolor{red}{63.45}\cr
+                x_{..}& \vdots & \vdots & \vdots & \vdots\cr
+                x_{m-1}&  12.34 & 13.32 & \ldots & 6.89\cr
+                x_{m}& 0     &   0     & \ldots & 0} $$
+
+  \end{frame}
+
 \section{Simulation}
-\subsection{Violation Parameter is IN the Model}
+\subsection{M.I. Violation Parameter is IN the Model}
 \begin{frame}
@@
 \end{frame}
-
+%--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{Simulation Setup}
+  Power and Type I error were examined across the following conditions.
+ \begin{itemize}
+   \item Three sample sizes ($n=120,\textcolor{red}{480},960$), 
+   \item Three numbers of categories ($m=4,8,12$), 
+   \item Violation magnitude: $d$ times the parameters' asymptotic 
+     standard errors (scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$ 
+   \item Violation Level: began at level $1+m/2$ of the auxiliary variable 
+         $V$
+ \end{itemize}
+   Statistics $\mathit{WDM}_o$, $\max \mathit{LM}_o$, $\mathit{LM}_{uo}$ were 
+   examined.
+\end{frame}
+%--------------------------------------------------------------------------
 \begin{frame}
@@
 \tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
-    parmaeter is $\psi_{11} \ldots \psi_{66}, \sigma_{12}$ and
+    parameters are $\psi_{11} \ldots \psi_{66}, \sigma_{12}$ and
     $\mu_{11} \ldots \mu_{66}$.}}
@@
 
-\subsection{Violation Parameter is OUT the Model}
+\subsection{M.I. Violation Parameter is OUT the Model}
 \begin{frame}
@@
 
+%--------------------------------------------------------------------------
 \section{Conclusion}
+
+\begin{frame}
+  \frametitle{Conclusion based on Simulation Studies}
+ \begin{itemize}
+   \item
+      The ordinal test statistic proposed have relatively-high power 
+      for detecting measurement 
+      invariance violation parameter specifically that are monotonic 
+      with the auxiliary ordinal variable. 
+   \item This specificity feature is robust as long as the violation 
+     parameter is included in the testing model.
+   \item The use of these tests is convenient without changing the 
+     focal psychometric model.
+   \end{itemize}
+\end{frame}
+%--------------------------------------------------------------------------
 \section{Application}

### 2014-03-19 frontiers/IMPSpres.Rnw
update the Rnw file.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@330 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{frame}
-  \frametitle{Cumulative Sum Score: Decorrelate and Order}
+  \frametitle{Empirical Cumulative Sum Score: Decorrelate and Order}
 \begin{equation} 
@@
      theorem can be
-   used to show that the fluctuation of the above cumulative sum score
+   used to show that the fluctuation of the above empirical cumulative sum 
+   score
    matrix ${\bm B}(\cdot; \hat {\bm \theta})$ converge in distribution to a 
@@
  \begin{frame}
-   \frametitle{Statistics Based on Cumulative Sum Socre}
+   \frametitle{Statistics Based on Empirical Cumulative Sum Socre}
    Multiple test statistics are available, depending on how one
-summarizes the behavior of the cumulative sum of scores. Below shows
- two statistics for ordinal auxiliary variable with $m$ levels,
+summarizes the behavior of the empirical cumulative sum of scores. Below 
+shows two statistics for ordinal auxiliary variable with $m$ levels,
 
@@
 where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$,
-$(t_{ell}$ is the cumulative proportions associated with level $\ell$, 
+$t_{\ell}$ is the cumulative proportions associated with level $\ell$, 
 ($\ell=1,\ldots,m-1)$.

### 2014-03-24 frontiers/frontiers.Rnw
set up presentation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@322 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor changes to frontiers manuscript.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@332 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 individual has $k$ scores describing the extent to which the fitted
-model describes that particular individual.  We can roughly interpret
+model describes that particular individual.  
+These scores are similar to residuals \cite<and, in fact, the tests
+can be applied directly to residuals in other contexts; see>{zeihor07}:
+we can roughly interpret
 scores near zero as providing a ``good'' description of an individual,
@@
 practice, however, one should generally adjust the significance level
-for the number of tests carried out.
+for the number of tests carried out.  Additionally, because
+measurement invariance researchers often have large sample sizes,
+cross-validation methods can be useful to help verify the test
+results.
 
@@
 @
-%
+
 Finally, we can get the results of a likelihood ratio test via the \code{anova()}
@@
 inference would then proceed in exactly the same manner as the calculation and
-inference illustrated in this paper.
+inference illustrated in this paper.  Study of the proposed tests'
+application to larger SEMs is warranted.
 
 In addition to alternative fit functions, the tests can be extended to other
-models estimated via ML.  Of primary interest to the topic of measurement
+models estimated via ML.  Of primary relevance to the topic of measurement
 invariance, the tests can be extended to item response models to examine

### 2014-03-29 frontiers/IMPSpres.Rnw
update score part.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@331 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
update presentation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@333 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{frame}
-\frametitle{Problem Scenario}
+\frametitle{Outline} 
+% Table of contents slide, comment this block out to remove it
+\tableofcontents 
+% Throughout your presentation, if you choose to use \section{} and 
+%\subsection{} commands, these will automatically be printed on this slide 
+%as an overview of your presentation
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+\frametitle{Measurement Invariance}
   \begin{itemize}
-  \item
- \emph{Data:} \\
- The data consist of responses to a gratitude scale from $n=1401$ youth 
- aged 10--19 years, dividing into six age groups: 10-11, 12-13, 14,
- 15, 16, 17-19. 
-  \item
-\emph{Question:} \\
-The researcher is interested in whether the scale
-is measurement invariant across these age groups.  
+    \item Measurement invariance: Sets of tests/items consistently
+      assigning scores across diverse groups of individuals.\\ \ \\
+    \item Notable examination of measurement invariance:
+      \begin{itemize}
+         \item SAT for different ethnic groups (Atkinson, 2001)
+         \item Gratitude for adolescents and adults (Froh JJ et al 2011)     
       \end{itemize}
+  \end{itemize}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{Example: Grade 1, 2, 3 vs. Grade 4, 5, 6}
+ \begin{figure}
+    \label{fig:pathinmodeltheory}
+\resizebox{8.0cm}{!}{
+\begin{tikzpicture}
+[latent/.style={circle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt, minimum width=20mm, minimum height=10mm},
+ manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt,minimum width=20mm, minimum height=10mm},
+ error/.style={circle,draw=black!50,fill=white!20,thick,
+                inner sep=0pt, minimum width=10mm},
+ post/.style={->,shorten >=1pt,>=stealth',semithick},
+ postred/.style={->,shorten >=1pt,>=stealth',semithick,red},
+ postdashed/.style={->,shorten >=1pt,>=stealth',semithick,red,dashed},
+ sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
+   auto,distance=20mm,semithick},
+ errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
+   auto,distance=10mm,semithick}]
+
+%position objects.
+
+\node[latent] (VERBAL) at (0,4) {VERBAL};
+\node[latent] (MATH)   at (0,0) {MATH};
+
+\node[manifest] (SCALE1) at(5,6) {SCALE1};
+\node[manifest] (SCALE2) at (5,4.5) {SCALE2};
+\node[manifest] (SCALE3) at (5,3) {SCALE3};
+\node[manifest] (SCALE4) at(5,1) {SCALE4};
+\node[manifest] (SCALE5) at (5,-0.5) {SCALE5};
+\node[manifest] (SCALE6) at (5,-2) {SCALE6};
+
+\node[error] (E1) at (7,6) {E1};
+\node[error] (E2) at (7,4.5) {E2};
+\node[error] (E3) at (7,3) {E3};
+\node[error] (E4) at (7,1) {E4};
+\node[error] (E5) at (7,-0.5) {E5};
+\node[error] (E6) at (7,-2) {E6};
+
+%arrows, slings and texts
+\node[latent] (VERBAL) at (0,4) {VERBAL}
+   edge [postred] node[above=0.01cm,text width=4cm]
+   {\scriptsize{$\lambda_{11}=4.92/7.92$}} (SCALE1)
+   edge [post] node[above=0.001mm,text width=2cm]
+   {\scriptsize{$\lambda_{21}=2.96$}} (SCALE2)
+   edge [post] node[above=0.1cm,text width=2cm]
+   {\scriptsize{$\lambda_{31}=5.96$}} (SCALE3)
+   edge [sling,bend right=100] node[auto,swap]{1}(VERBAL.-180)
+   edge [sling,bend right=70] node[auto,swap]
+   {$\phi_{12}=-0.48$}(MATH)  ;
+
+\node[latent] (MATH) at (0,0) {MATH}
+   edge [post] node[above=0.01cm,text width=2cm]
+   {\scriptsize{$\lambda_{42}=3.24$}}(SCALE4)
+   edge [post] node[above=0.01cm,text width=2cm]
+   {\scriptsize{$\lambda_{52}=4.32$}}(SCALE5)
+   edge [post] node[above=0.25cm,text width=2cm]
+   {\scriptsize{$\lambda_{62}=7.21$}}(SCALE6)
+   edge [sling,bend right=100] node[auto,swap]{1}(MATH.270);
+
+\node[error] (E1) at (7,6) {E1}
+   edge [post] node[auto,swap]{\small{1}}(SCALE1)
+   edge [errorsling, bend right=100]node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{11}=26.77$}} (E1.45);
+\node[error] (E2) at (7,4.5) {E2}
+   edge [post] node[auto,swap]{\small{1}} (SCALE2)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{22}=13.01$}}(E2.45);
+\node[error] (E3) at (7,3) {E3}
+   edge [post] node[auto,swap]{\small{1}}(SCALE3)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{33}=30.93$}}(E3.45);
+\node[error] (E4) at (7,1) {E4}
+   edge [post] node[auto,swap]{\small{1}}(SCALE4)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{44}=3.17$}}(E4.45);
+\node[error] (E5) at (7,-0.5) {E5}
+   edge [post] node[auto,swap]{\small{1}}(SCALE5)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{55}=8.82$}} (E5.45);
+\node[error] (E6) at (7,-2) {E6}
+   edge [post] node[auto,swap]{\small{1}}(SCALE6)
+   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
+   {\scriptsize{$\psi_{66}=22.5$}}(E6.45);
+\end{tikzpicture}}
+\end{figure}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{Hypotheses}
+ Hypothesis of  measurement invariance and alternative:
+      \begin{center}
+          $H_0: \bm{\theta}_i = \bm{\theta}_0$, \\
+          $H_1: \bm{\theta}_i = \bm{\theta}_{(v_{i})}$\\
+      \end{center}
+
+      $\bm{\theta}_i = (\lambda_{i, 11}, \dots, \psi_{i, 11}, 
+      \dots, \varphi_{i, 12})^\top$
+      is the full $k$-dimensional parameter vector for individual
+      $i$, $i=1,\ldots,n$.
+
+      \vspace{0.3in}
+      $V$ can be continuous, \textcolor{red}{ordinal} or 
+      categorical variable.
 
@@
 %--------------------------------------------------------------------------
-\begin{frame}
-  \frametitle{Similar Problem Scenarios and Generalization}
-  \begin{itemize}
-    \item Whether model parameter(s) hold(s) for people with different
-      income level, grade level, motivation level, etc.  
-    \item In general, the problem is whether measurement invariance(M.I.)
-      hypothesis holds with respect to an auxiliary ordinal variable. 
-    \end{itemize}
-\end{frame}
-%--------------------------------------------------------------------------
+%\begin{frame}
+%\frametitle{Problem Scenario}
+%\begin{itemize}
+%  \item
+% \emph{Data:} \\
+% The data consist of responses to a gratitude scale from $n=1401$ youth 
+% aged 10--19 years, dividing into six age groups: 10-11, 12-13, 14,
+% 15, 16, 17-19. 
+%  \item
+%\emph{Question:} \\
+%The researcher is interested in whether the scale
+%is measurement invariant across these age groups.  
+%\end{itemize}
+%
+%\end{frame}
+%%--------------------------------------------------------------------------
+%\begin{frame}
+%  \frametitle{Similar Problem Scenarios and Generalization}
+%  \begin{itemize}
+%    \item Whether model parameter(s) hold(s) for people with different
+%      income level, grade level, motivation level, etc.  
+%    \item In general, the problem is whether measurement invariance(M.I.)
+%      hypothesis holds with respect to an auxiliary ordinal variable. 
+%    \end{itemize}
+%\end{frame}
+%%--------------------------------------------------------------------------
 \begin{frame}
@@
 \end{frame}
-\begin{frame}
-\frametitle{Overview} 
-% Table of contents slide, comment this block out to remove it
-\tableofcontents 
-% Throughout your presentation, if you choose to use \section{} and 
-%\subsection{} commands, these will automatically be printed on this slide 
-%as an overview of your presentation
-\end{frame}
-%--------------------------------------------------------------------------
+
+%  --------------------------------------------------------------------------
 \section{Theoretical Background}
 \begin{frame}
-  \frametitle{Score for Each Observation}
-\begin{equation}
-\label{eq:score}
-s(\theta;x_{i}) = \left (
-  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{1}}, \dots,
-  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right
-  ),\ i=1,\ldots,n,
-\end{equation}
-
-\vspace{0.3in}
-where $\ell(\theta; x_i)$ is the likelihood associated with individual
-$i$ and $\theta$ is a $k$-dimensional parameter vector. The corresponding
-maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
-$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$. 
+  \frametitle{Score-based Tests}
+  \begin{itemize}
+    \item The score-based tests (Edgar Merkle and Achim Zeileis, 2013) 
+      rely on first derivatives of
+      the model's log-likelihood function.
+    \item We consider individual terms ({\em scores}) of the gradient.  
+      These scores tell us how well a particular parameter describes a 
+      particular individual.
+  % Equation of gradient vs scores for individuals
+  \end{itemize}
+\begin{eqnarray*}
+  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
+  {\bm 0},\text{ where} \\
+ \\
+  s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} 
+  \log
+\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
+\end{eqnarray*}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{Empirical Cumulative Sum Score: Decorrelate and Order}
+  \frametitle{Empirical Cumulative Sum Score: Decorrelate and
+    Systematic Order}
+ We need a way to aggregate scores across people so that we
+      can draw some general conclusions.
+      \begin{itemize}
+        \item Decorrelate gradient correlations.
+        \item Order individuals by an auxiliary variable. 
+      \end{itemize}
 \begin{equation} 
@@
 
-\vspace{0.3in}
+\vspace{0.2in}
 where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
 nt \rfloor$ is the integer part of $nt$, $t$ is the cumulative proportion 
-of $n$, and $x_{(i)}$ reflects the 
-individual with the $i$-th smallest value of the auxiliary variable $V$. 
+of $n$, $t \in (1/n, n)$.
   \end{frame}
@@
    Multiple test statistics are available, depending on how one
-summarizes the behavior of the empirical cumulative sum of scores. Below 
-shows two statistics for ordinal auxiliary variable with $m$ levels,
+summarizes the behavior of the empirical cumulative sum of scores. 
+\begin{itemize}
+\item Statistics for ordianl variable: \\
+Weighted Double Maximum ordinal ($\mathit{WDM}_o$),\\
+maximum Lagrange multiplier ordinal ($\max \mathit{LM}_o$). \\
+\item Statistics for categorical variable:\\
+$\mathit{LM}_{uo}$ \\
+\end{itemize}
+\end{frame}
+%\scriptsize{\begin{eqnarray}
+%    \label{eq:wdm}
+%      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+%      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
+%                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |\\
+%    \label{eq:maxlmo}
+%      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+%        \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+%          \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2
+%\end{eqnarray}}
+%where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$,
+%$t_{\ell}$ is the cumulative proportions associated with level $\ell$, 
+%($\ell=1,\ldots,m-1)$.
 
-\scriptsize{\begin{eqnarray}
-    \label{eq:wdm}
-      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
-      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
-                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |\\
-    \label{eq:maxlmo}
-      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
-        \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-          \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2
-\end{eqnarray}}
-where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$,
-$t_{\ell}$ is the cumulative proportions associated with level $\ell$, 
-($\ell=1,\ldots,m-1)$.
-   \end{frame}
 %-------------------------------------------------------------------------- 
-\begin{frame}
-\frametitle{What If the Auxiliary Variable is Indeed Categorical}
-\begin{equation}
-    \label{eq:lmuo}
-     \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
-    \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}
-      (\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
-\end{equation}
-
-\vspace{0.3in}
-This statistic is asymptotically equivalent to the likelihood
-ratio test statistic but only requires estimation of only one model 
-(the reduced model).
- \end{frame}
+%begin{frame}
+%frametitle{What If the Auxiliary Variable is Indeed Categorical}
+%begin{equation}
+%   \label{eq:lmuo}
+%    \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+%   \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}
+%     (\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
+%end{equation}
+%
+%vspace{0.3in}
+%his statistic is asymptotically equivalent to the likelihood
+%atio test statistic but only requires estimation of only one model 
+%the reduced model).
+%\end{frame}
 %--------------------------------------------------------------------------
@@
   \end{frame}
-
+%-------------------------------------------------------------------------   
 \section{Simulation}
-\subsection{M.I. Violation Parameter is IN the Model}
+\begin{frame}
+  \frametitle{Simulation}
+  What is the tests' power of locating the M.I.violation parameter? 
+  \begin{itemize}
+   \item The M.I. violation parameter is included in the model(IN). 
+   \item The M.I. violation parameter is \emph{not} in the model(OUT). 
+  \end{itemize}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{Simulation Setup}
+ \begin{itemize}
+   \item Three sample sizes ($n=120,\textcolor{red}{480},960$), 
+   \item Three numbers of categories ($m=4,8,12$), 
+   \item Violation magnitude: $d$ times the parameters' asymptotic 
+     standard errors (scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$ 
+   \item Violation Level: began at level $1+m/2$ of the auxiliary variable 
+         $V$
+ \end{itemize}
+   Statistics $\mathit{WDM}_o$, $\max \mathit{LM}_o$, $\mathit{LM}_{uo}$ were 
+   examined.
+\end{frame}
+%--------------------------------------------------------------------------
 \begin{frame}
@@
 %--------------------------------------------------------------------------
-\begin{frame}
-  \frametitle{Simulation Setup}
-  Power and Type I error were examined across the following conditions.
- \begin{itemize}
-   \item Three sample sizes ($n=120,\textcolor{red}{480},960$), 
-   \item Three numbers of categories ($m=4,8,12$), 
-   \item Violation magnitude: $d$ times the parameters' asymptotic 
-     standard errors (scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$ 
-   \item Violation Level: began at level $1+m/2$ of the auxiliary variable 
-         $V$
- \end{itemize}
-   Statistics $\mathit{WDM}_o$, $\max \mathit{LM}_o$, $\mathit{LM}_{uo}$ were 
-   examined.
-\end{frame}
-%--------------------------------------------------------------------------
+
 \begin{frame}
@@
 
-\subsection{M.I. Violation Parameter is OUT the Model}
 \begin{frame}
@@
 \begin{figure}
-\tiny{\caption{The parameter violating measurement invariance is the unmodeled
-loading. Testing parameters are $\lambda_{11}$ and $\psi_{11}$.}}
+\tiny{\caption{The parameter violating measurement invariance is the 
+    unmodeled loading. Testing parameters are $\lambda_{11}$ and $\psi_{11}$.}}
 \label{fig:sim21res}
@@
 \begin{figure}
-\tiny{\caption{The parameter violating measurement invariance is the unmodeled
-loading. Testing parameters are $\sigma_{12}$ and $\mu_{11}$.}}
+\tiny{\caption{The parameter violating measurement invariance is 
+    the unmodeled loading. Testing parameters are $\sigma_{12}$ 
+    and $\mu_{11}$.}}
 \label{fig:sim22res}
@@
 \section{Application}
+\begin{frame}
+  \frametitle{Application: Software Preparation}
+The tests can be carried out in R. Specifically, 
+\begin{itemize}
+\item Model estimation can be carried out in \emph{lavaan} package. 
+\item The test statistic and plot is got from \emph{strucchange} package.
+\end{itemize}
+\end{frame}
+%--------------------------------------------------------------------------
 \begin{frame}[fragile]
@@
   xlab = "Age group", ylab = "Weighted test statistics",
-  main = expression(WDM[o]), boundary = list(col = 1, lty = 2))
+  main = expression(WDM[o]), boundary = list(col = "red", lty = 1))
 @
@@
 \end{frame}
-
+%--------------------------------------------------------------------------
+\begin{frame}
+ \begin{itemize}
+ \item Thanks! 
+ \item Questions?
+ \end{itemize}
+\end{frame}
+%--------------------------------------------------------------------------
 \end{document}

### 2014-04-23 frontiers/frontiers.Rnw
Final revisions to frontiers paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@334 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-Many of the papers in this special issue focus on the topic of approximate
+Some of the papers in this special issue focus on the topic of approximate
 measurement invariance: we know that strict hypotheses of measurement invariance
@@
 corresponding tests of measurement invariance.  Under a Bayesian
-approach, we may utilize the idea of approximate invariance
-\cite<e.g.,>{muthen13}, whereby across-group equality constraints on
-parameters are replaced with informative prior distributions.  In this paper, we
+approach, we may implement the idea of approximate invariance
+\cite<e.g.,>{muthen13} by replacing across-group equality constraints on
+parameters with informative prior distributions.  In this paper, we
 describe an alternative approach: the development of test statistics that 
@@
 on the output of a traditional factor model \cite<see>[for
-further comparison of these methods]{MerZei13}.
+further comparison of these methods]{MerZei13}.  These methods all
+assume that the estimated model is correctly specified, save
+possibly for differences in parameter values between individuals.
 
@@
 software considerations for carrying out the tests, and test
-performance under model misspecification.
+performance under model misspecification.  The latter issue is
+particularly important because, in practice, all models are
+misspecified.  Hence, practically-useful tests of measurement
+invariance should be robust to model misspecification.
 %In order to fix the psychometric scale, we need to specify the violating parameter.  
@@
 %of the tests' performance under practical research scenarios.
+
 In the following sections, we first briefly review the  theoretical framework of
@@
 We focus on measurement invariance of the factor loadings associated with 
-the GQ-6 scale, using a one-factor model.  This scale consists of five
-items scored from 1--9.  While the ``age group'' variable against 
-which we are testing measurement invariance is best considered ordinal, 
+one of the scales in the dataset, the GQ-6 scale \cite{mccemm02}.  This
+scale consists of five Likert scale items (there is a sixth item that
+is omitted from analyses, following Froh et al.) with seven points
+each.  We fit a one-factor model to these items, examining whether the
+factor model parameters are invariant with respect to age group.  
+While the age group variable is best considered ordinal, 
 for demonstration we consider its treatment as categorical, continuous, and ordinal.  

### 2014-04-25 frontiers/frontiers.Rnw
Frontiers paper: Modified code file and added code appendix to ms
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@338 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \proglang{R}~code for replication of our results is
-available at \url{http://semtools.R-Forge.R-project.org/}.
+available at \url{http://semtools.R-Forge.R-project.org/} and also in
+an online supplement to this article.
 
 \bibliography{refs}
+
+\section{Appendix}
+
+The accompanying zip file contains R code for replication of all
+analyses and simulations from the article.
+
+\begin{itemize}
+  \item mz-frontiers.R: Model estimation functions for simulations.
+  \item sim-frontiers.R: Functions for data generation, power
+    evaluation, and power summaries.
+  \item replication-frontiers.R: Code for the tutorial and
+    simulations, utilizing the other two files.
+\end{itemize}
 

### 2014-06-05 frontiers/IMPSpres-knitr.Rnw
update Rnw file chunk options format.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@340 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2014-03-29 frontiers/IMPSpres.Rnw
@@
     \caption{Model used for the simulation.}
-    \label{fig:pathinmodel}
+    \label{fig:pathinmodel1}
 \resizebox{8.0cm}{!}{
@@
 \label{fig:sim11res}
-<<sim11res, fig=TRUE, height=5, width=9, echo=FALSE>>=
+<<sim11res, fig.height=5, fig.width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 \label{fig:sim12res}
-<<sim12res, fig=TRUE, height=5, width=9, echo=FALSE>>=
+<<sim12res, fig.height=5, fig.width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
     \caption{Model used for the simulation.}
-    \label{fig:pathinmodel}
+    \label{fig:pathinmodel2}
 \resizebox{8.0cm}{!}{
@@
 \label{fig:sim21res}
-<<sim21res, fig=TRUE, height=5, width=9, echo=FALSE>>=
+<<sim21res, fig.height=5, fig.width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 \label{fig:sim22res}
-<<sim22res, fig=TRUE, height=5, width=9, echo=FALSE>>=
+<<sim22res, fig.height=5, fig.width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 \label{fig:ordres}
-<<sctest-ord-graphics,fig=TRUE, height=3.3, width=5, echo=FALSE, results=hide>>=
+<<sctest-ord-graphics, fig.height=3.3, fig.width=5, echo=FALSE, results='hide'>>=
 sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",

### 2014-06-06 frontiers/IMPSpres-knitr.Rnw
update presentation to score matrix.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@341 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage[english]{babel}
+\setbeamercolor{block title}{use=structure,fg=black,bg=blue!10!white}
+\setbeamercolor{block body}{use=structure,fg=black,bg=blue!05!white}
 \usefonttheme{serif}
@@
 
-\begin{frame}
-\frametitle{Outline} 
-% Table of contents slide, comment this block out to remove it
-\tableofcontents 
-% Throughout your presentation, if you choose to use \section{} and 
-%\subsection{} commands, these will automatically be printed on this slide 
-%as an overview of your presentation
-\end{frame}
 %--------------------------------------------------------------------------
@@
       \begin{itemize}
-         \item SAT for different ethnic groups (Atkinson, 2001)
-         \item Gratitude for adolescents and adults (Froh JJ et al 2011)     
+         \item SAT for different ethnic groups (Atkinson, 2001, CITE)
+         \item Intelligence tests for different cohorts
+           (Jelte M. Wicherts, Conor V.Dolan, David J. Hessen CITE)     
       \end{itemize}
@@
  Hypothesis of  measurement invariance and alternative:
+  \begin{columns}
+    \begin{column}{0.48\textwidth}
+      \begin{block}{2 Groups}
+        \begin{align}
+          &H_0: \underset{i\in\left\{1:n\right\}}{\bm{\theta}_i} = 
+          \bm{\theta}_0,\\
+          &H_1:
+           \begin{cases}
+               \underset{i \in \left\{v_{1}\right\}}{\bm{\theta}_i} = 
+               \bm{\theta}_{v_{1}} \\
+               \underset{i \in \left\{v_{2}\right\}}{\bm{\theta}_i} = 
+               \bm{\theta}_{v_{2}}, 
+           \end{cases}
+        \end{align}
         \begin{center}
-          $H_0: \bm{\theta}_i = \bm{\theta}_0$, \\
-          $H_1: \bm{\theta}_i = \bm{\theta}_{(v_{i})}$\\
+         where $\left\{v_{1}\right\}+\left\{v_{2}\right\}=\left\{1:n\right\}$
         \end{center}
-
-      $\bm{\theta}_i = (\lambda_{i, 11}, \dots, \psi_{i, 11}, 
-      \dots, \varphi_{i, 12})^\top$
-      is the full $k$-dimensional parameter vector for individual
-      $i$, $i=1,\ldots,n$.
-
-      \vspace{0.3in}
-      $V$ can be continuous, \textcolor{red}{ordinal} or 
-      categorical variable.
+      \end{block}
+  \end{column}
+  \begin{column}{0.48\textwidth}
+    \begin{block}{G Groups}
+         \begin{align}
+          &H_0: \underset{i\in\left\{1:n\right\}}{\bm{\theta}_i} = 
+                \bm{\theta}_0,\\
+          &H_1: \underset{i \in \left\{v_{g}\right\}}{\bm{\theta}_i} = 
+                \bm{\theta}_{v_{g}}, 
+         \end{align}
+       \begin{center}
+        where $\sum_{g=1}^{G}\left\{v_{g}\right\}=\left\{1:n\right\}$
+       \end{center}   
+       $V_{g}$ can be categorical, \textit{ordinal} and continous.
+    \end{block}
+  \end{column}
+\end{columns}
 
@@
 \item Likelihood Ratio Test(LRT) for full model and reduced model(s).
-  \begin{itemize}
-    \item Ignore the nature of ordinal variable, treat it as
-      categorical varaible. 
-    \item Sensitive to sample size. Rejecte the reduced model(s) too often.
-    \item Rather tedious and time consuming to fit and compare all possible
-      models. e.g, for testing one parameter, you need to fit $2^{m-1}$ 
-      models, $m$ is the level of ordinal variable.
-    \end{itemize}
-\item Score-based Tests w.r.t. ordinal variable. 
+%  \begin{itemize}
+%    \item Ignore the nature of ordinal variable, treat it as
+%      categorical varaible. 
+%    \item Sensitive to sample size. Rejecte the reduced model(s) too often.
+%    \item Rather tedious and time consuming to fit and compare all possible
+%      models. e.g, for testing one parameter, you need to fit $2^{m-1}$ 
+%      models, $m$ is the level of ordinal variable.
+%    \end{itemize}
+\item Wald Test
+\item Lagrange Multiplier Test(Score Test)
+\item \textbf{Score-based Tests(more general) w.r.t. ordinal variable}. 
 \end{itemize}
 \end{frame}
-
+%  --------------------------------------------------------------------------
+\begin{frame}
+\frametitle{Outline} 
+% Table of contents slide, comment this block out to remove it
+\tableofcontents 
+% Throughout your presentation, if you choose to use \section{} and 
+%\subsection{} commands, these will automatically be printed on this slide 
+%as an overview of your presentation
+\end{frame}
 %  --------------------------------------------------------------------------
@@
   \begin{itemize}
-    \item The score-based tests (Edgar Merkle and Achim Zeileis, 2013) 
+    \item The score-based tests (CITE Edgar Merkle and Achim Zeileis, 2013) 
       rely on first derivatives of

### 2014-06-09 frontiers/IMPSpres-knitr.Rnw
update presentation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@342 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{beamerthemeSzeged,graphics,amsmath,bm,booktabs}
+\usepackage[notocbib]{apacite}
 \usepackage{tikz}
@@
       \begin{itemize}
-         \item SAT for different ethnic groups (Atkinson, 2001, CITE)
-         \item Intelligence tests for different cohorts
-           (Jelte M. Wicherts, Conor V.Dolan, David J. Hessen CITE)     
+         \item SAT for different ethnic groups \cite{atk01}
+         \item Intelligence tests for different cohorts. \cite{wich04}
       \end{itemize}
@@
   \begin{itemize}
-    \item The score-based tests (CITE Edgar Merkle and Achim Zeileis, 2013) 
+    \item The score-based tests \cite{MerZei13}
       rely on first derivatives of
@@
 \begin{frame}
+  \frametitle{Visual Explanation of the Score-based Test w.r.t. ordinal
+    auxiliary variable.}
+  \begin{columns}
+    \begin{column}{0.48\textwidth}
+      \begin{block}{Score}
+        \resizebox{5.0cm}{!}{
+        $$\bordermatrix{\text{}& \theta_{1} & \theta_{2} & \ldots 
+          &\textcolor{red}{\theta_{k}}\cr
+                x_{1}&  1.32  & -4.32 & \ldots & -2.34\cr
+                x_{2}&  3.37 & 8.32 & \ldots & 63.45\cr
+                x_{3}&  4.12 & 3.48 & \ldots & -3.11\cr
+                x_{..}& \vdots & \vdots & \vdots & \vdots\cr
+                x_{m-1}&  12.34 & 13.32 & \ldots & 6.89\cr
+                x_{m}& 0     &   0     & \ldots & 0} $$}
+        \end{block}
+     \end{column}
+     \begin{column}{0.48\textwidth}
+       \begin{block}{Cumulative score}
+         \resizebox{5.0cm}{!}{
+      $$\bordermatrix{\text{}& \theta_{1} & \theta_{2} & \ldots
+        &\textcolor{red}{\theta_{k}}\cr
+                x_{1}&  1.32  & -4.32 & \ldots & -2.34\cr
+                x_{2}&  4.69 & 4.00 & \ldots & 61.11\cr
+                x_{3}&  8.81 & 7.48 & \ldots & 58.00\cr
+                x_{..}& \vdots & \vdots & \vdots & \vdots\cr
+                x_{m-1}&  7.89 & 3.32 & \ldots & 4.91\cr
+                x_{m}& 0     &   0     & \ldots & 0} $$}
+        \end{block}
+      \end{column}
+   \end{columns}
+\end{frame}
+%--------------------------------------------------------------------------
+
+\begin{frame}
   \frametitle{Convergence Distribution}
@@
 \begin{itemize}
-\item Statistics for ordianl variable: \\
+\item Statistics for ordinal variable: \\
 Weighted Double Maximum ordinal ($\mathit{WDM}_o$),\\
@@
 \end{frame}
-%\scriptsize{\begin{eqnarray}
-%    \label{eq:wdm}
-%      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
-%      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
-%                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |\\
-%    \label{eq:maxlmo}
-%      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
-%        \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
-%          \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2
-%\end{eqnarray}}
-%where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$,
-%$t_{\ell}$ is the cumulative proportions associated with level $\ell$, 
-%($\ell=1,\ldots,m-1)$.
 
-%-------------------------------------------------------------------------- 
-%begin{frame}
-%frametitle{What If the Auxiliary Variable is Indeed Categorical}
-%begin{equation}
-%   \label{eq:lmuo}
-%    \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
-%   \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}
-%     (\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
-%end{equation}
 %
@@
 %\end{frame}
-%--------------------------------------------------------------------------
-\begin{frame}
-  \frametitle{Visual Explanation of the Score-based Test w.r.t. ordinal
-    auxiliary variable.}
-$$\bordermatrix{\text{}& \theta_{1} & \theta_{2} & \ldots &\theta_{k}\cr
-                x_{1}&  1.32  & -4.32 & \ldots & -2.34\cr
-                x_{2}&  3.37 & 8.32 & \ldots & \textcolor{red}{63.45}\cr
-                x_{..}& \vdots & \vdots & \vdots & \vdots\cr
-                x_{m-1}&  12.34 & 13.32 & \ldots & 6.89\cr
-                x_{m}& 0     &   0     & \ldots & 0} $$
 
-  \end{frame}
 %-------------------------------------------------------------------------   
@@
 \begin{frame}
-  \frametitle{Simulation}
-  What is the tests' power of locating the M.I.violation parameter? 
+  \frametitle{Study Motivation}
+  \begin{block}{Practice Consideration}
     \begin{itemize}
-   \item The M.I. violation parameter is included in the model(IN). 
-   \item The M.I. violation parameter is \emph{not} in the model(OUT). 
+      \item Model misspecification is inevitable in practice.  
+      \item M.I. is often studied with respect to \textbf{ordinal} variable.
+      \item Availability/Convenience of software implementation.
+    \end{itemize}
+  \end{block}
+\end{frame}
+%-------------------------------------------------------------------------   
+\begin{frame}
+  \frametitle{Simulation Motivation}
+  What is the tests' power under model misspecification? 
+  \begin{itemize}
+    \item The model missed out one parameter, which is \emph {not} the 
+      M.I. parameter.
+    \item The model missed out one parameter, which \emph{is} the M.I. 
+      parameter. 
   \end{itemize}
@@
   \begin{figure}
-    \caption{Model used for the simulation.}
+    \caption{Missed out parameter is not the M.I. parameter}
     \label{fig:pathinmodel1}
@@
  postred/.style={->,shorten >=1pt,>=stealth',semithick,red},
- postdashed/.style={->,shorten >=1pt,>=stealth',semithick,red,dashed},
+ postdashed/.style={->,shorten >=1pt,>=stealth',semithick,dashed},
  sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
@@
    {\scriptsize{$\lambda_{42}=3.24$}}(SCALE4)
+   edge [postdashed] node [below=0.7cm,text width=2cm, left=0.001cm]
+   {\scriptsize{unmodeled loading=$3$}}(SCALE1.210)
    edge [post] node[above=0.01cm,text width=2cm]
@@
     parameters are $\lambda_{11}$ and $\lambda_{11} \ldots \lambda_{62}$.}}
-\label{fig:sim11res}
-<<sim11res, fig.height=5, fig.width=9, echo=FALSE>>=
+\label{fig:sim21res}
+<<sim21res, fig.height=5, fig.width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "loading" & 
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+       subset = (parms == "extra+loading" & 
           pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
@@
     $\mu_{11} \ldots \mu_{66}$.}}
-\label{fig:sim12res}
-<<sim12res, fig.height=5, fig.width=9, echo=FALSE>>=
+\label{fig:sim22res}
+<<sim22res, fig.height=5, fig.width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
-mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
-xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
-       subset = (parms == "loading" & 
+mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
+xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim2,
+       subset = (parms == "extra+loading" & 
           pars %in% c("All Errors", "Covariance", "All Intercepts") 
@@
 \begin{figure}
-    \caption{Model used for the simulation.}
+    \caption{Missed out parameter is the M.I. parameter}
     \label{fig:pathinmodel2}
@@
     unmodeled loading. Testing parameters are $\lambda_{11}$ and $\psi_{11}$.}}
-\label{fig:sim21res}
-<<sim21res, fig.height=5, fig.width=9, echo=FALSE>>=
+\label{fig:sim23res}
+<<sim23res, fig.height=5, fig.width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
     and $\mu_{11}$.}}
-\label{fig:sim22res}
-<<sim22res, fig.height=5, fig.width=9, echo=FALSE>>=
+\label{fig:sim24res}
+<<sim24res, fig.height=5, fig.width=9, echo=FALSE>>=
 trellis.par.set(theme = canonical.theme(color = FALSE))
@@
 %--------------------------------------------------------------------------
-\section{Conclusion}
 
@@
       with the auxiliary ordinal variable. 
-   \item This specificity feature is robust as long as the violation 
+   \item This feature is robust as long as the violation 
       parameter is included in the testing model.
@@
 \end{enumerate}
+\begin{itemize}
+  \item Gratitude scale (GQ-6, 5 items) completed by six age
+    groups(10-11, 12-13, 14, 15, 16, 17-19) \cite{frofan11}
+\end{itemize}
 <<data,eval=FALSE>>=
 data("YouthGratitude", package = "psychotools")
-compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
+compcases <- apply(YouthGratitude[, 4:28], 1, 
+                   function(x) all(x %in% 1:9))
 yg <- YouthGratitude[compcases, ]
@@
   \suite
-\item Fit your model:no need to comparing various models.
+\item Fit your constrained model:no need to comparing various models.
 \end{enumerate}
@@
   restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-  data = yg, group = "agegroup", meanstructure = TRUE,
-  group.equal = "loadings")
+               data = yg, group = "agegroup", 
+               meanstructure = TRUE, group.equal = "loadings")
 @
@@
 \end{enumerate}
-<<sctest-ord2, eval=FALSE>>=
+<<sctest-ord2, eval=TRUE>>=
  library(strucchange)
-wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "WDMo",  plot = TRUE, ylim = c(0, 3.2),
-  xlab = "Age group", ylab = "Weighted test statistics")
+wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, 
+               vcov = "info", functional = "WDMo",  plot = FALSE, 
+               ylim = c(0, 3.2), xlab = "Age group", 
+               ylab = "Weighted test statistics")
+wdmo
 @ 
@@
 %--------------------------------------------------------------------------
+\begin{frame}[shrink=0.8]{References}
+\bibliographystyle{apacite}
+\bibliography{refs}
+\end{frame}
+%--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{Ordinal statistics equations}
+\scriptsize{\begin{eqnarray}
+    \label{eq:wdm}
+      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
+                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |\\
+    \label{eq:maxlmo}
+      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
+        \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+          \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2
+\end{eqnarray}}
+where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$,
+$t_{\ell}$ is the cumulative proportions associated with level $\ell$, 
+($\ell=1,\ldots,m-1)$.
+\end{frame}
+%-------------------------------------------------------------------------- 
+\begin{frame}
+\frametitle{Categorical statistics equation}
+\begin{equation}
+   \label{eq:lmuo}
+    \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+   \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}
+     (\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
+\end{equation}
+\end{frame}
 \end{document}

### 2014-06-09 frontiers/knit.R

@@ -1,3 +1,0 @@
-library("knitr")
-Sweave2knitr("IMPSpres.Rnw")
-knit("IMPSpres-knitr.Rnw")

### 2014-06-16 frontiers/IMPSpres-knitr.Rnw
update presentation according to Ed's comment and edit bib to fix captical letters.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@343 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \author[shortname]{Ting Wang\inst{1} \and Edgar C. Merkle\inst{1} \and 
-  Achim Zeileis \inst{2}}
+  Achim Zeileis\inst{2}}
 \institute[]{\inst{1} University of Missouri \and 
@@
 \renewcommand{\baselinestretch}{1.1}
+\def\bibfont{\footnotesize}
 \newcounter{sauvegardeenumi}
@@
         \end{align}
-        \begin{center}
-         where $\left\{v_{1}\right\}+\left\{v_{2}\right\}=\left\{1:n\right\}$
-        \end{center}
+       % \begin{center}
+       %  where $\left\{v_{1}\right\}+\left\{v_{2}\right\}=\left\{1:n\right\}$
+       % \end{center}
       \end{block}
@@
          \end{align}
-       \begin{center}
-        where $\sum_{g=1}^{G}\left\{v_{g}\right\}=\left\{1:n\right\}$
-       \end{center}   
+      % \begin{center}
+      %  where $\sum_{g=1}^{G}\left\{v_{g}\right\}=\left\{1:n\right\}$
+      % \end{center}   
        $V_{g}$ can be categorical, \textit{ordinal} and continous.
@@
 \begin{itemize}
-\item Likelihood Ratio Test(LRT) for full model and reduced model(s).
+\item Likelihood Ratio Test (LRT) (Require estimates under null
+  hypothesis and alternative).
 %  \begin{itemize}
@@
 %    \end{itemize}
-\item Wald Test
-\item Lagrange Multiplier Test(Score Test)
-\item \textbf{Score-based Tests(more general) w.r.t. ordinal variable}. 
+\item Wald Test (Require estimates under alternative). 
+\item Lagrange Multiplier Test (Score Test) (Require estimates under
+  null hypotheis)
+\item \textbf{Score-based Tests (more general) w.r.t. categorical,
+    ordinal, continuous variable}. 
 \end{itemize}
@@
   \begin{itemize}
-    \item The score-based tests \cite{MerZei13}
+    \item The score-based tests \cite{MerZei13, MerFanZei}
       rely on first derivatives of
@@
       \begin{itemize}
-        \item Decorrelate gradient correlations.
+        \item Decorrelate sores for each parameter.
         \item Order individuals by an auxiliary variable. 
@@
 \begin{frame}
-  \frametitle{Visual Explanation of the Score-based Test w.r.t. ordinal
-    auxiliary variable.}
+  \frametitle{Visual Explanation of the Score-based Test}
   \begin{columns}
     \begin{column}{0.48\textwidth}
-      \begin{block}{Score}
+      \begin{block}{Empirical Score}
         \resizebox{5.0cm}{!}{
@@
                 x_{..}& \vdots & \vdots & \vdots & \vdots\cr
-                x_{m-1}&  12.34 & 13.32 & \ldots & 6.89\cr
-                x_{m}& 0     &   0     & \ldots & 0} $$}
+                x_{n-1}&  12.34 & 13.32 & \ldots & 6.89\cr
+                x_{n}& 0     &   0     & \ldots & 0} $$}
         \end{block}
@@
      \begin{column}{0.48\textwidth}
-       \begin{block}{Cumulative score}
+       \begin{block}{Empirical Cumulative score}
          \resizebox{5.0cm}{!}{
@@
                 x_{..}& \vdots & \vdots & \vdots & \vdots\cr
-                x_{m-1}&  7.89 & 3.32 & \ldots & 4.91\cr
-                x_{m}& 0     &   0     & \ldots & 0} $$}
+                x_{n-1}&  7.89 & 3.32 & \ldots & 4.91\cr
+                x_{n}& 0     &   0     & \ldots & 0} $$}
         \end{block}
@@
    score
-   matrix ${\bm B}(\cdot; \hat {\bm \theta})$ converge in distribution to a 
+   matrix ${\bm B}(\cdot; \hat {\bm \theta})$ converges in distribution to a 
    k-dimensional Brownian bridge ${\bm B}^{0}(\cdot)$.
@@
    \frametitle{Statistics Based on Empirical Cumulative Sum Socre}
-   Multiple test statistics are available, depending on how one
-summarizes the behavior of the empirical cumulative sum of scores. 
 \begin{itemize}
-\item Statistics for ordinal variable: \\
+  \item Statistics for auxiliary categorical variable:\\
+maximum Lagrange multiplier categorical ($\mathit{LM}_{uo}$), 
+asymptotic equvialent to LRT.\\
+\item Statistics for auxiliary ordinal variable: \\
 Weighted Double Maximum ordinal ($\mathit{WDM}_o$),\\
-maximum Lagrange multiplier ordinal ($\max \mathit{LM}_o$). \\
-\item Statistics for categorical variable:\\
-$\mathit{LM}_{uo}$ \\
+maximum Lagrange multiplier ordinal ($\max\mathit{LM}_o$). \\
+\item Statistics for auxiliary continuous variable:\\
+Double maximumum statistic ($\mathit{DM}$)\\
+Cram{\'e}r-von Mises type statistic ($\mathit{CvM}$)\\
+maximum Lagrange multiplier ($\max\mathit{LM}$)\\
 \end{itemize}
@@
   \frametitle{Simulation Motivation}
-  What is the tests' power under model misspecification? 
+  \begin{block}{Simulation:}
+    What is the ordinal tests' power under model misspecification? 
     \begin{itemize}
-    \item The model missed out one parameter, which is \emph {not} the 
-      M.I. parameter.
-    \item The model missed out one parameter, which \emph{is} the M.I. 
-      parameter. 
+      \item The estimated model excludes one parameter, which is \emph
+        {not} the problematic parameter.
+      \item The estimated model excludes one parameter, which \emph{is}
+        the problematic parameter. 
     \end{itemize}
+  \end{block}
 \end{frame}
@@
      standard errors (scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$ 
-   \item Violation Level: began at level $1+m/2$ of the auxiliary variable 
-         $V$
+   \item Ordinal Violation: began at level $1+m/2$ of the auxiliary 
+     variable $V$.
  \end{itemize}
@@
 %--------------------------------------------------------------------------
+
 \begin{frame}
@@
 \tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
-    parameters are $\psi_{11} \ldots \psi_{66}, \sigma_{12}$ and
+    parameters are $\phi_{12}$, $\psi_{11} \ldots \psi_{66}$, and
     $\mu_{11} \ldots \mu_{66}$.}}
@@
 \begin{figure}
-    \caption{Missed out parameter is the M.I. parameter}
+    \caption{Excluded parameter is the problematic parameter}
     \label{fig:pathinmodel2}
@@
 \tiny{\caption{The parameter violating measurement invariance is 
-    the unmodeled loading. Testing parameters are $\sigma_{12}$ 
+    the unmodeled loading. Testing parameters are $\phi_{12}$ 
     and $\mu_{11}$.}}
@@
    \item
-      The ordinal test statistic proposed have relatively-high power 
+      The ordinal test statistic proposed have better power 
       for detecting measurement 
@@
 \begin{itemize}
-  \item Gratitude scale (GQ-6, 5 items) completed by six age
+  \item Gratitude scale (GQ-6) completed by six age
     groups(10-11, 12-13, 14, 15, 16, 17-19) \cite{frofan11}
+  \item sample size is 1327.
+  \item LRT is always significant when sample size is large. 
 \end{itemize}
-<<data,eval=FALSE>>=
-data("YouthGratitude", package = "psychotools")
-compcases <- apply(YouthGratitude[, 4:28], 1, 
-                   function(x) all(x %in% 1:9))
-yg <- YouthGratitude[compcases, ]
-@
-
 \end{frame}
@@
   \suite
-\item Fit your constrained model:no need to comparing various models.
+\item Fit your constrained model: no need to comparing various models.
 \end{enumerate}
@@
 wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, 
-               vcov = "info", functional = "WDMo",  plot = FALSE, 
-               ylim = c(0, 3.2), xlab = "Age group", 
-               ylab = "Weighted test statistics")
+               vcov = "info", functional = "WDMo")
 wdmo
@@
 sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "WDMo", plot = TRUE, ylim = c(0, 3.2),
-  xlab = "Age group", ylab = "Weighted test statistics",
-  main = expression(WDM[o]), boundary = list(col = "red", lty = 1))
+  functional = "WDMo", plot = TRUE, 
+               ylim = c(0, 3.2), xlab = "Age group", 
+               ylab = "Weighted test statistics")
 @
@@
 %--------------------------------------------------------------------------
-\begin{frame}[shrink=0.8]{References}
+\begin{frame}[shrink=0.3]{Reference}
 \bibliographystyle{apacite}
 \bibliography{refs}
+\end{frame}
+%--------------------------------------------------------------------------
+
+\begin{frame}
+\frametitle{Categorical statistics equation}
+\begin{equation}
+   \label{eq:lmuo}
+    \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
+   \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}
+     (\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
+\end{equation}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Categorical statistics equation}
-\begin{equation}
-   \label{eq:lmuo}
-    \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
-   \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}
-     (\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
-\end{equation}
+  \frametitle{Continuous statistics equations}
+\begin{eqnarray}
+    \label{eq:dmax}
+    \mathit{DM} & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} | \\
+        \label{eq:cvm}
+    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
+    \label{eq:maxlm}
+    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
+      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
+      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
+\end{eqnarray}
 \end{frame}

### 2014-06-26 frontiers/IMPSpres-knitr.Rnw
update presentation and lines. Make sure the time.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@346 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{beamerthemeSzeged,graphics,amsmath,bm,booktabs}
+\usepackage{appendixnumberbeamer}
 \usepackage[notocbib]{apacite}
@@
   null hypotheis)
-\item \textbf{Score-based Tests (more general) w.r.t. categorical,
-    ordinal, continuous variable}. 
+\item \textbf{Score-based Tests (more general) w.r.t. auxiliary
+    variable (categorical, ordinal, continuous)}. 
 \end{itemize}
@@
   \end{itemize}
-\begin{eqnarray*}
-  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
-  {\bm 0},\text{ where} \\
- \\
+
+  \begin{eqnarray}
    s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} 
   \log
-\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
-\end{eqnarray*}
+\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}\\
+  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
+  {\bm 0} 
+  \end{eqnarray}
+
 \end{frame}
@@
 \begin{frame}
-  \frametitle{Empirical Cumulative Sum Score: Decorrelate and
-    Systematic Order}
+  \frametitle{Empirical Cumulative Sum Score:  Systematic Order and 
+    Decorrelate}
  We need a way to aggregate scores across people so that we
@@
       \begin{itemize}
+        \item Order individuals by an auxiliary variable.
         \item Decorrelate sores for each parameter.    
-        \item Order individuals by an auxiliary variable. 
       \end{itemize}
@@
     \begin{column}{0.48\textwidth}
-      \begin{block}{Empirical Score}
+      \begin{block}{Score}
         \resizebox{5.0cm}{!}{
@@
                 x_{n-1}&  12.34 & 13.32 & \ldots & 6.89\cr
-                x_{n}& 0     &   0     & \ldots & 0} $$}
+                x_{n}& -5.78     &   3.50     & \ldots & 7.67} $$}
         \end{block}
@@
      \begin{column}{0.48\textwidth}
-       \begin{block}{Empirical Cumulative score}
+       \begin{block}{Empirical Cumulative Score}
          \resizebox{5.0cm}{!}{
@@
    \item Three sample sizes ($n=120,\textcolor{red}{480},960$), 
-   \item Three numbers of categories ($m=4,8,12$), 
+   \item Three numbers of ordinal levels ($m=4,8,12$), 
    \item Violation magnitude: $d$ times the parameters' asymptotic 
@@
   \begin{figure}
-    \caption{Missed out parameter is not the M.I. parameter}
+    \caption{Excluded out parameter is not the problematic parameter}
     \label{fig:pathinmodel1}
@@
 wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, 
-               vcov = "info", functional = "WDMo")
+               vcov = "info", functional = "WDMo", 
+               plot = FALSE)
 wdmo
@@
 %--------------------------------------------------------------------------
+\appendix
 \begin{frame}[shrink=0.3]{Reference}
@@
 \begin{frame}
-\frametitle{Categorical statistics equation}
+\frametitle{Categorical Statistics Equation}
 \begin{equation}
@@
 \begin{frame}
-  \frametitle{Ordinal statistics equations}
+  \frametitle{Ordinal Statistics Equations}
 \scriptsize{\begin{eqnarray}
@@
 \begin{frame}
-  \frametitle{Continuous statistics equations}
+  \frametitle{Continuous Statistics Equations}
 \begin{eqnarray}

### 2014-06-30 frontiers/IMPSpres-knitr.Rnw
Revised imps presentation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@347 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
                       \inst{2} Universit\"{a}t Innsbruck}
+\date{Supported by NSF grant SES-1061334}
 
@@
 \newcommand{\suite}{\setcounter{enumi}{\thesauvegardeenumi}}
+
+\renewcommand\bibliographytypesize{\tiny}
 
@@
       assigning scores across diverse groups of individuals.\\ \ \\
-    \item Notable examination of measurement invariance:
+    \item Applications illustrating the importance of measurement invariance:
       \begin{itemize}
          \item SAT for different ethnic groups \cite{atk01}
-         \item Intelligence tests for different cohorts. \cite{wich04}
+         \item Intelligence tests' use over time. \cite{wich04}
       \end{itemize}
@@
 \begin{frame}
-  \frametitle{Example: Grade 1, 2, 3 vs. Grade 4, 5, 6}
+  \frametitle{Example: Grade 5 vs. Grade 6}
  \begin{figure}
@@
 \begin{itemize}
-\item Likelihood Ratio Test (LRT) (Require estimates under null
-  hypothesis and alternative).
+\item Likelihood Ratio Test (LRT): Require estimates under null
+  hypothesis and alternative.
 %  \begin{itemize}
@@
 %    \end{itemize}
-\item Wald Test (Require estimates under alternative). 
-\item Lagrange Multiplier Test (Score Test) (Require estimates under
-  null hypotheis)
-\item \textbf{Score-based Tests (more general) w.r.t. auxiliary
-    variable (categorical, ordinal, continuous)}. 
+\item Wald Test: Require estimates under alternative. 
+\item Lagrange Multiplier Test (Score Test): Require estimates under
+  null hypotheis.
+\item \textbf{Score Test generalizations w.r.t. general auxiliary
+    variable (categorical, ordinal, or continuous)}. 
 \end{itemize}
@@
 %--------------------------------------------------------------------------
+\begin{frame}
+  \frametitle{Score Aggregation}
+  \begin{itemize}
+    \item Each individual has a score for each model parameter,
+      resulting in $nk$ total scores (where $k$ is the dimension of
+      $\bm{\theta}$).
+    \item To draw general conclusions about measurement invariance, we
+      need to aggregate across scores in specific manners.
+    \item When $V$ is ordinal or continuous, we can make use of the
+      ordering in $V$ to obtain novel test statistics.
+  \end{itemize}
+\end{frame}
+
+%--------------------------------------------------------------------------
 
 \begin{frame}
-  \frametitle{Empirical Cumulative Sum Score:  Systematic Order and 
-    Decorrelate}
- We need a way to aggregate scores across people so that we
-      can draw some general conclusions.
+  \frametitle{Cumulative Sum of Scores}
       \begin{itemize}
         \item Order individuals by an auxiliary variable.
-        \item Decorrelate sores for each parameter.    
+        \item Decorrelate scores, using an estimate of the information
+          matrix.    
       \end{itemize}
@@
 \begin{frame}
-  \frametitle{Convergence Distribution}
+  \frametitle{Distributional Theory}
   \begin{itemize}
@@
  \begin{frame}
-   \frametitle{Statistics Based on Empirical Cumulative Sum Socre}
+   \frametitle{Test Statistics}
 \begin{itemize}
-  \item Statistics for auxiliary categorical variable:\\
-maximum Lagrange multiplier categorical ($\mathit{LM}_{uo}$), 
-asymptotic equvialent to LRT.\\
-\item Statistics for auxiliary ordinal variable: \\
-Weighted Double Maximum ordinal ($\mathit{WDM}_o$),\\
-maximum Lagrange multiplier ordinal ($\max\mathit{LM}_o$). \\
-\item Statistics for auxiliary continuous variable:\\
-Double maximumum statistic ($\mathit{DM}$)\\
+  \item Categorical $V$: Maximum Lagrange multiplier test (asymptotically equvialent to LRT).
+\item Ordinal $V$: Weighted double maximum test ($\mathit{WDM}_o$);\\
+Maximum Lagrange multiplier test ($\max\mathit{LM}_o$).
+\item Continuous $V$: Double maximumum test ($\mathit{DM}$)\\
 Cram{\'e}r-von Mises type statistic ($\mathit{CvM}$)\\
@@
   \frametitle{Study Motivation}
-  \begin{block}{Practice Consideration}
+  \begin{block}{Practical Considerations}
     \begin{itemize}
       \item Model misspecification is inevitable in practice.  
-      \item M.I. is often studied with respect to \textbf{ordinal} variable.
+      \item Measurement invariance is often studied with respect to an
+        \textbf{ordinal} variable.
       \item Availability/Convenience of software implementation.
@@
   \begin{block}{Simulation:}
-    What is the ordinal tests' power under model misspecification? 
+    What are the ordinal tests' power under model misspecification? 
     \begin{itemize}
@@
   \begin{figure}
-    \caption{Excluded out parameter is not the problematic parameter}
+    \caption{Excluded parameter does not violate invariance}
     \label{fig:pathinmodel1}
@@
 \begin{figure}
-\tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
-    parameters are $\lambda_{11}$ and $\lambda_{11} \ldots \lambda_{62}$.}}
+\tiny{\caption{$\lambda_{11}$ violates measurement invariance.  Tested
+    parameters are $\lambda_{11}$ and \{$\lambda_{11}, \ldots, \lambda_{62}$.\}}}
 \label{fig:sim21res}
@@
 \begin{figure}
-\tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
-    parameters are $\phi_{12}$, $\psi_{11} \ldots \psi_{66}$, and
-    $\mu_{11} \ldots \mu_{66}$.}}
+\tiny{\caption{$\lambda_{11}$ violates measurement invariance.  Tested
+    parameters are $\phi_{12}$, \{$\psi_{11}, \ldots, \psi_{66}$\}, and
+    \{$\mu_{11}, \ldots, \mu_{66}$\}.}}
 \label{fig:sim22res}
@@
 \begin{figure}
-    \caption{Excluded parameter is the problematic parameter}
+    \caption{Excluded parameter does violate invariance}
     \label{fig:pathinmodel2}
@@
 \tiny{\caption{The parameter violating measurement invariance is the 
-    unmodeled loading. Testing parameters are $\lambda_{11}$ and $\psi_{11}$.}}
+    unmodeled loading. Tested parameters are $\lambda_{11}$ and $\psi_{11}$.}}
 \label{fig:sim23res}
@@
 \tiny{\caption{The parameter violating measurement invariance is 
-    the unmodeled loading. Testing parameters are $\phi_{12}$ 
+    the unmodeled loading. Tested parameters are $\phi_{12}$ 
     and $\mu_{11}$.}}
@@
    \item
-      The ordinal test statistic proposed have better power 
+      The ordinal test statistics have better power 
       for detecting measurement 
-      invariance violation parameter specifically that are monotonic 
-      with the auxiliary ordinal variable. 
-   \item This feature is robust as long as the violation 
+      invariance violations that are monotonic 
+      with $V$.
+   \item This feature is robust as long as the violating 
       parameter is included in the testing model.
-   \item The use of these tests is convenient without changing the 
-      focal psychometric model.
+   \item The use of these tests is convenient, requiring estimation of
+     only a reduced model.
    \end{itemize}
@@
 \begin{frame}
-  \frametitle{Application: Software Preparation}
+  \frametitle{Software Illustration}
 The tests can be carried out in R. Specifically, 
@@
     groups(10-11, 12-13, 14, 15, 16, 17-19) \cite{frofan11}
-  \item sample size is 1327.
+  \item $n=1327$.
   \item LRT is always significant when sample size is large. 
@@
   \suite
-  \item Test if M.I. is violated.
+  \item Test of measurement invariance
 \end{enumerate}
@@
 \appendix
-\begin{frame}[shrink=0.3]{Reference}
+\begin{frame}{Reference}
 \bibliographystyle{apacite}

### 2014-07-02 frontiers/IMPSpres-knitr.Rnw
update IMPSpres-knitr.Rnw and lines according to Ed's comment.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@348 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \renewcommand\bibliographytypesize{\tiny}
+\DeclareMathOperator*{\argmax}{\arg\!\max}
 
@@
   \begin{eqnarray}
+  \hat{\bm \theta}=\argmax_{\bm \theta} \ell(\bm \theta; x_{1}, \ldots, x_{n})\\
    s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} 
@@
 \end{enumerate}
-<<sctest-ord2, eval=TRUE>>=
+<<sctest-ord2, eval=FALSE>>=
  library(strucchange)
-wdmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, 
-               vcov = "info", functional = "WDMo", 
-               plot = FALSE)
-wdmo
+maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, 
+               vcov = "info", functional = "maxLMo", 
+               plot = TRUE)
+@
+<<sctest-ord2res, eval=TRUE>>=
+maxlmo
 @ 
@@
 \begin{figure}
-\tiny{\caption{Fluctuation processes for the $\mathit{WDM}_o$ statistic}}
+\tiny{\caption{Fluctuation processes for the $\mathit{LM}_o$ statistic}}
 \label{fig:ordres}
@@
 sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
-  functional = "WDMo", plot = TRUE, 
-               ylim = c(0, 3.2), xlab = "Age group", 
-               ylab = "Weighted test statistics")
+  functional = maxLMo, plot = TRUE, 
+               ylim = c(0, 14), xlab = "Age group", 
+               ylab = "LM statistics")
 @
@@
  \end{itemize}
+\vspace{0.3in}
+ Wang, T., Merkle, E. C., Zeileis, A. (2014). Score-based tests of measurement 
+ invariance: Use in practice. \emph{Quantitative Psychology and Measurement},
+ 5, 438. 
 \end{frame}

### 2014-07-07 frontiers/IMPSpres-knitr.Rnw
update presentation and lines.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@349 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
update presentation.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@350 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor changes to IMPS presentation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@351 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \end{frame}
-%--------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------
+
 \begin{frame}
@@
 \end{frame}
-%--------------------------------------------------------------------------
-%\begin{frame}
+
+% --------------------------------------------------------------------------
+
+% \begin{frame}
 %\frametitle{Problem Scenario}
@@
 %\end{frame}
+
 %%--------------------------------------------------------------------------
+
 \begin{frame}
@@
 \end{frame}
+
 % --------------------------------------------------------------------------
+
 \begin{frame}
@@
 \end{frame}
+
 % --------------------------------------------------------------------------
+
 \section{Theoretical Background}
@@
 
+\scriptsize{   
   \begin{eqnarray}
-  \hat{\bm \theta}=\argmax_{\bm \theta} \ell(\bm \theta; x_{1}, \ldots, x_{n})\\
+    \label{eq:scoreml}
     s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} 
-  \log
-\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}\\
-  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
-  {\bm 0} 
+    \log\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} =
+    \widehat{\bm{\theta}}}, \text{where}\\[1em]
+    \label{eq:ml}
+    \hat{\bm \theta}=\argmax_{\bm \theta} \log\text{L}(\bm \theta; x_{1}, 
+    \ldots, x_{n})\\
+    \label{eq:scoresumml}
+   \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) = {\bm 0}
   \end{eqnarray}
+}
 
 \end{frame}
-%--------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------
+
 \begin{frame}
@@
   \end{frame}
-%--------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------
+
 \begin{frame}
@@
 \end{frame}
-%--------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------
 
@@
    matrix ${\bm B}(\cdot; \hat {\bm \theta})$ converges in distribution to a 
-   k-dimensional Brownian bridge ${\bm B}^{0}(\cdot)$.
+   $k$-dimensional Brownian bridge ${\bm B}^{0}(\cdot)$.
    \begin{equation}
@@
  \end{frame}
-%--------------------------------------------------------------------------  
+
+% --------------------------------------------------------------------------
+
 \begin{frame}
@@
 %-------------------------------------------------------------------------
+
 \section{Simulation}
@@
 \end{frame}
-%-------------------------------------------------------------------------   
+
+% -------------------------------------------------------------------------
+
 \begin{frame}
@@
     \begin{itemize}
-      \item The estimated model excludes one parameter, which is \emph
-        {not} the problematic parameter.
+      \item The estimated model excludes one parameter, which 
+        {\em is not} the problematic parameter.
       \item The estimated model excludes one parameter, which \emph{is}
@@
    \item This feature is robust as long as the violating 
-      parameter is included in the testing model.
-   \item The use of these tests is convenient, requiring estimation of
+      parameter is included in the estimated model.
+   \item Use of these tests is convenient, requiring estimation of
      only a reduced model.
@@
 \end{frame}
-%--------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------
+
 \section{Application}
@@
 \item Model estimation can be carried out in \emph{lavaan} package. 
-\item The test statistic and plot is got from \emph{strucchange} package.
+\item The test statistic and plot is obtained from \emph{strucchange} package.
 \end{itemize}
 \end{frame}
-%--------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------
+
 \begin{frame}[fragile]
@@
 %--------------------------------------------------------------------------
+
 \begin{frame}[fragile]
@@
   \suite
-\item Fit your constrained model: no need to comparing various models.
+\item Fit the constrained model: no need to compare various models.
 \end{enumerate}
@@
 \end{frame}
-%--------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------
+
 \begin{frame}[fragile]
@@
  Wang, T., Merkle, E. C., Zeileis, A. (2014). Score-based tests of measurement 
- invariance: Use in practice. \emph{Quantitative Psychology and Measurement},
- 5, 438. 
+ invariance: Use in practice. \emph{Frontiers in Psychology}, {\em 5(438)}, 
+ 1--11. 
 \end{frame}
-%--------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------
+
 \appendix

### 2014-07-08 frontiers/IMPSpres-knitr.Rnw
update presentation according to Achim's comments.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@352 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{appendixnumberbeamer}
+\usepackage{blkarray}
 \usepackage[notocbib]{apacite}
@@
  \begin{itemize}
-    \item The score-based tests \cite{MerZei13, MerFanZei}
-      rely on first derivatives of
-      the model's log-likelihood function.
+  %  \item The score-based tests \cite{MerZei13, MerFanZei}
+  %    rely on first derivatives of
+  %    the model's log-likelihood function.
+    \item Access deviations in the model score \cite{MerZei13,
+      MerFanZei}, i.e., the first derivatives of the model's log
+      likelihoood function.  
     \item We consider individual terms ({\em scores}) of the gradient.  
@@
         \resizebox{5.0cm}{!}{ 
-        $$\bordermatrix{\text{}& \theta_{1} & \theta_{2} & \ldots 
-          &\textcolor{red}{\theta_{k}}\cr
-                x_{1}&  1.32  & -4.32 & \ldots & -2.34\cr
-                x_{2}&  3.37 & 8.32 & \ldots & 63.45\cr
-                x_{3}&  4.12 & 3.48 & \ldots & -3.11\cr
-                x_{..}& \vdots & \vdots & \vdots & \vdots\cr
-                x_{n-1}&  12.34 & 13.32 & \ldots & 6.89\cr
-                x_{n}& -5.78     &   3.50     & \ldots & 7.67} $$}
+          $
+            \begin{blockarray}{cccccc}
+              \text{}& \theta_{1} & \theta_{2} & \ldots 
+                     &\textcolor{red}{\theta_{k}} & \text{age} \\
+               \begin{block}{l(cccc)c}
+                 x_{1} &  1.32  & -4.32 & \ldots & -2.34 & 10.5\\
+                 x_{2} &  3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
+                 x_{3} &  4.12 & 3.48 & \ldots & -3.11 & 11.3\\
+                 x_{..} & \vdots & \vdots & \vdots & \vdots & 17.1\\ 
+                 x_{n-1} &  12.34 & 13.32 & \ldots & 6.89 & 18.6\\
+                 x_{n} & -5.78 & 3.50 & \ldots & 7.67 & 18.9\\
+               \end{block}
+            \end{blockarray} 
+           $
+          }
         \end{block}   
      \end{column}
+
      \begin{column}{0.48\textwidth}
@@
           \resizebox{5.0cm}{!}{
-      $$\bordermatrix{\text{}& \theta_{1} & \theta_{2} & \ldots
-        &\textcolor{red}{\theta_{k}}\cr
-                x_{1}&  1.32  & -4.32 & \ldots & -2.34\cr
-                x_{2}&  4.69 & 4.00 & \ldots & 61.11\cr
-                x_{3}&  8.81 & 7.48 & \ldots & 58.00\cr
-                x_{..}& \vdots & \vdots & \vdots & \vdots\cr
-                x_{n-1}&  7.89 & 3.32 & \ldots & 4.91\cr
-                x_{n}& 0     &   0     & \ldots & 0} $$}
+           $
+            \begin{blockarray}{cccccc}
+              \text{}& \theta_{1} & \theta_{2} & \ldots 
+                     &\textcolor{red}{\theta_{k}} & \text{age} \\
+               \begin{block}{l(cccc)c}
+                 x_{1} &  1.32  & -4.32 & \ldots & -2.34 & 10.5 \\
+                 x_{2} &  4.69 & 4.00 & \ldots & 61.11 & 11.0 \\
+                 x_{3} &  8.81 & 7.48 & \ldots & 58.00 & 11.3 \\
+                 x_{..} & \vdots & \vdots & \vdots & \vdots & 17.1 \\
+                 x_{n-1} &  7.89 & 3.32 & \ldots & 4.91 & 18.6 \\
+                 x_{n}& 0     &   0     & \ldots & 0 & 18.9 \\
+               \end{block}
+            \end{blockarray} 
+           $
+         }  
         \end{block}
@@
   \item Gratitude scale (GQ-6) completed by six age
-    groups(10-11, 12-13, 14, 15, 16, 17-19) \cite{frofan11}
+    groups (10--11, 12--13, 14, 15, 16, 17--19) \cite{frofan11}
   \item $n=1327$.
@@
 <<modeest,eval=FALSE>>=
- library(lavaan)
+ library("lavaan")
   restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
@@
 <<sctest-ord2, eval=FALSE>>=
- library(strucchange)
+ library("strucchange")
 maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4, 
@@
 \begin{figure}
-\tiny{\caption{Fluctuation processes for the $\mathit{LM}_o$ statistic}}
+\tiny{\caption{Fluctuation processes for $\max\mathit{LM}_o$ statistic}}
 \label{fig:ordres}
 <<sctest-ord-graphics, fig.height=3.3, fig.width=5, echo=FALSE, results='hide'>>=
+par(cex = 0.8)
 sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
@@
                ylim = c(0, 14), xlab = "Age group", 
-               ylab = "LM statistics")
+               ylab = "LM statistics", main = expression(maxLM[o]))
 @

### 2014-07-10 frontiers/IMPSpres-knitr.Rnw
update presentation according to Ed's comment.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@354 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   \begin{columns}
-    \begin{column}{0.48\textwidth}     
-      \begin{block}{Score}
-        \resizebox{5.0cm}{!}{ 
+    \begin{column}{0.33\textwidth}     
+      \begin{block}{Unordered Score}
+        \resizebox{3.3cm}{!}{ 
           $
-            \begin{blockarray}{cccccc}
-              \text{}& \theta_{1} & \theta_{2} & \ldots 
+            \begin{blockarray}{lccccc}
+              \text{obs} &\theta_{1} & \theta_{2} & \ldots 
+                     &\textcolor{red}{\theta_{k}} & \text{age} \\
+               \begin{block}{l(cccc)c}
+                 &4.12 & 3.48 & \ldots & -3.11 & 11.3\\
+                 &1.32  & -4.32 & \ldots & -2.34 & 10.5\\
+                 &3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
+                 &-5.78 & 3.50 & \ldots & 7.67 & 18.9\\
+                 &\vdots & \vdots & \vdots & \vdots & 17.1\\ 
+                 &12.34 & 13.32 & \ldots & 6.89 & 18.6\\              
+               \end{block}
+            \end{blockarray} 
+           $
+          }
+        \end{block}   
+     \end{column}
+
+    \begin{column}{0.33\textwidth}     
+      \begin{block}{Ordered Score}
+        \resizebox{3.3cm}{!}{ 
+          $
+            \begin{blockarray}{lccccc}
+              \text{obs}& \theta_{1} & \theta_{2} & \ldots 
                      &\textcolor{red}{\theta_{k}} & \text{age} \\
@@
 
-     \begin{column}{0.48\textwidth}
-          \begin{block}{Empirical Cumulative Score}
-          \resizebox{5.0cm}{!}{
+     \begin{column}{0.33\textwidth}
+          \begin{block}{\scriptsize{Empirical Cumulative Score}}
+          \resizebox{3.3cm}{!}{
            $
-            \begin{blockarray}{cccccc}
-              \text{}& \theta_{1} & \theta_{2} & \ldots 
+            \begin{blockarray}{lccccc}
+              \text{obs}& \theta_{1} & \theta_{2} & \ldots 
                      &\textcolor{red}{\theta_{k}} & \text{age} \\
@@
 \label{fig:ordres}
-<<sctest-ord-graphics, fig.height=3.3, fig.width=5, echo=FALSE, results='hide'>>=
-par(cex = 0.8)
+<<sctest-ord-graphics,  out.width="250px", out.height ="180px", echo=FALSE, results='hide'>>=
 sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",

### 2014-07-14 frontiers/IMPSpres-knitr.Rnw
change the empirical part name.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@355 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Revisions to IMPS slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@356 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
       % \end{center}   
-       $V_{g}$ can be categorical, \textit{ordinal} and continous.
+       $V_{g}$ can be categorical, \textit{ordinal}, or continuous.
     \end{block}
@@
 \begin{frame}
-  \frametitle{Cumulative Sum of Scores}
+  \frametitle{Empirical Cumulative Score Process}
       \begin{itemize}
@@
           matrix.    
-      \end{itemize}
 \begin{equation} 
@@
 of $n$, $t \in (1/n, n)$.
+  \end{itemize}
   \end{frame}
@@
      \begin{column}{0.33\textwidth}
-          \begin{block}{\scriptsize{Empirical Cumulative Score}}
+          \begin{block}{Cumulative Score}
           \resizebox{3.3cm}{!}{
@@
 \begin{itemize}
+    \itemsep8pt
   \item Categorical $V$: Maximum Lagrange multiplier test (asymptotically equvialent to LRT).
@@
 Maximum Lagrange multiplier test ($\max\mathit{LM}_o$).
-\item Continuous $V$: Double maximumum test ($\mathit{DM}$)\\
+\item Continuous $V$: Double maximum test ($\mathit{DM}$)\\
 Cram{\'e}r-von Mises type statistic ($\mathit{CvM}$)\\
@@
         \textbf{ordinal} variable.
-      \item Availability/Convenience of software implementation.
+      \item Availability/convenience of software implementation.
     \end{itemize}
@@
  \begin{itemize}
+     \itemsep8pt
    \item
@@
 \begin{itemize}
-\item Model estimation can be carried out in \emph{lavaan} package. 
-\item The test statistic and plot is obtained from \emph{strucchange} package.
+\item Model estimation via package \emph{lavaan}. 
+\item The test statistic and plot via package \emph{strucchange}.
 \end{itemize}
@@
 \label{fig:ordres}
-<<sctest-ord-graphics,  out.width="250px", out.height ="180px", echo=FALSE, results='hide'>>=
+<<sctest-ord-graphics, out.width="2.5in", out.height ="2.5in", echo=FALSE, results='hide', fig.height=5, fig.width=5>>=
 sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
@@
  \item Thanks! 
- \item Questions?
  \end{itemize}

### 2014-07-14 frontiers/IMPSpres-knitr.Rnw
update slide 10 three matrix.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@357 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \item Lagrange Multiplier Test (Score Test): Require estimates under
-  null hypotheis.
+  null hypothesis.
 \item \textbf{Score Test generalizations w.r.t. general auxiliary
@@
       MerFanZei}, i.e., the first derivatives of the model's log
-      likelihoood function.  
+      likelihood function.  
     \item We consider individual terms ({\em scores}) of the gradient.  
@@
 
-\begin{frame}
+\begin{frame}[t]
   \frametitle{Visual Explanation of the Score-based Test}
   \begin{columns}
-    \begin{column}{0.33\textwidth}     
-      \begin{block}{Unordered Score}
-        \resizebox{3.3cm}{!}{ 
+    \begin{column}{0.45\textwidth}     
+      \begin{block}{\scriptsize{Unordered Score}}
+        \resizebox{4.0 cm}{!}{ 
           $
@@
               \text{obs} &\theta_{1} & \theta_{2} & \ldots 
-                     &\textcolor{red}{\theta_{k}} & \text{age} \\
+                     &\textcolor{red}{\theta_{k}} & V \\
                \begin{block}{l(cccc)c}
-                 &4.12 & 3.48 & \ldots & -3.11 & 11.3\\
-                 &1.32  & -4.32 & \ldots & -2.34 & 10.5\\
-                 &3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
-                 &-5.78 & 3.50 & \ldots & 7.67 & 18.9\\
-                 &\vdots & \vdots & \vdots & \vdots & 17.1\\ 
-                 &12.34 & 13.32 & \ldots & 6.89 & 18.6\\              
+                x_{1} &4.12 & 3.48 & \ldots & -3.11 & 11.3\\
+                x_{2}  &1.32  & -4.32 & \ldots & -2.34 & 10.5\\
+                x_{3} &3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
+                x_{..} &\vdots & \vdots & \vdots & \vdots & \vdots\\ 
+                x_{n-1} &-5.78 & 3.50 & \ldots & 7.67 & 18.9\\            
+                x_{n} &12.34 & 13.32 & \ldots & 6.89 & 18.6\\              
                \end{block}
@@
 
-    \begin{column}{0.33\textwidth}     
-      \begin{block}{Ordered Score}
-        \resizebox{3.3cm}{!}{ 
+    \begin{column}{0.45\textwidth}     
+      \begin{block}{\scriptsize{Ordered Score}}
+        \resizebox{4.0 cm}{!}{ 
           $
@@
               \text{obs}& \theta_{1} & \theta_{2} & \ldots 
-                     &\textcolor{red}{\theta_{k}} & \text{age} \\
+                     &\textcolor{red}{\theta_{k}} & V \\
                \begin{block}{l(cccc)c}
-                 x_{1} &  1.32  & -4.32 & \ldots & -2.34 & 10.5\\
-                 x_{2} &  3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
-                 x_{3} &  4.12 & 3.48 & \ldots & -3.11 & 11.3\\
-                 x_{..} & \vdots & \vdots & \vdots & \vdots & 17.1\\ 
-                 x_{n-1} &  12.34 & 13.32 & \ldots & 6.89 & 18.6\\
-                 x_{n} & -5.78 & 3.50 & \ldots & 7.67 & 18.9\\
-               \end{block}
-            \end{blockarray} 
-           $
-          }
-        \end{block}   
-     \end{column}
-
-     \begin{column}{0.33\textwidth}
-          \begin{block}{Cumulative Score}
-          \resizebox{3.3cm}{!}{
-           $
-            \begin{blockarray}{lccccc}
-              \text{obs}& \theta_{1} & \theta_{2} & \ldots 
-                     &\textcolor{red}{\theta_{k}} & \text{age} \\
-               \begin{block}{l(cccc)c}
-                 x_{1} &  1.32  & -4.32 & \ldots & -2.34 & 10.5 \\
-                 x_{2} &  4.69 & 4.00 & \ldots & 61.11 & 11.0 \\
-                 x_{3} &  8.81 & 7.48 & \ldots & 58.00 & 11.3 \\
-                 x_{..} & \vdots & \vdots & \vdots & \vdots & 17.1 \\
-                 x_{n-1} &  7.89 & 3.32 & \ldots & 4.91 & 18.6 \\
-                 x_{n}& 0     &   0     & \ldots & 0 & 18.9 \\
+                 x_{(1)} &  1.32  & -4.32 & \ldots & -2.34 & 10.5\\
+                 x_{(2)} &  3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
+                 x_{(3)} &  4.12 & 3.48 & \ldots & -3.11 & 11.3\\
+                 x_{(..)} & \vdots & \vdots & \vdots & \vdots & \vdots\\ 
+                 x_{(n-1)} &  12.34 & 13.32 & \ldots & 6.89 & 18.6\\
+                 x_{(n)} & -5.78 & 3.50 & \ldots & 7.67 & 18.9\\
                \end{block}
@@
    \end{columns}
+
+   \begin{block}{\scriptsize{Empirical Cumulative Score}}
+          $\bm B(\cdot;\hat{\bm \theta}) = \hat{\bm I}^{-1/2}n^{-1/2} \times$
+            \resizebox{4.2cm}{!}{
+           $
+            \begin{blockarray}{lccccc}
+               \text{obs}& \theta_{1} & \theta_{2} & \ldots 
+                     &\textcolor{red}{\theta_{k}} & V \\
+               \begin{block}{l(cccc)c}
+                 x_{(1)} &  1.32  & -4.32 & \ldots & -2.34 & 10.5 \\
+                 x_{(2)} &  4.69 & 4.00 & \ldots & 61.11 & 11.0 \\
+                 x_{(3)} &  8.81 & 7.48 & \ldots & 58.00 & 11.3 \\
+                 x_{(..)} & \vdots & \vdots & \vdots & \vdots & \vdots\\
+                 x_{(n-1)} &  7.89 & 3.32 & \ldots & 4.91 & 18.6 \\
+                 x_{(n)}& 0     &   0     & \ldots & 0 & 18.9 \\
+               \end{block}
+            \end{blockarray} 
+           $
+         }
+        \end{block}
+
 \end{frame}
@@
     \itemsep8pt
-  \item Categorical $V$: Maximum Lagrange multiplier test (asymptotically equvialent to LRT).
+  \item Categorical $V$: Maximum Lagrange multiplier test (asymptotically equivalent to LRT).
 \item Ordinal $V$: Weighted double maximum test ($\mathit{WDM}_o$);\\

### 2014-11-12 ordinal/preprint.Rnw
added final publication info
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@360 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \fancyhf{}
-\fancyfoot[LO,LE]{\small This is a preprint of an article accepted for publication in \textit{Psychometrika}.\\ %%, \textbf{78}(1), 59--82.\\
-  Copyright~{\copyright} 2013 The Psychometric Society \url{http://www.psychometrika.org/}}
-  %% \doi{10.1007/S11336-012-9302-4}}
+\fancyfoot[LO,LE]{\small This is a preprint of an article published in \textit{Psychometrika}, \textbf{79}(4), 569--584.\\
+  Copyright~{\copyright} 2013 The Psychometric Society \url{http://www.psychometrika.org/}\\
+  \doi{10.1007/S11336-013-9376-7}}
 }

### 2015-07-22 ecpa2015/slides.Rnw
first draft for ECPA13 slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@377 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[11pt,compress,t]{beamer}
\usetheme{Z}
\usepackage{amsfonts,amstext,amsmath,bm}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

%% need no \usepackage{Sweave}
\definecolor{InputColor}{rgb}{0,0,0.3}
\definecolor{OutputColor}{rgb}{0.2,0.2,0.2}

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
set.seed(1090)
@

\begin{document}

\title{{\fontsize{14.7}{16}\selectfont Score-Based Tests of Measurement Invariance with Respect to Continuous and Ordinal Variables}}
\author{Achim Zeileis, Edgar C. Merkle, Ting Wang}
\URL{http://eeecon.uibk.ac.at/~zeileis/}
\lecture{Score-Based Tests of Measurement Invariance with Respect to Continuous and Ordinal Variables}

\subsection{Overview}

\begin{frame}
\frametitle{Overview}

\begin{itemize}
  \item Motivation
  \item Framework
  \item Score-based tests
  \begin{itemize}
    \item Continuous variables
    \item Ordinal variables
    \item Categorial variables
  \end{itemize}
  \item Illustration
\end{itemize}

\end{frame}

\subsection{Motivation}

\begin{frame}
\frametitle{Motivation}

\textbf{Psychometric models:} Typically measure latent scales based on certain
manifest variables, e.g., item response theory (IRT) models or confirmatory
factor analysis (CFA, today's focus).

\medskip

\textbf{Crucial assumption:} Measurement invariance (MI). Otherwise observed differences in
scores cannot be reliably attributed to the latent variable that the model
purports to measure.

\medskip

\textbf{Parameter stability:} In parametric models, the MI assumption
corresponds to stability of parameters across all possible subgroups.

\medskip

\textbf{Inference:} The typical approach for assessing MI is
\begin{itemize}
  \item to split the data into reference and focal groups,
  \item assess the stability of selected parameters (all or only a subset) across these groups
  \item by means of standard tests: likelihood ratio (LR), Wald, or Lagrange multiplier
    (LM or score) tests.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Motivation}

\textbf{Problems:}
\begin{itemize}
  \item Subgroups have to be formed in advance.
  \item Numerical variables are often categorized into groups in an ad hoc way
    (e.g., splitting at the median).
  \item In ordinal variables the ordering of the categories is often not exploited
    (assessing only if at least one group differs from the others).
  \item When likelihood ratio or Wald tests are employed, the model has to be
    fitted to each subgroup which can become numerically challenging and computationally
    intensive.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Motivation}

\textbf{Idea:}
\begin{itemize}
  \item Generalize the LM test (or score test).
  \item Thus, the model only has to be fitted once under the MI assumption to the full data set.
  \item Catpure model deviations along a variable that is suspected to cause MI violations.
  \item Exploit ordering to assess if there is (at least) one split so that the
    model parameters before and after the split differ.
  \item The split does \emph{not} have to be known or guessed in advance.
\end{itemize}

\medskip

\textbf{Illustration:} CFA for artificial data.
\begin{itemize}
  \item Model with two latent scales: Verbal and math score.
  \item Three manifest variables for each scale.
  \item Violation of MI for the math loadings along the age of the subjects.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Motivation: CFA for age $\le 16$}

\vspace*{-0.6cm}

\includegraphics[width=0.95\textwidth]{../psychoco2012/model_under16.pdf}

\end{frame}

\begin{frame}
\frametitle{Motivation: CFA for age $> 16$}

\vspace*{-0.6cm}

\includegraphics[width=0.95\textwidth]{../psychoco2012/model_over16.pdf}

\end{frame}

\subsection{Framework}

\begin{frame}
\frametitle{Framework}

\textbf{Model:} Based on log-likelihood $\ell(\cdot)$ for
$p$-dimensional observations $\bm{x}_i$ ($i = 1, \dots, n$)
based on $k$-dimensional parameter ${\bm \theta}$.

\medskip

\textbf{Estimation:} Maximum likelihood.
\[
  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i).
\]

\medskip

\textbf{Equivalently:} Solve first order conditions
\[
  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
\]
where the score function is the partial derivative of the casewise
likelihood contributions w.r.t.\ the parameters $\bm{\theta}$.
\[
  {\bm s}({\bm \theta}; \bm{x}_i) ~=~ \left(
    \frac{\partial \ell({\bm \theta}; \bm{x}_i)}{\partial \theta_1},
    \dots,
    \frac{\partial \ell({\bm \theta}; \bm{x}_i)}{\partial \theta_k}
  \right)^\top.
\]

\end{frame}

\begin{frame}
\frametitle{Framework}

\textbf{Assumption:} Distribution/likelihood of $\bm{x}_i$ depends
only on the latent scales (through the parameters $\bm{\theta}$) --
but not on any other variable $v_i$.

\medskip

\textbf{Alternative view:} Parameters $\bm{\theta}$ do not depend
any such variable $v_i$. Hence assess for $i = 1, \ldots, n$
\begin{eqnarray*}
    H_0: {\bm \theta}_i & = & {\bm \theta}_0, \\
    H_1: {\bm \theta}_i & = & {\bm \theta}(v_i).
\end{eqnarray*}

\medskip

\textbf{Special case:} Two subgroups resulting from one split point $\nu$.
\begin{eqnarray*}
  H_1^*: {\bm \theta}_i & = & \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu
  \end{array} \right.
\end{eqnarray*}

\medskip

\textbf{Tests:} LR/Wald/LM tests can be easily employed if pattern $\bm{\theta}(v_i)$
is known, specifically for $H_1^*$ with fixed split point $\nu$.

\end{frame}

\begin{frame}
\frametitle{Framework}

\textbf{For unknown split points:} Compute LR/Wald/LM tests for each possible
split point $v_1 \le v_2 \le \dots \le v_n$ and reject if the maximum statistic is large.

\medskip

\textbf{Caution:} By maximally selecting the test statistic different critical
values are required (not from a $\chi^2$ distribution)!

\medskip

\textbf{Illustration:} Assess all $k^* = 19$ model parameters from the artificial
CFA example along the continuous variable age ($v_i$).

\end{frame}

\begin{frame}
\frametitle{Framework}

\vspace*{-1.2cm}

\includegraphics[width=\textwidth]{../psychoco2012/paper-lm-lr-19.pdf}

\end{frame}

\begin{frame}
\frametitle{Framework}

\textbf{Note:} For the maxLM test the parameters $\hat{\bm \theta}$ only have
to be estimated once. Only the model scores ${\bm s}(\hat{\bm \theta}; \bm{x}_i)$
have to be aggregated differently for each split point.

\medskip

\textbf{More generally:} Consider a class of tests that
assesses whether the model ``deviations'' ${\bm s}(\hat{\bm \theta}; \bm{x}_i)$
depend on $v_i$. This can consider only a subset $k*$ of all $k$
parameters/scores or try to capture other patterns than $H_1^*$.

\end{frame}

\begin{frame}
\frametitle{Score-based tests}

\textbf{Fluctuation process:} Capture fluctuations in the cumulative
sum of the scores ordered by the variable $v$.
\[
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; \bm{x}_{(i)})
  \qquad (0 \le t \le 1).
\]
\begin{itemize}
  \item $\hat{\bm I}$ -- estimate of the information matrix.
  \item $t$ -- proportion of data ordered by $v$.
  \item $\lfloor n \cdot t \rfloor$ -- integer part of $n \cdot t$.
  \item $x_{(i)}$ -- observation with the $i$-th smallest value of the
    variable $v$.
\end{itemize}

\medskip

\textbf{Functional central limit theorem:} Under $H_0$ convergence to a (continuous) Brownian bridge process
${\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot)$,
from which critical values can be obtained -- either analytically or by simulation.

\end{frame}

\begin{frame}
\frametitle{Score-based tests: Continuous variables}

\textbf{Test statistics:} The empirical process can be viewed as a matrix
${\bm B}(\hat {\bm \theta})_{ij}$ with rows $i = 1, \dots, n$ (observations)
and columns $j = 1, \dots, k$ (parameters). This can be aggregated to scalar
test statistics along continuous the variable $v$.
\begin{eqnarray*}
    \mathit{DM}      & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} | \\
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
\end{eqnarray*}

\medskip

\textbf{Critical values:} Analytically for $\mathit{DM}$. Otherwise by
direct simulation or further refined simulation techniques.

\end{frame}

\begin{frame}
\frametitle{Score-based tests: Ordinal variables}

\textbf{Test statistics:} Aggregation along ordinal variables $v$ with $m$ levels.
\begin{eqnarray*}
      \mathit{WDM}_o     & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray*}
where $i_1, \dots, i_{m-1}$ are the numbers of observations in each category.

\medskip

\textbf{Critical values:} For $\mathit{WDM}_o$ directly from a multivariate normal distribution.
For $\max \mathit{LM}_o$ via simulation.

\end{frame}

\begin{frame}
\frametitle{Score-based tests: Categorical variables}

\textbf{Test statistic:} Aggregation within the $m$ (unordered) categories of $v$.
\[
  \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
    \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\]

\medskip

\textbf{Critical values:} From a $\chi^2$ distribution (as usual).

\medskip

\textbf{Asymptotically equivalent:} LR test.

\end{frame}

\subsection{Illustration}

\begin{frame}
\frametitle{Illustration}

\textbf{Software:} In R system for statistical computing.
\begin{itemize}
  \item \emph{strucchange} implements this general framwork for
    parameter instability tests.
  \item Object-oriented implementation that can be applied to many
    model classes, including \emph{lavaan} objects for CFA models.
\end{itemize}

\medskip

\textbf{Data:}
\begin{itemize}
  \item Application of adult gratitude scale to $n=1401$ youth aged 10--19 years.
  \item GQ-6 scale has five Likert scale items with seven points each.
  \item Assess the factor loadings of a one-factor model.
  \item Question: Measurement invariance across six age groups?
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Illustration}

\textbf{Packages:}

<<packages, results=hide>>=
library("lavaan")
library("strucchange")
@

\medskip

\textbf{Data:} Omitting incomplete cases.

<<data>>=
data("YouthGratitude", package = "psychotools")
compcases <- apply(YouthGratitude[, 4:28], 1,
  function(x) all(x %in% 1:9))
yg <- YouthGratitude[compcases, ]
@

\medskip

\textbf{Estimation:} One-factor CFA with loadings restricted to be equal
across age groups.

<<models>>=
gq6cfa <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE,
  group.equal = "loadings")
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Illustration}

\textbf{Measurement invariance tests:}

<<wdmo, strip.white=all, fig=TRUE, height=4.5, width=5, include=FALSE>>=
sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
  vcov = "info", functional = "WDMo", plot = TRUE)
@

<<maxlmo, strip.white=all, fig=TRUE, height=4.5, width=5, include=FALSE>>=
sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
  vcov = "info", functional = "maxLMo", plot = TRUE)
@

Both tests reflect only moderate parameter instability across age groups
and do not show significant violations of measurement invariance at 5\% level.

\end{frame}

\begin{frame}
\frametitle{Illustration}

\vspace*{-0.6cm}

\centering
\includegraphics[width=0.8\textwidth]{slides-wdmo}

\end{frame}

\begin{frame}
\frametitle{Illustration}

\vspace*{-0.6cm}

\centering
\includegraphics[width=0.8\textwidth]{slides-maxlmo}

\end{frame}

\subsection{Summary}

\begin{frame}
\frametitle{Summary}

\begin{itemize}
  \item General score-based test framework for assessing measurement
    invariance in parametric psychometric models.
  \item Assessment is along some variable $v$ which can be continuous,
    ordinal, or categorical.
  \item Tests can be seen as generalizations of the Lagrange multiplier
    (or score) test.
  \item Computation of critical values might require simulation from
    certain stochastic processes (Brownian bridges).
  \item Easy-to-use implementation available in R package \emph{strucchange}.
  \item Can be re-used in model-based recursive partitioning.
\end{itemize}

\end{frame}

\subsection{References}

\begin{frame}
\frametitle{References}

\small

Merkle EC, Zeileis A (2013).
  \dquote{Tests of Measurement Invariance without Subgroups: A Generalization of Classical Methods.}
  \emph{Psychometrika}, \textbf{78}(1), 59--82.
  \doi{10.1007/s11336-012-9302-4}

\medskip

Merkle EC, Fan J, Zeileis A (2014).
  \dquote{Testing for Measurement Invariance with Respect to an Ordinal Variable.}
  \emph{Psychometrika}, \textbf{79}(4), 569--584.
  \doi{10.1007/s11336-013-9376-7}

\medskip

Wang T, Merkle EC, Zeileis A (2014).
  \dquote{Score-Based Tests of Measurement Invariance: Use in Practice.}
  \emph{Frontiers in Psychology}, \textbf{5}(438).
  \doi{10.3389/fpsyg.2014.00438}

%% \medskip
%% 
%% Strobl C, Julia Kopf, Zeileis A (2015).
%%   \dquote{Rasch Trees: A New Method for Detecting Differential Item Functioning in the Rasch Model.}
%%   \textit{Psychometrika}. \textbf{80}(2), 289--316.
%%   \doi{10.1007/s11336-013-9388-3}

\end{frame}

\end{document}

### 2015-07-23 ecpa2015/slides.Rnw
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@378 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \textbf{Crucial assumption:} Measurement invariance (MI). Otherwise observed differences in
-scores cannot be reliably attributed to the latent variable that the model
+scales cannot be reliably attributed to the latent variable that the model
 purports to measure.
@@
   \item Subgroups have to be formed in advance.
-  \item Numerical variables are often categorized into groups in an ad hoc way
+  \item Continuous variables are often categorized into groups in an ad hoc way
     (e.g., splitting at the median).
@@
 \begin{itemize}
-  \item Generalize the LM test (or score test).
+  \item Generalize the LM test.
   \item Thus, the model only has to be fitted once under the MI assumption to the full data set.
@@
 \begin{itemize}
-  \item Model with two latent scales: Verbal and math score.
+  \item Model with two latent scales (verbal and math).
   \item Three manifest variables for each scale.
@@
   \item Tests can be seen as generalizations of the Lagrange multiplier
-    (or score) test.
+    test.
   \item Computation of critical values might require simulation from

### 2015-08-04 constraints/lav_constraints.Rmd
added details on errors related to lavaan's new equality constraint handling
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@380 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

---
title: "Score-based tests in lavaan 0.5-18"
author: "Ed Merkle"
date: "08/03/2015"
output: pdf_document
---

Consider the following example from Wang, Merkle, & Zeileis (2014):
```{r}
library("lavaan")
data("YouthGratitude", package = "psychotools")
compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
yg <- YouthGratitude[compcases, ]
restr <- cfa("f1 = ~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
   data = yg, group = "agegroup", meanstructure = TRUE,
   group.equal = "loadings")
```
The four loading parameters are constrained to be equal across groups, and we wish to carry out score-based tests related to these four parameters.  The score-based tests can be carried out via package *strucchange* when the above model is estimated in *lavaan* 0.5-17.  Starting in *lavaan* 0.5-18, equality constraints are handled differently.  As a result, the following `sctest()` call throws an error:
```{r, error=TRUE, message=FALSE}
library("strucchange")
sctest(restr, order.by = yg$agegroup, parm=1:4, vcov="info", functional="LMuo")
```
This error arises when `sctest()` tries to invert the result of `vcov()`.  It happens because *lavaan* no longer "reduces" parameter vectors with equality constraints, so that multiple columns of the variance/covariance matrix are associated with the same (equality-constrained) parameter.  The matrix cannot be inverted because it is less than full rank.  We can also see that there is a problem with the scores:
```{r}
scores <- lavScores(restr)
colSums(scores)[1:5]
```
In previous versions of *lavaan*, these sums were always 0 (reflecting the gradient evaluated at the ML estimates).  Now, some columns do not sum to zero because the scores associated with a single parameter are contained in multiple columns of `scores`.

## Possible solutions
Two possible solutions (that do not work) involve (i) obtaining the information matrix directly from *lavaan*, and (ii) using the *lavaan* 
 projection matrix `eq.constraints.K`, which projects a vector from the full parameter space to the reduced parameter space.

Pursuing the first solution, we can at least avoid the error from `sctest()` by supplying our own information matrix via `lavInspect`.  First, we define the custom function `lavinfo`:
```{r}
lavinfo <- function(fit, ...) lavInspect(fit, "information")
```
which obtains the information matrix directly from `lavaan` instead of relying on `strucchange` to do it.  We now modify our call to `sctest`:
```{r}
sctest(restr, order.by=yg$agegroup, parm=1:4, vcov=lavinfo, sandwich=FALSE, functional="LMuo")
```
which returns a plausible result, but the result still differs from previous versions (see Wang et al, 2014, p. 4, top first column).  This is because the argument `parm=1:4` gives us only the first group's loading parameters, and we already saw above that the scores in these columns do not sum to zero.  To fix this, we could find all the columns that correspond to these four loadings; these column names happen to be the only ones that have the string ".p" in them:
```{r}
sctest(restr, order.by=yg$agegroup, parm=grep(".p", names(coef(restr))), vcov=lavinfo, 
       sandwich=FALSE, functional="LMuo")
```
We now obtain a test statistic that is an order of magnitude too high; this is related to the fact that we chose 24 parameters of the original matrix (6 groups $\times$ 4 equality-constrained parameters).  In computing the test statistic, `sctest()` thinks we are testing 24 unique parameters instead of only 4, and the scores do not sum to zero in any of these columns.

Pursuing the second solution, we can define custom functions to project the original score vector and information matrix into the "reduced" dimension.  We can then use these in our call to `sctest()`:
```{r}
lavsc <- function(fit, ...) lavScores(fit) %*% fit@Model@eq.constraints.K
lavinfo <- function(fit, ...) t(fit@Model@eq.constraints.K) %*% 
                              lavInspect(fit, "information") %*% 
                              fit@Model@eq.constraints.K

sctest(restr, order.by=yg$agegroup, parm=1:4, scores=lavsc, vcov=lavinfo, 
       sandwich=FALSE, functional="LMuo")
```
This result is again implausible.  It is because, after we project the original score matrix into a lower dimension, each column of the matrix no longer corresponds to a single parameter.  Hence, the argument `parm=1:4` tests some function of the scores that no longer corresponds to the four loadings of interest.

## Preliminary solution
There exists a simple solution for models with simple equality constraints (i.e., multiple freely-estimated parameters constrained to be equal to one another), which should include all the models from Merkle & Zeileis, 2013, Merkle, Fan, & Zeileis, 2014, and Wang, Merkle, & Zeileis, 2014. These are (presumably) models that satisfy the following conditions
```{r}
model.works <- TRUE
if(restr@Model@eq.constraints){
  ones <- (restr@Model@ceq.JAC == 1 | restr@Model@ceq.JAC == -1)
  model.works <- all(restr@Model@ceq.rhs == 0) &
                 all(apply(restr@Model@ceq.JAC != 0, 1, sum) == 2) &
                 all(apply(ones, 1, sum) == 2) &
                 length(restr@Model@ceq.nonlinear.idx) == 0
}
```
These conditions say that the constraints all involve a simple difference between two parameters that must equal 0.

In these cases, the reduced score vector can be obtained by finding columns of the full score vector that contain equality-constrained parameters, then summing across these columns.  The reduced `vcov()` can be obtained by taking only one row/column that corresponds to an equality-constrained parameter, as opposed to taking all of them.  The equality-constrained parameters are not entirely straightforward to find because there currently do not appear to be any indicators in *lavaan*.  We find them via this code:<!-- Further, depending on how the equality constraints are set, equality-constrained parameters may or may not have the same labels.  We must search for duplicated parameter labels and search for equality constraints in the parameter table. -->
```{r}
## parameters with equality constraints:
lpt <- parTable(restr)
eq.rows <- which(lpt$op == "==")
eq.pars <- tapply(lpt$rhs[eq.rows], lpt$lhs[eq.rows], unique)
```
We can then make use of the results to obtain the reduced score matrix and information matrix:
```{r}
lavsc <- function(fit, ...){
  lpt <- parTable(fit)
  eq.rows <- which(lpt$op == "==")
  eq.pars <- tapply(lpt$rhs[eq.rows], lpt$lhs[eq.rows], unique)
  fsname <- lpt$plabel[lpt$free > 0]
  redscores <- lavScores(fit)
  rmcols <- which(fsname %in% unlist(eq.pars))
  for(i in 1:length(eq.pars)){
    sumcols <- which(fsname %in% eq.pars[[i]])
    eqloc <- match(names(eq.pars)[i], fsname)
    redscores[,eqloc] <- rowSums(redscores[,c(eqloc,sumcols)])
  }
  redscores <- redscores[,-rmcols]
  redscores
}

lavinfo <- function(fit, ...){
  lpt <- parTable(fit)
  eq.rows <- which(lpt$op == "==")
  eq.pars <- tapply(lpt$rhs[eq.rows], lpt$lhs[eq.rows], unique)
  fsname <- lpt$plabel[lpt$free > 0]
  rmcols <- which(fsname %in% unlist(eq.pars))
  fullvcov <- vcov(fit)
  redvcov <- fullvcov[-rmcols,-rmcols]
  solve(redvcov * sum(unlist(restr@SampleStats@nobs)))
}
```
Now, we can call `sctest()` on this new output:
```{r}
sctest(restr, order.by=yg$agegroup, parm=1:4, scores=lavsc, vcov=lavinfo, 
       sandwich=FALSE, functional="LMuo")
```
This matches closely with the result from Wang et al. (2014).  The result may differ slightly because *lavaan* is optimizing the model in a new way, so that the 0.5-18 ML estimates may slightly differ from the 0.5-17 ML estimates.

## Potential full solution
If we wish to apply the tests to models with more general constraints, Yves suggests that *lavaan* can give us Lagrange multiplier values associated with each constraint.  These could be used to augment the score matrix so that, if we then wish to test some equality-constrained parameters, we would presumably use the columns of the score matrix that correspond to the Lagrange multiplier values.  An augmented information matrix can already be obtained for these purposes via
```{r}
auginfo <- lavInspect(restr, "augmented.information")
```
Yves suggests looking at the `lavTestScore()` function for hints on augmenting the scores.  This option has not yet been explored.

### 2015-08-07 constraints/lav_constraints.Rmd
small change so code works with lavaan 0.5-18
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@381 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 ```{r}
-auginfo <- lavInspect(restr, "augmented.information")
+auginfo <- lavInspect(restr, "augmented.information", add.labels=FALSE)
 ```
+where the `add.labels` argument works around a bug that Yves has already fixed.
 Yves suggests looking at the `lavTestScore()` function for hints on augmenting the scores.  This option has not yet been explored.

### 2015-08-14 constraints/lav_constraints.Rmd
more detail on lavaan's new constraints
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@384 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Yves suggests looking at the `lavTestScore()` function for hints on augmenting the scores.  This option has not yet been explored.
+
+```{r}
+## comparison, solution?
+## gradient of likelihood, which is (- LM term)
+lavInspect(restr, "gradient")
+## matches scores
+lsc <- lavScores(restr)
+colSums(lsc)/nobs(restr)
+
+lsc2 <- t(lavInspect(restr,"gradient") + t(lsc))
+
+lavsc2 <- function(fit, ...){
+  t(lavInspect(fit,"gradient") + t(lavScores(fit)))
+}
+
+lavinfo <- function(fit, ...) lavInspect(fit, "information")
+
+sctest(restr, order.by=yg$agegroup, parm=c(1,16,46,61,76), scores=lavsc2, vcov=lavinfo, 
+       sandwich=FALSE, functional="LMuo")
+
+## find parameters affected by equality constraints
+apply(restr@Model@ceq.JAC==0,2,all)
+## find redundant parameters (only one non-zero entry, but
+## not for a+b+c == 3)
+apply(restr@Model@ceq.JAC!=0,2,sum)
+
+## use qr() to find non-redundant parameters?
+```

### 2015-08-17 constraints/lav_constraints.Rmd
new developments to handling constraints
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@385 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2015-08-04 constraints/lav_constraints.Rmd
@@
 ---
-title: "Score-based tests in lavaan 0.5-18"
+title: "Score-based tests in new lavaan versions"
 author: "Ed Merkle"
@@
 
-Consider the following example from Wang, Merkle, & Zeileis (2014):
+Consider the following example from Wang, Merkle, & Zeileis (2014) (note this is a custom lavaan version, as I removed the `eq.constraints.K` stuff from `lavScores()`):
 ```{r}
-library("lavaan")
+library("lavaan", lib.loc="~/dev_package_versions")
 data("YouthGratitude", package = "psychotools")
@@
 ## Possible solutions
-Two possible solutions (that do not work) involve (i) obtaining the information matrix directly from *lavaan*, and (ii) using the *lavaan* 
+Two possible solutions (that do not work) involve (i) obtaining the information matrix directly from *lavaan* and maintaining the "full" model, and (ii) using the *lavaan* 
  projection matrix `eq.constraints.K`, which projects a vector from the full parameter space to the reduced parameter space.
@@
 ```
-which obtains the information matrix directly from `lavaan` instead of relying on `strucchange` to do it.  We now modify our call to `sctest`:
+which obtains the information matrix directly from `lavaan` instead of relying on `strucchange` to do it.
+Second, we transform the score matrix so that each column again sums to zero.
+This involves obtaining the gradient of the likelihood at the ML estimates; note that the optimization function now contains a second term that involves Lagrange multipliers, so this gradient will not equal zero at the ML estimates.
+We apply this gradient to the scores, which accounts for the second term in the optimization function:
 ```{r}
-sctest(restr, order.by=yg$agegroup, parm=1:4, vcov=lavinfo, sandwich=FALSE, functional="LMuo")
+lavsc <- function(fit, ...) t(t(lavScores(fit)) + lavInspect(fit,"gradient"))
+```
+
+We now modify our call to `sctest`:
+```{r}
+sctest(restr, order.by=yg$agegroup, parm=1:4, scores=lavsc, 
+       vcov=lavinfo, sandwich=FALSE, functional="LMuo")
 ```
@@
 ```{r}
-sctest(restr, order.by=yg$agegroup, parm=grep(".p", names(coef(restr))), vcov=lavinfo, 
-       sandwich=FALSE, functional="LMuo")
+sctest(restr, order.by=yg$agegroup, parm=grep(".p", names(coef(restr))), 
+       scores=lavsc, vcov=lavinfo, sandwich=FALSE, functional="LMuo")
 ```
-We now obtain a test statistic that is an order of magnitude too high; this is related to the fact that we chose 24 parameters of the original matrix (6 groups $\times$ 4 equality-constrained parameters).
-In computing the test statistic, `sctest()` thinks we are testing 24 unique parameters instead of only 4, and the scores do not sum to zero in any of these columns.
+We now obtain a test statistic that is an order of magnitude too high; this seems related to the fact that we chose 24 parameters of the original matrix (6 groups $\times$ 4 equality-constrained parameters), with each score column summing to zero.
+In computing the test statistic, `sctest()` thinks we are testing 24 unique parameters instead of only 4, and the scores should not sum to zero in columns that reflect the same parameter.
+These results seem to imply that *strucchange* relies on the reduced model in order to produce the desired results.
+Related to this point, I can use the above approach with *nonnest2* and obtain the "correct" test statistics.
+I cannot obtain the correct $p$-values for statistics that have weighted sum of $\chi^2$ limiting distributions, because the algorithms implicitly use the number of model parameters to count degrees of freedom.
 
@@
 Hence, the argument `parm=1:4` tests some function of the scores that no longer corresponds to the four loadings of interest.
+It is unclear how we could test specific parameters using this projection.
 
@@
 These conditions say that the constraints all involve a simple difference between two parameters that must equal 0.
+In retrospect, I don't think the *strucchange* tests ever worked for models with more general equality constraints (I never considered them until now).
 
-In these cases, the reduced score vector can be obtained by finding columns of the full score vector that contain equality-constrained parameters, then summing across these columns.
+In the simple cases outlined above, the reduced score vector can be obtained by finding columns of the full score vector that contain equality-constrained parameters, then summing across these columns.
 The reduced `vcov()` can be obtained by taking only one row/column that corresponds to an equality-constrained parameter, as opposed to taking all of them.
@@
 ## Potential full solution
-If we wish to apply the tests to models with more general constraints, Yves suggests that *lavaan* can give us Lagrange multiplier values associated with each constraint.
-These could be used to augment the score matrix so that, if we then wish to test some equality-constrained parameters, we would presumably use the columns of the score matrix that correspond to the Lagrange multiplier values.
-An augmented information matrix can already be obtained for these purposes via
+If we wish to apply the tests to models with general equality constraints within a single group, it seems like we can use the first attempted solution described previously.
+This involves defining custom score and information matrix functions as we did at the start of the document:
 ```{r}
-auginfo <- lavInspect(restr, "augmented.information")
+lavinfo <- function(fit, ...) lavInspect(fit, "information")
+lavsc <- function(fit, ...) t(t(lavScores(fit)) + lavInspect(fit,"gradient"))
 ```
-Yves suggests looking at the `lavTestScore()` function for hints on augmenting the scores.  This option has not yet been explored.
+
+Then we may use these functions to test parameters involved in equality constraints, so long as we are careful not to test redundant parameters.  For example, if we use this model:
+```{r}
+hsm <- ' visual  =~ x1 + a*x2 + b*x3
+         textual =~ x4 + c*x5 + x6
+         speed   =~ x7 + x8 + x9
+         a+b+c == 3 '
+
+fit <- cfa(hsm, data=HolzingerSwineford1939)
+```
+then we would presumably only want to test two of the parameters {a,b,c}, possibly also  distributing the third parameter's scores to the other two parameters' scores.
+The results would change depending on which of the two parameters one chose, but this is a property of the tests; it is the same as getting different results when one constrains different parameters to one in a factor analysis.
+
+Beyond this, Yves suggested that *lavaan* can give us Lagrange multiplier values associated with each constraint.
+These could possibly be used to augment the score matrix so that, if we then wish to test some equality-constrained parameters, we would presumably use the columns of the score matrix that correspond to the Lagrange multiplier values.
+I am not optimistic that this approach will work because the other packages implicitly assume that the model is in reduced form (vs full form with constraints).
+
+Finally, I notice in the "equality constraints" document that `lav_partable_df()` does not yet deal with equality constraints.
+I think that the solution to this issue will be very relevant to the *strucchange* issues.
+
+## General solution
+Say that we have the parameter constraint $\theta_1 + \theta_2 + \theta_3 = 3$.  Using the full version of the model, we can get the gradient (and scores) associated with all three parameters:
+$$
+\begin{align}
+g(\theta_1) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | \bm{X})}{\partial \theta_1} \\
+g(\theta_2) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | \bm{X})}{\partial \theta_2} \\
+g(\theta_3) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | \bm{X})}{\partial \theta_3}.
+\end{align}
+$$
+
+Now, let's take $\theta_3 = (3 - \theta_1 - \theta_2)$, so that $\theta_3$ disappears from the model, allowing us to focus on a reduced model.
+Let the gradient of the reduced model be denoted $g^*()$ with reduced model parameters also having asterisks.
+We can (possibly) express the conditional log-likelihood of $\theta^*_1$ as a full model product, which splits the $\theta_1$ piece from the $\theta_3$ piece:
+$$
+L(\theta^*_1 | \theta_2, \theta_3, \theta_4, \ldots, \bm{X}) = f_1(\theta_1 | \theta_2, \ldots, \bm{X}) f_2(\theta_3 | \theta_2, \theta_4, \ldots, \bm{X}).
+$$
+To get the gradient associated with $\theta^*_1$, we can now use a product rule combined with a chain rule (since $\theta_3$ is a function of $\theta_1$):
+$$
+g^*(\theta^*_1 | \theta_2, \theta_3, \theta_4, \ldots, \bm{X}) = g(\theta_1 | \theta_2, \theta_3, \ldots, \bm{X}) + \left ( g(\theta_3 | \theta_1, \theta_2, \ldots, \theta_4, \bm{X}) \times \frac{\partial{(3 - \theta^*_1 - \theta^*_2)}{\partial \theta^*_1} \right ).
+$$
+So this would be $g(\theta_1 | \cdot) - g(\theta_3 | \cdot)$, both of which we can obtain from the full model `estfun()`.
+In general, we need to have the derivative of the constraint wrt to each parameter involved.
+I believe this can be obtained from `ceq.JAC` (for those in `ceq.linear.idx`) and `con.jac` (for those in `ceq.nonlinear.idx`)?
+
+Note this strategy covers the "simple equality constraint" solution that was previously discussed.
+
+```{r}
+hsm2 <- ' visual  =~ x1 + a*x2 + b*x3
+         textual =~ x4 + c*x5 + x6
+         speed   =~ x7 + x8 + x9
+         a*b == 1 '
+
+fit <- cfa(hsm2, data=HolzingerSwineford1939)
+
+ls <- lavScores(fit)
+## assuming b = 1/a, convert to reduced model a parameter:
+sum(ls[,1] - ls[,2]/(coef(fit)[1]^2))
+
+## how to get derivative generally? see where ceq.function
+## is written, mimic to get something similar that will
+## give numerical derivs?
+```

### 2015-08-17 constraints/lav_constraints.Rmd
further updates to constraints doc
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@386 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 author: "Ed Merkle"
-date: "08/03/2015"
+date: \today
 output: pdf_document
@@
 
-Consider the following example from Wang, Merkle, & Zeileis (2014) (note this is a custom lavaan version, as I removed the `eq.constraints.K` stuff from `lavScores()`):
+Consider the following example from Wang, Merkle, & Zeileis (2014) (note this is a custom lavaan version, as I removed the most recent changes to `lavScores()` involving the `eq.constraints.K` stuff):
 ```{r}
@@
 4, top first column).
-This is because the argument `parm=1:4` gives us only the first group's loading parameters, and we already saw above that the scores in these columns do not sum to zero.
+This is because the argument `parm=1:4` gives us only the first group's loading parameters.
 To fix this, we could find all the columns that correspond to these four loadings; these column names happen to be the only ones that have the string ".p" in them:
@@
 ## Preliminary solution
-There exists a simple solution for models with simple equality constraints (i.e., multiple freely-estimated parameters constrained to be equal to one another), which should include all the models from Merkle & Zeileis, 2013, Merkle, Fan, & Zeileis, 2014, and Wang, Merkle, & Zeileis, 2014.
+(*maybe you should skip to "general solution", because that supercedes this section*) There is a simple solution for models with simple equality constraints (i.e., multiple freely-estimated parameters constrained to be equal to one another), which should include all the models from Merkle & Zeileis, 2013, Merkle, Fan, & Zeileis, 2014, and Wang, Merkle, & Zeileis, 2014.
 These are (presumably) models that satisfy the following conditions
@@
 
-## Potential full solution
-If we wish to apply the tests to models with general equality constraints within a single group, it seems like we can use the first attempted solution described previously.
-This involves defining custom score and information matrix functions as we did at the start of the document:
-```{r}
-lavinfo <- function(fit, ...) lavInspect(fit, "information")
-lavsc <- function(fit, ...) t(t(lavScores(fit)) + lavInspect(fit,"gradient"))
-```
-
-Then we may use these functions to test parameters involved in equality constraints, so long as we are careful not to test redundant parameters.  For example, if we use this model:
-```{r}
-hsm <- ' visual  =~ x1 + a*x2 + b*x3
-         textual =~ x4 + c*x5 + x6
-         speed   =~ x7 + x8 + x9
-         a+b+c == 3 '
-
-fit <- cfa(hsm, data=HolzingerSwineford1939)
-```
-then we would presumably only want to test two of the parameters {a,b,c}, possibly also  distributing the third parameter's scores to the other two parameters' scores.
-The results would change depending on which of the two parameters one chose, but this is a property of the tests; it is the same as getting different results when one constrains different parameters to one in a factor analysis.
-
-Beyond this, Yves suggested that *lavaan* can give us Lagrange multiplier values associated with each constraint.
-These could possibly be used to augment the score matrix so that, if we then wish to test some equality-constrained parameters, we would presumably use the columns of the score matrix that correspond to the Lagrange multiplier values.
-I am not optimistic that this approach will work because the other packages implicitly assume that the model is in reduced form (vs full form with constraints).
-
-Finally, I notice in the "equality constraints" document that `lav_partable_df()` does not yet deal with equality constraints.
-I think that the solution to this issue will be very relevant to the *strucchange* issues.
-
 ## General solution
@@
 $$
-\begin{align}
-g(\theta_1) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | \bm{X})}{\partial \theta_1} \\
-g(\theta_2) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | \bm{X})}{\partial \theta_2} \\
-g(\theta_3) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | \bm{X})}{\partial \theta_3}.
-\end{align}
+\begin{aligned}
+g(\theta_1) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | X)}{\partial \theta_1} \\
+g(\theta_2) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | X)}{\partial \theta_2} \\
+g(\theta_3) &= \frac{\partial L(\theta_1, \theta_2, \theta_3, \ldots | X)}{\partial \theta_3}.
+\end{aligned}
 $$
@@
 Let the gradient of the reduced model be denoted $g^*()$ with reduced model parameters also having asterisks.
-We can (possibly) express the conditional log-likelihood of $\theta^*_1$ as a full model product, which splits the $\theta_1$ piece from the $\theta_3$ piece:
+We can express the full model log-likelihood as two sub-functions, one of which includes $\theta_1$ and one of which includes $\theta_3$:
 $$
-L(\theta^*_1 | \theta_2, \theta_3, \theta_4, \ldots, \bm{X}) = f_1(\theta_1 | \theta_2, \ldots, \bm{X}) f_2(\theta_3 | \theta_2, \theta_4, \ldots, \bm{X}).
+L(\theta^*_1, \theta_2, \theta_4, \ldots | X) = f_1(\theta_1, \theta_2, \theta_4, \ldots | X) f_2(\theta_3, \theta_2, \theta_4, \ldots | X).
 $$
@@
 $$
-g^*(\theta^*_1 | \theta_2, \theta_3, \theta_4, \ldots, \bm{X}) = g(\theta_1 | \theta_2, \theta_3, \ldots, \bm{X}) + \left ( g(\theta_3 | \theta_1, \theta_2, \ldots, \theta_4, \bm{X}) \times \frac{\partial{(3 - \theta^*_1 - \theta^*_2)}{\partial \theta^*_1} \right ).
+g^*(\theta^*_1) = g(\theta_1) + \left ( g(\theta_3) \times \frac{\partial (3 - \theta^*_1 - \theta^*_2)}{\partial \theta^*_1} \right ).
 $$
-So this would be $g(\theta_1 | \cdot) - g(\theta_3 | \cdot)$, both of which we can obtain from the full model `estfun()`.
-In general, we need to have the derivative of the constraint wrt to each parameter involved.
-I believe this can be obtained from `ceq.JAC` (for those in `ceq.linear.idx`) and `con.jac` (for those in `ceq.nonlinear.idx`)?
+So this would be $g(\theta_1) - g(\theta_3)$, both of which we can obtain from the full model `lavScores()` (AKA `estfun()`).
+In general, we can implement this if we are able express one parameter involved in the constraint as a function of other parameters.
+This is simple for equality constraints, simple summations, and products, but it becomes more difficult to do automatically for things like
+$$
+\begin{aligned}
+(\theta_1 + \theta_2)^2 &== 2 \\
+\sin(\theta_1\theta_2) &== .5.
+\end{aligned}
+$$
+For constraints like these, perhaps `lavScores()` could throw an error and ask the user to specify the constraint using a single parameter on the lhs?
+If there is a single parameter on the lhs, we can always use numerical derivatives to obtain what we need.
+Or perhaps there is a way to use the results from `ceq.JAC` (for those in `ceq.linear.idx`) and `con.jac` (for those in `ceq.nonlinear.idx`) to deal with this?
 
-Note this strategy covers the "simple equality constraint" solution that was previously discussed.
-
+Here is an example showing how we could obtain reduced model scores when we have a simple nonlinear constraint:
 ```{r}
 hsm2 <- ' visual  =~ x1 + a*x2 + b*x3
-         textual =~ x4 + c*x5 + x6
+          textual =~ x4 + x5 + x6
           speed   =~ x7 + x8 + x9
@@
 ls <- lavScores(fit)
-## assuming b = 1/a, convert to reduced model a parameter:
-sum(ls[,1] - ls[,2]/(coef(fit)[1]^2))
+## assuming b = 1/a, convert to reduced model a parameter
+## (knowing that db/da = -1/a^2)
+ls[,1] <- ls[,1] - ls[,2]/(coef(fit)[1]^2)
+ls <- ls[,-2]
+head(colSums(ls))
+```
 
-## how to get derivative generally? see where ceq.function
-## is written, mimic to get something similar that will
-## give numerical derivs?
-```
+## Etc
+Beyond this, Yves suggested that *lavaan* can give us Lagrange multiplier values associated with each constraint.
+These could possibly be used to augment the score matrix so that, if we then wish to test some equality-constrained parameters, we would presumably use the columns of the score matrix that correspond to the Lagrange multiplier values.
+I am not optimistic that this approach will work because the other packages implicitly assume that the model is in reduced form (vs full form with constraints) for counting parameters, degrees of freedom, etc.
+But I may not understand the suggestion.
+
+I also notice in the "equality constraints" document that `lav_partable_df()` does not yet deal with equality constraints.
+I think that the solution to this issue might be relevant to the "reduced model" issues here.

### 2015-08-17 constraints/lav_constraints.Rmd
and some further small changes to constraints
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@387 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Beyond this, Yves suggested that *lavaan* can give us Lagrange multiplier values associated with each constraint.
-These could possibly be used to augment the score matrix so that, if we then wish to test some equality-constrained parameters, we would presumably use the columns of the score matrix that correspond to the Lagrange multiplier values.
-I am not optimistic that this approach will work because the other packages implicitly assume that the model is in reduced form (vs full form with constraints) for counting parameters, degrees of freedom, etc.
-But I may not understand the suggestion.
+These could possibly be used to augment the score matrix so that there is presumably one column associated with each constraint (?).
+Perhaps this could work, but we would need a separate value for each observation in the sample (vs a single value across all observations).
+I'm not sure how this could be obtained, after looking at the new `lavTestScore()`.
 

### 2015-12-04 ucl2015/slides.Rnw
presentation materials for seminar at UCL
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@388 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2015-07-23 ecpa2015/slides.Rnw
@@
 options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
+library("lavaan")
+library("strucchange")
+library("psychotree")
+library("psychotools")
 set.seed(1090)
@@
   \end{itemize}
-  \item Illustration
+  \item Software
+  \item Illustrations
 \end{itemize}
@@
 manifest variables, e.g., item response theory (IRT) models or confirmatory
-factor analysis (CFA, today's focus).
+factor analysis (CFA).
 
@@
   \item In ordinal variables the ordering of the categories is often not exploited
-    (assessing only if at least one group differs from the others).
+    -- assessing only if at least one group differs from the others.
   \item When likelihood ratio or Wald tests are employed, the model has to be
@@
 $p$-dimensional observations $\bm{x}_i$ ($i = 1, \dots, n$)
-based on $k$-dimensional parameter ${\bm \theta}$.
+and $k$-dimensional parameter ${\bm \theta}$.
 
@@
 assesses whether the model ``deviations'' ${\bm s}(\hat{\bm \theta}; \bm{x}_i)$
-depend on $v_i$. This can consider only a subset $k*$ of all $k$
+depend on $v_i$. This can consider only a subset $k^*$ of all $k$
 parameters/scores or try to capture other patterns than $H_1^*$.
@@
 \begin{frame}
+\frametitle{Score-based tests: Continuous variables}
+
+<<gefp, echo=FALSE, results=hide>>=
+load("mz.rda")
+source("../www/mz.R")
+gefp_3_info <- gefp(mz_fit, fit = NULL, order.by = mz_data$age, parm = 1:3,
+  vcov = info.mzfit, sandwich = FALSE)
+gefp_19_info <- gefp(mz_fit, fit = NULL, order.by = mz_data$age,
+  vcov = info.mzfit, sandwich = FALSE)
+@
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<gefp-dmax, echo=FALSE, results=hide, fig=TRUE, height=4.5, width=9>>=
+par(mfcol = c(1, 2))
+plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
+plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
+@
+
+%plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
+%plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
+%plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
+%plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Score-based tests: Continuous variables}
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<gefp-cvm, echo=FALSE, results=hide, fig=TRUE, height=4.5, width=9>>=
+par(mfcol = c(1, 2))
+plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
+plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
+@
+
+%plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
+%plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Score-based tests: Continuous variables}
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<gefp-maxlm, echo=FALSE, results=hide, fig=TRUE, height=4.5, width=9>>=
+par(mfcol = c(1, 2))
+plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
+plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
+@
+
+\end{frame}
+
+\begin{frame}
 \frametitle{Score-based tests: Ordinal variables}
@@
 
-\subsection{Illustration}
+\subsection{Software}
 
 \begin{frame}
-\frametitle{Illustration}
+\frametitle{Software}
 
-\textbf{Software:} In R system for statistical computing.
+\textbf{R packages:}
 \begin{itemize}
@@
     model classes, including \emph{lavaan} objects for CFA models.
+  \item Other psychometric models that cooperate with \emph{strucchange}
+    are provided in \emph{psychotools}, e.g., IRT models (Rasch, partial
+    credit, rating scale), Bradley-Terry models for paired comparisons,
+    and multinomial processing tree models.
+  \item Model-based recursive partitioning based on the general
+    parameter instability tests are provided in \emph{partykit}.
+  \item Adaptation to psychometric models in \emph{psychotree}.
 \end{itemize}
+
+\end{frame}
+
+\subsection{CFA: Youth gratitude}
+
+\begin{frame}
+\frametitle{CFA: Youth gratitude}
+
+\textbf{Question:} Does measurement invariance hold across age groups
+when an adult gratitude scale is applied to youth subjects?
+
+\medskip
+
+\textbf{Source:} Froh \emph{et al.} (2011, \emph{Psychological Assessment}).
+  ``Measuring Gratitude in Youth: Assessing the Psychometric Properties of
+  Adult Gratitude Scales in Children and Adolescents.''
 
@@
 \begin{itemize}
-  \item Application of adult gratitude scale to $n=1401$ youth aged 10--19 years.
-  \item GQ-6 scale has five Likert scale items with seven points each.
+  \item GQ-6 gratitude scale with five Likert scale items of seven points each.
+  \item Application to $n=1401$ youth aged 10--19 years (six age groups).
   \item Assess the factor loadings of a one-factor model.
-  \item Question: Measurement invariance across six age groups?
 \end{itemize}
@@
 \begin{frame}[fragile]
-\frametitle{Illustration}
+\frametitle{CFA: Youth gratitude}
 
@@
 <<models>>=
-gq6cfa <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+gq6_cfa <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
   data = yg, group = "agegroup", meanstructure = TRUE,
@@
 <<wdmo, strip.white=all, fig=TRUE, height=4.5, width=5, include=FALSE>>=
-sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
+sctest(gq6_cfa, order.by = yg$agegroup, parm = 1:4,
   vcov = "info", functional = "WDMo", plot = TRUE)
@@
 <<maxlmo, strip.white=all, fig=TRUE, height=4.5, width=5, include=FALSE>>=
-sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
+sctest(gq6_cfa, order.by = yg$agegroup, parm = 1:4,
   vcov = "info", functional = "maxLMo", plot = TRUE)
@@
 \begin{frame}
-\frametitle{Illustration}
+\frametitle{CFA: Youth gratitude}
 
@@
 \includegraphics[width=0.8\textwidth]{slides-maxlmo}
+
+\end{frame}
+
+\subsection{IRT: Examining exams}
+
+\begin{frame}
+\frametitle{IRT: Examining exams}
+
+\textbf{Question:} Does measurement invariance hold in a Rasch model for single-choice exam results?
+
+\medskip
+
+\textbf{Source:} Mathematics for first-year business and economics students at Universit\"at Innsbruck.
+Online tests (conducted in OpenOLAT) and written exams for 500--1,000 students per semester.
+
+\medskip
+
+\textbf{Data:} Individual results from an end-term exam.
+\begin{itemize}
+  \item 729 students (out of 941 registered).
+  \item 13~single-choice items with five answer alternatives,
+    covering the basics of analysis, linear algebra, financial mathematics.
+  \item Two groups with partially different item pools (on the same topics). Individual versions of
+    items generated via \emph{exams}.
+  \item Correctly solved items yield 100\% of associated points. Items without correct solution can
+    either be unanswered (0\%) or with an incorrect answer ($-25\%$). Only considered as binary here.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{IRT: Examining exams}
+
+\textbf{Packages:}
+
+<<packages, results=hide>>=
+library("psychotools")
+library("psychotree")
+@
+
+\medskip
+
+\textbf{Data:} Load, select first group, and exclude extreme scorers.
+
+<<data>>=
+load("MathExam.rda")
+mex <- subset(MathExam, group == 1 & nsolved > 0 & nsolved < 13)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{IRT: Examining exams}
+
+<<solved-plot, eval=FALSE>>=
+plot(mex$solved)
+@
+
+\vspace*{-0.6cm}
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<solved-plot-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
+<<solved-plot>>
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{IRT: Examining exams}
+
+<<credits-plot, eval=FALSE>>=
+plot(mex$credits)
+@
+
+\vspace*{-0.6cm}
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<credits-plot-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
+<<credits-plot>>
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{IRT: Examining exams}
+
+<<raschmodel-plot, eval=FALSE>>=
+mex_rasch <- raschmodel(mex$solved)
+plot(mex_rasch, type = "profile")
+@
+
+\vspace*{-0.6cm}
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<raschmodel-plot-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
+<<raschmodel-plot>>
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{IRT: Examining exams}
+
+<<raschmodel-piplot, eval=FALSE>>=
+plot(mex_rasch, type = "piplot")
+@
+
+\vspace*{-0.4cm}
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<raschmodel-piplot-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
+<<raschmodel-piplot>>
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{IRT: Examining exams}
+
+\textbf{Measurement invariance tests:}
+
+<<raschmodel-maxlm, strip.white=all, fig=TRUE, height=6.5, width=7.5, include=FALSE>>=
+sctest(mex_rasch, order.by = jitter(mex$tests),
+  vcov = "info", functional = "maxLM", plot = TRUE)
+@
+
+<<raschmodel-maxlmo, strip.white=all, fig=TRUE, height=6.5, width=7.5, include=FALSE>>=
+mex$otests <- cut(mex$tests, breaks = c(0, 14:24, 26),
+  ordered = TRUE, labels = c("<= 14", 15:24, ">= 25"))
+sctest(mex_rasch, order.by = mex$otests,
+  vcov = "info", functional = "maxLMo", plot = TRUE)
+@
+
+Clear violation of measurement invariance: Students that performed poorly in the
+previous online tests have a different item profile.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{IRT: Examining exams}
+
+\vspace*{-0.6cm}
+
+\centering
+\includegraphics[width=0.8\textwidth]{slides-raschmodel-maxlm}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{IRT: Examining exams}
+
+\vspace*{-0.6cm}
+
+\centering
+\includegraphics[width=0.8\textwidth]{slides-raschmodel-maxlmo}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{IRT: Examining exams}
+
+<<raschtree>>=
+mex_tree <- raschtree(solved ~ otests + attempt + semester + study,
+  data = mex, vcov = "info", ordinal = "L2")
+@
+
+\vspace*{-0.3cm}
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<raschtree-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
+plot(mex_tree)
+@
+
+\end{frame}
+
+\subsection{Paired comparisons: Modeling topmodels}
+
+\begin{frame}[fragile]
+\frametitle{Paired comparisons: Modeling topmodels}
+
+\textbf{Question:} Does measurement invariance hold for a Bradley-Terry
+preference scaling of attractiveness?
+
+\bigskip
+
+\textbf{Source:} Strobl, Wickelmaier, Zeileis
+(2010, \textit{Journal of Educational and Behavioral Statistics}).
+``Accounting for Individual Differences in Bradley-Terry Models by Means of Recursive Partitioning.''
+
+\bigskip
+
+\textbf{Data:}
+\begin{itemize}
+  \item Paired comparisons of attractiveness for
+        \emph{Germany's Next Topmodel~2007} finalists:
+	Barbara, Anni, Hana, Fiona, Mandy, Anja.
+  \item Survey with 192~respondents at Universit{\"a}t T{\"u}bingen.
+  \item Available covariates: Gender, age, familiarty with the TV~show.
+  \item Familiarity assessed by yes/no questions:
+        (1)~Do you recognize the women?/Do you know the show?
+	(2)~Did you watch it regularly?
+	(3)~Did you watch the final show?/Do you know who won?
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Paired comparisons: Modeling topmodels}
+
+\vspace*{-0.7cm}
+\begin{center}
+\includegraphics[width=0.25\textwidth]{Barbara.jpg} \hspace*{0.1cm}
+\includegraphics[width=0.25\textwidth]{Anni.jpg} \hspace*{0.1cm}
+\includegraphics[width=0.25\textwidth]{Hana.jpg}
+
+\vspace*{0.1cm}
+
+\includegraphics[width=0.25\textwidth]{Fiona.jpg} \hspace*{0.1cm}
+\includegraphics[width=0.25\textwidth]{Mandy.jpg} \hspace*{0.1cm}
+\includegraphics[width=0.25\textwidth]{Anja.jpg}
+\end{center}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Paired comparisons: Modeling topmodels}
+
+<<Topmodel-MOB>>=
+data("Topmodel2007", package = "psychotree")
+tm_mob <- bttree(preference ~ ., data = Topmodel2007,
+  minsplit = 5, ref = "Barbara")
+@
+
+\vspace*{-0.3cm}
+
+\setkeys{Gin}{width=1.05\textwidth}
+\hspace*{-0.4cm}%
+<<Topmodel-plot, echo=FALSE, fig=TRUE, height=6.2, width=10>>=
+plot(tm_mob, abbreviate = 1, yscale = c(0, 0.5))
+@
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Paired comparisons: Modeling topmodels}
+
+\textbf{Recursively partitioned preferences:} Standardized ranking from Bradley-Terry model.
+
+\bigskip
+
+\begin{tabular}{|r|rrrrrr|} \hline
+ & Barbara & Anni & Hana & Fiona & Mandy & Anja \\ \hline
+<<Topmodel-table, echo=FALSE, results=tex>>=
+tm_tab <- round(worth(tm_mob), digits = 2)
+tm_tab <- cbind(rownames(tm_tab), format(tm_tab, nsmall = 2))
+cat(paste(apply(tm_tab, 1, paste, collapse = " & "), collapse = "\\\\\n"))
+@
+\\ \hline
+\end{tabular}
 
@@
   \item Easy-to-use implementation available in R package \emph{strucchange}.
-  \item Can be re-used in model-based recursive partitioning.
+  \item Can be re-used in model-based recursive partitioning in R packages
+    \emph{partykit} and \emph{psychotree}.
 \end{itemize}
+
+\bigskip
+
+\textbf{Acknowledgments:} This work was supported by National Science Foundation grant SES-1061334.
 
@@
 
-\small
+\footnotesize
 
@@
 
-%% \medskip
-%% 
-%% Strobl C, Julia Kopf, Zeileis A (2015).
-%%   \dquote{Rasch Trees: A New Method for Detecting Differential Item Functioning in the Rasch Model.}
-%%   \textit{Psychometrika}. \textbf{80}(2), 289--316.
-%%   \doi{10.1007/s11336-013-9388-3}
+\bigskip
+
+Strobl C, Julia Kopf, Zeileis A (2015).
+  \dquote{Rasch Trees: A New Method for Detecting Differential Item Functioning in the Rasch Model.}
+  \textit{Psychometrika}. \textbf{80}(2), 289--316.
+  \doi{10.1007/s11336-013-9388-3}
+
+\medskip
+
+Strobl C, Wickelmaier F, Zeileis A (2011).
+  \dquote{Accounting for Individual Differences in Bradley-Terry Models by Means of Recursive Partitioning.}
+  \textit{Journal of Educational and Behavioral Statistics}, \textbf{36}(2), 135--153.
+  \doi{10.3102/1076998609359791}
+
+\medskip
+
+Zeileis A, Hothorn T, Hornik K (2008).
+ \dquote{Model-Based Recursive Partitioning.}
+  \textit{Journal of Computational and Graphical Statistics},
+  \textbf{17}(2), 492--514.
+  \doi{10.1198/106186008X319331}
 

### 2015-12-04 ucl2015/slides.R

@@ -1,0 +1,117 @@
+### R code from vignette source 'slides.Rnw'
+
+###################################################
+### code chunk number 1: preliminaries
+###################################################
+options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
+set.seed(1090)
+
+
+###################################################
+### code chunk number 2: packages
+###################################################
+library("lavaan")
+library("strucchange")
+
+
+###################################################
+### code chunk number 3: data
+###################################################
+data("YouthGratitude", package = "psychotools")
+compcases <- apply(YouthGratitude[, 4:28], 1,
+  function(x) all(x %in% 1:9))
+yg <- YouthGratitude[compcases, ]
+
+
+###################################################
+### code chunk number 4: models
+###################################################
+gq6cfa <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
+  data = yg, group = "agegroup", meanstructure = TRUE,
+  group.equal = "loadings")
+
+
+###################################################
+### code chunk number 5: wdmo
+###################################################
+sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "WDMo", plot = TRUE)
+
+
+###################################################
+### code chunk number 6: maxlmo
+###################################################
+sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
+  vcov = "info", functional = "maxLMo", plot = TRUE)
+
+
+###################################################
+### code chunk number 7: packages
+###################################################
+library("psychotools")
+library("psychotree")
+
+
+###################################################
+### code chunk number 8: data
+###################################################
+load("MathExam.rda")
+mex <- subset(MathExam, nsolved > 0 & nsolved < 13 & group == 1)
+
+
+###################################################
+### code chunk number 9: classes
+###################################################
+sapply(mex, function(x) class(x)[1])
+
+
+###################################################
+### code chunk number 10: solved-plot (eval = FALSE)
+###################################################
+## plot(mex$solved)
+
+
+###################################################
+### code chunk number 11: solved-plot-fig
+###################################################
+plot(mex$solved)
+
+
+###################################################
+### code chunk number 12: raschmodel-plot (eval = FALSE)
+###################################################
+## ram <- raschmodel(mex$solved)
+## plot(ram, type = "profile")
+
+
+###################################################
+### code chunk number 13: raschmodel-plot-fig
+###################################################
+ram <- raschmodel(mex$solved)
[…]

### 2015-12-04 ucl2015/slides.R
tangled R script not needed under version control
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@389 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,117 +1,0 @@
-### R code from vignette source 'slides.Rnw'
-
-###################################################
-### code chunk number 1: preliminaries
-###################################################
-options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
-set.seed(1090)
-
-
-###################################################
-### code chunk number 2: packages
-###################################################
-library("lavaan")
-library("strucchange")
-
-
-###################################################
-### code chunk number 3: data
-###################################################
-data("YouthGratitude", package = "psychotools")
-compcases <- apply(YouthGratitude[, 4:28], 1,
-  function(x) all(x %in% 1:9))
-yg <- YouthGratitude[compcases, ]
-
-
-###################################################
-### code chunk number 4: models
-###################################################
-gq6cfa <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
-  data = yg, group = "agegroup", meanstructure = TRUE,
-  group.equal = "loadings")
-
-
-###################################################
-### code chunk number 5: wdmo
-###################################################
-sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "WDMo", plot = TRUE)
-
-
-###################################################
-### code chunk number 6: maxlmo
-###################################################
-sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
-  vcov = "info", functional = "maxLMo", plot = TRUE)
-
-
-###################################################
-### code chunk number 7: packages
-###################################################
-library("psychotools")
-library("psychotree")
-
-
-###################################################
-### code chunk number 8: data
-###################################################
-load("MathExam.rda")
-mex <- subset(MathExam, nsolved > 0 & nsolved < 13 & group == 1)
-
-
-###################################################
-### code chunk number 9: classes
-###################################################
-sapply(mex, function(x) class(x)[1])
-
-
-###################################################
-### code chunk number 10: solved-plot (eval = FALSE)
-###################################################
-## plot(mex$solved)
-
-
-###################################################
-### code chunk number 11: solved-plot-fig
-###################################################
-plot(mex$solved)
-
-
-###################################################
-### code chunk number 12: raschmodel-plot (eval = FALSE)
-###################################################
-## ram <- raschmodel(mex$solved)
-## plot(ram, type = "profile")
-
-
-###################################################
-### code chunk number 13: raschmodel-plot-fig
-###################################################
-ram <- raschmodel(mex$solved)
[…]
